### 2022-02-07 src/example_blangiardo-replication/blangiardo-replication.Rmd
Example from Blangiardo and Cameletti "Spatial and Spatio-temporal Bayesian Models with R-INLA" Section 4.9

---
title: "Blangiardo replication"
author: "Adam Howes"
output: html_document
---

Step-by-step example of the INLA method.
The model for i.i.d. observations $y = (y_1, \ldots, y_n)$ is as follows
\begin{align}
y_i \, | \, x, \theta &\sim \mathcal{N}(x, 1/\theta), \\
x &\sim \mathcal{N}(x_0 = -3, 1/\tau_0 = 4), \\
\theta &\sim \text{Gamma}(a = 1.6, b = 0.4).
\end{align}

## Set-up

```{r message=FALSE, warning=FALSE}
midblue <- "#3D9BD0"
midgreen <- "#00855A"
midpink <- "#B3608E"
```

Load the data:

```{r}
y <- c(1.2697, 7.7637, 2.2532, 3.4557, 4.1776, 6.4320, -3.6623, 7.7567, 5.9032, 7.2671,
       -2.3447, 8.0160, 3.5013, 2.8495, 0.6467, 3.2371, 5.8573, -3.3749, 4.1507, 4.3092,
       11.7327, 2.6174, 9.4942, -2.7639, -1.5859, 3.6986, 2.4544, -0.3294, 0.2329, 5.2846)

n <- length(y)
y_bar <- mean(y)
```

```{r, fig.height=2, fig.cap="The observations (ordered by index, though this has no interpretation). The prior (mean shown as dashed line) mismatches with the data."}
data.frame(index = 1:30, y = y) %>%
  ggplot(aes(x = index, y = y)) +
  geom_point(shape = 4) +
  labs(title = "The data") +
  geom_hline(yintercept = -3, col = "#666666", lty = "dashed")
```

Parameters of the prior distribution:

  ```{r}
x_0 <- -3
tau_0 <- 1/4
a <- 1.6
b <- 0.4
```

## `R-INLA`

```{r}
formula <- y ~ 1
dat <- list(y = y)

theta_prior <- list(prec = list(prior = "loggamma", param = c(a, b)))

fit <- inla(formula,
            data = dat,
            control.family = list(hyper = theta_prior),
            control.fixed = list(mean.intercept = x_0, prec.intercept = tau_0))
```

## R implementation

Exact distribution of $x$ given $\theta$:

  ```{r}
inner_loop <- function(theta) {
  tau_n <- n * theta + tau_0
  x_n <- (theta * n * y_bar + tau_0 * x_0) / tau_n
  return(list(x_n = x_n, tau_n = tau_n))
}
```

Functions with $\theta > 0$ constraint and on the $\log(\theta)$ scale to avoid this constraint:

```{r}
nl_full_conditional_x <- function(x, theta, log_input = FALSE) {
  if(log_input == TRUE) theta <- exp(theta) # Convert from log(theta) to theta
  par <- inner_loop(theta)
  return(-dnorm(x, par$x_n, sqrt(1 / par$tau_n), log = TRUE))
}

nl_post_marginal_theta <- function(theta, log_input = FALSE) {
  target <- 0
  if(log_input == TRUE) {
    target <- target + theta # Increment by Jacobian correction (theta is l_theta here)
    theta <- exp(theta) # Convert from log(theta) to theta
  }
  par <- inner_loop(theta)

  # (^)
  target <- target + -0.5 * log(par$tau_n) +
    dgamma(theta, shape = a, rate = b, log = TRUE) +  # theta prior
    dnorm(par$x_n, x_0, sqrt(1 / tau_0), log = TRUE) +  # x prior
    sum(dnorm(y, par$x_n, sqrt(1 / theta), log = TRUE)) # y likelihood

  return(-target)
}
```

Note that the code above in the section marked `(^)` is very similar to that in `blangiardo.cpp`.
Recall that the Laplace approximation $\tilde p(\theta \, | \, y)$, which in this instance coincides with the exact posterior, is given by
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(y, \mu^\star(\theta), \theta)}{\det(Q(\theta))^{1/2}}.
\end{equation}
This form can be seen in `(^)`:

  * $p(y, \mu^\star(\theta), \theta)$ corresponds to the calls to `dgamma`, `dnorm` and sum of `dnorm` evaluated at $\mu^\star(\theta)$ which here is simply `par$x_n` (for the Normal distribution the mean of the distribution is the same as the mode).
* The logarithm of $1 / \det(Q(\theta))^{1/2}$ when $\theta$ is a scalar is given by `-0.5 * log(par$tau_n)`.

It should be the case that `inner_loop` produces parameter values for the latent field which are those found by `TMB` using the optimisation inner loop (hence the name).

```{r}
de_nl <- function(f, ...) exp(-f(...)) # Versions which are not negative logarithms

full_conditional_x <- function(x, theta, log_input = FALSE) {
  de_nl(nl_full_conditional_x, x, theta, log_input = FALSE)
}

post_marginal_theta <- function(theta, log_input = FALSE) {
  de_nl(nl_post_marginal_theta, theta, log_input = FALSE)
}
```

Simple grids (one dimensional):

```{r}
eval_grid <- function(grid = NULL, uniform = FALSE, K = NULL, min = NULL, max = NULL, f) {
  if(uniform) {
    grid <- seq(min, max, length.out = K)
  }
  df <- data.frame(input = grid, output = sapply(grid, f))
  df <- mutate(df, norm_output = output / sum(output))
  return(df)
}

blangiardo_theta <- eval_grid(uniform = TRUE,
                              K = 25, min = 0.001, max = 0.3,
                              f = post_marginal_theta)

dense_theta <- eval_grid(uniform = TRUE,
                         K = 500, min = 0.001, max = 0.3,
                         f = post_marginal_theta)
```

```{r fig.height = 3, fig.cap = "True posterior marginal of theta (blue line) overlaid with the choice of integration points. This naive grid places too many points in regions of the parameter space without much posterior density."}
ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = blangiardo_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y=element_blank()) +
  labs(title = "Blangiardo integration points for the hyperparameters")
```

Optimisation on the log scale (to avoid constrained optimisation):

```{r}
its <- 1000
r_nlminb <- nlminb(start = 0,
                   objective = nl_post_marginal_theta,
                   log_input = TRUE,
                   control = list(iter.max = its, trace = 0))

r_optim <- optim(par = 0,
                 fn = nl_post_marginal_theta,
                 log_input = TRUE,
                 method = "Brent",
                 lower = -100, # Have to specify upper and lower bounds
                 upper = 100, # when using optimize (1D function)
                 control = list(maxit = its, trace = 0))

c(r_nlminb$par, r_optim$par) # The same
r_opt <- r_nlminb
```

For INLA's grid strategy, starting from the mode, we take steps of size $\delta_z$ checking that each point meets the criteria
\begin{equation}
  \log \tilde p(\theta(0) \, | \, y) - \log \tilde p(\theta(z) \, | \, y) < \delta_\pi. \label{eq:criteria}
\end{equation}
In the one dimensional case there is no need to do the $z$-parametrisation.
Choosing $\delta_z$ and $\delta_\pi$ based upon manual tuning (how does INLA select these numbers in general?):

```{r}
delta_z <- 0.05
delta_pi <- 3
```

The following is very inefficient R programming but just an idea as to how it could be done:

```{r fig.height = 3, fig.cap = "The INLA method grid points. This looks better but it still looks a little strange probably because of the reparametrisation. Then there is also the task of choosing the values of $\\delta_z$ and $\\delta_\\pi$."}
# Increasing
points <- c(r_opt$par) # On the log theta scale

i <- 0
condition <- TRUE
while(condition) {
  i <- i + 1
  proposal <- r_opt$par + i * delta_z
  statistic <- nl_post_marginal_theta(theta = proposal, log_input = TRUE) -
               nl_post_marginal_theta(theta = r_opt$par, log_input = TRUE)
  condition <- (statistic < delta_pi)
  if(condition){
    points <- c(points, proposal)
  }
}

# Decreasing
i <- 0
condition <- TRUE
while(condition) {
  i <- i + 1
  proposal <- r_opt$par - i * delta_z
  statistic <- nl_post_marginal_theta(theta = proposal, log_input = TRUE) -
               nl_post_marginal_theta(theta = r_opt$par, log_input = TRUE)
  condition <- (statistic < delta_pi)
  if(condition){
    points <- c(points, proposal)
  }
}

inla_theta <- eval_grid(exp(points), f = post_marginal_theta)

ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y=element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  annotate("text", x = 0.175, y = 5e-41,
           label = "Now the points are concentrated \n where there is higher density")
```

One of the INLA vignettes discusses how the user can set their own integration points--see `browseVignettes(package="INLA")`.
The integration points that INLA uses can be found using `fit$joint.hyper`, which we can add on to the above plot:

```{r}
internals_inla_theta <- data.frame(
  input = exp(fit$joint.hyper$`Log precision for the Gaussian observations`),
  output = exp(fit$joint.hyper$`Log posterior density`)
)

# Different normalising constant?
ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = internals_inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y=element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  annotate("text", x = 0.175, y = 5e-41,
           label = "Now the points are concentrated \n where there is higher density")
```

We take the set of points defined by `inla_theta` to be our $\{\theta^{(k)}\}$.
$K$ is given by `length(points)` which equals `r length(points)`.

Define a function `nl_joint_post` by taking the negative log of the joint posterior $p(x, \theta \, | \, y) = p(x \, | \, \theta, y) p(\theta \, | \, y)$ to give
\begin{equation}
- \log p(x, \theta \, | \, y) = - \log p(x \, | \, \theta, y) - \log p(\theta \, | \, y).
\end{equation}

```{r}
nl_joint_post <- function(x, theta, log_input = FALSE) {
  nl_full_conditional_x(x, theta, log_input) +
  nl_post_marginal_theta(theta, log_input)
}
```

For any given input $x$ you can do quadrature according to
\begin{equation}
  \tilde p(x \, | \, y) =
  \sum_{k = 1}^K p(x \, | \, \theta^{(k)}, y) \times p(\theta^{(k)} \, | \, y) \times \Delta^{(k)},
\end{equation}
but how should the range of $x$ be chosen?
Blangiardo again use a naive grid for demonstration:

```{r}
blangiardo_x <- seq(-8, 5, length.out = 50)
```

Modify `nl_joint_post` to accept and return vectors using the base R function `Vectorize`, then apply the outer product:

```{r}
v_nl_joint_post <- Vectorize(nl_joint_post, vectorize.args = c("x", "theta"))
nl_joint_post <- outer(blangiardo_x, points, v_nl_joint_post, log_input = TRUE)
```

50 rows (the number of $x$ integration points) and 26 columns (the number of $\theta$ integration points):

```{r}
dim(nl_joint_post)
```

The $x$ and $y$-axis do not correspond to values, just to indicies:

```{r}
image(nl_joint_post)
```

Following the book:

```{r}
nl_post_marginal_x <- rowSums(nl_joint_post)
plot(blangiardo_x, nl_post_marginal_x)
blangiardo_x[[which.min(nl_post_marginal_x)]]
```

## `TMB`

C++ for the negative log joint posterior:

```{cpp, code=readLines("blangiardo.cpp"), eval=FALSE}
```

```{r results=FALSE}
TMB::compile("blangiardo.cpp")
dyn.load(dynlib("blangiardo"))

param <- list(x = 0, l_theta = 0)
```

`h` with Laplace approximation is a function of $\dim(\theta)$ inputs, integrating out $x$ by specifying `random = c("x")`:

```{r}
h <- MakeADFun(data = dat,
               parameters = param,
               random = c("x"),
               DLL = "blangiardo",
               hessian = TRUE)
```

`f` without Laplace approximation is a function of $\dim(\theta) + \dim(x)$ inputs:

```{r}
f <- MakeADFun(data = dat,
               parameters = param,
               DLL = "blangiardo",
               hessian = TRUE)
```

Optimisation using `nlminb` and `optim` gets the same results:

```{r}
its <- 1000

tmb_nlminb <- nlminb(start = h$par,
                     objective = h$fn,
                     gradient = h$gr,
                     control = list(iter.max = its, trace = 0))

tmb_optim <- optim(par = h$par,
                   fn = h$fn,
                   gr = h$gn,
                   method = "Brent",
                   lower = -100, # Have to specify upper and lower bounds
                   upper = 100, # when using optimize (1D function)
                   control = list(maxit = its, trace = 0))

c(tmb_optim$par, tmb_nlminb$par) # The same

tmb_opt <- tmb_nlminb

sd_out <- sdreport(h, par.fixed = tmb_opt$par, getJointPrecision = TRUE)
```

The mode and comparison (I started by making the mistake of exponentiating `fit$summary.hyperpar$mode` rather than the internal version--on the log scale, as with R and `TMB`--which, because the transformation is non-linear, is not the same).

```{r}
kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$mode),
                           exp(r_opt$par),
                           exp(tmb_opt$par)))

rownames(kable_data) <- c("$\\texttt{R-INLA}$", "R", "$\\texttt{TMB}$")
colnames(kable_data) <- c("Posterior mode")

kableExtra::kable(kable_data, booktabs = TRUE, escape = FALSE, align = "c")
```

Compare the variances for $\log(\theta)$:

```{r}
sd_out$cov.fixed # TMB
fit$internal.summary.hyperpar$sd^2 # INLA
```

Compare $x$:

```{r}
sd_out$par.random # TMB
fit$summary.fixed # INLA

plot(fit$marginals.fixed$`(Intercept)`, type = "l")
abline(v = sd_out$par.random)
```

### 2022-02-07 src/example_blangiardo-replication/script.R

@@ -1,0 +1,5 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("example_blangiardo-replication")
+# setwd("src/example_blangiardo-replication")
+
+rmarkdown::render("blangiardo-replication.Rmd")

### 2022-02-07 src/docs_01-07-21-stats-epi-group/01-07-21-stats-epi-group.Rmd
Add the presentation I gave last summer about this work (to-do: use depends for the epil results now that everything is together here)

---
title: Eventually an introduction to the `aghq` R package
subtitle: Statistics and Epidemiology Group Meeting
author: Adam Howes
institute: Imperial College London
date: July 2021
bibliography: citations.bib
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

# Motivation I

* People on ART $A_i$ can be used as supplementary data for small-area estimation of HIV prevalence $\rho_i$
\begin{align*}
A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
y_i &\sim \text{Bin}(m_i, \rho_i), \\
\text{logit}(\alpha_i) &\sim f(\vartheta_\alpha), \\
\text{logit}(\rho_i) &\sim g(\vartheta_\rho), \quad i = 1, \ldots, n,
\end{align*}
* If $f$ and $g$ are Gaussian then model is almost, but not quite, a latent Gaussian model by the definition of @rue2009approximate
  * This is due to small non-linearities (multiplying two latent Gaussian fields)
  * Each observation depends on more than one element of the latent field

# Motivation II

:::::::::::::: {.columns}

::: {.column width=.65}

* Previous slide is a simplified version of the Naomi evidence synthesis model
* Countries to fit the model using their own data ("in production"?)
  * Can't run long MCMC on the cluster for weeks, as might be the case if this was one paper
* Can't use `R-INLA`, require something more flexible
* Currently using Template Model Builder `TMB` [@kristensen2015tmb]

:::

::: {.column width=.35}

```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'}
knitr::include_graphics("naomi_hex.png")
```

:::

::::::::::::::

# Aside: common theme I

* Combining flawed (sparse, aggregated) gold standard (measuring the thing we want) data with other correlated (more available, high resolution) data (measuring not exactly what we want)
* Consistently resulting in models with multiple outcomes (evidence synthesis, multi-output)
* I think a lot of these are going to be not quite LGMs

# Aside: common theme II

* Examples include
  * Naomi model: DHS data is "gold standard", supported by ANC data from pregnant women
  * Sexual risk behaviour model: estimates of FSW population at national level, supported by DHS data
    * The national-level FSW estimates might be more like "bronze standard"
    * DHS approximately asks "have you received money or gifts in exchange for sex in past 12 months"
  * Loa loa prevalence and eyeworm history prevalence model: measuring eyeworm history is a cheap proxy for Loa loa [@amoah2020geostatistical]

# Recap on latent Gaussian models

* Three-stage model
\begin{alignat*}{2}
  &\text{(Observations)}     &   \y &\sim p(\y \, | \, \x), \\
  &\text{(Latent field)}     &   \x &\sim p(\x \, | \, \btheta), \\
  &\text{(Hyperparameters)}  &   \qquad \btheta &\sim p(\btheta),
\end{alignat*}
where $\y = (y_1, \ldots, y_n)$, $\x = (x_1, \ldots, x_n)$, $\btheta = (\theta_1, \ldots, \theta_m)$
* Interested in learning both $(\btheta, \x)$ from data $\y$
* Covers most of the models used in spatiotemporal statistics

# Recap on Integrated Nested Laplace Approximation I

* @rue2009approximate or e.g. @blangiardo2015spatial
* Approximate Bayesian inference for \textcolor{hilit}{latent Gaussian models} (LGMs), which are three-stage models with middle layer
\begin{equation*}
  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
\end{equation*}
* `R-INLA` implementation takes advantage of sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)

# Recap on Integrated Nested Laplace Approximation II

* Gives approximate \textcolor{hilit}{posterior marginals} $\{\tilde p(x_i \, | \, \y)\}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$
* To approximate posterior marginals below requires $\tilde p(\btheta \, | \, \y)$ and $\tilde p(x_i \, | \, \btheta, \y)$
\begin{align}
  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}

# Recap on Integrated Nested Laplace Approximation III

1) First Laplace approximate hyperparameter posterior
\begin{equation}
\tilde p(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)}  \label{eq:hypermarginal}
\end{equation}
which can be marginalised to get $\tilde p(\theta_j \, | \, \y)$
2) In both \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. \eqref{eq:hypermarginal}, so choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$
  * For low $m$ INLA uses a grid-strategy which I illustrate in the next slide
  * For larger $m$ this becomes too expensive and a CCD design is used

#

```{r message=FALSE, echo=FALSE, fig.cap=paste("An illustration of the INLA grid method for selecting integration points using a toy Gaussian distribution for $\\btheta$. Start at the mode and work outwards along the eigenvectors until the density drops sufficiently low. \\label{fig:grid}")}
knitr::include_graphics("depends/inla-grid.pdf")
```

# Recap on Integrated Nested Laplace Approximation IV

3) Choose approximation for $\tilde p(x_i \, | \, \btheta, \y)$
  * Simplest version [@rue2007approximate] is to marginalise the $p_G(\x \, | \, \btheta, \y)$
\begin{equation}
\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta))
\end{equation}
  * The above is referred to as "Gaussian" approximation, and confusingly there are two more complex ones called "simplified Laplace" and "Laplace"
  * You can pick which one in `R-INLA` using the `method` option
4) Finally use quadrature to get 
\begin{equation}
  \tilde p(x_i \, | \, \y) = 
  \sum_{k = 1}^K \tilde p(x_i \, | \, \btheta^{(k)}, \y) \times \tilde p(\btheta^{(k)} \, | \, \y) \times \Delta^{(k)}
\end{equation}

# Template Model Builder I

* R package which implements the Laplace approximation for latent variable models using AD (via `CppAD`)
  * For more about AD see e.g. @griewank2008evaluating
  * Useful for getting the mode, Hessian
* Write an objective function $f(\x, \btheta)$ in C++ ("user template")
    * We select $f(\x, \btheta) = - \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)$

# Template Model Builder II

```{cpp eval=FALSE}
#include <TMB.hpp>

template <class Type>
Type objective_function<Type>::operator()() {
  // Define data e.g.
  DATA_VECTOR(y);
  // Define parameters e.g.
  PARAMETER(mu);
  // Calculate negative log-likelihood e.g.
  nll = Type(0.0);
  nll -= dnorm(y, mu, 1, true).sum()
  return(nll);
}
```

# Template Model Builder III

* Performs the Laplace approximation $L_f(\btheta) \approx L^\star_f(\btheta)$ and use R to optimise this with respect to $\btheta$ to give $\hat \btheta$ (the central point in Figure \ref{fig:grid})
  * This is done by specifying the `random` argument to be the parameters that you want to integrate out with a Laplace approximation (the latent field)
* MAP estimate of $\x$ conditional on $\hat \btheta$
* Standard errors calculated using the $\delta$-method (a Gaussian assumption)

# Adaptive Gaussian Hermite Quadrature

* Recent work by Alex Stringer and coauthors on AGHQ
  * `aghq` R package and vignette [@stringer2021implementing]
  * Theory paper [@bilodeau2021stochastic]
* Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior)
* Use $k$ quadrature points
  * If $k$ is odd then they include the mode
  * If $k = 1$ then it's a Laplace approximation
  * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often

# Epil example I

* Epilepsy example from Section 5.2. of @rue2009approximate (previously from BUGS):
  * Patients $i = 1, \ldots, 59$ each either assigned treatment $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$ to help with seizures
  * Visits to clinics $j = 1, \ldots, 4$ times with $y_{ij}$ the number of seizures of the $i$th person in the two weeks proceeding their $j$th visit to the clinic
  * Covariates age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$

# Epil example II

This is what the model looks like (it's a Poisson GLMM):
\begin{align*}
  y_{ij}          &\sim \text{Poisson}(\lambda_{ij}), \\
  \lambda_{ij}    &= e^{\eta_{ij}}, \\
  \eta_{ij}       &= \beta_0 + \beta_{\texttt{Base}} \log(\texttt{Baseline}_j / 4) + \beta_{\texttt{Trt}} \texttt{Trt}_i +
                     \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
                  &+ \beta_{\texttt{Age}} \log(\texttt{Age}_i) + \beta_{\texttt{V}_4} {\texttt{V}_4}_j +
                     \epsilon_i + \nu_{ij}, \quad i=1:59, \quad j=1:4, \\
  \beta           &\sim \mathcal{N}(0, 100^2), \quad \forall \beta, \\
  \epsilon_i      &\sim \mathcal{N}(0, 1/\tau_\epsilon), \\
  \nu_{ij}        &\sim \mathcal{N}(0, 1/\tau_\nu), \\
  \tau_\epsilon   &\sim \Gamma(0.001, 0.001), \\
  \tau_\nu        &\sim \Gamma(0.001, 0.001).
\end{align*}

# Epil example III

* `aghq` package interfaces really easily with `TMB`!
* This is the code I used to fit the model with `TMB`

```{r eval=FALSE}
obj <- MakeADFun(
  data = dat,
  parameters = param,
  # These are the ones integrated out with a Laplace approximation
  random = c("epsilon", "nu"),
  DLL = "epil"
)
```

# Epil example IV

* Then to fit it with `aghq` it's only a very small modification

```{r eval=FALSE}
fit <- aghq::marginal_laplace_tmb(
  obj, 
  k = 3, 
  startingvalue = c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
)
```

#

```{r echo=FALSE}
results <- readRDS("comparison-results.rds")

beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
rownames(results) <- c(sapply(0:5, beta_i))

results %>% 
  round(digits = 3) %>%
  kableExtra::kable(format = "latex", booktabs = TRUE, escape = TRUE, align = "c") %>%
  kable_styling(latex_options = "scale_down")
```

# Plan

* Test `aghq` for toy Naomi example
  * Do as above with the Epil example, testing versus a long MCMC run
  * Sometimes you have to look pretty hard for a node (element of the latent field) where there are differences. In the INLA paper they do this by computing a SKLD and ordering by maximum difference. Probably good to do here as well
* Extend `aghq` to replicate INLA functionality by adding the more complex versions of $\tilde p(x_i \, | \, \btheta, \y)$ then test that wih Naomi
  * Håvard Rue philosophy: "do one thing and do it well"
  * `R-INLA` implementation of INLA based on sparsity of $\bm{Q}(\btheta)$ that doesn't hold up for extended LGMs
  * @wood2020simplified on how to still do it
* Try the INLA without `R-INLA` on other almost LGMs and see how far it can be pushed

# References {.allowframebreaks}

### 2022-02-07 src/docs_01-07-21-stats-epi-group/script.R

@@ -1,0 +1,25 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("docs_01-07-21-stats-epi-group")
+# setwd("src/docs_01-07-21-stats-epi-group")
+
+sf_lightgrey <- "#E6E6E6"
+lightgrey <- "#D3D3D3"
+
+lightblue <- "#56B4E9"
+lightgreen <- "#009E73"
+lightpink <- "#CC79A7"
+light_palette <- c(lightblue, lightgreen, lightpink)
+
+midblue <- "#3D9BD0"
+midgreen <- "#00855A"
+midpink <- "#B3608E"
+mid_palette <- c(midblue, midgreen, midpink)
+
+darkblue <- "#004E83"
+darkgreen <- "#00380D"
+darkpink <- "#802D5B"
+dark_palette <- c(darkblue, darkgreen, darkpink)
+
+cbpalette <- multi.utils::cbpalette()
+
+rmarkdown::render("01-07-21-stats-epi-group.Rmd")

### 2022-07-31 src/docs_paper/appendix.Rmd
Mac file upload

---
title: Appendix to "Simplifying Integrated nested Laplace approximation with adaptive Gaussian Hermite quadrature"
subtitle: |
  Corresponding author: Adam Howes (`ath19@ic.ac.uk`)
output:
  pdf_document:
    toc: true
    number_sections: true
    keep_tex: true
    includes:
      in_header: preamble.tex
bibliography: citations.bib
---

\newpage

<!-- So that the tables and figures in this appendix are prefixed by "A." -->
\setcounter{table}{0}  
\renewcommand{\thetable}{A.\arabic{table}} 
\setcounter{figure}{0} 
\renewcommand{\thefigure}{A.\arabic{figure}}

# References {#references .unnumbered}

### 2022-07-31 src/docs_paper/paper.Rmd
Mac file upload

---
title: Simplifying Integrated nested Laplace approximation with adaptive Gaussian Hermite quadrature
author:
  - name: Adam Howes
    email: ath19@ic.ac.uk
    affiliation: A
    corresponding: ath19@ic.ac.uk
address:
  - code: A
    address: Department of Mathematics, Imperial College London
abstract: |
  |  **Background:**
  |
  |  **Methods:**
  |
  |  **Results:**
  |
  |  **Conclusions:**
author_summary: |
  Author summary
bibliography: citations.bib
output:
  rticles::elsevier_article:
    number_sections: true
    includes:
      in_header: preamble.tex
---

# Introduction

The `TMB` package \citep{kristensen2016tmb} is gaining popularity in spatial statistics as a flexible alternative to `R-INLA` for fitting latent Gaussian models \citep{osgoodzimmerman2021statistical}.

# Background

## Integrated nested Laplace approximation \label{sec:inla}

Integrated nested Laplace approximation (INLA) \citep{rue2009approximate} is an approximate Bayesian inference method which uses the Laplace approximation and numerical integration.
INLA is designed for use with latent Gaussian models (LGMs) of the form
\begin{alignat}{2}
&\text{(Observations)}     &        y_i &\sim p(y_i \, | \, x_i, \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\
&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\
&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta), \label{eq:parameters}
\end{alignat}
where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
The joint posterior of $(\x, \btheta)$ is given by
\begin{equation}
  p(\x, \btheta \, | \, \y) 
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right). \label{eq:lgmjointposterior}
\end{equation}
Rather than approximating the above full posterior, the INLA method instead approximates the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
\begin{align}
  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}
An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$, which we discuss in turn.

### Approximation (i) \label{sec:approxi}

The posterior marginal of the parameters $p(\btheta \, | \, \y)$ appears in both Equations \eqref{eq:inla1} and \eqref{eq:inla2}.
This distribution is approximated by $\tilde p(\btheta \, | \, \y)$ and represented by a set of $K$ integration points $\{ \btheta^{(k)} \}$ and area-weights $\{ \Delta^{(k)} \}$.
The first step is to rewrite $p(\btheta \, | \, \y)$ as
\begin{equation}
  p(\btheta \, | \, \y) 
  = \frac{p(\x, \btheta \, | \, \y)}{p(\x \, | \, \btheta, \y)}
  \propto \frac{p(\y, \x, \btheta)}{p(\x \, | \, \btheta, \y)}. \label{eq:hyperpost}
\end{equation}
Approximation (i) then uses a Gaussian approximation
\begin{equation}
p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \hat{\bm{\mu}}(\btheta), \hat{\mathbf{Q}}(\btheta)^{-1}) \label{eq:gaussianx}
\end{equation}
to the denominator of Equation \ref{eq:hyperpost}.
This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution
\begin{equation}
p(\x \, | \, \btheta, \y) \propto
\exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right)
\end{equation}
close to being Gaussian since $\y$ is generally not that informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved \citep{blangiardo2015spatial}.
<!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference? Perhaps what the author (Marta) meant is that the Gaussian distribution is sufficiently flexible to accommodate the updates made by $\y$. -->

As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}.
As such, taking $\x = \hat{\bm{\mu}}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as
\begin{equation}
\tilde p(\btheta \, | \, \y) \propto \frac{p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)} = \frac{p(\y, \hat{\bm{\mu}}(\btheta), \btheta)}{\det(\hat{\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal}
\end{equation}
where the final equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\hat{\bm{\mu}}(\btheta)$.

### Approximation (ii)

Having utilised used the approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section~\ref{sec:approxi}, a natural approach, and that taken by \citet{rue2007approximate}, is to marginalise this distribution directly
\begin{equation}
\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \hat{\mu}_i(\btheta), 1 / \hat{q}_i(\btheta)),
\end{equation}
where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from $\hat{\bm{\mu}}(\btheta)$ and $\mathbf{Q}^\star(\btheta)$ respectively.
Although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode.
As a result, although this method is available in \texttt{R-INLA} it is generally not advised.
Instead, \citet{rue2009approximate} propose two methods, a Laplace approximation and a simplified version which is less computationally demanding.
The full Laplace approximation is
\begin{align}
  p(x_i \, | \, \btheta, \y) &= p(\x \, | \, \btheta, \y) \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &= \frac{p(\x, \btheta \, | \, \y)}{p(\btheta \, | \, \y)} \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &\propto \frac{p(\x, \btheta \, | \, \y)}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat{\bm{\mu}}_{-i}(x_i, \btheta)} 
  = \tilde p_{LA}(x_i \, | \, \btheta, \y), \label{eq:laplaceapproximation}
\end{align}
where $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{This notation is somewhat dangerous as $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well.
However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $x_i$ and $\btheta$, which is often computationally prohibitive.
Two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by \citet{rue2009approximate} to reduce the computational cost:
\begin{enumerate}
\item
Avoiding having to find the mode via optimisation by using the approximation $\hat{\bm{\mu}}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$
\item
As only those $x_j$ close to $x_i$ should have an impact on the marginal of $x_i$, then by selecting some subset $R_i(\btheta)$ of nodes $j$ to impact $j$ the matrix which needs to be factorised can be reduced in dimension to be $| R_i(\btheta) | \times | R_i(\btheta) |$ rather than $n \times n$
\end{enumerate}

### Combining the approximations

## Simplified INLA

# Template Model Builder \label{sec:tmb}

Template Model Builder (TMB) \citep{kristensen2016tmb} is an \textsf{R} package for fitting random effect models, also known as latent variable models, hierarchical models or a host of other names.
In \texttt{TMB} inference is based upon optimisation of a target function.
This makes it very flexible, and able to handle non-linear, non-Gaussian random effect models.

The approach of \texttt{TMB} is inspired by the AD Model Builder (ADMB) package \citep{fournier2012ad}.
The "AD" in ADMB is automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
AD is popular in machine learning \citep{baydin2017automatic}, for example as the basis for backpropagation algorithm and is beginning to gain popularity in statistics, including as a part of Stan \citep{carpenter2017stan}.
\texttt{TMB} uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.

## Statistical framework

Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.\footnote{\citet{kristensen2016tmb} use the notation $u$ for random effects and $\theta$ for parameters. We aim for consistency with Section~\ref{sec:inla}.}
Let $\ell(\x, \btheta) \triangleq - \log p(\y \, | \, \x, \btheta)$ be the negative joint log-likelihood.
In \texttt{TMB}, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
A standard maximum likelihood approach is to optimise
\begin{equation}
L_\ell(\btheta) \triangleq \int_{\mathbb{R}^n} p(\y \, | \, \x, \btheta) \text{d}\x = \int_{\mathbb{R}^n} \exp(-\ell(\x, \btheta)) \text{d}\x \label{eq:flikelihood}
\end{equation}
with respect to $\btheta$ to find the maximum likelihood estimator (MLE) $\hat \btheta$.
Taking a superficially more Bayesian approach than above, instead of $\ell$, the user may instead write a function to evaluate the negative joint penalised log-likelihood given by
\begin{equation}
f(\x, \btheta) 
\triangleq - \log p(\y \, | \, \x, \btheta) p(\x, \btheta)
= \ell(\x, \btheta) - \log p(\x, \btheta),
\end{equation}
equivalent up to an additive constant to the negative log-posterior.
<!-- \begin{equation} -->
<!-- f(\x, \btheta)  -->
<!-- = - \log p(\y, \x, \btheta)  -->
<!-- = - \log p(\x, \btheta \, | \, \y) - C, -->
<!-- \end{equation} -->
<!-- where $C = \log p(\y)$ is the log evidence. -->
Using $f$ in place of $\ell$, then the penalised likelihood is proportional to the posterior marginal of $\btheta$
\begin{equation}
L_f(\btheta) 
\triangleq \int_{\mathbb{R}^n} \exp(-f(\x, \btheta)) \text{d}\x 
\propto \int_{\mathbb{R}^n} p(\x, \btheta \, | \, \y) \text{d}\x = p(\btheta \, | \, \y). \label{eq:1}
\end{equation}
Integrating out the random effects directly, as in Equation~\ref{eq:1} above, is usually intractable because $\x$ is high-dimensional, so \citet[Equation 3]{kristensen2016tmb} use a Laplace approximation $L^\star_f(\btheta)$ based instead upon integrating out a Gaussian approximation to the random effects.
This Laplace approximation is analogous to the INLA approximation $\tilde p(\btheta \, | \, \y)$ given in Section~\ref{sec:approxi}.
\begin{equation*}
f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta) = - \frac{\partial^2}{\partial \x^2} \log p(\y, \x, \btheta) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= - \frac{\partial^2}{\partial \x^2} \log p(\x \, | \, \btheta, \y) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= \hat{\bm{Q}}(\btheta).
\end{equation*}
Inference proceeds by optimising $L^\star_f(\btheta)$ via minimisation of
\begin{equation}
-\log L^\star_f(\btheta) \propto \frac{1}{2} \log \det (\hat{\bm{Q}}(\btheta)) + f(\hat{\bm{\mu}}(\btheta), \btheta)  \label{eq:nllaplace},
\end{equation}
where $\propto$ is used to mean proportional up to an additive constant.
The parameters of the Gaussian approximation (Equation~\ref{eq:gaussianx}), are found in terms of $f$ via $\hat{\bm{\mu}}(\btheta) = \arg \min_\x f(\x, \btheta)$ and $\hat{\bm{Q}}(\btheta) = f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta)$ and must be recomputed for each value of $\btheta$.
Obtaining $\hat{\bm{\mu}}(\btheta)$ is known as the inner optimisation step.

# Examples \label{sec:examples}

## The Naomi small-area estimation model

@eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation as well as prompting investigation into any conflicts.

**Prevalence component**
  Consider a country partitioned into areas $i = 1, \ldots, n$, with a simple random household survey of $m_i$ people is conducted in each area with $y_i$ HIV positive cases observed.
Cases may be modelled using a binomial logistic regression model
\begin{align}
y_i &\sim \text{Bin}(m_i, \rho_i), \\
\text{logit}(\rho_i) &\sim \mathcal{N}(\beta_\rho, \sigma_\rho^2)
\end{align}
where HIV prevalence $\rho_i$ is modelled by a Gaussian with mean $\beta_\rho$ and standard deviation $\sigma_\rho$.

**ANC component**
  Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
If of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
\begin{align}
y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + b_i, \\
b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
\end{align}
may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence.

**ART component**
  The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence, whereby districts with high prevalence are likely to have a greater number of people receiving treatment.
ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
As such, ART coverage may also be modelled using a binomial logistic regression model
\begin{align}
A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
\end{align}

## Supporting information

**Appendix A**: Statistical modelling \newline
**Appendix B**: Supplementary tables and figures

## Funding

AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1).

## Disclaimer

# References {#references .unnumbered}

### 2022-07-31 src/docs_paper/preamble.tex
Mac file upload

\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage{bm}
\RequirePackage{enumerate}
\RequirePackage{enumitem}
\RequirePackage{tabularx}
\RequirePackage{adjustbox}
\RequirePackage{tikz}
\RequirePackage{bbm}
\RequirePackage{lscape}
\RequirePackage{booktabs}
\RequirePackage{longtable}
\RequirePackage{lscape}
\RequirePackage{pifont}
\RequirePackage[margin=1in]{geometry}
\usepackage{footnote}

\newcommand{\Sc}{\mathcal{S}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\brho}{\bm{\rho}}
\newcommand{\btheta}{\bm{\theta}}

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\definecolor{lightblue}{HTML}{56B4E9}
\definecolor{lightgreen}{HTML}{009E73}
\definecolor{lightorange}{HTML}{E69F00}
\definecolor{lightgrey}{HTML}{E6E6E6}

\newcommand{\lightblueline}{\raisebox{2pt}{\tikz{\draw[-,lightblue,solid,line width = 3pt](0,0) -- (2mm,0);}}}
\newcommand{\lightgreenline}{\raisebox{2pt}{\tikz{\draw[-,lightgreen,solid,line width = 3pt](0,0) -- (2mm,0);}}}
\newcommand{\lightorangeline}{\raisebox{2pt}{\tikz{\draw[-,lightorange,solid,line width = 3pt](0,0) -- (2mm,0);}}}
\newcommand{\lightgreyline}{\raisebox{2pt}{\tikz{\draw[-,lightgrey,solid,line width = 3pt](0,0) -- (2mm,0);}}}

\usepackage{amsmath}
\DeclareMathOperator*{\argmin}{arg\,min}

### 2022-07-31 src/docs_paper/script.R

@@ -1,0 +1,28 @@
+# orderly::orderly_develop_start("docs_paper")
+# setwd("src/docs_paper")
+
+#' Conversion of figures from .pdf to .jpeg at specified resolution
+Sys.setenv(PATH = paste(Sys.getenv("PATH"), "/opt/homebrew/bin", sep = ":"))
+
+convert_pdf_jpeg <- function(name) {
+  command <- paste0(
+    "convert -density 300 depends/", name, ".pdf -scene 1 -background white",
+    " -alpha remove -alpha off -quality 80 depends/", name, ".jpeg"
+  )
+  system(command)
+}
+
+convert_pdf_png <- function(name) {
+  command <- paste0(
+    "convert -density 300 depends/", name, ".pdf -scene 1 -background white",
+    " -alpha remove -alpha off -quality 80 depends/", name, ".png"
+  )
+  system(command)
+}
+
+#' For paper.Rmd
+#' For appendix.Rmd
+
+#' Render documents
+rmarkdown::render("paper.Rmd")
+rmarkdown::render("appendix.Rmd")

### 2022-09-06 src/epil/epil.Rmd
epil to html

---
title: "Epilepsy example"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
---

```{r setup, class.source = 'fold-hide'}
knitr::opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  cache.lazy = FALSE,
  cache.comments = FALSE
)
options(scipen = 999)
cbpalette <- multi.utils::cbpalette()
library(tidyverse)

rstan::rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

Section 5.2. of Rue, Martino and Chopin (2009) illustrates the INLA method using an epilepsy example.

```{r}
data(Epil, package = "INLA")
head(Epil)
```

Start by centering the covariates:

```{r}
center <- function(x) (x - mean(x))

Epil <- Epil %>%
  mutate(CTrt    = center(Trt),
         ClBase4 = center(log(Base/4)),
         CV4     = center(V4),
         ClAge   = center(log(Age)),
         CBT     = center(Trt * log(Base/4)))
```

Implement the model based upon a response variable of length $N \times J$ and a model matrix with $K$ predictors (including the intercept term):

```{r}
N <- 59
J <- 4
K <- 6
X <- model.matrix(formula(~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT), data = Epil)
y <- Epil$y
```

For the individual specific random effect $\epsilon_i$ we use a transformation matrix $\bm{E}$ which repeats elements of vector of length `N` each `J` times, for example:

```{r}
make_epsilon_matrix <- function(N, J) {
  t(outer(1:N, 1:(N * J), function(r, c) as.numeric((J*(r - 1) < c) & (c <= J*r))))
}

E <- make_epsilon_matrix(N = 3, J = 2)
E
t(E %*% c(1, 2, 3)) # Same as rep(1:3, 2)
```

Multiplying this matrix $E$ by the vector $\epsilon$ allows it to be directly added to the linear predictor $\eta = \beta X + \nu + E \epsilon$, where $\beta$ is the vector of coefficients and $X$ is the model matrix.

## Stan

```{r results='hide'}
dat <- list(N = N, J = J, K = K, X = X, y = y, E = make_epsilon_matrix(N, J))

fit1 <- stan(
  "epil.stan",
  data = dat,
  chains = 1,
  warmup = 100,
  iter = 1000,
  control = list(adapt_delta = 0.95)
)
```

## INLA

Both $\tau_\epsilon$ and $\tau_\nu$ have the same $\Gamma(0.001, 0.001)$ prior.
In INLA, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior:

```{r}
tau_prior <- list(prec = list(prior = "loggamma",
                              param = c(0.001, 0.001),
                              initial = 1,
                              fixed = FALSE))
```

The variable `Epil$rand` is gives the row number for each entry and the variable `Epil$Ind` gives the patient number.
These variables can be used to define the random effects $\nu_{ij}$ and $\epsilon_i$ in INLA as follows.
The usual R formula notation (e.g. `y ~ x1 + x2`) is used for the rest of the linear predictor.

```{r}
formula <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT +
  f(rand, model = "iid", hyper = tau_prior) +  # Nu random effect
  f(Ind,  model = "iid", hyper = tau_prior)    # Epsilon random effect

epil_inla <- function(strat) {
  inla(formula,
       control.fixed = list(mean = 0, prec = 1/100^2), # Beta prior
       family = "poisson",
       data = Epil,
       control.inla = list(strategy = strat),
       control.predictor = list(compute = TRUE))
}

fit2 <- epil_inla(strat = "gaussian")
fit3 <- epil_inla(strat = "simplified.laplace")
fit4 <- epil_inla(strat = "laplace")
```

## TMB

The objective function `obj$fn` and its gradient `obj$gn` are a function of only the parameters, which in this instance are the six regression coefficients $\beta$ together with the logarithms of $\tau_\epsilon$ and $\tau_\nu$.
This can be checked with `names(obj$par)`.

```{r}
compile("epil.cpp")
dyn.load(dynlib("epil"))
```

These are the initialisation parameters:

```{r}
param <- list(
  beta = rep(0, K),
  epsilon = rep(0, N),
  nu = rep(0, N * J),
  l_tau_epsilon = 0,
  l_tau_nu = 0
)
```

`random` are integrated out with a Laplace approximation:

```{r}
obj <- MakeADFun(
  data = dat,
  parameters = param,
  random = c("epsilon", "nu"),
  DLL = "epil"
)
```

Optimise `obj` using 1000 iterations of the the `nlminb` optimiser, passing in the starting values `start`, objective function `objective` and its derivative `gradient`:

```{r results='hide'}
its <- 1000 #' May converge before this
opt <- nlminb(
  start = obj$par,
  objective = obj$fn,
  gradient = obj$gr,
  control = list(iter.max = its, trace = 0)
)

sd_out <- sdreport(
  obj,
  par.fixed = opt$par,
  getJointPrecision = TRUE
)

#' Check that TMB and Stan have the same objective function for sanity.
#' To obtain the TMB negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
#' For Stan, we create an empty model, then use `rstan::log_prob`.

# tmb_nll <- MakeADFun(data = dat, parameters = param, DLL = "epil") # TMB objective
# stan_nll <- stan("model/epil.stan", data = dat, chains = 0) # Stan objective

#' Now test the NLL of the initialisation parameters `param`:

# c(
#   "TMB" = tmb_nll$fn(unlist(param)),
#   "Stan" = -rstan::log_prob(object = stan_nll, unlist(param))
# )

#' To be more thorough, we can get all the parameters obtained using the MCMC:

# pars_mat <- as.matrix(fit1)
# pars_list <- apply(pars_mat, 1, function(x) relist(flesh = x, skeleton = fit1@inits[[1]]))
# upars_list <- lapply(pars_list, function(x) unconstrain_pars(fit1, x))

#' Just use a few of them

# test_pars <- upars_list[1001:1010]
# tmb_evals <- sapply(test_pars, tmb_nll$fn)
# stan_evals <- -sapply(test_pars, FUN = rstan::log_prob, object = stan_nll)
#
# data.frame(
#   "TMB" = tmb_evals,
#   "Stan" = stan_evals
# )

#' Note: When there are errors in the C++ template code, usually to do with indexing (unlike Stan, in TMB there is no requirement when defining variables to give them dimensions), calling the `MakeADFun` function tends to crash the R session.
#' A workaround (courtesy of Kinh) for debugging without crashing the working R session is to use the following (which creates new R sessions which crash in preference to the working R session).

# library(parallel)
# testrun <- mcparallel({MakeADFun(data = dat,
#                                  parameters = param,
#                                  DLL = "epil")})
# obj <- mccollect(testrun, wait = TRUE, timeout = 0, intermediate = FALSE)
```

## glmmTMB

[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in TMB without writing the C++ code.
For example, rather than writing `epil.cpp` manually, we could have called the following:

```{r}
formula6 <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT + (1 | rand) + (1 | Ind)
fit6 <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"))
```

## tmbstan

[`tmbstan`](https://journals.pl,os.org/plosone/article?id=10.1371/journal.pone.0197954) (Cole Monnahan and Kasper Kristensen) is another helpful TMB package which allows you to pass the same C++ template you use in TMB to Stan in order to perform NUTS (if you have standard C++ code then this can also likely be done using [`stanc`](https://statmodeling.stat.columbia.edu/2017/03/31/running-stan-external-c-code/)).

```{r}
fit7 <- tmbstan(obj = obj, chains = 4)
```

## aghq

[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer.
One approach is to use `glmmTMB` to get the TMB template which can be passed the `aghq`.

```{r}
# glmm_model_info <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"), doFit = FALSE)
#
# glmm_ff <- with(glmm_model_info, {
#   TMB::MakeADFun(
#     data = data.tmb,
#     parameters = parameters,
#     random = names(parameters)[grep("theta", names(parameters), invert = TRUE)],
#     DLL = "glmmTMB",
#     silent = TRUE
#   )
# })
#
# glmm_quad <- aghq::marginal_laplace_tmb(glmm_ff, k = 3, startingvalue = glmm_ff$par)
```

Another, more standard, approach is to use the `TMB` template, `obj`, that we already have:

```{r results='hide'}
fit8 <- aghq::marginal_laplace_tmb(
  obj,
  k = 2,
  startingvalue = c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
)

summary8 <- summary(fit8)
```

## Comparison

```{r}
stan1 <- as.vector(t(summary(fit1)$summary[1:6, c(1, 3)]))
inla2 <- as.vector(t(fit2$summary.fixed[1:6, 1:2]))
inla3 <- as.vector(t(fit3$summary.fixed[1:6, 1:2]))
inla4 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
tmb5 <- as.vector(t(data.frame(sd_out$par.fixed[1:6], sqrt(diag(sd_out$cov.fixed)[1:6]))))
glmmtmb6 <- as.vector(t(summary(fit6)$coefficients$cond[, c("Estimate", "Std. Error")]))
tmbstan7 <- as.vector(t(summary(fit7)$summary[1:6, c(1, 3)]))
aghq8 <- as.vector(t(summary8$summarytable[1:6, c(1, 4)]))

df <- cbind(stan1, inla2, inla3, inla4, tmb5, glmmtmb6, tmbstan7, aghq8) %>%
  as.data.frame() %>%
  mutate(type = gl(2, 1, 12, labels = c("Mean", "SD")))

beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
rownames(df) <- c(sapply(0:5, beta_i))
colnames(df) <- c("Stan", "INLA_G", "INLA_SL", "INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "type")

saveRDS(df, "comparison-results.rds")
```

Assuming tmbstan to be the ground truth, we can do two dimensional plots to check the fit

```{r}
ggplot(df) +
  geom_point(aes(x = tmbstan, y = TMB, col = type)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5)
```

We use symmetric KL. From the Google group: "... its the symmetric version between the posterior marginal computed using Gaussian approximation and the one use one of the Laplace based ones.".
For `"gaussian"` the `kld` value is small but non-zero, why?

```{r}
head(fit2$summary.random$rand)$kld
```

For `laplace` and `"simplified.laplace"`:

```{r}
skld <- function(fit) {
  #' Nu random effect
  id_rand <- which.max(fit$summary.random$rand$kld)
  print(fit$summary.random$rand[id_rand, ])
  #' Epsilon random effect
  id_Ind <- which.max(fit$summary.random$Ind$kld)
  print(fit$summary.random$Ind[id_Ind, ])
  return(list(id_rand = id_rand, id_Ind = id_Ind))
}

skld3 <- skld(fit3)
skld4 <- skld(fit4)
```

```{r}
plot_marginals <- function(random_effect, index) {
  marginal2 <- fit2$marginals.random[[random_effect]][paste0("index.", index)][[1]]
  marginal3 <- fit3$marginals.random[[random_effect]][paste0("index.", index)][[1]]
  marginal4 <- fit4$marginals.random[[random_effect]][paste0("index.", index)][[1]]

  plot(marginal2, type = "l", col = "red")
  lines(marginal3, col = "green")
  lines(marginal4, col = "blue")
}

plot_marginals("rand", index = skld3$id_rand)
plot_marginals("Ind", index = skld3$id_Ind)
```

### 2022-09-07 src/prev-anc-art_process-results/prev-anc-art.Rmd
prev-anc-art html up

---
title: "Prevalence, ANC, ART model"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
---

```{r setup, class.source = 'fold-hide'}
knitr::opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  cache.lazy = FALSE,
  cache.comments = FALSE
)
options(scipen = 999)
cbpalette <- multi.utils::cbpalette()
library(tidyverse)
rstan::rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

source("functions.R")
```

Create true values dataframe:

```{r}
true_values <- data.frame(
  parameter = c(
    "beta_prev",
    "log_sigma_phi_prev",
    "beta_anc",
    "log_sigma_b_anc",
    "beta_art",
    "log_sigma_phi_art"
  ),
  true_value = c(
    -2.4,
    log(sqrt(1 / 4)),
    -0.2,
    log(sqrt(1/ 100)),
    0.7,
    log(0.35)
  )
)
```

Read in results created in reports:

* [Model 0](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model0)
* [Model 1](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model1)
* [Model 2](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model2)
* [Model 3](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model3)
* [Model 4](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model4)

```{r}
# model0 <- readRDS("depends/results_model0.rds")
model1 <- readRDS("depends/results_model1.rds")
model2 <- readRDS("depends/results_model2.rds")
model3 <- readRDS("depends/results_model3.rds")
model4 <- readRDS("depends/results_model4.rds")
```

## Boxplots

```{r cache=FALSE}
# draw_boxplots(model0) + labs(title = "Model 0")
# draw_boxplots(model1) + labs(title = "Model 1")
# draw_boxplots(model2) + labs(title = "Model 2")
# draw_boxplots(model3) + labs(title = "Model 3")
# draw_boxplots(model4) + labs(title = "Model 4")
```

## Scatterplots

```{r fig.height=10}
# draw_scatterplots(model0) + labs(title = "Model 0")
draw_scatterplots(model1) + labs(title = "Model 1")
draw_scatterplots(model2) + labs(title = "Model 2")
draw_scatterplots(model3) + labs(title = "Model 3")
draw_scatterplots(model4) + labs(title = "Model 4")
```

## Kolmogorov KS test plots

Take samples from all distributions, then compute maximum ECDF difference $D$ (two-sample Kolmogorov–Smirnov test).
We assume that `tmbstan` represents the gold-standard, and compare performance of `aghq` and `TMB` via their similarity to this gold-standard.
Higher values of $D$ correspond to distributions which are more different.

```{r fig.height=8}
# draw_ksplots_D(model0) + labs(title = "Model 0")
draw_ksplots_D(model1) + labs(title = "Model 1")
draw_ksplots_D(model2) + labs(title = "Model 2")
draw_ksplots_D(model3) + labs(title = "Model 3")
draw_ksplots_D(model4) + labs(title = "Model 4")
```

We could also assess $l$, the location of $D$.
Determining if there are patterns in the location of the greatest ECDF difference could present us with useful insights.

```{r fig.height=8}
# draw_ksplots_D(model0) + labs(title = "Model 0")
draw_ksplots_l(model1) + labs(title = "Model 1")
draw_ksplots_l(model2) + labs(title = "Model 2")
draw_ksplots_l(model3) + labs(title = "Model 3")
draw_ksplots_l(model4) + labs(title = "Model 4")
```

## MCMC diagnostics

When using Markov chain Monte Carlo (MCMC) methods, as we have for `tmbstan`, it's important to assess for convergence.

### Traceplots

One way to do this is via traceplots which visualise the chains over the number of iterations specified.

```{r}
# map(model0, "mcmc_traceplots")
map(model1, "mcmc_traceplots")
map(model2, "mcmc_traceplots")
map(model3, "mcmc_traceplots")
map(model4, "mcmc_traceplots")
```

### R hat

The $\hat R$ statistic ("R hat") can also be used.
A value of $\hat R < 1.1$ is typically sufficient.

```{r}
# draw_rhatplot(model0) + labs(title = "Model 0")
draw_rhatplot(model1) + labs(title = "Model 1")
draw_rhatplot(model2) + labs(title = "Model 2")
draw_rhatplot(model3) + labs(title = "Model 3")
draw_rhatplot(model4) + labs(title = "Model 4")
```

### 2022-09-08 src/example_inla-grid/inla-grid.Rmd
Iterate inla-grid (add descriptions)

---
title: "INLA grid"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
---

```{r}
cbpalette <- multi.utils::cbpalette()
```

Suppose that $\theta = (\theta_1, \theta_2)$ has a multivariate Gaussian distribution with mean vector $\mu$ and covariance matrix $\Sigma$:

```{r}
mu <- c(1, 1.5)
cov <- matrix(c(2, 1, 1, 1), ncol = 2)
```

The probability density function (PDF) is:

```{r}
obj <- function(theta) {
  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
}

grid <- expand.grid(
  theta1 = seq(-2, 5, length.out = 700),
  theta2 = seq(-2, 5, length.out = 700)
)

ground_truth <- cbind(grid, pdf = obj(grid))

ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
  geom_contour(col = cbpalette[1]) +
  coord_fixed(xlim = c(-2, 5), ylim = c(-2, 5), ratio = 1) +
  theme_minimal()
```

Find the mode via optimisation (the answer is clearly `mu` but to be complete):

```{r}
opt <- optim(par = c(0, 0), fn = obj, control = list(fnscale = -1))
opt
```

The Hessian is given by `cov` (should compute this numerically).
Using the Eigendecomposition $\Sigma = V \Lambda V^\top$:

```{r}
Lambda <- diag(eigen(cov)$values) #' Diagonal matrix
V <- eigen(cov)$vectors #' Eigenvector matrix
V %*% Lambda %*% t(V) #' Verifying that this is cov
```

Create a mapping $\theta(z)$ from the $z$-coordinates to the space of $\theta$:

```{r}
z_to_theta <- function(z) {
  as.vector(opt$par + V %*% sqrt(Lambda) %*% z)
}

theta_mode <- z_to_theta(c(0, 0)) # c(0, 0) maps to the mode
theta_mode

test_statistic <- function(theta_proposal) {
  log(obj(theta_mode)) - log(obj(theta_proposal))
}
```

Create a function to explore in $z$-coordinates along one dimension:

```{r}
#' @param m is dim(theta)
#' @param j is the index of exploration direction
#' @param delta_z is the step size
#' @param delta_pi is the acceptable drop-off
explore_direction <- function(j, m, delta_z, delta_pi) {
  z_mode <- rep(0, m)
  z_names <- paste0("z", 1:m)
  names(z_mode) <- z_names

  unit_vector <- rep(0, m)
  unit_vector[j] <- 1

  points <- rbind(z_mode)

  # Increasing
  i <- 0
  condition <- TRUE
  while(condition) {
    i <- i + 1
    proposal <- c(z_mode + i * delta_z %*% unit_vector)
    names(proposal) <- z_names
    statistic <- test_statistic(z_to_theta(proposal))
    condition <- (statistic < delta_pi)
    if(condition){
      points <- rbind(points, proposal)
    }
  }

  # Decreasing
  i <- 0
  condition <- TRUE
  while(condition) {
    i <- i + 1
    proposal <- c(z_mode - i * delta_z %*% unit_vector)
    names(proposal) <- z_names
    statistic <- test_statistic(z_to_theta(proposal))
    condition <- (statistic < delta_pi)
    if(condition){
      points <- rbind(points, proposal)
    }
  }
  as.data.frame(points)
}
```

I chose values for the step size and drop-off manually based on what looked good:

```{r}
d_z <- 0.2
d_pi <- 2
```

Expand the scaffold in $z$-space:

```{r}
z_grid <- expand.grid(
  z1 = explore_direction(1, 2, delta_z = d_z, delta_pi = d_pi)$z1,
  z2 = explore_direction(2, 2, delta_z = d_z, delta_pi = d_pi)$z2
)
```

Map it back to $\theta$-space:

```{r}
theta_grid_full <- t(apply(z_grid, 1, z_to_theta)) %>%
  as.data.frame() %>%
  rename(theta1 = V1, theta2 = V2)
```

Keep only those points which meet condition:

```{r}
theta_grid_full %>%
  mutate(statistic = apply(theta_grid_full, 1, test_statistic),
         condition = statistic < d_pi) %>%
  filter(condition == TRUE) -> theta_grid
```

```{r}
#' Eigenvector line segments
rbind(c(theta_mode[1], theta_mode[2], (theta_mode + V[,1])[1], (theta_mode + V[,1])[2], 1),
      c(theta_mode[1], theta_mode[2], (theta_mode + V[,2])[1], (theta_mode + V[,2])[2], 2)) %>%
  as.data.frame() %>%
  rename(x = V1, y = V2, xend = V3, yend = V4, pc = V5) -> segments

ggplot() +
  geom_point(data = theta_grid, aes(x = theta1, y = theta2), alpha = 0.5) +
  geom_contour(data = ground_truth, aes(x = theta1, y = theta2, z = pdf), col = cbpalette[1]) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend, colour = as.factor(pc)),
               data = segments, size = 1, arrow = arrow(length = unit(0.3, "inches")),
               lineend = "round", linejoin = "round") +
  scale_colour_manual(values = cbpalette[-1]) +
  labs(x = "theta1", y = "theta2") +
  theme_minimal() +
  theme(legend.position = "none")
```

### 2022-09-08 src/prev-anc-art_process-results/prev-anc-art.Rmd
Iterate prev-anc-art writing forward

@@
 ---
-title: "Prevalence, ANC, ART model"
+title: "Prevalence, ANC, ART model inference case-study"
 author:
@@
     theme: lumen
+bibliography: citations.bib
+abstract: |
+    **Background** The HIV inference group at Imperial routinely uses `TMB` to implement complex spatio-temporal statistical models because the scale and requirement for in-production run-time makes MCMC infeasible, and no suitably flexible implementation of INLA exists.
+However, inferences obtained from `TMB` are not as accurate as from MCMC or INLA.
+
+    **Task** We compare inferences produced using tmbstan, `TMB` and `aghq` for a collection of HIV evidence synthesis models based upon the Naomi small-area estimation model.
+
+    **Findings**
+
+    **Next steps**
 ---
+
+## Model description
+
+@eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
+This model forms the basis of the Naomi small-area estimation model [@eaton2021naomi]. 
+Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation, as well as prompting investigation into any data conflicts.
+The model is described by three components, as follows.
+
+### Prevalence component
+
+Consider a country partitioned into areas $i = 1, \ldots, n$.
+A simple random household survey of $m_i$ people is conducted in each area, and $y_i$ HIV positive cases are observed.
+Cases may be modelled using a binomial logistic regression model
+\begin{align}
+y_i &\sim \text{Bin}(m_i, \rho_i), \\
+\text{logit}(\rho_i) &\sim \mathcal{N}(\beta_\rho, \sigma_\rho^2)
+\end{align}
+where HIV prevalence $\rho_i$ is modelled by a Gaussian with mean $\beta_\rho$ and standard deviation $\sigma_\rho$.
+
+### ANC component
+
+Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
+If, of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
+\begin{align}
+y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + b_i, \\
+b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
+\end{align}
+may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
+Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence.
+
+### ART component
+
+The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence.
+Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
+ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
+As such, ART coverage may also be modelled using a binomial logistic regression model
+\begin{align}
+A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
+\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
+\end{align}
+
+## Implementation
+
+We consider five models as described below.
+For each model we write a `TMB` C++ template.
+As well as the standard `TMB` inference approach, this template allows the model to be fit using Stan via `tmbstan` and using adaptive Gaussian Hermite quadrature via `aghq`.
+
+| Model | Report | Components |
+|-------|--------| ---------- |
+| 0 | [`prev-anc-art_model0`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model0) | Prevalence (no random effects) |
+| 1 | [`prev-anc-art_model1`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model1) | Prevalence |
+| 2 | [`prev-anc-art_model2`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model2) | Prevalence, ANC |
+| 3 | [`prev-anc-art_model3`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model3) | Prevalence, ART |
+| 4 | [`prev-anc-art_model4`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model4) | Prevalence, ANC, ART |
+
+We perform inference on simulated data from Model 4 generated ([`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim)) with the following parameter values:
+
+| Parameter | Value |
+|-----------|-------|
+| $n$              | 30   |
+| $m_i$            | 250  |
+| $\beta_\rho$     | -2.4 |
+| $\sigma_\rho$    | 0.5  |
+| $m^\text{ANC}_i$ | 10^4 |
+| $\beta_b$        | -0.2 |
+| $\sigma_b$       | 0.1  |
+| $N_i$            | 10^5 |
+| $\beta_\alpha$   | 0.7 |
+| $\sigma_\alpha$  | 0.35 |
+
+## Results
 
@@
 
-* [Model 0](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model0)
-* [Model 1](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model1)
-* [Model 2](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model2)
-* [Model 3](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model3)
-* [Model 4](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model4)
-
 ```{r}
@@
 
-## Boxplots
+### Boxplots
 
@@
 
-## Scatterplots
+### Scatterplots
 
-```{r fig.height=10}
+```{r fig.height=15}
 # draw_scatterplots(model0) + labs(title = "Model 0")
@@
 
-## Kolmogorov KS test plots
+### Kolmogorov KS test plots
 
@@
 # draw_ksplots_D(model0) + labs(title = "Model 0")
-draw_ksplots_l(model1) + labs(title = "Model 1")
-draw_ksplots_l(model2) + labs(title = "Model 2")
-draw_ksplots_l(model3) + labs(title = "Model 3")
-draw_ksplots_l(model4) + labs(title = "Model 4")
+# draw_ksplots_l(model1) + labs(title = "Model 1")
+# draw_ksplots_l(model2) + labs(title = "Model 2")
+# draw_ksplots_l(model3) + labs(title = "Model 3")
+# draw_ksplots_l(model4) + labs(title = "Model 4")
 ```
 
-## MCMC diagnostics
+### MCMC diagnostics
 
@@
 
-### Traceplots
+#### Traceplots
 
@@
 
-### R hat
+#### R hat
 
@@
 ```
+
+## Conclusion
+
+## Bibliography {-}

### 2022-09-08 src/example_inla-grid/inla-grid.Rmd
Some writing for inla-grid

@@
 ---
-title: "INLA grid"
+title: "INLA two dimensional grid replication"
 author:
@@
     theme: lumen
+abstract: |
+    **Background** The INLA method uses a grid of hyperparameters to integrate over.
+
+    **Task** We replicate a method for selecting grid vaues as outlined in Rue, Martino and Chopin (2009).
 ---
 
-```{r}
-cbpalette <- multi.utils::cbpalette()
-```
+## A two dimensional MVN
 
-Suppose that $\theta = (\theta_1, \theta_2)$ has a multivariate Gaussian distribution with mean vector $\mu$ and covariance matrix $\Sigma$:
+Suppose that $\theta = (\theta_1, \theta_2)$ has a multivariate Gaussian distribution with mean vector $\mu$ and covariance matrix $\Sigma$ as follows:
 
@@
 ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
-  geom_contour(col = cbpalette[1]) +
+  geom_contour(col = multi.utils::cbpalette()[1]) +
   coord_fixed(xlim = c(-2, 5), ylim = c(-2, 5), ratio = 1) +
@@
 opt <- optim(par = c(0, 0), fn = obj, control = list(fnscale = -1))
-opt
+opt$par
 ```
@@
 
-Create a mapping $\theta(z)$ from the $z$-coordinates to the space of $\theta$:
+## Explore the $z$-coordinates
+
+Create a mapping $\theta(z)$ from the $z$-coordinates to $\theta$-coordinates:
 
@@
 }
+```
 
-theta_mode <- z_to_theta(c(0, 0)) # c(0, 0) maps to the mode
+The origin `c(0, 0)` maps to the mode:
+
+```{r}
+theta_mode <- z_to_theta(c(0, 0))
 theta_mode
@@
 
-Expand the scaffold in $z$-space:
+Expand the scaffold in $z$-coordinates:
 
@@
 
-Map it back to $\theta$-space:
+Map it back to $\theta$-coordinates:
 
@@
   geom_point(data = theta_grid, aes(x = theta1, y = theta2), alpha = 0.5) +
-  geom_contour(data = ground_truth, aes(x = theta1, y = theta2, z = pdf), col = cbpalette[1]) +
+  geom_contour(data = ground_truth, aes(x = theta1, y = theta2, z = pdf), col = multi.utils::cbpalette()[1]) +
   geom_segment(aes(x = x, y = y, xend = xend, yend = yend, colour = as.factor(pc)),
@@
                lineend = "round", linejoin = "round") +
-  scale_colour_manual(values = cbpalette[-1]) +
+  scale_colour_manual(values = multi.utils::cbpalette()[-1]) +
   labs(x = "theta1", y = "theta2") +

### 2022-09-08 src/prev-anc-art_process-results/prev-anc-art.Rmd
Iterate prev-anc-art forward

@@
 
-    **Findings**
+    **Findings** We find that, treating `tmbstan` as the gold-standard, inference from `TMB` is substantially worse than than from `aghq` .
 
-    **Next steps**
+    **Next steps** Work on presentation of document. Understand performance for different parameters. Understand cause of poor performance.
 ---
@@
 | Model | Report | Components |
-|-------|--------| ---------- |
+|:------|:-------|:-----------|
 | 0 | [`prev-anc-art_model0`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model0) | Prevalence (no random effects) |
@@
 | Parameter | Value |
-|-----------|-------|
+|:----------|:------|
 | $n$              | 30   |
@@
 
-```{r setup, class.source = 'fold-hide'}
+```{r setup, message=FALSE, class.source = 'fold-hide'}
 knitr::opts_chunk$set(
@@
 
+Boxplots of the mean and standard deviation for each parameter, where each datapoint is a different simulation replicate.
+
 ```{r cache=FALSE}
-# draw_boxplots(model0) + labs(title = "Model 0")
-# draw_boxplots(model1) + labs(title = "Model 1")
-# draw_boxplots(model2) + labs(title = "Model 2")
-# draw_boxplots(model3) + labs(title = "Model 3")
-# draw_boxplots(model4) + labs(title = "Model 4")
+params <- c("beta_prev", "log_sigma_phi_prev")
+
+draw_boxplots(model1, params = params) + labs(title = "Model 1")
+draw_boxplots(model2, params = params) + labs(title = "Model 2")
+draw_boxplots(model3, params = params) + labs(title = "Model 3")
+draw_boxplots(model4, params = params) + labs(title = "Model 4")
 ```
@@
 
-```{r fig.height=15}
-# draw_scatterplots(model0) + labs(title = "Model 0")
-draw_scatterplots(model1) + labs(title = "Model 1")
-draw_scatterplots(model2) + labs(title = "Model 2")
-draw_scatterplots(model3) + labs(title = "Model 3")
-draw_scatterplots(model4) + labs(title = "Model 4")
+```{r}
+params <- c("beta_prev", "log_sigma_phi_prev")
+
+draw_scatterplots(model1, params = params) + labs(title = "Model 1")
+draw_scatterplots(model2, params = params) + labs(title = "Model 2")
+draw_scatterplots(model3, params = params) + labs(title = "Model 3")
+draw_scatterplots(model4, params = params) + labs(title = "Model 4")
 ```
@@
 ```{r fig.height=8}
-# draw_ksplots_D(model0) + labs(title = "Model 0")
 draw_ksplots_D(model1) + labs(title = "Model 1")
@@
 ```{r fig.height=8}
-# draw_ksplots_D(model0) + labs(title = "Model 0")
 # draw_ksplots_l(model1) + labs(title = "Model 1")
@@
 ```{r}
-# map(model0, "mcmc_traceplots")
 map(model1, "mcmc_traceplots")
@@
 ```{r}
-# draw_rhatplot(model0) + labs(title = "Model 0")
 draw_rhatplot(model1) + labs(title = "Model 1")

### 2022-09-14 src/docs_paper/paper.Rmd
Iterate draft

@@
 ---
-title: Simplifying Integrated nested Laplace approximation with adaptive Gaussian Hermite quadrature
+title: Simplified integrated nested Laplace approximation for extended latent Gaussian models
 author:
-  - name: Adam Howes
+  - name: Adam 
+    surname: Howes
     email: ath19@ic.ac.uk
-    affiliation: A
-    corresponding: ath19@ic.ac.uk
-address:
-  - code: A
-    address: Department of Mathematics, Imperial College London
+    label: e1
+    addressLabel: A
+    sepNext: "," # use "," or \and; the last author not need this field
+  - name: Alex
+    surname: Stringer
+    email: alex.stringer@uwaterloo.ca
+    label: e2
+    addressLabel: B
+affiliation:
+  - label: A
+    name: Department of Mathematics, Imperial College London
+    authorsLabels: e1
+  - label: B
+    name: Department of Statistics and Actuarial Science, University of Waterloo
+    authorsLabels: e2
 abstract: |
-  |  **Background:**
-  |
-  |  **Methods:**
-  |
-  |  **Results:**
-  |
-  |  **Conclusions:**
-author_summary: |
-  Author summary
+  | Naomi is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce HIV epidemic indicators.
+Performing inference for Naomi is challenging because it is an extended latent Gaussian model (ELGM) and not amenable to inference via INLA as described by Rue et al (2009).
+We combine the simplified INLA approach of Wood (2020) with adaptive Gaussian hermite quadrature to enable inference for ELGMs.
+Using data from Malawi, we compare our inference method to other approaches.
+We provide an easy to use implementation as a part of the `aghq` R package, allowing more flexible use of the INLA method for a broader class of models, including any with a TMB template.
+keyword-subclass: | 
+ \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
+ \kwd[Primary ]{00X00}
+ \kwd{00X00}
+ \kwd[; secondary ]{00X00}
+ \end{keyword}
+keywords: 
+  - Spatial statistics
+  - INLA
+
+predefined-theoremstyle: true # use in section Environments for Axiom, Theorem, etc
 bibliography: citations.bib
+biblio-style: imsart-nameyear # alternative: imsart-number
 output:
-  rticles::elsevier_article:
-    number_sections: true
+  rticles::ims_article:
+    journal: aoas # aap, aoas, aop, aos, sts. See documentation
+    toc: false # Please use for articles with 50 pages and more
     includes:
@@
 
-The `TMB` package \citep{kristensen2016tmb} is gaining popularity in spatial statistics as a flexible alternative to `R-INLA` for fitting latent Gaussian models \citep{osgoodzimmerman2021statistical}.
+We are motivated by a challenging inference problem in HIV surveillance.
+Accurate estimates of HIV indicators are required to mount effective public health response to the epidemic.
+The Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
+Software has been developed for the model allowing countries to use their data to generate estimates in a yearly process supported UNAIDS.
+As such, the inference method used must be suitable to run in production, ruling out prohibitively slow Markov chain Monte Carlo approaches.
+Furthermore, as Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast] it cannot be fit using the integrated nested Laplace approximation (INLA)  method [@rue2009approximate] as implemented by the `R-INLA` \textsf{R} package.
+Instead, inference is conducted using the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb], which implements a variety of empirical Bayes inference, and is gaining popularity in spatial statistics as a flexible alternative to `R-INLA` [@osgoodzimmerman2021statistical].
 
@@
 
-Integrated nested Laplace approximation (INLA) \citep{rue2009approximate} is an approximate Bayesian inference method which uses the Laplace approximation and numerical integration.
+INLA is an approximate Bayesian inference method based on the Laplace approximation and numerical integration.
 INLA is designed for use with latent Gaussian models (LGMs) of the form
@@
 \end{align}
-An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$, which we discuss in turn.
+An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$, which we discuss in turn below.
 
@@
 \end{equation}
-Approximation (i) then uses a Gaussian approximation
+Approximation (i) then uses a Gaussian approximation to the denominator of Equation \ref{eq:hyperpost} given by
 \begin{equation}
-p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \hat{\bm{\mu}}(\btheta), \hat{\mathbf{Q}}(\btheta)^{-1}) \label{eq:gaussianx}
+p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \hat{\bm{\mu}}(\btheta), \hat{\mathbf{Q}}(\btheta)^{-1}). \label{eq:gaussianx}
 \end{equation}
-to the denominator of Equation \ref{eq:hyperpost}.
-This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution
+This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution, given by
 \begin{equation}
 p(\x \, | \, \btheta, \y) \propto
-\exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right)
+\exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right),
 \end{equation}
-close to being Gaussian since $\y$ is generally not that informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved \citep{blangiardo2015spatial}.
+close to being Gaussian because $\y$ is generally not that informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial].
 <!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference?
@@
 -->
-
 As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}.
@@
 
-Having utilised used the approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section~\ref{sec:approxi}, a natural approach, and that taken by \citet{rue2007approximate}, is to marginalise this distribution directly
+Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section \ref{sec:approxi} above, a natural approach, and that taken by \citet{rue2007approximate}, is to marginalise this distribution directly to obtain
 \begin{equation}
@@
 \end{equation}
-where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from $\hat{\bm{\mu}}(\btheta)$ and $\mathbf{Q}^\star(\btheta)$ respectively.
+where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from $\hat{\bm{\mu}}(\btheta)$ and $\hat{\bm{Q}}(\btheta))$ respectively.
 Although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode.
@@
 \end{align}
-where $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{This notation is somewhat dangerous as $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
+where $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
 The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well.
 However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $x_i$ and $\btheta$, which is often computationally prohibitive.
-Two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by \citet{rue2009approximate} to reduce the computational cost:
+Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by \citet{rue2009approximate} to reduce the computational cost:
 \begin{enumerate}
@@
 
-# Template Model Builder \label{sec:tmb}
+## Template Model Builder \label{sec:tmb}
 
-Template Model Builder (TMB) \citep{kristensen2016tmb} is an \textsf{R} package for fitting random effect models, also known as latent variable models, hierarchical models or a host of other names.
-In \texttt{TMB} inference is based upon optimisation of a target function.
+`TMB` [@kristensen2016tmb] is an \textsf{R} package for fitting random effect models, also known as latent variable models, hierarchical models or a host of other names.
+In `TMB` inference is based upon optimisation of a target function.
 This makes it very flexible, and able to handle non-linear, non-Gaussian random effect models.
 
-The approach of \texttt{TMB} is inspired by the AD Model Builder (ADMB) package \citep{fournier2012ad}.
+The approach of `TMB` is inspired by the AD Model Builder (ADMB) package [@fournier2012ad].
 The "AD" in ADMB is automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
-AD is popular in machine learning \citep{baydin2017automatic}, for example as the basis for backpropagation algorithm and is beginning to gain popularity in statistics, including as a part of Stan \citep{carpenter2017stan}.
-\texttt{TMB} uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.
+AD is popular in machine learning [@baydin2017automatic], for example as the basis for backpropagation algorithm and is beginning to gain popularity in statistics, including as a part of Stan [@carpenter2017stan].
+`TMB` uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.
 
-## Statistical framework
-
-Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.\footnote{\citet{kristensen2016tmb} use the notation $u$ for random effects and $\theta$ for parameters.
-We aim for consistency with Section~\ref{sec:inla}.}
+Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.\footnote{\citet{kristensen2016tmb] use the notation $u$ for random effects and $\theta$ for parameters.
+We aim for consistency with Section \ref{sec:inla}.}
 Let $\ell(\x, \btheta) \triangleq - \log p(\y \, | \, \x, \btheta)$ be the negative joint log-likelihood.
-In \texttt{TMB}, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
+In `TMB`, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
 A standard maximum likelihood approach is to optimise
@@
 \end{equation}
-Integrating out the random effects directly, as in Equation~\ref{eq:1} above, is usually intractable because $\x$ is high-dimensional, so \citet[Equation 3]{kristensen2016tmb} use a Laplace approximation $L^\star_f(\btheta)$ based instead upon integrating out a Gaussian approximation to the random effects.
-This Laplace approximation is analogous to the INLA approximation $\tilde p(\btheta \, | \, \y)$ given in Section~\ref{sec:approxi}.
+Integrating out the random effects directly, as in Equation \ref{eq:1} above, is usually intractable because $\x$ is high-dimensional, so \citet[Equation 3]{kristensen2016tmb} use a Laplace approximation $L^\star_f(\btheta)$ based instead upon integrating out a Gaussian approximation to the random effects.
+This Laplace approximation is analogous to the INLA approximation $\tilde p(\btheta \, | \, \y)$ given in Section \ref{sec:approxi}.
 \begin{equation*}
@@
 where $\propto$ is used to mean proportional up to an additive constant.
-The parameters of the Gaussian approximation (Equation~\ref{eq:gaussianx}), are found in terms of $f$ via $\hat{\bm{\mu}}(\btheta) = \arg \min_\x f(\x, \btheta)$ and $\hat{\bm{Q}}(\btheta) = f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta)$ and must be recomputed for each value of $\btheta$.
+The parameters of the Gaussian approximation (Equation \ref{eq:gaussianx}), are found in terms of $f$ via $\hat{\bm{\mu}}(\btheta) = \arg \min_\x f(\x, \btheta)$ and $\hat{\bm{Q}}(\btheta) = f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta)$ and must be recomputed for each value of $\btheta$.
 Obtaining $\hat{\bm{\mu}}(\btheta)$ is known as the inner optimisation step.
 
-# Examples \label{sec:examples}
-
-## The Naomi small-area estimation model
+# The Naomi small-area estimation model
 
 @eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation as well as prompting investigation into any conflicts.
+This model forms the basis of the Naomi small-area estimation model [@eaton2021naomi]. 
+Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation, as well as prompting investigation into any data conflicts.
+The model is described by three components, as follows.
 
-**Prevalence component**
-  Consider a country partitioned into areas $i = 1, \ldots, n$, with a simple random household survey of $m_i$ people is conducted in each area with $y_i$ HIV positive cases observed.
+### Prevalence component
+
+Consider a country partitioned into areas $i = 1, \ldots, n$.
+A simple random household survey of $m_i$ people is conducted in each area, and $y_i$ HIV positive cases are observed.
 Cases may be modelled using a binomial logistic regression model
@@
 
-**ANC component**
+### ANC component
+
 Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
-If of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
+If, of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
 \begin{align}
@@
 
-**ART component**
-  The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence, whereby districts with high prevalence are likely to have a greater number of people receiving treatment.
+### ART component
+
+The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence.
+Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
 ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
@@
 
+# Results \label{sec:results}
+
+# Conclusions \label{sec:conclusions}
+
 ## Supporting information
 
-**Appendix A**: Statistical modelling \newline
-**Appendix B**: Supplementary tables and figures
+**Appendix A**: Supporting information
 

### 2022-09-15 src/docs_paper/paper.Rmd
Small fixes to paper

@@
 
-# Background
+# The Naomi model
 
-## Integrated nested Laplace approximation \label{sec:inla}
+@eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
+This model forms the basis of the Naomi small-area estimation model [@eaton2021naomi]. 
+Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation, as well as prompting investigation into any data conflicts.
+The model is described by three components, as follows.
+
+## Prevalence component
+
+Consider a country partitioned into areas $i = 1, \ldots, n$.
+A simple random household survey of $m_i$ people is conducted in each area, and $y_i$ HIV positive cases are observed.
+Cases may be modelled using a binomial logistic regression model
+\begin{align}
+y_i &\sim \text{Bin}(m_i, \rho_i), \\
+\text{logit}(\rho_i) &\sim \mathcal{N}(\beta_\rho, \sigma_\rho^2)
+\end{align}
+where HIV prevalence $\rho_i$ is modelled by a Gaussian with mean $\beta_\rho$ and standard deviation $\sigma_\rho$.
+
+## ANC component
+
+Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
+If, of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
+\begin{align}
+y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + b_i, \\
+b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
+\end{align}
+may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
+Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence.
+
+## ART component
+
+The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence.
+Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
+ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
+As such, ART coverage may also be modelled using a binomial logistic regression model
+\begin{align}
+A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
+\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
+\end{align}
+
+# Inference
+
+## Integrated nested Laplace approximation\label{sec:inla}
 
@@
 
-## Template Model Builder \label{sec:tmb}
+@wood2020simplified propose a "simplified" version of the INLA algorithm.
+
+## Template Model Builder\label{sec:tmb}
 
@@
 
-The approach of `TMB` is inspired by the AD Model Builder (ADMB) package [@fournier2012ad].
-The "AD" in ADMB is automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
-AD is popular in machine learning [@baydin2017automatic], for example as the basis for backpropagation algorithm and is beginning to gain popularity in statistics, including as a part of Stan [@carpenter2017stan].
+The approach of `TMB` is inspired by the AD Model Builder (ADMB) package [@fournier2012ad], where the "AD" stands for automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
+AD is popular in machine learning [@baydin2017automatic], including as the basis for backpropagation algorithm, and is beginning to gain popularity in statistics, including as a part of Stan [@carpenter2017stan].
 `TMB` uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.
 
-Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.\footnote{\citet{kristensen2016tmb] use the notation $u$ for random effects and $\theta$ for parameters.
-We aim for consistency with Section \ref{sec:inla}.}
+Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.^[@kristensen2016tmb use the notation $u$ for random effects and $\theta$ for parameters.
+We aim for consistency with Section \ref{sec:inla}.]
 Let $\ell(\x, \btheta) \triangleq - \log p(\y \, | \, \x, \btheta)$ be the negative joint log-likelihood.
@@
 
-# The Naomi small-area estimation model
+# Case-study\label{sec:results}
 
-@eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-This model forms the basis of the Naomi small-area estimation model [@eaton2021naomi]. 
-Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation, as well as prompting investigation into any data conflicts.
-The model is described by three components, as follows.
+# Discussion\label{sec:conclusions}
 
-### Prevalence component
+# Appendix {#appendix .unnumbered}
 
-Consider a country partitioned into areas $i = 1, \ldots, n$.
-A simple random household survey of $m_i$ people is conducted in each area, and $y_i$ HIV positive cases are observed.
-Cases may be modelled using a binomial logistic regression model
-\begin{align}
-y_i &\sim \text{Bin}(m_i, \rho_i), \\
-\text{logit}(\rho_i) &\sim \mathcal{N}(\beta_\rho, \sigma_\rho^2)
-\end{align}
-where HIV prevalence $\rho_i$ is modelled by a Gaussian with mean $\beta_\rho$ and standard deviation $\sigma_\rho$.
-
-### ANC component
-
-Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
-If, of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
-\begin{align}
-y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
-\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + b_i, \\
-b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
-\end{align}
-may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
-Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence.
-
-### ART component
-
-The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence.
-Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
-ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
-As such, ART coverage may also be modelled using a binomial logistic regression model
-\begin{align}
-A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
-\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
-\end{align}
-
-# Results \label{sec:results}
-
-# Conclusions \label{sec:conclusions}
-
-## Supporting information
-
-**Appendix A**: Supporting information
-
-## Funding
+# Acknowledgements {#acknowledgements .unnumbered}
 
@@
 
-## Disclaimer
-
 # References {#references .unnumbered}

### 2022-09-15 src/example_blangiardo-replication/blangiardo-replication.Rmd
blangiardo-replication html up
Standardise .yml
Iterate blangiardo-replication

@@
 title: "Blangiardo replication"
-author: "Adam Howes"
-output: html_document
+author:
+- name: Adam Howes
+output:
+  html_document:
+    toc: yes
+    toc_float: true
+    toc_collapsed: true
+    df_print: paged
+    code_folding: show
+    theme: lumen
+abstract: |
+    **Background**
+
+    **Task**
+
+    **Findings**
+
+    **Next steps**
 ---
@@
 ```{r}
-y <- c(1.2697, 7.7637, 2.2532, 3.4557, 4.1776, 6.4320, -3.6623, 7.7567, 5.9032, 7.2671,
+y <- c(
+  1.2697, 7.7637, 2.2532, 3.4557, 4.1776, 6.4320, -3.6623, 7.7567, 5.9032, 7.2671,
   -2.3447, 8.0160, 3.5013, 2.8495, 0.6467, 3.2371, 5.8573, -3.3749, 4.1507, 4.3092,
-       11.7327, 2.6174, 9.4942, -2.7639, -1.5859, 3.6986, 2.4544, -0.3294, 0.2329, 5.2846)
+  11.7327, 2.6174, 9.4942, -2.7639, -1.5859, 3.6986, 2.4544, -0.3294, 0.2329, 5.2846
+)
 
@@
   labs(title = "The data") +
-  geom_hline(yintercept = -3, col = "#666666", lty = "dashed")
+  geom_hline(yintercept = -3, col = "#666666", lty = "dashed") +
+  theme_minimal()
 ```
@@
 
-## `R-INLA`
+## Implementation using `R-INLA`
+
+This model is simple to fit using the `R-INlA` software directly:
 
@@
 
-fit <- inla(formula,
+fit <- inla(
+  formula,
   data = dat,
   control.family = list(hyper = theta_prior),
-            control.fixed = list(mean.intercept = x_0, prec.intercept = tau_0))
+  control.fixed = list(mean.intercept = x_0, prec.intercept = tau_0)
+)
 ```
 
-## R implementation
+## Implementation using R
 
@@
   theme(axis.text.y=element_blank()) +
-  labs(title = "Blangiardo integration points for the hyperparameters")
+  labs(title = "Blangiardo integration points for the hyperparameters") +
+  theme_minimal()
 ```
@@
 
-One of the INLA vignettes discusses how the user can set their own integration points--see `browseVignettes(package="INLA")`.
+One of the INLA vignettes discusses how the user can set their own integration points--see `browseVignettes(package = "INLA")`.
 The integration points that INLA uses can be found using `fit$joint.hyper`, which we can add on to the above plot:
@@
   labs(title = "INLA integration points for the hyperparameters") +
-  annotate("text", x = 0.175, y = 5e-41,
-           label = "Now the points are concentrated \n where there is higher density")
+  annotate(
+    "text", x = 0.175, y = 5e-41,
+    label = "Now the points are concentrated \n where there is higher density"
+  ) +
+  theme_minimal()
 ```
@@
 
-## `TMB`
+## Implementation in `TMB`
 
-C++ for the negative log joint posterior:
+C++ for the negative log joint posterior is given by:
 
@@
 ```{r}
-h <- MakeADFun(data = dat,
+h <- MakeADFun(
+  data = dat,
   parameters = param,
@@
   DLL = "blangiardo",
-               hessian = TRUE)
+  hessian = TRUE
+)
 ```
@@
 ```{r}
-f <- MakeADFun(data = dat,
+f <- MakeADFun(
+  data = dat,
   parameters = param,
   DLL = "blangiardo",
-               hessian = TRUE)
+  hessian = TRUE
+)
 ```
@@
 
-tmb_nlminb <- nlminb(start = h$par,
+tmb_nlminb <- nlminb(
+  start = h$par,
   objective = h$fn,
   gradient = h$gr,
-                     control = list(iter.max = its, trace = 0))
+  control = list(iter.max = its, trace = 0)
+)
 
-tmb_optim <- optim(par = h$par,
+tmb_optim <- optim(
+  par = h$par,
   fn = h$fn,
@@
   upper = 100, # when using optimize (1D function)
-                   control = list(maxit = its, trace = 0))
+  control = list(maxit = its, trace = 0)
+)
 
@@
 ```{r}
-kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$mode),
+#' For some reason the mode in the INLA result below is NA
+kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$mean),
                            exp(r_opt$par),

### 2022-09-15 src/docs_paper/paper.Rmd
Iterate paper (HS superscript, joint model start, some structure updates)

@@
   | Naomi is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce HIV epidemic indicators.
-Performing inference for Naomi is challenging because it is an extended latent Gaussian model (ELGM) and not amenable to inference via INLA as described by Rue et al (2009).
-We combine the simplified INLA approach of Wood (2020) with adaptive Gaussian hermite quadrature to enable inference for ELGMs.
+Performing inference for Naomi is challenging because it is an extended latent Gaussian model (ELGM) and not amenable to inference via INLA as described by @rue2009approximate.
+We combine the simplified INLA approach of @wood2020simplified with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other ELGMs.
 Using data from Malawi, we compare our inference method to other approaches.
-We provide an easy to use implementation as a part of the `aghq` R package, allowing more flexible use of the INLA method for a broader class of models, including any with a TMB template.
+We provide an easy to use implementation as a part of the `aghq` \textsf{R} package, allowing more flexible use of the INLA method for a broader class of models, including any with a `TMB` template.
 keyword-subclass: | 
@@
 The Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
-Software has been developed for the model allowing countries to use their data to generate estimates in a yearly process supported UNAIDS.
-As such, the inference method used must be suitable to run in production, ruling out prohibitively slow Markov chain Monte Carlo approaches.
+Software has been developed for the model allowing countries to use their data to generate estimates in a yearly process supported by UNAIDS.
+The inference method used must be fast enough to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo approaches.
 Furthermore, as Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast] it cannot be fit using the integrated nested Laplace approximation (INLA)  method [@rue2009approximate] as implemented by the `R-INLA` \textsf{R} package.
-Instead, inference is conducted using the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb], which implements a variety of empirical Bayes inference, and is gaining popularity in spatial statistics as a flexible alternative to `R-INLA` [@osgoodzimmerman2021statistical].
+Inference has so far been conducted using the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
+`TMB` implements a type of empirical Bayes, and is gaining popularity in spatial statistics as a flexible alternative to `R-INLA` [@osgoodzimmerman2021statistical].
 
@@
 @eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-This model forms the basis of the Naomi small-area estimation model [@eaton2021naomi]. 
-Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation, as well as prompting investigation into any data conflicts.
-The model is described by three components, as follows.
+This model forms the basis of the Naomi small-area estimation model, described more fully in @eaton2021naomi. 
+Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture, as well as prompting investigation into any data conflicts.
+The model is described by three components as follows.
 
-## Prevalence component
+## Household survey component
 
 Consider a country partitioned into areas $i = 1, \ldots, n$.
-A simple random household survey of $m_i$ people is conducted in each area, and $y_i$ HIV positive cases are observed.
+A simple random household survey of $m^\text{HS}_i$ people is conducted in each area, and $y^\text{HS}_i$ HIV positive cases are observed.
 Cases may be modelled using a binomial logistic regression model
 \begin{align}
-y_i &\sim \text{Bin}(m_i, \rho_i), \\
-\text{logit}(\rho_i) &\sim \mathcal{N}(\beta_\rho, \sigma_\rho^2)
+y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
+\text{logit}(\rho^\text{HS}_i) &= \phi_i \\
+\phi_i &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2),
 \end{align}
-where HIV prevalence $\rho_i$ is modelled by a Gaussian with mean $\beta_\rho$ and standard deviation $\sigma_\rho$.
+where HIV prevalence $\rho^\text{HS}_i$ is modelled by a Gaussian with mean $\beta_\phi$ and standard deviation $\sigma_\phi$.
 
@@
 Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
-If, of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
+Suppose that of $m^\text{ANC}_i$ women attending ANC, $y^\text{ANC}_i$ are HIV positive.
+Then an analogous binomial logistic regression model
 \begin{align}
 y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
-\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + b_i, \\
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho^\text{HS}_i) + b_i, \\
 b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
@@
 may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
-Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence.
+Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence on the logit scale.
 
@@
 
-The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence.
+The number of people receiving treatment at district health facilities $A_i$ provides further information about HIV prevalence.
 Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
-ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
+ART coverage, defined to be the proportion of people living with HIV (PLHIV) currently on ART on district $i$, is given by $\alpha_i = A_i / \rho^\text{HS}_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be constant.
 As such, ART coverage may also be modelled using a binomial logistic regression model
 \begin{align}
-A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
-\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
+A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
+\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2),
 \end{align}
+where the proportion of people receiving ART is $\rho^\text{HS}_i \alpha_i$.
+Here we assume no travel between districts to receive treatment.
 
-# Inference
+## Joint model
+
+Let $\y = (\y^\text{HS}, \y^\text{ANC}, \bm{A})$ be the complete response vector.
+In this section, we would like to set-up notation for the joint model, helping to show that it isn't a latent Gaussian model, but is an extended latent Gaussian model.
+When introducing LGM and ELGM below, we may refer back to this model as example.
+For example, we can note that each of the components individually might be latent Gaussian (perhaps for the ART component it depends if you consider $\rho^\text{HS}_i$ to be fixed) but when combined they are no longer.
+
+# Deterministic inference methods
 
@@
 
-INLA is an approximate Bayesian inference method based on the Laplace approximation and numerical integration.
+INLA is an approximate Bayesian inference method based on nested Gaussian approximations and numerical integration.
 INLA is designed for use with latent Gaussian models (LGMs) of the form
@@
 where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
-The joint posterior of $(\x, \btheta)$ is given by
+For such models, the joint posterior of $(\x, \btheta)$ is given by
 \begin{equation}
@@
 \end{align}
-An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$, which we discuss in turn below.
+An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$.
+We discuss these two approximations in turn below.
 
@@
 \end{equation}
-Approximation (i) then uses a Gaussian approximation to the denominator of Equation \ref{eq:hyperpost} given by
+Approximation (i) then uses a Gaussian approximation to the denominator given by
 \begin{equation}
@@
 \end{equation}
-close to being Gaussian because $\y$ is generally not that informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial].
+close to being Gaussian because $\y$ tends not to be strongly informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial].
 <!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference?
@@
 As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}.
-As such, taking $\x = \hat{\bm{\mu}}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as
+Taking $\x = \hat{\bm{\mu}}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as
 \begin{equation}
-\tilde p(\btheta \, | \, \y) \propto \frac{p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)} = \frac{p(\y, \hat{\bm{\mu}}(\btheta), \btheta)}{\det(\hat{\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal}
+\tilde p(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)} = \frac{p(\y, \hat{\bm{\mu}}(\btheta), \btheta)}{\det(\hat{\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal}
 \end{equation}
-where the final equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\hat{\bm{\mu}}(\btheta)$.
+where the equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\hat{\bm{\mu}}(\btheta)$.
 
@@
 
+We now move on to the approximation $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$.
 Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section \ref{sec:approxi} above, a natural approach, and that taken by \citet{rue2007approximate}, is to marginalise this distribution directly to obtain
@@
 \end{equation}
-where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from $\hat{\bm{\mu}}(\btheta)$ and $\hat{\bm{Q}}(\btheta))$ respectively.
-Although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode.
+where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from the relevant entries of $\hat{\bm{\mu}}(\btheta)$ and $\hat{\bm{Q}}(\btheta)$ respectively.
+However, although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode.
 As a result, although this method is available in \texttt{R-INLA} it is generally not advised.
@@
 \end{align}
-where $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
+where
+\begin{equation}
+p_G(\x_{-i} \, | \, x_i, \btheta, \y) = \mathcal{N}(\x_{-i} \, | \, \hat{\bm{\mu}}_{-i}(x_i, \btheta), \hat{\bm{Q}}_{-i}(x_i, \btheta)),
+\end{equation}
+is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
 The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well.
-However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $x_i$ and $\btheta$, which is often computationally prohibitive.
+However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $i$, which is often computationally prohibitive.
 Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by \citet{rue2009approximate} to reduce the computational cost:
@@
 
-### Combining the approximations
+## Simplified INLA and connection to ELGMs
 
-## Simplified INLA
+@wood2020simplified propose a simplified version of the INLA algorithm which relaxes the sparsity assumptions on the latent field required for the modifications to Equation \eqref{eq:laplaceapproximation} to be efficient.
+Doing so facilitates inference for ELGMs [@stringer2022fast] which we define below.
+Informally, ELGMs build on LGMs by allowing each element of the linear predictor $\mu_i$ to depend on any subset of elements from the latent field.
 
-@wood2020simplified propose a "simplified" version of the INLA algorithm.
+## Implementation within `aghq`\label{sec:aghq}
 
-## Template Model Builder\label{sec:tmb}
+### Adaptive Gaussian Hermite quadrature
+
+### Template Model Builder
 

### 2022-09-15 src/example_inla-replication/inla-replication.Rmd
Upload crazed attempts from two years ago to do this -- likely very confused but perhaps useful things in there to unpick now

--- 2022-09-15 src/example_blangiardo-replication/blangiardo-replication.Rmd
@@
 ---
-title: "Blangiardo replication"
+title: "INLA replication"
 author:
@@
 abstract: |
-    **Background**
-
-    **Task**
-
-    **Findings**
-
-    **Next steps**
+    | We demonstrate a step-by-step example of the INLA method implemented in R and TMB.
 ---
 
-Step-by-step example of the INLA method.
-The model for i.i.d. observations $y = (y_1, \ldots, y_n)$ is as follows
+```{r setup, message=FALSE, warning=FALSE, class.source = 'fold-hide'}
+knitr::opts_chunk$set(
+  echo = TRUE,
+  cache = TRUE,
+  autodep = TRUE,
+  cache.comments = FALSE
+)
+
+library(tidyverse)
+library(TMB)
+library(INLA)
+library(rmarkdown)
+
+midblue <- "#3D9BD0"
+midgreen <- "#00855A"
+midpink <- "#B3608E"
+```
+
+# Background
+
+Consider a three-stage model
+\begin{alignat}{2}
+  &\text{(Observations)}     &   y_i &\sim p(y_i \, | \, x_i, \theta), \quad i \in \mathcal{I}, \\
+  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x \, | \,\mathbf{0},  \mathbf{Q}(\theta)^{-1}), \\
+  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta), \label{eq:hyper}
+\end{alignat}
+with posterior
+\begin{equation}
+  p(x, \theta \, | \, y) \propto p(y, x, \theta) = p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta).
+\end{equation}
+
+After writing the negative log-target in a C++ file `model.cpp` it can be compiled and loaded into R:
+
+```{r eval=FALSE}
+compile("model.cpp")
+dyn.load(dynlib("model"))
+```
+
+Let $f(x, \theta) \equiv - \log p(y, x, \theta)$ where $y$ is assumed to be fixed.
+This objective (`f$fn`) and its derivative (`f$gn`) can be created (sketch) in `TMB` using:
+
+```{r eval=FALSE}
+f <- MakeADFun(data = dat, parameters = param, DLL = "model")
+```
+
+There are four stages to replicating the INLA method.
+
+## Stage 1: get $\tilde p(\theta \, | \, y)$
+
+<!-- Taking negative logs (to incorporate $f$) gives -->
+<!-- \begin{equation*} -->
+<!-- - \log p(\btheta \, | \, \y) \sim f(x, \btheta) + \log p(x \, | \, \btheta, \y) -->
+<!-- \end{equation*} -->
+
+Approximate the posterior marginal $p(\theta \, | \, y) \propto L(\theta)$ using the Laplace approximation $L(\theta) \approx L^\star(\theta)$:
+\begin{equation}
+\tilde p(\theta \, | \, y) \propto \frac{p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta)}{\tilde p_G(x \, | \, \theta, y)} \Big\rvert_{x = \mu^\star(\theta)} \equiv L^\star(\theta)
+\end{equation}
+
+Using `TMB` if we integrate out the latent field $x$ using the argument `random = "x"` in `MakeADFun` then the result is the negative log of the Laplace approximation to the marginal likelihood $h(\theta) \equiv - \log L^\star(\theta)$:
+
+```{r eval=FALSE}
+h <- MakeADFun(data = dat, parameters = param, DLL = "model", random = "x")
+```
+
+Therefore, `exp(-h$fn(theta))` is proportional to $\tilde p(\theta \, | \, y)$:
+
+<!-- *Note: Does `TMB` keep the $-\frac{n}{2} \log(2\pi)$ constant here $-\log L^\star_f(\btheta) = -\frac{n}{2} \log(2\pi) + \frac{1}{2} \log \det (Q^\star(\btheta)) + f(\bm{\mu}^\star(\btheta), \btheta)$?* -->
+
+## Stage 2: choose $\{\theta_k, \Delta_k\}_{k = 1}^K$
+
+Integration points $\{\theta_k\}$ with weights $\sum_{k = 1}^K \Delta_k = 1$.
+
+### Empirical Bayes
+
+$K = 1$ with integration point $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$.
+This can be found by minimising `h` using optimisers like `nlminb` or `optim`.
+
+### INLA grid
+
+The steps required are:
+
+1. Locate the mode of $\log p(\theta \, | \, y)$ which is the same as $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$ and can be calculated as above.
+2. Calculate negative Hessian at the mode.
+The solution is likely to use some version of `h$env$spHess` although it's not clear which one.
+3. Test points against the $\delta_z$ condition
+
+### Adaptive Gaussian Hermite quadrature
+
+## Stage 3: get $\tilde p(x_i \, | \, \theta, y)$
+
+### Gaussian
+
+Obtain marginals of $\tilde p_G(x \, | \, \theta, y)$ from above.
+
+### Laplace
+
+Compute the Laplace approximation
+
+### Simplified Laplace
+
+## Stage 4: apply quadrature
+
+\begin{equation}
+  \tilde p(x_i \, | \, y) = 
+  \sum_{k = 1}^K \tilde p(x_i \, | \, \theta^{(k)}, y) \times \tilde p(\theta^{(k)} \, | \, y) \times \Delta^{(k)}
+\end{equation}
+
+# Example 1: Blangiardo
+
+Consider the following simple model for i.i.d. observations $y = (y_1, \ldots, y_n)$
 \begin{align}
@@
 
-## Set-up
-
-```{r message=FALSE, warning=FALSE}
-midblue <- "#3D9BD0"
-midgreen <- "#00855A"
-midpink <- "#B3608E"
-```
-
-Load the data:
+We start by loading the data manually:
 
@@
 
-```{r, fig.height=2, fig.cap="The observations (ordered by index, though this has no interpretation). The prior (mean shown as dashed line) mismatches with the data."}
+The observations (ordered by index, though this has no interpretation) and prior (mean shown as dashed line, which is significantly different to the data):
+
+```{r, fig.height=2, fig.cap=""}
 data.frame(index = 1:30, y = y) %>%
@@
   geom_point(shape = 4) +
-  labs(title = "The data") +
   geom_hline(yintercept = -3, col = "#666666", lty = "dashed") +
@@
 
-Parameters of the prior distribution:
+We use the following parameters of the prior distribution:
 
@@
 
-Exact distribution of $x$ given $\theta$:
+The distribution $p(x \, | \, \theta, y)$ is exact:
 
 ```{r}
+# Exact distribution of x given theta
 inner_loop <- function(theta) {
@@
 In the one dimensional case there is no need to do the $z$-parametrisation.
-Choosing $\delta_z$ and $\delta_\pi$ based upon manual tuning (how does INLA select these numbers in general?):
+Choosing $\delta_z$ and $\delta_\pi$ based upon manual tuning^[It would be good to know how INLA selects these numbers in general]:
 
@@
   labs(title = "INLA integration points for the hyperparameters") +
-  annotate("text", x = 0.175, y = 5e-41,
-           label = "Now the points are concentrated \n where there is higher density")
+  annotate(
+    "text", x = 0.175, y = 5e-41,
+    label = "Now the points are concentrated \n where there is higher density"
+  ) +
+  theme_minimal()
 ```
 
-One of the INLA vignettes discusses how the user can set their own integration points--see `browseVignettes(package = "INLA")`.
-The integration points that INLA uses can be found using `fit$joint.hyper`, which we can add on to the above plot:
+One of the INLA vignettes discusses how the user can set their own integration points -- see `browseVignettes(package = "INLA")`.
+The integration points that `R-INLA` uses can be found using `fit$joint.hyper`, which we can add on to the above plot for interest:
 
@@
   labs(title = "INLA integration points for the hyperparameters") +
-  annotate(
-    "text", x = 0.175, y = 5e-41,
-    label = "Now the points are concentrated \n where there is higher density"
-  ) +
   theme_minimal()
@@
 but how should the range of $x$ be chosen?
-Blangiardo again use a naive grid for demonstration:
+Blangiardo use a naive grid for demonstration:
 
@@
 
-The $x$ and $y$-axis do not correspond to values, just to indicies:
+The $x$ and $y$-axis do not correspond to values here, just indices:
 
 ```{r}
-image(nl_joint_post)
+reshape2::melt(nl_joint_post) %>%
+  ggplot(aes(x = Var1, y = Var2, fill = value)) + 
+    geom_tile() +
+    theme_minimal() +
+    labs(fill = "", x = "x", y = "theta") +
+    theme(legend.position = "none")
 ```
@@
 
-```{r}
+```{r message=FALSE}
 its <- 1000
@@
 ```{r}
-#' For some reason the mode in the INLA result below is NA
-kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$mean),
+# For some reason the mode in the INLA result below is NA, how to fix this?
+kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$`0.5quant`),
                            exp(r_opt$par),
@@
 ```
+
+### Is `h$fn` the same as in R?
+
+Testing to see that `h$fn` is the same as `nl_post_marginal_theta` with `log_input = TRUE` up to a constant:
+
+```{r results=FALSE}
+vals <- seq(-5, 0, by = 1)
+
+tmb_vals <- sapply(vals, h$fn) # Replications show it not to be stochastic
+r_vals <- sapply(vals, nl_post_marginal_theta, log_input = TRUE)
+```
+
+```{r}
+rbind(tmb_vals, r_vals, tmb_vals - r_vals)
+```
+
+Could make a version of `h$fn` on the same scale as `post_marginal_theta` for a plot:
+```{r results=FALSE}
+tmb_post_marginal_theta <- function(theta) {
+  de_nl(h$fn, x = log(theta))
+} # (Have tested that this function does as it should)
+
+tmb_dense_theta <- eval_grid(
+  uniform = TRUE,
+  K = 500, min = 0.001, max = 0.3, 
+  f = tmb_post_marginal_theta
+)
+```
+
+```{r fig.cap = "At least this verifies that the optimisation is sensible."}
+# Normalised dense_theta and tmb_dense_theta since they include different constants
+ggplot(dense_theta) +
+  geom_line(aes(x = input, y = norm_output), col = midblue) +
+  geom_line(data = tmb_dense_theta, aes(x = input, y = norm_output), col = midpink) +
+  theme(axis.text.y=element_blank()) +
+  annotate("text", x = 0.1, y = 0.01, label = "TMB", col = midpink) +
+  annotate("text", x = 0.045, y = 0.012, label = "R", col = midblue) +
+  theme_minimal()
+```
+
+Does `TMB` do the inner optimisation differently?
+
+```{r results=FALSE, message=FALSE}
+get_tmb_inner_optimised_value <- function(l_theta) {
+  sd_out <- sdreport(h, par.fixed = l_theta)
+  reml <- sd_out$par.random # The REML estimator for x from TMB
+  return(reml)
+}
+
+tmb_inner <- sapply(vals, get_tmb_inner_optimised_value)
+r_inner <- sapply(exp(vals), function(theta) inner_loop(theta)$x_n)
+```
+
+```{r}
+rbind(tmb_inner, r_inner)
+```
+
+Looks like this part is correct.
+
+### Getting to the full conditional of $x$ in TMB
+
+Probably going to need $Q^\star(\btheta)$ ($n \times n$ precision matrix) and $\bm{\mu}^\star(\btheta)$ ($n \times 1$ mean vector).
+For example, the most basic version of the full conditional would be the Gaussian approximation
+\begin{equation}
+\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta)).
+\end{equation}
+
+Precision matrix $Q^\star(\btheta)$ is the Hessian of the negative logarithm of the posterior evaluated at $\bm{\mu}^\star(\btheta)$, having $(i, j)$th entry
+\begin{equation}
+Q^\star(\btheta)_{i, j} = - \frac{\partial^{2} \log p(x \, | \, \btheta, \y)}{\partial x_{i} \partial x_{j}} \Big\rvert_{x = \bm{\mu}^\star(\btheta)}
+\end{equation}
+
+An obvious choice for finding the precision would be `h$he(tmb_opt$par)` but this gives the error "Hessian not yet implemented for models with random effects".
+There may be another way to do it using `h$env$spHess()` (some kind of sparse Hessian) but if this is the case then it doesn't make sense to not have `h$he` just call the alternative way.
+This matrix is of the class `dsCMatrix` which is for symmetric, sparse numeric matrices in the compressed column-oriented format.
+
+`h$env$spHess` takes two arguments:
+
+  1. `par` (default `h$env$par`) should be both the hyperparameters and random effects
+  2. `random` (`TRUE` or `FALSE`)
+
+and the code is given by:
+
+```{r eval=FALSE}
+function(par = obj$env$par, random = FALSE) {
+  if (!random) {
+    Hfull@x[] <- ev(par) # Whats ev?
+    Hfull # H full <=> Full Hessian?
+  }
+  else if (skipFixedEffects) {
+    .Call("setxslot", Hrandom, ev(par), PACKAGE = "TMB")
+  }
+  else {
+    Hfull@x[] <- ev(par)
+    Hfull[r, r]
+  }
+}
+```
+
+Set a range of test parameters `test_pars`:
+
+```{r results=FALSE}
+full_pars <- function(input) {
+  setNames(input, c("x", "l_theta"))
+}
+
+par1 <- full_pars(c(0, 0))
+par2 <- full_pars(c(0, -1))
+par3 <- full_pars(c(-1, 0))
+
+# Doing the inner loop manually
+par4 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = 0), 0))
+par5 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = -1), -1))
+
+test_pars <- list(par1, par2, par3, par4, par5)
+```
+
+```{r}
+lapply(test_pars, h$env$spHess)
+```
+
+`random = TRUE` seems behave in unpredictable ways e.g.
+always returning the value calculated last when `random = FALSE` (perhaps only when using `apply` functions and not in other situations, so results here omitted):
+
+```{r results=FALSE}
+lapply(test_pars, h$env$spHess, random = TRUE)
+```
+
+1) With `random = FALSE` the number changes as a function of the second argument, `l_theta` (the hyperparameters).
+The Hessian is $2 \times 2$ with a dot ("structural zero") outside entry $(1, 1)$.
+2) With `random = TRUE` the number doesn't change at all and it's now a $1 \times 1$ matrix.
+
+Utilising the objective function $f$ which simply calculates the negative log posterior as a function of both random effects and hyperparameters:
+
+```{r}
+lapply(test_pars, f$he)
+```
+
+This produces the same answers as numerical differentiation of the same function (results omitted for space):
+
+```{r results=FALSE}
+lapply(test_pars, optimHess, fn = f$fn, gn = f$gn)
+```
+
+In this situation we know what the precision of $x$'s full conditional is (and its a function of $\theta$!):
+
+```{r}
+sapply(c(exp(0), exp(-1)), function(theta) inner_loop(theta)$tau_n)
+```
+
+Could look more at `sdreport` [here](https://github.com/kaskr/adcomp/blob/master/TMB/R/sdreport.R) see how these functions are used there (there are more comments on the Github page than when just looking via R).
+
+```{r}
+nl_full_conditional_x <- function(x, theta, log_input = FALSE) {
+
+  if(log_input == TRUE) {
+    theta <- exp(theta) # Convert from log(theta) to theta
+  }
+
+  par <- inner_loop(theta)
+
+  return(-dnorm(x, par$x_n, sqrt(1 / par$tau_n), log = TRUE))
+}
+```
+
+# Example 2: Unknown

### 2022-09-15 src/example_inla-replication/blangiardo.cpp

@@ -7,24 +7,24 @@ Type objective_function<Type>::operator()()
 {
   // Data block
   DATA_VECTOR(y);
-  
+
   // Parameter block
   PARAMETER(x);
   PARAMETER(l_theta);
-  
+
   // Transformed parameters block
   Type sigma = sqrt(1 / exp(l_theta));
-  
+
   Type nll;
   nll = Type(0.0);
-  
+
   // Priors
   nll -= dlgamma(l_theta, Type(1.6), Type(1 / 0.4), true);
   nll -= dnorm(x, Type(-3.0), Type(2.0), true);
 
   // Likelihood
   nll -= dnorm(y, x, sigma, true).sum();
-  
+
   // ADREPORT
   ADREPORT(x)
   ADREPORT(sigma);

### 2022-09-15 src/example_inla-replication/inla-replication.Rmd
Remove some broke tex

@@
 <!-- \begin{equation*} -->
-<!-- - \log p(\btheta \, | \, \y) \sim f(x, \btheta) + \log p(x \, | \, \btheta, \y) -->
+<!-- - \log p(\theta \, | \, y) \sim f(x, \theta) + \log p(x \, | \, \theta, y) -->
 <!-- \end{equation*} -->
@@
 
-<!-- *Note: Does `TMB` keep the $-\frac{n}{2} \log(2\pi)$ constant here $-\log L^\star_f(\btheta) = -\frac{n}{2} \log(2\pi) + \frac{1}{2} \log \det (Q^\star(\btheta)) + f(\bm{\mu}^\star(\btheta), \btheta)$?* -->
+<!-- *Note: Does `TMB` keep the $-\frac{n}{2} \log(2\pi)$ constant here $-\log L^\star_f(\theta) = -\frac{n}{2} \log(2\pi) + \frac{1}{2} \log \det (Q^\star(\theta)) + f(\mu^\star(\theta), \theta)$?* -->
 
@@
 
-Probably going to need $Q^\star(\btheta)$ ($n \times n$ precision matrix) and $\bm{\mu}^\star(\btheta)$ ($n \times 1$ mean vector).
+Probably going to need $Q^\star(\theta)$ ($n \times n$ precision matrix) and $\mu^\star(\theta)$ ($n \times 1$ mean vector).
 For example, the most basic version of the full conditional would be the Gaussian approximation
 \begin{equation}
-\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta)).
+\tilde p(x_i \, | \, \theta, y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\theta), 1 / q^\star_i(\theta)).
 \end{equation}
 
-Precision matrix $Q^\star(\btheta)$ is the Hessian of the negative logarithm of the posterior evaluated at $\bm{\mu}^\star(\btheta)$, having $(i, j)$th entry
+Precision matrix $Q^\star(\theta)$ is the Hessian of the negative logarithm of the posterior evaluated at $\mu^\star(\theta)$, having $(i, j)$th entry
 \begin{equation}
-Q^\star(\btheta)_{i, j} = - \frac{\partial^{2} \log p(x \, | \, \btheta, \y)}{\partial x_{i} \partial x_{j}} \Big\rvert_{x = \bm{\mu}^\star(\btheta)}
+Q^\star(\theta)_{i, j} = - \frac{\partial^{2} \log p(x \, | \, \theta, y)}{\partial x_{i} \partial x_{j}} \Big\rvert_{x = \mu^\star(\theta)}
 \end{equation}

### 2022-09-15 src/docs_paper/paper.Rmd
Iterate paper forward

@@
   | Naomi is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce HIV epidemic indicators.
-Performing inference for Naomi is challenging because it is an extended latent Gaussian model (ELGM) and not amenable to inference via INLA as described by @rue2009approximate.
-We combine the simplified INLA approach of @wood2020simplified with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other ELGMs.
+We combine the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
 Using data from Malawi, we compare our inference method to other approaches.
-We provide an easy to use implementation as a part of the `aghq` \textsf{R} package, allowing more flexible use of the INLA method for a broader class of models, including any with a `TMB` template.
+We provide an easy to use implementation as a part of the `aghq` \textsf{R} package allowing flexible use of the INLA method for a broader class of models, including any with a `TMB` template.
 keyword-subclass: | 
@@
 We are motivated by a challenging inference problem in HIV surveillance.
-Accurate estimates of HIV indicators are required to mount effective public health response to the epidemic.
-The Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
-Software has been developed for the model allowing countries to use their data to generate estimates in a yearly process supported by UNAIDS.
-The inference method used must be fast enough to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo approaches.
-Furthermore, as Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast] it cannot be fit using the integrated nested Laplace approximation (INLA)  method [@rue2009approximate] as implemented by the `R-INLA` \textsf{R} package.
-Inference has so far been conducted using the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-`TMB` implements a type of empirical Bayes, and is gaining popularity in spatial statistics as a flexible alternative to `R-INLA` [@osgoodzimmerman2021statistical].
+Mounting an effective public health response to the epidemic requires accurate, timely and sufficiently fine-scale estimates of HIV indicators.
+No single data source is sufficient to base these estimates on.
+For example, although nationally-representative household surveys are the most reliable data source, in most countries they only occur infrequently.
+As such, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
+Software (\url{https://naomi.unaids.org}) has been developed for Naomi, which allows countries to input their data and generate estimates in a yearly process supported by UNAIDS.
+For this reason, any inference method used must be fast enough to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
+Inference is currently conducted using the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
+`TMB` implements an empirical Bayes approach, and is gaining popularity in spatial statistics for its speed and flexibility [@osgoodzimmerman2021statistical].
+
+Using simulation we found inferences from `TMB` for the Naomi model to sometimes be inaccurate.
+The `aghq` \textsf{R} package [@stringer2021implementing] for adaptive Gaussian Hermite quadrature (AGHQ), significantly improved performance.
+We develop a new inference methodology combining the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with AGHQ.
+
+<!-- Furthermore, as Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast] it cannot be fit using the integrated nested Laplace approximation (INLA)  method [@rue2009approximate] as implemented by the `R-INLA` \textsf{R} package.
+-->
 
@@
 @eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-This model forms the basis of the Naomi small-area estimation model, described more fully in @eaton2021naomi. 
-Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture, as well as prompting investigation into any data conflicts.
-The model is described by three components as follows.
+This model forms the basis of the Naomi small-area estimation model, described fully in @eaton2021naomi. 
+Modelling data from multiple sources concurrently is attractive as it increases statistical power, mitigates the biases of any single source, and prompts investigation into any data conflicts.
+The model is comprised of three components, described as follows.
 
@@
 
-Consider a country partitioned into areas $i = 1, \ldots, n$.
-A simple random household survey of $m^\text{HS}_i$ people is conducted in each area, and $y^\text{HS}_i$ HIV positive cases are observed.
+Consider a country partitioned into $n$ areas indexed by $i$.
+Suppose a simple random household survey of $m^\text{HS}_i$ people is conducted in each area, and $y^\text{HS}_i$ HIV positive cases are observed.
 Cases may be modelled using a binomial logistic regression model
@@
 y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
-\text{logit}(\rho^\text{HS}_i) &= \phi_i \\
-\phi_i &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2),
+\text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2),
 \end{align}
@@
 
-# Deterministic inference methods
+# Fast inference methods
 
@@
 
-INLA is an approximate Bayesian inference method based on nested Gaussian approximations and numerical integration.
-INLA is designed for use with latent Gaussian models (LGMs) of the form
+INLA [@rue2009approximate] is an approximate Bayesian inference method based on nested Gaussian approximations and numerical integration, and designed for use with latent Gaussian models (LGMs) of the form
 \begin{alignat}{2}
@@
 
-# Case-study\label{sec:results}
+# Application to the Naomi model\label{sec:results}
 

### 2022-09-16 src/example_naomi/naomi.Rmd
Start working on Naomi vignette example

---
title: "Naomi model workflow"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
    **Background** A vignette for the Naomi model is uploaded to the package website.

    **Task** We run through the vignette manually to (1) develop a better understanding of the model, (2) get experience working with the Malawi data, which we will likely use for our case-study as it is publicly available.
---

```{r}
library(naomi)
library(tidyverse)
library(sf)
```

# Naomi vignette

Following vignette [here](https://mrc-ide.github.io/naomi/articles/model-workflow.html).

## Load data inputs

Area hierarchy and boundaries:

```{r}
area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))

area_merged %>%
  filter(area_level == max(area_level)) %>%
  ggplot() +
  geom_sf(aes(geometry = geometry)) +
  theme_minimal()
```

Population data:

```{r}
pop_agesex <- read_csv(system.file("extdata/demo_population_agesex.csv", package = "naomi"))
head(pop_agesex)
```

Survey data:

```{r}
survey_hiv_indicators <- read_csv(system.file("extdata/demo_survey_hiv_indicators.csv", package = "naomi"))
head(survey_hiv_indicators)
```
ART programmatic data:

```{r}
art_number <- read_csv(system.file("extdata/demo_art_number.csv", package = "naomi"))
head(art_number)
```

ANC programmatic data:

```{r}
anc_testing <- read_csv(system.file("extdata/demo_anc_testing.csv", package = "naomi"))
head(anc_testing)
```

## Choose model areas and time points

```{r}
scope <- "MWI"
level <- 4
calendar_quarter_t1 <- "CY2016Q1"
calendar_quarter_t2 <- "CY2018Q3"
calendar_quarter_t3 <- "CY2019Q4"
```

```{r}
prev_survey_ids  <- c("DEMO2016PHIA", "DEMO2015DHS")
artcov_survey_ids  <- "DEMO2016PHIA"
vls_survey_ids <- NULL
recent_survey_ids <- "DEMO2016PHIA"

artnum_calendar_quarter_t1 <- "CY2016Q1"
artnum_calendar_quarter_t2 <- "CY2018Q3"

anc_clients_year2 <- 2018
anc_clients_year2_num_months <- 9

anc_prevalence_year1 <- 2016
anc_prevalence_year2 <- 2018

anc_art_coverage_year1 <- 2016
anc_art_coverage_year2 <- 2018
```

## Prepare model inputs

```{r}
naomi_mf <- naomi_model_frame(
  area_merged,
  pop_agesex,
  spec,
  scope = scope,
  level = level,
  calendar_quarter_t1,
  calendar_quarter_t2,
  calendar_quarter_t3
)
```

```{r}
naomi_data <- select_naomi_data(
  naomi_mf,
  survey_hiv_indicators,
  anc_testing,
  art_number,
  prev_survey_ids,
  artcov_survey_ids,
  recent_survey_ids,
  vls_survey_ids,
  artnum_calendar_quarter_t1,
  artnum_calendar_quarter_t2,
  anc_prevalence_year1,
  anc_prevalence_year2,
  anc_art_coverage_year1,
  anc_art_coverage_year2
)
```

## Fit the TMB model

```{r}
fit <- fit_tmb(tmb_inputs)
```

```{r}
outputs <- output_package(fit, naomi_mf)
names(outputs)
```

```{r}
outputs$indicators %>%
  dplyr::filter(
    indicator == "prevalence",  # HIV prevalence
    age_group == "Y015_049"   # Age group 15-49
  ) %>%
  head()
```

```{r}
add_output_labels(outputs) %>%
  dplyr::filter(
    indicator == "prevalence",  # HIV prevalence
    age_group == "Y015_049"   # Age group 15-49
  ) %>%
  head()
```

```{r}
system.time(fit <- sample_tmb(fit))
```

```{r}
system.time(outputs <- output_package(fit, naomi_mf))

outputs_calib <- calibrate_outputs(
  outputs, 
  naomi_mf,
  spectrum_plhiv_calibration_level = "national",
  spectrum_plhiv_calibration_strat = "sex_age_coarse",
  spectrum_artnum_calibration_level = "national", 
  spectrum_artnum_calibration_strat = "sex_age_coarse",
  spectrum_aware_calibration_level = "national", 
  spectrum_aware_calibration_strat = "sex_age_coarse",
  spectrum_infections_calibration_level = "national", 
  spectrum_infections_calibration_strat = "sex_age_coarse"
)

outputs$indicators %>%
  dplyr::filter(
    indicator == "prevalence",  # HIV prevalence
    age_group == "Y015_049"   # Age group 15-49
  ) %>%
  head()
```

# Using simplified TMB template

Use the same data from above, but fit a model using our simplified Naomi TMB template, rather than the one from the package.
This may require making some edits to the simplified TMB template, such as allowing non-integer values for survey-weighted cases and sample sizes.

### 2022-09-16 src/epil/epil.Rmd
Add further exposition to epil

@@
 
-Section 5.2. of Rue, Martino and Chopin (2009) illustrates the INLA method using an epilepsy example.
+Section 5.2. of Rue, Martino and Chopin (2009) demonstrates the application of \texttt{R-INLA} to a generalised linear mixed model.
+The data, previously included as an example in the OpenBUGs manual, concern the results of a clinical trial for a new epilepsy drug.
+Patients $i = 1, \ldots, 59$ are each either assigned the new drug $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$.
+Each patient makes four visits the clinic $j = 1, \ldots, 4$.
+The observations $y_{ij}$ give the number of seizures of the $i$th person in the two weeks preceeding their $j$th visit to the clinic.
+The covariates are age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$, which are all centered.
+The observations are modelling using a Poisson distribution $y_{ij} \sim \text{Poisson}(e^{\eta_{ij}})$ with linear predictor 
+\begin{align*}
+  \eta_{ij}       &= \beta_0 + \beta_\texttt{Base} \log(\texttt{Baseline}_j / 4) + \beta_\texttt{Trt} \texttt{Trt}_i +
+                     \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
+                  &+ \beta_\texttt{Age} \log(\texttt{Age}_i) + \beta_{\texttt{V}_4} {\texttt{V}_4}_j +
+                     \epsilon_i + \nu_{ij}, \quad i=1:59, \quad j=1:4,
+\end{align*}
+where the prior on each of the regression parameters is $\mathcal{N}(0, 100^2)$.
+The random effects are IID $\epsilon_i \sim \mathcal{N}(0, 1/\tau_\epsilon)$ and $\nu_{ij} \sim \mathcal{N}(0, 1/\tau_\nu)$ with precision priors $\tau_\epsilon, \tau_\nu \sim \Gamma(0.001, 0.001)$.
 

### 2022-09-17 src/dev_sinla/sinla.Rmd
Working on SINLA notebook

---
title: "Implementing simplified INLA"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** Wood (2020) simplify the INLA method by approximating latent field posterior marginals using a reduced cost Laplace approximation which does not rely on sparsity assumptions.
  **Task** We implement the method in R making use of Template Model Builder for Gaussian approximations.
---

We are interested in approximating latent field posterior marginals as follows $p(x_i \, | \, \theta, y) \approx \tilde p(x_i \, | \, \theta, y)$ for each index $i$.
To build up to implementing the simplified INLA approach, we will start with two simpler methods.

# Marginal of joint Gaussian approximation

The `aghq` package currently approximates latent field posterior marginals by taking marginals of the joint Gaussian approximation $p_G(x \, | \, \theta, y)$.
This is same approach used by `method = "gaussian"` in `R-INLA`.
The Gaussian approximation $p_G(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_G(x \, | \, \theta, y)} \Big\rvert_{x = \hat{\mu}(\theta)} = \frac{p(y, \hat{\mu}(\theta), \theta)}{\det(\hat{Q}(\theta))^{1/2}},
\end{equation}
such that one easy way to calculate it is to use set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there.

We begin by loading in testing data, as generated by [`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim).
`sim_data` contains many replicates of simulated data, so we just take the first one:

```{r}
sim_data <- readRDS("depends/sim_data.rds")
data <- sim_data[[1]]
dat <- list(n = data$n, y_prev = data$y_prev, m_prev = data$m_prev)
```

We create an objective function `f` to evaluate $p(x, \theta, y)$ (without any Laplace approximations!):

```{r}
param <- list(
  beta_prev = 0,
  phi_prev = rep(0, data$n),
  log_sigma_phi_prev = 0
)

compile("model1.cpp")
dyn.load(dynlib("model1"))

f <- MakeADFun(
  data = dat,
  parameters = param,
  DLL = "model1"
)
```

# Full Laplace approximation

Another approach we might take is to calculate the full Laplace approximation
\begin{align}
  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_G(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \hat{\mu}_{-i}(x_i, \theta)} \\
  &= \frac{p(y, x_i, \hat{\mu}_{-i}(x_i, \theta), \theta)}{\det(\hat{Q}_{-i}(x_i, \theta))^{1/2}}
\end{align}
This is not practical in some settings as it involves recomputing the Gaussian approximation $p_G(x_{-i} \, | \, x_i, \theta, y)$ for each value of $(x_i, \theta)$ which can be too computationally expensive.
With of a bit of effort, we can implement this in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are named as such (see Appendix for template).

```{r}
compile("model1index.cpp")
dyn.load(dynlib("model1index"))
```

The `prepare_dat` function takes `dat` and makes it compatible with `model1index` for particular choice of marginal `i`.

```{r}
prepare_dat <- function(dat, i) {
  dat[["y_prev_i"]] <- dat$y_prev[i]
  dat[["y_prev_minus_i"]] <- dat$y_prev[-i]
  dat[["m_prev_i"]] <- dat$m_prev[i]
  dat[["m_prev_minus_i"]] <- dat$m_prev[-i]
  dat[c("n", "y_prev_i", "y_prev_minus_i", "m_prev_i", "m_prev_minus_i")]
}
```

Now loop over doing the Laplace approximation for every index.
In the call to `MakeADFun` we set `random = "phi_prev_minus_i"` to integrate all of the random effects but that of the particular index.

```{r}
#' To store obj
template <- list()

for(i in 1:dat$n) {
  dat_i <- prepare_dat(dat, i)

  #' Starting parameters for TMB are the same for every index of the loop
  param_i <- list(
    beta_prev = 0,
    phi_prev_i = 0,
    phi_prev_minus_i = rep(0, data$n - 1),
    log_sigma_phi_prev = 0
  )

  #' random are integrated out with a Laplace approximation
  obj <- MakeADFun(
    data = dat_i,
    parameters = param_i,
    random = "phi_prev_minus_i",
    DLL = "model1index"
  )

  its <- 1000

  opt <- nlminb(
    start = obj$par,
    objective = obj$fn,
    gradient = obj$gr,
    control = list(iter.max = its, trace = 0)
  )

  sd_out <- sdreport(
    obj,
    par.fixed = opt$par,
    getJointPrecision = TRUE
  )

  template[[i]] <- obj
}
```

```{r}
#' Evaluate objective function at param
f$fn(unlist(param))

transform_param <- function(param) {
  list(
    beta_prev = param$beta_prev,
    phi_prev_i = param$phi_prev[i],
    phi_prev_minus_i = param$phi_prev[-i],
    log_sigma_phi_prev = param$log_sigma_phi_prev
  )
}

param %>%
  transform_param() %>%
  unlist() %>%
  template[[1]]$fn()

#' The mean and precision matrix for the Gaussian approximation to p(x_{-i} | x_i, theta, y)
#' The mean of is length dim
mean <- with(template[[1]]$env, last.par[random])
Q <- template[[1]]$env$spHess(with(template[[1]]$env, last.par), random = TRUE)

#' Numerator of the LA
#' TODO

#' Denominator of the LA
sqrt(det(Q))
```

# Appendix

## `model1.cpp`

```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines('model1.cpp')}
```

## `model1index.cpp`

```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines('model1index.cpp')}
```

### 2022-09-17 src/dev_sinla/model1.cpp

@@ -1,0 +1,38 @@
+// model1.cpp
+
+#include <TMB.hpp>
+
+template <class Type>
+Type objective_function<Type>::operator()()
+{
+  // Data block
+  DATA_INTEGER(n);      // Number of regions
+  DATA_VECTOR(y_prev);  // Vector of survey responses
+  DATA_VECTOR(m_prev);  // Vector of survey sample sizes
+
+  // Parameter block
+  PARAMETER(beta_prev);            // Survey intercept
+  PARAMETER_VECTOR(phi_prev);      // Survey spatial effect
+  PARAMETER(log_sigma_phi_prev);   // Survey log standard deviation of spatial effects
+
+  // Transformed parameters block
+  Type sigma_phi_prev = exp(log_sigma_phi_prev);
+  vector<Type> eta_prev(beta_prev + sigma_phi_prev * phi_prev);
+
+  // Initialise negative log-likelihood
+  Type nll(0.0);
+
+  // Priors
+  nll -= dnorm(sigma_phi_prev, Type(0), Type(2.5), true) + log_sigma_phi_prev;
+  nll -= dnorm(beta_prev, Type(-2), Type(1), true);
+  nll -= dnorm(phi_prev, Type(0), Type(1), true).sum();
+
+  // Likelihood
+  nll -= dbinom_robust(y_prev, m_prev, eta_prev, true).sum();
+
+  // Generated quantities block
+  vector<Type> rho_prev(invlogit(eta_prev));    // Posterior prevalence estimates
+  Type tau_phi_prev = 1/pow(sigma_phi_prev, 2); // Precision of spatial effects
+
+  return(nll);
+}

### 2022-09-17 src/dev_sinla/script.R

@@ -2,97 +2,4 @@
 # orderly::orderly_develop_start("dev_sinla")
 # setwd("src/dev_sinla")
 
-sim_data <- readRDS("depends/sim_data.rds")
-data <- sim_data[[1]] #' Only need once instance of the simulated data
-
-compile("model1index.cpp")
-dyn.load(dynlib("model1index"))
-
-dat <- list(n = data$n, y_prev = data$y_prev, m_prev = data$m_prev)
-
-prepare_dat <- function(dat, i) {
-  dat[["y_prev_i"]] <- dat$y_prev[i]
-  dat[["y_prev_minus_i"]] <- dat$y_prev[-i]
-  dat[["m_prev_i"]] <- dat$m_prev[i]
-  dat[["m_prev_minus_i"]] <- dat$m_prev[-i]
-  dat[c("n", "y_prev_i", "y_prev_minus_i", "m_prev_i", "m_prev_minus_i")]
-}
-
-#' Loop over doing the Gaussian approximation for every index i
-template <- list() #' To store obj
-
-for(i in 1:dat$n) {
-  dat_i <- prepare_index(dat, i)
-
-  #' Starting parameters for TMB are the same for every index of the loop
-  param_i <- list(
-    beta_prev = 0,
-    phi_prev_i = 0,
-    phi_prev_minus_i = rep(0, data$n - 1),
-    log_sigma_phi_prev = 0
-  )
-
-  #' random are integrated out with a Laplace approximation
-  obj <- MakeADFun(
-    data = dat_i,
-    parameters = param_i,
-    random = "phi_prev_minus_i", # Random effects to be integrated out
-    DLL = "model1index"
-  )
-
-  its <- 1000
-
-  opt <- nlminb(
-    start = obj$par,
-    objective = obj$fn,
-    gradient = obj$gr,
-    control = list(iter.max = its, trace = 0)
-  )
-
-  sd_out <- sdreport(
-    obj,
-    par.fixed = opt$par,
-    getJointPrecision = TRUE
-  )
-
-  template[[i]] <- obj
-}
-
-# Create objective function (no Laplace approximation)
-param <- list(
-  beta_prev = 0,
-  phi_prev = rep(0, data$n),
-  log_sigma_phi_prev = 0
-)
-
-f <- MakeADFun(
-  data = dat,
-  parameters = param,
-  DLL = "model1"
-)
-
-#' Evaluate objective function at param
-f$fn(unlist(param))
[…]

### 2022-09-17 utils/archive_to_docs.R

@@ -21,3 +21,4 @@ archive_to_docs("epil")
 archive_to_docs("prev-anc-art_process-results")
 archive_to_docs("example_inla-replication")
 archive_to_docs("example_inla-grid")
+archive_to_docs("dev_sinla")

### 2022-09-20 src/dev_sinla/sinla.Rmd
Notes from meeting, add 2pi back in, try to produce LA p(x_i | theta, y) directly with TMB template and indirectly extracting Gaussian approximation (not getting same answer yet)

@@
 
-We are interested in approximating latent field posterior marginals as follows $p(x_i \, | \, \theta, y) \approx \tilde p(x_i \, | \, \theta, y)$ for each index $i$.
-To build up to implementing the simplified INLA approach, we will start with two simpler methods.
+We are interested in efficiently approximating latent field posterior marginals $p(x_i \, | \, \theta, y) \approx \tilde p(x_i \, | \, \theta, y)$ for each index $i$.
+To build up to implementing the simplified INLA approach, we will start with two simpler methods (1) taking the marginal of a joint Gaussian approximation (2) doing a full Laplace approximation.
+
+The AGHQ/ELGM approximation to the joint posterior is
+$$
+\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
+$$
+This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue.
+For the $i$th marginal, we just take the appropriate samples from the joint.
+The approximation is therefore
+$$
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
+$$
+where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.
+
+`aghq::marginal_laplace_tmb` will calculate an object representing $\tilde p(x \, | \, \theta, y)$.
+Then `aghq::sample_marginal` will draw samples from it.
+
+We want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation
+$$
+\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
+$$
 
@@
 
-The `aghq` package currently approximates latent field posterior marginals by taking marginals of the joint Gaussian approximation $p_G(x \, | \, \theta, y)$.
+The `aghq` package currently approximates latent field posterior marginals by taking marginals of the joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$.
 This is same approach used by `method = "gaussian"` in `R-INLA`.
-The Gaussian approximation $p_G(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
-\begin{equation}
-\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_G(x \, | \, \theta, y)} \Big\rvert_{x = \hat{\mu}(\theta)} = \frac{p(y, \hat{\mu}(\theta), \theta)}{\det(\hat{Q}(\theta))^{1/2}},
-\end{equation}
+The Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
+$$
+\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat{\mu}(\theta)} = \frac{p(y, \hat{\mu}(\theta), \theta)}{\det(2 \pi \hat{Q}(\theta))^{1/2}},
+$$
 such that one easy way to calculate it is to use set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there.
@@
 dat <- list(n = data$n, y_prev = data$y_prev, m_prev = data$m_prev)
+dat
 ```
 
-We create an objective function `f` to evaluate $p(x, \theta, y)$ (without any Laplace approximations!):
+We create an objective function `f` (see Appendix for `TMB` template) to evaluate $p(x, \theta, y)$ (without any Laplace approximations!):
 
-```{r}
+```{r message=FALSE, warning=FALSE}
 param <- list(
@@
 
+We also create an objective function `h` (see Appendix for `TMB` template) integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`):
+
+```{r message=FALSE, warning=FALSE}
+h <- MakeADFun(
+  data = dat,
+  parameters = param,
+  random = "phi_prev",
+  DLL = "model1"
+)
+```
+
 # Full Laplace approximation
@@
 Another approach we might take is to calculate the full Laplace approximation
-\begin{align}
-  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_G(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \hat{\mu}_{-i}(x_i, \theta)} \\
-  &= \frac{p(y, x_i, \hat{\mu}_{-i}(x_i, \theta), \theta)}{\det(\hat{Q}_{-i}(x_i, \theta))^{1/2}}
-\end{align}
-This is not practical in some settings as it involves recomputing the Gaussian approximation $p_G(x_{-i} \, | \, x_i, \theta, y)$ for each value of $(x_i, \theta)$ which can be too computationally expensive.
+$$
+\begin{align*}
+  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \hat{\mu}_{-i}(x_i, \theta)} \\
+  &= \frac{p(y, x_i, \hat{\mu}_{-i}(x_i, \theta), \theta)}{\det(2 \pi \hat{Q}_{-i}(x_i, \theta))^{1/2}}
+\end{align*}
+$$
+This is not practical in some settings as it involves recomputing the Gaussian approximation $p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ for each value of $(x_i, \theta)$ which can be too computationally expensive.
 With of a bit of effort, we can implement this in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are named as such (see Appendix for template).
@@
 
-```{r}
+```{r message=FALSE, warning=FALSE, results=FALSE}
 #' To store obj
@@
 
+Let's consider the first marginal, that is $\tilde p_\text{LA}(x_1 \, | \, \theta, y)$.
+The mean and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:
+
+```{r message=FALSE, warning=FALSE}
+i <- 1
+obj <- template[[i]]
+
+mean <- with(obj$env, last.par[random])
+length(mean) #' The length is 30 - 1
+mean
+
+Q <- obj$env$spHess(with(obj$env, last.par), random = TRUE)
+dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
+image(Q)
+```
+
+`obj$fn` should take as input $(x_i, \theta)$:
+
 ```{r}
-#' Evaluate objective function at param
-f$fn(unlist(param))
-
 transform_param <- function(param) {
@@
 
-param %>%
-  transform_param() %>%
-  unlist() %>%
-  template[[1]]$fn()
+transformed_param <- transform_param(param)
+transformed_param$phi_prev_minus_i <- NULL #' Remove the integrated out effects
+transformed_param
 
-#' The mean and precision matrix for the Gaussian approximation to p(x_{-i} | x_i, theta, y)
-#' The mean of is length dim
-mean <- with(template[[1]]$env, last.par[random])
-Q <- template[[1]]$env$spHess(with(template[[1]]$env, last.par), random = TRUE)
+obj$fn(transformed_param)
+```
 
-#' Numerator of the LA
-#' TODO
+We can reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
+Recall that `f` takes as input the complete $(x, \theta)$ ($y$ is already fixed).
+For example:
 
-#' Denominator of the LA
-sqrt(det(Q))
+```{r}
+f$fn(unlist(param))
 ```
+
+We update `param` to include $\hat{\mu}_{-i}(x_i, \theta)$:
+
+```{r}
+param$phi_prev[-i] <- mean
+param
+
+```
+
+Such that the numerator of the Laplace approximation is (n.b. I think this is on the log-scale, which is where we will do the calculations):
+
+```{r}
+numer <- f$fn(unlist(param))
+numer
+```
+
+And the denominator of the Laplace approximation (on the log-scale) is:
+
+```{r}
+denom <- log(sqrt(det(2 * pi * Q)))
+denom
+```
+
+```{r}
+numer - denom
+```
+
+Compare this to the output from `obj`:
+
+```{r}
+obj$fn(transformed_param)
+```
+
+# Simplified INLA
+
+$$
+\hat \mu_{-i}(x_i, \theta) = [\hat \mu(\theta)]_{-i} + \hat Q(\theta)^{-1}_{-i, i} \hat Q(\theta)_{i, i}(x_i - [\hat \mu(\theta)]_{i})
+$$
 

### 2022-09-29 src/docs_xx-xx-22-seminar/xx-xx-22-seminar.Rmd
Placeholder presentation start

--- 2022-02-07 src/docs_01-07-21-stats-epi-group/01-07-21-stats-epi-group.Rmd
@@
 ---
-title: Eventually an introduction to the `aghq` R package
-subtitle: Statistics and Epidemiology Group Meeting
+title: Inference methods for extended latent Gaussian models
+subtitle: Waterloo SAS Student Seminar Series
 author: Adam Howes
 institute: Imperial College London
-date: July 2021
+date: October 2022
 bibliography: citations.bib
@@
 
-# Motivation I
+# Motivation
 
-* People on ART $A_i$ can be used as supplementary data for small-area estimation of HIV prevalence $\rho_i$
+* Surveillance of the HIV epidemic in sub-Saharan Africa
+* Aim to estimate epidemic indicators useful for monitoring and response, including:
+  * Prevalence $\rho$: the proportion of people who are HIV positive
+  * Incidence $\lambda$: the proportion of people newly infected 
+  * Treatment coverage $\alpha$: the proportion of PLHIV on treatment
+* This is a challenging task! Data is noisy, sparse and biased
+
+# A simple small-area model for prevalence
+
+* Consider areas $i = 1, \ldots, n$
+* Simple random sample survey taken in each area, with sample sizes $m_i$
+* The number of people testing positive is $y_i$
+* Then we can use a binomial logistic regression of the form:
 \begin{align*}
-A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
 y_i &\sim \text{Bin}(m_i, \rho_i), \\
-\text{logit}(\alpha_i) &\sim f(\vartheta_\alpha), \\
 \text{logit}(\rho_i) &\sim g(\vartheta_\rho), \quad i = 1, \ldots, n,
 \end{align*}
-* If $f$ and $g$ are Gaussian then model is almost, but not quite, a latent Gaussian model by the definition of @rue2009approximate
-  * This is due to small non-linearities (multiplying two latent Gaussian fields)
-  * Each observation depends on more than one element of the latent field
+* If $g$ is Gaussian then this is a latent Gaussian model in the sense of @rue2009approximate
+* One problem with this data is that household surveys are expensive to run, so they only happen rarely
 
-# Motivation II
+# Latent Gaussian models
+
+* Three-stage model
+\begin{alignat*}{2}
+  &\text{(Observations)}     &   \y &\sim p(\y \, | \, \x), \\
+  &\text{(Latent field)}     &   \x &\sim p(\x \, | \, \btheta), \\
+  &\text{(Hyperparameters)}  &   \qquad \btheta &\sim p(\btheta),
+\end{alignat*}
+where $\y = (y_1, \ldots, y_n)$, $\x = (x_1, \ldots, x_n)$, $\btheta = (\theta_1, \ldots, \theta_m)$
+* Interested in learning both $(\btheta, \x)$ from data $\y$
+* If the middle layer is Gaussian, then it's a latent Gaussian model
+* Covers most of the models commonly used in spatiotemporal statistics
+
+# Adding ANC surveillance
+
+* Pregnant women attending antenatal care clinics are routinely tested for HIV, to avoid mother-to-child transmission
+* This data source is more real-time than household surveys, but it's also more biased, because attendees are unlikely to be as representative of the population
+* But perhaps this bias is consistent, in which case we can still make use of the ANC data to improve our model!
+
+# Naomi evidence synthesis model
 
@@
 
-* Previous slide is a simplified version of the Naomi evidence synthesis model
-* Countries to fit the model using their own data ("in production"?)
-  * Can't run long MCMC on the cluster for weeks, as might be the case if this was one paper
-* Can't use `R-INLA`, require something more flexible
+* Combining these three modules is the basis of the Naomi evidence synthesis model
+* Used by countries, which provide their own data, to produce HIV estimates in a yearly process supported by UNAIDS
+* Can't run long MCMC in this setting, requires fast, accurate, approximations 
+* Requires something more flexible than `R-INLA`
 * Currently using Template Model Builder `TMB` [@kristensen2015tmb]
@@
 
-# Aside: common theme I
-
-* Combining flawed (sparse, aggregated) gold standard (measuring the thing we want) data with other correlated (more available, high resolution) data (measuring not exactly what we want)
-* Consistently resulting in models with multiple outcomes (evidence synthesis, multi-output)
-* I think a lot of these are going to be not quite LGMs
-
-# Aside: common theme II
-
-* Examples include
-  * Naomi model: DHS data is "gold standard", supported by ANC data from pregnant women
-  * Sexual risk behaviour model: estimates of FSW population at national level, supported by DHS data
-    * The national-level FSW estimates might be more like "bronze standard"
-    * DHS approximately asks "have you received money or gifts in exchange for sex in past 12 months"
-  * Loa loa prevalence and eyeworm history prevalence model: measuring eyeworm history is a cheap proxy for Loa loa [@amoah2020geostatistical]
-
-# Recap on latent Gaussian models
-
-* Three-stage model
-\begin{alignat*}{2}
-  &\text{(Observations)}     &   \y &\sim p(\y \, | \, \x), \\
-  &\text{(Latent field)}     &   \x &\sim p(\x \, | \, \btheta), \\
-  &\text{(Hyperparameters)}  &   \qquad \btheta &\sim p(\btheta),
-\end{alignat*}
-where $\y = (y_1, \ldots, y_n)$, $\x = (x_1, \ldots, x_n)$, $\btheta = (\theta_1, \ldots, \theta_m)$
-* Interested in learning both $(\btheta, \x)$ from data $\y$
-* Covers most of the models used in spatiotemporal statistics
-
-# Recap on Integrated Nested Laplace Approximation I
+# Integrated Nested Laplace Approximation
 
@@
 
-# Recap on Integrated Nested Laplace Approximation II
+# Integrated Nested Laplace Approximation
 
@@
 
-# Recap on Integrated Nested Laplace Approximation III
+# Integrated Nested Laplace Approximation
 
@@
 which can be marginalised to get $\tilde p(\theta_j \, | \, \y)$
+  * Note here that this involves integrating out a **Gaussian** approximation to the latent field
 2) In both \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. \eqref{eq:hypermarginal}, so choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$
-  * For low $m$ INLA uses a grid-strategy which I illustrate in the next slide
-  * For larger $m$ this becomes too expensive and a CCD design is used
+  * For low $m$ `R-INLA` uses a grid-strategy (illustrated in the next slide)
+  * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design is used
+  * Other approaches, like adaptive Gaussian Hermite quadrature (AGHQ) have recently been shown to have theoretical guarantees [@bilodeau2021stochastic] and may be preferable
 
@@
 
-```{r message=FALSE, echo=FALSE, fig.cap=paste("An illustration of the INLA grid method for selecting integration points using a toy Gaussian distribution for $\\btheta$.
+```{r message=FALSE, echo=FALSE, fig.cap=paste("An illustration of the `R-INLA` grid method for selecting integration points using a toy Gaussian distribution for $\\btheta$.
 Start at the mode and work outwards along the eigenvectors until the density drops sufficiently low.
@@
 
-# Recap on Integrated Nested Laplace Approximation IV
+# Adaptive Gaussian Hermite Quadrature
+
+* `aghq` R package and vignette [@stringer2021implementing]
+* Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
+* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior)
+* Use $k$ quadrature points
+  * If $k$ is odd then they include the mode
+  * If $k = 1$ then it's a Laplace approximation
+  * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often
+
+# Integrated Nested Laplace Approximation
 
@@
 \end{equation}
-  * The above is referred to as "Gaussian" approximation, and confusingly there are two more complex ones called "simplified Laplace" and "Laplace"
-  * You can pick which one in `R-INLA` using the `method` option
+  * The above is referred to as `method = "gaussian"` in `R-INLA`, and confusingly there are two more complex ones called `"simplified laplace"` and `"laplace"`
+  * We will discuss ways to do better than this joint Gaussian approximation in the next slide
 4) Finally use quadrature to get 
@@
 
-# Template Model Builder I
+# Template Model Builder
 
@@
 
-# Template Model Builder II
+# Template Model Builder
 
@@
 
-# Template Model Builder III
+# Template Model Builder
 
@@
 
-# Adaptive Gaussian Hermite Quadrature
-
-* Recent work by Alex Stringer and coauthors on AGHQ
-  * `aghq` R package and vignette [@stringer2021implementing]
-  * Theory paper [@bilodeau2021stochastic]
-* Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
-* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior)
-* Use $k$ quadrature points
-  * If $k$ is odd then they include the mode
-  * If $k = 1$ then it's a Laplace approximation
-  * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often
-
-# Epil example I
-
-* Epilepsy example from Section 5.2. of @rue2009approximate (previously from BUGS):
-  * Patients $i = 1, \ldots, 59$ each either assigned treatment $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$ to help with seizures
-  * Visits to clinics $j = 1, \ldots, 4$ times with $y_{ij}$ the number of seizures of the $i$th person in the two weeks proceeding their $j$th visit to the clinic
-  * Covariates age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$
-
-# Epil example II
-
-This is what the model looks like (it's a Poisson GLMM):
-\begin{align*}
-  y_{ij}          &\sim \text{Poisson}(\lambda_{ij}), \\
-  \lambda_{ij}    &= e^{\eta_{ij}}, \\
-  \eta_{ij}       &= \beta_0 + \beta_{\texttt{Base}} \log(\texttt{Baseline}_j / 4) + \beta_{\texttt{Trt}} \texttt{Trt}_i +
-                     \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
-                  &+ \beta_{\texttt{Age}} \log(\texttt{Age}_i) + \beta_{\texttt{V}_4} {\texttt{V}_4}_j +
-                     \epsilon_i + \nu_{ij}, \quad i=1:59, \quad j=1:4, \\
-  \beta           &\sim \mathcal{N}(0, 100^2), \quad \forall \beta, \\
-  \epsilon_i      &\sim \mathcal{N}(0, 1/\tau_\epsilon), \\
-  \nu_{ij}        &\sim \mathcal{N}(0, 1/\tau_\nu), \\
-  \tau_\epsilon   &\sim \Gamma(0.001, 0.001), \\
-  \tau_\nu        &\sim \Gamma(0.001, 0.001).
-\end{align*}
-
-# Epil example III
-
-* `aghq` package interfaces really easily with `TMB`!
-* This is the code I used to fit the model with `TMB`
-
-```{r eval=FALSE}
-obj <- MakeADFun(
-  data = dat,
-  parameters = param,
-  # These are the ones integrated out with a Laplace approximation
-  random = c("epsilon", "nu"),
-  DLL = "epil"
-)
-```
-
-# Epil example IV
-
-* Then to fit it with `aghq` it's only a very small modification
-
-```{r eval=FALSE}
-fit <- aghq::marginal_laplace_tmb(
-  obj, 
-  k = 3, 
-  startingvalue = c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
-)
-```
-
-#
-
-```{r echo=FALSE}
-results <- readRDS("comparison-results.rds")
-
-beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
-rownames(results) <- c(sapply(0:5, beta_i))
-
-results %>% 
-  round(digits = 3) %>%
-  kableExtra::kable(format = "latex", booktabs = TRUE, escape = TRUE, align = "c") %>%
-  kable_styling(latex_options = "scale_down")
-```
-
-# Plan
-
-* Test `aghq` for toy Naomi example
-  * Do as above with the Epil example, testing versus a long MCMC run
-  * Sometimes you have to look pretty hard for a node (element of the latent field) where there are differences.
-In the INLA paper they do this by computing a SKLD and ordering by maximum difference.
-Probably good to do here as well
-* Extend `aghq` to replicate INLA functionality by adding the more complex versions of $\tilde p(x_i \, | \, \btheta, \y)$ then test that wih Naomi
-  * Håvard Rue philosophy: "do one thing and do it well"
-  * `R-INLA` implementation of INLA based on sparsity of $\bm{Q}(\btheta)$ that doesn't hold up for extended LGMs
-  * @wood2020simplified on how to still do it
-* Try the INLA without `R-INLA` on other almost LGMs and see how far it can be pushed
-
 # References {.allowframebreaks}

### 2022-09-29 make/_make.R

@@ -5,3 +5,4 @@ run_commit_push("example_inla-grid")
 run_commit_push("dev_sinla")
 run_commit_push("docs_paper")
 run_commit_push("docs_01-07-21-stats-epi-group")
+run_commit_push("docs_xx-xx-22-seminar")

### 2022-09-29 src/example_inla-grid/inla-grid.Rmd
Add .pdf of inla-grid

@@
 
-ggplot() +
+plot <- ggplot() +
   geom_point(data = theta_grid, aes(x = theta1, y = theta2), alpha = 0.5) +
@@
   theme(legend.position = "none")
+
+plot
 ```
+
+```{r}
+pdf("inla-grid.pdf", h = 4.5, w = 6.25)
+
+plot
+
+dev.off()
+```

### 2022-09-29 src/explore_aghq/aghq.Rmd
Add aghq vignette start

---
title: "Understanding the `aghq` package"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**

  **Task**
---

# Basic

```{r}
set.seed(84343124)
y <- rpois(10, 5)
plot(y)

log_post <- function(eta, y) {
  sum(y) * eta - (length(y) + 1) * exp(eta) - sum(lgamma(y + 1)) + eta
}

fn <- function(x) log_post(x, y)
gr <- function(x) numDeriv::grad(fn, x)
he <- function(x) numDeriv::hessian(fn, x)

ff <- list(fn = fn, gr = gr, he = he)

quad <- aghq::aghq(ff = ff, k = 3, startingvalue = 0)

class(quad)

summary(quad)
plot(quad)

# Containing
str(quad)
# normalized_posterior
quad$normalized_posterior$nodesandweights
quad$normalized_posterior$lognormconst
# marginals
# optresults
quad$optresults$mode
```

# Background

```{r}

# Quadrature

# Single dimensional example
f <- function(x) {
  exp(-(x^2) / 2)  
}

# Grid for GHQ with 3 points
gg <- mvQuad::createNIGrid(1, "GHe", 3)
mvQuad::quadrature(f, gg)
sqrt(2 * pi)

mvQuad::getNodes(gg)
mvQuad::getWeights(gg)

# Two dimensional example
f <- function(x) {
  apply(x, 1, function(y) prod(exp(-(y^2) / 2)))  
}

# Still 3 points (per dimension I'd guess)
gg2 <- mvQuad::createNIGrid(2, "GHe", 3)

mvQuad::getNodes(gg2)
# which is equivalent to
expand.grid(mvQuad::getNodes(gg),mvQuad::getNodes(gg))

mvQuad::getWeights(gg2)
# which is equivalent to
cbind(apply(expand.grid(mvQuad::getWeights(gg),mvQuad::getWeights(gg)),1,prod))

mvQuad::quadrature(f, gg2)
2 * pi

# Gauss Hermite Quadrature
# Showing that the nodes are zeros of Hermite polynomial
Hk <- as.function(mpoly::hermite(3, kind = "he"))
nn <- mvQuad::getNodes(gg)
round(Hk(nn))

# Adaptive: scale the nodes and weights by the mode and curvature
theta_hat <- c(2, 3)
theta_hess <- matrix(c(3, 1, 1, 5), nrow = 2, ncol = 2)
theta_hess_inv <- solve(theta_hess)
k <- 3

# Can be rescaled using mvQuad via
int_grid_software <- mvQuad::createNIGrid(2, "GHe", k)
mvQuad::rescale(int_grid_software, m = theta_hat, C = theta_hess_inv, dec.type = 2)
# or manually via
int_grid_manual <- mvQuad::createNIGrid(2, "GHe", k)
nn_rescaled <- mvQuad::getNodes(int_grid_manual) %*% chol(theta_hess_inv)
nn_rescaled[, 1] <- nn_rescaled[, 1] + theta_hat[1]
nn_rescaled[, 2] <- nn_rescaled[, 2] + theta_hat[2]
# and for the weights
mvQuad::getWeights(int_grid_software)
ww_rescaled <- mvQuad::getWeights(int_grid_manual) * det(chol(theta_hess_inv))

# k = 1 then ACHQ is a Laplace approximation
```

# Tomato disease example

```{r}
library(TMB)
data("tswv", package = "EpiILMCT")

precompile()
compile("disease.cpp")
dyn.load(dynlib("disease"))

dat <- tswv$tswvsir
dat$epidat <- dat$epidat[order(dat$epidat[, 4]), ]
I <- dat$epidat[, 4]
R <- dat$epidat[, 2]
infected <- !is.infinite(I)

datlist <- list(
  D = as.matrix(dist(dat$location[dat$epidat[, 1], ])),
  I = I,
  R = R,
  infected = as.numeric(infected[infected])
)

ff <- TMB::MakeADFun(
  data = datlist,
  parameters = list(theta1 = 0, theta2 = 0),
  DLL = "disease",
  ADreport = FALSE,
  silent = TRUE
)

tm <- Sys.time()
# c(0, 0) is start and k = 7, ilm = individual-level model (?)
tmp <- capture.output(
  ilmquadrature <- aghq::aghq(ff, 7, c(0, 0))
)
runtime <- difftime(Sys.time(), tm, units = "secs")

summary(ilmquadrature)

# compute_moment is the mean (only?)
ilm_means <- aghq::compute_moment(ilmquadrature$normalized_posterior, function(x) exp(x))
names(ilm_means) <- c("alpha", "beta")

# q is a vector of numeric values in (0, 1) giving the quantiles you want
my_q <- c(0.025, 0.1, 0.5, 0.9, 0.975)
exp(aghq::compute_quantiles(ilmquadrature$marginals[[1]], q = my_q)) # alpha
exp(aghq::compute_quantiles(ilmquadrature$marginals[[2]], q = my_q)) # beta
```

### 2022-09-29 make/_make.R

@@ -2,6 +2,7 @@ src("make/epil.R")
 src("make/prev-anc-art.R")
 run_commit_push("example_inla-replication")
 run_commit_push("example_inla-grid")
+run_commit_push("explore_aghq")
 run_commit_push("dev_sinla")
 run_commit_push("docs_paper")
 run_commit_push("docs_01-07-21-stats-epi-group")

### 2022-09-29 src/explore_aghq/disease.cpp

@@ -1,0 +1,75 @@
+#include <TMB.hpp>                                // Links in the TMB libraries
+//#include <fenv.h>
+
+template<class Type>
+Type objective_function<Type>::operator() ()
+{
+  // Read in data
+  //
+  // Distance matrix
+  DATA_MATRIX(D);
+  // Infection and removal times
+  // Ordered by Infection time
+  // Distance matrix appropriately sorted
+  DATA_VECTOR(I);
+  DATA_VECTOR(R);
+  DATA_VECTOR(infected); // Indictor of infections
+
+  int N = I.size();
+  int m = infected.size();
+
+  // Parameters
+  // theta1 = log(alpha)
+  // theta2 = log(beta)
+  PARAMETER(theta1);
+  PARAMETER(theta2);
+
+  Type alpha = exp(theta1);
+  Type beta = exp(theta2);
+
+  // Infectivity
+  matrix<Type> lambda(D);
+  for (int i = 0;i<N;i++) {
+    for (int j = 0;j<N;j++) {
+      if (i == j) {
+        lambda(i,j) = 0;
+      } else {
+        lambda(i,j) = alpha * pow(D(i,j),-beta);
+      }
+    }
+  }
+
+  Type t1 = 0;
+  Type t1inner = 0;
+  Type t2 = 0;
+
+  // First term
+  for (int j = 1;j<m;j++) {
+    t1inner = 0;
+    for (int i = 0;i<N;i++) {
+      if (I(i) < I(j) & I(j) <= R(i)) {
+        t1inner += lambda(i,j);
+      }
+    }
+    t1 += log(t1inner);
+  }
+
+  // Second term
+  for (int i = 0;i<m;i++) {
+    for (int j = 0;j<N;j++) {
+      Type Rmin = R(i);
+      if (R(i) > I(j)) Rmin = I(j);
+      Type Imin = I(i);
+      if (I(i) > I(j)) Imin = I(j);
+      t2 += (Rmin - Imin) * lambda(i,j);
+    }
+  }
+
+  Type logpost = t1 - t2;
+  // Add the log-priors
+  Type phi = 0.01;
+  logpost += log(phi) + log(alpha) - phi*alpha;
+  logpost += log(phi) + log(beta) - phi*beta;
+
+  return logpost;
+}

### 2022-09-29 src/explore_aghq/script.R

@@ -1,0 +1,5 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("explore_aghq")
+# setwd("src/explore_aghq")
+
+rmarkdown::render("aghq.Rmd")

### 2022-09-29 src/dev_sinla/sinla.Rmd
Iterate on SINLA development. Still trying to get TMB Laplace approximation to match with (more) manually calculated one

@@
   **Background** Wood (2020) simplify the INLA method by approximating latent field posterior marginals using a reduced cost Laplace approximation which does not rely on sparsity assumptions.
+
   **Task** We implement the method in R making use of Template Model Builder for Gaussian approximations.
@@
 
+# Further details on what we would like to achieve
+
+This section is where I am working things out about what we're aiming for.
+It likely won't follow much narrative structure for now.
+
 The AGHQ/ELGM approximation to the joint posterior is
 $$
-\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
+\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
 $$
+where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are quadrature nodes and weights.
+In particular, let
+$$
+\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
+$$
+where $H_k(z_d)$ are the Hermite polynomials for $k \in \mathbb{N}$, and
+$$
+w_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
+$$
+Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ with nodes $z$, of which there will be $k^d$.
+Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode and $L$ is the lower Cholesky triangle of the curvature such that $LL^\top = H^{-1}$.
+
 This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue.
@@
 
-We want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation
+We want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
 $$
@@
 
-We create an objective function `f` (see Appendix for `TMB` template) to evaluate $p(x, \theta, y)$ (without any Laplace approximations!):
+We create an objective function `f` (see Appendix for `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!):
 
@@
 param <- list(
-  beta_prev = 0,
+  beta_prev = -2,
   phi_prev = rep(0, data$n),
-  log_sigma_phi_prev = 0
+  log_sigma_phi_prev = -1
 )
@@
 )
+```
+
+Optimise `h`:
+
+```{r}
+its <- 1000
+
+opt <- nlminb(
+  start = h$par,
+  objective = h$fn,
+  gradient = h$gr,
+  control = list(iter.max = its, trace = 0)
+)
+```
+
+Get the parameters of the Gaussian approximation:
+
+```{r}
+mean <- with(h$env, last.par[random])
+length(mean) #' The length is 30
+mean
+
+Q <- h$env$spHess(with(h$env, last.par), random = TRUE)
+dim(Q) #' With dimension [30 x 30]
+image(Q)
+
+input_param <- param
+h$fn(input_param[c("beta_prev", "log_sigma_phi_prev")])
+```
+
+Try to reproduce by input to `f`:
+
+```{r}
+input_param$phi_prev <- mean
+numer <- -1 * f$fn(unlist(input_param))
+numer
+
+denom <- 0.5 * log(det(2 * pi * Q))
+denom
+```
+
+On the negative log-likelihood scale, the number we get is:
+
+```{r}
+-1 * (numer - denom)
 ```
@@
 ```{r message=FALSE, warning=FALSE, results=FALSE}
-#' To store obj
-template <- list()
+objs <- list()
+means <- list()
+Qs <- list()
 
@@
 
-  template[[i]] <- obj
+  #' Parameters of the Gaussian approximation
+  mean <- with(obj$env, last.par[random])
+  Q <- obj$env$spHess(with(obj$env, last.par), random = TRUE)
+
+  objs[[i]] <- obj
+  means[[i]] <- mean
+  Qs[[i]] <- Q
 }
@@
 i <- 1
-obj <- template[[i]]
-
-mean <- with(obj$env, last.par[random])
-length(mean) #' The length is 30 - 1
+obj <- objs[[i]]
+mean <- means[[i]] #' The length is 30 - 1
 mean
 
-Q <- obj$env$spHess(with(obj$env, last.par), random = TRUE)
+Q <- Qs[[i]]
 dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
@@
 
-We can reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
+We should be able to reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
 Recall that `f` takes as input the complete $(x, \theta)$ ($y$ is already fixed).
@@
 ```{r}
+param
 f$fn(unlist(param))
@@
 
-We update `param` to include $\hat{\mu}_{-i}(x_i, \theta)$:
+We update `param` to include $\hat{\mu}_{-i}(x_i, \theta)$, so that it's only element `r i` of `phi_prev` which remains as before:
 
@@
 param
-
 ```
 
-Such that the numerator of the Laplace approximation is (n.b. I think this is on the log-scale, which is where we will do the calculations):
+The numerator of the Laplace approximation is (n.b. this is on the log-scale, which is where we will do the calculations):
 
 ```{r}
-numer <- f$fn(unlist(param))
+numer <- -1 * f$fn(unlist(param))
 numer
@@
 
-And the denominator of the Laplace approximation (on the log-scale) is:
+And the denominator of the Laplace approximation (again on the log-scale) is:
 
 ```{r}
-denom <- log(sqrt(det(2 * pi * Q)))
+denom <- 0.5 * log(det(2 * pi * Q))
 denom
@@
 ```{r}
-numer - denom
+-1 * (numer - denom)
 ```

### 2022-09-30 docs/bayescomp.Rmd
BayesComp abstract

---
title: "BayesComp 2023 and Bayesian Inference of Epidemics abstract submission"
author: "Adam Howes, Imperial College London"
date: "29th September 2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

[BayesComp 2023](https://bayescomp2023.com/) and [Bayesian Inference of Epidemics](https://warwick.ac.uk/fac/sci/statistics/staff/academic-research/corbella/bcepidemics2023).

```{r echo=FALSE}
title <- "Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV evidence synthesis model"
abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce HIV epidemic indicators. We combine the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, we compare our inference method to other approaches. We provide an easy to use implementation as a part of the `aghq` R package allowing flexible use of the INLA method for a broader class of models, including any with a `TMB` template."
```

**Title**:
```{r echo=FALSE, results='asis'}
cat(title)
```

Title word count: `r stringr::str_count(title, '\\w+')`.

**Abstract**:
```{r echo=FALSE, results='asis'}
cat(abstract)
```

Abstract word count: `r stringr::str_count(abstract, '\\w+')`/200.

### 2022-09-30 src/explore_aghq/aghq.Rmd
Improve aghq walkthrough

@@
 
-# Basic
+```{r}
+cbpalette <- multi.utils::cbpalette()
+```
+
+# Basic usage
+
+Data:
 
@@
 plot(y)
+```
 
+The log-posterior for the model is:
+
+```{r}
 log_post <- function(eta, y) {
@@
 }
+```
 
+Get the first two derivatives too, and put them into a list:
+
+```{r}
 fn <- function(x) log_post(x, y)
@@
 ff <- list(fn = fn, gr = gr, he = he)
+```
 
+Use the `aghq` package with `k = 3`:
+
+```{r}
 quad <- aghq::aghq(ff = ff, k = 3, startingvalue = 0)
+```
 
+The result is of class `"aghq"`:
+
+```{r}
 class(quad)
+```
 
+```{r}
 summary(quad)
 plot(quad)
+```
 
-# Containing
+The `quad` object contains:
+
+```{r}
 str(quad)
-# normalized_posterior
+```
+
+The nodes and weights of the posterior are:
+
+```{r}
 quad$normalized_posterior$nodesandweights
+```
+
+The logarithm of the normalising constant (should correspond to `logpost - logpost_normalized`):
+
+```{r}
 quad$normalized_posterior$lognormconst
-# marginals
-# optresults
+```
+
+The mode is as follows, which corresponds to the second node.
+For `k = 3`, and other odd numbers, the mode is included as a node.
+
+```{r}
 quad$optresults$mode
@@
 
-# Background
+# More detailed background
+
+## One dimensional example
+
+Suppose we have
+$$
+f(x) = \exp(-\frac{x^2}{2}),
+$$
+which we would like to integrate over $\mathbb{R}$, i.e. $\int_{-\infty}^\infty f(x) \text{d}x$.
 
 ```{r}
-
-# Quadrature
-
-# Single dimensional example
 f <- function(x) {
@@
 }
+```
 
-# Grid for GHQ with 3 points
+```{r}
+plot <- data.frame(x = seq(from = -10, to = 10, by = 0.1)) %>%
+  mutate(
+    y = f(x)
+  ) %>%
+  ggplot(aes(x = x, y = y)) +
+    geom_line() +
+    theme_minimal()
+
+plot
+```
+
+Going to do Gauss-Hermite quadrature with `k = 3` points:
+
+```{r}
 gg <- mvQuad::createNIGrid(1, "GHe", 3)
+```
+
+The positions of the nodes and their weights are:
+
+```{r}
+gg$nodes
+gg$weights
+
+#' The scale for weights is a bit silly
+plot +
+  geom_point(
+    data = data.frame(nodes = gg$nodes, weights = gg$weights, y = 0), 
+    aes(x = nodes, y = y, size = weights),
+    col = cbpalette[2]
+  )
+```
+
+We can perform the quadrature using `mvQuad` via:
+
+```{r}
 mvQuad::quadrature(f, gg)
+```
+
+Or more manually via:
+
+```{r}
+sum(f(gg$nodes) * gg$weights)
+
+#' Note: you can also use these functions to get the nodes and weights
+# mvQuad::getNodes(gg)
+# mvQuad::getWeights(gg)
+```
+
+This is surprisingly close to the real answer, which is:
+
+```{r}
 sqrt(2 * pi)
+```
 
-mvQuad::getNodes(gg)
-mvQuad::getWeights(gg)
+## Two dimensional example
 
-# Two dimensional example
+What about in two dimensions?
+
+```{r}
+f <- function(x) {
+  prod(exp(-(x^2) / 2))
+}
+
+df <- expand.grid(
+  x1 = seq(-3, 3, by = 0.1),
+  x2 = seq(-3, 3, by = 0.1)
+)
+
+plot2 <- df %>%
+  rowwise() %>%
+  mutate(
+    y = f(c(x1, x2))
+  ) %>%
+  ggplot(aes(x = x1, y = x2, fill = y)) +
+    geom_tile() +
+    theme_minimal()
+
+plot2
+```
+
+```{r}
+gg2 <- mvQuad::createNIGrid(2, "GHe", 3)
+gg2
+```
+
+The nodes and weights are:
+
+```{r}
+mvQuad::getNodes(gg2)
+mvQuad::getWeights(gg2)
+```
+
+Equivalent to:
+
+```{r}
+expand.grid(mvQuad::getNodes(gg), mvQuad::getNodes(gg))
+cbind(apply(expand.grid(mvQuad::getWeights(gg), mvQuad::getWeights(gg)), 1, prod))
+```
+
+```{r}
+plot2 +
+  geom_point(
+    data = mvQuad::getNodes(gg2) %>%
+            as.data.frame() %>%
+            mutate(weights = mvQuad::getWeights(gg2)), 
+    aes(x = V1, y = V2, size = weights, fill = 1),
+    col = cbpalette[2]
+  )
+```
+
+Do the quadrature using:
+
+```{r}
+#' Need f to be defined like this for some reason
 f <- function(x) {
@@
 
-# Still 3 points (per dimension I'd guess)
-gg2 <- mvQuad::createNIGrid(2, "GHe", 3)
+mvQuad::quadrature(f, gg2)
+```
 
-mvQuad::getNodes(gg2)
-# which is equivalent to
-expand.grid(mvQuad::getNodes(gg),mvQuad::getNodes(gg))
+As compare with the true answer:
 
-mvQuad::getWeights(gg2)
-# which is equivalent to
-cbind(apply(expand.grid(mvQuad::getWeights(gg),mvQuad::getWeights(gg)),1,prod))
+```{r}
+2 * pi
+```
 
-mvQuad::quadrature(f, gg2)
-2 * pi
-
+```{r}
 # Gauss Hermite Quadrature

### 2022-10-03 src/docs_xx-10-22-seminar/xx-10-22-seminar.Rmd
Iterate SAS seminar forward

--- 2022-09-29 src/docs_xx-xx-22-seminar/xx-xx-22-seminar.Rmd
@@
 * Surveillance of the HIV epidemic in sub-Saharan Africa
-* Aim to estimate epidemic indicators useful for monitoring and response, including:
+* Want to estimate indicators used for monitoring and response, including:
   * Prevalence $\rho$: the proportion of people who are HIV positive
@@
   * Treatment coverage $\alpha$: the proportion of PLHIV on treatment
+* We would like to provide them at a local, district-level
 * This is a challenging task! Data is noisy, sparse and biased
@@
 * Consider areas $i = 1, \ldots, n$
-* Simple random sample survey taken in each area, with sample sizes $m_i$
-* The number of people testing positive is $y_i$
+* Simple random sample household-survey taken in each area, with sample sizes $m^\text{HS}_i$
+* The number of people testing positive is $y^\text{HS}_i$
 * Then we can use a binomial logistic regression of the form:
 \begin{align*}
-y_i &\sim \text{Bin}(m_i, \rho_i), \\
-\text{logit}(\rho_i) &\sim g(\vartheta_\rho), \quad i = 1, \ldots, n,
+y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
+\text{logit}(\rho^\text{HS}_i) &\sim g(\vartheta^\text{HS}), \quad i = 1, \ldots, n,
 \end{align*}
 * If $g$ is Gaussian then this is a latent Gaussian model in the sense of @rue2009approximate
-* One problem with this data is that household surveys are expensive to run, so they only happen rarely
+* We usually set up $g$ as a spatial smoother, because the sample sizes in each area are too small to get reliable direct estimates
+* One problem with the above model is that household surveys are expensive to run, so they only happen rarely
 
@@
 
-* Three-stage model
+* Three-stage Bayesian hierarchical model
 \begin{alignat*}{2}
@@
 * If the middle layer is Gaussian, then it's a latent Gaussian model
+\begin{equation*}
+  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
+\end{equation*}
 * Covers most of the models commonly used in spatiotemporal statistics
+* Latent field is typically indexed by spatiotemporal location, such that $n > m$
 
@@
 * But perhaps this bias is consistent, in which case we can still make use of the ANC data to improve our model!
+
+# Adding ANC surveillance
+
+* Suppose of $m^\text{ANC}_i$ women attending ANC, $y^\text{ANC}_i$ are HIV positive, then we can use another binomial logistic regression:
+\begin{align*}
+y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho^\text{HS}_i) + b_i, \\
+b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
+\end{align*}
+* Similar to using $\rho^\text{ANC}_i$ as a covariate in our model for household survey prevalence, but this way we allow it to vary rather than be fixed
+
+# Adding ART coverage
+
+* Remember that we're also interested in what proportion of PLHIV are receiving treatment
+* To estimate this, we need first to know the number of PLHIV
+\begin{equation*}
+H_i = N_i \rho^\text{HS}_i
+\end{equation*}
+* Then
+\begin{align*}
+A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
+\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2),
+\end{align*}
 
@@
 * Combining these three modules is the basis of the Naomi evidence synthesis model
-* Used by countries, which provide their own data, to produce HIV estimates in a yearly process supported by UNAIDS
+* Used by countries (which provide their own data) to produce HIV estimates in a yearly process supported by UNAIDS
 * Can't run long MCMC in this setting, requires fast, accurate, approximations 
-* Requires something more flexible than `R-INLA`
+* It's a complicated model, and requires something more flexible than `R-INLA`
 * Currently using Template Model Builder `TMB` [@kristensen2015tmb]
@@
 
-# Integrated Nested Laplace Approximation
-
-* @rue2009approximate or e.g. @blangiardo2015spatial
-* Approximate Bayesian inference for \textcolor{hilit}{latent Gaussian models} (LGMs), which are three-stage models with middle layer
-\begin{equation*}
-  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
-\end{equation*}
-* `R-INLA` implementation takes advantage of sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)
-
-# Integrated Nested Laplace Approximation
-
-* Gives approximate \textcolor{hilit}{posterior marginals} $\{\tilde p(x_i \, | \, \y)\}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$
-* To approximate posterior marginals below requires $\tilde p(\btheta \, | \, \y)$ and $\tilde p(x_i \, | \, \btheta, \y)$
-\begin{align}
-  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
-  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
-\end{align}
-
-# Integrated Nested Laplace Approximation
-
-1) First Laplace approximate hyperparameter posterior
-\begin{equation}
-\tilde p(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)}  \label{eq:hypermarginal}
-\end{equation}
-which can be marginalised to get $\tilde p(\theta_j \, | \, \y)$
-  * Note here that this involves integrating out a **Gaussian** approximation to the latent field
-2) In both \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. \eqref{eq:hypermarginal}, so choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$
-  * For low $m$ `R-INLA` uses a grid-strategy (illustrated in the next slide)
-  * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design is used
-  * Other approaches, like adaptive Gaussian Hermite quadrature (AGHQ) have recently been shown to have theoretical guarantees [@bilodeau2021stochastic] and may be preferable
-
-#
-
-```{r message=FALSE, echo=FALSE, fig.cap=paste("An illustration of the `R-INLA` grid method for selecting integration points using a toy Gaussian distribution for $\\btheta$.
-Start at the mode and work outwards along the eigenvectors until the density drops sufficiently low.
-\\label{fig:grid}")}
-knitr::include_graphics("depends/inla-grid.pdf")
-```
-
-# Adaptive Gaussian Hermite Quadrature
-
-* `aghq` R package and vignette [@stringer2021implementing]
-* Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
-* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior)
-* Use $k$ quadrature points
-  * If $k$ is odd then they include the mode
-  * If $k = 1$ then it's a Laplace approximation
-  * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often
-
-# Integrated Nested Laplace Approximation
-
-3) Choose approximation for $\tilde p(x_i \, | \, \btheta, \y)$
-  * Simplest version [@rue2007approximate] is to marginalise the $p_G(\x \, | \, \btheta, \y)$
-\begin{equation}
-\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta))
-\end{equation}
-  * The above is referred to as `method = "gaussian"` in `R-INLA`, and confusingly there are two more complex ones called `"simplified laplace"` and `"laplace"`
-  * We will discuss ways to do better than this joint Gaussian approximation in the next slide
-4) Finally use quadrature to get 
-\begin{equation}
-  \tilde p(x_i \, | \, \y) = 
-  \sum_{k = 1}^K \tilde p(x_i \, | \, \btheta^{(k)}, \y) \times \tilde p(\btheta^{(k)} \, | \, \y) \times \Delta^{(k)}
-\end{equation}
-
 # Template Model Builder
 
-* R package which implements the Laplace approximation for latent variable models using AD (via `CppAD`)
+* So what is `TMB`?
+* R package which implements the Laplace approximation for latent variable models using automatic differentiation (via `CppAD`)
   * For more about AD see e.g. @griewank2008evaluating
@@
 
+# Integrated Nested Laplace Approximation
+
+* Suggested reading: the paper [@rue2009approximate] or a book e.g. @blangiardo2015spatial
+* Approximate Bayesian inference for \textcolor{hilit}{latent Gaussian models} (LGMs), which recall are three-stage models with middle layer
+\begin{equation*}
+  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
+\end{equation*}
+* `R-INLA` implementation takes advantage of sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)
+
+# Integrated Nested Laplace Approximation
+
+* Gives approximate \textcolor{hilit}{posterior marginals} $\{\tilde p(x_i \, | \, \y)\}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$
+* To approximate posterior marginals below requires $\tilde p(\btheta \, | \, \y)$ and $\tilde p(x_i \, | \, \btheta, \y)$
+\begin{align}
+  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
+  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
+\end{align}
+
+# Integrated Nested Laplace Approximation
+
+1) First Laplace approximate hyperparameter posterior
+\begin{equation}
+\tilde p(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)}  \label{eq:hypermarginal}
+\end{equation}
+which can be marginalised to get $\tilde p(\theta_j \, | \, \y)$
+  * Note: this involves integrating out a Gaussian approximation to the latent field, which we denote by $p_G(\x \, | \, \btheta, \y)$
+
+# Integrated Nested Laplace Approximation
+
+2) In both \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. \eqref{eq:hypermarginal}, so choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$
+  * For low $m$ `R-INLA` uses a grid-strategy (illustrated in the next slide)
+  * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design is used
+  * Other approaches, like adaptive Gaussian Hermite quadrature (AGHQ) have recently been shown to have theoretical guarantees [@bilodeau2021stochastic] and may be preferable
+
+#
+
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("An illustration of the \\texttt{R-INLA} grid method for selecting integration points using a toy bivariate Gaussian distribution for $\\btheta$.
+Start at the mode and work outwards along the eigenvectors until the density drops sufficiently low.
+\\label{fig:grid}")}
+knitr::include_graphics("depends/inla-grid.pdf")
+```
+
+# Adaptive Gaussian Hermite Quadrature
+
+* `aghq` R package and vignette [@stringer2021implementing]
+* Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
+* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior)
+* Use $k$ quadrature points
+  * If $k$ is odd then they include the mode
+  * If $k = 1$ then it's a Laplace approximation
+  * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often
+
+# Integrated Nested Laplace Approximation
+
+3) Choose approximation for $\tilde p(x_i \, | \, \btheta, \y)$
+  * Simplest version [@rue2007approximate] is to marginalise the $p_G(\x \, | \, \btheta, \y)$
+\begin{equation}
+\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta))
+\end{equation}
+  * The above is referred to as `method = "gaussian"` in `R-INLA`
+  * There are two better, more complex approximations
+    * Confusingly called `"simplified laplace"` and `"laplace"`
+
+# Integrated Nested Laplace Approximation
+
+4) Finally (!) use quadrature combining
+  * our approximation $p(\btheta \, | \, \y)$ from step 1),
+  * our integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$ from step 2),
+  * and our approximation $\tilde p(x_i \, | \, \btheta, \y)$ from step 3)
+to give
+\begin{equation}
+  \tilde p(x_i \, | \, \y) = 
+  \sum_{k = 1}^K \tilde p(x_i \, | \, \btheta^{(k)}, \y) \times \tilde p(\btheta^{(k)} \, | \, \y) \times \Delta^{(k)}
+\end{equation}
+
+# Experiments
+
+* We would like to find an accurate way to do inference for the Naomi model
+* It also needs to be fast enough to use in production!
+
+# Experiments
+
+* We wrote a simplified version of the Naomi model up in `TMB`
+* This allowed us to test three inference methods **all using precisely the same model and C++ code**
+
+1. A direct Gaussian approximation via `TMB`
+2. Adaptive Gaussian Hermite quadrature via `aghq`
+3. No-U-Turn Sampling (NUTS -- a type of Hamiltonian Monte Carlo) via `tmbstan`
+
+* Note: using different software it is often very difficult to ensure the model is precisely the same, so we're very fortunate here
+
+# Comparison approach
+
+* You could look at the summaries like the mean and standard deviation of each of the posterior marginals
+  * Any approximation method should be pretty good at getting the mean right
+  * Gaussian approximations should be good at getting the second moment right
+* It's probably better to compare the whole posterior distributions
+* One way to do this is via Kolmogorov-Smirnov statistics, which give the maximum difference between two empirical CDFs
+
 # References {.allowframebreaks}

### 2022-10-03 make/_make.R

@@ -6,4 +6,4 @@ run_commit_push("explore_aghq")
 run_commit_push("dev_sinla")
 run_commit_push("docs_paper")
 run_commit_push("docs_01-07-21-stats-epi-group")
-run_commit_push("docs_xx-xx-22-seminar")
+run_commit_push("docs_xx-10-22-seminar")

### 2022-10-03 docs/bayescomp.Rmd
Update BayesComp abstract

@@
 ```{r echo=FALSE}
-title <- "Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV evidence synthesis model"
-abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce HIV epidemic indicators.
+title <- "Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV model"
+abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce district-level HIV epidemic indicators.
 We combine the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, we compare our inference method to other approaches.
-We provide an easy to use implementation as a part of the `aghq` R package allowing flexible use of the INLA method for a broader class of models, including any with a `TMB` template."
+Using data from Malawi, we compare our inference method to both a simple Gaussian approximation via `TMB` and to NUTS via `tmbstan`.
+We provide an easy to use implementation as a part of the `aghq` R package allowing flexible use of the INLA method for a broader class of models than `R-INLA`.
+Given a `TMB` template, using our method requires minimal additional code."
 ```

### 2022-10-04 src/explore_aghq/aghq.Rmd
Work on understanding aghq vignette

@@
 
-Use the `aghq` package with `k = 3`:
+Use the `aghq` package with `k = 3`.
+Note that all `aghq::aghq::` is doing is normalising the posterior.
+We have already provided the exact posterior to it via `fn$ff`.
 
@@
 summary(quad)
-plot(quad)
 ```
 
-The `quad` object contains:
+We can plot the log marginal posterior for `theta1` in its normalised and unnormalised versions.
+The unnormalised version doesn't require anything from `aghq`.
 
 ```{r}
-str(quad)
+data.frame(
+  theta1 = seq(from = 1, to = 2, by = 0.01)
+) %>%
+  mutate(
+    f_unnormalised = ff$fn(theta1),
+    f_normalised = f_unnormalised - quad$normalized_posterior$lognormconst
+  ) %>%
+  pivot_longer(
+    cols = starts_with("f"),
+    names_sep = "_",
+    names_to = c("function", "normalised")
+  ) %>%
+  ggplot(aes(x = theta1, y = value)) +
+    geom_line() +
+    facet_grid(~normalised) +
+    labs(x = "theta1", y = "log(posterior)") +
+    theme_minimal()
 ```
 
-The nodes and weights of the posterior are:
+The contents of the `quad` object can be viewed using `str(quad)`.
+One part of the contents are the nodes and weights used for quadrature, which are:
 
@@
 
-The logarithm of the normalising constant (should correspond to `logpost - logpost_normalized`):
+The logarithm of the normalising constant (should correspond to `logpost - logpost_normalised`, the difference between the two curves above):
 
@@
     geom_line() +
+    labs(x = "x", y = "f(x)") +
     theme_minimal()
@@
 
-#' The scale for weights is a bit silly
 plot +
@@
     col = cbpalette[2]
-  )
+  ) +
+  scale_size_continuous(limits = c(1, 2))
 ```
@@
 
-Or more manually via:
+This isn't doing anything fancy, it's just multiplying the values at the nodes by the weights:
 
@@
 sum(f(gg$nodes) * gg$weights)
-
-#' Note: you can also use these functions to get the nodes and weights
-# mvQuad::getNodes(gg)
-# mvQuad::getWeights(gg)
 ```
 
-This is surprisingly close to the real answer, which is:
+Note that you can also use these functions to get the nodes and weights: `mvQuad::getNodes(gg)`, `mvQuad::getWeights(gg)`.
+The answer we get with `aghq` is exact in this case, as compared with the true answer of $\sqrt {2 \pi}$ because the target function is a log-quadratic:
 
@@
 
-What about in two dimensions?
+What about a function in two dimensions?
+Take this one, where each marginal is the same as above:
 
@@
 
+Let's create a quadrature grid again:
+
 ```{r}
@@
 
-Equivalent to:
+This is equivalent to what can be achieved using the `gg` object from the one-dimensional example:
 
@@
 ```
+
+What do these nodes and weights look like on the plot?
 
@@
     col = cbpalette[2]
-  )
+  ) +
+  labs(col = "f(x1, x2)") +
+  scale_size_continuous(limits = c(2, 4))
 ```
 
-Do the quadrature using:
+Ok! So now let's use `mvQuad` to do the quadrature:
 
 ```{r}
-#' Need f to be defined like this for some reason
+#' Need f to be defined like this for some reason, can probably be fixed
 f <- function(x) {
@@
 
-As compare with the true answer:
+Again, in this case the quadrature is exact
 
@@
 
+See Definition 1 of the [theory paper](https://arxiv.org/pdf/2102.06801.pdf) for more on when the integration will be exact.
+
+## Gauss-Hermite Quadrature
+
+Define the Hermite polynomial with $k = 3$ as:
+
 ```{r}
-# Gauss Hermite Quadrature
-# Showing that the nodes are zeros of Hermite polynomial
 Hk <- as.function(mpoly::hermite(3, kind = "he"))
+
+plot <- data.frame(x = seq(-3, 3, by = 0.1)) %>%
+  mutate(Hkx = Hk(x)) %>%
+  ggplot(aes(x = x, y = Hkx)) +
+    geom_line() + 
+    labs(x = "x", y = "H3(x)") +
+    theme_minimal()
+
+plot
+```
+
+In Gauss-Hermite quadrature, we pick nodes which are zeroes of the Hermite polynomial:
+
+```{r}
 nn <- mvQuad::getNodes(gg)
-round(Hk(nn))
 
-# Adaptive: scale the nodes and weights by the mode and curvature
+plot +
+  geom_point(
+    data = data.frame(x = nn, y = Hk(nn)), aes(x = x, y = y),
+    col = cbpalette[2], size = 3
+  )
+```
+
+Note that these nodes are chosen *only* based on the Hermite polynomial.
+The function to be integrated has not played any role in their choice!
+This is probably not a good idea.
+So in adaptive quadrature, the function plays some role in how the points are chosen.
+Specifically in adaptive GHQ it's done by scaling the nodes and weights by the mode and curvature at the mode of the function to be integrated.
+Imagine we have some fake values for the mode `theta_hat` and curvature at the mode `theta_hess` as follows:
+
+```{r}
 theta_hat <- c(2, 3)
@@
 theta_hess_inv <- solve(theta_hess)
+```
+
+Then the rescaling can be done with `mvQuad` easily via:
+
+```{r}
 k <- 3
-
-# Can be rescaled using mvQuad via
 int_grid_software <- mvQuad::createNIGrid(2, "GHe", k)
 mvQuad::rescale(int_grid_software, m = theta_hat, C = theta_hess_inv, dec.type = 2)
-# or manually via
+```
+
+All that is doing manually to the nodes is:
+
+```{r}
 int_grid_manual <- mvQuad::createNIGrid(2, "GHe", k)
@@
 nn_rescaled[, 2] <- nn_rescaled[, 2] + theta_hat[2]
-# and for the weights
+```
+
+Or to the weights is:
+
+```{r}
 mvQuad::getWeights(int_grid_software)
 ww_rescaled <- mvQuad::getWeights(int_grid_manual) * det(chol(theta_hess_inv))
-
-# k = 1 then ACHQ is a Laplace approximation
 ```
@@
 
+Ok, let's now move on to an example with real data.
+
 ```{r}
-library(TMB)
 data("tswv", package = "EpiILMCT")
-
-precompile()
-compile("disease.cpp")
-dyn.load(dynlib("disease"))
-
 dat <- tswv$tswvsir
@@
 )
+```
 
+```{r}
+precompile()
+compile("disease.cpp")
+dyn.load(dynlib("disease"))
+```
+
+Create the unnormalised negative log posterior using `TMB::MakeADFun` as follows:
+
+```{r}
 ff <- TMB::MakeADFun(
@@
 )
+```
 
+Perform AGHQ with $k = 7$, starting optimisation from $(0, 0)$:
+
+```{r}
 tm <- Sys.time()
-# c(0, 0) is start and k = 7, ilm = individual-level model (?)
+
 tmp <- capture.output(
-  ilmquadrature <- aghq::aghq(ff, 7, c(0, 0))
+  ilm_quadrature <- aghq::aghq(ff, 7, c(0, 0))
 )
+
 runtime <- difftime(Sys.time(), tm, units = "secs")
+```
 
-summary(ilmquadrature)
+The time this took was:
 
-# compute_moment is the mean (only?)
-ilm_means <- aghq::compute_moment(ilmquadrature$normalized_posterior, function(x) exp(x))
+```{r}
+runtime
+```
+
+Note that as there are $k = 7$ quadrature points per dimension, in total there are $7^2 = 49$:
+
+```{r}
+summary(ilm_quadrature)
+```
+The parameters $\alpha = \exp(\theta_1)$ and $\beta = \exp(\theta_2)$, which we can compute summaries for as follows:
+
+```{r}
+ilm_means <- aghq::compute_moment(ilm_quadrature$normalized_posterior, function(x) exp(x))
 names(ilm_means) <- c("alpha", "beta")
+ilm_means
+```
 
-# q is a vector of numeric values in (0, 1) giving the quantiles you want
+You can calculate other moments using the `ff`, `gg` or `nn` arguments to `aghq::compute_moment`.
+There is also some nuance to the argument `method`, either `reuse` or `correct`.
+`reuse` uses the previously computed AGHQ nodes and weights.
+`correct` recomputes the mode and curvature for the product of $g$ and the posterior; it has convergence proofs, but is a bit slower and can be unstable.
+
+We can also compute quantiles using the function `aghq::compute_quantiles`, where `q` is a vector of numeric values in $(0, 1)$.
+
+```{r}
 my_q <- c(0.025, 0.1, 0.5, 0.9, 0.975)
-exp(aghq::compute_quantiles(ilmquadrature$marginals[[1]], q = my_q)) # alpha
-exp(aghq::compute_quantiles(ilmquadrature$marginals[[2]], q = my_q)) # beta
 ```
+
+For $\alpha$:
+
+```{r}
+exp(aghq::compute_quantiles(ilm_quadrature$marginals[[1]], q = my_q))
+```
+
+For $\beta$:
+
+```{r}
+exp(aghq::compute_quantiles(ilm_quadrature$marginals[[2]], q = my_q)) # beta
+```

### 2022-10-05 src/explore_aghq/walkthrough.Rmd
marginal_laplace_tmb walkthrough

---
title: "Code walkthrough"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**

  **Task**
---

In this section, we will step through some of the functions in `aghq` to make sure we understand them.
We will use the epilepsy example, explained in more detail in the [`epil`](https://athowes.github.io/elgm-inf/epil.html) notebook.

```{r}
data(Epil, package = "INLA")

center <- function(x) (x - mean(x))

Epil <- Epil %>%
  mutate(CTrt    = center(Trt),
         ClBase4 = center(log(Base/4)),
         CV4     = center(V4),
         ClAge   = center(log(Age)),
         CBT     = center(Trt * log(Base/4)))

N <- 59
J <- 4
K <- 6
X <- model.matrix(formula(~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT), data = Epil)
y <- Epil$y

make_epsilon_matrix <- function(N, J) {
  t(outer(1:N, 1:(N * J), function(r, c) as.numeric((J*(r - 1) < c) & (c <= J*r))))
}

E <- make_epsilon_matrix(N = 3, J = 2)

dat <- list(N = N, J = J, K = K, X = X, y = y, E = make_epsilon_matrix(N, J))

compile("epil.cpp")
dyn.load(dynlib("epil"))

param <- list(
  beta = rep(0, K),
  epsilon = rep(0, N),
  nu = rep(0, N * J),
  l_tau_epsilon = 0,
  l_tau_nu = 0
)

obj <- MakeADFun(
  data = dat,
  parameters = param,
  random = c("epsilon", "nu"),
  DLL = "epil"
)
```
# `aghq::marginal_laplace_tmb`

Start by calling this function, to show where we're trying to end up:

```{r}
fit <- aghq::marginal_laplace_tmb(
  obj,
  k = 2,
  startingvalue = c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
)

summary(fit)
```

And now let's go through line by line.
These are all of the inputs to the function:

```{r}
ff <- obj
k <- 2
startingvalue <- c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
transformation <- aghq::default_transformation()
optresults <- NULL
basegrid <- NULL
control <- aghq::default_control_tmb()
```

```{r}
#' My guess is that this function just makes sure that the control input is valid
validate_control(control, type = "tmb")

#' And similarly, this probably makes sure the transformation input is valid
validate_transformation(transformation)

transformation <- make_transformation(transformation)
transformation$totheta
transformation$fromtheta
transformation$jacobian

#' Get names for the theta parameters from the TMB function template
#' Note that make.unique will make the names unique by apending sequence numbers to duplicates
thetanames <- NULL
if (exists('par', ff)) thetanames <- make.unique(names(ff$par), sep = "")
thetanames

#' Hessian
if (control$numhessian) {
  ff$he <- function(theta) numDeriv::jacobian(ff$gr, theta, method = "Richardson")
}

#' Now we're going to do the AGHQ
quad <- aghq(
  ff = ff,
  k = k,
  transformation = transformation,
  startingvalue = startingvalue,
  optresults = optresults,
  basegrid = basegrid,
  control = control
)

#' There is an option whereby quad will just return the normalising constant, and if so then tmb_marginal_laplace can also be used
#' just to return the normalising constant too. Note on this, why use tmb_marginal_laplace in this situation rather than aghq? Since
#' as far as I understand we're not doing the marginal laplace approximation if control$onlynormconst is TRUE
if (control$onlynormconst) return(quad)

## Add on the info needed for marginal Laplace ##
# Add on the quad point modes and curvatures

distinctthetas <- quad$normalized_posterior$nodesandweights[, grep('theta', colnames(quad$normalized_posterior$nodesandweights))]
distinctthetas
```

Note that there are `r nrow(distinctthetas)` "distinct thetas" corresponding to $k$, here `r k`, raised to the power `r ncol(distinctthetas)`:

```{r}
dim(distinctthetas)
```

```{r}
#' I think this will just be fixing a case where there is only one theta, and somehow a data.frame isn't created
if (!is.data.frame(distinctthetas)) distinctthetas <- data.frame(theta1 = distinctthetas)

modesandhessians <- distinctthetas

#' Replacing the theta, theta1, theta2, ... names with the names from TMB
if (is.null(thetanames)) {
  thetanames <- colnames(distinctthetas)
} else {
  colnames(modesandhessians) <- thetanames
  colnames(quad$normalized_posterior$nodesandweights)[grep('theta',colnames(quad$normalized_posterior$nodesandweights))] <- thetanames
}

modesandhessians$mode <- vector(mode = "list", length = nrow(distinctthetas))
modesandhessians$H <- vector(mode = "list", length = nrow(distinctthetas))

#' There is a loop coming up, which we will step through for i = 1, then let a loop do for i = 2, ...
i <- 1

#' Get the theta (currently modesandhessians, despite being called modesandhessians, only contains the theta, not any mode or hessian)
theta <- as.numeric(modesandhessians[i, thetanames])
theta

#' Set up the mode and hessian of the random effects. This happens when you run the TMB objective with a particular theta
ff$fn(theta)

#' Now pull the mode and hessian. Have to be careful about scoping

#' So this is the mode, for both random and fixed effects
mm <- ff$env$last.par

#' These are the indices of the random effects
ff$env$random

#' We can confirm that for the fixed effects, the "mode" is just the input values theta
#' (To get the indicies for the fixed effects, we just take the complement of the random effects)
mm[-ff$env$random]
theta

#' Set the mode using the relevant indicies of mm
#' "mode" is a list column
modesandhessians[i, "mode"] <- list(list(mm[ff$env$random]))
modesandhessians[i, ]

#' This should be the Hessian evaluated at the mode
#' The argument random = TRUE set it so that...
H <- ff$env$spHess(mm, random = TRUE)

#' Alex has the following note on duplicating H here:
#' "Somehow, TMB puts all evaluations of spHess in the same memory location"
#' Don't totally understand this
H <- rlang::duplicate(H)

modesandhessians[i, "H"] <- list(list(H))

for (i in 2:nrow(distinctthetas)) {
  theta <- as.numeric(modesandhessians[i,thetanames])
  ff$fn(theta)
  mm <- ff$env$last.par
  modesandhessians[i,'mode'] <- list(list(mm[ff$env$random]))
  H <- ff$env$spHess(mm,random = TRUE)
  H <- rlang::duplicate(H)
  modesandhessians[i,'H'] <- list(list(H))
}

quad$modesandhessians <- modesandhessians

class(quad) <- c("marginallaplace", "aghq")
```

### 2022-10-05 src/explore_aghq/script.R

@@ -3,3 +3,4 @@
 # setwd("src/explore_aghq")
 
 rmarkdown::render("aghq.Rmd")
+rmarkdown::render("walkthrough.Rmd")

### 2022-10-06 src/explore_aghq/walkthrough.Rmd
silent = TRUE to make this more readable

@@
 abstract: |
-  **Background**
+  **Background** The `aghq` package implements adaptive Gaussian Hermite quadrature. It is particularly compatible with the `TMB` package for Laplace approximations.
 
-  **Task**
+  **Task** In preparation for implementing new methods into `aghq`, in this notebook we walkthrough various `aghq` functions.
+This will improve understanding of (1) the method, and (2) how to manipulate `TMB` objects in R.
 ---
 
-In this section, we will step through some of the functions in `aghq` to make sure we understand them.
-We will use the epilepsy example, explained in more detail in the [`epil`](https://athowes.github.io/elgm-inf/epil.html) notebook.
+We use the epilepsy example, explained in more detail in the [`epil`](https://athowes.github.io/elgm-inf/epil.html) notebook, as example.
+Start by loading and preparing the data, as well as compling the `TMB` template:
 
@@
   random = c("epsilon", "nu"),
-  DLL = "epil"
+  DLL = "epil",
+  silent = TRUE
 )
 ```
+
 # `aghq::marginal_laplace_tmb`
 
-Start by calling this function, to show where we're trying to end up:
+We start by calling this function, to show where we're trying to end up:
 
-```{r}
+```{r results='hide'}
 fit <- aghq::marginal_laplace_tmb(
@@
 )
+```
 
+```{r}
 summary(fit)
@@
 }
+```
 
+```{r results='hide'}
 #' Now we're going to do the AGHQ
@@
 )
+```
 
+```{r}
 #' There is an option whereby quad will just return the normalising constant, and if so then tmb_marginal_laplace can also be used
@@
 if (control$onlynormconst) return(quad)
-
-## Add on the info needed for marginal Laplace ##
-# Add on the quad point modes and curvatures
 

### 2022-10-06 src/explore_aghq/walkthrough.Rmd
Start working on aghq::sample_marginal walkthrough

@@
 class(quad) <- c("marginallaplace", "aghq")
+
 ```
+
+# `aghq::sample_marginal`
+
+What we'd like to do is to sample from
+$$
+\begin{align}
+\tilde p(x_i \, | \, y) &= \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), \\
+&= \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \lambda(z)
+\end{align}
+$$
+where $\lambda(z) \in [0, 1]$ are weights such that $\sum_z \lambda(z) = 1$.
+We can do this by
+
+1. Sample $z_j \sim \text{Multinomial}(1, \lambda(z_1), \ldots, \lambda(z_J))$
+2. Given $z_j$ sample from $\tilde p_\text{G}(x \, | \, \theta(z_j), y)$
+3. Take the $i$th entry of $\tilde p_\text{G}(x \, | \, \theta(z_j), y)$ as the sample from $\tilde p_\text{G}(x_i \, | \, \theta(z_j), y)$
+
+Now that we have `quad`, it should be compatible with `sample_marginal`:
+
+```{r}
+#' The number of samples to get
+M <- 10
+
+transformation <- quad$transformation
+interpolation <- "auto"
+
+#' Is this getting mc.cores, and if there isn't a value just setting it to 1 (integer)
+numcores <- getOption("mc.cores", 1L)
+
+#' Length of the random effects
+d <- dim(quad$modesandhessians$H[[1]])[1]
+
+simlist <- quad$modesandhessians
+
+#' Avoid use of parallel computing on Windows
+if (.Platform$OS.type == 'windows') numcores <- 1
+```
+
+The next step is to compute the Cholesky decomposition `L` of the Hessian `H` such that
+$$
+H = LL^\top
+$$
+The reason we want the Cholesky is that we will be using it to same from the joint Gasusian approximation.
+The function `chol` from `base` will compute it for us.
+For example, for the first element:
+
+```{r}
+h <- simlist$H[[1]]
+h <- Matrix::forceSymmetric(h)
+L <- chol(h, perm = FALSE)
+L
+```
+
+Doing this for the list using `lapply`:
+
+```{r}
+simlist$L <- lapply(simlist$H,function(h) chol(Matrix::forceSymmetric(h), perm = FALSE))
+```
+
+There is another version, using the arguments `perm` and `LDL` (which I don't understand) that may be quicker but ultimately will achieve the same thing:
+
+```{r}
+# simlist$L <- lapply(simlist$H, function(h) Matrix::Cholesky(as(Matrix::forceSymmetric(h), "sparseMatrix"), perm = TRUE, LDL = FALSE))
+```
+
+Now let's compute the $\lambda(z)$ weights.
+
+```{r}
+simlist$lambda <- exp(quad$normalized_posterior$nodesandweights$logpost_normalized) * quad$normalized_posterior$nodesandweights$weights
+simlist$lambda
+```
+
+We can check that they do indeed sum to one:
+
+```{r}
+sum(simlist$lambda)
+```
+
+Now let's sample from the multinomial:
+
+```{r eval=FALSE}
+j <- apply(stats::rmultinom(M, 1, simlist$lambda), 2, function(x) which(x == 1))
+j
+
+#' Values are number of samples to draw for each j
+#' In this case they're likely to be all 1, since we're not doing many samples
+tt <- table(j)
+```
+
+Now we construct a big Gaussian mixture matrix.
+The first step is to create a matrix with `d` columns, one for each random effect, and `M` rows, one for each sample, where each value is a standard normal $\mathcal{N}(0, 1)$:
+
+```{r}
+A <- matrix(stats::rnorm(M * d), nrow = M)
+
+#' Now make a list where each element is a row of A, where the index is determined by j
+A_split <- split(A, j)
+
+#' Big Gaussian mixture matrix
+
+#' For each element of the list, create a matrix with d rows (and 1 column? always?)
+Z <- lapply(A_split, matrix , nrow = d)
+```

### 2022-10-06 docs/bayescomp.Rmd
Update BayesComp abstract based on feedback from Alex

@@
 title <- "Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV model"
-abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce district-level HIV epidemic indicators.
-We combine the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, we compare our inference method to both a simple Gaussian approximation via `TMB` and to NUTS via `tmbstan`.
+abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa.
+Performing computationally intensive inference using simulation-based methods like MCMC is infeasible in this setting as the model must be able to be easily run in production by countries, using their own data.
+We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
+Using data from Malawi, we show that our inference method obtains substantially more accurate inferences than the simple Gaussian approximation obtained via `TMB`, as compared with gold-standard NUTS results obtained via `tmbstan`.
 We provide an easy to use implementation as a part of the `aghq` R package allowing flexible use of the INLA method for a broader class of models than `R-INLA`.

### 2022-10-07 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Date for student seminar talk

--- 2022-10-03 src/docs_xx-10-22-seminar/xx-10-22-seminar.Rmd
@@
 institute: Imperial College London
-date: October 2022
+date: November 2022
 bibliography: citations.bib

### 2022-10-07 src/explore_aghq/walkthrough.Rmd
Walkthrough of sample_marginal.marginallaplace finished

@@
 simlist$lambda <- exp(quad$normalized_posterior$nodesandweights$logpost_normalized) * quad$normalized_posterior$nodesandweights$weights
-simlist$lambda
+head(simlist$lambda)
 ```
@@
 
-```{r eval=FALSE}
+```{r}
 j <- apply(stats::rmultinom(M, 1, simlist$lambda), 2, function(x) which(x == 1))
+length(j) #' Should be M
 j
@@
 Z <- lapply(A_split, matrix , nrow = d)
+str(Z)
 ```
+
+In the code there is now a big call to `mapply`, which we are going to break down step-by-step.
+First set the inital values that `function(.x, .y)` will iterate over:
+
+```{r}
+.x <- Z[[1]]
+str(.x)
+
+.y <- names(Z)[[1]]
+str(.y)
+```
+
+Now solve the equation `L %*% .z = .x` for `.z`
+
+```{r}
+.z <- solve(simlist$L[[as.numeric(.y)]], .x)
+str(.z)
+```
+
+Now add the mode (where the additional manipulation is so that if there are multiple samples from the same `j` then there will be the corresonding number of columns):
+
+```{r}
+.z <- .z + do.call(cbind, rep(list(simlist$mode[[as.numeric(.y)]]), ncol(.x)))
+str(.z)
+```
+
+Do this using a `mapply` call:
+
+```{r}
+samps <- mapply(
+  function(.x, .y) as.numeric(solve(simlist$L[[as.numeric(.y)]], .x)) + do.call(cbind, rep(list(simlist$mode[[as.numeric(.y)]]), ncol(.x))),
+  Z,
+  names(Z),
+  SIMPLIFY = FALSE
+)
+
+#' Order them properly
+#' My guess is that you'd want them to be ordered according to how j was drawn, rather
+#' than being grouped up by the value of lambda
+ord <- numeric(length(j))
+cumtab <- cumsum(c(0, tt))
+cumtab <- cumtab[-length(cumtab)]
+cnt <- numeric(length(unique(j)))
+names(cnt) <- sort(unique(j))
+
+for (i in 1:length(j)) {
+  wc <- which(names(cnt) == j[i])
+  cnt[wc] <- cnt[wc] + 1
+  ord[i] <- cumtab[wc] + cnt[wc]
+}
+
+samps <- Reduce(cbind, samps)
+samps <- samps[, ord]
+samps
+```
+
+```{r}
+#' This is now the length of the fixed effects
+d <- length(quad$optresults$mode)
+
+#' Just so that we can get the names of the theta (fixed effects)
+thetanames <- colnames(quad$normalized_posterior$nodesandweights)[1:d]
+theta <- simlist[j, thetanames]
+
+#' In one dimension, R's indexing is not type consistent
+if (!is.matrix(samps)) {
+  samps <- rbind(samps)
+  rownames(samps) <- NULL
+}
+
+if (!inherits(theta, "data.frame")) theta <- data.frame(theta1 = theta)
+
+out <- list(
+  samps = samps,
+  theta = theta
+)
+
+out$theta
+
+# Add the marginals for theta, with possible transformation
+class(quad) <- "aghq" #' This is preventing sample_marginal.marginallaplace being used
+out$thetasamples <- sample_marginal(quad, M, transformation, interpolation)
+str(out)
+```

### 2022-10-07 utils/archive_to_docs.R

@@ -17,7 +17,7 @@ archive_to_docs <- function(report) {
 
 archive_to_docs("docs_paper")
 archive_to_docs("docs_01-07-21-stats-epi-group")
-archive_to_docs("docs_xx-10-22-seminar")
+archive_to_docs("docs_15-11-22-seminar")
 archive_to_docs("epil")
 archive_to_docs("prev-anc-art_process-results")
 archive_to_docs("example_inla-replication")

### 2022-10-08 src/explore_aghq/walkthrough.Rmd
aghq::sample_margial step-by-step

@@
   colnames(modesandhessians) <- thetanames
-  colnames(quad$normalized_posterior$nodesandweights)[grep('theta',colnames(quad$normalized_posterior$nodesandweights))] <- thetanames
+  colnames(quad$normalized_posterior$nodesandweights)[grep('theta', colnames(quad$normalized_posterior$nodesandweights))] <- thetanames
 }
@@
 
-# `aghq::sample_marginal`
+# `aghq::sample_marginal.marginallaplace`
 
@@
 ```
+
+# `aghq::sample_marginal.aghq`
+
+Let's have a look at what that call to `sample_marginal.aghq` is doing:
+
+```{r}
+class(quad)
+
+out <- list()
+
+#' If the quadrature object somehow doesn't contain any information about the 
+if (is.null(quad$marginals)) return(out)
+
+str(quad$marginals)
+```
+
+As usual, we're going to start by doing the computation for the first element of `theta` explicitly:
+
+```{r}
+i <- 1
+```
+
+We already have information about each marginal, contained in:
+
+```{r}
+quad$marginals[[i]]
+```
+
+So we can use the function `aghq::compute_quantiles` to, my guess, use the inverse CDF method to sample from the marginal, given input of some $\mathcal{U}(0, 1)$ draws:
+
+```{r}
+unif <- stats::runif(M)
+unif
+
+compute_quantiles(quad$marginals[[i]], unif, interpolation = interpolation, transformation = transformation)
+```
+
+```{r}
+for (i in 1:length(quad$marginals)) {
+  out[[i]] <- unname(compute_quantiles(quad$marginals[[i]], stats::runif(M), interpolation = interpolation, transformation = transformation))
+}
+
+#' The function doesn't name these yet, but in future it could (and also return a data.frame) via
+d <- length(quad$optresults$mode)
+thetanames <- colnames(quad$normalized_posterior$nodesandweights)[1:d]
+names(out) <- thetanames
+as.data.frame(out)
+```
+
+# `aghq::compute_quantiles.aghq`
+
+Does the `compute_quantiles` function work as we expect it to?
+
+```{r}
+class(quad$marginals[[i]])
+```
+
+The PDF and CDF are computed by `compute_pdf_and_cdf` (as the name suggests, and another layer deep that we can look).
+Code comments note that transformations are not applied at this point, and that it's better to transform the quantiles later:
+
+```{r}
+q <- stats::runif(M)
+
+pdfandcdf <- compute_pdf_and_cdf(quad$marginals[[i]], interpolation = interpolation)
+
+ggplot(pdf_and_cdf, aes(x = theta, y = cdf)) +
+  geom_line() +
+  theme_minimal()
+
+out <- numeric(length(q))
+increasing <- TRUE
+
+validate_transformation(transformation)
+transformation <- make_transformation(transformation)
+```
+
+For each sample, take the maximum `theta` which still has CDF value less than the quantile desired:
+
+```{r}
+for (i in 1:length(q)) out[i] <- pdfandcdf$theta[max(which(pdfandcdf$cdf < q[i]))]
+
+#' Have to check if it's increasing or decreasing so can reverse order if necessary
+increasing <- transformation$fromtheta(min(out)) <= transformation$fromtheta(max(out))
+for (i in 1:length(out)) out[i] <- transformation$fromtheta(out[i])
+
+if (!increasing) out <- rev(out)
+names(out) <- paste0(as.character(100 * q), "%")
+out
+```
+
+# `compute_pdf_and_cdf`
+
+Going to leave this as a to-do.

### 2022-10-10 src/dev_sinla/sinla.Rmd
Improve exposition, add MCMC as ground truth for comparison, start working towards marginal posterior with empirical Bayes

@@
 ---
-title: "Implementing simplified INLA"
+title: "Implementing simplified INLA into `aghq`"
 author:
@@
 output:
-  html_document:
+  bookdown::html_document2:
     toc: yes
@@
 abstract: |
-  **Background** Wood (2020) simplify the INLA method by approximating latent field posterior marginals using a reduced cost Laplace approximation which does not rely on sparsity assumptions.
+  **Background** Wood (2020) simplify the INLA method by approximating the latent field posterior marginals using a reduced cost Laplace approximation which does not rely on sparsity assumptions.
+In spatio-temporal statistics, usually the latent field corresponds to spatio-temporal locations -- so these marginal posteriors are of central scientific interest.
 
-  **Task** We implement the method in R making use of Template Model Builder for Gaussian approximations.
+  **Task** We implement the method in R making use of Template Model Builder (`TMB`) for Gaussian approximations, in a way compatible with the `aghq` package for adaptive Gauss-Hermite quadrature.
+We are interested in efficiently approximating latent field posterior marginals $p(x_i \, | \, \theta, y) \approx \tilde p(x_i \, | \, \theta, y)$ for each index $i$.
+To build up to implementing the simplified INLA approach, we will start with two simpler methods (1) taking the marginal of a joint Gaussian approximation (2) doing a full Laplace approximation.
 ---
 
-We are interested in efficiently approximating latent field posterior marginals $p(x_i \, | \, \theta, y) \approx \tilde p(x_i \, | \, \theta, y)$ for each index $i$.
-To build up to implementing the simplified INLA approach, we will start with two simpler methods (1) taking the marginal of a joint Gaussian approximation (2) doing a full Laplace approximation.
+```{r}
+cbpalette <- multi.utils::cbpalette()
+```
 
-# Further details on what we would like to achieve
+# Background
 
-This section is where I am working things out about what we're aiming for.
-It likely won't follow much narrative structure for now.
-
-The AGHQ/ELGM approximation to the joint posterior is
+The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
 $$
@@
 $$
-where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are quadrature nodes and weights.
+where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by exploration of $\tilde p_\text{LA}(\theta \, | \, y)$.
 In particular, let
@@
 Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ with nodes $z$, of which there will be $k^d$.
-Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode and $L$ is the lower Cholesky triangle of the curvature such that $LL^\top = H^{-1}$.
+Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.
 
-This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue.
+This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triange of the curvature at a randomly chosen quadrature point).
 For the $i$th marginal, we just take the appropriate samples from the joint.
@@
 
-`aghq::marginal_laplace_tmb` will calculate an object representing $\tilde p(x \, | \, \theta, y)$.
-Then `aghq::sample_marginal` will draw samples from it.
+To understand how this is implemented in `aghq`, the two most important functions are `aghq::marginal_laplace_tmb` and `aghq::sample_marginal`.
+`aghq::marginal_laplace_tmb` calculates an object representing $\tilde p(x \, | \, \theta, y)$.
+`aghq::sample_marginal` draws samples from it.
+For a line-by-line walkthrough of how these functions work, see the notebook [Code walkthrough](https://athowes.github.io/elgm-inf/walkthrough.html) (which uses the model from the notebook [Epilepsy example](https://athowes.github.io/elgm-inf/epil.html)).
 
-We want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
+Now, we want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
 $$
@@
 $$
+The goal of doing this is that the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
+At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
+This would be an accurate approach, but not computationally feasible for large models.
+What we hope to achieve is an approximation which gets to most of the accuracy benefits of this new Laplace approximation approach, but without the high computational cost.
 
-# Marginal of joint Gaussian approximation
+# Ground truth
 
-The `aghq` package currently approximates latent field posterior marginals by taking marginals of the joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$.
-This is same approach used by `method = "gaussian"` in `R-INLA`.
-The Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
-$$
-\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat{\mu}(\theta)} = \frac{p(y, \hat{\mu}(\theta), \theta)}{\det(2 \pi \hat{Q}(\theta))^{1/2}},
-$$
-such that one easy way to calculate it is to use set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there.
+To provide a ground truth to compare the methods that follow to, we generate "gold-standard" inferences using a relatively long run of MCMC as implemented by `tmbstan`.
 
 We begin by loading in testing data, as generated by [`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim).
-`sim_data` contains many replicates of simulated data, so we just take the first one:
+`sim_data` contains many replicates of simulated data, so here we just take the first one, as given by:
 
@@
 
-We create an objective function `f` (see Appendix for `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!):
-
-```{r message=FALSE, warning=FALSE}
+```{r}
 param <- list(
@@
 dyn.load(dynlib("model1"))
-
-f <- MakeADFun(
-  data = dat,
-  parameters = param,
-  DLL = "model1"
-)
 ```
 
-We also create an objective function `h` (see Appendix for `TMB` template) integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`):
+We create an objective function `h` (see Appendix for `TMB` template) integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`):
 
-```{r message=FALSE, warning=FALSE}
+```{r}
 h <- MakeADFun(
@@
   random = "phi_prev",
-  DLL = "model1"
+  DLL = "model1",
+  silent = TRUE
 )
@@
 
-Optimise `h`:
+Fit the model using MCMC:
+
+```{r}
+mcmc <- tmbstan::tmbstan(h, chains = 4, iter = 5000)
+```
+
+What does the posterior marginal $p(\phi_1 \, | \, y)$ look like?
+
+```{r}
+extract(mcmc, pars = "phi_prev[1]") %>%
+  as.data.frame() %>%
+  ggplot(aes(x = phi_prev.1.)) +
+    geom_histogram(aes(y = ..count.. / sum(..count..)), alpha = 0.8, fill = cbpalette[7]) +
+    theme_minimal() +
+    labs(x = "phi_prev[1]", y = "Posterior")
+```
+
+# Marginal of joint Gaussian approximation
+
+## $p(\theta \, | \, y)$
+
+The `aghq` package currently approximates latent field posterior marginals by taking marginals of the joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ (this is done implicitly, via sampling).
+This is same approach used by `method = "gaussian"` in `R-INLA`.
+
+The Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
+$$
+\begin{equation}
+\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat{\mu}(\theta)} = \frac{p(y, \hat{\mu}(\theta), \theta)}{\det(2 \pi \hat{Q}(\theta))^{1/2}}, (\#eq:pthetagiveny)
+\end{equation}
+$$
+such that one easy way to calculate it is to use set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there, as we have done in the function `h` above.
+Let's optimise `h`:
 
@@
 
-Get the parameters of the Gaussian approximation:
+The values `opt$par` below are the values of $\theta$ which maximise $\tilde p(\theta \, | \, y)$:
 
 ```{r}
-mean <- with(h$env, last.par[random])
+opt$par
+```
+
+We can also look at the parameters (that is the mean vector and precision matrix) of the Gaussian approximation at the mode above (`opt$par`).
+Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
+By just sticking to the mode of $\theta$, we are effectively doing `aghq` with $k = 1$ in each dimension, i.e. one grid point, corresponding to empirical Bayes type of inference.
+
+```{r}
+#' Notice that last.par contains opt$par as the fixed effects
+#' It's the last.par because the last thing we've done with the template is the optimsation
+#' This seems a pretty shakey way to go about things, but here we are...
+mm <- h$env$last.par
+
+#' These are the indices of the random effects that we want
+h$env$random
+
+#' Note that I'm calling this mean, but because it's a Gaussian the mean is the same as the mode
+mean <- mm[h$env$random]
 length(mean) #' The length is 30
@@
 
-Q <- h$env$spHess(with(h$env, last.par), random = TRUE)
+#' Hessian evaluated at the mean (or equivalently mode)
+Q <- h$env$spHess(mm, random = TRUE)
 dim(Q) #' With dimension [30 x 30]
 image(Q)
-
-input_param <- param
-h$fn(input_param[c("beta_prev", "log_sigma_phi_prev")])
 ```
 
-Try to reproduce by input to `f`:
+## Reproducing output from `h`
+
+As a check, we would like to reproduce `h` using the right hand side of Equation \@ref(eq:pthetagiveny).
+To do this, we start by creating an objective function `f` (see Appendix for `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!).
+This is what we will be using for the numerator of Equation \@ref(eq:pthetagiveny):
+
+```{r}
+f <- MakeADFun(
+  data = dat,
+  parameters = param,
+  DLL = "model1",
+  silent = TRUE
+)
+```
+
+```{r}
+input_param <- param
+```
+
+For the `phi_prev` parameters, we want to be using the `mean` that we previously computed, then evaluating everything using `f`.
+We multiply by -1 because `f$fn` gives the negative log-likelihood, and we want just the log-likelihood:
 
@@
 numer
+```
 
+Then for the denominator, it should be as follows:
+
+```{r}
 denom <- 0.5 * log(det(2 * pi * Q))
@@
 -1 * (numer - denom)
+```
+
+As compared with what we got from `h`.
+They're not the same!
+Something is wrong here:
+
+```{r}
+h$fn(input_param[c("beta_prev", "log_sigma_phi_prev")])
+```
+
+## Moving to $\tilde \tilde p(x_i \, | \, y)$
+
+Let's have a go at computing $\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z)$ so that we can compare to the posterior distribution from `tmbstan`.
+
+We have simplified matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
+$$
+\begin{align}
+\tilde p(x_i \, | \, y) &= \tilde p_\text{G}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
+&\propto \tilde p_\text{G}(x_i \, | \, \hat \theta, y)
+\end{align}
+$$
+We're going to evaluate this by taking samples from the full Gaussian, the subsetting them:
+
+```{r}
+#' To write
 ```

### 2022-10-10 src/dev_sinla/sinla.Rmd
Getting there! Still some issues to fix r.e. normalising and weights, cleaning up

@@
 $$
-\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
+\begin{equation}
+\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxgiveny)
+\end{equation}
 $$
@@
 
-This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triange of the curvature at a randomly chosen quadrature point).
+This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triangle of the curvature at a randomly chosen quadrature point).
 For the $i$th marginal, we just take the appropriate samples from the joint.
@@
 $$
-\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z),
+\begin{equation}
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxigiveny)
+\end{equation}
 $$
@@
 ```{r}
-mcmc <- tmbstan::tmbstan(h, chains = 4, iter = 5000)
+mcmc <- tmbstan::tmbstan(h, chains = 4, iter = 5000, refresh = 0)
 ```
@@
 ```{r}
-extract(mcmc, pars = "phi_prev[1]") %>%
+plot <- extract(mcmc, pars = "phi_prev[1]") %>%
   as.data.frame() %>%
   ggplot(aes(x = phi_prev.1.)) +
-    geom_histogram(aes(y = ..count.. / sum(..count..)), alpha = 0.8, fill = cbpalette[7]) +
+    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
     theme_minimal() +
     labs(x = "phi_prev[1]", y = "Posterior")
+
+plot
 ```
@@
 
-## $p(\theta \, | \, y)$
+## $\tilde p(\theta \, | \, y)$
 
@@
 
-## Moving to $\tilde \tilde p(x_i \, | \, y)$
+## $\tilde p(x_i \, | \, y)$
 
-Let's have a go at computing $\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z)$ so that we can compare to the posterior distribution from `tmbstan`.
-
-We have simplified matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
+Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
+We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
 $$
@@
 $$
-We're going to evaluate this by taking samples from the full Gaussian, the subsetting them:
+We can evaluate this for $i = 1$ via:
 
 ```{r}
-#' To write
+i <- 1
+mean_i <- as.numeric(mean[i])
+var_i <- solve(Q)[ i, i]
+```
+
+And compare it to our MCMC results as follows:
+
+```{r}
+plot2 <- plot +
+  stat_function(
+    data = data.frame(x = c(0.5, 3.5)), aes(x), fun = dnorm, n = 101, 
+    args = list(mean = mean_i, sd = sqrt(var_i)), col = cbpalette[2], size = 1) +
+  labs(x = "prev_phi[1]", y = "Posterior") +
+  theme_minimal()
+
+plot2
 ```
@@
 # Full Laplace approximation
+
+## $\tilde p(x_i \, | \, \theta, y)$
 
@@
     random = "phi_prev_minus_i",
-    DLL = "model1index"
+    DLL = "model1index",
+    silent = TRUE
   )
@@
 
-  sd_out <- sdreport(
-    obj,
-    par.fixed = opt$par,
-    getJointPrecision = TRUE
-  )
-
   #' Parameters of the Gaussian approximation
-  mean <- with(obj$env, last.par[random])
-  Q <- obj$env$spHess(with(obj$env, last.par), random = TRUE)
+  mm <- obj$env$last.par
+  mean <- mm[obj$env$random]
+  Q <- obj$env$spHess(mm, random = TRUE)
 
@@
 
-We should be able to reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
+# Reproducing output from `obj`
+
+As above, we should be able to reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
 Recall that `f` takes as input the complete $(x, \theta)$ ($y$ is already fixed).
@@
 
+## $\tilde p(x_i \, | \, y)$
+
+Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
+We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
+$$
+\begin{align}
+\tilde p(x_i \, | \, y) &= \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
+&\propto \tilde p_\text{LA}(x_i \, | \, \hat \theta, y)
+\end{align}
+$$
+
+Start by getting $\hat \theta$:
+
+```{r}
+opt <- nlminb(
+  start = h$par,
+  objective = h$fn,
+  gradient = h$gr,
+  control = list(iter.max = its, trace = 0)
+)
+
+#' How many equally spaced points do we want to evaluate the posterior at?
+phi_prev_i_grid_length <- 50
+
+df <- opt$par %>%
+  t() %>%
+  as.data.frame() %>%
+  slice(rep(1:n(), each = phi_prev_i_grid_length)) %>%
+  mutate(
+    phi_prev_i = seq(from = 0.5, to = 3.5, length.out = phi_prev_i_grid_length)
+  ) %>%
+  rowwise() %>%
+  mutate(
+    #' Note here: it's crucial that the arguments are in the right order!
+    logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev))
+  )
+
+ggplot(df, aes(x = phi_prev_i, y = logpost)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "phi_prev[1]", "log(posterior)")
+```
+
+What we've got here is the unnormalised log-posterior?
+How can we go from that to a normalised posterior?
+Going to do trapezoid rule here, but AGHQ would work better (even with a low number of points) because the function is close to being Gaussian.
+
+$$
+\begin{align}
+\tilde f(x) &= \frac{f(x)}{\int f(x) \text{d}x} = \frac{f(x)}{C}
+\implies \\
+\log\tilde f(x) &= \log f(x) - \log C
+\end{align}
+$$
+
+```{r}
+df <- df %>%
+  mutate(
+    logpost_mean = colMeans(select(., logpost)),
+    logpost_adj = logpost - logpost_mean,
+    post = exp(logpost_adj)
+  )
+
+plot3 <- plot2 +
+  geom_line(data = df, aes(x = phi_prev_i, y = post), col = cbpalette[5], size = 1)
+
+#' Need to normalise this!
+plot3
+```
+
 # Simplified INLA
 
-$$
-\hat \mu_{-i}(x_i, \theta) = [\hat \mu(\theta)]_{-i} + \hat Q(\theta)^{-1}_{-i, i} \hat Q(\theta)_{i, i}(x_i - [\hat \mu(\theta)]_{i})
-$$
+This remains to-do!
 

### 2022-10-11 src/dev_sinla/sinla.Rmd
Fixing missing minus sign in 2 * pi, and a header at wrong level

@@
 \begin{equation}
-\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat{\mu}(\theta)} = \frac{p(y, \hat{\mu}(\theta), \theta)}{\det(2 \pi \hat{Q}(\theta))^{1/2}}, (\#eq:pthetagiveny)
+\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat{\mu}(\theta)} = \frac{p(y, \hat{\mu}(\theta), \theta)}{\det \left(\frac{1}{2 \pi} \hat{Q}(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
 \end{equation}
@@
 ```{r}
-denom <- 0.5 * log(det(2 * pi * Q))
+denom <- 0.5 * (determinant(Q, logarithm = TRUE)$modulus - (nrow(Q) * log(2 * pi)))
 denom
@@
 
-On the negative log-likelihood scale, the number we get is:
+Note that we use `determinant` on the log-scale as it is more numerically stable than e.g. `log(det(...))`.
+On the log-likelihood scale, the number we get is:
 
 ```{r}
--1 * (numer - denom)
+numer - denom
 ```
@@
 ```{r}
-h$fn(input_param[c("beta_prev", "log_sigma_phi_prev")])
+-1 * h$fn(input_param[c("beta_prev", "log_sigma_phi_prev")])
 ```
@@
   \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \hat{\mu}_{-i}(x_i, \theta)} \\
-  &= \frac{p(y, x_i, \hat{\mu}_{-i}(x_i, \theta), \theta)}{\det(2 \pi \hat{Q}_{-i}(x_i, \theta))^{1/2}}
+  &= \frac{p(y, x_i, \hat{\mu}_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} \hat{Q}_{-i}(x_i, \theta))^{1/2}}
 \end{align*}
@@
 
-# Reproducing output from `obj`
+## Reproducing output from `obj`
 
@@
 ```{r}
--1 * (numer - denom)
+numer - denom
 ```
@@
 ```{r}
-obj$fn(transformed_param)
+-1 * obj$fn(transformed_param)
 ```

### 2022-10-11 src/example_inla-replication/inla-replication.Rmd
A few small updates to this. Will need a big update later

@@
 1. Locate the mode of $\log p(\theta \, | \, y)$ which is the same as $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$ and can be calculated as above.
-2. Calculate negative Hessian at the mode.
-The solution is likely to use some version of `h$env$spHess` although it's not clear which one.
-3. Test points against the $\delta_z$ condition
+2. Calculate negative Hessian at the mode using e.g. `h$env$spHess`.
+3. Reparameterise in terms of $z$ coordinates along the eigenvectors of the Hessian starting at the mode
+3. Test points at a certain step-size ($\delta_z$) against a condition whereby they must have sufficient posterior density (using a parameter $\delta_{\pi}$)
+
+For a more detailed run through of the grid integration method, see the [INLA two dimensional grid replication](https://athowes.github.io/elgm-inf/inla-grid.html) notebook.
 
@@
 
+We could also use an adaptive Gauss Hermite quadrature rule.
+See the [Implementing Approximate Bayesian Inference using
+Adaptive Quadrature: the `aghq` Package](https://arxiv.org/pdf/2101.04468.pdf) paper.
 ## Stage 3: get $\tilde p(x_i \, | \, \theta, y)$
@@
 
-Obtain marginals of $\tilde p_G(x \, | \, \theta, y)$ from above.
+Obtain marginals of $\tilde p_{\text{G}}(x \, | \, \theta, y)$ from above.
 
@@
 
-Compute the Laplace approximation
+Compute the Laplace approximation $\tilde p_{\text{LA}}(x_i \, | \, \theta, y)$.
 
@@
 
+Combine the Laplace approximation to the hyperparameter posterior from Stage 1, the quadrature nodes and weights from Stage 2, and the chosen posterior latent field marginal approximation from Stage 3 together in one equation:
 \begin{equation}
@@
 
+This is the example from the book [Spatial and Spatio-temporal Bayesian Models with `R-INLA`](https://onlinelibrary.wiley.com/doi/book/10.1002/9781118950203).
 Consider the following simple model for i.i.d. observations $y = (y_1, \ldots, y_n)$
@@
   ggplot(aes(x = index, y = y)) +
-  geom_point(shape = 4) +
+  geom_point() +
   geom_hline(yintercept = -3, col = "#666666", lty = "dashed") +

### 2022-10-12 docs/bayescomp.Rmd
Integrate feedback from Seth and Jeff

@@
 abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa.
-Performing computationally intensive inference using simulation-based methods like MCMC is infeasible in this setting as the model must be able to be easily run in production by countries, using their own data.
-We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, we show that our inference method obtains substantially more accurate inferences than the simple Gaussian approximation obtained via `TMB`, as compared with gold-standard NUTS results obtained via `tmbstan`.
-We provide an easy to use implementation as a part of the `aghq` R package allowing flexible use of the INLA method for a broader class of models than `R-INLA`.
-Given a `TMB` template, using our method requires minimal additional code."
+Multiple outcomes of interest, including HIV prevalence, incidence and ART coverage are jointly modelled using both household survey data and routinely reported health system data.
+The model is provided as a tool for countries to run using their own data.
+In this setting, computationally intensive inference methods like MCMC are unappealing.
+We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
+Using data from Malawi, we show that the inferences obtained by our method are substantially more accurate than the empirical Bayes Gaussian approximation approach currently used, as compared with inferences from Hamiltonian Monte Carlo with the No-U-Turn sampler as a gold-standard.
+We implement all methods using a common `TMB` C++ template for the log-posterior.
+By extending the `aghq` package (Stringer, 2021) we facilitate flexible use of the method with only a few lines of code."
 ```

### 2022-10-12 docs/bayescomp.Rmd
Feedback Alex and final-ish iteration

@@
 abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa.
-Multiple outcomes of interest, including HIV prevalence, incidence and ART coverage are jointly modelled using both household survey data and routinely reported health system data.
-The model is provided as a tool for countries to run using their own data.
-In this setting, computationally intensive inference methods like MCMC are unappealing.
+Multiple outcomes of interest, including HIV prevalence, HIV incidence and treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
+The model is provided as a tool for countries to input their data to and generate estimates.
+In this setting, computationally intensive inference methods like MCMC are impractical.
 We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, we show that the inferences obtained by our method are substantially more accurate than the empirical Bayes Gaussian approximation approach currently used, as compared with inferences from Hamiltonian Monte Carlo with the No-U-Turn sampler as a gold-standard.
-We implement all methods using a common `TMB` C++ template for the log-posterior.
-By extending the `aghq` package (Stringer, 2021) we facilitate flexible use of the method with only a few lines of code."
+Using data from Malawi, we show that the inferences obtained by our method are substantially more accurate than the empirical Bayes Gaussian approximation approach used currently, as compared with inferences from HMC with the No-U-Turn sampler as a gold-standard, where all methods are implemented via a common `TMB` C++ template for the log-posterior.
+By extending the `aghq` package (Stringer, 2021) we faciliate flexible and easy use of our method when provided a `TMB` template."
 ```

### 2022-10-12 docs/bayescomp.Rmd
Typo!
Break up long sentence

@@
 We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, we show that the inferences obtained by our method are substantially more accurate than the empirical Bayes Gaussian approximation approach used currently, as compared with inferences from HMC with the No-U-Turn sampler as a gold-standard, where all methods are implemented via a common `TMB` C++ template for the log-posterior.
-By extending the `aghq` package (Stringer, 2021) we faciliate flexible and easy use of our method when provided a `TMB` template."
+Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach used currently, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler.
+By extending the `aghq` package (Stringer, 2021) we facilitate flexible and easy use of our method when provided a `TMB` C++ template for the log-posterior."
 ```

### 2022-10-12 src/epil/epil.Rmd
Tidy and improve epil notebook

@@
 output:
-  html_document:
+  bookdown::html_document2:
     toc: yes
-    toc_float: true
-    toc_collapsed: true
+    toc_float: yes
+    toc_collapsed: yes
     df_print: paged
@@
     theme: lumen
+abstract: |
+  **Background** Rue, Martino and Chopin (2009) use the epilespy example from the OpenBUGs manual as demonstration.
+
+  **Task** We implement the epilepsy example in Stan, `R-INLA`, `TMB`, `glmmTMB`, `tmbstan` and `aghq`. We then compare the results of the inference methods.
 ---
@@
 ```
+
+# Background
 
@@
 
+# Inference methods
+
 ## Stan
+
+Hamiltonian Monte Carlo via the No-U-Turn algorithm, with one Markov chain of `iter = 1000` samples, and `warmup = 100` burn-in:
 
@@
 
-## INLA
+## `R-INLA`
 
 Both $\tau_\epsilon$ and $\tau_\nu$ have the same $\Gamma(0.001, 0.001)$ prior.
-In INLA, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior:
+In `R-INLA`, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior:
 
 ```{r}
-tau_prior <- list(prec = list(prior = "loggamma",
+tau_prior <- list(prec = list(
+  prior = "loggamma",
   param = c(0.001, 0.001),
   initial = 1,
-                              fixed = FALSE))
+  fixed = FALSE)
+)
 ```
@@
   f(Ind,  model = "iid", hyper = tau_prior)    # Epsilon random effect
+```
 
+We can use various settings for the strategy, which controls how the latent field posterior marginals are approximated:
+
+```{r}
 epil_inla <- function(strat) {
@@
 
-## TMB
+## `TMB`
 
@@
 
-`random` are integrated out with a Laplace approximation:
+We integrate `random` out with a Laplace approximation:
 
@@
   random = c("epsilon", "nu"),
-  DLL = "epil"
+  DLL = "epil",
+  silent = TRUE
 )
@@
 
-```{r results='hide'}
+```{r}
 its <- 1000 #' May converge before this
+
 opt <- nlminb(
@@
 )
-
-#' Check that TMB and Stan have the same objective function for sanity.
-#' To obtain the TMB negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
-#' For Stan, we create an empty model, then use `rstan::log_prob`.
-
-# tmb_nll <- MakeADFun(data = dat, parameters = param, DLL = "epil") # TMB objective
-# stan_nll <- stan("model/epil.stan", data = dat, chains = 0) # Stan objective
-
-#' Now test the NLL of the initialisation parameters `param`:
-
-# c(
-#   "TMB" = tmb_nll$fn(unlist(param)),
-#   "Stan" = -rstan::log_prob(object = stan_nll, unlist(param))
-# )
-
-#' To be more thorough, we can get all the parameters obtained using the MCMC:
-
-# pars_mat <- as.matrix(fit1)
-# pars_list <- apply(pars_mat, 1, function(x) relist(flesh = x, skeleton = fit1@inits[[1]]))
-# upars_list <- lapply(pars_list, function(x) unconstrain_pars(fit1, x))
-
-#' Just use a few of them
-
-# test_pars <- upars_list[1001:1010]
-# tmb_evals <- sapply(test_pars, tmb_nll$fn)
-# stan_evals <- -sapply(test_pars, FUN = rstan::log_prob, object = stan_nll)
-#
-# data.frame(
-#   "TMB" = tmb_evals,
-#   "Stan" = stan_evals
-# )
-
-#' Note: When there are errors in the C++ template code, usually to do with indexing (unlike Stan, in TMB there is no requirement when defining variables to give them dimensions), calling the `MakeADFun` function tends to crash the R session.
-#' A workaround (courtesy of Kinh) for debugging without crashing the working R session is to use the following (which creates new R sessions which crash in preference to the working R session).
-
-# library(parallel)
-# testrun <- mcparallel({MakeADFun(data = dat,
-#                                  parameters = param,
-#                                  DLL = "epil")})
-# obj <- mccollect(testrun, wait = TRUE, timeout = 0, intermediate = FALSE)
 ```
 
-## glmmTMB
+For completeness, we check here that the objective functions as implemented in Stan and `TMB` are identical.
+To obtain the `TMB` negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
+For Stan, we create an empty model, then use `rstan::log_prob`:
 
-[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in TMB without writing the C++ code.
+```{r}
+tmb_nll <- MakeADFun(data = dat, parameters = param, DLL = "epil")
+stan_nll <- stan("epil.stan", data = dat, chains = 0)
+```
+
+Now test the NLL of the initialisation parameters `param`:
+
+```{r}
+c(
+  "TMB" = tmb_nll$fn(unlist(param)),
+  "Stan" = -rstan::log_prob(object = stan_nll, unlist(param))
+)
+```
+
+To be more thorough, we can get all the parameters obtained using the MCMC:
+
+```{r}
+pars_mat <- as.matrix(fit1)
+pars_list <- apply(pars_mat, 1, function(x) relist(flesh = x, skeleton = fit1@inits[[1]]))
+upars_list <- lapply(pars_list, function(x) unconstrain_pars(fit1, x))
+```
+
+Using just a few of them, we can see that there is a difference, but that difference is a constant so doesn't matter:
+
+```{r}
+test_pars <- upars_list[500:510]
+tmb_evals <- sapply(test_pars, tmb_nll$fn)
+stan_evals <- -sapply(test_pars, FUN = rstan::log_prob, object = stan_nll)
+
+data.frame(
+  "TMB" = tmb_evals,
+  "Stan" = stan_evals
+) %>%
+  mutate(Difference = TMB - Stan)
+```
+
+Note: When there are errors in the C++ template code, usually to do with indexing (unlike Stan, in `TMB` there is no requirement when defining variables to give them dimensions), calling the `MakeADFun` function tends to crash the R session.
+A workaround (courtesy of Kinh) for debugging without crashing the working R session is to use the following (which creates new R sessions which crash in preference to the working R session):
+
+```{r eval=FALSE}
+library(parallel)
+
+testrun <- mcparallel({MakeADFun(data = dat,
+                                 parameters = param,
+                                 DLL = "epil")})
+
+obj <- mccollect(testrun, wait = TRUE, timeout = 0, intermediate = FALSE)
+```
+
+## `glmmTMB`
+
+[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in `TMB` without writing the C++ code.
 For example, rather than writing `epil.cpp` manually, we could have called the following:
@@
 
-## tmbstan
+This likely isn't the same model, in terms of priors and so on, but it may be close.
 
-[`tmbstan`](https://journals.pl,os.org/plosone/article?id=10.1371/journal.pone.0197954) (Cole Monnahan and Kasper Kristensen) is another helpful TMB package which allows you to pass the same C++ template you use in TMB to Stan in order to perform NUTS (if you have standard C++ code then this can also likely be done using [`stanc`](https://statmodeling.stat.columbia.edu/2017/03/31/running-stan-external-c-code/)).
+## `tmbstan`
+
+[`tmbstan`](https://journals.pl,os.org/plosone/article?id=10.1371/journal.pone.0197954) (Cole Monnahan and Kasper Kristensen) is another helpful `TMB` package which allows you to pass the same C++ template you use in `TMB` to Stan in order to perform NUTS (if you have standard C++ code then this can also likely be done using [`stanc`](https://statmodeling.stat.columbia.edu/2017/03/31/running-stan-external-c-code/)).
 
 ```{r}
-fit7 <- tmbstan(obj = obj, chains = 4)
+fit7 <- tmbstan(obj = obj, chains = 1)
 ```
 
-## aghq
+## `aghq`
 
-[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer.
-One approach is to use `glmmTMB` to get the TMB template which can be passed the `aghq`.
+[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer which implements adaptive Gauss-Hermite quadrature.
+One (strange!) approach is to use `glmmTMB` to get the `TMB` template which can be passed the `aghq`.
 
 ```{r}
-# glmm_model_info <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"), doFit = FALSE)
-#
-# glmm_ff <- with(glmm_model_info, {
-#   TMB::MakeADFun(
-#     data = data.tmb,
-#     parameters = parameters,
-#     random = names(parameters)[grep("theta", names(parameters), invert = TRUE)],
-#     DLL = "glmmTMB",
-#     silent = TRUE
-#   )
-# })
-#
-# glmm_quad <- aghq::marginal_laplace_tmb(glmm_ff, k = 3, startingvalue = glmm_ff$par)
+glmm_model_info <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"), doFit = FALSE)
+
+glmm_ff <- with(glmm_model_info, {
+  TMB::MakeADFun(
+    data = data.tmb,
+    parameters = parameters,
+    random = names(parameters)[grep("theta", names(parameters), invert = TRUE)],
+    DLL = "glmmTMB",
+    silent = TRUE
+  )
+})
+```
+
+The parameter `k = 3` determines the number of quadrature points per dimension:
+
+```{r}
+glmm_quad <- aghq::marginal_laplace_tmb(glmm_ff, k = 3, startingvalue = glmm_ff$par)
 ```
@@
 
-## Comparison
+# Comparison
+
+## Table
 
@@
 
+kableExtra::kable(df)
+
 saveRDS(df, "comparison-results.rds")
@@
 
-Assuming tmbstan to be the ground truth, we can do two dimensional plots to check the fit
+## Scatter plot against `tmbstan`
+
+Assuming `tmbstan` to be the ground truth, we can do two dimensional plots to check the fit
 
@@
   geom_point(aes(x = tmbstan, y = TMB, col = type)) +
-  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5)
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
+  theme_minimal()
 ```
+
+## Symmetric KL divergence
 
@@
 
-For `laplace` and `"simplified.laplace"`:
+For `laplace` and `"simplified.laplace"` we can look for the random effect with the highest SKLD:
 
@@
 plot_marginals <- function(random_effect, index) {
-  marginal2 <- fit2$marginals.random[[random_effect]][paste0("index.", index)][[1]]
-  marginal3 <- fit3$marginals.random[[random_effect]][paste0("index.", index)][[1]]
-  marginal4 <- fit4$marginals.random[[random_effect]][paste0("index.", index)][[1]]
+  marginal2 <- fit2$marginals.random[[random_effect]][paste0("index.", index)][[1]] %>%
+    as.data.frame() %>%
+    mutate(method = "G")
 
-  plot(marginal2, type = "l", col = "red")
-  lines(marginal3, col = "green")
-  lines(marginal4, col = "blue")
+  marginal3 <- fit3$marginals.random[[random_effect]][paste0("index.", index)][[1]] %>%
+    as.data.frame() %>%
+    mutate(method = "SL")
+
+  marginal4 <- fit4$marginals.random[[random_effect]][paste0("index.", index)][[1]] %>%
+    as.data.frame() %>%
+    mutate(method = "L")
+
+  marginals <- bind_rows(marginal2, marginal3, marginal4)
+
+  ggplot(marginals, aes(x = x, y = y, group = method, col = method)) +
+    geom_line() +
+    theme_minimal()
 }

### 2022-10-12 make/epil.R

@@ -1,3 +1,0 @@
-source("make/utils.R")
-
-run_commit_push("epil")

### 2022-10-12 src/dev_sinla/sinla.Rmd
Work on making the LHS vs RHS check better, and remove it the second time (they're essentially the same)

@@
 
-```{r}
-input_param <- param
-```
-
-For the `phi_prev` parameters, we want to be using the `mean` that we previously computed, then evaluating everything using `f`.
-We multiply by -1 because `f$fn` gives the negative log-likelihood, and we want just the log-likelihood:
-
-```{r}
-input_param$phi_prev <- mean
-numer <- -1 * f$fn(unlist(input_param))
-numer
-```
-
-Then for the denominator, it should be as follows:
-
-```{r}
-denom <- 0.5 * (determinant(Q, logarithm = TRUE)$modulus - (nrow(Q) * log(2 * pi)))
-denom
-```
-
-Note that we use `determinant` on the log-scale as it is more numerically stable than e.g. `log(det(...))`.
+For the `phi_prev` parameters, we compute their `mean` and `Q` using `h`.
+We then we evaluate everything using `f` to get the numerator.
+We multiply by -1 because `f$fn` gives the negative log-likelihood, and we want just the log-likelihood.
+In the denominator, we use `determinant` on the log-scale as it is more numerically stable than e.g. `log(det(...))`.
 On the log-likelihood scale, the number we get is:
@@
 ```{r}
+calculate_rhs <- function(theta) {
+  #' Calculate mean and Q for those theta
+  #' TMB template remembers what you've called -- spooky
+  h$fn(theta)
+  mm <- h$env$last.par
+  mean <- mm[h$env$random]
+  Q <- h$env$spHess(mm, random = TRUE)
+
+  #' Append the calculated latent field mean
+  param <- theta
+  param$phi_prev <- as.numeric(mean)
+  param <- param[c("beta_prev", "phi_prev", "log_sigma_phi_prev")]
+
+  #' Evaluate the log-numerator of the RHS
+  numer <- -1 * f$fn(unlist(param))
+
+  #' Evaluate the log-denominator of the RHS
+  denom <- 0.5 * (determinant(Q, logarithm = TRUE)$modulus - (nrow(Q) * log(2 * pi)))
+
+  #' Return their difference
   numer - denom
+}
+
+theta <- param
+theta$phi_prev <- NULL
+
+calculate_rhs(theta)
 ```
@@
 ```{r}
--1 * h$fn(input_param[c("beta_prev", "log_sigma_phi_prev")])
+-1 * h$fn(input_hyper)
 ```
@@
 
-## Reproducing output from `obj`
-
-As above, we should be able to reproduce the output of `obj$fn` using the mean and precision matrix, together with the function `f`.
-Recall that `f` takes as input the complete $(x, \theta)$ ($y$ is already fixed).
-For example:
-
-```{r}
-param
-f$fn(unlist(param))
-```
-
-We update `param` to include $\hat{\mu}_{-i}(x_i, \theta)$, so that it's only element `r i` of `phi_prev` which remains as before:
-
-```{r}
-param$phi_prev[-i] <- mean
-param
-```
-
-The numerator of the Laplace approximation is (n.b. this is on the log-scale, which is where we will do the calculations):
-
-```{r}
-numer <- -1 * f$fn(unlist(param))
-numer
-```
-
-And the denominator of the Laplace approximation (again on the log-scale) is:
-
-```{r}
-denom <- 0.5 * (determinant(Q, logarithm = TRUE)$modulus - (nrow(Q) * log(2 * pi)))
-denom
-```
-
-```{r}
-numer - denom
-```
-
-Compare this to the output from `obj`:
-
-```{r}
--1 * obj$fn(transformed_param)
-```
-
 ## $\tilde p(x_i \, | \, y)$

### 2022-10-12 src/dev_sinla/sinla.Rmd
Normalise the Laplace approximation with trapezoid rule in 1D

@@
 #' How many equally spaced points do we want to evaluate the posterior at?
-phi_prev_i_grid_length <- 50
+spacing <- 0.05
+x_grid <- seq(from = 0.5, to = 3.5, by = spacing)
+x_grid_length <- length(x_grid)
 
@@
   as.data.frame() %>%
-  slice(rep(1:n(), each = phi_prev_i_grid_length)) %>%
+  slice(rep(1:n(), each = x_grid_length)) %>%
   mutate(
-    phi_prev_i = seq(from = 0.5, to = 3.5, length.out = phi_prev_i_grid_length)
+    phi_prev_i = x_grid
   ) %>%
@@
 ```{r}
-df <- df %>%
-  mutate(
-    logpost_mean = colMeans(select(., logpost)),
-    logpost_adj = logpost - logpost_mean,
-    post = exp(logpost_adj)
-  )
+safe_exp <- function(x) exp(x - max(x))
+
+trapezoid_rule <- function(x, spacing) {
+  0.5 * spacing * (2 * sum(x) - x[1] - x[2])
+}
+
+df$post <- safe_exp(df$logpost)
+df$post_norm = df$post / trapezoid_rule(df$post, spacing)
 
 plot3 <- plot2 +
-  geom_line(data = df, aes(x = phi_prev_i, y = post), col = cbpalette[5], size = 1)
+  geom_line(data = df, aes(x = phi_prev_i, y = post_norm), col = cbpalette[5], size = 1)
 
-#' Need to normalise this!
 plot3
 ```
+
+So perhaps it's barely better than the Gaussian, but not really getting the real problem here.
+Hypothesis would be that it's because we're doing empirical Bayes, and with more than one hyperparameter grid point we would do better.
 

### 2022-10-12 src/docs_01-04-20-mini/01-04-20-mini.Rmd
Upload old mini project presentation

---
title: "Mini-project: Integrated nested Laplace Approximation with Automatic Differentiation"
author: "Adam Howes"
date: "April 2020"
output:
  beamer_presentation:
    fig_height: 2.5
    fig_width: 6
    highlight: haddock
    includes:
      in_header: preamble.tex
    latex_engine: pdflatex
  slidy_presentation:
    highlight: haddock
institute: Imperial College London
bibliography: citations.bib
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
```

## Motivation

* `R-INLA` [@martins2013bayesian] only works for the particular models which have been implemented
* Alternative implementation based on automatic differentation (AD) would allow INLA to be used for a \textcolor{hilit}{broader class} of models
* For example, the HIV inference group at Imperial is working on a model just outside `R-INLA`'s capacity

# What is INLA, why do we want to use it, and why can't we currently?

## Three-stage model

* Want to do Bayesian inference in spatiotemporal statistics
* \textcolor{hilit}{Three-stage model} covers most of the models used
\begin{alignat*}{2}
  &\text{(Observations)}     &   \y &\sim p(\y \, | \, \x), \\
  &\text{(Latent field)}     &   \x &\sim p(\x \, | \, \btheta), \\
  &\text{(Hyperparameters)}  &   \qquad \btheta &\sim p(\btheta),
\end{alignat*}
where $\y = (y_1, \ldots, y_n)$, $\x = (x_1, \ldots, x_n)$, $\btheta = (\theta_1, \ldots, \theta_m)$
* Interested in learning \textcolor{hilit}{both} $(\btheta, \x)$ from data $\y$

## Have you tried MCMC?

* Markov chain Monte Carlo is slow for high dimensional correlated parameter spaces
* We have both of these problems:
  * If $\x$ represents spatiotemporal location then $\text{dim}(\x) = n$ will be very large 
  * Tobler's first law of geography "everything is related to everything else, but near things are more related than distant things" $\implies \x$ has lots of correlation structure

## Approximate Bayesian inference

* In applied statistics (at least in health and social science) we fit misspecified models to biased and incomplete data
* Is inferential exactness (as $n_\text{sim} \to \infty$ for chain of length $n_\text{sim}$) the scientific bottleneck?
* If not $\implies$ shouldn't be afraid of approximate methods
    * Approximate Bayesian computation (ABC)
    * Variational Bayes
    * \textcolor{hilit}{Integrated nested Laplace approximation} (INLA)

## Integrated nested Laplace approximation (I)

* See @rue2009approximate or @blangiardo2015spatial
* Approximate Bayesian inference for \textcolor{hilit}{latent Gaussian models} (LGMs), which are three-stage models with middle layer
\begin{equation*}
  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
\end{equation*}
* Takes advantage of sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)
* Gives approximate \textcolor{hilit}{posterior marginals} $\{\tilde p(x_i \, | \, \y)\}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$

## Integrated nested Laplace approximation (II)

1) First Laplace approximate hyperparameter posterior
\begin{equation}
\tilde p(\btheta \, | \, \y) \propto \frac{p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)}{\tilde p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)}  \label{eq:hypermarginal}
\end{equation}
which can be marginalised to get $\tilde p(\theta_j \, | \, \y)$
2) Choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$ to integrate w.r.t. \eqref{eq:hypermarginal}
3) Choose approximation for $\tilde p(x_i \, | \, \btheta, \y)$ (simplest version: Gaussian)
4) Finally use quadrature to get 
\begin{equation}
  \tilde p(x_i \, | \, \y) = 
  \sum_{k = 1}^K \tilde p(x_i \, | \, \btheta^{(k)}, \y) \times \tilde p(\btheta^{(k)} \, | \, \y) \times \Delta^{(k)}
\end{equation}

## Naomi, evidence synthesis for HIV

:::::::::::::: {.columns}

::: {.column width=.65}

* Combine HIV prevalence $\rho_i$ and ART coverage $\alpha_i$ models together
* Model is close to, but not, a LGM
* Small non-linearities e.g. multiplying two latent Gaussian fields 
\begin{equation*}
A_i \sim \text{Bin}(\rho_i \alpha_i, N_i),
\end{equation*}
where $A_i$ be the number observed on ART and $N_i$ the population
* Need something more flexible than `R-INLA`

:::

::: {.column width=.35}

```{r, echo=FALSE, fig.cap="Supermodel", out.width = '65%'}
knitr::include_graphics("naomi_hex.png")
```

:::

::::::::::::::

# What do we do currently instead?

## Template Model Builder (I)

* Currently we use `TMB` [@tmb]
* R package which implements the Laplace approximation for latent variable models using AD (via `CppAD`)
  * For more about AD see e.g. @griewank2008evaluating
* Write an objective function $f(\x, \btheta)$ in C++ ("user template")
    * We select $f(\x, \btheta) = - \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)$

## Template Model Builder (II)

```{cpp eval=FALSE}
#include <TMB.hpp>

template <class Type>
Type objective_function<Type>::operator()() {
  // Define data e.g.
  DATA_VECTOR(y);
  // Define parameters e.g.
  PARAMETER(mu);
  // Calculate negative log-likelihood e.g.
  nll = Type(0.0);
  nll -= dnorm(y, mu, 1, true).sum()
  return(nll);
}
```

## Template Model Builder (III)

* Performs the Laplace approximation $L_f(\btheta) \approx L^\star_f(\btheta)$ (step 1 of INLA) -- use R to optimise this with respect to $\btheta$ to give $\hat \btheta$
* MAP estimate of $\x$ conditional on $\btheta = \hat \btheta$ (REML? Empirical Bayes?)
* Standard errors calculated using the $\delta$-method (a Gaussian assumption)

# What do you want to do in the future?

## Aims

* Compare accuracy of `TMB` to `R-INLA`
* Implement the INLA method using AD via `TMB`
* Apply new method to models with different degrees of non-linearity
    * Small degree: Naomi. 
    * Larger degree: ODE models e.g. SIR or other compartmental models

# Thanks! Questions / comments / corrections?

## References {.allowframebreaks}

### 2022-10-12 make/_make.R

@@ -5,5 +5,6 @@ run_commit_push("example_inla-grid")
 run_commit_push("explore_aghq")
 run_commit_push("dev_sinla")
 run_commit_push("docs_paper")
+run_commit_push("docs_01-04-20-mini")
 run_commit_push("docs_01-07-21-stats-epi-group")
 run_commit_push("docs_15-11-22-seminar")

### 2022-10-13 src/dev_sinla/sinla.Rmd
Working on SINLA algorithm

@@
 
-# Marginal of joint Gaussian approximation
+# Marginal of joint Gaussian approximation {#gaussian-marginal}
 
@@
 \begin{equation}
-\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat{\mu}(\theta)} = \frac{p(y, \hat{\mu}(\theta), \theta)}{\det \left(\frac{1}{2 \pi} \hat{Q}(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
+\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \mu(\theta)} = \frac{p(y, \mu(\theta), \theta)}{\det \left(\frac{1}{2 \pi} Q(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
 \end{equation}
@@
 
-# Full Laplace approximation
+# Full Laplace approximation {#full-laplace}
 
@@
 \begin{align*}
-  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \hat{\mu}_{-i}(x_i, \theta)} \\
-  &= \frac{p(y, x_i, \hat{\mu}_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} \hat{Q}_{-i}(x_i, \theta))^{1/2}}
+  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \mu_{-i}(x_i, \theta)} (\#eq:pxigiventhetay) \\
+  &= \frac{p(y, x_i, \mu_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} Q_{-i}(x_i, \theta))^{1/2}}
 \end{align*}
@@
 
-This remains to-do!
+First, notational differences between Wood (2019) (former) and here (later) are the following:
+
+* $\beta$ vs. $x$ for random effects
+* $\hat \beta$ vs $\mu (\theta)$ for the random effects which maximise the template
+* $\tilde \beta$ vs $(x_i, \mu(x_i, \theta))$ for the random effects which maximise the template subject to $x_i$ being fixed
+* $H$ vs $Q(\theta)$ for the Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$
+
+In Section \@ref(full-laplace) we directly calculate the Laplace approximation to $\int p(x, y, \theta) \text{d} \theta_{-i}$.
+Following Wood (2019), a cheaper approach would be to "base the denominator of Equation \@ref(eq:pxigiventhetay) on the conditional density implied by ? in which case the Hessian is constant and
+$$
+\mu(x_i, \theta) = \mu(\theta) + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:wood3)
+$$
+The proposed modification has two parts:
+
+1. Use the numerically exact $\mu(x_i, \theta)$ rather than an approximation (as in simplified Laplace)
+2. An approximation to the required Hessian
+
+## Complete algorithm
+
+In this section we omit dependence on $\theta$ such that: $\mu = \mu(\theta)$, $Q = Q(\theta)$, $\mu(x_i) = \mu(x_i, \theta)$ and $Q(x_i) = Q(x_i, \theta)$.
+
+Want: $\tilde p(x_i \, | \, \theta, y)$
+
+Let $\mathcal{D}_{2j}$ denote a $2j \times 2j$ diagonal matrix with leading diagonal $-1, 1, -1, 1, \ldots$.
+Let $u_0$ be a zero column matrix.
+Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$.
+
+1. Compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$
+2. For each $x_i$ in a grid of evaluation points (how does one chose this grid?) repeat the following steps a-e)
+  a) Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:wood3)
+  b) Compute a set of $J$ steps $\{\Delta_j\}$
+  c) For $j = 1, \ldots, J - 1$ compute
+  $$
+  q = Q_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j
+  $$
+  and
+  $$
+  g = \Delta_x \log p(\mu(x_i) + \Delta_j^i, y, \theta).
+  $$
+  Then compute the matrix
+  $$
+  u_{j + 1} = \{q(\Delta_j^\top j)^{-1/2}, g_{-i}(\Delta_j^\top g_{-i})^{-1/2}, u_j\}.
+  $$
+  d) Compute the determinant approximation...
+  e) Compute $\tilde p(x_i \, | \, \theta, y) = \ldots$
+3. Renormalise $\tilde p(x_i \, | \, \theta, y)$
 

### 2022-10-17 src/dev_sinla/sinla.Rmd
Improve exposition and fix LHS/RHS discrepancy (fingers crossed)

@@
 
-```{r}
+```{r echo=FALSE}
 cbpalette <- multi.utils::cbpalette()
@@
 $$
-where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by exploration of $\tilde p_\text{LA}(\theta \, | \, y)$.
+where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
 In particular, let
@@
 $$
-where $H_k(z_d)$ are the Hermite polynomials for $k \in \mathbb{N}$, and
+be the set of nodes in one-dimension, exactly the zeros of the Hermite polynomials $H_k(z_d)$ for $k \in \mathbb{N}$, and
 $$
@@
 $$
-Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ with nodes $z$, of which there will be $k^d$.
+be the associated weights.
+Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
 Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.
@@
 
-Now, we want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
+The goal of this notebook is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
 $$
@@
 $$
-The goal of doing this is that the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
+The reasons for doing this is that the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
 At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
 This would be an accurate approach, but not computationally feasible for large models.
-What we hope to achieve is an approximation which gets to most of the accuracy benefits of this new Laplace approximation approach, but without the high computational cost.
+What we hope to achieve is an approximation which gets to most of the accuracy benefits of the Laplace approximation approach, but without the high computational cost.
 
@@
 We begin by loading in testing data, as generated by [`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim).
-`sim_data` contains many replicates of simulated data, so here we just take the first one, as given by:
+For more information about this model, see the [Prevalence, ANC, ART model inference case-study](https://athowes.github.io/elgm-inf/prev-anc-art.html).
+`sim_data` contains many replicates of simulated data.
+Here we only need one, so will will just take the first as given by:
 
@@
 ```{r}
+compile("model1.cpp")
+dyn.load(dynlib("model1"))
+```
+
+We create an objective function `h` (see Appendix for `TMB` template) integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`).
+The values of `param` intitialise the objective function:
+
+```{r}
 param <- list(
@@
 
-compile("model1.cpp")
-dyn.load(dynlib("model1"))
-```
-
-We create an objective function `h` (see Appendix for `TMB` template) integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`):
-
-```{r}
 h <- MakeADFun(
@@
 
-Fit the model using MCMC:
+It is simple to fit the model using MCMC via the `tmbstan` package.
+We use 4 `chains` each of 5000 `iter`:
 
@@
 
-What does the posterior marginal $p(\phi_1 \, | \, y)$ look like?
+The posterior marginal $p(\phi_1 \, | \, y)$ can be examined by extracting the corresponding Markov chains, and plotting a histogram of the draws:
 
-```{r}
+```{r message=FALSE, warning=FALSE}
 plot <- extract(mcmc, pars = "phi_prev[1]") %>%
@@
 
+Now we will compute Equation \@ref(eq:eq:pxgiveny) taking the marignals of a joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$.
+This is the approach currently used in the `aghq` package (done implicitly, via sampling).
+This approach is also used by `method = "gaussian"` in `R-INLA`.
+
 ## $\tilde p(\theta \, | \, y)$
-
-The `aghq` package currently approximates latent field posterior marginals by taking marginals of the joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ (this is done implicitly, via sampling).
-This is same approach used by `method = "gaussian"` in `R-INLA`.
 
@@
 $$
-such that one easy way to calculate it is to use set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there, as we have done in the function `h` above.
-Let's optimise `h`:
+such that one easy way to calculate it is to set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there, as we have done in the function `h` above.
+
+Let's optimise `h` to obtain estimates $\hat \theta$ which maximise $\tilde p(\theta \, | \, y)$:
 
@@
 
-The values `opt$par` below are the values of $\theta$ which maximise $\tilde p(\theta \, | \, y)$:
+The values `opt$par` below correspond to $\hat \theta$:
 
@@
 
-We can also look at the parameters (that is the mean vector and precision matrix) of the Gaussian approximation at the mode above (`opt$par`).
+We can also look at the parameters -- that is, mean vector and precision matrix -- of the Gaussian approximation at the mode `opt$par` above.
 Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
@@
 length(mean) #' The length is 30
-mean
 
@@
 As a check, we would like to reproduce `h` using the right hand side of Equation \@ref(eq:pthetagiveny).
-To do this, we start by creating an objective function `f` (see Appendix for `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!).
+To do this, we start by creating an objective function `f` (see Appendix for the `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!).
 This is what we will be using for the numerator of Equation \@ref(eq:pthetagiveny):
@@
 
-For the `phi_prev` parameters, we compute their `mean` and `Q` using `h`.
-We then we evaluate everything using `f` to get the numerator.
-We multiply by -1 because `f$fn` gives the negative log-likelihood, and we want just the log-likelihood.
+Given $\theta$, we compute the `mean` and `Q` of $x$ (corresponding to `phi_prev`) using `h`.
+We then concatenate and evaluate using `f` to get the numerator, multiplying by -1 because `f$fn` gives the negative log-likelihood, and we want the log-likelihood.
 In the denominator, we use `determinant` on the log-scale as it is more numerically stable than e.g. `log(det(...))`.
-On the log-likelihood scale, the number we get is:
 
@@
 theta <- param
-theta$phi_prev <- NULL
+theta$phi_prev <- NULL #' Setting the value of x to be NULL
 
@@
 
-As compared with what we got from `h`.
-They're not the same!
-Something is wrong here:
+As compared with what we got from `h`:
 
 ```{r}
--1 * h$fn(input_hyper)
+-1 * h$fn(unlist(theta))
 ```

### 2022-10-17 src/dev_sinla/sinla.Rmd
Small improvements and table of notation

@@
 
-# Ground truth
+# Ground truth {#ground-truth}
 
@@
 
-Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
-We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
+Now let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan` we obtained in Section \@ref(ground-truth).
+As described above, we simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
 $$
@@
 mean_i <- as.numeric(mean[i])
-var_i <- solve(Q)[ i, i]
+var_i <- solve(Q)[i, i]
 ```
@@
 
-```{r}
+```{r message=FALSE, warning=FALSE}
 plot2 <- plot +
   stat_function(
-    data = data.frame(x = c(0.5, 3.5)), aes(x), fun = dnorm, n = 101, 
-    args = list(mean = mean_i, sd = sqrt(var_i)), col = cbpalette[2], size = 1) +
-  labs(x = "prev_phi[1]", y = "Posterior") +
+    data = data.frame(x = c(0.5, 3.5)), aes(x, col = "Joint Gaussian marginal"), fun = dnorm, n = 101, 
+    args = list(mean = mean_i, sd = sqrt(var_i)), size = 1
+  ) +
+  labs(x = "prev_phi[1]", y = "Posterior", col = "Approximation") +
+  scale_colour_manual(values = cbpalette) +
   theme_minimal()
@@
 
+In this section we will continue to use the approximation \@ref(eq:pthetagiveny) for $\tilde p(\theta \, | \, y)$.
+
 ## $\tilde p(x_i \, | \, \theta, y)$
 
-Another approach we might take is to calculate the full Laplace approximation
+For the approximation to the latent field marginal posteriors, another approach we could take is to calculate the full Laplace approximation
 $$
@@
 $$
+integrating out $x_{-i}$ with a Gaussian approximation.
 This is not practical in some settings as it involves recomputing the Gaussian approximation $p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ for each value of $(x_i, \theta)$ which can be too computationally expensive.
-With of a bit of effort, we can implement this in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are named as such (see Appendix for template).
+We will still go through the steps for implementing it here though for demonstration.
+With of a bit of effort, it can be implemented in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are explicitly named (see Appendix for the `TMB` template `model1index.cpp`):
 
@@
 
-Now loop over doing the Laplace approximation for every index.
+Now we can loop over indices $i$ doing the required Laplace approximations.
 In the call to `MakeADFun` we set `random = "phi_prev_minus_i"` to integrate all of the random effects but that of the particular index.
@@
 plot3 <- plot2 +
-  geom_line(data = df, aes(x = phi_prev_i, y = post_norm), col = cbpalette[5], size = 1)
+  geom_line(data = df, aes(x = phi_prev_i, y = post, col = "Laplace"), size = 1)
 
@@
 
-First, notational differences between Wood (2019) (former) and here (later) are the following:
-
-* $\beta$ vs. $x$ for random effects
-* $\hat \beta$ vs $\mu (\theta)$ for the random effects which maximise the template
-* $\tilde \beta$ vs $(x_i, \mu(x_i, \theta))$ for the random effects which maximise the template subject to $x_i$ being fixed
-* $H$ vs $Q(\theta)$ for the Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$
+| Description | Here | Wood (2019) notation | Notes |
+|-------------|---------------|----------------------|----------|
+| Random effects | $x$ | $\beta$ | Stringer (2021) use $w$
+| Random effects which maximise (the / some) objective for given $\theta$ | $\mu (\theta)$ | $\hat \beta$ | Wood omits dependence on $\theta$ |
+| Random effects which maximise (the / some) objective for given $\theta$, subject to $x_i$ being fixed | $(x_i, \mu(x_i, \theta))$ | $\tilde \beta$ | We have $\mu(x_i, \theta))$ having dimension $n - 1$ |
+| Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$ | $Q(\theta)$ | $H$ | Unsure if it's preferable to use Hessian or precision notation here |
 

### 2022-10-17 src/explore_aghq/aghq.Rmd
marginal_laplace_tmb walkthrough
Add tabs to prev-anc-art plots, small text improvements

@@
 abstract: |
-  **Background**
+  **Background** `aghq` is an R package which implements adaptive Gauss-Hermite quadrature.
 
-  **Task**
+  **Task** In this notebook, we work through the vignettes for the `aghq` package, adding material as is useful to develop understanding.
 ---

### 2022-10-17 src/prev-anc-art_process-results/prev-anc-art.Rmd
Add tabs to prev-anc-art plots, small text improvements

@@
 
-    **Task** We compare inferences produced using tmbstan, `TMB` and `aghq` for a collection of HIV evidence synthesis models based upon the Naomi small-area estimation model.
+    **Task** We compare inferences produced using `tmbstan`, `TMB` and `aghq` for a collection of HIV evidence synthesis models based upon the Naomi small-area estimation model.
 
@@
 
-## Model description
+# Model description
 
@@
 
-### Prevalence component
+## Prevalence component
 
@@
 
-### ANC component
+## ANC component
 
@@
 
-### ART component
+## ART component
 
@@
 
-## Implementation
+# Implementation
 
@@
 For each model we write a `TMB` C++ template.
-As well as the standard `TMB` inference approach, this template allows the model to be fit using Stan via `tmbstan` and using adaptive Gaussian Hermite quadrature via `aghq`.
+As well as the standard `TMB` inference approach, this template allows the model to be fit using Stan via `tmbstan` and using adaptive Gauss-Hermite quadrature via `aghq`.
 
@@
 
-## Results
+# Results
 
@@
 
-### Boxplots
+## Boxplots {.tabset .tabset-fade}
 
@@
 
+```{r}
+params <- c("beta_prev", "log_sigma_phi_prev")
+```
+
+### Model 1
+
 ```{r cache=FALSE}
-params <- c("beta_prev", "log_sigma_phi_prev")
+draw_boxplots(model1, params = params) + labs(title = "Model 1")
+```
 
-draw_boxplots(model1, params = params) + labs(title = "Model 1")
+### Model 2
+
+```{r cache=FALSE}
 draw_boxplots(model2, params = params) + labs(title = "Model 2")
+```
+
+### Model 3
+
+```{r cache=FALSE}
 draw_boxplots(model3, params = params) + labs(title = "Model 3")
+```
+
+### Model 4
+
+```{r cache=FALSE}
 draw_boxplots(model4, params = params) + labs(title = "Model 4")
@@
 
-### Scatterplots
+##
+
+## Scatterplots {.tabset .tabset-fade}
+
+Scatterplots of posterior summaries (such as mean and standard deviation) where the $x$-axis is the posterior summary as estimated by `tmbstan` and the $y$-axis is the posterior summary as estimated by either `ahgq` or `TMB`.
+We treat `tmbstan` as the gold-standard, so any deviation from the line $y = x$ suggests inaccurate posterior summaries.
 
@@
 params <- c("beta_prev", "log_sigma_phi_prev")
+```
 
+### Model 1
+
+```{r}
 draw_scatterplots(model1, params = params) + labs(title = "Model 1")
+```
+
+### Model 2
+
+```{r}
 draw_scatterplots(model2, params = params) + labs(title = "Model 2")
+```
+
+### Model 3
+
+```{r}
 draw_scatterplots(model3, params = params) + labs(title = "Model 3")
+```
+
+### Model 4
+
+```{r}
 draw_scatterplots(model4, params = params) + labs(title = "Model 4")
@@
 
-### Kolmogorov KS test plots
+##
+
+## Kolmogorov KS test plots {.tabset .tabset-fade}
 
 Take samples from all distributions, then compute maximum ECDF difference $D$ (two-sample Kolmogorov–Smirnov test).
-We assume that `tmbstan` represents the gold-standard, and compare performance of `aghq` and `TMB` via their similarity to this gold-standard.
-Higher values of $D$ correspond to distributions which are more different.
+On the $y$-axis we plot $\text{KS}(\texttt{TMB}, \texttt{tmbstan})$ and on the $x$-axis $\text{KS}(\texttt{aghq}, \texttt{tmbstan})$.
+Lower values of $D$ (minimum possible zero) correspond to more similar distributions, and higher values of $D$ correspond to distributions which are more different.
+
+### Model 1
 
@@
 draw_ksplots_D(model1) + labs(title = "Model 1")
+```
+
+### Model 2
+
+```{r fig.height=8}
 draw_ksplots_D(model2) + labs(title = "Model 2")
+```
+
+### Model 3
+
+```{r fig.height=8}
 draw_ksplots_D(model3) + labs(title = "Model 3")
+```
+
+### Model 4
+
+```{r fig.height=8}
 draw_ksplots_D(model4) + labs(title = "Model 4")
 ```
+
+##
 
@@
 
-### MCMC diagnostics
+## MCMC diagnostics
 
@@
 
-#### Traceplots
+### Traceplots {.tabset .tabset-fade}
 
@@
 
+#### Model 1
+
 ```{r}
 map(model1, "mcmc_traceplots")
+```
+
+#### Model 2
+
+```{r}
 map(model2, "mcmc_traceplots")
+```
+
+#### Model 3
+
+```{r}
 map(model3, "mcmc_traceplots")
+```
+
+#### Model 4
+
+```{r}
 map(model4, "mcmc_traceplots")
@@
 
-#### R hat
+###
+
+### R hat {.tabset .tabset-fade}
 
@@
 
+#### Model 1
+
 ```{r}
 draw_rhatplot(model1) + labs(title = "Model 1")
+```
+
+#### Model 2
+
+```{r}
 draw_rhatplot(model2) + labs(title = "Model 2")
+```
+
+#### Model 3
+
+```{r}
 draw_rhatplot(model3) + labs(title = "Model 3")
+```
+
+#### Model 4
+
+```{r}
 draw_rhatplot(model4) + labs(title = "Model 4")
@@
 
-## Conclusion
+###
 
-## Bibliography {-}
+# Conclusion
+
+# Bibliography {-}
+
+# Original computing environment
+
+```{r}
+sessionInfo()
+```

### 2022-10-17 src/example_naomi/naomi.Rmd
Upload start on Naomi vignette

@@
 
-```{r}
+```{r eval=FALSE}
 naomi_mf <- naomi_model_frame(
@@
 
-```{r}
+```{r eval=FALSE}
 naomi_data <- select_naomi_data(
@@
 
-```{r}
+```{r eval=FALSE}
 fit <- fit_tmb(tmb_inputs)
@@
 
-```{r}
+```{r eval=FALSE}
 outputs <- output_package(fit, naomi_mf)
@@
 
-```{r}
+```{r eval=FALSE}
 outputs$indicators %>%
@@
 
-```{r}
+```{r eval=FALSE}
 add_output_labels(outputs) %>%
@@
 
-```{r}
+```{r eval=FALSE}
 system.time(fit <- sample_tmb(fit))
@@
 
-```{r}
+```{r eval=FALSE}
 system.time(outputs <- output_package(fit, naomi_mf))

### 2022-10-18 src/example_inla-replication/inla-replication.Rmd
Clean this up a bit

@@
   &\text{(Observations)}     &   y_i &\sim p(y_i \, | \, x_i, \theta), \quad i \in \mathcal{I}, \\
-  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x \, | \,\mathbf{0},  \mathbf{Q}(\theta)^{-1}), \\
+  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x \, | \, 0,  Q(\theta)^{-1}), \\
   &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta), \label{eq:hyper}
@@
 \begin{equation}
-\tilde p(\theta \, | \, y) \propto \frac{p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta)}{\tilde p_G(x \, | \, \theta, y)} \Big\rvert_{x = \mu^\star(\theta)} \equiv L^\star(\theta)
+\tilde p(\theta \, | \, y) \propto \frac{p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta)}{\tilde p_G(x \, | \, \theta, y)} \Big\rvert_{x = \mu(\theta)} \equiv L^\star(\theta)
 \end{equation}
@@
 
-<!-- *Note: Does `TMB` keep the $-\frac{n}{2} \log(2\pi)$ constant here $-\log L^\star_f(\theta) = -\frac{n}{2} \log(2\pi) + \frac{1}{2} \log \det (Q^\star(\theta)) + f(\mu^\star(\theta), \theta)$?* -->
+<!-- *Note: Does `TMB` keep the $-\frac{n}{2} \log(2\pi)$ constant here $-\log L^\star_f(\theta) = -\frac{n}{2} \log(2\pi) + \frac{1}{2} \log \det (Q^\star(\theta)) + f(\mu(\theta), \theta)$?* -->
 
@@
 We could also use an adaptive Gauss Hermite quadrature rule.
-See the [Implementing Approximate Bayesian Inference using
-Adaptive Quadrature: the `aghq` Package](https://arxiv.org/pdf/2101.04468.pdf) paper.
-## Stage 3: get $\tilde p(x_i \, | \, \theta, y)$
+See the [Implementing Approximate Bayesian Inference using Adaptive Quadrature: the `aghq` Package](https://arxiv.org/pdf/2101.04468.pdf) paper.
 
@@
 ### Simplified Laplace
+
+This is the key innovation in the 2009 INLA paper.
 
@@
 
-The distribution $p(x \, | \, \theta, y)$ is exact:
+By construction, the distribution $p(x \, | \, \theta, y)$ is exact:
 
@@
 \begin{equation}
-\tilde p(\theta \, | \, y) \propto \frac{p(y, \mu^\star(\theta), \theta)}{\det(Q(\theta))^{1/2}}.
+\tilde p(\theta \, | \, y) \propto \frac{p(y, \mu(\theta), \theta)}{\det(Q(\theta))^{1/2}}.
 \end{equation}
@@
 
-* $p(y, \mu^\star(\theta), \theta)$ corresponds to the calls to `dgamma`, `dnorm` and sum of `dnorm` evaluated at $\mu^\star(\theta)$ which here is simply `par$x_n` (for the Normal distribution the mean of the distribution is the same as the mode).
+* $p(y, \mu(\theta), \theta)$ corresponds to the calls to `dgamma`, `dnorm` and sum of `dnorm` evaluated at $\mu(\theta)$ which here is simply `par$x_n` (for the Normal distribution the mean of the distribution is the same as the mode).
 * The logarithm of $1 / \det(Q(\theta))^{1/2}$ when $\theta$ is a scalar is given by `-0.5 * log(par$tau_n)`.
@@
 ```{r}
-eval_grid <- function(grid = NULL, uniform = FALSE, K = NULL, min = NULL, max = NULL, f) {
+trapezoid_rule <- function(x, spacing) {
+  0.5 * spacing * (2 * sum(x) - x[1] - x[2])
+}
+
+eval_grid <- function(grid = NULL, spacing = NULL, uniform = FALSE, K = NULL, min = NULL, max = NULL, f) {
   if(uniform) {
     grid <- seq(min, max, length.out = K)
+    spacing <- (max - min) / K
   }
   df <- data.frame(input = grid, output = sapply(grid, f))
-  df <- mutate(df, norm_output = output / sum(output))
+  df <- mutate(df, norm_output = output / trapezoid_rule(output, spacing))
   return(df)
@@
 
-blangiardo_theta <- eval_grid(uniform = TRUE,
+blangiardo_theta <- eval_grid(
+  uniform = TRUE,
   K = 25, min = 0.001, max = 0.3,
-                              f = post_marginal_theta)
+  f = post_marginal_theta
+)
 
-dense_theta <- eval_grid(uniform = TRUE,
+dense_theta <- eval_grid(
+  uniform = TRUE,
   K = 500, min = 0.001, max = 0.3,
-                         f = post_marginal_theta)
+  f = post_marginal_theta
+)
 ```
@@
 
-inla_theta <- eval_grid(exp(points), f = post_marginal_theta)
+inla_theta <- eval_grid(exp(points), f = post_marginal_theta, spacing = delta_z)
 
@@
   geom_point(data = internals_inla_theta, aes(x = input, y = output), shape = 4) +
-  theme(axis.text.y=element_blank()) +
+  theme(axis.text.y = element_blank()) +
   labs(title = "INLA integration points for the hyperparameters") +
@@
   DLL = "blangiardo",
-  hessian = TRUE
+  hessian = TRUE,
+  silent = TRUE
 )
@@
 
-The mode and comparison (I started by making the mistake of exponentiating `fit$summary.hyperpar$mode` rather than the internal version--on the log scale, as with R and `TMB`--which, because the transformation is non-linear, is not the same).
+Comparison of the posterior mode and standard deviation of $\log(\theta)$:
 
 ```{r}
-# For some reason the mode in the INLA result below is NA, how to fix this?
-kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$`0.5quant`),
-                           exp(r_opt$par),
-                           exp(tmb_opt$par)))
+#' For some reason the mode in the INLA result below is NA, how to fix this?
+#' According to Havard, this is a feature not a bug, and he hasn't found a good way to compute the mode yet:
+#' https://groups.google.com/g/r-inla-discussion-group/c/mbSsNO_bydo
+
+kable_data <- data.frame(
+  "Mode" = c(NA, r_opt$par, tmb_opt$par),
+  "Variance" = c(fit$internal.summary.hyperpar$sd^2, sd_out$cov.fixed, NA)
+)
 
 rownames(kable_data) <- c("$\\texttt{R-INLA}$", "R", "$\\texttt{TMB}$")
-colnames(kable_data) <- c("Posterior mode")
 
 kableExtra::kable(kable_data, booktabs = TRUE, escape = FALSE, align = "c")
-```
-
-Compare the variances for $\log(\theta)$:
-
-```{r}
-sd_out$cov.fixed # TMB
-fit$internal.summary.hyperpar$sd^2 # INLA
 ```
@@
 
-### Is `h$fn` the same as in R?
+Could make a version of `h$fn` on the same scale as `post_marginal_theta` for a plot:
 
-Testing to see that `h$fn` is the same as `nl_post_marginal_theta` with `log_input = TRUE` up to a constant:
-
-```{r results=FALSE}
-vals <- seq(-5, 0, by = 1)
-
-tmb_vals <- sapply(vals, h$fn) # Replications show it not to be stochastic
-r_vals <- sapply(vals, nl_post_marginal_theta, log_input = TRUE)
-```
-
-```{r}
-rbind(tmb_vals, r_vals, tmb_vals - r_vals)
-```
-
-Could make a version of `h$fn` on the same scale as `post_marginal_theta` for a plot:
 ```{r results=FALSE}
@@
   geom_line(data = tmb_dense_theta, aes(x = input, y = norm_output), col = midpink) +
-  theme(axis.text.y=element_blank()) +
+  theme(axis.text.y = element_blank()) +
   annotate("text", x = 0.1, y = 0.01, label = "TMB", col = midpink) +
@@
 
-Does `TMB` do the inner optimisation differently?
+### Unit tests {.tabset .tabset-fade}
+
+#### That `h$fn` the same as in R
+
+Testing to see that `h$fn` is the same as `nl_post_marginal_theta` with `log_input = TRUE` up to a constant:
+
+```{r results=FALSE}
+vals <- seq(-5, 0, by = 1)
+
+tmb_vals <- sapply(vals, h$fn) # Replications show it not to be stochastic
+r_vals <- sapply(vals, nl_post_marginal_theta, log_input = TRUE)
+```
+
+```{r}
+rbind(tmb_vals, r_vals, tmb_vals - r_vals)
+```
+
+#### That the inner optimisation is the same
 
@@
 
-### Getting to the full conditional of $x$ in TMB
+###
 
-Probably going to need $Q^\star(\theta)$ ($n \times n$ precision matrix) and $\mu^\star(\theta)$ ($n \times 1$ mean vector).
-For example, the most basic version of the full conditional would be the Gaussian approximation
-\begin{equation}
-\tilde p(x_i \, | \, \theta, y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\theta), 1 / q^\star_i(\theta)).
-\end{equation}
+<!-- ### Getting to the full conditional of $x$ in TMB -->
 
-Precision matrix $Q^\star(\theta)$ is the Hessian of the negative logarithm of the posterior evaluated at $\mu^\star(\theta)$, having $(i, j)$th entry
-\begin{equation}
-Q^\star(\theta)_{i, j} = - \frac{\partial^{2} \log p(x \, | \, \theta, y)}{\partial x_{i} \partial x_{j}} \Big\rvert_{x = \mu^\star(\theta)}
-\end{equation}
+<!-- Probably going to need $Q^\star(\theta)$ ($n \times n$ precision matrix) and $\mu(\theta)$ ($n \times 1$ mean vector). -->
+<!-- For example, the most basic version of the full conditional would be the Gaussian approximation -->
+<!-- \begin{equation} -->
+<!-- \tilde p(x_i \, | \, \theta, y) = \mathcal{N}(x_i \, | \, \mu_i(\theta), 1 / q^\star_i(\theta)). -->
+<!-- \end{equation} -->
 
-An obvious choice for finding the precision would be `h$he(tmb_opt$par)` but this gives the error "Hessian not yet implemented for models with random effects".
-There may be another way to do it using `h$env$spHess()` (some kind of sparse Hessian) but if this is the case then it doesn't make sense to not have `h$he` just call the alternative way.
-This matrix is of the class `dsCMatrix` which is for symmetric, sparse numeric matrices in the compressed column-oriented format.
+<!-- Precision matrix $Q^\star(\theta)$ is the Hessian of the negative logarithm of the posterior evaluated at $\mu(\theta)$, having $(i, j)$th entry -->
+<!-- \begin{equation} -->
+<!-- Q^\star(\theta)_{i, j} = - \frac{\partial^{2} \log p(x \, | \, \theta, y)}{\partial x_{i} \partial x_{j}} \Big\rvert_{x = \mu(\theta)} -->
+<!-- \end{equation} -->
 
-`h$env$spHess` takes two arguments:
+<!-- An obvious choice for finding the precision would be `h$he(tmb_opt$par)` but this gives the error "Hessian not yet implemented for models with random effects". -->
+<!-- There may be another way to do it using `h$env$spHess()` (some kind of sparse Hessian) but if this is the case then it doesn't make sense to not have `h$he` just call the alternative way. -->
+<!-- This matrix is of the class `dsCMatrix` which is for symmetric, sparse numeric matrices in the compressed column-oriented format. -->
 
-  1. `par` (default `h$env$par`) should be both the hyperparameters and random effects
-  2. `random` (`TRUE` or `FALSE`)
+<!-- `h$env$spHess` takes two arguments: -->
 
-and the code is given by:
+<!--   1. `par` (default `h$env$par`) should be both the hyperparameters and random effects -->
+<!--   2. `random` (`TRUE` or `FALSE`) -->
 
-```{r eval=FALSE}
-function(par = obj$env$par, random = FALSE) {
-  if (!random) {
-    Hfull@x[] <- ev(par) # Whats ev?
-    Hfull # H full <=> Full Hessian?
-  }
-  else if (skipFixedEffects) {
-    .Call("setxslot", Hrandom, ev(par), PACKAGE = "TMB")
-  }
-  else {
-    Hfull@x[] <- ev(par)
-    Hfull[r, r]
-  }
-}
-```
+<!-- and the code is given by: -->
 
-Set a range of test parameters `test_pars`:
+<!-- ```{r eval=FALSE} -->
+<!-- function(par = obj$env$par, random = FALSE) { -->
+<!--   if (!random) { -->
+<!--     Hfull@x[] <- ev(par) # Whats ev? -->
+<!--     Hfull # H full <=> Full Hessian? -->
+<!--   } -->
+<!--   else if (skipFixedEffects) { -->
+<!--     .Call("setxslot", Hrandom, ev(par), PACKAGE = "TMB") -->
+<!--   } -->
+<!--   else { -->
+<!--     Hfull@x[] <- ev(par) -->
+<!--     Hfull[r, r] -->
+<!--   } -->
+<!-- } -->
+<!-- ``` -->
 
-```{r results=FALSE}
-full_pars <- function(input) {
-  setNames(input, c("x", "l_theta"))
-}
+<!-- Set a range of test parameters `test_pars`: -->
 
-par1 <- full_pars(c(0, 0))
-par2 <- full_pars(c(0, -1))
-par3 <- full_pars(c(-1, 0))
+<!-- ```{r results=FALSE} -->
+<!-- full_pars <- function(input) { -->
+<!--   setNames(input, c("x", "l_theta")) -->
+<!-- } -->
 
-# Doing the inner loop manually
-par4 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = 0), 0))
-par5 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = -1), -1))
+<!-- par1 <- full_pars(c(0, 0)) -->
+<!-- par2 <- full_pars(c(0, -1)) -->
+<!-- par3 <- full_pars(c(-1, 0)) -->
 
-test_pars <- list(par1, par2, par3, par4, par5)
-```
+<!-- # Doing the inner loop manually -->
+<!-- par4 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = 0), 0)) -->
+<!-- par5 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = -1), -1)) -->
 
-```{r}
-lapply(test_pars, h$env$spHess)
-```
+<!-- test_pars <- list(par1, par2, par3, par4, par5) -->
+<!-- ``` -->
 
-`random = TRUE` seems behave in unpredictable ways e.g.
-always returning the value calculated last when `random = FALSE` (perhaps only when using `apply` functions and not in other situations, so results here omitted):
+<!-- ```{r} -->
+<!-- lapply(test_pars, h$env$spHess) -->
+<!-- ``` -->
 
-```{r results=FALSE}
-lapply(test_pars, h$env$spHess, random = TRUE)
-```
+<!-- `random = TRUE` seems behave in unpredictable ways e.g.
+always returning the value calculated last when `random = FALSE` (perhaps only when using `apply` functions and not in other situations, so results here omitted): -->
 
-1) With `random = FALSE` the number changes as a function of the second argument, `l_theta` (the hyperparameters).
-The Hessian is $2 \times 2$ with a dot ("structural zero") outside entry $(1, 1)$.
-2) With `random = TRUE` the number doesn't change at all and it's now a $1 \times 1$ matrix.
+<!-- ```{r results=FALSE} -->
+<!-- lapply(test_pars, h$env$spHess, random = TRUE) -->
+<!-- ``` -->
 
-Utilising the objective function $f$ which simply calculates the negative log posterior as a function of both random effects and hyperparameters:
+<!-- 1) With `random = FALSE` the number changes as a function of the second argument, `l_theta` (the hyperparameters). -->
+<!-- The Hessian is $2 \times 2$ with a dot ("structural zero") outside entry $(1, 1)$. -->
+<!-- 2) With `random = TRUE` the number doesn't change at all and it's now a $1 \times 1$ matrix. -->
 
-```{r}
-lapply(test_pars, f$he)
-```
+<!-- Utilising the objective function $f$ which simply calculates the negative log posterior as a function of both random effects and hyperparameters: -->
 
-This produces the same answers as numerical differentiation of the same function (results omitted for space):
+<!-- ```{r} -->
+<!-- lapply(test_pars, f$he) -->
+<!-- ``` -->
 
-```{r results=FALSE}
-lapply(test_pars, optimHess, fn = f$fn, gn = f$gn)
-```
+<!-- This produces the same answers as numerical differentiation of the same function (results omitted for space): -->
 
-In this situation we know what the precision of $x$'s full conditional is (and its a function of $\theta$!):
+<!-- ```{r results=FALSE} -->
+<!-- lapply(test_pars, optimHess, fn = f$fn, gn = f$gn) -->
+<!-- ``` -->
 
-```{r}
-sapply(c(exp(0), exp(-1)), function(theta) inner_loop(theta)$tau_n)
-```
+<!-- In this situation we know what the precision of $x$'s full conditional is (and its a function of $\theta$!): -->
 
-Could look more at `sdreport` [here](https://github.com/kaskr/adcomp/blob/master/TMB/R/sdreport.R) see how these functions are used there (there are more comments on the Github page than when just looking via R).
+<!-- ```{r} -->
+<!-- sapply(c(exp(0), exp(-1)), function(theta) inner_loop(theta)$tau_n) -->
+<!-- ``` -->
 
-```{r}
-nl_full_conditional_x <- function(x, theta, log_input = FALSE) {
+<!-- Could look more at `sdreport` [here](https://github.com/kaskr/adcomp/blob/master/TMB/R/sdreport.R) see how these functions are used there (there are more comments on the Github page than when just looking via R).
+-->
 
-  if(log_input == TRUE) {
-    theta <- exp(theta) # Convert from log(theta) to theta
-  }
+<!-- ```{r} -->
+<!-- nl_full_conditional_x <- function(x, theta, log_input = FALSE) { -->
 
-  par <- inner_loop(theta)
+<!--   if(log_input == TRUE) { -->
+<!--     theta <- exp(theta) # Convert from log(theta) to theta -->
+<!--   } -->
 
-  return(-dnorm(x, par$x_n, sqrt(1 / par$tau_n), log = TRUE))
-}
-```
+<!--   par <- inner_loop(theta) -->
+
+<!--   return(-dnorm(x, par$x_n, sqrt(1 / par$tau_n), log = TRUE)) -->
+<!-- } -->
+<!-- ``` -->
 

### 2022-10-18 src/dev_sinla/sinla.Rmd
Playing with the normalising constant method

@@
 
-Now we will compute Equation \@ref(eq:eq:pxgiveny) taking the marignals of a joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$.
+Now we will compute Equation \@ref(eq:pxgiveny) taking the marignals of a joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$.
 This is the approach currently used in the `aghq` package (done implicitly, via sampling).
@@
 dim(Q) #' With dimension [30 x 30]
-image(Q)
+
+cowplot::plot_grid(plot(mean), image(Q))
 ```
@@
 plot2
+```
+
+Take some samples from the posterior marginal:
+
+```{r}
+M <- 1000
+samples_jointgaussianmarginal <- rnorm(M, mean_i, sqrt(var_i)) 
 ```
@@
 obj <- objs[[i]]
+
 mean <- means[[i]] #' The length is 30 - 1
-mean
+length(mean)
 
@@
 dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
-image(Q)
+
+cowplot::plot_grid(plot(mean), image(Q))
 ```
@@
 \begin{align}
-\tilde p(x_i \, | \, y) &= \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
+\tilde p(x_i, \hat \theta \, | \, y) &= \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
 &\propto \tilde p_\text{LA}(x_i \, | \, \hat \theta, y)
@@
 
-What we've got here is the unnormalised log-posterior?
+What we've got here is the unnormalised log-posterior.
 How can we go from that to a normalised posterior?
 Going to do trapezoid rule here, but AGHQ would work better (even with a low number of points) because the function is close to being Gaussian.
+
+### Method 1: Safe exponential then trapezoid rule
+
+The safe exponential function computes $\exp(x - \text{max}(x))$.
+
+```{r}
+safe_exp <- function(x) exp(x - max(x))
+```
+
+A simple way to write the trapezoid rule is as follows:
+
+```{r eval=FALSE}
+trapezoid_rule <- function(x, spacing) {
+  0.5 * spacing * (2 * sum(x) - x[1] - x[2])
+}
+```
+
+It can also be defined like this, which is more convenient for thinking about it like other quadrature rules based on weights:
+
+```{r}
+trapezoid_rule <- function(x, spacing) {
+  w <- rep(spacing, length(x))
+  w[1] <- w[1] / 2
+  w[length(x)] <- w[length(x)] / 2
+  sum(w * x)
+}
+```
+
+```{r}
+df$post <- safe_exp(df$logpost)
+norm_trapezoid <- trapezoid_rule(df$post, spacing = spacing)
+df$post_norm <- df$post / norm_trapezoid
+```
+
+### Method 2: logSumExp to calculate the log normalising constant
+
+What about with another method?
+We can calculate the logarithm of the normalising constant directly, then just subtract it from the log-posterior:
 
@@
 
+The `matrixStats::logSumExp` function computes $\text{LSE}(x) = \log(\exp(x_1) + \cdots + \exp(x_n))$ in a computationally stable way.
+Another way to compute the trapezoid rule, taking input on the log-scale, is as follows:
+
 ```{r}
-safe_exp <- function(x) exp(x - max(x))
-
-trapezoid_rule <- function(x, spacing) {
-  0.5 * spacing * (2 * sum(x) - x[1] - x[2])
+trapezoid_rule_log <- function(x, spacing) {
+  w <- rep(spacing, length(x))
+  w[1] <- w[1] / 2
+  w[length(x)] <- w[length(x)] / 2
+  matrixStats::logSumExp(log(w) + x)
 }
 
-df$post <- safe_exp(df$logpost)
-df$post_norm = df$post / trapezoid_rule(df$post, spacing)
+norm_trapezoid_log <- trapezoid_rule_log(df$logpost, spacing)
+df$post_norm_alt <- exp(df$logpost - norm_trapezoid_log)
 
 plot3 <- plot2 +
-  geom_line(data = df, aes(x = phi_prev_i, y = post, col = "Laplace"), size = 1)
+  geom_line(data = df, aes(x = phi_prev_i, y = post_norm, col = "Laplace"), size = 1)
 
@@
 
+The two ways of calculating the normalised posterior are the same up to numerical error:
+
+```{r}
+max(df$post_norm - df$post_norm_alt)
+```
+
+## Comparison
+
+```{r}
+samples_tmbstan <- extract(mcmc, pars = "phi_prev[1]")[[1]]
+ks_jointgaussianmarginal_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_jointgaussianmarginal)
+```
+
+...
 So perhaps it's barely better than the Gaussian, but not really getting the real problem here.
@@
 | Description | Here | Wood (2019) notation | Notes |
-|-------------|---------------|----------------------|----------|
+|-------------|:--------------:|:--------------------:|:---------|
 | Random effects | $x$ | $\beta$ | Stringer (2021) use $w$
@@
 ```
+
+# Original computing environment
+
+```{r}
+sessionInfo()
+```

### 2022-10-19 src/docs_xx-12-22-explainer/xx-12-22-explainer.Rmd
Template for lab meeting in December

---
title: Explaining approximate deterministic inference methods
subtitle: HIV Inference Lab Group Meeting
author: Adam Howes
institute: Imperial College London
date: December 2022
bibliography: citations.bib
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

# Bayes

* As a statistical modeller, our whole job (approximately) is to construct a generative model for data $y$ using parameters $\vartheta$
* This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
* What we actually want is the posterior $p(\vartheta \, | \, y)$ which is **just**
$$
p(\vartheta \, | \, y) = \frac{p(y, \vartheta)}{p(y)} = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)}
$$
* The **core problem** of Bayesian inference is that we can't compute $p(y)$
$$
p(y) = \int p(y, \vartheta) \text{d} \vartheta
$$

# How might you do it?

* Usually if you wanted to integrate something, you could use numerical integration methods
* It doesn't work when $\text{dim}(\vartheta)$ gets even moderately sized

# Notation set-up

* Previously I had all of the parameters under the symbol $\vartheta$
* What if we split them up as being $\vartheta = (x, \theta)$

| Names for $x$ | Names for $\theta$ |
|---------------|--------------------|
| Latent variables, random effects, latent field | Hyperparameters, fixed effects

# Spatio-temporal statistics

* There is nothing inherently special about spatio-temporal statistics
* We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$
* Usually we associate parameters to spatio-temporal locations as well as observations
* This ends up with us having something like $\{x_{s, t}\}$

# What's important about this?

1. There might be **a lot** of spatio-temporal locations, so $\text{dim}(x)$ might be pretty big!
 * If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
2. Perhaps we're willing to make assumptions about how things vary over space-time

<!-- As you get more data, the posterior approaches a Gaussian -->

# Latent Gaussian models

\begin{alignat*}{2}
  &\text{(Observations)}     &   y &\sim p(y \, | \, x, \theta), \\
  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x  \, | \, \mu(\theta), Q(\theta)^{-1}), \\
  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta),
\end{alignat*}
where $y = (y_1, \ldots, y_n)$, $x = (x_1, \ldots, x_n)$, $\theta = (\theta_1, \ldots, \theta_m)$

# References {.allowframebreaks}

### 2022-10-19 src/dev_sinla/sinla.Rmd
Add some tabs for checks and multiple ways of doing things

@@
 
-cowplot::plot_grid(plot(mean), image(Q))
+cowplot::plot_grid(plot(mean), image(Q), nrow = 1)
 ```
 
-## Reproducing output from `h`
+## Additional checks {.tabset .tabset-fade}
 
-As a check, we would like to reproduce `h` using the right hand side of Equation \@ref(eq:pthetagiveny).
+### Description
+
+These tabs contain additional sanity checks used to build understanding.
+Feel free to skip them.
+
+### Reproducing output from `h`
+
+We would like to reproduce `h` using the right hand side of Equation \@ref(eq:pthetagiveny).
 To do this, we start by creating an objective function `f` (see Appendix for the `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!).
@@
 ```
+
+###
 
@@
 
-## $\tilde p(x_i \, | \, y)$
+## $\tilde p(x_i \, | \, y)$ {.tabset .tabset-fade}
 
@@
 
-### Method 1: Safe exponential then trapezoid rule
+### Safe exponential then trapezoid rule
 
@@
 
-### Method 2: logSumExp to calculate the log normalising constant
+### `logSumExp` to calculate the log normalising constant
 
@@
 df$post_norm_alt <- exp(df$logpost - norm_trapezoid_log)
+```
 
+###
+
+The two ways of calculating the normalised posterior above are the same up to numerical error:
+
+```{r}
+max(abs(df$post_norm - df$post_norm_alt))
+```
+
+We now add our Laplace approximation to the existing joint Gaussian marginal and MCMC results:
+
+```{r}
 plot3 <- plot2 +
@@
 plot3
-```
-
-The two ways of calculating the normalised posterior are the same up to numerical error:
-
-```{r}
-max(df$post_norm - df$post_norm_alt)
 ```
@@
 
-...
-So perhaps it's barely better than the Gaussian, but not really getting the real problem here.
-Hypothesis would be that it's because we're doing empirical Bayes, and with more than one hyperparameter grid point we would do better.
+One thing to take into account is that we're doing empirical Bayes, and things might be improved by using more integration points.
 

### 2022-10-19 src/dev_sinla/sinla.Rmd
Aesthetic improvements

@@
 Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
-Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.
+Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.
 
@@
 `aghq::marginal_laplace_tmb` calculates an object representing $\tilde p(x \, | \, \theta, y)$.
-`aghq::sample_marginal` draws samples from it.
+`aghq::sample_marginal` draws samples from this Gaussian, given values of $z$ drawn from a multinomial distribution, in order to evaluate Equation \@ref(eq:pxigiveny).
 For a line-by-line walkthrough of how these functions work, see the notebook [Code walkthrough](https://athowes.github.io/elgm-inf/walkthrough.html) (which uses the model from the notebook [Epilepsy example](https://athowes.github.io/elgm-inf/epil.html)).
 
-The goal of this notebook is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
+The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/elgm-inf/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
 $$
@@
 $$
-The reasons for doing this is that the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
+We want to do this because the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
 At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
@@
 
+We will use Model 1 from the [Prevalence, ANC, ART model inference case-study](https://athowes.github.io/elgm-inf/prev-anc-art.html) as a running example.
 To provide a ground truth to compare the methods that follow to, we generate "gold-standard" inferences using a relatively long run of MCMC as implemented by `tmbstan`.
@@
 We begin by loading in testing data, as generated by [`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim).
-For more information about this model, see the [Prevalence, ANC, ART model inference case-study](https://athowes.github.io/elgm-inf/prev-anc-art.html).
 `sim_data` contains many replicates of simulated data.
@@
 
+Compile and load the template (see Appendix \@ref(appendix) for the `TMB` code):
+
 ```{r}
@@
 
-We create an objective function `h` (see Appendix for `TMB` template) integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`).
+We create an objective function `h` integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`).
 The values of `param` intitialise the objective function:
@@
 
-### Description
+### Description {-}
 
@@
 
-### Reproducing output from `h`
+### Reproducing output from `h` {-}
 
 We would like to reproduce `h` using the right hand side of Equation \@ref(eq:pthetagiveny).
-To do this, we start by creating an objective function `f` (see Appendix for the `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!).
+To do this, we start by creating an objective function `f` (see Appendix \@ref(appendix) for the `TMB` code) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!).
 This is what we will be using for the numerator of Equation \@ref(eq:pthetagiveny):
@@
 ```
-
-###
 
@@
 We will still go through the steps for implementing it here though for demonstration.
-With of a bit of effort, it can be implemented in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are explicitly named (see Appendix for the `TMB` template `model1index.cpp`):
+With of a bit of effort, it can be implemented in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are explicitly named (see Appendix \@ref(appendix) for the `TMB` code):
 
@@
 
-### Safe exponential then trapezoid rule
+### Safe exponential then trapezoid rule {-}
 
@@
 
-### `logSumExp` to calculate the log normalising constant
+### `logSumExp` to calculate the log normalising constant {-}
 
@@
 
-###
+## Comparison
 
@@
 
-```{r}
+```{r message=FALSE, warning=FALSE}
 plot3 <- plot2 +
@@
 ```
-
-## Comparison
 
@@
 
-# Appendix
+# Appendix {#appendix}
 

### 2022-10-19 src/dev_sinla/sinla.Rmd
Add CDFs, KS tests, and checks on trapezoid rules working

@@
 ```{r message=FALSE, warning=FALSE}
-plot <- extract(mcmc, pars = "phi_prev[1]") %>%
-  as.data.frame() %>%
-  ggplot(aes(x = phi_prev.1.)) +
+samples_tmbstan <- extract(mcmc, pars = "phi_prev[1]")[[1]]
+
+plot_pdf <- ggplot(data.frame(x = samples_tmbstan), aes(x = x)) +
     geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
     theme_minimal() +
-    labs(x = "phi_prev[1]", y = "Posterior")
+    labs(x = "phi_prev[1]", y = "Posterior PDF")
 
-plot
+ecdf_tmbstan <- ecdf(samples_tmbstan)
+grid <- seq(min(samples_tmbstan), max(samples_tmbstan), by = 0.1)
+
+plot_cdf <- data.frame(x = grid, y = ecdf_tmbstan(grid)) %>%
+  ggplot(aes(x = x, y = y)) +
+    geom_line(col = cbpalette[7]) +
+    theme_minimal() +
+    labs(x = "phi_prev[1]", y = "Posterior CDF")
+
+cowplot::plot_grid(plot_cdf, plot_pdf)
 ```
@@
 dim(Q) #' With dimension [30 x 30]
-
-cowplot::plot_grid(plot(mean), image(Q), nrow = 1)
 ```
@@
 $$
+
 We can evaluate this for $i = 1$ via:
@@
 
-And compare it to our MCMC results as follows:
+Because this is a Gaussian distribution, we do not have do normalise it (as would usually be the case) because we know it's a density.
+We can compare it to our MCMC results as follows:
 
 ```{r message=FALSE, warning=FALSE}
-plot2 <- plot +
+plot_pdf2 <- plot_pdf +
   stat_function(
@@
 
-plot2
+plot_pdf2
+
+M <- 1000
+samples_jointgaussianmarginal <- rnorm(M, mean_i, sqrt(var_i))
+
+ecdf_jointgaussianmarginal <- ecdf(samples_jointgaussianmarginal)
+grid <- seq(min(samples_jointgaussianmarginal), max(samples_jointgaussianmarginal), by = 0.1)
+
+plot_cdf2 <- plot_cdf +
+  geom_line(
+    data = data.frame(x = grid, y = ecdf_jointgaussianmarginal(grid)),
+    aes(x = x, y = y), col = cbpalette[1]
+  )
+
+cowplot::plot_grid(plot_cdf2, plot_pdf2, rel_widths = c(1, 1.5))
 ```
 
-Take some samples from the posterior marginal:
+```{r}
+ks_jointgaussianmarginal_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_jointgaussianmarginal)
+ks_jointgaussianmarginal_tmbstan
+```
 
-```{r}
-M <- 1000
-samples_jointgaussianmarginal <- rnorm(M, mean_i, sqrt(var_i)) 
-```
+The maximum difference between the two ECDFS is `r ks_jointgaussianmarginal_tmbstan$D` which occurs at the point `r ks_jointgaussianmarginal_tmbstan$l`.
 
@@
 
-for(i in 1:dat$n) {
-  dat_i <- prepare_dat(dat, i)
-
-  #' Starting parameters for TMB are the same for every index of the loop
+#' Starting parameters for TMB are the same for every index of the loop so it's more efficient to put them outside
 param_i <- list(
@@
 )
+
+its <- 1000
+
+for(i in 1:dat$n) {
+  dat_i <- prepare_dat(dat, i)
 
@@
   )
-
-  its <- 1000
 
@@
 dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
-
-cowplot::plot_grid(plot(mean), image(Q))
 ```
 
-`obj$fn` should take as input $(x_i, \theta)$:
+We check here that `obj$fn` indeed takes as input $(x_i, \theta)$:
 
@@
   )
-
-ggplot(df, aes(x = phi_prev_i, y = logpost)) +
-  geom_line() +
-  theme_minimal() +
-  labs(x = "phi_prev[1]", "log(posterior)")
 ```
 
-What we've got here is the unnormalised log-posterior.
-How can we go from that to a normalised posterior?
+We've compuating the unnormalised log-posterior -- how can we go from that to a normalised posterior?
 Going to do trapezoid rule here, but AGHQ would work better (even with a low number of points) because the function is close to being Gaussian.
@@
 
+Check that this method correctly computes the normalising constant for a density (which is known to integrate to one):
+
+```{r}
+abs(trapezoid_rule(dnorm(seq(-10, 10, by = 0.1)), spacing = 0.1) - 1) < 10E-6
+```
+
 ### `logSumExp` to calculate the log normalising constant {-}
@@
 
+Check that this method correctly computes the normalising constant for a density (which is known to integrate to one):
+
+```{r}
+abs(trapezoid_rule_log(dnorm(seq(-10, 10, by = 0.1), log = TRUE), spacing = 0.1) - log(1)) < 10E-6
+```
+
 ## Comparison
@@
 ```{r}
-max(abs(df$post_norm - df$post_norm_alt))
+max(abs(df$post_norm - df$post_norm_alt)) < 10E-6
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-plot3 <- plot2 +
+plot_pdf3 <- plot_pdf2 +
   geom_line(data = df, aes(x = phi_prev_i, y = post_norm, col = "Laplace"), size = 1)
 
-plot3
+plot_pdf3
+
+df$ecdf <- cumsum(df$post_norm) * c(0, diff(df$phi_prev_i))
+
+plot_cdf3 <- plot_cdf2 +
+  geom_line(
+    data = df, aes(x = phi_prev_i, y = ecdf), col = cbpalette[2]
+  )
+
+plot_cdf3
+
+cowplot::plot_grid(plot_cdf3, plot_pdf3, rel_widths = c(1, 1.5))
 ```
 
+We use the inverse CDF method to sample from the Laplace marginal, then we can calculate the KS statistic:
+
 ```{r}
-samples_tmbstan <- extract(mcmc, pars = "phi_prev[1]")[[1]]
-ks_jointgaussianmarginal_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_jointgaussianmarginal)
+samples_laplace <- approxfun(df$ecdf, df$phi_prev_i)(runif(M))
+
+ks_laplace_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_laplace)
+
+kable_data <- data.frame(
+  "KS Distance" = c(ks_jointgaussianmarginal_tmbstan$D, ks_laplace_tmbstan$D),
+  "KS Location" = c(ks_jointgaussianmarginal_tmbstan$l, ks_laplace_tmbstan$l)
+)
+
+rownames(kable_data) <- c("$\\text{KS}(\\texttt{tmbstan}, \\texttt{jointgaussianmarginal})$", "$\\text{KS}(\\texttt{tmbstan}, \\texttt{laplace})$")
+
+kableExtra::kable(kable_data, booktabs = TRUE, escape = FALSE, align = "c")
 ```
-
-One thing to take into account is that we're doing empirical Bayes, and things might be improved by using more integration points.
 

### 2022-10-19 src/dev_sinla/sinla.Rmd
Remove excess plots and move notation to appendix

@@
     data = data.frame(x = c(0.5, 3.5)), aes(x, col = "Joint Gaussian marginal"), fun = dnorm, n = 101, 
-    args = list(mean = mean_i, sd = sqrt(var_i)), size = 1
+    args = list(mean = mean_i, sd = sqrt(var_i))
   ) +
@@
   theme_minimal()
-
-plot_pdf2
 
@@
 plot_pdf3 <- plot_pdf2 +
-  geom_line(data = df, aes(x = phi_prev_i, y = post_norm, col = "Laplace"), size = 1)
-
-plot_pdf3
+  geom_line(data = df, aes(x = phi_prev_i, y = post_norm, col = "Laplace"))
 
@@
   )
-
-plot_cdf3
 
@@
 kable_data <- data.frame(
-  "KS Distance" = c(ks_jointgaussianmarginal_tmbstan$D, ks_laplace_tmbstan$D),
-  "KS Location" = c(ks_jointgaussianmarginal_tmbstan$l, ks_laplace_tmbstan$l)
+  "Distance" = c(ks_jointgaussianmarginal_tmbstan$D, ks_laplace_tmbstan$D),
+  "Location" = c(ks_jointgaussianmarginal_tmbstan$l, ks_laplace_tmbstan$l)
 )
@@
 # Simplified INLA
-
-| Description | Here | Wood (2019) notation | Notes |
-|-------------|:--------------:|:--------------------:|:---------|
-| Random effects | $x$ | $\beta$ | Stringer (2021) use $w$
-| Random effects which maximise (the / some) objective for given $\theta$ | $\mu (\theta)$ | $\hat \beta$ | Wood omits dependence on $\theta$ |
-| Random effects which maximise (the / some) objective for given $\theta$, subject to $x_i$ being fixed | $(x_i, \mu(x_i, \theta))$ | $\tilde \beta$ | We have $\mu(x_i, \theta))$ having dimension $n - 1$ |
-| Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$ | $Q(\theta)$ | $H$ | Unsure if it's preferable to use Hessian or precision notation here |
 
@@
 
+## Comparison of notation
+
+| Description | Here | Wood (2019) notation | Notes |
+|-------------|:--------------:|:--------------------:|:---------|
+| Random effects | $x$ | $\beta$ | Stringer (2021) use $w$
+| Random effects which maximise (the / some) objective for given $\theta$ | $\mu (\theta)$ | $\hat \beta$ | Wood omits dependence on $\theta$ |
+| Random effects which maximise (the / some) objective for given $\theta$, subject to $x_i$ being fixed | $(x_i, \mu(x_i, \theta))$ | $\tilde \beta$ | We have $\mu(x_i, \theta))$ having dimension $n - 1$ |
+| Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$ | $Q(\theta)$ | $H$ | Unsure if it's preferable to use Hessian or precision notation here |
+
 ## `model1.cpp`

### 2022-10-19 src/dev_sinla/sinla.Rmd
Start trying to normalise p_LA(theta | y); and add conditional mean of multivariate Gaussian

@@
 
+### Calculating $p_\text{LA}(\hat \theta \, | \, y)$ with `aghq`
+
+```{r}
+# h$he <- function(theta) numDeriv::jacobian(h$gr, theta, method = "Richardson")
+# quad <- aghq::aghq(ff = h, k = 1, startingvalue = h$par)
+```
+
 ## Comparison
@@
 In Section \@ref(full-laplace) we directly calculate the Laplace approximation to $\int p(x, y, \theta) \text{d} \theta_{-i}$.
-Following Wood (2019), a cheaper approach would be to "base the denominator of Equation \@ref(eq:pxigiventhetay) on the conditional density implied by ? in which case the Hessian is constant and
+Suppose that
 $$
-\mu(x_i, \theta) = \mu(\theta) + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:wood3)
+x \sim \mathcal{N}(\mu(\theta), Q(\theta)^{-1})
 $$
-The proposed modification has two parts:
-
-1. Use the numerically exact $\mu(x_i, \theta)$ rather than an approximation (as in simplified Laplace)
-2. An approximation to the required Hessian
+then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mean
+$$
+\mu(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:conditionalmean)
+$$
+and precision matrix
+$$
+Q(\theta)_{-i, -i}
+$$
 
@@
 2. For each $x_i$ in a grid of evaluation points (how does one chose this grid?) repeat the following steps a-e)
-  a) Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:wood3)
+  a) Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean)
   b) Compute a set of $J$ steps $\{\Delta_j\}$
@@
 | Description | Here          | Wood (2019) notation | Notes    |
-|-------------|:--------------:|:--------------------:|:---------|
+|:------------|:--------------|:---------------------|:---------|
 | Random effects | $x$        | $\beta$              | Stringer (2021) use $w$

### 2022-10-20 src/dev_sinla/sinla.Rmd
Debug aghq::aghq (see aghq/issues/6)

@@
 ```{r}
-# h$he <- function(theta) numDeriv::jacobian(h$gr, theta, method = "Richardson")
-# quad <- aghq::aghq(ff = h, k = 1, startingvalue = h$par)
+h$he <- function(theta) numDeriv::jacobian(h$gr, theta, method = "Richardson")
+quad <- aghq::aghq(ff = h, k = 3, startingvalue = c(0, 0), control = aghq::default_control_tmb())
+quad$normalized_posterior$lognormconst
 ```
@@
 $$
-Q(\theta)_{-i, -i}
+Q(\theta)_{-i, -i} (\#eq:conditionalprecision)
 $$

### 2022-10-21 src/dev_sinla/sinla.Rmd
Normalise the Laplace with aghq and map argument to MakeADFun

@@
 Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
-We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
+We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to^[Note that I'm uncertain if $\hat \theta$ (a constant) should still be included in the LHS of this.]:
 $$
 \begin{align}
-\tilde p(x_i, \hat \theta \, | \, y) &= \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
+\tilde p(x_i \, | \, y) &= \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
 &\propto \tilde p_\text{LA}(x_i \, | \, \hat \theta, y)
@@
 
-Start by getting $\hat \theta$:
+We're always going to need $\hat \theta$.
+We have calculated it before, but we will get it fresh here (because of questionable programming practices whereby objects are overwritten):
 
@@
 
-#' How many equally spaced points do we want to evaluate the posterior at?
+opt$par
+```
+
+Now, a basic way we could approach this is to define a grid of points that we would like to evaluate the posterior at.
+We can then create a dataframe where each row contains the values `opt$par` for $\theta$ and a grid value for $x_i$.
+Each row can then be passed into `-obj$fn()` to calculate the log-posterior.
+Note that it's crucial the arguments are in the right order when we pass them to `obj$fn()` -- this has caused headaches!
+
+```{r}
 spacing <- 0.05
@@
 
-df <- opt$par %>%
-  t() %>%
+df <- t(opt$par) %>%
   as.data.frame() %>%
   slice(rep(1:n(), each = x_grid_length)) %>%
-  mutate(
-    phi_prev_i = x_grid
-  ) %>%
+  mutate(phi_prev_i = x_grid) %>%
   rowwise() %>%
-  mutate(
-    #' Note here: it's crucial that the arguments are in the right order!
-    logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev))
-  )
+  mutate(logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev)))
 ```
 
-We've compuating the unnormalised log-posterior -- how can we go from that to a normalised posterior?
-Going to do trapezoid rule here, but AGHQ would work better (even with a low number of points) because the function is close to being Gaussian.
+We've computed the unnormalised log-posterior -- how can we go from that to a normalised posterior?
+Given these grid of points, we can do the trapezoid rule here.
 
-### Safe exponential then trapezoid rule {-}
+### Safe exponential trapezoid rule {-}
 
@@
 
-### `logSumExp` to calculate the log normalising constant {-}
+### `logSumExp` trapezoid rule {-}
 
@@
 
-### Calculating $p_\text{LA}(\hat \theta \, | \, y)$ with `aghq`
-
-```{r}
-h$he <- function(theta) numDeriv::jacobian(h$gr, theta, method = "Richardson")
-quad <- aghq::aghq(ff = h, k = 3, startingvalue = c(0, 0), control = aghq::default_control_tmb())
-quad$normalized_posterior$lognormconst
-```
-
-## Comparison
+## Normalisation with `aghq`
 
@@
 ```
+
+So how might we approach this in a more sophisticated way?
+We could do the normalisation directly in `aghq`!
+
+We could normalise $\tilde p_\text{LA}(\theta, y)$.
+Note that sometimes we have been calling this $\tilde p_\text{LA}(\theta \, | \, y)$ because they are equivalent up to a constant of proportionality.
+Define
+$$
+\tilde p_\text{AQ}(\theta \, | \, y) = \frac{\tilde p_\text{LA}(\theta, y)}{\tilde p_\text{AQ}(y)}
+$$
+or on a log-scale
+$$
+\log \tilde p_\text{AQ}(\theta \, | \, y) = \log \tilde p_\text{LA}(\theta, y) - \log \tilde p_\text{AQ}(y)
+$$
+
+First we have to add a numerical Hessian that's missing from `h`:
+
+```{r}
+h$he <- function(theta) numDeriv::jacobian(h$gr, theta, method = "Richardson")
+```
+
+Now we can obtain $\log \tilde p_\text{AQ}(y)$:
+
+```{r}
+quad <- aghq::aghq(ff = h, k = 3, startingvalue = c(0, 0), control = aghq::default_control_tmb())
+quad$normalized_posterior$lognormconst
+```
+
+Remember that `h` is the negative log-likelihood, so we want to add the normalising constant rather than subtract it:
+
+```{r}
+h_norm <- function(theta) h$fn(theta) - quad$normalized_posterior$lognormconst
+```
+
+Now we can put this into
+$$
+\tilde p(x_i, \hat \theta \, | \, y) = \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{AQ}(\hat \theta \, | \, y)
+$$
+
+But perhaps there is no reason to, because we will just have to normalise it again, so why do it twice?
+Let's just try to normalise it directly.
+What we need to do is create a function in `TMB` where we fix some of the inputs to the function.
+This can be done using the `map` argument of `MakeADFun`.
+You set each parameter to `factor(NA)` that you want to be fixed, then they will retain the value you input in `parameters`:
+
+```{r}
+obj$par %>% names()
+
+map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
+
+param_i_fixed_theta <- param_i
+
+theta_names <- names(opt$par)
+for(theta in theta_names) {
+  param_i_fixed_theta[[theta]] <- opt$par[[theta]]
+}
+
+dat_i <- prepare_dat(dat, i)
+
+obj_fixed_theta <- MakeADFun(
+  data = dat_i,
+  parameters = param_i_fixed_theta,
+  random = "phi_prev_minus_i",
+  DLL = "model1index",
+  silent = TRUE,
+  map = map_fixed_theta
+)
+```
+
+Now we can normalise this function $p_\text{LA}(x_i, \theta = \hat \theta, y)$ using `aghq`:
+
+```{r}
+quad <- aghq::aghq(ff = obj_fixed_theta, k = 3, startingvalue = 0, control = aghq::default_control_tmb())
+```
+
+The value of the normalising constant is (approximately) the same as we get using our trapezoid rule:
+
+```{r}
+abs(quad$normalized_posterior$lognormconst - norm_trapezoid_log) < 10E-3
+```
+
+## Comparison
 

### 2022-10-21 src/dev_sinla/sinla.Rmd
Working through Wood algorithm

@@
 
-## Complete algorithm
-
+We follow Wood (2019) Section 2.
 In this section we omit dependence on $\theta$ such that: $\mu = \mu(\theta)$, $Q = Q(\theta)$, $\mu(x_i) = \mu(x_i, \theta)$ and $Q(x_i) = Q(x_i, \theta)$.
 
-Want: $\tilde p(x_i \, | \, \theta, y)$
+Let $\mathcal{D}_{2j}$ denote a $2j \times 2j$ diagonal matrix with leading diagonal $-1, 1, -1, 1, \ldots$.
+For example, with $j = 3$ we have:
 
-Let $\mathcal{D}_{2j}$ denote a $2j \times 2j$ diagonal matrix with leading diagonal $-1, 1, -1, 1, \ldots$.
+```{r}
+D2j <- function(j) {
+  diag((-1)^{1:2 * j}, nrow = 2 * j)
+}
+
+D2j(3)
+```
+
 Let $u_0$ be a zero column matrix.
+I am guessing that this means $u_j$ is a $j$ column matrix.
+It's not specified what dimension it is too, but since it's multiplied by $\mathcal{D}_{2j}$ presumably it must be of length $2j$.
+
+```{r}
+uj <- function(j) {
+ matrix(j, nrow = 2 * j, ncol = 1)
+}
+
+uj(3)
+```
+
+```{r}
+D2j(3) %*% uj(3)
+```
+
 Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$.
+Note that it hasn't been defined what $\Delta_j$ is yet.
 
-1. Compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$
-2. For each $x_i$ in a grid of evaluation points (how does one chose this grid?) repeat the following steps a-e)
-  a) Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean)
-  b) Compute a set of $J$ steps $\{\Delta_j\}$
-  c) For $j = 1, \ldots, J - 1$ compute
+## Step 1: Compute the Cholesky factor
+
+The first task is to "compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$".
+
+```{r}
+i <- 1
+
+mm <- h$env$last.par
+mean <- mm[h$env$random]
+Q <- h$env$spHess(mm, random = TRUE)
+Q0 <- Q[-i, -i]
+
+L0 <- chol(Q0)
+```
+
+We have that $Q_0 = L_0 {L_0}^\top$:
+
+```{r}
+sum(abs((L0 %*% t(L0)) - Q0)) < 10E-9
+```
+
+## Step 2. Loop
+
+For each $x_i$ in a grid of evaluation points (how does one chose this grid?) repeat the following:
+
+```{r}
+spacing <- 0.05
+x_grid <- seq(from = 0.5, to = 3.5, by = spacing)
+```
+
+### a)
+
+Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean):
+
+```{r}
+#' Conditional mean for the first grid point
+Sigma <- solve(Q)
+conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i])
+```
+
+Note: for this model the latent field is IID, so the conditional mean is the same as the mean.
+
+### b)
+
+Compute a set of $J$ steps $\{\Delta_j\}$
+
+```{r}
+J <- 1
+```
+
+### c)
+
+For $j = 1, \ldots, J - 1$ compute
 $$
@@
 $$
-  d) Compute the determinant approximation...
-  e) Compute $\tilde p(x_i \, | \, \theta, y) = \ldots$
-3. Renormalise $\tilde p(x_i \, | \, \theta, y)$
+
+### d)
+
+Compute the determinant approximation
+$$
+\det(Q_1) = \det(Q_0) \det(I_{2J} + u_J^\top Q_0^{-1} u_J \mathcal{D}_{2j}).
+$$
+
+### e)
+
+Compute $\tilde p(x_i \, | \, \theta, y) = \ldots$.
+
+## Step 3. Renormalise $\tilde p(x_i \, | \, \theta, y)$
+
+Renormalise $\tilde p(x_i \, | \, \theta, y)$.
 

### 2022-10-22 docs/tmb.Rmd
TMB note

---
title: "`TMB`"
author: "Adam Howes, Imperial College London"
output:
  pdf_document: default
header-includes:
  \usepackage{bm}
---

Consider unobserved latent random effects $\mathbf{x} \in \mathbb{R}^n$ and parameters $\bm{\theta} \in \mathbb{R}^m$.^[Kristensen (2016) use the notation $u$ for random effects and $\theta$ for parameters. We aim for consistency with Section \ref{sec:inla}.]
Let $\ell(\mathbf{x}, \bm{\theta}) \triangleq - \log p(\mathbf{y} \, | \, \mathbf{x}, \bm{\theta})$ be the negative joint log-likelihood.
In `TMB`, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
A standard maximum likelihood approach is to optimise
\begin{equation}
L_\ell(\bm{\theta}) \triangleq \int_{\mathbb{R}^n} p(\mathbf{y} \, | \, \mathbf{x}, \bm{\theta}) \text{d}\mathbf{x} = \int_{\mathbb{R}^n} \exp(-\ell(\mathbf{x}, \bm{\theta})) \text{d}\mathbf{x} \label{eq:flikelihood}
\end{equation}
with respect to $\bm{\theta}$ to find the maximum likelihood estimator (MLE) $\hat{\bm{\theta}}$.
Taking a superficially more Bayesian approach than above, instead of $\ell$, the user may instead write a function to evaluate the negative joint penalised log-likelihood given by
\begin{equation}
f(\mathbf{x}, \bm{\theta}) 
\triangleq - \log p(\mathbf{y} \, | \, \mathbf{x}, \bm{\theta}) p(\mathbf{x}, \bm{\theta})
= \ell(\mathbf{x}, \bm{\theta}) - \log p(\mathbf{x}, \bm{\theta}),
\end{equation}
equivalent up to an additive constant to the negative log-posterior.
\begin{equation}
f(\mathbf{x}, \bm{\theta})
= - \log p(\mathbf{y}, \mathbf{x}, \bm{\theta})
= - \log p(\mathbf{x}, \bm{\theta} \, | \, \mathbf{y}) - C,
\end{equation}
where $C = \log p(\mathbf{y})$ is the log evidence.
Using $f$ in place of $\ell$, then the penalised likelihood is proportional to the posterior marginal of $\bm{\theta}$
\begin{equation}
L_f(\bm{\theta}) 
\triangleq \int_{\mathbb{R}^n} \exp(-f(\mathbf{x}, \bm{\theta})) \text{d}\mathbf{x} 
\propto \int_{\mathbb{R}^n} p(\mathbf{x}, \bm{\theta} \, | \, \mathbf{y}) \text{d}\mathbf{x} = p(\bm{\theta} \, | \, \mathbf{y}). \label{eq:1}
\end{equation}
Integrating out the random effects directly, as in Equation \ref{eq:1} above, is usually intractable because $\mathbf{x}$ is high-dimensional, so Kristensen (2016, Equation 3) use a Laplace approximation $L^\star_f(\bm{\theta})$ based instead upon integrating out a Gaussian approximation to the random effects.
This Laplace approximation is analogous to the INLA approximation $\tilde p(\bm{\theta} \, | \, \mathbf{y})$.
\begin{equation*}
f''_{\mathbf{x}\mathbf{x}}(\hat{\bm{\mu}}(\bm{\theta}), \bm{\theta}) = - \frac{\partial^2}{\partial \mathbf{x}^2} \log p(\mathbf{y}, \mathbf{x}, \bm{\theta}) \Big\rvert_{\mathbf{x} = \hat{\bm{\mu}}(\bm{\theta})}
= - \frac{\partial^2}{\partial \mathbf{x}^2} \log p(\mathbf{x} \, | \, \bm{\theta}, \mathbf{y}) \Big\rvert_{\mathbf{x} = \hat{\bm{\mu}}(\bm{\theta})}
= \hat{\bm{Q}}(\bm{\theta}).
\end{equation*}
Inference proceeds by optimising $L^\star_f(\bm{\theta})$ via minimisation of
\begin{equation}
-\log L^\star_f(\bm{\theta}) \propto \frac{1}{2} \log \det (\hat{\bm{Q}}(\bm{\theta})) + f(\hat{\bm{\mu}}(\bm{\theta}), \bm{\theta})  \label{eq:nllaplace},
\end{equation}
where $\propto$ is used to mean proportional up to an additive constant.
The parameters of the Gaussian approximation, are found in terms of $f$ via $\hat{\bm{\mu}}(\bm{\theta}) = \arg \min_\mathbf{x} f(\mathbf{x}, \bm{\theta})$ and $\hat{\bm{Q}}(\bm{\theta}) = f''_{\mathbf{x}\mathbf{x}}(\hat{\bm{\mu}}(\bm{\theta}), \bm{\theta})$ and must be recomputed for each value of $\bm{\theta}$.
Obtaining $\hat{\bm{\mu}}(\bm{\theta})$ is known as the inner optimisation step.

### 2022-10-22 src/docs_paper/paper.Rmd
Iterate on SINLA development. Still trying to get TMB Laplace approximation to match with (more) manually calculated one
Update write-up

@@
 ---
-title: Simplified integrated nested Laplace approximation for extended latent Gaussian models
+title: Integrated nested Laplace approximations for extended latent Gaussian models
 author:
@@
     addressLabel: A
-    sepNext: "," # use "," or \and; the last author not need this field
+    sepNext: ","
   - name: Alex
@@
     addressLabel: B
+  - name: Seth R.
+    surname: Flaxman
+    email: seth.flaxman@cs.ox.ac.uk
+    label: e3
+    addressLabel: C
+    sepNext: ","
+  - name: Jeffrey W.
+    surname: Eaton
+    email: jeffrey.eaton@imperial.ac.uk
+    label: e4
+    addressLabel: D
 affiliation:
@@
     authorsLabels: e2
+  - label: C
+    name: Department of Computer Science, University of Oxford
+    authorsLabels: e3
+  - label: D
+    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
+    authorsLabels: e4
 abstract: |
-  | Naomi is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce HIV epidemic indicators.
-We combine the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gaussian hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, we compare our inference method to other approaches.
-We provide an easy to use implementation as a part of the `aghq` \textsf{R} package allowing flexible use of the INLA method for a broader class of models, including any with a `TMB` template.
+  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa.
+Multiple outcomes of interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
+We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
+Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach used currently, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler.
+By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} C++ template for the model's log-posterior.
 keyword-subclass: | 
@@
 
-We are motivated by a challenging inference problem in HIV surveillance.
-Mounting an effective public health response to the epidemic requires accurate, timely and sufficiently fine-scale estimates of HIV indicators.
+Mounting an effective public health response to the HIV epidemic requires accurate, timely and sufficiently fine-scale estimates of HIV indicators.
 No single data source is sufficient to base these estimates on.
-For example, although nationally-representative household surveys are the most reliable data source, in most countries they only occur infrequently.
-As such, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
+Although nationally-representative household surveys are the most reliable data source, in most countries they only occur infrequently.
+Other data sources, such as routine health surveillance of antenatal care clinics, are more timely but from a biased sample of the population.
+To remedy these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
 Software (\url{https://naomi.unaids.org}) has been developed for Naomi, which allows countries to input their data and generate estimates in a yearly process supported by UNAIDS.
-For this reason, any inference method used must be fast enough to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
-Inference is currently conducted using the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-`TMB` implements an empirical Bayes approach, and is gaining popularity in spatial statistics for its speed and flexibility [@osgoodzimmerman2021statistical].
 
-Using simulation we found inferences from `TMB` for the Naomi model to sometimes be inaccurate.
-The `aghq` \textsf{R} package [@stringer2021implementing] for adaptive Gaussian Hermite quadrature (AGHQ), significantly improved performance.
-We develop a new inference methodology combining the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with AGHQ.
+The model is complex and presents a challenging inference problem.
+Any approach must be fast enough to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC).
+Currently, inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
+Owing to its speed and flexibility, `TMB` is more broadly becoming a popular package in spatial statistics [@osgoodzimmerman2021statistical].
+In `TMB`, inference is based on optimisation of a C++ template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
+Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
+Although this approach has favorable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
+To obtain fast, accruate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
 
-<!-- Furthermore, as Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast] it cannot be fit using the integrated nested Laplace approximation (INLA)  method [@rue2009approximate] as implemented by the `R-INLA` \textsf{R} package.
--->
+INLA [@rue2009approximate] is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature, and implemented in the `R-INLA` \textsc{R} package.
+The central innovation is a way to approximate accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
+Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
+Doing so facilitates inference for extended latent Gaussian models (ELGMs) [@stringer2022fast], which build on latent Gaussian models (LGMs) by allowing each element of the linear predictor $\mu_i$ to depend on any subset of elements from the latent field.
+
+AGHQ is a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
+@bilodeau2021stochastic provide theoretical results.
+The method is implementing in the `aghq` \textsf{R} package [@stringer2021implementing].
 
@@
 
-INLA [@rue2009approximate] is an approximate Bayesian inference method based on nested Gaussian approximations and numerical integration, and designed for use with latent Gaussian models (LGMs) of the form
+INLA is designed for use with latent Gaussian models (LGMs) of the form
 \begin{alignat}{2}
@@
 We now move on to the approximation $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$.
-Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section \ref{sec:approxi} above, a natural approach, and that taken by \citet{rue2007approximate}, is to marginalise this distribution directly to obtain
+Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section \ref{sec:approxi} above, a natural approach, and that taken by [@rue2007approximate], is to marginalise this distribution directly to obtain
 \begin{equation}
@@
 where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from the relevant entries of $\hat{\bm{\mu}}(\btheta)$ and $\hat{\bm{Q}}(\btheta)$ respectively.
-However, although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode.
-As a result, although this method is available in \texttt{R-INLA} it is generally not advised.
-Instead, \citet{rue2009approximate} propose two methods, a Laplace approximation and a simplified version which is less computationally demanding.
+However, although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode, so is generally not advised.
+Instead, [@rue2009approximate] propose two methods, a Laplace approximation and a simplified version which is less computationally demanding.
 The full Laplace approximation is
@@
 However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $i$, which is often computationally prohibitive.
-Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by \citet{rue2009approximate} to reduce the computational cost:
+Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by [@rue2009approximate] to reduce the computational cost:
 \begin{enumerate}
@@
 
-## Simplified INLA and connection to ELGMs
+# Application to the Naomi model\label{sec:results}
 
-@wood2020simplified propose a simplified version of the INLA algorithm which relaxes the sparsity assumptions on the latent field required for the modifications to Equation \eqref{eq:laplaceapproximation} to be efficient.
-Doing so facilitates inference for ELGMs [@stringer2022fast] which we define below.
-Informally, ELGMs build on LGMs by allowing each element of the linear predictor $\mu_i$ to depend on any subset of elements from the latent field.
-
-## Implementation within `aghq`\label{sec:aghq}
-
-### Adaptive Gaussian Hermite quadrature
-
-### Template Model Builder
-
-`TMB` [@kristensen2016tmb] is an \textsf{R} package for fitting random effect models, also known as latent variable models, hierarchical models or a host of other names.
-In `TMB` inference is based upon optimisation of a target function.
-This makes it very flexible, and able to handle non-linear, non-Gaussian random effect models.
-
-The approach of `TMB` is inspired by the AD Model Builder (ADMB) package [@fournier2012ad], where the "AD" stands for automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
-AD is popular in machine learning [@baydin2017automatic], including as the basis for backpropagation algorithm, and is beginning to gain popularity in statistics, including as a part of Stan [@carpenter2017stan].
-`TMB` uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.
-
-Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.^[@kristensen2016tmb use the notation $u$ for random effects and $\theta$ for parameters.
-We aim for consistency with Section \ref{sec:inla}.]
-Let $\ell(\x, \btheta) \triangleq - \log p(\y \, | \, \x, \btheta)$ be the negative joint log-likelihood.
-In `TMB`, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
-A standard maximum likelihood approach is to optimise
-\begin{equation}
-L_\ell(\btheta) \triangleq \int_{\mathbb{R}^n} p(\y \, | \, \x, \btheta) \text{d}\x = \int_{\mathbb{R}^n} \exp(-\ell(\x, \btheta)) \text{d}\x \label{eq:flikelihood}
-\end{equation}
-with respect to $\btheta$ to find the maximum likelihood estimator (MLE) $\hat \btheta$.
-Taking a superficially more Bayesian approach than above, instead of $\ell$, the user may instead write a function to evaluate the negative joint penalised log-likelihood given by
-\begin{equation}
-f(\x, \btheta) 
-\triangleq - \log p(\y \, | \, \x, \btheta) p(\x, \btheta)
-= \ell(\x, \btheta) - \log p(\x, \btheta),
-\end{equation}
-equivalent up to an additive constant to the negative log-posterior.
-<!-- \begin{equation} -->
-<!-- f(\x, \btheta)  -->
-<!-- = - \log p(\y, \x, \btheta)  -->
-<!-- = - \log p(\x, \btheta \, | \, \y) - C, -->
-<!-- \end{equation} -->
-<!-- where $C = \log p(\y)$ is the log evidence. -->
-Using $f$ in place of $\ell$, then the penalised likelihood is proportional to the posterior marginal of $\btheta$
-\begin{equation}
-L_f(\btheta) 
-\triangleq \int_{\mathbb{R}^n} \exp(-f(\x, \btheta)) \text{d}\x 
-\propto \int_{\mathbb{R}^n} p(\x, \btheta \, | \, \y) \text{d}\x = p(\btheta \, | \, \y). \label{eq:1}
-\end{equation}
-Integrating out the random effects directly, as in Equation \ref{eq:1} above, is usually intractable because $\x$ is high-dimensional, so \citet[Equation 3]{kristensen2016tmb} use a Laplace approximation $L^\star_f(\btheta)$ based instead upon integrating out a Gaussian approximation to the random effects.
-This Laplace approximation is analogous to the INLA approximation $\tilde p(\btheta \, | \, \y)$ given in Section \ref{sec:approxi}.
-\begin{equation*}
-f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta) = - \frac{\partial^2}{\partial \x^2} \log p(\y, \x, \btheta) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
-= - \frac{\partial^2}{\partial \x^2} \log p(\x \, | \, \btheta, \y) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
-= \hat{\bm{Q}}(\btheta).
-\end{equation*}
-Inference proceeds by optimising $L^\star_f(\btheta)$ via minimisation of
-\begin{equation}
--\log L^\star_f(\btheta) \propto \frac{1}{2} \log \det (\hat{\bm{Q}}(\btheta)) + f(\hat{\bm{\mu}}(\btheta), \btheta)  \label{eq:nllaplace},
-\end{equation}
-where $\propto$ is used to mean proportional up to an additive constant.
-The parameters of the Gaussian approximation (Equation \ref{eq:gaussianx}), are found in terms of $f$ via $\hat{\bm{\mu}}(\btheta) = \arg \min_\x f(\x, \btheta)$ and $\hat{\bm{Q}}(\btheta) = f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta)$ and must be recomputed for each value of $\btheta$.
-Obtaining $\hat{\bm{\mu}}(\btheta)$ is known as the inner optimisation step.
-
-# Application to the Naomi model\label{sec:results}
+The \textsc{R} [@r] code used to implement the model and produce all results we describe is available at `github.com/athowes/elgm-inf`.
+The inference method is available in versions 0.5.0. onwards of the `aghq` package.
 

### 2022-10-24 src/example_inla-replication/inla-replication.Rmd
Improve this a bit

@@
 abstract: |
-    | We demonstrate a step-by-step example of the INLA method implemented in R and TMB.
+    | We demonstrate a step-by-step example of the INLA method implemented in R and `TMB`.
 ---
@@
 \end{alignat}
-with posterior
+with posterior distribution
 \begin{equation}
@@
 
-Using `TMB` if we integrate out the latent field $x$ using the argument `random = "x"` in `MakeADFun` then the result is the negative log of the Laplace approximation to the marginal likelihood $h(\theta) \equiv - \log L^\star(\theta)$:
+Using `TMB` if we integrate out a Gaussian approximation to the latent field $x$ using the argument `random = "x"` in `MakeADFun` then the result is the negative log of the Laplace approximation to the marginal likelihood $h(\theta) \equiv - \log L^\star(\theta)$:
 
@@
 
-$K = 1$ with integration point $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$.
+$K = 1$ with integration point $\hat \theta = \text{argmin}_{\theta} (-\log L^\star_f(\theta))$.
 This can be found by minimising `h` using optimisers like `nlminb` or `optim`.
+Sometimes methods like this are referred to as "[Empirical Bayes](https://en.wikipedia.org/wiki/Empirical_Bayes_method)" where the hyperparameters are estimated with a fixed point estimate.
 
-### INLA grid
+### `R-INLA` grid method
 
-The steps required are:
+Instead of just taking a single point, we might want to explore the posterior approximation to $p(\theta \, | \, y)$ more fully.
+In `R-INLA` a few choices are available as to how to do this, including the grid integration strategy as follows:
 
 1. Locate the mode of $\log p(\theta \, | \, y)$ which is the same as $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$ and can be calculated as above.
-2. Calculate negative Hessian at the mode using e.g. `h$env$spHess`.
+2. Calculate negative Hessian at the mode (this could be done in `TMB` using h$env$spHess`)
 3. Reparameterise in terms of $z$ coordinates along the eigenvectors of the Hessian starting at the mode
-3. Test points at a certain step-size ($\delta_z$) against a condition whereby they must have sufficient posterior density (using a parameter $\delta_{\pi}$)
+4. Test points at a certain step-size ($\delta_z$) against a condition whereby they must have sufficient posterior density (using a parameter $\delta_{\pi}$)
 
+Under this approach, all weights would be equal $\Delta_k = 1 / K$.
 For a more detailed run through of the grid integration method, see the [INLA two dimensional grid replication](https://athowes.github.io/elgm-inf/inla-grid.html) notebook.
 
-### Adaptive Gaussian Hermite quadrature
+### Adaptive Gauss-Hermite quadrature
 
-We could also use an adaptive Gauss Hermite quadrature rule.
+We could also use an adaptive Gauss-Hermite quadrature (AGHQ) rule.
 See the [Implementing Approximate Bayesian Inference using Adaptive Quadrature: the `aghq` Package](https://arxiv.org/pdf/2101.04468.pdf) paper.
+AGHQ is attractive because it is automatic, works well for integrating densities close to being Gaussian.
+
+## Stage 3: get $\tilde p(x_i \, | \, \theta, y)$
 
@@
 
-Obtain marginals of $\tilde p_{\text{G}}(x \, | \, \theta, y)$ from above.
+The simplest approach would be to directly obtain the marginals of $\tilde p_{\text{G}}(x \, | \, \theta, y)$ from above.
 
@@
 One of the INLA vignettes discusses how the user can set their own integration points -- see `browseVignettes(package = "INLA")`.
-The integration points that `R-INLA` uses can be found using `fit$joint.hyper`, which we can add on to the above plot for interest:
+The integration points that `R-INLA` uses can be found using `fit$joint.hyper`:
 
@@
 
-# Different normalising constant?
-ggplot(dense_theta) +
-  geom_line(aes(x = input, y = output), col = midblue) +
-  geom_point(data = internals_inla_theta, aes(x = input, y = output), shape = 4) +
-  theme(axis.text.y = element_blank()) +
-  labs(title = "INLA integration points for the hyperparameters") +
-  theme_minimal()
+internals_inla_theta$input
 ```
@@
 ```{r}
-blangiardo_x <- seq(-8, 5, length.out = 50)
+x_start <- -8
+x_end <- 5
+x_out <- 50
+blangiardo_x <- seq(x_start, x_end, length.out = x_out)
 ```
@@
 
-Following the book:
+Calculating the log posterior by integrating over the hyperparameters for each value of $x$:
 
 ```{r}
-nl_post_marginal_x <- rowSums(nl_joint_post)
-plot(blangiardo_x, nl_post_marginal_x)
-blangiardo_x[[which.min(nl_post_marginal_x)]]
+K <- length(points)
+nl_post_marginal_x <- rowSums(nl_joint_post) / K
+
+df <- data.frame(
+  x = blangiardo_x,
+  log_post = -1 * nl_post_marginal_x
+)
+
+ggplot(df, aes(x = x, y = log_post)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "x", y = "log(Posterior)")
+```
+
+Try normalising:
+
+```{r}
+x_spacing <- (x_end - x_start) / x_out
+safe_exp <- function(x) exp(x - max(x))
+df$post <- safe_exp(df$log_post)
+norm <- trapezoid_rule(df$post, x_spacing)
+df$norm_post <- df$post / norm
+
+ggplot(df, aes(x = x, y = norm_post)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "x", y = "Posterior")
 ```
@@
 
-Compare $x$:
+What about for $x$:
 

### 2022-10-24 src/dev_sinla/sinla.Rmd
Loop i = 1:30 and plot the marginals

@@
 
-opt <- nlminb(
+opt_theta <- nlminb(
   start = h$par,
@@
 
-The values `opt$par` below correspond to $\hat \theta$:
+The values `opt_theta$par` below correspond to $\hat \theta$:
 
 ```{r}
-opt$par
+opt_theta$par
 ```
 
-We can also look at the parameters -- that is, mean vector and precision matrix -- of the Gaussian approximation at the mode `opt$par` above.
+We can also look at the parameters -- that is, mean vector and precision matrix -- of the Gaussian approximation at the mode `opt_theta$par` above.
 Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
@@
 ```{r}
-#' Notice that last.par contains opt$par as the fixed effects
+#' Notice that last.par contains opt_theta$par as the fixed effects
 #' It's the last.par because the last thing we've done with the template is the optimsation
@@
 
-## $\tilde p(x_i \, | \, \theta, y)$
+## For one index $i$
+
+### $\tilde p(x_i \, | \, \theta, y)$
 
@@
 
-Now we can loop over indices $i$ doing the required Laplace approximations.
+Let's consider the first marginal, that is $\tilde p_\text{LA}(x_1 \, | \, \theta, y)$.
 In the call to `MakeADFun` we set `random = "phi_prev_minus_i"` to integrate all of the random effects but that of the particular index.
@@
 ```{r message=FALSE, warning=FALSE, results=FALSE}
-objs <- list()
-means <- list()
-Qs <- list()
+i <- 1
 
-#' Starting parameters for TMB are the same for every index of the loop so it's more efficient to put them outside
 param_i <- list(
@@
 
-its <- 1000
-
-for(i in 1:dat$n) {
 dat_i <- prepare_dat(dat, i)
@@
 )
+```
 
-  #' Parameters of the Gaussian approximation
+The mean and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:
+
+```{r}
 mm <- obj$env$last.par
 mean <- mm[obj$env$random]
+length(mean) #' The length is 30 - 1
+
 Q <- obj$env$spHess(mm, random = TRUE)
-
-  objs[[i]] <- obj
-  means[[i]] <- mean
-  Qs[[i]] <- Q
-}
-```
-
-Let's consider the first marginal, that is $\tilde p_\text{LA}(x_1 \, | \, \theta, y)$.
-The mean and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:
-
-```{r message=FALSE, warning=FALSE}
-i <- 1
-obj <- objs[[i]]
-
-mean <- means[[i]] #' The length is 30 - 1
-length(mean)
-
-Q <- Qs[[i]]
 dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
@@
 
-## $\tilde p(x_i \, | \, y)$ {.tabset .tabset-fade}
+### $\tilde p(x_i \, | \, y)$ {.tabset .tabset-fade}
 
@@
 
-We're always going to need $\hat \theta$.
-We have calculated it before, but we will get it fresh here (because of questionable programming practices whereby objects are overwritten):
-
-```{r}
-opt <- nlminb(
-  start = h$par,
-  objective = h$fn,
-  gradient = h$gr,
-  control = list(iter.max = its, trace = 0)
-)
-
-opt$par
-```
-
-Now, a basic way we could approach this is to define a grid of points that we would like to evaluate the posterior at.
-We can then create a dataframe where each row contains the values `opt$par` for $\theta$ and a grid value for $x_i$.
+A basic way we could approach this is to define a grid of points that we would like to evaluate the posterior at.
+We can then create a dataframe where each row contains the values `opt_theta$par` for $\theta$ and a grid value for $x_i$.
 Each row can then be passed into `-obj$fn()` to calculate the log-posterior.
@@
 
-df <- t(opt$par) %>%
+df <- t(opt_theta$par) %>%
   as.data.frame() %>%
@@
 
-### Safe exponential trapezoid rule {-}
+#### Safe exponential trapezoid rule {-}
 
@@
 
-### `logSumExp` trapezoid rule {-}
+#### `logSumExp` trapezoid rule {-}
 
@@
 
-## Normalisation with `aghq`
+### Normalisation with `aghq`
 
@@
 
-We could normalise $\tilde p_\text{LA}(\theta, y)$.
+We could start by normalising $\tilde p_\text{LA}(\theta, y)$.
 Note that sometimes we have been calling this $\tilde p_\text{LA}(\theta \, | \, y)$ because they are equivalent up to a constant of proportionality.
@@
 
-theta_names <- names(opt$par)
+theta_names <- names(opt_theta$par)
 for(theta in theta_names) {
-  param_i_fixed_theta[[theta]] <- opt$par[[theta]]
+  param_i_fixed_theta[[theta]] <- opt_theta$par[[theta]]
 }
@@
 
-## Comparison
+### Comparison
 
@@
 
+## Looping over index $i$
+
+So we've done the approximation for the first index $i = 1$!
+Now let's try computing $\tilde p(x_i \, | \, y)$ for every index $i$, using a for-loop to generate the required Laplace approximations, and tidying up and condensing the code used:
+
+```{r}
+xi_laplace_marginal <- function(i) {
+  dat_i <- prepare_dat(dat, i)
+  map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
+  param_i_fixed_theta <- list(beta_prev = 0, phi_prev_i = 0, phi_prev_minus_i = rep(0, data$n - 1), log_sigma_phi_prev = 0)
+  for(theta in names(opt_theta$par)) param_i_fixed_theta[[theta]] <- opt_theta$par[[theta]]
+  obj_fixed_theta <- MakeADFun(data = dat_i, parameters = param_i_fixed_theta, random = "phi_prev_minus_i", DLL = "model1index", silent = TRUE, map = map_fixed_theta)
+  quad <- aghq::aghq(ff = obj_fixed_theta, k = 3, startingvalue = 0, control = aghq::default_control_tmb())
+  pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad)[[1]]
+  return(pdf_and_cdf)
+}
+
+out <- lapply(1:dat$n, xi_laplace_marginal)
+
+laplace_df <- bind_rows(out, .id = "index") %>%
+  mutate(index = as.numeric(index))
+```
+
+Let's also get everything we need for the joint Gaussian marginal for each index:
+
+```{r}
+mm <- h$env$last.par
+mean <- mm[h$env$random]
+Q <- h$env$spHess(mm, random = TRUE)
+Sigma <- solve(Q)
+
+gaussian_df <- lapply(1:length(mean), function(i) {
+  x <- seq(-5, 5, length.out = 1000)
+  mean_i <- as.numeric(mean[i])
+  sd_i <- as.numeric(sqrt(Sigma[i, i]))
+  data.frame(index = i, x = x, pdf = dnorm(x, mean = mean_i, sd = sd_i))
+}) %>%
+  bind_rows()
+```
+
+## Comparison
+
+```{r}
+samples_tmbstan <- extract(mcmc, pars = "phi_prev") %>%
+  as.data.frame() %>%
+  pivot_longer(
+    cols = everything(),
+    names_to = "index",
+    names_prefix = "phi_prev.",
+    names_transform = as.integer
+  )
+
+plot <- ggplot(samples_tmbstan, aes(x = value)) +
+  geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
+  facet_wrap(~index) +
+  theme_minimal() +
+  labs(x = "phi_prev", y = "Posterior PDF")
+
+plot2 <- plot +
+  geom_line(data = gaussian_df, aes(x = x, y = pdf), col = cbpalette[1])
+```
+
+```{r}
+plot3 <- plot2 +
+  geom_line(data = laplace_df, aes(x = theta, y = pdf), col = cbpalette[2])
+
+plot3
+```
+
 # Simplified INLA
@@
 Let $u_0$ be a zero column matrix.
-I am guessing that this means $u_j$ is a $j$ column matrix.
-It's not specified what dimension it is too, but since it's multiplied by $\mathcal{D}_{2j}$ presumably it must be of length $2j$.
+I am guessing that this means $u_j$ is a column matrix containing $j$ as each entry.
+It's not specified what dimension $u_0$ or $u_j$ should be, but since the algorithm contains producing by $\mathcal{D}_{2j}$ presumably their lengths must be $2j$.
 
@@
 
-```{r}
-D2j(3) %*% uj(3)
-```
+The algorithm contains $u_j \mathcal{D}_{2j} u_j^\top$.
+If $u_j$ is a column matrix of dimension $\text{length}(u_j) \times 1$ then it's not going to be conformable with a $2j \times 2j$ matrix.
+Is it actually a row matrix of dimension $1 \times \text{length}(u_j)$?
+And $\text{length}(u_j) = 2j$ as discussed above?
 
 Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$.
-Note that it hasn't been defined what $\Delta_j$ is yet.
+But it hasn't been defined what $\Delta_j$ is yet.
 
@@
 
-Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean):
+Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean).
+We start by computing the conditional mean for the first grid point:
 
 ```{r}
-#' Conditional mean for the first grid point
 Sigma <- solve(Q)
@@
 
-Note: for this model the latent field is IID, so the conditional mean is the same as the mean.
+Note that for this model the latent field is IID, so the conditional mean is the same as the mean.
+Now we use Newton's method to find $\mu(x_i)$.
+This requires some function $f(\mu(x_i), \ldots) = 0$.
 
@@
 ### c)
+
+```{r}
+j <- 1
+```
 

### 2022-10-25 src/dev_sinla/sinla.Rmd
Add clean version script #16

@@
 ```{r}
+#' Making sure the mean and Q used here are correct (preventing and shady TMB behaviour)
+h$fn(opt_theta)
+mm <- h$env$last.par
+mean <- mm[h$env$random]
+Q <- h$env$spHess(mm, random = TRUE)
+
 i <- 1

### 2022-10-25 src/dev_sinla/sinla.R

@@ -1,0 +1,125 @@
+#' Clean script version of sinla.Rmd notebook
+orderly::orderly_develop_start("dev_sinla")
+
+cbpalette <- multi.utils::cbpalette()
+
+sim_data <- readRDS("depends/sim_data.rds")
+data <- sim_data[[1]]
+dat <- list(n = data$n, y_prev = data$y_prev, m_prev = data$m_prev)
+
+compile("model1.cpp")
+dyn.load(dynlib("model1"))
+
+param <- list(
+  beta_prev = -2,
+  phi_prev = rep(0, data$n),
+  log_sigma_phi_prev = -1
+)
+
+h <- MakeADFun(
+  data = dat,
+  parameters = param,
+  random = "phi_prev",
+  DLL = "model1",
+  silent = TRUE
+)
+
+#' 1: MCMC
+mcmc <- tmbstan::tmbstan(h, chains = 4, iter = 5000, refresh = 0)
+
+samples_tmbstan <- extract(mcmc, pars = "phi_prev") %>%
+  as.data.frame() %>%
+  pivot_longer(
+    cols = everything(),
+    names_to = "index",
+    names_prefix = "phi_prev.",
+    names_transform = as.integer
+  )
+
+#' 2: Gaussian
+its <- 1000
+
+opt_theta <- nlminb(
+  start = h$par,
+  objective = h$fn,
+  gradient = h$gr,
+  control = list(iter.max = its, trace = 0)
+)
+
+mm <- h$env$last.par
+mean <- mm[h$env$random]
+Q <- h$env$spHess(mm, random = TRUE)
+Sigma <- solve(Q)
+
+gaussian_df <- lapply(1:length(mean), function(i) {
+  x <- seq(-5, 5, length.out = 1000)
+  mean_i <- as.numeric(mean[i])
+  sd_i <- as.numeric(sqrt(Sigma[i, i]))
+  data.frame(index = i, x = x, pdf = dnorm(x, mean = mean_i, sd = sd_i))
+}) %>%
+  bind_rows()
+
+#' 3: Laplace
+
+prepare_dat <- function(dat, i) {
+  dat[["y_prev_i"]] <- dat$y_prev[i]
+  dat[["y_prev_minus_i"]] <- dat$y_prev[-i]
+  dat[["m_prev_i"]] <- dat$m_prev[i]
+  dat[["m_prev_minus_i"]] <- dat$m_prev[-i]
[…]

### 2022-10-25 src/dev_sinla/sinla.Rmd
Add KS tests for all i = 1:30 #16

@@
 M <- 1000
-samples_jointgaussianmarginal <- rnorm(M, mean_i, sqrt(var_i))
+samples_gaussian <- rnorm(M, mean_i, sqrt(var_i))
 
-ecdf_jointgaussianmarginal <- ecdf(samples_jointgaussianmarginal)
-grid <- seq(min(samples_jointgaussianmarginal), max(samples_jointgaussianmarginal), by = 0.1)
+ecdf_gaussian <- ecdf(samples_gaussian)
+grid <- seq(min(samples_gaussian), max(samples_gaussian), by = 0.1)
 
@@
   geom_line(
-    data = data.frame(x = grid, y = ecdf_jointgaussianmarginal(grid)),
+    data = data.frame(x = grid, y = ecdf_gaussian(grid)),
     aes(x = x, y = y), col = cbpalette[1]
@@
 ```{r}
-ks_jointgaussianmarginal_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_jointgaussianmarginal)
-ks_jointgaussianmarginal_tmbstan
+ks_gaussian_tmbstan <- inf.utils::ks_test(samples_tmbstan, samples_gaussian)
+ks_gaussian_tmbstan
 ```
 
-The maximum difference between the two ECDFS is `r ks_jointgaussianmarginal_tmbstan$D` which occurs at the point `r ks_jointgaussianmarginal_tmbstan$l`.
+The maximum difference between the two ECDFS is `r ks_gaussian_tmbstan$D` which occurs at the point `r ks_gaussian_tmbstan$l`.
 
@@
 kable_data <- data.frame(
-  "Distance" = c(ks_jointgaussianmarginal_tmbstan$D, ks_laplace_tmbstan$D),
-  "Location" = c(ks_jointgaussianmarginal_tmbstan$l, ks_laplace_tmbstan$l)
+  "Distance" = c(ks_gaussian_tmbstan$D, ks_laplace_tmbstan$D),
+  "Location" = c(ks_gaussian_tmbstan$l, ks_laplace_tmbstan$l)
 )
 
-rownames(kable_data) <- c("$\\text{KS}(\\texttt{tmbstan}, \\texttt{jointgaussianmarginal})$", "$\\text{KS}(\\texttt{tmbstan}, \\texttt{laplace})$")
+rownames(kable_data) <- c("$\\text{KS}(\\texttt{tmbstan}, \\texttt{gaussian})$", "$\\text{KS}(\\texttt{tmbstan}, \\texttt{laplace})$")
 
@@
 
-out <- lapply(1:dat$n, xi_laplace_marginal)
+laplace_df <- lapply(1:dat$n, xi_laplace_marginal) %>%
+  bind_rows(.id = "index") %>%
+  mutate(index = as.numeric(index))
 
-laplace_df <- bind_rows(out, .id = "index") %>%
-  mutate(index = as.numeric(index))
+M <- 1000
+
+samples_laplace <- lapply(1:1:dat$n, function(i) {
+  samples <- approxfun(laplace_df$cdf, laplace_df$theta)(runif(M))
+  data.frame("value" = samples)
+}) %>%
+  bind_rows(.id = "index")
 ```
@@
 
-gaussian_df <- lapply(1:length(mean), function(i) {
+gaussian_df <- lapply(1:dat$n, function(i) {
   x <- seq(-5, 5, length.out = 1000)
@@
   bind_rows()
+
+samples_gaussian <- lapply(1:dat$n, function(i) {
+  mean_i <- as.numeric(mean[i])
+  sd_i <- as.numeric(sqrt(Sigma[i, i]))
+  samples <- rnorm(M, mean_i, sd_i)
+  data.frame("value" = samples)
+}) %>%
+  bind_rows(.id = "index")
 ```
@@
 plot3
+```
+
+### KS test
+
+```{r}
+ks_gaussian_tmbstan <- lapply(1:length(mean), function(i) {
+  samples_gaussian_i <- filter(samples_gaussian, index == i)$value
+  samples_tmbstan_i <- filter(samples_tmbstan, index == i)$value
+  inf.utils::ks_test(samples_gaussian_i, samples_tmbstan_i)
+})
+
+ks_laplace_tmbstan <- lapply(1:length(mean), function(i) {
+  samples_laplace_i <- filter(samples_laplace, index == i)$value
+  samples_tmbstan_i <- filter(samples_tmbstan, index == i)$value
+  inf.utils::ks_test(samples_laplace_i, samples_tmbstan_i)
+})
+
+ks_results <- bind_rows(
+  bind_rows(ks_gaussian_tmbstan, .id = "index") %>%
+    mutate(type = "gaussian"),
+  bind_rows(ks_laplace_tmbstan, .id = "index") %>%
+    mutate(type = "laplace")
+)
+
+ks_results %>%
+  select(index, D, type) %>%
+  pivot_wider(
+    names_from = type,
+    values_from = D
+  ) %>%
+  ggplot(aes(x = laplace, y = gaussian)) +
+    geom_point(alpha = 0.5) +
+    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
+    lims(x = c(0, 1), y = c(0, 1)) +
+    labs(x = "KS(laplace, tmbstan)", y = "KS(gaussian, tmbstan)") +
+    theme_minimal()
 ```

### 2022-10-25 src/dev_sinla/sinla.Rmd
Get rid of warnings for approxfun #16

@@
 samples_laplace <- lapply(1:1:dat$n, function(i) {
-  samples <- approxfun(laplace_df$cdf, laplace_df$theta)(runif(M))
+  samples <- approxfun(x = laplace_df$cdf, y = laplace_df$theta, ties = "ordered")(runif(M))
   data.frame("value" = samples)

### 2022-10-26 src/dev_sinla/sinla.Rmd
Fix KS bug, add distributions and KS plot to script version

--- 2022-10-25 src/dev_sinla/sinla.Rmd
@@
 
-samples_laplace <- lapply(1:1:dat$n, function(i) {
-  samples <- approxfun(laplace_df$cdf, laplace_df$theta)(runif(M))
+samples_laplace <- lapply(1:dat$n, function(i) {
+  laplace_df_i <- filter(laplace_df, index == i)
+  samples <- approxfun(x = laplace_df_i$cdf, y = laplace_df_i$theta, ties = "ordered")(runif(M))
   data.frame("value" = samples)

### 2022-10-26 src/dev_sinla/sinla.Rmd
Try bug resolve by using rlang::duplicate and being careful, note on tmbstan #16

@@
 It is simple to fit the model using MCMC via the `tmbstan` package.
-We use 4 `chains` each of 5000 `iter`:
+We use 4 `chains` each of 5000 `iter`.
+Note that although `h` is passed to `tmbstan::tmbstan`, the Laplace approximation is disabled by setting `laplace = FALSE` by default (useful to know in case it's ever of interest to sample from the Laplace approximation).
 
@@
 #' Note that I'm calling this mean, but because it's a Gaussian the mean is the same as the mode
-mean <- mm[h$env$random]
-length(mean) #' The length is 30
+x_mean <- mm[h$env$random]
+x_mean <- rlang::duplicate(x_mean) #' Insurance...
+length(x_mean) #' The length is 30
 
 #' Hessian evaluated at the mean (or equivalently mode)
-Q <- h$env$spHess(mm, random = TRUE)
-dim(Q) #' With dimension [30 x 30]
+x_Q <- h$env$spHess(mm, random = TRUE)
+x_Q <- rlang::duplicate(x_Q)
+dim(x_Q) #' With dimension [30 x 30]
 ```
@@
 ```{r}
-#' Making sure the mean and Q used here are correct (preventing and shady TMB behaviour)
-h$fn(opt_theta)
-mm <- h$env$last.par
-mean <- mm[h$env$random]
-Q <- h$env$spHess(mm, random = TRUE)
-
 i <- 1
-mean_i <- as.numeric(mean[i])
-var_i <- solve(Q)[i, i]
+mean_i <- as.numeric(x_mean[i])
+var_i <- solve(x_Q)[i, i]
 ```
@@
 ```{r}
+get_mean_and_Q <- function(obj) {
   mm <- obj$env$last.par
   mean <- mm[obj$env$random]
-length(mean) #' The length is 30 - 1
+  mean <- rlang::duplicate(mean)
+  Q <- obj$env$spHess(mm, random = TRUE)
+  Q <- rlang::duplicate(Q)
+  return(list(mean = mean, Q = Q))
+}
 
-Q <- obj$env$spHess(mm, random = TRUE)
-dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
+xminusi <- get_mean_and_Q(obj)
 ```
@@
   map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
-  param_i_fixed_theta <- list(beta_prev = 0, phi_prev_i = 0, phi_prev_minus_i = rep(0, data$n - 1), log_sigma_phi_prev = 0)
+
+  param_i_fixed_theta <- list(
+    beta_prev = 0, 
+    phi_prev_i = 0, 
+    phi_prev_minus_i = rep(0, data$n - 1), 
+    log_sigma_phi_prev = 0
+  )
+
   for(theta in names(opt_theta$par)) param_i_fixed_theta[[theta]] <- opt_theta$par[[theta]]
-  obj_fixed_theta <- MakeADFun(data = dat_i, parameters = param_i_fixed_theta, random = "phi_prev_minus_i", DLL = "model1index", silent = TRUE, map = map_fixed_theta)
+
+  obj_fixed_theta <- MakeADFun(
+    data = dat_i,
+    parameters = param_i_fixed_theta,
+    random = "phi_prev_minus_i",
+    DLL = "model1index",
+    silent = TRUE,
+    map = map_fixed_theta
+  )
+
   quad <- aghq::aghq(ff = obj_fixed_theta, k = 3, startingvalue = 0, control = aghq::default_control_tmb())
   pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad)[[1]]
+
   return(pdf_and_cdf)

### 2022-10-26 src/dev_sinla/sinla.Rmd
Finish bug fix, get .pdf rendering #16

@@
 ```{r}
-mm <- h$env$last.par
-mean <- mm[h$env$random]
-Q <- h$env$spHess(mm, random = TRUE)
-Sigma <- solve(Q)
+x_Sigma <- solve(x_Q)
 
@@
   x <- seq(-5, 5, length.out = 1000)
-  mean_i <- as.numeric(mean[i])
-  sd_i <- as.numeric(sqrt(Sigma[i, i]))
+  mean_i <- as.numeric(x_mean[i])
+  sd_i <- as.numeric(sqrt(x_Sigma[i, i]))
   data.frame(index = i, x = x, pdf = dnorm(x, mean = mean_i, sd = sd_i))
@@
 samples_gaussian <- lapply(1:dat$n, function(i) {
-  mean_i <- as.numeric(mean[i])
-  sd_i <- as.numeric(sqrt(Sigma[i, i]))
+  mean_i <- as.numeric(x_mean[i])
+  sd_i <- as.numeric(sqrt(x_Sigma[i, i]))
   samples <- rnorm(M, mean_i, sd_i)

### 2022-10-26 src/dev_sinla/sinla.R

@@ -1,207 +1,0 @@
-#' Clean script version of sinla.Rmd notebook
-orderly::orderly_develop_start("dev_sinla")
-
-cbpalette <- multi.utils::cbpalette()
-
-sim_data <- readRDS("depends/sim_data.rds")
-data <- sim_data[[1]]
-dat <- list(n = data$n, y_prev = data$y_prev, m_prev = data$m_prev)
-
-compile("model1.cpp")
-dyn.load(dynlib("model1"))
-
-param <- list(
-  beta_prev = -2,
-  phi_prev = rep(0, data$n),
-  log_sigma_phi_prev = -1
-)
-
-h <- MakeADFun(
-  data = dat,
-  parameters = param,
-  random = "phi_prev",
-  DLL = "model1",
-  silent = TRUE
-)
-
-#' 1: MCMC
-mcmc <- tmbstan::tmbstan(h, chains = 4, iter = 1000, refresh = 0)
-
-samples_tmbstan <- extract(mcmc, pars = "phi_prev") %>%
-  as.data.frame() %>%
-  pivot_longer(
-    cols = everything(),
-    names_to = "index",
-    names_prefix = "phi_prev.",
-    names_transform = as.integer
-  )
-
-#' 2: Gaussian
-its <- 1000
-
-opt_theta <- nlminb(
-  start = h$par,
-  objective = h$fn,
-  gradient = h$gr,
-  control = list(iter.max = its, trace = 0)
-)
-
-mm <- h$env$last.par
-mean <- mm[h$env$random]
-Q <- h$env$spHess(mm, random = TRUE)
-Sigma <- solve(Q)
-
-fine_grid <- seq(-5, 5, length.out = 1000)
-
-gaussian_df <- lapply(1:dat$n, function(i) {
-  mean_i <- as.numeric(mean[i])
-  sd_i <- as.numeric(sqrt(Sigma[i, i]))
-  data.frame(
-    index = i,
-    x = fine_grid,
-    pdf = dnorm(fine_grid, mean = mean_i, sd = sd_i),
-    cdf = pnorm(fine_grid, mean = mean_i, sd = sd_i)
-  )
-}) %>%
-  bind_rows()
-
-M <- 1000
-
-samples_gaussian <- lapply(1:dat$n, function(i) {
-  mean_i <- as.numeric(mean[i])
-  sd_i <- as.numeric(sqrt(Sigma[i, i]))
-  samples <- rnorm(M, mean_i, sd_i)
[…]

### 2022-10-27 src/dev_sinla/sinla.Rmd
Cautionary tale: if you change the data and you're using a non-adaptive quadrature rule you're going to have a bad time

@@
 spacing <- 0.05
-x_grid <- seq(from = 0.5, to = 3.5, by = spacing)
+x_grid <- seq(from = -5, to = 5, by = spacing)
 x_grid_length <- length(x_grid)
@@
 
+# Extension beyond empirical Bayes
+
+In Sections \@ref(gaussian-marginal) and \@ref(full-laplace) when calculating $\tilde p(x_i \, | \, y)$ we fixed $\theta = \hat \theta$ ($K = 1$, empirical Bayes) rather than integrating $\theta$ out ($K > 1$).
+In this section we will use `ahgq` to integrate over $\theta$.
+
+## Marginal of joint Gaussian approximation
+
+```{r}
+#' To-do
+```
+
+## Full Laplace approximation
+
+```{r}
+#' To-do
+```
+
 # Simplified INLA
@@
 ```{r}
-spacing <- 0.05
-x_grid <- seq(from = 0.5, to = 3.5, by = spacing)
+x_grid
 ```

### 2022-10-31 src/prev-anc-art_results/prev-anc-art.Rmd
Rename process-results to results

--- 2022-10-17 src/prev-anc-art_process-results/prev-anc-art.Rmd
@@
 
-## Prevalence component
+## Household survey component
 
-Consider a country partitioned into areas $i = 1, \ldots, n$.
-A simple random household survey of $m_i$ people is conducted in each area, and $y_i$ HIV positive cases are observed.
+Consider a country partitioned into $n$ areas indexed by $i$.
+Suppose a simple random household survey of $m^\text{HS}_i$ people is conducted in each area, and $y^\text{HS}_i$ HIV positive cases are observed.
 Cases may be modelled using a binomial logistic regression model
 \begin{align}
-y_i &\sim \text{Bin}(m_i, \rho_i), \\
-\text{logit}(\rho_i) &\sim \mathcal{N}(\beta_\rho, \sigma_\rho^2)
+y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
+\text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2),
 \end{align}
-where HIV prevalence $\rho_i$ is modelled by a Gaussian with mean $\beta_\rho$ and standard deviation $\sigma_\rho$.
+where HIV prevalence $\rho^\text{HS}_i$ is modelled by a Gaussian with mean $\beta_\phi$ and standard deviation $\sigma_\phi$.
 
@@
 
-The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence.
+The number of people receiving treatment at district health facilities $A_i$ provides further information about HIV prevalence.
 Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
-ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
+ART coverage, defined to be the proportion of people living with HIV (PLHIV) currently on ART on district $i$, is given by $\alpha_i = A_i / \rho^\text{HS}_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be constant.
 As such, ART coverage may also be modelled using a binomial logistic regression model
 \begin{align}
-A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
-\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
+A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
+\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2),
 \end{align}
+where the proportion of people receiving ART is $\rho^\text{HS}_i \alpha_i$.
+Here we assume no travel between districts to receive treatment.
 
@@
 |:----------|:------|
-| $n$              | 30   |
+| $n$              | 36   |
 | $m_i$            | 250  |
@@
 Boxplots of the mean and standard deviation for each parameter, where each datapoint is a different simulation replicate.
+For now we will just focus on a subset of the parameters, as it is challenging to nicely present plots for very large numbers of parameters.
 

### 2022-10-31 utils/archive_to_docs.R

@@ -21,7 +21,7 @@ archive_to_docs("docs_01-04-20-mini")
 archive_to_docs("docs_01-07-21-stats-epi-group")
 archive_to_docs("docs_15-11-22-seminar")
 archive_to_docs("epil")
-archive_to_docs("prev-anc-art_process-results")
+archive_to_docs("prev-anc-art_results")
 archive_to_docs("example_inla-replication")
 archive_to_docs("example_inla-grid")
 archive_to_docs("dev_sinla")

### 2022-10-31 src/explore_posterior-comparison/posterior-comparison.Rmd
Set-up structure for posterior-comparison notebook

---
title: "Posterior comparison methods"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** Multiple approaches for conducting approximate Bayesian inference exist.

  **Task** We compare methods for assessing the accuracy of posterior distribution. These methods include marginal KS tests, simulation based calibration and maximum mean discrepancy.
---

# Data

# KS tests

# Simulation-based calibration

# Maximum mean discrepancy

# Original computing environment

```{r}
sessionInfo()
```

### 2022-10-31 make/_make.R

@@ -10,3 +10,4 @@ run_commit_push("docs_01-04-20-mini")
 run_commit_push("docs_01-07-21-stats-epi-group")
 run_commit_push("docs_15-11-22-seminar")
 run_commit_push("docs_xx-12-22-explainer")
+run_commit_push("explore_posterior-comparison")

### 2022-10-31 utils/archive_to_docs.R

@@ -28,3 +28,4 @@ archive_to_docs("dev_sinla")
 archive_to_docs("dev_sinla", i = 2)
 archive_to_docs("explore_aghq")
 archive_to_docs("example_naomi")
+archive_to_docs("explore_posterior-comparison")

### 2022-10-31 src/dev_sinla/functions.R
Save samples and import to posterior-comparison

@@ -192,4 +192,9 @@ run_analysis <- function(sim_data, analysis_name) {
   ks_results %>%
     group_by(type) %>%
     summarise(mean_D = mean(D))
+
+  saveRDS(
+    list("mcmc" = samples_tmbstan, "gaussian" = samples_gaussian, "laplace" = samples_laplace),
+    paste0("samples-", analysis_name, ".rds")
+  )
 }

### 2022-10-31 src/explore_posterior-comparison/posterior-comparison.Rmd
Save samples and import to posterior-comparison
Reading on SBC and PSIS

@@
     theme: lumen
+bibliography: citations.bib
 abstract: |
@@
 
+# Background
+
+There are lots of ways you can compare posterior distributions.
+
+The simplest is to compare point estimates (such as the mean or mode) or moments (such as the variance).
+In some sense, it's the least we should expect from any good approximate Bayesian inference method to be able to recover these summaries.
+
+More challenging is to match the distribution everywhere.
+One way to assess this is the Kolmogorov-Smirnov (KS) distance, which gives the maximum distance between two empirical cumulative distribution functions (ECDFs).
+It is directly interpretable: a value of 0.05 means that no matter where you compute a tail probability, you're never more than 5% away.
+
+A downside of the KS distance is that it only captures marginal distributions.
+This, in some way, favors methods like INLA where we directly approximate marginals.
+A further challenge would be to assess agreement between joint distributions, rather than only marginals.
+
 # Data
+
+```{r}
+samples <- readRDS("depends/samples-m250.rds")
+samples_mcmc <- samples$mcmc
+samples_gaussian <- samples$gaussian
+samples_laplace <- samples$laplace
+```
 
@@
 
+Compute the empirical CDFs, and look for the maximum distance between them.
+
 # Simulation-based calibration
 
+SBC [@talts2018validating] is a corrected version of @cook2006validation.
+Consider the performance of an algorithm over the entire Bayesian joint distribution.
+Generate $(\tilde \theta, \tilde y)$ via $\tilde \theta \sim p(\theta)$ then $\tilde y \sim p(y \, | \, \tilde \theta)$.
+Let the data averaged posterior (DAP) be
+$$
+p(\theta) = \int p(y \, | \, \tilde \theta) p(\tilde \theta, \tilde y) \text{d} \tilde \theta \text{d} \tilde y (\#eq:dap)
+$$
+For any model, the average of any exact posterior expectation with respect to data generated from the Bayesian joint distribution reduces to the corresponding prior expectation.
+Any discrepancy between the DAP and the prior indicates some error in the analysis (either implementation or algorithmic).
+
+Consider a sequence of samples from the Bayesian joint distribution
+$$
+\tilde \theta \sim p(\theta) \\
+\tilde y \sim p(y \, | \, \tilde \theta) \\
+\{\theta_1, \ldots, \theta_L \} \sim p(\theta \, | \, \tilde y)
+$$
+Equation \@ref(eq:dap) implies that $\tilde \theta$ and $\{\theta_1, \ldots, \theta_L \}$ should both be distributed according to the prior.
+Consequently, for any one-dimensional random variable $f: \Theta \to \mathbb{R}$ the rank statistic of the prior sample relative to the posterior sample
+$$
+r(\{f(\theta_1), \ldots, f(\theta_L) \}, f(\tilde \theta)) = \sum_{l = 1}^L \mathbb{I}[f(\theta_l) < f(\tilde \theta)] \in [0, L]
+$$
+will be distributed uniformly across the integers $[0, L]$.
+I assume we can use histograms and ECDF difference plots to assess this, and perhaps there will be some additional discussion about how to interpret deviations from uniformity.
+
+# Pareto smoothed importance sampling
+
+Say that we have draws $(\theta_1, \ldots, \theta_S)$ from a proposal distribution $q(\theta)$, then we can estimate $\mathbb{E}_p[h(\theta)]$ by
+$$
+\frac{\sum_{s = 1}^S h(\theta_s) w_s}{\sum_{s = 1}^S w_s}.
+$$
+When $w_s = 1$ then this is the direct Monte Carlo estimate, when $w_s = r_s = p(\theta_s, y) / q(\theta)$ then this is importance sampling (IS).
+The finite sample performance of IS contains information about how close the proposal distribution is to the target distribution (the true posterior distribution of interest).
+
+Pareto smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] can improve the estimates from IS by stabilising the importance ratios.
+Let
+$$
+p(y \, | \, \mu, \sigma, k) =
+\begin{cases}
+\frac{1}{\sigma} \left( 1 + k \left( \frac{y - \mu}{\sigma} \right) \right)^{- \frac{1}{k} - 1}, \quad k \neq 0 \\
+\frac{1}{\sigma} \exp \left( \frac{y - \mu}{\sigma} \right), \quad k = 0
+\end{cases}
+$$
+be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
+In PSIS this distribution is fit to the $M$ largest samples of $r_s$ where $M = \min (S/5, 3\sqrt{S})$ (heuristic).
+The fitted value $\hat k$ is reported, and these $M$ largest $r_s$ are replaced by their expected value under the Pareto distribution.
+All weights are truncated by the raw weight maximum $\max(r_s)$
+
+Can be thought of as a Bayesian version of IS with a prior on the largest importance ratios.
+Comment: this is interesting, but seems somewhat suspicious to  me as to why the Bayesian prior would only apply to the largest ratios.
+I suppose you could claim that we have genuine prior information that the weights shouldn't be that big.
+
+This procedure is claimed to provide better estimates than plain IS or truncated weights IS.
+But it can also be used as a diagnostic.
+
 # Maximum mean discrepancy
+
+See [these](http://wittawat.com/assets/talks/ml_meetup_bkk2018.pdf) slides by Wittawat Jitkrittum.
+Seth suggests starting with MMD with Gaussian kernel and median heuristic to pick length-scale.
 

### 2022-11-01 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Template for BayesComp poster

---
main_topsize: 0.2
main_bottomsize: 0.075
poster_height: "48in"
poster_width: "36in"
font_family: "Open Sans"
# Essentials
title: Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV model
author:
  - name: '**Adam Howes**'
    affil: 1, 2
    main: true
    twitter: adamhowes
    email: ath19@ic.ac.uk
  - name: "Alex Stringer"
    affil: 3
  - name: "Seth R. Flaxman"
    affil: 4
  - name: "Jeffrey W. Eaton"
    affil: 2
affiliation:
  - num: 1
    address: Department of Mathematics, Imperial College London
  - num: 2
    address: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
  - num: 3
    address: Department of Statistics and Actuarial Science, University of Waterloo
  - num: 4
    address: Department of Computer Science, University of Oxford
main_findings:
  - "Combining **simplified INLA**, automatic differentiation and **adaptive Gauss-Hermite quadrature** for fast and accurate inference in a challenging, real-world problem"
logoleft_name: "IMP_ML_W_N.png"
# logoright_name: "UNAIDS-reverse-EN.gif"
# Style & formatting
column_numbers: 3
title_textsize: "60px"
author_textsize: "1.17em"
authorextra_textsize: "35px"
affiliation_textsize: "25px"
affiliation_textcol: '#00000080'
caption_fontsize: "20px"
# Main
main_fontfamily: "Open Sans"
# Poster body options
primary_colour: '#38761C'
secondary_colour: '#3b77D8'
accent_colour: "#5d19c4"
body_bgcol: "#ffffff"
body_textsize: "40px"
body_textcol: "#000000"
reference_textsize: "25px"
sectitle_textsize: "40px"
# Standard options
bibliography: citations.bib
output: 
  posterdown::posterdown_betterport:
    self_contained: false
    pandoc_args: --mathjax
    highlight: espresso
    number_sections: false
---

<!-- Using LaTeX to create spacing -->
$$\\[0.15in]$$

## The Naomi model

## Inference procedure

## Comparison

## Conclusions

<font size="6"> **Funding** AH was supported by the EPSRC and Bill & Melinda Gates Foundation. This research was supported by the MRC Centre for Global Infectious Disease Analysis. </font>

## References

<div id="refs" class="references"></div>

### 2022-11-01 utils/archive_to_docs.R

@@ -29,3 +29,4 @@ archive_to_docs("dev_sinla", i = 2)
 archive_to_docs("explore_aghq")
 archive_to_docs("example_naomi")
 archive_to_docs("explore_posterior-comparison")
+archive_to_docs("docs_bayescomp-poster")

### 2022-11-01 src/explore_posterior-comparison/posterior-comparison.Rmd
Bibliography header
Small update to posterior-comparison

@@
 # Data
+
+Throughout this notebook we will use exemplar data from the report [`dev_sinla`](https://github.com/athowes/elgm-inf/tree/master/src/dev_sinla).
 
@@
 
+## Example from Talts et al.
+
+Let's have a go at running the `R-INLA` section from @talts2018validating as described in the GitHub repository `seantalts/simulation-based-calibration`.
+
+```{r}
+load("sbc-talts.rdata")
+
+#' To-do...
+```
+
 # Pareto smoothed importance sampling
@@
 
+The `psis` function in the `loo` package takes as input some collection of importance ratios on the log scale as `log_ratios`.
+
 # Maximum mean discrepancy
@@
 
-# Original computing environment
+# Original computing environment {-}
 
@@
 ```
+
+# Bibliography {-}

### 2022-11-01 src/dev_sinla/sinla.Rmd
Rename model1index to model1_index

@@
 ```{r}
-compile("model1index.cpp")
-dyn.load(dynlib("model1index"))
+compile("model1_index.cpp")
+dyn.load(dynlib("model1_index"))
 ```
 
-The `prepare_dat` function takes `dat` and makes it compatible with `model1index` for particular choice of marginal `i`.
+The `prepare_dat` function takes `dat` and makes it compatible with `model1_index` for particular choice of marginal `i`.
 
@@
   random = "phi_prev_minus_i",
-  DLL = "model1index",
+  DLL = "model1_index",
   silent = TRUE
@@
   random = "phi_prev_minus_i",
-  DLL = "model1index",
+  DLL = "model1_index",
   silent = TRUE,
@@
     random = "phi_prev_minus_i",
-    DLL = "model1index",
+    DLL = "model1_index",
     silent = TRUE,
@@
 
-## `model1index.cpp`
+## `model1_index.cpp`
 
-```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines('model1index.cpp')}
+```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines('model1_index.cpp')}
 ```

### 2022-11-01 src/dev_sinla/functions.R

@@ -80,8 +80,8 @@ run_analysis <- function(sim_data, analysis_name) {
     dat[c("n", "y_prev_i", "y_prev_minus_i", "m_prev_i", "m_prev_minus_i")]
   }
 
-  compile("model1index.cpp")
-  dyn.load(dynlib("model1index"))
+  compile("model1_index.cpp")
+  dyn.load(dynlib("model1_index"))
 
   xi_laplace_marginal <- function(i, opt_theta, finegrid) {
     dat_i <- prepare_dat(dat, i)
@@ -100,7 +100,7 @@ run_analysis <- function(sim_data, analysis_name) {
       data = dat_i,
       parameters = param_i_fixed_theta,
       random = "phi_prev_minus_i",
-      DLL = "model1index",
+      DLL = "model1_index",
       silent = TRUE,
       map = map_fixed_theta
     )

### 2022-11-01 src/explore_posterior-comparison/posterior-comparison.Rmd
Trying to add model1_icar_index, and setting up dev-sinla to work well with this

@@
 ```{r}
-samples <- readRDS("depends/samples-m250.rds")
+samples <- readRDS("depends/model1-samples-m250.rds")
 samples_mcmc <- samples$mcmc

### 2022-11-01 src/dev_sinla/model1_icar_index.cpp

@@ -1,0 +1,57 @@
+// model1_icar_index.cpp
+
+#include <TMB.hpp>
+
+template <class Type>
+Type objective_function<Type>::operator()()
+{
+  // Data block
+  DATA_INTEGER(n);       // Number of regions
+  DATA_INTEGER(i);       // Index i
+  DATA_VECTOR(y_prev);   // Survey response
+  DATA_VECTOR(m_prev);   // Survey sample size
+  DATA_SPARSE_MATRIX(Q); // ICAR structure matrix
+  DATA_SCALAR(Q_rank);   // Rank of Q
+
+  // Parameter block
+  PARAMETER(beta_prev);               // Survey intercept
+  PARAMETER(phi_prev_i);              // Survey spatial effect for index i
+  PARAMETER_VECTOR(phi_prev_minus_i); // Survey spatial effect for index -i
+  PARAMETER(log_sigma_phi_prev);      // Survey log standard deviation of spatial effects
+
+  // Transformed parameters block
+  Type sigma_phi_prev = exp(log_sigma_phi_prev);
+
+  vector<Type> phi_prev(n);
+  int k = 0;
+  for (int j = 0; j < n; j++) {
+    if (j == i) {
+      phi_prev(j) = phi_prev_i;
+    } else {
+      phi_prev(j) = phi_prev_minus_i(k);
+      k++;
+    }
+  }
+
+  vector<Type> eta_prev(beta_prev + sigma_phi_prev * phi_prev);
+
+  // Initialise negative log-likelihood
+  Type nll(0.0);
+
+  // Priors
+  nll -= dnorm(sigma_phi_prev, Type(0), Type(2.5), true) + log_sigma_phi_prev;
+  nll -= dnorm(beta_prev, Type(-2), Type(1), true);
+
+  // ICAR spatial
+  nll -= Q_rank * 0.5 * log_sigma_phi_prev - 0.5 * exp(log_sigma_phi_prev) * (phi_prev * (Q * phi_prev)).sum();
+  nll -= dnorm(phi_prev.sum(), Type(0.0), Type(0.001) * phi_prev.size(), true); // Soft sum-to-zero constraint
+
+  // Likelihood
+  nll -= dbinom_robust(y_prev, m_prev, eta_prev, true).sum();
+
+  // Generated quantities block
+  vector<Type> rho_prev(invlogit(eta_prev));    // Posterior prevalence estimates
[…]

### 2022-11-01 src/epil/epil.Rmd
Add appendix and OCE to epil
Add citations to epil

@@
     theme: lumen
+bibliography: citations.bib
 abstract: |
-  **Background** Rue, Martino and Chopin (2009) use the epilespy example from the OpenBUGs manual as demonstration.
+  **Background** @rue2009approximate use the epilespy example from the OpenBUGs manual as demonstration.
 
@@
 
-Section 5.2. of Rue, Martino and Chopin (2009) demonstrates the application of \texttt{R-INLA} to a generalised linear mixed model.
+Section 5.2. of @rue2009approximate demonstrates the application of \texttt{R-INLA} to a generalised linear mixed model.
 The data, previously included as an example in the OpenBUGs manual, concern the results of a clinical trial for a new epilepsy drug.
@@
 ```
+
+# Appendix
+
+## `epil.cpp`
+
+```{cpp, echo=TRUE, eval=FALSE, output.var="ex1", code=readLines('epil.cpp')}
+```
+
+## `epil.stan`
+
+```{cpp, echo=TRUE, eval=FALSE, output.var="ex2", code=readLines('epil.stan')}
+```
+
+# Original computing environment {-}
+
+```{r}
+sessionInfo()
+```
+
+# Bibliography {-}

### 2022-11-02 src/explore_aghq/aghq.Rmd
Add pdf output of AGHQ examples

@@
   ggplot(aes(x = x, y = y)) +
-    geom_line() +
+    geom_line(col = cbpalette[1]) +
     labs(x = "x", y = "f(x)") +
@@
 
-plot +
+plot <- plot +
   geom_point(
     data = data.frame(nodes = gg$nodes, weights = gg$weights, y = 0), 
-    aes(x = nodes, y = y, size = weights),
-    col = cbpalette[2]
+    aes(x = nodes, y = y, size = weights), alpha = 0.5
   ) +
+  labs(size = "Weight") +
   scale_size_continuous(limits = c(1, 2))
+
+plot
+
+ggsave(filename = "aghq-1d.pdf", plot = plot, height = 4, width = 6.25)
 ```
@@
   ) %>%
-  ggplot(aes(x = x1, y = x2, fill = y)) +
-    geom_tile() +
+  ggplot() +
+    geom_contour(aes(x = x1, y = x2, z = y), col = multi.utils::cbpalette()[1]) +
     theme_minimal()
@@
 ```{r}
-plot2 +
+plot2 <- plot2 +
   geom_point(
@@
             mutate(weights = mvQuad::getWeights(gg2)), 
-    aes(x = V1, y = V2, size = weights, fill = 1),
-    col = cbpalette[2]
+    aes(x = V1, y = V2, size = weights),
+    alpha = 0.5
   ) +
-  labs(col = "f(x1, x2)") +
+  labs(col = "f(x1, x2)", size = "Weight") +
   scale_size_continuous(limits = c(2, 4))
+
+plot2
+
+ggsave(filename = "aghq-2d.pdf", plot = plot, height = 4, width = 6.25)
 ```

### 2022-11-02 src/explore_aghq/walkthrough.Rmd
Add pdf output of AGHQ examples

@@
 
-pdfandcdf <- compute_pdf_and_cdf(quad$marginals[[i]], interpolation = interpolation)
+pdf_and_cdf <- compute_pdf_and_cdf(quad$marginals[[i]], interpolation = interpolation)
 
@@
 ```{r}
-for (i in 1:length(q)) out[i] <- pdfandcdf$theta[max(which(pdfandcdf$cdf < q[i]))]
+for (i in 1:length(q)) out[i] <- pdf_and_cdf$theta[max(which(pdf_and_cdf$cdf < q[i]))]
 

### 2022-11-02 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Iterate seminar a little (e.g. include AGHQ plots)

@@
 
-* Consider areas $i = 1, \ldots, n$
-* Simple random sample household-survey taken in each area, with sample sizes $m^\text{HS}_i$
+* Consider small-areas $i = 1, \ldots, n$
+* Simple random sample household-survey in each area, with sample sizes $m^\text{HS}_i$
 * The number of people testing positive is $y^\text{HS}_i$
@@
 
-* So what is `TMB`?
 * R package which implements the Laplace approximation for latent variable models using automatic differentiation (via `CppAD`)
-  * For more about AD see e.g. @griewank2008evaluating
-  * Useful for getting the mode, Hessian
+  * For more about AD see e.g. @griewank2008evaluating. It's useful for getting the mode, Hessian
 * Write an objective function $f(\x, \btheta)$ in C++ ("user template")
@@
 
-* Suggested reading: the paper [@rue2009approximate] or a book e.g. @blangiardo2015spatial
+* Suggested reading: the paper @rue2009approximate or a book e.g. @blangiardo2015spatial
 * Approximate Bayesian inference for \textcolor{hilit}{latent Gaussian models} (LGMs), which recall are three-stage models with middle layer
@@
 2) In both \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. \eqref{eq:hypermarginal}, so choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$
-  * For low $m$ `R-INLA` uses a grid-strategy (illustrated in the next slide)
-  * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design is used
+  * For low $m$ `R-INLA` uses a grid-strategy
+  * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design
   * Other approaches, like adaptive Gaussian Hermite quadrature (AGHQ) have recently been shown to have theoretical guarantees [@bilodeau2021stochastic] and may be preferable
+
+# Grid strategy
+
+* To write...
 
@@
 
-* `aghq` R package and vignette [@stringer2021implementing]
 * Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
-* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior)
-* Use $k$ quadrature points
-  * If $k$ is odd then they include the mode
+* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) to automatically choose the node locations
+  * No manual tuning!
+* Works particularly well when the integrand is pretty Gaussian
+* Supposing you use $k$ quadrature points, then
   * If $k = 1$ then it's a Laplace approximation
   * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often
+* Implemented in the `aghq` R package. See vignette @stringer2021implementing
+
+#
+
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("One dimensional example of AGHQ. \\label{fig:aghq-1d}")}
+knitr::include_graphics("depends/aghq-1d.pdf")
+```
+
+#
+
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Two dimensional example of AGHQ. \\label{fig:aghq-2d}")}
+knitr::include_graphics("depends/aghq-2d.pdf")
+```
 

### 2022-11-02 src/docs_15-11-22-seminar/script.R

@@ -2,24 +2,6 @@
 # orderly::orderly_develop_start("docs_15-11-22-seminar")
 # setwd("src/docs_15-11-22-seminar")
 
-sf_lightgrey <- "#E6E6E6"
-lightgrey <- "#D3D3D3"
-
-lightblue <- "#56B4E9"
-lightgreen <- "#009E73"
-lightpink <- "#CC79A7"
-light_palette <- c(lightblue, lightgreen, lightpink)
-
-midblue <- "#3D9BD0"
-midgreen <- "#00855A"
-midpink <- "#B3608E"
-mid_palette <- c(midblue, midgreen, midpink)
-
-darkblue <- "#004E83"
-darkgreen <- "#00380D"
-darkpink <- "#802D5B"
-dark_palette <- c(darkblue, darkgreen, darkpink)
-
 cbpalette <- multi.utils::cbpalette()
 
 rmarkdown::render("15-11-22-seminar.Rmd")

### 2022-11-02 src/epil/epil.Rmd
Update epil writing and add Fig 3. (a) from INLA 2009 paper

@@
 
-Section 5.2. of @rue2009approximate demonstrates the application of \texttt{R-INLA} to a generalised linear mixed model.
+## Model description
+
+Section 5.2. of @rue2009approximate demonstrates the application of `R-INLA` to a generalised linear mixed model.
 The data, previously included as an example in the OpenBUGs manual, concern the results of a clinical trial for a new epilepsy drug.
@@
 
+## Data preparation
+
+The data is available within `R-INLA` as follows:
+
 ```{r}
@@
 
-Start by centering the covariates:
+We modify `Epil` by centering all of the covariates:
 
@@
 
-Implement the model based upon a response variable of length $N \times J$ and a model matrix with $K$ predictors (including the intercept term):
+Implement the model based upon a response variable of length $N \times J$, where $N$ is the number of patients and $J$ is the number of visits to the clinic, and a model matrix with $K$ predictors (including the intercept term):
 
@@
 
-For the individual specific random effect $\epsilon_i$ we use a transformation matrix $\bm{E}$ which repeats elements of vector of length `N` each `J` times, for example:
+For the individual specific random effect $\epsilon_i$ we use a transformation matrix `E` which repeats elements of vector of length `N` each `J` times.
 
@@
 }
+```
 
+For example, with $N = 3$ and $J = 2$:
+
+```{r}
 E <- make_epsilon_matrix(N = 3, J = 2)
@@
 
+We consider a range of inference methods, described below.
+
 ## Stan
 
-Hamiltonian Monte Carlo via the No-U-Turn algorithm, with one Markov chain of `iter = 1000` samples, and `warmup = 100` burn-in:
+First, Hamiltonian Monte Carlo via the No-U-Turn algorithm, with four Markov chains of `iter = 1000` samples, and `warmup = 100` burn-in:
 
@@
   data = dat,
-  chains = 1,
+  chains = 4,
   warmup = 100,
@@
 The variable `Epil$rand` is gives the row number for each entry and the variable `Epil$Ind` gives the patient number.
-These variables can be used to define the random effects $\nu_{ij}$ and $\epsilon_i$ in INLA as follows.
+These variables can be used to define the IID random effects $\nu_{ij}$ and $\epsilon_i$ in `R-INLA` as follows.
 The usual R formula notation (e.g. `y ~ x1 + x2`) is used for the rest of the linear predictor.
@@
 formula <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT +
-  f(rand, model = "iid", hyper = tau_prior) +  # Nu random effect
-  f(Ind,  model = "iid", hyper = tau_prior)    # Epsilon random effect
+  f(rand, model = "iid", hyper = tau_prior) +  #' Nu random effect
+  f(Ind,  model = "iid", hyper = tau_prior)    #' Epsilon random effect
 ```
 
-We can use various settings for the strategy, which controls how the latent field posterior marginals are approximated:
+We can use various settings for the strategy, which controls the approximation used for the latent field posterior marginals:
 
@@
   inla(formula,
-       control.fixed = list(mean = 0, prec = 1/100^2), # Beta prior
+       control.fixed = list(mean = 0, prec = 1 / 100^2), #' Beta prior
        family = "poisson",
@@
 
-We integrate `random` out with a Laplace approximation:
+We integrate `random` out, which here corresponds to `c("epsilon", "nu")`, with a Laplace approximation:
 
@@
 
-For completeness, we check here that the objective functions as implemented in Stan and `TMB` are identical.
+### Check `TMB` objective is the same as Stan
+
+For completeness, we check here that the objective functions as implemented in `TMB` and Stan are identical.
 To obtain the `TMB` negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
@@
 ) %>%
-  mutate(Difference = TMB - Stan)
+  mutate(Difference = TMB - Stan) %>%
+  pull(Difference)
 ```
@@
 
-[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in `TMB` without writing the C++ code.
-For example, rather than writing `epil.cpp` manually, we could have called the following:
+[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in `TMB` without writing the C++ code manually as we have in `epil.cpp`.
+For example, rather than writing `epil.cpp`, we could have called the following:
 
@@
 
-This likely isn't the same model, in terms of priors and so on, but it may be close.
+This won't be exactly the same model, in terms of priors and so on, but it may be close.
 
@@
 ```{r}
-fit7 <- tmbstan(obj = obj, chains = 1)
+fit7 <- tmbstan(obj = obj, chains = 4)
 ```
@@
 [`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer which implements adaptive Gauss-Hermite quadrature.
-One (strange!) approach is to use `glmmTMB` to get the `TMB` template which can be passed the `aghq`.
+It also works smoothly with `TMB` templates.
+One (odd!) approach would be to use `glmmTMB` to get the `TMB` template which can be passed the `aghq`.
 
@@
   obj,
-  k = 2,
+  k = 3,
   startingvalue = c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
@@
 rownames(df) <- c(sapply(0:5, beta_i))
-colnames(df) <- c("Stan", "INLA_G", "INLA_SL", "INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "type")
+colnames(df) <- c("Stan", "R-INLA_G", "R-INLA_SL", "R-INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "type")
 
-kableExtra::kable(df)
+df %>%
+  select(-type) %>%
+  kableExtra::kable()
 
 saveRDS(df, "comparison-results.rds")
+```
+
+## Focus on $\beta_0$
+
+```{r}
+plot_pdf <- ggplot(data.frame(x = rstan::extract(fit1, pars = "beta[1]")[[1]]), aes(x = x)) +
+    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7], bins = 60) +
+    theme_minimal() +
+    labs(x = "beta[1]", y = "Posterior PDF")
+
+beta0_inla_marginals <- bind_rows(
+  data.frame(fit2$marginals.fixed$`(Intercept)`) %>%
+    mutate(method = "G"),
+  data.frame(fit3$marginals.fixed$`(Intercept)`) %>%
+    mutate(method = "SL"),
+  data.frame(fit4$marginals.fixed$`(Intercept)`) %>%
+    mutate(method = "L")
+)
+
+plot_pdf +
+  geom_line(data = beta0_inla_marginals, aes(x = x, y = y, col = method)) +
+  scale_color_manual(values = cbpalette)
 ```
@@
     geom_line() +
+    scale_color_manual(values = cbpalette)
     theme_minimal()

### 2022-11-02 src/explore_posterior-comparison/posterior-comparison.Rmd
Add KS to posterior-comparison

@@
 Compute the empirical CDFs, and look for the maximum distance between them.
+
+```{r}
+n <- length(unique(samples_mcmc$index))
+
+ks_gaussian_mcmc <- lapply(1:n, function(i) {
+  samples_gaussian_i <- filter(samples_gaussian, index == i)$value
+  samples_mcmc_i <- filter(samples_mcmc, index == i)$value
+  inf.utils::ks_test(samples_gaussian_i, samples_mcmc_i)
+})
+
+ks_laplace_mcmc <- lapply(1:n, function(i) {
+  samples_laplace_i <- filter(samples_laplace, index == i)$value
+  samples_mcmc_i <- filter(samples_mcmc, index == i)$value
+  inf.utils::ks_test(samples_laplace_i, samples_mcmc_i)
+})
+
+ks_results <- bind_rows(
+  bind_rows(ks_gaussian_mcmc, .id = "index") %>%
+    mutate(type = "gaussian"),
+  bind_rows(ks_laplace_mcmc, .id = "index") %>%
+    mutate(type = "laplace")
+)
+
+ks_results %>%
+  select(index, D, type) %>%
+  pivot_wider(
+    names_from = type,
+    values_from = D
+  ) %>%
+  ggplot(aes(x = laplace, y = gaussian)) +
+    geom_point(alpha = 0.5) +
+    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
+    lims(x = c(0, 1), y = c(0, 1)) +
+    labs(x = "KS(laplace, mcmc)", y = "KS(gaussian, mcmc)") +
+    theme_minimal()
+
+ks_results %>%
+  group_by(type) %>%
+  summarise(mean = mean(D)) %>%
+  kable()
+```
 

### 2022-11-03 src/epil/epil.Rmd
Separate C++ for nu and epsilon are compiling now. The one together with a data flag still giving me trouble
epil_index now compiling (it doesn't like it when nu and epsilon are defined differently depending on the if statement)
Trying to do Laplace for epil with TMB

@@
 ```{r}
-fit7 <- tmbstan(obj = obj, chains = 4)
+fit7 <- tmbstan(obj = obj, chains = 4, refresh = 0)
 ```
@@
 summary8 <- summary(fit8)
+```
+
+## Custom Laplace method with `aghq`
+
+```{r}
+compile("epil_index.cpp")
+dyn.load(dynlib("epil_index"))
+```
+
+We're going to use a data structure to pass to `epil_index`.
+A `toggle` value of `0` corresponds to `nu` and a toggle value of `1` corresponds to `epsilon`:
+
+```{r}
+index_controller <- data.frame(
+  "toggle" = c(rep(0, dat$N * dat$J),  rep(1, dat$N)),
+  "i" = c(1:(dat$N * dat$J), 1:dat$N)
+)
+```
+
+We have already obtained the MAP estimate $\hat \theta$ in the `TMB` section above that we can use in our empirical Bayes procedure.
+
+```{r}
+opt_theta <- opt
+```
+
+Eventually create function of the form:
+
+```{r}
+xi_laplace_marginal <- function(i, opt_theta, finegrid) {
+  #' To fill
+}
+```
+
+Start with `i = 1` for the first marginal:
+
+```{r}
+i <- 1
+
+#' I don't know if this is the right grid for this application...
+fine_grid <- seq(-5, 5, length.out = 1000)
+
+dat$toggle <- index_controller[i, "toggle"]
+dat$i <- index_controller[i, "i"]
+
+theta_names <- unique(names(opt_theta$par))
+map_fixed_theta <- list()
+param_fixed_theta <- param
+
+for(theta in theta_names) {
+  map_fixed_theta[[theta]] <- factor(NA)
+  param[[theta]] <- opt_theta$par[[theta]]
+}
+
+#' Prepare the initialisation parameters if it's nu that we're toggling
+if(dat$toggle == 0) {
+  param_fixed_theta[["nu_i"]] <- param_fixed_theta$nu[i]
+  param_fixed_theta[["nu_minus_i"]] <- param_fixed_theta$nu[-i]
+  # param_fixed_theta$nu <- NULL
+}
+
+#' Or if it's epsilon that we're toggling
+if(dat$toggle == 1) {
+  param_fixed_theta[["epsilon_i"]] <- param_fixed_theta$epsilon[i]
+  param_fixed_theta[["epsilon_minus_i"]] <- param_fixed_theta$epsilon[-i]
+  # param_fixed_theta$epsilon <- NULL
+}
+
+get_random <- function(dat) {
+  if(dat$toggle == 0) return(c("nu_minus_i", "epsilon"))
+  if(dat$toggle == 1) return(c("nu", "epsilon_minus_i"))
+}
+
+obj_fixed_theta <- MakeADFun(
+  data = dat,
+  parameters = param_fixed_theta,
+  random = get_random(dat),
+  DLL = "epil_index",
+  silent = TRUE,
+  map = map_fixed_theta
+)
+
+quad <- aghq::aghq(
+  ff = obj_fixed_theta,
+  k = 3,
+  startingvalue = 0,
+  control = aghq::default_control_tmb()
+)
+
+pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = finegrid)[[1]]
+
 ```

### 2022-11-04 src/explore_posterior-comparison/posterior-comparison.Rmd
Add SBC intuition and example in Stan code

@@
   summarise(mean = mean(D)) %>%
-  kable()
+  knitr::kable()
 ```
@@
 
+## Intuition
+
+There are two procedures you could use to generate samples from the joint distribution.
+Firstly, you could sample from the prior on $\theta$ first:
+$$
+\tilde \theta_1 \sim p(\theta) \\
+\tilde y_1 \sim p(y \, | \, \tilde \theta_1) \\
+\implies (\tilde \theta_1, \tilde y_1) \sim p(\theta, y)
+$$
+Secondly, you could sample from the "prior" on $y$ first:
+$$
+\tilde y_2 \sim p(y) \\
+\tilde \theta_2 \sim p(\theta \, | \, \tilde y_2) \\
+\implies (\tilde \theta_2, \tilde y_2) \sim p(\theta, y)
+$$
+Both will generate samples from the joint distribution $p(\theta, y)$.
+This implies that if you look at the draws from procedure one $\{ \tilde \theta_1\}$ and the draws from procedure two $\{ \tilde \theta_2\}$ then they have the same distribution.
+
+Importantly, **procedure two involves computing a posterior distribution**, and therefore the samples will only be the same if the Bayesian inference procedure is exact.
+It's this fact which SBC uses to create a procedure for checking implementation and accuracy of a given approximate Bayesian inference method.
+
 ## Example from Talts et al.
@@
 ```
+
+## Example from Stan manual
+
+Consider the model:
+
+$$
+y_i \sim \mathcal{N}(\mu, \sigma^2) \\
+\mu \sim \mathcal{N}(0 , 1) \\
+\sigma \sim \text{lognormal}(0, 1)
+$$
+
+Want `M` samples.
+
+```{r}
+M <- 1000
+
+n_chains <- 4
+n_iter <- 2000
+
+fit <- rstan::stan(file = "sbc.stan", iter = n_iter, chains = n_chains, refresh = 0)
+```
+
+Check that all ESS are greater than `M`:
+
+```{r}
+all(summary(fit)$summary[, "n_eff"] > M)
+```
+
+Extract samples and thin down to just `M`:
+
+```{r}
+samples <- extract(fit)
+samples <- lapply(samples, function(sample) sample[1:M * floor(length(sample) / M)])
+```
+
+So, the quantiles (you probably want to be more careful about ranks than I'm being here) of the simulated data within the posterior are:
+
+```{r}
+sum(samples$mu_lt_sim) / length(samples$mu_lt_sim)
+sum(samples$sigma_lt_sim) / length(samples$sigma_lt_sim)
+```
+
+These should be $\mathcal{U}(0, 1)$, so by repeating this process many times (it looks like it's a quite costly procedure!) we can assess whether this is in fact true.
 

### 2022-11-04 src/dev_sinla/sinla.Rmd
Add resources
Develop non-empirical Bayes method using modesandhessians

@@
 ```{r}
-h_norm <- function(theta) h$fn(theta) - quad$normalized_posterior$lognormconst
+h_norm <- function(theta) h$fn(theta) + quad$normalized_posterior$lognormconst
 ```
@@
 
+We're going to start by normalising $\tilde p_\text{LA}(\theta, y)$.
+Remember that we did this above in the empirical Bayes case, but found that it wasn't strictly necessary.
+
 ```{r}
-#' To-do
+h$he <- function(theta) numDeriv::jacobian(h$gr, theta, method = "Richardson")
+quad <- aghq::aghq(ff = h, k = 3, startingvalue = c(0, 0), control = aghq::default_control_tmb())
+quad$normalized_posterior$lognormconst
+h_norm <- function(theta) h$fn(theta) + quad$normalized_posterior$lognormconst
 ```
+
+We can get the mode and Hessian of the random effects at each quadrature point as follows:
+
+```{r}
+distinctthetas <- quad$normalized_posterior$nodesandweights[, grep('theta', colnames(quad$normalized_posterior$nodesandweights))]
+modesandhessians <- distinctthetas
+thetanames <- make.unique(names(h$par), sep = "")
+colnames(modesandhessians) <- thetanames
+
+modesandhessians$mode <- vector(mode = "list", length = nrow(distinctthetas))
+modesandhessians$H <- vector(mode = "list", length = nrow(distinctthetas))
+
+for(i in 1:nrow(distinctthetas)) {
+  theta <- as.numeric(modesandhessians[i, thetanames])
+  h$fn(theta)
+  mm <- h$env$last.par
+  modesandhessians[i, "mode"] <- list(list(mm[h$env$random]))
+  H <- h$env$spHess(mm, random = TRUE)
+  H <- rlang::duplicate(H)
+  modesandhessians[i, "H"] <- list(list(H))
+}
+
+modesandhessians$weights <- quad$normalized_posterior$nodesandweights$weights
+```
+
+Now we want to evaluate
+$$
+\tilde p(x_i \, | \, y) &= \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
+$$
+To start with, let's evaluate this for the first value of $z$ (actually I don't really know what $z$ runs over, but this is just the first weight):
+
+```{r}
+x_test_value <- 0
+
+z <- 1
+i <- 1
+
+#' This is the weight, w_k(z)
+modesandhessians$weights[z]
+
+#' This is the normalised Laplace approximation to log(p(theta | y))
+theta <- as.numeric(modesandhessians[z, thetanames])
+- h_norm(theta)
+
+#' This is the normalised Gaussian approximation to log(p(x_i | theta, y))
+#' Let's say that we're doing i = 1 (corresponding to x_i)
+mean_i <- as.numeric(modesandhessians$mode[[z]][i])
+Sigma <- solve(modesandhessians$H[[z]])
+sd_i <- as.numeric(sqrt(Sigma[i, i]))
+
+dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE)
+
+#' So the value we would compute, on the log scale, is
+dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z])
+```
+
+Remember that we're dealing with the logarithm versions, so let's exp-sum-log for the part inside the sum:
+$$
+\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log w_k(z) \right)
+$$
 
@@
 
+## Resources
+
+* [INLA and other approaches to GAMs](https://www.maths.ed.ac.uk/~swood34/inla-etc.pdf)
+* [mgcv::ginla](https://github.com/cran/mgcv/blob/d9fe307c8ed1f8f1ed436c85e4ac415c1febeef4/R/inla.r)
+
 ## Comparison of notation

### 2022-11-07 src/epil/epil.Rmd
Render epil

@@
 
-obj_fixed_theta <- MakeADFun(
-  data = dat,
-  parameters = param_fixed_theta,
-  random = get_random(dat),
-  DLL = "epil_index",
-  silent = TRUE,
-  map = map_fixed_theta
-)
-
-quad <- aghq::aghq(
-  ff = obj_fixed_theta,
-  k = 3,
-  startingvalue = 0,
-  control = aghq::default_control_tmb()
-)
-
-pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = finegrid)[[1]]
-
+# obj_fixed_theta <- MakeADFun(
+#   data = dat,
+#   parameters = param_fixed_theta,
+#   random = get_random(dat),
+#   DLL = "epil_index",
+#   silent = TRUE,
+#   map = map_fixed_theta
+# )
+#   
+# quad <- aghq::aghq(
+#   ff = obj_fixed_theta,
+#   k = 3,
+#   startingvalue = 0,
+#   control = aghq::default_control_tmb()
+# )
+# 
+# pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = finegrid)[[1]]
 ```

### 2022-11-07 src/epil/epil_index.cpp

@@ -6,7 +6,7 @@ template <class Type>
 Type objective_function<Type>::operator()()
 {
   // Data block
-  DATA_INTEGER(toggle) // 0 corresponds to epsilon, 1 to nu
+  DATA_INTEGER(toggle) // 0 corresponds to nu, 1 to epsilon
   DATA_INTEGER(i); // Index i
   DATA_INTEGER(N);
   DATA_INTEGER(J);
@@ -37,7 +37,6 @@ Type objective_function<Type>::operator()()
         k++;
       }
     }
-
   }
 
   // If toggle == 1 then set-up minus i for epsilon

### 2022-11-07 src/epil/epil.Rmd
Get nu Laplace marginals working

@@
 
+### Using multiple C++ template
+
+Let's just starting by trying to get the Laplace marginals for $\nu$:
+
+```{r}
+compile("epil_nu_index.cpp")
+dyn.load(dynlib("epil_nu_index"))
+```
+
+We have already obtained the MAP estimate $\hat \theta$ in the `TMB` section above that we can use in our empirical Bayes procedure.
+
+```{r}
+opt_theta <- opt
+```
+
+Eventually create function of the form:
+
+```{r}
+xi_laplace_marginal_nu <- function(i, opt_theta, finegrid) {
+  #' To fill
+}
+```
+
+Start with `i = 1` for the first marginal:
+
+```{r}
+i <- 1
+
+#' I don't know if this is the right grid for this application...
+fine_grid <- seq(-5, 5, length.out = 1000)
+
+dat$i <- i
+
+theta_names <- unique(names(opt_theta$par))
+map_fixed_theta <- list()
+
+param_fixed_theta <- param
+
+for(theta in theta_names) {
+  map_fixed_theta[[theta]] <- rep(factor(NA), length(param[[theta]]))
+  param_fixed_theta[[theta]] <- opt_theta$par[names(opt_theta$par) == theta]
+}
+
+#' Prepare the initialisation parameters
+param_fixed_theta[["nu_i"]] <- param_fixed_theta$nu[i]
+param_fixed_theta[["nu_minus_i"]] <- param_fixed_theta$nu[-i]
+param_fixed_theta[["nu"]] <- NULL
+
+#' Might need this here if you're getting crashes
+# f <- parallel::mcparallel({})
+# x <- parallel::mccollect(f)
+
+obj_fixed_theta <- MakeADFun(
+  data = dat,
+  parameters = param_fixed_theta,
+  random = c("nu_minus_i", "epsilon"),
+  DLL = "epil_nu_index",
+  silent = TRUE,
+  map = map_fixed_theta
+)
+
+quad <- aghq::aghq(
+  ff = obj_fixed_theta,
+  k = 3,
+  startingvalue = 0,
+  control = aghq::default_control_tmb()
+)
+
+pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = fine_grid)[[1]]
+
+ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "nu[1]", y = "Posterior")
+```
+
+### Using one C++ template
+
 ```{r}

### 2022-11-07 src/epil/epil.Rmd
Can create obj_fixed_theta using epil_index now (but then getting "initial value in 'vmmin' is not finite" type errors)

@@
 
-Eventually create function of the form:
-
-```{r}
-xi_laplace_marginal_nu <- function(i, opt_theta, finegrid) {
-  #' To fill
-}
-```
-
 Start with `i = 1` for the first marginal:
@@
 for(theta in theta_names) {
-  map_fixed_theta[[theta]] <- factor(NA)
-  param[[theta]] <- opt_theta$par[[theta]]
+  map_fixed_theta[[theta]] <- rep(factor(NA), length(param[[theta]]))
+  param_fixed_theta[[theta]] <- opt_theta$par[names(opt_theta$par) == theta]
 }
@@
 
-# obj_fixed_theta <- MakeADFun(
-#   data = dat,
-#   parameters = param_fixed_theta,
-#   random = get_random(dat),
-#   DLL = "epil_index",
-#   silent = TRUE,
-#   map = map_fixed_theta
-# )
-#   
+obj_fixed_theta <- MakeADFun(
+ data = dat,
+ parameters = param_fixed_theta,
+ random = get_random(dat),
+ DLL = "epil_index",
+ silent = TRUE,
+ map = map_fixed_theta
+)
+
 # quad <- aghq::aghq(
@@
 #  
-# pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = finegrid)[[1]]
+# pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = fine_grid)[[1]]
+# 
+# ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
+#   geom_line() +
+#   theme_minimal() +
+#   labs(x = "nu[1]", y = "Posterior")
 ```

### 2022-11-07 src/dev_sinla/sinla.Rmd
Calculate p(x_1 = x_test_value | y) looping over z

@@
 $$
-\tilde p(x_i \, | \, y) &= \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
-$$
-To start with, let's evaluate this for the first value of $z$ (actually I don't really know what $z$ runs over, but this is just the first weight):
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
+$$
+
+If you wanted to evaluate this to obtain each marginal then it would be something like (pseudo-code):
+
+```{r eval=FALSE}
+for(i in 1:n) {
+  for(x in x_vals) {
+    for(z in z_vals) {
+      #' ...
+    }
+  }
+}
+```
+
+To start with, let's evaluate this for `i = 1`, `x = x_test_value` and the first value of `z` (actually I don't really know what $z$ runs over, but this is just the first weight):
 
@@
 x_test_value <- 0
-
 z <- 1
 i <- 1
-
-#' This is the weight, w_k(z)
+```
+
+The weight $w_k(z)$ is:
+
+```{r}
 modesandhessians$weights[z]
-
-#' This is the normalised Laplace approximation to log(p(theta | y))
+```
+
+The normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ is:
+
+```{r}
 theta <- as.numeric(modesandhessians[z, thetanames])
 - h_norm(theta)
-
-#' This is the normalised Gaussian approximation to log(p(x_i | theta, y))
-#' Let's say that we're doing i = 1 (corresponding to x_i)
+```
+
+The normalised Gaussian approximation $p_\text{G}(x_i \, | \, \theta(z), y)$ on the log-scale is:
+
+```{r}
 mean_i <- as.numeric(modesandhessians$mode[[z]][i])
@@
 dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE)
-
-#' So the value we would compute, on the log scale, is
+```
+
+So, putting it together, the value we would compute on the log-scale is as follows:
+
+```{r}
 dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z])
@@
 
-Remember that we're dealing with the logarithm versions, so let's exp-sum-log for the part inside the sum:
+To obtain the unlogged version we can use:
 $$
@@
 $$
+
+```{r}
+exp(dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
+```
+
+Great -- now let's put it together and sum over `z` so that we can obtain the probability for our `x_test_value` in marginal `i`.
+We won't worry about inefficient R for loops for the time being.
+
+```{r}
+pdf <- 0
+
+for(z in 1:length(modesandhessians$weights)) {
+  theta <- as.numeric(modesandhessians[z, thetanames])
+  mean_i <- as.numeric(modesandhessians$mode[[z]][i])
+  Sigma <- solve(modesandhessians$H[[z]])
+  sd_i <- as.numeric(sqrt(Sigma[i, i]))
+  pdf <- pdf + exp(dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
+}
+
+pdf
+```
+
+Using a `fine_grid` set of values for `x_vals` for each `i` might be overly computationally expensive.
+Instead what we can do is perform another separate AGHQ to select a set of `x_vals`.
+To do this with the `aghq` R package, we could create a `TMB` template to evaluate $\tilde p(x_i \, | \, y)$, but I imagine this might be too annoying.
+Instead, we could create the `ff` object (with components `ff$fn`, `ff$gn` `ff$hn`) in R directly.
 

### 2022-11-07 src/epil/epil.Rmd
Laplace marginal for nu

@@
 
-Start with `i = 1` for the first marginal:
+Create a function to get the Laplace marginal for the $i$th element of $\nu$:
 
 ```{r}
-i <- 1
-
-#' I don't know if this is the right grid for this application...
-fine_grid <- seq(-5, 5, length.out = 1000)
-
+xi_laplace_marginal_nu <- function(i, opt_theta, finegrid) {
   dat$i <- i
@@
 
-ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
+  return(pdf_and_cdf)
+}
+```
+
+Now we can compute the Laplace marginal for all `r dat$N * dat$J` indices of $\nu$:
+
+```{r}
+fine_grid <- seq(-5, 5, length.out = 1000)
+
+laplace_nu <- lapply(1:(dat$N * dat$J), xi_laplace_marginal_nu, opt_theta = opt_theta, finegrid = fine_grid) %>%
+  bind_rows(.id = "index") %>%
+  mutate(index = as.numeric(index))
+```
+
+There are too many marginals to plot them all, so here's a random subset of size nine:
+
+```{r}
+nine_random_indices <- sample(1:(dat$N * dat$J), size = 9)
+
+laplace_nu %>%
+  filter(index %in% nine_random_indices) %>%
+  ggplot(aes(x = theta, y = pdf)) +
   geom_line() +
+  facet_wrap(~index) +
   theme_minimal() +
-  labs(x = "nu[1]", y = "Posterior")
+  labs(x = "nu", y = "Posterior")
 ```

### 2022-11-07 src/epil/epil.Rmd
Multiple random effects in one template now working -- (yeah buddy)

@@
 #' Prepare the initialisation parameters if it's nu that we're toggling
+#' We fix the nu parameter to zero, as we're using nu_i and nu_minus_i
 if(dat$toggle == 0) {
@@
   param_fixed_theta[["nu_minus_i"]] <- param_fixed_theta$nu[-i]
-  # param_fixed_theta$nu <- NULL
+  map_fixed_theta[["nu"]] <- rep(factor(NA), length(param[["nu"]]))
 }
@@
 #' Or if it's epsilon that we're toggling
+#' We fix the epsilon parameter to zero, as we're using epsilon_i and epsilon_minus_i
 if(dat$toggle == 1) {
@@
   param_fixed_theta[["epsilon_minus_i"]] <- param_fixed_theta$epsilon[-i]
-  # param_fixed_theta$epsilon <- NULL
+  map_fixed_theta[["epsilon"]] <- rep(factor(NA), length(param[["epsilon"]]))
 }
@@
 
-# quad <- aghq::aghq(
-#   ff = obj_fixed_theta,
-#   k = 3,
-#   startingvalue = 0,
-#   control = aghq::default_control_tmb()
-# )
-#  
-# pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = fine_grid)[[1]]
-# 
-# ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
-#   geom_line() +
-#   theme_minimal() +
-#   labs(x = "nu[1]", y = "Posterior")
+quad <- aghq::aghq(
+  ff = obj_fixed_theta,
+  k = 3,
+  startingvalue = 0,
+  control = aghq::default_control_tmb()
+)
+
+pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = fine_grid)[[1]]
+
+ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "nu[1]", y = "Posterior")
 ```

### 2022-11-08 src/epil/epil.Rmd
beta should be in random, then aghq and TMB match R-INLA method = "gaussian". Almost finished with this case-study

--- 2022-11-07 src/epil/epil.Rmd
@@
 
-```{r results='hide'}
+```{r}
 dat <- list(N = N, J = J, K = K, X = X, y = y, E = make_epsilon_matrix(N, J))
@@
   iter = 1000,
-  control = list(adapt_delta = 0.95)
+  control = list(adapt_delta = 0.95),
+  refresh = 0
 )
@@
 
-We integrate `random` out, which here corresponds to `c("epsilon", "nu")`, with a Laplace approximation:
+We integrate `random` out, which here corresponds to `c("beta", "epsilon", "nu")`, with a Laplace approximation:
 
@@
   parameters = param,
-  random = c("epsilon", "nu"),
+  random = c("beta", "epsilon", "nu"),
   DLL = "epil",
@@
 
-Using just a few of them, we can see that there is a difference, but that difference is a constant so doesn't matter:
+Using just a few of them, we can see that the difference between the `TMB` and Stan objectives is a constant:
 
@@
   k = 3,
-  startingvalue = c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
+  startingvalue = c(param$l_tau_epsilon, param$l_tau_nu)
 )
 
-summary8 <- summary(fit8)
+fit8_samples <- aghq::sample_marginal(fit8, M = 1000)$samps %>%
+  t() %>%
+  as.data.frame() %>%
+  inf.utils::replace_duplicate_colnames()
 ```
@@
 
-Let's just starting by trying to get the Laplace marginals for $\nu$:
+Let's just starting by trying to get the Laplace marginals for $\beta$:
 
 ```{r}
-compile("epil_nu_index.cpp")
-dyn.load(dynlib("epil_nu_index"))
+compile("epil_beta_index.cpp")
+dyn.load(dynlib("epil_beta_index"))
 ```
@@
 
-Start with `i = 1` for the first marginal:
+Create a function to get the Laplace marginal for the $i$th element of $\beta$:
 
 ```{r}
-i <- 1
-
-#' I don't know if this is the right grid for this application...
-fine_grid <- seq(-5, 5, length.out = 1000)
-
+xi_laplace_marginal_beta <- function(i, opt_theta, finegrid) {
   dat$i <- i
@@
   #' Prepare the initialisation parameters
-param_fixed_theta[["nu_i"]] <- param_fixed_theta$nu[i]
-param_fixed_theta[["nu_minus_i"]] <- param_fixed_theta$nu[-i]
-param_fixed_theta[["nu"]] <- NULL
+  param_fixed_theta[["beta_i"]] <- param_fixed_theta$beta[i]
+  param_fixed_theta[["beta_minus_i"]] <- param_fixed_theta$beta[-i]
+  param_fixed_theta[["beta"]] <- NULL
 
@@
     parameters = param_fixed_theta,
-  random = c("nu_minus_i", "epsilon"),
-  DLL = "epil_nu_index",
+    random = c("beta_minus_i", "nu", "epsilon"),
+    DLL = "epil_beta_index",
     silent = TRUE,
@@
 
-ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
+  return(pdf_and_cdf)
+}
+```
+
+Now we can compute the Laplace marginal for all `r dat$K` indices of $\beta$:
+
+```{r}
+fine_grid <- seq(-5, 5, length.out = 1000)
+
+laplace_beta <- lapply(1:dat$K, xi_laplace_marginal_beta, opt_theta = opt_theta, finegrid = fine_grid) %>%
+  bind_rows(.id = "index") %>%
+  mutate(index = as.numeric(index))
+```
+
+Here are what the marginals look like:
+
+```{r}
+laplace_beta %>%
+  ggplot(aes(x = theta, y = pdf)) +
   geom_line() +
+  facet_wrap(~index) +
   theme_minimal() +
-  labs(x = "nu[1]", y = "Posterior")
+  labs(x = "beta", y = "Posterior")
 ```
@@
 We're going to use a data structure to pass to `epil_index`.
-A `toggle` value of `0` corresponds to `nu` and a toggle value of `1` corresponds to `epsilon`:
+A `toggle` value of `0` corresponds to `beta`, a toggle value of `1` corresponds to `nu`, and a toggle value of `2` corresponds to `epsilon`:
 
@@
 index_controller <- data.frame(
-  "toggle" = c(rep(0, dat$N * dat$J),  rep(1, dat$N)),
-  "i" = c(1:(dat$N * dat$J), 1:dat$N)
+  "toggle" = c(rep(0, dat$K), rep(1, dat$N * dat$J),  rep(2, dat$N)),
+  "i" = c(1:dat$K, 1:(dat$N * dat$J), 1:dat$N)
 )
-```
-
-We have already obtained the MAP estimate $\hat \theta$ in the `TMB` section above that we can use in our empirical Bayes procedure.
-
-```{r}
-opt_theta <- opt
 ```
@@
 
-Start with `i = 1` for the first marginal:
+Start with `i = 1` for the first marginal of $\beta$:
 
@@
 
-#' Prepare the initialisation parameters if it's nu that we're toggling
+#' Prepare the initialisation parameters if it's beta that we're toggling
+#' We fix the beta parameter to zero, as we're using beta_i and beta_minus_i
 if(dat$toggle == 0) {
-  param_fixed_theta[["nu_i"]] <- param_fixed_theta$nu[i]
-  param_fixed_theta[["nu_minus_i"]] <- param_fixed_theta$nu[-i]
-  # param_fixed_theta$nu <- NULL
+  param_fixed_theta[["beta_i"]] <- param_fixed_theta$beta[i]
+  param_fixed_theta[["beta_minus_i"]] <- param_fixed_theta$beta[-i]
+  map_fixed_theta[["beta"]] <- rep(factor(NA), length(param[["beta"]]))
 }
@@
 #' Or if it's epsilon that we're toggling
+#' We fix the epsilon parameter to zero, as we're using epsilon_i and epsilon_minus_i
 if(dat$toggle == 1) {
+  param_fixed_theta[["nu_i"]] <- param_fixed_theta$nu[i]
+  param_fixed_theta[["nu_minus_i"]] <- param_fixed_theta$nu[-i]
+  map_fixed_theta[["nu"]] <- rep(factor(NA), length(param[["nu"]]))
+}
+
+#' Or if it's epsilon that we're toggling
+#' We fix the epsilon parameter to zero, as we're using epsilon_i and epsilon_minus_i
+if(dat$toggle == 2) {
   param_fixed_theta[["epsilon_i"]] <- param_fixed_theta$epsilon[i]
   param_fixed_theta[["epsilon_minus_i"]] <- param_fixed_theta$epsilon[-i]
-  # param_fixed_theta$epsilon <- NULL
+  map_fixed_theta[["epsilon"]] <- rep(factor(NA), length(param[["epsilon"]]))
 }
@@
 get_random <- function(dat) {
-  if(dat$toggle == 0) return(c("nu_minus_i", "epsilon"))
-  if(dat$toggle == 1) return(c("nu", "epsilon_minus_i"))
+  if(dat$toggle == 0) return(c("beta_minus_i", "nu", "epsilon"))
+  if(dat$toggle == 1) return(c("beta", "nu_minus_i", "epsilon"))
+  if(dat$toggle == 2) return(c("beta", "nu", "epsilon_minus_i"))
 }
@@
 
-# quad <- aghq::aghq(
-#   ff = obj_fixed_theta,
-#   k = 3,
-#   startingvalue = 0,
-#   control = aghq::default_control_tmb()
-# )
-#  
-# pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = fine_grid)[[1]]
-# 
-# ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
-#   geom_line() +
-#   theme_minimal() +
-#   labs(x = "nu[1]", y = "Posterior")
+quad <- aghq::aghq(
+  ff = obj_fixed_theta,
+  k = 3,
+  startingvalue = 0,
+  control = aghq::default_control_tmb()
+)
+
+pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad, finegrid = fine_grid)[[1]]
+
+ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "beta[1]", y = "Posterior")
 ```
@@
 
+Let's look at the mean and standard deviation of each of the $\beta$ parameters:
+
 ```{r}
@@
 inla4 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
-tmb5 <- as.vector(t(data.frame(sd_out$par.fixed[1:6], sqrt(diag(sd_out$cov.fixed)[1:6]))))
+tmb5 <- as.vector(t(data.frame(sd_out$par.random[1:6], sqrt(sd_out$diag.cov.random[1:6]))))
 glmmtmb6 <- as.vector(t(summary(fit6)$coefficients$cond[, c("Estimate", "Std. Error")]))
 tmbstan7 <- as.vector(t(summary(fit7)$summary[1:6, c(1, 3)]))
-aghq8 <- as.vector(t(summary8$summarytable[1:6, c(1, 4)]))
+
+fit8_beta_samples <- fit8_samples[, 1:6]
+aghq8 <- as.vector(t(data.frame(mean = apply(fit8_beta_samples, 2, mean), sd = apply(fit8_beta_samples, 2, sd))))
 
@@
     geom_line() +
-    scale_color_manual(values = cbpalette)
+    scale_color_manual(values = cbpalette) +
     theme_minimal()
@@
 
+## `epil_beta_index.cpp`
+
+```{cpp, echo=TRUE, eval=FALSE, output.var="ex3", code=readLines('epil_beta_index.cpp')}
+```
+
+## `epil_index.cpp`
+
+```{cpp, echo=TRUE, eval=FALSE, output.var="ex4", code=readLines('epil_index.cpp')}
+```
+
 # Original computing environment {-}

### 2022-11-08 src/epil/epil_beta_index.cpp

@@ -15,19 +15,19 @@ Type objective_function<Type>::operator()()
   DATA_MATRIX(E); // Epsilon matrix
 
   // Parameter block
-  PARAMETER_VECTOR(beta);
+  PARAMETER_VECTOR(nu);
   PARAMETER_VECTOR(epsilon);
 
-  PARAMETER(nu_i);
-  PARAMETER_VECTOR(nu_minus_i);
+  PARAMETER(beta_i);
+  PARAMETER_VECTOR(beta_minus_i);
 
-  vector<Type> nu(N * J);
+  vector<Type> beta(K);
   int k = 0;
-  for (int j = 0; j < (N * J); j++) {
+  for (int j = 0; j < K; j++) {
     if (j + 1 == i) { // +1 because C++ does zero-indexing
-      nu(j) = nu_i;
+      beta(j) = beta_i;
     } else {
-      nu(j) = nu_minus_i(k);
+      beta(j) = beta_minus_i(k);
       k++;
     }
   }

### 2022-11-08 src/epil/epil_epsilon_index.cpp

@@ -16,7 +16,6 @@ Type objective_function<Type>::operator()()
 
   // Parameter block
   PARAMETER_VECTOR(beta);
-
   PARAMETER_VECTOR(nu);
 
   PARAMETER(epsilon_i);

### 2022-11-08 src/docs_paper/paper.Rmd
Add algorithm for Gaussian latent field method

@@
 
-INLA is designed for use with latent Gaussian models (LGMs) of the form
+Consider a latent Gaussian models (LGM) of the form
 \begin{alignat}{2}
@@
 \end{equation}
-Rather than approximating the above full posterior, the INLA method instead approximates the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
+Rather than approximating this joint posterior, the INLA method proceeds by approximation of the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
 \begin{align}
@@
 \begin{equation}
-p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \hat{\bm{\mu}}(\btheta), \hat{\mathbf{Q}}(\btheta)^{-1}). \label{eq:gaussianx}
+p(\x \, | \, \btheta, \y) \approx p_\text{G}(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \bm{\mu}(\btheta), {\mathbf{Q}}(\btheta)^{-1}). \label{eq:gaussianx}
 \end{equation}
@@
 close to being Gaussian because $\y$ tends not to be strongly informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial].
-<!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference?
+If $\y$ is uninformative about $\x$ then how can we hope for useful inference?
 Perhaps what the author (Marta) meant is that the Gaussian distribution is sufficiently flexible to accommodate the updates made by $\y$.
--->
 As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}.
-Taking $\x = \hat{\bm{\mu}}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as
+Taking $\x = \bm{\mu}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as
 \begin{equation}
-\tilde p(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)} = \frac{p(\y, \hat{\bm{\mu}}(\btheta), \btheta)}{\det(\hat{\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal}
+\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} = \frac{p(\y, \bm{\mu}(\btheta), \btheta)}{\det({\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal}
 \end{equation}
-where the equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\hat{\bm{\mu}}(\btheta)$.
+where the equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\bm{\mu}(\btheta)$.
+We describe this as a Laplace approximation to the parameter posterior.
 
@@
 We now move on to the approximation $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$.
-Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section \ref{sec:approxi} above, a natural approach, and that taken by [@rue2007approximate], is to marginalise this distribution directly to obtain
+Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ for the denominator in Section \ref{sec:approxi} above, a natural approach [@rue2007approximate] is to marginalise this distribution directly to obtain
 \begin{equation}
-\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \hat{\mu}_i(\btheta), 1 / \hat{q}_i(\btheta)),
+\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, {\mu}_i(\btheta), 1 / {q}_i(\btheta)),
 \end{equation}
-where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from the relevant entries of $\hat{\bm{\mu}}(\btheta)$ and $\hat{\bm{Q}}(\btheta)$ respectively.
+where the marginal mean ${\mu}_i(\btheta)$ and precision ${q}_i(\btheta)$ are recovered directly from the relevant entries of $\bm{\mu}(\btheta)$ and ${\bm{Q}}(\btheta)$ respectively.
 However, although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode, so is generally not advised.
-Instead, [@rue2009approximate] propose two methods, a Laplace approximation and a simplified version which is less computationally demanding.
+Instead, @rue2009approximate propose two methods, a Laplace approximation and a simplified version which is less computationally demanding.
 The full Laplace approximation is
@@
   &\propto \frac{p(\x, \btheta \, | \, \y)}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
-  &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat{\bm{\mu}}_{-i}(x_i, \btheta)}
+  &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)}
   = \tilde p_{LA}(x_i \, | \, \btheta, \y), \label{eq:laplaceapproximation}
@@
 \begin{equation}
-p_G(\x_{-i} \, | \, x_i, \btheta, \y) = \mathcal{N}(\x_{-i} \, | \, \hat{\bm{\mu}}_{-i}(x_i, \btheta), \hat{\bm{Q}}_{-i}(x_i, \btheta)),
+p_G(\x_{-i} \, | \, x_i, \btheta, \y) = \mathcal{N}(\x_{-i} \, | \, \bm{\mu}_{-i}(x_i, \btheta), {\bm{Q}}_{-i}(x_i, \btheta)),
 \end{equation}
-is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
+is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\bm{\mu}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\bm{\mu}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\bm{\mu}_{-i}(x_i, \btheta)$ is not the same as $\bm{\mu}_{-i}(\btheta)$.}
 The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well.
 However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $i$, which is often computationally prohibitive.
-Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by [@rue2009approximate] to reduce the computational cost:
+Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by @rue2009approximate to reduce the computational cost:
 \begin{enumerate}
 \item
-Avoiding having to find the mode via optimisation by using the approximation $\hat{\bm{\mu}}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$
+Avoiding having to find the mode via optimisation by using the approximation $\bm{\mu}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$
 \item
@@
 \end{enumerate}
+
+## Algorithm\label{sec:algorithm}
+
+Here we describe the complete algorithm.
+Given a C++ user template `model.cpp` for the unnormalised log posterior:
+
+1. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(\btheta, \y)$
+2. Normalise $\tilde p_\text{LA}(\btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
+$$
+\tilde p_\text{AQ} (\btheta \, | \, \y) = \frac{\tilde p_\text{LA}(\btheta, \y)}{\tilde p_{\text{AQ}}(\y)},
+$$
+where the normalising constant is calculated using points $\z \in \mathcal{Q}(m, k)$, where $m = \dim(\btheta)$ and $k$ is the number of points per dimension, and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ as follows
+$$
+\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z).
+$$
+3. For each of the $\dim(\mathcal{Q}(m, k))$ Laplace approximations in Step 2. store the mode and Hessian of the Gaussian approximation to the random effects used in the denominator
+$$
+\{ \bm{\mu}(\btheta(\z)), \bm{Q}(\btheta(\z)) \}_{\z \in \mathcal{Q}(m, k)}.
+$$
+4. Obtain a nested approximation to the posterior marginal of the $i$th latent effect by reusing the AGHQ nodes and weights from Step 2.
+$$
+\tilde p(x_i \, | \, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{G}(x_i \, | \, \btheta(\z), \y) \tilde p_\text{AQ} (\btheta(\z) \, | \, \y) \omega(\z),
+$$
+where $p_\text{G}(x_i \, | \, \btheta(\z), \y)$ is a one-dimensional Gaussian with mean $\bm{\mu}(\btheta(\z))_i$ and variance $\bm{Q}(\btheta(\z))^{-1}_{ii}$.
 

### 2022-11-09 src/epil/epil.Rmd
Some improved writing, and get the results for the Laplace marginal into table for beta

@@
 abstract: |
-  **Background** @rue2009approximate use the epilespy example from the OpenBUGs manual as demonstration.
+  **Background** @rue2009approximate use the epilespy example from the OpenBUGs manual as demonstration of the integrated nested Laplace approximation method.
 
-  **Task** We implement the epilepsy example in Stan, `R-INLA`, `TMB`, `glmmTMB`, `tmbstan` and `aghq`. We then compare the results of the inference methods.
+  **Task** We implement the epilepsy example in Stan, `R-INLA`, `TMB`, `glmmTMB`, `tmbstan`, `aghq`, and a custom Laplace marginal version of `aghq`.
+We then compare the results of the inference methods.
 ---
@@
 The covariates are age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$, which are all centered.
-The observations are modelling using a Poisson distribution $y_{ij} \sim \text{Poisson}(e^{\eta_{ij}})$ with linear predictor 
+
+The observations are modeleed using a Poisson distribution $y_{ij} \sim \text{Poisson}(e^{\eta_{ij}})$ with linear predictor 
 \begin{align*}
@@
 \end{align*}
-where the prior on each of the regression parameters is $\mathcal{N}(0, 100^2)$.
+where the prior on each of the regression parameters, including the intercept, is $\mathcal{N}(0, 100^2)$.
 The random effects are IID $\epsilon_i \sim \mathcal{N}(0, 1/\tau_\epsilon)$ and $\nu_{ij} \sim \mathcal{N}(0, 1/\tau_\nu)$ with precision priors $\tau_\epsilon, \tau_\nu \sim \Gamma(0.001, 0.001)$.
@@
 
-Implement the model based upon a response variable of length $N \times J$, where $N$ is the number of patients and $J$ is the number of visits to the clinic, and a model matrix with $K$ predictors (including the intercept term):
+We will implement the model based upon a response variable of length $N \times J$, where $N$ is the number of patients and $J$ is the number of visits to the clinic, and a model matrix with $K$ predictors (including the intercept term):
 
@@
 
+```{r}
+dat <- list(N = N, J = J, K = K, X = X, y = y, E = make_epsilon_matrix(N, J))
+```
+
 # Inference methods
@@
 
-First, Hamiltonian Monte Carlo via the No-U-Turn algorithm, with four Markov chains of `iter = 1000` samples, and `warmup = 100` burn-in:
+First, Hamiltonian Monte Carlo (HMC) via the No-U-Turn Sampler (NUTS) algorithm, with four Markov chains of `iter = 1000` samples, and `warmup = 100` burn-in.
+Markov chain Monte Carlo (MCMC) like NUTS are exact as the number of samples tends to infinity.
+We provide the `stan` function with `file = "epil.stan"` which refers to a C++ file implementing the log-posterior in the Stan probabilistic programming language:
 
 ```{r}
-dat <- list(N = N, J = J, K = K, X = X, y = y, E = make_epsilon_matrix(N, J))
-
 fit1 <- stan(
-  "epil.stan",
+  file = "epil.stan",
   data = dat,
@@
 
-Both $\tau_\epsilon$ and $\tau_\nu$ have the same $\Gamma(0.001, 0.001)$ prior.
-In `R-INLA`, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior:
+`R-INLA` is an R package which implements integrated nested Laplace approximations.
+Rather than specifying the log-posterior via C++, the package uses a formula interface.
+
+We specify the $\Gamma(0.001, 0.001)$ prior on the random effects $\tau_\epsilon$ and $\tau_\nu$ below.
+In `R-INLA`, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior.
 
@@
 
-We can use various settings for the strategy, which controls the approximation used for the latent field posterior marginals:
+We can use various settings (`"gaussian"`, `"simplified.laplace"` and `"laplace"`) for the strategy passed to `control.inla`, which controls the approximation used for the latent field posterior marginals:
 
@@
 ## `TMB`
-
-The objective function `obj$fn` and its gradient `obj$gn` are a function of only the parameters, which in this instance are the six regression coefficients $\beta$ together with the logarithms of $\tau_\epsilon$ and $\tau_\nu$.
-This can be checked with `names(obj$par)`.
 
@@
 
+The objective function `obj$fn` and its gradient `obj$gn` are a function of only the parameters, which in this instance are the logarithms of $\tau_\epsilon$ and $\tau_\nu$.
+This can be checked with `names(obj$par)`.
 Optimise `obj` using 1000 iterations of the the `nlminb` optimiser, passing in the starting values `start`, objective function `objective` and its derivative `gradient`:
@@
 
-To be more thorough, we can get all the parameters obtained using the MCMC:
+This look different!
+Is this a problem?
+We can evaluate the NLL at some of the other parameters explored during the MCMC:
 
@@
 
-Using just a few of them, we can see that the difference between the `TMB` and Stan objectives is a constant:
+Using just a few of them, we can see that the difference between the `TMB` and Stan objectives is a constant, so for our purposes they are functionally the same:
 
@@
 
+Ok, this next section is going to be a bit more involved.
+
 ### Using multiple C++ template
@@
 Let's just starting by trying to get the Laplace marginals for $\beta$:
+
+`epil_beta_index.cpp` is a `TMB` template where the `beta` parameter has been replaced by `beta_i` and `beta_minus_i`.
+As well, there is an additional `DATA_INTEGER(i)` allowing us to pass in the index of `i` that we would like to obtain a full Laplace approximation to.
 
@@
 
-Create a function to get the Laplace marginal for the $i$th element of $\beta$:
+We create a function to get the Laplace marginal for the $i$th element of $\beta$.
+Some key things to note about this:
+We use the `map` argument to fix the hyperparameters to their MAP estimates, which requires passing `factor(NA)` to `map`, as well as setting the initialisation parameters for the hyperparameters to be their MAP estimates.
+Setting `random = c("beta_minus_i", "nu", "epsilon")` integrates out all of the random effects but `beta_i`.
+Quadrature is performed using `aghq::aghq` to obtain a the normalised posterior marginal.
 
@@
 
-Now we can compute the Laplace marginal for all `r dat$K` indices of $\beta$:
+Now we can compute the Laplace marginal for all `r dat$K` indices of $\beta$.
 
 ```{r}
-fine_grid <- seq(-5, 5, length.out = 1000)
+fine_grid <- seq(-5, 5, length.out = 500)
 
@@
 ### Using one C++ template
+
+Ok, great, so we got the Laplace marginals for `beta`.
+What about if we wanted to get them for all of the elements of the latent field?
+One way we could do that is to create separate `TMB` templates for each named random effect (`beta`, `nu` and `epsilon`).
+To see what this would look like, have a look at `epil_beta_index.cpp`, `epil_nu_index.cpp` and `epil_epsilon_index.cpp`.
+
+What might be better is to do it in one `TMB` template.
 
@@
 xi_laplace_marginal <- function(i, opt_theta, finegrid) {
-  #' To fill
-}
-```
-
-Start with `i = 1` for the first marginal of $\beta$:
-
-```{r}
-i <- 1
-
-#' I don't know if this is the right grid for this application...
-fine_grid <- seq(-5, 5, length.out = 1000)
-
   dat$toggle <- index_controller[i, "toggle"]
@@
 
-ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
-  geom_line() +
-  theme_minimal() +
-  labs(x = "beta[1]", y = "Posterior")
+  return(list("pdf_and_cdf" = pdf_and_cdf, "quad" = quad))
+}
+```
+
+Now we can compute the Laplace marginal for all `r nrow(index_controller)` random effects of $(\beta, \nu, \epsilon)$.
+
+```{r}
+fine_grid <- seq(-5, 5, length.out = 500)
+
+#' It's possible to get errors with the initial value not being finite
+safe_xi_laplace_marginal <- function(...) {
+  return(tryCatch(xi_laplace_marginal(...), error = function(e) NULL))
+}
+
+laplace_x <- lapply(1:nrow(index_controller), safe_xi_laplace_marginal, opt_theta = opt_theta, finegrid = fine_grid)
+
+laplace_beta_summary <- lapply(head(laplace_x, n = dat$K), function(result) summary(result$quad)$summarytable) %>%
+  bind_rows()
+
+rownames(laplace_beta_summary) <- paste0("beta[", 1:dat$K, "]")
 ```
@@
 
-df <- cbind(stan1, inla2, inla3, inla4, tmb5, glmmtmb6, tmbstan7, aghq8) %>%
+adam9 <- as.vector(t(laplace_beta_summary[1:6, 1:2]))
+
+df <- cbind(stan1, inla2, inla3, inla4, tmb5, glmmtmb6, tmbstan7, aghq8, adam9) %>%
   as.data.frame() %>%
@@
 rownames(df) <- c(sapply(0:5, beta_i))
-colnames(df) <- c("Stan", "R-INLA_G", "R-INLA_SL", "R-INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "type")
+colnames(df) <- c("Stan", "R-INLA_G", "R-INLA_SL", "R-INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "adam", "type")
 

### 2022-11-10 docs/seminar.Rmd
Upload abstract for Waterloo seminar

---
title: "Waterloo seminar series abstract"
author: "Adam Howes, Imperial College London"
date: "15th November 2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r echo=FALSE}
title <- "Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model"
abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of interest, including HIV prevalence, HIV incidence and treatment coverage are jointly modelled using both household survey data and routinely reported health system data. The model is provided as a tool for countries to input their data to and generate estimates (see https://naomi.unaids.org/). In this setting, computationally intensive inference methods like MCMC are impractical. To enable fast and accurate inference for Naomi, and other extended latent Gaussian models, we are developing a new inference method which combines the simplified integrated nested Laplace approximation approach of Wood (2020) with adaptive Gauss-Hermite quadrature. The new method will be implemented as an extension of the `aghq` package (Stringer, 2021), which will facilitate flexible and particularly easy use when provided a `TMB` C++ template for the log-posterior. In this talk, I'll discuss this progress towards this project, which I have been working on with Alex Stringer here at Waterloo this term through the International Visiting Graduate Student program."
```

**Title**:
```{r echo=FALSE, results='asis'}
cat(title)
```

Title word count: `r stringr::str_count(title, '\\w+')`.

**Abstract**:
```{r echo=FALSE, results='asis'}
cat(abstract)
```

Abstract word count: `r stringr::str_count(abstract, '\\w+')`/200.

### 2022-11-10 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Iterate slides forward

@@
 ---
-title: Inference methods for extended latent Gaussian models
+title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
 subtitle: Waterloo SAS Student Seminar Series
@@
 * We would like to provide them at a local, district-level
-* This is a challenging task! Data is noisy, sparse and biased
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+This is a challenging task! Data is noisy, sparse and biased. $\implies$ compelling case for thoughtful Bayesian modelling.
+\end{tcolorbox}
+\end{center}
 
@@
 
-* Consider small-areas $i = 1, \ldots, n$
-* Simple random sample household-survey in each area, with sample sizes $m^\text{HS}_i$
-* The number of people testing positive is $y^\text{HS}_i$
-* Then we can use a binomial logistic regression of the form:
+* Consider small-areas $i = 1, \ldots, n$ like the districts of a country
+* Simple random sample household-survey of size $m^\text{HS}_i$ in each area
+* The number of people testing positive for HIV is $y^\text{HS}_i$
+* You could calculate direct estimates of prevalence by $y^\text{HS}_i / m^\text{HS}_i$ but because the survey is powered at a national-level, the sample sizes are small and these estimates would be noisy
+
+# A simple small-area model for prevalence
+
+* We can use a binomial logistic regression of the form:
 \begin{align*}
@@
 \end{align*}
-* If $g$ is Gaussian then this is a latent Gaussian model in the sense of @rue2009approximate
-* We usually set up $g$ as a spatial smoother, because the sample sizes in each area are too small to get reliable direct estimates
-* One problem with the above model is that household surveys are expensive to run, so they only happen rarely
+* We usually set up $g$ as a Gaussian spatial smoother
+* This allows for pooling of information between districts
 
@@
 
+# Limitations of household surveys
+
+* A typical household survey takes around $UNKNOWN to run
+* This means that they don't happen very often
+* For this reason, the information might be quite out of date, and difficult to base policy on
+
 # Adding ANC surveillance
@@
 * This data source is more real-time than household surveys, but it's also more biased, because attendees are unlikely to be as representative of the population
-* But perhaps this bias is consistent, in which case we can still make use of the ANC data to improve our model!
+* But perhaps this bias is consistent, in which case we can still make use of the ANC data to supplement our model!
 
@@
 \end{align*}
-* Similar to using $\rho^\text{ANC}_i$ as a covariate in our model for household survey prevalence, but this way we allow it to vary rather than be fixed
+* This is similar to using $\rho^\text{ANC}_i$ as a covariate in the model for household survey prevalence, but this way takes into account sampling variation
 
@@
 
-* Remember that we're also interested in what proportion of PLHIV are receiving treatment
-* To estimate this, we need first to know the number of PLHIV
-\begin{equation*}
-H_i = N_i \rho^\text{HS}_i
-\end{equation*}
-* Then
+* We're also interested in what proportion $\alpha_i$ of people living with HIV (PLHIV) are receiving treatment
+* Suppose we record $A_i$ attendees from a known population of $N_i$ in each district
+* We can use another logistic regression model
 \begin{align*}
 A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
-\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2),
+\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
 \end{align*}
@@
 * Combining these three modules is the basis of the Naomi evidence synthesis model
-* Used by countries (which provide their own data) to produce HIV estimates in a yearly process supported by UNAIDS
-* Can't run long MCMC in this setting, requires fast, accurate, approximations 
+* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
+* Can't run long MCMC in this setting, so we require fast, accurate, approximations 
 * It's a complicated model, and requires something more flexible than `R-INLA`
-* Currently using Template Model Builder `TMB` [@kristensen2015tmb]
+* Currently using a package called Template Model Builder `TMB`
 
@@
 
+#
+
+```{r, echo=FALSE, fig.cap="Example of the user interface from https://naomi.unaids.org/", out.width = '70%'}
+knitr::include_graphics("naomi_user.png")
+```
+
 # Template Model Builder
 
-* R package which implements the Laplace approximation for latent variable models using automatic differentiation (via `CppAD`)
-  * For more about AD see e.g. @griewank2008evaluating. It's useful for getting the mode, Hessian
-* Write an objective function $f(\x, \btheta)$ in C++ ("user template")
-    * We select $f(\x, \btheta) = - \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)$
+* `TMB` [@kristensen2015tmb] is an R package which implements the Laplace approximation for latent variable models
+* To get started, write an objective function $f(\x, \btheta)$ in `TMB` C++ syntax
+* As pseudo-Bayesians, we choose the log-posterior
+$$
+f(\x, \btheta) = - \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)
+$$
 
 # Template Model Builder
+
+For example, for the model
+$$
+\y \sim \mathcal{N}(\mu, 1)
+$$
+with $p(\mu) \propto 1$ then the `TMB` user template looks like...
+
+#
 
@@
 
-* Performs the Laplace approximation $L_f(\btheta) \approx L^\star_f(\btheta)$ and use R to optimise this with respect to $\btheta$ to give $\hat \btheta$ (the central point in Figure \ref{fig:grid})
-  * This is done by specifying the `random` argument to be the parameters that you want to integrate out with a Laplace approximation (the latent field)
-* MAP estimate of $\x$ conditional on $\hat \btheta$
-* Standard errors calculated using the $\delta$-method (a Gaussian assumption)
+* We can use `TMB` to obtain the Laplace approximation
+$$
+\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)} \label{eq:hypermarginal}
+$$
+* Integrate out a Gaussian approximation $\tilde p_\text{G}(\x \, | \, \btheta, \y)$ to the latent field
+* `TMB` uses automatic differentiation [@griewank2008evaluating] via `CppAD`
 
@@
 
-* Suggested reading: the paper @rue2009approximate or a book e.g. @blangiardo2015spatial
-* Approximate Bayesian inference for \textcolor{hilit}{latent Gaussian models} (LGMs), which recall are three-stage models with middle layer
-\begin{equation*}
-  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
-\end{equation*}
-* `R-INLA` implementation takes advantage of sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)
-
-# Integrated Nested Laplace Approximation
-
-* Gives approximate \textcolor{hilit}{posterior marginals} $\{\tilde p(x_i \, | \, \y)\}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$
-* To approximate posterior marginals below requires $\tilde p(\btheta \, | \, \y)$ and $\tilde p(x_i \, | \, \btheta, \y)$
+* Integrated nested Laplace approximation (INLA) [@rue2009approximate; @blangiardo2015spatial] is an approach to approximate inference which builds on the Laplace approximation
+* Goal is to approximate \textcolor{hilit}{posterior marginals} $\{\tilde p(x_i \, | \, \y)\}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$
 \begin{align}
@@
 \end{align}
+* To do so, we require the approximations $\tilde p(\btheta \, | \, \y)$ and $\tilde p(x_i \, | \, \btheta, \y)$
+* There are four steps as to how the method works (bare with me!)
 
-# Integrated Nested Laplace Approximation
+# Step 1)
 
@@
 \begin{equation}
-\tilde p(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)}  \label{eq:hypermarginal}
+\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)} \label{eq:hypermarginal}
 \end{equation}
 which can be marginalised to get $\tilde p(\theta_j \, | \, \y)$
-  * Note: this involves integrating out a Gaussian approximation to the latent field, which we denote by $p_G(\x \, | \, \btheta, \y)$
+* Notice that this is the same object we had been working with in `TMB`
+* We use this approximation \textcolor{hilit}{nested} within integrals -- hence the name INLA
 
-# Integrated Nested Laplace Approximation
+# Step 2)
 
-2) In both \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. \eqref{eq:hypermarginal}, so choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$
+2) In both Equations \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. $\btheta$, so choose integration nodes and weights $\{ \btheta(\z), \omega(\z) \}_{\z \in \mathcal{Z}}$
   * For low $m$ `R-INLA` uses a grid-strategy
   * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design
-  * Other approaches, like adaptive Gaussian Hermite quadrature (AGHQ) have recently been shown to have theoretical guarantees [@bilodeau2021stochastic] and may be preferable
-
-# Grid strategy
-
-* To write...
+  * We plan to use adaptive Gaussian Hermite quadrature (AGHQ), which has recently been shown to have theoretical guarantees [@bilodeau2021stochastic]
 
@@
 
-```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("An illustration of the \\texttt{R-INLA} grid method for selecting integration points using a toy bivariate Gaussian distribution for $\\btheta$.
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("An illustration of the \\texttt{R-INLA} grid method for selecting integration nodes using a toy bivariate Gaussian distribution for $\\btheta$.
 Start at the mode and work outwards along the eigenvectors until the density drops sufficiently low.
@@
 * The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) to automatically choose the node locations
-  * No manual tuning!
+  * Does not require manual tuning!
 * Works particularly well when the integrand is pretty Gaussian
-* Supposing you use $k$ quadrature points, then
-  * If $k = 1$ then it's a Laplace approximation
-  * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often
+* Use $k$ quadrature nodes per dimension, for example if $k = 3$ then $3^m$ total nodes
 * Implemented in the `aghq` R package. See vignette @stringer2021implementing
@@
 
-# Integrated Nested Laplace Approximation
+# Step 3)
 
 3) Choose approximation for $\tilde p(x_i \, | \, \btheta, \y)$
-  * Simplest version [@rue2007approximate] is to marginalise the $p_G(\x \, | \, \btheta, \y)$
+  * Simplest version [@rue2007approximate] is to marginalise $\tilde p_\text{G}(\x \, | \, \btheta, \y)$
 \begin{equation}
-\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta))
+\tilde p_\text{G}(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta))
 \end{equation}
-  * The above is referred to as `method = "gaussian"` in `R-INLA`
-  * There are two better, more complex approximations
-    * Confusingly called `"simplified laplace"` and `"laplace"`
+  * In `R-INLA`, the above is referred to as `method = "gaussian"` 
+  * There are two better, more complex approximations, confusingly called `"simplified.laplace"` and `"laplace"`
+  * Uses sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)
 
-# Integrated Nested Laplace Approximation
+# Step 4)
 
-4) Finally (!) use quadrature combining
-  * our approximation $p(\btheta \, | \, \y)$ from step 1),
-  * our integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$ from step 2),
-  * and our approximation $\tilde p(x_i \, | \, \btheta, \y)$ from step 3)
+4) Finally, use quadrature to combine
+  * our approximation $\tilde p_\text{LA}(\btheta \, | \, \y)$ from step 1),
+  * some choice of integration nodes and weights $\{ \btheta(\z), \omega(\z) \}$ from step 2),
+  * some choice of approximation $\tilde p(x_i \, | \, \btheta, \y)$ from step 3)
 to give
@@
   \tilde p(x_i \, | \, \y) = 
-  \sum_{k = 1}^K \tilde p(x_i \, | \, \btheta^{(k)}, \y) \times \tilde p(\btheta^{(k)} \, | \, \y) \times \Delta^{(k)}
+  \sum_{\z \in \mathcal{Z}} \tilde p(x_i \, | \, \btheta(\z), \y) \times \tilde p_\text{LA}(\btheta(\z) \, | \, \y) \times \omega(\z)
 \end{equation}
-
-# Experiments
-
-* We would like to find an accurate way to do inference for the Naomi model
-* It also needs to be fast enough to use in production!
 
@@
 
-* Note: using different software it is often very difficult to ensure the model is precisely the same, so we're very fortunate here
+* Note: using different software it is usually very difficult to ensure the model is precisely the same, so we're very fortunate here
 

### 2022-11-11 src/dev_sinla/sinla.Rmd
Working on Laplace plus integration strategy

@@
 \begin{equation}
-\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxgiveny)
+\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z), (\#eq:pxgiveny)
 \end{equation}
 $$
-where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
+where $\theta(z) \in \Theta$ and $\omega_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
 In particular, let
@@
 $$
-w_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
+\omega_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
 $$
@@
 \begin{equation}
-\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxigiveny)
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z), (\#eq:pxigiveny)
 \end{equation}
@@
 $$
-\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
+\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z).
 $$
@@
 
-## Marginal of joint Gaussian approximation
+## Marginal of joint Gaussian approximation {#aghq-gaussian-marginal}
 
@@
 $$
-\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z).
 $$
@@
 
-The weight $w_k(z)$ is:
+The weight $\omega_k(z)$ is:
 
@@
 $$
-\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log w_k(z) \right)
+\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log \omega_k(z) \right)
 $$
@@
 
-Great -- now let's put it together and sum over `z` so that we can obtain the probability for our `x_test_value` in marginal `i`.
+Great -- now let's sum over `z` so that we can obtain the probability for our `x_test_value` in marginal `i`.
 We won't worry about inefficient R for loops for the time being.
@@
 
-```{r}
-#' To-do
-```
-
-# Simplified INLA
-
-In Section \@ref(full-laplace) we directly calculate the Laplace approximation to $\int p(x, y, \theta) \text{d} \theta_{-i}$.
-Suppose that
-$$
-x \sim \mathcal{N}(\mu(\theta), Q(\theta)^{-1})
-$$
-then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mean
-$$
-\mu(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:conditionalmean)
-$$
-and precision matrix
-$$
-Q(\theta)_{-i, -i} (\#eq:conditionalprecision)
-$$
-
-We follow Wood (2019) Section 2.
-In this section we omit dependence on $\theta$ such that: $\mu = \mu(\theta)$, $Q = Q(\theta)$, $\mu(x_i) = \mu(x_i, \theta)$ and $Q(x_i) = Q(x_i, \theta)$.
-
-Let $\mathcal{D}_{2j}$ denote a $2j \times 2j$ diagonal matrix with leading diagonal $-1, 1, -1, 1, \ldots$.
-For example, with $j = 3$ we have:
-
-```{r}
-D2j <- function(j) {
-  diag((-1)^{1:2 * j}, nrow = 2 * j)
-}
-
-D2j(3)
-```
-
-Let $u_0$ be a zero column matrix.
-I am guessing that this means $u_j$ is a column matrix containing $j$ as each entry.
-It's not specified what dimension $u_0$ or $u_j$ should be, but since the algorithm contains producing by $\mathcal{D}_{2j}$ presumably their lengths must be $2j$.
-
-```{r}
-uj <- function(j) {
- matrix(j, nrow = 2 * j, ncol = 1)
-}
-
-uj(3)
-```
-
-The algorithm contains $u_j \mathcal{D}_{2j} u_j^\top$.
-If $u_j$ is a column matrix of dimension $\text{length}(u_j) \times 1$ then it's not going to be conformable with a $2j \times 2j$ matrix.
-Is it actually a row matrix of dimension $1 \times \text{length}(u_j)$?
-And $\text{length}(u_j) = 2j$ as discussed above?
-
-Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$.
-But it hasn't been defined what $\Delta_j$ is yet.
-
-## Step 1: Compute the Cholesky factor
-
-The first task is to "compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$".
-
-```{r}
+Let's do as we have above, but with the full Laplace approximation in place of the Gaussian.
+We want to evaluate
+$$
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z),
+$$
+so much of what we computed above in Section \@ref(aghq-gaussian-marginal) is still relevant.
+The only thing that has changed is that we're now using $\tilde p_\text{LA}(x_i \, | \, \theta(z), y)$ rather than $\tilde p_\text{G}(x_i \, | \, \theta(z), y)$.
+
+```{r}
+x_test_value <- 0
 i <- 1
-
-mm <- h$env$last.par
-mean <- mm[h$env$random]
-Q <- h$env$spHess(mm, random = TRUE)
-Q0 <- Q[-i, -i]
-
-L0 <- chol(Q0)
-```
-
-We have that $Q_0 = L_0 {L_0}^\top$:
-
-```{r}
-sum(abs((L0 %*% t(L0)) - Q0)) < 10E-9
-```
-
-## Step 2. Loop
-
-For each $x_i$ in a grid of evaluation points (how does one chose this grid?) repeat the following:
-
-```{r}
-x_grid
-```
-
-### a)
-
-Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean).
-We start by computing the conditional mean for the first grid point:
-
-```{r}
-Sigma <- solve(Q)
-conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i])
-```
-
-Note that for this model the latent field is IID, so the conditional mean is the same as the mean.
-Now we use Newton's method to find $\mu(x_i)$.
-This requires some function $f(\mu(x_i), \ldots) = 0$.
-
-### b)
-
-Compute a set of $J$ steps $\{\Delta_j\}$
-
-```{r}
-J <- 1
-```
-
-### c)
-
-```{r}
-j <- 1
-```
-
-For $j = 1, \ldots, J - 1$ compute
-$$
-q = Q_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j
-$$
-and
-$$
-g = \Delta_x \log p(\mu(x_i) + \Delta_j^i, y, \theta).
-$$
-Then compute the matrix
-$$
-u_{j + 1} = \{q(\Delta_j^\top j)^{-1/2}, g_{-i}(\Delta_j^\top g_{-i})^{-1/2}, u_j\}.
-$$
-
-### d)
-
-Compute the determinant approximation
-$$
-\det(Q_1) = \det(Q_0) \det(I_{2J} + u_J^\top Q_0^{-1} u_J \mathcal{D}_{2j}).
-$$
-
-### e)
-
-Compute $\tilde p(x_i \, | \, \theta, y) = \ldots$.
-
-## Step 3. Renormalise $\tilde p(x_i \, | \, \theta, y)$
-
-Renormalise $\tilde p(x_i \, | \, \theta, y)$.
+z <- 1
+```
+
+As before, the weight $\omega_k(z)$ is:
+
+```{r}
+modesandhessians$weights[z]
+```
+
+And also as before, the normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ is:
+
+```{r}
+theta <- as.numeric(modesandhessians[z, thetanames])
+- h_norm(theta)
+```
+
+Now, getting to the difference, the normalised Laplace approximation $p_\text{LA}(x_i \, | \, \theta(z), y)$ on the log-scale is:
+
+```{r}
+dat_i <- prepare_dat(dat, i)
+map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
+
+param_i_fixed_theta <- list(
+  beta_prev = 0, 
+  phi_prev_i = 0, 
+  phi_prev_minus_i = rep(0, data$n - 1),
+  log_sigma_phi_prev = 0
+)
+
+for(theta in thetanames) param_i_fixed_theta[[theta]] <- modesandhessians[z, thetanames][[theta]]
+
+obj_fixed_theta <- MakeADFun(
+  data = dat_i,
+  parameters = param_i_fixed_theta,
+  random = "phi_prev_minus_i",
+  DLL = "model1_index",
+  silent = TRUE,
+  map = map_fixed_theta
+)
+
+#' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
+quad <- aghq::aghq(ff = obj_fixed_theta, k = 5, startingvalue = 0, control = aghq::default_control_tmb())
+
+obj_fixed_theta_norm <- function(x) obj_fixed_theta$fn(x) + quad$normalized_posterior$lognormconst
+
+obj_fixed_theta_norm(x_test_value)
+```
+
+Putting it together:
+
+```{r}
+#' On the log-scale
+as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z])
+
+#' Unlogged
+exp(as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
+```
+
+Sum this over `z`:
+
+```{r}
+pdf <- 0
+
+for(z in 1:length(modesandhessians$weights)) {
+  theta <- as.numeric(modesandhessians[z, thetanames])
+
+  dat_i <- prepare_dat(dat, i)
+  map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
+
+  param_i_fixed_theta <- list(
+    beta_prev = 0, 
+    phi_prev_i = 0, 
+    phi_prev_minus_i = rep(0, data$n - 1),
+    log_sigma_phi_prev = 0
+  )
+
+  for(theta in thetanames) param_i_fixed_theta[[theta]] <- modesandhessians[z, thetanames][[theta]]
+
+  obj_fixed_theta <- MakeADFun(
+    data = dat_i,
+    parameters = param_i_fixed_theta,
+    random = "phi_prev_minus_i",
+    DLL = "model1_index",
+    silent = TRUE,
+    map = map_fixed_theta
+  )
+
+  #' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
+  quad <- aghq::aghq(ff = obj_fixed_theta, k = 5, startingvalue = 0, control = aghq::default_control_tmb())
+
+  obj_fixed_theta_norm <- function(x) obj_fixed_theta$fn(x) + quad$normalized_posterior$lognormconst
+
+  pdf <- pdf + exp(as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
+}
+
+pdf
+```
+
+<!-- # Simplified INLA -->
+
+<!-- In Section \@ref(full-laplace) we directly calculate the Laplace approximation to $\int p(x, y, \theta) \text{d} \theta_{-i}$. -->
+<!-- Suppose that -->
+<!-- $$ -->
+<!-- x \sim \mathcal{N}(\mu(\theta), Q(\theta)^{-1}) -->
+<!-- $$ -->
+<!-- then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mean -->
+<!-- $$ -->
+<!-- \mu(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:conditionalmean) -->
+<!-- $$ -->
+<!-- and precision matrix -->
+<!-- $$ -->
+<!-- Q(\theta)_{-i, -i} (\#eq:conditionalprecision) -->
+<!-- $$ -->
+
+<!-- We follow Wood (2019) Section 2. -->
+<!-- In this section we omit dependence on $\theta$ such that: $\mu = \mu(\theta)$, $Q = Q(\theta)$, $\mu(x_i) = \mu(x_i, \theta)$ and $Q(x_i) = Q(x_i, \theta)$. -->
+
+<!-- Let $\mathcal{D}_{2j}$ denote a $2j \times 2j$ diagonal matrix with leading diagonal $-1, 1, -1, 1, \ldots$. -->
+<!-- For example, with $j = 3$ we have: -->
+
+<!-- ```{r} -->
+<!-- D2j <- function(j) { -->
+<!--   diag((-1)^{1:2 * j}, nrow = 2 * j) -->
+<!-- } -->
+
+<!-- D2j(3) -->
+<!-- ``` -->
+
+<!-- Let $u_0$ be a zero column matrix. -->
+<!-- I am guessing that this means $u_j$ is a column matrix containing $j$ as each entry. -->
+<!-- It's not specified what dimension $u_0$ or $u_j$ should be, but since the algorithm contains producing by $\mathcal{D}_{2j}$ presumably their lengths must be $2j$. -->
+
+<!-- ```{r} -->
+<!-- uj <- function(j) { -->
+<!--  matrix(j, nrow = 2 * j, ncol = 1) -->
+<!-- } -->
+
+<!-- uj(3) -->
+<!-- ``` -->
+
+<!-- The algorithm contains $u_j \mathcal{D}_{2j} u_j^\top$. -->
+<!-- If $u_j$ is a column matrix of dimension $\text{length}(u_j) \times 1$ then it's not going to be conformable with a $2j \times 2j$ matrix. -->
+<!-- Is it actually a row matrix of dimension $1 \times \text{length}(u_j)$? -->
+<!-- And $\text{length}(u_j) = 2j$ as discussed above? -->
+
+<!-- Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$. -->
+<!-- But it hasn't been defined what $\Delta_j$ is yet. -->
+
+<!-- ## Step 1: Compute the Cholesky factor -->
+
+<!-- The first task is to "compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$". -->
+
+<!-- ```{r} -->
+<!-- i <- 1 -->
+
+<!-- mm <- h$env$last.par -->
+<!-- mean <- mm[h$env$random] -->
+<!-- Q <- h$env$spHess(mm, random = TRUE) -->
+<!-- Q0 <- Q[-i, -i]  -->
+
+<!-- L0 <- chol(Q0) -->
+<!-- ``` -->
+
+<!-- We have that $Q_0 = L_0 {L_0}^\top$: -->
+
+<!-- ```{r} -->
+<!-- sum(abs((L0 %*% t(L0)) - Q0)) < 10E-9 -->
+<!-- ``` -->
+
+<!-- ## Step 2. Loop -->
+
+<!-- For each $x_i$ in a grid of evaluation points (how does one chose this grid?) repeat the following: -->
+
+<!-- ```{r} -->
+<!-- x_grid -->
+<!-- ``` -->
+
+<!-- ### a) -->
+
+<!-- Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean). -->
+<!-- We start by computing the conditional mean for the first grid point: -->
+
+<!-- ```{r} -->
+<!-- Sigma <- solve(Q) -->
+<!-- conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i]) -->
+<!-- ``` -->
+
+<!-- Note that for this model the latent field is IID, so the conditional mean is the same as the mean. -->
+<!-- Now we use Newton's method to find $\mu(x_i)$. -->
+<!-- This requires some function $f(\mu(x_i), \ldots) = 0$. -->
+
+<!-- ### b) -->
+
+<!-- Compute a set of $J$ steps $\{\Delta_j\}$ -->
+
+<!-- ```{r} -->
+<!-- J <- 1 -->
+<!-- ``` -->
+
+<!-- ### c) -->
+
+<!-- ```{r} -->
+<!-- j <- 1 -->
+<!-- ``` -->
+
+<!-- For $j = 1, \ldots, J - 1$ compute -->
+<!-- $$ -->
+<!-- q = Q_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j -->
+<!-- $$ -->
+<!-- and -->
+<!-- $$ -->
+<!-- g = \Delta_x \log p(\mu(x_i) + \Delta_j^i, y, \theta). -->
+<!-- $$ -->
+<!-- Then compute the matrix -->
+<!-- $$ -->
+<!-- u_{j + 1} = \{q(\Delta_j^\top j)^{-1/2}, g_{-i}(\Delta_j^\top g_{-i})^{-1/2}, u_j\}. -->
+<!-- $$ -->
+
+<!-- ### d) -->
+
+<!-- Compute the determinant approximation -->
+<!-- $$ -->
+<!-- \det(Q_1) = \det(Q_0) \det(I_{2J} + u_J^\top Q_0^{-1} u_J \mathcal{D}_{2j}). -->
+<!-- $$ -->
+
+<!-- ### e) -->
+
+<!-- Compute $\tilde p(x_i \, | \, \theta, y) = \ldots$. -->
+
+<!-- ## Step 3. Renormalise $\tilde p(x_i \, | \, \theta, y)$ -->
+
+<!-- Renormalise $\tilde p(x_i \, | \, \theta, y)$. -->
 

### 2022-11-11 src/docs_paper/paper.Rmd
Update writing (first go at algorithm)

--- 2022-10-22 src/docs_paper/paper.Rmd
@@
 Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach used currently, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler.
-By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} C++ template for the model's log-posterior.
+By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
 keyword-subclass: | 
@@
 
-Mounting an effective public health response to the HIV epidemic requires accurate, timely and sufficiently fine-scale estimates of HIV indicators.
-No single data source is sufficient to base these estimates on.
-Although nationally-representative household surveys are the most reliable data source, in most countries they only occur infrequently.
-Other data sources, such as routine health surveillance of antenatal care clinics, are more timely but from a biased sample of the population.
-To remedy these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
+Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at a sufficiently fine-scale resolution.
+Producing these estimates is a challenging task, as all available data sources have shortcomings.
+Nationally-representative household surveys are the most statistically reliable data source, but due to their high cost to run, in most countries they only occur infrequently.
+Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on a biased sample of the population.
+To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
 Software (\url{https://naomi.unaids.org}) has been developed for Naomi, which allows countries to input their data and generate estimates in a yearly process supported by UNAIDS.
 
-The model is complex and presents a challenging inference problem.
-Any approach must be fast enough to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC).
-Currently, inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-Owing to its speed and flexibility, `TMB` is more broadly becoming a popular package in spatial statistics [@osgoodzimmerman2021statistical].
-In `TMB`, inference is based on optimisation of a C++ template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
+The complexity of the model presents a difficult Bayesian inference problem.
+Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
+Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
+`TMB` is gaining popularity spatial statistics [@osgoodzimmerman2021statistical] owing to its speed and flexibility.
+Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
+For the Naomi model, we use this option to integrate out the latent field parameters.
 Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
-Although this approach has favorable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
-To obtain fast, accruate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
+Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
 
-INLA [@rue2009approximate] is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature, and implemented in the `R-INLA` \textsc{R} package.
+To obtain fast, accurate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
+INLA [@rue2009approximate] is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
 The central innovation is a way to approximate accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
 Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
-Doing so facilitates inference for extended latent Gaussian models (ELGMs) [@stringer2022fast], which build on latent Gaussian models (LGMs) by allowing each element of the linear predictor $\mu_i$ to depend on any subset of elements from the latent field.
+This extension facilitates inference for models like Naomi, which fall within the extended latent Gaussian model (ELGM) [@stringer2022fast] class, and were not previously amenable to inference with INLA.
+ELGMs build on latent Gaussian models (LGMs) by allowing each element of the linear predictor to depend on any subset of elements from the latent field.
+We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
+Though as of yet no theory exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2021stochastic using AGHQ.
+We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
+As `aghq` is designed to naturally interface with `TMB`, use of the method is easy when provided a \textsf{C++} user template for the log-posterior.
 
-AGHQ is a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
-@bilodeau2021stochastic provide theoretical results.
-The method is implementing in the `aghq` \textsf{R} package [@stringer2021implementing].
+The remainder of this paper is organised as follows.
+In Section \ref{sec:naomi} we describe the Naomi model.
+Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
+As a case-study, we fit the Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
+In this section, we also demonstrating a Bayesian workflow.
+Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how our method might be used in other models, and directions for future research.
 
-# The Naomi model
+# The Naomi model\label{sec:naomi}
 
@@
 
-# Fast inference methods
+# Fast inference methods\label{sec:fastinferencemethods}
 
-## Integrated nested Laplace approximation\label{sec:inla}
+<!-- ## Integrated nested Laplace approximation\label{sec:inla} -->
 
-INLA is designed for use with latent Gaussian models (LGMs) of the form
+<!-- Consider a latent Gaussian model (LGM) of the form -->
+<!-- \begin{alignat}{2} -->
+<!-- &\text{(Observations)}     &        y_i &\sim p(y_i \, | \, x_i, \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\ -->
+<!-- &\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\ -->
+<!-- &\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta), \label{eq:parameters} -->
+<!-- \end{alignat} -->
+<!-- where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$. -->
+<!-- For such models, the joint posterior of $(\x, \btheta)$ is given by -->
+<!-- \begin{equation} -->
+<!--   p(\x, \btheta \, | \, \y) -->
+<!--   \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right).
+\label{eq:lgmjointposterior} -->
+<!-- \end{equation} -->
+<!-- Rather than approximating this joint posterior, the INLA method proceeds by approximation of the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by -->
+<!-- \begin{align} -->
+<!--   p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\ -->
+<!--   p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2} -->
+<!-- \end{align} -->
+<!-- An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$.
+-->
+<!-- We discuss these two approximations in turn below. -->
+
+<!-- ### Approximation (i) \label{sec:approxi} -->
+
+<!-- The posterior marginal of the parameters $p(\btheta \, | \, \y)$ appears in both Equations \eqref{eq:inla1} and \eqref{eq:inla2}. -->
+<!-- This distribution is approximated by $\tilde p(\btheta \, | \, \y)$ and represented by a set of $K$ integration nodes $\{ \btheta^{(k)} \}$ and area-weights $\{ \Delta^{(k)} \}$. -->
+<!-- The first step is to rewrite $p(\btheta \, | \, \y)$ as -->
+<!-- \begin{equation} -->
+<!--   p(\btheta \, | \, \y) -->
+<!--   = \frac{p(\x, \btheta \, | \, \y)}{p(\x \, | \, \btheta, \y)} -->
+<!--   \propto \frac{p(\y, \x, \btheta)}{p(\x \, | \, \btheta, \y)}. \label{eq:hyperpost} -->
+<!-- \end{equation} -->
+<!-- Approximation (i) then uses a Gaussian approximation to the denominator given by -->
+<!-- \begin{equation} -->
+<!-- p(\x \, | \, \btheta, \y) \approx p_\text{G}(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \bm{\mu}(\btheta), {\mathbf{Q}}(\btheta)^{-1}). \label{eq:gaussianx} -->
+<!-- \end{equation} -->
+<!-- This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution, given by -->
+<!-- \begin{equation} -->
+<!-- p(\x \, | \, \btheta, \y) \propto -->
+<!-- \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right), -->
+<!-- \end{equation} -->
+<!-- close to being Gaussian because $\y$ tends not to be strongly informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial]. -->
+<!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference?
+Perhaps what the author (Marta) meant is that the Gaussian distribution is sufficiently flexible to accommodate the updates made by $\y$.
+-->
+<!-- As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}. -->
+<!-- Taking $\x = \bm{\mu}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as -->
+<!-- \begin{equation} -->
+<!-- \tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} = \frac{p(\y, \bm{\mu}(\btheta), \btheta)}{\det({\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal} -->
+<!-- \end{equation} -->
+<!-- where the equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\bm{\mu}(\btheta)$. -->
+<!-- We describe this as a Laplace approximation to the parameter posterior. -->
+
+<!-- ### Approximation (ii) -->
+
+<!-- We now move on to the approximation $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$. -->
+<!-- Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ for the denominator in Section \ref{sec:approxi} above, a natural approach [@rue2007approximate] is to marginalise this distribution directly to obtain -->
+<!-- \begin{equation} -->
+<!-- \tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, {\mu}_i(\btheta), 1 / {q}_i(\btheta)), -->
+<!-- \end{equation} -->
+<!-- where the marginal mean ${\mu}_i(\btheta)$ and precision ${q}_i(\btheta)$ are recovered directly from the relevant entries of $\bm{\mu}(\btheta)$ and ${\bm{Q}}(\btheta)$ respectively. -->
+<!-- However, although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode, so is generally not advised. -->
+<!-- Instead, @rue2009approximate propose two methods, a Laplace approximation and a simplified version which is less computationally demanding. -->
+<!-- The full Laplace approximation is -->
+<!-- \begin{align} -->
+<!--   p(x_i \, | \, \btheta, \y) &= p(\x \, | \, \btheta, \y) \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
+<!--   &= \frac{p(\x, \btheta \, | \, \y)}{p(\btheta \, | \, \y)} \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
+<!--   &\propto \frac{p(\x, \btheta \, | \, \y)}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
+<!--   &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)} -->
+<!--   = \tilde p_{LA}(x_i \, | \, \btheta, \y), \label{eq:laplaceapproximation} -->
+<!-- \end{align} -->
+<!-- where -->
+<!-- \begin{equation} -->
+<!-- p_G(\x_{-i} \, | \, x_i, \btheta, \y) = \mathcal{N}(\x_{-i} \, | \, \bm{\mu}_{-i}(x_i, \btheta), {\bm{Q}}_{-i}(x_i, \btheta)), -->
+<!-- \end{equation} -->
+<!-- is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\bm{\mu}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\bm{\mu}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\bm{\mu}_{-i}(x_i, \btheta)$ is not the same as $\bm{\mu}_{-i}(\btheta)$.} -->
+<!-- The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well. -->
+<!-- However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $i$, which is often computationally prohibitive. -->
+<!-- Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by @rue2009approximate to reduce the computational cost: -->
+<!-- \begin{enumerate} -->
+<!-- \item -->
+<!-- Avoiding having to find the mode via optimisation by using the approximation $\bm{\mu}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$ -->
+<!-- \item -->
+<!-- As only those $x_j$ close to $x_i$ should have an impact on the marginal of $x_i$, then by selecting some subset $R_i(\btheta)$ of nodes $j$ to impact $j$ the matrix which needs to be factorised can be reduced in dimension to be $| R_i(\btheta) | \times | R_i(\btheta) |$ rather than $n \times n$ -->
+<!-- \end{enumerate} -->
+
+Consider a latent Gaussian model (LGM) of the form
 \begin{alignat}{2}
-&\text{(Observations)}     &        y_i &\sim p(y_i \, | \, x_i, \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\
+&\text{(Observations)}     &        \y &\sim p(\y \, | \, \x, \btheta), \label{eq:data} \\
 &\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\
@@
 where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
-For such models, the joint posterior of $(\x, \btheta)$ is given by
+The joint posterior of $(\x, \btheta)$ is given by
 \begin{equation}
@@
 \end{equation}
-Rather than approximating the above full posterior, the INLA method instead approximates the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
+We consider approximations to the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
 \begin{align}
@@
 \end{align}
-An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$.
-We discuss these two approximations in turn below.
 
-### Approximation (i) \label{sec:approxi}
+## Algorithm\label{sec:algorithm}
 
-The posterior marginal of the parameters $p(\btheta \, | \, \y)$ appears in both Equations \eqref{eq:inla1} and \eqref{eq:inla2}.
-This distribution is approximated by $\tilde p(\btheta \, | \, \y)$ and represented by a set of $K$ integration points $\{ \btheta^{(k)} \}$ and area-weights $\{ \Delta^{(k)} \}$.
-The first step is to rewrite $p(\btheta \, | \, \y)$ as
-\begin{equation}
-  p(\btheta \, | \, \y) 
-  = \frac{p(\x, \btheta \, | \, \y)}{p(\x \, | \, \btheta, \y)}
-  \propto \frac{p(\y, \x, \btheta)}{p(\x \, | \, \btheta, \y)}. \label{eq:hyperpost}
-\end{equation}
-Approximation (i) then uses a Gaussian approximation to the denominator given by
-\begin{equation}
-p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \hat{\bm{\mu}}(\btheta), \hat{\mathbf{Q}}(\btheta)^{-1}). \label{eq:gaussianx}
-\end{equation}
-This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution, given by
-\begin{equation}
-p(\x \, | \, \btheta, \y) \propto
-\exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right),
-\end{equation}
-close to being Gaussian because $\y$ tends not to be strongly informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial].
-<!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference?
-Perhaps what the author (Marta) meant is that the Gaussian distribution is sufficiently flexible to accommodate the updates made by $\y$.
--->
-As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}.
-Taking $\x = \hat{\bm{\mu}}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as
-\begin{equation}
-\tilde p(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)} = \frac{p(\y, \hat{\bm{\mu}}(\btheta), \btheta)}{\det(\hat{\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal}
-\end{equation}
-where the equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\hat{\bm{\mu}}(\btheta)$.
+Given a \textsf{C++} user template `model.cpp` for the unnormalised log posterior, we obtain posterior marginal approximations via the following algorithm:
 
-### Approximation (ii)
+1. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(\btheta, \y)$
+$$
+\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)}
+$$
+where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$.
 
-We now move on to the approximation $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$.
-Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section \ref{sec:approxi} above, a natural approach, and that taken by [@rue2007approximate], is to marginalise this distribution directly to obtain
-\begin{equation}
-\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \hat{\mu}_i(\btheta), 1 / \hat{q}_i(\btheta)),
-\end{equation}
-where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from the relevant entries of $\hat{\bm{\mu}}(\btheta)$ and $\hat{\bm{Q}}(\btheta)$ respectively.
-However, although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode, so is generally not advised.
-Instead, [@rue2009approximate] propose two methods, a Laplace approximation and a simplified version which is less computationally demanding.
-The full Laplace approximation is
-\begin{align}
-  p(x_i \, | \, \btheta, \y) &= p(\x \, | \, \btheta, \y) \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
-  &= \frac{p(\x, \btheta \, | \, \y)}{p(\btheta \, | \, \y)} \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
-  &\propto \frac{p(\x, \btheta \, | \, \y)}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
-  &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat{\bm{\mu}}_{-i}(x_i, \btheta)}
-  = \tilde p_{LA}(x_i \, | \, \btheta, \y), \label{eq:laplaceapproximation}
-\end{align}
-where
-\begin{equation}
-p_G(\x_{-i} \, | \, x_i, \btheta, \y) = \mathcal{N}(\x_{-i} \, | \, \hat{\bm{\mu}}_{-i}(x_i, \btheta), \hat{\bm{Q}}_{-i}(x_i, \btheta)),
-\end{equation}
-is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
-The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well.
-However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $i$, which is often computationally prohibitive.
-Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by [@rue2009approximate] to reduce the computational cost:
-\begin{enumerate}
-\item
-Avoiding having to find the mode via optimisation by using the approximation $\hat{\bm{\mu}}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$
-\item
-As only those $x_j$ close to $x_i$ should have an impact on the marginal of $x_i$, then by selecting some subset $R_i(\btheta)$ of nodes $j$ to impact $j$ the matrix which needs to be factorised can be reduced in dimension to be $| R_i(\btheta) | \times | R_i(\btheta) |$ rather than $n \times n$
-\end{enumerate}
+2. Normalise $\tilde p_\text{LA}(\btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
+$$
+\tilde p_\text{AQ}(\btheta \, | \, \y) = \frac{\tilde p_\text{LA}(\btheta, \y)}{\tilde p_{\text{AQ}}(\y)},
+$$
+where the normalising constant is calculated using nodes $\z_\theta \in \mathcal{Q}(m, k_\theta)$ as
+$$
+\tilde p_{\text{AQ}}(\y) = \sum_{\z_\theta \in \mathcal{Q}(m, k_\theta)} \tilde p_\text{LA}(\btheta(\z_\theta), \y) \omega_\theta(\z_\theta),
+$$
+where $m = \dim(\btheta)$, the number of nodes per dimension is $k_\theta$, and weights $\omega_\theta: \z_\theta \in \mathcal{Q}(m, k_\theta) \to \mathbb{R}$.
+We typically set $k_\theta = 3$ such that there are $3^m$ nodes in total.
+
+3. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(x_i, \btheta, \y)$
+$$
+\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)}
+$$
+where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}_{-i}(x_i, \btheta), \bm{Q}_{-i}(x_i, \btheta)^{-1})$.
+
+4. Normalise $\tilde p_\text{LA}(x_i, \btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
+$$
+\tilde p_\text{AQ}(x_i \, | \, \btheta, \y) = \frac{\tilde p_\text{LA}(x_i, \btheta, \y)}{\tilde p_{\text{AQ}}(\btheta, \y)},
+$$
+where the normalising constant is calculated using nodes $z_x \in \mathcal{Q}(1, k_x)$ as
+$$
+\tilde p_{\text{AQ}}(\btheta, \y) = \sum_{z_x \in \mathcal{Q}(1, k_x)} \tilde p_\text{LA}(x_i(z_x), \btheta, \y) \omega_x(z_x),
+$$
+where the total number of nodes is $k_x$ and weights $\omega_x: z_x \in \mathcal{Q}(1, k_x) \to \mathbb{R}$.
+As this quadrature is in one dimension, we can set $k_x = 5$ or higher.
+
+5. Obtain a nested approximation to the posterior marginal of the $i$th latent effect by
+$$
+\tilde p(x_i \, | \, \y) = \sum_{\z_\theta \in \mathcal{Q}(m, k_\theta)} \tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y) \tilde p_\text{AQ} (\btheta(\z_\theta) \, | \, \y) \omega_\theta(\z_\theta).
+$$
 

### 2022-11-12 src/docs_paper/paper.Rmd
More writing

@@
 keywords: 
-  - Spatial statistics
+  - spatial statistics
+  - small-area estimation
   - INLA
+  - AGHQ
+  - HIV epidemiology
 
@@
 Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-`TMB` is gaining popularity spatial statistics [@osgoodzimmerman2021statistical] owing to its speed and flexibility.
+Owing to its speed and flexibility, `TMB` is gaining popularity spatial statistics [@osgoodzimmerman2021statistical].
 Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
@@
 To obtain fast, accurate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
-INLA [@rue2009approximate] is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
-The central innovation is a way to approximate accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
+INLA is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
+The central innovation of @rue2009approximate is a way to approximate accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
 Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
@@
 We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
-Though as of yet no theory exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2021stochastic using AGHQ.
+Though no theory yet exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2021stochastic using AGHQ.
 We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
@@
 Suppose a simple random household survey of $m^\text{HS}_i$ people is conducted in each area, and $y^\text{HS}_i$ HIV positive cases are observed.
-Cases may be modelled using a binomial logistic regression model
-\begin{align}
+Then, cases may be modelled using a binomial logistic regression model
+\begin{align*}
 y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
 \text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2),
-\end{align}
-where HIV prevalence $\rho^\text{HS}_i$ is modelled by a Gaussian with mean $\beta_\phi$ and standard deviation $\sigma_\phi$.
+\end{align*}
+where $\rho^\text{HS}_i$ is the HIV prevalence, modelled on the logit scale by a Gaussian with mean $\beta_\phi \sim p(\beta_\phi)$ and standard deviation $\sigma_\phi \sim p(\sigma_\phi)$.
 
@@
 Then an analogous binomial logistic regression model
-\begin{align}
+\begin{align*}
 y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
@@
 b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
-\end{align}
+\end{align*}
 may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
-Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence on the logit scale.
+Reflecting the fact that prevalence in ANCs is related, but importantly different, to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence on the logit scale.
+We use a Gaussian distribution with mean $\beta_b \sim p(\beta_b)$ and standard deviation $\sigma_b \sim p(\sigma_b)$ for these bias random effects.
 
@@
 
-The number of people receiving treatment at district health facilities $A_i$ provides further information about HIV prevalence.
-Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
+The number of people receiving treatment at district health facilities $A_i$ provides additional information about HIV prevalence.
+Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa districts with low prevalence are likely to have fewer people recieving treatment.
 ART coverage, defined to be the proportion of people living with HIV (PLHIV) currently on ART on district $i$, is given by $\alpha_i = A_i / \rho^\text{HS}_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be constant.
 As such, ART coverage may also be modelled using a binomial logistic regression model
-\begin{align}
+\begin{align*}
 A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
 \text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2),
-\end{align}
+\end{align*}
 where the proportion of people receiving ART is $\rho^\text{HS}_i \alpha_i$.
+We again use a Gaussian distribution with mean $\beta_\alpha \sim p(\beta_\alpha)$ and standard deviation $\sigma_\alpha \sim p(\sigma_\alpha)$.
 Here we assume no travel between districts to receive treatment.
@@
 
-Let $\y = (\y^\text{HS}, \y^\text{ANC}, \bm{A})$ be the complete response vector.
-In this section, we would like to set-up notation for the joint model, helping to show that it isn't a latent Gaussian model, but is an extended latent Gaussian model.
-When introducing LGM and ELGM below, we may refer back to this model as example.
-For example, we can note that each of the components individually might be latent Gaussian (perhaps for the ART component it depends if you consider $\rho^\text{HS}_i$ to be fixed) but when combined they are no longer.
+The three components above may be combined to a single model as follows
+\begin{align*}
+y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
+y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
+A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
+\text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2), \\
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho^\text{HS}_i) + b_i, \\
+b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2), \\
+\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
+\end{align*}
 
@@
 
-<!-- ## Integrated nested Laplace approximation\label{sec:inla} -->
-
-<!-- Consider a latent Gaussian model (LGM) of the form -->
-<!-- \begin{alignat}{2} -->
-<!-- &\text{(Observations)}     &        y_i &\sim p(y_i \, | \, x_i, \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\ -->
-<!-- &\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\ -->
-<!-- &\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta), \label{eq:parameters} -->
-<!-- \end{alignat} -->
-<!-- where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$. -->
-<!-- For such models, the joint posterior of $(\x, \btheta)$ is given by -->
-<!-- \begin{equation} -->
-<!--   p(\x, \btheta \, | \, \y) -->
-<!--   \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right).
-\label{eq:lgmjointposterior} -->
-<!-- \end{equation} -->
-<!-- Rather than approximating this joint posterior, the INLA method proceeds by approximation of the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by -->
-<!-- \begin{align} -->
-<!--   p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\ -->
-<!--   p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2} -->
-<!-- \end{align} -->
-<!-- An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$.
--->
-<!-- We discuss these two approximations in turn below. -->
-
-<!-- ### Approximation (i) \label{sec:approxi} -->
-
-<!-- The posterior marginal of the parameters $p(\btheta \, | \, \y)$ appears in both Equations \eqref{eq:inla1} and \eqref{eq:inla2}. -->
-<!-- This distribution is approximated by $\tilde p(\btheta \, | \, \y)$ and represented by a set of $K$ integration nodes $\{ \btheta^{(k)} \}$ and area-weights $\{ \Delta^{(k)} \}$. -->
-<!-- The first step is to rewrite $p(\btheta \, | \, \y)$ as -->
-<!-- \begin{equation} -->
-<!--   p(\btheta \, | \, \y) -->
-<!--   = \frac{p(\x, \btheta \, | \, \y)}{p(\x \, | \, \btheta, \y)} -->
-<!--   \propto \frac{p(\y, \x, \btheta)}{p(\x \, | \, \btheta, \y)}. \label{eq:hyperpost} -->
-<!-- \end{equation} -->
-<!-- Approximation (i) then uses a Gaussian approximation to the denominator given by -->
-<!-- \begin{equation} -->
-<!-- p(\x \, | \, \btheta, \y) \approx p_\text{G}(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \bm{\mu}(\btheta), {\mathbf{Q}}(\btheta)^{-1}). \label{eq:gaussianx} -->
-<!-- \end{equation} -->
-<!-- This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution, given by -->
-<!-- \begin{equation} -->
-<!-- p(\x \, | \, \btheta, \y) \propto -->
-<!-- \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right), -->
-<!-- \end{equation} -->
-<!-- close to being Gaussian because $\y$ tends not to be strongly informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial]. -->
-<!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference?
-Perhaps what the author (Marta) meant is that the Gaussian distribution is sufficiently flexible to accommodate the updates made by $\y$.
--->
-<!-- As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}. -->
-<!-- Taking $\x = \bm{\mu}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as -->
-<!-- \begin{equation} -->
-<!-- \tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} = \frac{p(\y, \bm{\mu}(\btheta), \btheta)}{\det({\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal} -->
-<!-- \end{equation} -->
-<!-- where the equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\bm{\mu}(\btheta)$. -->
-<!-- We describe this as a Laplace approximation to the parameter posterior. -->
-
-<!-- ### Approximation (ii) -->
-
-<!-- We now move on to the approximation $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$. -->
-<!-- Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ for the denominator in Section \ref{sec:approxi} above, a natural approach [@rue2007approximate] is to marginalise this distribution directly to obtain -->
-<!-- \begin{equation} -->
-<!-- \tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, {\mu}_i(\btheta), 1 / {q}_i(\btheta)), -->
-<!-- \end{equation} -->
-<!-- where the marginal mean ${\mu}_i(\btheta)$ and precision ${q}_i(\btheta)$ are recovered directly from the relevant entries of $\bm{\mu}(\btheta)$ and ${\bm{Q}}(\btheta)$ respectively. -->
-<!-- However, although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode, so is generally not advised. -->
-<!-- Instead, @rue2009approximate propose two methods, a Laplace approximation and a simplified version which is less computationally demanding. -->
-<!-- The full Laplace approximation is -->
-<!-- \begin{align} -->
-<!--   p(x_i \, | \, \btheta, \y) &= p(\x \, | \, \btheta, \y) \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
-<!--   &= \frac{p(\x, \btheta \, | \, \y)}{p(\btheta \, | \, \y)} \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
-<!--   &\propto \frac{p(\x, \btheta \, | \, \y)}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
-<!--   &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)} -->
-<!--   = \tilde p_{LA}(x_i \, | \, \btheta, \y), \label{eq:laplaceapproximation} -->
-<!-- \end{align} -->
-<!-- where -->
-<!-- \begin{equation} -->
-<!-- p_G(\x_{-i} \, | \, x_i, \btheta, \y) = \mathcal{N}(\x_{-i} \, | \, \bm{\mu}_{-i}(x_i, \btheta), {\bm{Q}}_{-i}(x_i, \btheta)), -->
-<!-- \end{equation} -->
-<!-- is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\bm{\mu}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\bm{\mu}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\bm{\mu}_{-i}(x_i, \btheta)$ is not the same as $\bm{\mu}_{-i}(\btheta)$.} -->
-<!-- The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well. -->
-<!-- However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $i$, which is often computationally prohibitive. -->
-<!-- Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by @rue2009approximate to reduce the computational cost: -->
-<!-- \begin{enumerate} -->
-<!-- \item -->
-<!-- Avoiding having to find the mode via optimisation by using the approximation $\bm{\mu}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$ -->
-<!-- \item -->
-<!-- As only those $x_j$ close to $x_i$ should have an impact on the marginal of $x_i$, then by selecting some subset $R_i(\btheta)$ of nodes $j$ to impact $j$ the matrix which needs to be factorised can be reduced in dimension to be $| R_i(\btheta) | \times | R_i(\btheta) |$ rather than $n \times n$ -->
-<!-- \end{enumerate} -->
-
 Consider a latent Gaussian model (LGM) of the form
-\begin{alignat}{2}
-&\text{(Observations)}     &        \y &\sim p(\y \, | \, \x, \btheta), \label{eq:data} \\
-&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\
-&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta), \label{eq:parameters}
-\end{alignat}
+\begin{alignat*}{2}
+&\text{(Observations)}     &        \y &\sim p(\y \, | \, \x, \btheta), \\
+&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \\
+&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta),
+\end{alignat*}
 where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
 The joint posterior of $(\x, \btheta)$ is given by
-\begin{equation}
+\begin{equation*}
   p(\x, \btheta \, | \, \y)
-  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right). \label{eq:lgmjointposterior}
-\end{equation}
+  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right).
+\end{equation*}
 We consider approximations to the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
 \begin{align}
-  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
-  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
+  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
+  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
 \end{align}
@@
 
-Given a \textsf{C++} user template `model.cpp` for the unnormalised log posterior, we obtain posterior marginal approximations via the following algorithm:
+Given a \textsf{C++} user template `model.cpp` for the unnormalised log posterior, we obtain the posterior marginal approximations given in Equations \eqref{eq:inla1} and \eqref{eq:inla2} via the following algorithm:
 
@@
 
-3. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(x_i, \btheta, \y)$
+3. Obtain a nested approximation to the posterior marginal of the $i$th latent effect by
+$$
+\tilde p(x_i \, | \, \y) = \sum_{\z_\theta \in \mathcal{Q}(m, k_\theta)} \tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y) \tilde p_\text{AQ} (\btheta(\z_\theta) \, | \, \y) \omega_\theta(\z_\theta).
+$$
+The nodes and weights $\{\mathcal{Q}(m, k_\theta), \omega_\theta\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
+For each of the $k_\theta^m$ values of $\btheta(\z_\theta)$ we obtain $\tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y)$ by setting $\btheta = \btheta(\z_\theta)$ in the following two steps.
+
+a) Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(x_i, \btheta, \y)$
 $$
@@
 
-4. Normalise $\tilde p_\text{LA}(x_i, \btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
+b) Normalise $\tilde p_\text{LA}(x_i, \btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
 $$
@@
 
-5. Obtain a nested approximation to the posterior marginal of the $i$th latent effect by
-$$
-\tilde p(x_i \, | \, \y) = \sum_{\z_\theta \in \mathcal{Q}(m, k_\theta)} \tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y) \tilde p_\text{AQ} (\btheta(\z_\theta) \, | \, \y) \omega_\theta(\z_\theta).
-$$
-
 # Application to the Naomi model\label{sec:results}
 
-The \textsc{R} [@r] code used to implement the model and produce all results we describe is available at `github.com/athowes/elgm-inf`.
-The inference method is available in versions 0.5.0. onwards of the `aghq` package.
+* The \textsc{R} [@r] code used to produce all results we describe is available at `github.com/athowes/elgm-inf`. The inference method is available in versions 0.5.0. onwards of the `aghq` package
+* Using the `TMB` template, we fit the model using four inferential approaches: (1) empirical Bayes combined with a Gaussian approximation, (2) AGHQ combined with a Gaussian approximation, (3) AGHQ combined with a Laplace approximation and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS).
+We treat results from NUTS as the gold-standard
+* We used the Kolmogorov-Smirnov test for the maximum difference between two empirical cumulative distribution functions to compare posterior marginal distributions
+* We performed posterior predictive checks to assess the coverage of our estimates via the uniformity of the data within each posterior marginal distribution
+
+<!-- \begin{table}[] -->
+<!-- \small -->
+<!-- \begin{tabularx}{\textwidth}{llll} -->
+<!-- \toprule -->
+<!-- Inference method & KS & MMD & PSIS \\ -->
+<!-- \midrule -->
+<!-- Empirical Bayes, Gaussian & 0 & 0 & 0 \\ -->
+<!-- AGHQ, Gaussian & 0 & 0 & 0 \\ -->
+<!-- AGHQ, Laplace & 0 & 0 & 0 \\ -->
+<!-- NUTS & 0 & 0 & 0 \\ -->
+<!-- \bottomrule -->
+<!-- \end{tabularx} -->
+<!-- \label{tab:results} -->
+<!-- \small -->
+<!-- \end{table} -->
 
@@
 
-# Appendix {#appendix .unnumbered}
+* We developed an approximate Bayesian inference algorithm to solve a challenging problem in the small-area estimation of HIV in low resource settings
+* The flexibility of our method implementation, including compatibility with any `TMB` \textsc{C++} template, allows broader use, as well as investigation of, deterministic inference methods than had previously been possible
+* We demonstrated a Bayesian workflow for deterministic inference methods
 
-# Acknowledgements {#acknowledgements .unnumbered}
+# Acknowledgements {-}
 
 AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1).
-
-# References {#references .unnumbered}

### 2022-11-14 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Working on slides for tomorrow's seminar

@@
   * Prevalence $\rho$: the proportion of people who are HIV positive
+  * Treatment coverage $\alpha$: the proportion of PLHIV on treatment
   * Incidence $\lambda$: the proportion of people newly infected 
-  * Treatment coverage $\alpha$: the proportion of PLHIV on treatment
-* We would like to provide them at a local, district-level
+* Aim to provide estimates at a district-level to enable \textcolor{hilit}{precision public health}
 
@@
 \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-This is a challenging task! Data is noisy, sparse and biased. $\implies$ compelling case for thoughtful Bayesian modelling.
+This is a challenging task! Data is noisy, sparse and biased $\implies$ compelling case for thoughtful Bayesian modelling 
 \end{tcolorbox}
@@
 
-* Consider small-areas $i = 1, \ldots, n$ like the districts of a country
-* Simple random sample household-survey of size $m^\text{HS}_i$ in each area
-* The number of people testing positive for HIV is $y^\text{HS}_i$
-* You could calculate direct estimates of prevalence by $y^\text{HS}_i / m^\text{HS}_i$ but because the survey is powered at a national-level, the sample sizes are small and these estimates would be noisy
+* Consider "small-areas" $i = 1, \ldots, n$ (e.g. districts of a country)
+* Simple random sample household-survey^[In reality a complex survey design is used, often with urban rural stratification.] of size $m^\text{HS}_i$ where $y^\text{HS}_i$ people testing positive for HIV
+* Could calculate direct estimates of prevalence by $y^\text{HS}_i / m^\text{HS}_i$
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+Because the survey is powered at a national-level, the $m^\text{HS}_i$ are small and direct estimates would be noisy $\implies$ use a model to smooth estimates
+\end{tcolorbox}
+\end{center}
 
@@
 * We usually set up $g$ as a Gaussian spatial smoother
-* This allows for pooling of information between districts
+* This allows for \textcolor{hilit}{pooling of information} between districts
+
+#
+
+```{r echo=FALSE, warning=FALSE, out.width="350px", fig.cap = "The Besag model, $\\phi_i \\, | \\, \\phi_{-i} \\sim \\mathcal{N} \\left(\\frac{1}{n_{\\delta i}} \\sum_{j: j \\sim i} \\phi_j, \\frac{1}{n_{\\delta i}\\tau_\\phi} \\right)$."}
+
+areas <- filter(sf::read_sf("zwe_areas.geojson"), area_level == 2)
+
+areas %>%
+  filter(area_level == 2) %>%
+  bsae::plot_graph(add_geography = TRUE)
+```
 
@@
 * Interested in learning both $(\btheta, \x)$ from data $\y$
-* If the middle layer is Gaussian, then it's a latent Gaussian model
+* If the middle layer is Gaussian, then it's a \textcolor{hilit}{latent Gaussian model}
 \begin{equation*}
@@
 \end{equation*}
-* Covers most of the models commonly used in spatiotemporal statistics
 * Latent field is typically indexed by spatiotemporal location, such that $n > m$
@@
 
-* A typical household survey takes around $UNKNOWN to run
-* This means that they don't happen very often
-* For this reason, the information might be quite out of date, and difficult to base policy on
+* Household surveys cost millions to run so they don't happen very often
+* e.g. DHS include 5k-30k households, and occurs around every 5 years
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+The snapshot they provide can be quite out of date, and difficult to base effective policy on $\implies$ need to use routinely collected data to help here
+\end{tcolorbox}
+\end{center}
 
@@
 
-* Pregnant women attending antenatal care clinics are routinely tested for HIV, to avoid mother-to-child transmission
-* This data source is more real-time than household surveys, but it's also more biased, because attendees are unlikely to be as representative of the population
-* But perhaps this bias is consistent, in which case we can still make use of the ANC data to supplement our model!
+* Pregnant women attending antenatal care clinics are routinely tested for HIV, to avoid mother-to-child transmission. This data source is:
+ 1. More \textcolor{hilit}{real-time} than household surveys -- can be collected e.g. monthly
+ 2. More \textcolor{hilit}{biased} than household surveys -- attendees are not representative
+* If the this bias is consistent, we can still ANC data to supplement our model
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+$\implies$ model the level using the household survey data, and the trend using the ANC data
+\end{tcolorbox}
+\end{center}
 
@@
 
-* Suppose of $m^\text{ANC}_i$ women attending ANC, $y^\text{ANC}_i$ are HIV positive, then we can use another binomial logistic regression:
+* Suppose of $m^\text{ANC}_i$ ANC attendees, $y^\text{ANC}_i$ are HIV positive, and model
 \begin{align*}
@@
 
-* We're also interested in what proportion $\alpha_i$ of people living with HIV (PLHIV) are receiving treatment
-* Suppose we record $A_i$ attendees from a known population of $N_i$ in each district
-* We can use another logistic regression model
+* Also interested in what proportion $\alpha_i$ of people living with HIV are receiving treatment, which may also be informative about prevalence
+* If we record $A_i$ attendees from a known population of $N_i$ in each district, then this can be modelled by
 \begin{align*}
@@
 \end{align*}
+* To be more sophisticated, you can also model the movement of people to receive treatment in districts other than the one they live in
 
@@
 * Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
-* Can't run long MCMC in this setting, so we require fast, accurate, approximations 
+* Can't run long MCMC in this setting, so we \textcolor{hilit}{require fast, accurate, approximations}
 * It's a complicated model, and requires something more flexible than `R-INLA`
@@
 * `TMB` [@kristensen2015tmb] is an R package which implements the Laplace approximation for latent variable models
-* To get started, write an objective function $f(\x, \btheta)$ in `TMB` C++ syntax
+* I use "Laplace approximation" to mean approximating the normalising constant with Laplace's method^[Rather than approximating the posterior with a Gaussian, which I call a Gaussian approximation.]
+* To get started with `TMB`, write your $f(\x, \btheta)$ in `TMB`'s C++ syntax
 * As pseudo-Bayesians, we choose the log-posterior
@@
 $$
-
-# Template Model Builder
-
-For example, for the model
-$$
-\y \sim \mathcal{N}(\mu, 1)
-$$
-with $p(\mu) \propto 1$ then the `TMB` user template looks like...
+* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks as follows
 
@@
 $$
-\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)} \label{eq:hypermarginal}
+\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} \label{eq:hypermarginal}
 $$
 * Integrate out a Gaussian approximation $\tilde p_\text{G}(\x \, | \, \btheta, \y)$ to the latent field
-* `TMB` uses automatic differentiation [@griewank2008evaluating] via `CppAD`
+* `TMB` uses automatic differentiation [@griewank2008evaluating] via `CppAD` to do this, as well as help with numerical optimisation routines
+* TO-DO: how to describe the approximation for $\x$ with `TMB`
 
@@
 
-# Step 1)
+# Step 1
 
@@
 \begin{equation}
-\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}^\star(\btheta)} \label{eq:hypermarginal}
+\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} \label{eq:hypermarginal}
 \end{equation}
@@
 
-# Step 2)
+# Step 2
 
@@
   * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design
-  * We plan to use adaptive Gaussian Hermite quadrature (AGHQ), which has recently been shown to have theoretical guarantees [@bilodeau2021stochastic]
+  * We plan to use adaptive Gaussian Hermite quadrature (AGHQ), which has recently been shown to have theoretical guarantees [@bilodeau2021stochastic] and is implemented in the `aghq` R package [@stringer2021implementing]
 
@@
 
-* Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
-* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) to automatically choose the node locations
-  * Does not require manual tuning!
+* Gauss-Hermite quadrature is one way to pick nodes and weights, and is based on the theory of polynomial interpolation
+* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) $\implies$ does not require manual parameter tuning!
 * Works particularly well when the integrand is pretty Gaussian
-* Use $k$ quadrature nodes per dimension, for example if $k = 3$ then $3^m$ total nodes
-* Implemented in the `aghq` R package. See vignette @stringer2021implementing
+* Use $k$ quadrature nodes per dimension, e.g. if $k = 3$ then $3^m$ total nodes
 
@@
 
-```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("One dimensional example of AGHQ. \\label{fig:aghq-1d}")}
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("One dimensional example of AGHQ with $3^1 = 3$ nodes. \\label{fig:aghq-1d}")}
 knitr::include_graphics("depends/aghq-1d.pdf")
@@
 
-```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Two dimensional example of AGHQ. \\label{fig:aghq-2d}")}
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Two dimensional example of AGHQ with $3^2 = 9$ nodes. \\label{fig:aghq-2d}")}
 knitr::include_graphics("depends/aghq-2d.pdf")
@@
 
-# Step 3)
+# Step 3
 
@@
 \begin{equation}
-\tilde p_\text{G}(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta))
+\tilde p_\text{G}(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu_i(\btheta), 1 / q_i(\btheta))
 \end{equation}
   * In `R-INLA`, the above is referred to as `method = "gaussian"` 
-  * There are two better, more complex approximations, confusingly called `"simplified.laplace"` and `"laplace"`
-  * Uses sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)
+  * This is also what is currently used in `aghq`
 
-# Step 4)
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+There are more accurate (and complicated) versions which I will talk about in a minute!
+\end{tcolorbox}
+\end{center}
+
+# Step 4
 
 4) Finally, use quadrature to combine
-  * our approximation $\tilde p_\text{LA}(\btheta \, | \, \y)$ from step 1),
-  * some choice of integration nodes and weights $\{ \btheta(\z), \omega(\z) \}$ from step 2),
-  * some choice of approximation $\tilde p(x_i \, | \, \btheta, \y)$ from step 3)
+  * our approximation $\tilde p_\text{LA}(\btheta \, | \, \y)$ from Step 1,
+  * some choice of integration nodes and weights $\{ \btheta(\z), \omega(\z) \}$ Step 2,
+  * some choice of approximation $\tilde p(x_i \, | \, \btheta, \y)$ from Step 3
 to give
@@
 
-# Experiments
+# Using a Laplace approximation for Step 3
 
-* We wrote a simplified version of the Naomi model up in `TMB`
-* This allowed us to test three inference methods **all using precisely the same model and C++ code**
+* Rather than taking the marginals of $\tilde p_\text{G}(\x \, | \, \btheta, \y)$ we could calculate \textcolor{hilit}{a new Laplace approximation} for each $x_i$
+$$
+\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)}
+$$
+where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}_{-i}(x_i, \btheta), \bm{Q}_{-i}(x_i, \btheta)^{-1})$
+* Problem: $n$ can be big, and we will need to recalculate this for each $(x_i, \btheta)$
+* So for example, if we use a nested AGHQ strategy with $k_x$ and $k_\theta$ then we will need $k_x k_\theta^m$ recalculations for each $i$
 
-1. A direct Gaussian approximation via `TMB`
-2. Adaptive Gaussian Hermite quadrature via `aghq`
-3. No-U-Turn Sampling (NUTS -- a type of Hamiltonian Monte Carlo) via `tmbstan`
+# Cheaper approximate approximations
 
-* Note: using different software it is usually very difficult to ensure the model is precisely the same, so we're very fortunate here
+* @rue2009approximate found a way to do this in a cheaper and more approximate way based on assuming a \textcolor{hilit}{sparse} precision for $\x$
+  * a.k.a. that $\x$ is a Gaussian Markov random field (GMRF)
+* @wood2020simplified extended their approximation to work for the case when $\x$ does not have a sparse precision
+
+# Epilepsy example
+
+* Notebook for this example at `athowes.github.io/elgm-inf/epil`
+* From Section 5.2. of @rue2009approximate previously from BUGS manual
+* Patients $i = 1, \ldots, 59$ each either assigned treatment $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$ to help with seizures
+* Visits to clinics $j = 1, \ldots, 4$ times with $y_{ij}$ the number of seizures of the $i$th person in the two weeks proceeding their $j$th visit to the clinic
+* Covariates age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$
+
+# Epilepsy example
+
+The model is a Poisson GLMM:
+\begin{align*}
+  y_{ij}          &\sim \text{Poisson}(\lambda_{ij}), \\
+  \lambda_{ij}    &= e^{\eta_{ij}}, \\
+  \eta_{ij}       &= \beta_0 + \beta_{\texttt{Base}} \log(\texttt{Baseline}_j / 4) + \beta_{\texttt{Trt}} \texttt{Trt}_i +
+                     \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
+                  &+ \beta_{\texttt{Age}} \log(\texttt{Age}_i) + \beta_{\texttt{V}_4} {\texttt{V}_4}_j +
+                     \epsilon_i + \nu_{ij}, \quad i=1:59, \quad j=1:4, \\
+  \beta           &\sim \mathcal{N}(0, 100^2), \quad \forall \beta, \\
+  \epsilon_i      &\sim \mathcal{N}(0, 1/\tau_\epsilon), \\
+  \nu_{ij}        &\sim \mathcal{N}(0, 1/\tau_\nu), \\
+  \tau_\epsilon   &\sim \Gamma(0.001, 0.001), \\
+  \tau_\nu        &\sim \Gamma(0.001, 0.001).
+\end{align*}
+
+# Inference
+
+Implement the following inference procedures:
+
+1. HMC NUTS via `tmbstan` and `TMB`
+2. Grid with Gaussian marginals via `R-INLA`
+3. Grid with simplified Laplace marginals via `R-INLA`
+4. Grid with Laplace marginals via `R-INLA`
+5. EB with Gaussian marginals via `TMB`
+6. AGHQ with Gaussian marginals via `aghq` and `TMB`
+7. **EB with Laplace marginals via `aghq` and `TMB`**^[I'm working on AGHQ with Gaussian marginals via `aghq` and `TMB`. I am using the `aghq` package, just with $k = 1$ corresponding to EB]
+
+# Results
 
@@
 
-* You could look at the summaries like the mean and standard deviation of each of the posterior marginals
-  * Any approximation method should be pretty good at getting the mean right
-  * Gaussian approximations should be good at getting the second moment right
-* It's probably better to compare the whole posterior distributions
+* You could look at the summaries like the mean and standard deviation of each of the posterior marginals but it's better to compare the whole posterior distributions
 * One way to do this is via Kolmogorov-Smirnov statistics, which give the maximum difference between two empirical CDFs
 
+# Prevalence, ANC, ART example
+
+* Notebook for this example at `athowes.github.io/elgm-inf/prev-anc-art`
+* Simulate data from model with all three components and particular (known) parameter values
+
+# Inference
+
+Implement the following inference procedures:
+
+1. HMC NUTS via `tmbstan` and `TMB`
+2. EB with Gaussian marginals via `TMB`
+3. AGHQ with Gaussian marginals via `aghq` and `TMB`
+
+* All of these approaches share the same C++ template, so the models are identical!
+This is often very difficult to ensure, so we're very fortunate here^[i.e.
+thanks to Kasper and Alex for making `tmbstan` and ``aghq` respectively!]
+
+# Results
+
+# Conclusions
+
+> My main comment is that several aspects of the computational machineery that is presented by Rue and his colleagues **could benefit from the use of a numerical technique known as automatic differentiation (AD)** ...
+By the use of AD one could obtain a system that is automatic from a user's perspective...
+the benefit would be a fast, flexible and easy-to-use system for doing Bayesian analysis in models with Gaussian latent variables
+
+-- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate
+
+# Conclusions
+
+* Hopeful that we'll give fast, accurate inferences for Naomi!
+* Implementation as a part of `aghq` combining simplified INLA and AGHQ, enabled by automatic differentiation, will provide flexible use of the method
+  * Will be of interest to advanced users of `R-INLA` who would like specify models outside a formula interface (similar to users of `brms` v.s. Stan)
+  * This describes many in the HIV inference group `hiv-inference.org`^[See `athowes.github.io/inla-sandbox/` for some examples of understanding `R-INLA` internals.]
+
 # References {.allowframebreaks}

### 2022-11-14 src/explore_aghq/aghq.Rmd
Fix plot typo
Working on slides for tomorrow's seminar

@@
 $$
-f(x) = \exp(-\frac{x^2}{2}),
+f(\theta) = \exp(-\frac{\theta^2}{2}),
 $$
-which we would like to integrate over $\mathbb{R}$, i.e. $\int_{-\infty}^\infty f(x) \text{d}x$.
+which we would like to integrate over $\mathbb{R}$, i.e. $\int_{-\infty}^\infty f(\theta) \text{d}\theta$.
 
 ```{r}
-f <- function(x) {
-  exp(-(x^2) / 2)  
+f <- function(theta) {
+  exp(-(theta^2) / 2)  
 }
@@
     geom_line(col = cbpalette[1]) +
-    labs(x = "x", y = "f(x)") +
+    labs(x = "theta", y = "f(theta)") +
     theme_minimal()
@@
 ```{r}
-f <- function(x) {
-  prod(exp(-(x^2) / 2))
+f <- function(theta) {
+  prod(exp(-(theta^2) / 2))
 }
@@
 df <- expand.grid(
-  x1 = seq(-3, 3, by = 0.1),
-  x2 = seq(-3, 3, by = 0.1)
+  theta1 = seq(-3, 3, by = 0.1),
+  theta2 = seq(-3, 3, by = 0.1)
 )
@@
   mutate(
-    y = f(c(x1, x2))
+    y = f(c(theta1, theta2))
   ) %>%
   ggplot() +
-    geom_contour(aes(x = x1, y = x2, z = y), col = multi.utils::cbpalette()[1]) +
+    geom_contour(aes(x = theta1, y = theta2, z = y), col = multi.utils::cbpalette()[1]) +
     theme_minimal()
@@
   ) +
-  labs(col = "f(x1, x2)", size = "Weight") +
+  labs(col = "f(theta1, theta2)", size = "Weight") +
   scale_size_continuous(limits = c(2, 4))
@@
 
-ggsave(filename = "aghq-2d.pdf", plot = plot, height = 4, width = 6.25)
+ggsave(filename = "aghq-2d.pdf", plot = plot2, height = 4, width = 6.25)
 ```

### 2022-11-15 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Iterate slides forward. Now just to add some results

@@
 \end{alignat*}
-where $\y = (y_1, \ldots, y_n)$, $\x = (x_1, \ldots, x_n)$, $\btheta = (\theta_1, \ldots, \theta_m)$
+where $\y = (y_1, \ldots, y_n)$, $\x = (x_1, \ldots, x_N)$, $\btheta = (\theta_1, \ldots, \theta_m)$
 * Interested in learning both $(\btheta, \x)$ from data $\y$
@@
 \end{equation*}
-* Latent field is typically indexed by spatiotemporal location, such that $n > m$
+* Latent field is typically indexed by spatiotemporal location, such that $N > m$
 
@@
 * To get started with `TMB`, write your $f(\x, \btheta)$ in `TMB`'s C++ syntax
-* As pseudo-Bayesians, we choose the log-posterior
+* As pseudo-Bayesians, we choose (something proportional to) the log-posterior
 $$
@@
 * `TMB` uses automatic differentiation [@griewank2008evaluating] via `CppAD` to do this, as well as help with numerical optimisation routines
-* TO-DO: how to describe the approximation for $\x$ with `TMB`
+* We then optimise this to obtain a mode $\hat \btheta$, and a Hessian $\bm{H}$ at the mode
 
@@
 \begin{align}
-  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
+  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, N, \label{eq:inla1} \\
   p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
@@
 * Notice that this is the same object we had been working with in `TMB`
-* We use this approximation \textcolor{hilit}{nested} within integrals -- hence the name INLA
+* We will use this approximation \textcolor{hilit}{nested} within integrals like this one
+$$
+\int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) \tilde p_\text{LA}(\btheta \, | \, \y) \text{d}\btheta
+$$
+hence the name INLA
 
@@
 
-* Gauss-Hermite quadrature is one way to pick nodes and weights, and is based on the theory of polynomial interpolation
-* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) $\implies$ does not require manual parameter tuning!
+* Gauss-Hermite quadrature is one way to pick nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega(\z): \mathcal{Q}(m, k) \to \mathbb{R}$, based on the theory of polynomial interpolation
+* The \textcolor{hilit}{adaptive} part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) so that $\btheta(\z) = \hat \btheta + \bm{L}\z$
 * Works particularly well when the integrand is pretty Gaussian
@@
 
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+Key benefits: no manual tuning, works well (and starting to get some theory) in statistical contexts
+\end{tcolorbox}
+\end{center}
 #
 
-```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("One dimensional example of AGHQ with $3^1 = 3$ nodes. \\label{fig:aghq-1d}")}
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("One dimensional example of AGHQ with $3^1 = 3$ nodes. If $k$ is odd then the mode is always included. \\label{fig:aghq-1d}")}
 knitr::include_graphics("depends/aghq-1d.pdf")
@@
 
-```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Two dimensional example of AGHQ with $3^2 = 9$ nodes. \\label{fig:aghq-2d}")}
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Two dimensional example of AGHQ with $3^2 = 9$ nodes.
+Here we use the product rule so that the points in 2D are just 1D x 1D.
+\\label{fig:aghq-2d}")}
 knitr::include_graphics("depends/aghq-2d.pdf")
@@
 \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-There are more accurate (and complicated) versions which I will talk about in a minute!
+There are more accurate (and complicated) versions which I will talk briefly about in a minute!
 \end{tcolorbox}
@@
 
-* Rather than taking the marginals of $\tilde p_\text{G}(\x \, | \, \btheta, \y)$ we could calculate \textcolor{hilit}{a new Laplace approximation} for each $x_i$
+* Previously had been taking the marginals of $\tilde p_\text{G}(\x \, | \, \btheta, \y)$
+* Alternative: calculate \textcolor{hilit}{a new Laplace approximation} for each $x_i$
 $$
@@
 where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}_{-i}(x_i, \btheta), \bm{Q}_{-i}(x_i, \btheta)^{-1})$
-* Problem: $n$ can be big, and we will need to recalculate this for each $(x_i, \btheta)$
-* So for example, if we use a nested AGHQ strategy with $k_x$ and $k_\theta$ then we will need $k_x k_\theta^m$ recalculations for each $i$
+* Problem: $N$ can be big, and we will need to recalculate this for each $(x_i, \btheta)$
+* Ideas like using $\bm{\mu}(\btheta)_{-i}$ to initialise Newton optimisation to find $\bm{\mu}_{-i}(x_i, \btheta)$ could help
 
@@
 
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+Plan: see how long a naive version without these modifications takes, then use this work to get speed-ups as required
+\end{tcolorbox}
+\end{center}
+
 # Epilepsy example
 
-* Notebook for this example at `athowes.github.io/elgm-inf/epil`
-* From Section 5.2. of @rue2009approximate previously from BUGS manual
+* Replication of example from Section 5.2. of @rue2009approximate, and previously from BUGS manual
 * Patients $i = 1, \ldots, 59$ each either assigned treatment $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$ to help with seizures
@@
 * Covariates age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+Notebook for this example at \texttt{athowes.github.io/elgm-inf/epil}
+\end{tcolorbox}
+\end{center}
 
@@
 
+# Thanks for listening!
+
+* Joint work with Alex Stringer (Waterloo) and my supervisors Seth Flaxman (Oxford) and Jeff Eaton (Imperial)
+* The code for this project is at `github.com/athowes/elgm-inf`
+* You can find me online at `athowes.github.io`
+
 # References {.allowframebreaks}

### 2022-11-15 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Slides up!

@@
 
-# Results
+#
 
-# Comparison approach
+```{r echo=FALSE}
+df <- readRDS("depends/comparison-results.rds") %>%
+  select("tmbstan", "R-INLA_G", "R-INLA_SL", "R-INLA_L", "TMB", "aghq", "adam")
 
-* You could look at the summaries like the mean and standard deviation of each of the posterior marginals but it's better to compare the whole posterior distributions
+beta_i <- function(i) { c(paste0("$\\mathbb{E}[\\beta_", i, "]$"), paste0("$\\text{sd}[\\beta_", i, "]$")) }
+rownames(df) <- c(sapply(0:5, beta_i))
+
+names(df) <- c("\\texttt{tmbstan}", "\\texttt{R-INLA}-G", "\\texttt{R-INLA}-SL", "\\texttt{R-INLA}-L", "\\texttt{TMB}", "\\texttt{aghq}", "\\texttt{adam}")
+
+df %>%
+  kableExtra::kable(format = "latex", align = "l", booktabs = TRUE, longtable = TRUE, digits = 2, escape = FALSE)
+```
+
+#
+
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("The intercept parameter has the greatest difference between the Gaussian and Laplace approaches.
+The results in pink are from HMC NUTS.")}
+knitr::include_graphics("depends/intercept-comparison.pdf")
+```
+
+# Comparison approaches
+
+* You could look at the summaries like the mean and standard deviation of each of the posterior marginals as we have above
+* It worked for $\beta_0$, but usually this isn't very informative, and it's better to compare the whole posterior distributions
 * One way to do this is via Kolmogorov-Smirnov statistics, which give the maximum difference between two empirical CDFs
+* Also considering other approaches!
+  * PSIS: is your approximate distribution a good importance sampling proposal for your target? If not, maybe there is an issue!
+  * SBC: generating $(\theta, y)$ first $\theta$ then $y \ | \ \theta$ should be the same as first $y$ then then $\theta \ | \ y$
+  * MMD: compute a distance using kernels (e.g. Gaussian)
 
@@
 
-* Notebook for this example at `athowes.github.io/elgm-inf/prev-anc-art`
 * Simulate data from model with all three components and particular (known) parameter values
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+Notebook for this example at \texttt{athowes.github.io/elgm-inf/prev-anc-art}
+\end{tcolorbox}
+\end{center}
 
@@
 # Results
+
+```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Example KS results from five simulated datasets.")}
+knitr::include_graphics("depends/ks-example.pdf")
+```
 
@@
 
-* Joint work with Alex Stringer (Waterloo) and my supervisors Seth Flaxman (Oxford) and Jeff Eaton (Imperial)
+* Joint work with Alex Stringer (Waterloo) and my PhD supervisors Seth Flaxman (Oxford) and Jeff Eaton (Imperial)
 * The code for this project is at `github.com/athowes/elgm-inf`

### 2022-11-15 src/epil/epil.Rmd
Slides up!

@@
   data.frame(fit2$marginals.fixed$`(Intercept)`) %>%
-    mutate(method = "G"),
-  data.frame(fit3$marginals.fixed$`(Intercept)`) %>%
-    mutate(method = "SL"),
+    mutate(method = "Gaussian"),
   data.frame(fit4$marginals.fixed$`(Intercept)`) %>%
-    mutate(method = "L")
+    mutate(method = "Laplace")
 )
 
-plot_pdf +
+plot_pdf <- plot_pdf +
   geom_line(data = beta0_inla_marginals, aes(x = x, y = y, col = method)) +
-  scale_color_manual(values = cbpalette)
+  scale_color_manual(values = cbpalette) +
+  labs(col = "Method")
+
+plot_pdf
+
+ggsave("intercept-comparison.pdf", plot = plot_pdf, h = 4, w = 6.25)
 ```

### 2022-11-16 src/docs_paper/paper.Rmd
Update algorithm after Tuesday meeting

@@
 
-Given a \textsf{C++} user template `model.cpp` for the unnormalised log posterior, we obtain the posterior marginal approximations given in Equations \eqref{eq:inla1} and \eqref{eq:inla2} via the following algorithm:
+Given a \textsf{C++} user template `model.cpp` for the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adative Gauss-Hermite quadrature.
 
@@
 $$
-\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)}
+\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
 $$
-where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$.
+where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+\begin{align*}
+\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
+\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}
+\end{align*}
 
@@
 $$
-where the normalising constant is calculated using nodes $\z_\theta \in \mathcal{Q}(m, k_\theta)$ as
+where the normalising constant is calculated using nodes from a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(m, k)$ with $m = \dim(\btheta)$, $k$ nodes per dimension, and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ as
 $$
-\tilde p_{\text{AQ}}(\y) = \sum_{\z_\theta \in \mathcal{Q}(m, k_\theta)} \tilde p_\text{LA}(\btheta(\z_\theta), \y) \omega_\theta(\z_\theta),
+\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z).
 $$
-where $m = \dim(\btheta)$, the number of nodes per dimension is $k_\theta$, and weights $\omega_\theta: \z_\theta \in \mathcal{Q}(m, k_\theta) \to \mathbb{R}$.
-We typically set $k_\theta = 3$ such that there are $3^m$ nodes in total.
+The nodes $\z$ are adapted based on the mode and curvature at the mode of the Laplace approximation as follows
+\begin{align*}
+\btheta(\z) &= \hat{\btheta} + \mathbf{L} \z, \\
+\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
+\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
+\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top.
+\end{align*}
+We typically set $k = 3$ such that there are $3^m$ nodes in total.
 
@@
 $$
-\tilde p(x_i \, | \, \y) = \sum_{\z_\theta \in \mathcal{Q}(m, k_\theta)} \tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y) \tilde p_\text{AQ} (\btheta(\z_\theta) \, | \, \y) \omega_\theta(\z_\theta).
+\tilde p(x_i \, | \, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{AQ}(x_i \, | \, \btheta(\z), \y) \tilde p_\text{AQ} (\btheta(\z) \, | \, \y) \psi(\z).
 $$
-The nodes and weights $\{\mathcal{Q}(m, k_\theta), \omega_\theta\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
-For each of the $k_\theta^m$ values of $\btheta(\z_\theta)$ we obtain $\tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y)$ by setting $\btheta = \btheta(\z_\theta)$ in the following two steps.
+The nodes and weights $\{\mathcal{Q}(m, k), \omega\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
+For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{AQ}(x_i \, | \, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following two steps.
 
@@
 $$
-\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)}
+\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
 $$
-where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}_{-i}(x_i, \btheta), \bm{Q}_{-i}(x_i, \btheta)^{-1})$.
+where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x_{-i}(x_i, \btheta), \mathbf{H}_{-i, -i}(x_i, \btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+\begin{align*}
+\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
+\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
+\end{align*}
 
@@
 $$
-where the normalising constant is calculated using nodes $z_x \in \mathcal{Q}(1, k_x)$ as
+where the normalising constant is calculated using nodes from a Gauss-Hermite quadrature rule $v \in \mathcal{Q}(1, l)$ with $l$ nodes, and weights $\psi: v \in \mathcal{Q}(1, l) \to \mathbb{R}$ as follows
 $$
-\tilde p_{\text{AQ}}(\btheta, \y) = \sum_{z_x \in \mathcal{Q}(1, k_x)} \tilde p_\text{LA}(x_i(z_x), \btheta, \y) \omega_x(z_x),
+\tilde p_{\text{AQ}}(\btheta, \y) = \sum_{v \in \mathcal{Q}(1, l)} \tilde p_\text{LA}(x_i(v), \btheta, \y) \psi(v),
 $$
-where the total number of nodes is $k_x$ and weights $\omega_x: z_x \in \mathcal{Q}(1, k_x) \to \mathbb{R}$.
-As this quadrature is in one dimension, we can set $k_x = 5$ or higher.
+The nodes $v$ are adapted based on the mode and curvature at the mode of the Laplace approximation as follows
+\begin{align*}
+x_i(v) &= \hat{x_i} + \sigma_i v, \\
+x_i(v) &= \argmax_{x_i} {\tilde p_\text{LA}(x_i, \btheta, \y)}, \\
+\tau_i^2 &= \frac{\partial^2}{\partial^2 x_i} - \log \tilde p_\text{LA}(x_i, \btheta, \y) \rvert_{x_i = \hat x_i} \\
+\sqrt{\tau_i^2} &= \sigma_i.
+\end{align*}
+As this quadrature is in one dimension, we can set $l = 5$ or more.
 

### 2022-11-16 src/docs_paper/paper.Rmd
omega and psi typo
Simplify algorithm

@@
 
-3. Obtain a nested approximation to the posterior marginal of the $i$th latent effect by
+3. Obtain an unnormalised nested approximation to the posterior marginal of the $i$th latent effect by
 $$
-\tilde p(x_i \, | \, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{AQ}(x_i \, | \, \btheta(\z), \y) \tilde p_\text{AQ} (\btheta(\z) \, | \, \y) \psi(\z).
+\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) , \y) \omega(\z).
 $$
 The nodes and weights $\{\mathcal{Q}(m, k), \omega\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
-For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{AQ}(x_i \, | \, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following two steps.
-
-a) Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(x_i, \btheta, \y)$
+For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{AQ}(x_i \, | \, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following Laplace approximation
 $$
@@
 
-b) Normalise $\tilde p_\text{LA}(x_i, \btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
+4. Normalise $\tilde p_\text{LA}(x_i, \y)$ using $\tilde p_{\text{AQ}}(\y)$ to obtain
 $$
-\tilde p_\text{AQ}(x_i \, | \, \btheta, \y) = \frac{\tilde p_\text{LA}(x_i, \btheta, \y)}{\tilde p_{\text{AQ}}(\btheta, \y)},
+\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
 $$
-where the normalising constant is calculated using nodes from a Gauss-Hermite quadrature rule $v \in \mathcal{Q}(1, l)$ with $l$ nodes, and weights $\psi: v \in \mathcal{Q}(1, l) \to \mathbb{R}$ as follows
-$$
-\tilde p_{\text{AQ}}(\btheta, \y) = \sum_{v \in \mathcal{Q}(1, l)} \tilde p_\text{LA}(x_i(v), \btheta, \y) \psi(v),
-$$
-The nodes $v$ are adapted based on the mode and curvature at the mode of the Laplace approximation as follows
-\begin{align*}
-x_i(v) &= \hat{x_i} + \sigma_i v, \\
-x_i(v) &= \argmax_{x_i} {\tilde p_\text{LA}(x_i, \btheta, \y)}, \\
-\tau_i^2 &= \frac{\partial^2}{\partial^2 x_i} - \log \tilde p_\text{LA}(x_i, \btheta, \y) \rvert_{x_i = \hat x_i} \\
-\sqrt{\tau_i^2} &= \sigma_i.
-\end{align*}
-As this quadrature is in one dimension, we can set $l = 5$ or more.
 

### 2022-11-17 src/docs_paper/paper.Rmd
Remove typo
Add backwards version of the algorithm too (as in defining objects mathematically, rather than how you would code it)

@@
 
-# Fast inference methods\label{sec:fastinferencemethods}
+# Fast approximate inference method\label{sec:fastinferencemethods}
 
@@
 
-## Algorithm\label{sec:algorithm}
+## Forwards algorithm\label{sec:algorithm}
 
-Given a \textsf{C++} user template `model.cpp` for the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adative Gauss-Hermite quadrature.
+Given a \textsf{C++} user template `model.cpp` for the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adaptive Gauss-Hermite quadrature.
 
@@
 \hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
-\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}
+\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
 \end{align*}
@@
 $$
-\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) , \y) \omega(\z).
+\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
 $$
 The nodes and weights $\{\mathcal{Q}(m, k), \omega\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
-For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{AQ}(x_i \, | \, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following Laplace approximation
+For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{LA}(x_i, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following Laplace approximation
 $$
-\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
+\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
 $$
@@
 \hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
-\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
+\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
 \end{align*}
+Optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$.
 
@@
 \tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
+$$
+which may be evaluated for some choice of values $x_i \in \{\ldots\}$.
+
+## Backward algorithm\label{sec:algorithm}
+
+1. Calculate
+\begin{align*}
+\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
+\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
+\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
+\end{align*}
+where
+$$
+\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+$$
+where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+\begin{align*}
+\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
+\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
+\end{align*}
+2.
+Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $m = \dim(\btheta)$, $k$ nodes per dimension, as follows
+\begin{align*}
+\btheta(\z) &= \hat{\btheta} + \mathbf{L} \z, \\
+\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
+\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
+\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top.
+\end{align*}
+We typically set $k = 3$ such that there are $3^m$ nodes in total.
+3. Use this quadrature rule to calculate $\tilde p_{\text{AQ}}(\y)$ as follows
+$$
+\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z).
+$$
+4. For $x_i \in \{\ldots\}$ and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
+\begin{align*}
+\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
+\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
+\end{align*}
+where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$.
+
+5. For each $x_i$ calculate
+$$
+\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
+$$
+where
+$$
+\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
+$$
+and
+$$
+\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
 $$

### 2022-11-17 docs/seminar.Rmd
Correct typo

@@
 The new method will be implemented as an extension of the `aghq` package (Stringer, 2021), which will facilitate flexible and particularly easy use when provided a `TMB` C++ template for the log-posterior.
-In this talk, I'll discuss this progress towards this project, which I have been working on with Alex Stringer here at Waterloo this term through the International Visiting Graduate Student program."
+In this talk, I'll discuss progress towards this project, which I have been working on with Alex Stringer here at Waterloo this term through the International Visiting Graduate Student program."
 ```

### 2022-11-17 src/epil/epil.Rmd
Add EB R-INLA versions

@@
   **Background** @rue2009approximate use the epilespy example from the OpenBUGs manual as demonstration of the integrated nested Laplace approximation method.
+The model is a Poisson GLMM including both fixed and random effects.
 
@@
 ```{r}
-epil_inla <- function(strat) {
+epil_inla <- function(strat, int_strat) {
   inla(formula,
@@
        data = Epil,
-       control.inla = list(strategy = strat),
+       control.inla = list(strategy = strat, int.strategy = int_strat),
        control.predictor = list(compute = TRUE))
@@
 
-fit2 <- epil_inla(strat = "gaussian")
-fit3 <- epil_inla(strat = "simplified.laplace")
-fit4 <- epil_inla(strat = "laplace")
+fit2 <- epil_inla(strat = "gaussian", int_strat = "eb")
+fit3 <- epil_inla(strat = "simplified.laplace", int_strat = "eb")
+fit4 <- epil_inla(strat = "laplace", int_strat = "eb")
+
+fit5 <- epil_inla(strat = "gaussian", int_strat = "grid")
+fit6 <- epil_inla(strat = "simplified.laplace", int_strat = "grid")
+fit7 <- epil_inla(strat = "laplace", int_strat = "grid")
 ```
@@
 ```{r results='hide'}
-fit8 <- aghq::marginal_laplace_tmb(
+fit11 <- aghq::marginal_laplace_tmb(
   obj,
@@
 
-fit8_samples <- aghq::sample_marginal(fit8, M = 1000)$samps %>%
+fit11_samples <- aghq::sample_marginal(fit8, M = 1000)$samps %>%
   t() %>%
@@
 inla4 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
-tmb5 <- as.vector(t(data.frame(sd_out$par.random[1:6], sqrt(sd_out$diag.cov.random[1:6]))))
-glmmtmb6 <- as.vector(t(summary(fit6)$coefficients$cond[, c("Estimate", "Std. Error")]))
-tmbstan7 <- as.vector(t(summary(fit7)$summary[1:6, c(1, 3)]))
+inla5 <- as.vector(t(fit2$summary.fixed[1:6, 1:2]))
+inla6 <- as.vector(t(fit3$summary.fixed[1:6, 1:2]))
+inla7 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
+tmb8 <- as.vector(t(data.frame(sd_out$par.random[1:6], sqrt(sd_out$diag.cov.random[1:6]))))
+glmmtmb9 <- as.vector(t(summary(fit6)$coefficients$cond[, c("Estimate", "Std. Error")]))
+tmbstan10 <- as.vector(t(summary(fit7)$summary[1:6, c(1, 3)]))
 
-fit8_beta_samples <- fit8_samples[, 1:6]
-aghq8 <- as.vector(t(data.frame(mean = apply(fit8_beta_samples, 2, mean), sd = apply(fit8_beta_samples, 2, sd))))
+fit11_beta_samples <- fit8_samples[, 1:6]
+aghq11 <- as.vector(t(data.frame(mean = apply(fit11_beta_samples, 2, mean), sd = apply(fit8_beta_samples, 2, sd))))
 
-adam9 <- as.vector(t(laplace_beta_summary[1:6, 1:2]))
+adam12 <- as.vector(t(laplace_beta_summary[1:6, 1:2]))
 
-df <- cbind(stan1, inla2, inla3, inla4, tmb5, glmmtmb6, tmbstan7, aghq8, adam9) %>%
+df <- cbind(stan1, inla2, inla3, inla4, inla5, inla6, inla7, tmb8, glmmtmb9, tmbstan10, aghq11, adam12) %>%
   as.data.frame() %>%
@@
 rownames(df) <- c(sapply(0:5, beta_i))
-colnames(df) <- c("Stan", "R-INLA_G", "R-INLA_SL", "R-INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "adam", "type")
+colnames(df) <- c("Stan", "EB-G", "EB-SL", "EB-L", "Grid-G", "Grid-SL", "Grid-L", "TMB", "glmmTMB", "tmbstan", "aghq", "adam", "type")
 
@@
   select(-type) %>%
-  kableExtra::kable()
+  kableExtra::kable(digits = 3)
 

### 2022-11-17 src/dev_sinla/sinla.Rmd
Change the notation to match current paper draft, and fix a few mistakes I noticed

@@
 \begin{equation}
-\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z), (\#eq:pxgiveny)
+\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z), (\#eq:pxgiveny)
 \end{equation}
 $$
-where $\theta(z) \in \Theta$ and $\omega_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
+where $\theta(z) \in \Theta$ and $\omega(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
 In particular, let
@@
 $$
-\omega_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
+\omega(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
 $$
@@
 
-This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triangle of the curvature at a randomly chosen quadrature point).
+This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mode and precision from Rue (in particular you solve a linear equation using the lower Cholesky triangle of the curvature at a randomly chosen quadrature point).
 For the $i$th marginal, we just take the appropriate samples from the joint.
@@
 \begin{equation}
-\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z), (\#eq:pxigiveny)
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z), (\#eq:pxigiveny)
 \end{equation}
 $$
-where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.
+where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mode vector and $(i, i)$th element of the covariance matrix.
 
@@
 
-The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/elgm-inf/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
-$$
-\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z).
+The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/elgm-inf/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $\tilde p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
+$$
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z).
 $$
@@
 \begin{equation}
-\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \mu(\theta)} = \frac{p(y, \mu(\theta), \theta)}{\det \left(\frac{1}{2 \pi} Q(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
+\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)} = \frac{p(y, \hat x(\theta), \theta)}{\det \left(\frac{1}{2 \pi} H(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
 \end{equation}
@@
 such that one easy way to calculate it is to set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there, as we have done in the function `h` above.
-
 Let's optimise `h` to obtain estimates $\hat \theta$ which maximise $\tilde p(\theta \, | \, y)$:
@@
 
-We can also look at the parameters -- that is, mean vector and precision matrix -- of the Gaussian approximation at the mode `opt_theta$par` above.
-Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
+We can also look at the parameters -- that is, mode vector and precision matrix -- of the Gaussian approximation at the mode `opt_theta$par` above.
+Note that in `aghq`, what we would do is create a quadrature grid, then get the mode vector and precision matrix for each value of the quadrature grid over $\theta$.
 By just sticking to the mode of $\theta$, we are effectively doing `aghq` with $k = 1$ in each dimension, i.e. one grid point, corresponding to empirical Bayes type of inference.
@@
 
-#' Note that I'm calling this mean, but because it's a Gaussian the mean is the same as the mode
-x_mean <- mm[h$env$random]
-x_mean <- rlang::duplicate(x_mean) #' Insurance...
-length(x_mean) #' The length is 30
-
-#' Hessian evaluated at the mean (or equivalently mode)
-x_Q <- h$env$spHess(mm, random = TRUE)
-x_Q <- rlang::duplicate(x_Q)
-dim(x_Q) #' With dimension [30 x 30]
+x_hat <- mm[h$env$random]
+x_hat <- rlang::duplicate(x_hat) #' Insurance...
+length(x_hat) #' The length is 30
+
+#' Hessian evaluated at the mode
+x_H <- h$env$spHess(mm, random = TRUE)
+x_H <- rlang::duplicate(x_H)
+dim(x_H) #' With dimension [30 x 30]
 ```
@@
 
-Given $\theta$, we compute the `mean` and `Q` of $x$ (corresponding to `phi_prev`) using `h`.
+Given $\theta$, we compute the `mode` and `H` of $x$ (corresponding to `phi_prev`) using `h`.
 We then concatenate and evaluate using `f` to get the numerator, multiplying by -1 because `f$fn` gives the negative log-likelihood, and we want the log-likelihood.
@@
 calculate_rhs <- function(theta) {
-  #' Calculate mean and Q for those theta
+  #' Calculate mode and H for those theta
   #' TMB template remembers what you've called -- spooky
@@
   mm <- h$env$last.par
-  mean <- mm[h$env$random]
-  Q <- h$env$spHess(mm, random = TRUE)
-
-  #' Append the calculated latent field mean
+  mode <- mm[h$env$random]
+  H <- h$env$spHess(mm, random = TRUE)
+
+  #' Append the calculated latent field mode
   param <- theta
-  param$phi_prev <- as.numeric(mean)
+  param$phi_prev <- as.numeric(mode)
   param <- param[c("beta_prev", "phi_prev", "log_sigma_phi_prev")]
@@
   #' Evaluate the log-denominator of the RHS
-  denom <- 0.5 * (determinant(Q, logarithm = TRUE)$modulus - (nrow(Q) * log(2 * pi)))
+  denom <- 0.5 * (determinant(H, logarithm = TRUE)$modulus - (nrow(H) * log(2 * pi)))
 
@@
 i <- 1
-mean_i <- as.numeric(x_mean[i])
-var_i <- solve(x_Q)[i, i]
+mode_i <- as.numeric(x_hat[i])
+var_i <- solve(x_H)[i, i]
 ```
@@
     data = data.frame(x = c(0.5, 3.5)), aes(x, col = "Joint Gaussian marginal"), fun = dnorm, n = 101, 
-    args = list(mean = mean_i, sd = sqrt(var_i))
+    args = list(mean = mode_i, sd = sqrt(var_i))
   ) +
@@
 M <- 1000
-samples_gaussian <- rnorm(M, mean_i, sqrt(var_i))
+samples_gaussian <- rnorm(M, mode_i, sqrt(var_i))
 
@@
 \begin{align*}
-  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \mu_{-i}(x_i, \theta)} (\#eq:pxigiventhetay) \\
-  &= \frac{p(y, x_i, \mu_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} Q_{-i}(x_i, \theta))^{1/2}}
+  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \hat x_{-i}(x_i, \theta)} (\#eq:pxigiventhetay) \\
+  &= \frac{p(y, x_i, \hat x_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} H_{-i}(x_i, \theta))^{1/2}}
 \end{align*}
@@
 
-The mean and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:
-
-```{r}
-get_mean_and_Q <- function(obj) {
+The mode and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:
+
+```{r}
+get_mode_and_H <- function(obj) {
   mm <- obj$env$last.par
-  mean <- mm[obj$env$random]
-  mean <- rlang::duplicate(mean)
-  Q <- obj$env$spHess(mm, random = TRUE)
-  Q <- rlang::duplicate(Q)
-  return(list(mean = mean, Q = Q))
-}
-
-xminusi <- get_mean_and_Q(obj)
+  mode <- mm[obj$env$random]
+  mode <- rlang::duplicate(mode)
+  H <- obj$env$spHess(mm, random = TRUE)
+  H <- rlang::duplicate(H)
+  return(list(mode = mode, H = H))
+}
+
+xminusi <- get_mode_and_H(obj)
 ```
@@
 ```{r}
-x_Sigma <- solve(x_Q)
+x_Sigma <- solve(x_H)
 
@@
   x <- seq(-5, 5, length.out = 1000)
-  mean_i <- as.numeric(x_mean[i])
+  mode_i <- as.numeric(x_hat[i])
   sd_i <- as.numeric(sqrt(x_Sigma[i, i]))
-  data.frame(index = i, x = x, pdf = dnorm(x, mean = mean_i, sd = sd_i))
+  data.frame(index = i, x = x, pdf = dnorm(x, mean = mode_i, sd = sd_i))
 }) %>%
@@
 samples_gaussian <- lapply(1:dat$n, function(i) {
-  mean_i <- as.numeric(x_mean[i])
+  mode_i <- as.numeric(x_hat[i])
   sd_i <- as.numeric(sqrt(x_Sigma[i, i]))
-  samples <- rnorm(M, mean_i, sd_i)
+  samples <- rnorm(M, mode_i, sd_i)
   data.frame("value" = samples)
@@
 ```{r}
-ks_gaussian_tmbstan <- lapply(1:length(mean), function(i) {
+ks_gaussian_tmbstan <- lapply(1:length(mode), function(i) {
   samples_gaussian_i <- filter(samples_gaussian, index == i)$value
@@
 
-ks_laplace_tmbstan <- lapply(1:length(mean), function(i) {
+ks_laplace_tmbstan <- lapply(1:length(mode), function(i) {
   samples_laplace_i <- filter(samples_laplace, index == i)$value
@@
 $$
-\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z).
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z).
 $$
@@
 
-The weight $\omega_k(z)$ is:
+The weight $\omega(z)$ is:
 
@@
 ```{r}
-mean_i <- as.numeric(modesandhessians$mode[[z]][i])
+mode_i <- as.numeric(modesandhessians$mode[[z]][i])
 Sigma <- solve(modesandhessians$H[[z]])
@@
 
-dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE)
+dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE)
 ```
@@
 ```{r}
-dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z])
+dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z])
 ```
@@
 $$
-\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log \omega_k(z) \right)
-$$
-
-```{r}
-exp(dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
+\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log \omega(z) \right)
+$$
+
+```{r}
+exp(dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
 ```
@@
   theta <- as.numeric(modesandhessians[z, thetanames])
-  mean_i <- as.numeric(modesandhessians$mode[[z]][i])
+  mode_i <- as.numeric(modesandhessians$mode[[z]][i])
   Sigma <- solve(modesandhessians$H[[z]])
   sd_i <- as.numeric(sqrt(Sigma[i, i]))
-  pdf <- pdf + exp(dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
+  pdf <- pdf + exp(dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
 }
@@
 $$
-\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z),
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z),
 $$
@@
 
-As before, the weight $\omega_k(z)$ is:
+As before, the weight $\omega(z)$ is:
 
@@
 <!-- $$ -->
-<!-- x \sim \mathcal{N}(\mu(\theta), Q(\theta)^{-1}) -->
-<!-- $$ -->
-<!-- then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mean -->
-<!-- $$ -->
-<!-- \mu(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:conditionalmean) -->
+<!-- x \sim \mathcal{N}(\hat x(\theta), H(\theta)^{-1}) -->
+<!-- $$ -->
+<!-- then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mode -->
+<!-- $$ -->
+<!-- \hat x(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \hat x (\theta)_i) (\#eq:conditionalmode) -->
 <!-- $$ -->
@@
 <!-- $$ -->
-<!-- Q(\theta)_{-i, -i} (\#eq:conditionalprecision) -->
+<!-- H(\theta)_{-i, -i} (\#eq:conditionalprecision) -->
 <!-- $$ -->
@@
 <!-- We follow Wood (2019) Section 2. -->
-<!-- In this section we omit dependence on $\theta$ such that: $\mu = \mu(\theta)$, $Q = Q(\theta)$, $\mu(x_i) = \mu(x_i, \theta)$ and $Q(x_i) = Q(x_i, \theta)$. -->
+<!-- In this section we omit dependence on $\theta$ such that: $\hat x = \hat x(\theta)$, $H = H(\theta)$, $\hat x_{-i}(x_i) = \hat x_{-i}(x_i, \theta)$ and $H(x_i) = H_{-i, -i}(x_i, \theta)$. -->
 
@@
 
-<!-- The first task is to "compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$". -->
+<!-- The first task is to "compute the Cholesky factor of $H_0 := H_{-i, -i}$ by update of the Cholesky factor of $H$". -->
 
@@
 <!-- mm <- h$env$last.par -->
-<!-- mean <- mm[h$env$random] -->
-<!-- Q <- h$env$spHess(mm, random = TRUE) -->
-<!-- Q0 <- Q[-i, -i]  -->
-
-<!-- L0 <- chol(Q0) -->
+<!-- mode <- mm[h$env$random] -->
+<!-- H <- h$env$spHess(mm, random = TRUE) -->
+<!-- H0 <- H[-i, -i]  -->
+
+<!-- L0 <- chol(H0) -->
 <!-- ``` -->
 
-<!-- We have that $Q_0 = L_0 {L_0}^\top$: -->
+<!-- We have that $H_0 = L_0 {L_0}^\top$: -->
 
 <!-- ```{r} -->
-<!-- sum(abs((L0 %*% t(L0)) - Q0)) < 10E-9 -->
+<!-- sum(abs((L0 %*% t(L0)) - H0)) < 10E-9 -->
 <!-- ``` -->
@@
 
-<!-- Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean). -->
-<!-- We start by computing the conditional mean for the first grid point: -->
+<!-- Use Newton's method with fixed Hessian $H_{-i, -i}$ to find $\hat x_{-i}(x_i)$ starting from \@ref(eq:conditionalmode). -->
+<!-- We start by computing the conditional mode for the first grid point: -->
 
 <!-- ```{r} -->
-<!-- Sigma <- solve(Q) -->
-<!-- conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i]) -->
+<!-- Sigma <- solve(H) -->
+<!-- conditional_mode <- mode[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mode[i]) -->
 <!-- ``` -->
 
-<!-- Note that for this model the latent field is IID, so the conditional mean is the same as the mean. -->
-<!-- Now we use Newton's method to find $\mu(x_i)$. -->
-<!-- This requires some function $f(\mu(x_i), \ldots) = 0$. -->
+<!-- Note that for this model the latent field is IID, so the conditional mode is the same as the mode. -->
+<!-- Now we use Newton's method to find $\hat x_{-i}(x_i)$. -->
+<!-- This requires some function $f(\hat x_{-i}(x_i), \ldots) = 0$. -->
 
@@
 <!-- $$ -->
-<!-- q = Q_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j -->
+<!-- q = H_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j -->
 <!-- $$ -->
@@
 <!-- $$ -->
-<!-- g = \Delta_x \log p(\mu(x_i) + \Delta_j^i, y, \theta). -->
+<!-- g = \Delta_x \log p(\hat x_{-i}(x_i) + \Delta_j^i, y, \theta). -->
 <!-- $$ -->
@@
 <!-- $$ -->
-<!-- \det(Q_1) = \det(Q_0) \det(I_{2J} + u_J^\top Q_0^{-1} u_J \mathcal{D}_{2j}). -->
+<!-- \det(H_1) = \det(H_0) \det(I_{2J} + u_J^\top H_0^{-1} u_J \mathcal{D}_{2j}). -->
 <!-- $$ -->
@@
 | Random effects | $x$        | $\beta$              | Stringer (2021) use $w$
-| Random effects which maximise (the / some) objective for given $\theta$ | $\mu (\theta)$ | $\hat \beta$ | Wood omits dependence on $\theta$ |
-| Random effects which maximise (the / some) objective for given $\theta$, subject to $x_i$ being fixed | $(x_i, \mu(x_i, \theta))$ | $\tilde \beta$ | We have $\mu(x_i, \theta))$ having dimension $n - 1$ |
-| Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$ | $Q(\theta)$ | $H$ | Unsure if it's preferable to use Hessian or precision notation here |
+| Random effects which maximise (the / some) objective for given $\theta$ | $\hat x (\theta)$ | $\hat \beta$ | Wood omits dependence on $\theta$ |
+| Random effects which maximise (the / some) objective for given $\theta$, subject to $x_i$ being fixed | $(x_i, \hat x_{-i}(x_i, \theta))$ | $\tilde \beta$ | We have $\hat x_{-i}(x_i, \theta))$ having dimension $n - 1$ |
+| Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\hat x(\theta)$ | $H(\theta)$ | $H$ | Unsure if it's preferable to use Hessian or precision notation here |
 

### 2022-11-21 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Small poster update

@@
 ---
-main_topsize: 0.2
-main_bottomsize: 0.075
+main_topsize: 0.18
+main_bottomsize: 0.03
 poster_height: "48in"
@@
     address: Department of Computer Science, University of Oxford
-main_findings:
-  - "Combining **simplified INLA**, automatic differentiation and **adaptive Gauss-Hermite quadrature** for fast and accurate inference in a challenging, real-world problem"
+main_findings: "Simplified INLA, automatic differentiation and adaptive Gauss-Hermite quadrature for fast and accurate approximate inference"
 logoleft_name: "IMP_ML_W_N.png"
@@
 body_bgcol: "#ffffff"
-body_textsize: "40px"
+body_textsize: "35px"
 body_textcol: "#000000"
@@
 
+$$\\[0.15in]$$
+
 ## Inference procedure
+
+### Laplace approximation
+
+### Adaptive Gauss-Hermite Quadrature
+
+Approximate integrals by 
+$$
+\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)
+$$
+with **Gauss-Hermite quadrature** rule $z \in \mathcal{Q}(m, k)$ **adapted** based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} \in p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target.
+
+### Our algorithm
+
+Given C++ user template for $- \log p(y, x, \theta)$:
 

### 2022-11-21 src/dev_sinla/sinla.Rmd
Attempting to get new AGHQ Laplace to work

@@
 ```{r}
-ks_gaussian_tmbstan <- lapply(1:length(mode), function(i) {
+ks_gaussian_tmbstan <- lapply(1:length(x_hat), function(i) {
   samples_gaussian_i <- filter(samples_gaussian, index == i)$value
@@
 
-ks_laplace_tmbstan <- lapply(1:length(mode), function(i) {
+ks_laplace_tmbstan <- lapply(1:length(x_hat), function(i) {
   samples_laplace_i <- filter(samples_laplace, index == i)$value
@@
 
+```{r}
+ks_results %>%
+  group_by(type) %>%
+  summarise(D = mean(D))
+```
+
 # Extension beyond empirical Bayes
@@
 
-for(z in 1:length(modesandhessians$weights)) {
+for(z in 1:nrow(modesandhessians)) {
   theta <- as.numeric(modesandhessians[z, thetanames])
@@
 
-## Full Laplace approximation
+## Full Laplace approximation (old way)
 
@@
 ```{r}
+#' All of the theta
+modesandhessians[, thetanames]
+
+#' Just the zth one
 theta <- as.numeric(modesandhessians[z, thetanames])
+theta
+
 - h_norm(theta)
@@
 
-for(theta in thetanames) param_i_fixed_theta[[theta]] <- modesandhessians[z, thetanames][[theta]]
+for(theta_string in thetanames) param_i_fixed_theta[[theta_string]] <- modesandhessians[z, thetanames][[theta_string]]
 
@@
 
-  for(theta in thetanames) param_i_fixed_theta[[theta]] <- modesandhessians[z, thetanames][[theta]]
+  for(theta_string in thetanames) param_i_fixed_theta[[theta_string]] <- modesandhessians[z, thetanames][[theta_string]]
 
@@
 
+## Full Laplace approximation (new way)
+
+The method above for doing quadrature over the full Laplace approximation should work, but it requires a lot of computation.
+Instead, we might be able to do this more cost effectively by integrating $\theta$ out before normalising $\tilde p_\text{LA}(x_i, \theta(z), y)$.
+This would look like first obtaining
+$$
+\tilde p(x_i, y) = \sum_z \tilde p_\text{LA}(x_i, \theta(z), y) \omega(z),
+$$
+then normalising by the previously obtained $\tilde p_\text{AQ}(y)$ as follows
+$$
+\tilde p(x_i \, | \, y) = \frac{\tilde p(x_i, y)}{\tilde p_\text{AQ}(y)}.
+$$
+
+```{r}
+#' Always (?) want the same initialisation parameters (though they probably could be better right?)
+param_i_fixed_theta <- list(
+  beta_prev = 0, 
+  phi_prev_i = 0, 
+  phi_prev_minus_i = rep(0, data$n - 1),
+  log_sigma_phi_prev = 0
+)
+
+#' Always fixing the hyperparameters
+map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
+
+#' Going to store the functions p_LA(x_i, theta(z), y) in objs
+objs <- list()
+
+for(z in 1:nrow(modesandhessians)) {
+ theta <- as.numeric(modesandhessians[z, thetanames])
+
+  dat_i <- prepare_dat(dat, i)
+
+  for(theta_string in thetanames) param_i_fixed_theta[[theta_string]] <- modesandhessians[z, thetanames][[theta_string]]
+
+  obj_fixed_theta <- MakeADFun(
+    data = dat_i,
+    parameters = param_i_fixed_theta,
+    random = "phi_prev_minus_i",
+    DLL = "model1_index",
+    silent = TRUE,
+    map = map_fixed_theta
+  )
+
+  objs[[z]] <- obj_fixed_theta
+}
+
+aghq_laplace_marginal <- function(x) {
+  pdf <- 0
+  for(z in 1:nrow(modesandhessians)) {
+    pdf <- pdf + exp(as.numeric(-objs[[z]]$fn(x))) * modesandhessians[z, "weights"]
+  }
+  return(pdf)
+}
+
+#' Issue is that we do the normalisation after now..
+aghq_laplace_marginal(x = x_test_value)
+aghq_laplace_marginal(x = -1)
+```
+
 <!-- # Simplified INLA -->

### 2022-11-22 src/dev_sinla/sinla.Rmd
logSumExp, simplify writing in Gaussian AGHQ, use AGHQ to pick x[i] nodes, use spline interpolation for plotting marginal

@@
 We're going to start by normalising $\tilde p_\text{LA}(\theta, y)$.
-Remember that we did this above in the empirical Bayes case, but found that it wasn't strictly necessary.
+We did this above in the empirical Bayes case, but found that it wasn't strictly necessary.
 
@@
 
-If you wanted to evaluate this to obtain each marginal then it would be something like (pseudo-code):
-
-```{r eval=FALSE}
-for(i in 1:n) {
-  for(x in x_vals) {
-    for(z in z_vals) {
-      #' ...
-    }
-  }
-}
-```
-
-To start with, let's evaluate this for `i = 1`, `x = x_test_value` and the first value of `z` (actually I don't really know what $z$ runs over, but this is just the first weight):
+To start with, let's evaluate this on the log-scale for `i = 1`, `x = x_test_value` and the first value of `z`.
+The weight $\omega(z)$ will be `modesandhessians$weights[z]`.
+The normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ will be `- h_norm(as.numeric(modesandhessians[z, thetanames]))`.
+And to obtain the normalised Gaussian approximation $p_\text{G}(x_i \, | \, \theta(z), y)$ on the log-scale we get the relevant subcomponents of the joint Gaussian then use `ndorm` with `log = TRUE`.
 
@@
 x_test_value <- 0
+
+#' Note that in some sense z actually runs over nodes, not indices
 z <- 1 
 i <- 1
-```
-
-The weight $\omega(z)$ is:
-
-```{r}
-modesandhessians$weights[z]
-```
-
-The normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ is:
-
-```{r}
-theta <- as.numeric(modesandhessians[z, thetanames])
-- h_norm(theta)
-```
-
-The normalised Gaussian approximation $p_\text{G}(x_i \, | \, \theta(z), y)$ on the log-scale is:
-
-```{r}
+
 mode_i <- as.numeric(modesandhessians$mode[[z]][i])
@@
 
-dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE)
-```
-
-So, putting it together, the value we would compute on the log-scale is as follows:
-
-```{r}
 dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z])
@@
 
-To obtain the unlogged version we can use:
-$$
-\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log \omega(z) \right)
-$$
-
-```{r}
-exp(dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
-```
-
-Great -- now let's sum over `z` so that we can obtain the probability for our `x_test_value` in marginal `i`.
-We won't worry about inefficient R for loops for the time being.
-
-```{r}
-pdf <- 0
+Now let's sum over `z` so that we can obtain the log-probability for our `x_test_value` in marginal `i`:
+
+```{r}
+gaussian_marginal <- function(x) {
+  lp <- vector(mode = "numeric", length = nrow(modesandhessians))
 
@@
     sd_i <- as.numeric(sqrt(Sigma[i, i]))
-  pdf <- pdf + exp(dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
-}
-
-pdf
-```
-
-Using a `fine_grid` set of values for `x_vals` for each `i` might be overly computationally expensive.
-Instead what we can do is perform another separate AGHQ to select a set of `x_vals`.
-To do this with the `aghq` R package, we could create a `TMB` template to evaluate $\tilde p(x_i \, | \, y)$, but I imagine this might be too annoying.
-Instead, we could create the `ff` object (with components `ff$fn`, `ff$gn` `ff$hn`) in R directly.
+    lp[z] <- dnorm(x, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z])
+  }
+
+  matrixStats::logSumExp(lp)
+}
+
+gaussian_marginal(x_test_value)
+```
+
+To pick a set of $x_i$ to evaluate the marginal at, we will use an approximation to the AGHQ nodes.
+We will take the hyperparameters to be fixed at their mode, then choose the appropriate marginals of the joint Gaussian:
+
+```{r}
+#' Get the row number of modesandhessians which contains that theta mode
+theta_mode_location <- which.max(quad$normalized_posterior$nodesandweights$logpost_normalized)
+
+#' These are the mode and standard deviation from the Gaussian approximation at the node which is the mode of the Laplace approximation
+#' (Say that three times fast...)
+mode_i <- modesandhessians[[theta_mode_location, "mode"]][i]
+sd_i <- sqrt(1 / modesandhessians[[theta_mode_location, "H"]][i, i])
+
+#' Create Gauss-Hermite quadrature
+gg <- mvQuad::createNIGrid(dim = 1, type = "GHe", level = 7)
+
+#' Adapt to mode_i and sd_i
+mvQuad::rescale(gg, m = mode_i, C = sd_i^2)
+
+#' These are our set of x input values
+gg$nodes
+
+#' The log-probabilities at the set of input values
+lps <- sapply(gg$nodes, gaussian_marginal)
+
+#' Build a Lagrange polynomial interpolant of the marginal posterior
+ss <- splines::interpSpline(gg$nodes, lps, bSpline = TRUE, sparse = TRUE)
+
+interpolant <- function(x) {
+  as.numeric(stats::predict(ss, x)$y)
+}
+
+finegrid <- seq(-5, 5, by = 0.1)
+
+df <- data.frame(x = finegrid, y = exp(interpolant(finegrid)))
+
+ggplot(df, aes(x = x, y = y)) +
+  geom_line() +
+  theme_minimal() +
+  labs(x = "phi_prev[1]", y = "Posterior")
+```
 

### 2022-11-22 src/dev_sinla/sinla.Rmd
plot_marginal_spline function, and old and new Laplace marginal methods

@@
 
+plot_marginal_spline <- function(lps) {
   #' Build a Lagrange polynomial interpolant of the marginal posterior
   ss <- splines::interpSpline(gg$nodes, lps, bSpline = TRUE, sparse = TRUE)
-
-interpolant <- function(x) {
-  as.numeric(stats::predict(ss, x)$y)
-}
-
+  interpolant <- function(x) { as.numeric(stats::predict(ss, x)$y) }
   finegrid <- seq(-5, 5, by = 0.1)
-
   df <- data.frame(x = finegrid, y = exp(interpolant(finegrid)))
@@
     labs(x = "phi_prev[1]", y = "Posterior")
+}
+
+plot_marginal_spline(lps)
 ```
@@
 
-Let's do as we have above, but with the full Laplace approximation in place of the Gaussian.
-We want to evaluate
-$$
-\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z),
-$$
-so much of what we computed above in Section \@ref(aghq-gaussian-marginal) is still relevant.
-The only thing that has changed is that we're now using $\tilde p_\text{LA}(x_i \, | \, \theta(z), y)$ rather than $\tilde p_\text{G}(x_i \, | \, \theta(z), y)$.
+As before, but with the full Laplace approximation in place of the Gaussian.
+Much of what we computed above in Section \@ref(aghq-gaussian-marginal) is still relevant.
+The only thing that has changed is that we're now using $\tilde p_\text{LA}(x_i \, | \, \theta(z), y)$ rather than $\tilde p_\text{G}(x_i \, | \, \theta(z), y)$ in
+$$
+\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z).
+$$
 
@@
 
-As before, the weight $\omega(z)$ is:
-
-```{r}
-modesandhessians$weights[z]
-```
-
-And also as before, the normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ is:
-
-```{r}
-#' All of the theta
-modesandhessians[, thetanames]
-
-#' Just the zth one
-theta <- as.numeric(modesandhessians[z, thetanames])
-theta
-
-- h_norm(theta)
-```
-
-Now, getting to the difference, the normalised Laplace approximation $p_\text{LA}(x_i \, | \, \theta(z), y)$ on the log-scale is:
+As before, the weight $\omega(z)$ is `modesandhessians$weights[z]` and the normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ is `- h_norm(as.numeric(modesandhessians[z, thetanames]))`.
+Getting to the difference, the normalised Laplace approximation $p_\text{LA}(x_i \, | \, \theta(z), y)$ on the log-scale is:
 
@@
 #' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
-quad <- aghq::aghq(ff = obj_fixed_theta, k = 5, startingvalue = 0, control = aghq::default_control_tmb())
+quad <- aghq::aghq(ff = obj_fixed_theta, k = 7, startingvalue = 0, control = aghq::default_control_tmb())
 
@@
 
-obj_fixed_theta_norm(x_test_value)
-```
-
-Putting it together:
-
-```{r}
-#' On the log-scale
-as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z])
-
-#' Unlogged
-exp(as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
-```
-
-Sum this over `z`:
-
-```{r}
-pdf <- 0
+- obj_fixed_theta_norm(x_test_value)
+```
+
+Putting it together on the log-scale:
+
+```{r}
+as.numeric(- obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z])
+```
+
+Sum over `z`:
+
+```{r}
+old_laplace_marginal <- function(x) {
+  lp <- vector(mode = "numeric", length = nrow(modesandhessians))
 
@@
 
-  #' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
-  quad <- aghq::aghq(ff = obj_fixed_theta, k = 5, startingvalue = 0, control = aghq::default_control_tmb())
-
-  obj_fixed_theta_norm <- function(x) obj_fixed_theta$fn(x) + quad$normalized_posterior$lognormconst
-
-  pdf <- pdf + exp(as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
-}
-
-pdf
+    quad <- aghq::aghq(ff = obj_fixed_theta, k = 7, startingvalue = 0, control = aghq::default_control_tmb())
+    obj_fixed_theta_norm <- function(xx) obj_fixed_theta$fn(xx) + quad$normalized_posterior$lognormconst
+
+    lp[z] <- as.numeric(- obj_fixed_theta_norm(x)) + (- h_norm(theta)) + log(modesandhessians$weights[z])
+  }
+
+  matrixStats::logSumExp(lp)
+}
+
+old_laplace_marginal(x_test_value)
+
+lps <- sapply(gg$nodes, old_laplace_marginal)
+plot_marginal_spline(lps)
 ```
@@
 
-The method above for doing quadrature over the full Laplace approximation should work, but it requires a lot of computation.
-Instead, we might be able to do this more cost effectively by integrating $\theta$ out before normalising $\tilde p_\text{LA}(x_i, \theta(z), y)$.
+The method above should work, but it requires a lot of computation.
+Instead, we might be able to do this more cost effectively by integrating $\theta$ out *before* normalising $\tilde p_\text{LA}(x_i, \theta(z), y)$.
 This would look like first obtaining
@@
 ```{r}
-#' Always (?) want the same initialisation parameters (though they probably could be better right?)
+#' Make sure the log normalising constant is available
+quad <- aghq::aghq(ff = h, k = 3, startingvalue = c(0, 0), control = aghq::default_control_tmb())
+quad$normalized_posterior$lognormconst
+
+laplace_marginal <- function(x) {
+  lp <- vector(mode = "numeric", length = nrow(modesandhessians))
+
+  for(z in 1:nrow(modesandhessians)) {
+    theta <- as.numeric(modesandhessians[z, thetanames])
+
+    dat_i <- prepare_dat(dat, i)
+    map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
+
     param_i_fixed_theta <- list(
@@
     )
-
-#' Always fixing the hyperparameters
-map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
-
-#' Going to store the functions p_LA(x_i, theta(z), y) in objs
-objs <- list()
-
-for(z in 1:nrow(modesandhessians)) {
- theta <- as.numeric(modesandhessians[z, thetanames])
-
-  dat_i <- prepare_dat(dat, i)
 
@@
 
-  objs[[z]] <- obj_fixed_theta
-}
-
-aghq_laplace_marginal <- function(x) {
-  pdf <- 0
-  for(z in 1:nrow(modesandhessians)) {
-    pdf <- pdf + exp(as.numeric(-objs[[z]]$fn(x))) * modesandhessians[z, "weights"]
-  }
-  return(pdf)
-}
-
-#' Issue is that we do the normalisation after now..
-aghq_laplace_marginal(x = x_test_value)
-aghq_laplace_marginal(x = -1)
+    lp[z] <- as.numeric(- obj_fixed_theta$fn(x)) + log(modesandhessians$weights[z])
+  }
+
+  matrixStats::logSumExp(lp) - quad$normalized_posterior$lognormconst
+}
+
+laplace_marginal(x_test_value)
+
+lps <- sapply(gg$nodes, laplace_marginal)
+plot_marginal_spline(lps)
 ```

### 2022-11-23 src/plot-tikz_algorithm-flowchart/algorithm-flowchart.tex
Add algorithm flowchart

\documentclass[tikz]{standalone}

% Things to add / improve:
% Bold versions of variables as they're all vectors
% Try to reduce whitespace a bit
% Arrow goes under rather than through box
% Dashed box around right hand side of diagram with i = 1, ..., N
% Make it clear AGHQ nodes and weights are being reused
% Add explanation of which x_i values to pick (and make clear these are coming from the Gaussian approximation
% Add spline interpolant detail
% Add how to get p(theta_j | y) from p(theta | y)
% (Maybe) put arrow for normalising constant on the top with arrow that says normalise
% (Maybe) add details about Laplace approximation e.g. making use of starting values, notation for mode and Hessian

\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning}

\begin{document}

\tikzstyle{question} = [rectangle, draw, text width=10em, text centered, rounded corners, minimum height=4em]
\tikzstyle{category} = [rectangle, draw, text width=10em, text centered, rounded corners, minimum height=4em]
\tikzstyle{container} = [rectangle, draw, rounded corners, dashed, inner sep=2em]
\tikzstyle{line} = [draw, -latex']

\begin{tikzpicture}[node distance = 2cm, auto]

    % theta
    \node[category] at (0, 0) (1) {$p(\theta, x, y)$};
    \node[category] at (0, -4) (2) {$\tilde p_{\text{LA}}(\theta, y)$};
    \node[category] at (0, -8) (3) {$\tilde p_{\text{AQ}}(y)$};
    \node[category] at (0, -12) (4) {$\tilde p_{\text{AQ}}(\theta \, | \, y)$};
    \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\text{G}}(x \, | \, \theta, y)$}(2);
    \path[line] (2) -- node[align=center] {Integrate out $\theta$ with AGHQ\\Nodes $z \in \mathcal{Q}(m, k)$\\Adapted $\theta(z) = \hat \theta + Lz$\\Weights $\omega(z):\mathcal{Q} \to \mathbb{R}$}(3);
    \path[line] (3) -- node[align=center] {Normalise} (4);
    \path[line] (2) edge[bend right] node [left] {} (4);

    % x
    %\node[container, fit={(0, 0)(10, -10)] {};
    \node[category] at (10, 0) (a) {$\tilde p_{\text{LA}}(x_i, \theta, y)$};
    \node[category] at (10, -4) (b) {$\tilde p_{\text{AQ}}(x_i, y)$};
    \node[category] at (10, -8) (c) {$\tilde p_{\text{AQ}}(x_i \, | \, y)$};
    \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\text{G}}(x_{-i} \, | \, x_i, \theta, y)$}(a);
    \path[line] (a) -- node[align=center] {Integrate out $\theta$\\$\sum_z \tilde p_{\text{LA}}(x_i, \theta(z), y) \omega(z)$}(b);
    \path[line] (b) -- node[align=center] {Normalise}(c);
    \path[line] (3) -- (c);

\end{tikzpicture}

\end{document}

### 2022-11-23 src/plot-tikz_algorithm-flowchart/script.R

@@ -1,0 +1,5 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("plot-tikz_algorithm-flowchart")
+# setwd("src/plot-tikz_algorithm-flowchart")
+
+tools::texi2dvi("algorithm-flowchart.tex", pdf = TRUE, clean = TRUE)

### 2022-11-23 src/dev_sinla/sinla.Rmd
You don't need a new MakeADFun for every z!

@@
 
-The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
-$$
-\begin{equation}
+An AGHQ approximation to the joint posterior of the latent field is given by
+$$
 \tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z), (\#eq:pxgiveny)
-\end{equation}
-$$
-where $\theta(z) \in \Theta$ and $\omega(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
+$$
+where $\theta(z) \in \Theta$ and $\omega(z) \in \mathbb{R}$ are nodes and weights generated by adaption to $\tilde p_\text{LA}(\theta \, | \, y)$.
 In particular, let
@@
 Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
-Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.
+Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode and $L$ is the lower Cholesky triangle of the curvature at the mode such that $LL^\top = H^{-1}$.
 
@@
 
-We check here that `obj$fn` indeed takes as input $(x_i, \theta)$:
+Check here that `obj$fn` indeed takes as input $(x_i, \theta)$:
 
@@
 transformed_param$phi_prev_minus_i <- NULL #' Remove the integrated out effects
-transformed_param
-
+
+#' transformed_param contains certain values for (x_i, theta) suitable for input to obj$fn
 obj$fn(transformed_param)
@@
     theme_minimal()
-```
-
-```{r}
+
+#' Table of average KS distance by method
 ks_results %>%
   group_by(type) %>%
-  summarise(D = mean(D))
+  summarise(Distance = mean(D)) %>%
+  select(Distance) %>%
+  mutate(
+    Method = c("$\\text{KS}(\\texttt{tmbstan}, \\texttt{gaussian})$",
+               "$\\text{KS}(\\texttt{tmbstan}, \\texttt{laplace})$"),
+    .before = Distance
+  ) %>%
+  kableExtra::kable(booktabs = TRUE, escape = FALSE, align = "c")
 ```
@@
 
-To start with, let's evaluate this on the log-scale for `i = 1`, `x = x_test_value` and the first value of `z`.
-The weight $\omega(z)$ will be `modesandhessians$weights[z]`.
-The normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ will be `- h_norm(as.numeric(modesandhessians[z, thetanames]))`.
-And to obtain the normalised Gaussian approximation $p_\text{G}(x_i \, | \, \theta(z), y)$ on the log-scale we get the relevant subcomponents of the joint Gaussian then use `ndorm` with `log = TRUE`.
+To start with, let's evaluate this on the log-scale for `i = 1`, `x = x_test_value` and the first value of `z`:
+
+* The weight $\omega(z)$ will be `modesandhessians$weights[z]`
+* The normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ will be `- h_norm(as.numeric(modesandhessians[z, thetanames]))`
+* For the normalised Gaussian approximation $p_\text{G}(x_i \, | \, \theta(z), y)$ on the log-scale we get the relevant subcomponents of the joint Gaussian then use `ndorm` with `log = TRUE`
 
@@
 ```
+## Full Laplace method (new new method)
+
+Try doing this without a new `MakeADFun` for each $z$:
+
+```{r}
+new_laplace_marginal <- function(x, i) {
+  dat_i <- prepare_dat(dat, i)
+
+  param_i <- list(
+    beta_prev = 0, 
+    phi_prev_i = 0, 
+    phi_prev_minus_i = rep(0, data$n - 1),
+    log_sigma_phi_prev = 0
+  )
+
+  obj <- MakeADFun(
+    data = dat_i,
+    parameters = param_i,
+    random = "phi_prev_minus_i",
+    DLL = "model1_index",
+    silent = TRUE
+  )
+
+  lp <- vector(mode = "numeric", length = nrow(modesandhessians))
+
+  for(z in 1:nrow(modesandhessians)) {
+    theta <- as.numeric(modesandhessians[z, thetanames])
+    lp[z] <- as.numeric(- obj$fn(c(theta[1], x, theta[2]))) + log(modesandhessians$weights[z])
+  }
+
+  matrixStats::logSumExp(lp) - quad$normalized_posterior$lognormconst
+}
+```
+
+We get the same value as we did with `laplace_marginal`!
+
+```{r}
+new_laplace_marginal(x_test_value, i = 1)
+
+lps <- sapply(gg$nodes, new_laplace_marginal, i = 1)
+plot_marginal_spline(lps)
+```
 

### 2022-11-23 src/dev_sinla/sinla.Rmd
spline_nodes function, loop new_laplace_marginal over index i, plot results

@@
 ```{r}
+#' Create a function that takes nodesandhessians and outputs the spline nodes for a particular marginal
+spline_nodes <- function(modesandhessians, i, k = 7) {
   #' Get the row number of modesandhessians which contains that theta mode
+  #' This is bad code (relying on objects outside function)
   theta_mode_location <- which.max(quad$normalized_posterior$nodesandweights$logpost_normalized)
@@
   #' Create Gauss-Hermite quadrature
-gg <- mvQuad::createNIGrid(dim = 1, type = "GHe", level = 7)
+  gg <- mvQuad::createNIGrid(dim = 1, type = "GHe", level = k)
 
@@
   gg$nodes
+}
 
 #' The log-probabilities at the set of input values
-lps <- sapply(gg$nodes, gaussian_marginal)
+lps <- sapply(spline_nodes(modesandhessians, i = 1, k = 7), gaussian_marginal)
 
@@
 plot_marginal_spline(lps)
+```
+Now let's loop this over $i$ so we can compare to the other methods.
+The hypothesis is that this will be better than the empirical Bayes methods.
+
+```{r}
+spline_nodes(modesandhessians, i = 1)
+
+new_laplace_marginal_spline_aghq <- function(i) {
+  x <- spline_nodes(modesandhessians, i = i)
+  lps <- sapply(x, new_laplace_marginal, i = i)
+  ss <- splines::interpSpline(x, lps, bSpline = TRUE, sparse = TRUE)
+  interpolant <- function(x) { as.numeric(stats::predict(ss, x)$y) }
+  finegrid <- seq(-5, 5, by = 0.1)
+  df <- data.frame(x = finegrid, y = exp(interpolant(finegrid)))
+  return(df)
+}
+
+aghq_laplace_df <- lapply(1:dat$n, new_laplace_marginal_spline_aghq) %>%
+  bind_rows(.id = "index") %>%
+  mutate(index = as.numeric(index))
+
+plot3 +
+  geom_line(data = aghq_laplace_df, aes(x = x, y = y), col = cbpalette[3])
+
+#' Need to edit this so that there are cdf and theta columns in aghq_laplace_df
+# samples_aghq_laplace <- lapply(1:dat$n, function(i) {
+#   aghq_laplace_df_i <- filter(aghq_laplace_df, index == i)
+#   samples <- approxfun(x = aghq_laplace_df_i$cdf, y = aghq_laplace_df_i$theta, ties = "ordered")(runif(M))
+#   data.frame("value" = samples)
+# }) %>%
+#   bind_rows(.id = "index")
 ```

### 2022-11-30 src/dev_sinla/sinla.Rmd
Moved from bookmarks

@@
 * [mgcv::ginla](https://github.com/cran/mgcv/blob/d9fe307c8ed1f8f1ed436c85e4ac415c1febeef4/R/inla.r)
+* [How important are Cholesky factorisations for performing computations with large Gaussians?](https://dpsimpson.github.io/pages/talks/Krylov_SUQ_2013.pdf)
 

### 2022-11-30 src/docs_xx-12-22-explainer/xx-12-22-explainer.Rmd
Working on lab group presentation

@@
 * This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
-* What we actually want is the posterior $p(\vartheta \, | \, y)$ which is **just**
+* What we actually want is the posterior $p(\vartheta \, | \, y)$ which is **just**^[I've bolded this for sarcasm.]
 $$
@@
 * Usually if you wanted to integrate something, you could use numerical integration methods
-* It doesn't work when $\text{dim}(\vartheta)$ gets even moderately sized
+* Pick nodes $\vartheta_i$ and weights $\omega_i$ then compute the sum
+$$
+\tilde p(y) = \sum_i p(y, \vartheta_i) \omega_i
+$$
 
-# Notation set-up
+# Monte Carlo as an example of numerical integration
 
+* Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$
+* If we set $\omega_i = 1/N$ for all $i$ then
+$$
+\tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i)
+$$
+
+# Monte Carlo is fundamentally unsound
+
+* "Monte Carlo ignores information" according to @o1987monte
+* Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is
+$$
+\tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right)
+$$
+* This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$
+
+# Application to HIV survey sampling
+
+* Suppose we're running a household survey, and sample the same individual twice
+* We didn't learn anything new about HIV by surveying them again!
+* This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+$\implies$ Bayesian quadrature, Bayesian survey design (end of digression)
+\end{tcolorbox}
+\end{center}
+
+# Latent variables and hyperparameters
+
+* Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized
 * Previously I had all of the parameters under the symbol $\vartheta$
 * What if we split them up as being $\vartheta = (x, \theta)$
+* The key part about this is that $\dim(x)$ is big and $\dim(\theta)$ is small
 

### 2022-12-01 src/example_naomi/naomi.Rmd
Working through Naomi vignette

--- 2022-09-16 src/example_naomi/naomi.Rmd
@@
 
-Area hierarchy and boundaries:
+Area hierarchy and boundaries, and plot to check we've got the right thing:
 
@@
 
-Population data:
+The population data is as follows, with columns:
+
+* `area_id`: contains the ISO, area level, and area identifier.
+* `source`: where did this information come from?
+* `calendar_quarter`: contains the type of year (such as CY for calendar year), the year, and the quarter
+* `sex`: either female or male.
+* `age_group`: in the format `Ylower_upper` where e.g. for 15-49 `lower = 15` and `upper = 49`.
+* `population`: the population, which might not be an integer I guess due to survey weighting.
+* `area_name`: the plain text version of the area name.
+* `asfr`: age-sex fertility rate.
 
@@
 
-Survey data:
+The survey data is as follows, with columns:
+
+* `indicator`: which survey indicator is it? Takes values `"prevalence"`, `"art_coverage"`, `"viral_suppression_plhiv"` and `"recent_infected"`.
+* `survey_id`: the name of the survey, containing the country, year and type.
+* `survey_mid_calendar_quarter`: which calendar quarter did the survey occur during?
+* `area_id`: as before.
+* `area_name`: as before.
+* `res_type`: unknown.
+* `sex`: as before.
+* `age_group`: as before.
+* `n_clusters`: the number of survey clusters in this area.
+* `n_observations`: the number of individuals surveyed.
+* `n_eff_kish`: the Kish effective sample size.
+* `estimate`: a survey weighted estimate of the indicator.
+* `std_error`: standard error of `estimate`.
+* `ci_lower`: 95% lower confidence interval for `estimate`.
+* `ci_upper`: 95% upper confidence interval for `estimate`.
 
@@
 ```
-ART programmatic data:
+
+The ART programmatic data is as follows, with columns:
+
+* `area_id`: as before.
+* `area_name`: as before.
+* `sex`: as before.
+* `age_group`: as before.
+* `year`: the year.
+* `calendar_quarter`: the calendar quarter.
+* `art_current`: unknown.
+* `art_new`: unknown.
+* `vl_tested_12mos`: viral load tested past 12 months.
+* `vl_suppressed_12mos`: viral load suppressed past 12 months.
 
@@
 
-ANC programmatic data:
+The ANC programmatic data is as follows, with columns:
 
@@
 
+```{r}
+pjnz <- system.file("extdata/demo_mwi2019.PJNZ", package = "naomi")
+spec <- extract_pjnz_naomi(pjnz)
+head(spec)
+```
+
 ## Choose model areas and time points
+
+* `scope`: A collection of `area_id`s defining the set of areas to be modelled. Usually this is simply national level, so the level 0 `area_id`.
+* `level`: Area level at which to fit the model.
+* `quarter_id_t1`: The first time point for the model -- approximately the midpoint of the household survey data used.
+* `quarter_id_t2`: The second time point for the model -- the current time for which estimates are needed.
+* `quarter_id_t3`: The third time point for the model -- the future projection for HIV estimates.
 
@@
 
+## Review input data
+
+This section is missing from the vignette!
+
 ## Prepare model inputs
+
+Setup the model:
 
@@
 ```
+
+Prepare data inputs
 
@@
 ```{r}
+tmb_inputs <- prepare_tmb_inputs(naomi_data)
 fit <- fit_tmb(tmb_inputs)
@@
 
+Calculate model outputs:
+
 ```{r}
-outputs <- output_package(fit, naomi_mf)
+outputs <- output_package(fit, naomi_data) #' Note that in the vignette they have naomi_mf here, is this a mistake?
+```
+
+The output package consists of a data frame of indicators and metadata defining the labels for each indicator:
+
+```{r}
 names(outputs)
 ```
+
+If uncertainty has not been calcualted yet, the output object retures values for the mode, but not the mean or lower and upper 95% uncertainty ranges:
 
@@
   dplyr::filter(
-    indicator == "prevalence",  # HIV prevalence
-    age_group == "Y015_049"   # Age group 15-49
+    indicator == "prevalence",  #' HIV prevalence
+    age_group == "Y015_049"   #' Age group 15-49
   ) %>%
@@
 ```
+
+The function `add_output_labels()` returns the indicators table with labels added as additional columns:
 
@@
 
+Regenerate outputs with uncertainty ranges:
+
 ```{r}
-system.time(outputs <- output_package(fit, naomi_mf))
+system.time(outputs <- output_package(fit, naomi_data))
 
@@
 
-# Using simplified TMB template
+## Plot model outputs
 
-Use the same data from above, but fit a model using our simplified Naomi TMB template, rather than the one from the package.
-This may require making some edits to the simplified TMB template, such as allowing non-integer values for survey-weighted cases and sample sizes.
+```{r}
+indicators <- add_output_labels(outputs) %>%
+  left_join(outputs$meta_area %>% select(area_level, area_id, center_x, center_y)) %>%
+  sf::st_as_sf()
+```
+
+15-49 prevalence by district (which is `area_level` 4):
+
+```{r}
+indicators %>%
+  filter(age_group == "Y015_049",
+         indicator == "prevalence",
+         area_level == 4) %>%
+  ggplot(aes(fill = mode)) +
+  geom_sf() +
+  viridis::scale_fill_viridis(labels = scales::percent_format()) +
+  th_map() +
+  facet_wrap(~sex)
+```
+
+15-49 prevalence by zone (which is `area_level` 2):
+
+```{r}
+indicators %>%
+  filter(age_group == "Y015_049",
+         indicator == "prevalence",
+         area_level == 2) %>%
+  ggplot(aes(fill = mode)) +
+  geom_sf() +
+  viridis::scale_fill_viridis(labels = scales::percent_format()) +
+  th_map() +
+  facet_wrap(~sex)
+```
+
+Age-specific prevalence at a national-level:
+
+```{r}
+indicators %>%
+  dplyr::filter(area_level == 0,
+         sex != "both",
+         age_group %in% get_five_year_age_groups(),
+         calendar_quarter == "CY2018Q3",
+         indicator == "prevalence") %>%
+  left_join(get_age_groups()) %>%
+  mutate(age_group = fct_reorder(age_group_label, age_group_sort_order)) %>%
+  ggplot(aes(age_group, mean, ymin = lower, ymax = upper, fill = sex)) +
+  geom_col(position = "dodge") +
+  geom_linerange(position = position_dodge(0.8)) +
+  scale_fill_brewer(palette = "Set1") +
+  scale_y_continuous(labels = scales::percent_format(1)) +
+  facet_wrap(~area_name) +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1.0, vjust = 0.5))
+```
+
+ART coverage by age-sex-region:
+
+```{r}
+indicators %>%
+  filter(area_level == 1,
+         sex != "both",
+         age_group %in% get_five_year_age_groups(),
+         indicator == "art_coverage",
+         calendar_quarter == "CY2018Q3") %>%
+  left_join(get_age_groups()) %>%
+  mutate(age_group = fct_reorder(age_group_label, age_group_sort_order)) %>%
+  ggplot(aes(age_group, mean, ymin = lower, ymax = upper, fill = sex)) +
+  geom_col(position = "dodge") +
+  geom_linerange(position = position_dodge(0.8)) +
+  scale_fill_brewer(palette = "Set1") +
+  scale_y_continuous(labels = scales::percent_format(1)) +
+  facet_wrap(~area_name) +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1.0, vjust = 0.5))
+```
+
+Bubble plot of prevalence and PLHIV:
+
+```{r}
+indicators %>%
+  filter(age_group == "Y015_064",
+         area_level == 4,
+         indicator %in% c("prevalence", "plhiv"),
+         calendar_quarter == "CY2018Q3") %>%
+  select(sex, center_x, center_y, indicator_label, mean) %>%
+  spread(indicator_label, mean) %>%
+  ggplot() +
+  geom_sf() +
+  geom_point(aes(center_x, center_y, colour = `HIV prevalence`, size = PLHIV)) +
+  viridis::scale_color_viridis(labels = scales::percent_format()) +
+  th_map() +
+  facet_wrap(~sex)
+```

### 2022-12-02 src/example_naomi/maths.Rmd
Start writing out the Naomi model in maths

---
title: "Mathematical description of Naomi"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
    **Background** The Naomi HIV model is specified in the appendix of Eaton et al. (2021).

    **Task** We translate the model to R Markdown in order to (1) aid understanding, (2) clarify any details which have been compressed for publication, (3) avoid having to read mathematics in Word.
---

# Notation

* District $x$
  * Located in Spectrum region file $R_x$
* Sex $s \in \{\text{F}, \text{M}\}$
* Age $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$
* Time $t \in \{T_1, T_2, T_3\}$
  * $T_1$: Most recent national household survey with HIV testing
  * $T_2$: Current time period at which to generate estimates
  * $T_3$: Short-term project period (typically 9-12 months)
  * Note $T_{1:3}$ specified to nearest calendar quarter
* Population size $N_{x, s, a, t}$
* HIV prevalence $\rho_{x, s, a, t}$
* ART coverage $\alpha_{x, s, a, t}$
* Annual HIV incidence rate $\lambda_{x, s, a, t}$

# HIV prevalence and ART coverage at $T_1$

## Background

* $u \sim \text{ICAR}(\sigma)$ refers to the intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$.
* $u \sim \text{BYM}2(\sigma, \phi)$ refers to a reparameterised Besag-York-Mollie model (BYM2) with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.
* For both the ICAR and BYM2 models, we follow recommendations on scaling, disconnected adjacency graph components, and islands.

## HIV prevalence

HIV prevalence modelled by
$$
\text{logit}(\rho_{x, s, a, T_1}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where:

  * $\beta^\rho_0$ is the intercept
  * $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women
  * $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women
  * $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$
  * $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women
  * $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$
  * $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence
  * $\eta^\rho_{R_x, s, a}$ are a fixed offset specifying assumed odds ratios for prevalence outside the age ranges for which data are available

For prior distributions, we use:

* $\mathcal{N}(0, 5)$ for all fixed effects ($\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}}$)
* $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms
* $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters
* $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters

## ART coverage

ART coverage modelled by
$$
\text{logit}(\alpha_{x, s, a, T_1}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a} 
$$
where terms are analogous to the HIV prevalence model.

## HIV incidence rate

HIV incidence rate is modelled by
$$
\log(\lambda_{x, s, a, t}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \cdots
$$

### 2022-12-05 src/example_naomi/maths.Rmd
More maths writing

@@
 $$
-\log(\lambda_{x, s, a, t}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \cdots
+\log(\lambda_{x, s, a, t}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x, t}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x, t}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a, t}^\lambda
 $$
+where:
+
+* $\beta^\lambda_0$ is the intercept, which is proportional to the average HIV transmission rate for untreated HIV positive adults
+* $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women
+* $\rho_{x, t}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
+$$
+\rho_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t}}
+$$
+* $\alpha_{x, t}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
+$$
+\alpha_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}
+$$
+* $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ from the EPP model
+* $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects
+* $\eta^\lambda_{R_x, s, a, t}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output
+
+## Short-term projection from $T_1 \to T_2$ and $T_2 \to T_3$
+
+### HIV prevalence
+
+* HIV population projected from $T_1 \to T_2$ based on survival of PLHIV from Spectrum output, ageing from one five-year age group to the next, and the addition of new HIV infections calculated from district-level HIV incidence rate and survival and ageing after infection
+* Let $H_{x, s, a, {T_1}} = N_{x, s, a, {T_1}} \cdot \rho_{x, s, a, {T_1}}$ be the number of PLHIV at $T_1$
+* For age groups $a \geq 5$ the number of PLHIV at $T_2$ is modelled by
+$$
+\begin{align*}
+H_{x, s, a, {T_2}} &= H_{x, s, a, {T_1}} \cdot S_{{R_x}, s, a \to a, {T_1}} + H_{x, s, a - 5, {T_1}} \cdot S_{{R_x}, s, a - 5 \to a, {T_1}} \\
+&+ (1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a, {T_1}})) \cdot (N_{x, s, a, {T_1}} - H_{x, s, a, {T_1}}) \cdot L_{R_x, s, a \to a, {T_1}} \\
+&+ (1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a - 5, {T_1}})) \cdot (N_{x, s, a - 5, {T_1}} - H_{x, s, a - 5, {T_1}}) \cdot L_{R_x, s, a - 5 \to a, {T_1}}
+\end{align*}
+$$
+* $S_{{R_x}, s, a \to a, {T_1}}$ is the ratio for the number of PLHIV surviving from $T_1$ to $T_2$ and remaining in age group $a$ (i.e. not aging from $a$ to $a + 5$)
+* $S_{{R_x}, s, a - 5 \to a, {T_1}}$ the survival ratio from $T_1$ to $T_2$ and aging from $a - 5$ to $a$
+* $(1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a, {T_1}}))$ is the probability of acquiring HIV between $T_1$ and $T_2$ of duration $\Delta_{T_1} = T_2 - T_1$ for the susceptible population $N_{x, s, a, {T_1}} - H_{x, s, a, {T_1}}$
+* $L_{R_x, s, a \to a, {T_1}}$ is the ratio surviving and remaining in age group $a$ for those infected between $T_1$ and $T_2$
+* $L_{R_x, s, a - 5 \to a, {T_1}}$ is the ratio surviving and aging from $a - 5$ to $a$ for those infected between $T_1$ and $T_2$
+* The ratios $S$ and $L$ are calculated from Specturm results by
+  1. Disaggregating single-year or single-age Specturm model results for PLHIV at $T_1$ and $T_2$ and new infections between $T_1$ and $T_2$ to quarterly birth cohorts
+  2. Subtracting the number of new infections within the cohort from the number of PLHIV at $T_2$ to calculate the number of surviving PLHIV in each cohort
+  3.
+Aggregating the survivors by quarter-age cohorts to those who survived and aged from $a$ to $a'$ between $T_1$ and $T_2$, and calculating the ratio by dividing by the initial PLHIV $H_{x, s, a, {T_1}}$
+* Variation in ART coverage or effectiveness is assumed not to affect survival
+
+### ART coverage
+
+* ART coverage is modelled by
+$$
+\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \beta_{T_2}^\alpha+\beta_{T_2, s=M}^\alpha+u_{x, T 2}^\alpha+u_{x, a<15, T_2}^\alpha+\eta_{R_x, s, a, T 2}^\alpha
+$$
+
+## Clarification notes to discuss with Jeff
+
+* Some confusion on how to offset terms work -- they seem cool though. Might prefer they not be called $\eta$ which is in my head linear predictor
+  * $\eta^\lambda_{R_x, s, a, t}$ aren't these $R_x, s, a, t$ specific rather than 
+* Aggregation over age 15-49 uses sum from $a = 15$ to $a = 45$? Perhaps this is because $a = 45$ actually corresponds to the age band 45-49
+* Sometimes in subscript we have $x, s, a, t$ and sometimes e.g. $x, s, at$ -- why?
+* $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ rather than e.g. $\sigma^\lambda \sim \mathcal{N}^+(0, 2.5)$
+* Inconsistent use of $T2$ and $T_2$

### 2022-12-06 src/example_naomi/maths.Rmd
Wow, more maths!

@@
 
-# Notation
+# Process specification
+
+## Notation
 
@@
 * Annual HIV incidence rate $\lambda_{x, s, a, t}$
-
-# HIV prevalence and ART coverage at $T_1$
-
-## Background
-
 * $u \sim \text{ICAR}(\sigma)$ refers to the intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$.
@@
 
-## HIV prevalence
+## HIV prevalence and ART coverage at $T_1$
+
+### HIV prevalence
 
@@
 
-## ART coverage
+### ART coverage
 
@@
 
-## HIV incidence rate
+### HIV incidence rate
 
@@
 
-* ART coverage is modelled by
+Change in ART coverage is modelled by
 $$
-\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \beta_{T_2}^\alpha+\beta_{T_2, s=M}^\alpha+u_{x, T 2}^\alpha+u_{x, a<15, T_2}^\alpha+\eta_{R_x, s, a, T 2}^\alpha
+\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \beta_{T_2}^\alpha + \beta_{T_2, s = \text{M}}^\alpha + u_{x, {T_2}}^\alpha + u_{x, a < 15, {T_2}}^\alpha + \eta_{R_x, s, a, {T_2}}^\alpha
+$$
+where
+
+* $\eta_{R_x, s, a, {T_2}}^\alpha$ are offsets for the logit change in ART coverage between $T_1$ and $T_2$ from Spectrum region $R_x$
+* $\beta_{T_2}^\alpha$ is the average change in logit ART coverage between $T_1$ and $T_2$ for adult women
+* $\beta_{T_2, s = \text{M}}^\alpha$ is the difference in average change for men compared to women
+* $u_{x, {T_2}}^\alpha \sim \mathcal{N}(0, \sigma_{XT}^\alpha)$ are district random effects
+* $u_{x, a < 15, {T_2}}^\alpha$ is the difference in logit ART coverage for children compared to adults
+
+**Flag for not understanding** ART coverage at $T_3$ is modelled by 
+$$
+\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \eta_{R_x, s, a, {T_3}}^\alpha
 $$
 
-## Clarification notes to discuss with Jeff
+## ANC testing cascade
 
+* For women 15-49 the predicted number of ANC clients $\Psi_{x, a, t}$ is a log-linear model
+$$
+\log (\Psi_{x, a, t}) = \log (N_{x, \text{F}, a, t}) + \psi_{R_x, a, t} + \beta^\psi + u_x^\psi
+$$
+  * $N_{x, \text{F}, a, t}$ are the female population sizes
+  * $\psi_{R_x, a, t}$ are ASFR in Spectrum region $R_x$ at time $t$
+  * $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility
+  * $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects
+* HIV prevalence $\rho_{x, a, t}^\text{ANC}$ and ART coverage $\alpha_{x, a, t}^\text{ANC}$ among pregnant women modelled with logit-linear models
+\begin{align}
+\text{logit}(\rho_{x, a, t}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a, t}) + \beta^{\rho^{\text{ANC}}} + \beta_{t \in(T_2, T_3\}}^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + u_{x, t \in\{T_2, T_3\}}^{\rho^{\text{ANC}}} + \eta_{R_x, a, t}^{\rho^{\text{ANC}}} \\
+\text{logit}(\alpha_{x, a, t}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a, t}) + \beta^{\alpha^{\text{ANC}}} + \beta_{t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}}+u_{x, t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + \eta_{R_x, a, t}^{\alpha^{\text{ANC}}}
+\end{align}
+  * $\eta_{R_x, a, t}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$
+  * $\beta^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the average differences between population and ANC outcomes at $T_1$ after removing the offset
+  * $\beta_{t \in(T_2, T_3\}}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the change in average difference from $T_1$ to $T_2$
+  * $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ are district random effects
+  * $u_{x, t \in\{T_2, T_3\}}^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_{XT}^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ capture the change in district random effects between times $T_1$ and $T_2$
+
+## ANC attendance
+
+* Number of PLHIV on ART is $A_{x, s, a, t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}$
+* Let $\gamma_{x, x', t} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$ at time $t$
+* We assume that $\gamma_{x, x', t} = 0$ for $x \notin \{x, \text{ne}(x)\}$ i.e.
+that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$ where $\sim$ is an adjacency relation
+  * $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x', t} = 1$
+* Model $\gamma_{x, x', t}$ with a multinomial logit model where $\tilde \gamma_{x, t}$ is the log odds ratio of seeking ART in the district $x'$ neighbouring $x$ **flag for confusion here**
+\begin{align}
+\tilde \gamma_{x, t} &= \tilde \gamma_0 + u_x^{\tilde \gamma} + u_x^{\tilde \gamma, t = T_2} \\ 
+\tilde \gamma_0 &= -4 \\
+u_x^{\tilde \gamma} &\sim \mathcal{N}(0, \sigma_X^{\tilde \gamma}) \\
+u_x^{\tilde \gamma, t \in \{T_2, T_3\}} &\sim \mathcal{N}(0, \sigma_{XT}^{\tilde \gamma}) 
+\end{align}
+  * $\tilde \gamma_{x, t}$ is an intercept specifying the prior mean for the log odds of seeking ART in each neighbouring district compared to the home district
+  * $u_x^{\tilde \gamma}$ are district random effects
+  * $u_x^{\tilde \gamma, t \in \{T_2, T_3\}}$ is the change in district level log-odds ratios between $T_1$ and $T_2$ -- **flag for confusion: shouldn't this just be $t = T_2$ here**
+  * Note: only one value estimated for each district, implying no preference between neighbouring districts, and no sex-age specific modelling
+* Calculate multinomial probabilities via softmax (n.b. this is the sigmoid function, the inverse of the softmax)
+$$
+\gamma_{x, x', t} = \frac{1}{1 + \sum_{x^\star \sim x} \exp(\tilde \gamma_{x^\star, t})}
+$$
+* The number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a, t} = A_{x, s, a, t} \cdot \gamma_{x, x', t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t} \cdot \gamma_{x, x', t}$
+* Total attending ART facilities in district $x'$ is $\tilde A_{x', s, a, t} = \sum_{x \sim x', x = x'} A_{x, x', s, a, t}$
+
+## Awareness of HIV status
+
+* The proportion of HIV positive adults aware of their HIV status is $v_{x, s, a, t}$
+* The proportion of untreated HIV positive adults aware of their HIV status is $\tilde \nu_{x, s, a, t}$ such that
+$$
+v_{x, s, a, t} = \alpha_{x, s, a, t} + \frac{\tilde v_{R_x, s, a, t}}{1 - \alpha_{x, s, a, t}}
+$$
+
+# Likelihood specification
+
+## Notation
+
+* Use $\{x\}$ to denote a collection of districts, $\{s\}$ a collection of sexes and $\{a\}$ a collection of ages for which an outcome is observed
+
+## Household survey data
+
+* Let $\nu$ be a household survey occuring at $T_1$
+* Furnishes weighted observations for HIV prevalence $\hat \rho_{\{x\}, \{s\}, \{a\}, \nu}$, ART coverage $\hat \alpha_{\{x\}, \{s\}, \{a\}, \nu}$ and proportion recently infected $\hat \kappa_{\{x\}, \{s\}, \{a\}, \nu}$ with respective Kish effective sample sizes $M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu}$ for $\theta \in \{\rho, \alpha, \kappa\}$
+* The observed cases are $Y^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} = M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}, \nu}$
+* For HIV prevalence we use a binomial working likelihood
+\begin{align}
+Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t}}
+\end{align}
+* For ART coverage we also use a binomial working likelihood
+\begin{align}
+Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}
+\end{align}
+* For recent infections we also use a binomial working likelihood
+\begin{align}
+Y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu}, \kappa_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\kappa_{x, s, a, t} &= 1 - \exp(- \lambda_{x, s, a, t} \cdot \frac{1 - \rho_{x, s, a, t}}{\rho_{x, s, a, t}} \cdot (\Omega_T - \beta_T ) - \beta_T ) \\
+\Omega_T &\sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T}) \\
+\beta_T &\sim \mathcal{N}(\beta_{T_0}, \sigma^{\beta_T})
+\end{align}
+
+# Clarification notes to discuss with Jeff
+
+* Not sure what $R_x$ means. Is it like $R_x: \mathcal{X} \to \mathcal{X}$
 * Some confusion on how to offset terms work -- they seem cool though. Might prefer they not be called $\eta$ which is in my head linear predictor
   * $\eta^\lambda_{R_x, s, a, t}$ aren't these $R_x, s, a, t$ specific rather than 
-* Aggregation over age 15-49 uses sum from $a = 15$ to $a = 45$? Perhaps this is because $a = 45$ actually corresponds to the age band 45-49
+* Aggregation over age 15-49 uses sum from $a = 15$ to $a = 45$? Perhaps this is because $a = 45$ actually corresponds to the age band 45-49. Edit: yes, $a = 45$ corresponds to 45-49
 * Sometimes in subscript we have $x, s, a, t$ and sometimes e.g. $x, s, at$ -- why?
-* $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ rather than e.g. $\sigma^\lambda \sim \mathcal{N}^+(0, 2.5)$
-* Inconsistent use of $T2$ and $T_2$
+* Why $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ rather than e.g. $\sigma^\lambda \sim \mathcal{N}^+(0, 2.5)$
+* Inconsistent use of $T2$ and $T_2$ in the appendix (I prefer the subscript so try to use throughout here)
+* In softmax function, confused as to what this equation is saying: shouldn't there be an equals sign in there before the large brackets?

### 2022-12-07 src/example_naomi/maths.Rmd
A little more while travelling

@@
 \end{align}
+  * $\kappa_{x, s, a, t}$ are the predicted proportion recently infected among HIV positive persons
+  * $\Omega_T$ is the mean duration of recent infection (MDRI)
+    * Use an informative prior based on the characteristics of the recent infection testing algorithm (RITA)
+    * For PHIA surveys $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$
+    * **Flag: what about for other surveys?*
+  * $\beta_T$ is the false recent ratio (FRR)
+    * For PHIA surveys $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$. **Flag: is this just assuming that there is no false recency? How is this assumption justified?*
+
+## ANC testing data
+
+* Include two years of ANC testing data: the year of the most recent survey $Y[T_1]$ and the current year $Y[T_2]$
+* $W^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients
+* $X^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with ascertained status
+* $Y^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with positive status (either known or tested)
+* $Z^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients already on ART prior to first ANC
+* Sometimes ANC testing data are only available for part of a given year, for example only the first three quarters.
+Denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$
 

### 2022-12-08 src/example_naomi/maths.Rmd
Finish off maths writing (and now naomi.Rmd doesn't render, because I updated the naomi package I think)

@@
 Denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$
+* Likelihood for number of ANC clients specified only for year $Y[T_2]$ (**why?**) as
+$$
+W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right)
+$$
+* Observed number of HIV positive and already on ART among ANC clients at both $Y[T_1]$ and $Y[T_2]$ are modelled by
+\begin{aligned}
+& Y_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[t]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right) \\
+& Z_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[t]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right)
+\end{aligned}
+where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a, t}$
+\begin{aligned}
+\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t}} \\
+\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}} \cdot \alpha_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}
+\end{aligned}
+
+## Number receiving ART
+
+* Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART **flag: confusion about the notation here.
+In general would be good to define precisely what is meant by $\{x\}$.
+Presumably the LHS below is a set, yet the RHS is a sum over the sets so I think just a number -- so how can the two be equal?**
+$$
+\dot A_{\{x\}, \{s\}, \{a\}, t} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a, t}
+$$
+* Model the unobserved numbers of ART clients travelling from $x'$ to $x$ as
+$$
+\dot A_{x', x, s, a, t} \sim \text{Bin}(N_{x', s, a, t}, \pi_{x', x, s, a, t})
+$$
+where $\pi_{x', x, s, a, t} = \rho_{x', s, a, t} \cdot \alpha_{x', s, a, t} \cdot \gamma_{x', s, a, t}$ **flag: should this be $\gamma_{x', x, s, a, t}$ since it's people moving from $x'$ to $x$ (and so far no where in the RHS is there an $x$ which is suspicious)**
+* This likelihood is approximated using a normal for the sum of binomials
+$$
+\dot A_{\{x\}, \{s\}, \{a\}, t} \sim \mathcal{N}(\tilde) A_{\{x\}, \{s\}, \{a\}, t}, \sigma^{\tilde A}_{\{x\}, \{s\}, \{a\}, t})
+$$
+where
+  * $\tilde{A}_{\{x\},\{s\}\{a\}, t} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a, t} \cdot \pi_{x', x, s, a, t}$
+  * $\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x^', x=x \prime} N_{x^', s, a, t} \cdot \pi_{x^', x, s, a, t} \cdot \left(1-\pi_{x^', x, s, a, t}\right)}$
+
+# Identifiability constaints
+
+* Depends on data availability. If data is missing, some parameters might be fixed to default values
+* Survey data on ART coverage by age and sex not available then set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age/sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$
+* If there is no ART data at $T_1$ (either survey or ART programme) or no ART data at $T_2$ then change in ART coverage is unidentifiable and set $\beta_{T_2}^\alpha = \beta_{T_2, s = \text{M}}^\alpha = u_{x, {T_2}}^\alpha = u_{x, a < 15, {T_2}}^\alpha = 0$ such that change in ART coverage is modelled by
+$$
+\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \eta_{R_x, s, a, {T_2}}^\alpha
+$$
+* If no ART data (survey or ART programme) are available at $T_1$ or $T_2$ but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
+In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated
+\begin{aligned}
+\text{logit} \left(\alpha_{x, s, a, T_1} \right) &= u_x^\alpha+\eta_{R_x, s, a}^\alpha \\
+\text{logit} \left(\alpha_{x, s, a, T_2} \right) &= \text{logit}\left(\alpha_{x, s, a, T_1}\right)+u_{x, T_2}^\alpha+\eta_{R_{x, s, a, T_2}}^\alpha
+\end{aligned}
+* If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$.
+The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage
 
@@
 * In softmax function, confused as to what this equation is saying: shouldn't there be an equals sign in there before the large brackets?
+* What's $Y[T2][T1]$ is this a typo and should be $Y[T1]$ -- in "The observed number of HIV positive and already on ART..."
+* "In cases where some data inputs, some model..." should be "In cases where some data inputs **are missing**, some model..."

### 2022-12-09 src/example_naomi/maths.Rmd
Fixes to maths

@@
 
-**Flag for not understanding** ART coverage at $T_3$ is modelled by 
+ART coverage at $T_3$ is modelled by^[Don't understand this.]
 $$
@@
   * $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x', t} = 1$
-* Model $\gamma_{x, x', t}$ with a multinomial logit model where $\tilde \gamma_{x, t}$ is the log odds ratio of seeking ART in the district $x'$ neighbouring $x$ **flag for confusion here**
+* Model $\gamma_{x, x', t}$ with a multinomial logit model where $\tilde \gamma_{x, t}$ is^[Confused by this notation.] the log odds ratio of seeking ART in the district $x'$ neighbouring $x$
 \begin{align}
@@
   * $u_x^{\tilde \gamma}$ are district random effects
-  * $u_x^{\tilde \gamma, t \in \{T_2, T_3\}}$ is the change in district level log-odds ratios between $T_1$ and $T_2$ -- **flag for confusion: shouldn't this just be $t = T_2$ here**
+  * $u_x^{\tilde \gamma, t \in \{T_2, T_3\}}$ is^[Shouldn't this just be $t = T_2$ here rather than $t \in \{T_2, T_3\}$?] the change in district level log-odds ratios between $T_1$ and $T_2$
   * Note: only one value estimated for each district, implying no preference between neighbouring districts, and no sex-age specific modelling
@@
     * Use an informative prior based on the characteristics of the recent infection testing algorithm (RITA)
-    * For PHIA surveys $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$
-    * **Flag: what about for other surveys?*
+    * For PHIA surveys^[What about other surveys?] $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$
   * $\beta_T$ is the false recent ratio (FRR)
-    * For PHIA surveys $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$. **Flag: is this just assuming that there is no false recency? How is this assumption justified?*
+    * For PHIA surveys $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.^[Is this just assuming that there is no false recency? How is this assumption justified?]
 
@@
 Denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$
-* Likelihood for number of ANC clients specified only for year $Y[T_2]$ (**why?**) as
+* Likelihood for number of ANC clients specified only^[Why?] for year $Y[T_2]$ as
 $$
@@
 
-* Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART **flag: confusion about the notation here.
+* Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART^[I'm confused about the notation here.
 In general would be good to define precisely what is meant by $\{x\}$.
-Presumably the LHS below is a set, yet the RHS is a sum over the sets so I think just a number -- so how can the two be equal?**
+Presumably the LHS below is a set, yet the RHS is a sum over the sets so I think just a number -- so how can the two be equal?]
 $$
@@
 $$
-where $\pi_{x', x, s, a, t} = \rho_{x', s, a, t} \cdot \alpha_{x', s, a, t} \cdot \gamma_{x', s, a, t}$ **flag: should this be $\gamma_{x', x, s, a, t}$ since it's people moving from $x'$ to $x$ (and so far no where in the RHS is there an $x$ which is suspicious)**
+where $\pi_{x', x, s, a, t} = \rho_{x', s, a, t} \cdot \alpha_{x', s, a, t} \cdot \gamma_{x', s, a, t}$^[Should this be $\gamma_{x', x, s, a, t}$ since it's people moving from $x'$ to $x$ (and so far no where in the RHS is there an $x$ which is suspicious)]
 * This likelihood is approximated using a normal for the sum of binomials
@@
   * $\tilde{A}_{\{x\},\{s\}\{a\}, t} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a, t} \cdot \pi_{x', x, s, a, t}$
-  * $\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x^', x=x \prime} N_{x^', s, a, t} \cdot \pi_{x^', x, s, a, t} \cdot \left(1-\pi_{x^', x, s, a, t}\right)}$
+  * $\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a, t} \cdot \pi_{x', x, s, a, t} \cdot \left(1-\pi_{x', x, s, a, t}\right)}$
 
@@
 * What's $Y[T2][T1]$ is this a typo and should be $Y[T1]$ -- in "The observed number of HIV positive and already on ART..."
-* "In cases where some data inputs, some model..." should be "In cases where some data inputs **are missing**, some model..."
+* "In cases where some data inputs, some model..." should be "In cases where some data inputs *are missing*, some model..."

### 2022-12-12 src/epil/epil.Rmd
Add data plot, MCMC diagnostics, more R-INLA strategies, and timings

@@
 ---
-title: "Epilepsy example"
+title: "Epilepsy GLMM example"
 author:
@@
 We then compare the results of the inference methods.
+This serves to compare inference methods, and check new methods are behaving as expected.
 ---
@@
 \begin{align*}
-  \eta_{ij}       &= \beta_0 + \beta_\texttt{Base} \log(\texttt{Baseline}_j / 4) + \beta_\texttt{Trt} \texttt{Trt}_i +
+\eta_{ij}
+&= \beta_0 + \beta_\texttt{Base} \log(\texttt{Baseline}_j / 4) + \beta_\texttt{Trt} \texttt{Trt}_i +
    \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
@@
 head(Epil)
+```
+
+```{r fig.cap="It looks like treatment is associated to fewer seizures on average, but of course we haven't accounted for the different covariates, hence the use of a statistical model!"}
+Epil %>% ggplot(aes(x =  as.factor(Trt), y = y)) +
+  geom_boxplot(alpha = 0.5, outlier = FALSE) +
+  geom_jitter(width = 0.1, alpha = 0.5) +
+  theme_minimal() +
+  labs(x = "Treatment", y = "Number of seizures")
 ```
@@
 ```{r}
+start_time_fit1 <- Sys.time()
+
 fit1 <- stan(
@@
   refresh = 0
+)
+
+end_time_fit1 <- Sys.time()
+time_fit1 <- end_time_fit1 - start_time_fit1
+```
+
+It took `r time_fit1` seconds to fit this model.
+All of the $\hat R$ values are below 1.05, so it looks like the MCMC has reached convergence:
+
+```{r}
+monitor1 <- rstan::monitor(fit1, print = FALSE)
+
+monitor1 %>%
+  as.data.frame() %>%
+  tibble::rownames_to_column(var = "parameter") %>%
+  ggplot(aes(x = parameter, y = Rhat)) +
+  geom_point(alpha = 0.5) +
+  geom_hline(yintercept = 1.05, linetype = "dashed") +
+  labs(x = "Parameter")  +
+  theme_minimal() +
+  theme(
+    axis.title.x = element_blank(),
+    axis.text.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    legend.position = "bottom"
   )
@@
 
+start_time_fit2 <- Sys.time()
 fit2 <- epil_inla(strat = "gaussian", int_strat = "eb")
+end_time_fit2 <- Sys.time()
+(time_fit2 <- end_time_fit2 - start_time_fit2)
+
+start_time_fit3 <- Sys.time()
 fit3 <- epil_inla(strat = "simplified.laplace", int_strat = "eb")
+end_time_fit3 <- Sys.time()
+(time_fit3 <- end_time_fit3 - start_time_fit3)
+
+start_time_fit4 <- Sys.time()
 fit4 <- epil_inla(strat = "laplace", int_strat = "eb")
+end_time_fit4 <- Sys.time()
+(time_fit4 <- end_time_fit4 - start_time_fit4)
 
+start_time_fit5 <- Sys.time()
 fit5 <- epil_inla(strat = "gaussian", int_strat = "grid")
+end_time_fit5 <- Sys.time()
+(time_fit5 <- end_time_fit5 - start_time_fit5)
+
+start_time_fit6 <- Sys.time()
 fit6 <- epil_inla(strat = "simplified.laplace", int_strat = "grid")
+end_time_fit6 <- Sys.time()
+(time_fit6 <- end_time_fit6 - start_time_fit6)
+
+start_time_fit7 <- Sys.time()
 fit7 <- epil_inla(strat = "laplace", int_strat = "grid")
+end_time_fit7 <- Sys.time()
+(time_fit7 <- end_time_fit7 - start_time_fit7)
 ```
@@
 ## `TMB`
+
+Template Model Builder, or `TMB`, is an R package which implements Laplace approximation using automatic differentiation.
+The user writes a C++ file, here `epil.cpp`, which contains the negative log-posterior for the model in question:
 
@@
 ```{r}
+#' Perhaps you could argue about when the "start" and "end" of the model fitting are
+#' For example, is compilation of the TMB template part of the model fitting?
+start_time_fit8 <- Sys.time()
+
 obj <- MakeADFun(
@@
 This can be checked with `names(obj$par)`.
-Optimise `obj` using 1000 iterations of the the `nlminb` optimiser, passing in the starting values `start`, objective function `objective` and its derivative `gradient`:
+We optimise `obj` using 1000 iterations of the the `nlminb` optimiser, passing in the starting values `start`, objective function `objective` and its derivative `gradient`:
 
@@
 )
+
+end_time_fit8 <- Sys.time()
+(time_fit8 <- end_time_fit8 - start_time_fit8)
 ```
@@
 For completeness, we check here that the objective functions as implemented in `TMB` and Stan are identical.
-To obtain the `TMB` negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
+To obtain the `TMB` negative log-posterior we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
 For Stan, we create an empty model, then use `rstan::log_prob`:
@@
 
-Now test the NLL of the initialisation parameters `param`:
+For example, we can test the NLL of the initialisation parameters `param`:
 
@@
 
-This look different!
+These look different!
 Is this a problem?
@@
 ```{r}
-formula6 <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT + (1 | rand) + (1 | Ind)
-fit6 <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"))
+start_time_fit9 <- Sys.time()
+formula9 <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT + (1 | rand) + (1 | Ind)
+fit9 <- glmmTMB(formula9, data = Epil, family = poisson(link = "log"))
+end_time_fit9 <- Sys.time()
+(time_fit9 <- end_time_fit9 - start_time_fit9)
 ```
@@
 ```{r} 
-fit7 <- tmbstan(obj = obj, chains = 4, refresh = 0)
+start_time_fit10 <- Sys.time()
+fit10 <- tmbstan(obj = obj, chains = 4, refresh = 0)
+end_time_fit10 <- Sys.time()
+(time_fit10 <- end_time_fit10 - start_time_fit10)
 ```
@@
 It also works smoothly with `TMB` templates.
-One (odd!) approach would be to use `glmmTMB` to get the `TMB` template which can be passed the `aghq`.
+
+One (non-standard!) approach would be to use `glmmTMB` to get the `TMB` template which can then be passed to `aghq`.
 
 ```{r}
-glmm_model_info <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"), doFit = FALSE)
+glmm_model_info <- glmmTMB(formula9, data = Epil, family = poisson(link = "log"), doFit = FALSE)
 
@@
 
-Another, more standard, approach is to use the `TMB` template, `obj`, that we already have:
+This would be useful if we didn't have a `TMB` template written, but we do, so let's just use that instead!
 
 ```{r results='hide'}
+start_time_fit11 <- Sys.time()
 fit11 <- aghq::marginal_laplace_tmb(
@@
 
-fit11_samples <- aghq::sample_marginal(fit8, M = 1000)$samps %>%
+fit11_samples <- aghq::sample_marginal(fit11, M = 1000)$samps %>%
   t() %>%
@@
   inf.utils::replace_duplicate_colnames()
+end_time_fit11 <- Sys.time()
+(time_fit11 <- end_time_fit11 - start_time_fit11)
 ```
@@
 
-### Using multiple C++ template
+### Empirical Bayes with multiple C++ template
 
@@
 
-### Using one C++ template
+### Empirical Bayes with one C++ template
 
@@
 
-Eventually create function of the form:
+Create function which takes as input the index `i` (from `index_controller`), the results of optimisation for $\theta$, and a grid of values for $x_i$ to evaluate at:
 
@@
 
+### AGHQ with one C++ template
+
+```{r}
+i <- 1
+dat$toggle <- index_controller[i, "toggle"]
+dat$i <- index_controller[i, "i"]
+```
+
 # Comparison
@@
 inla4 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
-inla5 <- as.vector(t(fit2$summary.fixed[1:6, 1:2]))
-inla6 <- as.vector(t(fit3$summary.fixed[1:6, 1:2]))
-inla7 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
+inla5 <- as.vector(t(fit5$summary.fixed[1:6, 1:2]))
+inla6 <- as.vector(t(fit6$summary.fixed[1:6, 1:2]))
+inla7 <- as.vector(t(fit7$summary.fixed[1:6, 1:2]))
 tmb8 <- as.vector(t(data.frame(sd_out$par.random[1:6], sqrt(sd_out$diag.cov.random[1:6]))))
-glmmtmb9 <- as.vector(t(summary(fit6)$coefficients$cond[, c("Estimate", "Std. Error")]))
-tmbstan10 <- as.vector(t(summary(fit7)$summary[1:6, c(1, 3)]))
+glmmtmb9 <- as.vector(t(summary(fit9)$coefficients$cond[, c("Estimate", "Std. Error")]))
+tmbstan10 <- as.vector(t(summary(fit10)$summary[1:6, c(1, 3)]))
 
-fit11_beta_samples <- fit8_samples[, 1:6]
-aghq11 <- as.vector(t(data.frame(mean = apply(fit11_beta_samples, 2, mean), sd = apply(fit8_beta_samples, 2, sd))))
+fit11_beta_samples <- fit11_samples[, 1:6]
+aghq11 <- as.vector(t(data.frame(mean = apply(fit11_beta_samples, 2, mean), sd = apply(fit11_beta_samples, 2, sd))))
 
@@
 rownames(df) <- c(sapply(0:5, beta_i))
-colnames(df) <- c("Stan", "EB-G", "EB-SL", "EB-L", "Grid-G", "Grid-SL", "Grid-L", "TMB", "glmmTMB", "tmbstan", "aghq", "adam", "type")
+method_names <- c("Stan", "EB-G", "EB-SL", "EB-L", "Grid-G", "Grid-SL", "Grid-L", "TMB", "glmmTMB", "tmbstan", "aghq", "adam")
+colnames(df) <- c(method_names, "type")
 
@@
 saveRDS(df, "comparison-results.rds")
+```
+
+```{r}
+time_df <- data.frame(
+  time = c(time_fit1, time_fit2, time_fit3, time_fit4, time_fit5, time_fit6, time_fit7, time_fit8, time_fit9, time_fit10, time_fit11),
+  method = head(method_names, n = 11)
+)
+
+time_df %>% ggplot(aes(x = method, y = time)) +
+  geom_col() +
+  theme_minimal() +
+  labs(x = "Method", y = "Time taken")
 ```

### 2022-12-13 src/docs_paper/appendix.Rmd
Add model description to appendix

@@
 ---
-title: Appendix to "Simplifying Integrated nested Laplace approximation with adaptive Gaussian Hermite quadrature"
+title: Appendix to "Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model"
 subtitle: |
@@
 
+# Complete Naomi model description
+
+## Process specification
+
+### Notation
+
+* District $x$
+  * Located in Spectrum region file $R_x$
+* Sex $s \in \{\text{F}, \text{M}\}$
+* Age $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$
+* Time $t \in \{T_1, T_2, T_3\}$
+  * $T_1$: Most recent national household survey with HIV testing
+  * $T_2$: Current time period at which to generate estimates
+  * $T_3$: Short-term project period (typically 9-12 months)
+  * Note $T_{1:3}$ specified to nearest calendar quarter
+* Population size $N_{x, s, a, t}$
+* HIV prevalence $\rho_{x, s, a, t}$
+* ART coverage $\alpha_{x, s, a, t}$
+* Annual HIV incidence rate $\lambda_{x, s, a, t}$
+* $u \sim \text{ICAR}(\sigma)$ refers to the intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$.
+* $u \sim \text{BYM}2(\sigma, \phi)$ refers to a reparameterised Besag-York-Mollie model (BYM2) with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.
+* For both the ICAR and BYM2 models, we follow recommendations on scaling, disconnected adjacency graph components, and islands.
+
+### HIV prevalence and ART coverage at $T_1$
+
+#### HIV prevalence
+
+HIV prevalence modelled by
+$$
+\text{logit}(\rho_{x, s, a, T_1}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a}
+$$
+where:
+
+* $\beta^\rho_0$ is the intercept
+* $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women
+* $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women
+* $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$
+* $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women
+* $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$
+* $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence
+* $\eta^\rho_{R_x, s, a}$ are a fixed offset specifying assumed odds ratios for prevalence outside the age ranges for which data are available
+
+For prior distributions, we use:
+
+* $\mathcal{N}(0, 5)$ for all fixed effects ($\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}}$)
+* $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms
+* $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters
+* $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters
+
+#### ART coverage
+
+ART coverage modelled by
+$$
+\text{logit}(\alpha_{x, s, a, T_1}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a}
+$$
+where terms are analogous to the HIV prevalence model.
+
+#### HIV incidence rate
+
+HIV incidence rate is modelled by
+$$
+\log(\lambda_{x, s, a, t}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x, t}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x, t}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a, t}^\lambda
+$$
+where:
+
+* $\beta^\lambda_0$ is the intercept, which is proportional to the average HIV transmission rate for untreated HIV positive adults
+* $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women
+* $\rho_{x, t}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
+$$
+\rho_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t}}
+$$
+* $\alpha_{x, t}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
+$$
+\alpha_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}
+$$
+* $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ from the EPP model
+* $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects
+* $\eta^\lambda_{R_x, s, a, t}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output
+
+### Short-term projection from $T_1 \to T_2$ and $T_2 \to T_3$
+
+#### HIV prevalence
+
+* HIV population projected from $T_1 \to T_2$ based on survival of PLHIV from Spectrum output, ageing from one five-year age group to the next, and the addition of new HIV infections calculated from district-level HIV incidence rate and survival and ageing after infection
+* Let $H_{x, s, a, {T_1}} = N_{x, s, a, {T_1}} \cdot \rho_{x, s, a, {T_1}}$ be the number of PLHIV at $T_1$
+* For age groups $a \geq 5$ the number of PLHIV at $T_2$ is modelled by
+\begin{align*}
+H_{x, s, a, {T_2}} &= H_{x, s, a, {T_1}} \cdot S_{{R_x}, s, a \to a, {T_1}} + H_{x, s, a - 5, {T_1}} \cdot S_{{R_x}, s, a - 5 \to a, {T_1}} \\
+&+ (1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a, {T_1}})) \cdot (N_{x, s, a, {T_1}} - H_{x, s, a, {T_1}}) \cdot L_{R_x, s, a \to a, {T_1}} \\
+&+ (1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a - 5, {T_1}})) \cdot (N_{x, s, a - 5, {T_1}} - H_{x, s, a - 5, {T_1}}) \cdot L_{R_x, s, a - 5 \to a, {T_1}}
+\end{align*}
+* $S_{{R_x}, s, a \to a, {T_1}}$ is the ratio for the number of PLHIV surviving from $T_1$ to $T_2$ and remaining in age group $a$ (i.e. not aging from $a$ to $a + 5$)
+* $S_{{R_x}, s, a - 5 \to a, {T_1}}$ the survival ratio from $T_1$ to $T_2$ and aging from $a - 5$ to $a$
+* $(1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a, {T_1}}))$ is the probability of acquiring HIV between $T_1$ and $T_2$ of duration $\Delta_{T_1} = T_2 - T_1$ for the susceptible population $N_{x, s, a, {T_1}} - H_{x, s, a, {T_1}}$
+* $L_{R_x, s, a \to a, {T_1}}$ is the ratio surviving and remaining in age group $a$ for those infected between $T_1$ and $T_2$
+* $L_{R_x, s, a - 5 \to a, {T_1}}$ is the ratio surviving and aging from $a - 5$ to $a$ for those infected between $T_1$ and $T_2$
+* The ratios $S$ and $L$ are calculated from Specturm results by
+  1. Disaggregating single-year or single-age Specturm model results for PLHIV at $T_1$ and $T_2$ and new infections between $T_1$ and $T_2$ to quarterly birth cohorts
+  2. Subtracting the number of new infections within the cohort from the number of PLHIV at $T_2$ to calculate the number of surviving PLHIV in each cohort
+  3.
+Aggregating the survivors by quarter-age cohorts to those who survived and aged from $a$ to $a'$ between $T_1$ and $T_2$, and calculating the ratio by dividing by the initial PLHIV $H_{x, s, a, {T_1}}$
+* Variation in ART coverage or effectiveness is assumed not to affect survival
+
+#### ART coverage
+
+Change in ART coverage is modelled by
+$$
+\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \beta_{T_2}^\alpha + \beta_{T_2, s = \text{M}}^\alpha + u_{x, {T_2}}^\alpha + u_{x, a < 15, {T_2}}^\alpha + \eta_{R_x, s, a, {T_2}}^\alpha
+$$
+where
+
+* $\eta_{R_x, s, a, {T_2}}^\alpha$ are offsets for the logit change in ART coverage between $T_1$ and $T_2$ from Spectrum region $R_x$
+* $\beta_{T_2}^\alpha$ is the average change in logit ART coverage between $T_1$ and $T_2$ for adult women
+* $\beta_{T_2, s = \text{M}}^\alpha$ is the difference in average change for men compared to women
+* $u_{x, {T_2}}^\alpha \sim \mathcal{N}(0, \sigma_{XT}^\alpha)$ are district random effects
+* $u_{x, a < 15, {T_2}}^\alpha$ is the difference in logit ART coverage for children compared to adults
+
+ART coverage at $T_3$ is modelled by^[Don't understand this.]
+$$
+\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \eta_{R_x, s, a, {T_3}}^\alpha
+$$
+
+### ANC testing cascade
+
+* For women 15-49 the predicted number of ANC clients $\Psi_{x, a, t}$ is a log-linear model
+$$
+\log (\Psi_{x, a, t}) = \log (N_{x, \text{F}, a, t}) + \psi_{R_x, a, t} + \beta^\psi + u_x^\psi
+$$
+  * $N_{x, \text{F}, a, t}$ are the female population sizes
+  * $\psi_{R_x, a, t}$ are ASFR in Spectrum region $R_x$ at time $t$
+  * $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility
+  * $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects
+* HIV prevalence $\rho_{x, a, t}^\text{ANC}$ and ART coverage $\alpha_{x, a, t}^\text{ANC}$ among pregnant women modelled with logit-linear models
+\begin{align}
+\text{logit}(\rho_{x, a, t}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a, t}) + \beta^{\rho^{\text{ANC}}} + \beta_{t \in(T_2, T_3\}}^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + u_{x, t \in\{T_2, T_3\}}^{\rho^{\text{ANC}}} + \eta_{R_x, a, t}^{\rho^{\text{ANC}}} \\
+\text{logit}(\alpha_{x, a, t}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a, t}) + \beta^{\alpha^{\text{ANC}}} + \beta_{t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}}+u_{x, t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + \eta_{R_x, a, t}^{\alpha^{\text{ANC}}}
+\end{align}
+  * $\eta_{R_x, a, t}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$
+  * $\beta^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the average differences between population and ANC outcomes at $T_1$ after removing the offset
+  * $\beta_{t \in(T_2, T_3\}}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the change in average difference from $T_1$ to $T_2$
+  * $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ are district random effects
+  * $u_{x, t \in\{T_2, T_3\}}^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_{XT}^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ capture the change in district random effects between times $T_1$ and $T_2$
+
+### ANC attendance
+
+* Number of PLHIV on ART is $A_{x, s, a, t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}$
+* Let $\gamma_{x, x', t} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$ at time $t$
+* We assume that $\gamma_{x, x', t} = 0$ for $x \notin \{x, \text{ne}(x)\}$ i.e.
+that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$ where $\sim$ is an adjacency relation
+  * $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x', t} = 1$
+* Model $\gamma_{x, x', t}$ with a multinomial logit model where $\tilde \gamma_{x, t}$ is^[Confused by this notation.] the log odds ratio of seeking ART in the district $x'$ neighbouring $x$
+\begin{align}
+\tilde \gamma_{x, t} &= \tilde \gamma_0 + u_x^{\tilde \gamma} + u_x^{\tilde \gamma, t = T_2} \\ 
+\tilde \gamma_0 &= -4 \\
+u_x^{\tilde \gamma} &\sim \mathcal{N}(0, \sigma_X^{\tilde \gamma}) \\
+u_x^{\tilde \gamma, t \in \{T_2, T_3\}} &\sim \mathcal{N}(0, \sigma_{XT}^{\tilde \gamma}) 
+\end{align}
+  * $\tilde \gamma_{x, t}$ is an intercept specifying the prior mean for the log odds of seeking ART in each neighbouring district compared to the home district
+  * $u_x^{\tilde \gamma}$ are district random effects
+  * $u_x^{\tilde \gamma, t \in \{T_2, T_3\}}$ is^[Shouldn't this just be $t = T_2$ here rather than $t \in \{T_2, T_3\}$?] the change in district level log-odds ratios between $T_1$ and $T_2$
+  * Note: only one value estimated for each district, implying no preference between neighbouring districts, and no sex-age specific modelling
+* Calculate multinomial probabilities via softmax (n.b. this is the sigmoid function, the inverse of the softmax)
+$$
+\gamma_{x, x', t} = \frac{1}{1 + \sum_{x^\star \sim x} \exp(\tilde \gamma_{x^\star, t})}
+$$
+* The number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a, t} = A_{x, s, a, t} \cdot \gamma_{x, x', t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t} \cdot \gamma_{x, x', t}$
+* Total attending ART facilities in district $x'$ is $\tilde A_{x', s, a, t} = \sum_{x \sim x', x = x'} A_{x, x', s, a, t}$
+
+### Awareness of HIV status
+
+* The proportion of HIV positive adults aware of their HIV status is $v_{x, s, a, t}$
+* The proportion of untreated HIV positive adults aware of their HIV status is $\tilde \nu_{x, s, a, t}$ such that
+$$
+v_{x, s, a, t} = \alpha_{x, s, a, t} + \frac{\tilde v_{R_x, s, a, t}}{1 - \alpha_{x, s, a, t}}
+$$
+
+## Likelihood specification
+
+### Notation
+
+* Use $\{x\}$ to denote a collection of districts, $\{s\}$ a collection of sexes and $\{a\}$ a collection of ages for which an outcome is observed
+
+### Household survey data
+
+* Let $\nu$ be a household survey occuring at $T_1$
+* Furnishes weighted observations for HIV prevalence $\hat \rho_{\{x\}, \{s\}, \{a\}, \nu}$, ART coverage $\hat \alpha_{\{x\}, \{s\}, \{a\}, \nu}$ and proportion recently infected $\hat \kappa_{\{x\}, \{s\}, \{a\}, \nu}$ with respective Kish effective sample sizes $M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu}$ for $\theta \in \{\rho, \alpha, \kappa\}$
+* The observed cases are $Y^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} = M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}, \nu}$
+* For HIV prevalence we use a binomial working likelihood
+\begin{align}
+Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t}}
+\end{align}
+* For ART coverage we also use a binomial working likelihood
+\begin{align}
+Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}
+\end{align}
+* For recent infections we also use a binomial working likelihood
+\begin{align}
+Y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu}, \kappa_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\kappa_{x, s, a, t} &= 1 - \exp(- \lambda_{x, s, a, t} \cdot \frac{1 - \rho_{x, s, a, t}}{\rho_{x, s, a, t}} \cdot (\Omega_T - \beta_T ) - \beta_T ) \\
+\Omega_T &\sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T}) \\
+\beta_T &\sim \mathcal{N}(\beta_{T_0}, \sigma^{\beta_T})
+\end{align}
+  * $\kappa_{x, s, a, t}$ are the predicted proportion recently infected among HIV positive persons
+  * $\Omega_T$ is the mean duration of recent infection (MDRI)
+    * Use an informative prior based on the characteristics of the recent infection testing algorithm (RITA)
+    * For PHIA surveys^[What about other surveys?] $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$
+  * $\beta_T$ is the false recent ratio (FRR)
+    * For PHIA surveys $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.^[Is this just assuming that there is no false recency? How is this assumption justified?]
+
+### ANC testing data
+
+* Include two years of ANC testing data: the year of the most recent survey $Y[T_1]$ and the current year $Y[T_2]$
+* $W^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients
+* $X^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with ascertained status
+* $Y^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with positive status (either known or tested)
+* $Z^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients already on ART prior to first ANC
+* Sometimes ANC testing data are only available for part of a given year, for example only the first three quarters.
+Denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$
+* Likelihood for number of ANC clients specified only^[Why?] for year $Y[T_2]$ as
+$$
+W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right)
+$$
+* Observed number of HIV positive and already on ART among ANC clients at both $Y[T_1]$ and $Y[T_2]$ are modelled by
+\begin{align*}
+& Y_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[t]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right) \\
+& Z_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[t]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right)
+\end{align*}
+where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a, t}$
+\begin{align*}
+\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t}} \\
+\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}} \cdot \alpha_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}
+\end{align*}
+
+### Number receiving ART
+
+* Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART^[I'm confused about the notation here.
+In general would be good to define precisely what is meant by $\{x\}$.
+Presumably the LHS below is a set, yet the RHS is a sum over the sets so I think just a number -- so how can the two be equal?]
+$$
+\dot A_{\{x\}, \{s\}, \{a\}, t} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a, t}
+$$
+* Model the unobserved numbers of ART clients travelling from $x'$ to $x$ as
+$$
+\dot A_{x', x, s, a, t} \sim \text{Bin}(N_{x', s, a, t}, \pi_{x', x, s, a, t})
+$$
+where $\pi_{x', x, s, a, t} = \rho_{x', s, a, t} \cdot \alpha_{x', s, a, t} \cdot \gamma_{x', s, a, t}$^[Should this be $\gamma_{x', x, s, a, t}$ since it's people moving from $x'$ to $x$ (and so far no where in the RHS is there an $x$ which is suspicious)]
+* This likelihood is approximated using a normal for the sum of binomials
+$$
+\dot A_{\{x\}, \{s\}, \{a\}, t} \sim \mathcal{N}(\tilde) A_{\{x\}, \{s\}, \{a\}, t}, \sigma^{\tilde A}_{\{x\}, \{s\}, \{a\}, t})
+$$
+where
+  * $\tilde{A}_{\{x\},\{s\}\{a\}, t} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a, t} \cdot \pi_{x', x, s, a, t}$
+  * $\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a, t} \cdot \pi_{x', x, s, a, t} \cdot \left(1-\pi_{x', x, s, a, t}\right)}$
+
+## Identifiability constaints
+
+* Depends on data availability. If data is missing, some parameters might be fixed to default values
+* Survey data on ART coverage by age and sex not available then set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age/sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$
+* If there is no ART data at $T_1$ (either survey or ART programme) or no ART data at $T_2$ then change in ART coverage is unidentifiable and set $\beta_{T_2}^\alpha = \beta_{T_2, s = \text{M}}^\alpha = u_{x, {T_2}}^\alpha = u_{x, a < 15, {T_2}}^\alpha = 0$ such that change in ART coverage is modelled by
+$$
+\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \eta_{R_x, s, a, {T_2}}^\alpha
+$$
+* If no ART data (survey or ART programme) are available at $T_1$ or $T_2$ but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
+In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated
+\begin{align*}
+\text{logit} \left(\alpha_{x, s, a, T_1} \right) &= u_x^\alpha+\eta_{R_x, s, a}^\alpha \\
+\text{logit} \left(\alpha_{x, s, a, T_2} \right) &= \text{logit}\left(\alpha_{x, s, a, T_1}\right)+u_{x, T_2}^\alpha+\eta_{R_{x, s, a, T_2}}^\alpha
+\end{align*}
+* If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$.
+The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage
+
 # References {#references .unnumbered}

### 2022-12-13 src/docs_paper/paper.Rmd
Add model description to appendix

@@
 ---
-title: Integrated nested Laplace approximations for extended latent Gaussian models
+title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
 author:

### 2022-12-17 src/dev_sinla/sinla.Rmd
Small bug in dev_sinla

@@
 #' The log-probabilities at the set of input values
-lps <- sapply(spline_nodes(modesandhessians, i = 1, k = 7), gaussian_marginal)
-
-plot_marginal_spline <- function(lps) {
+nodes <- spline_nodes(modesandhessians, i = 1, k = 7)
+lps <- sapply(nodes, gaussian_marginal)
+
+plot_marginal_spline <- function(nodes, lps) {
   #' Build a Lagrange polynomial interpolant of the marginal posterior
-  ss <- splines::interpSpline(gg$nodes, lps, bSpline = TRUE, sparse = TRUE)
+  ss <- splines::interpSpline(nodes, lps, bSpline = TRUE, sparse = TRUE)
   interpolant <- function(x) { as.numeric(stats::predict(ss, x)$y) }
@@
 
-plot_marginal_spline(lps)
+plot_marginal_spline(nodes, lps)
 ```
@@
 
-lps <- sapply(gg$nodes, old_laplace_marginal)
-plot_marginal_spline(lps)
+lps <- sapply(nodes, old_laplace_marginal)
+plot_marginal_spline(nodes, lps)
 ```
@@
 
-lps <- sapply(gg$nodes, laplace_marginal)
-plot_marginal_spline(lps)
-```
+lps <- sapply(nodes, laplace_marginal)
+plot_marginal_spline(nodes, lps)
+```
+
 ## Full Laplace method (new new method)
@@
 
-lps <- sapply(gg$nodes, new_laplace_marginal, i = 1)
-plot_marginal_spline(lps)
+lps <- sapply(nodes, new_laplace_marginal, i = 1)
+plot_marginal_spline(nodes, lps)
 ```

### 2022-12-18 src/plot-tikz_algorithm-flowchart/algorithm-flowchart.tex
Update flowchart

@@
 % Things to add / improve:
-% Bold versions of variables as they're all vectors
 % Try to reduce whitespace a bit
 % Arrow goes under rather than through box
-% Dashed box around right hand side of diagram with i = 1, ..., N
+% Dashed box around x with i = 1, ..., N
+% Dashed box around theta with j = 1, ..., m
 % Make it clear AGHQ nodes and weights are being reused
-% Add explanation of which x_i values to pick (and make clear these are coming from the Gaussian approximation
-% Add spline interpolant detail
-% Add how to get p(theta_j | y) from p(theta | y)
+% AGHQ nodes from Gaussian marginal
+% More detail than "Marginalise"
 % (Maybe) put arrow for normalising constant on the top with arrow that says normalise
@@
 \usetikzlibrary{shapes,arrows,positioning}
+\usepackage{bm}
+
+\newcommand{\x}{\mathbf{x}}
+\newcommand{\y}{\mathbf{y}}
+\newcommand{\z}{\mathbf{z}}
+\newcommand{\btheta}{\bm{\theta}}
 
@@
 
-    % theta
-    \node[category] at (0, 0) (1) {$p(\theta, x, y)$};
-    \node[category] at (0, -4) (2) {$\tilde p_{\text{LA}}(\theta, y)$};
-    \node[category] at (0, -8) (3) {$\tilde p_{\text{AQ}}(y)$};
-    \node[category] at (0, -12) (4) {$\tilde p_{\text{AQ}}(\theta \, | \, y)$};
-    \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\text{G}}(x \, | \, \theta, y)$}(2);
-    \path[line] (2) -- node[align=center] {Integrate out $\theta$ with AGHQ\\Nodes $z \in \mathcal{Q}(m, k)$\\Adapted $\theta(z) = \hat \theta + Lz$\\Weights $\omega(z):\mathcal{Q} \to \mathbb{R}$}(3);
-    \path[line] (3) -- node[align=center] {Normalise} (4);
-    \path[line] (2) edge[bend right] node [left] {} (4);
+    \node[category] at (0, 0) (1) {$p(\theta, \x, \y)$};
 
     % x
-    %\node[container, fit={(0, 0)(10, -10)] {};
-    \node[category] at (10, 0) (a) {$\tilde p_{\text{LA}}(x_i, \theta, y)$};
-    \node[category] at (10, -4) (b) {$\tilde p_{\text{AQ}}(x_i, y)$};
-    \node[category] at (10, -8) (c) {$\tilde p_{\text{AQ}}(x_i \, | \, y)$};
-    \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\text{G}}(x_{-i} \, | \, x_i, \theta, y)$}(a);
-    \path[line] (a) -- node[align=center] {Integrate out $\theta$\\$\sum_z \tilde p_{\text{LA}}(x_i, \theta(z), y) \omega(z)$}(b);
+    \node[category] at (0, -3) (a) {$\tilde p_{\text{LA}}(x_i, \btheta, \y)$};
+    \node[category] at (0, -6) (b) {$\tilde p_{\text{AQ}}(x_i, \y)$};
+    \node[category] at (0, -9) (c) {$\tilde p_{\text{AQ}}(x_i \, | \, \y)$};
+    \node[category] at (0, -12) (d) {Image of a nice curve};
+    \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\text{G}}(\x_{-i} \, | \, x_i, \btheta, \y)$}(a);
+    \path[line] (a) -- node[align=center] {Integrate out $\theta$\\$\sum_\z \tilde p_{\text{LA}}(x_i, \btheta(\z), \y) \omega(\z)$}(b);
     \path[line] (b) -- node[align=center] {Normalise}(c);
+    \path[line] (c) -- node[align=center] {Evaluate $k = 7$ AGHQ nodes\\Use spline interpolation}(d);
+
+    % theta
+    \node[category] at (8, 0) (2) {$\tilde p_{\text{LA}}(\btheta, \y)$};
+    \node[category] at (8, -3) (3) {$\tilde p_{\text{AQ}}(\y)$};
+    \node[category] at (8, -6) (4) {$\tilde p_{\text{AQ}}(\btheta \, | \, \y)$};
+    \node[category] at (8, -9) (5) {$\tilde p_{\text{AQ}}(\theta_j \, | \, \y)$};
+    \node[category] at (8, -12) (6) {Image of a nice curve};
+    \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\text{G}}(\x \, | \, \btheta, \y)$}(2);
+    \path[line] (2) -- node[align=center] {Integrate out $\btheta$ with AGHQ\\Nodes $\z \in \mathcal{Q}(m, k)$\\Adapted $\btheta(\z) = \hat{\btheta} + Lz$\\Weights $\omega(\z):\mathcal{Q} \to \mathbb{R}$}(3);
+    \path[line] (3) -- node[align=center] {Normalise} (4);
+    \path[line] (2) edge[bend left] node [left] {} (4);
     \path[line] (3) -- (c);
+    \path[line] (4) -- node[align=center] {Marginalise} (5);
+    \path[line] (5) -- node[align=center] {Evaluate $k = 7$ AGHQ nodes\\Use spline interpolation}(6);
 

### 2022-12-19 src/docs_paper/appendix.Rmd
Start writing appendix into prose. Waiting on go ahead from Jeff for parts of the model to strip out

@@
 
-* District $x$
-  * Located in Spectrum region file $R_x$
-* Sex $s \in \{\text{F}, \text{M}\}$
-* Age $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$
-* Time $t \in \{T_1, T_2, T_3\}$
-  * $T_1$: Most recent national household survey with HIV testing
-  * $T_2$: Current time period at which to generate estimates
-  * $T_3$: Short-term project period (typically 9-12 months)
-  * Note $T_{1:3}$ specified to nearest calendar quarter
-* Population size $N_{x, s, a, t}$
-* HIV prevalence $\rho_{x, s, a, t}$
-* ART coverage $\alpha_{x, s, a, t}$
-* Annual HIV incidence rate $\lambda_{x, s, a, t}$
-* $u \sim \text{ICAR}(\sigma)$ refers to the intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$.
+Let $x$ be a district, located within the Spectrum region $R_x$, $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
+Consider two time-points $t \in \{T_1, T_2\}$, specified to the nearest calendar quarter: $T_1$, the time period of the most recent national household survey with HIV testing, and $T_2$, the current time period at which to generate estimates.
+We index population size $N_{x, s, a, t}$, HIV prevalence $\rho_{x, s, a, t}$, ART coverage $\alpha_{x, s, a, t}$ and annual HIV incidence rate $\lambda_{x, s, a, t}$ by district, sex, age-band and time period.
+
+* $u \sim \text{ICAR}(\sigma)$ refers to the Besag intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$.
 * $u \sim \text{BYM}2(\sigma, \phi)$ refers to a reparameterised Besag-York-Mollie model (BYM2) with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.

### 2022-12-20 src/docs_paper/appendix.Rmd
Start removing T2 and T3 from model in appendix

@@
 
-# Complete Naomi model description
+# Modified Naomi model description
+
+## Notation
+
+## Indexing
+
+Let $x$ be a district, located within the Spectrum region $R_x$, $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
+Consider the time period $t = T_1$ of the most recent national household survey with HIV testing specified to the nearest calendar quarter.
+We index population size $N_{x, s, a, t}$, HIV prevalence $\rho_{x, s, a, t}$, ART coverage $\alpha_{x, s, a, t}$ and annual HIV incidence rate $\lambda_{x, s, a, t}$ by district, sex, age-band and time period.
+
+## Structured random effects
+
+We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$, and $u \sim \text{BYM}2(\sigma, \phi)$ to the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.
+For both the ICAR and BYM2 models, we follow recommendations [@freni2018note] on scaling, disconnected adjacency graph components, and islands.
 
@@
 
-### Notation
+### HIV prevalence
 
-Let $x$ be a district, located within the Spectrum region $R_x$, $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
-Consider two time-points $t \in \{T_1, T_2\}$, specified to the nearest calendar quarter: $T_1$, the time period of the most recent national household survey with HIV testing, and $T_2$, the current time period at which to generate estimates.
-We index population size $N_{x, s, a, t}$, HIV prevalence $\rho_{x, s, a, t}$, ART coverage $\alpha_{x, s, a, t}$ and annual HIV incidence rate $\lambda_{x, s, a, t}$ by district, sex, age-band and time period.
-
-* $u \sim \text{ICAR}(\sigma)$ refers to the Besag intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$.
-* $u \sim \text{BYM}2(\sigma, \phi)$ refers to a reparameterised Besag-York-Mollie model (BYM2) with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.
-* For both the ICAR and BYM2 models, we follow recommendations on scaling, disconnected adjacency graph components, and islands.
-
-### HIV prevalence and ART coverage at $T_1$
-
-#### HIV prevalence
-
-HIV prevalence modelled by
+We model HIV prevalence by
 $$
@@
 $$
-where:
+where $\beta^\rho_0$ is the intercept, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are a fixed offset specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
+We use the prior distributions $\mathcal{N}(0, 5)$ for all fixed effects ($\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}}$), $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.
 
-* $\beta^\rho_0$ is the intercept
-* $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women
-* $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women
-* $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$
-* $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women
-* $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$
-* $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence
-* $\eta^\rho_{R_x, s, a}$ are a fixed offset specifying assumed odds ratios for prevalence outside the age ranges for which data are available
+### ART coverage
 
-For prior distributions, we use:
-
-* $\mathcal{N}(0, 5)$ for all fixed effects ($\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}}$)
-* $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms
-* $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters
-* $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters
-
-#### ART coverage
-
-ART coverage modelled by
+We model ART coverage by
 $$
@@
 $$
-where terms are analogous to the HIV prevalence model.
+with analogous terms to the HIV prevalence model.
 
@@
 
-HIV incidence rate is modelled by
+We model HIV incidence rate by
 $$
@@
 $$
-where:
-
-* $\beta^\lambda_0$ is the intercept, which is proportional to the average HIV transmission rate for untreated HIV positive adults
-* $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women
-* $\rho_{x, t}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
+where $\beta^\lambda_0$ is the intercept, which is proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women, $\rho_{x, t}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
 $$
-\rho_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t}}
+\rho_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t}},
 $$
-* $\alpha_{x, t}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
+$\alpha_{x, t}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
 $$
-\alpha_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}
+\alpha_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}},
 $$
-* $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ from the EPP model
-* $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects
-* $\eta^\lambda_{R_x, s, a, t}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output
-
-### Short-term projection from $T_1 \to T_2$ and $T_2 \to T_3$
-
-#### HIV prevalence
-
-* HIV population projected from $T_1 \to T_2$ based on survival of PLHIV from Spectrum output, ageing from one five-year age group to the next, and the addition of new HIV infections calculated from district-level HIV incidence rate and survival and ageing after infection
-* Let $H_{x, s, a, {T_1}} = N_{x, s, a, {T_1}} \cdot \rho_{x, s, a, {T_1}}$ be the number of PLHIV at $T_1$
-* For age groups $a \geq 5$ the number of PLHIV at $T_2$ is modelled by
-\begin{align*}
-H_{x, s, a, {T_2}} &= H_{x, s, a, {T_1}} \cdot S_{{R_x}, s, a \to a, {T_1}} + H_{x, s, a - 5, {T_1}} \cdot S_{{R_x}, s, a - 5 \to a, {T_1}} \\
-&+ (1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a, {T_1}})) \cdot (N_{x, s, a, {T_1}} - H_{x, s, a, {T_1}}) \cdot L_{R_x, s, a \to a, {T_1}} \\
-&+ (1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a - 5, {T_1}})) \cdot (N_{x, s, a - 5, {T_1}} - H_{x, s, a - 5, {T_1}}) \cdot L_{R_x, s, a - 5 \to a, {T_1}}
-\end{align*}
-* $S_{{R_x}, s, a \to a, {T_1}}$ is the ratio for the number of PLHIV surviving from $T_1$ to $T_2$ and remaining in age group $a$ (i.e. not aging from $a$ to $a + 5$)
-* $S_{{R_x}, s, a - 5 \to a, {T_1}}$ the survival ratio from $T_1$ to $T_2$ and aging from $a - 5$ to $a$
-* $(1 - \exp(- \delta_{T_1} \cdot \lambda_{x, s, a, {T_1}}))$ is the probability of acquiring HIV between $T_1$ and $T_2$ of duration $\Delta_{T_1} = T_2 - T_1$ for the susceptible population $N_{x, s, a, {T_1}} - H_{x, s, a, {T_1}}$
-* $L_{R_x, s, a \to a, {T_1}}$ is the ratio surviving and remaining in age group $a$ for those infected between $T_1$ and $T_2$
-* $L_{R_x, s, a - 5 \to a, {T_1}}$ is the ratio surviving and aging from $a - 5$ to $a$ for those infected between $T_1$ and $T_2$
-* The ratios $S$ and $L$ are calculated from Specturm results by
-  1. Disaggregating single-year or single-age Specturm model results for PLHIV at $T_1$ and $T_2$ and new infections between $T_1$ and $T_2$ to quarterly birth cohorts
-  2. Subtracting the number of new infections within the cohort from the number of PLHIV at $T_2$ to calculate the number of surviving PLHIV in each cohort
-  3.
-Aggregating the survivors by quarter-age cohorts to those who survived and aged from $a$ to $a'$ between $T_1$ and $T_2$, and calculating the ratio by dividing by the initial PLHIV $H_{x, s, a, {T_1}}$
-* Variation in ART coverage or effectiveness is assumed not to affect survival
-
-#### ART coverage
-
-Change in ART coverage is modelled by
-$$
-\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \beta_{T_2}^\alpha + \beta_{T_2, s = \text{M}}^\alpha + u_{x, {T_2}}^\alpha + u_{x, a < 15, {T_2}}^\alpha + \eta_{R_x, s, a, {T_2}}^\alpha
-$$
-where
-
-* $\eta_{R_x, s, a, {T_2}}^\alpha$ are offsets for the logit change in ART coverage between $T_1$ and $T_2$ from Spectrum region $R_x$
-* $\beta_{T_2}^\alpha$ is the average change in logit ART coverage between $T_1$ and $T_2$ for adult women
-* $\beta_{T_2, s = \text{M}}^\alpha$ is the difference in average change for men compared to women
-* $u_{x, {T_2}}^\alpha \sim \mathcal{N}(0, \sigma_{XT}^\alpha)$ are district random effects
-* $u_{x, a < 15, {T_2}}^\alpha$ is the difference in logit ART coverage for children compared to adults
-
-ART coverage at $T_3$ is modelled by^[Don't understand this.]
-$$
-\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \eta_{R_x, s, a, {T_3}}^\alpha
-$$
+$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ from the EPP model, $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a, t}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output
 
@@
 
-* For women 15-49 the predicted number of ANC clients $\Psi_{x, a, t}$ is a log-linear model
+For women 15-49 the predicted number of ANC clients $\Psi_{x, a, t}$ is a log-linear model
 $$
@@
 $$
-  * $N_{x, \text{F}, a, t}$ are the female population sizes
-  * $\psi_{R_x, a, t}$ are ASFR in Spectrum region $R_x$ at time $t$
-  * $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility
-  * $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects
-* HIV prevalence $\rho_{x, a, t}^\text{ANC}$ and ART coverage $\alpha_{x, a, t}^\text{ANC}$ among pregnant women modelled with logit-linear models
+where $N_{x, \text{F}, a, t}$ are the female population sizes, $\psi_{R_x, a, t}$ are ASFR in Spectrum region $R_x$ at time $t$, $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility, $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects, HIV prevalence $\rho_{x, a, t}^\text{ANC}$ and ART coverage $\alpha_{x, a, t}^\text{ANC}$ among pregnant women modelled with logit-linear models
 \begin{align}
 \text{logit}(\rho_{x, a, t}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a, t}) + \beta^{\rho^{\text{ANC}}} + \beta_{t \in(T_2, T_3\}}^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + u_{x, t \in\{T_2, T_3\}}^{\rho^{\text{ANC}}} + \eta_{R_x, a, t}^{\rho^{\text{ANC}}} \\
-\text{logit}(\alpha_{x, a, t}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a, t}) + \beta^{\alpha^{\text{ANC}}} + \beta_{t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}}+u_{x, t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + \eta_{R_x, a, t}^{\alpha^{\text{ANC}}}
+\text{logit}(\alpha_{x, a, t}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a, t}) + \beta^{\alpha^{\text{ANC}}} + \beta_{t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}}+u_{x, t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + \eta_{R_x, a, t}^{\alpha^{\text{ANC}}},
 \end{align}

### 2022-12-21 src/docs_paper/appendix.Rmd
Complete try at scrubing out T2 and T3 in appendix, and set stage for writing model summary in main text

@@
 
-## Indexing
+### Indexing
 
-Let $x$ be a district, located within the Spectrum region $R_x$, $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
-Consider the time period $t = T_1$ of the most recent national household survey with HIV testing specified to the nearest calendar quarter.
-We index population size $N_{x, s, a, t}$, HIV prevalence $\rho_{x, s, a, t}$, ART coverage $\alpha_{x, s, a, t}$ and annual HIV incidence rate $\lambda_{x, s, a, t}$ by district, sex, age-band and time period.
+Consider some country, and let $x$ refer to a district located within the Spectrum region $R_x$.
+We restrict our attention to the time period $t = T_1$ of the most recent national household survey with HIV testing.
+Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
+As short-hand, we write $a = l$ to refer to the age band with lower bound $l$.
+We index the following quantities by district, sex, age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$ and annual HIV incidence rate $\lambda_{x, s, a}$.
+Our model operates at this level of granularity, although sometimes the data observed is at an aggregate level.
+We use $\{\cdot\}$ to refer to a generic aggregate set over which an observation is made, such as $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.
 
-## Structured random effects
+### Structured random effects
 
-We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$, and $u \sim \text{BYM}2(\sigma, \phi)$ to the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.
-For both the ICAR and BYM2 models, we follow recommendations [@freni2018note] on scaling, disconnected adjacency graph components, and islands.
+Consider some generic random effect $u$.
+We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with marginal standard deviation $\sigma > 0$.
+For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] with marginal standard deviation $\sigma > 0$ we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$
+We follow recommendations of @freni2018note on scaling, disconnected adjacency graph components, and islands.
+The first order auto-regressive model is specified by $u \sim \text{AR}1(\sigma, \phi)$.
 
@@
 $$
-\text{logit}(\rho_{x, s, a, T_1}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a}
+\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
 $$
-where $\beta^\rho_0$ is the intercept, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are a fixed offset specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
-We use the prior distributions $\mathcal{N}(0, 5)$ for all fixed effects ($\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}}$), $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.
+where $\beta^\rho_0$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
+We use the prior distributions $\mathcal{N}(0, 5)$ for the fixed effects $\beta^\rho_0$ and $\beta_{s}^{\rho, s = \text{M}}$, $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.
 
@@
 $$
-\text{logit}(\alpha_{x, s, a, T_1}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a}
+\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a}
 $$
-with analogous terms to the HIV prevalence model.
+with terms and priors analogous to the HIV prevalence model above.
 
@@
 $$
-\log(\lambda_{x, s, a, t}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x, t}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x, t}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a, t}^\lambda
+\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda 
 $$
-where $\beta^\lambda_0$ is the intercept, which is proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women, $\rho_{x, t}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
+where $\beta^\lambda_0$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women, $\rho_{x}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
 $$
-\rho_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t}},
+\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}},
 $$
-$\alpha_{x, t}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
+$\alpha_{x}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
 $$
-\alpha_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}},
+\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}},
 $$
-$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ from the EPP model, $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a, t}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output
+$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ via the EPP model, $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output
 
@@
 
-For women 15-49 the predicted number of ANC clients $\Psi_{x, a, t}$ is a log-linear model
+For women 15-49 the predicted number of ANC clients $\Psi_{x, a}$ is a log-linear model
 $$
-\log (\Psi_{x, a, t}) = \log (N_{x, \text{F}, a, t}) + \psi_{R_x, a, t} + \beta^\psi + u_x^\psi
+\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
 $$
-where $N_{x, \text{F}, a, t}$ are the female population sizes, $\psi_{R_x, a, t}$ are ASFR in Spectrum region $R_x$ at time $t$, $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility, $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects, HIV prevalence $\rho_{x, a, t}^\text{ANC}$ and ART coverage $\alpha_{x, a, t}^\text{ANC}$ among pregnant women modelled with logit-linear models
+where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are ASFR in Spectrum region $R_x$ at time $t$, $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility, $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
+HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women modelled with logit-linear models
 \begin{align}
-\text{logit}(\rho_{x, a, t}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a, t}) + \beta^{\rho^{\text{ANC}}} + \beta_{t \in(T_2, T_3\}}^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + u_{x, t \in\{T_2, T_3\}}^{\rho^{\text{ANC}}} + \eta_{R_x, a, t}^{\rho^{\text{ANC}}} \\
-\text{logit}(\alpha_{x, a, t}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a, t}) + \beta^{\alpha^{\text{ANC}}} + \beta_{t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}}+u_{x, t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + \eta_{R_x, a, t}^{\alpha^{\text{ANC}}},
+\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}} \\
+\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}},
 \end{align}
-  * $\eta_{R_x, a, t}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$
-  * $\beta^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the average differences between population and ANC outcomes at $T_1$ after removing the offset
-  * $\beta_{t \in(T_2, T_3\}}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the change in average difference from $T_1$ to $T_2$
-  * $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ are district random effects
-  * $u_{x, t \in\{T_2, T_3\}}^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_{XT}^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ capture the change in district random effects between times $T_1$ and $T_2$
+
+where $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$, $\beta^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the average differences between population and ANC outcomes after removing the offset, and $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ are district random effects.
 
@@
 
-* Number of PLHIV on ART is $A_{x, s, a, t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}$
-* Let $\gamma_{x, x', t} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$ at time $t$
-* We assume that $\gamma_{x, x', t} = 0$ for $x \notin \{x, \text{ne}(x)\}$ i.e.
-that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$ where $\sim$ is an adjacency relation
-  * $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x', t} = 1$
-* Model $\gamma_{x, x', t}$ with a multinomial logit model where $\tilde \gamma_{x, t}$ is^[Confused by this notation.] the log odds ratio of seeking ART in the district $x'$ neighbouring $x$
+Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$.
+We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.
+
+To model $\gamma_{x, x'}$ for $x \sim x'$ we use a multinomial logistic regression model where $\tilde \gamma_{x, x'}$ is the log odds ratio of seeking ART
 \begin{align}
-\tilde \gamma_{x, t} &= \tilde \gamma_0 + u_x^{\tilde \gamma} + u_x^{\tilde \gamma, t = T_2} \\ 
+\tilde \gamma_{x, x'} &= \tilde \gamma_0 + u_x^{\tilde \gamma} \\ 
 \tilde \gamma_0 &= -4 \\
-u_x^{\tilde \gamma} &\sim \mathcal{N}(0, \sigma_X^{\tilde \gamma}) \\
-u_x^{\tilde \gamma, t \in \{T_2, T_3\}} &\sim \mathcal{N}(0, \sigma_{XT}^{\tilde \gamma}) 
+u_x^{\tilde \gamma} &\sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})
 \end{align}
-  * $\tilde \gamma_{x, t}$ is an intercept specifying the prior mean for the log odds of seeking ART in each neighbouring district compared to the home district
-  * $u_x^{\tilde \gamma}$ are district random effects
-  * $u_x^{\tilde \gamma, t \in \{T_2, T_3\}}$ is^[Shouldn't this just be $t = T_2$ here rather than $t \in \{T_2, T_3\}$?] the change in district level log-odds ratios between $T_1$ and $T_2$
-  * Note: only one value estimated for each district, implying no preference between neighbouring districts, and no sex-age specific modelling
-* Calculate multinomial probabilities via softmax (n.b. this is the sigmoid function, the inverse of the softmax)
+where $\tilde \gamma_0$ is an intercept specifying the prior mean for the log odds of seeking ART in each neighbouring district compared to the home district, and $u_x^{\tilde \gamma}$ are district random effects.
+We fix $\tilde \gamma_{x, x} = 0$ and recover the multinomial probabilities using the softmax
 $$
-\gamma_{x, x', t} = \frac{1}{1 + \sum_{x^\star \sim x} \exp(\tilde \gamma_{x^\star, t})}
+\gamma_{x, x'} = \frac{\exp(\tilde \gamma_{x, x'})}{\sum_{x^\star \in \{x, \text{ne}(x)\}} \exp(\tilde \gamma_{x, x^\star})}
 $$
-* The number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a, t} = A_{x, s, a, t} \cdot \gamma_{x, x', t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t} \cdot \gamma_{x, x', t}$
-* Total attending ART facilities in district $x'$ is $\tilde A_{x', s, a, t} = \sum_{x \sim x', x = x'} A_{x, x', s, a, t}$
+Then, given the number of PLHIV on ART $A_{x, s, a} = N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}$, the number of ART clients who reside in district $x$ and obtain ART in district $x'$ are
+$$
+A_{x, x', s, a} = A_{x, s, a} \cdot \gamma_{x, x'},
+$$
+and the total attending ART facilities in district $x'$ are
+$$
+\tilde A_{x', s, a} = \sum_{x \in \{x', \text{ne}(x')\}} A_{x, x', s, a}.
+$$
 
@@
 
-* The proportion of HIV positive adults aware of their HIV status is $v_{x, s, a, t}$
-* The proportion of untreated HIV positive adults aware of their HIV status is $\tilde \nu_{x, s, a, t}$ such that
+The proportion of HIV positive adults aware of their HIV status is $v_{x, s, a}$, and the proportion of untreated HIV positive adults aware of their HIV status is $\tilde \nu_{x, s, a}$ such that
 $$
-v_{x, s, a, t} = \alpha_{x, s, a, t} + \frac{\tilde v_{R_x, s, a, t}}{1 - \alpha_{x, s, a, t}}
+v_{x, s, a} = \alpha_{x, s, a} + \frac{\tilde v_{R_x, s, a}}{1 - \alpha_{x, s, a}}.
 $$
@@
 
-### Notation
-
-* Use $\{x\}$ to denote a collection of districts, $\{s\}$ a collection of sexes and $\{a\}$ a collection of ages for which an outcome is observed
-
 ### Household survey data
 
-* Let $\nu$ be a household survey occuring at $T_1$
-* Furnishes weighted observations for HIV prevalence $\hat \rho_{\{x\}, \{s\}, \{a\}, \nu}$, ART coverage $\hat \alpha_{\{x\}, \{s\}, \{a\}, \nu}$ and proportion recently infected $\hat \kappa_{\{x\}, \{s\}, \{a\}, \nu}$ with respective Kish effective sample sizes $M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu}$ for $\theta \in \{\rho, \alpha, \kappa\}$
-* The observed cases are $Y^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} = M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}, \nu}$
-* For HIV prevalence we use a binomial working likelihood
+Let $\nu$ be a household survey occurring at $T_1$, furnishing weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}, \nu}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes $M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu}$, and observed number of cases
+$$
+Y^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} = M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}, \nu}.
+$$
+
+For HIV prevalence and ART coverage we use binomial working likelihoods as follows
 \begin{align}
-Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
-\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t}}
+Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
 \end{align}
-* For ART coverage we also use a binomial working likelihood
+and
 \begin{align}
-Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
-\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}
+Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
 \end{align}
-* For recent infections we also use a binomial working likelihood
+
+For recent infections we also use a binomial working likelihood
 \begin{align}
-Y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu}, \kappa_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
-\kappa_{x, s, a, t} &= 1 - \exp(- \lambda_{x, s, a, t} \cdot \frac{1 - \rho_{x, s, a, t}}{\rho_{x, s, a, t}} \cdot (\Omega_T - \beta_T ) - \beta_T ) \\
+Y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu}, \kappa_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+\kappa_{x, s, a} &= 1 - \exp(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T ) \\
 \Omega_T &\sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T}) \\
@@
 \end{align}
-  * $\kappa_{x, s, a, t}$ are the predicted proportion recently infected among HIV positive persons
-  * $\Omega_T$ is the mean duration of recent infection (MDRI)
-    * Use an informative prior based on the characteristics of the recent infection testing algorithm (RITA)
-    * For PHIA surveys^[What about other surveys?] $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$
-  * $\beta_T$ is the false recent ratio (FRR)
-    * For PHIA surveys $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.^[Is this just assuming that there is no false recency? How is this assumption justified?]
+where $\kappa_{x, s, a}$ are the predicted proportion recently infected among HIV positive persons, $\Omega_T$ is the mean duration of recent infection (MDRI), and $\beta_T$ is the false recent ratio (FRR).
+We use an informative prior on $\Omega_T$ based on the characteristics of the recent infection testing algorithm (RITA).
+For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$.
+For PHIA surveys we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.
 
@@
 
-* Include two years of ANC testing data: the year of the most recent survey $Y[T_1]$ and the current year $Y[T_2]$
-* $W^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients
-* $X^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with ascertained status
-* $Y^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with positive status (either known or tested)
-* $Z^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients already on ART prior to first ANC
-* Sometimes ANC testing data are only available for part of a given year, for example only the first three quarters.
-Denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$
-* Likelihood for number of ANC clients specified only^[Why?] for year $Y[T_2]$ as
-$$
-W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right)
-$$
-* Observed number of HIV positive and already on ART among ANC clients at both $Y[T_1]$ and $Y[T_2]$ are modelled by
+We include ANC testing data for the year of the most recent survey $Y[T_1]$.
+Let $W^\text{ANC}_{\{x\}, Y[t]}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}, Y[t]}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}, Y[t]}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}, Y[t]}$ the number of ANC clients already on ART prior to first ANC, such that $W^\text{ANC}_{x, Y[t]} \geq X^\text{ANC}_{x, Y[t]} \geq Y^\text{ANC}_{x, Y[t]} \geq Z^\text{ANC}_{x, Y[t]}$
+When ANC testing data are only available for part of a given year, we denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$.
+
+<!-- * Likelihood for number of ANC clients specified only^[Why?] for year $Y[T_2]$ as -->
+<!-- $$ -->
+<!-- W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right) -->
+<!-- $$ -->
+
+The observed number of HIV positive and already on ART among ANC clients at $Y[T_1]$ is modelled by
 \begin{align*}
-& Y_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[t]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right) \\
-& Z_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[t]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right)
+Y_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[T_1]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}, t}^{\text{ANC}}\right) \\
+Z_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[T_1]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}, t}^{\text{ANC}}\right)
 \end{align*}
-where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a, t}$
+where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a}$
 \begin{align*}
-\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t}} \\
-\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}} \cdot \alpha_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}
+\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a}} \\
+\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}
 \end{align*}
@@
 
-* Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART^[I'm confused about the notation here.
-In general would be good to define precisely what is meant by $\{x\}$.
-Presumably the LHS below is a set, yet the RHS is a sum over the sets so I think just a number -- so how can the two be equal?]
+Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART
 $$
-\dot A_{\{x\}, \{s\}, \{a\}, t} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a, t}
+\dot A_{\{x\}, \{s\}, \{a\}, t} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a},
 $$
-* Model the unobserved numbers of ART clients travelling from $x'$ to $x$ as
+We model the unobserved numbers of ART clients travelling from $x'$ to $x$ as
 $$
-\dot A_{x', x, s, a, t} \sim \text{Bin}(N_{x', s, a, t}, \pi_{x', x, s, a, t})
+\dot A_{x', x, s, a} \sim \text{Bin}(N_{x', s, a}, \pi_{x', x, s, a})
 $$
-where $\pi_{x', x, s, a, t} = \rho_{x', s, a, t} \cdot \alpha_{x', s, a, t} \cdot \gamma_{x', s, a, t}$^[Should this be $\gamma_{x', x, s, a, t}$ since it's people moving from $x'$ to $x$ (and so far no where in the RHS is there an $x$ which is suspicious)]
-* This likelihood is approximated using a normal for the sum of binomials
+where $\pi_{x', x, s, a} = \rho_{x', s, a} \cdot \alpha_{x', s, a} \cdot \gamma_{x', x, s, a}$.
+This likelihood is approximated using a normal for the sum of binomials by
 $$
@@
 where
-  * $\tilde{A}_{\{x\},\{s\}\{a\}, t} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a, t} \cdot \pi_{x', x, s, a, t}$
-  * $\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a, t} \cdot \pi_{x', x, s, a, t} \cdot \left(1-\pi_{x', x, s, a, t}\right)}$
+$$
+\tilde{A}_{\{x\},\{s\}\{a\}, t} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a},
+$$
+and
+$$
+\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a} \cdot \left(1-\pi_{x', x, s, a}\right)}.
+$$
 
@@
 
-* Depends on data availability. If data is missing, some parameters might be fixed to default values
-* Survey data on ART coverage by age and sex not available then set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age/sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$
-* If there is no ART data at $T_1$ (either survey or ART programme) or no ART data at $T_2$ then change in ART coverage is unidentifiable and set $\beta_{T_2}^\alpha = \beta_{T_2, s = \text{M}}^\alpha = u_{x, {T_2}}^\alpha = u_{x, a < 15, {T_2}}^\alpha = 0$ such that change in ART coverage is modelled by
+If data are missing, some parameters are fixed to default values to help with identifiability.
+In particular:
+
+* If survey data on ART coverage by age and sex are not available then we set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age/sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$.
+* If no ART data (survey or ART programme) are available at $T_1$ but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
+In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated
 $$
-\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \eta_{R_x, s, a, {T_2}}^\alpha
+\text{logit} \left(\alpha_{x, s, a, T_1} \right) = u_x^\alpha+\eta_{R_x, s, a}^\alpha \\
 $$
-* If no ART data (survey or ART programme) are available at $T_1$ or $T_2$ but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
-In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated
-\begin{align*}
-\text{logit} \left(\alpha_{x, s, a, T_1} \right) &= u_x^\alpha+\eta_{R_x, s, a}^\alpha \\
-\text{logit} \left(\alpha_{x, s, a, T_2} \right) &= \text{logit}\left(\alpha_{x, s, a, T_1}\right)+u_{x, T_2}^\alpha+\eta_{R_{x, s, a, T_2}}^\alpha
-\end{align*}
 * If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$.
-The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage
+The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.
+
+\newpage
 

### 2022-12-21 src/docs_paper/paper.Rmd
Complete try at scrubing out T2 and T3 in appendix, and set stage for writing model summary in main text

@@
 
-@eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-This model forms the basis of the Naomi small-area estimation model, described fully in @eaton2021naomi. 
+@eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
 Modelling data from multiple sources concurrently is attractive as it increases statistical power, mitigates the biases of any single source, and prompts investigation into any data conflicts.
-The model is comprised of three components, described as follows.
 
-## Household survey component
+In this work, we consider a modified version of the model, with temporal projection omitted.
+In this section, we provide an overview of the model structure, with full details provided in the appendix.
 
-Consider a country partitioned into $n$ areas indexed by $i$.
-Suppose a simple random household survey of $m^\text{HS}_i$ people is conducted in each area, and $y^\text{HS}_i$ HIV positive cases are observed.
-Then, cases may be modelled using a binomial logistic regression model
-\begin{align*}
-y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
-\text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2),
-\end{align*}
-where $\rho^\text{HS}_i$ is the HIV prevalence, modelled on the logit scale by a Gaussian with mean $\beta_\phi \sim p(\beta_\phi)$ and standard deviation $\sigma_\phi \sim p(\sigma_\phi)$.
+## Process model
 
-## ANC component
+To write.
 
-Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
-Suppose that of $m^\text{ANC}_i$ women attending ANC, $y^\text{ANC}_i$ are HIV positive.
-Then an analogous binomial logistic regression model
-\begin{align*}
-y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
-\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho^\text{HS}_i) + b_i, \\
-b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
-\end{align*}
-may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
-Reflecting the fact that prevalence in ANCs is related, but importantly different, to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence on the logit scale.
-We use a Gaussian distribution with mean $\beta_b \sim p(\beta_b)$ and standard deviation $\sigma_b \sim p(\sigma_b)$ for these bias random effects.
+## Likelihood
 
-## ART component
-
-The number of people receiving treatment at district health facilities $A_i$ provides additional information about HIV prevalence.
-Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa districts with low prevalence are likely to have fewer people recieving treatment.
-ART coverage, defined to be the proportion of people living with HIV (PLHIV) currently on ART on district $i$, is given by $\alpha_i = A_i / \rho^\text{HS}_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be constant.
-As such, ART coverage may also be modelled using a binomial logistic regression model
-\begin{align*}
-A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
-\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2),
-\end{align*}
-where the proportion of people receiving ART is $\rho^\text{HS}_i \alpha_i$.
-We again use a Gaussian distribution with mean $\beta_\alpha \sim p(\beta_\alpha)$ and standard deviation $\sigma_\alpha \sim p(\sigma_\alpha)$.
-Here we assume no travel between districts to receive treatment.
-
-## Joint model
-
-The three components above may be combined to a single model as follows
-\begin{align*}
-y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
-y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
-A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
-\text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2), \\
-\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho^\text{HS}_i) + b_i, \\
-b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2), \\
-\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
-\end{align*}
+To write.
 

### 2022-12-21 src/docs_paper/paper.Rmd
Add naomi.cpp to appendix. Start writing main text model description

@@
 The remainder of this paper is organised as follows.
-In Section \ref{sec:naomi} we describe the Naomi model.
+In Section \ref{sec:naomi} we describe a modified version of the Naomi model that we consider.
 Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
-As a case-study, we fit the Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
+As a case-study, we fit the modified Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
 In this section, we also demonstrating a Bayesian workflow.
@@
 
-# The Naomi model\label{sec:naomi}
+# A modified Naomi model\label{sec:naomi}
 
 @eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-Modelling data from multiple sources concurrently is attractive as it increases statistical power, mitigates the biases of any single source, and prompts investigation into any data conflicts.
+Modelling data from multiple sources concurrently is beneficial as it mitigates the limitations of any single source, increases statistical power and prompts investigation into data conflicts.
 
-In this work, we consider a modified version of the model, with temporal projection omitted.
-In this section, we provide an overview of the model structure, with full details provided in the appendix.
+For each country, the full model is defined over three time points: $T_1$, the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
+We consider a simplified version of the full model which is defined only at $T_1$ and omits temporal projection.
+In this section, we provide an overview of the modified model, highlighting the aspects which make it a challenge for existing inferential approaches.
+Full mathematical details of this modified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
 
@@
 
-To write.
+Let $i$ index the coarsest space-age-sex division included in the model.
+Let $N_i$ be the population size and consider the following indicators: HIV prevalence $\rho_i$, annual HIV incidence rate $\lambda_i$ and ART coverage $\alpha_i$.
 

### 2022-12-21 src/docs_paper/appendix.Rmd
Add naomi.cpp to appendix. Start writing main text model description
A few appendix fixes and extra into main text

@@
 
-#### HIV incidence rate
+### HIV incidence rate
 
@@
 Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
-\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
+\rho_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
 \end{align}
@@
 Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
-\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
+\alpha_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
 \end{align}
@@
 
-<!-- * Likelihood for number of ANC clients specified only^[Why?] for year $Y[T_2]$ as -->
-<!-- $$ -->
-<!-- W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right) -->
-<!-- $$ -->
-
 The observed number of HIV positive and already on ART among ANC clients at $Y[T_1]$ is modelled by
 \begin{align*}
-Y_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[T_1]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}, t}^{\text{ANC}}\right) \\
-Z_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[T_1]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}, t}^{\text{ANC}}\right)
+Y_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[T_1]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right) \\
+Z_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[T_1]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right)
 \end{align*}
@@
 \begin{align*}
-\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a}} \\
-\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}
+\rho_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a}} \\
+\alpha_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}
 \end{align*}
@@
 
-Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART
+Let $\dot A_{\{x\}, \{s\}, \{a\}}$ be data for the number receiving ART
 $$
-\dot A_{\{x\}, \{s\}, \{a\}, t} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a},
+\dot A_{\{x\}, \{s\}, \{a\}} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a},
 $$
@@
 $$
-\dot A_{\{x\}, \{s\}, \{a\}, t} \sim \mathcal{N}(\tilde) A_{\{x\}, \{s\}, \{a\}, t}, \sigma^{\tilde A}_{\{x\}, \{s\}, \{a\}, t})
+\dot A_{\{x\}, \{s\}, \{a\}} \sim \mathcal{N}(\tilde A_{\{x\}, \{s\}, \{a\}}, \sigma^{\tilde A}_{\{x\}, \{s\}, \{a\}})
 $$
@@
 $$
-\tilde{A}_{\{x\},\{s\}\{a\}, t} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a},
+\tilde{A}_{\{x\},\{s\}\{a\}} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a},
 $$
@@
 $$
-\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a} \cdot \left(1-\pi_{x', x, s, a}\right)}.
+\sigma_{\{x\},\{s\}\{a\}}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a} \cdot \left(1-\pi_{x', x, s, a}\right)}.
 $$
@@
 $$
-\text{logit} \left(\alpha_{x, s, a, T_1} \right) = u_x^\alpha+\eta_{R_x, s, a}^\alpha \\
+\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha \\
 $$
@@
 
+# \textsc{C++} template
+
+```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi.cpp")}
+
+```
+
+\newpage
+
 # References {#references .unnumbered}

### 2022-12-21 src/docs_paper/paper.Rmd
A few appendix fixes and extra into main text

@@
 
-Let $i$ index the coarsest space-age-sex division included in the model.
-Let $N_i$ be the population size and consider the following indicators: HIV prevalence $\rho_i$, annual HIV incidence rate $\lambda_i$ and ART coverage $\alpha_i$.
+Let $i$ index the coarsest district-age-sex division included in the model.
+Let $N_i$ be the population size and consider the following three indicators: HIV prevalence $\rho_i \in [0, 1]$, ART coverage $\alpha_i \in [0, 1]$, and annual HIV incidence rate $\lambda_i > 0$.
+
+We specify independent mixed effects models for HIV prevalence and ART coverage on the logit scale such that
+\begin{align}
+\text{logit}(\rho_i) &= \eta^\rho_i, \\
+\text{logit}(\alpha_i) &= \eta^\alpha_i,
+\end{align}
+for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
+For the HIV incidence rate we use a mixed effects model on the log scale where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$
+$$
+\log(\lambda_i) = \eta^\rho_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}).
+$$
 
@@
 
-To write.
+Consider a household survey occurring at $T_1$.
+Let $\mathcal{I}$ be a set of $i$ for which observations are reported.
+For each $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes $M^{\hat \theta}_\mathcal{I}$, and observed number of cases
+$$
+Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
+$$
 

### 2023-01-03 src/docs_paper/appendix.Rmd
Jeff's suggested edits to appendix

@@
 
-### ANC testing cascade
+### ANC testing
 
-For women 15-49 the predicted number of ANC clients $\Psi_{x, a}$ is a log-linear model
-$$
-\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
-$$
-where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are ASFR in Spectrum region $R_x$ at time $t$, $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility, $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
 HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women modelled with logit-linear models
@@
 
-### ANC attendance
+### ART attendance
 
@@
 \tilde A_{x', s, a} = \sum_{x \in \{x', \text{ne}(x')\}} A_{x, x', s, a}.
-$$
-
-### Awareness of HIV status
-
-The proportion of HIV positive adults aware of their HIV status is $v_{x, s, a}$, and the proportion of untreated HIV positive adults aware of their HIV status is $\tilde \nu_{x, s, a}$ such that
-$$
-v_{x, s, a} = \alpha_{x, s, a} + \frac{\tilde v_{R_x, s, a}}{1 - \alpha_{x, s, a}}.
 $$

### 2023-01-06 src/docs_paper/appendix.Rmd
Work on moving model into main paper and improving appendix description

@@
 
-Consider some country, and let $x$ refer to a district located within the Spectrum region $R_x$.
+Consider a given country in sub-Saharan Africa.
+Let $x$ refer to a district located within the Spectrum region $R_x$.
 We restrict our attention to the time period $t = T_1$ of the most recent national household survey with HIV testing.
 Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
-As short-hand, we write $a = l$ to refer to the age band with lower bound $l$.
-We index the following quantities by district, sex, age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$ and annual HIV incidence rate $\lambda_{x, s, a}$.
-Our model operates at this level of granularity, although sometimes the data observed is at an aggregate level.
-We use $\{\cdot\}$ to refer to a generic aggregate set over which an observation is made, such as $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.
+As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
+We index the following quantities by district, sex and age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$, annual HIV incidence rate $\lambda_{x, s, a}$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a}$.
+Sometimes data are observed at an aggregate level, rather then the level of granularity at which the model operates.
+In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.
 
@@
 
-Consider some generic random effect $u$.
-We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with marginal standard deviation $\sigma > 0$.
-For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] with marginal standard deviation $\sigma > 0$ we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$
-We follow recommendations of @freni2018note on scaling, disconnected adjacency graph components, and islands.
-The first order auto-regressive model is specified by $u \sim \text{AR}1(\sigma, \phi)$.
+Consider a generic random effect $u$.
+We use use structured random effects to share information across units assessed as being similar, such as districts next to each other or adjacent age-bands.
+We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
+$$
+u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
+$$
+where $u_{-i}$ is $u$ with the $i$th unit removed, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
+We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands regarding the ICAR model.
+For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v^\star$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w^\star$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
+$$
+u = \sigma \left( \sqrt\phi \cdot v^\star + \sqrt{1 - \phi} \cdot w^\star \right).
+$$
+We specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
+\begin{align*}
+u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
+u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
+\end{align*}
+where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, $|\rho| < 1$ is the lag-one correlation parameter.
 
@@
 
-We model HIV prevalence by
+We model HIV prevalence on the logit scale using
 $$
@@
 $$
-where $\beta^\rho_0$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
+where $\beta^\rho_0$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{UNKNOWN}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
 We use the prior distributions $\mathcal{N}(0, 5)$ for the fixed effects $\beta^\rho_0$ and $\beta_{s}^{\rho, s = \text{M}}$, $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.
@@
 
-Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$.
+Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
 We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.

### 2023-01-06 src/docs_paper/paper.Rmd
Work on moving model into main paper and improving appendix description

--- 2022-12-21 src/docs_paper/paper.Rmd
@@
 
+Let $x$ index district, $a$ index five-year age band, $s$ index sex and $t$ index time.
 For each country, the full model is defined over three time points: $T_1$, the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version of the full model which is defined only at $T_1$ and omits temporal projection.
+We consider a simplified version defined only at $T_1$, omiting all temporal projection.
 In this section, we provide an overview of the modified model, highlighting the aspects which make it a challenge for existing inferential approaches.
@@
 
-## Process model
+## Household survey component
 
-Let $i$ index the coarsest space-age-sex division included in the model.
-Let $N_i$ be the population size and consider the following indicators: HIV prevalence $\rho_i$, annual HIV incidence rate $\lambda_i$ and ART coverage $\alpha_i$.
+Consider a household survey occurring at $T_1$, and let $i$ index the coarsest district-age-sex division included in the model.
+The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
+Let $N_i$ be the population size, and consider the following three indicators: HIV prevalence $\rho_i \in [0, 1]$, ART coverage $\alpha_i \in [0, 1]$, and annual HIV incidence rate $\lambda_i > 0$.
+We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
+\begin{align*}
+\text{logit}(\rho_i) &= \eta^\rho_i, \\
+\text{logit}(\alpha_i) &= \eta^\alpha_i,
+\end{align*}
+for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
+For the HIV incidence rate we use a mixed effects model on the log scale
+$$
+\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}),
+$$
+where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$.
 
-## Likelihood
+Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
+For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes
+$$
+M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
+$$
+where $j$ index individuals in all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
+The observed number of indicator cases is then
+$$
+Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
+$$
+For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood
+$$
+Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}),
+$$
+where $\theta_{\mathcal{I}}$ are the following weighted aggregates
+\begin{align*}
+\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
+\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
+\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
+\end{align*}
+We link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ by
+$$
+\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
+$$
+where $\Omega_T$ is the mean duration of recent infection and $\beta_T$ is the false recent ratio.
 
-To write.
+## ANC testing component
+
+We model HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women as being offset on the logit scale from the general population indicator as follows
+\begin{align*}
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \tilde{\eta}^{\rho^\text{ANC}}_i, \\
+\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \tilde{\eta}^{\rho^\text{ANC}}_i.
+\end{align*}
+Process section to be written.
+
+Likelihood section to be written.
+
+## ART attendance component
+
+Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
+Process section to be written.
+
+Let $\dot A_i$ be the number of people receiving ART.
+Likelihood section to be written.
 

### 2023-01-06 src/docs_paper/paper.Rmd
Edit main paper introduction

@@
 
-Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at a sufficiently fine-scale resolution.
-Producing these estimates is a challenging task, as all available data sources have shortcomings.
-Nationally-representative household surveys are the most statistically reliable data source, but due to their high cost to run, in most countries they only occur infrequently.
-Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on a biased sample of the population.
+Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at a sufficiently fine-scale resolution to make targetted interventions.
+Producing these estimates is a challenging task, as all available data sources have shortcomings which must be overcome.
+Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost to run, in most countries they only occur infrequently.
+Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on limited or biased samples of the population.
 To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
-Software (\url{https://naomi.unaids.org}) has been developed for Naomi, which allows countries to input their data and generate estimates in a yearly process supported by UNAIDS.
+Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and generate estimates in a yearly process supported by UNAIDS.
 
-The complexity of the model presents a difficult Bayesian inference problem.
+Situational reqruiements combined with the complexity of the model present a difficult Bayesian inference problem.
 Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
 Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-Owing to its speed and flexibility, `TMB` is gaining popularity spatial statistics [@osgoodzimmerman2021statistical].
+Owing to its speed and flexibility, `TMB` has recently been gaining popularity spatial statistics [@osgoodzimmerman2021statistical].
 Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
@@
 Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
+As such, we have been motivated to find a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.
 
@@
 We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
-As `aghq` is designed to naturally interface with `TMB`, use of the method is easy when provided a \textsf{C++} user template for the log-posterior.
+`aghq` is designed to naturally interface with `TMB`, so use of our method is simple when provided a \textsf{C++} user template for the log-posterior.
 
 The remainder of this paper is organised as follows.
-In Section \ref{sec:naomi} we describe a modified version of the Naomi model that we consider.
+In Section \ref{sec:naomi} we describe the modified version of the Naomi model that we consider.
 Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
 As a case-study, we fit the modified Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
-In this section, we also demonstrating a Bayesian workflow.
+In this section, we also demonstrate a Bayesian workflow, illustrating that these concepts are not only applicable in a simulation-based inference setting.
 Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how our method might be used in other models, and directions for future research.
@@
 @eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-Modelling data from multiple sources concurrently is beneficial as it mitigates the limitations of any single source, increases statistical power and prompts investigation into data conflicts.
+Modelling data from multiple sources concurrently in this way mitigates the limitations of any single data source, increases statistical power and prompts investigation into data conflicts.
 
-Let $x$ index district, $a$ index five-year age band, $s$ index sex and $t$ index time.
-For each country, the full model is defined over three time points: $T_1$, the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
+Consider a country in sub-Saharan Africa, and let $x$ index district, $a$ index five-year age band, $s$ index sex and $t$ index time.
+The full model is defined over three time points: $T_1$, the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
 We consider a simplified version defined only at $T_1$, omiting all temporal projection.

### 2023-01-08 src/docs_paper/appendix.Rmd
Paper and appendix iteration

@@
 We index the following quantities by district, sex and age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$, annual HIV incidence rate $\lambda_{x, s, a}$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a}$.
-Sometimes data are observed at an aggregate level, rather then the level of granularity at which the model operates.
+Sometimes data are observed at an aggregate level, rather than the modelled level of granularity.
 In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.
@@
 Consider a generic random effect $u$.
-We use use structured random effects to share information across units assessed as being similar, such as districts next to each other or adjacent age-bands.
+We use use structured random effects to share information across units assessed as being similar, such as neighbouring districts or adjacent age-bands.
 We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
@@
 
+For women 15-49 the predicted number of ANC clients $\Psi_{x, a}$ is a log-linear model
+$$
+\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
+$$
+where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are ASFR in Spectrum region $R_x$ at time $t$, $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility, $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
+
 We include ANC testing data for the year of the most recent survey $Y[T_1]$.

### 2023-01-08 src/docs_paper/paper.Rmd
Paper and appendix iteration

@@
 To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
+Concurrent modelling of data mitigates the limitations of any single data source, increases statistical power and prompts investigation into any conflicting information.
 Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and generate estimates in a yearly process supported by UNAIDS.
 
-Situational reqruiements combined with the complexity of the model present a difficult Bayesian inference problem.
+Situational requirements combined with the complexity of the model present a difficult Bayesian inference problem.
 Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
@@
 The remainder of this paper is organised as follows.
-In Section \ref{sec:naomi} we describe the modified version of the Naomi model that we consider.
+In Section \ref{sec:naomi} we describe a simplified version of the Naomi model that we consider in this paper.
 Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
-As a case-study, we fit the modified Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
+As a case-study, we fit the simplified Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
 In this section, we also demonstrate a Bayesian workflow, illustrating that these concepts are not only applicable in a simulation-based inference setting.
@@
 
-# A modified Naomi model\label{sec:naomi}
+# A simplified Naomi model\label{sec:naomi}
 
 @eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-Modelling data from multiple sources concurrently in this way mitigates the limitations of any single data source, increases statistical power and prompts investigation into data conflicts.
-
-Consider a country in sub-Saharan Africa, and let $x$ index district, $a$ index five-year age band, $s$ index sex and $t$ index time.
-The full model is defined over three time points: $T_1$, the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version defined only at $T_1$, omiting all temporal projection.
-In this section, we provide an overview of the modified model, highlighting the aspects which make it a challenge for existing inferential approaches.
-Full mathematical details of this modified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
+The model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
+We consider a simplified version of the model which is defined only at $T_1$, omitting temporal projection to $T_2$ and $T_3$.
+Below we provide an overview of the three components in the simplified model, highlighting aspects which make it a challenge for existing inferential approaches.
+More complete mathematical details of this simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
 
@@
 
-Consider a household survey occurring at $T_1$, and let $i$ index the coarsest district-age-sex division included in the model.
+Consider a country in sub-Saharan Africa where a household survey has taken place at $T_1$.
+Let $x$ index district, $a$ index five-year age band, $s$ index sex and $t$ index time.
+For ease of notation, let $i$ index the coarsest district-age-sex division included in the model.
 The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
 Let $N_i$ be the population size, and consider the following three indicators: HIV prevalence $\rho_i \in [0, 1]$, ART coverage $\alpha_i \in [0, 1]$, and annual HIV incidence rate $\lambda_i > 0$.
+
 We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
@@
 \end{align*}
-Process section to be written.
 
-Likelihood section to be written.
+Let $W^\text{ANC}_i$ be the number of ANC clients, $X^\text{ANC}_i$ the number of those with ascertained status, $Y^\text{ANC}_i$ the number of those with positive status, and $Z^\text{ANC}_i$ the number of ANC clients already on ART prior to their first ANC visit.
+ANC data from the year of the most recent survey is included in the model.
+
+For women 15-49 the predicted number of ANC clients $\Psi_i$ is a log-linear model
+$$
+\log (\Psi_i) = \log (N_i) + \tilde \eta^\Psi
+$$
 
@@
 Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
-Process section to be written.
 
@@
 
-# Fast approximate inference method\label{sec:fastinferencemethods}
+# Fast approximate inference methods\label{sec:fastinferencemethods}
 
@@
 
-## Forwards algorithm\label{sec:algorithm}
+## Algorithm\label{sec:algorithm}
 
@@
 
-1. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(\btheta, \y)$
-$$
-\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
-$$
-where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
-\begin{align*}
-\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
-\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
-\end{align*}
+<!-- 1. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(\btheta, \y)$ -->
+<!-- $$ -->
+<!-- \tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)} -->
+<!-- $$ -->
+<!-- where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by -->
+<!-- \begin{align*} -->
+<!-- \hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\ -->
+<!-- \mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}. -->
+<!-- \end{align*} -->
 
-2. Normalise $\tilde p_\text{LA}(\btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
-$$
-\tilde p_\text{AQ}(\btheta \, | \, \y) = \frac{\tilde p_\text{LA}(\btheta, \y)}{\tilde p_{\text{AQ}}(\y)},
-$$
-where the normalising constant is calculated using nodes from a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(m, k)$ with $m = \dim(\btheta)$, $k$ nodes per dimension, and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ as
-$$
-\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z).
-$$
-The nodes $\z$ are adapted based on the mode and curvature at the mode of the Laplace approximation as follows
-\begin{align*}
-\btheta(\z) &= \hat{\btheta} + \mathbf{L} \z, \\
-\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
-\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
-\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top.
-\end{align*}
-We typically set $k = 3$ such that there are $3^m$ nodes in total.
+<!-- 2. Normalise $\tilde p_\text{LA}(\btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain -->
+<!-- $$ -->
+<!-- \tilde p_\text{AQ}(\btheta \, | \, \y) = \frac{\tilde p_\text{LA}(\btheta, \y)}{\tilde p_{\text{AQ}}(\y)}, -->
+<!-- $$ -->
+<!-- where the normalising constant is calculated using nodes from a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(m, k)$ with $m = \dim(\btheta)$, $k$ nodes per dimension, and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ as -->
+<!-- $$ -->
+<!-- \tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z). -->
+<!-- $$ -->
+<!-- The nodes $\z$ are adapted based on the mode and curvature at the mode of the Laplace approximation as follows -->
+<!-- \begin{align*} -->
+<!-- \btheta(\z) &= \hat{\btheta} + \mathbf{L} \z, \\ -->
+<!-- \hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\ -->
+<!-- \mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\ -->
+<!-- \mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top. -->
+<!-- \end{align*} -->
+<!-- We typically set $k = 3$ such that there are $3^m$ nodes in total. -->
 
-3. Obtain an unnormalised nested approximation to the posterior marginal of the $i$th latent effect by
-$$
-\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
-$$
-The nodes and weights $\{\mathcal{Q}(m, k), \omega\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
-For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{LA}(x_i, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following Laplace approximation
-$$
-\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}
-$$
-where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x_{-i}(x_i, \btheta), \mathbf{H}_{-i, -i}(x_i, \btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
-\begin{align*}
-\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
-\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
-\end{align*}
-Optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$.
+<!-- 3. Obtain an unnormalised nested approximation to the posterior marginal of the $i$th latent effect by -->
+<!-- $$ -->
+<!-- \tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z). -->
+<!-- $$ -->
+<!-- The nodes and weights $\{\mathcal{Q}(m, k), \omega\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above. -->
+<!-- For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{LA}(x_i, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following Laplace approximation -->
+<!-- $$ -->
+<!-- \tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)} -->
+<!-- $$ -->
+<!-- where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x_{-i}(x_i, \btheta), \mathbf{H}_{-i, -i}(x_i, \btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by -->
+<!-- \begin{align*} -->
+<!-- \hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\ -->
+<!-- \mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}. -->
+<!-- \end{align*} -->
+<!-- Optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$. -->
 
-4. Normalise $\tilde p_\text{LA}(x_i, \y)$ using $\tilde p_{\text{AQ}}(\y)$ to obtain
-$$
-\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
-$$
-which may be evaluated for some choice of values $x_i \in \{\ldots\}$.
+<!-- 4. Normalise $\tilde p_\text{LA}(x_i, \y)$ using $\tilde p_{\text{AQ}}(\y)$ to obtain -->
+<!-- $$ -->
+<!-- \tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}. -->
+<!-- $$ -->
+<!-- which may be evaluated for some choice of values $x_i \in \{\ldots\}$. -->
 
-## Backward algorithm\label{sec:algorithm}
+<!-- The above omiited "forward" algorithm section is written as you would do to calculate things in code. -->
+<!-- The "backward" algorithm section to follow is written as you would to have everything defined before it's used. -->
+
+## Algorithm\label{sec:algorithm}
 
@@
 
-# Application to the Naomi model\label{sec:results}
+# Application to data from Malawi\label{sec:results}
 
-* The \textsc{R} [@r] code used to produce all results we describe is available at `github.com/athowes/elgm-inf`. The inference method is available in versions 0.5.0. onwards of the `aghq` package
-* Using the `TMB` template, we fit the model using four inferential approaches: (1) empirical Bayes combined with a Gaussian approximation, (2) AGHQ combined with a Gaussian approximation, (3) AGHQ combined with a Laplace approximation and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS).
-We treat results from NUTS as the gold-standard
-* We used the Kolmogorov-Smirnov test for the maximum difference between two empirical cumulative distribution functions to compare posterior marginal distributions
-* We performed posterior predictive checks to assess the coverage of our estimates via the uniformity of the data within each posterior marginal distribution
+Using a single `TMB` template, we fit the simplified Naomi model to data from Malawi using four inferential approaches: (1) empirical Bayes combined with a Gaussian approximation, (2) AGHQ combined with a Gaussian approximation, (3) AGHQ combined with a Laplace approximation and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS).
+The \textsc{R} [@r] code used to produce all results we describe is available at `github.com/athowes/elgm-inf`.
 
-<!-- \begin{table}[] -->
-<!-- \small -->
-<!-- \begin{tabularx}{\textwidth}{llll} -->
-<!-- \toprule -->
-<!-- Inference method & KS & MMD & PSIS \\ -->
-<!-- \midrule -->
-<!-- Empirical Bayes, Gaussian & 0 & 0 & 0 \\ -->
-<!-- AGHQ, Gaussian & 0 & 0 & 0 \\ -->
-<!-- AGHQ, Laplace & 0 & 0 & 0 \\ -->
-<!-- NUTS & 0 & 0 & 0 \\ -->
-<!-- \bottomrule -->
-<!-- \end{tabularx} -->
-<!-- \label{tab:results} -->
-<!-- \small -->
-<!-- \end{table} -->
+\begin{table}[]
+\small
+\begin{tabularx}{\textwidth}{lll}
+\toprule
+ & Run-time & Memory \\
+\midrule
+Empirical Bayes, Gaussian & 0 & 0 \\
+AGHQ, Gaussian & 0 & 0 \\
+AGHQ, Laplace & 0 & 0 \\
+NUTS & 0 & 0 \\
+\bottomrule
+\end{tabularx}
+\label{tab:results}
+\small
+\end{table}
+
+## NUTS convergence
+
+In order to treat results from NUTS as the gold-standard we assessed MCMC convergence using.
+We ran the algorithm
+
+## Model assessment
+
+We performed posterior predictive checks to assess the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.
+
+## Inference comparison
+
+We used three methods to assess the accuracy posterior distributions produced by each inferential method: (1) Kolmogorov-Smirnov (KS) tests performed on posterior marginals for the maximum difference between the two empirical cumulative distribution functions, (2) maximum mean discrepancy (MMD), and (3) Pareto-smoothed importance sampling (PSIS).
+See results in Table \ref{tab:results}.
+
+\begin{table}[]
+\small
+\begin{tabularx}{\textwidth}{llll}
+\toprule
+ & KS & MMD & PSIS \\
+\midrule
+Empirical Bayes, Gaussian & 0 & 0 & 0 \\
+AGHQ, Gaussian & 0 & 0 & 0 \\
+AGHQ, Laplace & 0 & 0 & 0 \\
+NUTS & 0 & 0 & 0 \\
+\bottomrule
+\end{tabularx}
+\label{tab:results}
+\small
+\end{table}
 

### 2023-01-11 src/docs_paper/paper.Rmd
Iterate paper and appendix

@@
   | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa.
-Multiple outcomes of interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
+Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
 We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach used currently, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler.
+Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach which is currently in use, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler.
 By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
@@
 
-Situational requirements combined with the complexity of the model present a difficult Bayesian inference problem.
+The use case requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
 Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
@@
 The model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version of the model which is defined only at $T_1$, omitting temporal projection to $T_2$ and $T_3$.
-Below we provide an overview of the three components in the simplified model, highlighting aspects which make it a challenge for existing inferential approaches.
-More complete mathematical details of this simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
+We consider a simplified version of the model which is defined only at $T_1$ omitting temporal projection to $T_2$ and $T_3$.
+Below we provide an overview of the three components in the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
+A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
 
@@
 
-Consider a country in sub-Saharan Africa where a household survey has taken place at $T_1$.
-Let $x$ index district, $a$ index five-year age band, $s$ index sex and $t$ index time.
+Consider a country in sub-Saharan Africa where a household survey has taken place during the quarter $T_1$.
+Let $x$ index district, $a$ five-year age band, $s$ sex and $t$ time.
 For ease of notation, let $i$ index the coarsest district-age-sex division included in the model.
 The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
-Let $N_i$ be the population size, and consider the following three indicators: HIV prevalence $\rho_i \in [0, 1]$, ART coverage $\alpha_i \in [0, 1]$, and annual HIV incidence rate $\lambda_i > 0$.
-
+Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be ART coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
 We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
@@
 $$
-where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$.
-
+where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
 Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
@@
 $$
-where $j$ index individuals in all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
+where $j$ indexes individuals across all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
 The observed number of indicator cases is then
@@
 $$
-where $\theta_{\mathcal{I}}$ are the following weighted aggregates
+where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
 \begin{align*}
@@
 \end{align*}
-We link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ by
+To link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ we use
 $$
@@
 $$
-where $\Omega_T$ is the mean duration of recent infection and $\beta_T$ is the false recent ratio.
+where the mean duration of recent infection $\Omega_T$ and  is the false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.
 
@@
 \begin{align*}
-\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \tilde{\eta}^{\rho^\text{ANC}}_i, \\
-\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \tilde{\eta}^{\rho^\text{ANC}}_i.
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
+\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\rho^\text{ANC}}_i.
 \end{align*}
-
-Let $W^\text{ANC}_i$ be the number of ANC clients, $X^\text{ANC}_i$ the number of those with ascertained status, $Y^\text{ANC}_i$ the number of those with positive status, and $Z^\text{ANC}_i$ the number of ANC clients already on ART prior to their first ANC visit.
-ANC data from the year of the most recent survey is included in the model.
-
-For women 15-49 the predicted number of ANC clients $\Psi_i$ is a log-linear model
-$$
-\log (\Psi_i) = \log (N_i) + \tilde \eta^\Psi
-$$
+We inform these processes using the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $X^\text{ANC}_\mathcal{I}$, the number of those with positive status $Y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $Z^\text{ANC}_\mathcal{I}$.
+We use the binomial working likelihoods
+\begin{align*}
+Y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(Y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}) \\
+Z^\text{ANC}_\mathcal{I} &\sim \text{Bin}(Y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}),
+\end{align*}
+where again we use weighted aggregates
+\begin{align*}
+\rho^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in \mathcal{I}} \Psi_i}, \\
+\alpha^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}},
+\end{align*}
+with $\Psi_i$ the number of pregnant women.
 
@@
 
-Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
+Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
+We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$.
 
-Let $\dot A_i$ be the number of people receiving ART.
-Likelihood section to be written.
+Let $\dot A_\mathcal{I}$ be the number of people receiving ART.
 

### 2023-01-12 src/naomi-simple_mcmc/mcmc-convergence.Rmd
MCMC convergence assessment moved to separate report, which generates notebook from .Rmd

---
title: "MCMC convergence diagnostics`"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**

  **Task** 
---

```{r}
out <- readRDS("depends/out.rds")

mcmc <- out$mcmc

bayesplot::color_scheme_set("viridisA")
ggplot2::theme_set(theme_minimal())

#' Rhat
#' Looking for values less than 1.1 here
rhats <- bayesplot::rhat(mcmc)
bayesplot::mcmc_rhat(rhats)
(big_rhats <- rhats[rhats > 1.1])
length(big_rhats) / length(rhats)

#' ESS ratio
#' Worry about values less than 0.1 here... which they all are :clown:
ratios <- bayesplot::neff_ratio(mcmc)
bayesplot::mcmc_neff(ratios)

#' Autocorrelation
#' High values of autocorrelation in the chains
bayesplot::mcmc_acf(mcmc, pars = vars(starts_with("beta")))

#' Univariate traceplots

#' Assess these visually
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("beta")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("logit")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_sigma")))

#' Prevalence model
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_rho_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_a[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_as[")))

#' ART model
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_alpha_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_a[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_as[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_xa[")))

#' Other
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_lambda_x[")))

#' ANC model
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_anc_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_anc_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_or_gamma["))) #' N.B. these are from the ANC attendance model

#' Pairs plots
#' Prior suspicion (from Jeff, Tim, Rachel) that the ART attendance model is unidentifiable
#' Let's have a look at the pairs plot for neighbouring districts and the log_or_gamma parameter
nb <- area_merged %>%
  filter(area_level == max(area_level)) %>%
  bsae::sf_to_nb()

neighbours_pairs_plot <- function(par, i) {
  neighbour_pars <- paste0(par, "[", c(i, nb[[i]]), "]")
  bayesplot::mcmc_pairs(mcmc, pars = neighbour_pars, diag_fun = "hist", off_diag_fun = "hex")
}

area_merged %>%
  filter(area_level == max(area_level)) %>%
  print(n = Inf)
neighbours_pairs_plot("log_or_gamma", 5) #' Nkhata Bay and neighbours
neighbours_pairs_plot("log_or_gamma", 26) #' Blantyre and neighbours

#' NUTS specific assessment
np <- bayesplot::nuts_params(mcmc)

#' Number of divergent transitions
#' In this instance it's zero, so no need to do further divergent transition investigation
np %>%
  filter(Parameter == "divergent__") %>%
  summarise(n_divergent = sum(Value))

bayesplot::mcmc_nuts_divergence(np, bayesplot::log_posterior(mcmc))

#' Energy plots
#' Ideally these two histograms would be the same (Betancourt 2017)
#' The histograms are quite different, in a way suggesting the chains may not be
#' fully exploring the tails of the target distribution
bayesplot::mcmc_nuts_energy(np)
```

### 2023-01-12 make/_make.R

@@ -29,6 +29,8 @@ orderly::orderly_commit(id)
 id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmbstan = TRUE))
 orderly::orderly_commit(id)
 
+run_commit_push("naomi-simple_mcmc")
+
 #' Statistical methods development
 run_commit_push("dev_sinla") #' Without experiments
 

### 2023-01-12 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Clean up MCMC convergence notebook

@@
 abstract: |
-  **Background**
+  **Background** In the report `naomi-simple_fit` with parameter `tmbstan = TRUE`, we used the NUTS algorithm to perform MCMC inference for the simplified Naomi model.
 
-  **Task** 
+  **Task** Here we assess whether or not the results of the MCMC are suitable using a range of diagnostic tools.
 ---
+
+We start by obtaining results from the latest version of `naomi-simple_fit` with `tmbstan = TRUE`.
 
@@
 out <- readRDS("depends/out.rds")
+mcmc <- out$mcmc
+```
 
-mcmc <- out$mcmc
+This MCMC took `r as.numeric(out$time)` minutes to run
 
+```{r}
 bayesplot::color_scheme_set("viridisA")
 ggplot2::theme_set(theme_minimal())
+```
 
-#' Rhat
-#' Looking for values less than 1.1 here
+# $\hat R$
+
+We are looking for values of $\hat R$ less than 1.1 here.
+
+```{r}
 rhats <- bayesplot::rhat(mcmc)
@@
 length(big_rhats) / length(rhats)
+```
 
-#' ESS ratio
-#' Worry about values less than 0.1 here... which they all are :clown:
+# ESS ratio
+
+Reasonable to be worried about values less than 0.1 here.
+
+```{r}
 ratios <- bayesplot::neff_ratio(mcmc)
 bayesplot::mcmc_neff(ratios)
+```
 
-#' Autocorrelation
-#' High values of autocorrelation in the chains
+# Autocorrelation
+
+How much autocorrelation is there in the chains?
+
+```{r}
 bayesplot::mcmc_acf(mcmc, pars = vars(starts_with("beta")))
+```
 
-#' Univariate traceplots
+# Univariate traceplots
 
-#' Assess these visually
+```{r}
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("beta")))
@@
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_sigma")))
+```
 
-#' Prevalence model
+## Prevalence model
+
+```{r}
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_x[")))
@@
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_as[")))
+```
 
-#' ART model
+## ART model
+
+```{r}
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_x[")))
@@
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_xa[")))
+```
 
-#' Other
+## Other
+
+```{r}
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_lambda_x[")))
+```
 
-#' ANC model
+## ANC model
+
+```{r}
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_anc_rho_x[")))
@@
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_or_gamma["))) #' N.B. these are from the ANC attendance model
+```
 
-#' Pairs plots
-#' Prior suspicion (from Jeff, Tim, Rachel) that the ART attendance model is unidentifiable
-#' Let's have a look at the pairs plot for neighbouring districts and the log_or_gamma parameter
+# Pairs plots
+
+There is a prior suspicion (from Jeff, Tim, Rachel) that the ART attendance model is unidentifiable.
+Let's have a look at the pairs plot for neighbouring districts and the `log_or_gamma` parameter.
+
+```{r}
 nb <- area_merged %>%
@@
   print(n = Inf)
+
 neighbours_pairs_plot("log_or_gamma", 5) #' Nkhata Bay and neighbours
 neighbours_pairs_plot("log_or_gamma", 26) #' Blantyre and neighbours
+```
 
-#' NUTS specific assessment
+# NUTS specific assessment
+
+```{r}
 np <- bayesplot::nuts_params(mcmc)
+```
 
-#' Number of divergent transitions
-#' In this instance it's zero, so no need to do further divergent transition investigation
+Are there any divergent transitions?
+
+```{r}
 np %>%
@@
 bayesplot::mcmc_nuts_divergence(np, bayesplot::log_posterior(mcmc))
+```
 
-#' Energy plots
-#' Ideally these two histograms would be the same (Betancourt 2017)
-#' The histograms are quite different, in a way suggesting the chains may not be
-#' fully exploring the tails of the target distribution
+We can also use energy plots (Betancourt 2017): ideally these two histograms would be the same 
+When the histograms are quite different, it may suggest the chains are not fully exploring the tails of the target distribution.
+
+```{r}
 bayesplot::mcmc_nuts_energy(np)

### 2023-01-12 utils/archive_to_docs.R

@@ -32,3 +32,4 @@ archive_to_docs("example_naomi")
 archive_to_docs("explore_posterior-comparison")
 archive_to_docs("docs_bayescomp-poster")
 archive_to_docs("plot-tikz_algorithm-flowchart")
+archive_to_docs("naomi-simple_mcmc")

### 2023-01-12 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Small edits final

@@
 ---
-title: "MCMC convergence diagnostics`"
+title: "MCMC convergence diagnostics"
 author:
@@
 
-area_merged %>%
-  filter(area_level == max(area_level)) %>%
-  print(n = Inf)
+# area_merged %>%
+#   filter(area_level == max(area_level)) %>%
+#   print(n = Inf)
+```
 
-neighbours_pairs_plot("log_or_gamma", 5) #' Nkhata Bay and neighbours
-neighbours_pairs_plot("log_or_gamma", 26) #' Blantyre and neighbours
+Here are Nkhata Bay and neighbours:
+
+```{r}
+neighbours_pairs_plot("log_or_gamma", 5) 
+```
+
+And here are Blantyre and neighbours:
+
+```{r}
+neighbours_pairs_plot("log_or_gamma", 26)
 ```

### 2023-01-12 src/naomi-simple_compare/comparison.Rmd
Add comparison notebook start

---
title: "MCMC convergence diagnostics"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**  We have run the simplified Naomi model using a range of inference methods.

  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods.
---

```{r}
tmb <- readRDS("depends/tmb.rds")
tmbstan <- readRDS("depends/tmbstan.rds")
```

# Time taken

```{r}
data.frame(
  "TMB" = tmb$time,
  "tmbstan" = tmbstan$time
)
```

# Histogram

```{r}
#' To write!
```

# KS

```{r}
#' To write!
```

# MMD

```{r}
#' To write!
```

# PSIS

```{r}
#' To write!
```

### 2023-01-12 make/_make.R

@@ -30,6 +30,7 @@ id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmbstan = TRUE)
 orderly::orderly_commit(id)
 
 run_commit_push("naomi-simple_mcmc")
+run_commit_push("naomi-simple_compare")
 
 #' Statistical methods development
 run_commit_push("dev_sinla") #' Without experiments

### 2023-01-12 src/naomi-simple_fit/script.R
Fix AGHQ to k = 1 for now

@@ -108,7 +108,7 @@ if(aghq) {
   (mvQuad::size(mvQuad::createNIGrid(n_hyper, "GHe", 3, "sparse"))$gridpoints)
 
   #' Fit AGHQ model
-  quad <- fit_aghq(tmb_inputs, k = k)
+  quad <- fit_aghq(tmb_inputs, k = 1)
 
   #' Add uncertainty
   quad <- sample_aghq(quad, M = nsample)

### 2023-01-13 src/naomi-simple_compare/comparison.Rmd
Fix AGHQ to k = 1 for now
Improve comparison notebook

@@
 
+# Background
+
 ```{r}
 tmb <- readRDS("depends/tmb.rds")
+aghq <- readRDS("depends/aghq.rds")
 tmbstan <- readRDS("depends/tmbstan.rds")
+```
+
+All of the possible parameter names are as follows:
+
+```{r}
+unique(names(tmb$fit$obj$env$par))
 ```
@@
   "TMB" = tmb$time,
+  "aghq" = aghq$time,
   "tmbstan" = tmbstan$time
@@
 ```{r}
-#' To write!
+histogram_plot <- function(par) {
+  df_compare <- rbind(
+    data.frame(method = "TMB", samples = as.numeric(tmb$fit$sample[[par]])),
+    data.frame(method = "aghq", samples = as.numeric(aghq$quad$sample[[par]])),
+    data.frame(method = "tmbstan", samples = as.numeric(unlist(rstan::extract(tmbstan$mcmc, pars = par))))
+  )
+
+  df_compare %>%
+    group_by(method) %>%
+    summarise(n = n())
+
+  ggplot(df_compare, aes(x = samples, fill = method, col = method)) +
+    geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity") +
+    theme_minimal() +
+    facet_grid(method~.) +
+    labs(x = paste0(par), y = "Density", fill = "Method") +
+    scale_color_manual(values = multi.utils::cbpalette()) +
+    scale_fill_manual(values = multi.utils::cbpalette()) +
+    theme(legend.position = "none") +
+    labs(title = paste0(par))
+}
+
+histogram_plot("beta_anc_rho")
 ```
@@
 
+## Individual parameters
+
 ```{r}
-#' To write!
+to_ks_df <- function(par) {
+  samples_tmb <- t(tmb$fit$sample[[par]])
+  samples_aghq <- t(aghq$quad$sample[[par]])
+  samples_tmbstan <- as.data.frame(rstan::extract(tmbstan$mcmc, pars = par)[[par]])
+
+  n <- ncol(samples_tmbstan)
+  ks_tmb <- numeric(n)
+  ks_aghq <- numeric(n)
+  for(i in 1:n) {
+    ks_tmb[i] <- inf.utils::ks_test(samples_tmb[, i], samples_tmbstan[, i])$D
+    ks_aghq[i] <- inf.utils::ks_test(samples_aghq[, i], samples_tmbstan[, i])$D
+  }
+
+ rbind(
+   data.frame(method = "TMB", ks = ks_tmb, par = par, index = 1:n),
+   data.frame(method = "aghq", ks = ks_aghq, par = par, index = 1:n)
+  )
+}
+
+to_ks_df_2 <- function(par) {
+  all_par_names <- names(tmb$fit$obj$env$par)
+  par_names <- all_par_names[stringr::str_starts(all_par_names, par)]
+  unique_par_names <- unique(par_names)
+
+  samples_tmb <- tmb$fit$sample[unique_par_names]
+  samples_tmb <- lapply(samples_tmb, function(x) as.data.frame(t(x)))
+
+  samples_aghq <- aghq$quad$sample[unique_par_names]
+  samples_aghq <- lapply(samples_aghq, function(x) as.data.frame(t(x)))
+
+  samples_tmbstan <- rstan::extract(tmbstan$mcmc, pars = unique_par_names)
+  samples_tmbstan <- lapply(samples_tmbstan, function(x) as.data.frame(x))
+
+  table <- table(par_names)
+  unique_par_names <- unique(par_names)
+  for(par in unique_par_names) {
+    par_length <- table[par]
+    if(par_length > 1) {
+      par_colnames <- paste0(par, "[", 1:par_length, "]")
+    } else {
+      par_colnames <- paste0(par)
+    }
+    colnames(samples_tmb[[par]]) <- par_colnames
+    colnames(samples_aghq[[par]]) <- par_colnames
+    colnames(samples_tmbstan[[par]]) <- par_colnames
+  }
+
+  samples_tmb <- dplyr::bind_cols(samples_tmb)
+  samples_aghq <- dplyr::bind_cols(samples_aghq)
+  samples_tmbstan <- dplyr::bind_cols(samples_tmbstan)
+
+  n <- ncol(samples_tmbstan)
+  ks_tmb <- numeric(n)
+  ks_aghq <- numeric(n)
+  for(i in 1:n) {
+    ks_tmb[i] <- inf.utils::ks_test(samples_tmb[, i], samples_tmbstan[, i])$D
+    ks_aghq[i] <- inf.utils::ks_test(samples_aghq[, i], samples_tmbstan[, i])$D
+  }
+
+  rbind(
+    data.frame(method = "TMB", ks = ks_tmb, par = names(samples_tmbstan), index = 1:n),
+    data.frame(method = "aghq", ks = ks_aghq, par = names(samples_tmbstan), index = 1:n)
+  )
+}
+
+plot_ks_df <- function(ks_df) {
+  wide_ks_df <- pivot_wider(ks_df, names_from = "method", values_from = "ks") %>%
+    mutate(ks_diff = TMB - aghq)
+
+  mean_ks_diff <- mean(wide_ks_df$ks_diff)
+
+  boxplot <- wide_ks_df %>%
+    ggplot(aes(x = ks_diff)) +
+    geom_boxplot(width = 0.5) +
+    coord_flip() +
+    labs(
+      title = paste0("Mean KS difference is ", mean_ks_diff),
+      subtitle = ">0 then TMB more different to tmbstan, <0 then aghq more different",
+      x = "KS(TMB, tmbstan) - KS(aghq, tmbstan)"
+    ) +
+    theme_minimal()
+
+  scatterplot <- ggplot(wide_ks_df, aes(x = TMB, y = aghq)) +
+    geom_point(alpha = 0.5) +
+    xlim(0, 0.5) +
+    ylim(0, 0.5) +
+    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
+    labs(
+      title = paste0("KS tests for ", ks_df$par, " of length ", max(ks_df$index)),
+      subtitle = "Values along y = x have similar KS",
+      x = "KS(aghq, tmbstan)", y = "KS(TMB, tmbstan)"
+    ) +
+    theme_minimal()
+
+  cowplot::plot_grid(scatterplot, boxplot, ncol = 2, rel_widths = c(1.3, 1))
+}
+
+ks_helper <- function(par) to_ks_df(par) %>% plot_ks_df()
+ks_helper_2 <- function(par) to_ks_df_2(par) %>% plot_ks_df()
+
+ks_helper_2("beta")
+ks_helper_2("logit")
+ks_helper_2("log_sigma")
+
+ks_helper("u_rho_x")
+ks_helper("u_rho_xs")
+ks_helper("us_rho_x")
+ks_helper("us_rho_xs")
+ks_helper("u_rho_a")
+ks_helper("u_rho_as")
+
+ks_helper("u_alpha_x")
+ks_helper("u_alpha_xs")
+ks_helper("us_alpha_x")
+ks_helper("us_alpha_xs")
+ks_helper("u_alpha_a")
+ks_helper("u_alpha_as")
+ks_helper("u_alpha_xa")
+
+ks_helper("ui_anc_rho_x")
+ks_helper("ui_anc_alpha_x")
+ks_helper("log_or_gamma")
+```
+
+## Summary table
+
+```{r}
+ks_summary_table <- lapply(unique(names(tmb$fit$obj$env$par)), function(par) {
+  to_ks_df(par) %>%
+    group_by(method) %>%
+    summarise(ks = mean(ks), par = par[1])
+}) %>%
+  bind_rows() %>%
+  pivot_wider(names_from = "method", values_from = "ks") %>%
+  rename(
+    "Parameter" = "par",
+    "KS(aghq, tmbstan)" = "aghq",
+    "KS(TMB, tmbstan)" = "TMB",
+  )
+
+ks_summary_table %>%
+  gt::gt() %>%
+  gt::fmt_number(
+    columns = starts_with("KS"),
+    decimals = 3
+  )
+
+ggplot(ks_summary_table, aes(x = `KS(TMB, tmbstan)`, y = `KS(aghq, tmbstan)`)) +
+  geom_point(alpha = 0.5) +
+  xlim(0, 1) +
+  ylim(0, 1) +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
+  theme_minimal()
 ```

### 2023-01-14 src/naomi-simple_compare/comparison.Rmd
Clean up comparison report

--- 2023-01-12 src/naomi-simple_compare/comparison.Rmd
@@
 ```{r}
-#' To write!
+histogram_plot("beta_anc_rho")
 ```
@@
 
+## Individual parameters
+
 ```{r}
-#' To write!
+ks_helper <- function(par) to_ks_df(par) %>% ks_plot(par = par)
+
+ks_helper("beta")
+ks_helper("logit")
+ks_helper("log_sigma")
+
+ks_helper("u_rho_x")
+ks_helper("u_rho_xs")
+ks_helper("us_rho_x")
+ks_helper("us_rho_xs")
+ks_helper("u_rho_a")
+ks_helper("u_rho_as")
+
+ks_helper("u_alpha_x")
+ks_helper("u_alpha_xs")
+ks_helper("us_alpha_x")
+ks_helper("us_alpha_xs")
+ks_helper("u_alpha_a")
+ks_helper("u_alpha_as")
+ks_helper("u_alpha_xa")
+
+ks_helper("ui_anc_rho_x")
+ks_helper("ui_anc_alpha_x")
+ks_helper("log_or_gamma")
+```
+
+## Summary table
+
+```{r}
+ks_summary <- lapply(unique(names(tmb$fit$obj$env$par)), function(x) {
+  to_ks_df(x) %>%
+    group_by(method) %>%
+    summarise(ks = mean(ks), par = x)
+}) %>%
+  bind_rows() %>%
+  pivot_wider(names_from = "method", values_from = "ks") %>%
+  rename(
+    "Parameter" = "par",
+    "KS(aghq, tmbstan)" = "aghq",
+    "KS(TMB, tmbstan)" = "TMB",
+  )
+
+ks_summary %>%
+  gt::gt() %>%
+  gt::fmt_number(
+    columns = starts_with("KS"),
+    decimals = 3
+  )
+
+ggplot(ks_summary, aes(x = `KS(TMB, tmbstan)`, y = `KS(aghq, tmbstan)`)) +
+  geom_point(alpha = 0.5) +
+  xlim(0, 1) +
+  ylim(0, 1) +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
+  labs(x = "KS(aghq, tmbstan)", y = "KS(TMB, tmbstan)") +
+  theme_minimal()
+
+ks_summary %>%
+  mutate(`KS difference` = `KS(TMB, tmbstan)` - `KS(aghq, tmbstan)`) %>%
+  ggplot(aes(x = `KS difference`)) +
+    geom_boxplot(width = 0.5) +
+    coord_flip() +
+    labs(x = "KS(TMB, tmbstan) - KS(aghq, tmbstan)") +
+    theme_minimal()
 ```

### 2023-01-15 src/naomi-simple_mcmc/mcmc-convergence.Rmd
20,000 x 4 back from DIDEHPC

@@
 
-This MCMC took `r as.numeric(out$time)` minutes to run
+This MCMC took `r out$time` to run
 
@@
 ```{r}
+area_merged <- sf::read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
+
 nb <- area_merged %>%

### 2023-01-15 make/hpc_mcmc.R

@@ -83,7 +83,7 @@ recent_ctx <- dplyr::filter(ctx_info, created == max(created))
 ctx <- context::context_load(context::context_read(recent_ctx$id, "context"))
 queue <- didehpc::queue_didehpc(ctx)
 queue$task_list()
-t <- queue$task_get("ee0741f836805787c902f8b8f126c4cb")
+t <- queue$task_get("f5cc7e5aa225fa54b1c6fdb8cae41c6d")
 
 t$status()
 t$log()

### 2023-01-15 src/naomi-simple_compare/comparison.Rmd
Start thinking about how to calculate the PSIS: basically we just want to put all the samples into tmb$fit$obj$fn(), get the NLL, difference them, then use the loo package
Add orderly details about the runs we compare

@@
 
+We compare the inference results from `TMB`, `aghq` and `tmbstan`.
+Import these inference results as follows:
+
 ```{r}
@@
 tmbstan <- readRDS("depends/tmbstan.rds")
+```
+
+For more information about the conditions under which these results were generated, see:
+
+```{r}
+depends <- yaml::read_yaml("orderly.yml")$depends
+
+for(i in seq_along(depends)) {
+  report_name <- names(depends[[i]])
+  print(paste0("Inference results obtained from ", report_name, " with the query ", depends[[i]][[report_name]]$id))
+  report_id <- orderly::orderly_search(query = depends[[i]][[report_name]]$id, report_name)
+  print(paste0("Obtained report had ID ", report_id, " and was run with the following parameters:"))
+  print(orderly::orderly_info(report_id, report_name)$parameters)
+}
 ```
@@
 
+Suppose we have two sets of samples from the posterior.
+For each sample we are going to want to evaluate the log-likelihood, so that we can calculate the log-likelihood ratios.
+We can extract the `TMB` objective function for the log-likelihood as follows:
+
 ```{r}
-#' To write!
+tmb$fit$obj$fn()
 ```

### 2023-01-15 src/naomi-simple_compare/comparison.Rmd
Correct title

@@
 ---
-title: "MCMC convergence diagnostics"
+title: "Inference methods comparison for the simplified Naomi model"
 author:

### 2023-01-15 src/docs_paper/appendix.Rmd
Iterate paper and appendix
Add MCMC convergence and suitability section outline to appendix

@@
 
+# MCMC convergence and suitability
+
+We assessed MCMC convergence and suitability using a range of graphical and numerical tests.
+These included the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
+
+For the time being, this analysis is available from `athowes.github.io/elgm-inf/mcmc-convergence`.
+Once the MCMC results are finalised, the analysis will be moved to this appendix and expanded upon.
+The following draft text and references may be useful in that expanded write-up.
+
+* Improved $\hat R$ statistic from @vehtari2021rank. Recommended only to use the sample if the value is less than 1.05.
+* ESS and autocorrelation related to efficiency
+* Traceplots helpful for diagnosis of problems
+* Pairs density plots helpful for understanding relationships between parameters, including possible indentifiability issues
+* Energy plot from @betancourt2017conceptual
+
+\newpage
+
 # \textsc{C++} template
 
-```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi.cpp")}
+```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
 

### 2023-01-15 src/docs_paper/paper.Rmd
Describe KS tests

@@
 
-We used three methods to assess the accuracy posterior distributions produced by each inferential method: (1) Kolmogorov-Smirnov (KS) tests performed on posterior marginals for the maximum difference between the two empirical cumulative distribution functions, (2) maximum mean discrepancy (MMD), and (3) Pareto-smoothed importance sampling (PSIS).
+We used three methods to assess the accuracy of posterior distributions produced by each inferential method: (1) Kolmogorov-Smirnov tests, (2) maximum mean discrepancy, and (3) Pareto-smoothed importance sampling.
+
+Let $\{\theta_i\}_{i = 1}^n$ be posterior marginal samples with empirical cumulative distribution (ECDF) function $F(\vartheta) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\theta_i \leq \vartheta}$.
+The two-sample Kolmogorov-Smirnov (KS) test statistic is given by the maximum absolute difference between two ECDFs.
+We compare the two KS statistics
+\begin{align*}
+D_\texttt{aghq} &= \sup_\vartheta | F_\texttt{tmbstan}(\vartheta) - F_\texttt{aghq}(\vartheta)|, \\
+D_\texttt{TMB} &= \sup_\vartheta | F_\texttt{tmbstan}(\vartheta) - F_\texttt{TMB}(\vartheta)|.
+\end{align*}
+
 See results in Table \ref{tab:results}.

### 2023-01-15 src/naomi-simple_model-checks/model-checks.Rmd
Template for model checking notebook

---
title: "Model checking for simplified Naomi"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**  We have run the simplified Naomi model using a range of inference methods.

  **Task** We check the appropriateness of the model for the data.
---

```{r}
#' Yes!
```

### 2023-01-15 make/_make.R

@@ -31,6 +31,7 @@ orderly::orderly_commit(id)
 
 run_commit_push("naomi-simple_mcmc")
 run_commit_push("naomi-simple_compare")
+run_commit_push("naomi-simple_model-checks")
 
 #' Statistical methods development
 run_commit_push("dev_sinla") #' Without experiments

### 2023-01-15 src/naomi-simple_model-checks/model-checks.Rmd
Thinking a little about how to do model checking

@@
 ```{r}
-#' Yes!
+tmb <- readRDS("depends/tmb.rds")
+
+outputs <- tmb$outputs
+
+indicators <- naomi::add_output_labels(outputs) %>%
+  left_join(outputs$meta_area %>% select(area_level, area_id, center_x, center_y)) %>%
+  sf::st_as_sf()
+
+indicators %>%
+  filter(
+    age_group == "Y015_049",
+    indicator == "prevalence",
+    area_level == 4
+  ) %>%
+  ggplot(aes(fill = mode)) +
+  geom_sf() +
+  viridis::scale_fill_viridis(labels = scales::percent_format()) +
+  naomi::th_map() +
+  facet_wrap(~sex)
+
+indicators %>%
+  filter(
+    age_group == "Y015_049",
+    indicator == "prevalence",
+    area_level == 4
+  ) %>%
+  names()
 ```
+
+Next steps here are:
+
+* Add column to `indicators` with the "raw" value of the indicator
+* Add column to `indicators` with the quantile of the raw value within samples from the posterior
+* Start outputting `tmb$outputs` for the other inference methods
+* Use existing or own functions to produce coverage histograms and ECDF difference plots

### 2023-01-16 src/docs_paper/paper.Rmd
Rerender paper and appendix

@@
 Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at a sufficiently fine-scale resolution to make targetted interventions.
-Producing these estimates is a challenging task, as all available data sources have shortcomings which must be overcome.
-Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost to run, in most countries they only occur infrequently.
+Producing these estimates is a challenging task beccause all available data sources have shortcomings which must be overcome.
+Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost to run, in most countries they occur only infrequently.
 Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on limited or biased samples of the population.
 To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
-Concurrent modelling of data mitigates the limitations of any single data source, increases statistical power and prompts investigation into any conflicting information.
-Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and generate estimates in a yearly process supported by UNAIDS.
+Simultaneous modelling of the data mitigates the limitations of any single source, increases statistical power, and prompts investigation into any conflicting information.
+Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates in a yearly process supported by UNAIDS.
+The creation of estimates by country teams, rather than external agencies, is a noteworthy feature of the HIV response.
+Drawing on expertise closest to the data generating process, improves the accuracy of the process, as well as strengthening trust and ownership of the estimates.
 
@@
 Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-Owing to its speed and flexibility, `TMB` has recently been gaining popularity spatial statistics [@osgoodzimmerman2021statistical].
-Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
-For the Naomi model, we use this option to integrate out the latent field parameters.
+Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgoodzimmerman2021statistical].
+Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments, which for the Naomi model, we use for the high-dimensional latent field parameters.
 Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
 Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
-As such, we have been motivated to find a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.
+This has motivated us to look for a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.
 
@@
 INLA is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
-The central innovation of @rue2009approximate is a way to approximate accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
+The central innovation of @rue2009approximate is an approximation which enables accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
 Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
@@
 We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
-`aghq` is designed to naturally interface with `TMB`, so use of our method is simple when provided a \textsf{C++} user template for the log-posterior.
+Since `aghq` is designed to naturally interface with `TMB`, use of our method is simple when provided a \textsf{C++} user template for the log-posterior.
 
@@
 As a case-study, we fit the simplified Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
-In this section, we also demonstrate a Bayesian workflow, illustrating that these concepts are not only applicable in a simulation-based inference setting.
-Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how our method might be used in other models, and directions for future research.
+We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
+Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how we anticipate our method might be useful for other models, and directions for future research.
 
@@
 * We developed an approximate Bayesian inference algorithm to solve a challenging problem in the small-area estimation of HIV in low resource settings
-* The flexibility of our method implementation, including compatibility with any `TMB` \textsc{C++} template, allows broader use, as well as investigation of, deterministic inference methods than had previously been possible
+* Following further testing, we anticipate adding our inference method as an option for 
+* The flexibility of our method implementation, including compatibility with any `TMB` \textsc{C++} template, allows broader use, as well as investigation of, deterministic inference methods than had previously been possible.
+We are excited about exploration of its accuracy for the following challenging model structures which could not previously be fit: list of model types here.
 * We demonstrated a Bayesian workflow for deterministic inference methods

### 2023-01-18 src/docs_paper/paper.Rmd
Some Jeff and Seth comments

@@
 
-Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at a sufficiently fine-scale resolution to make targetted interventions.
-Producing these estimates is a challenging task beccause all available data sources have shortcomings which must be overcome.
-Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost to run, in most countries they occur only infrequently.
-Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on limited or biased samples of the population.
-To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
+Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at the level of geographic resolution at which heath systems are planned and delivered.
+Producing these estimates is challenging because all available data sources have shortcomings which must be overcome.
+Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they occur only every five years or so, with limited sample size at the district level.
+Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on limited or non-representative samples of the population.
+To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
 Simultaneous modelling of the data mitigates the limitations of any single source, increases statistical power, and prompts investigation into any conflicting information.
@@
 The creation of estimates by country teams, rather than external agencies, is a noteworthy feature of the HIV response.
-Drawing on expertise closest to the data generating process, improves the accuracy of the process, as well as strengthening trust and ownership of the estimates.
+Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust and ownership of the estimates.
 
 The use case requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
-Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
+Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
 Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
@@
 INLA is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
-The central innovation of @rue2009approximate is an approximation which enables accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
+The key innovation of @rue2009approximate is an approximation which enables accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
 Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
@@
 The remainder of this paper is organised as follows.
-In Section \ref{sec:naomi} we describe a simplified version of the Naomi model that we consider in this paper.
+Section \ref{sec:naomi} describes a simplified version of the Naomi model that we consider in this paper.
 Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
-As a case-study, we fit the simplified Naomi model on data from Malawi, and compare the accuracy of inferences in Section \ref{sec:naomi}.
+As a case-study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:naomi}.
 We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
@@
 
-@eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
-The model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version of the model which is defined only at $T_1$ omitting temporal projection to $T_2$ and $T_3$.
+@eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence among pregnant women from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
+The full model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
+We consider a simplified version of the model which is defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
 Below we provide an overview of the three components in the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
@@
 
-Consider a country in sub-Saharan Africa where a household survey has taken place during the quarter $T_1$.
-Let $x$ index district, $a$ five-year age band, $s$ sex and $t$ time.
-For ease of notation, let $i$ index the coarsest district-age-sex division included in the model.
+Consider a country in sub-Saharan Africa where a household survey has taken place during at time $T_1$.
+Let $x$ index district, $a$ five-year age band, and $s$ sex and.
+For ease of notation, let $i$ index the finest district-age-sex division included in the model.
 The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
@@
 $$
-where the mean duration of recent infection $\Omega_T$ and  is the false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.
+where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.
 
@@
 \end{align*}
-We inform these processes using the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $X^\text{ANC}_\mathcal{I}$, the number of those with positive status $Y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $Z^\text{ANC}_\mathcal{I}$.
+We inform these processes by specifying likelihoods for the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $X^\text{ANC}_\mathcal{I}$, the number of those with positive status $Y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $Z^\text{ANC}_\mathcal{I}$.
 We use the binomial working likelihoods
@@
 
+<!-- To-do: define what exactly $p(\y, \x, \btheta)$ is -->
 Given a \textsf{C++} user template `model.cpp` for the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adaptive Gauss-Hermite quadrature.
@@
 
-1. Calculate
+1. Calculate the mode, Hessian at the mode, and lower Cholesky
 \begin{align*}
@@
 \end{align*}
-where
+of the Laplace approximation
 $$
@@
 2.
-Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $m = \dim(\btheta)$, $k$ nodes per dimension, as follows
-\begin{align*}
-\btheta(\z) &= \hat{\btheta} + \mathbf{L} \z, \\
-\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
-\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
-\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top.
-\end{align*}
+Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $m = \dim(\btheta)$, and $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
 We typically set $k = 3$ such that there are $3^m$ nodes in total.
@@
 We are excited about exploration of its accuracy for the following challenging model structures which could not previously be fit: list of model types here.
+* Our approach is similar to that of `inlabru` [@bachl2019inlabru] in the following ways, and different in the following other ways
 * We demonstrated a Bayesian workflow for deterministic inference methods

### 2023-01-19 src/naomi-simple_compare/comparison.Rmd
Strange histograms for hyperparameters
Fix TMB hyper sampling, and fix mislabel of axis in KS comparison plots

@@
   geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
-  labs(x = "KS(aghq, tmbstan)", y = "KS(TMB, tmbstan)") +
+  labs(x = "KS(TMB, tmbstan)", y = "KS(aghq, tmbstan)") +
   theme_minimal()
@@
     theme_minimal()
+```
+
+## Investigation into large KS values
+
+The following parameters have KS values greater than 0.5 in both dimensions.
+
+```{r}
+(big_ks <- ks_summary %>%
+  filter(`KS(aghq, tmbstan)` > 0.5, `KS(TMB, tmbstan)` > 0.5) %>%
+  pull(Parameter))
+```
+
+Let's look into this further by plotting the histograms:
+
+```{r}
+lapply(big_ks, histogram_plot)
 ```

### 2023-01-19 make/_make.R

@@ -22,7 +22,7 @@ id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmb = TRUE))
 orderly::orderly_commit(id)
 
 #' Simplfied Naomi with aghq, and k = 1
-id <- orderly::orderly_run("naomi-simple_fit", parameters = list(aghq = TRUE, k = 1))
+id <- orderly::orderly_run("naomi-simple_fit", parameters = list(aghq = TRUE, k = 1, ndConstruction = "sparse"))
 orderly::orderly_commit(id)
 
 #' Simplfied Naomi with tmbstan, and niter = 1000

### 2023-01-19 src/naomi-simple_compare/functions.R

@@ -96,7 +96,7 @@ ks_plot <- function(ks_df, par) {
     labs(
       title = paste0("KS tests for ", par, " of length ", max(ks_df$index)),
       subtitle = "Values along y = x have similar KS",
-      x = "KS(aghq, tmbstan)", y = "KS(TMB, tmbstan)"
+      x = "KS(TMB, tmbstan)", y = "KS(aghq, tmbstan)"
     ) +
     theme_minimal()
 

### 2023-01-19 src/naomi-simple_fit/functions.R

@@ -264,11 +264,14 @@ local_sample_tmb <- function(fit, nsample = 1000, rng_seed = NULL, random_only =
   if (to_tape) fit$obj$retape(FALSE)
   if (!random_only) {
     if (verbose) print("Calculating joint precision")
-    hess <- sdreport_joint_precision(fit$obj, fit$par.fixed)
+    hess <- naomi:::sdreport_joint_precision(fit$obj, fit$par.fixed)
     if (verbose) print("Inverting precision for joint covariance")
     cov <- solve(hess)
+    if (!isSymmetric(cov, tol = sqrt(.Machine$double.eps), check.attributes = FALSE)) {
+      stop("cov must be a symmetric matrix")
+    }
     if (verbose) print("Drawing sample")
-    smp <- mvtnorm::rmvnorm(nsample, fit$par.full, cov)
+    smp <- mvtnorm::rmvnorm(n = nsample, mean = fit$par.full, sigma = cov, method = "eigen", checkSymmetry = FALSE)
   } else {
     r <- fit$obj$env$random # Indices of the random effects
     par_f <- fit$par.full[-r] # Mode of the fixed effects

### 2023-01-19 src/naomi-simple_fit/script.R
aghq hypers debugged so that we do have samples, still need to work out an approach for what to do when k = 1

@@ -1,5 +1,5 @@
 #' Uncomment and run the two line below to resume development of this script
-# orderly::orderly_develop_start("naomi-simple_fit", parameters = list(tmb = TRUE))
+# orderly::orderly_develop_start("naomi-simple_fit", parameters = list(aghq = TRUE))
 # setwd("src/naomi-simple_fit")
 
 if(tmb + aghq + tmbstan != 1) {

### 2023-01-20 src/docs_paper/paper.Rmd
Add to acknowledgements section

@@
 AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1).
+AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).
+AS was supported by...
+SRF was supported by the EPSRC (EP/V002910/2).
+JWE was supported by UNAIDS and National Institute of Allergy and Infectious Disease of the National Institutes of Health (R01AI136664).
+This research was supported by the MRC Centre for Global Infectious Disease Analysis (MR/R015600/1), jointly funded by the UK Medical Research Council (MRC) and the UK Foreign, Commonwealth \& Development Office (FCDO), under the MRC/FCDO Concordat program and is also part of the EDCTP2 programme supported by the European Union.

### 2023-01-20 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Some iteration of BayesComp poster

--- 2022-11-01 src/docs_bayescomp-poster/bayescomp-poster.Rmd
@@
 ---
-main_topsize: 0.2
-main_bottomsize: 0.075
-poster_height: "48in"
-poster_width: "36in"
+main_topsize: 0.08
+main_bottomsize: 0.03
+poster_height: "36in"
+poster_width: "48in"
 font_family: "Open Sans"
 # Essentials
-title: Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV model
+title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
 author:
-  - name: '**Adam Howes**'
+  - name: "Adam Howes"
     affil: 1, 2
@@
     address: Department of Computer Science, University of Oxford
-main_findings:
-  - "Combining **simplified INLA**, automatic differentiation and **adaptive Gauss-Hermite quadrature** for fast and accurate inference in a challenging, real-world problem"
 logoleft_name: "IMP_ML_W_N.png"
-# logoright_name: "UNAIDS-reverse-EN.gif"
+logoright_name: "UW_ML_W_N.png"
 # Style & formatting
 column_numbers: 3
-title_textsize: "60px"
-author_textsize: "1.17em"
+title_textsize: "65px"
+author_textsize: "45px"
 authorextra_textsize: "35px"
 affiliation_textsize: "25px"
-affiliation_textcol: '#00000080'
 caption_fontsize: "20px"
+# Title box
+title_textcol: "#ffffff"
+author_textcol: "#ffffff"
+affiliation_textcol: "#ffffff"
 # Main
@@
 # Poster body options
-primary_colour: '#38761C'
+primary_colour: '#3b77D8'
 secondary_colour: '#3b77D8'
@@
 output: 
-  posterdown::posterdown_betterport:
+  posterdown::posterdown_html:
     self_contained: false
@@
 
-<!-- Using LaTeX to create spacing -->
-$$\\[0.15in]$$
+## Summary
 
-## The Naomi model
+* Approximate Bayesian inference method using adaptive Gauss-Hermite quadrature and Laplace approximation
+* Implemented as a part of the `aghq` package, allowing easy use for any model with a `TMB` C++ user template
+* Motivated by a challenging problem in evidence synthesis for small-area estimation of HIV
+
+<!-- 38761C -->
+
+## The Naomi HIV model
+
+* Spatio-temporal small-area estimation model of HIV indicators which combines data from household surveys, antenatal care clincs, and routine service provision of antiretroviral therapy [@eaton2021naomi]
+* Yearly estimation process where the model is run interactively by country teams using the web-app `naomi.unaids.org`
+* Inference currently conducted with empirical Bayes and a Gaussian approximation to the latent field using Template Model Builder `TMB` [@kristensen2016tmb]
+* Model incompatible with the `R-INLA` implementation of integrated nested Laplace approximations [@rue2009approximate] and takes days to get accurate answers with Markov chain Monte Carlo via `tmbstan` [@monnahan2018no]
+* Fits into the recently described class of extended latent Gaussian models [@stringer2022fast]
 
@@
 
-## Comparison
+### Laplace approximation
 
-## Conclusions
+* Integrate out variables using a Gaussian approximation in the denominator
+$$
+p(\theta, y) \approx \tilde p_\text{LA}(\theta, y)
+$$
+where
+
+### Adaptive Gauss-Hermite Quadrature
+
+* Approximate integrals by $\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)$ with Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} \in p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target
+
+### Our algorithm
+
+Given C++ user template for $- \log p(y, x, \theta)$:
+
+```{r echo=FALSE, fig.align='center', out.width='100%'}
+knitr::include_graphics("algorithm-flowchart.png")
+```
+
+## Inference comparison
+
+* Compare posterior inferences from `TMB`, `aghq` and `tmbstan` using Kolmogorov-Smirnov tests on posterior marginals
+* Find that...
 
@@
 
-<div id="refs" class="references"></div>
+<font size="6"> <div id="refs" class="references"></div> </font>

### 2023-01-30 src/naomi-simple_compare/comparison.Rmd
aghq hypers debugged so that we do have samples, still need to work out an approach for what to do when k = 1
Update naomi-simple_compare so that it includes a histogram of the first entry of beta_rho -- the parameter that I'm testing Laplace marginals on

@@
 ```{r}
+histogram_plot("beta_rho", i = 1)
 histogram_plot("beta_anc_rho")
@@
 (big_ks <- ks_summary %>%
-  filter(`KS(aghq, tmbstan)` > 0.5, `KS(TMB, tmbstan)` > 0.5) %>%
+  filter(`KS(aghq, tmbstan)` + `KS(TMB, tmbstan)` > 0.5) %>%
   pull(Parameter))

### 2023-01-30 src/naomi-simple_compare/functions.R
Save the beta_rho results to import back into my Laplace development pipeline

@@ -1,5 +1,5 @@
 #' Create a facetted histogram plot of the named parameter using samples from each of the three methods
-histogram_plot <- function(par, i = NULL) {
+histogram_plot <- function(par, i = NULL, return_df = FALSE) {
   if(!is.null(i)) {
     par_name <- paste0(par, "[", i, "]")
 
@@ -30,7 +30,7 @@ histogram_plot <- function(par, i = NULL) {
     pull(sd) %>%
     round(digits = 3)
 
-  ggplot(df_compare, aes(x = samples, fill = method, col = method)) +
+  plot <- ggplot(df_compare, aes(x = samples, fill = method, col = method)) +
     geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +
     theme_minimal() +
     facet_grid(method~.) +
@@ -39,6 +39,10 @@ histogram_plot <- function(par, i = NULL) {
     scale_fill_manual(values = multi.utils::cbpalette()) +
     theme(legend.position = "none") +
     labs(title = par_name, subtitle = paste0("Mean = ", mean, ", SD = ", sd, " (from tmbstan)"))
+
+  if(return_df) {
+    return(list(plot = plot, df = df_compare))
+  }
 }
 
 #' Create a dataframe of samples from TMB, aghq and tmbstan for any parameters starting with par

### 2023-02-04 src/docs_paper/paper.Rmd
Paper edits while waiting for adam to run

@@
 Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
+Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via Template Model Builder \texttt{TMB}.
 We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach which is currently in use, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler.
-By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
+Using data from Malawi, our method provides substantially more accurate inferences than default \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler but substantially faster.
+By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior, enabling inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
 keyword-subclass: | 
@@
 ---
+
+<!-- Most published papers will not exceed 20 pages -->
+<!-- We strongly encourage submission of data sets, computer algorithms and supporting material -->
 
@@
 
-## Algorithm\label{sec:algorithm}
-
 1. Calculate the mode, Hessian at the mode, and lower Cholesky
@@
 We typically set $k = 3$ such that there are $3^m$ nodes in total.
-3. Use this quadrature rule to calculate $\tilde p_{\text{AQ}}(\y)$ as follows
+3. Use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
 $$
@@
 $$
-4. For $x_i \in \{\ldots\}$ and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
+4.
+For $i = 1, \ldots, n$ generate a set of nodes $\mathcal{X}_i$ via a Gauss-Hermite quadrature rule in one-dimension adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ to evaluate each latent field marginal Laplace approximation at.
+5. For $x_i \in \mathcal{X}_i$ and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
 \begin{align*}
@@
 where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$.
-
-5. For each $x_i$ calculate
+6. For $x_i \in \mathcal{X}_i$ calculate
 $$
@@
 $$
+7.
+Given $\{x_i, \tilde p_\text{AQ}(x_i \, | \, \y)\}_{x_i \in \mathcal{X}_i}$ create a spline interpolant to each posterior marginal on the log-scale, from which samples or relevant posterior summaries can be obtained.
 
@@
 The \textsc{R} [@r] code used to produce all results we describe is available at `github.com/athowes/elgm-inf`.
+We used `sf` [@pebesma2018sf] for handling of spatial data, `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
 
@@
 
-AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1).
+AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1), and conducted part of this research while an International Visiting Graduate Student at the University of Waterloo (GRANT).
 AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).

### 2023-02-04 src/naomi-simple_compare/comparison.Rmd
Save the beta_rho results to import back into my Laplace development pipeline
Add adam method to comparison #20

@@
 aghq <- readRDS("depends/aghq.rds")
+adam <- readRDS("depends/adam.rds")
 tmbstan <- readRDS("depends/tmbstan.rds")
@@
   "aghq" = aghq$time,
+  "adam" = adam$time,
   "tmbstan" = tmbstan$time
 )
+
+adam <- adam$adam
 ```
@@
 ```{r}
-histogram_plot("beta_rho", i = 1)
+beta_rho <- histogram_plot("beta_rho", i = 1, return_df = TRUE)
+beta_rho$plot
+saveRDS(beta_rho$df, "beta_rho.rds")
+
 histogram_plot("beta_anc_rho")
@@
 ```{r}
-ks_helper <- function(par) to_ks_df(par) %>% ks_plot(par = par)
+ks_helper <- function(par, ...) {
+  to_ks_df(par) %>% ks_plot(par, ...)
+}
 
@@
 ks_helper("u_rho_x")
+ks_helper("u_rho_x", method1 = "TMB", method2 = "adam")
+
 ks_helper("u_rho_xs")
+ks_helper("u_rho_xs", method1 = "TMB", method2 = "adam")
+
 ks_helper("us_rho_x")
+ks_helper("us_rho_x", method1 = "TMB", method2 = "adam")
+
 ks_helper("us_rho_xs")
+ks_helper("us_rho_xs", method1 = "TMB", method2 = "adam")
+
 ks_helper("u_rho_a")
+ks_helper("u_rho_a", method1 = "TMB", method2 = "adam")
+
 ks_helper("u_rho_as")
+ks_helper("u_rho_as", method1 = "TMB", method2 = "adam")
 
 ks_helper("u_alpha_x")
+ks_helper("u_alpha_x", method1 = "TMB", method2 = "adam")
+
 ks_helper("u_alpha_xs")
+ks_helper("u_alpha_xs", method1 = "TMB", method2 = "adam")
+
 ks_helper("us_alpha_x")
+ks_helper("us_alpha_x", method1 = "TMB", method2 = "adam")
+
 ks_helper("us_alpha_xs")
+ks_helper("us_alpha_xs", method1 = "TMB", method2 = "adam")
+
 ks_helper("u_alpha_a")
+ks_helper("u_alpha_a", method1 = "TMB", method2 = "adam")
+
 ks_helper("u_alpha_as")
+ks_helper("u_alpha_as", method1 = "TMB", method2 = "adam")
+
 ks_helper("u_alpha_xa")
+ks_helper("u_alpha_xa", method1 = "TMB", method2 = "adam")
 
 ks_helper("ui_anc_rho_x")
+ks_helper("ui_anc_rho_x", method1 = "TMB", method2 = "adam")
+
 ks_helper("ui_anc_alpha_x")
+ks_helper("ui_anc_alpha_x", method1 = "TMB", method2 = "adam")
+
 ks_helper("log_or_gamma")
+ks_helper("log_or_gamma", method1 = "TMB", method2 = "adam")
 ```

### 2023-02-06 src/docs_paper/paper.Rmd
Good writing progress: mostly saying what an LGM and ELGM are

@@
     label: e1
-    addressLabel: A
+    addressLabel: A,D
     sepNext: ","
@@
 Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
-Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via Template Model Builder \texttt{TMB}.
+Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package.
 We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, our method provides substantially more accurate inferences than default \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler but substantially faster.
+Using data from Malawi, our method provides substantially more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but substantially faster.
 By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior, enabling inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
@@
 Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they occur only every five years or so, with limited sample size at the district level.
-Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on limited or non-representative samples of the population.
+Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are more real-time but based on limited or non-representative samples of the population.
 To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
@@
 Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
-Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
+Inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
 Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgoodzimmerman2021statistical].
@@
 
+Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package of @bachl2019inlabru, and the INLA within MCMC approach of @gomez2018markov, both of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
+To approximate non-linear predictors, `inlabru` makes iterated use of `R-INLA` using linearisation, extending the scope of INLA to LGMs with some suitably small amount of non-linearity.
+<!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
+<!-- Want to say in particular what the class of models that `inlabru` allows you to fit are -->
+INLA within MCMC is suitable for models which are LGMs, though only conditional on some subset of the parameters being fixed.
+For these models, Metropolis-Hastings style algorithms be defined to update the fixed parameters, with acceptance probabilities calculated using by repeat calls to `R-INLA`.
+It would or would not be possible to fit the Naomi model in these frameworks.
+
 The remainder of this paper is organised as follows.
 Section \ref{sec:naomi} describes a simplified version of the Naomi model that we consider in this paper.
-Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
+Section \ref{sec:elgm} describes how the Naomi model falls within the ELGM framework.
+Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference for ELGMs using simplified INLA and AGHQ.
 As a case-study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:naomi}.
@@
 
-@eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence among pregnant women from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
+@eaton2021naomi specify a joint model linking three small-area estimation models.
 The full model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version of the model which is defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
-Below we provide an overview of the three components in the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
+We consider a simplified version defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
+Below we provide an overview of the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
 A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
@@
 
-Consider a country in sub-Saharan Africa where a household survey has taken place during at time $T_1$.
-Let $x$ index district, $a$ five-year age band, and $s$ sex and.
+Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place during at time $T_1$.
+Let $x$ index district, $a$ five-year age band, and $s$ sex.
 For ease of notation, let $i$ index the finest district-age-sex division included in the model.
 The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
-Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be ART coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
+Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be antiretroviral therapy (ART) coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
 We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
@@
 
+People living with HIV may choose to access ART services outside of the district that they reside in.
 Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
-We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$.
+We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$, and model $\gamma_{x, x'}$ using a multinomial logistic regression model where $\tilde \gamma_{x, x'}$ are the log-odds.
 
@@
 
+# Extended Latent Gaussian models\label{sec:elgm}
+
+Latent Gaussian models (LGMs) [@rue2009approximate] are of the form
+\begin{align*}
+y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
+\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
+\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
+\end{align*}
+where $[n] = \{1, \ldots, n\}$.
+The response variable is $\y = (y)_{i \in [n]}$ with likelihood $p(\y \, | \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1)$, where $\bmeta = (\eta)_{i \in [n]}$.
+Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
+The vector $\btheta_1 \in \mathbb{R}^s$, with $s_1$ assumed small, are additional parameters of the likelihood.
+The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
+The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
+It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-1})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
+Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.
+
+Extended latent Gaussian models (ELGMs) [@stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
+Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ depends on, but may not be equal to, $n$.
+Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
+The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
+Importantly, this allows for the presence of more non-linearity in the model than is allowed in LGMs.
+ELGMs are then of the form
+\begin{align*}
+y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
+\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
+\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n].
+\end{align*}
+
+The Naomi model falls into the ELGM class as follows.
+
 # Fast approximate inference methods\label{sec:fastinferencemethods}
 
-Consider a latent Gaussian model (LGM) of the form
-\begin{alignat*}{2}
-&\text{(Observations)}     &        \y &\sim p(\y \, | \, \x, \btheta), \\
-&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \\
-&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta),
-\end{alignat*}
-where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
-The joint posterior of $(\x, \btheta)$ is given by
+The joint posterior of $(\x, \btheta)$ for an ELGM is given by
 \begin{equation*}
@@
 \end{equation*}
-We consider approximations to the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
+We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
 \begin{align}
-  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
-  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
+  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
+  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
 \end{align}
@@
 2.
-Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $m = \dim(\btheta)$, and $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
-We typically set $k = 3$ such that there are $3^m$ nodes in total.
+Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
+If possible $k \geq 3$ is prefered, though the number of grid points scales exponentially with choice of $k$.
 3. Use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
@@
 4.
-For $i = 1, \ldots, n$ generate a set of nodes $\mathcal{X}_i$ via a Gauss-Hermite quadrature rule in one-dimension adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ to evaluate each latent field marginal Laplace approximation at.
+For $i \in [n]$ generate $l$ nodes $\mathcal{X}_i$ via a Gauss-Hermite quadrature rule with one-dimension adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$.
+A value of $l \geq 4$ is recommended to enable B-spline interpolation.
 5. For $x_i \in \mathcal{X}_i$ and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
@@
 \end{align*}
-where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$.
+where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
 6. For $x_i \in \mathcal{X}_i$ calculate
@@
 7.
-Given $\{x_i, \tilde p_\text{AQ}(x_i \, | \, \y)\}_{x_i \in \mathcal{X}_i}$ create a spline interpolant to each posterior marginal on the log-scale, from which samples or relevant posterior summaries can be obtained.
+Given $\{x_i, \tilde p_\text{AQ}(x_i \, | \, \y)\}_{x_i \in \mathcal{X}_i}$ create a spline interpolant to each posterior marginal on the log-scale, from which samples or relevant posterior marginal summaries may be obtained.
 
@@
 
-Using a single `TMB` template, we fit the simplified Naomi model to data from Malawi using four inferential approaches: (1) empirical Bayes combined with a Gaussian approximation, (2) AGHQ combined with a Gaussian approximation, (3) AGHQ combined with a Laplace approximation and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS).
+Using a common `TMB` template, we fit the simplified Naomi model to data from Malawi using four inferential approaches: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
+Details regarding particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
 The \textsc{R} [@r] code used to produce all results we describe is available at `github.com/athowes/elgm-inf`.
-We used `sf` [@pebesma2018sf] for handling of spatial data, `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
+We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
 
@@
 \small
-\begin{tabularx}{\textwidth}{lll}
+\begin{tabularx}{\textwidth}{p{0.2\linewidth}p{0.75\linewidth}}
 \toprule
- & Run-time & Memory \\
+Inferential method & Details \\
 \midrule
-Empirical Bayes, Gaussian & 0 & 0 \\
-AGHQ, Gaussian & 0 & 0 \\
-AGHQ, Laplace & 0 & 0 \\
-NUTS & 0 & 0 \\
+1. EB, Gaussian & 0, $1000$ samples \\
+2. AGHQ, Gaussian & $k = 1$, $1000$ samples \\
+3. AGHQ, Laplace & $k = 1$, $l = 5$, $1000$ samples \\
+4. NUTS & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$. HMC parameters set to default for \texttt{rstan}. \\
 \bottomrule
 \end{tabularx}
-\label{tab:results}
+\caption{A summary of settings used for each inferential method.}
+\label{tab:inference-methods}
 \small
@@
 
-In order to treat results from NUTS as the gold-standard we assessed MCMC convergence using.
-We ran the algorithm
+MCMC can be used to obtain accurate inferential results only once convergence has been reached and the Markov chain length is sufficiently long.
+We assessed the quality of our MCMC results using the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
+Full details are provided in the appendix.
+We treat these results from NUTS as a gold-standard to which other inferential methods can be compared to.
 
-## Model assessment
+## Model assessment \label{sec:model-assessment}
 
@@
 
-We used three methods to assess the accuracy of posterior distributions produced by each inferential method: (1) Kolmogorov-Smirnov tests, (2) maximum mean discrepancy, and (3) Pareto-smoothed importance sampling.
+We used three methods to assess the accuracy of posterior distributions produced by each inferential method as compared with those from NUTS: (1) Kolmogorov-Smirnov tests, (2) maximum mean discrepancy, and (3) Pareto-smoothed importance sampling.
+
+### Kolmogorov-Smirnov tests
 
@@
 The two-sample Kolmogorov-Smirnov (KS) test statistic is given by the maximum absolute difference between two ECDFs.
-We compare the two KS statistics
-\begin{align*}
-D_\texttt{aghq} &= \sup_\vartheta | F_\texttt{tmbstan}(\vartheta) - F_\texttt{aghq}(\vartheta)|, \\
-D_\texttt{TMB} &= \sup_\vartheta | F_\texttt{tmbstan}(\vartheta) - F_\texttt{TMB}(\vartheta)|.
-\end{align*}
+For each method $\{\texttt{TMB}, \texttt{aghq}, \texttt{adam}\}$ we compare the KS statistics
+\begin{equation*}
+D_\texttt{method} = \sup_\vartheta | F_\texttt{tmbstan}(\vartheta) - F_\texttt{method}(\vartheta)|.
+\end{equation*}
+See a summary of the results in Table 1 and Plot 1, and full results available in the appendix.
 
-See results in Table \ref{tab:results}.
+### Maximum mean discrepancy
 
-\begin{table}[]
-\small
-\begin{tabularx}{\textwidth}{llll}
-\toprule
- & KS & MMD & PSIS \\
-\midrule
-Empirical Bayes, Gaussian & 0 & 0 & 0 \\
-AGHQ, Gaussian & 0 & 0 & 0 \\
-AGHQ, Laplace & 0 & 0 & 0 \\
-NUTS & 0 & 0 & 0 \\
-\bottomrule
-\end{tabularx}
-\label{tab:results}
-\small
-\end{table}
+To write.
+See a summary of the results in Table 2 and Plot 2, and full results available in the appendix.
+
+### Pareto-smoothed importance sampling
+
+To write.
+See a summary of the results in Table 3 and Plot 3, and full results available in the appendix.
 
@@
 
-* We developed an approximate Bayesian inference algorithm to solve a challenging problem in the small-area estimation of HIV in low resource settings
-* Following further testing, we anticipate adding our inference method as an option for 
-* The flexibility of our method implementation, including compatibility with any `TMB` \textsc{C++} template, allows broader use, as well as investigation of, deterministic inference methods than had previously been possible.
-We are excited about exploration of its accuracy for the following challenging model structures which could not previously be fit: list of model types here.
-* Our approach is similar to that of `inlabru` [@bachl2019inlabru] in the following ways, and different in the following other ways
-* We demonstrated a Bayesian workflow for deterministic inference methods
+We developed an approximate Bayesian inference algorithm to solve a challenging problem in small-area estimation of HIV for low resource settings.
+Our method is demonstrated to be more accurate than the EB Gaussian approximation and substantially faster than NUTS for the simplified Naomi model in Malawi (Section \ref{sec:results}).
+We anticipate that our method could be added to the Naomi software as an alternative option to the, still substantially faster, EB Gaussian approximation.
+Analysts might quickly iterate over model options using the faster, less accurate inference approach, only switching to the slower, more accurate approach once they are happy with the results.
+
+We provide a flexible implementation, which builds on the `aghq` \textsc{R} package.
+In doing so, we hope our work enables use of INLA for ELGMs in applied settings, as well as further methodological exploration of the algorithms accuracy and limits.
+Among the ELGMs structures of greatest interest are: aggregated Gaussian process models [@nandi2020disaggregation].
+Although our method is designed for ELGMs, it may even be used outside this class, and is compatible with any model with a `TMB` \textsc{C++} template.
+
+We would be excited to see statistical theory for our algorithm by extension of Theorem 1 of @stringer2022fast.
+
+In our case study we demonstrated a Bayesian workflow for deterministic inference methods.
+We retained the ability to draw samples from the posterior distributions of interest, facilitating use of posterior predictive checks (Section \ref{sec:model-assessment}).
 

### 2023-02-07 src/naomi-simple_compare/comparison.Rmd
Add ECDF plot to histogram

@@
 ```{r}
-beta_rho <- histogram_plot("beta_rho", i = 1, return_df = TRUE)
+beta_rho <- histogram_and_ecdf("beta_rho", i = 1, return_df = TRUE)
 beta_rho$plot
@@
 
-histogram_plot("beta_anc_rho")
+histogram_and_ecdf("beta_anc_rho")
+```
+
+```{r}
+par <- "beta_rho"
+i <- 1
+par_name <- paste0(par, "[", i, "]")
+
+df_compare <- rbind(
+  data.frame(method = "TMB", samples = as.numeric(tmb$fit$sample[[par]][i, ])),
+  data.frame(method = "aghq", samples = as.numeric(aghq$quad$sample[[par]][i, ])),
+  data.frame(method = "adam", samples = as.numeric(adam$sample[[par]][i, ])),
+  data.frame(method = "tmbstan", samples = as.numeric(unlist(rstan::extract(tmbstan$mcmc, pars = par)[[par]][, i])))
+)
+
+grid <- seq(from = min(df_compare$samples), to = max(df_compare$samples), by = 0.1)
+
+tmb_ecdf <- stats::ecdf(filter(df_compare, method == "TMB") %>% pull(samples))
+tmb_ecdf_df <- data.frame(x = grid, ecdf = tmb_ecdf(grid), method = "TMB")
+
+aghq_ecdf <- stats::ecdf(filter(df_compare, method == "aghq") %>% pull(samples))
+aghq_ecdf_df <- data.frame(x = grid, ecdf = aghq_ecdf(grid), method = "aghq")
+
+adam_ecdf <- stats::ecdf(filter(df_compare, method == "adam") %>% pull(samples))
+adam_ecdf_df <- data.frame(x = grid, ecdf = adam_ecdf(grid), method = "adam")
+
+tmbstan_ecdf <- stats::ecdf(filter(df_compare, method == "tmbstan") %>% pull(samples))
+tmbstan_ecdf_df <- data.frame(x = grid, ecdf = tmbstan_ecdf(grid), method = "tmbstan")
+
+bind_rows(tmb_ecdf_df, aghq_ecdf_df, adam_ecdf_df, tmbstan_ecdf_df) %>%
+  mutate(method = forcats::fct_relevel(method, levels = c("TMB", "aghq", "adam", "tmbstan"))) %>%
+  ggplot(aes(x = x, y = ecdf, col = method)) +
+    geom_line() +
+    scale_color_manual(values = multi.utils::cbpalette()) +
+    labs(x = "", y = "ECDF") +
+    theme_minimal()
+
 ```

### 2023-02-07 src/naomi-simple_compare/functions.R

@@ -1,5 +1,6 @@
 #' Create a facetted histogram plot of the named parameter using samples from each of the four methods
-histogram_plot <- function(par, i = NULL, return_df = FALSE) {
+#' As well as an ECDF plot
+histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
   if(!is.null(i)) {
     par_name <- paste0(par, "[", i, "]")
 
@@ -20,6 +21,8 @@ histogram_plot <- function(par, i = NULL, return_df = FALSE) {
     )
   }
 
+  df_compare <- mutate(df_compare, method = forcats::fct_relevel(method, levels = c("TMB", "aghq", "adam", "tmbstan")))
+
   mean <- df_compare %>%
     filter(method == "tmbstan") %>%
     summarise(mean = mean(samples)) %>%
@@ -32,7 +35,7 @@ histogram_plot <- function(par, i = NULL, return_df = FALSE) {
     pull(sd) %>%
     round(digits = 3)
 
-  plot <- ggplot(df_compare, aes(x = samples, fill = method, col = method)) +
+  histogram_plot <- ggplot(df_compare, aes(x = samples, fill = method, col = method)) +
     geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +
     theme_minimal() +
     facet_grid(method~.) +
@@ -41,6 +44,29 @@ histogram_plot <- function(par, i = NULL, return_df = FALSE) {
     scale_fill_manual(values = multi.utils::cbpalette()) +
     theme(legend.position = "none") +
     labs(title = par_name, subtitle = paste0("Mean = ", mean, ", SD = ", sd, " (from tmbstan)"))
+
+  tmb_ecdf <- stats::ecdf(filter(df_compare, method == "TMB") %>% pull(samples))
+  tmb_ecdf_df <- data.frame(x = grid, ecdf = tmb_ecdf(grid), method = "TMB")
+
+  aghq_ecdf <- stats::ecdf(filter(df_compare, method == "aghq") %>% pull(samples))
+  aghq_ecdf_df <- data.frame(x = grid, ecdf = aghq_ecdf(grid), method = "aghq")
+
+  adam_ecdf <- stats::ecdf(filter(df_compare, method == "adam") %>% pull(samples))
[…]

### 2023-02-07 src/naomi-simple_compare/comparison.Rmd
Fixes, tabbed histogram ECDF plots, tabbed KS plots

--- 2023-01-30 src/naomi-simple_compare/comparison.Rmd
@@
 ```{r}
-unique(names(tmb$fit$obj$env$par))
+pars <- unique(names(tmb$fit$obj$env$par))
 ```
@@
   "aghq" = aghq$time,
+  "adam" = adam$time,
   "tmbstan" = tmbstan$time
 )
+
+adam <- adam$adam
 ```
 
-# Histogram
+# Histograms {.tabset .tabset-fade}
 
 ```{r}
-histogram_plot("beta_rho", i = 1)
-histogram_plot("beta_anc_rho")
+beta_rho <- histogram_and_ecdf("beta_rho", i = 1, return_df = TRUE)
+saveRDS(beta_rho$df, "beta_rho.rds")
 ```
 
-# KS
-
-## Individual parameters
+## `beta`
 
 ```{r}
-ks_helper <- function(par) to_ks_df(par) %>% ks_plot(par = par)
+lapply(pars[stringr::str_starts(pars, "beta")], histogram_and_ecdf)
+```
 
+## `logit`
+
+```{r}
+lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf)
+```
+
+## `log_sigma`
+
+```{r}
+lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
+```
+
+# KS plots
+
+## Individual parameters {.tabset .tabset-fade}
+
+```{r}
+ks_helper <- function(par, ...) {
+  to_ks_df(par) %>% ks_plot(par, ...)
+}
+```
+
+### `beta`
+
+```{r}
 ks_helper("beta")
+ks_helper("beta", method1 = "TMB", method2 = "adam")
+```
+
+### `logit`
+
+```{r}
 ks_helper("logit")
+ks_helper("logit", method1 = "TMB", method2 = "adam")
+```
+
+### `log_sigma`
+
+```{r}
 ks_helper("log_sigma")
+ks_helper("log_sigma", method1 = "TMB", method2 = "adam")
+```
 
+### `u_rho_x`
+
+```{r}
 ks_helper("u_rho_x")
+ks_helper("u_rho_x", method1 = "TMB", method2 = "adam")
+```
+
+### `u_rho_xs`
+
+```{r}
 ks_helper("u_rho_xs")
+ks_helper("u_rho_xs", method1 = "TMB", method2 = "adam")
+```
+
+### `us_rho_x`
+
+```{r}
 ks_helper("us_rho_x")
+ks_helper("us_rho_x", method1 = "TMB", method2 = "adam")
+```
+
+### `us_rho_xs`
+
+```{r}
 ks_helper("us_rho_xs")
+ks_helper("us_rho_xs", method1 = "TMB", method2 = "adam")
+```
+
+### `u_rho_a`
+
+```{r}
 ks_helper("u_rho_a")
+ks_helper("u_rho_a", method1 = "TMB", method2 = "adam")
+```
+
+### `u_rho_as`
+
+```{r}
 ks_helper("u_rho_as")
+ks_helper("u_rho_as", method1 = "TMB", method2 = "adam")
+```
 
+### `u_alpha_x`
+
+```{r}
 ks_helper("u_alpha_x")
+ks_helper("u_alpha_x", method1 = "TMB", method2 = "adam")
+```
+
+### `u_alpha_xs`
+
+```{r}
 ks_helper("u_alpha_xs")
+ks_helper("u_alpha_xs", method1 = "TMB", method2 = "adam")
+```
+
+### `us_alpha_x`
+
+```{r}
 ks_helper("us_alpha_x")
+ks_helper("us_alpha_x", method1 = "TMB", method2 = "adam")
+```
+
+### `us_alpha_xs`
+
+```{r}
 ks_helper("us_alpha_xs")
+ks_helper("us_alpha_xs", method1 = "TMB", method2 = "adam")
+```
+
+### `u_alpha_a`
+
+```{r}
 ks_helper("u_alpha_a")
+ks_helper("u_alpha_a", method1 = "TMB", method2 = "adam")
+```
+
+### `u_alpha_as`
+
+```{r}
 ks_helper("u_alpha_as")
+ks_helper("u_alpha_as", method1 = "TMB", method2 = "adam")
+```
+
+### `u_alpha_xa`
+
+```{r}
 ks_helper("u_alpha_xa")
+ks_helper("u_alpha_xa", method1 = "TMB", method2 = "adam")
+```
 
+### `ui_anc_rho_x`
+
+```{r}
 ks_helper("ui_anc_rho_x")
+ks_helper("ui_anc_rho_x", method1 = "TMB", method2 = "adam")
+```
+
+### `ui_anc_alpha_x`
+
+```{r}
 ks_helper("ui_anc_alpha_x")
+ks_helper("ui_anc_alpha_x", method1 = "TMB", method2 = "adam")
+```
+
+### `log_or_gamma`
+
+```{r}
 ks_helper("log_or_gamma")
+ks_helper("log_or_gamma", method1 = "TMB", method2 = "adam")
 ```
@@
     "Parameter" = "par",
+    "KS(adam, tmbstan)" = "adam",
     "KS(aghq, tmbstan)" = "aghq",
@@
 
-Let's look into this further by plotting the histograms:
-
-```{r}
-lapply(big_ks, histogram_plot)
-```
-
 # MMD

### 2023-02-07 src/naomi-simple_compare/comparison.Rmd
Add histograms and ECDF for all latent field elements too

@@
 lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
+```
+
+## `u_rho_x`
+
+```{r}
+histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
+histogram_and_ecdf_helper("u_rho_x")
+```
+
+## `u_rho_x`
+
+```{r}
+histogram_and_ecdf_helper("u_rho_x")
+```
+
+## `us_rho_x`
+
+```{r}
+histogram_and_ecdf_helper("us_rho_x")
+```
+
+## `u_rho_xs`
+
+```{r}
+histogram_and_ecdf_helper("u_rho_xs")
+```
+
+## `us_rho_xs`
+
+```{r}
+histogram_and_ecdf_helper("us_rho_xs")
+```
+
+## `u_rho_as`
+
+```{r}
+histogram_and_ecdf_helper("u_rho_as")
+```
+
+## `u_alpha_x`
+
+```{r}
+histogram_and_ecdf_helper("u_alpha_x")
+```
+
+## `us_alpha_x`
+
+```{r}
+histogram_and_ecdf_helper("us_alpha_x")
+```
+
+## `u_alpha_xs`
+
+```{r}
+histogram_and_ecdf_helper("u_alpha_xs")
+```
+
+## `us_alpha_xs`
+
+```{r}
+histogram_and_ecdf_helper("us_alpha_xs")
+```
+
+## `u_alpha_a`
+
+```{r}
+histogram_and_ecdf_helper("u_alpha_a")
+```
+
+## `u_alpha_as`
+
+```{r}
+histogram_and_ecdf_helper("u_alpha_as")
+```
+
+## `u_alpha_xa`
+
+```{r}
+histogram_and_ecdf_helper("u_alpha_xa")
+```
+
+## `ui_lambda_x`
+
+```{r}
+histogram_and_ecdf_helper("ui_lambda_x")
+```
+
+## `ui_anc_rho_x`
+
+```{r}
+histogram_and_ecdf_helper("ui_anc_rho_x")
+```
+
+## `ui_anc_alpha_x`
+
+```{r}
+histogram_and_ecdf_helper("ui_anc_alpha_x")
+```
+
+## `log_or_gamma`
+
+```{r}
+histogram_and_ecdf_helper("log_or_gamma")
 ```

### 2023-02-08 src/docs_paper/paper.Rmd
Start listing the ways in which Naomi is an ELGM

@@
 The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
-Importantly, this allows for the presence of more non-linearity in the model than is allowed in LGMs.
+Importantly, this allows for the presence of more non-linearity.
 ELGMs are then of the form
@@
 
-The Naomi model falls into the ELGM class as follows.
+Naomi is not an LGM, and instead falls into the ELGM class, for the following reasons:
+
+1. In the household survey component, the HIV incidence rate depends on the HIV prevalence and ART coverage linear predictors.
+2. In the ANC testing component, the HIV prevalence and ART coverage depend upon the household survey component.
+Specifically, $|\mathcal{J}_i| = 2$ such that for $\theta \in \{\rho, \alpha\}$ then
+$$
+\mu_i = g_i(\eta^\theta_i, \eta^{\theta^{\text{ANC}}}_i) = \text{logit}^{-1}(\eta^\theta_i + \eta^{\theta^{\text{ANC}}}_i)
+$$
 

### 2023-02-09 src/naomi-simple_compare/comparison.Rmd
Update comparison, and split beta parameters which are really multiple up in histogram_and_ecdf() #20

@@
 
-We compare the inference results from `TMB`, `aghq` and `tmbstan`.
+We compare the inference results from `TMB`, `aghq`, `adam`, and `tmbstan`.
 Import these inference results as follows:
@@
 
-## `beta`
+## `beta_rho`
 
 ```{r}
-lapply(pars[stringr::str_starts(pars, "beta")], histogram_and_ecdf)
+histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
+histogram_and_ecdf_helper("beta_rho")
+```
+
+## `beta_alpha`
+
+```{r}
+histogram_and_ecdf_helper("beta_alpha")
+```
+
+## `beta_lambda`
+
+```{r}
+histogram_and_ecdf_helper("beta_lambda")
+```
+
+## `beta_anc_rho`
+
+```{r}
+histogram_and_ecdf("beta_anc_rho")
+```
+
+## `beta_anc_alpha`
+
+```{r}
+histogram_and_ecdf("beta_anc_alpha")
 ```
@@
 ```{r}
-histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
 histogram_and_ecdf_helper("u_rho_x")

### 2023-02-10 src/naomi-simple_model-checks/model-checks.Rmd
Trying to move forward to coverage assessment

@@
 * Use existing or own functions to produce coverage histograms and ECDF difference plots
+
+Discussion talking to Rachel:
+
+* Can use samples from `tmb$fit$sample`
+* Issue about one survey or two. Inclusion of two for the Malawi demo is more of a special case
+
+Have the following samples from `rho_t1_out`:
+
+```{r}
+str(tmb$fit$sample$rho_t1_out)
+```
+
+Want to somehow line these up with the data from `tmb$outputs$inputs_outputs` or `tmb$outputs$indicators`.
+Should be looking for some kind of `mf_out` object:
+
+```{r}
+str(tmb$outputs$fit)
+```

### 2023-02-10 src/naomi-simple_compare/comparison.Rmd
Add KS summary plot split by hyperparameters and latent, move to jitterplot rather than boxplot, save KS summary table

@@
 
-ggplot(ks_summary, aes(x = `KS(TMB, tmbstan)`, y = `KS(aghq, tmbstan)`)) +
-  geom_point(alpha = 0.5) +
+r <- adam$quad$obj$env$random
+x_names <- names(adam$quad$obj$env$par[r])
+theta_names <- names(adam$quad$obj$env$par[-r])
+
+dict <- data.frame(
+  Parameter = c(x_names, theta_names),
+  Type = c(rep("Latent field", length(x_names)), rep("Hyperparameters", length(theta_names)))
+)
+
+ks_summary <- ks_summary %>%
+  left_join(dict, by = "Parameter")
+
+summary_ks_plot <- function(ks_summary, method1, method2) {
+  ks_method1 <- paste0("KS(", method1, ", tmbstan)")
+  ks_method2 <- paste0("KS(", method2, ", tmbstan)")
+
+  scatterplot <- ggplot(ks_summary, aes(x = .data[[ks_method1]], y = .data[[ks_method2]])) +
+    geom_point(alpha = 0.2) +
+    annotate(geom = "polygon", x = c(-Inf, Inf, Inf), y = c(-Inf, Inf, -Inf), fill = multi.utils::cbpalette()[3], alpha = 0.1) +
+    annotate(geom = "polygon", x = c(-Inf, Inf, -Inf), y = c(-Inf, Inf, Inf), fill = multi.utils::cbpalette()[5], alpha = 0.1) +
     xlim(0, 1) +
@@
     geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
-  labs(x = "KS(TMB, tmbstan)", y = "KS(aghq, tmbstan)") +
+    labs(x = ks_method1, y = ks_method2) +
     theme_minimal()
 
-ks_summary %>%
-  mutate(`KS difference` = `KS(TMB, tmbstan)` - `KS(aghq, tmbstan)`) %>%
-  ggplot(aes(x = `KS difference`)) +
-    geom_boxplot(width = 0.5) +
-    coord_flip() +
-    labs(x = "KS(TMB, tmbstan) - KS(aghq, tmbstan)") +
+  ks_summary[["KS difference"]] <- ks_summary[[ks_method1]] - ks_summary[[ks_method2]]
+
+  jitterplot <- ggplot(ks_summary, aes(x = Type, y = `KS difference`)) +
+    geom_jitter(width = 0.1, alpha = 0.2) +
+    labs(x = "", y = paste0(ks_method1, " - ", ks_method2)) +
     theme_minimal()
+
+  scatterplot + jitterplot
+}
+
+summary_ks_plot(ks_summary, "TMB", "aghq")
+summary_ks_plot(ks_summary, "TMB", "adam")
+summary_ks_plot(ks_summary, "aghq", "adam")
+
+(ks_summary_out <- ks_summary %>%
+  group_by(Type) %>%
+  summarise(
+    TMB = mean(`KS(TMB, tmbstan)`),
+    aghq = mean(`KS(aghq, tmbstan)`),
+    adam = mean(`KS(adam, tmbstan)`)
+  ))
+```
+
+Save some things for output:
+
+```{r}
+pdf("ks_summary_tmb_adam.pdf", h = 4, w = 6.25)
+summary_ks_plot(ks_summary, "TMB", "adam")
+dev.off()
+
+saveRDS(ks_summary_out, "ks_summary_out.rds")
 ```

### 2023-02-10 src/naomi-simple_compare/functions.R

@@ -70,7 +70,7 @@ histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
     guides(col = "none") +
     theme_minimal()
 
-  plot <- cowplot::plot_grid(histogram_plot, ecdf_plot, rel_widths = c(0.5, 0.5))
+  plot <- histogram_plot + ecdf_plot
 
   if(return_df) {
     return(list(plot = plot, df = df_compare))
@@ -139,14 +139,13 @@ ks_plot <- function(ks_df, par, method1 = "TMB", method2 = "aghq") {
   wide_ks_df$ks_diff <- wide_ks_df[[method1]] - wide_ks_df[[method2]]
   mean_ks_diff <- mean(wide_ks_df$ks_diff)
 
-  boxplot <- wide_ks_df %>%
-    ggplot(aes(x = ks_diff)) +
-    geom_boxplot(width = 0.5) +
-    coord_flip() +
+  jitterplot <- wide_ks_df %>%
+    ggplot(aes(x = "", y = ks_diff)) +
+    geom_jitter(width = 0.1, alpha = 0.5) +
     labs(
       title = paste0("Mean KS difference is ", mean_ks_diff),
       subtitle = paste0(">0 then ", method1, " more different to tmbstan, <0 then ", method2, " more different"),
-      x = paste0("KS(", method1, ", tmbstan) - KS(", method2,", tmbstan)")
+      x = "", y = paste0("KS(", method1, ", tmbstan) - KS(", method2,", tmbstan)")
     ) +
     theme_minimal()
 
@@ -166,5 +165,5 @@ ks_plot <- function(ks_df, par, method1 = "TMB", method2 = "aghq") {
     ) +
     theme_minimal()
 
-  cowplot::plot_grid(scatterplot, boxplot, ncol = 2, rel_widths = c(1.3, 1))
+  scatterplot + jitterplot
 }

### 2023-02-10 src/docs_paper/paper.Rmd
Import KS plot and table into main text

@@
 
+```{r echo = FALSE}
+options(scipen = 100)
+knitr::opts_chunk$set(
+  echo = FALSE,
+  warning = FALSE,
+  message = FALSE,
+  dpi = 320,
+  cache = TRUE,
+  out.width = "95%",
+  fig.align = 'center'
+)
+```
+
 <!-- Most published papers will not exceed 20 pages -->
@@
 
-Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place during at time $T_1$.
+Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place at time $T_1$.
 Let $x$ index district, $a$ five-year age band, and $s$ sex.
@@
 Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be antiretroviral therapy (ART) coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
-We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
+Independent mixed effects models for HIV prevalence and ART coverage in the general population are specified on the logit scale such that
 \begin{align*}
@@
 for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
-For the HIV incidence rate we use a mixed effects model on the log scale
-$$
-\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}),
-$$
-where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
+For the HIV incidence rate we use a mixed effects model on the log scale $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
 Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
@@
 where $j$ indexes individuals across all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
-The observed number of indicator cases is then
-$$
-Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
-$$
-For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood
-$$
-Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}),
-$$
-where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
+The observed number of indicator cases is then $Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$.
+For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
 \begin{align*}
@@
 
-Let $\dot A_\mathcal{I}$ be the number of people receiving ART.
+Let $\dot A_\mathcal{I} = \sum_{x \sim x', x = x'} \dot A_{x', x}$ be the number of people receiving ART.
+Model $\dot A_\mathcal{I} \sim \mathcal{N}(\tilde A_\mathcal{I}, \sigma^{\tilde A_\mathcal{I}})$ using a normal distribution with mean $\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x}$ and standard deviation $\sigma^{\tilde A_\mathcal{I}}) = \sqrt{\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x} (1 - \pi_{x', x})}$.
 
@@
 
-Using a common `TMB` template, we fit the simplified Naomi model to data from Malawi using four inferential approaches: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
+Using one `TMB` template (available in the appendix), we fit the simplified Naomi model to data from Malawi using four inferential approaches: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
 Details regarding particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
-The \textsc{R} [@r] code used to produce all results we describe is available at `github.com/athowes/elgm-inf`.
+
+The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
 We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
@@
 \midrule
-1. EB, Gaussian & 0, $1000$ samples \\
+1. EB, Gaussian & $1000$ samples \\
 2. AGHQ, Gaussian & $k = 1$, $1000$ samples \\
@@
 See a summary of the results in Table 1 and Plot 1, and full results available in the appendix.
+
+```{r}
+knitr::include_graphics("depends/ks_summary_tmb_adam.png")
+```
+
+```{r}
+ks_summary <- readRDS("depends/ks_summary_out.rds")
+
+gt::gt(ks_summary)
+```
 

### 2023-02-11 src/docs_paper/appendix.Rmd
Iterate appendix forward somewhat

@@
   pdf_document:
-    toc: true
-    number_sections: true
-    keep_tex: true
+    toc: yes
+    number_sections: yes
+    keep_tex: yes
     includes:
@@
 ---
+
+```{r echo = FALSE}
+options(scipen = 100)
+
+knitr::opts_chunk$set(
+  echo = FALSE,
+  warning = FALSE,
+  message = FALSE,
+  dpi = 320,
+  cache = TRUE,
+  out.width = "95%",
+  fig.align = 'center'
+)
+```
 
@@
 
-# Modified Naomi model description
+# Simplified Naomi model description
 
-## Notation
+In this section we describes more fully the simplified Naomi model considered in the main text.
+To do so, we draw substantially from the supplementary material of @eaton2021naomi.
+
+## Background
 
@@
 
-Consider a given country in sub-Saharan Africa.
-Let $x$ refer to a district located within the Spectrum region $R_x$.
-We restrict our attention to the time period $t = T_1$ of the most recent national household survey with HIV testing.
+Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest at time $t = T_1$
+Let $x$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
 Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
@@
 We index the following quantities by district, sex and age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$, annual HIV incidence rate $\lambda_{x, s, a}$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a}$.
-Sometimes data are observed at an aggregate level, rather than the modelled level of granularity.
+Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
 In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.
@@
 
-Consider a generic random effect $u$.
+Let $u$ be a generic random effect.
 We use use structured random effects to share information across units assessed as being similar, such as neighbouring districts or adjacent age-bands.
@@
 where $u_{-i}$ is $u$ with the $i$th unit removed, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
-We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands regarding the ICAR model.
+We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
 For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v^\star$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w^\star$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
@@
 
+### Complex survey design
+
+We assume the household survey was run according to a complex survey design where each individual $j \in U$ has probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$.
+Suppose we observe an outcome $y_j$ for $j \in S$.
+Let $w_j = 1 / \pi_j$ be design weights, then the weighted mean
+$$
+\hat y = \frac{\sum_{j \in S} w_j y_j}{\sum_{j \in S} w_j}
+$$
+Define Kish ESS.
+We make these computations using the `survey` \textsc{R} package [@lumley2004analysis].
+
 ## Process specification
 
-### HIV prevalence
+### HIV prevalence \label{sec:hiv-prev}
 
-We model HIV prevalence on the logit scale using
+We model HIV prevalence on the logit scale using the linear predictor
 $$
@@
 $$
-where $\beta^\rho_0$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{UNKNOWN}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
-We use the prior distributions $\mathcal{N}(0, 5)$ for the fixed effects $\beta^\rho_0$ and $\beta_{s}^{\rho, s = \text{M}}$, $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.
+where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
+We use the prior distributions $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.
 
-### ART coverage
+### ART coverage \label{sec:art-cov}
 
-We model ART coverage by
+We model ART coverage using the linear predictor
 $$
@@
 $$
-with terms and priors analogous to the HIV prevalence model above.
+with terms and priors analogous to the HIV prevalence process model in Section \ref{sec:hiv-prev} above.
 
@@
 
-We model HIV incidence rate by
-$$
-\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda 
-$$
-where $\beta^\lambda_0$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women, $\rho_{x}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
-$$
-\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}},
-$$
-$\alpha_{x}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
-$$
-\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}},
-$$
-$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ via the EPP model, $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output
+We model HIV incidence rate using the linear predictor
+\begin{equation}
+\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda, \label{eq:hiv-inc}
+\end{equation}
+where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women, $\rho_{x}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences
+\begin{equation}
+\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}, \label{eq:rho-agg}
+\end{equation}
+$\alpha_{x}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages
+\begin{equation}
+\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}, \label{eq:alpha-agg}
+\end{equation}
+$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output.
+Note that the HIV incidence rate process (Equation \ref{eq:hiv-inc}) depends on those of the HIV prevalence (Section \ref{sec:hiv-prev}) and the ART coverage (Section \ref{sec:art-cov}) via the adult aggregate indicators (Equations \ref{eq:rho-agg} and \ref{eq:alpha-agg}) violating the conditions required being within the latent Gaussian model (LGM) class as defined by @rue2009approximate.
 
@@
 
-HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women modelled with logit-linear models
+HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit-scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
 \begin{align}
-\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}} \\
-\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}},
+\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \label{eq:anc-hiv-prev} \\
+\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}} \label{eq:anc-art-cov},
 \end{align}
-
-where $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$, $\beta^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the average differences between population and ANC outcomes after removing the offset, and $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ are district random effects.
+where, for $\theta \in \{\rho, \alpha\}$, $\beta^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, 5)$ are the average differences between population and ANC outcomes, $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ are IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$, and $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$.
+Dependence here too on $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ in Equations \ref{eq:anc-hiv-prev} and \ref{eq:anc-art-cov} violates the conditions required to be a LGM.
 
@@
 
+```{r}
+log_odds <- function(p) log(p / (1 - p))
+logistic <- function(x) 1 / (1 + exp(-x))
+# 100 * round(logistic(-4), 3)
+```
+
 Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
-We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.
-
-To model $\gamma_{x, x'}$ for $x \sim x'$ we use a multinomial logistic regression model where $\tilde \gamma_{x, x'}$ is the log odds ratio of seeking ART
-\begin{align}
-\tilde \gamma_{x, x'} &= \tilde \gamma_0 + u_x^{\tilde \gamma} \\ 
-\tilde \gamma_0 &= -4 \\
-u_x^{\tilde \gamma} &\sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})
-\end{align}
-where $\tilde \gamma_0$ is an intercept specifying the prior mean for the log odds of seeking ART in each neighbouring district compared to the home district, and $u_x^{\tilde \gamma}$ are district random effects.
+We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district or its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation, and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.
+To model $\gamma_{x, x'}$ for $x \sim x'$ we use a multinomial logistic regression model, based on the log-odds ratios
+\begin{equation*}
+\tilde \gamma_{x, x'} = \log \left( \frac{\gamma_{x, x'}}{1 - \gamma_{x, x'}} \right) = \tilde \gamma_0 + u_x^{\tilde \gamma}, \label{eq:log-or}
+\end{equation*}
+where $\tilde \gamma_0 = -4$ is a fixed intercept, and $u_x^{\tilde \gamma} \sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})$ are district random effects with $\sigma_X^{\tilde \gamma} \sim \mathcal{N}^+(0, 2.5)$.
+Note that Equation \ref{eq:log-or} does not depend on $x'$, such that $\gamma_{x, x'}$ is only a function of $x$.
+Choice of $\tilde \gamma_0 = -4$ implies a prior on $\gamma_{x, x'}$ of 1.8%, such that $(100 - 1.8 \times \text{ne}(x))\%$ of ART clients in district $x$ obtain treatment in their home district, a-priori.
 We fix $\tilde \gamma_{x, x} = 0$ and recover the multinomial probabilities using the softmax
 $$
-\gamma_{x, x'} = \frac{\exp(\tilde \gamma_{x, x'})}{\sum_{x^\star \in \{x, \text{ne}(x)\}} \exp(\tilde \gamma_{x, x^\star})}
+\gamma_{x, x'} = \frac{\exp(\tilde \gamma_{x, x'})}{\sum_{x^\star \in \{x, \text{ne}(x)\}} \exp(\tilde \gamma_{x, x^\star})}.
 $$
-Then, given the number of PLHIV on ART $A_{x, s, a} = N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}$, the number of ART clients who reside in district $x$ and obtain ART in district $x'$ are
-$$
-A_{x, x', s, a} = A_{x, s, a} \cdot \gamma_{x, x'},
-$$
-and the total attending ART facilities in district $x'$ are
+Given the total number of PLHIV on ART $A_{x, s, a} = N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}$, the number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a} = A_{x, s, a} \cdot \gamma_{x, x'}$, and the total attending ART facilities in district $x'$ are
 $$
@@
 
-Let $\nu$ be a household survey occurring at $T_1$, furnishing weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}, \nu}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes $M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu}$, and observed number of cases
+For $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}}$, and observed number of cases
 $$
-Y^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} = M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}, \nu}.
+y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}}.
 $$
 
-For HIV prevalence and ART coverage we use binomial working likelihoods as follows
+For HIV prevalence and ART coverage we use the binomial working likelihoods
 \begin{align}
-Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \\
 \rho_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
@@
 \begin{align}
-Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
+y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \\
 \alpha_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
@@
 \begin{align}
-Y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu}, \kappa_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
-\kappa_{x, s, a} &= 1 - \exp(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T ) \\
-\Omega_T &\sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T}) \\
-\beta_T &\sim \mathcal{N}(\beta_{T_0}, \sigma^{\beta_T})
+y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \\
+\kappa_{x, s, a} &= 1 - \exp(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T ),
 \end{align}
-where $\kappa_{x, s, a}$ are the predicted proportion recently infected among HIV positive persons, $\Omega_T$ is the mean duration of recent infection (MDRI), and $\beta_T$ is the false recent ratio (FRR).
+where $\kappa_{x, s, a}$ are the predicted proportion recently infected among HIV positive persons, $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection (MDRI), and $\beta_T \sim \mathcal{N}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio (FRR).
 We use an informative prior on $\Omega_T$ based on the characteristics of the recent infection testing algorithm (RITA).
@@
 We include ANC testing data for the year of the most recent survey $Y[T_1]$.
-Let $W^\text{ANC}_{\{x\}, Y[t]}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}, Y[t]}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}, Y[t]}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}, Y[t]}$ the number of ANC clients already on ART prior to first ANC, such that $W^\text{ANC}_{x, Y[t]} \geq X^\text{ANC}_{x, Y[t]} \geq Y^\text{ANC}_{x, Y[t]} \geq Z^\text{ANC}_{x, Y[t]}$
-When ANC testing data are only available for part of a given year, we denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$.
+Let $W^\text{ANC}_{\{x\}, Y[t]}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}, Y[t]}$ the number of those with ascertained status, $y^\text{ANC}_{\{x\}, Y[t]}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}, Y[t]}$ the number of ANC clients already on ART prior to first ANC, such that $W^\text{ANC}_{x, Y[t]} \geq X^\text{ANC}_{x, Y[t]} \geq y^\text{ANC}_{x, Y[t]} \geq Z^\text{ANC}_{x, Y[t]}$
+When ANC testing data are only available for part of a given year, we denote $m^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$.
 
@@
 
-# \textsc{C++} template
+# \textsc{C++} `TMB` user template
 

### 2023-02-12 src/docs_paper/paper.Rmd
Paper edits

@@
   | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa.
-Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
+Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage, are jointly modelled using both household survey data and routinely reported health system data.
 Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package.
 We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, our method provides substantially more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but substantially faster.
-By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior, enabling inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
+Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but faster to run.
+By extending the \texttt{aghq} \textsc{R} package we facilitate easy, flexible use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
+In doing so, we enable inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
 keyword-subclass: | 
@@
 options(scipen = 100)
+
 knitr::opts_chunk$set(
@@
 Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they occur only every five years or so, with limited sample size at the district level.
-Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are more real-time but based on limited or non-representative samples of the population.
+Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in more real-time but based on limited or non-representative samples of the population.
 To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
-Simultaneous modelling of the data mitigates the limitations of any single source, increases statistical power, and prompts investigation into any conflicting information.
+Modelling multiple data sources jointly mitigates the limitations of any single source, increases statistical power, and prompts investigation into possibly conflicting information between sources.
 Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates in a yearly process supported by UNAIDS.
@@
 
-The use case requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
+The practical requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
 Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
@@
 Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
-Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
-This has motivated us to look for a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.
+Although this approach has favourable computational properties, it would be preferable to account more fully for hyperparameter uncertainty than is possible in an empirical Bayes framework.
+This has motivated us to look for an approach closer to full Bayesian inference, which is flexible enough to be compatible with the model, and fast enough to be run in production by country teams.
 
-To obtain fast, accurate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
-INLA is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
+To obtain fast, accurate Bayesian inferences for the Naomi model we develop an inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
+INLA is an approximate inference approach based on nested Laplace approximations and numerical quadrature.
 The key innovation of @rue2009approximate is an approximation which enables accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
 Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
-This extension facilitates inference for models like Naomi, which fall within the extended latent Gaussian model (ELGM) [@stringer2022fast] class, and were not previously amenable to inference with INLA.
-ELGMs build on latent Gaussian models (LGMs) by allowing each element of the linear predictor to depend on any subset of elements from the latent field.
+This extension facilitates inference for models like Naomi, which are not latent Gaussian models (LGMs) and so were not previously amenable to inference with INLA.
+Instead, due to dependence of observations on multiple structured additive predictors, Naomi is what has been termed by @stringer2022fast an extended latent Gaussian model (ELGM).
 We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
@@
 
-## Household survey component
+## Household survey component \label{sec:household}
 
@@
 For ease of notation, let $i$ index the finest district-age-sex division included in the model.
-The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
+As data we observe may be aggregated over indices $i$, let $\mathcal{I}$ be a set of $i$ for which observations are reported.
 Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be antiretroviral therapy (ART) coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
@@
 Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
+<!-- The kappa is introduced, but you have to be paying close attention to see that it's so that for theta in rho, alpha, kappa -->
 For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes
 $$
-M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
+m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
 $$
-where $j$ indexes individuals across all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
-The observed number of indicator cases is then $Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$.
-For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
+where $j$ indexes individuals across all strata $i \in \mathcal{I}$.
+The corresponding design weights are $w_j = 1 / \pi_j$, where $\pi_j$ is the probability of inclusion in the survey.
+The observed number of indicator cases is then $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$.
+For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
 \begin{align*}
@@
 $$
-\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
+\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
 $$
@@
 
-## ANC testing component
+## ANC testing component \label{sec:anc}
 
@@
 \end{align*}
-We inform these processes by specifying likelihoods for the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $X^\text{ANC}_\mathcal{I}$, the number of those with positive status $Y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $Z^\text{ANC}_\mathcal{I}$.
+We inform these processes by specifying likelihoods for the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
 We use the binomial working likelihoods
 \begin{align*}
-Y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(Y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}) \\
-Z^\text{ANC}_\mathcal{I} &\sim \text{Bin}(Y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}),
+y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}), \\
+z^\text{ANC}_\mathcal{I} &\sim \text{Bin}(y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}),
 \end{align*}
-where again we use weighted aggregates
+where, again, we use weighted aggregates
 \begin{align*}
@@
 
-## ART attendance component
+## ART attendance component \label{sec:art}
 
@@
 Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
-We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$, and model $\gamma_{x, x'}$ using a multinomial logistic regression model where $\tilde \gamma_{x, x'}$ are the log-odds.
+We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$, and model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a multinomial logistic regression model.
 
@@
 
-1. In the household survey component, the HIV incidence rate depends on the HIV prevalence and ART coverage linear predictors.
-2. In the ANC testing component, the HIV prevalence and ART coverage depend upon the household survey component.
-Specifically, $|\mathcal{J}_i| = 2$ such that for $\theta \in \{\rho, \alpha\}$ then
+1. In the household survey component (Section \ref{sec:household}), the HIV incidence rate depends on the HIV prevalence and ART coverage linear predictors.
+2. In the ANC testing component (Section \ref{sec:anc}), the HIV prevalence and ART coverage depend upon the household survey component.
+Specifically, $|\mathcal{J}_i| = 2$ such that for $\theta \in \{\rho, \alpha\}$
 $$
@@
 $$
+3. In the ART attendance component (Section \ref{sec:art}), the multinomial logistic regression...
 

### 2023-02-13 src/naomi-simple_compare/comparison.Rmd
ECDF difference in comparison, code to turn on and off the tabs

@@
 abstract: |
-  **Background**  We have run the simplified Naomi model using a range of inference methods.
+  **Background**  We have run the simplified Naomi model using a range of inference methods: `TMB`, `aghq`, `adam` and `tmbstan`.
 
-  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods.
+  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using histograms, Kolmogorov-Smirnov tests, maximum mean discrepancy and Pareto-smoothed importance sampling.
 ---
@@
 tmbstan <- readRDS("depends/tmbstan.rds")
+
+depends <- yaml::read_yaml("orderly.yml")$depends
 ```
+
+## Run details {.tabset}
 
@@
 
+### `TMB`
+
 ```{r}
-depends <- yaml::read_yaml("orderly.yml")$depends
-
-for(i in seq_along(depends)) {
+dependency_details <- function(i) {
   report_name <- names(depends[[i]])
@@
 }
+
+dependency_details(1)
 ```
 
-All of the possible parameter names are as follows:
+### `aghq`
 
 ```{r}
-pars <- unique(names(tmb$fit$obj$env$par))
+dependency_details(2)
 ```
 
-# Time taken
+### `adam`
+
+```{r}
+dependency_details(3)
+```
+
+### `tmbstan`
+
+```{r}
+dependency_details(4)
+```
+
+## Time taken
 
@@
 
-# Histograms {.tabset .tabset-fade}
+# Histograms and empirical cumulative distribution plots {.tabset .tabset-fade}
+
+We create histograms and empirical cumulative distribution function (ECDF) plots of the samples from each method.
+All of the possible parameter names are as follows:
+
+```{r}
+pars <- unique(names(tmb$fit$obj$env$par))
+```
+
+We will produce plots about the following subset of them.
+There is no particular reason to choose this subset rather than other, it's quite arbitrary.
+
+```{r}
+pars_eval <- pars %in% c("beta_rho", "beta_alpha", "beta_lambda", "beta_anc_rho", "beta_anc_alpha", "u_rho_x", "us_rho_x", "u_rho_xs")
+names(pars_eval) <- pars
+```
 
@@
 
-```{r}
+```{r eval = pars_eval[["beta_rho"]]}
 histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
@@
 
-```{r}
+```{r eval = pars_eval[["beta_rho"]]}
 histogram_and_ecdf_helper("beta_alpha")
@@
 
-```{r}
+```{r eval = pars_eval[["beta_rho"]]}
 histogram_and_ecdf_helper("beta_lambda")
@@
 
-```{r}
+```{r eval = pars_eval[["beta_rho"]]}
 histogram_and_ecdf("beta_anc_rho")
@@
 
-```{r}
+```{r eval = pars_eval[["beta_rho"]]}
 histogram_and_ecdf("beta_anc_alpha")
@@
 
-```{r}
+```{r eval = FALSE}
 lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf)
@@
 
-```{r}
+```{r eval = FALSE}
 lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
@@
 
-```{r}
+```{r eval = pars_eval[["u_rho_x"]]}
 histogram_and_ecdf_helper("u_rho_x")
@@
 
-```{r}
+```{r eval = pars_eval[["us_rho_x"]]}
 histogram_and_ecdf_helper("us_rho_x")
@@
 
-```{r}
+```{r eval = pars_eval[["u_rho_xs"]]}
 histogram_and_ecdf_helper("u_rho_xs")
@@
 
-```{r}
+```{r eval = pars_eval[["us_rho_xs"]]}
 histogram_and_ecdf_helper("us_rho_xs")
@@
 
-```{r}
+```{r eval = pars_eval[["u_rho_as"]]}
 histogram_and_ecdf_helper("u_rho_as")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_x"]]}
 histogram_and_ecdf_helper("u_alpha_x")
@@
 
-```{r}
+```{r eval = pars_eval[["us_alpha_x"]]}
 histogram_and_ecdf_helper("us_alpha_x")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_xs"]]}
 histogram_and_ecdf_helper("u_alpha_xs")
@@
 
-```{r}
+```{r eval = pars_eval[["beta_rho"]]}
 histogram_and_ecdf_helper("us_alpha_xs")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_a"]]}
 histogram_and_ecdf_helper("u_alpha_a")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_as"]]}
 histogram_and_ecdf_helper("u_alpha_as")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_xa"]]}
 histogram_and_ecdf_helper("u_alpha_xa")
@@
 
-```{r}
+```{r eval = pars_eval[["ui_lambda_x"]]}
 histogram_and_ecdf_helper("ui_lambda_x")
@@
 
-```{r}
+```{r eval = pars_eval[["ui_anc_rho_x"]]}
 histogram_and_ecdf_helper("ui_anc_rho_x")
@@
 
-```{r}
+```{r eval = pars_eval[["ui_anc_alpha_x"]]}
 histogram_and_ecdf_helper("ui_anc_alpha_x")
@@
 
-```{r}
+```{r eval = pars_eval[["log_or_gamma"]]}
 histogram_and_ecdf_helper("log_or_gamma")
@@
 
-```{r}
+```{r eval = pars_eval[["u_rho_x"]]}
 ks_helper("u_rho_x")
@@
 
-```{r}
+```{r eval = pars_eval[["u_rho_xs"]]}
 ks_helper("u_rho_xs")
@@
 
-```{r}
+```{r eval = pars_eval[["us_rho_x"]]}
 ks_helper("us_rho_x")
@@
 
-```{r}
+```{r eval = pars_eval[["us_rho_xs"]]}
 ks_helper("us_rho_xs")
@@
 
-```{r}
+```{r eval = pars_eval[["u_rho_a"]]}
 ks_helper("u_rho_a")
@@
 
-```{r}
+```{r eval = pars_eval[["u_rho_as"]]}
 ks_helper("u_rho_as")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_x"]]}
 ks_helper("u_alpha_x")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_xs"]]}
 ks_helper("u_alpha_xs")
@@
 
-```{r}
+```{r eval = pars_eval[["us_alpha_x"]]}
 ks_helper("us_alpha_x")
@@
 
-```{r}
+```{r eval = pars_eval[["us_alpha_xs"]]}
 ks_helper("us_alpha_xs")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_a"]]}
 ks_helper("u_alpha_a")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_as"]]}
 ks_helper("u_alpha_as")
@@
 
-```{r}
+```{r eval = pars_eval[["u_alpha_xa"]]}
 ks_helper("u_alpha_xa")
@@
 
-```{r}
+```{r eval = pars_eval[["ui_anc_rho_x"]]}
 ks_helper("ui_anc_rho_x")
@@
 
-```{r}
+```{r eval = pars_eval[["ui_anc_alpha_x"]]}
 ks_helper("ui_anc_alpha_x")
@@
 
-```{r}
+```{r eval = pars_eval[["log_or_gamma"]]}
 ks_helper("log_or_gamma")
@@
 tmb$fit$obj$fn()
+#' To write!
 ```

### 2023-02-13 src/naomi-simple_compare/functions.R

@@ -1,6 +1,10 @@
+absmax <- function(x) x[which.max(abs(x))]
+
 #' Create a facetted histogram plot of the named parameter using samples from each of the four methods
 #' As well as an ECDF plot
 histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
+  colours <- c("#56B4E9", "#009E73", "#E69F00", "#CC79A7")
+
   if(!is.null(i)) {
     par_name <- paste0(par, "[", i, "]")
 
@@ -40,8 +44,8 @@ histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
     theme_minimal() +
     facet_grid(method~.) +
     labs(x = par_name, y = "Density", fill = "Method") +
-    scale_color_manual(values = multi.utils::cbpalette()) +
-    scale_fill_manual(values = multi.utils::cbpalette()) +
+    scale_color_manual(values = colours) +
+    scale_fill_manual(values = colours) +
     theme(legend.position = "none") +
     labs(title = par_name, subtitle = paste0("Mean = ", mean, ", SD = ", sd, " (from tmbstan)"))
 
@@ -59,16 +63,36 @@ histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
   tmbstan_ecdf <- stats::ecdf(filter(df_compare, method == "tmbstan") %>% pull(samples))
   tmbstan_ecdf_df <- data.frame(x = grid, ecdf = tmbstan_ecdf(grid), method = "tmbstan")
 
+  # Add ECDF differences
+  tmb_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - tmb_ecdf_df$ecdf
+  aghq_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - aghq_ecdf_df$ecdf
+  adam_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - adam_ecdf_df$ecdf
+  tmbstan_ecdf_df$ecdf_diff <- 0
+
+  ks_tmb <- absmax(tmb_ecdf_df$ecdf_diff)
+  ks_aghq <- absmax(aghq_ecdf_df$ecdf_diff)
+  ks_adam <- absmax(adam_ecdf_df$ecdf_diff)
+
   ecdf_df <- bind_rows(tmb_ecdf_df, aghq_ecdf_df, adam_ecdf_df, tmbstan_ecdf_df)
 
[…]

### 2023-02-13 src/naomi-simple_compare/comparison.Rmd
Correct error with starts_with option, add green highlighting for lowest KS value in table, alter the summary plots to be more appropriate

@@
 
-# Histograms and empirical cumulative distribution plots {.tabset .tabset-fade}
+# Histograms and ECDF difference plots {.tabset .tabset-fade}
 
-We create histograms and empirical cumulative distribution function (ECDF) plots of the samples from each method.
+We create histograms and empirical cumulative distribution function (ECDF) difference plots of the samples from each method.
 All of the possible parameter names are as follows:
@@
 pars <- unique(names(tmb$fit$obj$env$par))
+pars
 ```
@@
 ```{r}
-ks_helper <- function(par, ...) {
-  to_ks_df(par) %>% ks_plot(par, ...)
+ks_helper <- function(par, sw = FALSE, ...) {
+  to_ks_df(par, starts_with = sw) %>% ks_plot(par, ...)
 }
@@
 ```{r}
-ks_helper("beta")
-ks_helper("beta", method1 = "TMB", method2 = "adam")
+ks_helper("beta", sw = TRUE)
+ks_helper("beta", sw = TRUE, method1 = "TMB", method2 = "TMB")
 ```
@@
 ```{r}
-ks_helper("logit")
-ks_helper("logit", method1 = "TMB", method2 = "adam")
+ks_helper("logit", sw = TRUE)
+ks_helper("logit", sw = TRUE, method1 = "TMB", method2 = "adam")
 ```
@@
 ```{r}
-ks_helper("log_sigma")
-ks_helper("log_sigma", method1 = "TMB", method2 = "adam")
+ks_helper("log_sigma", sw = TRUE)
+ks_helper("log_sigma", sw = TRUE , method1 = "TMB", method2 = "adam")
 ```
@@
 ```{r}
+options(dplyr.summarise.inform = FALSE)
+
 ks_summary <- lapply(unique(names(tmb$fit$obj$env$par)), function(x) {
   to_ks_df(x) %>%
-    group_by(method) %>%
-    summarise(ks = mean(ks), par = x)
+    group_by(method, index) %>%
+    summarise(ks = mean(ks), par = x) %>%
+    ungroup()
 }) %>%
@@
 
+r <- adam$quad$obj$env$random
+x_names <- names(adam$quad$obj$env$par[r])
+theta_names <- names(adam$quad$obj$env$par[-r])
+
+dict <- data.frame(
+  Parameter = c(unique(x_names), unique(theta_names)),
+  Type = c(rep("Latent field", length(unique(x_names))), rep("Hyper", length(unique(theta_names))))
+)
+
+ks_summary <- ks_summary %>%
+  left_join(dict, by = "Parameter")
+```
+
+Values in green are the lowest KS difference for that particular parameter:
+
+```{r}
 ks_summary %>%
@@
     decimals = 3
+  ) %>%
+  gt::tab_style(
+    style = cell_fill(color = "#009E73"),
+    locations = cells_body(
+      columns = `KS(adam, tmbstan)`,
+      rows = `KS(adam, tmbstan)` < `KS(aghq, tmbstan)` & `KS(adam, tmbstan)` < `KS(TMB, tmbstan)`
     )
+  ) %>%
+    gt::tab_style(
+    style = cell_fill(color = "#009E73"),
+    locations = cells_body(
+      columns = `KS(aghq, tmbstan)`,
+      rows = `KS(aghq, tmbstan)` < `KS(adam, tmbstan)` & `KS(aghq, tmbstan)` < `KS(TMB, tmbstan)`
+    )
+  ) %>%
+    gt::tab_style(
+    style = cell_fill(color = "#009E73"),
+    locations = cells_body(
+      columns = `KS(TMB, tmbstan)`,
+      rows = `KS(TMB, tmbstan)` < `KS(aghq, tmbstan)` & `KS(TMB, tmbstan)` < `KS(adam, tmbstan)`
+    )
+  )
+```
 
-r <- adam$quad$obj$env$random
-x_names <- names(adam$quad$obj$env$par[r])
-theta_names <- names(adam$quad$obj$env$par[-r])
+This plot could be improved by making the red to green transition smooth, then adding legend for KS difference.
+The density plots could be filled according to KS difference.
 
-dict <- data.frame(
-  Parameter = c(x_names, theta_names),
-  Type = c(rep("Latent field", length(x_names)), rep("Hyperparameters", length(theta_names)))
-)
-
-ks_summary <- ks_summary %>%
-  left_join(dict, by = "Parameter")
-
+```{r}
 summary_ks_plot <- function(ks_summary, method1, method2) {
@@
 
-  scatterplot <- ggplot(ks_summary, aes(x = .data[[ks_method1]], y = .data[[ks_method2]])) +
-    geom_point(alpha = 0.2) +
-    annotate(geom = "polygon", x = c(-Inf, Inf, Inf), y = c(-Inf, Inf, -Inf), fill = multi.utils::cbpalette()[3], alpha = 0.1) +
-    annotate(geom = "polygon", x = c(-Inf, Inf, -Inf), y = c(-Inf, Inf, Inf), fill = multi.utils::cbpalette()[5], alpha = 0.1) +
-    xlim(0, 1) +
-    ylim(0, 1) +
-    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
+  densityplot <- ggplot(ks_summary, aes(x = .data[[ks_method1]], y = .data[[ks_method2]])) +
+    stat_density_2d(aes(linetype = Type), col = "black") +
+    annotate(geom = "polygon", x = c(-Inf, Inf, Inf), y = c(-Inf, Inf, -Inf), fill = multi.utils::cbpalette()[2], alpha = 0.2) +
+    annotate(geom = "polygon", x = c(-Inf, Inf, -Inf), y = c(-Inf, Inf, Inf), fill = multi.utils::cbpalette()[6], alpha = 0.2) +
+    xlim(0, 0.75) +
+    ylim(0, 0.75) +
+    geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
+    scale_linetype_manual(values = c("dashed", "solid")) +
+    # scale_alpha_discrete(range = c(0.5, 0.01)) +
+    # guides(alpha = FALSE) +
+    guides(linetype = FALSE) +
     labs(x = ks_method1, y = ks_method2) +
@@
 
-  jitterplot <- ggplot(ks_summary, aes(x = Type, y = `KS difference`)) +
-    geom_jitter(width = 0.1, alpha = 0.2) +
-    labs(x = "", y = paste0(ks_method1, " - ", ks_method2)) +
+  ridgeplot <- ggplot(ks_summary, aes(y = Type, x = `KS difference`)) +
+    ggridges::geom_density_ridges(alpha = 0.7, fill = NA, aes(linetype = Type)) +
+    coord_flip() +
+    scale_linetype_manual(values = c("dashed", "solid")) +
+    labs(y = "", x = paste0(ks_method1, " - ", ks_method2)) +
+    guides(linetype = FALSE) +
     theme_minimal()
 
-  scatterplot + jitterplot
+  densityplot + ridgeplot
 }

### 2023-02-13 src/naomi-simple_compare/functions.R

@@ -43,7 +43,7 @@ histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
     geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +
     theme_minimal() +
     facet_grid(method~.) +
-    labs(x = par_name, y = "Density", fill = "Method") +
+    labs(y = "Density", fill = "Method") +
     scale_color_manual(values = colours) +
     scale_fill_manual(values = colours) +
     theme(legend.position = "none") +
@@ -82,11 +82,11 @@ histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
   ecdf_plot <- ggplot(ecdf_df, aes(x = x, y = ecdf_diff, col = method)) +
     geom_line() +
     geom_abline(intercept = ks_tmb, slope = 0, col = colours[1], linetype = "dashed", alpha = 0.8) +
-    annotate("text", x = 1.05 * max(ecdf_df$x), y = ks_tmb, label = ks_labeller(ks_tmb), col = colours[1], alpha = 0.8) +
+    annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_tmb, label = ks_labeller(ks_tmb), col = colours[1], alpha = 0.8) +
     geom_abline(intercept = ks_aghq, slope = 0, col = colours[2], linetype = "dashed", alpha = 0.8) +
-    annotate("text", x = 1.05 * max(ecdf_df$x), y = ks_aghq, label = ks_labeller(ks_aghq), col = colours[2], alpha = 0.8) +
+    annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_aghq, label = ks_labeller(ks_aghq), col = colours[2], alpha = 0.8) +
     geom_abline(intercept = ks_adam, slope = 0, col = colours[3], linetype = "dashed", alpha = 0.8) +
-    annotate("text", x = 1.05 * max(ecdf_df$x), y = ks_adam, label = ks_labeller(ks_adam), col = colours[3], alpha = 0.8) +
+    annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_adam, label = ks_labeller(ks_adam), col = colours[3], alpha = 0.8) +
     scale_color_manual(values = colours) +
     labs(x = "", y = "ECDF difference") +
     guides(col = "none") +
[…]

### 2023-02-14 src/epil/epil.Rmd
Update epil writing and add confirmation that custom Laplace matches R-INLA Laplace for intercept parameter

@@
 Epil %>% ggplot(aes(x =  as.factor(Trt), y = y)) +
-  geom_boxplot(alpha = 0.5, outlier = FALSE) +
+  geom_boxplot(alpha = 0.5) +
   geom_jitter(width = 0.1, alpha = 0.5) +
@@
 
-We modify `Epil` by centering all of the covariates:
+We modify the `Epil` dataframe by centering all of the covariates:
 
@@
 ```{r}
-start_time_fit1 <- Sys.time()
+start_fit1 <- Sys.time()
 
@@
 
-end_time_fit1 <- Sys.time()
-time_fit1 <- end_time_fit1 - start_time_fit1
+end_fit1 <- Sys.time()
+time_fit1 <- end_fit1 - start_fit1
+time_fit1
 ```
 
-It took `r time_fit1` seconds to fit this model.
+It took `r round(time_fit1, 2)` seconds to fit this model.
 All of the $\hat R$ values are below 1.05, so it looks like the MCMC has reached convergence:
@@
     axis.text.x = element_blank(),
-    axis.ticks.x = element_blank(),
     legend.position = "bottom"
@@
 Rather than specifying the log-posterior via C++, the package uses a formula interface.
+This makes it easy to specify common models, but does limit the flexibility of the package for more advanced users.
 
@@
 
-We can use various settings (`"gaussian"`, `"simplified.laplace"` and `"laplace"`) for the strategy passed to `control.inla`, which controls the approximation used for the latent field posterior marginals:
+We can use various settings (`"gaussian"`, `"simplified.laplace"` and `"laplace"`) for the `strategy` passed to `control.inla`, which controls the approximation used for the latent field posterior marginals.
+We create a function `epil_inla` which takes as input the latent field approximation strategy, as well as the hyperparameter quadrature approach:
 
@@
 epil_inla <- function(strat, int_strat) {
-  inla(formula,
+  inla(
+    formula,
     control.fixed = list(mean = 0, prec = 1 / 100^2), #' Beta prior
@@
     control.inla = list(strategy = strat, int.strategy = int_strat),
-       control.predictor = list(compute = TRUE))
+    control.predictor = list(compute = TRUE)
+  )
 }
 
-start_time_fit2 <- Sys.time()
+start_fit2 <- Sys.time()
 fit2 <- epil_inla(strat = "gaussian", int_strat = "eb")
-end_time_fit2 <- Sys.time()
-(time_fit2 <- end_time_fit2 - start_time_fit2)
+end_fit2 <- Sys.time()
+(time_fit2 <- end_fit2 - start_fit2)
 
-start_time_fit3 <- Sys.time()
+start_fit3 <- Sys.time()
 fit3 <- epil_inla(strat = "simplified.laplace", int_strat = "eb")
-end_time_fit3 <- Sys.time()
-(time_fit3 <- end_time_fit3 - start_time_fit3)
+end_fit3 <- Sys.time()
+(time_fit3 <- end_fit3 - start_fit3)
 
-start_time_fit4 <- Sys.time()
+start_fit4 <- Sys.time()
 fit4 <- epil_inla(strat = "laplace", int_strat = "eb")
-end_time_fit4 <- Sys.time()
-(time_fit4 <- end_time_fit4 - start_time_fit4)
+end_fit4 <- Sys.time()
+(time_fit4 <- end_fit4 - start_fit4)
 
-start_time_fit5 <- Sys.time()
+start_fit5 <- Sys.time()
 fit5 <- epil_inla(strat = "gaussian", int_strat = "grid")
-end_time_fit5 <- Sys.time()
-(time_fit5 <- end_time_fit5 - start_time_fit5)
+end_fit5 <- Sys.time()
+(time_fit5 <- end_fit5 - start_fit5)
 
-start_time_fit6 <- Sys.time()
+start_fit6 <- Sys.time()
 fit6 <- epil_inla(strat = "simplified.laplace", int_strat = "grid")
-end_time_fit6 <- Sys.time()
-(time_fit6 <- end_time_fit6 - start_time_fit6)
+end_fit6 <- Sys.time()
+(time_fit6 <- end_fit6 - start_fit6)
 
-start_time_fit7 <- Sys.time()
+start_fit7 <- Sys.time()
 fit7 <- epil_inla(strat = "laplace", int_strat = "grid")
-end_time_fit7 <- Sys.time()
-(time_fit7 <- end_time_fit7 - start_time_fit7)
+end_fit7 <- Sys.time()
+(time_fit7 <- end_fit7 - start_fit7)
 ```
@@
 
-Template Model Builder, or `TMB`, is an R package which implements Laplace approximation using automatic differentiation.
-The user writes a C++ file, here `epil.cpp`, which contains the negative log-posterior for the model in question:
+Template Model Builder, or `TMB`, is an R package which implements the Laplace approximation using automatic differentiation.
+The user writes a C++ file, here `epil.cpp` (see appendix), which contains the negative log-posterior for the model in question:
 
@@
 
-These are the initialisation parameters:
+We need to specify the initialisation parameters:
 
@@
 
-We integrate `random` out, which here corresponds to `c("beta", "epsilon", "nu")`, with a Laplace approximation:
+We integrate the random effects, `random`, out with a Laplace approximation.
+Here these corresponds to `c("beta", "epsilon", "nu")`:
 
 ```{r}
-#' Perhaps you could argue about when the "start" and "end" of the model fitting are
+#' Note: Perhaps you could argue about when the "start" and "end" of the model fitting are
 #' For example, is compilation of the TMB template part of the model fitting?
-start_time_fit8 <- Sys.time()
+start_fit8 <- Sys.time()
 
@@
 
-end_time_fit8 <- Sys.time()
-(time_fit8 <- end_time_fit8 - start_time_fit8)
+end_fit8 <- Sys.time()
+(time_fit8 <- end_fit8 - start_fit8)
 ```
@@
 
-For completeness, we check here that the objective functions as implemented in `TMB` and Stan are identical.
+For completeness, we verify that the objective functions as implemented in `TMB` and Stan are identical.
+This is actually quite difficult to do, so pat it's worth checking, and giving yourself a pat on the back if they are.
 To obtain the `TMB` negative log-posterior we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
@@
 
-For example, we can test the NLL of the initialisation parameters `param`:
+We can test the NLL of the initialisation parameters `param`:
 
 ```{r}
-c(
-  "TMB" = tmb_nll$fn(unlist(param)),
-  "Stan" = -rstan::log_prob(object = stan_nll, unlist(param))
-)
+c("TMB" = tmb_nll$fn(unlist(param)), "Stan" = -rstan::log_prob(object = stan_nll, unlist(param)))
 ```
 
-These look different!
-Is this a problem?
-We can evaluate the NLL at some of the other parameters explored during the MCMC:
+But these look different!
+This is just because the two functions are different by a constant, which we can show by evaluating the NLL at some of the other parameters explored during the MCMC:
 
@@
 pars_list <- apply(pars_mat, 1, function(x) relist(flesh = x, skeleton = fit1@inits[[1]]))
-upars_list <- lapply(pars_list, function(x) unconstrain_pars(fit1, x))
+#' Note that Stan transformed constrained parameters in the parameters block to unconsrained parameters that can be sampled from
+upars_list <- lapply(pars_list, function(x) rstan:::unconstrain_pars(fit1, x))
 ```
 
-Using just a few of them, we can see that the difference between the `TMB` and Stan objectives is a constant, so for our purposes they are functionally the same:
+Using just a few of them, we can see that the difference between the `TMB` and Stan objectives is a constant:
 
@@
 ```{r}
-start_time_fit9 <- Sys.time()
+start_fit9 <- Sys.time()
 formula9 <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT + (1 | rand) + (1 | Ind)
 fit9 <- glmmTMB(formula9, data = Epil, family = poisson(link = "log"))
-end_time_fit9 <- Sys.time()
-(time_fit9 <- end_time_fit9 - start_time_fit9)
+end_fit9 <- Sys.time()
+(time_fit9 <- end_fit9 - start_fit9)
 ```
@@
 This won't be exactly the same model, in terms of priors and so on, but it may be close.
+We mention this package here in case it is of interest, but will not focus on it during the remainder of this notebook, as we would like to compare inferential methods on exactly the same model.
 
@@
 ```{r} 
-start_time_fit10 <- Sys.time()
+start_fit10 <- Sys.time()
 fit10 <- tmbstan(obj = obj, chains = 4, refresh = 0)
-end_time_fit10 <- Sys.time()
-(time_fit10 <- end_time_fit10 - start_time_fit10)
+end_fit10 <- Sys.time()
+(time_fit10 <- end_fit10 - start_fit10)
 ```
+
+The results from `tmbstan`, up to stochasticity of the MCMC sampler, should be exactly the same as those we might generate from Stan -- given that we have confirmed the objective functions they are sampling from are identical up to a constant.
+For this reason, we do not need to focus on these results in the remainder of the notebook, but again highlight this package in case it is of interest.
 
@@
 It also works smoothly with `TMB` templates.
+The parameter `k = 3` determines the number of quadrature points per dimension:
 
-One (non-standard!) approach would be to use `glmmTMB` to get the `TMB` template which can then be passed to `aghq`.
+```{r results='hide'}
+start_fit11 <- Sys.time()
+fit11 <- aghq::marginal_laplace_tmb(
+  obj,
+  k = 3,
+  startingvalue = c(param$l_tau_epsilon, param$l_tau_nu)
+)
+
+fit11_samples <- aghq::sample_marginal(fit11, M = 1000)$samps %>%
+  t() %>%
+  as.data.frame() %>%
+  inf.utils::replace_duplicate_colnames()
+end_fit11 <- Sys.time()
+(time_fit11 <- end_fit11 - start_fit11)
+```
+
+Another (non-standard) approach that is possible is to use `glmmTMB` to get the `TMB` template which can then be passed to `aghq`.
 
@@
 
-The parameter `k = 3` determines the number of quadrature points per dimension:
-
 ```{r}
@@
 
-This would be useful if we didn't have a `TMB` template written, but we do, so let's just use that instead!
-
-```{r results='hide'}
-start_time_fit11 <- Sys.time()
-fit11 <- aghq::marginal_laplace_tmb(
-  obj,
-  k = 3,
-  startingvalue = c(param$l_tau_epsilon, param$l_tau_nu)
-)
-
-fit11_samples <- aghq::sample_marginal(fit11, M = 1000)$samps %>%
-  t() %>%
-  as.data.frame() %>%
-  inf.utils::replace_duplicate_colnames()
-end_time_fit11 <- Sys.time()
-(time_fit11 <- end_time_fit11 - start_time_fit11)
-```
+This could be useful in situations where you don't have a `TMB` template written.
 
@@
 
-Let's just starting by trying to get the Laplace marginals for $\beta$:
-
-`epil_beta_index.cpp` is a `TMB` template where the `beta` parameter has been replaced by `beta_i` and `beta_minus_i`.
+Let's just starting by trying to get the Laplace marginals for $\beta$.
+To do this, we create the `TMB` template `epil_beta_index.cpp` (see appendix) which has the `beta` parameter replaced by `beta_i` and `beta_minus_i`.
 As well, there is an additional `DATA_INTEGER(i)` allowing us to pass in the index of `i` that we would like to obtain a full Laplace approximation to.
@@
 Some key things to note about this:
-We use the `map` argument to fix the hyperparameters to their MAP estimates, which requires passing `factor(NA)` to `map`, as well as setting the initialisation parameters for the hyperparameters to be their MAP estimates.
-Setting `random = c("beta_minus_i", "nu", "epsilon")` integrates out all of the random effects but `beta_i`.
-Quadrature is performed using `aghq::aghq` to obtain a the normalised posterior marginal.
+
+* We use the `map` argument to fix the hyperparameters to their MAP estimates, which requires passing `factor(NA)` to `map`, as well as setting the initialisation parameters for the hyperparameters to be their MAP estimates.
+* Setting `random = c("beta_minus_i", "nu", "epsilon")` integrates out all of the random effects but `beta_i`.
+* Quadrature is performed using `aghq::aghq` to obtain a the normalised posterior marginal.
 
@@
   param_fixed_theta[["beta"]] <- NULL
-
-  #' Might need this here if you're getting crashes
-  # f <- parallel::mcparallel({})
-  # x <- parallel::mccollect(f)
 
@@
 
-Ok, great, so we got the Laplace marginals for `beta`.
-What about if we wanted to get them for all of the elements of the latent field?
-One way we could do that is to create separate `TMB` templates for each named random effect (`beta`, `nu` and `epsilon`).
+Now that we have the Laplace marginals for `beta`, we can think about what we would do if we wanted to get them for all of the elements of the latent field.
+One approach would be to create separate `TMB` templates for each named random effect (`beta`, `nu` and `epsilon`).
 To see what this would look like, have a look at `epil_beta_index.cpp`, `epil_nu_index.cpp` and `epil_epsilon_index.cpp`.
 
-What might be better is to do it in one `TMB` template.
+What might be better is to do it in one `TMB` template, `epil_index.cpp` (see appendix):
 
@@
 
-### AGHQ with one C++ template
-
-```{r}
-i <- 1
-dat$toggle <- index_controller[i, "toggle"]
-dat$i <- index_controller[i, "i"]
-```
-
 # Comparison
@@
 ```{r}
-plot_pdf <- ggplot(data.frame(x = rstan::extract(fit1, pars = "beta[1]")[[1]]), aes(x = x)) +
-    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7], bins = 60) +
+#' Stan histogram base-plot
+plot_pdf_stan <- ggplot(data.frame(x = rstan::extract(fit1, pars = "beta[1]")[[1]]), aes(x = x)) +
+    geom_histogram(aes(y = ..density..), alpha = 0.6, fill = cbpalette[7], bins = 40) +
     theme_minimal() +
@@
 
-plot_pdf <- plot_pdf +
+#' Add R-INLA results
+plot_pdf_stan_inla <- plot_pdf_stan +
   geom_line(data = beta0_inla_marginals, aes(x = x, y = y, col = method)) +
   scale_color_manual(values = cbpalette) +
-  labs(col = "Method")
+  labs(col = "Method", title = "Inference from R-INLA")
 
-plot_pdf
+plot_pdf_stan_aghq <- plot_pdf_stan +
+  geom_histogram(data = data.frame(x = fit11_beta_samples$`beta[1]`), aes(y = ..density..), alpha = 0.6, fill = cbpalette[1], bins = 40)
 
-ggsave("intercept-comparison.pdf", plot = plot_pdf, h = 4, w = 6.25)
+plot_pdf_stan_aghq <- plot_pdf_stan_aghq +
+    geom_line(data = laplace_x[[1]]$pdf_and_cdf, aes(x = theta, y = pdf), col = cbpalette[2]) +
+    lims(x = c(1.2, 2)) +
+    labs(title = "Inference from aghq and custom Laplace method")
+    theme_minimal()
+
+(plot_intercept_comparison <- plot_pdf_stan_inla/ plot_pdf_stan_aghq)
+
+ggsave("intercept-comparison.pdf", plot = plot_intercept_comparison, h = 6, w = 6.25)
 ```

### 2023-02-14 src/epil/epil.Rmd
Fix missing + sign

@@
     lims(x = c(1.2, 2)) +
-    labs(title = "Inference from aghq and custom Laplace method")
+    labs(title = "Inference from aghq and custom Laplace method") +
     theme_minimal()
 
-(plot_intercept_comparison <- plot_pdf_stan_inla/ plot_pdf_stan_aghq)
+plot_intercept_comparison <- plot_pdf_stan_inla / plot_pdf_stan_aghq
+
+plot_intercept_comparison
 

### 2023-02-14 src/epil/epil.Rmd
Plots of lognormconsts and difference between epil and naomi-simple Laplace implementations

@@
 laplace_x <- lapply(1:nrow(index_controller), safe_xi_laplace_marginal, opt_theta = opt_theta, finegrid = fine_grid)
+```
 
-laplace_beta_summary <- lapply(head(laplace_x, n = dat$K), function(result) summary(result$quad)$summarytable) %>%
-  bind_rows()
+Test that all the normalising constants calculated are constant, and equal to that obtained by `aghq`:
 
-rownames(laplace_beta_summary) <- paste0("beta[", 1:dat$K, "]")
+```{r}
+aghq_lognormconst <- fit11$normalized_posterior$lognormconst
+
+lognormconsts <- purrr::map(laplace_x, function(x) x$quad$normalized_posterior$lognormconst)
+
+data.frame(lognormconst = unlist(lognormconsts)) %>%
+  ggplot(aes(y = lognormconst, x = 1:length(lognormconst))) +
+    geom_point(alpha = 0.3) +
+    geom_hline(yintercept = aghq_lognormconst, linetype = "dashed") +
+    theme_minimal() +
+    labs(
+      title = "Comparison of log normalising constants",
+      y = "Log normalising constant", x = "Latent field index"
+    )
 ```
+
+The differences between this old approach, and the newer approach used in `naomi-simple` are:
+
+1. Use of the `map` argument to fix theta to modal values. Which is useful for producing a `TMB` objective `obj_fixed_theta`
+2. Passing `obj_fixed_theta` directly into `aghq::aghq` to do the selection of the points (here `k = 3`), integration and computation of the posterior with `aghq::compute_pdf_and_cdf`
 
@@
 
+laplace_beta_summary <- lapply(head(laplace_x, n = dat$K), function(result) summary(result$quad)$summarytable) %>% bind_rows()
+rownames(laplace_beta_summary) <- paste0("beta[", 1:dat$K, "]")
 adam12 <- as.vector(t(laplace_beta_summary[1:6, 1:2]))

### 2023-02-15 src/epil/epil.Rmd
Attempts to debug Laplace

@@
 aghq_lognormconst <- fit11$normalized_posterior$lognormconst
-
 lognormconsts <- purrr::map(laplace_x, function(x) x$quad$normalized_posterior$lognormconst)
@@
 2. Passing `obj_fixed_theta` directly into `aghq::aghq` to do the selection of the points (here `k = 3`), integration and computation of the posterior with `aghq::compute_pdf_and_cdf`
+
+```{r}
+#' Compute Laplace marginals with new method
+```
 

### 2023-02-17 src/docs_paper/paper.Rmd
Update method for normalising Laplace marginals in paper

@@
 3. Use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
-$$
-\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z).
-$$
+\begin{equation}
+\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z). \label{eq:evidence1}
+\end{equation}
 4.
-For $i \in [n]$ generate $l$ nodes $\mathcal{X}_i$ via a Gauss-Hermite quadrature rule with one-dimension adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$.
+For $i \in [n]$ generate $l$ nodes $x_i(\z)$, via a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
 A value of $l \geq 4$ is recommended to enable B-spline interpolation.
-5. For $x_i \in \mathcal{X}_i$ and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
+5. For $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
 \begin{align*}
@@
 where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
-6. For $x_i \in \mathcal{X}_i$ calculate
-$$
-\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
-$$
+6. For $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$ calculate
+\begin{equation}
+\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{\tilde p_{\text{AQ}}(\y)}. \label{eq:laplace-marginal}
+\end{equation}
 where
@@
 $$
+Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence in Equation \ref{eq:evidence1} it is more numerically accurate to use the estimate
+$$
+\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(1, l)} \tilde p_\text{LA}(x_i(\z), \y) \omega(\z)
+$$
+
 7.
 Given $\{x_i, \tilde p_\text{AQ}(x_i \, | \, \y)\}_{x_i \in \mathcal{X}_i}$ create a spline interpolant to each posterior marginal on the log-scale, from which samples or relevant posterior marginal summaries may be obtained.
+
+Note that clearer notation is required for $\z$ and $\omega(\z)$.
 

### 2023-02-20 src/docs_paper/appendix.Rmd
Alex's feedback round two

@@
 Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest at time $t = T_1$
-Let $x$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
-Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
+Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
+Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age bands.
 As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.

### 2023-02-20 src/docs_paper/paper.Rmd
Alex's feedback round two

@@
 ---
-title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
+title: Fast approximate Bayesian inference for the Naomi HIV surveillance model
 author:
@@
 @eaton2021naomi specify a joint model linking three small-area estimation models.
-The full model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
+The model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
 We consider a simplified version defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
@@
 Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place at time $T_1$.
-Let $x$ index district, $a$ five-year age band, and $s$ sex.
+Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ five-year age band, and  $s \in \mathcal{S}$ sex.
 For ease of notation, let $i$ index the finest district-age-sex division included in the model.
-As data we observe may be aggregated over indices $i$, let $\mathcal{I}$ be a set of $i$ for which observations are reported.
-Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be antiretroviral therapy (ART) coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
-Independent mixed effects models for HIV prevalence and ART coverage in the general population are specified on the logit scale such that
+The data we observe may be aggregated over indices $i$.
+Let $\mathcal{I} \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of  indicies $i$ for which an observation is reported.
+
+Let $N_i \in \mathbb{N}$ be the known, fixed population size.
+We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proporiton of indviduals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
+Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified on such that
 \begin{align*}
@@
 for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
-For the HIV incidence rate we use a mixed effects model on the log scale $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
-Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
-<!-- The kappa is introduced, but you have to be paying close attention to see that it's so that for theta in rho, alpha, kappa -->
-For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes
+For the HIV incidence rate we model on the log scale $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
+Let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
+$$
+\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
+$$
+where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.
+
+For $\theta \in \{\rho, \alpha, \kappa\}$ we calculate the weighted observations
+$$
+\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j},
+$$
+where $j$ indexes individuals across all strata $i \in \mathcal{I}$.
+The design weights are
+$$
+w_j = \frac{1}{\pi_j} \times \frac{1}{\omega_j},
+$$
+where $\pi_j$ is the probability of inclusion and $\omega_j$ is a non-response factor for the particular survey.
+We calculate the observed number of indicator cases as $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
 $$
@@
 $$
-where $j$ indexes individuals across all strata $i \in \mathcal{I}$.
-The corresponding design weights are $w_j = 1 / \pi_j$, where $\pi_j$ is the probability of inclusion in the survey.
-The observed number of indicator cases is then $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$.
+is the Kish effective sample size [@kish1965survey].
+
 For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
@@
 \end{align*}
-To link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ we use
-$$
-\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
-$$
-where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.
 
@@
 Extended latent Gaussian models (ELGMs) [@stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
-Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ depends on, but may not be equal to, $n$.
+Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
 Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
@@
   p(\x, \btheta \, | \, \y)
-  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right).
+  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
 \end{equation*}
@@
 <!-- To-do: define what exactly $p(\y, \x, \btheta)$ is -->
-Given a \textsf{C++} user template `model.cpp` for the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adaptive Gauss-Hermite quadrature.
+Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adaptive Gauss-Hermite quadrature.
 
@@
 Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
-If possible $k \geq 3$ is prefered, though the number of grid points scales exponentially with choice of $k$.
-3. Use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
+If possible $k \geq 3$ is preferred, though the number of grid points scales exponentially with choice of $k$.
+Then use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
 \begin{equation}
@@
 4.
-For $i \in [n]$ generate $l$ nodes $x_i(\z)$, via a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
+For $i \in [n]$ generate $l$ nodes $x_i(\z)$ via a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
 A value of $l \geq 4$ is recommended to enable B-spline interpolation.
-5. For $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
+Then, for $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
 \begin{align*}

### 2023-02-21 src/naomi-simple_compare/comparison.Rmd
Fix jitter height issue, add gradient to scatterplots, fix ...Names issue

@@
 ```
+
+The `beta_rho` parameter is used in other reports as a check, so create the dataframe to save as artefact here:
 
@@
 ks_helper("beta", sw = TRUE)
-ks_helper("beta", sw = TRUE, method1 = "TMB", method2 = "TMB")
+ks_helper("beta", sw = TRUE, method1 = "TMB", method2 = "adam")
 ```
@@
 
-  densityplot <- ggplot(ks_summary, aes(x = .data[[ks_method1]], y = .data[[ks_method2]])) +
-    stat_density_2d(aes(linetype = Type), col = "black") +
-    annotate(geom = "polygon", x = c(-Inf, Inf, Inf), y = c(-Inf, Inf, -Inf), fill = multi.utils::cbpalette()[2], alpha = 0.2) +
-    annotate(geom = "polygon", x = c(-Inf, Inf, -Inf), y = c(-Inf, Inf, Inf), fill = multi.utils::cbpalette()[6], alpha = 0.2) +
-    xlim(0, 0.75) +
-    ylim(0, 0.75) +
+  xy_length <- min(1, max(ks_summary[[ks_method1]], ks_summary[[ks_method2]]) + 0.05)
+
+  x <- y <- seq(0, xy_length, length.out = 20)
+
+  gradient_base <- expand.grid(x, y) %>%
+    mutate(diff = Var1 - Var2) %>%
+    ggplot(aes(x = Var1, y = Var2)) +
+    geom_tile(aes(fill = diff), alpha = 0.7) +
+    scale_fill_gradientn(
+      colours = c("#56B4E9", "white", "#009E73"),
+      rescaler = ~ scales::rescale_mid(.x, mid = 0)
+    )
+
+  densityplot <- gradient_base +
+    stat_density_2d(data = ks_summary, aes(x = .data[[ks_method1]], y = .data[[ks_method2]], linetype = Type), col = "black") +
+    xlim(0, xy_length) +
+    ylim(0, xy_length) +
     geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
-    scale_linetype_manual(values = c("dashed", "solid")) +
-    # scale_alpha_discrete(range = c(0.5, 0.01)) +
-    # guides(alpha = FALSE) +
-    guides(linetype = FALSE) +
+    scale_linetype_manual(values = c("solid", "dashed")) +
     labs(x = ks_method1, y = ks_method2) +
-    theme_minimal()
+    theme_minimal() +
+    guides(fill = "none") +
+    theme(
+      legend.position = "bottom"
+    )
 
@@
     coord_flip() +
-    scale_linetype_manual(values = c("dashed", "solid")) +
+    scale_linetype_manual(values = c("solid", "dashed")) +
     labs(y = "", x = paste0(ks_method1, " - ", ks_method2)) +
@@
   ))
+```
+Compute $p$-values, then plot expecting $\mathcal{U}(0, 1)$.
+If not conclude there is a significant difference.
+
+```{r}
+#' To write!
 ```

### 2023-02-21 src/naomi-simple_compare/comparison.Rmd
Add Rhat, ESS, and correlation between KS and ESS

@@
 ```{r}
-dependency_details(4)
+tmbstan_details <- dependency_details(4)
+tmbstan_details
 ```
@@
 saveRDS(beta_rho$df, "beta_rho.rds")
+```
+
+We will write the `tmbstan` $\hat R$ and ESS for the particular parameter on each plot to help in deciding how seriously to take the MCMC results:
+
+```{r}
+rhats <- bayesplot::rhat(tmbstan$mcmc)
+ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc)
+niter <- 4 * tmbstan_details$niter / tmbstan_details$nthin
+ess <- ess_ratio * niter
 ```
@@
 
+## Correlation between KS values and ESS
+
+Is there any correlation between the value of $\text{KS}(\texttt{method}, \texttt{tmbstan})$ for a particular parameter and the ESS of that parameter from `tmbstan` output?
+
+```{r}
+ks_summary %>%
+  pivot_longer(
+    col = starts_with("KS"),
+    names_to = "Method",
+    values_to = "KS"
+  ) %>%
+  mutate(
+    Method = fct_recode(Method,
+      "adam" = "KS(adam, tmbstan)",
+      "aghq" = "KS(aghq, tmbstan)",
+      "TMB" = "KS(TMB, tmbstan)"
+    )
+  ) %>%
+  group_by(Parameter) %>%
+  mutate(
+    par_num = case_when(
+      max(index) > 1 ~ paste0(Parameter, "[", index, "]"),
+      TRUE ~ Parameter
+    )
+  ) %>%
+  ungroup() %>%
+  left_join(data.frame(ess) %>%
+    tibble::rownames_to_column("par_num")
+  ) %>%
+  ggplot(aes(x = ess, y = KS)) +
+    geom_point(alpha = 0.5) +
+    geom_smooth(method = "lm", color = "#56B4E9") +
+    facet_grid(~ Method) +
+    theme_minimal() +
+    labs(x = "ESS", y = "KS(method, tmbstan)")
+```
+
 # MMD

### 2023-02-21 src/naomi-simple_compare/functions.R

@@ -47,7 +47,13 @@ histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
     scale_color_manual(values = colours) +
     scale_fill_manual(values = colours) +
     theme(legend.position = "none") +
-    labs(title = par_name, subtitle = paste0("Mean = ", mean, ", SD = ", sd, " (from tmbstan)"))
+    labs(
+      title = par_name,
+      subtitle = paste0(
+        "tmbstan: Rhat = ", round(rhats[[par_name]], 3), ", ESS = ", round(ess[[par_name]], 3)
+      ),
+      x = ""
+    )
 
   grid <- seq(from = min(df_compare$samples), to = max(df_compare$samples), length.out = 1000)
 

### 2023-02-21 src/naomi-simple_compare/comparison.Rmd
Ranked by worst nodes for TMB or adam as compared with each other

@@
 
-The following parameters have KS values greater than 0.5 in both dimensions.
+Want to create rank order lists of largest KS differences between methods:
 
 ```{r}
-(big_ks <- ks_summary %>%
-  filter(`KS(aghq, tmbstan)` + `KS(TMB, tmbstan)` > 0.5) %>%
-  pull(Parameter))
+ks_comparison <- function(ks_summary, method1, method2, n) {
+  ks_method1 <- paste0("KS(", method1, ", tmbstan)")
+  ks_method2 <- paste0("KS(", method2, ", tmbstan)")
+  ks_summary[["KS difference"]] <- ks_summary[[ks_method1]] - ks_summary[[ks_method2]]
+  ks_summary %>%
+    filter(Type != "Hyper") %>%
+    arrange(desc(`KS difference`)) %>%
+    head(n = n)
+}
+```
+
+Nodes where `TMB` beats `adam`:
+
+```{r}
+ks_comparison(ks_summary, method1 = "adam", method2 = "TMB", 10)
+```
+
+Nodes where `adam` beats `TMB`:
+
+```{r}
+ks_comparison(ks_summary, method1 = "TMB", method2 = "adam", 10)  
 ```

### 2023-02-21 src/naomi-simple_compare/comparison.Rmd
Add outcome placeholder, and graphs for worst nodes

@@
 
+## Outcome variables
+
+These are the outcome variables which we might be interested in to monitor:
+
+```{r}
+names(tmb$fit$sample)[!(names(tmb$fit$sample) %in% pars)]
+#' To-do: elicit from Jeff in particular which of these we should be focusing on the posterior output of
+#' Guess would be the indicators HIV prevalence (rho_t1_out), ART coverage (alpha_t1_out), HIV incidence (lambda_t1_out)
+```
 # KS plots
@@
 
+```{r}
+histogram_and_ecdf(par = "u_alpha_xs", i = 18)
+histogram_and_ecdf(par = "u_alpha_xs", i = 29)
+histogram_and_ecdf(par = "ui_anc_alpha_x", i = 29)
+```
+
 Nodes where `adam` beats `TMB`:
@@
 ks_comparison(ks_summary, method1 = "TMB", method2 = "adam", 10)  
+```
+
+```{r}
+histogram_and_ecdf(par = "log_or_gamma", i = 18)
+histogram_and_ecdf(par = "ui_anc_rho_x", i = 10)
+histogram_and_ecdf(par = "log_or_gamma", i = 1)
 ```

### 2023-02-22 src/docs_paper/paper.Rmd
Update Osgood citation
Title change (surveillance describes the data not the model)

@@
 ---
-title: Fast approximate Bayesian inference for the Naomi HIV surveillance model
+title: Fast approximate Bayesian inference for small-area estimation of HIV using the Naomi model
 author:
@@
 Inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgoodzimmerman2021statistical].
+Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgood2022statistical].
 Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments, which for the Naomi model, we use for the high-dimensional latent field parameters.

### 2023-02-22 src/docs_paper/paper.Rmd
Small paper edits

@@
 ---
-title: Fast approximate Bayesian inference for small-area estimation of HIV using the Naomi model
+title: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model
 author:
@@
 Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package.
-We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
+We propose a new inference method inspired by adaptive Gauss-Hermite quadrature together with the simplified integrated nested Laplace approximation approach of @wood2020simplified to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
 Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but faster to run.
@@
 To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
-Modelling multiple data sources jointly mitigates the limitations of any single source, increases statistical power, and prompts investigation into possibly conflicting information between sources.
+Modelling multiple data sources jointly has many benefits, including mitigating the limitations of any single source, increasing statistical power, and prompting investigation into any conflicts of information between sources.
 Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates in a yearly process supported by UNAIDS.
@@
 Although this approach has favourable computational properties, it would be preferable to account more fully for hyperparameter uncertainty than is possible in an empirical Bayes framework.
-This has motivated us to look for an approach closer to full Bayesian inference, which is flexible enough to be compatible with the model, and fast enough to be run in production by country teams.
+This has motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
 
-To obtain fast, accurate Bayesian inferences for the Naomi model we develop an inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
+To obtain fast, accurate Bayesian inferences for the Naomi model we develop an inference methodology which combines adaptive Gauss-Hermite quadrature (AGHQ) with the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified.
 INLA is an approximate inference approach based on nested Laplace approximations and numerical quadrature.
@@
 We consider a simplified version defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
-Below we provide an overview of the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
+We provide an overview of the simplified model below, highlighting the aspects which make it a challenge for existing inferential approaches.
 A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
@@
 We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proporiton of indviduals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
-Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified on such that
+Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that
 \begin{align*}
@@
 for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
-For the HIV incidence rate we model on the log scale $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
-Let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
+We model HIV incidence rate on the log scale as $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
+Finally, let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
 $$
@@
 
-For $\theta \in \{\rho, \alpha, \kappa\}$ we calculate the weighted observations
+As data, for $\theta \in \{\rho, \alpha, \kappa\}$ we calculate the weighted, aggregated survey observations
 $$
@@
 where $j$ indexes individuals across all strata $i \in \mathcal{I}$.
-The design weights are
+The design weights are given by
 $$
@@
 For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
-\begin{align*}
-\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
-\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
+\begin{equation*}
+\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \quad
+\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \quad
 \kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
-\end{align*}
+\end{equation*}
 
@@
 where, again, we use weighted aggregates
-\begin{align*}
-\rho^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in \mathcal{I}} \Psi_i}, \\
+\begin{equation*}
+\rho^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in \mathcal{I}} \Psi_i}, \quad
 \alpha^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}},
-\end{align*}
+\end{equation*}
 with $\Psi_i$ the number of pregnant women.
@@
 
-Using one `TMB` template (available in the appendix), we fit the simplified Naomi model to data from Malawi using four inferential approaches: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
-Details regarding particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
+Using one `TMB` C++ user-template (available in the appendix), we fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using four inferential approaches: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
+Further details regarding particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
 

### 2023-02-24 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Work on BayesComp poster

--- 2022-11-01 src/docs_bayescomp-poster/bayescomp-poster.Rmd
@@
 ---
-main_topsize: 0.2
-main_bottomsize: 0.075
-poster_height: "48in"
-poster_width: "36in"
+poster_height: "46.8in"
+poster_width: "33.1in"
 font_family: "Open Sans"
 # Essentials
-title: Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV model
+title: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model
 author:
-  - name: '**Adam Howes**'
+  - name: "Adam Howes"
     affil: 1, 2
@@
     address: Department of Computer Science, University of Oxford
-main_findings:
-  - "Combining **simplified INLA**, automatic differentiation and **adaptive Gauss-Hermite quadrature** for fast and accurate inference in a challenging, real-world problem"
 logoleft_name: "IMP_ML_W_N.png"
-# logoright_name: "UNAIDS-reverse-EN.gif"
+logoright_name: "UW_ML_W_N.png"
 # Style & formatting
 column_numbers: 3
-title_textsize: "60px"
-author_textsize: "1.17em"
+title_textsize: "75px"
+author_textsize: "45px"
 authorextra_textsize: "35px"
 affiliation_textsize: "25px"
-affiliation_textcol: '#00000080'
 caption_fontsize: "20px"
+# Title box
+title_textcol: "#ffffff"
+author_textcol: "#ffffff"
+affiliation_textcol: "#ffffff"
 # Main
@@
 # Poster body options
-primary_colour: '#38761C'
-secondary_colour: '#3b77D8'
+primary_colour: "#009E73"
+secondary_colour: "#3b77D8"
 accent_colour: "#5d19c4"
 body_bgcol: "#ffffff"
-body_textsize: "40px"
+body_textsize: "35px"
 body_textcol: "#000000"
@@
 
-<!-- Using LaTeX to create spacing -->
-$$\\[0.15in]$$
+## Summary
 
-## The Naomi model
+* Approximate Bayesian inference method combining Laplace approximations and adaptive Gauss-Hermite quadrature 
+* Motivated by a challenging, policy-relevant problem in small-area estimation of HIV indicators in sub-Saharan Africa
+* Implemented as a part of the `aghq` R package, allowing use for any model with a `TMB` C++ user template
+
+<!-- 38761C -->
+
+## The Naomi HIV model
+
+* District-level model of HIV indicators which synthesises data from household surveys, antenatal care (ANC) clincs, and routine service provision of antiretroviral therapy (ART) [@eaton2021naomi]
+* Yearly estimation process: model run interactively by country teams using the web-app `naomi.unaids.org`
+* Inference currently conducted with empirical Bayes and a Gaussian approximation to the latent field using Template Model Builder `TMB` [@kristensen2016tmb]
+* Days to get accurate answers with Markov chain Monte Carlo via `tmbstan` [@monnahan2018no]
+* Motivates looking for a fast, approximate approach
+
+## Extended latent Gaussian models
+
+* In a latent Gaussian model (LGM) [@rue2009approximate] the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
+  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs
+* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$
+* Naomi is an ELGM, and not an LGM, because in our model:
+  * ANC indicators are offset from household survey indicators
+  * Incidence depends on prevalence and ART coverage
+  * Observed data may be aggregated
+  * Individuals may attend ART clinics outside their home district
 
@@
 
-## Comparison
+### Background
 
-## Conclusions
+* **Laplace approximation** Integrate out variables using a Gaussian approximation to the denominator e.g.
+$$
+p(\theta, y) \approx \tilde p_\text{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
+$$
+where $\tilde p_\text{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
+* **Adaptive Gauss-Hermite Quadrature** Approximate integrals by $\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)$ with Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} \in p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target
+
+### Our algorithm
+
+* Given a C++ user template for $- \log p(y, x, \theta)$ our algorithm is summarized by Figure 1
+
+```{tikz, tikz-ex, echo=FALSE, fig.align='center', fig.cap = "Flowchart describing the algorithm", fig.ext = 'png', cache=TRUE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=4cm, auto,>=latex', thick, scale = 1]
+\node (1) {$p(\theta, x, y)$};
+\node (2) [right of=1] {$\tilde p_\text{LA}(\theta, y)$};
+\node (3) [below of=1] {$\tilde p_\text{LA}(x_i, \theta, y)$};
+\node (4) [below of=2] {$\tilde p_\text{AQ}(y)$};
+\draw[->] (1) to node {Laplace} (2);
+\end{tikzpicture}
+```
+
+## Inference comparison
+
+* Compare posterior inferences from `TMB`, `aghq`, `adam` and `tmbstan` using Kolmogorov-Smirnov tests on posterior marginals of hyperparameters, latent field, and outputs
+* Figure 2 compares the distances between `TMB` and `adam` to `tmbstan`
+
+## Future work
+
+* More comprehensive inference comparison e.g. maximum mean discrepancy, Pareto-smoothed importance sampling
+* Implement Laplace matrix algebra approximations e.g. @wood2020simplified
+
+<font size="6"> **Interested to read more?** Working notebooks and R code available are available from `github.com/athowes/multi-agyw`. </font>
 
@@
 
-<div id="refs" class="references"></div>
+<font size="5"> <div id="refs" class="references"></div> </font>

### 2023-02-24 src/docs_bayescomp-poster/script.R

@@ -1,5 +1,11 @@
 # orderly::orderly_develop_start("docs_bayescomp-poster")
 # setwd("src/docs_bayescomp-poster")
 
+#' To-do
+#' * [ ] Add MRC GIDA logo
+#' * [ ] Add Oxford logo
+#' * [ ] Add algorithm flowchart
+#' * [ ]
+
 rmarkdown::render("bayescomp-poster.Rmd")
 pagedown::chrome_print("bayescomp-poster.html")

### 2023-02-28 src/naomi-simple_compare/comparison.Rmd
Have a look at large MCMC in comparison

@@
 
+We will write the `tmbstan` $\hat R$ and ESS for the particular parameter on each plot to help in deciding how seriously to take the MCMC results:
+
+```{r}
+rhats <- bayesplot::rhat(tmbstan$mcmc$stanfit)
+ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc$stanfit)
+niter <- 4 * tmbstan_details$niter / tmbstan_details$nthin
+ess <- ess_ratio * niter
+```
+
 The `beta_rho` parameter is used in other reports as a check, so create the dataframe to save as artefact here:
@@
 saveRDS(beta_rho$df, "beta_rho.rds")
-```
-
-We will write the `tmbstan` $\hat R$ and ESS for the particular parameter on each plot to help in deciding how seriously to take the MCMC results:
-
-```{r}
-rhats <- bayesplot::rhat(tmbstan$mcmc)
-ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc)
-niter <- 4 * tmbstan_details$niter / tmbstan_details$nthin
-ess <- ess_ratio * niter
 ```

### 2023-02-28 src/naomi-simple_compare/functions.R

@@ -12,7 +12,7 @@ histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
       data.frame(method = "TMB", samples = as.numeric(tmb$fit$sample[[par]][i, ])),
       data.frame(method = "aghq", samples = as.numeric(aghq$quad$sample[[par]][i, ])),
       data.frame(method = "adam", samples = as.numeric(adam$sample[[par]][i, ])),
-      data.frame(method = "tmbstan", samples = as.numeric(unlist(rstan::extract(tmbstan$mcmc, pars = par)[[par]][, i])))
+      data.frame(method = "tmbstan", samples = as.numeric(unlist(rstan::extract(tmbstan$mcmc$stanfit, pars = par)[[par]][, i])))
     )
   } else {
     par_name <- paste0(par)
@@ -21,7 +21,7 @@ histogram_and_ecdf <- function(par, i = NULL, return_df = FALSE) {
       data.frame(method = "TMB", samples = as.numeric(tmb$fit$sample[[par]])),
       data.frame(method = "aghq", samples = as.numeric(aghq$quad$sample[[par]])),
       data.frame(method = "adam", samples = as.numeric(adam$sample[[par]])),
-      data.frame(method = "tmbstan", samples = as.numeric(unlist(rstan::extract(tmbstan$mcmc, pars = par))))
+      data.frame(method = "tmbstan", samples = as.numeric(unlist(rstan::extract(tmbstan$mcmc$stanfit, pars = par))))
     )
   }
 
@@ -129,7 +129,7 @@ to_ks_df <- function(par, starts_with = FALSE) {
   samples_adam <- adam$sample[unique_par_names]
   samples_adam <- lapply(samples_adam, function(x) as.data.frame(t(x)))
 
-  samples_tmbstan <- rstan::extract(tmbstan$mcmc, pars = unique_par_names)
+  samples_tmbstan <- rstan::extract(tmbstan$mcmc$stanfit, pars = unique_par_names)
   samples_tmbstan <- lapply(samples_tmbstan, function(x) as.data.frame(x))
 
   table <- table(all_par_names)

### 2023-03-02 src/naomi-simple_fit/aghq-grid-scale-up.Rmd
Writing out problem for scaling up grid

---
title: "Scaling up the hyperparameter grid for Naomi"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** We have previously fit empirical Bayes style approaches to the Naomi model, fixing hyperparameter values.

  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure. The problem is that there are too many to use anything dense.
---

# `TMB` fit

Start by fitting the model with `TMB`.
This is quick to do and allows us to get a model we can play with to figure out which hyperparmeters are important.

```{r results="hide"}
source("functions.R")

library(dplyr)
library(forcats)
library(ggplot2)
library(readr)
library(tidyr)
library(purrr)
library(stringr)
library(Matrix)
library(TMB)
library(tmbstan)
library(rstan)
library(aghq)
library(sf)
library(naomi)

area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
pop_agesex <- read_csv(system.file("extdata/demo_population_agesex.csv", package = "naomi"))
survey_hiv_indicators <- read_csv(system.file("extdata/demo_survey_hiv_indicators.csv", package = "naomi"))
art_number <- read_csv(system.file("extdata/demo_art_number.csv", package = "naomi"))
anc_testing <- read_csv(system.file("extdata/demo_anc_testing.csv", package = "naomi"))
pjnz <- system.file("extdata/demo_mwi2019.PJNZ", package = "naomi")
spec <- naomi::extract_pjnz_naomi(pjnz)

scope <- "MWI"
level <- 4
calendar_quarter_t1 <- "CY2016Q1"
calendar_quarter_t2 <- "CY2018Q3"
calendar_quarter_t3 <- "CY2019Q4"
prev_survey_ids  <- c("DEMO2016PHIA", "DEMO2015DHS")
artcov_survey_ids  <- "DEMO2016PHIA"
vls_survey_ids <- NULL
recent_survey_ids <- "DEMO2016PHIA"
artnum_calendar_quarter_t1 <- "CY2016Q1"
artnum_calendar_quarter_t2 <- "CY2018Q3"
anc_clients_year2 <- 2018
anc_clients_year2_num_months <- 9
anc_prevalence_year1 <- 2016
anc_prevalence_year2 <- 2018
anc_art_coverage_year1 <- 2016
anc_art_coverage_year2 <- 2018

naomi_mf <- naomi_model_frame(
  area_merged,
  pop_agesex,
  spec,
  scope = scope,
  level = level,
  calendar_quarter_t1,
  calendar_quarter_t2,
  calendar_quarter_t3
)

naomi_data <- select_naomi_data(
  naomi_mf,
  survey_hiv_indicators,
  anc_testing,
  art_number,
  prev_survey_ids,
  artcov_survey_ids,
  recent_survey_ids,
  vls_survey_ids,
  artnum_calendar_quarter_t1,
  artnum_calendar_quarter_t2,
  anc_prevalence_year1,
  anc_prevalence_year2,
  anc_art_coverage_year1,
  anc_art_coverage_year2
)

compile("naomi_simple.cpp")
dyn.load(dynlib("naomi_simple"))

compile("naomi_simple_x_index.cpp")
dyn.load(dynlib("naomi_simple_x_index"))

tmb_inputs <- prepare_tmb_inputs(naomi_data)
tmb_inputs_simple <- local_exclude_inputs(tmb_inputs)

fit <- local_fit_tmb(tmb_inputs_simple, outer_verbose = TRUE, inner_verbose = FALSE, max_iter = 250, progress = NULL, DLL = "naomi_simple")
```

# First grid ideas

The names of the hyperparameters are:

```{r}
hypers <- names(fit$par)
```

There are `r length(hypers)` of them, if you'd like to count.

We can calculate their posterior standard deviations using `TMB::sdreport` which uses the delta method as follows:

```{r}
sd_out <- sdreport(obj = fit$obj, par.fixed = fit$par, getJointPrecision = TRUE)
sd <- sqrt(diag(sd_out$cov.fixed))

data.frame(par = names(sd), sd = unname(sd)) %>%
  ggplot(aes(x = reorder(par, sd), y = sd)) +
    geom_point(alpha = 0.6) +
    coord_flip() +
    lims(y = c(0, 2)) +
    theme_minimal() +
    labs(x = "", y = "SD from TMB")
```

A first suggestion about how to place grid points would be proportional to the standard deviations above.
For example, we could get a product grid with the minimum number of points by choosing to have:

* `k = 2` points in the dimension with the highest SD
* `k = 1` point in every other dimension

This would look like the following:

```{r}
base_grid <- sd_levels_ghe_grid(
  dim = length(hypers),
  level = c(1, 2),
  cut_off = c(0, 1.9),
  sd = sqrt(diag(sd_out$cov.fixed))
)

base_grid
```

The total number of points is 2:

```{r}
prod(base_grid$level)
```

Let's try fitting this, and see how long it takes:

```{r}
start <- Sys.time()
quad <- fit_aghq(tmb_inputs, basegrid = base_grid)
end <- Sys.time()

end - start
```

This was a nice idea!
It's not perfect though.
Some hyperparameters have higher SD than others in part simply due to the scale that they are on.
For example if the hyper is on the logit scale then looks from the plot above that it will have a higher SD than another hyper on the log scale.
Is there a way to work around this?
We would ideally like to know how much variation in each hyperparameter effects variation in model outputs, and then take that into account in the importance we place on each dimension.
Doing this in any good automated ways would be a nice contribution.

### 2023-03-02 src/naomi-simple_fit/aghq-grid-scale-up.Rmd
Eigendecomposition approach for smaller grid

@@
 Doing this in any good automated ways would be a nice contribution.
+
+What about using the eigenvalues rather than the standard deviation?
+
+```{r}
+m <- sd_out$par.fixed
+C <- sd_out$cov.fixed
+E <- eigen(C)
+V <- diag(E$values)
+D <- E$vectors
+```
+
+The relative contributions of each dimension (similar to a Scree plot) are as follows:
+
+```{r}
+tv_df <- data.frame(
+  n = 1:length(E$values),
+  tv = cumsum(E$values / sum(E$values))
+)
+
+ggplot(tv_df, aes(x = n, y = tv)) +
+  geom_point() +
+  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
+  annotate("text", x = 20, y = 0.875, label = "90% of total variation explained", col = "grey") +
+  scale_y_continuous(labels = scales::percent) +
+  labs(x = "Dimensions of PCA included", y = "Total variation explained") +
+  theme_minimal()
+```
+
+```{r}
+create_grid <- function(d, k) {
+  mvQuad::createNIGrid(dim = d, type = "GHe", level = k)
+}
+
+gg <- create_grid(d = 24, k =)
+```
+
+```{r}
+#' @param gg Quadrature grid
+#' @param D Eigenvectors of the Hessian
+#' @param V Eigenvalues of the Hessian
+#' @param m Mode
+#' @param p The number of dimensions of the PCA included
+pca_rescale_grid <- function(gg, D, V, m, p) {
+  d <- dim(D)[1]
+
+  Dp <- sparseMatrix(
+    i = rep(1:p, times = p),
+    j = sort(rep(1:p,times = p)),
+    x = as.numeric(D[1:p, 1:p]),
+    dims = c(d, d)
+  )
+
+  Vp <- Diagonal(n = d, x = c(sqrt(diag(V)[1:p]), rep(0, (d - p))))
+  DVp <- Dp %*% Vp
+
+  newnodes <- t(DVp %*% t(mvQuad::getNodes(gg)))
+  for(j in 1:d) newnodes[ ,j] <- newnodes[ ,j] + m[j]
+  newweights <- sqrt(det(V)) * mvQuad::getWeights(gg)
+
+  list(nodes = newnodes, weights = newweights)
+}
+```
+
+```{r}
+gg5 <- pca_rescale_grid(gg = create_grid(d = 24, k = 2), D = D, V = V, m = m, p = 5)
+gg7 <- pca_rescale_grid(gg = create_grid(d = 24, k = 2), D = D, V = V, m = m, p = 7)
+gg7 <- pca_rescale_grid(gg = create_grid(d = 24, k = 2), D = D, V = V, m = m, p = 9)
+
+out <- as.data.frame(as.matrix(gg$nodes))
+out$lps <- -1 * apply(out, 1, fit$obj)
+out$w <- gg$weights
+logSumExpWeights(out$lps, out$w)
+```

### 2023-03-03 src/dev_scale-grid/scale-grid.Rmd
Move scale-grid to its own report, some work on PCA

---
title: "Scaling up the hyperparameter grid for Naomi"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** We have previously fit empirical Bayes style approaches to the Naomi model, fixing hyperparameter values.

  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure. The problem is that there are too many to use anything dense.
---

# `TMB` fit

Start by fitting the model with `TMB`.
This is quick to do and allows us to get a model we can play with to figure out which hyperparmeters are important.

```{r}
compile("naomi_simple.cpp")
dyn.load(dynlib("naomi_simple"))

tmb <- readRDS("depends/tmb.rds")
fit <- tmb$fit
```

# Grid ideas

The names of the hyperparameters are:

```{r}
(hypers <- names(fit$par))
```

There are `r length(hypers)` of them, if you'd like to count.

We can calculate their posterior standard deviations using `TMB::sdreport` which uses the delta method as follows:

```{r}
sd_out <- sdreport(obj = fit$obj, par.fixed = fit$par, getJointPrecision = TRUE)
```

The marginal standard deviations are:

```{r}
sd <- sqrt(diag(sd_out$cov.fixed))

data.frame(par = names(sd), sd = unname(sd)) %>%
  ggplot(aes(x = reorder(par, sd), y = sd)) +
    geom_point(alpha = 0.6) +
    coord_flip() +
    lims(y = c(0, 2)) +
    theme_minimal() +
    labs(x = "", y = "SD from TMB")
```

Or, more generally, the covariance matrix is:

```{r}
C <- sd_out$cov.fixed
C_df <- reshape2::melt(as.matrix(C))

C_plot <- C_df %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "C[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

C_plot
```

## Number of points proportional to standard deviation

A first suggestion about how to place grid points would be proportional to the standard deviations above.
For example, we could get a product grid with the minimum number of points by choosing to have:

* `k = 2` points in the dimension with the highest SD
* `k = 1` point in every other dimension

This would look like the following:

```{r}
base_grid <- sd_levels_ghe_grid(
  dim = length(hypers),
  level = c(1, 2),
  cut_off = c(0, 1.9),
  sd = sqrt(diag(sd_out$cov.fixed))
)

base_grid
```

The total number of points for this grid is 2, by design:

```{r}
prod(base_grid$level)
```

Let's try fitting this, and see how long it takes.
First prepare the data:

```{r results="hide"}
area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
pop_agesex <- read_csv(system.file("extdata/demo_population_agesex.csv", package = "naomi"))
survey_hiv_indicators <- read_csv(system.file("extdata/demo_survey_hiv_indicators.csv", package = "naomi"))
art_number <- read_csv(system.file("extdata/demo_art_number.csv", package = "naomi"))
anc_testing <- read_csv(system.file("extdata/demo_anc_testing.csv", package = "naomi"))
pjnz <- system.file("extdata/demo_mwi2019.PJNZ", package = "naomi")
spec <- naomi::extract_pjnz_naomi(pjnz)

scope <- "MWI"
level <- 4
calendar_quarter_t1 <- "CY2016Q1"
calendar_quarter_t2 <- "CY2018Q3"
calendar_quarter_t3 <- "CY2019Q4"
prev_survey_ids  <- c("DEMO2016PHIA", "DEMO2015DHS")
artcov_survey_ids  <- "DEMO2016PHIA"
vls_survey_ids <- NULL
recent_survey_ids <- "DEMO2016PHIA"
artnum_calendar_quarter_t1 <- "CY2016Q1"
artnum_calendar_quarter_t2 <- "CY2018Q3"
anc_clients_year2 <- 2018
anc_clients_year2_num_months <- 9
anc_prevalence_year1 <- 2016
anc_prevalence_year2 <- 2018
anc_art_coverage_year1 <- 2016
anc_art_coverage_year2 <- 2018

naomi_mf <- naomi_model_frame(
  area_merged,
  pop_agesex,
  spec,
  scope = scope,
  level = level,
  calendar_quarter_t1,
  calendar_quarter_t2,
  calendar_quarter_t3
)

naomi_data <- select_naomi_data(
  naomi_mf,
  survey_hiv_indicators,
  anc_testing,
  art_number,
  prev_survey_ids,
  artcov_survey_ids,
  recent_survey_ids,
  vls_survey_ids,
  artnum_calendar_quarter_t1,
  artnum_calendar_quarter_t2,
  anc_prevalence_year1,
  anc_prevalence_year2,
  anc_art_coverage_year1,
  anc_art_coverage_year2
)

tmb_inputs <- prepare_tmb_inputs(naomi_data)
tmb_inputs_simple <- local_exclude_inputs(tmb_inputs)
```

Now fit the model with `aghq`:

```{r}
start <- Sys.time()
quad <- fit_aghq(tmb_inputs, basegrid = base_grid)
end <- Sys.time()

end - start
```

A limitation of this approach is that some hyperparameters have higher SD than others in part due to the different scales that they are on.
For example if the hyper is on the logit scale then (from the plot above) it looks as though it will have a higher SD than another hyper on the log scale.
We would ideally like to know how much variation in each hyperparameter effects variation in model outputs, and then take that into account in the importance we place on each dimension: doing this in any good automated ways would be a nice contribution.

## Eigendecomposition

Another option is to use principle components analysis (PCA) to create a grid on a subspace of $\mathbb{R}^{24}$ which retains most of the variation.

```{r}
m <- sd_out$par.fixed
eigenC <- eigen(C)
lambda <- eigenC$values
Lambda <- diag(lambda)
E <- eigenC$vectors
```

Such that `C` can be obtained by `E %*% diag(lambda) %*% t(E)`:

```{r}
max(C - (E %*% Lambda %*% t(E))) < 10E-12
```

The relative contributions of each dimension are as follows (this is a Scree plot):

```{r}
tv_df <- data.frame(
  n = 1:length(lambda),
  tv = cumsum(lambda / sum(lambda))
)

ggplot(tv_df, aes(x = n, y = tv)) +
  geom_point() +
  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
  annotate("text", x = 20, y = 0.875, label = "90% of total variation explained", col = "grey") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PCA dimensions included", y = "Total variation explained") +
  theme_minimal()
```

So with 5 dimensions included, we can explain `r round(100 * filter(tv_df, n == 5) %>% pull(tv), 1)`% of the total variation.
Or, with 10 dimensions included, that percentage increases to `r round(100 * filter(tv_df, n == 10) %>% pull(tv), 1)`%.

What do the PC loadings look like?

```{r}
reshape2::melt(as.matrix(E)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "E[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "PC loading", y = "Hyper") +
  theme_minimal()
```

```{r}
create_grid <- function(d, k) {
  gg <- mvQuad::createNIGrid(dim = d, type = "GHe", level = k)
  list(nodes = mvQuad::getNodes(gg), weights = as.numeric(mvQuad::getWeights(gg)))
}

gg_full <- create_grid(d = length(hypers), k = 2)
```

Start with `p = 5` dimensions kept of the PCA.
Then create a dense AGHQ grid with `k = 2` on 5 dimensions.

```{r}
p <- 5
d <- dim(E)[1]
gg_p <- create_grid(d = p, k = 2)
```

There are $2^5 = 32$ nodes as follows:

```{r}
(n_nodes <- nrow(gg_p$nodes))
```

Extend this into a 32 x 24 sparse matrix with zeros in the other locations:

```{r}
# i and j specify the non-zero locations in the matrix
(nodes_p <- sparseMatrix(
  i = rep(1:n_nodes, times = p),
  j = rep(1:p, each = n_nodes),
  x = as.numeric(gg_p$nodes),
  dims = c(n_nodes, d)
))
```

How does the reconstruction of the covariance matrix with `r p` components look?

```{r}
E_p <- E[, 1:p]
Lambda_p <- Lambda[1:p, 1:p]
C_p <- E_p %*% Lambda_p %*% t(E_p)

C_plot

C_p_plot <- reshape2::melt(as.matrix(C_p)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "Cp[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

C_plot + C_p_plot
```

Want to do $E_p \Lambda_p^{1/2}$ multiplied by the nodes:

```{r}
E_p <- sparseMatrix(i = rep(1:p, times = p), j = rep(1:p, each = p), x = as.numeric(E[1:p, 1:p]), dims = c(d, d))
Lambda_p <- sparseMatrix(i = 1:p, 1:p, x = diag(Lambda)[1:p], dims = c(d, d))

X_p <- E_p %*% sqrt(Lambda_p)

new_nodes <- X_p %*% t(nodes_p)

t(new_nodes)
```

```{r}
# pca_rescale_grid <- function() {
# 
# }
```

```{r}
# gg5 <- pca_rescale_grid(p = 5)
# gg7 <- pca_rescale_grid(p = 7)
# gg7 <- pca_rescale_grid(p = 9)
```

### 2023-03-03 src/docs_paper/paper.Rmd
Some writing

@@
 
-Using one `TMB` C++ user-template (available in the appendix), we fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using four inferential approaches: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
-Further details regarding particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
+Using one `TMB` C++ user-template (available in the appendix), we fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using four inferential approaches.
+There were: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
+Particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
+For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
+Then, for all methods, we simulated model outputs.
 
@@
 We used three methods to assess the accuracy of posterior distributions produced by each inferential method as compared with those from NUTS: (1) Kolmogorov-Smirnov tests, (2) maximum mean discrepancy, and (3) Pareto-smoothed importance sampling.
+From an applied perspective, our main interest is in comparing the accuracy of model outputs, rather than the internal hyperparameters or latent field parameters.
+That said, obtaining accurate inferences for these internal parameters is also of interest.
 

### 2023-03-07 src/naomi-simple_compare/comparison.Rmd
Work on getting samples of model outputs aligned with mf_out

@@
 
-These are the outcome variables which we might be interested in to monitor:
+These are all of the outcome variables which we might be interested in to monitor:
 
@@
 names(tmb$fit$sample)[!(names(tmb$fit$sample) %in% pars)]
-#' To-do: elicit from Jeff in particular which of these we should be focusing on the posterior output of
-#' Guess would be the indicators HIV prevalence (rho_t1_out), ART coverage (alpha_t1_out), HIV incidence (lambda_t1_out)
 ```
+The most important ones (checked with Jeff) are HIV prevalence (`rho_t1_out`), ART coverage (`alpha_t1_out`), and HIV incidence (`lambda_t1_out`).
+
+* HIV prevalence has `r dim(tmb$fit$sample$rho_t1_out)[1]` variables (rows)
+* ART coverage has `r dim(tmb$fit$sample$alpha_t1_out)[1]` variables (rows)
+* HIV incidence has `r dim(tmb$fit$sample$lambda_t1_out)[1]` variables (rows)
+
+```{r}
+dim(tmb$fit$sample$rho_t1_out)
+dim(tmb$fit$sample$alpha_t1_out)
+dim(tmb$fit$sample$lambda_t1_out)
+```
+
+Within `tmb$naomi_data` there is a dataframe called `mf_out` (modelframe output) which contains the area, age, sex mapping these `r dim(tmb$fit$sample$rho_t1_out)[1]` rows.
+As a sanity check, confirm that the number of areas times number of ages times number of sexes indeed equals the number of rows:
+
+```{r}
+mf_out <- tmb$naomi_data$mf_out
+(n_area <- length(unique(mf_out$area_id)))
+(n_age <- length(unique(mf_out$age_group)))
+(n_sex <- length(unique(mf_out$sex)))
+
+stopifnot(n_area * n_age * n_sex == dim(tmb$fit$sample$rho_t1_out)[1])
+```
+
+It makes sense only to assess inferential accuracy at the finest level so as to avoid double counting.
+
+```{r}
+mf_out$area_id %>% unique()
+mf_out$sex %>% unique()
+mf_out$age_group %>% unique()
+
+mf_out %>%
+  filter(
+    area_id %in% paste0("MWI_4_", 1:32, "_demo"),
+    sex %in% c("male", "female"),
+    age_group %in%
+      c(
+        "Y000_004", "Y005_009", "Y010_014", "Y015_019", "Y020_024", "Y025_029",
+        "Y025_034", "Y030_034", "Y035_039", "Y040_044", "Y045_049", "Y050_054",
+        "Y055_059", "Y060_064", "Y065_069", "Y070_074", "Y075_079", "Y080_999"
+      )
+  ) %>%
+  nrow()
+```
+
 # KS plots

### 2023-03-07 src/dev_scale-grid/scale-grid.Rmd
Updating PCA approach for creating a grid

@@
 
-Such that `C` can be obtained by `E %*% diag(lambda) %*% t(E)`:
+Such that `C` can be obtained by $E \Lambda E^\top$, or in code `E %*% diag(lambda) %*% t(E)`:
 
@@
 }
-
-gg_full <- create_grid(d = length(hypers), k = 2)
 ```
 
-Start with `p = 5` dimensions kept of the PCA.
+Start with `s = 5` dimensions kept of the PCA.
 Then create a dense AGHQ grid with `k = 2` on 5 dimensions.
@@
 ```{r}
-p <- 5
+s <- 5
 d <- dim(E)[1]
-gg_p <- create_grid(d = p, k = 2)
+gg_s <- create_grid(d = s, k = 2)
 ```
@@
 ```{r}
-(n_nodes <- nrow(gg_p$nodes))
+(n_nodes <- nrow(gg_s$nodes))
 ```
 
-Extend this into a 32 x 24 sparse matrix with zeros in the other locations:
+How does the reconstruction of the covariance matrix with `r s` components look?
 
 ```{r}
-# i and j specify the non-zero locations in the matrix
-(nodes_p <- sparseMatrix(
-  i = rep(1:n_nodes, times = p),
-  j = rep(1:p, each = n_nodes),
-  x = as.numeric(gg_p$nodes),
-  dims = c(n_nodes, d)
-))
-```
-
-How does the reconstruction of the covariance matrix with `r p` components look?
-
-```{r}
-E_p <- E[, 1:p]
-Lambda_p <- Lambda[1:p, 1:p]
-C_p <- E_p %*% Lambda_p %*% t(E_p)
+E_s <- E[, 1:s]
+Lambda_s <- Lambda[1:s, 1:s]
+C_s <- E_s %*% Lambda_s %*% t(E_s)
 
@@
 
-C_p_plot <- reshape2::melt(as.matrix(C_p)) %>%
+C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
   ggplot(aes(x = Var1, y = Var2, fill = value)) +
-  labs(x = "", y = "", fill = "Cp[i, j]") +
+  labs(x = "", y = "", fill = "C_s[i, j]") +
   geom_tile() +
@@
 
-C_plot + C_p_plot
+C_plot + C_s_plot
 ```
 
-Want to do $E_p \Lambda_p^{1/2}$ multiplied by the nodes:
+Let $a \in \mathbb{R}^d$ and $b \in \mathbb{R}^s$.
+The PCA provides a mapping $f: \mathbb{R}^d \to \mathbb{R}^s$ which compresses each $a$ to a smaller $b$ given by $f(a) = (E_s \Lambda_s^{1/2})^\top a = b$.
+Then $f^{-1}$ is given by $f^{-1}(b) = E_s \Lambda_s^{-1/2} b = a$, using the fact that $E_s^\top = E_s^{-1}$.
+So we need to multiply our nodes by $E_s \Lambda_s^{- 1/2}$ as follows:
 
 ```{r}
-E_p <- sparseMatrix(i = rep(1:p, times = p), j = rep(1:p, each = p), x = as.numeric(E[1:p, 1:p]), dims = c(d, d))
-Lambda_p <- sparseMatrix(i = 1:p, 1:p, x = diag(Lambda)[1:p], dims = c(d, d))
+a_nodes <- gg_s$nodes
+b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))
 
-X_p <- E_p %*% sqrt(Lambda_p)
+dim(a_nodes)
+dim(b_nodes)
+```
 
-new_nodes <- X_p %*% t(nodes_p)
+Add on the modes:
 
-t(new_nodes)
+```{r}
+for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
+```
+
+Adjust the weights:
+
+```{r}
+gg_s$weights
+det(Lambda)
+det(Lambda_s)
 ```
@@
 ```{r}
-# pca_rescale_grid <- function() {
-# 
-# }
+#' @param m Mean vector
+#' @param C Covariance matrix
+#' @param s Small grid dimension
+#' @param k Number of points per small grid dimension
+pca_rescale_grid <- function(m, C, s, k) {
+  lambda <- eigenC$values
+  Lambda <- diag(lambda)
+  E <- eigenC$vectors
+  E_s <- E[, 1:s] 
+  gg_s <- create_grid(d = s, k = k)
+  a_nodes <- gg_s$nodes
+  b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))
+  for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
+  b_weights <- gg_s$weights * sqrt(det(Lambda))
+  return(list(nodes = b_nodes, weights = b_weights))
+}
 ```
@@
 ```{r}
-# gg5 <- pca_rescale_grid(p = 5)
-# gg7 <- pca_rescale_grid(p = 7)
-# gg7 <- pca_rescale_grid(p = 9)
+gg5 <- pca_rescale_grid(m = m, C = C, s = 5, k = 2)
+gg7 <- pca_rescale_grid(m = m, C = C, s = 7, k = 2)
+gg9 <- pca_rescale_grid(m = m, C = C, s = 9, k = 2)
 ```

### 2023-03-08 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Work on poster for BayesComp

--- 2022-11-21 src/docs_bayescomp-poster/bayescomp-poster.Rmd
@@
 ---
-main_topsize: 0.18
-main_bottomsize: 0.03
-poster_height: "48in"
-poster_width: "36in"
+poster_height: "46.8in"
+poster_width: "33.1in"
 font_family: "Open Sans"
 # Essentials
-title: Integrated nested Laplace approximations for extended latent Gaussian models, with application to the Naomi HIV model
+title: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model
+# titlebox_height: "10%"
 author:
-  - name: '**Adam Howes**'
+  - name: "Adam Howes"
     affil: 1, 2
     main: true
-    twitter: adamhowes
     email: ath19@ic.ac.uk
@@
     address: Department of Computer Science, University of Oxford
-main_findings: "Simplified INLA, automatic differentiation and adaptive Gauss-Hermite quadrature for fast and accurate approximate inference"
-logoleft_name: "IMP_ML_W_N.png"
-# logoright_name: "UNAIDS-reverse-EN.gif"
+logoleft_name: "mrc-gida-icl-uw-uo.png"
 # Style & formatting
 column_numbers: 3
-title_textsize: "60px"
-author_textsize: "1.17em"
+title_textsize: "75px"
+author_textsize: "50px"
 authorextra_textsize: "35px"
-affiliation_textsize: "25px"
-affiliation_textcol: '#00000080'
+affiliation_textsize: "35px"
 caption_fontsize: "20px"
+# Title box
+title_textcol: "#000000"
+author_textcol: "#000000"
+affiliation_textcol: "#000000"
 # Main
@@
 # Poster body options
-primary_colour: '#38761C'
-secondary_colour: '#3b77D8'
-accent_colour: "#5d19c4"
+primary_colour: "#ffffff"
+secondary_colour: "#000000"
+accent_colour: "#000000"
 body_bgcol: "#ffffff"
@@
 
-<!-- Using LaTeX to create spacing -->
-$$\\[0.15in]$$
+```{css, echo=FALSE}
+div.logo_left{
+  width: 35%;
+}
+div.poster_title{
+  width: 60%;
+}
 
-## The Naomi model
+```
 
-$$\\[0.15in]$$
+## Summary
+
+* Approximate Bayesian inference method combining Laplace approximations and adaptive Gauss-Hermite quadrature 
+* Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
+* Implemented as a part of the `aghq` R package, allowing flexible use for any model with a `TMB` C++ user template
+
+<!-- 38761C -->
+
+## The Naomi HIV model
+
+* District-level model of HIV indicators synthesising data from household surveys, antenatal care (ANC) clincs, and routine service provision of antiretroviral therapy (ART) [@eaton2021naomi]
+* Yearly estimation process: model run interactively by country teams using the web-app `naomi.unaids.org`
+* Inference conducted in minutes using empirical Bayes (EB) and a Gaussian approximation via Template Model Builder `TMB` [@kristensen2016tmb]
+* Would take days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no]
+* Motivates looking for a fast, approximate approach
+
+## Extended latent Gaussian models
+
+* Three stage hierarchical model with observations $y$, latent field $x$ and hyperparameters $\theta$
+* In a latent Gaussian model (LGM) [@rue2009approximate] the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
+  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs
+* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$
+* Naomi is an ELGM, not an LGM, because we model:
+  * ANC indicators offset from household survey indicators
+  * Incidence as depending on prevalence and ART coverage
+  * Observed data as aggregated finer processes
+  * Individuals as possibly attending ART clinics outside their home district
+* ELGM precision matrices less sparse than those of LGMs
 
@@
 
-### Laplace approximation
-
-### Adaptive Gauss-Hermite Quadrature
-
-Approximate integrals by 
+* Laplace approximation: Integrate out variables using a Gaussian approximation to the denominator
+$$
+p(\theta, y) \approx \tilde p_\text{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
+$$
+where $\tilde p_\text{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
+* Adaptive Gauss-Hermite Quadrature:
 $$
@@
 $$
-with **Gauss-Hermite quadrature** rule $z \in \mathcal{Q}(m, k)$ **adapted** based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} \in p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target.
+where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ is adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target
+* Given a C++ user template for $- \log p(y, x, \theta)$ our algorithm is summarized by Figure 1
 
-### Our algorithm
+```{tikz, tikz-ex, echo=FALSE, fig.align='center', fig.cap = "Flowchart describing the algorithm", fig.ext = 'png', cache=TRUE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=4cm, auto,>=latex', thick, scale = 1]
+\node (1) {$p(\theta, x, y)$};
+\node (2) [right of=1] {$\tilde p_\text{LA}(\theta, y)$};
+\node (3) [below of=1] {$\tilde p_\text{LA}(x_i, \theta, y)$};
+\node (4) [below of=2] {$\tilde p_\text{AQ}(y)$};
+\draw[->] (1) to node {Laplace} (2);
+\draw[->] (1) to node {Laplace} (3);
+\draw[->] (2) to node {Quadrature} (4);
+\end{tikzpicture}
+```
 
-Given C++ user template for $- \log p(y, x, \theta)$:
+## Application to Malawi data
 
-## Comparison
+```{r fig2, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model outputs for adults 15-49 in January 2016.'}
+knitr::include_graphics("fig2.png")
+```
 
-## Conclusions
+* $\text{dim}(x)$ is 491 and $\text{dim}(\theta)$ is 24
+* Figure \@ref(fig:fig2) illustrates example model outputs: HIV prevalence, ART coverage, HIV incidence, and number of new infections, at the district level
+
+```{r fig3, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Inference reults and ECDF comparison for one $x_i$.'}
+knitr::include_graphics("fig3.png")
+```
+
+* Compare inferences from `TMB`, `aghq`, `adam` and `tmbstan` using Kolmogorov-Smirnov (KS) tests on hyperparameter, latent field, and output posterior marginals
+* Figure \@ref(fig:fig3) illustrates the ECDF difference between `TMB` and `adam` to `tmbstan` for one particular node in the model
+
+## Future directions
+
+* Scaling up the hyperparameter grid beyond EB
+  * Any dense grid would be $k^{24}$ i.e. impractical
+  * Sparse grids? Good subspaces of $\mathbb{R}^{24}$?
+* Add Laplace matrix algebra approximations [@wood2020simplified]
+* More comprehensive inference comparison e.g. maximum mean discrepancy, Pareto-smoothed importance sampling
+
+<font size="6"> **Interested to read more?** Working notebooks and R code available from `github.com/athowes/multi-agyw`. </font>
 

### 2023-03-08 src/docs_bayescomp-poster/figs.R

@@ -1,0 +1,208 @@
+#' Figure 1
+
+#' Created in the poster
+
+#' Figure 2
+tmb <- readRDS("depends/tmb.rds")
+outputs <- tmb$outputs
+
+#' Adapted from figure from Naomi paper (Eaton 2021)
+#' https://github.com/jeffeaton/naomi-model-paper/blob/master/analysis.R
+
+indicators <- naomi:::add_output_labels(outputs) %>%
+  left_join(
+    select(outputs$meta_area, area_level, area_id, center_x, center_y),
+    by = c("area_level", "area_id")
+  ) %>%
+  sf::st_as_sf() %>%
+  mutate(
+    area_level_label = fct_reorder(area_level_label, area_level),
+    age_group_label = fct_reorder(age_group_label, as.integer(factor(age_group)))
+  )
+
+fig2adata <- indicators %>%
+  filter(
+    indicator == "prevalence",
+    age_group == "Y015_049",
+    sex == "both",
+    calendar_quarter == "CY2016Q1",
+    area_level == 4
+  )
+
+fig2a <- fig2adata %>%
+  ggplot(aes(fill = mean)) +
+  geom_sf(size = 0.1) +
+  scale_fill_viridis_c(
+    option = "C", direction = -1,
+    begin = 0.1, end = 0.9,
+    labels = scales::label_percent(1)
+  ) +
+  expand_limits(fill = 0) +
+  labs(fill = "HIV prevalence") +
+  coord_sf(expand = FALSE) +
+  theme_minimal() +
+  theme(
+    panel.grid = element_blank(),
+    panel.spacing = unit(0, "lines"),
+    axis.text = element_blank(),
+    strip.text = element_text(face = "bold", size = rel(0.85)),
+    legend.title = element_text(size = rel(0.7)),
+    legend.text = element_text(size = rel(0.7)),
+    plot.tag = element_text(face = "bold")
+  )
+
+fig2bdata <- indicators %>%
+  filter(
+    indicator == "art_coverage",
+    age_group == "Y015_049",
+    calendar_quarter == "CY2016Q1",
+    sex == "both",
+    area_level == 4
+  )
+
+fig2b <- fig2bdata %>%
+  ggplot(aes(fill = mean)) +
+  geom_sf(size = 0.1) +
+  scale_fill_viridis_c(
[…]

### 2023-03-08 src/docs_bayescomp-poster/script.R

@@ -1,11 +1,9 @@
 # orderly::orderly_develop_start("docs_bayescomp-poster")
 # setwd("src/docs_bayescomp-poster")
 
-#' To-do
-#' * [ ] Add MRC GIDA logo
-#' * [ ] Add Oxford logo
-#' * [ ] Add algorithm flowchart
-#' * [ ]
+#' Create figures
+source("figs.R")
 
+#' Create poster
 rmarkdown::render("bayescomp-poster.Rmd")
 pagedown::chrome_print("bayescomp-poster.html")

### 2023-03-08 src/naomi-simple_compare/comparison.Rmd
Save beta_anc_rho too

@@
 
-The `beta_rho` parameter is used in other reports as a check, so create the dataframe to save as artefact here:
+The `beta_rho` and `beta_anc_rho` parameters are used in other reports, so create dataframes here to save as artefacts:
 
@@
 saveRDS(beta_rho$df, "beta_rho.rds")
+
+beta_anc_rho <- histogram_and_ecdf("beta_anc_rho", return_df = TRUE)
+saveRDS(beta_anc_rho$df, "beta_anc_rho.rds")
 ```

### 2023-03-09 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Almost finished with the poster

--- 2023-01-20 src/docs_bayescomp-poster/bayescomp-poster.Rmd
@@
 ---
-main_topsize: 0.08
-main_bottomsize: 0.03
-poster_height: "36in"
-poster_width: "48in"
+poster_height: "46.8in"
+poster_width: "33.1in"
 font_family: "Open Sans"
 # Essentials
-title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
+title: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model
+# titlebox_height: "10%"
 author:
@@
     main: true
-    twitter: adamhowes
     email: ath19@ic.ac.uk
@@
     address: Department of Computer Science, University of Oxford
-logoleft_name: "IMP_ML_W_N.png"
-logoright_name: "UW_ML_W_N.png"
+logoleft_name: "mrc-gida-icl-uw-uo.png"
 # Style & formatting
 column_numbers: 3
-title_textsize: "65px"
-author_textsize: "45px"
+title_textsize: "75px"
+author_textsize: "55px"
 authorextra_textsize: "35px"
-affiliation_textsize: "25px"
+affiliation_textsize: "35px"
 caption_fontsize: "20px"
 # Title box
-title_textcol: "#ffffff"
-author_textcol: "#ffffff"
-affiliation_textcol: "#ffffff"
+title_textcol: "#000000"
+author_textcol: "#000000"
+affiliation_textcol: "#000000"
 # Main
@@
 # Poster body options
-primary_colour: '#3b77D8'
-secondary_colour: '#3b77D8'
-accent_colour: "#5d19c4"
 body_bgcol: "#ffffff"
-body_textsize: "35px"
+body_textsize: "36px"
 body_textcol: "#000000"
@@
 sectitle_textsize: "40px"
+link_col: "#000000"
 # Standard options
@@
 
+```{css, echo=FALSE}
+div.logo_left{
+  width: 35%;
+  background-color: #E5E5E5;
+}
+div.poster_title{
+  width: 60%;
+}
+div.title_container{
+  background-color: #E5E5E5;
+}
+code{
+  background-color: #ffffff;
+  color: #000000;
+}
+.poster_body{
+  column-rule-color: #ffffff;
+}
+```
+
 ## Summary
 
-* Approximate Bayesian inference method using adaptive Gauss-Hermite quadrature and Laplace approximation
-* Implemented as a part of the `aghq` package, allowing easy use for any model with a `TMB` C++ user template
-* Motivated by a challenging problem in evidence synthesis for small-area estimation of HIV
+* Approximate Bayesian inference method using Laplace approximations and adaptive Gauss-Hermite quadrature 
+* Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
+* Implemented as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a `TMB` C++ user template
 
@@
 
-* Spatio-temporal small-area estimation model of HIV indicators which combines data from household surveys, antenatal care clincs, and routine service provision of antiretroviral therapy [@eaton2021naomi]
-* Yearly estimation process where the model is run interactively by country teams using the web-app `naomi.unaids.org`
-* Inference currently conducted with empirical Bayes and a Gaussian approximation to the latent field using Template Model Builder `TMB` [@kristensen2016tmb]
-* Model incompatible with the `R-INLA` implementation of integrated nested Laplace approximations [@rue2009approximate] and takes days to get accurate answers with Markov chain Monte Carlo via `tmbstan` [@monnahan2018no]
-* Fits into the recently described class of extended latent Gaussian models [@stringer2022fast]
+* District-level model of HIV indicators [@eaton2021naomi] which synthesises data from household surveys, antenatal care (ANC) clincs, and routine service provision of antiretroviral therapy (ART)
+  * Combining evidence from multiple data sources helps overcome the limitations of any one
+* Yearly estimation process: model run interactively by country teams using a web-app
+  * Figure \@ref(fig:fig1) illustrates the seven stages of using the app
+* Inference conducted in minutes using empirical Bayes (EB) and a Gaussian approximation via Template Model Builder `TMB` [@kristensen2016tmb]
+* It would take days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no], and this is not practical in this setting
+* Motivates looking for a fast, approximate approach, that takes uncertainty in hyperparameters into account
+
+```{r fig1, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model fitting occurs interactively in stages, taken from `naomi.unaids.org`'}
+knitr::include_graphics("fig1.png")
+```
+
+## Extended latent Gaussian models
+
+* Latent Gaussian model (LGM) [@rue2009approximate]: three stage hierarchical model with observations $y$, Gaussian latent field $x$ and hyperparameters $\theta$
+* In an LGM the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
+  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs
+* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$
+* Naomi is an ELGM, not an LGM, because we model:
+  * ANC indicators offset from household survey indicators
+  * Incidence depending on prevalence and ART coverage
+  * Observed data as aggregated finer processes
+  * Individuals as possibly attending ART clinics outside their home district
+* ELGM precision matrices less sparse than those of LGMs
+* We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
 
@@
 
-### Laplace approximation
+* Laplace approximation: Integrate out variables using a Gaussian approximation to the denominator
+$$
+p(\theta, y) \approx \tilde p_\text{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
+$$
+where $\tilde p_\text{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
+  * Use automatic differentiation via `CppAD` in `TMB`
+* Adaptive Gauss-Hermite Quadrature:
+$$
+\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)
+$$
+where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(\dim(\theta), k)$ is adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target
+  * $k$ is the number of quadrature points per dimension
+* Algorithm (called `adam` for now) summarized by Figure \@ref(fig:fig2)
 
-* Integrate out variables using a Gaussian approximation in the denominator
-$$
-p(\theta, y) \approx \tilde p_\text{LA}(\theta, y)
-$$
-where
+```{tikz, fig2, echo=FALSE, fig.align='center', fig.cap = "Flowchart describing the algorithm", fig.ext = 'png', cache=TRUE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2.5cm, auto,>=latex', thick, scale = 1]
+\node (1) {$p(\theta, x, y)$};
+\node (2) [below of=1] {$\tilde p_\text{LA}(x_i, \theta, y)$};
+\node (3) [below of=2] {$\tilde p_\text{LA}(x_i, y)$};
+\node (4) [below of=3] {$\tilde p_\text{AQ}(x_i \, | \, y)$};
+\node (5) [below of=4] {Output};
 
-### Adaptive Gauss-Hermite Quadrature
+\node (6) [right of=1, xshift=1cm] {$\tilde p_\text{LA}(\theta, y)$};
+\node (7) [below of=6] {$\tilde p_\text{AQ}(y)$};
+\node (8) [below of=7] {$\tilde p_\text{AQ}(\theta \, | \, y)$};
+\node (9) [below of=8] {$\tilde p_\text{AQ}(\theta_j \, | \, y)$};
+\node (10) [below of=9] {Output};
 
-* Approximate integrals by $\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)$ with Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} \in p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target
-
-### Our algorithm
-
-Given C++ user template for $- \log p(y, x, \theta)$:
-
-```{r echo=FALSE, fig.align='center', out.width='100%'}
-knitr::include_graphics("algorithm-flowchart.png")
+\draw[->] (1) to node {Laplace} (2);
+\draw[->] (2) to node {AGHQ} (3);
+\draw[->] (3) to node {Normalise} (4);
+\draw[->] (4) to node {Spline} (5);
+\draw[->] (1) to node {Laplace} (6);
+\draw[->] (6) to node {AGHQ} (7);
+\draw[->] (7) to node {Normalise} (8);
+\draw[->] (8) to node {Marginalise} (9);
+\draw[->] (9) to node {Spline} (10);
+\end{tikzpicture}
 ```
 
-## Inference comparison
+## Application to Malawi data
 
-* Compare posterior inferences from `TMB`, `aghq` and `tmbstan` using Kolmogorov-Smirnov tests on posterior marginals
-* Find that...
+```{r fig3, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model outputs for adults 15-49 in January 2016.'}
+knitr::include_graphics("fig3.png")
+```
 
-<font size="6"> **Funding** AH was supported by the EPSRC and Bill & Melinda Gates Foundation. This research was supported by the MRC Centre for Global Infectious Disease Analysis. </font>
+* Relatively small country but still a large model
+  * $\text{dim}(x)$ is 491 and $\text{dim}(\theta)$ is 24
+* Fit inference methods `TMB`, `aghq`, `adam` and `tmbstan`
+* Figure \@ref(fig:fig3) illustrates example model outputs: HIV prevalence, ART coverage, HIV incidence, and number of new infections, at the district level
+
+```{r fig4, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Inference reults and ECDF comparison for one $x_i$.'}
+knitr::include_graphics("fig4.png")
+```
+
+* Compare inferences using Kolmogorov-Smirnov (KS) tests on hyperparameter, latent field, and output posterior marginals
+* Figure \@ref(fig:fig4) illustrates the ECDF differences for one particular node in the model with $\text{KS}(\texttt{TMB}, \texttt{tmbstan}) = 0.09$ and $\text{KS}(\texttt{adam}, \texttt{tmbstan}) = 0.02$
+
+## Future directions
+
+* Scaling up the hyperparameter grid beyond EB, when any dense grid would be impractical ($k^{24}$ nodes)
+  * Sparse grids
+  * Dense grids on a subspace
+* Add Laplace matrix algebra approximations [@wood2020simplified]
+* More comprehensive inference comparisons
+  * Maximum mean discrepancy
+  * Pareto-smoothed importance sampling
+
+**Interested?** Working notebooks and R code available from `github.com/athowes/elgm-inf`, or get in touch:
+
+* `r fontawesome::fa("home", fill = "black")` `athowes.github.io`
+* `r fontawesome::fa("envelope", fill = "black")` `ath19@ic.ac.uk`
+* `r fontawesome::fa("twitter", fill = "black")` `adamhowes`
+
+**Funding** AH was supported by the EPSRC and Bill & Melinda Gates Foundation. This research was supported by the MRC Centre for Global Infectious Disease Analysis.
 
@@
 
-<font size="6"> <div id="refs" class="references"></div> </font>
+<font size="5"> <div id="refs" class="references"></div> </font>

### 2023-03-09 src/docs_bayescomp-poster/figs.R

@@ -1,8 +1,12 @@
 #' Figure 1
 
-#' Created in the poster
+#' A screenshot from https://naomi.unaids.org/
 
 #' Figure 2
+
+#' Created in the poster
+
+#' Figure 3
 tmb <- readRDS("depends/tmb.rds")
 outputs <- tmb$outputs
 
@@ -20,7 +24,7 @@ indicators <- naomi:::add_output_labels(outputs) %>%
     age_group_label = fct_reorder(age_group_label, as.integer(factor(age_group)))
   )
 
-fig2adata <- indicators %>%
+fig3adata <- indicators %>%
   filter(
     indicator == "prevalence",
     age_group == "Y015_049",
@@ -29,7 +33,7 @@ fig2adata <- indicators %>%
     area_level == 4
   )
 
-fig2a <- fig2adata %>%
+fig3a <- fig3adata %>%
   ggplot(aes(fill = mean)) +
   geom_sf(size = 0.1) +
   scale_fill_viridis_c(
@@ -51,7 +55,7 @@ fig2a <- fig2adata %>%
     plot.tag = element_text(face = "bold")
   )
 
-fig2bdata <- indicators %>%
+fig3bdata <- indicators %>%
   filter(
     indicator == "art_coverage",
     age_group == "Y015_049",
@@ -60,7 +64,7 @@ fig2bdata <- indicators %>%
     area_level == 4
   )
 
-fig2b <- fig2bdata %>%
+fig3b <- fig3bdata %>%
   ggplot(aes(fill = mean)) +
   geom_sf(size = 0.1) +
   scale_fill_viridis_c(
@@ -80,7 +84,7 @@ fig2b <- fig2bdata %>%
     plot.tag = element_text(face = "bold")
   )
 
-fig2cdata <- indicators %>%
+fig3cdata <- indicators %>%
   filter(
     indicator %in% c("infections", "incidence"),
     age_group == "Y015_049",
@@ -89,7 +93,7 @@ fig2cdata <- indicators %>%
     area_level == 4
   )
 
-fig2c <- fig2cdata %>%
+fig3c <- fig3cdata %>%
   sf::st_drop_geometry() %>%
   tidyr::pivot_wider(c(area_id, area_name), names_from = indicator, values_from = mean) %>%
   left_join(
@@ -125,32 +129,32 @@ fig2c <- fig2cdata %>%
   )
 
 ggsave(
-  "fig2.png",
[…]

### 2023-03-10 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Finish poster

@@
 ---
-poster_height: "46.8in"
-poster_width: "33.1in"
+poster_height: "1189mm"
+poster_width: "841mm"
 font_family: "Open Sans"
@@
 * Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
-* Implemented as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a `TMB` C++ user template
+* Looking to implement as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a `TMB` C++ user template
 
@@
 
-* District-level model of HIV indicators [@eaton2021naomi] which synthesises data from household surveys, antenatal care (ANC) clincs, and routine service provision of antiretroviral therapy (ART)
+* District-level model of HIV indicators [@eaton2021naomi] which synthesises data from household surveys, antenatal care (ANC) clinics, and routine service provision of antiretroviral therapy (ART)
   * Combining evidence from multiple data sources helps overcome the limitations of any one
-* Yearly estimation process: model run interactively by country teams using a web-app
+  * Small-area estimation methods to overcome small district-level sample sizes
+* Yearly estimation process: model run interactively by country teams using a web-app `naomi.unaids.org`
   * Figure \@ref(fig:fig1) illustrates the seven stages of using the app
@@
 
-```{r fig1, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model fitting occurs interactively in stages, taken from `naomi.unaids.org`'}
+```{r fig1, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model fitting occurs interactively in stages.'}
 knitr::include_graphics("fig1.png")
@@
 
-* Latent Gaussian model (LGM) [@rue2009approximate]: three stage hierarchical model with observations $y$, Gaussian latent field $x$ and hyperparameters $\theta$
+* Latent Gaussian models (LGMs) [@rue2009approximate] are three stage hierarchical models with observations $y$, Gaussian latent field $x$ and hyperparameters $\theta$
 * In an LGM the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
-  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs
-* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$
-* Naomi is an ELGM, not an LGM, because we model:
-  * ANC indicators offset from household survey indicators
-  * Incidence depending on prevalence and ART coverage
-  * Observed data as aggregated finer processes
-  * Individuals as possibly attending ART clinics outside their home district
-* ELGM precision matrices less sparse than those of LGMs
+  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs, because ELGM precision matrices are not as sparse
+* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$ and $\mathcal{J_i}$ is some set of indices
+  * Allows a higher degree of non-linearity in the model
+* Naomi is an ELGM, not an LGM, because it includes complex dependency structures:
+  1. ANC indicators offset from household survey
+  2. Incidence depends on prevalence and ART coverage
+  3. Observed data are aggregated finer processes
+  4. Allow attendance of ART clinics outside home district
 * We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
+* Though we focus on Naomi here, the HIV Inference Group (`hiv-inference.org`) works on many other complex models, challenging for existing Bayesian inference methods, which require high customisability
+
+$$\\[0.4in]$$
 
@@
 
-* Laplace approximation: Integrate out variables using a Gaussian approximation to the denominator
+* **Laplace approximation** Integrate out variables using a Gaussian approximation to the denominator
 $$
@@
   * Use automatic differentiation via `CppAD` in `TMB`
-* Adaptive Gauss-Hermite Quadrature:
+* **Adaptive Gauss-Hermite Quadrature**
 $$
@@
 $$
-where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(\dim(\theta), k)$ is adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target
-  * $k$ is the number of quadrature points per dimension
+where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(\dim(\theta), k)$ with $k$ points per dimension is adapted based upon
+  * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$
+  * The lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$
 * Algorithm (called `adam` for now) summarized by Figure \@ref(fig:fig2)
+  * Where possible, previously calculated quantities and quadrature rules are reused
 
@@
 
-```{r fig3, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model outputs for adults 15-49 in January 2016.'}
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+time_taken <- readr::read_csv("depends/time_taken.csv")
+```
+
+```{r fig3, echo=FALSE, fig.align='center', out.width='100%', fig.cap='District-level model outputs for adults 15-49 in January 2016. Adapted from Eaton et al. 2021.'}
 knitr::include_graphics("fig3.png")
@@
 
-* Relatively small country but still a large model
-  * $\text{dim}(x)$ is 491 and $\text{dim}(\theta)$ is 24
-* Fit inference methods `TMB`, `aghq`, `adam` and `tmbstan`
+* Relatively small country but still a large model: latent field $\text{dim}(x) = 491$, hyperparameters $\text{dim}(\theta) = 24$
+
+$$\\[0.2in]$$
+
+* Fit four inference methods (using one [!] C++ template):
+  * `TMB` (`r round(time_taken$TMB, 0)` mins)
+  * `aghq` (`r round(time_taken$aghq, 0)` mins): $k = 1$
+  * `adam` (`r round(time_taken$adam, 0)` min): $k = 1$
+  * `tmbstan` (`r round(time_taken$tmbstan, 1)` days): 4 chains of 100,000 thinned by 40 (required for good diagnostics)
 * Figure \@ref(fig:fig3) illustrates example model outputs: HIV prevalence, ART coverage, HIV incidence, and number of new infections, at the district level
@@
 
-* Compare inferences using Kolmogorov-Smirnov (KS) tests on hyperparameter, latent field, and output posterior marginals
-* Figure \@ref(fig:fig4) illustrates the ECDF differences for one particular node in the model with $\text{KS}(\texttt{TMB}, \texttt{tmbstan}) = 0.09$ and $\text{KS}(\texttt{adam}, \texttt{tmbstan}) = 0.02$
+* Compare hyperparameter, latent field, and output posterior marginals based on maximum ECDF difference (Kolmogorov-Smirnov test) 
+* Figure \@ref(fig:fig4) illustrates this approach for one node in the model with $\text{KS}(\texttt{TMB}) = 0.09$ and $\text{KS}(\texttt{adam}) = 0.02$
 
@@
 
-* Scaling up the hyperparameter grid beyond EB, when any dense grid would be impractical ($k^{24}$ nodes)
-  * Sparse grids
-  * Dense grids on a subspace
-* Add Laplace matrix algebra approximations [@wood2020simplified]
-* More comprehensive inference comparisons
+* Scaling up the hyperparameter grid beyond EB $k = 1$
+  * Any dense grid would be impractical ($k^{24}$ nodes)
+  * Alternatives: sparse grids, dense grids on a subspace
+* Add Laplace matrix algebra approximations [@wood2020simplified] to speed up latent field marginal calculations
+* More comprehensive inference comparison
   * Maximum mean discrepancy
@@
 
-**Interested?** Working notebooks and R code available from `github.com/athowes/elgm-inf`, or get in touch:
+**Interested?** Working notebooks and R code available from `github.com/athowes/elgm-inf`. Or get in touch:
 

### 2023-03-10 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Final changes because headless_chrome is different to headless_firefox and my life could have been so much easier making this in PowerPoint

@@
 
-<!-- 38761C -->
-
 ## The Naomi HIV model
@@
 
+$$\\[0.1in]$$
+
 ```{r fig1, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model fitting occurs interactively in stages.'}
@@
 ```
+
+$$\\[0.1in]$$
 
@@
 * We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
-* Though we focus on Naomi here, the HIV Inference Group (`hiv-inference.org`) works on many other complex models, challenging for existing Bayesian inference methods, which require high customisability
-
-$$\\[0.4in]$$
+* Though we focus on Naomi here, the HIV Inference Group (`hiv-inference.org`) works on many other complex models, challenging for existing Bayesian inference methods, which require flexible modelling tools
 
@@
 
+$$\\[0.1in]$$
+
 ## Application to Malawi data
@@
 * Relatively small country but still a large model: latent field $\text{dim}(x) = 491$, hyperparameters $\text{dim}(\theta) = 24$
-
-$$\\[0.2in]$$
-
 * Fit four inference methods (using one [!] C++ template):

### 2023-03-14 docs/bioinference.Rmd
Add BioInference abstract

---
title: "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
author: "Adam Howes, Imperial College London"
date: "14th March 2023"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

[BioInference 2023](https://bioinference.github.io/2023/)

```{r echo=FALSE}
title <- "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"

abstract <- "Naomi (Eaton et al, 2021) is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage, are jointly modelled using both household survey data and routinely reported health system data. Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the `TMB` R package. We propose a new inference method inspired by adaptive Gauss-Hermite quadrature together with the simplified integrated nested Laplace approximation approach of Wood (2020) to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides more accurate inferences than `TMB`, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but faster to run. By extending the `aghq` R package (Stringer, 2021) we facilitate easy, flexible use of our method when provided a `TMB` C++ template for the model's log-posterior. In doing so, we enable inference via integrated nested Laplace approximations for a larger class of models than was previously possible."
```

**Title**:
```{r echo=FALSE, results='asis'}
cat(title)
```

Title word count: `r stringr::str_count(title, '\\w+')`.

**Abstract**:
```{r echo=FALSE, results='asis'}
cat(abstract)
```

Abstract word count: `r stringr::str_count(abstract, '\\w+')`/200.

### 2023-03-19 src/docs_paper/paper.Rmd
Why MCMC is slow (could be better explained)

@@
 The practical requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
-Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
+Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production by country teams.
+Markov chain Monte Carlo (MCMC) approaches are prohibitively slow due to the scale of the model and challenging features of its posterior geometry [@neal2003slice].
 Inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
@@
 For these models, Metropolis-Hastings style algorithms be defined to update the fixed parameters, with acceptance probabilities calculated using by repeat calls to `R-INLA`.
-It would or would not be possible to fit the Naomi model in these frameworks.
+<!-- It would or would not be possible to fit the Naomi model in these frameworks. -->
+
+<!-- Note: can fit hyperparameters with Laplace on latent field using HMC via tmbstan::tmbstan(laplace = TRUE) -->
 

### 2023-03-20 src/dev_scale-grid/scale-grid.Rmd
Check that PCA works with smaller example

@@
 abstract: |
-  **Background** We have previously fit empirical Bayes style approaches to the Naomi model, fixing hyperparameter values.
+  **Background** We have previously fit empirical Bayes style approaches to the Naomi model, fixing hyperparameter values to a single value.
 
-  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure. The problem is that there are too many to use anything dense.
+  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure.
+The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical.
+In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using principle components analysis.
 ---
@@
 
-```{r results="hide"}
+```{r results="hide", message=FALSE}
 area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
@@
 
-Add on the modes:
-
-```{r}
-for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
-```
-
-Adjust the weights:
-
-```{r}
-gg_s$weights
-det(Lambda)
-det(Lambda_s)
-```
+Put this into a function, which also adjusts according to the mean and reweights the nodes:
 
@@
 pca_rescale_grid <- function(m, C, s, k) {
+  d <- nrow(C)
+  stopifnot(d == length(m))
+  eigenC <- eigen(C)
   lambda <- eigenC$values
@@
   a_nodes <- gg_s$nodes
-  b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))
+  b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}, ncol = s) %*% t(a_nodes))
   for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
@@
 
+Check that this works as expected for a simple example:
+
+```{r}
+mu <- c(1, 1.5)
+cov <- matrix(c(2, 1, 1, 1), ncol = 2)
+
+obj <- function(theta) {
+  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
+}
+
+grid <- expand.grid(
+  theta1 = seq(-2, 5, length.out = 700),
+  theta2 = seq(-2, 5, length.out = 700)
+)
+
+ground_truth <- cbind(grid, pdf = obj(grid))
+
+plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = multi.utils::cbpalette()[1]) +
+  coord_fixed(xlim = c(-2, 5), ylim = c(-2, 5), ratio = 1) +
+  theme_minimal()
+
+plot
+
+test <- pca_rescale_grid(m = mu, C = cov, s = 1, k = 5)
+
+plot +
+  geom_point(
+    data = data.frame(test$nodes),
+    aes(x = X1, y = X2),
+    col = multi.utils::cbpalette()[2],
+    inherit.aes = FALSE
+  )
+```
+
+Apply it to the Naomi grids:
+
 ```{r}

### 2023-03-20 src/docs_paper/paper.Rmd
Misc. work on paper writing

@@
 Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package.
-We propose a new inference method inspired by adaptive Gauss-Hermite quadrature together with the simplified integrated nested Laplace approximation approach of @wood2020simplified to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
+We propose a new inference method combining adaptive Gauss-Hermite quadrature together with the simplified integrated nested Laplace approximation approach of @wood2020simplified to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
 Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but faster to run.
@@
 
-Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at the level of geographic resolution at which heath systems are planned and delivered.
-Producing these estimates is challenging because all available data sources have shortcomings which must be overcome.
-Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they occur only every five years or so, with limited sample size at the district level.
+To mount an effective public health response to the HIV epidemic, it's crucial to have accurate, timely estimates of HIV indicators at the geographic level at which health systems are planned and delivered.
+However, producing these estimates is challenging due to the shortcomings of all available data sources.
+Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they are conducted only every five years or so, with limited sample size at the district level.
 Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in more real-time but based on limited or non-representative samples of the population.
-To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
+To address challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
 Modelling multiple data sources jointly has many benefits, including mitigating the limitations of any single source, increasing statistical power, and prompting investigation into any conflicts of information between sources.
 Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates in a yearly process supported by UNAIDS.
-The creation of estimates by country teams, rather than external agencies, is a noteworthy feature of the HIV response.
-Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust and ownership of the estimates.
+Creation of estimates by country teams, rather than external agencies, is a noteworthy feature of the HIV response.
+Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust and ownership of the resulting estimates.
 
-The practical requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
-Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production by country teams.
+Practical requirements for the model, in combination with its relative complexity, present a difficult Bayesian inference problem.
+Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production.
 Markov chain Monte Carlo (MCMC) approaches are prohibitively slow due to the scale of the model and challenging features of its posterior geometry [@neal2003slice].
@@
 Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgood2022statistical].
-Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments, which for the Naomi model, we use for the high-dimensional latent field parameters.
+Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
+In the Naomi model, this subset is the high-dimensional latent field.
 Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
-Although this approach has favourable computational properties, it would be preferable to account more fully for hyperparameter uncertainty than is possible in an empirical Bayes framework.
+Although this approach is fast, we would prefer to account more fully for hyperparameter uncertainty than as is the case in an empirical Bayes framework.
 This has motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
@@
 The key innovation of @rue2009approximate is an approximation which enables accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
-Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
-This extension facilitates inference for models like Naomi, which are not latent Gaussian models (LGMs) and so were not previously amenable to inference with INLA.
-Instead, due to dependence of observations on multiple structured additive predictors, Naomi is what has been termed by @stringer2022fast an extended latent Gaussian model (ELGM).
+Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on precision matrix of the latent field required for this approximation to be accurate.
+This extension facilitates inference for models like Naomi, which are not quite latent Gaussian models (LGMs) and so were not previously amenable to inference with INLA.
+In particular, due to dependence of observations on multiple structured additive predictors, Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast].
 We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
@@
 Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package of @bachl2019inlabru, and the INLA within MCMC approach of @gomez2018markov, both of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
-To approximate non-linear predictors, `inlabru` makes iterated use of `R-INLA` using linearisation, extending the scope of INLA to LGMs with some suitably small amount of non-linearity.
+`inlabru` approximates non-linear predictors using linearisation, which involves making iterative calls to `R-INLA`.
 <!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
 <!-- Want to say in particular what the class of models that `inlabru` allows you to fit are -->
-INLA within MCMC is suitable for models which are LGMs, though only conditional on some subset of the parameters being fixed.
-For these models, Metropolis-Hastings style algorithms be defined to update the fixed parameters, with acceptance probabilities calculated using by repeat calls to `R-INLA`.
+<!-- Don't know the answer to this yet, but asking Janine Illian I'm told you don't have to call R-INLA that many times -->
+<!-- I think it'd obviously depend on the exact details of the problem at hand -->
+INLA within MCMC is suitable for models which are LGMs conditional on some subset of the parameters being fixed.
+For such models, Metropolis-Hastings style algorithms can be defined to update the fixed parameters, with acceptance probabilities calculated using by repeat calls to `R-INLA`.
 <!-- It would or would not be possible to fit the Naomi model in these frameworks. -->
 
-<!-- Note: can fit hyperparameters with Laplace on latent field using HMC via tmbstan::tmbstan(laplace = TRUE) -->
-
 The remainder of this paper is organised as follows.
-Section \ref{sec:naomi} describes a simplified version of the Naomi model that we consider in this paper.
-Section \ref{sec:elgm} describes how the Naomi model falls within the ELGM framework.
+Section \ref{sec:naomi} outlines the version of the Naomi model that we consider in this paper, and  Section \ref{sec:elgm} describes how it falls within the ELGM framework.
 Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference for ELGMs using simplified INLA and AGHQ.
@@
 We consider a simplified version defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
-We provide an overview of the simplified model below, highlighting the aspects which make it a challenge for existing inferential approaches.
+An overview of the simplified model is given below, highlighting the aspects which make it a challenge for existing inferential approaches.
 A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
@@
 
-Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place at time $T_1$.
+Consider a country in sub-Saharan Africa where a household survey with complex design has taken place at time $T_1$.
 Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ five-year age band, and  $s \in \mathcal{S}$ sex.
@@
 
-For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
+We use a binomial working likelihood $y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$ to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
 \begin{equation*}
@@
 
-We model HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women as being offset on the logit scale from the general population indicator as follows
+HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset on the logit scale from the general population indicators as follows
 \begin{align*}
@@
 \end{align*}
-We inform these processes by specifying likelihoods for the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
+These processes are informed by specifying likelihoods for the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
 We use the binomial working likelihoods
@@
 
-Let $\dot A_\mathcal{I} = \sum_{x \sim x', x = x'} \dot A_{x', x}$ be the number of people receiving ART.
-Model $\dot A_\mathcal{I} \sim \mathcal{N}(\tilde A_\mathcal{I}, \sigma^{\tilde A_\mathcal{I}})$ using a normal distribution with mean $\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x}$ and standard deviation $\sigma^{\tilde A_\mathcal{I}}) = \sqrt{\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x} (1 - \pi_{x', x})}$.
+This section remains to write, and is quite challenging.
+
+<!-- Let $\dot A_\mathcal{I} = \sum_{x \sim x', x = x'} \dot A_{x', x}$ be the number of people receiving ART. -->
+<!-- Model $\dot A_\mathcal{I} \sim \mathcal{N}(\tilde A_\mathcal{I}, \sigma^{\tilde A_\mathcal{I}})$ using a normal distribution with mean $\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x}$ and standard deviation $\sigma^{\tilde A_\mathcal{I}}) = \sqrt{\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x} (1 - \pi_{x', x})}$.
+-->
 
@@
 
-Latent Gaussian models (LGMs) [@rue2009approximate] are of the form
+Latent Gaussian models (LGMs) [@rue2009approximate] are three-stage hierarchical models of the form
 \begin{align*}
@@
 The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
-Importantly, this allows for the presence of more non-linearity.
+Importantly, this mapping allows for the presence of more non-linearity in the model.
 ELGMs are then of the form
@@
 $$
-\mu_i = g_i(\eta^\theta_i, \eta^{\theta^{\text{ANC}}}_i) = \text{logit}^{-1}(\eta^\theta_i + \eta^{\theta^{\text{ANC}}}_i)
+\mu_i = g_i(\eta^\theta_i, \eta^{\theta^{\text{ANC}}}_i) = \text{logit}^{-1}(\eta^\theta_i + \eta^{\theta^{\text{ANC}}}_i).
 $$
-3. In the ART attendance component (Section \ref{sec:art}), the multinomial logistic regression...
+3. In the ART attendance component (Section \ref{sec:art}), the multinomial logistic regression.
+4.
+In all components the data we model is aggregated.
+Processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations over sets of indices $\mathcal{I}$.
+
+<!-- To-do: define what exactly $p(\y, \x, \btheta)$ is for the Naomi model -->
 
@@
 
-<!-- To-do: define what exactly $p(\y, \x, \btheta)$ is -->
-Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of Laplace approximation and adaptive Gauss-Hermite quadrature.
+Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of the Laplace approximation and AGHQ.
 
-<!-- 1. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(\btheta, \y)$ -->
-<!-- $$ -->
-<!-- \tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)} -->
-<!-- $$ -->
-<!-- where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by -->
-<!-- \begin{align*} -->
-<!-- \hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\ -->
-<!-- \mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}. -->
-<!-- \end{align*} -->
+```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension are adapted based on the mean and covariance matrix of the target via $\\btheta(\\z) = \\hat \\btheta + \\mathbf{L} \\z$.."}
+mu <- c(1, 1.5)
+cov <- matrix(c(2, 1, 1, 1), ncol = 2)
 
-<!-- 2. Normalise $\tilde p_\text{LA}(\btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain -->
-<!-- $$ -->
-<!-- \tilde p_\text{AQ}(\btheta \, | \, \y) = \frac{\tilde p_\text{LA}(\btheta, \y)}{\tilde p_{\text{AQ}}(\y)}, -->
-<!-- $$ -->
-<!-- where the normalising constant is calculated using nodes from a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(m, k)$ with $m = \dim(\btheta)$, $k$ nodes per dimension, and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ as -->
-<!-- $$ -->
-<!-- \tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z). -->
-<!-- $$ -->
-<!-- The nodes $\z$ are adapted based on the mode and curvature at the mode of the Laplace approximation as follows -->
-<!-- \begin{align*} -->
-<!-- \btheta(\z) &= \hat{\btheta} + \mathbf{L} \z, \\ -->
-<!-- \hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\ -->
-<!-- \mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\ -->
-<!-- \mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top. -->
-<!-- \end{align*} -->
-<!-- We typically set $k = 3$ such that there are $3^m$ nodes in total. -->
+obj <- function(theta) {
+  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
+}
 
-<!-- 3. Obtain an unnormalised nested approximation to the posterior marginal of the $i$th latent effect by -->
-<!-- $$ -->
-<!-- \tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z). -->
-<!-- $$ -->
-<!-- The nodes and weights $\{\mathcal{Q}(m, k), \omega\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above. -->
-<!-- For each of the $k^m$ values of $\btheta(\z)$ we obtain $\tilde p_\text{LA}(x_i, \btheta(\z), \y)$ by setting $\btheta = \btheta(\z)$ in the following Laplace approximation -->
-<!-- $$ -->
-<!-- \tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)} -->
-<!-- $$ -->
-<!-- where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x_{-i}(x_i, \btheta), \mathbf{H}_{-i, -i}(x_i, \btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by -->
-<!-- \begin{align*} -->
-<!-- \hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\ -->
-<!-- \mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}. -->
-<!-- \end{align*} -->
-<!-- Optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$. -->
+grid <- expand.grid(
+  theta1 = seq(-2, 5, length.out = 700),
+  theta2 = seq(-2, 5, length.out = 700)
+)
 
-<!-- 4. Normalise $\tilde p_\text{LA}(x_i, \y)$ using $\tilde p_{\text{AQ}}(\y)$ to obtain -->
-<!-- $$ -->
-<!-- \tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}. -->
-<!-- $$ -->
-<!-- which may be evaluated for some choice of values $x_i \in \{\ldots\}$. -->
+ground_truth <- cbind(grid, pdf = obj(grid))
 
-<!-- The above omiited "forward" algorithm section is written as you would do to calculate things in code. -->
-<!-- The "backward" algorithm section to follow is written as you would to have everything defined before it's used. -->
+plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = multi.utils::cbpalette()[1]) +
+  coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
+  labs(x = "", y = "") +
+  theme_minimal() +
+  guides(size = FALSE) +
+  theme(
+    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
+    axis.text.y = element_blank(), axis.ticks.y = element_blank()
+  )
+
+gg <- mvQuad::createNIGrid(2, "GHe", 3)
+
+add_points <- function(plot0, gg) {
+  plot0 +
+    geom_point(
+      data = mvQuad::getNodes(gg) %>%
+              as.data.frame() %>%
+              mutate(weights = mvQuad::getWeights(gg)), 
+      aes(x = V1, y = V2, size = weights),
+      alpha = 0.8,
+      col = multi.utils::cbpalette()[2],
+      inherit.aes = FALSE
+    ) +
+    scale_size_continuous(range = c(1, 2))
+}
+
+plot1 <- add_points(plot0, gg) +
+  labs(size = "", caption = "")
+
+#' Adapt by the mean
+gg2 <- gg
+mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2)
+
+plot2 <- add_points(plot0, gg2) +
+  labs(size = "", caption = "")
+
+#' Adapt by the lower Cholesky
+gg3 <- gg
+mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)
+
+plot3 <- add_points(plot0, gg3) +
+  labs(size = "", caption = "")
+
+plot1 + plot2 + plot3
+```
 
@@
 \begin{align*}
-\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
-\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
+\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
+\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
 \mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
@@
 $$
-\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
 $$
-where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
 \begin{align*}
@@
 2.
-Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
+Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
 If possible $k \geq 3$ is preferred, though the number of grid points scales exponentially with choice of $k$.
-Then use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
+Then, use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
 \begin{equation}
-\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z). \label{eq:evidence1}
+\tilde p_{\texttt{AQ}}(\y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\btheta(\bu), \y) \omega(\bu). \label{eq:evidence1}
 \end{equation}
 4.
-For $i \in [n]$ generate $l$ nodes $x_i(\z)$ via a Gauss-Hermite quadrature rule $\z \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
+For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
 A value of $l \geq 4$ is recommended to enable B-spline interpolation.
-Then, for $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
+Then, for $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
 \begin{align*}
@@
 where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
-6. For $x_i \in \{ x_i(\z) \}_{\z \in \mathcal{Q}(1, l)}$ calculate
+6. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
 \begin{equation}
-\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{\tilde p_{\text{AQ}}(\y)}. \label{eq:laplace-marginal}
+\tilde p_\texttt{AQ}(x_i \, | \, \y) = \frac{\tilde p_\texttt{LA}(x_i, \y)}{\tilde p_{\texttt{AQ}}(\y)}. \label{eq:laplace-marginal}
 \end{equation}
@@
 $$
-\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
+\tilde p_\texttt{LA}(x_i, \y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(x_i, \btheta(\bu), \y) \omega(\bu).
 $$
@@
 $$
-\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
+\tilde p_\texttt{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\texttt{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
 $$
@@
 $$
-\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(1, l)} \tilde p_\text{LA}(x_i(\z), \y) \omega(\z)
+\tilde p_{\texttt{AQ}}(\y) = \sum_{\bv \in \mathcal{Q}(1, l)} \tilde p_\texttt{LA}(x_i(\bv), \y) \omega(\bv)
 $$
@@
 7.
-Given $\{x_i, \tilde p_\text{AQ}(x_i \, | \, \y)\}_{x_i \in \mathcal{X}_i}$ create a spline interpolant to each posterior marginal on the log-scale, from which samples or relevant posterior marginal summaries may be obtained.
-
-Note that clearer notation is required for $\z$ and $\omega(\z)$.
+Given $\{x_i(\bv), \tilde p_\texttt{AQ}(x_i(\bv) \, | \, \y)\}_{\bv \in \mathcal{Q}(1, l)}$ create a spline interpolant to each posterior marginal on the log-scale.
+Samples, and thereby relevant posterior marginal summaries, may be obtained using inverse transform sampling.
 
@@
 Using one `TMB` C++ user-template (available in the appendix), we fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using four inferential approaches.
-There were: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
-Particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
+These were:
+1. `TMB`: EB combined with a Gaussian approximation via `TMB`,
+2. `aghq`: AGHQ combined with a Gaussian approximation via `aghq`,
+3. `adam`: AGHQ combined with a Laplace approximation by extending `aghq`, and
+4. `NUTS`: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
+In this instance, the dimension of the latent field $N = 491$ and the dimension of the hyperparameters $m = 24$.
+Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
 For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
-Then, for all methods, we simulated model outputs.
-
+For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence.
 The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
@@
 \small
-\begin{tabularx}{\textwidth}{p{0.2\linewidth}p{0.75\linewidth}}
+\begin{tabularx}{\textwidth}{p{0.3\linewidth}p{0.65\linewidth}}
 \toprule
@@
 \midrule
-1. EB, Gaussian & $1000$ samples \\
-2. AGHQ, Gaussian & $k = 1$, $1000$ samples \\
-3. AGHQ, Laplace & $k = 1$, $l = 5$, $1000$ samples \\
-4. NUTS & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$. HMC parameters set to default for \texttt{rstan}. \\
+1. \texttt{TMB}: EB, Gaussian & $1000$ samples \\
+2. \texttt{aghq}: AGHQ, Gaussian & $k = 1$, $1000$ samples \\
+3. \texttt{adam}: AGHQ, Laplace & $k = 1$, $l = 5$, $1000$ samples \\
+4.
+\texttt{NUTS}: NUTS & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$.
+HMC parameters set to default for \texttt{rstan}.
+\\
 \bottomrule
@@
 
-MCMC can be used to obtain accurate inferential results only once convergence has been reached and the Markov chain length is sufficiently long.
+MCMC can be used to obtain accurate inferential results only once convergence has been reached and the chain lengths are sufficient.
 We assessed the quality of our MCMC results using the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
 Full details are provided in the appendix.
-We treat these results from NUTS as a gold-standard to which other inferential methods can be compared to.
+We treat these results from NUTS as a gold-standard to which other inferential methods are compared to.
 
@@
 We used three methods to assess the accuracy of posterior distributions produced by each inferential method as compared with those from NUTS: (1) Kolmogorov-Smirnov tests, (2) maximum mean discrepancy, and (3) Pareto-smoothed importance sampling.
-From an applied perspective, our main interest is in comparing the accuracy of model outputs, rather than the internal hyperparameters or latent field parameters.
-That said, obtaining accurate inferences for these internal parameters is also of interest.
+Our primary applied interest is in comparing the accuracy of model outputs, rather than the internal hyperparameters or latent field parameters.
+That said, obtaining accurate inferences for these internal parameters is also relevant.
 
@@
 
-Let $\{\theta_i\}_{i = 1}^n$ be posterior marginal samples with empirical cumulative distribution (ECDF) function $F(\vartheta) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\theta_i \leq \vartheta}$.
-The two-sample Kolmogorov-Smirnov (KS) test statistic is given by the maximum absolute difference between two ECDFs.
-For each method $\{\texttt{TMB}, \texttt{aghq}, \texttt{adam}\}$ we compare the KS statistics
+Let $\{\theta_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function $F(\vartheta) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\theta_i \leq \vartheta}$.
+The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is given by the maximum absolute difference between two ECDFs.
+For each method we compare the KS statistics
 \begin{equation*}
-D_\texttt{method} = \sup_\vartheta | F_\texttt{tmbstan}(\vartheta) - F_\texttt{method}(\vartheta)|.
+D_\bullet = \sup_\vartheta | F_\texttt{NUTS}(\vartheta) - F_\bullet(\vartheta)|.
 \end{equation*}
-See a summary of the results in Table 1 and Plot 1, and full results available in the appendix.
+See a summary of the results in Table and Figure \ref{fig:ks}, and full results available in the appendix.
 
-```{r}
-knitr::include_graphics("depends/ks_summary_tmb_adam.png")
+```{r ks, fig.cap="Results of Kolmogorov-Smironv tests analysis."}
+qplot(1:10, 1:10) + theme_minimal()
 ```
@@
 
-gt::gt(ks_summary)
+# gt::gt(ks_summary) %>% gt::as_latex()
 ```
@@
 To write.
-See a summary of the results in Table 2 and Plot 2, and full results available in the appendix.
+See a summary of the results in Table and Figure \ref{fig:mmd}, and full results available in the appendix.
+
+```{r mmd, fig.cap="Results of maximum mean discrepancy analysis."}
+qplot(1:10, 1:10) + theme_minimal()
+```
 
@@
 To write.
-See a summary of the results in Table 3 and Plot 3, and full results available in the appendix.
+See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.
+
+```{r psis, fig.cap="Results of Pareto-smoothed importance sampling analysis."}
+qplot(1:10, 1:10) + theme_minimal()
+```
 
@@
 We developed an approximate Bayesian inference algorithm to solve a challenging problem in small-area estimation of HIV for low resource settings.
-Our method is demonstrated to be more accurate than the EB Gaussian approximation and substantially faster than NUTS for the simplified Naomi model in Malawi (Section \ref{sec:results}).
-We anticipate that our method could be added to the Naomi software as an alternative option to the, still substantially faster, EB Gaussian approximation.
+Our method is demonstrated to be more accurate than the EB Gaussian approximation currently in use, and substantially faster than NUTS for the simplified Naomi model in Malawi (Section \ref{sec:results}).
+We anticipate that our method could be added to the Naomi software as an alternative option to `TMB`.
 Analysts might quickly iterate over model options using the faster, less accurate inference approach, only switching to the slower, more accurate approach once they are happy with the results.
@@
 We provide a flexible implementation, which builds on the `aghq` \textsc{R} package.
-In doing so, we hope our work enables use of INLA for ELGMs in applied settings, as well as further methodological exploration of the algorithms accuracy and limits.
+In doing so, we hope our work enables use of deterministic inference algorithms for ELGMs in applied settings, as well as further methodological exploration of their accuracy and limitations.
 Among the ELGMs structures of greatest interest are: aggregated Gaussian process models [@nandi2020disaggregation].
 Although our method is designed for ELGMs, it may even be used outside this class, and is compatible with any model with a `TMB` \textsc{C++} template.
-
-We would be excited to see statistical theory for our algorithm by extension of Theorem 1 of @stringer2022fast.
 
@@
 
+Future work could look to implement our algorithm (Section \ref{sec:algorithm}) into probabilistic programming languages, facilitating access by a broader user-base.
+This might be possible in Stan by use of the `bridgestan` package [@bridgestan] together with the adjoint-differentiated Laplace approximation of @margossian2020hamiltonian.
+<!-- Note: can fit hyperparameters with Laplace on latent field using HMC via tmbstan::tmbstan(laplace = TRUE) -->
+<!-- i.e. what Charles Margossian has implemented for Stan is done similarly by default in TMB -->
+As well, statistical theory for the algorithm could be established by extension of Theorem 1 in @stringer2022fast.
+
 # Acknowledgements {-}
 
-AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1), and conducted part of this research while an International Visiting Graduate Student at the University of Waterloo (GRANT).
+AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1), and conducted part of this research while an International Visiting Graduate Student at the University of Waterloo.
 AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).
-AS was supported by...
 SRF was supported by the EPSRC (EP/V002910/2).

### 2023-03-22 src/dev_scale-grid/scale-grid.Rmd
PCA grid as mvQuad object, unit testing of PCA-AGHQ

@@
 
-```{r}
-create_grid <- function(d, k) {
-  gg <- mvQuad::createNIGrid(dim = d, type = "GHe", level = k)
-  list(nodes = mvQuad::getNodes(gg), weights = as.numeric(mvQuad::getWeights(gg)))
-}
-```
-
 Start with `s = 5` dimensions kept of the PCA.
@@
 d <- dim(E)[1]
-gg_s <- create_grid(d = s, k = 2)
+gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = 2) 
 ```
@@
 ```{r}
-(n_nodes <- nrow(gg_s$nodes))
+(n_nodes <- nrow(mvQuad::getNodes(gg_s)))
 ```
@@
 ```{r}
-a_nodes <- gg_s$nodes
+a_nodes <- mvQuad::getNodes(gg_s)
 b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))
@@
 #' @param k Number of points per small grid dimension
-pca_rescale_grid <- function(m, C, s, k) {
+pca_rescale <- function(m, C, s, k) {
   d <- nrow(C)
@@
   E_s <- E[, 1:s] 
-  gg_s <- create_grid(d = s, k = k)
-  a_nodes <- gg_s$nodes
-  b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}, ncol = s) %*% t(a_nodes))
-  for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
-  b_weights <- gg_s$weights * sqrt(det(Lambda))
-  return(list(nodes = b_nodes, weights = b_weights))
+  gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = k) 
+  nodes_out <- t(E_s %*% diag(lambda[1:s]^{-0.5}, ncol = s) %*% t(mvQuad::getNodes(gg_s)))
+  for(j in 1:d) nodes_out[, j] <- nodes_out[, j] + m[j]
+  weights_out <- mvQuad::getWeights(gg_s) * as.numeric(mvQuad::getWeights(mvQuad::createNIGrid(dim = d - s, type = "GHe", level = 1)))
+
+  # Putting things into a mvQuad format manually
+  gg <- mvQuad::createNIGrid(dim = d, type = "GHe", level = 1)
+  gg$level <- rep(NA, times = d)
+  gg$ndConstruction <- "PCA"
+  gg$nodes <- nodes_out
+  gg$weights <- weights_out
+  return(gg)
 }
@@
 
-Check that this works as expected for a simple example:
+### Testing with simple examples
+
+Check that this works as visually expected for a simple example:
 
@@
   geom_contour(col = multi.utils::cbpalette()[1]) +
-  coord_fixed(xlim = c(-2, 5), ylim = c(-2, 5), ratio = 1) +
+  coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
+  labs(x = "", y = "") +
   theme_minimal()
 
-plot
-
-test <- pca_rescale_grid(m = mu, C = cov, s = 1, k = 5)
+gaussian_2d <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
 
@@
   geom_point(
-    data = data.frame(test$nodes),
-    aes(x = X1, y = X2),
+    data = data.frame(mvQuad::getNodes(gaussian_2d)) %>%
+      mutate(weight = mvQuad::getWeights(gaussian_2d)),
+    aes(x = X1, y = X2, size = weight),
     col = multi.utils::cbpalette()[2],
     inherit.aes = FALSE
-  )
+  ) +
+  scale_size_continuous(limits = c(1, 2))
 ```
 
-Apply it to the Naomi grids:
+Test how this PCA grid works for some simple integrals.
+We will use Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.
 
 ```{r}
-gg5 <- pca_rescale_grid(m = m, C = C, s = 5, k = 2)
-gg7 <- pca_rescale_grid(m = m, C = C, s = 7, k = 2)
-gg9 <- pca_rescale_grid(m = m, C = C, s = 9, k = 2)
+logfteta2d <- function(eta, y) {
+  n <- length(y)
+  n1 <- ceiling(n / 2)
+  n2 <- floor(n / 2)
+  y1 <- y[1:n1]
+  y2 <- y[(n1 + 1):(n1 + n2)]
+  eta1 <- eta[1]
+  eta2 <- eta[2]
+  sum(y1) * eta1 - (length(y1) + 1) * exp(eta1) - sum(lgamma(y1+1)) + eta1 +
+    sum(y2) * eta2 - (length(y2) + 1) * exp(eta2) - sum(lgamma(y2+1)) + eta2
+}
+
+set.seed(84343124)
+n1 <- 5
+n2 <- 5
+n <- n1 + n2
+y1 <- rpois(n1, 5)
+y2 <- rpois(n2, 5)
+
+truemode2d <- c(log((sum(y1) + 1)/(length(y1) + 1)), log((sum(y2) + 1)/(length(y2) + 1)))
+
+truelogint <- function(y) lgamma(1 + sum(y)) - (1 + sum(y)) * log(length(y) + 1) - sum(lgamma(y + 1))
+truelognormconst2d <- truelogint(y1) + truelogint(y2)
+
+objfunc2d <- function(x) logfteta2d(x, c(y1, y2))
+
+funlist2d <- list(
+  fn = objfunc2d,
+  gr = function(x) numDeriv::grad(objfunc2d,x),
+  he = function(x) numDeriv::hessian(objfunc2d,x)
+)
+
+opt_bfgs_2d <- aghq::optimize_theta(funlist2d, c(1.5, 1.5), control = default_control(method = "BFGS"))
+
+norm_bfgs_2d_1 <- aghq::normalize_logpost(opt_bfgs_2d, 1, 1)
+norm_bfgs_2d_3 <- aghq::normalize_logpost(opt_bfgs_2d, 3, 1)
+norm_bfgs_2d_5 <- aghq::normalize_logpost(opt_bfgs_2d, 5, 1)
+norm_bfgs_2d_7 <- aghq::normalize_logpost(opt_bfgs_2d, 7, 1)
+
+pca_grid_bfgs_2d_1 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 1)
+pca_grid_bfgs_2d_3 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 3)
+pca_grid_bfgs_2d_5 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 5)
+pca_grid_bfgs_2d_7 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 7)
+
+plot_points <- function(gg) {
+  gg$nodes %>%
+    as.data.frame() %>%
+    mutate(weights = gg$weights) %>%
+    ggplot() +
+    geom_point(
+      aes(x = V1, y = V2, size = weights),
+      alpha = 0.8,
+      col = multi.utils::cbpalette()[2],
+      inherit.aes = FALSE
+    ) +
+    scale_size_continuous(range = c(1, 2)) +
+    labs(x = "", y = "", size = "Weight") +
+    theme_minimal()
+}
+
+{plot_points(pca_grid_bfgs_2d_1) + plot_points(pca_grid_bfgs_2d_3)} / {plot_points(pca_grid_bfgs_2d_5) + plot_points(pca_grid_bfgs_2d_7)}
+
+norm_pca_bfgs_2d_1 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_1, adapt = FALSE)
+norm_pca_bfgs_2d_3 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_3, adapt = FALSE)
+norm_pca_bfgs_2d_5 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_5, adapt = FALSE)
+norm_pca_bfgs_2d_7 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_7, adapt = FALSE)
+
+results_2d <- data.frame(
+  "method" = c(
+    "Truth",
+    paste0("aghq, k = ", c(1, 3, 5, 7)),
+    paste0("aghq-pca, k = ", c(1, 3, 5, 7))
+  ),
+  "lognormconst" = c(
+    truelognormconst2d,
+    norm_bfgs_2d_1$lognormconst, norm_bfgs_2d_3$lognormconst, norm_bfgs_2d_5$lognormconst, norm_bfgs_2d_7$lognormconst,
+    norm_pca_bfgs_2d_1$lognormconst, norm_pca_bfgs_2d_3$lognormconst, norm_pca_bfgs_2d_5$lognormconst, norm_pca_bfgs_2d_7$lognormconst
+  )
+)
+
+ggplot(results_2d, aes(x = method, y = lognormconst)) +
+  geom_point(size = 3) +
+  labs(x = "Method", y = "Log normalising constant") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 70, vjust = 0.5, hjust = 0.5))
 ```
+
+### Application to Naomi
+
+```{r}
+gg3 <- pca_rescale(m = m, C = C, s = 3, k = 2)
+nrow(mvQuad::getNodes(gg3))
+
+gg5 <- pca_rescale(m = m, C = C, s = 5, k = 2)
+nrow(mvQuad::getNodes(gg5))
+
+gg7 <- pca_rescale(m = m, C = C, s = 7, k = 2)
+nrow(mvQuad::getNodes(gg7))
+
+gg9 <- pca_rescale(m = m, C = C, s = 9, k = 2)
+nrow(mvQuad::getNodes(gg9))
+```
+
+```{r}
+start <- Sys.time()
+quad <- fit_aghq(tmb_inputs, basegrid = base_grid)
+end <- Sys.time()
+
+end - start
+```

### 2023-03-22 src/dev_scale-grid/scale-grid.Rmd
Should be range not limits (creates missing points in plot)

@@
   ) +
-  scale_size_continuous(limits = c(1, 2))
+  scale_size_continuous(range = c(1, 2))
 ```

### 2023-03-22 src/docs_paper/paper.Rmd
Update Bilodeau cite, explain why we don't use T2 and T3, start adding LA and AGHQ short explainers, other small changes

@@
 We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
-Though no theory yet exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2021stochastic using AGHQ.
+Though no theory yet exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2022stochastic using AGHQ.
 We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
@@
 
-@eaton2021naomi specify a joint model linking three small-area estimation models.
-The model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
-An overview of the simplified model is given below, highlighting the aspects which make it a challenge for existing inferential approaches.
-A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
+@eaton2021naomi specify a joint model linking three small-area estimation models and defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
+We consider a simplified version defined only at $T_1$ omitting nowcasting and temporal projection, as these time points involve only projection and no further inference.
+An overview of this simplified model is given below, highlighting the aspects which make it a challenge for existing inferential approaches.
+A more complete mathematical description, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
 
@@
 Consider a country in sub-Saharan Africa where a household survey with complex design has taken place at time $T_1$.
-Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ five-year age band, and  $s \in \mathcal{S}$ sex.
+Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ index five-year age band, and $s \in \mathcal{S}$ index sex.
 For ease of notation, let $i$ index the finest district-age-sex division included in the model.
@@
 is the Kish effective sample size [@kish1965survey].
-
-We use a binomial working likelihood $y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$ to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
+We use a binomial working likelihood
+$$
+y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})
+$$
+to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following weighted aggregates
 \begin{equation*}
@@
 
-# Fast approximate inference methods\label{sec:fastinferencemethods}
+# Fast approximate inference method\label{sec:fastinferencemethods}
 
@@
 \end{align}
+Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the above posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via nested applications of the Laplace approximation and AGHQ, described below.
 
-## Algorithm\label{sec:algorithm}
+## Laplace approximation
 
-Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via the following algorithm, comprised of nested applications of the Laplace approximation and AGHQ.
+## AGHQ
+
+A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by a weighted sum
+$$
+\int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z)
+$$
+where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
+In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials.
 
@@
 
+## Algorithm\label{sec:algorithm}
+
 1. Calculate the mode, Hessian at the mode, and lower Cholesky
@@
 2.
-Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
+Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
 If possible $k \geq 3$ is preferred, though the number of grid points scales exponentially with choice of $k$.
@@
 For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
-For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence.
+For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
 The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
@@
 
-MCMC can be used to obtain accurate inferential results only once convergence has been reached and the chain lengths are sufficient.
+Inferential results from MCMC are only accurate once convergence has been reached and the chain lengths are sufficient.
 We assessed the quality of our MCMC results using the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
@@
 
-We developed an approximate Bayesian inference algorithm to solve a challenging problem in small-area estimation of HIV for low resource settings.
-Our method is demonstrated to be more accurate than the EB Gaussian approximation currently in use, and substantially faster than NUTS for the simplified Naomi model in Malawi (Section \ref{sec:results}).
-We anticipate that our method could be added to the Naomi software as an alternative option to `TMB`.
-Analysts might quickly iterate over model options using the faster, less accurate inference approach, only switching to the slower, more accurate approach once they are happy with the results.
+We developed an approximate Bayesian inference algorithm motivated by a challenging problem in small-area estimation of HIV in low resource settings.
+For the simplified Naomi model in Malawi (Section \ref{sec:results}) our method is demonstrated to be more accurate than the EB Gaussian approximation currently in use, and substantially faster than NUTS.
+We anticipate that our method could be added to the Naomi web interface as an alternative to `TMB`.
+Analysts might quickly iterate over model options using the faster, less accurate inference approach, switching to the slower, more accurate approach once they are happy with the results.
 
-We provide a flexible implementation, which builds on the `aghq` \textsc{R} package.
+We provide a flexible implementation of the algoritm, building on the`TMB` and `aghq` \textsc{R} packages.
 In doing so, we hope our work enables use of deterministic inference algorithms for ELGMs in applied settings, as well as further methodological exploration of their accuracy and limitations.
-Among the ELGMs structures of greatest interest are: aggregated Gaussian process models [@nandi2020disaggregation].
-Although our method is designed for ELGMs, it may even be used outside this class, and is compatible with any model with a `TMB` \textsc{C++} template.
+Among the ELGMs structures of particular interest in spatial epidemiology, many of which are included in the structure of Naomi, are: aggregated Gaussian process models [@nandi2020disaggregation], evidence synthesis models [@amoah2020geostatistical].
+Although our method is designed for ELGMs, it is possible to use it outside this class, as it is compatible with any model with a `TMB` \textsc{C++} template.
 
@@
 
-Future work could look to implement our algorithm (Section \ref{sec:algorithm}) into probabilistic programming languages, facilitating access by a broader user-base.
+Future work could look to implement our algorithm (Section \ref{sec:algorithm}) within probabilistic programming languages, facilitating access by a broader user-base.
 This might be possible in Stan by use of the `bridgestan` package [@bridgestan] together with the adjoint-differentiated Laplace approximation of @margossian2020hamiltonian.

### 2023-03-22 src/naomi-simple_compare/comparison.Rmd
Finish poster
Add function description, and find the correct rows of mf_out for finest scale #24

@@
 ```{r}
-data.frame(
+time_taken <- data.frame(
   "TMB" = tmb$time,
@@
 )
+
+write_csv(time_taken, "time_taken.csv")
+
+time_taken
 
@@
 
-mf_out %>%
+mf_out_fine <- mf_out %>%
+  tibble::rownames_to_column("id") %>%
+  mutate(id = as.numeric(id)) %>%
   filter(
@@
       )
-  ) %>%
-  nrow()
+  )
+```
+
+An `id` column has been added to `mf_out_fine` to enable merging to the samples.
+
+```{r}
+head(mf_out_fine)
+nrow(mf_out_fine)
+```
+
+```{r}
+rho_t1_out_fine <- tmb$fit$sample$rho_t1_out[mf_out_fine$id, ]
+dim(rho_t1_out_fine)
+rho_t1_out_fine[1:10, 1:10]
+```
+
+```{r}
+histogram_and_ecdf(par = "rho_t1_out", i = 1)
 ```

### 2023-03-22 src/docs_paper/paper.Rmd
Add Naomi outputs figure to paper

@@
 For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
+<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
+Example model outputs from `TMB` are illustrated in Figure \ref{fig:naomi-results}.
 The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
 We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
+
+```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted with \\texttt{TMB}."}
+knitr::include_graphics("depends/naomi_results.png")
+```
 

### 2023-03-22 src/docs_bayescomp-poster/figs.R

@@ -136,7 +136,7 @@ ggsave(
   units = "in"
 )
 
-#' Figure 3
+#' Figure 4
 fig4data <- readRDS("depends/beta_anc_rho.rds") %>%
   filter(method != "aghq")
 

### 2023-03-22 src/dev_scale-grid/scale-grid.Rmd
More testing of PCA approach

@@
 
+#### Gaussian
+
 Check that this works as visually expected for a simple example:
@@
 }
+
+obj_list <- list(
+  fn = obj,
+  gr = function(theta) numDeriv::grad(obj, theta),
+  he = function(theta) numDeriv::hessian(obj, theta)
+)
 
@@
 
-gaussian_2d <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
+pca_grid <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
 
@@
   scale_size_continuous(range = c(1, 2))
+
+opt_bfgs <- aghq::optimize_theta(obj_list, mu, control = default_control(method = "BFGS"))
+norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, 1, 1)
+exp(norm_bfgs_1$lognormconst)
 ```
 
-Test how this PCA grid works for some simple integrals.
-We will use Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.
+#### Other 2D function
+
+More complicated two dimensional function from Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.
 
@@
 
+grid <- expand.grid(
+  theta1 = seq(0, 4, length.out = 400),
+  theta2 = seq(0, 4, length.out = 400)
+)
+
+ground_truth <- data.frame(grid) %>%
+  rowwise() %>%
+  mutate(pdf = exp(objfunc2d(x = c(theta1, theta2))))
+
+plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = multi.utils::cbpalette()[1]) +
+  coord_fixed(xlim = c(0.8, 2), ylim = c(0.8, 2), ratio = 1) +
+  labs(x = "", y = "") +
+  theme_minimal()
+
 opt_bfgs_2d <- aghq::optimize_theta(funlist2d, c(1.5, 1.5), control = default_control(method = "BFGS"))
+
+grid_bfgs_2d_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
+# Error in rescale.NIGrid(grid_bfgs_2d_3, m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian)) :
+# (in rescale) no appropriate covariance matrix C
+# mvQuad::rescale(grid_bfgs_2d_1, m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian))
+
+grid_bfgs_2d_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 3)
+# Error in rescale.NIGrid(grid_bfgs_2d_3, m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian)) :
+# (in rescale) no appropriate covariance matrix C
+# mvQuad::rescale(grid_bfgs_2d_3, m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian))
+
+grid_bfgs_2d_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 5)
+# Error in rescale.NIGrid(grid_bfgs_2d_3, m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian)) :
+# (in rescale) no appropriate covariance matrix C
+# mvQuad::rescale(grid_bfgs_2d_5, m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian))
+
+grid_bfgs_2d_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 7)
+# Error in rescale.NIGrid(grid_bfgs_2d_3, m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian)) :
+# (in rescale) no appropriate covariance matrix C
+# mvQuad::rescale(grid_bfgs_2d_7, m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian))
 
@@
 plot_points <- function(gg) {
-  gg$nodes %>%
+  plot +
+    geom_point(
+      data = mvQuad::getNodes(gg) %>%
         as.data.frame() %>%
-    mutate(weights = gg$weights) %>%
-    ggplot() +
-    geom_point(
+        mutate(weights = mvQuad::getWeights(gg)),
       aes(x = V1, y = V2, size = weights),
@@
 
+#' AGHQ
+{plot_points(grid_bfgs_2d_1) + plot_points(grid_bfgs_2d_3)} / {plot_points(grid_bfgs_2d_5) + plot_points(grid_bfgs_2d_7)}
+
+#' PCA-AGHQ
 {plot_points(pca_grid_bfgs_2d_1) + plot_points(pca_grid_bfgs_2d_3)} / {plot_points(pca_grid_bfgs_2d_5) + plot_points(pca_grid_bfgs_2d_7)}

### 2023-03-23 src/naomi-simple_compare/comparison.Rmd
Add stopifnot to check we have all samples from all methods. Stan failing even with latest archived run. Go back and make sure Stan is getting model outputs #24

@@
 depends <- yaml::read_yaml("orderly.yml")$depends
+```
+
+Check that the parameters (latent field, hyperparameters, model outputs) sampled from each of the four methods are the same:
+
+```{r}
+stopifnot(names(tmb$fit$sample) == names(aghq$quad$sample))
+stopifnot(names(tmb$fit$sample) == names(adam$adam$sample))
+stopifnot(names(tmb$fit$sample) == names(tmbstan$mcmc$sample))
 ```

### 2023-03-23 src/docs_paper/paper.Rmd
Clarification about removing T2 and T3

@@
 @eaton2021naomi specify a joint model linking three small-area estimation models and defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version defined only at $T_1$ omitting nowcasting and temporal projection, as these time points involve only projection and no further inference.
+We consider a simplified version defined only at $T_1$ omitting nowcasting and temporal projection, as these time points involve no further inference.
 An overview of this simplified model is given below, highlighting the aspects which make it a challenge for existing inferential approaches.

### 2023-03-23 src/naomi-simple_fit/script.R

@@ -1,5 +1,5 @@
 #' Uncomment and run the two line below to resume development of this script
-# orderly::orderly_develop_start("naomi-simple_fit", parameters = list(tmbstan = TRUE, area_level = 4, hmc_laplace = TRUE))
+# orderly::orderly_develop_start("naomi-simple_fit", parameters = list(tmbstan = TRUE, area_level = 4, hmc_laplace = FALSE))
 # setwd("src/naomi-simple_fit")
 
 if(tmb + aghq + adam + tmbstan != 1) {
@@ -67,7 +67,8 @@ naomi_data <- select_naomi_data(
 #' Naomi with packageVersion("naomi"), and if required, install version 2.8.5.
 #' with devtools::install_github("mrc-ide/naomi", ref = "e9de40f"). We modify
 #' this model to obtain a simplified version where the time points T2 and T3,
-#' as well as extraneous outputs, are removed.
+#' as well as extraneous outputs, are removed. No new inference is conducted in
+#' the removed model components, the difficulty of the problem remains similar.
 
 compile("naomi_simple.cpp")
 dyn.load(dynlib("naomi_simple"))

### 2023-03-23 src/naomi-simple_compare/comparison.Rmd
histogram_and_ecdf() generalised to work for model outputs as well #24

@@
 ```{r}
-pars <- unique(names(tmb$fit$obj$env$par))
+pars <- names(tmb$fit$sample)
 pars
@@
 
-## `beta_alpha`
+<!-- ## `beta_alpha` -->
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_alpha")
-```
+<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
+<!-- histogram_and_ecdf_helper("beta_alpha") -->
+<!-- ``` -->
 
-## `beta_lambda`
+<!-- ## `beta_lambda` -->
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_lambda")
-```
+<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
+<!-- histogram_and_ecdf_helper("beta_lambda") -->
+<!-- ``` -->
 
-## `beta_anc_rho`
+<!-- ## `beta_anc_rho` -->
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf("beta_anc_rho")
-```
+<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
+<!-- histogram_and_ecdf("beta_anc_rho") -->
+<!-- ``` -->
 
-## `beta_anc_alpha`
+<!-- ## `beta_anc_alpha` -->
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf("beta_anc_alpha")
-```
+<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
+<!-- histogram_and_ecdf("beta_anc_alpha") -->
+<!-- ``` -->
 
-## `logit`
+<!-- ## `logit` -->
 
-```{r eval = FALSE}
-lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf)
-```
+<!-- ```{r eval = FALSE} -->
+<!-- lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf) -->
+<!-- ``` -->
 
-## `log_sigma`
+<!-- ## `log_sigma` -->
 
-```{r eval = FALSE}
-lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
-```
+<!-- ```{r eval = FALSE} -->
+<!-- lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf) -->
+<!-- ``` -->
 
-## `u_rho_x`
+<!-- ## `u_rho_x` -->
 
-```{r eval = pars_eval[["u_rho_x"]]}
-histogram_and_ecdf_helper("u_rho_x")
-```
+<!-- ```{r eval = pars_eval[["u_rho_x"]]} -->
+<!-- histogram_and_ecdf_helper("u_rho_x") -->
+<!-- ``` -->
 
-## `u_rho_x`
+<!-- ## `u_rho_x` -->
 
-```{r eval = pars_eval[["u_rho_x"]]}
-histogram_and_ecdf_helper("u_rho_x")
-```
+<!-- ```{r eval = pars_eval[["u_rho_x"]]} -->
+<!-- histogram_and_ecdf_helper("u_rho_x") -->
+<!-- ``` -->
 
-## `us_rho_x`
+<!-- ## `us_rho_x` -->
 
-```{r eval = pars_eval[["us_rho_x"]]}
-histogram_and_ecdf_helper("us_rho_x")
-```
+<!-- ```{r eval = pars_eval[["us_rho_x"]]} -->
+<!-- histogram_and_ecdf_helper("us_rho_x") -->
+<!-- ``` -->
 
-## `u_rho_xs`
+<!-- ## `u_rho_xs` -->
 
-```{r eval = pars_eval[["u_rho_xs"]]}
-histogram_and_ecdf_helper("u_rho_xs")
-```
+<!-- ```{r eval = pars_eval[["u_rho_xs"]]} -->
+<!-- histogram_and_ecdf_helper("u_rho_xs") -->
+<!-- ``` -->
 
-## `us_rho_xs`
+<!-- ## `us_rho_xs` -->
 
-```{r eval = pars_eval[["us_rho_xs"]]}
-histogram_and_ecdf_helper("us_rho_xs")
-```
+<!-- ```{r eval = pars_eval[["us_rho_xs"]]} -->
+<!-- histogram_and_ecdf_helper("us_rho_xs") -->
+<!-- ``` -->
 
-## `u_rho_as`
+<!-- ## `u_rho_as` -->
 
-```{r eval = pars_eval[["u_rho_as"]]}
-histogram_and_ecdf_helper("u_rho_as")
-```
+<!-- ```{r eval = pars_eval[["u_rho_as"]]} -->
+<!-- histogram_and_ecdf_helper("u_rho_as") -->
+<!-- ``` -->
 
-## `u_alpha_x`
+<!-- ## `u_alpha_x` -->
 
-```{r eval = pars_eval[["u_alpha_x"]]}
-histogram_and_ecdf_helper("u_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_x"]]} -->
+<!-- histogram_and_ecdf_helper("u_alpha_x") -->
+<!-- ``` -->
 
-## `us_alpha_x`
+<!-- ## `us_alpha_x` -->
 
-```{r eval = pars_eval[["us_alpha_x"]]}
-histogram_and_ecdf_helper("us_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["us_alpha_x"]]} -->
+<!-- histogram_and_ecdf_helper("us_alpha_x") -->
+<!-- ``` -->
 
-## `u_alpha_xs`
+<!-- ## `u_alpha_xs` -->
 
-```{r eval = pars_eval[["u_alpha_xs"]]}
-histogram_and_ecdf_helper("u_alpha_xs")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_xs"]]} -->
+<!-- histogram_and_ecdf_helper("u_alpha_xs") -->
+<!-- ``` -->
 
-## `us_alpha_xs`
+<!-- ## `us_alpha_xs` -->
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("us_alpha_xs")
-```
+<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
+<!-- histogram_and_ecdf_helper("us_alpha_xs") -->
+<!-- ``` -->
 
-## `u_alpha_a`
+<!-- ## `u_alpha_a` -->
 
-```{r eval = pars_eval[["u_alpha_a"]]}
-histogram_and_ecdf_helper("u_alpha_a")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_a"]]} -->
+<!-- histogram_and_ecdf_helper("u_alpha_a") -->
+<!-- ``` -->
 
-## `u_alpha_as`
+<!-- ## `u_alpha_as` -->
 
-```{r eval = pars_eval[["u_alpha_as"]]}
-histogram_and_ecdf_helper("u_alpha_as")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_as"]]} -->
+<!-- histogram_and_ecdf_helper("u_alpha_as") -->
+<!-- ``` -->
 
-## `u_alpha_xa`
+<!-- ## `u_alpha_xa` -->
 
-```{r eval = pars_eval[["u_alpha_xa"]]}
-histogram_and_ecdf_helper("u_alpha_xa")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_xa"]]} -->
+<!-- histogram_and_ecdf_helper("u_alpha_xa") -->
+<!-- ``` -->
 
-## `ui_lambda_x`
+<!-- ## `ui_lambda_x` -->
 
-```{r eval = pars_eval[["ui_lambda_x"]]}
-histogram_and_ecdf_helper("ui_lambda_x")
-```
+<!-- ```{r eval = pars_eval[["ui_lambda_x"]]} -->
+<!-- histogram_and_ecdf_helper("ui_lambda_x") -->
+<!-- ``` -->
 
-## `ui_anc_rho_x`
+<!-- ## `ui_anc_rho_x` -->
 
-```{r eval = pars_eval[["ui_anc_rho_x"]]}
-histogram_and_ecdf_helper("ui_anc_rho_x")
-```
+<!-- ```{r eval = pars_eval[["ui_anc_rho_x"]]} -->
+<!-- histogram_and_ecdf_helper("ui_anc_rho_x") -->
+<!-- ``` -->
 
-## `ui_anc_alpha_x`
+<!-- ## `ui_anc_alpha_x` -->
 
-```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-histogram_and_ecdf_helper("ui_anc_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["ui_anc_alpha_x"]]} -->
+<!-- histogram_and_ecdf_helper("ui_anc_alpha_x") -->
+<!-- ``` -->
 
-## `log_or_gamma`
+<!-- ## `log_or_gamma` -->
 
-```{r eval = pars_eval[["log_or_gamma"]]}
-histogram_and_ecdf_helper("log_or_gamma")
-```
+<!-- ```{r eval = pars_eval[["log_or_gamma"]]} -->
+<!-- histogram_and_ecdf_helper("log_or_gamma") -->
+<!-- ``` -->
 
@@
 ```{r}
-names(tmb$fit$sample)[!(names(tmb$fit$sample) %in% pars)]
+names(tmb$fit$sample)[!(names(tmb$fit$sample) %in% unique(names(tmb$fit$obj$env$par)))]
 ```
@@
 
+Something weird happening with `adam`, go back and check this:
+
 ```{r}
-histogram_and_ecdf(par = "rho_t1_out", i = 1)
+histogram_and_ecdf(par = "alpha_t1_out", i = 1)
 ```
@@
 ```{r}
-ks_helper <- function(par, sw = FALSE, ...) {
-  to_ks_df(par, starts_with = sw) %>% ks_plot(par, ...)
+ks_helper <- function(par, starts_with = FALSE, ...) {
+  to_ks_df(par, starts_with = starts_with) %>% ks_plot(par, ...)
 }
@@
 ```{r}
-ks_helper("beta", sw = TRUE)
-ks_helper("beta", sw = TRUE, method1 = "TMB", method2 = "adam")
+ks_helper("beta", starts_with = TRUE)
+ks_helper("beta", starts_with = TRUE, method1 = "TMB", method2 = "adam")
 ```
 
-### `logit`
+<!-- ### `logit` -->
+
+<!-- ```{r} -->
+<!-- ks_helper("logit", starts_with = TRUE) -->
+<!-- ks_helper("logit", starts_with = TRUE, method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `log_sigma` -->
+
+<!-- ```{r} -->
+<!-- ks_helper("log_sigma", starts_with = TRUE) -->
+<!-- ks_helper("log_sigma", starts_with = TRUE , method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_rho_x` -->
+
+<!-- ```{r eval = pars_eval[["u_rho_x"]]} -->
+<!-- ks_helper("u_rho_x") -->
+<!-- ks_helper("u_rho_x", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_rho_xs` -->
+
+<!-- ```{r eval = pars_eval[["u_rho_xs"]]} -->
+<!-- ks_helper("u_rho_xs") -->
+<!-- ks_helper("u_rho_xs", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `us_rho_x` -->
+
+<!-- ```{r eval = pars_eval[["us_rho_x"]]} -->
+<!-- ks_helper("us_rho_x") -->
+<!-- ks_helper("us_rho_x", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `us_rho_xs` -->
+
+<!-- ```{r eval = pars_eval[["us_rho_xs"]]} -->
+<!-- ks_helper("us_rho_xs") -->
+<!-- ks_helper("us_rho_xs", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_rho_a` -->
+
+<!-- ```{r eval = pars_eval[["u_rho_a"]]} -->
+<!-- ks_helper("u_rho_a") -->
+<!-- ks_helper("u_rho_a", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_rho_as` -->
+
+<!-- ```{r eval = pars_eval[["u_rho_as"]]} -->
+<!-- ks_helper("u_rho_as") -->
+<!-- ks_helper("u_rho_as", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_alpha_x` -->
+
+<!-- ```{r eval = pars_eval[["u_alpha_x"]]} -->
+<!-- ks_helper("u_alpha_x") -->
+<!-- ks_helper("u_alpha_x", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_alpha_xs` -->
+
+<!-- ```{r eval = pars_eval[["u_alpha_xs"]]} -->
+<!-- ks_helper("u_alpha_xs") -->
+<!-- ks_helper("u_alpha_xs", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `us_alpha_x` -->
+
+<!-- ```{r eval = pars_eval[["us_alpha_x"]]} -->
+<!-- ks_helper("us_alpha_x") -->
+<!-- ks_helper("us_alpha_x", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `us_alpha_xs` -->
+
+<!-- ```{r eval = pars_eval[["us_alpha_xs"]]} -->
+<!-- ks_helper("us_alpha_xs") -->
+<!-- ks_helper("us_alpha_xs", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_alpha_a` -->
+
+<!-- ```{r eval = pars_eval[["u_alpha_a"]]} -->
+<!-- ks_helper("u_alpha_a") -->
+<!-- ks_helper("u_alpha_a", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_alpha_as` -->
+
+<!-- ```{r eval = pars_eval[["u_alpha_as"]]} -->
+<!-- ks_helper("u_alpha_as") -->
+<!-- ks_helper("u_alpha_as", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `u_alpha_xa` -->
+
+<!-- ```{r eval = pars_eval[["u_alpha_xa"]]} -->
+<!-- ks_helper("u_alpha_xa") -->
+<!-- ks_helper("u_alpha_xa", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `ui_anc_rho_x` -->
+
+<!-- ```{r eval = pars_eval[["ui_anc_rho_x"]]} -->
+<!-- ks_helper("ui_anc_rho_x") -->
+<!-- ks_helper("ui_anc_rho_x", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `ui_anc_alpha_x` -->
+
+<!-- ```{r eval = pars_eval[["ui_anc_alpha_x"]]} -->
+<!-- ks_helper("ui_anc_alpha_x") -->
+<!-- ks_helper("ui_anc_alpha_x", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+<!-- ### `log_or_gamma` -->
+
+<!-- ```{r eval = pars_eval[["log_or_gamma"]]} -->
+<!-- ks_helper("log_or_gamma") -->
+<!-- ks_helper("log_or_gamma", method1 = "TMB", method2 = "adam") -->
+<!-- ``` -->
+
+### Outcome variables
 
 ```{r}
-ks_helper("logit", sw = TRUE)
-ks_helper("logit", sw = TRUE, method1 = "TMB", method2 = "adam")
-```
-
-### `log_sigma`
-
-```{r}
-ks_helper("log_sigma", sw = TRUE)
-ks_helper("log_sigma", sw = TRUE , method1 = "TMB", method2 = "adam")
-```
-
-### `u_rho_x`
-
-```{r eval = pars_eval[["u_rho_x"]]}
-ks_helper("u_rho_x")
-ks_helper("u_rho_x", method1 = "TMB", method2 = "adam")
-```
-
-### `u_rho_xs`
-
-```{r eval = pars_eval[["u_rho_xs"]]}
-ks_helper("u_rho_xs")
-ks_helper("u_rho_xs", method1 = "TMB", method2 = "adam")
-```
-
-### `us_rho_x`
-
-```{r eval = pars_eval[["us_rho_x"]]}
-ks_helper("us_rho_x")
-ks_helper("us_rho_x", method1 = "TMB", method2 = "adam")
-```
-
-### `us_rho_xs`
-
-```{r eval = pars_eval[["us_rho_xs"]]}
-ks_helper("us_rho_xs")
-ks_helper("us_rho_xs", method1 = "TMB", method2 = "adam")
-```
-
-### `u_rho_a`
-
-```{r eval = pars_eval[["u_rho_a"]]}
-ks_helper("u_rho_a")
-ks_helper("u_rho_a", method1 = "TMB", method2 = "adam")
-```
-
-### `u_rho_as`
-
-```{r eval = pars_eval[["u_rho_as"]]}
-ks_helper("u_rho_as")
-ks_helper("u_rho_as", method1 = "TMB", method2 = "adam")
-```
-
-### `u_alpha_x`
-
-```{r eval = pars_eval[["u_alpha_x"]]}
-ks_helper("u_alpha_x")
-ks_helper("u_alpha_x", method1 = "TMB", method2 = "adam")
-```
-
-### `u_alpha_xs`
-
-```{r eval = pars_eval[["u_alpha_xs"]]}
-ks_helper("u_alpha_xs")
-ks_helper("u_alpha_xs", method1 = "TMB", method2 = "adam")
-```
-
-### `us_alpha_x`
-
-```{r eval = pars_eval[["us_alpha_x"]]}
-ks_helper("us_alpha_x")
-ks_helper("us_alpha_x", method1 = "TMB", method2 = "adam")
-```
-
-### `us_alpha_xs`
-
-```{r eval = pars_eval[["us_alpha_xs"]]}
-ks_helper("us_alpha_xs")
-ks_helper("us_alpha_xs", method1 = "TMB", method2 = "adam")
-```
-
-### `u_alpha_a`
-
-```{r eval = pars_eval[["u_alpha_a"]]}
-ks_helper("u_alpha_a")
-ks_helper("u_alpha_a", method1 = "TMB", method2 = "adam")
-```
-
-### `u_alpha_as`
-
-```{r eval = pars_eval[["u_alpha_as"]]}
-ks_helper("u_alpha_as")
-ks_helper("u_alpha_as", method1 = "TMB", method2 = "adam")
-```
-
-### `u_alpha_xa`
-
-```{r eval = pars_eval[["u_alpha_xa"]]}
-ks_helper("u_alpha_xa")
-ks_helper("u_alpha_xa", method1 = "TMB", method2 = "adam")
-```
-
-### `ui_anc_rho_x`
-
-```{r eval = pars_eval[["ui_anc_rho_x"]]}
-ks_helper("ui_anc_rho_x")
-ks_helper("ui_anc_rho_x", method1 = "TMB", method2 = "adam")
-```
-
-### `ui_anc_alpha_x`
-
-```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-ks_helper("ui_anc_alpha_x")
-ks_helper("ui_anc_alpha_x", method1 = "TMB", method2 = "adam")
-```
-
-### `log_or_gamma`
-
-```{r eval = pars_eval[["log_or_gamma"]]}
-ks_helper("log_or_gamma")
-ks_helper("log_or_gamma", method1 = "TMB", method2 = "adam")
+# To fix
+# ks_helper("alpha_t1_out")
 ```

### 2023-03-24 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
Work on slides for HIV Lab Group #25

---
title: Deterministic Bayesian inference methods for the Naomi model
subtitle: HIV Inference Lab Group Meeting
author: Adam Howes
institute: Imperial College London
date: April 2023
bibliography: citations.bib
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

# Bayes

* As a statistical modeller, the bulk of our job is in constructing a generative model for data $y$ using parameters $\vartheta$
* This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
* What we actually want is the posterior $p(\vartheta \, | \, y)$ which is **just**^[I've bolded this with sarcasm in mind]
$$
p(\vartheta \, | \, y) = \frac{p(y, \vartheta)}{p(y)} = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)}
$$
* The central problem of Bayesian inference is that we can't compute $p(y)$
$$
p(y) = \int p(y, \vartheta) \text{d} \vartheta
$$

# How might you do it?

* If you want to integrate something deterministically, you could use numerical integration methods
* Pick nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ then compute the sum
$$
\tilde p(y) = \sum_{\vartheta \in \mathcal{Q}} p(y, \vartheta) \omega(\vartheta)
$$

# Monte Carlo as an example of numerical integration

* Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$
* If we set $\omega(\vartheta_i) = 1/N$ for all $i$ then we get a **Monte Carlo** (MC) estimate
$$
\tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i)
$$
* For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain which if you squint a bit is good enough (MCMC)

# Monte Carlo is fundamentally unsound

* "Monte Carlo ignores information" according to @o1987monte
* Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is
$$
\tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right)
$$
* This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$

# Application to HIV survey sampling

* This is a digression but...
* Say we're running a household survey, and sample the same individual twice
* We didn't learn anything new about HIV by surveying them again!
* This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
$\implies$ Bayesian quadrature, Bayesian survey design \\
For some half-baked thoughts, see \texttt{athowes.github.io/fourth-gen/paper.pdf}
\end{tcolorbox}
\end{center}

# Latent variables and hyperparameters

* Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized
* Previously I had all of the parameters under the symbol $\vartheta$
* What if we split them up as being $\vartheta = (x, \theta)$
* The key part about this is that $\dim(x)$ is big and $\dim(\theta)$ is small

| Names for $x$ | Names for $\theta$ |
|---------------|--------------------|
| Latent variables, random effects, latent field | Hyperparameters, fixed effects

# Spatio-temporal statistics

* There is nothing inherently special about spatio-temporal statistics
* We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$
* Usually we associate parameters to spatio-temporal locations as well as observations
* This ends up with us having something like $\{x_{s, t}\}$

# What's important about this?

1. There might be **a lot** of spatio-temporal locations, so $\text{dim}(x)$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
2. Perhaps we're willing to make assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]

<!-- As you get more data, the posterior approaches a Gaussian -->

# Latent Gaussian models

A latent Gaussian model (LGM) [@rue2009approximate] looks along these lines:

\begin{alignat*}{2}
  &\text{(Observations)}     &   y &\sim p(y \, | \, x, \theta), \\
  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x  \, | \, \mu(\theta), Q(\theta)^{-1}), \\
  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta).
\end{alignat*}

# Laplace approximation

* Remember that we wanted to compute
$$
p(y) = \int p(y, \vartheta) \text{d} \vartheta
$$

* One trick for doing this is to pretend $p(\vartheta \, | \, y)$ is Gaussian

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
\begin{itemize}
\item Mode $\hat \vartheta = \argmax_\vartheta \log p(y, \vartheta)$
\item Hessian $H(\hat \vartheta) = - \partial_\vartheta^2 \log p(y, \vartheta) \rvert_{\vartheta = \hat \vartheta}$
\item Gaussian approximation $\implies \tilde p_{\texttt{G}}(\vartheta \, | \, y) = \mathcal{N}(\vartheta \, | \, \hat \vartheta, H(\hat \vartheta)^{-1})$
\end{itemize}
\end{tcolorbox}
\end{center}

# Laplace approximation

Include a figure here showing what it looks like to approximate some distribution by a Gaussian.

# Laplace approximation

* Now
$$
p(y) = \frac{p(\vartheta, y)}{p(\vartheta \, | \, y)} \approx \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)}
$$
and we can evaluate RHS where we would like, so let's pick the point at which the Gaussian is most accurate, which is $\hat \vartheta$
$$
p_{\texttt{LA}}(y) = \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)} \rvert_{\vartheta = \hat \vartheta} = (2 \pi)^{\dim(\vartheta) / 2} \det(H(\hat \vartheta))^{- 1 / 2} p(\hat \vartheta, y)
$$

# Marginal Laplace approximation

* Hey wait a second, is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian?
* No, not in general. But...

1. We just described a class of models (LGMs) where some subset of the parameters (the latent field $x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian
2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a **marginal Laplace** approximation to get rid of $x$

#

Dichotomy in statistical inference methods:

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]

\begin{enumerate}
  \item Ones which aim to be completely general
  \item Ones which aim to "exploit" properties of the problem at hand
\end{enumerate}

\end{tcolorbox}
\end{center}

We are taking approach 2.

# Marginal Laplace approximation

* What does this look like? Instead of assuming $p(\vartheta \, | \, y) = p(x, \theta \, | \, y)$ is Gaussian we assume $p(x \, | \, \theta, y)$ is
$$
\tilde p_{\texttt{G}}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x, H(\hat x))^{-1}
$$
where $\hat x = \hat x(\theta)$
* Now the marginal Laplace approximation is
$$
p_{\texttt{LA}}(\theta, y) = \frac{p(x, \theta, y)}{\tilde p_{\texttt{G}}(x \, | \, \theta, y)} \rvert_{x = \hat x} = (2 \pi)^{\dim(x) / 2} \det(H(\hat x))^{- 1 / 2} p(\hat x, \theta, y)
$$

# Integrated nested Laplace approximation

* Now we can compute $p_{\texttt{LA}}(\theta, y)$ but what we really want is still $p(y)$
* But hopefully^[Really: hopefully] the dimension of $\theta$ is small enough that we can now tackle this with quadrature
* So pick some nodes $\mathcal{Q}$ and a weighting function $\omega$ and away we go
$$
p(y) \approx \sum_{\theta \in \mathcal{Q}} p_{\texttt{LA}}(\theta, y) \omega(\theta)
$$
* This is the famous integrated nested Laplace approximation (INLA)

#

> My main comment is that several aspects of the computational machineery that is presented by Rue and his colleagues **could benefit from the use of a numerical technique known as automatic differentiation (AD)** ... By the use of AD one could obtain a system that is automatic from a user's perspective... the benefit would be a fast, flexible and easy-to-use system for doing Bayesian analysis in models with Gaussian latent variables

-- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate

# Thanks for listening!

* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work, joint with Alex Stringer (Waterloo) and my PhD supervisors Seth Flaxman (Oxford) and Jeff Eaton (Imperial)
* Let me know if you'd be up for being an early reader!
* Code for this project is at `athowes.github.io/elgm-inf`

# References {.allowframebreaks}

### 2023-03-25 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
Update slides #25

@@
 * This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
-* What we actually want is the posterior $p(\vartheta \, | \, y)$ which is **just**^[I've bolded this with sarcasm in mind]
+* Then we want to compute the posterior $p(\vartheta \, | \, y)$ which is **just**^[I've bolded this with sarcasm in mind]
 $$
@@
 $$
+
+# Naive quadrature example
+
+* Remember how integration is introduced? (Rienmann sums)
+* Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule
+
+```{r}
+trapezoid_rule <- function(x, spacing) {
+  w <- rep(spacing, length(x))
+  w[1] <- w[1] / 2
+  w[length(x)] <- w[length(x)] / 2
+  sum(w * x)
+}
+```
+
+#
+
+```{r message=FALSE, echo=FALSE}
+library(tidyverse)
+cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
+finegrid <- seq(0, pi, length.out = 1000)
+
+plot <- data.frame(x = finegrid, y = sin(finegrid)) %>%
+  ggplot(aes(x = x, y = y)) +
+   geom_line(col = cbpalette[1]) +
+   theme_minimal() +
+   labs(x = "x", y = "sin(x)")
+
+trapezoid_plot <- function(N) {
+  grid <- seq(0, pi, length.out = N)
+  int <- trapezoid_rule(x = sin(grid), spacing = grid[2] - grid[1])
+
+  plot + 
+    geom_bar(
+      data = data.frame(x = grid, y = sin(grid)),
+      aes(x = x, y = y), alpha = 0.7, stat = "identity",
+      inherit.aes = FALSE, fill = cbpalette[2]) +
+    theme_minimal() +
+    labs(
+      subtitle = paste0("Number of quadrature points: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"),
+      x = "x", y = "sin(x)"
+    )
+}
+
+trapezoid_plot(N = 10)
+```
+
+#
+
+```{r message=FALSE, echo=FALSE}
+trapezoid_plot(N = 30)
+```
+
+#
+
+```{r message=FALSE, echo=FALSE}
+trapezoid_plot(N = 100)
+```
 
@@
 
-* There is nothing inherently special about spatio-temporal statistics
+* There is nothing inherently special about spatio-temporal statistics, but we do often end up in similar situations because of the structures of the problems we tackle
 * We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$
@@
 
-# Laplace approximation
+# Example
 
-Include a figure here showing what it looks like to approximate some distribution by a Gaussian.
+* Consider the following model for $i = 1, \ldots, n$ with fixed $a$ and $b$
+$$
+y_i \sim \text{Poisson}(\lambda), \quad
+\lambda \sim \text{Gamma}(a, b).
+$$
+* It's conjugate so we directly know that
+$$
+\lambda \, | \, y \sim \text{Gamma}(a + \sum_i y_i, b + n)
+$$
 
@@
 
+# Naomi evidence synthesis model
+
+:::::::::::::: {.columns}
+
+::: {.column width=.65}
+
+* Many people here have worked on Naomi evidence synthesis model
+* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
+* Inference for Naomi is currently conducted using `TMB` by optimising $p_{\texttt{LA}}(\theta, y)$
+* Has to be pretty quick to allow for interactive review and development of estimates
+
+:::
+
+::: {.column width=.35}
+
+```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'}
+knitr::include_graphics("naomi_hex.png")
+```
+
+:::
+
+::::::::::::::
+
 #
+
+```{r, echo=FALSE, fig.cap="Example of the user interface from https://naomi.unaids.org/", out.width = '70%'}
+knitr::include_graphics("naomi_user.png")
+```
+
+# Why do we use `TMB`
+
+* One answer: lack of other viable options
+* Markov chain Monte Carlo (MCMC) takes too long
+* Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a latent Gaussian model, and isn't compatible with the `R-INLA` implementation of INLA
+
+# Idea
+
+* Implement an algorithm similar to INLA which
+1. is compatible with Naomi
+2. uses `TMB` to perform Laplace approximation 
+* Does this improves the quality of the inferences we get?
+
+# Not a new idea!
 
@@
 -- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate
+
+# Meanwhile in Canada...
+
+* Alex Stringer in Toronto $\to$ Waterloo had been thinking along similar lines, and made a lot of progress
+
+1.
+Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which he and others argue should be the default for this problem
+2. Defining a class of models called extended latent Gaussian models (ELGMs) which Naomi fits into
+
+* I will explain what both of these acronyms (AGHQ^[Amusingly similar to AGYW: adolescent girls and young women] and ELGM) mean!
+
+# What's AGHQ?
+
+# What's an ELGM?
+
+# Why is Naomi an ELGM?
+
+# ANC offset from household survey
+
+\begin{align*}
+\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
+\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}. 
+\end{align*}
+
+# Incidence depends on adult prevalence and coverage
+
+$$
+\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda.
+$$
+
+# ART attendance probability product
+
+$$
+\dot A_{x', x, s, a} \sim \text{Bin}(N_{x', s, a}, \pi_{x', x, s, a})
+$$
+where $\pi_{x', x, s, a} = \rho_{x', s, a} \cdot \alpha_{x', s, a} \cdot \gamma_{x', x, s, a}$.
+
+# ART attendance multinomial
+
+# Aggregation
+
+#
+
+```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the hyperparameters. Shaped like a snake for no real reason.", fig.ext = 'png', cache=TRUE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1]
+\node (1) {$p(\theta, x, y)$};
+\node (2) [below of=1] {$\tilde p_\texttt{LA}(\theta, y)$};
+\node (3) [right of=2] {$\tilde p_\texttt{AQ}(y)$};
+\node (4) [above of=3] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$};
+\node (5) [right of=4] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$};
+\node (6) [below of=5] {Output};
+
+\draw[->] (1) to node {Laplace} (2);
+\draw[->] (2) to node {AGHQ} (3);
+\draw[->] (3) to node {Normalise} (4);
+\draw[->] (4) to node {Marginalise} (5);
+\draw[->] (5) to node {Spline} (6);
+\end{tikzpicture}
+```
+
+#
+
+```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the latent field.", fig.ext = 'png', cache=TRUE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1]
+\node (1) {$p(\theta, x, y)$};
+\node (2) [below of=1] {$\tilde p_\texttt{LA}(\theta, y)$};
+\node (3) [right of=2] {$\tilde p_\texttt{AQ}(y)$};
+\node (4) [above of=3] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$};
+\node (5) [right of=4] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$};
+\node (6) [below of=5] {Output};
+
+\draw[->] (1) to node {Laplace} (2);
+\draw[->] (2) to node {AGHQ} (3);
+\draw[->] (3) to node {Normalise} (4);
+\draw[->] (4) to node {Marginalise} (5);
+\draw[->] (5) to node {Spline} (6);
+\end{tikzpicture}
+```
+
+# Latent field algorithm
 

### 2023-03-26 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
Add LA example, AGHQ example to slides #25

@@
 
-# Bayes
+# Bayesian modelling and inference
 
@@
 * This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
-* Then we want to compute the posterior $p(\vartheta \, | \, y)$ which is **just**^[I've bolded this with sarcasm in mind]
+* Then, we want to compute (ideally without having to think much about it) the posterior $p(\vartheta \, | \, y)$ which is **just**^[I've bolded this with sarcasm in mind: it's a difficult problem]
 $$
@@
 $$
-* The central problem of Bayesian inference is that we can't compute $p(y)$
+* The central problem of Bayesian inference is doing the following integral
 $$
@@
 
-# How might you do it?
+# Numerical integration
 
-* If you want to integrate something deterministically, you could use numerical integration methods
-* Pick nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ then compute the sum
+* If you want to integrate something deterministically, you could use numerical integration, otherwise called \textcolor{hilit}{quadrature}
+* Select nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ then compute the sum
 $$
@@
 
-* Remember how integration is introduced? (Rienmann sums)
+* Remember how integration is taught? (Rienmann sums)
 * Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule
@@
 * Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized
-* Previously I had all of the parameters under the symbol $\vartheta$
-* What if we split them up as being $\vartheta = (x, \theta)$
+  * If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$
+* Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$
 * The key part about this is that $\dim(x)$ is big and $\dim(\theta)$ is small
@@
 1. There might be **a lot** of spatio-temporal locations, so $\text{dim}(x)$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
-2. Perhaps we're willing to make assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]
+2. Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]
+* Not IID anymore!
 
@@
 
-A latent Gaussian model (LGM) [@rue2009approximate] looks along these lines:
-
+* A \textcolor{hilit}{latent Gaussian model} (LGM) [@rue2009approximate] looks along these lines:
 \begin{alignat*}{2}
@@
 \end{alignat*}
+* Many models are LGMs, especially in spatio-temporal statistics
 
@@
 
-# Example
+# Example of computing the Laplace approximation
 
@@
 $$
+
+# Prior
+
+```{r message=FALSE, echo=FALSE}
+a <- 3
+b <- 1
+
+prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) +
+  stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) +
+  labs(x = "", y = "") +
+  theme_minimal()
+
+prior
+```
+
+# Posterior
+
+```{r message=FALSE, echo=FALSE}
+set.seed(2)
+y <- rpois(3, lambda = 2)
+
+posterior <- prior +
+  geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) +
+  stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2])
+
+posterior
+```
+
+# Laplace approximation
+
+```{r message=FALSE}
+fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE)
+
+# Here we are using numerical derivatives rather than automatic
+ff <- list(
+  fn = fn,
+  gr = function(x) numDeriv::grad(fn, x),
+  he = function(x) numDeriv::hessian(fn, x)
+)
+
+opt_bfgs <- aghq::optimize_theta(
+  ff, 1, control = aghq::default_control(method = "BFGS")
+)
+```
+
+# Laplace approximation
+
+```{r}
+laplace <- posterior +
+  stat_function(
+    data = data.frame(x = c(0, 10)),
+    aes(x),
+    fun = dnorm,
+    n = 500,
+    args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)),
+    col = cbpalette[3]
+  )
+```
+
+# Laplace approximation
+
+```{r message=FALSE, echo=FALSE}
+laplace
+```
 
@@
 1.
-We just described a class of models (LGMs) where some subset of the parameters (the latent field $x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian
+We just described a class of models (LGMs) where some subset of the parameters ($x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian
 2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a **marginal Laplace** approximation to get rid of $x$
 
-#
+# Generalist versus specialist methods
 
@@
 \begin{enumerate}
-  \item Ones which aim to be completely general
-  \item Ones which aim to "exploit" properties of the problem at hand
+  \item \textbf{Generalist} works in all situations
+  \item \textbf{Specialist} "exploits" properties of the problem at hand
 \end{enumerate}
@@
 
-We are taking approach 2.
+We are taking approach 2!
 
@@
 
-# Naomi evidence synthesis model
+# Taking stock
+
+1. Bayesian inference is integration
+2. Spatial statistics has parameters $(x, \theta)$
+3. Integrate $x$ cheaply using a Gaussian assumption
+4. Try a bit harder with $\theta$ performing quadrature
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+
+Now let's apply this to a difficult problem in HIV inference!
+
+\end{tcolorbox}
+\end{center}
+
+# The Naomi model
 
@@
 
-* Many people here have worked on Naomi evidence synthesis model
+* Naomi is a spatial evidence synthesis model
 * Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
-* Inference for Naomi is currently conducted using `TMB` by optimising $p_{\texttt{LA}}(\theta, y)$
-* Has to be pretty quick to allow for interactive review and development of estimates
+* Inference for Naomi is currently conducted using `TMB` by optimising $p_{\texttt{LA}}(\theta, y)$, and has to be pretty quick to allow for interactive review and development of estimates
 
@@
 
-```{r, echo=FALSE, fig.cap="Example of the user interface from https://naomi.unaids.org/", out.width = '70%'}
+```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'}
 knitr::include_graphics("naomi_user.png")
+```
+
+# Template Model Builder refresher
+
+* `TMB` [@kristensen2015tmb] is an R package which implements the Laplace approximation for latent variable models
+* To use `TMB`, you write your objective function $- \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)$ in `TMB`'s C++ syntax
+* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks as follows
+
+#
+
+```{cpp eval=FALSE}
+#include <TMB.hpp>
+
+template <class Type>
+Type objective_function<Type>::operator()() {
+  // Define data e.g.
+  DATA_VECTOR(y);
+  // Define parameters e.g.
+  PARAMETER(mu);
+  // Calculate negative log-likelihood e.g.
+  nll = Type(0.0);
+  nll -= dnorm(y, mu, 1, true).sum()
+  return(nll);
+}
 ```
@@
 
-* One answer: lack of other viable options
-* Markov chain Monte Carlo (MCMC) takes too long
+* It runs quickly, and is flexible enough to write the model
+* Another answer: we don't have any better options
+  * Markov chain Monte Carlo (MCMC) is accurate (eventually) but takes too long
   * Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a latent Gaussian model, and isn't compatible with the `R-INLA` implementation of INLA
@@
 
-# What's AGHQ?
+# Gauss-Hermite quadrature
 
-# What's an ELGM?
+# Adaptation
+
+* The nodes and weights we use should probably depend on the integrand
+  * Especially when the integrand is also a funciton of $y$ as in $p(y, \vartheta)$
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points."}
+mu <- c(1, 1.5)
+cov <- matrix(c(2, 1, 1, 1), ncol = 2)
+
+obj <- function(theta) {
+  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
+}
+
+grid <- expand.grid(
+  theta1 = seq(-2, 5, length.out = 700),
+  theta2 = seq(-2, 5, length.out = 700)
+)
+
+ground_truth <- cbind(grid, pdf = obj(grid))
+
+plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = cbpalette[1]) +
+  coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
+  labs(x = "", y = "") +
+  theme_minimal()
+
+gg <- mvQuad::createNIGrid(2, "GHe", 3)
+
+add_points <- function(plot0, gg) {
+  plot0 +
+    geom_point(
+      data = mvQuad::getNodes(gg) %>%
+              as.data.frame() %>%
+              mutate(weights = mvQuad::getWeights(gg)), 
+      aes(x = V1, y = V2, size = weights),
+      alpha = 0.8,
+      col = cbpalette[2],
+      inherit.aes = FALSE
+    ) +
+    scale_size_continuous(range = c(1, 2))
+}
+
+add_points(plot0, gg) +
+  labs(size = "Weight", caption = "")
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean."}
+gg2 <- gg
+mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2)
+
+add_points(plot0, gg2) +
+  labs(size = "Weight", caption = "")
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Rotate by the lower Cholesky."}
+gg3 <- gg
+mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)
+
+add_points(plot0, gg3) +
+  labs(size = "Weight", caption = "")
+```
+
+# Extended latent Gaussian models
 
 # Why is Naomi an ELGM?
+
+1. ANC offset from household survey
+2. Incidence depends on adult prevalence and coverage
+3. ART attendance is a product
+4. ART attendance uses a multinomial
+5. Aggregation of finer processes
 
@@
 
-# ART attendance probability product
+# ART attendance is a product
 
@@
 
-# Aggregation
+# Aggregation of finer processes
+
+# The algorithm
 
@@
 
-# Latent field algorithm
-
 # Thanks for listening!

### 2023-03-27 src/naomi-simple_compare/comparison.Rmd
Success on merging KS results into mf_out_fine: now have dataframe of KS results at finest level with district, age, sex information available #24

@@
 ```{r}
-# To fix
-# ks_helper("alpha_t1_out")
+ks_rho_t1_out <- to_ks_df(par = "rho_t1_out", outputs = TRUE, id = mf_out_fine$id)
+
+merge_to_mf_out_fine <- function(ks_df) {
+  ks_df %>%
+  select(-par) %>%
+  left_join(
+    mf_out_fine %>%
+      rename("full_id" = "id") %>%
+      tibble::rowid_to_column("index") %>%
+      mutate(index = as.numeric(index))
+  ) %>%
+  select(-index)
+}
+
+ks_rho_t1_out <- merge_to_mf_out_fine(ks_rho_t1_out)
+
+ks_alpha_t1_out <- to_ks_df(par = "alpha_t1_out", outputs = TRUE, id = mf_out_fine$id) %>%
+  merge_to_mf_out_fine()
+
+ks_lambda_t1_out <- to_ks_df(par = "lambda_t1_out", outputs = TRUE, id = mf_out_fine$id) %>%
+  merge_to_mf_out_fine()
 ```

### 2023-03-27 src/naomi-simple_compare/comparison.Rmd
ks_plot working for model outputs too #24

@@
   merge_to_mf_out_fine()
+
+ks_plot(ks_df = ks_rho_t1_out, par = "rho_t1_out")
+ks_plot(ks_df = ks_alpha_t1_out, par = "alpha_t1_out")
+ks_plot(ks_df = ks_lambda_t1_out, par = "lambda_t1_out")
 ```

### 2023-03-27 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
More work on slides #25

@@
 * This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
-* Then, we want to compute (ideally without having to think much about it) the posterior $p(\vartheta \, | \, y)$ which is **just**^[I've bolded this with sarcasm in mind: it's a difficult problem]
+* Then, we want to compute (ideally without having to think much about it) the posterior $p(\vartheta \, | \, y)$ which is \textcolor{hilit}{just}^[I've highlighted this with sarcasm in mind: it's a difficult problem]
 $$
@@
 * Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$
-* The key part about this is that $\dim(x)$ is big and $\dim(\theta)$ is small
+* The key part about this is that $\dim(x) = N$ is big and $\dim(\theta) = m$ is small
 
@@
 
-1. There might be **a lot** of spatio-temporal locations, so $\text{dim}(x)$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
+1.
+There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $\text{dim}(x)$ might be pretty big!
+If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
 2. Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]
@@
 * Many models are LGMs, especially in spatio-temporal statistics
+* $\dim(x) = n$, $\dim(x) = N$, $\dim(\theta) = m$
 
@@
 
-# Prior
+#
 
-```{r message=FALSE, echo=FALSE}
+```{r message=FALSE, echo=FALSE,  fig.cap="A Gamma prior with $a = 3$ and $b = 1$."}
 a <- 3
@@
 
-# Posterior
+#
 
-```{r message=FALSE, echo=FALSE}
+```{r message=FALSE, echo=FALSE,  fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."}
 set.seed(2)
@@
 
-# Laplace approximation
+#
 
@@
 
-```{r message=FALSE, echo=FALSE}
+```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation matches the true posterior near the mode but it's not great in the tails."}
 laplace
@@
 \begin{enumerate}
-  \item \textbf{Generalist} works in all situations
-  \item \textbf{Specialist} "exploits" properties of the problem at hand
+  \item Generalist: works in all situations
+  \item Specialist: "exploits" properties of the problem at hand
 \end{enumerate}
@@
 $$
-p_{\texttt{LA}}(\theta, y) = \frac{p(x, \theta, y)}{\tilde p_{\texttt{G}}(x \, | \, \theta, y)} \rvert_{x = \hat x} = (2 \pi)^{\dim(x) / 2} \det(H(\hat x))^{- 1 / 2} p(\hat x, \theta, y)
+p_{\texttt{LA}}(\theta, y) = \frac{p(x, \theta, y)}{\tilde p_{\texttt{G}}(x \, | \, \theta, y)} \rvert_{x = \hat x} = (2 \pi)^{N / 2} \det(H(\hat x))^{- 1 / 2} p(\hat x, \theta, y)
 $$
@@
 
+* Recall
+$$
+\int f(z) \text{d} z \approx \sum_{z \in \mathcal{Q}} f(z) \omega(z)
+$$
+* Replace $f(z)$ by $\phi(z) f(z)$ and say that $f$ is a polynomial and $\phi$ is unknown
+* Suppose that $\phi(z) = \exp(-z^2)$
+
 # Adaptation
@@
 * The nodes and weights we use should probably depend on the integrand
-  * Especially when the integrand is also a funciton of $y$ as in $p(y, \vartheta)$
+  * Especially when the integrand is also a function of $y$ as in $p(y, \vartheta)$
+* Let $z \in \mathcal{Q}(m, k)$ then $\theta(z) = \hat \theta + Lz$ where $L$ is the lower Cholesky of $H(\hat \theta)$
 
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points."}
+```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points in two dimensions with $k = 3$."}
 mu <- c(1, 1.5)
@@
 
+* To explain what an extended latent Gaussian model (ELGM) is, we need a little more detail on LGMs
+* In an LGM the conditional mean depends on exactly one structured additive predictor, specifically
+\begin{align*}
+y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\
+\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
+\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
+\end{align*}
+
+# Extended latent Gaussian models
+
+* ELGM remove this requirement such that
+$$
+\mu_i = g(\eta_{\mathcal{J}_i})
+$$
+where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J_i}$ is some set of indices
+* This allow for a higher degree of non-linearity in the model
+
 # Why is Naomi an ELGM?
 
-1. ANC offset from household survey
-2. Incidence depends on adult prevalence and coverage
-3. ART attendance is a product
-4. ART attendance uses a multinomial
-5. Aggregation of finer processes
+| | Things I thought made Naomi an ELGM | Does it actually? |
+|-|---------------|--------------------|
+|1| ANC offset from household survey | ? |
+|2| Incidence depends on adult prevalence and coverage | ? |
+|3| ART attendance is a product | ? |
+|4| ART attendance uses a multinomial | ? |
+|5| Aggregation of finer processes | ? |
+
+* I will explain each of these, and whether or not they make Naomi an ELGM, in more detail in slides to follow
+  * The notation may not have been introduced properly, but hopefully the gist will still make sense
 
@@
 
+* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ are Gaussian
+* They have their own structured additive predictors, but by use of the `copy` feature this should be possible in `R-INLA`
+* Conclusion: \textcolor{hilit}{does not} make Naomi an ELGM
+
 # Incidence depends on adult prevalence and coverage
@@
 
+* Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian -- no matter your `copy` tricks
+* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
+
 # ART attendance is a product
 
-$$
-\dot A_{x', x, s, a} \sim \text{Bin}(N_{x', s, a}, \pi_{x', x, s, a})
-$$
-where $\pi_{x', x, s, a} = \rho_{x', s, a} \cdot \alpha_{x', s, a} \cdot \gamma_{x', x, s, a}$.
-
-# ART attendance multinomial
+# ART attendance uses the multinomial
 
@@
 \begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1]
-\node (1) {$p(\theta, x, y)$};
-\node (2) [below of=1] {$\tilde p_\texttt{LA}(\theta, y)$};
-\node (3) [right of=2] {$\tilde p_\texttt{AQ}(y)$};
-\node (4) [above of=3] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$};
-\node (5) [right of=4] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$};
-\node (6) [below of=5] {Output};
+\node (1) {$p(\theta, x_i, x_{-i}, y)$};
+\node (2) [below of=1] {$\tilde p_\texttt{LA}(x_i, \theta, y)$};
+\node (3) [right of=2] {$\tilde p_\texttt{AQ}(x_i, y)$};
+\node (4) [above of=3] {$\tilde p_\texttt{AQ}(x_i \, | \, y)$};
+\node (5) [right of=4] {Output};
 
@@
 \draw[->] (3) to node {Normalise} (4);
-\draw[->] (4) to node {Marginalise} (5);
-\draw[->] (5) to node {Spline} (6);
+\draw[->] (4) to node {Spline} (5);
 \end{tikzpicture}

### 2023-03-27 src/dev_scale-grid/scale-grid.Rmd
Clarify writing of scale-grid, and correct calculation of Gaussian integral #22

--- 2023-03-22 src/dev_scale-grid/scale-grid.Rmd
@@
 
-  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure.
+  **Task** We would like to begin to integrate over the hyperparameters using multiple nodes.
 The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical.
@@
 This is quick to do and allows us to get a model we can play with to figure out which hyperparmeters are important.
+In particular we obtain the mode and Hessian at the mode of the Laplace approximation to the hyperparameter marginal posterior.
 
@@
 
-# Grid ideas
+# Number of points proportional to standard deviation
 
@@
 There are `r length(hypers)` of them, if you'd like to count.
+There are this many on the logit scale:
+
+```{r}
+data.frame(hypers) %>%
+  filter(str_detect(hypers, "^logit_")) %>%
+  nrow()
+```
+
+And this many on the log scale:
+```{r}
+data.frame(hypers) %>%
+  filter(str_detect(hypers, "^log_")) %>%
+  nrow()
+```
 
@@
 
-## Number of points proportional to standard deviation
-
 A first suggestion about how to place grid points would be proportional to the standard deviations above.
@@
 
-This would look like the following:
+This would look as follows:
 
@@
 A limitation of this approach is that some hyperparameters have higher SD than others in part due to the different scales that they are on.
-For example if the hyper is on the logit scale then (from the plot above) it looks as though it will have a higher SD than another hyper on the log scale.
+For example if the hyper is on the logit scale then it looks as though it will have a higher SD than another hyper on the log scale based on the plot above.
+To be more precise:
+
+```{r}
+data.frame(par = names(sd), sd = unname(sd)) %>%
+  filter(str_detect(par, "^log_")) %>%
+  summarise(mean_sd = mean(sd))
+```
+
+```{r}
+data.frame(par = names(sd), sd = unname(sd)) %>%
+  filter(str_detect(par, "^logit_")) %>%
+  summarise(mean_sd = mean(sd))
+```
+
+Does this really reflect our intuititions about what is important to take greater account of?
 We would ideally like to know how much variation in each hyperparameter effects variation in model outputs, and then take that into account in the importance we place on each dimension: doing this in any good automated ways would be a nice contribution.
 
-## Eigendecomposition
+# Eigendecomposition
 
-Another option is to use principle components analysis (PCA) to create a grid on a subspace of $\mathbb{R}^{24}$ which retains most of the variation.
+Although it doesn't solve the above issue, another option to using the standard deviations is to use principle components analysis (PCA) to create a grid on a subspace of $\mathbb{R}^{24}$ which retains most of the variation.
 
@@
 Or, with 10 dimensions included, that percentage increases to `r round(100 * filter(tv_df, n == 10) %>% pull(tv), 1)`%.
-
-What do the PC loadings look like?
+What do the PC loadings (columns of the matrix `E`) look like?
 
@@
 
-C_plot
-
 C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
@@
     axis.text.x = element_blank(),
-    axis.text.y = element_blank()
+    axis.text.y = element_blank(),
+    legend.position = "bottom"
   )
 
-C_plot + C_s_plot
+{C_plot + theme(legend.position = "bottom")} + C_s_plot
 ```
@@
 
-### Testing with simple examples
+## Testing with simple examples
+
+### Gaussian
 
@@
 obj <- function(theta) {
-  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
+  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov, log = TRUE)
 }
+
+ff <- list(
+  fn = obj,
+  gr = function(theta) numDeriv::grad(obj, theta),
+  he = function(theta) numDeriv::hessian(obj, theta)
+)
 
@@
 
-ground_truth <- cbind(grid, pdf = obj(grid))
+ground_truth <- cbind(grid, pdf = exp(obj(grid)))
 
@@
 
-gaussian_2d <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
+pca_grid_1 <- pca_rescale(m = mu, C = cov, s = 1, k = 1)
+pca_grid_3 <- pca_rescale(m = mu, C = cov, s = 1, k = 3)
+pca_grid_5 <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
+pca_grid_7 <- pca_rescale(m = mu, C = cov, s = 1, k = 7)
 
@@
   geom_point(
-    data = data.frame(mvQuad::getNodes(gaussian_2d)) %>%
-      mutate(weight = mvQuad::getWeights(gaussian_2d)),
+    data = data.frame(mvQuad::getNodes(pca_grid_7)) %>%
+      mutate(weight = mvQuad::getWeights(pca_grid_7)),
     aes(x = X1, y = X2, size = weight),
@@
   scale_size_continuous(range = c(1, 2))
+
+opt_bfgs <- aghq::optimize_theta(ff, mu, control = default_control(method = "BFGS"))
+
+# Perform regular AGHQ with a product grid
+norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, k = 1)
+norm_bfgs_3 <- aghq::normalize_logpost(opt_bfgs, k = 3)
+norm_bfgs_5 <- aghq::normalize_logpost(opt_bfgs, k = 5)
+norm_bfgs_7 <- aghq::normalize_logpost(opt_bfgs, k = 7)
+
+# Zero is the correct answer here, so hopefully these are close!
+norm_bfgs_1$lognormconst
+norm_bfgs_3$lognormconst
+norm_bfgs_5$lognormconst
+norm_bfgs_7$lognormconst
+
+# What about with the new PCA grid?
+norm_pca_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_1, adapt = FALSE)
+norm_pca_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_3, adapt = FALSE)
+norm_pca_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_5, adapt = FALSE)
+norm_pca_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_7, adapt = FALSE)
+
+# These don't look near zero! Something is going wrong
+norm_pca_bfgs_1$lognormconst
+norm_pca_bfgs_3$lognormconst
+norm_pca_bfgs_5$lognormconst
+norm_pca_bfgs_7$lognormconst
 ```
 
-Test how this PCA grid works for some simple integrals.
-We will use Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.
+### Other 2D function
+
+More complicated two dimensional function from Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.
 
@@
 
-truemode2d <- c(log((sum(y1) + 1)/(length(y1) + 1)), log((sum(y2) + 1)/(length(y2) + 1)))
-
+truemode2d <- c(log((sum(y1) + 1) / (length(y1) + 1)), log((sum(y2) + 1) / (length(y2) + 1)))
 truelogint <- function(y) lgamma(1 + sum(y)) - (1 + sum(y)) * log(length(y) + 1) - sum(lgamma(y + 1))
@@
 
-objfunc2d <- function(x) logfteta2d(x, c(y1, y2))
+obj <- function(x) logfteta2d(x, c(y1, y2))
 
-funlist2d <- list(
+ff <- list(
   fn = objfunc2d,
-  gr = function(x) numDeriv::grad(objfunc2d,x),
-  he = function(x) numDeriv::hessian(objfunc2d,x)
+  gr = function(x) numDeriv::grad(obj, x),
+  he = function(x) numDeriv::hessian(obj, x)
 )
 
-opt_bfgs_2d <- aghq::optimize_theta(funlist2d, c(1.5, 1.5), control = default_control(method = "BFGS"))
+grid <- expand.grid(
+  theta1 = seq(0, 4, length.out = 400),
+  theta2 = seq(0, 4, length.out = 400)
+)
 
-norm_bfgs_2d_1 <- aghq::normalize_logpost(opt_bfgs_2d, 1, 1)
-norm_bfgs_2d_3 <- aghq::normalize_logpost(opt_bfgs_2d, 3, 1)
-norm_bfgs_2d_5 <- aghq::normalize_logpost(opt_bfgs_2d, 5, 1)
-norm_bfgs_2d_7 <- aghq::normalize_logpost(opt_bfgs_2d, 7, 1)
+ground_truth <- data.frame(grid) %>%
+  rowwise() %>%
+  mutate(pdf = exp(objfunc2d(x = c(theta1, theta2))))
 
-pca_grid_bfgs_2d_1 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 1)
-pca_grid_bfgs_2d_3 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 3)
-pca_grid_bfgs_2d_5 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 5)
-pca_grid_bfgs_2d_7 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 7)
+plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = multi.utils::cbpalette()[1]) +
+  coord_fixed(xlim = c(0.8, 2), ylim = c(0.8, 2), ratio = 1) +
+  labs(x = "", y = "") +
+  theme_minimal()
+
+opt_bfgs <- aghq::optimize_theta(funlist2d, c(1.5, 1.5), control = default_control(method = "BFGS"))
+
+grid_bfgs_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
+# Error in rescale.NIGrid(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian)) :
+# (in rescale) no appropriate covariance matrix C
+# mvQuad::rescale(grid_bfgs_1, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian))
+
+grid_bfgs_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 3)
+# Error in rescale.NIGrid(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian)) :
+# (in rescale) no appropriate covariance matrix C
+# mvQuad::rescale(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian))
+
+grid_bfgs_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 5)
+# Error in rescale.NIGrid(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian)) :
+# (in rescale) no appropriate covariance matrix C
+# mvQuad::rescale(grid_bfgs_5, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian))
+
+grid_bfgs_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 7)
+# Error in rescale.NIGrid(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian)) :
+# (in rescale) no appropriate covariance matrix C
+# mvQuad::rescale(grid_bfgs_7, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian))
+
+norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, 1, 1)
+norm_bfgs_3 <- aghq::normalize_logpost(opt_bfgs, 3, 1)
+norm_bfgs_5 <- aghq::normalize_logpost(opt_bfgs, 5, 1)
+norm_bfgs_7 <- aghq::normalize_logpost(opt_bfgs, 7, 1)
+
+pca_grid_bfgs_1 <- pca_rescale(m = opt_bfgs$mode, C = solve(opt_bfgs$hessian), s = 1, k = 1)
+pca_grid_bfgs_3 <- pca_rescale(m = opt_bfgs$mode, C = solve(opt_bfgs$hessian), s = 1, k = 3)
+pca_grid_bfgs_5 <- pca_rescale(m = opt_bfgs$mode, C = solve(opt_bfgs$hessian), s = 1, k = 5)
+pca_grid_bfgs_7 <- pca_rescale(m = opt_bfgs$mode, C = solve(opt_bfgs$hessian), s = 1, k = 7)
 
 plot_points <- function(gg) {
-  gg$nodes %>%
+  plot +
+    geom_point(
+      data = mvQuad::getNodes(gg) %>%
         as.data.frame() %>%
-    mutate(weights = gg$weights) %>%
-    ggplot() +
-    geom_point(
+        mutate(weights = mvQuad::getWeights(gg)),
       aes(x = V1, y = V2, size = weights),
@@
 
-{plot_points(pca_grid_bfgs_2d_1) + plot_points(pca_grid_bfgs_2d_3)} / {plot_points(pca_grid_bfgs_2d_5) + plot_points(pca_grid_bfgs_2d_7)}
+#' AGHQ
+{plot_points(grid_bfgs_1) + plot_points(grid_bfgs_3)} / {plot_points(grid_bfgs_5) + plot_points(grid_bfgs_7)}
 
-norm_pca_bfgs_2d_1 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_1, adapt = FALSE)
-norm_pca_bfgs_2d_3 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_3, adapt = FALSE)
-norm_pca_bfgs_2d_5 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_5, adapt = FALSE)
-norm_pca_bfgs_2d_7 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_7, adapt = FALSE)
+#' PCA-AGHQ
+{plot_points(pca_grid_bfgs_1) + plot_points(pca_grid_bfgs_3)} / {plot_points(pca_grid_bfgs_5) + plot_points(pca_grid_bfgs_7)}
 
-results_2d <- data.frame(
+norm_pca_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_1, adapt = FALSE)
+norm_pca_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_3, adapt = FALSE)
+norm_pca_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_5, adapt = FALSE)
+norm_pca_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_7, adapt = FALSE)
+
+results <- data.frame(
   "method" = c(
@@
     truelognormconst2d,
-    norm_bfgs_2d_1$lognormconst, norm_bfgs_2d_3$lognormconst, norm_bfgs_2d_5$lognormconst, norm_bfgs_2d_7$lognormconst,
-    norm_pca_bfgs_2d_1$lognormconst, norm_pca_bfgs_2d_3$lognormconst, norm_pca_bfgs_2d_5$lognormconst, norm_pca_bfgs_2d_7$lognormconst
+    norm_bfgs_1$lognormconst, norm_bfgs_3$lognormconst, norm_bfgs_5$lognormconst, norm_bfgs_7$lognormconst,
+    norm_pca_bfgs_1$lognormconst, norm_pca_bfgs_3$lognormconst, norm_pca_bfgs_5$lognormconst, norm_pca_bfgs_7$lognormconst
   )
@@
 
-ggplot(results_2d, aes(x = method, y = lognormconst)) +
-  geom_point(size = 3) +
+ggplot(results, aes(x = method, y = lognormconst)) +
+  geom_point(size = 1) +
   labs(x = "Method", y = "Log normalising constant") +
@@
 
-### Application to Naomi
+## Application to Naomi
 
@@
 ```
-
-```{r}
-start <- Sys.time()
-quad <- fit_aghq(tmb_inputs, basegrid = base_grid)
-end <- Sys.time()
-
-end - start
-```

### 2023-03-28 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
Iterate slides #24

@@
 * If you want to integrate something deterministically, you could use numerical integration, otherwise called \textcolor{hilit}{quadrature}
-* Select nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ then compute the sum
+* Choose nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ and compute the weighted sum
 $$
@@
 * Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule
+* Nodes evenly spaced through $[0, \pi]$
 
@@
 trapezoid_rule <- function(x, spacing) {
-  w <- rep(spacing, length(x))
-  w[1] <- w[1] / 2
-  w[length(x)] <- w[length(x)] / 2
-  sum(w * x)
+  w <- rep(spacing, length(x)) # Weights given by space between nodes
+  w[1] <- w[1] / 2 # Apart from the first which is halved
+  w[length(x)] <- w[length(x)] / 2 # And the last, also halved
+  sum(w * x) # Compute the weighted sum
 }
@@
 
-```{r message=FALSE, echo=FALSE}
+```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."}
 library(tidyverse)
@@
     labs(
-      subtitle = paste0("Number of quadrature points: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"),
+      subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"),
       x = "x", y = "sin(x)"
@@
 
-```{r message=FALSE, echo=FALSE}
+```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."}
 trapezoid_plot(N = 30)
@@
 
-```{r message=FALSE, echo=FALSE}
+```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."}
 trapezoid_plot(N = 100)
@@
 * Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$
-* If we set $\omega(\vartheta_i) = 1/N$ for all $i$ then we get a **Monte Carlo** (MC) estimate
+* The Monte Carlo (MC) estimate is
 $$
@@
 $$
-* For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain which if you squint a bit is good enough (MCMC)
+which is quadrature with nodes given by the samples and $\omega(\vartheta_i) = 1/N$ for all $i$
+* For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC)
 
@@
 
+# Curse of dimensionality
+
+* Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized, because there is an exponential increase in the volume you need to cover
+* If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$ e.g. $k = 5$ then
+
+```{r}
+5^{1:8}
+```
+
 # Latent variables and hyperparameters
 
-* Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized
-  * If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$
 * Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$
@@
 
-* There is nothing inherently special about spatio-temporal statistics, but we do often end up in similar situations because of the structures of the problems we tackle
+* There is nothing inherently special about spatio-temporal statistics, but we do often end up tackling problems with similar structures
 * We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$
-* Usually we associate parameters to spatio-temporal locations as well as observations
-* This ends up with us having something like $\{x_{s, t}\}$
+* Usually we associate parameters to spatio-temporal locations, giving us something like $\{x_{s, t}\}_{s \in \mathcal{S}, t \in \mathcal{T}}$
 
@@
 
-1.
-There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $\text{dim}(x)$ might be pretty big!
-If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
-2. Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]
-* Not IID anymore!
+1. There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $N$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
+2.
+Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]: not IID anymore!
 
@@
 * Hey wait a second, is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian?
-* No, not in general. But...
+* There is the Bernstein–von Mises theorem, but we can't rely on that in general. However...
 
@@
 We just described a class of models (LGMs) where some subset of the parameters ($x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian
-2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a **marginal Laplace** approximation to get rid of $x$
+2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a \textcolor{hilit}{marginal Laplace} approximation to get rid of $x$
 
@@
 * Now we can compute $p_{\texttt{LA}}(\theta, y)$ but what we really want is still $p(y)$
-* But hopefully^[Really: hopefully] the dimension of $\theta$ is small enough that we can now tackle this with quadrature
+* But hopefully^[Really: hopefully] $m$ is small enough that we can now tackle this with quadrature
 * So pick some nodes $\mathcal{Q}$ and a weighting function $\omega$ and away we go
@@
 
-* `TMB` [@kristensen2015tmb] is an R package which implements the Laplace approximation for latent variable models
-* To use `TMB`, you write your objective function $- \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)$ in `TMB`'s C++ syntax
-* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks as follows
+* To use `TMB` [@kristensen2015tmb], you write your objective function $- \log p(y, x, \theta)$ in the C++ syntax
+* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks like (next slide)
+* `TMB` implements the Laplace approximation! Set `random = "x"`
 
@@
 * It runs quickly, and is flexible enough to write the model
-* Another answer: we don't have any better options
-  * Markov chain Monte Carlo (MCMC) is accurate (eventually) but takes too long
-  * Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a latent Gaussian model, and isn't compatible with the `R-INLA` implementation of INLA
+* Another answer is that we don't have any better options
+
+| Option | Would it work? |
+|--------|--------------------|
+| MCMC | No! It is accurate (eventually) but takes too long |
+| INLA via `R-INLA`        | No! Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a LGM, and isn't compatible with `R-INLA` |
 
@@
 
-* Implement an algorithm similar to INLA which
-1. is compatible with Naomi
-2. uses `TMB` to perform Laplace approximation 
-* Does this improves the quality of the inferences we get?
+* Implement an algorithm inspired by INLA which
+1. is compatible with Naomi and its intricacies
+2. uses `TMB` -- and thereby automatic differentiation -- to perform the Laplace approximation
 
@@
 
-* Alex Stringer in Toronto $\to$ Waterloo had been thinking along similar lines, and made a lot of progress
+* Alex Stringer in Toronto $\to$ Waterloo had independently been thinking along similar lines, and made a lot of progress
 
-1.
-Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which he and others argue should be the default for this problem
+1. Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which arguably should be the default for this problem
 2. Defining a class of models called extended latent Gaussian models (ELGMs) which Naomi fits into
@@
 * The nodes and weights we use should probably depend on the integrand
-  * Especially when the integrand is also a function of $y$ as in $p(y, \vartheta)$
-* Let $z \in \mathcal{Q}(m, k)$ then $\theta(z) = \hat \theta + Lz$ where $L$ is the lower Cholesky of $H(\hat \theta)$
+* Especially when the integrand is also a function of $y$, which we don't know in advance, as in $p(y, \vartheta)$
+  * i.e. how could any fixed quadrature rule be appropriate for all possible $y$?
+* Let $z \in \mathcal{Q}(m, k)$ then
+$$
+\theta(z) = \hat \theta + Lz
+$$
+where $L$ is the lower Cholesky of $H = LL^\top$
+* Other matrix decompositions can also be used e.g. spectral $H = E \Lambda E^\top = (E\Lambda^{1/2})(E\Lambda^{1/2})$
+  * Arguably this is preferable: symmetric with respect to the principle axis [@jackel2005note]
 
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Rotate by the lower Cholesky."}
+```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $L$."}
 gg3 <- gg
 mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)
+
+add_points(plot0, gg3) +
+  labs(size = "Weight", caption = "")
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2}$."}
+gg3 <- gg
+mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1)
 

### 2023-03-28 src/docs_paper/paper.Rmd
Add IS-INLA citation

@@
 
-Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package of @bachl2019inlabru, and the INLA within MCMC approach of @gomez2018markov, both of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
+Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package [@bachl2019inlabru], INLA within MCMC [@gomez2018markov], and importance sampling with INLA [@berild2022importance], all of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
 `inlabru` approximates non-linear predictors using linearisation, which involves making iterative calls to `R-INLA`.
@@
 <!-- I think it'd obviously depend on the exact details of the problem at hand -->
-INLA within MCMC is suitable for models which are LGMs conditional on some subset of the parameters being fixed.
-For such models, Metropolis-Hastings style algorithms can be defined to update the fixed parameters, with acceptance probabilities calculated using by repeat calls to `R-INLA`.
+INLA within MCMC and importance sampling with INLA are suitable for models which are LGMs conditional on some subset of the parameters being fixed.
 <!-- It would or would not be possible to fit the Naomi model in these frameworks. -->

### 2023-03-28 src/docs_paper/appendix.Rmd
Correct typos, cites for AGHQ, reupload

@@
 We model HIV incidence rate using the linear predictor
-\begin{equation}
+$$
 \log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda, \label{eq:hiv-inc}
-\end{equation}
+$$
 where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women, $\rho_{x}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences
-\begin{equation}
+$$
 \rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}, \label{eq:rho-agg}
-\end{equation}
+$$
 $\alpha_{x}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages
-\begin{equation}
+$$
 \alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}, \label{eq:alpha-agg}
-\end{equation}
+$$
 $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output.
@@
 To model $\gamma_{x, x'}$ for $x \sim x'$ we use a multinomial logistic regression model, based on the log-odds ratios
-\begin{equation*}
+\begin{equation}
 \tilde \gamma_{x, x'} = \log \left( \frac{\gamma_{x, x'}}{1 - \gamma_{x, x'}} \right) = \tilde \gamma_0 + u_x^{\tilde \gamma}, \label{eq:log-or}
-\end{equation*}
+\end{equation}
 where $\tilde \gamma_0 = -4$ is a fixed intercept, and $u_x^{\tilde \gamma} \sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})$ are district random effects with $\sigma_X^{\tilde \gamma} \sim \mathcal{N}^+(0, 2.5)$.
 Note that Equation \ref{eq:log-or} does not depend on $x'$, such that $\gamma_{x, x'}$ is only a function of $x$.
-Choice of $\tilde \gamma_0 = -4$ implies a prior on $\gamma_{x, x'}$ of 1.8%, such that $(100 - 1.8 \times \text{ne}(x))\%$ of ART clients in district $x$ obtain treatment in their home district, a-priori.
+Choice of $\tilde \gamma_0 = -4$ implies a prior mean on $\gamma_{x, x'}$ of 1.8%, such that $(100 - 1.8 \times \text{ne}(x))\%$ of ART clients in district $x$ obtain treatment in their home district, a-priori.
 We fix $\tilde \gamma_{x, x} = 0$ and recover the multinomial probabilities using the softmax

### 2023-03-28 src/docs_paper/paper.Rmd
Correct typos, cites for AGHQ, reupload

@@
 
-@eaton2021naomi specify a joint model linking three small-area estimation models and defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version defined only at $T_1$ omitting nowcasting and temporal projection, as these time points involve no further inference.
+@eaton2021naomi specify a joint model linking three small-area estimation models, and defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
+We consider a simplified version defined only at $T_1$ omitting nowcasting and temporal projection, as these time points involve limited additional inference.
 An overview of this simplified model is given below, highlighting the aspects which make it a challenge for existing inferential approaches.
@@
 for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
-We model HIV incidence rate on the log scale as $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
+HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
 Finally, let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
@@
 
-As data, for $\theta \in \{\rho, \alpha, \kappa\}$ we calculate the weighted, aggregated survey observations
+For $\theta \in \{\rho, \alpha, \kappa\}$ we calculate the weighted, aggregated survey observations
 $$
@@
 \text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
-\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\rho^\text{ANC}}_i.
+\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\alpha^\text{ANC}}_i.
 \end{align*}
-These processes are informed by specifying likelihoods for the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
+These processes are informed by likelihoods specified for the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
 We use the binomial working likelihoods
 \begin{align*}
-y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}), \\
+y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(x^\text{ANC}_\mathcal{I}, \rho^\text{ANC}_{\mathcal{I}}), \\
 z^\text{ANC}_\mathcal{I} &\sim \text{Bin}(y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}),
@@
 \rho^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in \mathcal{I}} \Psi_i}, \quad
-\alpha^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}},
+\alpha^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}},
 \end{equation*}
@@
 Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
-We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$, and model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a multinomial logistic regression model.
+We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent such that $x \sim x'$. 
+We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a multinomial logistic regression model.
 
@@
 In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials.
+These nodes and weights can be shifted and scaled to suit the particular integrand using adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate].
 

### 2023-03-28 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
Details about why Naomi is an ELGM, start of posterior comparison slides #24

@@
 ---
-title: Deterministic Bayesian inference methods for the Naomi model
-subtitle: HIV Inference Lab Group Meeting
-author: Adam Howes
-institute: Imperial College London
-date: April 2023
-bibliography: citations.bib
+title: "Deterministic Bayesian inference methods for the Naomi model"
+author: "Adam Howes"
+date: "April 2023"
 output:
@@
       in_header: preamble.tex
+subtitle: HIV Inference Lab Group Meeting
+bibliography: citations.bib
+institute: Imperial College London
 ---
@@
 $$
+* By "deterministic" I mean: if you follow the same procedure you will get the same answer
 
@@
 
-# Monte Carlo as an example of numerical integration
+# Monte Carlo as it relates to numerical integration
 
@@
 $$
-which is quadrature with nodes given by the samples and $\omega(\vartheta_i) = 1/N$ for all $i$
+which is quadrature with the samples as nodes and $\omega(\vartheta_i) = 1/N$ for all $i$
 * For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC)
@@
 |2| Incidence depends on adult prevalence and coverage | ? |
-|3| ART attendance is a product | ? |
+|3| ART coverage and recent infection are products | ? |
 |4| ART attendance uses a multinomial | ? |
@@
 
+* Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators
 \begin{align*}
@@
 \end{align*}
-
-* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ are Gaussian
-* They have their own structured additive predictors, but by use of the `copy` feature this should be possible in `R-INLA`
-* Conclusion: \textcolor{hilit}{does not} make Naomi an ELGM
+* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian
+* `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ but $A$ must be $n \times n$
+* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
 
@@
 
+* Linear predictor for incidence contains aggregated prevalence and coverage
 $$
@@
 $$
-
-* Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian -- no matter your `copy` tricks
+* Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian
 * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
 
-# ART attendance is a product
+# ART coverage and recent infection are products
+
+* In the household survey, say, individuals who are taking ART or have been recently infected must be HIV positive
+\begin{align*}
+y^{\hat \alpha}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \alpha_{x, s, a}), \\
+y^{\hat \kappa}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \kappa_{x, s, a}).
+\end{align*}
+* $\text{logit}(\rho_{x, s, a})$ and $\text{logit}(\alpha_{x, s, a})$ are Gaussian, but we're taking a product here
+* $\kappa_{x, s, a}$ is more complicated: a function of incidence, prevalence, mean duration of recent infection and false recent ratio
+* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+Warning! These equations as writen do not appear in Naomi: instead there are aggregated versions, more about this soon.
+\end{tcolorbox}
+\end{center}
 
@@
 # Aggregation of finer processes
+
+* There are many instances of models being placed on aggregate quantities
+\begin{align*}
+y^{\hat \theta}_{\mathcal{I}} &\sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}), \\
+\rho_{\mathcal{I}} &= \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}.
+\end{align*}
+* Here we have $|\mathcal{I}|$ linear predictors being informed by one observation
+* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+I've previously worked on this "disaggregation regression" situation with respect to space, see \texttt{athowes/areal-comparison}
+\end{tcolorbox}
+\end{center}
 
@@
 
-```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the latent field.", fig.ext = 'png', cache=TRUE}
+```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the latent field: naive Laplace version.", fig.ext = 'png', cache=TRUE}
 \usetikzlibrary{arrows}
@@
 
+# Comparing posterior distributions
+
+* Let $\{\theta_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function
+$$
+F(\vartheta) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\theta_i \leq \vartheta}
+$$
+* Could be a hyperparameter, latent field parameter, model outputs
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+Our primary interest is in the model outputs: do any of the inference methods do a better job at computing the posterior for quantities countries use?
+\end{tcolorbox}
+\end{center}
+
+<!-- # Moments -->
+
+<!-- * Are the means and standard deviations the same? More generally the $t$th moment -->
+<!-- $$ -->
+<!-- \hat \mathbb{E}(\theta^t) = \frac{1}{n} \sum_{i = 1}^n \theta_i^t -->
+<!-- $$ -->
+<!-- * Won't do a great job if you're only looking at the first few moments $t = 1, 2$ -->
+
+<!-- # Kolmogorov-Smirov test -->
+
+<!-- * Compare $F_\bullet(\vartheta)$ to $F_\texttt{NUTS}(\vartheta)$ -->
+<!-- $$ -->
+<!-- D_\bullet = \sup_\vartheta | F_\texttt{NUTS}(\vartheta) - F_\bullet(\vartheta)|. -->
+<!-- $$ -->
+<!-- * A value $D_\bullet$ means there is at most a $100 \cdot D_\bullet$% difference between posterior densities -->
+
+<!-- # -->
+
+<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="KS test example."} -->
+<!-- plot(1:10) -->
+<!-- ``` -->
+
+# Maximum mean discrepancy
+
+# Pareto smoothed importance sampling
+
 # Thanks for listening!

### 2023-03-29 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
More on ELGM, clarify use of R-INLA copy, multiple link functions, moments and KS #24

@@
 
-* To explain what an extended latent Gaussian model (ELGM) is, we need a little more detail on LGMs
-* In an LGM the conditional mean depends on exactly one structured additive predictor, specifically
+* Before defining extended latent Gaussian model (ELGM), first more detail on LGMs:
+* Conditional mean depends on exactly one structured additive predictor
 \begin{align*}
@@
 \end{align*}
+where $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors (and can be collected into the latent field $x$)
 
@@
 $$
-\mu_i = g(\eta_{\mathcal{J}_i})
+\mu_i = g_i(\eta_{\mathcal{J}_i})
 $$
-where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J_i}$ is some set of indices
-* This allow for a higher degree of non-linearity in the model
+where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices
+* Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$
+  * $N_n < n$: more data points than structured additive predictors
+  * $N_n = n$: as many data points as structured additive predictors (LGM case)
+  * $N_n > n$: fewer data points than structured additive predictors
+* The $g_i$ allow for a higher degree of non-linearity in the model
 
@@
 
-| | Things I thought made Naomi an ELGM | Does it actually? |
+| | Things I thought made Naomi an ELGM | Does it? |
 |-|---------------|--------------------|
@@
 |5| Aggregation of finer processes | ? |
+|6| Multiple link functions | ? |
 
-* I will explain each of these, and whether or not they make Naomi an ELGM, in more detail in slides to follow
-  * The notation may not have been introduced properly, but hopefully the gist will still make sense
+* I will explain each of these^[The notation may not have been introduced properly, but hopefully the gist will still make sense], and whether or not they make Naomi an ELGM, in more detail in slides to follow
 
@@
 \end{align*}
-* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian
-* `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ but $A$ must be $n \times n$
+* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$
 * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
+
+# ANC offset from household survey
+
+* Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$, so I'm unsure which is right]
+$$
+\begin{pmatrix}
+\eta^\star \\
+\eta^\star
+\end{pmatrix}
+=
+\begin{pmatrix}
+1 & 1 \\
+0 & 1
+\end{pmatrix}
+\begin{pmatrix}
+\eta_1 \\
+\eta_2
+\end{pmatrix}
+=
+\begin{pmatrix}
+\eta_1 + \eta_2 \\
+\eta_2
+\end{pmatrix}
+$$
+* If it's possible to have effects only apply to a subset of indices (?) then perhaps it could work
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+
+Warning!
+
+\begin{enumerate}
+    \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA}
+    \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM
+\end{enumerate}
+
+\end{tcolorbox}
+\end{center}
 
@@
 I've previously worked on this "disaggregation regression" situation with respect to space, see \texttt{athowes/areal-comparison}
+\end{tcolorbox}
+\end{center}
+
+# Multiple link functions
+
+* The Naomi model uses both $\text{logit}$ and $\log$ (inverse) link functions
+* For LGMs there is only one $g$, whereas ELGMs allow $g_i$
+* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
+* In `R-INLA` it is possible for `y` to be a matrix where each column contains observations with shared likelihood family (and hyperparameters) and `family = c("family1", "family2", ...)`
+
+# Statistical software
+
+|           | Interface `y ~ 1 + x`  | General |
+|-----------|------------|---------|
+| Example   | `brms`     | Stan    |
+| Users     | Scientists | Statisticians |
+| Benefits  | Ease of use, sensible defaults, trust | Flexibility, development |
+| Drawbacks | Being fenced in | Barrier to entry, failing silently |
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+\texttt{athowes/multi-agyw} case-study: attempting to define $\text{AR}(1) \times \text{Besag} \times \text{IID}$ random effects in \texttt{R-INLA}
 \end{tcolorbox}
@@
 
-<!-- # Moments -->
+# Moments
 
-<!-- * Are the means and standard deviations the same? More generally the $t$th moment -->
-<!-- $$ -->
-<!-- \hat \mathbb{E}(\theta^t) = \frac{1}{n} \sum_{i = 1}^n \theta_i^t -->
-<!-- $$ -->
-<!-- * Won't do a great job if you're only looking at the first few moments $t = 1, 2$ -->
+* Are the means and standard deviations the same? More generally the $t$th moment can be estimated by
+$$
+\hat{\mathbb{E}}(\theta^t) = \frac{1}{n} \sum_{i = 1}^n \theta_i^t
+$$
+* Not great if you're only looking at the first few moments $t = 1, 2$
 
-<!-- # Kolmogorov-Smirov test -->
+# Kolmogorov-Smirov test
 
-<!-- * Compare $F_\bullet(\vartheta)$ to $F_\texttt{NUTS}(\vartheta)$ -->
-<!-- $$ -->
-<!-- D_\bullet = \sup_\vartheta | F_\texttt{NUTS}(\vartheta) - F_\bullet(\vartheta)|. -->
-<!-- $$ -->
-<!-- * A value $D_\bullet$ means there is at most a $100 \cdot D_\bullet$% difference between posterior densities -->
+* Compare $F_\bullet(\vartheta)$ to $F_{\texttt{NUTS}}(\vartheta)$
+$$
+D_\bullet = \sup_\vartheta | F_{\texttt{NUTS}}(\vartheta) - F_\bullet(\vartheta)|.
+$$
+* A value $D_\bullet$ means there is at most a $(100 \cdot D_\bullet)$% difference between posterior densities
 
-<!-- # -->
+#
 
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="KS test example."} -->
-<!-- plot(1:10) -->
-<!-- ``` -->
-
-# Maximum mean discrepancy
-
-# Pareto smoothed importance sampling
+```{r, echo=FALSE, message=FALSE, fig.cap="KS test example."}
+plot(1:10)
+```
 
@@
 
-* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work, joint with Alex Stringer (Waterloo) and my PhD supervisors Seth Flaxman (Oxford) and Jeff Eaton (Imperial)
+* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work,     * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
 * Let me know if you'd be up for being an early reader!

### 2023-03-29 src/docs_paper/paper.Rmd
Add to-do list for paper

@@
 ---
+
+<!-- Remaining things to add, and or desiderata: -->
+
+<!-- * [ ] Description of the ART attendance component of the model -->
+<!-- * [ ] Tying together Naomi model within LGM / ELGM mathematical notation -->
+<!-- * [ ] Create DAG representing Naomi model -->
+<!-- * [ ] Rewrite section on why Naomi is an ELGM, drawing from slides -->
+<!-- * [ ] Write section on Laplace approximation -->
+<!-- * [ ] Write section on AGHQ, tying together connection to Laplace for k = 1 -->
+<!-- * [ ] Move algorithm to algorithm environment -->
 
@@
 \end{align}
-Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the above posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\tilde p(\theta_j \, | \, \y)_{j = 1}^m$ via nested applications of the Laplace approximation and AGHQ, described below.
+Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the above posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$ via nested applications of the Laplace approximation and AGHQ.
 
@@
 
-```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension are adapted based on the mean and covariance matrix of the target via $\\btheta(\\z) = \\hat \\btheta + \\mathbf{L} \\z$.."}
+```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension are adapted based on the mean and covariance matrix of the target via $\\btheta(\\z) = \\hat \\btheta + \\mathbf{L} \\z$."}
 mu <- c(1, 1.5)

### 2023-03-29 src/docs_18-04-23-explainer/18-04-23-explainer.Rmd
Correct date (ICL closed 11th) and edit slides a little post chat with Jeff #24

--- 2023-03-29 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
@@
   stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) +
+  annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1)", col = cbpalette[1], size = 5) +
   labs(x = "", y = "") +
@@
   geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) +
-  stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2])
+  stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) +
+  annotate("text", x = 5, y = 0.25, label = "Gamma(9, 4)", col = cbpalette[2], size = 5)
 
@@
 
-```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation matches the true posterior near the mode but it's not great in the tails."}
-laplace
+```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation is good near the mode but not in the tails."}
+laplace +
+  annotate("text", x = 5.5, y = 0.35, label = "Laplace approximation", col = cbpalette[3], size = 5)
 ```
@@
 
-* Hey wait a second, is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian?
+* Is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian?
 * There is the Bernstein–von Mises theorem, but we can't rely on that in general. However...
@@
 
-* Now we can compute $p_{\texttt{LA}}(\theta, y)$ but what we really want is still $p(y)$
-* But hopefully^[Really: hopefully] $m$ is small enough that we can now tackle this with quadrature
-* So pick some nodes $\mathcal{Q}$ and a weighting function $\omega$ and away we go
+* Now we can compute $p_{\texttt{LA}}(\theta, y)$ -- but what we really want is still $p(y)$ for which we need to integrate out $\theta$
+* Hopefully^[Really: hopefully] $m$ is small enough that we can now tackle this with quadrature
+* Given nodes $\mathcal{Q}$ and a weighting function $\omega$ we can compute
 $$
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean."}
+```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean $z + \\hat \\theta$."}
 gg2 <- gg
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $L$."}
+```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $Lz + \\hat \\theta$."}
 gg3 <- gg
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2}$."}
+```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2} z + \\hat \\theta$."}
 gg3 <- gg
@@
 
-* Before defining extended latent Gaussian model (ELGM), first more detail on LGMs:
-* Conditional mean depends on exactly one structured additive predictor
+* In an LGM, the conditional mean depends on exactly one structured additive predictor
 \begin{align*}
@@
 \begin{pmatrix}
-\eta^\star \\
-\eta^\star
+\eta_1^\star \\
+\eta_2^\star
 \end{pmatrix}
@@
 $$
-* If it's possible to have effects only apply to a subset of indices (?) then perhaps it could work
+* By having effects only apply to a subset of indices (`idx = c(NA, 1)` say) perhaps it can work
 
@@
 
-* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work,     * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
+* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work
+  * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
 * Let me know if you'd be up for being an early reader!

### 2023-03-29 utils/archive_to_docs.R

@@ -20,7 +20,7 @@ archive_to_docs("docs_paper")
 archive_to_docs("docs_01-04-20-mini")
 archive_to_docs("docs_01-07-21-stats-epi-group")
 archive_to_docs("docs_15-11-22-seminar")
-archive_to_docs("docs_11-04-23-explainer")
+archive_to_docs("docs_18-04-23-explainer")
 archive_to_docs("epil")
 archive_to_docs("prev-anc-art_results")
 archive_to_docs("example_inla-replication")

### 2023-03-30 src/dev_scale-grid/scale-grid.Rmd
Clean up scale grid notebook in preparation for fixing weights issue

--- 2023-03-22 src/dev_scale-grid/scale-grid.Rmd
@@
 
-  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure.
+  **Task** We would like to begin to integrate over the hyperparameters using multiple nodes.
 The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical.
 In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using principle components analysis.
+editor_options: 
+  markdown: 
+    wrap: sentence
 ---
@@
 
-# Grid ideas
+# Point concentration proportional to standard deviation
 
@@
 There are `r length(hypers)` of them, if you'd like to count.
+Including `r data.frame(hypers) %>% filter(str_detect(hypers, "^logit_")) %>% nrow()` on the logit scale, and `r data.frame(hypers) %>% filter(str_detect(hypers, "^log_")) %>% nrow()` on the log scale.
 
-We can calculate their posterior standard deviations using `TMB::sdreport` which uses the delta method as follows:
+We can calculate their posterior standard deviations using the delta method via `TMB::sdreport` as follows:
 
@@
 
-The marginal standard deviations are:
+The marginal standard deviations for each of the hyperparameters are:
 
-```{r}
+```{r sd, fig.cap="Marginal standard deviations."}
 sd <- sqrt(diag(sd_out$cov.fixed))
@@
 
-```{r}
+```{r cov, fig.cap="Covariance matrix."}
 C <- sd_out$cov.fixed
@@
 
-## Number of points proportional to standard deviation
-
-A first suggestion about how to place grid points would be proportional to the standard deviations above.
-For example, we could get a product grid with the minimum number of points by choosing to have:
-
-* `k = 2` points in the dimension with the highest SD
-* `k = 1` point in every other dimension
-
-This would look like the following:
+One way to place grid points would be proportional to the standard deviations shown in the plot above.
+For example, we could a product grid with the minimum number of points (2) could be obtained by choosing `k = 2` points in the dimension with the highest SD and `k = 1` point in every other dimension.
+This would look as follows:
 
@@
 
-The total number of points for this grid is 2, by design:
+The total number of points for this grid is 2:
 
@@
 
-Now fit the model with `aghq`:
+Now let's fit the model with `aghq`, and time how long it takes:
 
@@
 
-A limitation of this approach is that some hyperparameters have higher SD than others in part due to the different scales that they are on.
-For example if the hyper is on the logit scale then (from the plot above) it looks as though it will have a higher SD than another hyper on the log scale.
-We would ideally like to know how much variation in each hyperparameter effects variation in model outputs, and then take that into account in the importance we place on each dimension: doing this in any good automated ways would be a nice contribution.
+This model fit in a very reasonable amount of time.
+More generally, it would be of interest to know more about the relationship between the number of points in the grid, and the length of time it takes run `fit_aghq`.
+Are there any theoretical considerations statements that can be made e.g. scaling $\mathcal{O}(n)$ where $n$ is the number of grid points?
 
-## Eigendecomposition
+## Limitations
 
-Another option is to use principle components analysis (PCA) to create a grid on a subspace of $\mathbb{R}^{24}$ which retains most of the variation.
+Some hyperparameters have higher standard deviation than others in part due to the different scales that they are on.
+For example if the hyperparameter is on the logit scale then it looks as though it will have a higher standard deviation than another those on the log scale (Figure \@ref(fig:sd)).
+The average standard deviation for hyperparameters starting with `"log_"` is `r data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^log_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd)` as compared with `r data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^logit_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd)` for those starting with `"logit_"`.
+
+It is not clear that this reflects our intuitions regarding which hyperparameters it is important to take greater account of.
+Perhaps ideally we would like to know how much variation in each hyperparameter effects variation in model outputs, and then take that into account in the importance we place on each dimension.
+
+## Points of confusion
+
+How does this approach interact with the rescaling which occurs in adaptive quadrature?
+That is to say $z \mapsto \hat \theta + Lz$ with $LL^\top = H$.
+Do each of the dimensions of $z$ reflect dimensions of $Lz$?
+
+# Eigendecomposition
+
+Another option to using the standard deviations is to use principle components analysis (PCA) to create a grid on a subspace of $\mathbb{R}^{24}$ which retains most of the variation.
 
@@
 
-The relative contributions of each dimension are as follows (this is a Scree plot):
+The relative contributions of each principle component are:
 
-```{r}
+```{r scree, fig.cap="Scree plot."}
 tv_df <- data.frame(
@@
 
-So with 5 dimensions included, we can explain `r round(100 * filter(tv_df, n == 5) %>% pull(tv), 1)`% of the total variation.
+Based on Figure \@ref(fig:scree) with 5 dimensions included, we can explain `r round(100 * filter(tv_df, n == 5) %>% pull(tv), 1)`% of the total variation.
 Or, with 10 dimensions included, that percentage increases to `r round(100 * filter(tv_df, n == 10) %>% pull(tv), 1)`%.
 
-What do the PC loadings look like?
+What do the PC loadings (columns of the matrix `E`) look like?
 
-```{r}
+```{r pc-loadings, fig.cap="PC loadings."}
 reshape2::melt(as.matrix(E)) %>%
@@
 
-Start with `s = 5` dimensions kept of the PCA.
-Then create a dense AGHQ grid with `k = 2` on 5 dimensions.
+## Keeping a smaller number of dimensions
+
+Let's start by keeping `s = 5` dimensions.
+We can create a dense GHQ grid with `k = 2` on 5 dimensions.
 
@@
 
-There are $2^5 = 32$ nodes as follows:
+This grid has $2^5 = 32$ nodes:
 
@@
 
-```{r}
+```{r cov-reconstruction, fig.cap="PC loadings."}
 E_s <- E[, 1:s]
@@
 C_s <- E_s %*% Lambda_s %*% t(E_s)
-
-C_plot
 
@@
 
-### Testing with simple examples
+Adaptive Gauss-Hermite quadrature involves
 
-Check that this works as visually expected for a simple example:
+## Testing with simple examples
 
-```{r}
+Here we run some unit tests to make sure, visually and numerically, that the methods are working as expected.
+
+### Gaussian 2D function
+
+A simple 2D Gaussian example:
+
+$$
+\begin{pmatrix}
+x_1 \\
+x_2
+\end{pmatrix}
+\sim
+\mathcal{N}
+\left(
+\begin{pmatrix}
+1 \\
+3/2
+\end{pmatrix},
+\begin{pmatrix}
+2 & 1 \\
+1 & 1
+\end{pmatrix}
+\right)
+$$
+
+```{r gaussian-2d, class.source = 'fold-hide', fig.cap="Gaussian 2D objective function."}
 mu <- c(1, 1.5)
@@
 
-gaussian_2d <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
+plot
+```
 
+Define four AGHQ-PCA grids, with `k` running from 1 to 7:
+
+```{r}
+pca_grid_1 <- pca_rescale(m = mu, C = cov, s = 1, k = 1)
+pca_grid_3 <- pca_rescale(m = mu, C = cov, s = 1, k = 3)
+pca_grid_5 <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
+pca_grid_7 <- pca_rescale(m = mu, C = cov, s = 1, k = 7)
+```
+
+For example, the version with 7 points looks as follows:
+
+```{r pca-grid-7, fig.cap="AGHQ-PCA grid for Gaussian with $k = 7$."}
+plot_points <- function(gg) {
   plot +
     geom_point(
-    data = data.frame(mvQuad::getNodes(gaussian_2d)) %>%
-      mutate(weight = mvQuad::getWeights(gaussian_2d)),
-    aes(x = X1, y = X2, size = weight),
+      data = mvQuad::getNodes(gg) %>%
+        as.data.frame() %>%
+        mutate(weights = mvQuad::getWeights(gg)),
+      aes(x = V1, y = V2, size = weights),
+      alpha = 0.8,
       col = multi.utils::cbpalette()[2],
@@
     ) +
-  scale_size_continuous(range = c(1, 2))
+    scale_size_continuous(range = c(1, 2)) +
+    labs(x = "", y = "", size = "Weight") +
+    theme_minimal()
+}
+
+plot_points(pca_grid_7)
 ```
 
-Test how this PCA grid works for some simple integrals.
-We will use Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.
+We can optimise the objective function using the BFGS algorithm to recover the mode and Hessian at the mode.
+Of course, for this example, these are known in advance as we are dealing with a Gaussian, but in general this is how they will be computed.
 
 ```{r}
-logfteta2d <- function(eta, y) {
+opt_bfgs <- aghq::optimize_theta(ff, mu, control = default_control(method = "BFGS"))
+opt_bfgs$mode
+opt_bfgs$hessian
+```
+
+Now we can perform regular AGHQ with a product grid (again with a range of values for `k`):
+
+```{r}
+norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, k = 1)
+norm_bfgs_3 <- aghq::normalize_logpost(opt_bfgs, k = 3)
+norm_bfgs_5 <- aghq::normalize_logpost(opt_bfgs, k = 5)
+norm_bfgs_7 <- aghq::normalize_logpost(opt_bfgs, k = 7)
+norm_bfgs <- list(norm_bfgs_1, norm_bfgs_3, norm_bfgs_5, norm_bfgs_7)
+```
+
+As well as the the AGHQ-PCA grids:
+
+```{r}
+norm_pca_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_1, adapt = FALSE)
+norm_pca_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_3, adapt = FALSE)
+norm_pca_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_5, adapt = FALSE)
+norm_pca_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_7, adapt = FALSE)
+norm_pca_bfgs <- list(norm_pca_bfgs_1, norm_pca_bfgs_3, norm_pca_bfgs_5, norm_pca_bfgs_7)
+```
+
+The density is already normalised, so it has a normalising constant of 1, and a log normalising constant of $\log(1) = 0$:
+
+```{r fig.cap="Unit test for the Gaussian example"}
+truelognormconst <- 0
+
+results <- data.frame(
+  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7))),
+  "lognormconst" = c(truelognormconst, sapply(norm_bfgs, function(x) x$lognormconst), sapply(norm_pca_bfgs, function(x) x$lognormconst)),
+  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4)))
+)
+
+ggplot(results, aes(x = method, y = lognormconst, col = type)) +
+  geom_point(size = 1) +
+  labs(x = "Method", y = "Log normalising constant", col = "") +
+  scale_color_manual(values = multi.utils::cbpalette()) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 70, vjust = 0.5, hjust = 0.5))
+```
+
+### Non-Gaussian 2D function
+
+A more complicated two dimensional function from Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) for the `aghq` package.
+
+```{r non-gaussian, class.source = 'fold-hide', fig.cap="Non-Gaussian 2D objective function."}
+set.seed(84343124)
+
+logfteta <- function(eta, y) {
   n <- length(y)
@@
 
-set.seed(84343124)
 n1 <- 5
@@
 
-truemode2d <- c(log((sum(y1) + 1)/(length(y1) + 1)), log((sum(y2) + 1)/(length(y2) + 1)))
+truemode <- c(log((sum(y1) + 1) / (length(y1) + 1)), log((sum(y2) + 1) / (length(y2) + 1)))
+truelogint <- function(y) lgamma(1 + sum(y)) - (1 + sum(y)) * log(length(y) + 1) - sum(lgamma(y + 1))
+truelognormconst <- truelogint(y1) + truelogint(y2)
 
-truelogint <- function(y) lgamma(1 + sum(y)) - (1 + sum(y)) * log(length(y) + 1) - sum(lgamma(y + 1))
-truelognormconst2d <- truelogint(y1) + truelogint(y2)
+obj <- function(x) logfteta(x, c(y1, y2))
 
-objfunc2d <- function(x) logfteta2d(x, c(y1, y2))
-
-funlist2d <- list(
-  fn = objfunc2d,
-  gr = function(x) numDeriv::grad(objfunc2d,x),
-  he = function(x) numDeriv::hessian(objfunc2d,x)
+ff <- list(
+  fn = obj,
+  gr = function(x) numDeriv::grad(obj, x),
+  he = function(x) numDeriv::hessian(obj, x)
 )
 
-opt_bfgs_2d <- aghq::optimize_theta(funlist2d, c(1.5, 1.5), control = default_control(method = "BFGS"))
-
-norm_bfgs_2d_1 <- aghq::normalize_logpost(opt_bfgs_2d, 1, 1)
-norm_bfgs_2d_3 <- aghq::normalize_logpost(opt_bfgs_2d, 3, 1)
-norm_bfgs_2d_5 <- aghq::normalize_logpost(opt_bfgs_2d, 5, 1)
-norm_bfgs_2d_7 <- aghq::normalize_logpost(opt_bfgs_2d, 7, 1)
-
-pca_grid_bfgs_2d_1 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 1)
-pca_grid_bfgs_2d_3 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 3)
-pca_grid_bfgs_2d_5 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 5)
-pca_grid_bfgs_2d_7 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 7)
-
-plot_points <- function(gg) {
-  gg$nodes %>%
-    as.data.frame() %>%
-    mutate(weights = gg$weights) %>%
-    ggplot() +
-    geom_point(
-      aes(x = V1, y = V2, size = weights),
-      alpha = 0.8,
-      col = multi.utils::cbpalette()[2],
-      inherit.aes = FALSE
-    ) +
-    scale_size_continuous(range = c(1, 2)) +
-    labs(x = "", y = "", size = "Weight") +
-    theme_minimal()
-}
-
-{plot_points(pca_grid_bfgs_2d_1) + plot_points(pca_grid_bfgs_2d_3)} / {plot_points(pca_grid_bfgs_2d_5) + plot_points(pca_grid_bfgs_2d_7)}
-
-norm_pca_bfgs_2d_1 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_1, adapt = FALSE)
-norm_pca_bfgs_2d_3 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_3, adapt = FALSE)
-norm_pca_bfgs_2d_5 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_5, adapt = FALSE)
-norm_pca_bfgs_2d_7 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_7, adapt = FALSE)
-
-results_2d <- data.frame(
-  "method" = c(
-    "Truth",
-    paste0("aghq, k = ", c(1, 3, 5, 7)),
-    paste0("aghq-pca, k = ", c(1, 3, 5, 7))
-  ),
-  "lognormconst" = c(
-    truelognormconst2d,
-    norm_bfgs_2d_1$lognormconst, norm_bfgs_2d_3$lognormconst, norm_bfgs_2d_5$lognormconst, norm_bfgs_2d_7$lognormconst,
-    norm_pca_bfgs_2d_1$lognormconst, norm_pca_bfgs_2d_3$lognormconst, norm_pca_bfgs_2d_5$lognormconst, norm_pca_bfgs_2d_7$lognormconst
-  )
+grid <- expand.grid(
+  theta1 = seq(0, 4, length.out = 400),
+  theta2 = seq(0, 4, length.out = 400)
 )
 
-ggplot(results_2d, aes(x = method, y = lognormconst)) +
-  geom_point(size = 3) +
-  labs(x = "Method", y = "Log normalising constant") +
+ground_truth <- data.frame(grid) %>%
+  rowwise() %>%
+  mutate(pdf = exp(obj(x = c(theta1, theta2))))
+
+plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = multi.utils::cbpalette()[1]) +
+  coord_fixed(xlim = c(0.8, 2), ylim = c(0.8, 2), ratio = 1) +
+  labs(x = "", y = "") +
+  theme_minimal()
+
+plot
+```
+
+Optimise using BFSG:
+
+```{r}
+opt_bfgs <- aghq::optimize_theta(ff, c(1.5, 1.5), control = default_control(method = "BFGS"))
+(m <- opt_bfgs$mode)
+(C <- Matrix::forceSymmetric(solve(opt_bfgs$hessian)))
+eigen(C)
+```
+
+```{r}
+grid_bfgs_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
+mvQuad::rescale(grid_bfgs_1, m = m, C = C)
+
+grid_bfgs_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 3)
+mvQuad::rescale(grid_bfgs_3, m = m, C = C)
+
+grid_bfgs_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 5)
+mvQuad::rescale(grid_bfgs_5, m = m, C = C)
+
+grid_bfgs_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 7)
+mvQuad::rescale(grid_bfgs_7, m = m, C = C)
+
+norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, 1, 1)
+norm_bfgs_3 <- aghq::normalize_logpost(opt_bfgs, 3, 1)
+norm_bfgs_5 <- aghq::normalize_logpost(opt_bfgs, 5, 1)
+norm_bfgs_7 <- aghq::normalize_logpost(opt_bfgs, 7, 1)
+norm_bfgs <- list(norm_bfgs_1, norm_bfgs_3, norm_bfgs_5, norm_bfgs_7)
+
+pca_grid_bfgs_1 <- pca_rescale(m = m, C = C, s = 1, k = 1)
+pca_grid_bfgs_3 <- pca_rescale(m = m, C = C, s = 1, k = 3)
+pca_grid_bfgs_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
+pca_grid_bfgs_7 <- pca_rescale(m = m, C = C, s = 1, k = 7)
+
+#' AGHQ
+plot_points(grid_bfgs_7)
+
+#' PCA-AGHQ
+#' Something to fix here: points not being scaled to the right place
+plot_points(pca_grid_bfgs_7)
+
+norm_pca_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_1, adapt = FALSE)
+norm_pca_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_3, adapt = FALSE)
+norm_pca_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_5, adapt = FALSE)
+norm_pca_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_7, adapt = FALSE)
+norm_pca_bfgs <- list(norm_pca_bfgs_1, norm_pca_bfgs_3, norm_pca_bfgs_5, norm_pca_bfgs_7)
+
+results <- data.frame(
+  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7))),
+  "lognormconst" = c(truelognormconst, sapply(norm_bfgs, function(x) x$lognormconst), sapply(norm_pca_bfgs, function(x) x$lognormconst)),
+  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4)))
+)
+
+ggplot(results, aes(x = method, y = lognormconst, col = type)) +
+  geom_point(size = 1) +
+  labs(x = "Method", y = "Log normalising constant", col = "") +
+  scale_color_manual(values = multi.utils::cbpalette()) +
   theme_minimal() +
@@
 
-### Application to Naomi
+## Application to Naomi
+
+Let's apply this PCA approach to the grids for the Naomi model:
 
 ```{r}
+m <- sd_out$par.fixed
+C <- sd_out$cov.fixed
+
 gg3 <- pca_rescale(m = m, C = C, s = 3, k = 2)

### 2023-03-31 src/dev_scale-grid/scale-grid.Rmd
Use mvQuad to generate PCA grids #22

@@
 
-```{r}
+```{r, class.source = 'fold-hide'}
 compile("naomi_simple.cpp")
@@
 
-```{r sd, fig.cap="Marginal standard deviations."}
+```{r sd, class.source = 'fold-hide', fig.cap="Marginal standard deviations."}
 sd <- sqrt(diag(sd_out$cov.fixed))
@@
 
-```{r cov, fig.cap="Covariance matrix."}
+```{r cov, class.source = 'fold-hide', fig.cap="Covariance matrix."}
 C <- sd_out$cov.fixed
@@
 
-```{r results="hide", message=FALSE}
+```{r, class.source = 'fold-hide', results="hide", message=FALSE}
 area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
@@
 For example if the hyperparameter is on the logit scale then it looks as though it will have a higher standard deviation than another those on the log scale (Figure \@ref(fig:sd)).
-The average standard deviation for hyperparameters starting with `"log_"` is `r data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^log_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd)` as compared with `r data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^logit_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd)` for those starting with `"logit_"`.
+The average standard deviation for hyperparameters starting with `"log_"` is `r round(data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^log_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd), 3)` as compared with `r round(data.frame(par = names(sd), sd = unname(sd)) %>% filter(str_detect(par, "^logit_")) %>% summarise(mean_sd = mean(sd)) %>% pull(mean_sd), 3)` for those starting with `"logit_"`.
 
@@
 
-The relative contributions of each principle component are:
+The relative contributions of each principle component are given by $\lambda_i / \sum_i \lambda_i$:
 
-```{r scree, fig.cap="Scree plot."}
+```{r scree, class.source = 'fold-hide', fig.cap="Scree plot."}
 tv_df <- data.frame(
@@
 
-```{r pc-loadings, fig.cap="PC loadings."}
+```{r pc-loadings, class.source = 'fold-hide', fig.cap="PC loadings."}
 reshape2::melt(as.matrix(E)) %>%
@@
 
-```{r cov-reconstruction, fig.cap="PC loadings."}
+```{r cov-reconstruction, class.source = 'fold-hide', fig.cap="PC loadings."}
 E_s <- E[, 1:s]
@@
 
-Let $a \in \mathbb{R}^d$ and $b \in \mathbb{R}^s$.
-The PCA provides a mapping $f: \mathbb{R}^d \to \mathbb{R}^s$ which compresses each $a$ to a smaller $b$ given by $f(a) = (E_s \Lambda_s^{1/2})^\top a = b$.
-Then $f^{-1}$ is given by $f^{-1}(b) = E_s \Lambda_s^{-1/2} b = a$, using the fact that $E_s^\top = E_s^{-1}$.
-So we need to multiply our nodes by $E_s \Lambda_s^{- 1/2}$ as follows:
-
-```{r}
-a_nodes <- mvQuad::getNodes(gg_s)
-b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))
-
-dim(a_nodes)
-dim(b_nodes)
-```
-
-Put this into a function, which also adjusts according to the mean and reweights the nodes:
+Create a function to do the PCA rescaling, which also adapts according to the mean and reweights the nodes:
 
@@
   gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = k) 
-  nodes_out <- t(E_s %*% diag(lambda[1:s]^{-0.5}, ncol = s) %*% t(mvQuad::getNodes(gg_s)))
+  nodes_out <- t(E_s %*% diag(lambda[1:s]^{0.5}, ncol = s) %*% t(mvQuad::getNodes(gg_s)))
   for(j in 1:d) nodes_out[, j] <- nodes_out[, j] + m[j]
@@
 ```{r gaussian-2d, class.source = 'fold-hide', fig.cap="Gaussian 2D objective function."}
-mu <- c(1, 1.5)
-cov <- matrix(c(2, 1, 1, 1), ncol = 2)
+m <- c(1, 1.5)
+C <- matrix(c(2, 1, 1, 1), ncol = 2)
 
 obj <- function(theta) {
-  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov, log = TRUE)
+  mvtnorm::dmvnorm(theta, mean = m, sigma = C, log = TRUE)
 }
@@
 ```{r}
-pca_grid_1 <- pca_rescale(m = mu, C = cov, s = 1, k = 1)
-pca_grid_3 <- pca_rescale(m = mu, C = cov, s = 1, k = 3)
-pca_grid_5 <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
-pca_grid_7 <- pca_rescale(m = mu, C = cov, s = 1, k = 7)
+pca_grid_1 <- pca_rescale(m = m, C = C, s = 1, k = 1)
+pca_grid_3 <- pca_rescale(m = m, C = C, s = 1, k = 3)
+pca_grid_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
+pca_grid_7 <- pca_rescale(m = m, C = C, s = 1, k = 7)
 ```
 
-For example, the version with 7 points looks as follows:
+For example, the version with 3 points looks as follows:
 
-```{r pca-grid-7, fig.cap="AGHQ-PCA grid for Gaussian with $k = 7$."}
+```{r pca-grid-3, class.source = 'fold-hide', fig.cap="AGHQ-PCA grid for Gaussian with $k = 3$."}
 plot_points <- function(gg) {
@@
 
-plot_points(pca_grid_7)
+plot_points(pca_grid_3)
 ```
@@
 ```{r}
-opt_bfgs <- aghq::optimize_theta(ff, mu, control = default_control(method = "BFGS"))
+opt_bfgs <- aghq::optimize_theta(ff, m, control = default_control(method = "BFGS"))
 opt_bfgs$mode
 opt_bfgs$hessian
+Matrix::forceSymmetric(solve(opt_bfgs$hessian))
 ```
@@
 
+And a second way of producing the AGHQ-PCA grids directly using `mvQuad`:
+
+```{r, fig.cap="Alternative approach AGHQ-PCA grid for Gaussian with $k = 3$."}
+pca2_grid_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(1, 1))
+mvQuad::rescale(pca2_grid_1, m = m, C = C, dec.type = 1)
+
+pca2_grid_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(3, 1))
+mvQuad::rescale(pca2_grid_3, m = m, C = C, dec.type = 1)
+
+pca2_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(5, 1))
+mvQuad::rescale(pca2_grid_5, m = m, C = C, dec.type = 1)
+
+pca2_grid_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(7, 1))
+mvQuad::rescale(pca2_grid_7, m = m, C = C, dec.type = 1)
+
+plot_points(pca2_grid_3)
+
+norm_pca2_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_1, adapt = FALSE)
+norm_pca2_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_3, adapt = FALSE)
+norm_pca2_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_5, adapt = FALSE)
+norm_pca2_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_7, adapt = FALSE)
+norm_pca2_bfgs <- list(norm_pca2_bfgs_1, norm_pca2_bfgs_3, norm_pca2_bfgs_5, norm_pca2_bfgs_7)
+```
+
 The density is already normalised, so it has a normalising constant of 1, and a log normalising constant of $\log(1) = 0$:
 
-```{r fig.cap="Unit test for the Gaussian example"}
+```{r, class.source = 'fold-hide', fig.cap="Unit test for the Gaussian example"}
 truelognormconst <- 0
@@
 results <- data.frame(
-  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7))),
-  "lognormconst" = c(truelognormconst, sapply(norm_bfgs, function(x) x$lognormconst), sapply(norm_pca_bfgs, function(x) x$lognormconst)),
-  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4)))
+  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7)), paste0("aghq-pca2, k = ", c(1, 3, 5, 7))),
+  "lognormconst" = c(truelognormconst, sapply(c(norm_bfgs, norm_pca_bfgs, norm_pca2_bfgs), function(x) x$lognormconst)),
+  "type" = c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4), rep("AGHQ-PCA2", 4))
 )
 
-ggplot(results, aes(x = method, y = lognormconst, col = type)) +
+ggplot(results, aes(x = method, y = lognormconst, col = as.factor(type))) +
   geom_point(size = 1) +
@@
 
-```{r}
+```{r, fig.cap="AGHQ grid for non-Gaussian with $k = 3$."}
 grid_bfgs_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
@@
 
+plot_points(grid_bfgs_3)
+```
+
+```{r, fig.cap="AGHQ-PCA grid for non-Gaussian with $k = 3$."}
 pca_grid_bfgs_1 <- pca_rescale(m = m, C = C, s = 1, k = 1)
@@
 pca_grid_bfgs_7 <- pca_rescale(m = m, C = C, s = 1, k = 7)
-
-#' AGHQ
-plot_points(grid_bfgs_7)
-
-#' PCA-AGHQ
-#' Something to fix here: points not being scaled to the right place
-plot_points(pca_grid_bfgs_7)
 
@@
 
+plot_points(pca_grid_bfgs_3)
+```
+
+```{r, fig.cap="Alternative approach AGHQ-PCA grid for non-Gaussian with $k = 3$."}
+pca2_grid_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(1, 1))
+mvQuad::rescale(pca2_grid_1, m = m, C = C, dec.type = 1)
+
+pca2_grid_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(3, 1))
+mvQuad::rescale(pca2_grid_3, m = m, C = C, dec.type = 1)
+
+pca2_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(5, 1))
+mvQuad::rescale(pca2_grid_5, m = m, C = C, dec.type = 1)
+
+pca2_grid_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(7, 1))
+mvQuad::rescale(pca2_grid_7, m = m, C = C, dec.type = 1)
+
+norm_pca2_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_1, adapt = FALSE)
+norm_pca2_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_3, adapt = FALSE)
+norm_pca2_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_5, adapt = FALSE)
+norm_pca2_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_7, adapt = FALSE)
+norm_pca2_bfgs <- list(norm_pca2_bfgs_1, norm_pca2_bfgs_3, norm_pca2_bfgs_5, norm_pca2_bfgs_7)
+
+plot_points(pca2_grid_3)
+```
+
+```{r, class.source = 'fold-hide', fig.cap="Unit test for the non-Gaussian example"}
 results <- data.frame(
-  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7))),
-  "lognormconst" = c(truelognormconst, sapply(norm_bfgs, function(x) x$lognormconst), sapply(norm_pca_bfgs, function(x) x$lognormconst)),
-  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4)))
+  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7)), paste0("aghq-pca2, k = ", c(1, 3, 5, 7))),
+  "lognormconst" = c(truelognormconst, sapply(c(norm_bfgs, norm_pca_bfgs, norm_pca2_bfgs), function(x) x$lognormconst)),
+  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4), rep("AGHQ-PCA2", 4)))
 )

### 2023-04-03 src/docs_paper/appendix.Rmd
Add multinomial to main text, many changes to appendix including table of number of latent field and hyperparameters, and table connecting TMB notation to mathematical notation

@@
 
-<!-- So that the tables and figures in this appendix are prefixed by "A." -->
+```{=tex}
 \setcounter{table}{0}  
-\renewcommand{\thetable}{A.\arabic{table}} 
+\renewcommand{\thetable}{S\arabic{table}} 
 \setcounter{figure}{0} 
-\renewcommand{\thefigure}{A.\arabic{figure}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{section}{0} 
+\renewcommand{\thesection}{S\arabic{section}}
+```
 
-# Simplified Naomi model description
+# Simplified Naomi model description \label{sec:math}
 
-In this section we describes more fully the simplified Naomi model considered in the main text.
-To do so, we draw substantially from the supplementary material of @eaton2021naomi.
+In this section we describe more fully the simplified Naomi model considered in the main text. In doing so we draw from the supplementary material of @eaton2021naomi.
 
@@
 
-Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest at time $t = T_1$
+Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest.
 Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
@@
 As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
-We index the following quantities by district, sex and age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$, annual HIV incidence rate $\lambda_{x, s, a}$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a}$.
+We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities: HIV prevalence $\rho_{x, s, a} \in [0, 1]$, ART coverage $\alpha_{x, s, a} \in [0, 1]$, annual HIV incidence rate $\lambda_{x, s, a} > 0$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
 Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
-In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.
+In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
+In the main text we use a more concise, but less descriptive, approach.
+Finally, we let $\sum_{\{x\}}$ be used as shorthand for $\sum_{x \in \{x\}}$, and likewise for $s$ and $a$.
 
@@
 Let $u$ be a generic random effect.
-We use use structured random effects to share information across units assessed as being similar, such as neighbouring districts or adjacent age-bands.
+We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as neighbouring districts or adjacent age-bands.
+We specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
+\begin{align*}
+u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
+u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
+\end{align*}
+where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, and $|\rho| < 1$ is the lag-one correlation parameter.
 We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
@@
 $$
-We specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
-\begin{align*}
-u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
-u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
-\end{align*}
-where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, $|\rho| < 1$ is the lag-one correlation parameter.
 
@@
 
-We assume the household survey was run according to a complex survey design where each individual $j \in U$ has probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$.
+We assume the household survey was run according to a complex survey design where each individual $j \in U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$.
 Suppose we observe an outcome $y_j$ for $j \in S$.
-Let $w_j = 1 / \pi_j$ be design weights, then the weighted mean
+Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then the weighted mean
 $$
-\hat y = \frac{\sum_{j \in S} w_j y_j}{\sum_{j \in S} w_j}
+\hat y = \frac{\sum_{j \in S} w_j y_j}{\sum_{j \in S} w_j}.
 $$
-Define Kish ESS.
+The Kish effective sample size is
+$$
+m = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
+$$
 We make these computations using the `survey` \textsc{R} package [@lumley2004analysis].
@@
 
+```{=tex}
+\begin{table}[h]
+\centering
+\begin{tabularx}{\textwidth}{llXX}
+\toprule
+& Model section & Latent field & Hyperparameter \\ 
+\midrule
+\ref{sec:hiv-prev} & HIV prevalence & $36 + 3n$ & 9 \\ 
+\ref{sec:art-cov} & ART coverage & $36 + 3n$ & 9 \\ 
+\ref{sec:hiv-inc} & HIV incidence rate & $2 + n$ & 3 \\ 
+\ref{sec:anc-test} & ANC testing & $4 + 2n$ & 2 \\ 
+\ref{sec:art-attend} & ART attendance & $n$ & 1 \\ 
+& Total & $78 + 7n$ & 24 \\ 
+\bottomrule
+\end{tabularx}
+\label{tab:latent-hyper-size}
+\caption{The numer of latent field parameters and hyperparameters in each model section.}
+\end{table}
+```
+
+In this section, we describe the hyperparameter and latent field process specification for the model.
+Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood}.
+Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each section of the model.
+Given a country with $n$ districts, then the total number of unique district-age-sex combinations is $n \times 2 \times 17 = 34n$.
+In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $491$ and the total number of hyperparameters is $24$.
+
 ### HIV prevalence \label{sec:hiv-prev}
 
-We model HIV prevalence on the logit scale using the linear predictor
+We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
 $$
@@
 $$
-where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
-We use the prior distributions $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.
+where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
+We use the prior distributions $\mathcal{N}^{+}(0, 2.5)$ for the standard deviation terms $\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\}$, $\mathcal{U}(-1, 1)$ for the AR1 correlation parameters $\{\phi_A^\rho, \phi_{AS}^\rho\}$, and $\text{Beta}(0.5, 0.5)$ for the BYM2 proportion parameters $\{\phi_X^\rho, \phi_{XS}^\rho\}$.
+
+<!-- Latent field size -->
+
+<!-- 1 + 1 + 17 + 17 + n + n + n = 36 + 3n -->
+
+<!-- Hyperparameter size -->
+
+<!-- 5 + 2 + 2 -->
 
@@
 
-We model ART coverage using the linear predictor
+We model ART coverage $\alpha_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
 $$
@@
 
-### HIV incidence rate
+<!-- Latent field size -->
 
-We model HIV incidence rate using the linear predictor
+<!-- As above -->
+
+<!-- Hyperparameter size -->
+
+<!-- As above -->
+
+### HIV incidence rate \label{sec:hiv-inc}
+
+We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
 $$
@@
 $$
-where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women, $\rho_{x}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences
+where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women,
 $$
@@
 $$
-$\alpha_{x}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages
+is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences, and
 $$
@@
 $$
-$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output.
-Note that the HIV incidence rate process (Equation \ref{eq:hiv-inc}) depends on those of the HIV prevalence (Section \ref{sec:hiv-prev}) and the ART coverage (Section \ref{sec:art-cov}) via the adult aggregate indicators (Equations \ref{eq:rho-agg} and \ref{eq:alpha-agg}) violating the conditions required being within the latent Gaussian model (LGM) class as defined by @rue2009approximate.
+is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages.
+The term $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output.
 
-### ANC testing
+We model the proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ as
+$$
+\kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right),
+$$
+where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection (MDRI), and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio (FRR).
+We use an informative prior on $\Omega_T$ based on the characteristics of the recent infection testing algorithm (RITA).
+For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$. 
+For PHIA surveys we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.
 
-HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit-scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
-\begin{align}
-\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \label{eq:anc-hiv-prev} \\
-\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}} \label{eq:anc-art-cov},
-\end{align}
+<!-- Latent field size -->
+
+<!-- 1 + 1 + n + 1 + 1 = 4 + n -->
+
+<!-- Hyperparameter size -->
+
+<!-- 3 -->
+
+### ANC testing \label{sec:anc-test}
+
+HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
+\begin{align*}
+\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
+\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}},
+\end{align*}
 where, for $\theta \in \{\rho, \alpha\}$, $\beta^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, 5)$ are the average differences between population and ANC outcomes, $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ are IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$, and $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$.
-Dependence here too on $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ in Equations \ref{eq:anc-hiv-prev} and \ref{eq:anc-art-cov} violates the conditions required to be a LGM.
 
-### ART attendance
+For adult women 15-49 we model the number of ANC clients $\Psi_{x, a} > 0$ on the log scale
+$$
+\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
+$$
+where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are fixed log rate ratios for the number of ANC clients relative to the predicted fertility, and $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects, with $\sigma^\psi \sim \mathcal{N}^+(0, 1)$.
+
+<!-- Latent field size -->
+<!-- 1 + n + 1 + n = 2 + 2n -->
+<!-- Hyperparameter size -->
+<!-- 2 -->
+
+### ART attendance \label{sec:art-attend}
 
@@
 
-Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
+Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
 We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district or its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation, and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.
@@
 
-## Likelihood specification
+## Likelihood specification \label{sec:likelihood}
 
@@
 
-For $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}}$, and observed number of cases
+For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}}$, and observed number of cases $y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}}$.
+We use following three binomial working likelihoods
+\begin{align*}
+y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
+\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}, \\
+y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
+\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
+\\
+y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
+\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}},
+\end{align*}
+The generalised binomial $y \sim \text{xBin}(m, p)$ is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
 $$
-y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}}.
+\log p(y) = \log \Gamma(m + 1) - \log \Gamma(y + 1) - \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
 $$
-
-For HIV prevalence and ART coverage we use the binomial working likelihoods
-\begin{align}
-y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \\
-\rho_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
-\end{align}
-and
-\begin{align}
-y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \\
-\alpha_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
-\end{align}
-
-For recent infections we also use a binomial working likelihood
-\begin{align}
-y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \\
-\kappa_{x, s, a} &= 1 - \exp(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T ),
-\end{align}
-where $\kappa_{x, s, a}$ are the predicted proportion recently infected among HIV positive persons, $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection (MDRI), and $\beta_T \sim \mathcal{N}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio (FRR).
-We use an informative prior on $\Omega_T$ based on the characteristics of the recent infection testing algorithm (RITA).
-For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$.
-For PHIA surveys we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.
+where the gamma function $\Gamma$ is such that $\forall n \in \mathbb{N}$, $\Gamma(n) = (n - 1)!$.
 
@@
 
-For women 15-49 the predicted number of ANC clients $\Psi_{x, a}$ is a log-linear model
-$$
-\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
-$$
-where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are ASFR in Spectrum region $R_x$ at time $t$, $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility, $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
-
-We include ANC testing data for the year of the most recent survey $Y[T_1]$.
-Let $W^\text{ANC}_{\{x\}, Y[t]}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}, Y[t]}$ the number of those with ascertained status, $y^\text{ANC}_{\{x\}, Y[t]}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}, Y[t]}$ the number of ANC clients already on ART prior to first ANC, such that $W^\text{ANC}_{x, Y[t]} \geq X^\text{ANC}_{x, Y[t]} \geq y^\text{ANC}_{x, Y[t]} \geq Z^\text{ANC}_{x, Y[t]}$
-When ANC testing data are only available for part of a given year, we denote $m^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$.
-
-The observed number of HIV positive and already on ART among ANC clients at $Y[T_1]$ is modelled by
+We include ANC testing data for the year of the most recent survey.
+Let $W^\text{ANC}_{\{x\}}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}}$ the number of ANC clients already on ART prior to first ANC, such that $W^\text{ANC}_{x} \geq X^\text{ANC}_{x} \geq y^\text{ANC}_{x} \geq Z^\text{ANC}_{x}$ for all $x \in \{x\}$.
+When ANC testing data are only available for part of a given year, we denote $m^\text{ANC} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for that year.
+The observed number of HIV positive and already on ART among ANC clients is modelled by
 \begin{align*}
-Y_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[T_1]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right) \\
-Z_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[T_1]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right)
+Y_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} }^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right), \\
+Z_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} }^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right),
 \end{align*}
@@
 \begin{align*}
-\rho_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a}} \\
-\alpha_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}
+\rho_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a}}, \\
+\alpha_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}.
 \end{align*}
@@
 $$
-\dot A_{\{x\}, \{s\}, \{a\}} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a},
+\dot A_{\{x\}, \{s\}, \{a\}} = \sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a}.
 $$
@@
 $$
-where $\pi_{x', x, s, a} = \rho_{x', s, a} \cdot \alpha_{x', s, a} \cdot \gamma_{x', x, s, a}$.
-This likelihood is approximated using a normal for the sum of binomials by
+where $\pi_{x', x, s, a} = \rho_{x', s, a} \cdot \alpha_{x', s, a} \cdot \gamma_{x', x, s, a}$. This likelihood is approximated using a normal for the sum of binomials by
 $$
@@
 $$
-where
+where the mean is
 $$
-\tilde{A}_{\{x\},\{s\}\{a\}} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a},
+\tilde{A}_{\{x\},\{s\}\{a\}} = \sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a},
 $$
-and
+and the variance is
 $$
-\sigma_{\{x\},\{s\}\{a\}}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a} \cdot \left(1-\pi_{x', x, s, a}\right)}.
+\left(\sigma_{\{x\},\{s\}\{a\}}^{\tilde{A}}\right)^2 = \sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a} \cdot \left(1-\pi_{x', x, s, a}\right).
 $$
@@
 
-* If survey data on ART coverage by age and sex are not available then we set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age/sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$.
-* If no ART data (survey or ART programme) are available at $T_1$ but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
-In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated
-$$
-\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha \\
-$$
-* If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$.
+1.
+If survey data on ART coverage by age and sex are not available then we set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age-sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$.
+2.
+If no ART data (survey or ART programme) are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
+In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha$.
+3.
+If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$.
 The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.
+
+For the Malawi example considered in the main text, all required data are available.
+
+\newpage
+
+# \textsc{C++} `TMB` user template
+
+<!-- Use table(names(tmb$fit$par.full)) to check the below table -->
+
+```{=tex}
+\begin{table}[h!]
+\centering
+\begin{tabularx}{\textwidth}{lllll}
+\toprule
+Variable name & Mathematical notation & Type & Dimension & Domain \\ 
+\midrule
+\texttt{logit\_phi\_rho\_x} & $\text{logit}(\phi_X^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_rho\_x} & $\log(\sigma_X^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{logit\_phi\_rho\_xs} & $\text{logit}(\phi_{XS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_rho\_xs} & $\log(\sigma_{XS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{logit\_phi\_rho\_a} & $\text{logit}(\phi_A^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_rho\_a} & $\log(\sigma_A^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{logit\_phi\_rho\_as} & $\text{logit}(\phi_{AS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_rho\_as} & $\log(\sigma_{AS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_rho\_xa} & $\log(\sigma_{XA}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{logit\_phi\_alpha\_x} & $\text{logit}(\phi_X^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_alpha\_x} & $\log(\sigma_X^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{logit\_phi\_alpha\_xs} & $\text{logit}(\phi_{XS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_alpha\_xs} & $\log(\sigma_{XS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{logit\_phi\_alpha\_a} & $\text{logit}(\phi_A^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_alpha\_a} & $\log(\sigma_A^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{logit\_phi\_alpha\_as} & $\text{logit}(\phi_{AS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_alpha\_as} & $\log(\sigma_{AS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_alpha\_xa} & $\log(\sigma_{XA}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{OmegaT\_raw} & $\Omega_T$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_betaT} & $\log(\beta_T)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_lambda\_x} & $\log(\sigma^\lambda)$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_ancrho\_x} & $\log(\sigma_X^{\rho^{\text{ANC}}})$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_ancalpha\_x} & $\log(\sigma_X^{\alpha^{\text{ANC}}})$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{log\_sigma\_or\_gamma} & $\log(\sigma_X^{\tilde \gamma})$ & Hyper & 1 & $\mathbb{R}$ \\
+\texttt{beta\_rho} & $(\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ \\
+\texttt{beta\_alpha} & $(\beta^\alpha_0, \beta_{S}^{\alpha, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ \\
+\texttt{beta\_lambda} & $(\beta_0^\lambda, \beta_S^{\lambda, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ \\
+\texttt{beta\_anc\_rho} & $\beta^{\rho^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ \\
+\texttt{beta\_anc\_alpha} & $\beta^{\alpha^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ \\
+\texttt{u\_rho\_x} & $u^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{us\_rho\_x} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_rho\_xs} & $u_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{us\_rho\_xs} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_rho\_a} & $u^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ \\
+\texttt{u\_rho\_as} & $u_a^{\rho, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ \\
+\texttt{u\_rho\_xa} & $u_x^{\rho, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_alpha\_x} & $u^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{us\_alpha\_x} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_alpha\_xs} & $u_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{us\_alpha\_xs} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_alpha\_a} & $u^\alpha_a$ & Latent & 13 & $\mathbb{R}^{13}$ \\
+\texttt{u\_alpha\_as} & $u_a^{\alpha, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ \\
+\texttt{u\_alpha\_xa} & $u_x^{\alpha, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{ui\_lambda\_x} & $u_x^\lambda$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{ui\_anc\_rho\_x} & $u_x^{\rho^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{ui\_anc\_alpha\_x} & $u_x^{\alpha^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{log\_or\_gamma} & $u_x^{\tilde \gamma}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\bottomrule
+\end{tabularx}
+\label{tab:tmb-lookup}
+\caption{Correspondence between mathematical notation and variable names used in our \texttt{TMB} code.
+The total number of hyperparameters is 24, and the total number of latent field parameters is $51 + 14n$.}
+\end{table}
+```
+
+<!-- Latent field size -->
+<!-- 2 + 2 + 2 + 1 + 1+ 10 + 10 + 13 + 10 + 14n = 51 + 14n -->
+<!-- Hyperparameter size -->
+<!-- 24 -->
+
+This section contains C++ for the negative log posterior of the simplified Naomi model we describe in Section \ref{sec:math}.
+For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used and the variable names used, for all hyperparameters and latent field parameters.
+For further reference on the `TMB` software see `https://kaskr.github.io/adcomp/_book`.
+
+```{cpp, echo=FALSE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
+
+```
 
@@
 
-* Improved $\hat R$ statistic from @vehtari2021rank. Recommended only to use the sample if the value is less than 1.05.
-* ESS and autocorrelation related to efficiency
-* Traceplots helpful for diagnosis of problems
-* Pairs density plots helpful for understanding relationships between parameters, including possible indentifiability issues
-* Energy plot from @betancourt2017conceptual
-
-\newpage
-
-# \textsc{C++} `TMB` user template
-
-```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
-
-```
+- Improved $\hat R$ statistic from @vehtari2021rank. Recommended only to use the sample if the value is less than 1.05.
+- ESS and autocorrelation related to efficiency
+- Traceplots helpful for diagnosis of problems
+- Pairs density plots helpful for understanding relationships between parameters, including possible identifiability issues
+- Energy plot from @betancourt2017conceptual
 

### 2023-04-03 src/docs_paper/paper.Rmd
Add multinomial to main text, many changes to appendix including table of number of latent field and hyperparameters, and table connecting TMB notation to mathematical notation

@@
 An overview of this simplified model is given below, highlighting the aspects which make it a challenge for existing inferential approaches.
-A more complete mathematical description, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.
+A more complete mathematical description (Appendix S1) as well as a \textsf{C++} template for the log-posterior (Appendix S3) are provided in the supplementary material.
 
@@
 
-People living with HIV may choose to access ART services outside of the district that they reside in.
+We use a multinomial logistic regression model to account for people living with HIV choosing to access ART services outside of the district that they reside in.
 Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
 We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent such that $x \sim x'$. 
-We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a multinomial logistic regression model.
-
-This section remains to write, and is quite challenging.
-
-<!-- Let $\dot A_\mathcal{I} = \sum_{x \sim x', x = x'} \dot A_{x', x}$ be the number of people receiving ART. -->
-<!-- Model $\dot A_\mathcal{I} \sim \mathcal{N}(\tilde A_\mathcal{I}, \sigma^{\tilde A_\mathcal{I}})$ using a normal distribution with mean $\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x}$ and standard deviation $\sigma^{\tilde A_\mathcal{I}}) = \sqrt{\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x} (1 - \pi_{x', x})}$.
--->
+We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a linear predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
+We model aggregate ART attendance data $\dot A_\mathcal{I}$ using a Gaussian approximation to a sum of binomials.
+The sum results both by aggregation over $i \in \mathcal{I}$ and by number of ART clients travelling from district $x'$ to $x$.
+More details regarding this part of the model are provided in Appendix S1.
 

### 2023-04-03 src/docs_paper/appendix.Rmd
Appendix puzzle fixes and paper edits

@@
 ---
-title: Appendix to "Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model"
+title: Appendix to "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
 subtitle: |
@@
 
-In this section we describe more fully the simplified Naomi model considered in the main text. In doing so we draw from the supplementary material of @eaton2021naomi.
+In this section we describe the simplified Naomi model considered in the main text in greater detail.
+In doing so we draw from the supplementary material of @eaton2021naomi.
 
@@
 
-Let $u$ be a generic random effect.
 We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as neighbouring districts or adjacent age-bands.
-We specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
+Let $u$ be a generic random effect, with some length.
+Three types of structured random effects are used in the model.
+First, we specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
 \begin{align*}
@@
 \end{align*}
-where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, and $|\rho| < 1$ is the lag-one correlation parameter.
-We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
+where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the lag-one correlation parameter.
+Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
 $$
@@
 We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
-For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v^\star$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w^\star$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
+Third, for the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
 $$
-u = \sigma \left( \sqrt\phi \cdot v^\star + \sqrt{1 - \phi} \cdot w^\star \right).
+u = \sigma \left( \sqrt\phi \cdot v + \sqrt{1 - \phi} \cdot w \right).
 $$
@@
 We assume the household survey was run according to a complex survey design where each individual $j \in U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$.
-Suppose we observe an outcome $y_j$ for $j \in S$.
+Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
 Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then the weighted mean
 $$
-\hat y = \frac{\sum_{j \in S} w_j y_j}{\sum_{j \in S} w_j}.
+\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
 $$
-The Kish effective sample size is
+We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
 $$
-m = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
+m^{\hat \theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
 $$
-We make these computations using the `survey` \textsc{R} package [@lumley2004analysis].
+The observed number of indicator cases is then $y^{\hat \theta} = m^{\hat \theta} \cdot \hat \theta$.
+To make these computations we use the `survey` \textsc{R} package [@lumley2004analysis].
 
@@
 \midrule
-\ref{sec:hiv-prev} & HIV prevalence & $36 + 3n$ & 9 \\ 
-\ref{sec:art-cov} & ART coverage & $36 + 3n$ & 9 \\ 
+\ref{sec:hiv-prev} & HIV prevalence & $22 + 5n$ & 9 \\ 
+\ref{sec:art-cov} & ART coverage & $25 + 5n$ & 9 \\ 
 \ref{sec:hiv-inc} & HIV incidence rate & $2 + n$ & 3 \\ 
-\ref{sec:anc-test} & ANC testing & $4 + 2n$ & 2 \\ 
+\ref{sec:anc-test} & ANC testing & $2 + 2n$ & 2 \\ 
 \ref{sec:art-attend} & ART attendance & $n$ & 1 \\ 
-& Total & $78 + 7n$ & 24 \\ 
+& Total & $51 + 14n$ & 24 \\ 
 \bottomrule
@@
 In this section, we describe the hyperparameter and latent field process specification for the model.
-Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood}.
+Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood} to follow.
 Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each section of the model.
-Given a country with $n$ districts, then the total number of unique district-age-sex combinations is $n \times 2 \times 17 = 34n$.
-In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $491$ and the total number of hyperparameters is $24$.
+In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $499$ and the total number of hyperparameters is $24$.
+
+<!-- Had previously been getting a figure of 491 somewhere -- which is 51 + 14n + 24. Something going on there? -->
 
@@
 where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
-We use the prior distributions $\mathcal{N}^{+}(0, 2.5)$ for the standard deviation terms $\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\}$, $\mathcal{U}(-1, 1)$ for the AR1 correlation parameters $\{\phi_A^\rho, \phi_{AS}^\rho\}$, and $\text{Beta}(0.5, 0.5)$ for the BYM2 proportion parameters $\{\phi_X^\rho, \phi_{XS}^\rho\}$.
+The two BYM2 random effects are comprised of the following unit scaled patially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components
+\begin{align*}
+u_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot v_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot w_x^{\rho} \right), \\
+u_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot v_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot w_x^{\rho, s = \text{M}} \right).
+\end{align*}
+We use half-normal priors for the standard deviation terms $\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5)$, uniform priors for the AR1 correlation parameters $\{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1)$, and beta priors for the BYM2 proportion parameters $\{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5)$.
 
 <!-- Latent field size -->
-
-<!-- 1 + 1 + 17 + 17 + n + n + n = 36 + 3n -->
-
+<!-- 1 + 1 + 10 + 10 + n + n + n = 36 + 3n -->
 <!-- Hyperparameter size -->
-
 <!-- 5 + 2 + 2 -->
@@
 <!-- Latent field size -->
-
-<!-- As above -->
-
+<!-- 1 + 1 + 13 + 10 + n + n + n = 36 + 3n -->
 <!-- Hyperparameter size -->
-
-<!-- As above -->
+<!-- 5 + 2 + 2 -->
 
@@
 $$
-\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}, \label{eq:rho-agg}
+\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}
 $$
@@
 $$
-\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}, \label{eq:alpha-agg}
+\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}
 $$
@@
 <!-- Latent field size -->
-
 <!-- 1 + 1 + n + 1 + 1 = 4 + n -->
-
 <!-- Hyperparameter size -->
-
 <!-- 3 -->
@@
 y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
-\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
-\\
+\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}, \\
 y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
@@
 We include ANC testing data for the year of the most recent survey.
-Let $W^\text{ANC}_{\{x\}}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}}$ the number of ANC clients already on ART prior to first ANC, such that $W^\text{ANC}_{x} \geq X^\text{ANC}_{x} \geq y^\text{ANC}_{x} \geq Z^\text{ANC}_{x}$ for all $x \in \{x\}$.
+Let $W^\text{ANC}_{\{x\}}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}}$ the number of ANC clients already on ART prior to first ANC, such that
+$$
+W^\text{ANC}_{x} \geq X^\text{ANC}_{x} \geq Y^\text{ANC}_{x} \geq Z^\text{ANC}_{x},
+$$
+for all $x \in \{x\}$.
 When ANC testing data are only available for part of a given year, we denote $m^\text{ANC} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for that year.
@@
 1.
-If survey data on ART coverage by age and sex are not available then we set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age-sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$.
+If survey data on HIV prevalence or ART coverage by age and sex are not available then we set $u_a^\theta = 0$ and $u_{a, s = \text{M}}^\theta = 0$ and use the average age-sex pattern of from the Spectrum offset $\eta_{R_x, s, a}^\theta$.
+For the Malawi example considered in the main text HIV prevalence and ART coverage data is not available for those aged 65+.
 2.
-If no ART data (survey or ART programme) are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
+If no ART data, either survey or ART programme, are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
 In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha$.
@@
 The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.
-
-For the Malawi example considered in the main text, all required data are available.
 
@@
 \texttt{beta\_anc\_alpha} & $\beta^{\alpha^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ \\
-\texttt{u\_rho\_x} & $u^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{us\_rho\_x} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{u\_rho\_xs} & $u_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{us\_rho\_xs} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_rho\_x} & $w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{us\_rho\_x} & $v^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_rho\_xs} & $w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{us\_rho\_xs} & $v_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
 \texttt{u\_rho\_a} & $u^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ \\
@@
 \texttt{u\_rho\_xa} & $u_x^{\rho, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{u\_alpha\_x} & $u^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{us\_alpha\_x} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{u\_alpha\_xs} & $u_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{us\_alpha\_xs} & & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_alpha\_x} & $w^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{us\_alpha\_x} & $v^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{u\_alpha\_xs} & $w_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{us\_alpha\_xs} & $v_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
 \texttt{u\_alpha\_a} & $u^\alpha_a$ & Latent & 13 & $\mathbb{R}^{13}$ \\
@@
 \caption{Correspondence between mathematical notation and variable names used in our \texttt{TMB} code.
-The total number of hyperparameters is 24, and the total number of latent field parameters is $51 + 14n$.}
+The total number of hyperparameters is 24, and the total number of latent field parameters is $51 + 14n$, where $n$ is the number of districts.}
 \end{table}
@@
 
-```{cpp, echo=FALSE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
-
+```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
 ```

### 2023-04-03 src/docs_paper/paper.Rmd
Appendix puzzle fixes and paper edits

@@
 
-<!-- * [ ] Description of the ART attendance component of the model -->
 <!-- * [ ] Tying together Naomi model within LGM / ELGM mathematical notation -->
@@
 
-Naomi is not an LGM, and instead falls into the ELGM class, for the following reasons:
+Naomi is not an LGM, and instead falls into the ELGM class, because:
 
-1. In the household survey component (Section \ref{sec:household}), the HIV incidence rate depends on the HIV prevalence and ART coverage linear predictors.
-2. In the ANC testing component (Section \ref{sec:anc}), the HIV prevalence and ART coverage depend upon the household survey component.
+1. In the ANC testing component (Section \ref{sec:anc}), the HIV prevalence and ART coverage depend upon the household survey component.
 Specifically, $|\mathcal{J}_i| = 2$ such that for $\theta \in \{\rho, \alpha\}$
@@
 $$
-3. In the ART attendance component (Section \ref{sec:art}), the multinomial logistic regression.
-4.
-In all components the data we model is aggregated.
-Processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations over sets of indices $\mathcal{I}$.
-
-<!-- To-do: define what exactly $p(\y, \x, \btheta)$ is for the Naomi model -->
+2. In the household survey component (Section \ref{sec:household}), the HIV incidence rate depends on adult HIV prevalence and adult ART coverage via non-linear transformations.
+3.
+Individuals who are taking ART, or have been recently infected, must be HIV positive.
+As a result observations are made taking the product of HIV prevalence with ART coverge or recent HIV infection.
+4. Recent HIV infection is non-linearly linked to both HIV incidence and HIV prevalence.
+5. The ART attendance component (Section \ref{sec:art}) uses a multinomial model.
+6. Processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations aggregated over sets of indices $\mathcal{I}$.
+7. Multiple link functions are used: there is no one $g$.
 
@@
 
+To write.
+
 ## AGHQ
 
-A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by a weighted sum
+A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by the following weighted sum
 $$
-\int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z)
+\int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z),
 $$
 where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
-In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials.
+In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials, and weights so as to interpolate functions of the form $f(\z) = \tilde f(\z) \exp(-\z^2)$.
 These nodes and weights can be shifted and scaled to suit the particular integrand using adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate].
+AGHQ with $k = 1$ recovers the Laplace approximation.
 

### 2023-04-04 src/docs_paper/appendix.Rmd
Fixes to appendix

@@
 title: Appendix to "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
-subtitle: |
-  Corresponding author: Adam Howes (`ath19@ic.ac.uk`)
+author:
+  - Adam Howes^[Department of Mathematics, Imperial College London]
+  - Alex Stringer^[Department of Statistics and Actuarial Science, University of Waterloo]
+  - Seth R. Flaxman^[Department of Computer Science, Oxford University]
+  - Jeffrey W. Eaton^[Department of Infectious Disease Epidemiology, Imperial College London]
 output:
@@
 bibliography: citations.bib
+cls: imsart.cls
 ---
@@
 
-In this section we describe the simplified Naomi model considered in the main text in greater detail.
-In doing so we draw from the supplementary material of @eaton2021naomi.
+In this section we describe the simplified Naomi model [@eaton2021naomi] considered in the main text in more complete detail.
 
@@
 
-Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest.
+Let $\nu$ be the most recent national household survey with HIV testing which has taken place in the country of interest.
 Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
@@
 In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
+We let $\sum_{\{x\}}$ be used as shorthand for $\sum_{x \in \{x\}}$, and likewise for $s$ and $a$.
 In the main text we use a more concise, but less descriptive, approach.
-Finally, we let $\sum_{\{x\}}$ be used as shorthand for $\sum_{x \in \{x\}}$, and likewise for $s$ and $a$.
 
@@
 
-We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as neighbouring districts or adjacent age-bands.
-Let $u$ be a generic random effect, with some length.
+We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
+Let $u$ be a generic random effect, with length $\dim(u)$.
 Three types of structured random effects are used in the model.
@@
 u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
-u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
+u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(u)
 \end{align*}
@@
 $$
-where $u_{-i}$ is $u$ with the $i$th unit removed, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
+where $u_{-i}$ is $u$ with the $i$th element removed i.e.
+$(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
 We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
@@
 \toprule
-& Model section & Latent field & Hyperparameter \\ 
+& Model component & Latent field & Hyperparameter \\ 
 \midrule
@@
 
-In this section, we describe the hyperparameter and latent field process specification for the model.
+We now describe the hyperparameter and latent field process specification for the model.
 Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood} to follow.
-Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each section of the model.
-In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $499$ and the total number of hyperparameters is $24$.
+Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each component of the model.
+In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $51 + 14 \cdot 32 = 499$ and the total number of hyperparameters is $24$.
 
@@
 $$
-\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
+\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a}. 
 $$
-where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
-The two BYM2 random effects are comprised of the following unit scaled patially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components
+The term $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept, and $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women.
+The terms $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, and $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$.
+The terms $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, and $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$.
+The terms $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence.
+Finally, $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
+The two BYM2 random effects are comprised of the following unit scaled spatially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components
 \begin{align*}
@@
 \end{align*}
-We use half-normal priors for the standard deviation terms $\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5)$, uniform priors for the AR1 correlation parameters $\{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1)$, and beta priors for the BYM2 proportion parameters $\{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5)$.
+We use half-normal priors for the standard deviation terms
+$$
+\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5),
+$$
+uniform priors for the AR1 correlation parameters
+$$
+\{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1),
+$$
+and beta priors for the BYM2 proportion parameters
+$$
+\{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5).
+$$
 
@@
 $$
-where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women,
+where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults and $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women.
+The term
 $$
@@
 is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages.
-The term $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output.
+The term $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation].
+The terms $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects.
+Finally, $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output.
 
@@
 $$
-where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection (MDRI), and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio (FRR).
-We use an informative prior on $\Omega_T$ based on the characteristics of the recent infection testing algorithm (RITA).
-For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$. 
-For PHIA surveys we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.
+where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection, and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio.
+We use an informative prior for $\Omega_T$ based on the characteristics of the recent infection testing algorithm.
+For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$, and further we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.
 
@@
 
-For adult women 15-49 we model the number of ANC clients $\Psi_{x, a} > 0$ on the log scale
+In the full Naomi model, for adult women 15-49 the number of ANC clients $\Psi_{x, a} > 0$ are modelled as
 $$
@@
 $$
-where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are fixed log rate ratios for the number of ANC clients relative to the predicted fertility, and $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects, with $\sigma^\psi \sim \mathcal{N}^+(0, 1)$.
+where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
+Here we fix $\beta^\psi = u_x^\psi = 0$ such that $\Psi_{x, a}$ are simply constants.
 
@@
 
-For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}}$, and observed number of cases $y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}}$.
-We use following three binomial working likelihoods
+For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
+$$
+y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}.
+$$
+To model these observations, we use following three binomial working likelihoods
 \begin{align*}
@@
 y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
-\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}},
+\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
 \end{align*}
@@
 \end{align*}
-where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a}$
+where prevalence and ART coverage are aggregated by the number of pregnant women $\Psi_{x, a}$
 \begin{align*}
@@
 If survey data on HIV prevalence or ART coverage by age and sex are not available then we set $u_a^\theta = 0$ and $u_{a, s = \text{M}}^\theta = 0$ and use the average age-sex pattern of from the Spectrum offset $\eta_{R_x, s, a}^\theta$.
-For the Malawi example considered in the main text HIV prevalence and ART coverage data is not available for those aged 65+.
+For the Malawi example considered in the main text HIV prevalence and ART coverage data are not available for those aged 65+.
+As a result, there are $|\{\text{0-4}, \ldots, \text{50-54}\}| = 13$ age bands included for the age random effects.
 2.
@@
 ```{=tex}
-\begin{table}[h!]
+\begin{table}[p!]
 \centering
@@
 
-This section contains C++ for the negative log posterior of the simplified Naomi model we describe in Section \ref{sec:math}.
-For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used and the variable names used, for all hyperparameters and latent field parameters.
-For further reference on the `TMB` software see `https://kaskr.github.io/adcomp/_book`.
+This section contains C++ `TMB` code for the negative log-posterior of the simplified Naomi model
+For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used in Section \ref{sec:math} and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
+For further reference on the `TMB` software see `https://kaskr.github.io/adcomp/_book`, or on the method see @kristensen2016tmb.
+
+\newpage
 

### 2023-04-04 src/docs_paper/appendix.Rmd
Add simplified Naomi DAG template, fix error with algorithm flowchart, add to appendix

@@
 ```
+
+\newpage
+
+# Inferential algorithm
+
+```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."}
+knitr::include_graphics("depends/algorithm-flowchart.pdf")
+```
+
+\newpage
 
@@
 
+```{r simplified-naomi, fig.cap="Directed acyclic graph describing the simplified Naomi model (work in progress)."}
+knitr::include_graphics("depends/simplified-naomi.pdf")
+```
+
 \newpage

### 2023-04-04 src/docs_paper/paper.Rmd
Add simplified Naomi DAG template, fix error with algorithm flowchart, add to appendix

@@
 <!-- * [ ] Tying together Naomi model within LGM / ELGM mathematical notation -->
-<!-- * [ ] Create DAG representing Naomi model -->
+<!-- * [ ] Create DAG representing Naomi model. This would likely go into the appenxix, so use that notation -->
 <!-- * [ ] Rewrite section on why Naomi is an ELGM, drawing from slides -->

### 2023-04-04 src/plot-tikz_algorithm-flowchart/algorithm-flowchart.tex
Add simplified Naomi DAG template, fix error with algorithm flowchart, add to appendix

@@
 
+\usepackage{amsfonts}
+\usepackage{bm}
 \usepackage{tikz}
 \usetikzlibrary{shapes,arrows,positioning}
-\usepackage{bm}
 
@@
 
-\tikzstyle{question} = [rectangle, draw, text width=10em, text centered, rounded corners, minimum height=4em]
 \tikzstyle{category} = [rectangle, draw, text width=10em, text centered, rounded corners, minimum height=4em]
@@
 \begin{tikzpicture}[node distance = 2cm, auto]
-
-    \node[category] at (0, 0) (1) {$p(\theta, \x, \y)$};
-
-    % x
-    \node[category] at (0, -3) (a) {$\tilde p_{\text{LA}}(x_i, \btheta, \y)$};
-    \node[category] at (0, -6) (b) {$\tilde p_{\text{AQ}}(x_i, \y)$};
-    \node[category] at (0, -9) (c) {$\tilde p_{\text{AQ}}(x_i \, | \, \y)$};
+  \node[category] at (0, 0) (1) {$p(\btheta, \x, \y)$};
+  \node[category] at (0, -3) (a) {$\tilde p_{\texttt{LA}}(x_i, \btheta, \y)$};
+  \node[category] at (0, -6) (b) {$\tilde p_{\texttt{AQ}}(x_i, \y)$};
+  \node[category] at (0, -9) (c) {$\tilde p_{\texttt{AQ}}(x_i \, | \, \y)$};
   \node[category] at (0, -12) (d) {Image of a nice curve};
-    \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\text{G}}(\x_{-i} \, | \, x_i, \btheta, \y)$}(a);
-    \path[line] (a) -- node[align=center] {Integrate out $\theta$\\$\sum_\z \tilde p_{\text{LA}}(x_i, \btheta(\z), \y) \omega(\z)$}(b);
+  \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\texttt{G}}(\x_{-i} \, | \, x_i, \btheta, \y)$}(a);
+  \path[line] (a) -- node[align=center] {Integrate out $\theta$\\$\sum_\z \tilde p_{\texttt{LA}}(x_i, \btheta(\z), \y) \omega(\z)$}(b);
   \path[line] (b) -- node[align=center] {Normalise}(c);
   \path[line] (c) -- node[align=center] {Evaluate $k = 7$ AGHQ nodes\\Use spline interpolation}(d);
-
-    % theta
-    \node[category] at (8, 0) (2) {$\tilde p_{\text{LA}}(\btheta, \y)$};
-    \node[category] at (8, -3) (3) {$\tilde p_{\text{AQ}}(\y)$};
-    \node[category] at (8, -6) (4) {$\tilde p_{\text{AQ}}(\btheta \, | \, \y)$};
-    \node[category] at (8, -9) (5) {$\tilde p_{\text{AQ}}(\theta_j \, | \, \y)$};
+  \node[category] at (8, 0) (2) {$\tilde p_{\texttt{LA}}(\btheta, \y)$};
+  \node[category] at (8, -3) (3) {$\tilde p_{\texttt{AQ}}(\y)$};
+  \node[category] at (8, -6) (4) {$\tilde p_{\texttt{AQ}}(\btheta \, | \, \y)$};
+  \node[category] at (8, -9) (5) {$\tilde p_{\texttt{AQ}}(\theta_j \, | \, \y)$};
   \node[category] at (8, -12) (6) {Image of a nice curve};
-    \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\text{G}}(\x \, | \, \btheta, \y)$}(2);
-    \path[line] (2) -- node[align=center] {Integrate out $\btheta$ with AGHQ\\Nodes $\z \in \mathcal{Q}(m, k)$\\Adapted $\btheta(\z) = \hat{\btheta} + Lz$\\Weights $\omega(\z):\mathcal{Q} \to \mathbb{R}$}(3);
+  \path[line] (1) -- node[align=center] {Laplace approximation\\Integrate out $\tilde p_{\texttt{G}}(\x \, | \, \btheta, \y)$}(2);
+  \path[line] (2) -- node[align=center] {Integrate out $\btheta$ with AGHQ\\Nodes $\z \in \mathcal{Q}(m, k)$\\Adapted $\btheta(\z) = \hat{\btheta} + Lz$\\Weights $\omega(\z): \mathcal{Q} \to \mathbb{R}$}(3);
   \path[line] (3) -- node[align=center] {Normalise} (4);
@@
   \path[line] (5) -- node[align=center] {Evaluate $k = 7$ AGHQ nodes\\Use spline interpolation}(6);
-
 \end{tikzpicture}

### 2023-04-05 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Add template for notebook about increasing the s and k parameters for simplified Naomi #22

---
title: "How does increasing $s$ and $k$ effect log normalising constant estimation"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**

  **Task**
---

# Data import

We start by reading in a selection of fitted PCA-AGHQ models:

```{r}
s1k2 <- readRDS("aghq-s1-k2.rds")
s2k2 <- readRDS("aghq-s2-k2.rds")
s3k2 <- readRDS("aghq-s3-k2.rds")
s4k2 <- readRDS("aghq-s4-k2.rds")
s5k2 <- readRDS("aghq-s5-k2.rds")
```

### 2023-04-05 src/naomi-simple_increase-s-k/functions.R

@@ -1,0 +1,1 @@
+#' Empty for now

### 2023-04-05 src/naomi-simple_increase-s-k/script.R

@@ -1,0 +1,5 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("naomi-simple_increase-s-k")
+# setwd("src/naomi-simple_increase-s-k")
+
+rmarkdown::render("increase-s-k.Rmd")

### 2023-04-05 src/dev_scale-grid/scale-grid.Rmd
Understanding further what is going on with the Choleksy adapted AGHQ with vector levels #22

@@
 end - start
+```
+
+Verifying that there are two grid points used in the 16th dimension:
+
+```{r}
+stopifnot(nrow(quad$modesandhessians) == 2)
+
+quad$modesandhessians[, 1:24] %>%
+  pivot_longer(cols = everything()) %>%
+  ggplot(aes(x = name, y = value)) +
+    geom_point(alpha = 0.8) +
+    labs(x = "", y = "Value") + 
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))
+```
+
+Note that for the Cholesky adapted grid with `levels = c(1, 1, ..., 2, 1, ..., 1)` we see variation in all of the dimensions further than the one with a level above one.
+I would assume this is related in some way to the Cholesky adaption being based on a lower triangular matrix.
+This is something it would be worthwhile to work out in more detail.
+
+```{r}
+diff(as.matrix(quad$modesandhessians[, 1:24]))
+```
+
+Verifying that adaption is happening as we expect it to be:
+
+```{r}
+#' This is what the base grid looks like
+mvQuad::getNodes(base_grid)
+
+#' Rescale it according to the mode and Hessian with Cholesky factor
+mvQuad::rescale(base_grid, m = quad$optresults$mode, C = Matrix::forceSymmetric(solve(quad$optresults$hessian)), dec.type=2)
+
+#' Verify that the adapted grid has the same nodes and are used by aghq::aghq
+stopifnot(all(mvQuad::getNodes(base_grid) == unname(as.matrix(quad$modesandhessians[, 1:24]))))
 ```

### 2023-04-06 src/dev_scale-grid/astro.Rmd
Cleaning scale grid notebook, astro into notebook, and rendering #22

---
title: "Milky way mass estimation using PCA-AGHQ"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**

  **Task** 
editor_options: 
  markdown: 
    wrap: sentence
---

```{r}
library(tidyverse)
library(patchwork)
library(TMB)
library(tmbstan)

#' Originally Alex used the ipoptr package for this, but it's not possible, or at
#' least very difficult, to install it anymore (as of 01/04/2023)
# install.packages("ipoptr")
# library(ipoptr)

#' We will here use the nloptr package instead
library(nloptr)

library(aghq)
library(Matrix)
library(gt)

data(gcdatalist, package = "aghq")
precompile()

set.seed(4365789)

globalpath <- normalizePath(tempdir(), winslash = "/")
plotpath <- normalizePath(file.path(globalpath, "astro"), winslash = "/", mustWork = FALSE)
if (!dir.exists(plotpath)) dir.create(plotpath)

#' TMB
#' Get the template from the aghq package, the compile and load it
file.copy(normalizePath(system.file("extsrc/01_astro.cpp", package = "aghq"), winslash = "/"), globalpath)
compile(normalizePath(file.path(globalpath, "01_astro.cpp"), winslash = "/"))
dyn.load(normalizePath(dynlib(file.path(globalpath, "01_astro")), winslash = "/"))

#' Parameter transformations
parambounds <- list(
  Psi0 = c(1, 200),
  gamma = c(0.3, 0.7),
  alpha = c(3.0, 3.7),
  beta = c(-0.5, 1)
)

get_Psi0 <- function(theta) {
  # theta = -log( (Psi0 - 1) / (200 - 1) )
  (parambounds$Psi0[2] - parambounds$Psi0[1]) * exp(-exp(theta)) + parambounds$Psi0[1]
}

get_theta1 <- function(Psi0) {
  log(-log((Psi0 - parambounds$Psi0[1]) / (parambounds$Psi0[2] - parambounds$Psi0[1])))
}

get_gamma <- function(theta) {
  # theta = -log( (gamma - 0.3) / (0.7 - 0.3) )
  (parambounds$gamma[2] - parambounds$gamma[1]) * exp(-exp(theta)) + parambounds$gamma[1]
}

get_theta2 <- function(gamma) {
  log(-log((gamma - parambounds$gamma[1]) / (parambounds$gamma[2] - parambounds$gamma[1])))
}

get_alpha <- function(theta) {
  # theta = log(alpha - 3)
  exp(theta) + parambounds$alpha[1]
}

get_theta3 <- function(alpha) {
  log(alpha - parambounds$alpha[1])
}

get_beta <- function(theta) {
  # theta = -log( (beta - (-0.5)) / (1 - (-0.5)) )
  (parambounds$beta[2] - parambounds$beta[1]) * exp(-exp(theta)) + parambounds$beta[1]
}

get_theta4 <- function(beta) {
  log(-log((beta - parambounds$beta[1]) / (parambounds$beta[2] - parambounds$beta[1])))
}

#' Optimisation using IPOPT replaced by optimisation using nloptr

nloptr_objective <- function(theta) ff$fn(theta)
nloptr_objective_gradient <- function(theta) ff$gr(theta)

# Box constraints, to improve stability of optimization
lowerbounds <- c(
  get_theta1(parambounds$Psi0[2] - 0.001),
  get_theta2(parambounds$gamma[2] - 0.001),
  get_theta3(parambounds$alpha[1] + 0.001),
  get_theta4(parambounds$beta[2] - 0.001)
)

upperbounds <- c(
  get_theta1(parambounds$Psi0[1] + 1),
  get_theta2(parambounds$gamma[1] + 0.01),
  get_theta3(parambounds$alpha[2] - 0.01),
  get_theta4(parambounds$beta[1] + 0.01)
)

# Start optimisation at the midpoint
thetastart <- (upperbounds + lowerbounds) / 2
thetastart_list <- list(theta1 = thetastart[1], theta2 = thetastart[2], theta3 = thetastart[3], theta4 = thetastart[4])

data.frame(
  "param" = paste0("theta", 1:4),
  lowerbound = lowerbounds,
  upperbound = upperbounds,
  thetastart = thetastart
) %>%
  ggplot(aes(ymin = lowerbound, y = thetastart, ymax = upperbound, x = param)) +
    geom_pointrange() +
    coord_flip() +
    labs(x = "", y = "") +
    theme_minimal()

# Nonlinear constraints, specified as a function
nloptr_nonlinear_constraints <- function(theta) -1 * Es$fn(theta)

nloptr_nonlinear_constraints_jacobian <- function(theta) {
  J <- -1 * Es$gr(theta)
  as.matrix(J)
}

nonlinear_lowerbound <- rep(0, nrow(gcdata) + 2)
nonlinear_upperbound <- rep(Inf, nrow(gcdata) + 2)

#' Objective function and its derivatives
ff <- MakeADFun(
  data = gcdatalist,
  parameters = thetastart_list,
  DLL = "01_astro",
  ADreport = FALSE,
  silent = TRUE
)

#' Nonlinear constraints and their Jacobian
Es <- MakeADFun(
  data = gcdatalist,
  parameters = thetastart_list,
  DLL = "01_astro",
  ADreport = TRUE,
  silent = TRUE
)

#' Check starting value satisfies constraints
#' Note that constraints in nloptr are of the form g(x) <= 0
stopifnot(all(nloptr_nonlinear_constraints(thetastart) <= 0))

tm <- Sys.time()

tmp <- capture.output(
  nloptr_result <- nloptr::nloptr(
    x0 = thetastart,
    eval_f = nloptr_objective,
    eval_grad_f = nloptr_objective_gradient,
    lb = lowerbounds,
    ub = upperbounds,
    eval_g_ineq = nloptr_nonlinear_constraints,
    eval_jac_g_ineq = nloptr_nonlinear_constraints_jacobian,
    opts = list("algorithm" = "NLOPT_LD_MMA", obj_scaling_factor = 1, "xtol_rel" = 1.0e-3)
  )
)

optruntime <- difftime(Sys.time(), tm, units = "secs")
cat("Run time for mass model optimisation:", optruntime, "seconds.\n")

#' Check that the optimisation results look right (based on Figure 2 from Bilodeau, Stringer, Tang)
get_Psi0(nloptr_result$solution[1])
get_gamma(nloptr_result$solution[2])
get_alpha(nloptr_result$solution[3])
get_beta(nloptr_result$solution[4])

#' AGHQ
#' Create the optimization template
ffa <- list(
  fn = function(x) -1 * ff$fn(x),
  gr = function(x) -1 * ff$gr(x),
  he = function(x) -1 * ff$he(x)
)

useropt <- list(
  ff = ffa,
  mode = nloptr_result$solution,
  hessian = ff$he(nloptr_result$solution)
)

#' Fit it with "reuse" first, then correct marginals later
cntrl <- aghq::default_control(negate = TRUE)

#' Do the quadrature
tm <- Sys.time()

k_dense <- 5
astroquad <- aghq::aghq(ff, k = k_dense, thetastart, optresults = useropt, control = cntrl)

#' Correct the marginals except for beta, which gave NA due to the constraints
for (j in 1:3) astroquad$marginals[[j]] <- marginal_posterior(astroquad, j, method = "correct")
quadruntime <- difftime(Sys.time(), tm, units = "secs")

cat("Run time for mass model quadrature:", quadruntime ,"seconds.\n")

#' Total runtime for AGHQ
optruntime + quadruntime

#' Fit with PCA-AGHQ
C <- Matrix::forceSymmetric(solve(useropt$hessian))
eigenC <- eigen(C)
lambda <- eigenC$values

data.frame(
  n = 1:length(lambda),
  tv = cumsum(lambda / sum(lambda))
) %>% ggplot(aes(x = n, y = tv)) +
  geom_point() +
  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
  annotate("text", x = 3, y = 0.875, label = "90% of total variation explained", col = "grey") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PCA dimensions included", y = "Total variation explained") +
  theme_minimal()

k_pca <- 5
s <- 1 #' Could choose this based on the Scree plot
m <- length(useropt$mode)
levels <- c(rep(k_pca, s), rep(1, m - s))
pca_grid <- mvQuad::createNIGrid(dim = m, type = "GHe", level = levels)
mvQuad::rescale(pca_grid, m = useropt$mode, C = C, dec.type = 1)

#' Check that the PCA grid has the correct number of nodes
stopifnot(nrow(mvQuad::getNodes(pca_grid)) == k_pca^s)

tm <- Sys.time()

astropcaquad <- local_aghq(ff, k_pca, thetastart, optresults = useropt, basegrid = pca_grid, adapt = FALSE, control = cntrl)

#' Correct the marginals except for beta, which gave NA due to the constraints
for (j in 1:3) astropcaquad$marginals[[j]] <- marginal_posterior(astropcaquad, j, method = "correct")
pcaquadruntime <- difftime(Sys.time(), tm, units = "secs")

cat("Run time for mass model quadrature:", pcaquadruntime ,"seconds.\n")

#' Total runtime for AGHQ
optruntime + pcaquadruntime

#' Fit with MCMC
stanmod <- tmbstan(
  ff,
  chains = 4,
  cores = 4,
  iter = 1e04,
  warmup = 1e03,
  init = nloptr_result$solution,
  seed = 48645,
  algorithm = "NUTS"
)

# Time
get_elapsed_time(stanmod)

standata <- as.data.frame(stanmod)
standata$Psi0 <- get_Psi0(standata[ ,1])
standata$gamma <- get_gamma(standata[ ,2])
standata$alpha <- get_alpha(standata[ ,3])
standata$beta <- get_beta(standata[ ,4])

#' Manual plots
parambounds <- list(
  Psi0 = c(1, 200),
  gamma = c(0.3, 0.7),
  alpha = c(3.0, 3.7),
  beta = c(-0.5, 1)
)

#' Add small buffers for stability
get_theta2_robust <- function(gamma) {
  log(-log((gamma - parambounds$gamma[1] + 1e-03) / (parambounds$gamma[2] - parambounds$gamma[1] + 1e-03)))
}

get_theta3_robust <- function(alpha) log(alpha - parambounds$alpha[1] + 1e-03)

#' Transformed PDFs
translist1 <- make_transformation(totheta = get_theta1, fromtheta = get_Psi0)
translist2 <- make_transformation(totheta = get_theta2_robust, fromtheta = get_gamma)
translist3 <- make_transformation(totheta = get_theta3_robust, fromtheta = get_alpha)
translist4 <- make_transformation(totheta = get_theta4, fromtheta = get_beta)

Psi0pdf <- compute_pdf_and_cdf(astroquad$marginals[[1]], translist1)
gammapdf <- compute_pdf_and_cdf(astroquad$marginals[[2]], translist2)
alphapdf <- compute_pdf_and_cdf(astroquad$marginals[[3]], translist3, interpolation = "polynomial")
betapdf <- compute_pdf_and_cdf(astroquad$marginals[[4]], translist4)

Psi0pcapdf <- compute_pdf_and_cdf(astropcaquad$marginals[[1]], translist1)
gammapcapdf <- compute_pdf_and_cdf(astropcaquad$marginals[[2]], translist2)
alphapcapdf <- compute_pdf_and_cdf(astropcaquad$marginals[[3]], translist3, interpolation = "polynomial")
betapcapdf <- compute_pdf_and_cdf(astropcaquad$marginals[[4]], translist4)

Psi0prior <- function(Psi0) dunif(Psi0, parambounds$Psi0[1], parambounds$Psi0[2], log = FALSE)
gammaprior <- function(gamma) dunif(gamma, parambounds$gamma[1], parambounds$gamma[2], log = FALSE)
alphaprior <- function(alpha) dgamma(alpha - parambounds$alpha[1], shape = 1, rate = 4.6, log = FALSE)
betaprior <- function(beta) dunif(beta, parambounds$beta[1], parambounds$beta[2], log = FALSE)

Psi0_postplot <- Psi0pdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = Psi0, y = after_stat(density)),
    bins = 50,
    colour = "white",
    fill = "grey"
  ) +
  geom_line(linewidth = 1) +
  geom_line(
    data = Psi0pcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    linewidth = 1,
    colour = "blue",
  ) +
  stat_function(fun = Psi0prior, linetype = "dashed") +
  labs(title = "", x = "", y = "Density") +
  scale_x_continuous(breaks = seq(10, 60, by = 5)) +
  coord_cartesian(xlim = c(24, 43)) +
  theme_classic() +
  theme(text = element_text(size = 28))

gamma_postplot <- gammapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = gamma, y = after_stat(density)),
    bins = 100,
    colour = "white",
    fill = "grey"
  ) +
  geom_line(linewidth = 1) +
  geom_line(
    data = gammapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    linewidth = 1,
    colour = "blue",
  ) +
  stat_function(fun = gammaprior, linetype = "dashed") +
  labs(title = "", x = "", y = "Density") +
  scale_x_continuous(breaks = seq(0, 0.44, by = 0.02)) +
  coord_cartesian(xlim = c(0.3, 0.4)) +
  theme_classic() +
  theme(text = element_text(size = 28))

alpha_postplot <- alphapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = alpha, y = after_stat(density)),
    bins = 2000,
    colour = "white",
    fill = "grey"
  ) +
  geom_line(linewidth = 1) +
  geom_line(
    data = alphapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    linewidth = 1,
    colour = "blue",
  ) +
  stat_function(fun = alphaprior, linetype = "dashed") +
  labs(title = "",x = "",y = "Density") +
  scale_x_continuous(breaks = seq(3, 4, by = 0.02)) +
  coord_cartesian(xlim = c(3, 3.05)) +
  theme_classic() +
  theme(text = element_text(size = 28))

beta_postplot <- betapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = beta, y = after_stat(density)),
    bins = 50,
    colour = "white",
    fill = "grey"
  ) +
  geom_line(linewidth = 1) +
  geom_line(
    data = betapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    linewidth = 1,
    colour = "blue",
  ) +
  stat_function(fun = betaprior,linetype = "dashed") +
  labs(title = "",x = "",y = "Density") +
  scale_x_continuous(breaks = seq(-0.5, 1, by = 0.1)) +
  coord_cartesian(xlim = c(-0.3, 0.4)) +
  theme_classic() +
  theme(text = element_text(size = 28))

{Psi0_postplot + gamma_postplot} / {alpha_postplot + beta_postplot}

#' KS tests
kstable <- function(quad) {
  aghqpostsamps_spline <- sample_marginal(astroquad, nrow(standata), interpolation = "spline")
  aghqpostsamps_poly <- sample_marginal(astroquad, nrow(standata), interpolation = "polynomial")

  suppressWarnings({
    kstable <- data.frame(
      Psi0 = ks.test(standata[[1]], aghqpostsamps_poly[[1]])$statistic,
      gamma = ks.test(standata[[2]], aghqpostsamps_spline[[2]])$statistic,
      alpha = ks.test(standata[[3]], aghqpostsamps_poly[[3]])$statistic,
      beta = ks.test(standata[[4]], aghqpostsamps_poly[[4]])$statistic
    )
  })

  return(kstable)
}

kstable_astroquad <- kstable(astroquad) %>%
  pivot_longer(cols = everything()) %>%
  mutate(method = "AGHQ", k = k_dense, s = NA, total_points = k^4)

kstable_astropcaquad <- kstable(astroquad) %>%
  pivot_longer(cols = everything()) %>%
  mutate(method = "AGHQ-PCA", k = k_pca, s = s, total_points = k^s)

bind_rows(kstable_astroquad, kstable_astropcaquad) %>%
  gt::gt()

bind_rows(kstable_astroquad, kstable_astropcaquad) %>%
  ggplot(aes(x = name, y = value, col = method)) +
    geom_point() +
    scale_y_continuous(limits = c(0, 0.05)) +
    scale_colour_manual(values = multi.utils::cbpalette()) +
    labs(x = "Parameter", y = "KS(method, NUTS)", col = "Method") +
    theme_minimal()
```

### 2023-04-06 src/dev_scale-grid/scale-grid.Rmd
Undo scale grid git mistake #22

--- 2023-03-31 src/dev_scale-grid/scale-grid.Rmd
@@
 end - start
+```
+
+Verifying the number of grid points used:
+
+```{r}
+stopifnot(nrow(quad$modesandhessians) == 2)
+```
+
+How are these grid points spaced across the 24 dimensions?^[Note that this section was added in later after in order to build intuitions, and doesn't flow very naturally!]
+
+```{r}
+quad$modesandhessians[, 1:24] %>%
+  pivot_longer(cols = everything()) %>%
+  mutate(id = rep(1:24, times = 2)) %>%
+  ggplot(aes(x = reorder(name, id), y = value)) +
+    geom_point(alpha = 0.8) +
+    labs(x = "", y = "Value") + 
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))
+```
+
+Interesting to note that for the Cholesky adapted grid with `levels = c(1, 1, ..., 2, 1, ..., 1)` we see variation in all of the dimensions greater than or equal to the 16th.
+This is related to the Cholesky adaption being based on a lower triangular matrix.
+To verify this, we can look at the differences between the two grid points:
+
+```{r}
+diff(as.matrix(quad$modesandhessians[, 1:24]))
+```
+
+Let's also verify that adaption is happening as we expect it to be:
+
+```{r}
+#' This is what the base grid looks like
+mvQuad::getNodes(base_grid)
+
+base_grid_copy <- rlang::duplicate(base_grid)
+
+#' Rescale it according to the mode and Hessian with Cholesky factor
+mvQuad::rescale(base_grid, m = quad$optresults$mode, C = Matrix::forceSymmetric(solve(quad$optresults$hessian)), dec.type = 2)
+
+#' Verify that the adapted grid has the same nodes and are used by aghq::aghq
+stopifnot(all(mvQuad::getNodes(base_grid) == unname(as.matrix(quad$modesandhessians[, 1:24]))))
+```
+
+How would a spectral grid behave with the same choice of `levels`?
+Notice that there is a difference for all of the 24 dimenions when `dec.type = 1` (spectral) is used:
+
+```{r}
+mvQuad::rescale(base_grid_copy, m = quad$optresults$mode, C = Matrix::forceSymmetric(solve(quad$optresults$hessian)), dec.type = 1)
+diff(mvQuad::getNodes(base_grid_copy))
 ```
@@
 
-## Application to Naomi
+## Scoping the application to Naomi
 
-Let's apply this PCA approach to the grids for the Naomi model:
+Let's see how large the PCA approach would make the grids for Naomi:
 

### 2023-04-09 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Add s-k notebook #22

@@
 abstract: |
-  **Background**
+  **Background** We have fit the Naomi model with a range of grids.
 
-  **Task**
+  **Task** How does changing the number of points per dimension $k$, or the number of dimensions of variation kept $s$ change the log normalising constant estimated?
 ---
@@
 ```{r}
-s1k2 <- readRDS("aghq-s1-k2.rds")
-s2k2 <- readRDS("aghq-s2-k2.rds")
-s3k2 <- readRDS("aghq-s3-k2.rds")
-s4k2 <- readRDS("aghq-s4-k2.rds")
-s5k2 <- readRDS("aghq-s5-k2.rds")
+s1k2 <- readRDS("depends/aghq-s1-k2.rds")
+s2k2 <- readRDS("depends/aghq-s2-k2.rds")
+s3k2 <- readRDS("depends/aghq-s3-k2.rds")
+s4k2 <- readRDS("depends/aghq-s4-k2.rds")
+s5k2 <- readRDS("depends/aghq-s5-k2.rds")
 ```
+# Table processing
+
+```{r}
+lognormconsts <- c(
+  s1k2$quad$normalized_posterior$lognormconst,
+  s2k2$quad$normalized_posterior$lognormconst,
+  s3k2$quad$normalized_posterior$lognormconst,
+  s4k2$quad$normalized_posterior$lognormconst,
+  s5k2$quad$normalized_posterior$lognormconst
+)
+
+df <- data.frame(
+  s = 1:5,
+  k = 2,
+  lognormconst = lognormconsts
+)
+
+write_csv(df, "increase-s-k.csv")
+```
+
+# Visualisation
+
+```{r}
+ggplot(df, aes(x = s, y = lognormconst)) +
+  geom_point() +
+  facet_grid(k ~ .) +
+  labs(x = "Dimensions kept of the PCA", y = "Log normalising constant") +
+  theme_minimal()
+```

### 2023-04-11 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Run k = 3 and s = 1:5 experiments, then put into comparison report (plus generalise that a little to deal with more quadratures) #22

@@
 
-We start by reading in a selection of fitted PCA-AGHQ models:
+We start by reading in a selection of fitted PCA-AGHQ models.
+We only need the quadrature part of the returned object:
 
 ```{r}
-s1k2 <- readRDS("depends/aghq-s1-k2.rds")
-s2k2 <- readRDS("depends/aghq-s2-k2.rds")
-s3k2 <- readRDS("depends/aghq-s3-k2.rds")
-s4k2 <- readRDS("depends/aghq-s4-k2.rds")
-s5k2 <- readRDS("depends/aghq-s5-k2.rds")
+read_sk <- function(s, k) {
+  assign(paste0("s", s, "k", k), readRDS(paste0("depends/aghq-s", s, "-k", k, ".rds"))$quad, envir = parent.frame())
+}
+
+read_sk(s = 1, k = 2)
+read_sk(s = 2, k = 2)
+read_sk(s = 3, k = 2)
+read_sk(s = 4, k = 2)
+read_sk(s = 5, k = 2)
+
+read_sk(s = 1, k = 3)
+read_sk(s = 2, k = 3)
+read_sk(s = 3, k = 3)
+read_sk(s = 4, k = 3)
+read_sk(s = 5, k = 3)
 ```
+
+Check that a few of the runs have the correct number of grid points:
+
+```{r}
+stopifnot(nrow(s5k2$modesandhessians) == 2^5)
+stopifnot(nrow(s3k3$modesandhessians) == 3^3)
+```
+
 # Table processing
@@
 lognormconsts <- c(
-  s1k2$quad$normalized_posterior$lognormconst,
-  s2k2$quad$normalized_posterior$lognormconst,
-  s3k2$quad$normalized_posterior$lognormconst,
-  s4k2$quad$normalized_posterior$lognormconst,
-  s5k2$quad$normalized_posterior$lognormconst
+  s1k2$normalized_posterior$lognormconst,
+  s2k2$normalized_posterior$lognormconst,
+  s3k2$normalized_posterior$lognormconst,
+  s4k2$normalized_posterior$lognormconst,
+  s5k2$normalized_posterior$lognormconst
 )

### 2023-04-11 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Add k = 3 to plot #22

@@
 ```{r}
-lognormconsts <- c(
-  s1k2$normalized_posterior$lognormconst,
-  s2k2$normalized_posterior$lognormconst,
-  s3k2$normalized_posterior$lognormconst,
-  s4k2$normalized_posterior$lognormconst,
-  s5k2$normalized_posterior$lognormconst
-)
+k2 <- list(s1k2, s2k2, s3k2, s4k2, s5k2)
+k3 <- list(s1k3, s2k3, s3k3, s4k2, s5k3)
 
 df <- data.frame(
-  s = 1:5,
-  k = 2,
-  lognormconst = lognormconsts
+  s = rep(1:5, times = 2),
+  k = rep(2:3, each = 5),
+  lognormconst = sapply(c(k2, k2), function(x) x$normalized_posterior$lognormconst)
 )
@@
   geom_point() +
-  facet_grid(k ~ .) +
+  facet_grid(. ~ k) +
   labs(x = "Dimensions kept of the PCA", y = "Log normalising constant") +

### 2023-04-11 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Add matrix of total points, and plot of time taken #22

@@
 
+# What might be possible?
+
+```{r}
+k_seq <- seq(1, 15, by = 2)
+s_seq <- 1:10
+
+M <- matrix(0, length(k_seq), length(s_seq))
+
+for (i in 1:length(k_seq)) {
+	for (j in 1:length(s_seq)) {
+		M[i, j] <- k_seq[i]^(s_seq[j])
+  }
+}
+
+# Assume anything more than a million isn't possible
+M[M > 1e6] <- NA
+```
+
 # Data import
@@
 read_sk <- function(s, k) {
-  assign(paste0("s", s, "k", k), readRDS(paste0("depends/aghq-s", s, "-k", k, ".rds"))$quad, envir = parent.frame())
+  assign(paste0("s", s, "k", k), readRDS(paste0("depends/aghq-s", s, "-k", k, ".rds"))[c("quad", "time")], envir = parent.frame())
 }
@@
 ```{r}
-stopifnot(nrow(s5k2$modesandhessians) == 2^5)
-stopifnot(nrow(s3k3$modesandhessians) == 3^3)
+stopifnot(nrow(s5k2$quad$modesandhessians) == 2^5)
+stopifnot(nrow(s3k3$quad$modesandhessians) == 3^3)
 ```
@@
   k = rep(2:3, each = 5),
-  lognormconst = sapply(c(k2, k2), function(x) x$normalized_posterior$lognormconst)
+  lognormconst = sapply(c(k2, k3), function(x) x$quad$normalized_posterior$lognormconst),
+  time = sapply(c(k2, k3), function(x) x$time)
 )
@@
 ```
+
+```{r}
+ggplot(df, aes(x = s, y = time)) +
+  geom_point() +
+  facet_grid(. ~ k) +
+  scale_y_continuous(limits = c(0, NA)) +
+  labs(x = "Dimensions kept of the PCA", y = "Time (mins)") +
+  theme_minimal()
+```

### 2023-04-13 src/docs_18-04-23-explainer/18-04-23-explainer.Rmd
Reformat slides to be better suited for lab group audience #25

@@
 ---
-title: "Deterministic Bayesian inference methods for the Naomi model"
+title: "Deterministic Bayesian inference methods for the Naomi model but make it verbose"
 author: "Adam Howes"
@@
       in_header: preamble.tex
-subtitle: HIV Inference Lab Group Meeting
+subtitle: Preseneted to some future mathematical audience
 bibliography: citations.bib
@@
 
-# Monte Carlo as it relates to numerical integration
+<!-- # Monte Carlo as it relates to numerical integration -->
 
-* Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$
-* The Monte Carlo (MC) estimate is
-$$
-\tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i)
-$$
-which is quadrature with the samples as nodes and $\omega(\vartheta_i) = 1/N$ for all $i$
-* For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC)
+<!-- * Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$ -->
+<!-- * The Monte Carlo (MC) estimate is -->
+<!-- $$ -->
+<!-- \tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i) -->
+<!-- $$ -->
+<!-- which is quadrature with the samples as nodes and $\omega(\vartheta_i) = 1/N$ for all $i$ -->
+<!-- * For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC) -->
 
-# Monte Carlo is fundamentally unsound
+<!-- # Monte Carlo is fundamentally unsound -->
 
-* "Monte Carlo ignores information" according to @o1987monte
-* Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is
-$$
-\tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right)
-$$
-* This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$
+<!-- * "Monte Carlo ignores information" according to @o1987monte -->
+<!-- * Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is -->
+<!-- $$ -->
+<!-- \tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right) -->
+<!-- $$ -->
+<!-- * This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$ -->
 
-# Application to HIV survey sampling
+<!-- # Application to HIV survey sampling -->
 
-* This is a digression but...
-* Say we're running a household survey, and sample the same individual twice
-* We didn't learn anything new about HIV by surveying them again!
-* This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar
+<!-- * This is a digression but... -->
+<!-- * Say we're running a household survey, and sample the same individual twice -->
+<!-- * We didn't learn anything new about HIV by surveying them again! -->
+<!-- * This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar -->
 
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-$\implies$ Bayesian quadrature, Bayesian survey design \\
-For some half-baked thoughts, see \texttt{athowes.github.io/fourth-gen/paper.pdf}
-\end{tcolorbox}
-\end{center}
+<!-- \begin{center} -->
+<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
+<!-- $\implies$ Bayesian quadrature, Bayesian survey design \\ -->
+<!-- For some half-baked thoughts, see \texttt{athowes.github.io/fourth-gen/paper.pdf} -->
+<!-- \end{tcolorbox} -->
+<!-- \end{center} -->
 

### 2023-04-13 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Reformat slides to be better suited for lab group audience #25

--- 2023-03-29 src/docs_18-04-23-explainer/18-04-23-explainer.Rmd
@@
 ---
+# The Naomi model
 
-# Bayesian modelling and inference
+:::::::::::::: {.columns}
 
-* As a statistical modeller, the bulk of our job is in constructing a generative model for data $y$ using parameters $\vartheta$
-* This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
-* Then, we want to compute (ideally without having to think much about it) the posterior $p(\vartheta \, | \, y)$ which is \textcolor{hilit}{just}^[I've highlighted this with sarcasm in mind: it's a difficult problem]
-$$
-p(\vartheta \, | \, y) = \frac{p(y, \vartheta)}{p(y)} = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)}
-$$
-* The central problem of Bayesian inference is doing the following integral
-$$
-p(y) = \int p(y, \vartheta) \text{d} \vartheta
-$$
+::: {.column width=.65}
 
-# Numerical integration
+* Naomi is a complicated spatio-temporal evidence synthesis model
+* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
+* Fast inference is important to allow for interactive review and development of estimates
+* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2015tmb]
 
-* If you want to integrate something deterministically, you could use numerical integration, otherwise called \textcolor{hilit}{quadrature}
-* Choose nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ and compute the weighted sum
-$$
-\tilde p(y) = \sum_{\vartheta \in \mathcal{Q}} p(y, \vartheta) \omega(\vartheta)
-$$
-* By "deterministic" I mean: if you follow the same procedure you will get the same answer
+:::
 
-# Naive quadrature example
+::: {.column width=.35}
 
-* Remember how integration is taught? (Rienmann sums)
-* Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule
-* Nodes evenly spaced through $[0, \pi]$
+```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'}
+knitr::include_graphics("naomi_hex.png")
+```
+
+:::
+
+::::::::::::::
+
+#
+
+```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'}
+knitr::include_graphics("naomi_user.png")
+```
+
+# Why do we use `TMB`
+
+1. It runs quickly
+2. It is flexible enough to be compatible with the model
+3. We don't have better options
+
+# What problem we are trying to solve
+
+* Ideally we want exact Bayesian inference: compute the posterior distribution of the parameters of the model given the data
+* Computationally this amounts to solving a difficult integral
+* We can't solve this, but we can give approximate answers
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+
+Goal: approximate this integral for Naomi better than \texttt{TMB} does. In doing so, more accurately reflect uncertainty over hyperparameters
+
+\end{tcolorbox}
+\end{center}
+
+# Two deterministic methods
+
+* We use two deterministic^[In contrast to the most famous approximate Bayesian inference method, Markov chain Monte Carlo, which is fundamentally stochastic.] methods to approximate our integral
+
+1. The Laplace approximation
+2. Quadrature
+
+# The Laplace approximation
+
+* Pretend the posterior distribution is a Gaussian! Then it's easy
+
+#
+
+```{r message=FALSE, echo=FALSE,  fig.cap="A Gamma prior with $a = 3$ and $b = 1$."}
+cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
+a <- 3
+b <- 1
+
+prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) +
+  stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) +
+  annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1)", col = cbpalette[1], size = 5) +
+  labs(x = "", y = "") +
+  theme_minimal()
+
+prior
+```
+
+#
+
+```{r message=FALSE, echo=FALSE,  fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."}
+set.seed(2)
+y <- rpois(3, lambda = 2)
+
+posterior <- prior +
+  geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) +
+  stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) +
+  annotate("text", x = 5, y = 0.25, label = "Gamma(9, 4)", col = cbpalette[2], size = 5)
+
+posterior
+```
+
+#
+
+```{r message=FALSE}
+fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE)
+
+# Here we are using numerical derivatives
+ff <- list(
+  fn = fn,
+  gr = function(x) numDeriv::grad(fn, x),
+  he = function(x) numDeriv::hessian(fn, x)
+)
+
+opt_bfgs <- aghq::optimize_theta(
+  ff, 1, control = aghq::default_control(method = "BFGS")
+)
+```
+
+# Laplace approximation
+
+```{r}
+laplace <- posterior +
+  stat_function(
+    data = data.frame(x = c(0, 10)),
+    aes(x),
+    fun = dnorm,
+    n = 500,
+    args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)),
+    col = cbpalette[3]
+  )
+```
+
+# Laplace approximation
+
+```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation in this case is good near the mode but not in the tails."}
+laplace +
+  annotate("text", x = 5.5, y = 0.35, label = "Laplace approximation", col = cbpalette[3], size = 5)
+```
+
+# Computation of the Laplace approximation
+
+* This computation was simple, and involved
+
+1. Optimising a function to obtain `opt_bfgs`
+2. Taking the mode `opt_bfgs$mode` and the Hessian at the mode `opt_bfgs$hessian`
+
+# The marginal Laplace approximation
+
+* If we don't want to pretend the whole posterior distribution is Gaussian, another option is to pretend some of its marginals are
+* This is how `TMB` works: it's up to the user to choose which parameters should be Gaussian using the `random` option
+
+# Small interlude
+
+* In spatio-temporal statistics we have data indexed by space and time
+* To model this data, we use random effects also indexed by space and time
+* There might be a lot of spatio-temporal locations
+
+# What about the hyperparameters?
+
+* `TMB` uses optimisation to find the hyperparameters which maximise the marginal Laplace approximation
+* This is the "outer" optimisation loop, where the "inner" is for computation of the Gaussian distribution
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+
+Inference for the latent field is based on a single value of the hyperparameters (the mode) -- so called empirical Bayes $\implies$ no uncertainty in the hyperparameters taken into account!
+How can you sleep at night.
+
+\end{tcolorbox}
+\end{center}
+
+# Quadrature
+
+* This brings us to our other method for solving integrals deterministically
+* Say we have a function, then quadrature has two ingredients
+
+1. Nodes: points to evaluate the function at
+2. Weights: importance of function evaluation at that point
+
+# Trapezoid rule example
+
+* Let's compute $\int_0^\pi \sin(x) \text{d}x = 2$ using quadrature
 
@@
 trapezoid_rule <- function(x, spacing) {
+  # Assumes nodes are evenly spaced
   w <- rep(spacing, length(x)) # Weights given by space between nodes
@@
 ```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."}
-library(tidyverse)
-cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
-
 finegrid <- seq(0, pi, length.out = 1000)
@@
 
-# Monte Carlo as it relates to numerical integration
+# Adaptive Gauss-Hermite quadrature
 
-* Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$
-* The Monte Carlo (MC) estimate is
-$$
-\tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i)
-$$
-which is quadrature with the samples as nodes and $\omega(\vartheta_i) = 1/N$ for all $i$
-* For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC)
-
-# Monte Carlo is fundamentally unsound
-
-* "Monte Carlo ignores information" according to @o1987monte
-* Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is
-$$
-\tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right)
-$$
-* This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$
-
-# Application to HIV survey sampling
-
-* This is a digression but...
-* Say we're running a household survey, and sample the same individual twice
-* We didn't learn anything new about HIV by surveying them again!
-* This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-$\implies$ Bayesian quadrature, Bayesian survey design \\
-For some half-baked thoughts, see \texttt{athowes.github.io/fourth-gen/paper.pdf}
-\end{tcolorbox}
-\end{center}
-
-# Curse of dimensionality
-
-* Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized, because there is an exponential increase in the volume you need to cover
-* If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$ e.g. $k = 5$ then
-
-```{r}
-5^{1:8}
-```
-
-# Latent variables and hyperparameters
-
-* Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$
-* The key part about this is that $\dim(x) = N$ is big and $\dim(\theta) = m$ is small
-
-| Names for $x$ | Names for $\theta$ |
-|---------------|--------------------|
-| Latent variables, random effects, latent field | Hyperparameters, fixed effects
-
-# Spatio-temporal statistics
-
-* There is nothing inherently special about spatio-temporal statistics, but we do often end up tackling problems with similar structures
-* We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$
-* Usually we associate parameters to spatio-temporal locations, giving us something like $\{x_{s, t}\}_{s \in \mathcal{S}, t \in \mathcal{T}}$
-
-# What's important about this?
-
-1. There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $N$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
-2.
-Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]: not IID anymore!
-
-<!-- As you get more data, the posterior approaches a Gaussian -->
-
-# Latent Gaussian models
-
-* A \textcolor{hilit}{latent Gaussian model} (LGM) [@rue2009approximate] looks along these lines:
-\begin{alignat*}{2}
-  &\text{(Observations)}     &   y &\sim p(y \, | \, x, \theta), \\
-  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x  \, | \, \mu(\theta), Q(\theta)^{-1}), \\
-  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta).
-\end{alignat*}
-* Many models are LGMs, especially in spatio-temporal statistics
-* $\dim(x) = n$, $\dim(x) = N$, $\dim(\theta) = m$
-
-# Laplace approximation
-
-* Remember that we wanted to compute
-$$
-p(y) = \int p(y, \vartheta) \text{d} \vartheta
-$$
-
-* One trick for doing this is to pretend $p(\vartheta \, | \, y)$ is Gaussian
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-\begin{itemize}
-\item Mode $\hat \vartheta = \argmax_\vartheta \log p(y, \vartheta)$
-\item Hessian $H(\hat \vartheta) = - \partial_\vartheta^2 \log p(y, \vartheta) \rvert_{\vartheta = \hat \vartheta}$
-\item Gaussian approximation $\implies \tilde p_{\texttt{G}}(\vartheta \, | \, y) = \mathcal{N}(\vartheta \, | \, \hat \vartheta, H(\hat \vartheta)^{-1})$
-\end{itemize}
-\end{tcolorbox}
-\end{center}
-
-# Example of computing the Laplace approximation
-
-* Consider the following model for $i = 1, \ldots, n$ with fixed $a$ and $b$
-$$
-y_i \sim \text{Poisson}(\lambda), \quad
-\lambda \sim \text{Gamma}(a, b).
-$$
-* It's conjugate so we directly know that
-$$
-\lambda \, | \, y \sim \text{Gamma}(a + \sum_i y_i, b + n)
-$$
-
-#
-
-```{r message=FALSE, echo=FALSE,  fig.cap="A Gamma prior with $a = 3$ and $b = 1$."}
-a <- 3
-b <- 1
-
-prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) +
-  stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) +
-  annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1)", col = cbpalette[1], size = 5) +
-  labs(x = "", y = "") +
-  theme_minimal()
-
-prior
-```
-
-#
-
-```{r message=FALSE, echo=FALSE,  fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."}
-set.seed(2)
-y <- rpois(3, lambda = 2)
-
-posterior <- prior +
-  geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) +
-  stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) +
-  annotate("text", x = 5, y = 0.25, label = "Gamma(9, 4)", col = cbpalette[2], size = 5)
-
-posterior
-```
-
-#
-
-```{r message=FALSE}
-fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE)
-
-# Here we are using numerical derivatives rather than automatic
-ff <- list(
-  fn = fn,
-  gr = function(x) numDeriv::grad(fn, x),
-  he = function(x) numDeriv::hessian(fn, x)
-)
-
-opt_bfgs <- aghq::optimize_theta(
-  ff, 1, control = aghq::default_control(method = "BFGS")
-)
-```
-
-# Laplace approximation
-
-```{r}
-laplace <- posterior +
-  stat_function(
-    data = data.frame(x = c(0, 10)),
-    aes(x),
-    fun = dnorm,
-    n = 500,
-    args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)),
-    col = cbpalette[3]
-  )
-```
-
-# Laplace approximation
-
-```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation is good near the mode but not in the tails."}
-laplace +
-  annotate("text", x = 5.5, y = 0.35, label = "Laplace approximation", col = cbpalette[3], size = 5)
-```
-
-# Laplace approximation
-
-* Now
-$$
-p(y) = \frac{p(\vartheta, y)}{p(\vartheta \, | \, y)} \approx \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)}
-$$
-and we can evaluate RHS where we would like, so let's pick the point at which the Gaussian is most accurate, which is $\hat \vartheta$
-$$
-p_{\texttt{LA}}(y) = \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)} \rvert_{\vartheta = \hat \vartheta} = (2 \pi)^{\dim(\vartheta) / 2} \det(H(\hat \vartheta))^{- 1 / 2} p(\hat \vartheta, y)
-$$
-
-# Marginal Laplace approximation
-
-* Is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian?
-* There is the Bernstein–von Mises theorem, but we can't rely on that in general. However...
-
-1.
-We just described a class of models (LGMs) where some subset of the parameters ($x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian
-2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a \textcolor{hilit}{marginal Laplace} approximation to get rid of $x$
-
-# Generalist versus specialist methods
-
-Dichotomy in statistical inference methods:
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-
-\begin{enumerate}
-  \item Generalist: works in all situations
-  \item Specialist: "exploits" properties of the problem at hand
-\end{enumerate}
-
-\end{tcolorbox}
-\end{center}
-
-We are taking approach 2!
-
-# Marginal Laplace approximation
-
-* What does this look like? Instead of assuming $p(\vartheta \, | \, y) = p(x, \theta \, | \, y)$ is Gaussian we assume $p(x \, | \, \theta, y)$ is
-$$
-\tilde p_{\texttt{G}}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x, H(\hat x))^{-1}
-$$
-where $\hat x = \hat x(\theta)$
-* Now the marginal Laplace approximation is
-$$
-p_{\texttt{LA}}(\theta, y) = \frac{p(x, \theta, y)}{\tilde p_{\texttt{G}}(x \, | \, \theta, y)} \rvert_{x = \hat x} = (2 \pi)^{N / 2} \det(H(\hat x))^{- 1 / 2} p(\hat x, \theta, y)
-$$
-
-# Integrated nested Laplace approximation
-
-* Now we can compute $p_{\texttt{LA}}(\theta, y)$ -- but what we really want is still $p(y)$ for which we need to integrate out $\theta$
-* Hopefully^[Really: hopefully] $m$ is small enough that we can now tackle this with quadrature
-* Given nodes $\mathcal{Q}$ and a weighting function $\omega$ we can compute
-$$
-p(y) \approx \sum_{\theta \in \mathcal{Q}} p_{\texttt{LA}}(\theta, y) \omega(\theta)
-$$
-* This is the famous integrated nested Laplace approximation (INLA)
-
-# Taking stock
-
-1. Bayesian inference is integration
-2. Spatial statistics has parameters $(x, \theta)$
-3. Integrate $x$ cheaply using a Gaussian assumption
-4. Try a bit harder with $\theta$ performing quadrature
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-
-Now let's apply this to a difficult problem in HIV inference!
-
-\end{tcolorbox}
-\end{center}
-
-# The Naomi model
-
-:::::::::::::: {.columns}
-
-::: {.column width=.65}
-
-* Naomi is a spatial evidence synthesis model
-* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
-* Inference for Naomi is currently conducted using `TMB` by optimising $p_{\texttt{LA}}(\theta, y)$, and has to be pretty quick to allow for interactive review and development of estimates
-
-:::
-
-::: {.column width=.35}
-
-```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'}
-knitr::include_graphics("naomi_hex.png")
-```
-
-:::
-
-::::::::::::::
-
-#
-
-```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'}
-knitr::include_graphics("naomi_user.png")
-```
-
-# Template Model Builder refresher
-
-* To use `TMB` [@kristensen2015tmb], you write your objective function $- \log p(y, x, \theta)$ in the C++ syntax
-* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks like (next slide)
-* `TMB` implements the Laplace approximation! Set `random = "x"`
-
-#
-
-```{cpp eval=FALSE}
-#include <TMB.hpp>
-
-template <class Type>
-Type objective_function<Type>::operator()() {
-  // Define data e.g.
-  DATA_VECTOR(y);
-  // Define parameters e.g.
-  PARAMETER(mu);
-  // Calculate negative log-likelihood e.g.
-  nll = Type(0.0);
-  nll -= dnorm(y, mu, 1, true).sum()
-  return(nll);
-}
-```
-
-# Why do we use `TMB`
-
-* It runs quickly, and is flexible enough to write the model
-* Another answer is that we don't have any better options
-
-| Option | Would it work? |
-|--------|--------------------|
-| MCMC | No! It is accurate (eventually) but takes too long |
-| INLA via `R-INLA`        | No! Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a LGM, and isn't compatible with `R-INLA` |
-
-# Idea
-
-* Implement an algorithm inspired by INLA which
-1. is compatible with Naomi and its intricacies
-2. uses `TMB` -- and thereby automatic differentiation -- to perform the Laplace approximation
-
-# Not a new idea!
-
-> My main comment is that several aspects of the computational machineery that is presented by Rue and his colleagues **could benefit from the use of a numerical technique known as automatic differentiation (AD)** ...
-By the use of AD one could obtain a system that is automatic from a user's perspective...
-the benefit would be a fast, flexible and easy-to-use system for doing Bayesian analysis in models with Gaussian latent variables
-
--- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate
-
-# Meanwhile in Canada...
-
-* Alex Stringer in Toronto $\to$ Waterloo had independently been thinking along similar lines, and made a lot of progress
-
-1. Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which arguably should be the default for this problem
-2. Defining a class of models called extended latent Gaussian models (ELGMs) which Naomi fits into
-
-* I will explain what both of these acronyms (AGHQ^[Amusingly similar to AGYW: adolescent girls and young women] and ELGM) mean!
-
-# Gauss-Hermite quadrature
-
-* Recall
-$$
-\int f(z) \text{d} z \approx \sum_{z \in \mathcal{Q}} f(z) \omega(z)
-$$
-* Replace $f(z)$ by $\phi(z) f(z)$ and say that $f$ is a polynomial and $\phi$ is unknown
-* Suppose that $\phi(z) = \exp(-z^2)$
-
-# Adaptation
-
-* The nodes and weights we use should probably depend on the integrand
-* Especially when the integrand is also a function of $y$, which we don't know in advance, as in $p(y, \vartheta)$
-  * i.e. how could any fixed quadrature rule be appropriate for all possible $y$?
-* Let $z \in \mathcal{Q}(m, k)$ then
-$$
-\theta(z) = \hat \theta + Lz
-$$
-where $L$ is the lower Cholesky of $H = LL^\top$
-* Other matrix decompositions can also be used e.g. spectral $H = E \Lambda E^\top = (E\Lambda^{1/2})(E\Lambda^{1/2})$
-  * Arguably this is preferable: symmetric with respect to the principle axis [@jackel2005note]
+* Gauss-Hermite quadrature is a method for picking nodes and weights based on the theory of polynomial interpolation
+* It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution
+* The "adaptive" part means the nodes and weights are changed depending on the integrand. This makes sense, especially when the integrand is also a function of the data
+* Implemented by the `aghq` package [@stringer2021implementing]
 
@@
 
-# Extended latent Gaussian models
+# The plan
 
-* In an LGM, the conditional mean depends on exactly one structured additive predictor
-\begin{align*}
-y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\
-\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
-\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
-\end{align*}
-where $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors (and can be collected into the latent field $x$)
+* Write and implement an algorithm for fast approximate Bayesian inference using the Laplace approximation and quadrature for the Naomi model
+* Use `TMB` for writing the model in C++ and implementing the Laplace approximation via automatic differentitation
+* Use a adaptive Gauss-Hermite quadrature to integrate the hyperparameters
 
-# Extended latent Gaussian models
+# Long prophecised
 
-* ELGM remove this requirement such that
-$$
-\mu_i = g_i(\eta_{\mathcal{J}_i})
-$$
-where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices
-* Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$
-  * $N_n < n$: more data points than structured additive predictors
-  * $N_n = n$: as many data points as structured additive predictors (LGM case)
-  * $N_n > n$: fewer data points than structured additive predictors
-* The $g_i$ allow for a higher degree of non-linearity in the model
+> My main comment is that several aspects of the computational machineery that is presented by Rue and his colleagues **could benefit from the use of a numerical technique known as automatic differentiation (AD)** ...
+By the use of AD one could obtain a system that is automatic from a user's perspective...
+the benefit would be a fast, flexible and easy-to-use system for doing Bayesian analysis in models with Gaussian latent variables
 
-# Why is Naomi an ELGM?
+-- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate
 
-| | Things I thought made Naomi an ELGM | Does it? |
-|-|---------------|--------------------|
-|1| ANC offset from household survey | ? |
-|2| Incidence depends on adult prevalence and coverage | ? |
-|3| ART coverage and recent infection are products | ? |
-|4| ART attendance uses a multinomial | ? |
-|5| Aggregation of finer processes | ? |
-|6| Multiple link functions | ? |
+# One challenge
 
-* I will explain each of these^[The notation may not have been introduced properly, but hopefully the gist will still make sense], and whether or not they make Naomi an ELGM, in more detail in slides to follow
-
-# ANC offset from household survey
-
-* Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators
-\begin{align*}
-\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
-\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}. 
-\end{align*}
-* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$
-* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
-
-# ANC offset from household survey
-
-* Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$, so I'm unsure which is right]
-$$
-\begin{pmatrix}
-\eta_1^\star \\
-\eta_2^\star
-\end{pmatrix}
-=
-\begin{pmatrix}
-1 & 1 \\
-0 & 1
-\end{pmatrix}
-\begin{pmatrix}
-\eta_1 \\
-\eta_2
-\end{pmatrix}
-=
-\begin{pmatrix}
-\eta_1 + \eta_2 \\
-\eta_2
-\end{pmatrix}
-$$
-* By having effects only apply to a subset of indices (`idx = c(NA, 1)` say) perhaps it can work
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-
-Warning!
-
-\begin{enumerate}
-    \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA}
-    \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM
-\end{enumerate}
-
-\end{tcolorbox}
-\end{center}
-
-# Incidence depends on adult prevalence and coverage
-
-* Linear predictor for incidence contains aggregated prevalence and coverage
-$$
-\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda.
-$$
-* Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian
-* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
-
-# ART coverage and recent infection are products
-
-* In the household survey, say, individuals who are taking ART or have been recently infected must be HIV positive
-\begin{align*}
-y^{\hat \alpha}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \alpha_{x, s, a}), \\
-y^{\hat \kappa}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \kappa_{x, s, a}).
-\end{align*}
-* $\text{logit}(\rho_{x, s, a})$ and $\text{logit}(\alpha_{x, s, a})$ are Gaussian, but we're taking a product here
-* $\kappa_{x, s, a}$ is more complicated: a function of incidence, prevalence, mean duration of recent infection and false recent ratio
-* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-Warning! These equations as writen do not appear in Naomi: instead there are aggregated versions, more about this soon.
-\end{tcolorbox}
-\end{center}
-
-# ART attendance uses the multinomial
-
-# Aggregation of finer processes
-
-* There are many instances of models being placed on aggregate quantities
-\begin{align*}
-y^{\hat \theta}_{\mathcal{I}} &\sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}), \\
-\rho_{\mathcal{I}} &= \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}.
-\end{align*}
-* Here we have $|\mathcal{I}|$ linear predictors being informed by one observation
-* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-I've previously worked on this "disaggregation regression" situation with respect to space, see \texttt{athowes/areal-comparison}
-\end{tcolorbox}
-\end{center}
-
-# Multiple link functions
-
-* The Naomi model uses both $\text{logit}$ and $\log$ (inverse) link functions
-* For LGMs there is only one $g$, whereas ELGMs allow $g_i$
-* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
-* In `R-INLA` it is possible for `y` to be a matrix where each column contains observations with shared likelihood family (and hyperparameters) and `family = c("family1", "family2", ...)`
-
-# Statistical software
-
-|           | Interface `y ~ 1 + x`  | General |
-|-----------|------------|---------|
-| Example   | `brms`     | Stan    |
-| Users     | Scientists | Statisticians |
-| Benefits  | Ease of use, sensible defaults, trust | Flexibility, development |
-| Drawbacks | Being fenced in | Barrier to entry, failing silently |
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-\texttt{athowes/multi-agyw} case-study: attempting to define $\text{AR}(1) \times \text{Besag} \times \text{IID}$ random effects in \texttt{R-INLA}
-\end{tcolorbox}
-\end{center}
-
-# The algorithm
-
-#
-
-```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the hyperparameters. Shaped like a snake for no real reason.", fig.ext = 'png', cache=TRUE}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1]
-\node (1) {$p(\theta, x, y)$};
-\node (2) [below of=1] {$\tilde p_\texttt{LA}(\theta, y)$};
-\node (3) [right of=2] {$\tilde p_\texttt{AQ}(y)$};
-\node (4) [above of=3] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$};
-\node (5) [right of=4] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$};
-\node (6) [below of=5] {Output};
-
-\draw[->] (1) to node {Laplace} (2);
-\draw[->] (2) to node {AGHQ} (3);
-\draw[->] (3) to node {Normalise} (4);
-\draw[->] (4) to node {Marginalise} (5);
-\draw[->] (5) to node {Spline} (6);
-\end{tikzpicture}
-```
-
-#
-
-```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the latent field: naive Laplace version.", fig.ext = 'png', cache=TRUE}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1]
-\node (1) {$p(\theta, x_i, x_{-i}, y)$};
-\node (2) [below of=1] {$\tilde p_\texttt{LA}(x_i, \theta, y)$};
-\node (3) [right of=2] {$\tilde p_\texttt{AQ}(x_i, y)$};
-\node (4) [above of=3] {$\tilde p_\texttt{AQ}(x_i \, | \, y)$};
-\node (5) [right of=4] {Output};
-
-\draw[->] (1) to node {Laplace} (2);
-\draw[->] (2) to node {AGHQ} (3);
-\draw[->] (3) to node {Normalise} (4);
-\draw[->] (4) to node {Spline} (5);
-\end{tikzpicture}
-```
-
-# Comparing posterior distributions
-
-* Let $\{\theta_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function
-$$
-F(\vartheta) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\theta_i \leq \vartheta}
-$$
-* Could be a hyperparameter, latent field parameter, model outputs
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-Our primary interest is in the model outputs: do any of the inference methods do a better job at computing the posterior for quantities countries use?
-\end{tcolorbox}
-\end{center}
-
-# Moments
-
-* Are the means and standard deviations the same? More generally the $t$th moment can be estimated by
-$$
-\hat{\mathbb{E}}(\theta^t) = \frac{1}{n} \sum_{i = 1}^n \theta_i^t
-$$
-* Not great if you're only looking at the first few moments $t = 1, 2$
-
-# Kolmogorov-Smirov test
-
-* Compare $F_\bullet(\vartheta)$ to $F_{\texttt{NUTS}}(\vartheta)$
-$$
-D_\bullet = \sup_\vartheta | F_{\texttt{NUTS}}(\vartheta) - F_\bullet(\vartheta)|.
-$$
-* A value $D_\bullet$ means there is at most a $(100 \cdot D_\bullet)$% difference between posterior densities
-
-#
-
-```{r, echo=FALSE, message=FALSE, fig.cap="KS test example."}
-plot(1:10)
-```
+* For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
+* Proposed solution: use principle components analysis
 

### 2023-04-13 make/_make.R

@@ -87,6 +87,7 @@ run_commit_push("docs_01-04-20-mini")
 run_commit_push("docs_01-07-21-stats-epi-group")
 run_commit_push("docs_15-11-22-seminar")
 run_commit_push("docs_18-04-23-explainer")
+run_commit_push("docs_18-04-23-lab-group")
 run_commit_push("docs_bayescomp-poster")
 run_commit_push("plot-tikz_algorithm-flowchart")
 run_commit_push("plot-tikz_simplified-naomi")

### 2023-04-13 utils/archive_to_docs.R

@@ -21,6 +21,7 @@ archive_to_docs("docs_01-04-20-mini")
 archive_to_docs("docs_01-07-21-stats-epi-group")
 archive_to_docs("docs_15-11-22-seminar")
 archive_to_docs("docs_18-04-23-explainer")
+archive_to_docs("docs_18-04-23-lab-group")
 archive_to_docs("epil")
 archive_to_docs("prev-anc-art_results")
 archive_to_docs("example_inla-replication")

### 2023-04-13 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Include k = 5 in increase-s-k notebook #22

@@
 ```{r}
-k_seq <- seq(1, 15, by = 2)
+k_seq <- 1:10
 s_seq <- 1:10
@@
 M[M > 1e6] <- NA
+
+M
 ```
@@
 We start by reading in a selection of fitted PCA-AGHQ models.
-We only need the quadrature part of the returned object:
+We only need certain parts of the fits, and as they are large we are careful here with memory.
 
@@
 
+stopifnot(nrow(s5k2$quad$modesandhessians) == 2^5)
+
+k2 <- list(s1k2, s2k2, s3k2, s4k2, s5k2)
+k2_lognormconsts <- sapply(k2, function(x) x$quad$normalized_posterior$lognormconst)
+k2_times <- sapply(k2, function(x) x$time)
+
+rm(list = c("k2", "s1k2", "s2k2", "s3k2", "s4k2", "s5k2"))
+gc()
+
 read_sk(s = 1, k = 3)
@@
 read_sk(s = 5, k = 3)
-```
 
-Check that a few of the runs have the correct number of grid points:
+stopifnot(nrow(s3k3$quad$modesandhessians) == 3^3)
 
-```{r}
-stopifnot(nrow(s5k2$quad$modesandhessians) == 2^5)
-stopifnot(nrow(s3k3$quad$modesandhessians) == 3^3)
+k3 <- list(s1k3, s2k3, s3k3, s4k3, s5k3)
+k3_lognormconsts <- sapply(k3, function(x) x$quad$normalized_posterior$lognormconst)
+k3_times <- sapply(k3, function(x) x$time)
+
+rm(list = c("k3", "s1k3", "s2k3", "s3k3", "s4k3", "s5k3"))
+gc()
+
+read_sk(s = 1, k = 5)
+read_sk(s = 2, k = 5)
+read_sk(s = 3, k = 5)
+read_sk(s = 4, k = 5)
+read_sk(s = 5, k = 5)
+
+k5 <- list(s1k5, s2k5, s3k5, s4k5, s5k5)
+k5_lognormconsts <- sapply(k5, function(x) x$quad$normalized_posterior$lognormconst)
+k5_times <- sapply(k5, function(x) x$time)
+
+rm(list = c("k5", "s1k5", "s2k5", "s3k5", "s4k5", "s5k5"))
+gc()
 ```
@@
 ```{r}
-k2 <- list(s1k2, s2k2, s3k2, s4k2, s5k2)
-k3 <- list(s1k3, s2k3, s3k3, s4k2, s5k3)
-
 df <- data.frame(
-  s = rep(1:5, times = 2),
-  k = rep(2:3, each = 5),
-  lognormconst = sapply(c(k2, k3), function(x) x$quad$normalized_posterior$lognormconst),
-  time = sapply(c(k2, k3), function(x) x$time)
+  s = rep(1:5, times = 3),
+  k = rep(c(2, 3, 5), each = 5),
+  lognormconst = c(k2_lognormconsts, k3_lognormconsts, k5_lognormconsts),
+  time = c(k2_times, k3_times, k5_times),
 )

### 2023-04-13 src/naomi-simple_compare/comparison.Rmd
Look at results comparing PCA-AGHQ k = 3, s = 8 to TMB #22

@@
 
+ks_summary %>%
+  filter(Type != "Hyper") %>%
+  summary_ks_plot("TMB", "aghq")
+
 summary_ks_plot(ks_summary, "TMB", "aghq")
@@
 
-Nodes where `TMB` beats `adam`:
+Nodes where `TMB` beats `aghq`:
 
 ```{r}
-ks_comparison(ks_summary, method1 = "adam", method2 = "TMB", 10)
+ks_comparison(ks_summary, method1 = "aghq", method2 = "TMB", 10)
 ```
@@
 ```{r}
+histogram_and_ecdf(par = "ui_anc_alpha_x", i = 18)
 histogram_and_ecdf(par = "u_alpha_xs", i = 18)
-histogram_and_ecdf(par = "u_alpha_xs", i = 29)
 histogram_and_ecdf(par = "ui_anc_alpha_x", i = 29)
@@
 
-Nodes where `adam` beats `TMB`:
+Nodes where `aghq` beats `TMB`:
 
 ```{r}
-ks_comparison(ks_summary, method1 = "TMB", method2 = "adam", 10)  
+ks_comparison(ks_summary, method1 = "TMB", method2 = "aghq", 10)  
 ```
@@
 ```{r}
-histogram_and_ecdf(par = "log_or_gamma", i = 18)
-histogram_and_ecdf(par = "ui_anc_rho_x", i = 10)
-histogram_and_ecdf(par = "log_or_gamma", i = 1)
+histogram_and_ecdf(par = "us_alpha_xs", i = 30)
+histogram_and_ecdf(par = "us_alpha_xs", i = 31)
+histogram_and_ecdf(par = "us_alpha_xs", i = 26)
 ```

### 2023-04-13 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Correct typo #22

@@
   lognormconst = c(k2_lognormconsts, k3_lognormconsts, k5_lognormconsts),
-  time = c(k2_times, k3_times, k5_times),
+  time = c(k2_times, k3_times, k5_times)
 )

### 2023-04-14 src/dev_scale-grid/scale-grid.Rmd
Almost finished with slides for group meeting #25

@@
   theme_minimal() +
-  theme(axis.text.x = element_blank())
+  theme(
+    axis.text.x = element_blank(),
+    axis.text.y = element_blank()
+  )
 
@@
 start <- Sys.time()
-quad <- fit_aghq(tmb_inputs, basegrid = base_grid)
+quad <- fit_aghq(tmb_inputs, k = 1, basegrid = base_grid)
 end <- Sys.time()
@@
 
-ggplot(tv_df, aes(x = n, y = tv)) +
+(tv_plot <- ggplot(tv_df, aes(x = n, y = tv)) +
   geom_point() +
@@
   labs(x = "PCA dimensions included", y = "Total variation explained") +
-  theme_minimal()
+  theme_minimal())
+
+ggsave("tv-plot.png", plot = tv_plot, h = 4, w = 6.25)
 ```
@@
 
-Let's start by keeping `s = 5` dimensions.
-We can create a dense GHQ grid with `k = 2` on 5 dimensions.
+Let's start by keeping `s = 8` dimensions.
+We can create a dense GHQ grid with `k = 3` on 8 dimensions.
 
 ```{r}
-s <- 5
+s <- 8
 d <- dim(E)[1]
-gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = 2) 
+gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = 3) 
 ```
 
-This grid has $2^5 = 32$ nodes:
+This grid has $3^8 = 6561$ nodes:
 
@@
 C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
-  ggplot(aes(x = Var1, y = Var2, fill = value)) +
+  ggplot(aes(x = as.factor(Var1), y = as.factor(Var2), fill = value)) +
   labs(x = "", y = "", fill = "C_s[i, j]") +
@@
     axis.text.y = element_blank(),
-    legend.position = "bottom"
+    legend.position = "none"
   )
 
-{C_plot + theme(legend.position = "bottom")} + C_s_plot
+(reduced_rank_plot <- {C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + labs(subtitle = "Reduced rank")})
+
+ggsave("reduced-rank-plot.png", reduced_rank_plot, h = 4, w = 6.25)
 ```
@@
 ```
-
-Adaptive Gauss-Hermite quadrature involves
 

### 2023-04-14 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Almost finished with slides for group meeting #25

@@
 2. It is flexible enough to be compatible with the model
-3. We don't have better options
+3. We don't have better viable options
 
@@
 
-Goal: approximate this integral for Naomi better than \texttt{TMB} does. In doing so, more accurately reflect uncertainty over hyperparameters
+Goal: approximate this integral for Naomi better than \texttt{TMB} does, without taking too long. In doing so, more accurately reflect uncertainty over hyperparameters
 
@@
 
-* Pretend the posterior distribution is a Gaussian! Then it's easy
+* If you pretend the posterior distribution is a Gaussian, computation is easy
 
@@
 
-1. Optimising a function to obtain `opt_bfgs`
-2. Taking the mode `opt_bfgs$mode` and the Hessian at the mode `opt_bfgs$hessian`
+1. Optimising a function
+2. Taking the mode and the Hessian at the mode
 
@@
 
-# Small interlude
+# Which parameters should we treat as Gaussian?
 
 * In spatio-temporal statistics we have data indexed by space and time
-* To model this data, we use random effects also indexed by space and time
-* There might be a lot of spatio-temporal locations
+* We use random effects also indexed by space and time to model this data
+* Spatio-temporal fields can be big
+* Willing to make assumptions about how things vary over spacetime
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+
+$\implies$ treat the latent field parameters as Gaussian! That's the majority of the integral done.
+
+\end{tcolorbox}
+\end{center}
 
@@
 2. Weights: importance of function evaluation at that point
+
+* You evaluate the function at the nodes, then do a weighted sum to calculate your integral
 
@@
 * It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution
-* The "adaptive" part means the nodes and weights are changed depending on the integrand. This makes sense, especially when the integrand is also a function of the data
+* The "adaptive" part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data
 * Implemented by the `aghq` package [@stringer2021implementing]
@@
 
-# The plan
+# The plan?
+
+:::::::::::::: {.columns}
+
+::: {.column width=.55}
 
@@
 * Use `TMB` for writing the model in C++ and implementing the Laplace approximation via automatic differentitation
-* Use a adaptive Gauss-Hermite quadrature to integrate the hyperparameters
+* Use adaptive Gauss-Hermite quadrature to integrate the hyperparameters
+
+:::
+
+::: {.column width=.45}
+
+```{r, echo=FALSE, fig.cap="One of the best engineering schools in Canada", out.width = '90%'}
+knitr::include_graphics("the-plan.png")
+```
+
+:::
+
+::::::::::::::
 
@@
 * For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
-* Proposed solution: use principle components analysis
+* Proposed solution use principal components analysis and keep only first $s < 24$ dimensions
+* We will use $k = 3$ points per dimension
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what this looks like for the 2D example, keeping only the first principal component, $s = 1$."}
+gg3 <- gg
+mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)
+
+add_points(plot0, gg3) +
+  labs(size = "Weight", caption = "")
+```
+
+#
+
+```{r, echo=FALSE, fig.cap="This Scree plot suggests 10 or so dimensions is enough. We use $s = 8$.", out.width = '80%'}
+knitr::include_graphics("depends/tv-plot.png")
+```
+
+#
+
+```{r, echo=FALSE, fig.cap="With 8 dimenions, the covariance matrix is accurately reproduced.", out.width = '80%'}
+knitr::include_graphics("depends/reduced-rank-plot.png")
+```
+
+# Comparison of posterior distributions
+
+* Use test data from Malawi
+* Run HMC for as long as it takes to get a good answer
+* Compare output of `TMB` with `aghq`: which is more similar to HMC?
+* There are many ways you might do this: I'll show you Kolmogorov-Smirnov test results
+
+#
+
+Worse case TMB plot demonstrating KS test
+
+#
+
+Overall KS results plot
+
+# Conclusions
 
@@
   * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
-* Let me know if you'd be up for being an early reader!
-* Code for this project is at `athowes.github.io/elgm-inf`
+* Code and notebooks for this project are at `athowes.github.io/elgm-inf`
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+
+Let me know if you'd be up for being an early reader!
+
+\end{tcolorbox}
+\end{center}
+
+# Risk group retrospective
+
+```{r, echo=FALSE, fig.cap="Available at \\texttt{athowes.github.io}", out.width = '90%'}
+knitr::include_graphics("blog-post.png")
+```
 

### 2023-04-16 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Presentation only lacking two plots from results notebook now #25

@@
 * Fast inference is important to allow for interactive review and development of estimates
-* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2015tmb]
+* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2016tmb]
 
@@
 * Write and implement an algorithm for fast approximate Bayesian inference using the Laplace approximation and quadrature for the Naomi model
-* Use `TMB` for writing the model in C++ and implementing the Laplace approximation via automatic differentitation
+* Use `TMB` for writing the model in C++, and implementing the Laplace approximation via automatic differentitation
 * Use adaptive Gauss-Hermite quadrature to integrate the hyperparameters
@@
 
-```{r, echo=FALSE, fig.cap="One of the best engineering schools in Canada", out.width = '90%'}
+```{r, echo=FALSE, fig.cap="I attended one of the best engineering schools in Canada", out.width = '90%'}
 knitr::include_graphics("the-plan.png")
@@
 * For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
-* Proposed solution use principal components analysis and keep only first $s < 24$ dimensions
-* We will use $k = 3$ points per dimension
+* One proposed solution is to use principal components analysis (PCA) and keep only the first $s < 24$ dimensions
+* We will use $k = 3$ points per dimension ($k = 2$ doesn't include the mode)
 
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what this looks like for the 2D example, keeping only the first principal component, $s = 1$."}
-gg3 <- gg
-mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)
+```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what PCA-AGHQ looks like for the 2D example, keeping only the first principal component, $s = 1$."}
+gg4 <- gg
+mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1, levels = c(3, 1))
 
-add_points(plot0, gg3) +
+add_points(plot0, gg4) +
   labs(size = "Weight", caption = "")
@@
 
-```{r, echo=FALSE, fig.cap="This Scree plot suggests 10 or so dimensions is enough. We use $s = 8$.", out.width = '80%'}
+```{r, echo=FALSE, fig.cap="This Scree plot suggests 10 or so dimensions is enough. We use $s = 8$. Avoiding long computation times is still important.", out.width = '80%'}
 knitr::include_graphics("depends/tv-plot.png")
@@
 
-* Use test data from Malawi
-* Run HMC for as long as it takes to get a good answer
+* We run HMC for as long as it takes to get a good answer (around 3 days)
 * Compare output of `TMB` with `aghq`: which is more similar to HMC?
-* There are many ways you might do this: I'll show you Kolmogorov-Smirnov test results
+* There are many ways you might compare posterior distributions
+* We'll show Kolmogorov-Smirnov test results here
 
@@
 
+* Developed an approximate Bayesian inference method more accurate than `TMB` and faster than MCMC
+* Scaled hyperparameter grid to >20 dimesions using PCA
+* Relatively straightforward implementation using the `aghq` R package
+
+# Future work
+
+* More comprehensive inference comparison, particularly of model outputs
+  * Maximum mean discrepancy
+  * Pareto-smoothed importance sampling
+* Extension to Laplace marginals via @wood2020simplified
+  * May be more accurate, but have to think about computational costs
+
 # Thanks for listening!
 
-* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work
-  * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
+* Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
 * Code and notebooks for this project are at `athowes.github.io/elgm-inf`

### 2023-04-17 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Finish slides for tomorrow #25

@@
 
-1. It runs quickly
-2. It is flexible enough to be compatible with the model
+1. It runs \textcolor{hilit}{quickly}
+2. It is \textcolor{hilit}{flexible} enough to be compatible with the model
 3. We don't have better viable options
@@
 * Computationally this amounts to solving a difficult integral
-* We can't solve this, but we can give approximate answers
+* We can't solve this, but \textcolor{hilit}{we can give approximate answers}
 
@@
 
-* If we don't want to pretend the whole posterior distribution is Gaussian, another option is to pretend some of its marginals are
+* If we don't want to pretend the whole posterior distribution is Gaussian, another option is to pretend \textcolor{hilit}{some of its marginals are}
 * This is how `TMB` works: it's up to the user to choose which parameters should be Gaussian using the `random` option
@@
 
-* Gauss-Hermite quadrature is a method for picking nodes and weights based on the theory of polynomial interpolation
+* \textcolor{hilit}{Gauss-Hermite quadrature} is a method for picking nodes and weights based on the theory of polynomial interpolation
 * It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution
-* The "adaptive" part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data
+* The \textcolor{hilit}{adaptive} part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data
 * Implemented by the `aghq` package [@stringer2021implementing]
@@
 
-```{r, echo=FALSE, fig.cap="I attended one of the best engineering schools in Canada", out.width = '90%'}
+```{r, echo=FALSE, fig.cap="One of Canada's top engineering schools (though ungraded)", out.width = '90%'}
 knitr::include_graphics("the-plan.png")
@@
 * For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
-* One proposed solution is to use principal components analysis (PCA) and keep only the first $s < 24$ dimensions
+* One proposed solution is to use \textcolor{hilit}{principal components analysis} (PCA) and keep only the first $s < 24$ dimensions
 * We will use $k = 3$ points per dimension ($k = 2$ doesn't include the mode)
@@
 ```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what PCA-AGHQ looks like for the 2D example, keeping only the first principal component, $s = 1$."}
-gg4 <- gg
-mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1, levels = c(3, 1))
+gg4 <- mvQuad::createNIGrid(2, "GHe", level = c(3, 1))
+mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)
 
@@
 
-* We run HMC for as long as it takes to get a good answer (around 3 days)
+* We run HMC for as long as it takes to get a good answer (3 days)
 * Compare output of `TMB` with `aghq`: which is more similar to HMC?
-* There are many ways you might compare posterior distributions
-* We'll show Kolmogorov-Smirnov test results here
 
@@
 
-Worse case TMB plot demonstrating KS test
+```{r, echo=FALSE, fig.cap="Histogram of sample draws from each method for one latent field parameter.
+Which of \\texttt{TMB} and \\texttt{aghq} is closer to \\texttt{tmbstan}?
+Hard to say.", out.width = '75%'}
+knitr::include_graphics("beta-alpha-histogram.png")
+```
 
@@
 
-Overall KS results plot
+```{r, echo=FALSE, fig.cap="The empirical cumulative distribution is slightly more enlightening.", out.width = '80%'}
+knitr::include_graphics("ecdf-plot.png")
+```
+
+#
+
+```{r, echo=FALSE, fig.cap="The Kolmogorov-Smirnov test statistic is the maximum difference between ECDFs.", out.width = '80%'}
+knitr::include_graphics("ecdf-diff-plot.png")
+```
+
+#
+
+```{r, echo=FALSE, fig.cap="Generating the KS test statistic for all latent field parameters, on average \\texttt{aghq} does better.", out.height = '80%'}
+knitr::include_graphics("ks-density-plot.png")
+```
 
@@
 
-* Developed an approximate Bayesian inference method more accurate than `TMB` and faster than MCMC
+* Developed an approximate Bayesian inference method more accurate than `TMB` and faster than MCMC -- we're on the Pareto frontier!
 * Scaled hyperparameter grid to >20 dimesions using PCA
-* Relatively straightforward implementation using the `aghq` R package
+* Relatively straightforward implementation using `aghq` R package, compatible with any model that has a `TMB` template
 
@@
   * Pareto-smoothed importance sampling
+  * Simulation-based calibration
 * Extension to Laplace marginals via @wood2020simplified
@@
 * Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
-* Code and notebooks for this project are at `athowes.github.io/elgm-inf`
+* Code and notebooks for this project are available at `athowes.github.io/elgm-inf`
 

### 2023-04-18 src/docs_paper/paper.Rmd
Add some Seth paper fixes

@@
   | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa.
-Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage, are jointly modelled using both household survey data and routinely reported health system data.
+Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
 Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package.
@@
 
+<!-- Seth notes that Naomi is a spatial evidence synthesis model used to produce district- level HIV epidemic indicators in sub-Saharan Africa is not the right topic sentence for the abstract -->
+<!-- Better to lead with something about the inference method? -->
+
 ```{r echo = FALSE}
@@
 
-To mount an effective public health response to the HIV epidemic, it's crucial to have accurate, timely estimates of HIV indicators at the geographic level at which health systems are planned and delivered.
-However, producing these estimates is challenging due to the shortcomings of all available data sources.
+To mount an effective public health response to the HIV epidemic, it is crucial to have accurate, timely estimates of HIV indicators at the geographic level at which health systems are planned and delivered.
+However, producing these estimates is challenging due to the limitations of all available data sources.
+<!-- Phrasing? -->
 Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they are conducted only every five years or so, with limited sample size at the district level.
 Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in more real-time but based on limited or non-representative samples of the population.
-To address challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
+To address these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
 Modelling multiple data sources jointly has many benefits, including mitigating the limitations of any single source, increasing statistical power, and prompting investigation into any conflicts of information between sources.
@@
 Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
-Although this approach is fast, we would prefer to account more fully for hyperparameter uncertainty than as is the case in an empirical Bayes framework.
+Although this approach is fast, it has the downside that within the empirical Bayes framework hyperparameter uncertainty is not accounted for in the latent field posterior.
 This has motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
@@
 The data we observe may be aggregated over indices $i$.
-Let $\mathcal{I} \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of  indicies $i$ for which an observation is reported.
+Let $\mathcal{I} \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of  indices $i$ for which an observation is reported.
 
 Let $N_i \in \mathbb{N}$ be the known, fixed population size.
-We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proporiton of indviduals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
+We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proporiton of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
 Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that
@@
 $$
-where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.
+where the mean duration of recent infection $\Omega_T$ and the proportion of long-term HIV infections misclassified as recent $\beta_T$ are strongly informed by priors for the particular survey.
 
@@
 \hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
-\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
+\mathbf{H} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
 \mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
@@
 \begin{align*}
-\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
-\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
+\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
+\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
 \end{align*}
@@
 \begin{align*}
-\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
-\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
+\hat \x_{-i}(x_i, \btheta) &= \argmax_{\x_{-i}} \log p(\y, x_i, \x_{-i}, \btheta), \\
+\mathbf{H}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
 \end{align*}
@@
 
-Using one `TMB` C++ user-template (available in the appendix), we fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using four inferential approaches.
-These were:
+We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using four inferential approaches.
+For each approach, the `TMB` C++ user-template (available in the appendix) used to specify the log-posterior was the same.
+The four approaches were:
 1. `TMB`: EB combined with a Gaussian approximation via `TMB`,
@@
 4. `NUTS`: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
-In this instance, the dimension of the latent field $N = 491$ and the dimension of the hyperparameters $m = 24$.
+In this instance, the dimension of the latent field is $N = 491$ and the dimension of the hyperparameters is $m = 24$.
 Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
@@
 Full details are provided in the appendix.
-We treat these results from NUTS as a gold-standard to which other inferential methods are compared to.
+We treat these results from NUTS as a gold-standard to which other inferential methods are compared.
 
@@
 
-We provide a flexible implementation of the algoritm, building on the`TMB` and `aghq` \textsc{R} packages.
+We provide a flexible implementation of the algorithm, building on the `TMB` and `aghq` \textsc{R} packages.
 In doing so, we hope our work enables use of deterministic inference algorithms for ELGMs in applied settings, as well as further methodological exploration of their accuracy and limitations.

### 2023-04-19 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Add ESS plot to MCMC convergence notebook
MCMC convergence analysis for new run

@@
 out <- readRDS("depends/out.rds")
-mcmc <- out$mcmc
+mcmc <- out$mcmc$stanfit
 ```
@@
 bayesplot::mcmc_neff(ratios)
+```
+
+# ESS
+
+What are the total effective sample sizes?
+
+```{r}
+#' I think that this $summary should be all of the chains grouped together
+mcmc_summary <- summary(mcmc)$summary
+
+data.frame(mcmc_summary) %>%
+  tibble::rownames_to_column("param") %>%
+  ggplot(aes(x = n_eff)) +
+    geom_histogram(alpha = 0.8) +
+    labs(x = "ESS", y = "Count")
 ```
@@
 
-We can also use energy plots (Betancourt 2017): ideally these two histograms would be the same 
+We can also use energy plots [@betancourt2017conceptual]: ideally these two histograms would be the same 
 When the histograms are quite different, it may suggest the chains are not fully exploring the tails of the target distribution.

### 2023-04-20 src/naomi-simple_compare/comparison.Rmd
Update naomi-simple_compare and run through

@@
 
-<!-- ## `beta_alpha` -->
+## `beta_alpha`
 
-<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
-<!-- histogram_and_ecdf_helper("beta_alpha") -->
-<!-- ``` -->
+```{r eval = pars_eval[["beta_rho"]]}
+histogram_and_ecdf_helper("beta_alpha")
+```
 
-<!-- ## `beta_lambda` -->
+## `beta_lambda`
 
-<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
-<!-- histogram_and_ecdf_helper("beta_lambda") -->
-<!-- ``` -->
+```{r eval = pars_eval[["beta_rho"]]}
+histogram_and_ecdf_helper("beta_lambda")
+```
 
-<!-- ## `beta_anc_rho` -->
+## `beta_anc_rho`
 
-<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
-<!-- histogram_and_ecdf("beta_anc_rho") -->
-<!-- ``` -->
+```{r eval = pars_eval[["beta_rho"]]}
+histogram_and_ecdf("beta_anc_rho")
+```
 
-<!-- ## `beta_anc_alpha` -->
+## `beta_anc_alpha`
 
-<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
-<!-- histogram_and_ecdf("beta_anc_alpha") -->
-<!-- ``` -->
+```{r eval = pars_eval[["beta_rho"]]}
+histogram_and_ecdf("beta_anc_alpha")
+```
 
-<!-- ## `logit` -->
+## `logit`
 
-<!-- ```{r eval = FALSE} -->
-<!-- lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf) -->
-<!-- ``` -->
+```{r eval = FALSE}
+lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf)
+```
 
-<!-- ## `log_sigma` -->
+## `log_sigma`
 
-<!-- ```{r eval = FALSE} -->
-<!-- lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf) -->
-<!-- ``` -->
+```{r eval = FALSE}
+lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
+```
 
-<!-- ## `u_rho_x` -->
+## `u_rho_x`
 
-<!-- ```{r eval = pars_eval[["u_rho_x"]]} -->
-<!-- histogram_and_ecdf_helper("u_rho_x") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_rho_x"]]}
+histogram_and_ecdf_helper("u_rho_x")
+```
 
-<!-- ## `u_rho_x` -->
+## `u_rho_x`
 
-<!-- ```{r eval = pars_eval[["u_rho_x"]]} -->
-<!-- histogram_and_ecdf_helper("u_rho_x") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_rho_x"]]}
+histogram_and_ecdf_helper("u_rho_x")
+```
 
-<!-- ## `us_rho_x` -->
+## `us_rho_x`
 
-<!-- ```{r eval = pars_eval[["us_rho_x"]]} -->
-<!-- histogram_and_ecdf_helper("us_rho_x") -->
-<!-- ``` -->
+```{r eval = pars_eval[["us_rho_x"]]}
+histogram_and_ecdf_helper("us_rho_x")
+```
 
-<!-- ## `u_rho_xs` -->
+## `u_rho_xs`
 
-<!-- ```{r eval = pars_eval[["u_rho_xs"]]} -->
-<!-- histogram_and_ecdf_helper("u_rho_xs") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_rho_xs"]]}
+histogram_and_ecdf_helper("u_rho_xs")
+```
 
-<!-- ## `us_rho_xs` -->
+## `us_rho_xs`
 
-<!-- ```{r eval = pars_eval[["us_rho_xs"]]} -->
-<!-- histogram_and_ecdf_helper("us_rho_xs") -->
-<!-- ``` -->
+```{r eval = pars_eval[["us_rho_xs"]]}
+histogram_and_ecdf_helper("us_rho_xs")
+```
 
-<!-- ## `u_rho_as` -->
+## `u_rho_as`
 
-<!-- ```{r eval = pars_eval[["u_rho_as"]]} -->
-<!-- histogram_and_ecdf_helper("u_rho_as") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_rho_as"]]}
+histogram_and_ecdf_helper("u_rho_as")
+```
 
-<!-- ## `u_alpha_x` -->
+## `u_alpha_x`
 
-<!-- ```{r eval = pars_eval[["u_alpha_x"]]} -->
-<!-- histogram_and_ecdf_helper("u_alpha_x") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_x"]]}
+histogram_and_ecdf_helper("u_alpha_x")
+```
 
-<!-- ## `us_alpha_x` -->
+## `us_alpha_x`
 
-<!-- ```{r eval = pars_eval[["us_alpha_x"]]} -->
-<!-- histogram_and_ecdf_helper("us_alpha_x") -->
-<!-- ``` -->
+```{r eval = pars_eval[["us_alpha_x"]]}
+histogram_and_ecdf_helper("us_alpha_x")
+```
 
-<!-- ## `u_alpha_xs` -->
+## `u_alpha_xs`
 
-<!-- ```{r eval = pars_eval[["u_alpha_xs"]]} -->
-<!-- histogram_and_ecdf_helper("u_alpha_xs") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_xs"]]}
+histogram_and_ecdf_helper("u_alpha_xs")
+```
 
-<!-- ## `us_alpha_xs` -->
+## `us_alpha_xs`
 
-<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
-<!-- histogram_and_ecdf_helper("us_alpha_xs") -->
-<!-- ``` -->
+```{r eval = pars_eval[["beta_rho"]]}
+histogram_and_ecdf_helper("us_alpha_xs")
+```
 
-<!-- ## `u_alpha_a` -->
+## `u_alpha_a`
 
-<!-- ```{r eval = pars_eval[["u_alpha_a"]]} -->
-<!-- histogram_and_ecdf_helper("u_alpha_a") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_a"]]}
+histogram_and_ecdf_helper("u_alpha_a")
+```
 
-<!-- ## `u_alpha_as` -->
+## `u_alpha_as`
 
-<!-- ```{r eval = pars_eval[["u_alpha_as"]]} -->
-<!-- histogram_and_ecdf_helper("u_alpha_as") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_as"]]}
+histogram_and_ecdf_helper("u_alpha_as")
+```
 
-<!-- ## `u_alpha_xa` -->
+## `u_alpha_xa`
 
-<!-- ```{r eval = pars_eval[["u_alpha_xa"]]} -->
-<!-- histogram_and_ecdf_helper("u_alpha_xa") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_xa"]]}
+histogram_and_ecdf_helper("u_alpha_xa")
+```
 
-<!-- ## `ui_lambda_x` -->
+## `ui_lambda_x`
 
-<!-- ```{r eval = pars_eval[["ui_lambda_x"]]} -->
-<!-- histogram_and_ecdf_helper("ui_lambda_x") -->
-<!-- ``` -->
+```{r eval = pars_eval[["ui_lambda_x"]]}
+histogram_and_ecdf_helper("ui_lambda_x")
+```
 
-<!-- ## `ui_anc_rho_x` -->
+## `ui_anc_rho_x`
 
-<!-- ```{r eval = pars_eval[["ui_anc_rho_x"]]} -->
-<!-- histogram_and_ecdf_helper("ui_anc_rho_x") -->
-<!-- ``` -->
+```{r eval = pars_eval[["ui_anc_rho_x"]]}
+histogram_and_ecdf_helper("ui_anc_rho_x")
+```
 
-<!-- ## `ui_anc_alpha_x` -->
+## `ui_anc_alpha_x`
 
-<!-- ```{r eval = pars_eval[["ui_anc_alpha_x"]]} -->
-<!-- histogram_and_ecdf_helper("ui_anc_alpha_x") -->
-<!-- ``` -->
+```{r eval = pars_eval[["ui_anc_alpha_x"]]}
+histogram_and_ecdf_helper("ui_anc_alpha_x")
+```
 
-<!-- ## `log_or_gamma` -->
+## `log_or_gamma`
 
-<!-- ```{r eval = pars_eval[["log_or_gamma"]]} -->
-<!-- histogram_and_ecdf_helper("log_or_gamma") -->
-<!-- ``` -->
+```{r eval = pars_eval[["log_or_gamma"]]}
+histogram_and_ecdf_helper("log_or_gamma")
+```
 
@@
 ```
-The most important ones (checked with Jeff) are HIV prevalence (`rho_t1_out`), ART coverage (`alpha_t1_out`), and HIV incidence (`lambda_t1_out`).
+The most important ones (this has been checked with Jeff) are HIV prevalence (`rho_t1_out`), ART coverage (`alpha_t1_out`), and HIV incidence (`lambda_t1_out`).
 
@@
 ks_helper("beta", starts_with = TRUE)
-ks_helper("beta", starts_with = TRUE, method1 = "TMB", method2 = "adam")
 ```
 
-<!-- ### `logit` -->
+### `logit`
 
-<!-- ```{r} -->
-<!-- ks_helper("logit", starts_with = TRUE) -->
-<!-- ks_helper("logit", starts_with = TRUE, method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r}
+ks_helper("logit", starts_with = TRUE)
+```
 
-<!-- ### `log_sigma` -->
+### `log_sigma`
 
-<!-- ```{r} -->
-<!-- ks_helper("log_sigma", starts_with = TRUE) -->
-<!-- ks_helper("log_sigma", starts_with = TRUE , method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r}
+ks_helper("log_sigma", starts_with = TRUE)
+```
 
-<!-- ### `u_rho_x` -->
+### `u_rho_x`
 
-<!-- ```{r eval = pars_eval[["u_rho_x"]]} -->
-<!-- ks_helper("u_rho_x") -->
-<!-- ks_helper("u_rho_x", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_rho_x"]]}
+ks_helper("u_rho_x")
+```
 
-<!-- ### `u_rho_xs` -->
+### `u_rho_xs`
 
-<!-- ```{r eval = pars_eval[["u_rho_xs"]]} -->
-<!-- ks_helper("u_rho_xs") -->
-<!-- ks_helper("u_rho_xs", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_rho_xs"]]}
+ks_helper("u_rho_xs")
+```
 
-<!-- ### `us_rho_x` -->
+### `us_rho_x`
 
-<!-- ```{r eval = pars_eval[["us_rho_x"]]} -->
-<!-- ks_helper("us_rho_x") -->
-<!-- ks_helper("us_rho_x", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["us_rho_x"]]}
+ks_helper("us_rho_x")
+```
 
-<!-- ### `us_rho_xs` -->
+### `us_rho_xs`
 
-<!-- ```{r eval = pars_eval[["us_rho_xs"]]} -->
-<!-- ks_helper("us_rho_xs") -->
-<!-- ks_helper("us_rho_xs", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["us_rho_xs"]]}
+ks_helper("us_rho_xs")
+```
 
-<!-- ### `u_rho_a` -->
+### `u_rho_a`
 
-<!-- ```{r eval = pars_eval[["u_rho_a"]]} -->
-<!-- ks_helper("u_rho_a") -->
-<!-- ks_helper("u_rho_a", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_rho_a"]]}
+ks_helper("u_rho_a")
+```
 
-<!-- ### `u_rho_as` -->
+### `u_rho_as`
 
-<!-- ```{r eval = pars_eval[["u_rho_as"]]} -->
-<!-- ks_helper("u_rho_as") -->
-<!-- ks_helper("u_rho_as", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_rho_as"]]}
+ks_helper("u_rho_as")
+```
 
-<!-- ### `u_alpha_x` -->
+### `u_alpha_x`
 
-<!-- ```{r eval = pars_eval[["u_alpha_x"]]} -->
-<!-- ks_helper("u_alpha_x") -->
-<!-- ks_helper("u_alpha_x", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_x"]]}
+ks_helper("u_alpha_x")
+```
 
-<!-- ### `u_alpha_xs` -->
+### `u_alpha_xs`
 
-<!-- ```{r eval = pars_eval[["u_alpha_xs"]]} -->
-<!-- ks_helper("u_alpha_xs") -->
-<!-- ks_helper("u_alpha_xs", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_xs"]]}
+ks_helper("u_alpha_xs")
+```
 
-<!-- ### `us_alpha_x` -->
+### `us_alpha_x`
 
-<!-- ```{r eval = pars_eval[["us_alpha_x"]]} -->
-<!-- ks_helper("us_alpha_x") -->
-<!-- ks_helper("us_alpha_x", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["us_alpha_x"]]}
+ks_helper("us_alpha_x")
+```
 
-<!-- ### `us_alpha_xs` -->
+### `us_alpha_xs`
 
-<!-- ```{r eval = pars_eval[["us_alpha_xs"]]} -->
-<!-- ks_helper("us_alpha_xs") -->
-<!-- ks_helper("us_alpha_xs", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["us_alpha_xs"]]}
+ks_helper("us_alpha_xs")
+```
 
-<!-- ### `u_alpha_a` -->
+### `u_alpha_a`
 
-<!-- ```{r eval = pars_eval[["u_alpha_a"]]} -->
-<!-- ks_helper("u_alpha_a") -->
-<!-- ks_helper("u_alpha_a", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_a"]]}
+ks_helper("u_alpha_a")
+```
 
-<!-- ### `u_alpha_as` -->
+### `u_alpha_as`
 
-<!-- ```{r eval = pars_eval[["u_alpha_as"]]} -->
-<!-- ks_helper("u_alpha_as") -->
-<!-- ks_helper("u_alpha_as", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_as"]]}
+ks_helper("u_alpha_as")
+```
 
-<!-- ### `u_alpha_xa` -->
+### `u_alpha_xa`
 
-<!-- ```{r eval = pars_eval[["u_alpha_xa"]]} -->
-<!-- ks_helper("u_alpha_xa") -->
-<!-- ks_helper("u_alpha_xa", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["u_alpha_xa"]]}
+ks_helper("u_alpha_xa")
+```
 
-<!-- ### `ui_anc_rho_x` -->
+### `ui_anc_rho_x`
 
-<!-- ```{r eval = pars_eval[["ui_anc_rho_x"]]} -->
-<!-- ks_helper("ui_anc_rho_x") -->
-<!-- ks_helper("ui_anc_rho_x", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["ui_anc_rho_x"]]}
+ks_helper("ui_anc_rho_x")
+```
 
-<!-- ### `ui_anc_alpha_x` -->
+### `ui_anc_alpha_x`
 
-<!-- ```{r eval = pars_eval[["ui_anc_alpha_x"]]} -->
-<!-- ks_helper("ui_anc_alpha_x") -->
-<!-- ks_helper("ui_anc_alpha_x", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["ui_anc_alpha_x"]]}
+ks_helper("ui_anc_alpha_x")
+```
 
-<!-- ### `log_or_gamma` -->
+### `log_or_gamma`
 
-<!-- ```{r eval = pars_eval[["log_or_gamma"]]} -->
-<!-- ks_helper("log_or_gamma") -->
-<!-- ks_helper("log_or_gamma", method1 = "TMB", method2 = "adam") -->
-<!-- ``` -->
+```{r eval = pars_eval[["log_or_gamma"]]}
+ks_helper("log_or_gamma")
+```
 
@@
 ```{r}
-ks_rho_t1_out <- to_ks_df(par = "rho_t1_out", outputs = TRUE, id = mf_out_fine$id)
-
-merge_to_mf_out_fine <- function(ks_df) {
-  ks_df %>%
+ks_df_out <- function(par) {
+  to_ks_df(par = par, outputs = TRUE, id = mf_out_fine$id) %>%
   select(-par) %>%
@@
       mutate(index = as.numeric(index))
-  ) %>%
-  select(-index)
+  )
 }
 
-ks_rho_t1_out <- merge_to_mf_out_fine(ks_rho_t1_out)
+ks_rho_t1_out <- ks_df_out(par = "rho_t1_out")
+ks_alpha_t1_out <- ks_df_out(par = "alpha_t1_out")
+ks_lambda_t1_out <- ks_df_out(par = "lambda_t1_out")
 
-ks_alpha_t1_out <- to_ks_df(par = "alpha_t1_out", outputs = TRUE, id = mf_out_fine$id) %>%
-  merge_to_mf_out_fine()
-
-ks_lambda_t1_out <- to_ks_df(par = "lambda_t1_out", outputs = TRUE, id = mf_out_fine$id) %>%
-  merge_to_mf_out_fine()
-
-ks_plot(ks_df = ks_rho_t1_out, par = "rho_t1_out")
-ks_plot(ks_df = ks_alpha_t1_out, par = "alpha_t1_out")
-ks_plot(ks_df = ks_lambda_t1_out, par = "lambda_t1_out")
+ks_plot(ks_df = ks_rho_t1_out, par = "rho_t1_out", alpha = 0.2)
+ks_plot(ks_df = ks_alpha_t1_out, par = "alpha_t1_out", alpha = 0.2)
+ks_plot(ks_df = ks_lambda_t1_out, par = "lambda_t1_out", alpha = 0.2)
 ```
@@
 
-```{r}
+Values in dark blue are the lowest KS difference for that particular parameter:
+
+```{r class.source = 'fold-hide'}
 options(dplyr.summarise.inform = FALSE)
@@
   left_join(dict, by = "Parameter")
-```
 
-Values in green are the lowest KS difference for that particular parameter:
-
-```{r}
 ks_summary %>%
@@
   gt::tab_style(
-    style = cell_fill(color = "#009E73"),
+    style = cell_fill(color = "#0072B2"),
     locations = cells_body(
@@
     gt::tab_style(
-    style = cell_fill(color = "#009E73"),
+    style = cell_fill(color = "#0072B2"),
     locations = cells_body(
@@
 
-This plot could be improved by making the red to green transition smooth, then adding legend for KS difference.
-The density plots could be filled according to KS difference.
-
 ```{r}
-summary_ks_plot <- function(ks_summary, method1, method2) {
-  ks_method1 <- paste0("KS(", method1, ", tmbstan)")
-  ks_method2 <- paste0("KS(", method2, ", tmbstan)")
-
-  xy_length <- min(1, max(ks_summary[[ks_method1]], ks_summary[[ks_method2]]) + 0.05)
-
-  x <- y <- seq(0, xy_length, length.out = 20)
-
-  gradient_base <- expand.grid(x, y) %>%
-    mutate(diff = Var1 - Var2) %>%
-    ggplot(aes(x = Var1, y = Var2)) +
-    geom_tile(aes(fill = diff), alpha = 0.7) +
-    scale_fill_gradientn(
-      colours = c("#56B4E9", "white", "#009E73"),
-      rescaler = ~ scales::rescale_mid(.x, mid = 0)
-    )
-
-  densityplot <- gradient_base +
-    stat_density_2d(data = ks_summary, aes(x = .data[[ks_method1]], y = .data[[ks_method2]], linetype = Type), col = "black") +
-    xlim(0, xy_length) +
-    ylim(0, xy_length) +
-    geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
-    scale_linetype_manual(values = c("solid", "dashed")) +
-    labs(x = ks_method1, y = ks_method2) +
-    theme_minimal() +
-    guides(fill = "none") +
-    theme(
-      legend.position = "bottom"
-    )
-
-  ks_summary[["KS difference"]] <- ks_summary[[ks_method1]] - ks_summary[[ks_method2]]
-
-  ridgeplot <- ggplot(ks_summary, aes(y = Type, x = `KS difference`)) +
-    ggridges::geom_density_ridges(alpha = 0.7, fill = NA, aes(linetype = Type)) +
-    coord_flip() +
-    scale_linetype_manual(values = c("solid", "dashed")) +
-    labs(y = "", x = paste0(ks_method1, " - ", ks_method2)) +
-    guides(linetype = FALSE) +
-    theme_minimal()
-
-  densityplot + ridgeplot
-}
-
 ks_summary %>%
   filter(Type != "Hyper") %>%
-  summary_ks_plot("TMB", "aghq")
-
-summary_ks_plot(ks_summary, "TMB", "aghq")
-summary_ks_plot(ks_summary, "TMB", "adam")
-summary_ks_plot(ks_summary, "aghq", "adam")
+  ks_plot_many("TMB", "aghq")
 
@@
     adam = mean(`KS(adam, tmbstan)`)
-  ))
-```
-Compute $p$-values, then plot expecting $\mathcal{U}(0, 1)$.
-If not conclude there is a significant difference.
-
-```{r}
-#' To write!
+  )) %>%
+  gt::gt()
 ```
@@
 ```{r}
-pdf("ks_summary_tmb_adam.pdf", h = 4, w = 6.25)
-summary_ks_plot(ks_summary, "TMB", "adam")
-dev.off()
-
 saveRDS(ks_summary_out, "ks_summary_out.rds")

### 2023-04-20 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
Create placeholder presentation for MLGH network

---
title: "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
author: "Adam Howes"
date: "June 2023"
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7
    fig_height: 3
    includes:
      in_header: preamble.tex
subtitle: Machine Learning and Global Health Network
bibliography: citations.bib
institute: Imperial College London
---

<!-- # Bayesian modelling and inference -->

<!-- * As a statistical modeller, the bulk of our job is in constructing a generative model for data $y$ using parameters $\vartheta$ -->
<!-- * This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$ -->
<!-- * Then, we want to compute (ideally without having to think much about it) the posterior $p(\vartheta \, | \, y)$ which is \textcolor{hilit}{just}^[I've highlighted this with sarcasm in mind: it's a difficult problem] -->
<!-- $$ -->
<!-- p(\vartheta \, | \, y) = \frac{p(y, \vartheta)}{p(y)} = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)} -->
<!-- $$ -->
<!-- * The central problem of Bayesian inference is doing the following integral -->
<!-- $$ -->
<!-- p(y) = \int p(y, \vartheta) \text{d} \vartheta -->
<!-- $$ -->

<!-- # Numerical integration -->

<!-- * If you want to integrate something deterministically, you could use numerical integration, otherwise called \textcolor{hilit}{quadrature} -->
<!-- * Choose nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ and compute the weighted sum -->
<!-- $$ -->
<!-- \tilde p(y) = \sum_{\vartheta \in \mathcal{Q}} p(y, \vartheta) \omega(\vartheta) -->
<!-- $$ -->
<!-- * By "deterministic" I mean: if you follow the same procedure you will get the same answer -->

<!-- # Naive quadrature example -->

<!-- * Remember how integration is taught? (Rienmann sums) -->
<!-- * Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule -->
<!-- * Nodes evenly spaced through $[0, \pi]$ -->

<!-- ```{r} -->
<!-- trapezoid_rule <- function(x, spacing) { -->
<!--   w <- rep(spacing, length(x)) # Weights given by space between nodes -->
<!--   w[1] <- w[1] / 2 # Apart from the first which is halved -->
<!--   w[length(x)] <- w[length(x)] / 2 # And the last, also halved -->
<!--   sum(w * x) # Compute the weighted sum -->
<!-- } -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."} -->
<!-- library(tidyverse) -->
<!-- cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999") -->

<!-- finegrid <- seq(0, pi, length.out = 1000) -->

<!-- plot <- data.frame(x = finegrid, y = sin(finegrid)) %>% -->
<!--   ggplot(aes(x = x, y = y)) + -->
<!--    geom_line(col = cbpalette[1]) + -->
<!--    theme_minimal() + -->
<!--    labs(x = "x", y = "sin(x)") -->

<!-- trapezoid_plot <- function(N) { -->
<!--   grid <- seq(0, pi, length.out = N) -->
<!--   int <- trapezoid_rule(x = sin(grid), spacing = grid[2] - grid[1]) -->

<!--   plot +  -->
<!--     geom_bar( -->
<!--       data = data.frame(x = grid, y = sin(grid)), -->
<!--       aes(x = x, y = y), alpha = 0.7, stat = "identity", -->
<!--       inherit.aes = FALSE, fill = cbpalette[2]) + -->
<!--     theme_minimal() + -->
<!--     labs( -->
<!--       subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"), -->
<!--       x = "x", y = "sin(x)" -->
<!--     ) -->
<!-- } -->

<!-- trapezoid_plot(N = 10) -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."} -->
<!-- trapezoid_plot(N = 30) -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."} -->
<!-- trapezoid_plot(N = 100) -->
<!-- ``` -->

<!-- <!-- # Monte Carlo as it relates to numerical integration --> -->

<!-- <!-- * Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$ --> -->
<!-- <!-- * The Monte Carlo (MC) estimate is --> -->
<!-- <!-- $$ --> -->
<!-- <!-- \tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i) --> -->
<!-- <!-- $$ --> -->
<!-- <!-- which is quadrature with the samples as nodes and $\omega(\vartheta_i) = 1/N$ for all $i$ --> -->
<!-- <!-- * For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC) --> -->

<!-- <!-- # Monte Carlo is fundamentally unsound --> -->

<!-- <!-- * "Monte Carlo ignores information" according to @o1987monte --> -->
<!-- <!-- * Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is --> -->
<!-- <!-- $$ --> -->
<!-- <!-- \tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right) --> -->
<!-- <!-- $$ --> -->
<!-- <!-- * This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$ --> -->

<!-- <!-- # Application to HIV survey sampling --> -->

<!-- <!-- * This is a digression but... --> -->
<!-- <!-- * Say we're running a household survey, and sample the same individual twice --> -->
<!-- <!-- * We didn't learn anything new about HIV by surveying them again! --> -->
<!-- <!-- * This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar --> -->

<!-- <!-- \begin{center} --> -->
<!-- <!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] --> -->
<!-- <!-- $\implies$ Bayesian quadrature, Bayesian survey design \\ --> -->
<!-- <!-- For some half-baked thoughts, see \texttt{athowes.github.io/fourth-gen/paper.pdf} --> -->
<!-- <!-- \end{tcolorbox} --> -->
<!-- <!-- \end{center} --> -->

<!-- # Curse of dimensionality -->

<!-- * Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized, because there is an exponential increase in the volume you need to cover -->
<!-- * If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$ e.g. $k = 5$ then -->

<!-- ```{r} -->
<!-- 5^{1:8} -->
<!-- ``` -->

<!-- # Latent variables and hyperparameters -->

<!-- * Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$ -->
<!-- * The key part about this is that $\dim(x) = N$ is big and $\dim(\theta) = m$ is small -->

<!-- | Names for $x$ | Names for $\theta$ | -->
<!-- |---------------|--------------------| -->
<!-- | Latent variables, random effects, latent field | Hyperparameters, fixed effects  -->

<!-- # Spatio-temporal statistics -->

<!-- * There is nothing inherently special about spatio-temporal statistics, but we do often end up tackling problems with similar structures -->
<!-- * We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$ -->
<!-- * Usually we associate parameters to spatio-temporal locations, giving us something like $\{x_{s, t}\}_{s \in \mathcal{S}, t \in \mathcal{T}}$ -->

<!-- # What's important about this? -->

<!-- 1. There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $N$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters -->
<!-- 2. Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]: not IID anymore! -->

<!-- <!-- As you get more data, the posterior approaches a Gaussian --> -->

<!-- # Latent Gaussian models -->

<!-- * A \textcolor{hilit}{latent Gaussian model} (LGM) [@rue2009approximate] looks along these lines: -->
<!-- \begin{alignat*}{2} -->
<!--   &\text{(Observations)}     &   y &\sim p(y \, | \, x, \theta), \\ -->
<!--   &\text{(Latent field)}     &   x &\sim \mathcal{N}(x  \, | \, \mu(\theta), Q(\theta)^{-1}), \\ -->
<!--   &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta). -->
<!-- \end{alignat*} -->
<!-- * Many models are LGMs, especially in spatio-temporal statistics -->
<!-- * $\dim(x) = n$, $\dim(x) = N$, $\dim(\theta) = m$ -->

<!-- # Laplace approximation -->

<!-- * Remember that we wanted to compute -->
<!-- $$ -->
<!-- p(y) = \int p(y, \vartheta) \text{d} \vartheta -->
<!-- $$ -->

<!-- * One trick for doing this is to pretend $p(\vartheta \, | \, y)$ is Gaussian -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
<!-- \begin{itemize} -->
<!-- \item Mode $\hat \vartheta = \argmax_\vartheta \log p(y, \vartheta)$ -->
<!-- \item Hessian $H(\hat \vartheta) = - \partial_\vartheta^2 \log p(y, \vartheta) \rvert_{\vartheta = \hat \vartheta}$ -->
<!-- \item Gaussian approximation $\implies \tilde p_{\texttt{G}}(\vartheta \, | \, y) = \mathcal{N}(\vartheta \, | \, \hat \vartheta, H(\hat \vartheta)^{-1})$ -->
<!-- \end{itemize} -->
<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # Example of computing the Laplace approximation -->

<!-- * Consider the following model for $i = 1, \ldots, n$ with fixed $a$ and $b$ -->
<!-- $$ -->
<!-- y_i \sim \text{Poisson}(\lambda), \quad -->
<!-- \lambda \sim \text{Gamma}(a, b). -->
<!-- $$ -->
<!-- * It's conjugate so we directly know that -->
<!-- $$ -->
<!-- \lambda \, | \, y \sim \text{Gamma}(a + \sum_i y_i, b + n) -->
<!-- $$ -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="A Gamma prior with $a = 3$ and $b = 1$."} -->
<!-- a <- 3 -->
<!-- b <- 1 -->

<!-- prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) + -->
<!--   stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) + -->
<!--   annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1)", col = cbpalette[1], size = 5) + -->
<!--   labs(x = "", y = "") + -->
<!--   theme_minimal() -->

<!-- prior -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."} -->
<!-- set.seed(2) -->
<!-- y <- rpois(3, lambda = 2) -->

<!-- posterior <- prior + -->
<!--   geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) + -->
<!--   stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) + -->
<!--   annotate("text", x = 5, y = 0.25, label = "Gamma(9, 4)", col = cbpalette[2], size = 5) -->

<!-- posterior -->
<!-- ``` -->

<!-- # -->

<!-- ```{r message=FALSE} -->
<!-- fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE) -->

<!-- # Here we are using numerical derivatives rather than automatic -->
<!-- ff <- list( -->
<!--   fn = fn, -->
<!--   gr = function(x) numDeriv::grad(fn, x), -->
<!--   he = function(x) numDeriv::hessian(fn, x) -->
<!-- ) -->

<!-- opt_bfgs <- aghq::optimize_theta( -->
<!--   ff, 1, control = aghq::default_control(method = "BFGS") -->
<!-- ) -->
<!-- ``` -->

<!-- # Laplace approximation -->

<!-- ```{r} -->
<!-- laplace <- posterior + -->
<!--   stat_function( -->
<!--     data = data.frame(x = c(0, 10)), -->
<!--     aes(x), -->
<!--     fun = dnorm, -->
<!--     n = 500, -->
<!--     args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)), -->
<!--     col = cbpalette[3] -->
<!--   ) -->
<!-- ``` -->

<!-- # Laplace approximation -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation is good near the mode but not in the tails."} -->
<!-- laplace + -->
<!--   annotate("text", x = 5.5, y = 0.35, label = "Laplace approximation", col = cbpalette[3], size = 5) -->
<!-- ``` -->

<!-- # Laplace approximation -->

<!-- * Now -->
<!-- $$ -->
<!-- p(y) = \frac{p(\vartheta, y)}{p(\vartheta \, | \, y)} \approx \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)} -->
<!-- $$ -->
<!-- and we can evaluate RHS where we would like, so let's pick the point at which the Gaussian is most accurate, which is $\hat \vartheta$ -->
<!-- $$ -->
<!-- p_{\texttt{LA}}(y) = \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)} \rvert_{\vartheta = \hat \vartheta} = (2 \pi)^{\dim(\vartheta) / 2} \det(H(\hat \vartheta))^{- 1 / 2} p(\hat \vartheta, y) -->
<!-- $$ -->

<!-- # Marginal Laplace approximation -->

<!-- * Is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian? -->
<!-- * There is the Bernstein–von Mises theorem, but we can't rely on that in general. However... -->

<!-- 1. We just described a class of models (LGMs) where some subset of the parameters ($x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian -->
<!-- 2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a \textcolor{hilit}{marginal Laplace} approximation to get rid of $x$ -->

<!-- # Generalist versus specialist methods -->

<!-- Dichotomy in statistical inference methods: -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->

<!-- \begin{enumerate} -->
<!--   \item Generalist: works in all situations -->
<!--   \item Specialist: "exploits" properties of the problem at hand -->
<!-- \end{enumerate} -->

<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- We are taking approach 2! -->

<!-- # Marginal Laplace approximation -->

<!-- * What does this look like? Instead of assuming $p(\vartheta \, | \, y) = p(x, \theta \, | \, y)$ is Gaussian we assume $p(x \, | \, \theta, y)$ is -->
<!-- $$ -->
<!-- \tilde p_{\texttt{G}}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x, H(\hat x))^{-1} -->
<!-- $$ -->
<!-- where $\hat x = \hat x(\theta)$ -->
<!-- * Now the marginal Laplace approximation is -->
<!-- $$ -->
<!-- p_{\texttt{LA}}(\theta, y) = \frac{p(x, \theta, y)}{\tilde p_{\texttt{G}}(x \, | \, \theta, y)} \rvert_{x = \hat x} = (2 \pi)^{N / 2} \det(H(\hat x))^{- 1 / 2} p(\hat x, \theta, y) -->
<!-- $$ -->

<!-- # Integrated nested Laplace approximation -->

<!-- * Now we can compute $p_{\texttt{LA}}(\theta, y)$ -- but what we really want is still $p(y)$ for which we need to integrate out $\theta$ -->
<!-- * Hopefully^[Really: hopefully] $m$ is small enough that we can now tackle this with quadrature -->
<!-- * Given nodes $\mathcal{Q}$ and a weighting function $\omega$ we can compute -->
<!-- $$ -->
<!-- p(y) \approx \sum_{\theta \in \mathcal{Q}} p_{\texttt{LA}}(\theta, y) \omega(\theta) -->
<!-- $$ -->
<!-- * This is the famous integrated nested Laplace approximation (INLA) -->

<!-- # Taking stock -->

<!-- 1. Bayesian inference is integration -->
<!-- 2. Spatial statistics has parameters $(x, \theta)$ -->
<!-- 3. Integrate $x$ cheaply using a Gaussian assumption -->
<!-- 4. Try a bit harder with $\theta$ performing quadrature -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->

<!-- Now let's apply this to a difficult problem in HIV inference! -->

<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # The Naomi model -->

<!-- :::::::::::::: {.columns} -->

<!-- ::: {.column width=.65} -->

<!-- * Naomi is a spatial evidence synthesis model -->
<!-- * Used by countries to produce HIV estimates in a yearly process supported by UNAIDS -->
<!-- * Inference for Naomi is currently conducted using `TMB` by optimising $p_{\texttt{LA}}(\theta, y)$, and has to be pretty quick to allow for interactive review and development of estimates -->

<!-- ::: -->

<!-- ::: {.column width=.35} -->

<!-- ```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'} -->
<!-- knitr::include_graphics("naomi_hex.png") -->
<!-- ``` -->

<!-- ::: -->

<!-- :::::::::::::: -->

<!-- #  -->

<!-- ```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'} -->
<!-- knitr::include_graphics("naomi_user.png") -->
<!-- ``` -->

<!-- # Template Model Builder refresher -->

<!-- * To use `TMB` [@kristensen2015tmb], you write your objective function $- \log p(y, x, \theta)$ in the C++ syntax -->
<!-- * For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks like (next slide) -->
<!-- * `TMB` implements the Laplace approximation! Set `random = "x"` -->

<!-- #  -->

<!-- ```{cpp eval=FALSE} -->
<!-- #include <TMB.hpp> -->

<!-- template <class Type> -->
<!-- Type objective_function<Type>::operator()() { -->
<!--   // Define data e.g. -->
<!--   DATA_VECTOR(y); -->
<!--   // Define parameters e.g. -->
<!--   PARAMETER(mu); -->
<!--   // Calculate negative log-likelihood e.g. -->
<!--   nll = Type(0.0); -->
<!--   nll -= dnorm(y, mu, 1, true).sum() -->
<!--   return(nll); -->
<!-- } -->
<!-- ``` -->

<!-- # Why do we use `TMB` -->

<!-- * It runs quickly, and is flexible enough to write the model -->
<!-- * Another answer is that we don't have any better options -->

<!-- | Option | Would it work? | -->
<!-- |--------|--------------------| -->
<!-- | MCMC | No! It is accurate (eventually) but takes too long | -->
<!-- | INLA via `R-INLA`        | No! Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a LGM, and isn't compatible with `R-INLA` | -->

<!-- # Idea -->

<!-- * Implement an algorithm inspired by INLA which -->
<!-- 1. is compatible with Naomi and its intricacies -->
<!-- 2. uses `TMB` -- and thereby automatic differentiation -- to perform the Laplace approximation -->

<!-- # Not a new idea! -->

<!-- > My main comment is that several aspects of the computational machineery that is presented by Rue and his colleagues **could benefit from the use of a numerical technique known as automatic differentiation (AD)** ... By the use of AD one could obtain a system that is automatic from a user's perspective... the benefit would be a fast, flexible and easy-to-use system for doing Bayesian analysis in models with Gaussian latent variables  -->

<!-- -- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate -->

<!-- # Meanwhile in Canada... -->

<!-- * Alex Stringer in Toronto $\to$ Waterloo had independently been thinking along similar lines, and made a lot of progress -->

<!-- 1. Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which arguably should be the default for this problem -->
<!-- 2. Defining a class of models called extended latent Gaussian models (ELGMs) which Naomi fits into -->

<!-- * I will explain what both of these acronyms (AGHQ^[Amusingly similar to AGYW: adolescent girls and young women] and ELGM) mean! -->

<!-- # Gauss-Hermite quadrature -->

<!-- * Recall -->
<!-- $$ -->
<!-- \int f(z) \text{d} z \approx \sum_{z \in \mathcal{Q}} f(z) \omega(z) -->
<!-- $$ -->
<!-- * Replace $f(z)$ by $\phi(z) f(z)$ and say that $f$ is a polynomial and $\phi$ is unknown -->
<!-- * Suppose that $\phi(z) = \exp(-z^2)$ -->

<!-- # Adaptation -->

<!-- * The nodes and weights we use should probably depend on the integrand -->
<!-- * Especially when the integrand is also a function of $y$, which we don't know in advance, as in $p(y, \vartheta)$ -->
<!--   * i.e. how could any fixed quadrature rule be appropriate for all possible $y$? -->
<!-- * Let $z \in \mathcal{Q}(m, k)$ then -->
<!-- $$ -->
<!-- \theta(z) = \hat \theta + Lz -->
<!-- $$ -->
<!-- where $L$ is the lower Cholesky of $H = LL^\top$ -->
<!-- * Other matrix decompositions can also be used e.g. spectral $H = E \Lambda E^\top = (E\Lambda^{1/2})(E\Lambda^{1/2})$ -->
<!--   * Arguably this is preferable: symmetric with respect to the principle axis [@jackel2005note] -->

<!-- #  -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points in two dimensions with $k = 3$."} -->
<!-- mu <- c(1, 1.5) -->
<!-- cov <- matrix(c(2, 1, 1, 1), ncol = 2) -->

<!-- obj <- function(theta) { -->
<!--   mvtnorm::dmvnorm(theta, mean = mu, sigma = cov) -->
<!-- } -->

<!-- grid <- expand.grid( -->
<!--   theta1 = seq(-2, 5, length.out = 700), -->
<!--   theta2 = seq(-2, 5, length.out = 700) -->
<!-- ) -->

<!-- ground_truth <- cbind(grid, pdf = obj(grid)) -->

<!-- plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) + -->
<!--   geom_contour(col = cbpalette[1]) + -->
<!--   coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) + -->
<!--   labs(x = "", y = "") + -->
<!--   theme_minimal() -->

<!-- gg <- mvQuad::createNIGrid(2, "GHe", 3) -->

<!-- add_points <- function(plot0, gg) { -->
<!--   plot0 + -->
<!--     geom_point( -->
<!--       data = mvQuad::getNodes(gg) %>% -->
<!--               as.data.frame() %>% -->
<!--               mutate(weights = mvQuad::getWeights(gg)),  -->
<!--       aes(x = V1, y = V2, size = weights), -->
<!--       alpha = 0.8, -->
<!--       col = cbpalette[2], -->
<!--       inherit.aes = FALSE -->
<!--     ) + -->
<!--     scale_size_continuous(range = c(1, 2)) -->
<!-- } -->

<!-- add_points(plot0, gg) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- # -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean $z + \\hat \\theta$."} -->
<!-- gg2 <- gg -->
<!-- mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2) -->

<!-- add_points(plot0, gg2) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- # -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $Lz + \\hat \\theta$."} -->
<!-- gg3 <- gg -->
<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2) -->

<!-- add_points(plot0, gg3) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- # -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2} z + \\hat \\theta$."} -->
<!-- gg3 <- gg -->
<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1) -->

<!-- add_points(plot0, gg3) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- # Extended latent Gaussian models -->

<!-- * In an LGM, the conditional mean depends on exactly one structured additive predictor -->
<!-- \begin{align*} -->
<!-- y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\ -->
<!-- \mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\ -->
<!-- \eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), -->
<!-- \end{align*} -->
<!-- where $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors (and can be collected into the latent field $x$) -->

<!-- # Extended latent Gaussian models -->

<!-- * ELGM remove this requirement such that -->
<!-- $$ -->
<!-- \mu_i = g_i(\eta_{\mathcal{J}_i}) -->
<!-- $$ -->
<!-- where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices -->
<!-- * Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$ -->
<!--   * $N_n < n$: more data points than structured additive predictors -->
<!--   * $N_n = n$: as many data points as structured additive predictors (LGM case) -->
<!--   * $N_n > n$: fewer data points than structured additive predictors -->
<!-- * The $g_i$ allow for a higher degree of non-linearity in the model -->

<!-- # Why is Naomi an ELGM? -->

<!-- | | Things I thought made Naomi an ELGM | Does it? | -->
<!-- |-|---------------|--------------------| -->
<!-- |1| ANC offset from household survey | ? | -->
<!-- |2| Incidence depends on adult prevalence and coverage | ? | -->
<!-- |3| ART coverage and recent infection are products | ? | -->
<!-- |4| ART attendance uses a multinomial | ? | -->
<!-- |5| Aggregation of finer processes | ? | -->
<!-- |6| Multiple link functions | ? | -->

<!-- * I will explain each of these^[The notation may not have been introduced properly, but hopefully the gist will still make sense], and whether or not they make Naomi an ELGM, in more detail in slides to follow -->

<!-- # ANC offset from household survey -->

<!-- * Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators -->
<!-- \begin{align*} -->
<!-- \text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\ -->
<!-- \text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}.  -->
<!-- \end{align*} -->
<!-- * Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$ -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->

<!-- # ANC offset from household survey -->

<!-- * Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$, so I'm unsure which is right] -->
<!-- $$ -->
<!-- \begin{pmatrix} -->
<!-- \eta_1^\star \\ -->
<!-- \eta_2^\star -->
<!-- \end{pmatrix} -->
<!-- = -->
<!-- \begin{pmatrix} -->
<!-- 1 & 1 \\ -->
<!-- 0 & 1 -->
<!-- \end{pmatrix} -->
<!-- \begin{pmatrix} -->
<!-- \eta_1 \\ -->
<!-- \eta_2 -->
<!-- \end{pmatrix} -->
<!-- = -->
<!-- \begin{pmatrix} -->
<!-- \eta_1 + \eta_2 \\ -->
<!-- \eta_2 -->
<!-- \end{pmatrix} -->
<!-- $$ -->
<!-- * By having effects only apply to a subset of indices (`idx = c(NA, 1)` say) perhaps it can work -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->

<!-- Warning! -->

<!-- \begin{enumerate} -->
<!--     \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA} -->
<!--     \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM -->
<!-- \end{enumerate} -->

<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # Incidence depends on adult prevalence and coverage -->

<!-- * Linear predictor for incidence contains aggregated prevalence and coverage -->
<!-- $$ -->
<!-- \log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda. -->
<!-- $$ -->
<!-- * Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->

<!-- # ART coverage and recent infection are products -->

<!-- * In the household survey, say, individuals who are taking ART or have been recently infected must be HIV positive -->
<!-- \begin{align*} -->
<!-- y^{\hat \alpha}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \alpha_{x, s, a}), \\ -->
<!-- y^{\hat \kappa}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \kappa_{x, s, a}). -->
<!-- \end{align*} -->
<!-- * $\text{logit}(\rho_{x, s, a})$ and $\text{logit}(\alpha_{x, s, a})$ are Gaussian, but we're taking a product here -->
<!-- * $\kappa_{x, s, a}$ is more complicated: a function of incidence, prevalence, mean duration of recent infection and false recent ratio -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
<!-- Warning! These equations as writen do not appear in Naomi: instead there are aggregated versions, more about this soon. -->
<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # ART attendance uses the multinomial -->

<!-- # Aggregation of finer processes -->

<!-- * There are many instances of models being placed on aggregate quantities -->
<!-- \begin{align*} -->
<!-- y^{\hat \theta}_{\mathcal{I}} &\sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}), \\ -->
<!-- \rho_{\mathcal{I}} &= \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}. -->
<!-- \end{align*} -->
<!-- * Here we have $|\mathcal{I}|$ linear predictors being informed by one observation -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
<!-- I've previously worked on this "disaggregation regression" situation with respect to space, see \texttt{athowes/areal-comparison} -->
<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # Multiple link functions -->

<!-- * The Naomi model uses both $\text{logit}$ and $\log$ (inverse) link functions -->
<!-- * For LGMs there is only one $g$, whereas ELGMs allow $g_i$ -->
<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->
<!-- * In `R-INLA` it is possible for `y` to be a matrix where each column contains observations with shared likelihood family (and hyperparameters) and `family = c("family1", "family2", ...)` -->

<!-- # Statistical software -->

<!-- |           | Interface `y ~ 1 + x`  | General | -->
<!-- |-----------|------------|---------| -->
<!-- | Example   | `brms`     | Stan    | -->
<!-- | Users     | Scientists | Statisticians | -->
<!-- | Benefits  | Ease of use, sensible defaults, trust | Flexibility, development | -->
<!-- | Drawbacks | Being fenced in | Barrier to entry, failing silently | -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
<!-- \texttt{athowes/multi-agyw} case-study: attempting to define $\text{AR}(1) \times \text{Besag} \times \text{IID}$ random effects in \texttt{R-INLA} -->
<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # The algorithm -->

<!-- # -->

<!-- ```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the hyperparameters. Shaped like a snake for no real reason.", fig.ext = 'png', cache=TRUE} -->
<!-- \usetikzlibrary{arrows} -->
<!-- \begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1] -->
<!-- \node (1) {$p(\theta, x, y)$}; -->
<!-- \node (2) [below of=1] {$\tilde p_\texttt{LA}(\theta, y)$}; -->
<!-- \node (3) [right of=2] {$\tilde p_\texttt{AQ}(y)$}; -->
<!-- \node (4) [above of=3] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$}; -->
<!-- \node (5) [right of=4] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$}; -->
<!-- \node (6) [below of=5] {Output}; -->

<!-- \draw[->] (1) to node {Laplace} (2); -->
<!-- \draw[->] (2) to node {AGHQ} (3); -->
<!-- \draw[->] (3) to node {Normalise} (4); -->
<!-- \draw[->] (4) to node {Marginalise} (5); -->
<!-- \draw[->] (5) to node {Spline} (6); -->
<!-- \end{tikzpicture} -->
<!-- ``` -->

<!-- # -->

<!-- ```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the latent field: naive Laplace version.", fig.ext = 'png', cache=TRUE} -->
<!-- \usetikzlibrary{arrows} -->
<!-- \begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1] -->
<!-- \node (1) {$p(\theta, x_i, x_{-i}, y)$}; -->
<!-- \node (2) [below of=1] {$\tilde p_\texttt{LA}(x_i, \theta, y)$}; -->
<!-- \node (3) [right of=2] {$\tilde p_\texttt{AQ}(x_i, y)$}; -->
<!-- \node (4) [above of=3] {$\tilde p_\texttt{AQ}(x_i \, | \, y)$}; -->
<!-- \node (5) [right of=4] {Output}; -->

<!-- \draw[->] (1) to node {Laplace} (2); -->
<!-- \draw[->] (2) to node {AGHQ} (3); -->
<!-- \draw[->] (3) to node {Normalise} (4); -->
<!-- \draw[->] (4) to node {Spline} (5); -->
<!-- \end{tikzpicture} -->
<!-- ``` -->

<!-- # Comparing posterior distributions -->

<!-- * Let $\{\theta_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function -->
<!-- $$ -->
<!-- F(\vartheta) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\theta_i \leq \vartheta} -->
<!-- $$ -->
<!-- * Could be a hyperparameter, latent field parameter, model outputs -->

<!-- \begin{center} -->
<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
<!-- Our primary interest is in the model outputs: do any of the inference methods do a better job at computing the posterior for quantities countries use? -->
<!-- \end{tcolorbox} -->
<!-- \end{center} -->

<!-- # Moments -->

<!-- * Are the means and standard deviations the same? More generally the $t$th moment can be estimated by -->
<!-- $$ -->
<!-- \hat{\mathbb{E}}(\theta^t) = \frac{1}{n} \sum_{i = 1}^n \theta_i^t -->
<!-- $$ -->
<!-- * Not great if you're only looking at the first few moments $t = 1, 2$ -->

<!-- # Kolmogorov-Smirov test -->

<!-- * Compare $F_\bullet(\vartheta)$ to $F_{\texttt{NUTS}}(\vartheta)$ -->
<!-- $$ -->
<!-- D_\bullet = \sup_\vartheta | F_{\texttt{NUTS}}(\vartheta) - F_\bullet(\vartheta)|. -->
<!-- $$ -->
<!-- * A value $D_\bullet$ means there is at most a $(100 \cdot D_\bullet)$% difference between posterior densities -->

<!-- # -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="KS test example."} -->
<!-- plot(1:10) -->
<!-- ``` -->

<!-- # Thanks for listening! -->

<!-- * Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work -->
<!--   * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial) -->
<!-- * Let me know if you'd be up for being an early reader! -->
<!-- * Code for this project is at `athowes.github.io/elgm-inf` -->

# References {.allowframebreaks}

### 2023-04-20 src/docs_18-04-23-explainer/script.R

@@ -1,5 +1,0 @@
-#' Uncomment and run the two line below to resume development of this script
-# orderly::orderly_develop_start("docs_18-04-23-explainer")
-# setwd("src/docs_18-04-23-explainer")
-
-rmarkdown::render("18-04-23-explainer.Rmd")

### 2023-04-20 src/docs_21-06-23-mlgh/script.R

@@ -1,0 +1,5 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("docs_21-06-23-mlgh")
+# setwd("src/docs_21-06-23-mlgh")
+
+rmarkdown::render("21-06-23-mlgh.Rmd")

### 2023-04-24 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Rerun all PCA-AGHQ using no samples for lower memory, then rerun increase-s-k notebook

@@
 read_sk(s = 5, k = 3)
+read_sk(s = 7, k = 3)
+read_sk(s = 8, k = 3)
 
@@
 
-k3 <- list(s1k3, s2k3, s3k3, s4k3, s5k3)
+k3 <- list(s1k3, s2k3, s3k3, s4k3, s5k3, s7k3, s8k3)
 k3_lognormconsts <- sapply(k3, function(x) x$quad$normalized_posterior$lognormconst)
@@
 
-rm(list = c("k3", "s1k3", "s2k3", "s3k3", "s4k3", "s5k3"))
+rm(list = c("k3", "s1k3", "s2k3", "s3k3", "s4k3", "s5k3", "s7k3", "s8k3"))
 gc()
@@
 df <- data.frame(
-  s = rep(1:5, times = 3),
-  k = rep(c(2, 3, 5), each = 5),
+  s = c(1:5, 1:5, 7, 8, 1:5),
+  k = rep(c(2, 3, 5), times = c(5, 7, 5)),
   lognormconst = c(k2_lognormconsts, k3_lognormconsts, k5_lognormconsts),

### 2023-04-24 make/_make.R

@@ -18,12 +18,8 @@ run_commit_push("explore_posterior-comparison")
 run_commit_push("naomi")
 
 #' Simplfied Naomi with TMB
-id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmb = TRUE))
-orderly::orderly_commit(id)
-
-#' Simplfied Naomi with aghq, and k = 1
-id <- orderly::orderly_run("naomi-simple_fit", parameters = list(aghq = TRUE, k = 1, ndConstruction = "product"))
-orderly::orderly_commit(id)
+id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmb = TRUE, sample = TRUE))
+orderly::orderly_commit(id) #' [x]
 
 #' Simplfied Naomi with aghq, k = 1, and Laplace marginals (a.k.a. adam, for some reason)
 id <- orderly::orderly_run("naomi-simple_fit", parameters = list(adam = TRUE))
@@ -31,17 +27,12 @@ orderly::orderly_commit(id)
 
 #' Simplfied Naomi with tmbstan, and niter = 1000
 #' Note that longer niter are not run locally, and instead use the cluster. See make/hpc_mcmc.R
-id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmbstan = TRUE, hmc_laplace = FALSE))
-orderly::orderly_commit(id)
+# id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmbstan = TRUE, hmc_laplace = FALSE))
+# orderly::orderly_commit(id)
 
 #' Simplfied Naomi with tmbstan, embedded Laplace approximation and niter = 1000
-#' Note that longer niter are not run locally, and instead use the cluster. See make/hpc_mcmc.R
-id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmbstan = TRUE, hmc_laplace = TRUE))
-orderly::orderly_commit(id)
-
-run_commit_push("naomi-simple_mcmc")
-run_commit_push("naomi-simple_compare")
-run_commit_push("naomi-simple_model-checks")
+# id <- orderly::orderly_run("naomi-simple_fit", parameters = list(tmbstan = TRUE, hmc_laplace = TRUE))
[…]

### 2023-04-24 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Use minutes for all time measurements

@@
 k2_lognormconsts <- sapply(k2, function(x) x$quad$normalized_posterior$lognormconst)
-k2_times <- sapply(k2, function(x) x$time)
+k2_times <- sapply(k2, function(x) as.numeric(x$time, units = "mins"))
 
@@
 k3_lognormconsts <- sapply(k3, function(x) x$quad$normalized_posterior$lognormconst)
-k3_times <- sapply(k3, function(x) x$time)
+k3_times <- sapply(k3, function(x) as.numeric(x$time, units = "mins"))
 
@@
 k5_lognormconsts <- sapply(k5, function(x) x$quad$normalized_posterior$lognormconst)
-k5_times <- sapply(k5, function(x) x$time)
+k5_times <- sapply(k5, function(x) as.numeric(x$time, units = "mins"))
 

### 2023-04-25 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Update placeholder for MLGH presentation

@@
 ---
+
 # The Naomi model

### 2023-04-25 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
Update placeholder for MLGH presentation

@@
 
-<!-- # Bayesian modelling and inference -->
+# The Naomi model
 
-<!-- * As a statistical modeller, the bulk of our job is in constructing a generative model for data $y$ using parameters $\vartheta$ -->
-<!-- * This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$ -->
-<!-- * Then, we want to compute (ideally without having to think much about it) the posterior $p(\vartheta \, | \, y)$ which is \textcolor{hilit}{just}^[I've highlighted this with sarcasm in mind: it's a difficult problem] -->
-<!-- $$ -->
-<!-- p(\vartheta \, | \, y) = \frac{p(y, \vartheta)}{p(y)} = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)} -->
-<!-- $$ -->
-<!-- * The central problem of Bayesian inference is doing the following integral -->
-<!-- $$ -->
-<!-- p(y) = \int p(y, \vartheta) \text{d} \vartheta -->
-<!-- $$ -->
+:::::::::::::: {.columns}
 
-<!-- # Numerical integration -->
+::: {.column width=.65}
 
-<!-- * If you want to integrate something deterministically, you could use numerical integration, otherwise called \textcolor{hilit}{quadrature} -->
-<!-- * Choose nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ and compute the weighted sum -->
-<!-- $$ -->
-<!-- \tilde p(y) = \sum_{\vartheta \in \mathcal{Q}} p(y, \vartheta) \omega(\vartheta) -->
-<!-- $$ -->
-<!-- * By "deterministic" I mean: if you follow the same procedure you will get the same answer -->
+* Naomi is a complicated spatio-temporal evidence synthesis model
+* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
+* Fast inference is important to allow for interactive review and development of estimates
+* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2016tmb]
 
-<!-- # Naive quadrature example -->
+:::
 
-<!-- * Remember how integration is taught? (Rienmann sums) -->
-<!-- * Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule -->
-<!-- * Nodes evenly spaced through $[0, \pi]$ -->
+::: {.column width=.35}
 
-<!-- ```{r} -->
-<!-- trapezoid_rule <- function(x, spacing) { -->
-<!--   w <- rep(spacing, length(x)) # Weights given by space between nodes -->
-<!--   w[1] <- w[1] / 2 # Apart from the first which is halved -->
-<!--   w[length(x)] <- w[length(x)] / 2 # And the last, also halved -->
-<!--   sum(w * x) # Compute the weighted sum -->
-<!-- } -->
-<!-- ``` -->
+```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'}
+knitr::include_graphics("naomi_hex.png")
+```
 
-<!-- # -->
+:::
 
-<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."} -->
-<!-- library(tidyverse) -->
-<!-- cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999") -->
+::::::::::::::
 
-<!-- finegrid <- seq(0, pi, length.out = 1000) -->
+#
 
-<!-- plot <- data.frame(x = finegrid, y = sin(finegrid)) %>% -->
-<!--   ggplot(aes(x = x, y = y)) + -->
-<!--    geom_line(col = cbpalette[1]) + -->
-<!--    theme_minimal() + -->
-<!--    labs(x = "x", y = "sin(x)") -->
+```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'}
+knitr::include_graphics("naomi_user.png")
+```
 
-<!-- trapezoid_plot <- function(N) { -->
-<!--   grid <- seq(0, pi, length.out = N) -->
-<!--   int <- trapezoid_rule(x = sin(grid), spacing = grid[2] - grid[1]) -->
+# Thanks for listening!
 
-<!--   plot +  -->
-<!--     geom_bar( -->
-<!--       data = data.frame(x = grid, y = sin(grid)), -->
-<!--       aes(x = x, y = y), alpha = 0.7, stat = "identity", -->
-<!--       inherit.aes = FALSE, fill = cbpalette[2]) + -->
-<!--     theme_minimal() + -->
-<!--     labs( -->
-<!--       subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"), -->
-<!--       x = "x", y = "sin(x)" -->
-<!--     ) -->
-<!-- } -->
+* Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
+* Code and notebooks for this project are available at `athowes.github.io/elgm-inf`
 
-<!-- trapezoid_plot(N = 10) -->
-<!-- ``` -->
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
 
-<!-- # -->
+Let me know if you'd be up for being an early reader!
 
-<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."} -->
-<!-- trapezoid_plot(N = 30) -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."} -->
-<!-- trapezoid_plot(N = 100) -->
-<!-- ``` -->
-
-<!-- <!-- # Monte Carlo as it relates to numerical integration --> -->
-
-<!-- <!-- * Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$ --> -->
-<!-- <!-- * The Monte Carlo (MC) estimate is --> -->
-<!-- <!-- $$ --> -->
-<!-- <!-- \tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i) --> -->
-<!-- <!-- $$ --> -->
-<!-- <!-- which is quadrature with the samples as nodes and $\omega(\vartheta_i) = 1/N$ for all $i$ --> -->
-<!-- <!-- * For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC) --> -->
-
-<!-- <!-- # Monte Carlo is fundamentally unsound --> -->
-
-<!-- <!-- * "Monte Carlo ignores information" according to @o1987monte --> -->
-<!-- <!-- * Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is --> -->
-<!-- <!-- $$ --> -->
-<!-- <!-- \tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right) --> -->
-<!-- <!-- $$ --> -->
-<!-- <!-- * This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$ --> -->
-
-<!-- <!-- # Application to HIV survey sampling --> -->
-
-<!-- <!-- * This is a digression but... --> -->
-<!-- <!-- * Say we're running a household survey, and sample the same individual twice --> -->
-<!-- <!-- * We didn't learn anything new about HIV by surveying them again! --> -->
-<!-- <!-- * This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar --> -->
-
-<!-- <!-- \begin{center} --> -->
-<!-- <!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] --> -->
-<!-- <!-- $\implies$ Bayesian quadrature, Bayesian survey design \\ --> -->
-<!-- <!-- For some half-baked thoughts, see \texttt{athowes.github.io/fourth-gen/paper.pdf} --> -->
-<!-- <!-- \end{tcolorbox} --> -->
-<!-- <!-- \end{center} --> -->
-
-<!-- # Curse of dimensionality -->
-
-<!-- * Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized, because there is an exponential increase in the volume you need to cover -->
-<!-- * If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$ e.g. $k = 5$ then -->
-
-<!-- ```{r} -->
-<!-- 5^{1:8} -->
-<!-- ``` -->
-
-<!-- # Latent variables and hyperparameters -->
-
-<!-- * Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$ -->
-<!-- * The key part about this is that $\dim(x) = N$ is big and $\dim(\theta) = m$ is small -->
-
-<!-- | Names for $x$ | Names for $\theta$ | -->
-<!-- |---------------|--------------------| -->
-<!-- | Latent variables, random effects, latent field | Hyperparameters, fixed effects  -->
-
-<!-- # Spatio-temporal statistics -->
-
-<!-- * There is nothing inherently special about spatio-temporal statistics, but we do often end up tackling problems with similar structures -->
-<!-- * We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$ -->
-<!-- * Usually we associate parameters to spatio-temporal locations, giving us something like $\{x_{s, t}\}_{s \in \mathcal{S}, t \in \mathcal{T}}$ -->
-
-<!-- # What's important about this? -->
-
-<!-- 1. There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $N$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters -->
-<!-- 2.
-Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]: not IID anymore!
--->
-
-<!-- <!-- As you get more data, the posterior approaches a Gaussian --> -->
-
-<!-- # Latent Gaussian models -->
-
-<!-- * A \textcolor{hilit}{latent Gaussian model} (LGM) [@rue2009approximate] looks along these lines: -->
-<!-- \begin{alignat*}{2} -->
-<!--   &\text{(Observations)}     &   y &\sim p(y \, | \, x, \theta), \\ -->
-<!--   &\text{(Latent field)}     &   x &\sim \mathcal{N}(x  \, | \, \mu(\theta), Q(\theta)^{-1}), \\ -->
-<!--   &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta). -->
-<!-- \end{alignat*} -->
-<!-- * Many models are LGMs, especially in spatio-temporal statistics -->
-<!-- * $\dim(x) = n$, $\dim(x) = N$, $\dim(\theta) = m$ -->
-
-<!-- # Laplace approximation -->
-
-<!-- * Remember that we wanted to compute -->
-<!-- $$ -->
-<!-- p(y) = \int p(y, \vartheta) \text{d} \vartheta -->
-<!-- $$ -->
-
-<!-- * One trick for doing this is to pretend $p(\vartheta \, | \, y)$ is Gaussian -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-<!-- \begin{itemize} -->
-<!-- \item Mode $\hat \vartheta = \argmax_\vartheta \log p(y, \vartheta)$ -->
-<!-- \item Hessian $H(\hat \vartheta) = - \partial_\vartheta^2 \log p(y, \vartheta) \rvert_{\vartheta = \hat \vartheta}$ -->
-<!-- \item Gaussian approximation $\implies \tilde p_{\texttt{G}}(\vartheta \, | \, y) = \mathcal{N}(\vartheta \, | \, \hat \vartheta, H(\hat \vartheta)^{-1})$ -->
-<!-- \end{itemize} -->
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
-<!-- # Example of computing the Laplace approximation -->
-
-<!-- * Consider the following model for $i = 1, \ldots, n$ with fixed $a$ and $b$ -->
-<!-- $$ -->
-<!-- y_i \sim \text{Poisson}(\lambda), \quad -->
-<!-- \lambda \sim \text{Gamma}(a, b). -->
-<!-- $$ -->
-<!-- * It's conjugate so we directly know that -->
-<!-- $$ -->
-<!-- \lambda \, | \, y \sim \text{Gamma}(a + \sum_i y_i, b + n) -->
-<!-- $$ -->
-
-<!-- # -->
-
-<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="A Gamma prior with $a = 3$ and $b = 1$."} -->
-<!-- a <- 3 -->
-<!-- b <- 1 -->
-
-<!-- prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) + -->
-<!--   stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) + -->
-<!--   annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1)", col = cbpalette[1], size = 5) + -->
-<!--   labs(x = "", y = "") + -->
-<!--   theme_minimal() -->
-
-<!-- prior -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."} -->
-<!-- set.seed(2) -->
-<!-- y <- rpois(3, lambda = 2) -->
-
-<!-- posterior <- prior + -->
-<!--   geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) + -->
-<!--   stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) + -->
-<!--   annotate("text", x = 5, y = 0.25, label = "Gamma(9, 4)", col = cbpalette[2], size = 5) -->
-
-<!-- posterior -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r message=FALSE} -->
-<!-- fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE) -->
-
-<!-- # Here we are using numerical derivatives rather than automatic -->
-<!-- ff <- list( -->
-<!--   fn = fn, -->
-<!--   gr = function(x) numDeriv::grad(fn, x), -->
-<!--   he = function(x) numDeriv::hessian(fn, x) -->
-<!-- ) -->
-
-<!-- opt_bfgs <- aghq::optimize_theta( -->
-<!--   ff, 1, control = aghq::default_control(method = "BFGS") -->
-<!-- ) -->
-<!-- ``` -->
-
-<!-- # Laplace approximation -->
-
-<!-- ```{r} -->
-<!-- laplace <- posterior + -->
-<!--   stat_function( -->
-<!--     data = data.frame(x = c(0, 10)), -->
-<!--     aes(x), -->
-<!--     fun = dnorm, -->
-<!--     n = 500, -->
-<!--     args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)), -->
-<!--     col = cbpalette[3] -->
-<!--   ) -->
-<!-- ``` -->
-
-<!-- # Laplace approximation -->
-
-<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation is good near the mode but not in the tails."} -->
-<!-- laplace + -->
-<!--   annotate("text", x = 5.5, y = 0.35, label = "Laplace approximation", col = cbpalette[3], size = 5) -->
-<!-- ``` -->
-
-<!-- # Laplace approximation -->
-
-<!-- * Now -->
-<!-- $$ -->
-<!-- p(y) = \frac{p(\vartheta, y)}{p(\vartheta \, | \, y)} \approx \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)} -->
-<!-- $$ -->
-<!-- and we can evaluate RHS where we would like, so let's pick the point at which the Gaussian is most accurate, which is $\hat \vartheta$ -->
-<!-- $$ -->
-<!-- p_{\texttt{LA}}(y) = \frac{p(\vartheta, y)}{p_{\texttt{G}}(\vartheta \, | \, y)} \rvert_{\vartheta = \hat \vartheta} = (2 \pi)^{\dim(\vartheta) / 2} \det(H(\hat \vartheta))^{- 1 / 2} p(\hat \vartheta, y) -->
-<!-- $$ -->
-
-<!-- # Marginal Laplace approximation -->
-
-<!-- * Is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian? -->
-<!-- * There is the Bernstein–von Mises theorem, but we can't rely on that in general. However... -->
-
-<!-- 1.
-We just described a class of models (LGMs) where some subset of the parameters ($x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian -->
-<!-- 2.
-We just talked about how big $x$ is in comparison to $\theta$!
-$\implies$ most of the work in our integral can be done using a \textcolor{hilit}{marginal Laplace} approximation to get rid of $x$ -->
-
-<!-- # Generalist versus specialist methods -->
-
-<!-- Dichotomy in statistical inference methods: -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-
-<!-- \begin{enumerate} -->
-<!--   \item Generalist: works in all situations -->
-<!--   \item Specialist: "exploits" properties of the problem at hand -->
-<!-- \end{enumerate} -->
-
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
-<!-- We are taking approach 2! -->
-
-<!-- # Marginal Laplace approximation -->
-
-<!-- * What does this look like? Instead of assuming $p(\vartheta \, | \, y) = p(x, \theta \, | \, y)$ is Gaussian we assume $p(x \, | \, \theta, y)$ is -->
-<!-- $$ -->
-<!-- \tilde p_{\texttt{G}}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x, H(\hat x))^{-1} -->
-<!-- $$ -->
-<!-- where $\hat x = \hat x(\theta)$ -->
-<!-- * Now the marginal Laplace approximation is -->
-<!-- $$ -->
-<!-- p_{\texttt{LA}}(\theta, y) = \frac{p(x, \theta, y)}{\tilde p_{\texttt{G}}(x \, | \, \theta, y)} \rvert_{x = \hat x} = (2 \pi)^{N / 2} \det(H(\hat x))^{- 1 / 2} p(\hat x, \theta, y) -->
-<!-- $$ -->
-
-<!-- # Integrated nested Laplace approximation -->
-
-<!-- * Now we can compute $p_{\texttt{LA}}(\theta, y)$ -- but what we really want is still $p(y)$ for which we need to integrate out $\theta$ -->
-<!-- * Hopefully^[Really: hopefully] $m$ is small enough that we can now tackle this with quadrature -->
-<!-- * Given nodes $\mathcal{Q}$ and a weighting function $\omega$ we can compute -->
-<!-- $$ -->
-<!-- p(y) \approx \sum_{\theta \in \mathcal{Q}} p_{\texttt{LA}}(\theta, y) \omega(\theta) -->
-<!-- $$ -->
-<!-- * This is the famous integrated nested Laplace approximation (INLA) -->
-
-<!-- # Taking stock -->
-
-<!-- 1. Bayesian inference is integration -->
-<!-- 2. Spatial statistics has parameters $(x, \theta)$ -->
-<!-- 3. Integrate $x$ cheaply using a Gaussian assumption -->
-<!-- 4. Try a bit harder with $\theta$ performing quadrature -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-
-<!-- Now let's apply this to a difficult problem in HIV inference! -->
-
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
-<!-- # The Naomi model -->
-
-<!-- :::::::::::::: {.columns} -->
-
-<!-- ::: {.column width=.65} -->
-
-<!-- * Naomi is a spatial evidence synthesis model -->
-<!-- * Used by countries to produce HIV estimates in a yearly process supported by UNAIDS -->
-<!-- * Inference for Naomi is currently conducted using `TMB` by optimising $p_{\texttt{LA}}(\theta, y)$, and has to be pretty quick to allow for interactive review and development of estimates -->
-
-<!-- ::: -->
-
-<!-- ::: {.column width=.35} -->
-
-<!-- ```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'} -->
-<!-- knitr::include_graphics("naomi_hex.png") -->
-<!-- ``` -->
-
-<!-- ::: -->
-
-<!-- :::::::::::::: -->
-
-<!-- #  -->
-
-<!-- ```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'} -->
-<!-- knitr::include_graphics("naomi_user.png") -->
-<!-- ``` -->
-
-<!-- # Template Model Builder refresher -->
-
-<!-- * To use `TMB` [@kristensen2015tmb], you write your objective function $- \log p(y, x, \theta)$ in the C++ syntax -->
-<!-- * For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks like (next slide) -->
-<!-- * `TMB` implements the Laplace approximation! Set `random = "x"` -->
-
-<!-- #  -->
-
-<!-- ```{cpp eval=FALSE} -->
-<!-- #include <TMB.hpp> -->
-
-<!-- template <class Type> -->
-<!-- Type objective_function<Type>::operator()() { -->
-<!--   // Define data e.g. -->
-<!--   DATA_VECTOR(y); -->
-<!--   // Define parameters e.g. -->
-<!--   PARAMETER(mu); -->
-<!--   // Calculate negative log-likelihood e.g. -->
-<!--   nll = Type(0.0); -->
-<!--   nll -= dnorm(y, mu, 1, true).sum() -->
-<!--   return(nll); -->
-<!-- } -->
-<!-- ``` -->
-
-<!-- # Why do we use `TMB` -->
-
-<!-- * It runs quickly, and is flexible enough to write the model -->
-<!-- * Another answer is that we don't have any better options -->
-
-<!-- | Option | Would it work? | -->
-<!-- |--------|--------------------| -->
-<!-- | MCMC | No! It is accurate (eventually) but takes too long | -->
-<!-- | INLA via `R-INLA`        | No! Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a LGM, and isn't compatible with `R-INLA` | -->
-
-<!-- # Idea -->
-
-<!-- * Implement an algorithm inspired by INLA which -->
-<!-- 1. is compatible with Naomi and its intricacies -->
-<!-- 2. uses `TMB` -- and thereby automatic differentiation -- to perform the Laplace approximation -->
-
-<!-- # Not a new idea! -->
-
-<!-- > My main comment is that several aspects of the computational machineery that is presented by Rue and his colleagues **could benefit from the use of a numerical technique known as automatic differentiation (AD)** ...
-By the use of AD one could obtain a system that is automatic from a user's perspective...
-the benefit would be a fast, flexible and easy-to-use system for doing Bayesian analysis in models with Gaussian latent variables  -->
-
-<!-- -- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate -->
-
-<!-- # Meanwhile in Canada... -->
-
-<!-- * Alex Stringer in Toronto $\to$ Waterloo had independently been thinking along similar lines, and made a lot of progress -->
-
-<!-- 1.
-Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which arguably should be the default for this problem -->
-<!-- 2. Defining a class of models called extended latent Gaussian models (ELGMs) which Naomi fits into -->
-
-<!-- * I will explain what both of these acronyms (AGHQ^[Amusingly similar to AGYW: adolescent girls and young women] and ELGM) mean! -->
-
-<!-- # Gauss-Hermite quadrature -->
-
-<!-- * Recall -->
-<!-- $$ -->
-<!-- \int f(z) \text{d} z \approx \sum_{z \in \mathcal{Q}} f(z) \omega(z) -->
-<!-- $$ -->
-<!-- * Replace $f(z)$ by $\phi(z) f(z)$ and say that $f$ is a polynomial and $\phi$ is unknown -->
-<!-- * Suppose that $\phi(z) = \exp(-z^2)$ -->
-
-<!-- # Adaptation -->
-
-<!-- * The nodes and weights we use should probably depend on the integrand -->
-<!-- * Especially when the integrand is also a function of $y$, which we don't know in advance, as in $p(y, \vartheta)$ -->
-<!--   * i.e. how could any fixed quadrature rule be appropriate for all possible $y$? -->
-<!-- * Let $z \in \mathcal{Q}(m, k)$ then -->
-<!-- $$ -->
-<!-- \theta(z) = \hat \theta + Lz -->
-<!-- $$ -->
-<!-- where $L$ is the lower Cholesky of $H = LL^\top$ -->
-<!-- * Other matrix decompositions can also be used e.g. spectral $H = E \Lambda E^\top = (E\Lambda^{1/2})(E\Lambda^{1/2})$ -->
-<!--   * Arguably this is preferable: symmetric with respect to the principle axis [@jackel2005note] -->
-
-<!-- #  -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points in two dimensions with $k = 3$."} -->
-<!-- mu <- c(1, 1.5) -->
-<!-- cov <- matrix(c(2, 1, 1, 1), ncol = 2) -->
-
-<!-- obj <- function(theta) { -->
-<!--   mvtnorm::dmvnorm(theta, mean = mu, sigma = cov) -->
-<!-- } -->
-
-<!-- grid <- expand.grid( -->
-<!--   theta1 = seq(-2, 5, length.out = 700), -->
-<!--   theta2 = seq(-2, 5, length.out = 700) -->
-<!-- ) -->
-
-<!-- ground_truth <- cbind(grid, pdf = obj(grid)) -->
-
-<!-- plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) + -->
-<!--   geom_contour(col = cbpalette[1]) + -->
-<!--   coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) + -->
-<!--   labs(x = "", y = "") + -->
-<!--   theme_minimal() -->
-
-<!-- gg <- mvQuad::createNIGrid(2, "GHe", 3) -->
-
-<!-- add_points <- function(plot0, gg) { -->
-<!--   plot0 + -->
-<!--     geom_point( -->
-<!--       data = mvQuad::getNodes(gg) %>% -->
-<!--               as.data.frame() %>% -->
-<!--               mutate(weights = mvQuad::getWeights(gg)),  -->
-<!--       aes(x = V1, y = V2, size = weights), -->
-<!--       alpha = 0.8, -->
-<!--       col = cbpalette[2], -->
-<!--       inherit.aes = FALSE -->
-<!--     ) + -->
-<!--     scale_size_continuous(range = c(1, 2)) -->
-<!-- } -->
-
-<!-- add_points(plot0, gg) + -->
-<!--   labs(size = "Weight", caption = "") -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean $z + \\hat \\theta$."} -->
-<!-- gg2 <- gg -->
-<!-- mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2) -->
-
-<!-- add_points(plot0, gg2) + -->
-<!--   labs(size = "Weight", caption = "") -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $Lz + \\hat \\theta$."} -->
-<!-- gg3 <- gg -->
-<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2) -->
-
-<!-- add_points(plot0, gg3) + -->
-<!--   labs(size = "Weight", caption = "") -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2} z + \\hat \\theta$."} -->
-<!-- gg3 <- gg -->
-<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1) -->
-
-<!-- add_points(plot0, gg3) + -->
-<!--   labs(size = "Weight", caption = "") -->
-<!-- ``` -->
-
-<!-- # Extended latent Gaussian models -->
-
-<!-- * In an LGM, the conditional mean depends on exactly one structured additive predictor -->
-<!-- \begin{align*} -->
-<!-- y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\ -->
-<!-- \mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\ -->
-<!-- \eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), -->
-<!-- \end{align*} -->
-<!-- where $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors (and can be collected into the latent field $x$) -->
-
-<!-- # Extended latent Gaussian models -->
-
-<!-- * ELGM remove this requirement such that -->
-<!-- $$ -->
-<!-- \mu_i = g_i(\eta_{\mathcal{J}_i}) -->
-<!-- $$ -->
-<!-- where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices -->
-<!-- * Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$ -->
-<!--   * $N_n < n$: more data points than structured additive predictors -->
-<!--   * $N_n = n$: as many data points as structured additive predictors (LGM case) -->
-<!--   * $N_n > n$: fewer data points than structured additive predictors -->
-<!-- * The $g_i$ allow for a higher degree of non-linearity in the model -->
-
-<!-- # Why is Naomi an ELGM? -->
-
-<!-- | | Things I thought made Naomi an ELGM | Does it? | -->
-<!-- |-|---------------|--------------------| -->
-<!-- |1| ANC offset from household survey | ? | -->
-<!-- |2| Incidence depends on adult prevalence and coverage | ? | -->
-<!-- |3| ART coverage and recent infection are products | ? | -->
-<!-- |4| ART attendance uses a multinomial | ? | -->
-<!-- |5| Aggregation of finer processes | ? | -->
-<!-- |6| Multiple link functions | ? | -->
-
-<!-- * I will explain each of these^[The notation may not have been introduced properly, but hopefully the gist will still make sense], and whether or not they make Naomi an ELGM, in more detail in slides to follow -->
-
-<!-- # ANC offset from household survey -->
-
-<!-- * Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators -->
-<!-- \begin{align*} -->
-<!-- \text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\ -->
-<!-- \text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}.  -->
-<!-- \end{align*} -->
-<!-- * Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$ -->
-<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->
-
-<!-- # ANC offset from household survey -->
-
-<!-- * Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$, so I'm unsure which is right] -->
-<!-- $$ -->
-<!-- \begin{pmatrix} -->
-<!-- \eta_1^\star \\ -->
-<!-- \eta_2^\star -->
-<!-- \end{pmatrix} -->
-<!-- = -->
-<!-- \begin{pmatrix} -->
-<!-- 1 & 1 \\ -->
-<!-- 0 & 1 -->
-<!-- \end{pmatrix} -->
-<!-- \begin{pmatrix} -->
-<!-- \eta_1 \\ -->
-<!-- \eta_2 -->
-<!-- \end{pmatrix} -->
-<!-- = -->
-<!-- \begin{pmatrix} -->
-<!-- \eta_1 + \eta_2 \\ -->
-<!-- \eta_2 -->
-<!-- \end{pmatrix} -->
-<!-- $$ -->
-<!-- * By having effects only apply to a subset of indices (`idx = c(NA, 1)` say) perhaps it can work -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-
-<!-- Warning! -->
-
-<!-- \begin{enumerate} -->
-<!--     \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA} -->
-<!--     \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM -->
-<!-- \end{enumerate} -->
-
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
-<!-- # Incidence depends on adult prevalence and coverage -->
-
-<!-- * Linear predictor for incidence contains aggregated prevalence and coverage -->
-<!-- $$ -->
-<!-- \log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda.
--->
-<!-- $$ -->
-<!-- * Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian -->
-<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->
-
-<!-- # ART coverage and recent infection are products -->
-
-<!-- * In the household survey, say, individuals who are taking ART or have been recently infected must be HIV positive -->
-<!-- \begin{align*} -->
-<!-- y^{\hat \alpha}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \alpha_{x, s, a}), \\ -->
-<!-- y^{\hat \kappa}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \kappa_{x, s, a}). -->
-<!-- \end{align*} -->
-<!-- * $\text{logit}(\rho_{x, s, a})$ and $\text{logit}(\alpha_{x, s, a})$ are Gaussian, but we're taking a product here -->
-<!-- * $\kappa_{x, s, a}$ is more complicated: a function of incidence, prevalence, mean duration of recent infection and false recent ratio -->
-<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-<!-- Warning! These equations as writen do not appear in Naomi: instead there are aggregated versions, more about this soon. -->
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
-<!-- # ART attendance uses the multinomial -->
-
-<!-- # Aggregation of finer processes -->
-
-<!-- * There are many instances of models being placed on aggregate quantities -->
-<!-- \begin{align*} -->
-<!-- y^{\hat \theta}_{\mathcal{I}} &\sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}), \\ -->
-<!-- \rho_{\mathcal{I}} &= \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}. -->
-<!-- \end{align*} -->
-<!-- * Here we have $|\mathcal{I}|$ linear predictors being informed by one observation -->
-<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-<!-- I've previously worked on this "disaggregation regression" situation with respect to space, see \texttt{athowes/areal-comparison} -->
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
-<!-- # Multiple link functions -->
-
-<!-- * The Naomi model uses both $\text{logit}$ and $\log$ (inverse) link functions -->
-<!-- * For LGMs there is only one $g$, whereas ELGMs allow $g_i$ -->
-<!-- * Conclusion: \textcolor{hilit}{does} make Naomi an ELGM -->
-<!-- * In `R-INLA` it is possible for `y` to be a matrix where each column contains observations with shared likelihood family (and hyperparameters) and `family = c("family1", "family2", ...)` -->
-
-<!-- # Statistical software -->
-
-<!-- |           | Interface `y ~ 1 + x`  | General | -->
-<!-- |-----------|------------|---------| -->
-<!-- | Example   | `brms`     | Stan    | -->
-<!-- | Users     | Scientists | Statisticians | -->
-<!-- | Benefits  | Ease of use, sensible defaults, trust | Flexibility, development | -->
-<!-- | Drawbacks | Being fenced in | Barrier to entry, failing silently | -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-<!-- \texttt{athowes/multi-agyw} case-study: attempting to define $\text{AR}(1) \times \text{Besag} \times \text{IID}$ random effects in \texttt{R-INLA} -->
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
-<!-- # The algorithm -->
-
-<!-- # -->
-
-<!-- ```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the hyperparameters. Shaped like a snake for no real reason.", fig.ext = 'png', cache=TRUE} -->
-<!-- \usetikzlibrary{arrows} -->
-<!-- \begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1] -->
-<!-- \node (1) {$p(\theta, x, y)$}; -->
-<!-- \node (2) [below of=1] {$\tilde p_\texttt{LA}(\theta, y)$}; -->
-<!-- \node (3) [right of=2] {$\tilde p_\texttt{AQ}(y)$}; -->
-<!-- \node (4) [above of=3] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$}; -->
-<!-- \node (5) [right of=4] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$}; -->
-<!-- \node (6) [below of=5] {Output}; -->
-
-<!-- \draw[->] (1) to node {Laplace} (2); -->
-<!-- \draw[->] (2) to node {AGHQ} (3); -->
-<!-- \draw[->] (3) to node {Normalise} (4); -->
-<!-- \draw[->] (4) to node {Marginalise} (5); -->
-<!-- \draw[->] (5) to node {Spline} (6); -->
-<!-- \end{tikzpicture} -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{tikz, echo=FALSE, fig.align='center', fig.cap = "Inference for the latent field: naive Laplace version.", fig.ext = 'png', cache=TRUE} -->
-<!-- \usetikzlibrary{arrows} -->
-<!-- \begin{tikzpicture}[node distance=3.5cm, auto,>=latex', thick, scale = 1] -->
-<!-- \node (1) {$p(\theta, x_i, x_{-i}, y)$}; -->
-<!-- \node (2) [below of=1] {$\tilde p_\texttt{LA}(x_i, \theta, y)$}; -->
-<!-- \node (3) [right of=2] {$\tilde p_\texttt{AQ}(x_i, y)$}; -->
-<!-- \node (4) [above of=3] {$\tilde p_\texttt{AQ}(x_i \, | \, y)$}; -->
-<!-- \node (5) [right of=4] {Output}; -->
-
-<!-- \draw[->] (1) to node {Laplace} (2); -->
-<!-- \draw[->] (2) to node {AGHQ} (3); -->
-<!-- \draw[->] (3) to node {Normalise} (4); -->
-<!-- \draw[->] (4) to node {Spline} (5); -->
-<!-- \end{tikzpicture} -->
-<!-- ``` -->
-
-<!-- # Comparing posterior distributions -->
-
-<!-- * Let $\{\theta_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function -->
-<!-- $$ -->
-<!-- F(\vartheta) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\theta_i \leq \vartheta} -->
-<!-- $$ -->
-<!-- * Could be a hyperparameter, latent field parameter, model outputs -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-<!-- Our primary interest is in the model outputs: do any of the inference methods do a better job at computing the posterior for quantities countries use? -->
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
-<!-- # Moments -->
-
-<!-- * Are the means and standard deviations the same? More generally the $t$th moment can be estimated by -->
-<!-- $$ -->
-<!-- \hat{\mathbb{E}}(\theta^t) = \frac{1}{n} \sum_{i = 1}^n \theta_i^t -->
-<!-- $$ -->
-<!-- * Not great if you're only looking at the first few moments $t = 1, 2$ -->
-
-<!-- # Kolmogorov-Smirov test -->
-
-<!-- * Compare $F_\bullet(\vartheta)$ to $F_{\texttt{NUTS}}(\vartheta)$ -->
-<!-- $$ -->
-<!-- D_\bullet = \sup_\vartheta | F_{\texttt{NUTS}}(\vartheta) - F_\bullet(\vartheta)|. -->
-<!-- $$ -->
-<!-- * A value $D_\bullet$ means there is at most a $(100 \cdot D_\bullet)$% difference between posterior densities -->
-
-<!-- # -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="KS test example."} -->
-<!-- plot(1:10) -->
-<!-- ``` -->
-
-<!-- # Thanks for listening! -->
-
-<!-- * Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work -->
-<!--   * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial) -->
-<!-- * Let me know if you'd be up for being an early reader! -->
-<!-- * Code for this project is at `athowes.github.io/elgm-inf` -->
+\end{tcolorbox}
+\end{center}
 

### 2023-04-25 utils/archive_to_docs.R

@@ -22,6 +22,7 @@ archive_to_docs("docs_01-07-21-stats-epi-group")
 archive_to_docs("docs_15-11-22-seminar")
 archive_to_docs("docs_18-04-23-explainer")
 archive_to_docs("docs_18-04-23-lab-group")
+archive_to_docs("docs_21-06-23-mlgh")
 archive_to_docs("epil")
 archive_to_docs("prev-anc-art_results")
 archive_to_docs("example_inla-replication")

### 2023-04-27 src/naomi-simple_compare/comparison.Rmd
Two small errors
Check issue with KS == 1 -- related to specific age groups (infants and >80) #30

@@
 
+```{r}
+histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
+```
+
 ## `beta_rho`
@@
 ```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
 histogram_and_ecdf_helper("beta_rho")
@@
 lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
-```
-
-## `u_rho_x`
-
-```{r eval = pars_eval[["u_rho_x"]]}
-histogram_and_ecdf_helper("u_rho_x")
 ```
@@
 ks_plot(ks_df = ks_alpha_t1_out, par = "alpha_t1_out", alpha = 0.2)
-ks_plot(ks_df = ks_lambda_t1_out, par = "lambda_t1_out", alpha = 0.2)
+ks_plot(ks_df = ks_lambda_t1_out %>% filter(!(age_group %in% c("Y005_009", "Y010_014", "Y080_999"))), par = "lambda_t1_out", alpha = 0.2)
+```
+
+What is going on with those very high values of KS?
+
+```{r}
+ks_lambda_t1_out %>%
+  filter(ks > 0.99) %>%
+  group_by(method) %>%
+  summarise(n = n())
+
+ks_lambda_t1_out %>%
+  filter(ks > 0.99) %>%
+  pull(age_group) %>%
+  unique()
 ```

### 2023-04-27 src/naomi-simple_ks/ks.Rmd
Rename "compare" to "ks" in preparation for more different types of posterior comparisons

--- 2023-04-27 src/naomi-simple_compare/comparison.Rmd
@@
 
-  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using histograms, Kolmogorov-Smirnov tests, maximum mean discrepancy and Pareto-smoothed importance sampling.
+  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using histograms and Kolmogorov-Smirnov tests.
 ---

### 2023-04-27 src/naomi-simple_compare/script.R

@@ -1,5 +1,0 @@
-#' Uncomment and run the two line below to resume development of this script
-# orderly::orderly_develop_start("naomi-simple_compare")
-# setwd("src/naomi-simple_compare")
-
-rmarkdown::render("comparison.Rmd")

### 2023-04-27 src/naomi_fit_exceedance/exceedance.Rmd
Template for #35

---
title: "Exceedance probability comparison for the simplified Naomi model"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**  We have run the simplified Naomi model using a range of inference methods: `TMB`, `aghq`, `adam` and `tmbstan`.

  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using exceedance probabilities.
---

# Background

We compare the inference results from `TMB`, `aghq`, `adam`, and `tmbstan`.
Import these inference results as follows:

```{r}
tmb <- readRDS("depends/tmb.rds")
aghq <- readRDS("depends/aghq.rds")
adam <- readRDS("depends/adam.rds")
tmbstan <- readRDS("depends/tmbstan.rds")
```

### 2023-04-27 src/docs_bioinference-poster/bioinference-poster.Rmd
Template for #36

--- 2023-03-10 src/docs_bayescomp-poster/bayescomp-poster.Rmd
@@
   4. Allow attendance of ART clinics outside home district
+  5. ART attendance probability as product of prevalence and coverage 
 * We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
@@
 $$
-p(\theta, y) \approx \tilde p_\text{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
+p(\theta, y) \approx \tilde p_\texttt{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\texttt{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
 $$
-where $\tilde p_\text{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
+where $\tilde p_\texttt{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
   * Use automatic differentiation via `CppAD` in `TMB`
@@
 where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(\dim(\theta), k)$ with $k$ points per dimension is adapted based upon
-  * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$
+  * The mode $\hat \theta = \texttt{argmax}_{\theta \in \Theta} p(\theta)$
   * The lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$
@@
 \node (1) {$p(\theta, x, y)$};
-\node (2) [below of=1] {$\tilde p_\text{LA}(x_i, \theta, y)$};
-\node (3) [below of=2] {$\tilde p_\text{LA}(x_i, y)$};
-\node (4) [below of=3] {$\tilde p_\text{AQ}(x_i \, | \, y)$};
+\node (2) [below of=1] {$\tilde p_\texttt{LA}(x_i, \theta, y)$};
+\node (3) [below of=2] {$\tilde p_\texttt{LA}(x_i, y)$};
+\node (4) [below of=3] {$\tilde p_\texttt{AQ}(x_i \, | \, y)$};
 \node (5) [below of=4] {Output};
@@
 \node (6) [right of=1, xshift=1cm] {$\tilde p_\text{LA}(\theta, y)$};
-\node (7) [below of=6] {$\tilde p_\text{AQ}(y)$};
-\node (8) [below of=7] {$\tilde p_\text{AQ}(\theta \, | \, y)$};
-\node (9) [below of=8] {$\tilde p_\text{AQ}(\theta_j \, | \, y)$};
+\node (7) [below of=6] {$\tilde p_\texttt{AQ}(y)$};
+\node (8) [below of=7] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$};
+\node (9) [below of=8] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$};
 \node (10) [below of=9] {Output};

### 2023-04-27 make/_make.R

@@ -90,5 +90,6 @@ run_commit_push("docs_18-04-23-explainer")
 run_commit_push("docs_18-04-23-lab-group")
 run_commit_push("docs_21-06-23-mlgh")
 run_commit_push("docs_bayescomp-poster")
+run_commit_push("docs_bioinference-poster")
 run_commit_push("plot-tikz_algorithm-flowchart")
 run_commit_push("plot-tikz_simplified-naomi")

### 2023-05-01 src/naomi-simple_ks/ks.Rmd
Add ESS for Naomi outputs, close #29

@@
 adam <- readRDS("depends/adam.rds")
+adam_time <- adam$time
+adam <- adam$adam
 tmbstan <- readRDS("depends/tmbstan.rds")
@@
   "aghq" = aghq$time,
-  "adam" = adam$time,
+  "adam" = adam_time,
   "tmbstan" = tmbstan$time
@@
 time_taken
-
-adam <- adam$adam
 ```
@@
 names(pars_eval) <- pars
-```
-
-We will write the `tmbstan` $\hat R$ and ESS for the particular parameter on each plot to help in deciding how seriously to take the MCMC results:
-
-```{r}
-rhats <- bayesplot::rhat(tmbstan$mcmc$stanfit)
-ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc$stanfit)
-niter <- 4 * tmbstan_details$niter / tmbstan_details$nthin
-ess <- ess_ratio * niter
 ```
@@
 ```{r}
+rhats <- bayesplot::rhat(tmbstan$mcmc$stanfit)
+ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc$stanfit)
+niter <- 4 * tmbstan_details$niter / tmbstan_details$nthin
+ess <- ess_ratio * niter
+
 ks_summary %>%

### 2023-05-02 src/docs_paper/paper.Rmd
Small paper edit

@@
 $$
-where $j$ indexes individuals across all strata $i \in \mathcal{I}$.
-The design weights are given by
-$$
-w_j = \frac{1}{\pi_j} \times \frac{1}{\omega_j},
-$$
-where $\pi_j$ is the probability of inclusion and $\omega_j$ is a non-response factor for the particular survey.
-We calculate the observed number of indicator cases as $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
+where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
+<!-- The design weights are given by -->
+<!-- $$ -->
+<!-- w_j = \frac{1}{\pi_j} \times \frac{1}{\omega_j}, -->
+<!-- $$ -->
+<!-- where $\pi_j$ is the probability of inclusion and $\omega_j$ is a non-response factor for the particular survey. -->
+We calculate the observed number of outcomes as $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
 $$

### 2023-05-02 src/naomi-simple_ks/ks.Rmd
Remove MMD and PSIS boilerplate from KS report
Fix #30

@@
 ```{r}
-histogram_and_ecdf(par = "alpha_t1_out", i = 1)
+histogram_and_ecdf(par = "rho_t1_out", i = 1)
 ```
@@
 ks_plot(ks_df = ks_alpha_t1_out, par = "alpha_t1_out", alpha = 0.2)
-ks_plot(ks_df = ks_lambda_t1_out %>% filter(!(age_group %in% c("Y005_009", "Y010_014", "Y080_999"))), par = "lambda_t1_out", alpha = 0.2)
+ks_plot(ks_df = ks_lambda_t1_out, par = "lambda_t1_out", alpha = 0.2)
 ```
 
-What is going on with those very high values of KS?
+What is going on with those very low values of KS?
+It's that there are new new infections in those age groups, and the posterior samples from each method are exactly the same, in particular draws of only zero.
 
@@
 ks_lambda_t1_out %>%
-  filter(ks > 0.99) %>%
+  filter(ks == 0) %>%
   group_by(method) %>%
@@
 ks_lambda_t1_out %>%
-  filter(ks > 0.99) %>%
+  filter(ks == 0) %>%
   pull(age_group) %>%
@@
 ```
-
-# MMD
-
-```{r}
-#' To write!
-```
-
-# PSIS
-
-Suppose we have two sets of samples from the posterior.
-For each sample we are going to want to evaluate the log-likelihood, so that we can calculate the log-likelihood ratios.
-We can extract the `TMB` objective function for the log-likelihood as follows:
-
-```{r}
-tmb$fit$obj$fn()
-#' To write!
-```

### 2023-05-04 src/docs_paper/appendix.Rmd
Add note as to why certain age groups new infections might have to be filtered out of inference results #30

@@
 The terms $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects.
-Finally, $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output.
+Finally, $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output^[Note that these outputs are the only source of age structure for this part of the model.
+Furthermore, Spectrum assumes that there are no new infections in children apart from via mother to child transmission.
+One result is that it is possible for the posterior over new infections in certain age groups to be exactly zero by definition.
+Where this occurs, we remove these age groups from any inference comparisons.]
 

### 2023-05-04 src/docs_paper/appendix.Rmd
Iterate text on Spectrum age structure #30

@@
 Finally, $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output^[Note that these outputs are the only source of age structure for this part of the model.
-Furthermore, Spectrum assumes that there are no new infections in children apart from via mother to child transmission.
-One result is that it is possible for the posterior over new infections in certain age groups to be exactly zero by definition.
-Where this occurs, we remove these age groups from any inference comparisons.]
+Furthermore, Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80.
+A consequence is that the posterior over new infections in these age groups is exactly zero, by definition.
+We remove these identically zero posteriors from any inference comparisons.]
 

### 2023-05-04 src/docs_paper/appendix.Rmd
Add obvious direct dependences to table #33

@@
 \centering
-\begin{tabularx}{\textwidth}{lllll}
+\begin{tabularx}{\textwidth}{llllllll}
 \toprule
-Variable name & Mathematical notation & Type & Dimension & Domain \\ 
+Variable name & Notation & Type & Size & Domain & $\rho$ input? & $\alpha$ input? & $\lambda$ input? \\
 \midrule
-\texttt{logit\_phi\_rho\_x} & $\text{logit}(\phi_X^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_rho\_x} & $\log(\sigma_X^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{logit\_phi\_rho\_xs} & $\text{logit}(\phi_{XS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_rho\_xs} & $\log(\sigma_{XS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{logit\_phi\_rho\_a} & $\text{logit}(\phi_A^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_rho\_a} & $\log(\sigma_A^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{logit\_phi\_rho\_as} & $\text{logit}(\phi_{AS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_rho\_as} & $\log(\sigma_{AS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_rho\_xa} & $\log(\sigma_{XA}^\rho)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{logit\_phi\_alpha\_x} & $\text{logit}(\phi_X^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_alpha\_x} & $\log(\sigma_X^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{logit\_phi\_alpha\_xs} & $\text{logit}(\phi_{XS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_alpha\_xs} & $\log(\sigma_{XS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{logit\_phi\_alpha\_a} & $\text{logit}(\phi_A^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_alpha\_a} & $\log(\sigma_A^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{logit\_phi\_alpha\_as} & $\text{logit}(\phi_{AS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_alpha\_as} & $\log(\sigma_{AS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_alpha\_xa} & $\log(\sigma_{XA}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{OmegaT\_raw} & $\Omega_T$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_betaT} & $\log(\beta_T)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_lambda\_x} & $\log(\sigma^\lambda)$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_ancrho\_x} & $\log(\sigma_X^{\rho^{\text{ANC}}})$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_ancalpha\_x} & $\log(\sigma_X^{\alpha^{\text{ANC}}})$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{log\_sigma\_or\_gamma} & $\log(\sigma_X^{\tilde \gamma})$ & Hyper & 1 & $\mathbb{R}$ \\
-\texttt{beta\_rho} & $(\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ \\
-\texttt{beta\_alpha} & $(\beta^\alpha_0, \beta_{S}^{\alpha, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ \\
-\texttt{beta\_lambda} & $(\beta_0^\lambda, \beta_S^{\lambda, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ \\
-\texttt{beta\_anc\_rho} & $\beta^{\rho^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ \\
-\texttt{beta\_anc\_alpha} & $\beta^{\alpha^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ \\
-\texttt{u\_rho\_x} & $w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{us\_rho\_x} & $v^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{u\_rho\_xs} & $w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{us\_rho\_xs} & $v_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{u\_rho\_a} & $u^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ \\
-\texttt{u\_rho\_as} & $u_a^{\rho, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ \\
-\texttt{u\_rho\_xa} & $u_x^{\rho, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{u\_alpha\_x} & $w^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{us\_alpha\_x} & $v^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{u\_alpha\_xs} & $w_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{us\_alpha\_xs} & $v_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{u\_alpha\_a} & $u^\alpha_a$ & Latent & 13 & $\mathbb{R}^{13}$ \\
-\texttt{u\_alpha\_as} & $u_a^{\alpha, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ \\
-\texttt{u\_alpha\_xa} & $u_x^{\alpha, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{ui\_lambda\_x} & $u_x^\lambda$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{ui\_anc\_rho\_x} & $u_x^{\rho^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{ui\_anc\_alpha\_x} & $u_x^{\alpha^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
-\texttt{log\_or\_gamma} & $u_x^{\tilde \gamma}$ & Latent & $n$ & $\mathbb{R}^{n}$ \\
+\texttt{logit\_phi\_rho\_x} & $\text{logit}(\phi_X^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{log\_sigma\_rho\_x} & $\log(\sigma_X^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{logit\_phi\_rho\_xs} & $\text{logit}(\phi_{XS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{log\_sigma\_rho\_xs} & $\log(\sigma_{XS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{logit\_phi\_rho\_a} & $\text{logit}(\phi_A^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{log\_sigma\_rho\_a} & $\log(\sigma_A^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{logit\_phi\_rho\_as} & $\text{logit}(\phi_{AS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{log\_sigma\_rho\_as} & $\log(\sigma_{AS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{log\_sigma\_rho\_xa} & $\log(\sigma_{XA}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
+\texttt{logit\_phi\_alpha\_x} & $\text{logit}(\phi_X^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{log\_sigma\_alpha\_x} & $\log(\sigma_X^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{logit\_phi\_alpha\_xs} & $\text{logit}(\phi_{XS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{log\_sigma\_alpha\_xs} & $\log(\sigma_{XS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{logit\_phi\_alpha\_a} & $\text{logit}(\phi_A^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{log\_sigma\_alpha\_a} & $\log(\sigma_A^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{logit\_phi\_alpha\_as} & $\text{logit}(\phi_{AS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{log\_sigma\_alpha\_as} & $\log(\sigma_{AS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{log\_sigma\_alpha\_xa} & $\log(\sigma_{XA}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
+\texttt{OmegaT\_raw} & $\Omega_T$ & Hyper & 1 & $\mathbb{R}$ & & & \cmark \\
+\texttt{log\_betaT} & $\log(\beta_T)$ & Hyper & 1 & $\mathbb{R}$ & & & \cmark \\
+\texttt{log\_sigma\_lambda\_x} & $\log(\sigma^\lambda)$ & Hyper & 1 & $\mathbb{R}$ & & & \cmark \\
+\texttt{log\_sigma\_ancrho\_x} & $\log(\sigma_X^{\rho^{\text{ANC}}})$ & Hyper & 1 & $\mathbb{R}$ & & & \\
+\texttt{log\_sigma\_ancalpha\_x} & $\log(\sigma_X^{\alpha^{\text{ANC}}})$ & Hyper & 1 & $\mathbb{R}$ & & & \\
+\texttt{log\_sigma\_or\_gamma} & $\log(\sigma_X^{\tilde \gamma})$ & Hyper & 1 & $\mathbb{R}$ & & & \\
+\texttt{beta\_rho} & $(\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ & \cmark & & \\
+\texttt{beta\_alpha} & $(\beta^\alpha_0, \beta_{S}^{\alpha, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ & & \cmark & \\
+\texttt{beta\_lambda} & $(\beta_0^\lambda, \beta_S^{\lambda, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ & & & \cmark \\
+\texttt{beta\_anc\_rho} & $\beta^{\rho^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ & & & \\
+\texttt{beta\_anc\_alpha} & $\beta^{\alpha^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ & & & \\
+\texttt{u\_rho\_x} & $w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{us\_rho\_x} & $v^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{u\_rho\_xs} & $w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{us\_rho\_xs} & $v_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{u\_rho\_a} & $u^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
+\texttt{u\_rho\_as} & $u_a^{\rho, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
+\texttt{u\_rho\_xa} & $u_x^{\rho, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{u\_alpha\_x} & $w^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ && \cmark & \\
+\texttt{us\_alpha\_x} & $v^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
+\texttt{u\_alpha\_xs} & $w_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
+\texttt{us\_alpha\_xs} & $v_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
+\texttt{u\_alpha\_a} & $u^\alpha_a$ & Latent & 13 & $\mathbb{R}^{13}$ & & \cmark & \\
+\texttt{u\_alpha\_as} & $u_a^{\alpha, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ & & \cmark & \\
+\texttt{u\_alpha\_xa} & $u_x^{\alpha, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
+\texttt{ui\_lambda\_x} & $u_x^\lambda$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \cmark \\
+\texttt{ui\_anc\_rho\_x} & $u_x^{\rho^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
+\texttt{ui\_anc\_alpha\_x} & $u_x^{\alpha^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
+\texttt{log\_or\_gamma} & $u_x^{\tilde \gamma}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
 \bottomrule
@@
 \caption{Correspondence between mathematical notation and variable names used in our \texttt{TMB} code.
-The total number of hyperparameters is 24, and the total number of latent field parameters is $51 + 14n$, where $n$ is the number of districts.}
+The total number of hyperparameters is 24, and the total number of latent field parameters is $51 + 14n$, where $n$ is the number of districts.
+We use the notation \cmark \ to refer to direct dependence of the parameter on the variable, \xmark \ to refer to no dependence, and a blank entry to refer to dependence conditional on the data.}
 \end{table}

### 2023-05-04 src/naomi-simple_ks/ks.Rmd
Use random_only TMB and reorganise KS comparison notebook #27

@@
 ---
-title: "Inference methods comparison for the simplified Naomi model"
+title: "Inference methods comparison for the simplified Naomi model using the Kolmogorov-Smirnov test statistic"
 author:
@@
 We create histograms and empirical cumulative distribution function (ECDF) difference plots of the samples from each method.
-All of the possible parameter names are as follows:
+All of the possible latent field and hyperparamter names are given by `pars`:
 
@@
 pars <- names(tmb$fit$sample)
-pars
 ```
 
-We will produce plots about the following subset of them.
+There are also all of the Naomi outcome variables:
+
+```{r}
+names(tmb$fit$sample)[!(names(tmb$fit$sample) %in% unique(names(tmb$fit$obj$env$par)))]
+```
+
+We will especially focus on the outputs:
+
+* HIV prevalence (`rho_t1_out`) which has `r dim(tmb$fit$sample$rho_t1_out)[1]` variables (rows)
+* ART coverage (`alpha_t1_out`) which has `r dim(tmb$fit$sample$alpha_t1_out)[1]` variables (rows)
+* HIV incidence (`lambda_t1_out`) which has `r dim(tmb$fit$sample$lambda_t1_out)[1]` variables (rows)
+
+These outputs are at an aggregate resolution as well as at the finest resolution.
+It seems to make sense to only to assess inferential accuracy at the finest resolution so as to avoid double counting.
+As such, we need to find the subset of these variables at the finest resolution.
+Within `tmb$naomi_data` there is a dataframe called `mf_out` (model frame output) which contains the area, age, sex mapping for these `r dim(tmb$fit$sample$rho_t1_out)[1]` rows.
+As a sanity check, confirm that the number of areas times number of ages times number of sexes indeed equals the number of rows:
+
+```{r}
+mf_out <- tmb$naomi_data$mf_out
+(n_area <- length(unique(mf_out$area_id)))
+(n_age <- length(unique(mf_out$age_group)))
+(n_sex <- length(unique(mf_out$sex)))
+
+stopifnot(n_area * n_age * n_sex == dim(tmb$fit$sample$rho_t1_out)[1])
+```
+
+The unique levels for area, age and sex are:
+
+```{r}
+mf_out$area_id %>% unique()
+mf_out$age_group %>% unique()
+mf_out$sex %>% unique()
+```
+
+Filter to only the finest resolution, adding an `id` column has to enable merging to the samples:
+
+```{r}
+mf_out_fine <- mf_out %>%
+  tibble::rownames_to_column("id") %>%
+  mutate(id = as.numeric(id)) %>%
+  filter(
+    area_id %in% paste0("MWI_4_", 1:32, "_demo"),
+    sex %in% c("male", "female"),
+    age_group %in%
+      c(
+        "Y000_004", "Y005_009", "Y010_014", "Y015_019", "Y020_024", "Y025_029",
+        "Y025_034", "Y030_034", "Y035_039", "Y040_044", "Y045_049", "Y050_054",
+        "Y055_059", "Y060_064", "Y065_069", "Y070_074", "Y075_079", "Y080_999"
+      )
+  )
+```
+
+We will produce plots about the a small subset of parameters for the time being.
 There is no particular reason to choose this subset rather than other, it's quite arbitrary.
@@
 ```{r}
-pars_eval <- pars %in% c("beta_rho", "beta_alpha", "beta_lambda", "beta_anc_rho", "beta_anc_alpha", "u_rho_x", "us_rho_x", "u_rho_xs")
+pars_eval <- pars %in% c("beta_rho", "u_rho_x")
 names(pars_eval) <- pars
-```
-
-The `beta_rho` and `beta_anc_rho` parameters are used in other reports, so create dataframes here to save as artefacts:
-
-```{r}
-beta_rho <- histogram_and_ecdf("beta_rho", i = 1, return_df = TRUE)
-saveRDS(beta_rho$df, "beta_rho.rds")
-
-beta_anc_rho <- histogram_and_ecdf("beta_anc_rho", return_df = TRUE)
-saveRDS(beta_anc_rho$df, "beta_anc_rho.rds")
 ```
@@
 
-## Outcome variables
+## `rho_t1_out`
 
-These are all of the outcome variables which we might be interested in to monitor:
+There are too many variables to plot this here!
 
-```{r}
-names(tmb$fit$sample)[!(names(tmb$fit$sample) %in% unique(names(tmb$fit$obj$env$par)))]
-```
-The most important ones (this has been checked with Jeff) are HIV prevalence (`rho_t1_out`), ART coverage (`alpha_t1_out`), and HIV incidence (`lambda_t1_out`).
+## `alpha_t1_out`
 
-* HIV prevalence has `r dim(tmb$fit$sample$rho_t1_out)[1]` variables (rows)
-* ART coverage has `r dim(tmb$fit$sample$alpha_t1_out)[1]` variables (rows)
-* HIV incidence has `r dim(tmb$fit$sample$lambda_t1_out)[1]` variables (rows)
+There are too many variables to plot this here!
 
-```{r}
-dim(tmb$fit$sample$rho_t1_out)
-dim(tmb$fit$sample$alpha_t1_out)
-dim(tmb$fit$sample$lambda_t1_out)
-```
+## `lambda_t1_out`
 
-Within `tmb$naomi_data` there is a dataframe called `mf_out` (modelframe output) which contains the area, age, sex mapping these `r dim(tmb$fit$sample$rho_t1_out)[1]` rows.
-As a sanity check, confirm that the number of areas times number of ages times number of sexes indeed equals the number of rows:
-
-```{r}
-mf_out <- tmb$naomi_data$mf_out
-(n_area <- length(unique(mf_out$area_id)))
-(n_age <- length(unique(mf_out$age_group)))
-(n_sex <- length(unique(mf_out$sex)))
-
-stopifnot(n_area * n_age * n_sex == dim(tmb$fit$sample$rho_t1_out)[1])
-```
-
-It makes sense only to assess inferential accuracy at the finest level so as to avoid double counting.
-
-```{r}
-mf_out$area_id %>% unique()
-mf_out$sex %>% unique()
-mf_out$age_group %>% unique()
-
-mf_out_fine <- mf_out %>%
-  tibble::rownames_to_column("id") %>%
-  mutate(id = as.numeric(id)) %>%
-  filter(
-    area_id %in% paste0("MWI_4_", 1:32, "_demo"),
-    sex %in% c("male", "female"),
-    age_group %in%
-      c(
-        "Y000_004", "Y005_009", "Y010_014", "Y015_019", "Y020_024", "Y025_029",
-        "Y025_034", "Y030_034", "Y035_039", "Y040_044", "Y045_049", "Y050_054",
-        "Y055_059", "Y060_064", "Y065_069", "Y070_074", "Y075_079", "Y080_999"
-      )
-  )
-```
-
-An `id` column has been added to `mf_out_fine` to enable merging to the samples.
-
-```{r}
-head(mf_out_fine)
-nrow(mf_out_fine)
-```
-
-```{r}
-rho_t1_out_fine <- tmb$fit$sample$rho_t1_out[mf_out_fine$id, ]
-dim(rho_t1_out_fine)
-rho_t1_out_fine[1:10, 1:10]
-```
-
-Something weird happening with `adam`, go back and check this:
-
-```{r}
-histogram_and_ecdf(par = "rho_t1_out", i = 1)
-```
+There are too many variables to plot this here!
 
@@
 }
+
+ks_df_out <- function(par) {
+  to_ks_df(par = par, outputs = TRUE, id = mf_out_fine$id) %>%
+  select(-par) %>%
+  left_join(
+    mf_out_fine %>%
+      rename("full_id" = "id") %>%
+      tibble::rowid_to_column("index") %>%
+      mutate(index = as.numeric(index))
+  )
+}
+
 ```
@@
 
-### Outcome variables
+### `rho_t1_out`
 
 ```{r}
-ks_df_out <- function(par) {
-  to_ks_df(par = par, outputs = TRUE, id = mf_out_fine$id) %>%
-  select(-par) %>%
-  left_join(
-    mf_out_fine %>%
-      rename("full_id" = "id") %>%
-      tibble::rowid_to_column("index") %>%
-      mutate(index = as.numeric(index))
-  )
-}
-
-ks_rho_t1_out <- ks_df_out(par = "rho_t1_out")
-ks_alpha_t1_out <- ks_df_out(par = "alpha_t1_out")
-ks_lambda_t1_out <- ks_df_out(par = "lambda_t1_out")
-
-ks_plot(ks_df = ks_rho_t1_out, par = "rho_t1_out", alpha = 0.2)
-ks_plot(ks_df = ks_alpha_t1_out, par = "alpha_t1_out", alpha = 0.2)
-ks_plot(ks_df = ks_lambda_t1_out, par = "lambda_t1_out", alpha = 0.2)
+ks_df_out(par = "rho_t1_out") %>%
+  ks_plot(par = "rho_t1_out", alpha = 0.2)
 ```
 
-What is going on with those very low values of KS?
-It's that there are new new infections in those age groups, and the posterior samples from each method are exactly the same, in particular draws of only zero.
+### `alpha_t1_out`
+
+```{r}
+ks_df_out(par = "alpha_t1_out") %>%
+  ks_plot(par = "rho_t1_out", alpha = 0.2)
+```
+
+### `lambda_t1_out`
+
+```{r}
+ks_df_out(par = "rho_t1_out") %>%
+  filter(!age_group %in% c("Y005_009", "Y010_014", "Y080_999")) %>%
+  ks_plot(par = "lambda_t1_out", alpha = 0.2)
+```
+
+Filter out the age group 5-9, 10-14 and 80+ here becausethere are new new infections in those age groups, and the posterior samples from each method are exactly the same: in particular draws of identically zero.
 
@@
 
-## Summary table
+## Summary {.tabset .tabset-fade}
 
@@
 
+saveRDS(ks_summary_out, "ks_summary_out.rds")
+
 ks_summary %>%
-  gt::gt() %>%
-  gt::fmt_number(
-    columns = starts_with("KS"),
-    decimals = 3
-  ) %>%
-  gt::tab_style(
-    style = cell_fill(color = "#0072B2"),
-    locations = cells_body(
-      columns = `KS(adam, tmbstan)`,
-      rows = `KS(adam, tmbstan)` < `KS(aghq, tmbstan)` & `KS(adam, tmbstan)` < `KS(TMB, tmbstan)`
-    )
-  ) %>%
-    gt::tab_style(
-    style = cell_fill(color = "#0072B2"),
-    locations = cells_body(
-      columns = `KS(aghq, tmbstan)`,
-      rows = `KS(aghq, tmbstan)` < `KS(adam, tmbstan)` & `KS(aghq, tmbstan)` < `KS(TMB, tmbstan)`
-    )
-  ) %>%
-    gt::tab_style(
-    style = cell_fill(color = "#0072B2"),
-    locations = cells_body(
-      columns = `KS(TMB, tmbstan)`,
-      rows = `KS(TMB, tmbstan)` < `KS(aghq, tmbstan)` & `KS(TMB, tmbstan)` < `KS(adam, tmbstan)`
-    )
-  )
+  filter(Type != "Hyper") %>%
+  ks_plot_many("TMB", "aghq")
 ```
@@
 ```{r}
-ks_summary %>%
-  filter(Type != "Hyper") %>%
-  ks_plot_many("TMB", "aghq")
-
 (ks_summary_out <- ks_summary %>%
@@
 
-Save some things for output:
+### Latent field
 
 ```{r}
-saveRDS(ks_summary_out, "ks_summary_out.rds")
+ks_table(ks_summary, type = "Latent field")
+```
+
+### Hyperparameters
+
+```{r}
+ks_table(ks_summary, type = "Latent field")
+```
+
+### Outputs
+
+```{r}
+#' To-do
 ```
@@
 
-Nodes where `TMB` beats `aghq`:
+### Nodes where `TMB` beats `aghq`
 
@@
 ks_comparison(ks_summary, method1 = "aghq", method2 = "TMB", 10)
+
+histogram_and_ecdf(par = "us_rho_x", i = 19)
+histogram_and_ecdf(par = "ui_anc_rho_x", i = 2)
+histogram_and_ecdf(par = "u_rho_x", i = 2)
 ```
 
-```{r}
-histogram_and_ecdf(par = "ui_anc_alpha_x", i = 18)
-histogram_and_ecdf(par = "u_alpha_xs", i = 18)
-histogram_and_ecdf(par = "ui_anc_alpha_x", i = 29)
-```
-
-Nodes where `aghq` beats `TMB`:
+### Nodes where `aghq` beats `TMB`
 
@@
 ks_comparison(ks_summary, method1 = "TMB", method2 = "aghq", 10)
-```
 
-```{r}
+histogram_and_ecdf(par = "us_alpha_xs", i = 18)
 histogram_and_ecdf(par = "us_alpha_xs", i = 30)
-histogram_and_ecdf(par = "us_alpha_xs", i = 31)
-histogram_and_ecdf(par = "us_alpha_xs", i = 26)
+histogram_and_ecdf(par = "u_alpha_xa", i = 10)
 ```

### 2023-05-04 src/docs_19-05-23-turing/19-05-23-turing.Rmd
Placeholder for #37

--- 2023-04-25 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
@@
       in_header: preamble.tex
-subtitle: HIV Inference Lab Group Meeting
+subtitle: PhD Student Presentations & Networking Event
 bibliography: citations.bib

### 2023-05-04 make/_make.R

@@ -85,18 +85,23 @@ run_commit_push("naomi-simple_mcmc")
 run_commit_push("naomi-simple_ks")
 run_commit_push("naomi-simple_model-checks")
 
+#' Checks
+run_commit_push("check_sd-estimation")
+
 #' Documentation and plots
 run_commit_push("docs_paper")
+run_commit_push("plot-tikz_algorithm-flowchart")
+run_commit_push("plot-tikz_simplified-naomi")
+
+#' Presentations
 run_commit_push("docs_01-04-20-mini")
 run_commit_push("docs_01-07-21-stats-epi-group")
 run_commit_push("docs_15-11-22-seminar")
 run_commit_push("docs_18-04-23-explainer")
 run_commit_push("docs_18-04-23-lab-group")
+run_commit_push("docs_19-05-23-turing")
 run_commit_push("docs_21-06-23-mlgh")
+
+#' Posters
 run_commit_push("docs_bayescomp-poster")
 run_commit_push("docs_bioinference-poster")
-run_commit_push("plot-tikz_algorithm-flowchart")
-run_commit_push("plot-tikz_simplified-naomi")
-
-#' Checks
-run_commit_push("check_sd-estimation")

### 2023-05-05 src/naomi-simple_mmd/mmd.Rmd
Templates for #41 and #42

--- 2023-05-04 src/naomi-simple_ks/ks.Rmd
@@
 ---
-title: "Inference methods comparison for the simplified Naomi model using the Kolmogorov-Smirnov test statistic"
+title: "Inference methods comparison for the simplified Naomi model using maximum mean discrepancy"
 author:
@@
 
-  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using histograms and Kolmogorov-Smirnov tests.
+  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using maximum mean discrepancy.
 ---
@@
 
-## Time taken
+# Maximum mean discrepancy
 
 ```{r}
-time_taken <- data.frame(
-  "TMB" = tmb$time,
-  "aghq" = aghq$time,
-  "adam" = adam_time,
-  "tmbstan" = tmbstan$time
-)
-
-write_csv(time_taken, "time_taken.csv")
-
-time_taken
+#' To write!
 ```
-
-# Histograms and ECDF difference plots {.tabset .tabset-fade}
-
-We create histograms and empirical cumulative distribution function (ECDF) difference plots of the samples from each method.
-All of the possible latent field and hyperparamter names are given by `pars`:
-
-```{r}
-pars <- names(tmb$fit$sample)
-```
-
-There are also all of the Naomi outcome variables:
-
-```{r}
-names(tmb$fit$sample)[!(names(tmb$fit$sample) %in% unique(names(tmb$fit$obj$env$par)))]
-```
-
-We will especially focus on the outputs:
-
-* HIV prevalence (`rho_t1_out`) which has `r dim(tmb$fit$sample$rho_t1_out)[1]` variables (rows)
-* ART coverage (`alpha_t1_out`) which has `r dim(tmb$fit$sample$alpha_t1_out)[1]` variables (rows)
-* HIV incidence (`lambda_t1_out`) which has `r dim(tmb$fit$sample$lambda_t1_out)[1]` variables (rows)
-
-These outputs are at an aggregate resolution as well as at the finest resolution.
-It seems to make sense to only to assess inferential accuracy at the finest resolution so as to avoid double counting.
-As such, we need to find the subset of these variables at the finest resolution.
-Within `tmb$naomi_data` there is a dataframe called `mf_out` (model frame output) which contains the area, age, sex mapping for these `r dim(tmb$fit$sample$rho_t1_out)[1]` rows.
-As a sanity check, confirm that the number of areas times number of ages times number of sexes indeed equals the number of rows:
-
-```{r}
-mf_out <- tmb$naomi_data$mf_out
-(n_area <- length(unique(mf_out$area_id)))
-(n_age <- length(unique(mf_out$age_group)))
-(n_sex <- length(unique(mf_out$sex)))
-
-stopifnot(n_area * n_age * n_sex == dim(tmb$fit$sample$rho_t1_out)[1])
-```
-
-The unique levels for area, age and sex are:
-
-```{r}
-mf_out$area_id %>% unique()
-mf_out$age_group %>% unique()
-mf_out$sex %>% unique()
-```
-
-Filter to only the finest resolution, adding an `id` column has to enable merging to the samples:
-
-```{r}
-mf_out_fine <- mf_out %>%
-  tibble::rownames_to_column("id") %>%
-  mutate(id = as.numeric(id)) %>%
-  filter(
-    area_id %in% paste0("MWI_4_", 1:32, "_demo"),
-    sex %in% c("male", "female"),
-    age_group %in%
-      c(
-        "Y000_004", "Y005_009", "Y010_014", "Y015_019", "Y020_024", "Y025_029",
-        "Y025_034", "Y030_034", "Y035_039", "Y040_044", "Y045_049", "Y050_054",
-        "Y055_059", "Y060_064", "Y065_069", "Y070_074", "Y075_079", "Y080_999"
-      )
-  )
-```
-
-We will produce plots about the a small subset of parameters for the time being.
-There is no particular reason to choose this subset rather than other, it's quite arbitrary.
-
-```{r}
-pars_eval <- pars %in% c("beta_rho", "u_rho_x")
-names(pars_eval) <- pars
-```
-
-```{r}
-histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
-```
-
-## `beta_rho`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_rho")
-```
-
-## `beta_alpha`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_alpha")
-```
-
-## `beta_lambda`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_lambda")
-```
-
-## `beta_anc_rho`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf("beta_anc_rho")
-```
-
-## `beta_anc_alpha`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf("beta_anc_alpha")
-```
-
-## `logit`
-
-```{r eval = FALSE}
-lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf)
-```
-
-## `log_sigma`
-
-```{r eval = FALSE}
-lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
-```
-
-## `u_rho_x`
-
-```{r eval = pars_eval[["u_rho_x"]]}
-histogram_and_ecdf_helper("u_rho_x")
-```
-
-## `us_rho_x`
-
-```{r eval = pars_eval[["us_rho_x"]]}
-histogram_and_ecdf_helper("us_rho_x")
-```
-
-## `u_rho_xs`
-
-```{r eval = pars_eval[["u_rho_xs"]]}
-histogram_and_ecdf_helper("u_rho_xs")
-```
-
-## `us_rho_xs`
-
-```{r eval = pars_eval[["us_rho_xs"]]}
-histogram_and_ecdf_helper("us_rho_xs")
-```
-
-## `u_rho_as`
-
-```{r eval = pars_eval[["u_rho_as"]]}
-histogram_and_ecdf_helper("u_rho_as")
-```
-
-## `u_alpha_x`
-
-```{r eval = pars_eval[["u_alpha_x"]]}
-histogram_and_ecdf_helper("u_alpha_x")
-```
-
-## `us_alpha_x`
-
-```{r eval = pars_eval[["us_alpha_x"]]}
-histogram_and_ecdf_helper("us_alpha_x")
-```
-
-## `u_alpha_xs`
-
-```{r eval = pars_eval[["u_alpha_xs"]]}
-histogram_and_ecdf_helper("u_alpha_xs")
-```
-
-## `us_alpha_xs`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("us_alpha_xs")
-```
-
-## `u_alpha_a`
-
-```{r eval = pars_eval[["u_alpha_a"]]}
-histogram_and_ecdf_helper("u_alpha_a")
-```
-
-## `u_alpha_as`
-
-```{r eval = pars_eval[["u_alpha_as"]]}
-histogram_and_ecdf_helper("u_alpha_as")
-```
-
-## `u_alpha_xa`
-
-```{r eval = pars_eval[["u_alpha_xa"]]}
-histogram_and_ecdf_helper("u_alpha_xa")
-```
-
-## `ui_lambda_x`
-
-```{r eval = pars_eval[["ui_lambda_x"]]}
-histogram_and_ecdf_helper("ui_lambda_x")
-```
-
-## `ui_anc_rho_x`
-
-```{r eval = pars_eval[["ui_anc_rho_x"]]}
-histogram_and_ecdf_helper("ui_anc_rho_x")
-```
-
-## `ui_anc_alpha_x`
-
-```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-histogram_and_ecdf_helper("ui_anc_alpha_x")
-```
-
-## `log_or_gamma`
-
-```{r eval = pars_eval[["log_or_gamma"]]}
-histogram_and_ecdf_helper("log_or_gamma")
-```
-
-## `rho_t1_out`
-
-There are too many variables to plot this here!
-
-## `alpha_t1_out`
-
-There are too many variables to plot this here!
-
-## `lambda_t1_out`
-
-There are too many variables to plot this here!
-
-# KS plots
-
-## Individual parameters {.tabset .tabset-fade}
-
-```{r}
-ks_helper <- function(par, starts_with = FALSE, ...) {
-  to_ks_df(par, starts_with = starts_with) %>% ks_plot(par, ...)
-}
-
-ks_df_out <- function(par) {
-  to_ks_df(par = par, outputs = TRUE, id = mf_out_fine$id) %>%
-  select(-par) %>%
-  left_join(
-    mf_out_fine %>%
-      rename("full_id" = "id") %>%
-      tibble::rowid_to_column("index") %>%
-      mutate(index = as.numeric(index))
-  )
-}
-
-```
-
-### `beta`
-
-```{r}
-ks_helper("beta", starts_with = TRUE)
-```
-
-### `logit`
-
-```{r}
-ks_helper("logit", starts_with = TRUE)
-```
-
-### `log_sigma`
-
-```{r}
-ks_helper("log_sigma", starts_with = TRUE)
-```
-
-### `u_rho_x`
-
-```{r eval = pars_eval[["u_rho_x"]]}
-ks_helper("u_rho_x")
-```
-
-### `u_rho_xs`
-
-```{r eval = pars_eval[["u_rho_xs"]]}
-ks_helper("u_rho_xs")
-```
-
-### `us_rho_x`
-
-```{r eval = pars_eval[["us_rho_x"]]}
-ks_helper("us_rho_x")
-```
-
-### `us_rho_xs`
-
-```{r eval = pars_eval[["us_rho_xs"]]}
-ks_helper("us_rho_xs")
-```
-
-### `u_rho_a`
-
-```{r eval = pars_eval[["u_rho_a"]]}
-ks_helper("u_rho_a")
-```
-
-### `u_rho_as`
-
-```{r eval = pars_eval[["u_rho_as"]]}
-ks_helper("u_rho_as")
-```
-
-### `u_alpha_x`
-
-```{r eval = pars_eval[["u_alpha_x"]]}
-ks_helper("u_alpha_x")
-```
-
-### `u_alpha_xs`
-
-```{r eval = pars_eval[["u_alpha_xs"]]}
-ks_helper("u_alpha_xs")
-```
-
-### `us_alpha_x`
-
-```{r eval = pars_eval[["us_alpha_x"]]}
-ks_helper("us_alpha_x")
-```
-
-### `us_alpha_xs`
-
-```{r eval = pars_eval[["us_alpha_xs"]]}
-ks_helper("us_alpha_xs")
-```
-
-### `u_alpha_a`
-
-```{r eval = pars_eval[["u_alpha_a"]]}
-ks_helper("u_alpha_a")
-```
-
-### `u_alpha_as`
-
-```{r eval = pars_eval[["u_alpha_as"]]}
-ks_helper("u_alpha_as")
-```
-
-### `u_alpha_xa`
-
-```{r eval = pars_eval[["u_alpha_xa"]]}
-ks_helper("u_alpha_xa")
-```
-
-### `ui_anc_rho_x`
-
-```{r eval = pars_eval[["ui_anc_rho_x"]]}
-ks_helper("ui_anc_rho_x")
-```
-
-### `ui_anc_alpha_x`
-
-```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-ks_helper("ui_anc_alpha_x")
-```
-
-### `log_or_gamma`
-
-```{r eval = pars_eval[["log_or_gamma"]]}
-ks_helper("log_or_gamma")
-```
-
-### `rho_t1_out`
-
-```{r}
-ks_df_out(par = "rho_t1_out") %>%
-  ks_plot(par = "rho_t1_out", alpha = 0.2)
-```
-
-### `alpha_t1_out`
-
-```{r}
-ks_df_out(par = "alpha_t1_out") %>%
-  ks_plot(par = "rho_t1_out", alpha = 0.2)
-```
-
-### `lambda_t1_out`
-
-```{r}
-ks_df_out(par = "rho_t1_out") %>%
-  filter(!age_group %in% c("Y005_009", "Y010_014", "Y080_999")) %>%
-  ks_plot(par = "lambda_t1_out", alpha = 0.2)
-```
-
-Filter out the age group 5-9, 10-14 and 80+ here becausethere are new new infections in those age groups, and the posterior samples from each method are exactly the same: in particular draws of identically zero.
-
-```{r}
-ks_lambda_t1_out %>%
-  filter(ks == 0) %>%
-  group_by(method) %>%
-  summarise(n = n())
-
-ks_lambda_t1_out %>%
-  filter(ks == 0) %>%
-  pull(age_group) %>%
-  unique()
-```
-
-## Summary {.tabset .tabset-fade}
-
-Values in dark blue are the lowest KS difference for that particular parameter:
-
-```{r class.source = 'fold-hide'}
-options(dplyr.summarise.inform = FALSE)
-
-ks_summary <- lapply(unique(names(tmb$fit$obj$env$par)), function(x) {
-  to_ks_df(x) %>%
-    group_by(method, index) %>%
-    summarise(ks = mean(ks), par = x) %>%
-    ungroup()
-}) %>%
-  bind_rows() %>%
-  pivot_wider(names_from = "method", values_from = "ks") %>%
-  rename(
-    "Parameter" = "par",
-    "KS(adam, tmbstan)" = "adam",
-    "KS(aghq, tmbstan)" = "aghq",
-    "KS(TMB, tmbstan)" = "TMB",
-  )
-
-r <- adam$quad$obj$env$random
-x_names <- names(adam$quad$obj$env$par[r])
-theta_names <- names(adam$quad$obj$env$par[-r])
-
-dict <- data.frame(
-  Parameter = c(unique(x_names), unique(theta_names)),
-  Type = c(rep("Latent field", length(unique(x_names))), rep("Hyper", length(unique(theta_names))))
-)
-
-ks_summary <- ks_summary %>%
-  left_join(dict, by = "Parameter")
-
-saveRDS(ks_summary_out, "ks_summary_out.rds")
-
-ks_summary %>%
-  filter(Type != "Hyper") %>%
-  ks_plot_many("TMB", "aghq")
-```
-
-```{r}
-(ks_summary_out <- ks_summary %>%
-  group_by(Type) %>%
-  summarise(
-    TMB = mean(`KS(TMB, tmbstan)`),
-    aghq = mean(`KS(aghq, tmbstan)`),
-    adam = mean(`KS(adam, tmbstan)`)
-  )) %>%
-  gt::gt()
-```
-
-### Latent field
-
-```{r}
-ks_table(ks_summary, type = "Latent field")
-```
-
-### Hyperparameters
-
-```{r}
-ks_table(ks_summary, type = "Latent field")
-```
-
-### Outputs
-
-```{r}
-#' To-do
-```
-
-## Investigation into large KS values
-
-Want to create rank order lists of largest KS differences between methods:
-
-```{r}
-ks_comparison <- function(ks_summary, method1, method2, n) {
-  ks_method1 <- paste0("KS(", method1, ", tmbstan)")
-  ks_method2 <- paste0("KS(", method2, ", tmbstan)")
-  ks_summary[["KS difference"]] <- ks_summary[[ks_method1]] - ks_summary[[ks_method2]]
-  ks_summary %>%
-    filter(Type != "Hyper") %>%
-    arrange(desc(`KS difference`)) %>%
-    head(n = n)
-}
-```
-
-### Nodes where `TMB` beats `aghq`
-
-```{r}
-ks_comparison(ks_summary, method1 = "aghq", method2 = "TMB", 10)
-
-histogram_and_ecdf(par = "us_rho_x", i = 19)
-histogram_and_ecdf(par = "ui_anc_rho_x", i = 2)
-histogram_and_ecdf(par = "u_rho_x", i = 2)
-```
-
-### Nodes where `aghq` beats `TMB`
-
-```{r}
-ks_comparison(ks_summary, method1 = "TMB", method2 = "aghq", 10)
-
-histogram_and_ecdf(par = "us_alpha_xs", i = 18)
-histogram_and_ecdf(par = "us_alpha_xs", i = 30)
-histogram_and_ecdf(par = "u_alpha_xa", i = 10)
-```
-
-## Correlation between KS values and ESS
-
-Is there any correlation between the value of $\text{KS}(\texttt{method}, \texttt{tmbstan})$ for a particular parameter and the ESS of that parameter from `tmbstan` output?
-
-```{r}
-rhats <- bayesplot::rhat(tmbstan$mcmc$stanfit)
-ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc$stanfit)
-niter <- 4 * tmbstan_details$niter / tmbstan_details$nthin
-ess <- ess_ratio * niter
-
-ks_summary %>%
-  pivot_longer(
-    col = starts_with("KS"),
-    names_to = "Method",
-    values_to = "KS"
-  ) %>%
-  mutate(
-    Method = fct_recode(Method,
-      "adam" = "KS(adam, tmbstan)",
-      "aghq" = "KS(aghq, tmbstan)",
-      "TMB" = "KS(TMB, tmbstan)"
-    )
-  ) %>%
-  group_by(Parameter) %>%
-  mutate(
-    par_num = case_when(
-      max(index) > 1 ~ paste0(Parameter, "[", index, "]"),
-      TRUE ~ Parameter
-    )
-  ) %>%
-  ungroup() %>%
-  left_join(data.frame(ess) %>%
-    tibble::rownames_to_column("par_num")
-  ) %>%
-  ggplot(aes(x = ess, y = KS)) +
-    geom_point(alpha = 0.5) +
-    geom_smooth(method = "lm", color = "#56B4E9") +
-    facet_grid(~ Method) +
-    theme_minimal() +
-    labs(x = "ESS", y = "KS(method, tmbstan)")
-```

### 2023-05-05 src/naomi-simple_psis/psis.Rmd
Templates for #41 and #42

--- 2023-04-20 src/naomi-simple_compare/comparison.Rmd
@@
 ---
-title: "Inference methods comparison for the simplified Naomi model"
+title: "Inference methods comparison for the simplified Naomi model using Pareto-smoothed importance sampling"
 author:
@@
 
-  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using histograms, Kolmogorov-Smirnov tests, maximum mean discrepancy and Pareto-smoothed importance sampling.
+  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using Pareto-smoothed importance sampling.
 ---
@@
 
-## Time taken
-
-```{r}
-time_taken <- data.frame(
-  "TMB" = tmb$time,
-  "aghq" = aghq$time,
-  "adam" = adam$time,
-  "tmbstan" = tmbstan$time
-)
-
-write_csv(time_taken, "time_taken.csv")
-
-time_taken
-
-adam <- adam$adam
-```
-
-# Histograms and ECDF difference plots {.tabset .tabset-fade}
-
-We create histograms and empirical cumulative distribution function (ECDF) difference plots of the samples from each method.
-All of the possible parameter names are as follows:
-
-```{r}
-pars <- names(tmb$fit$sample)
-pars
-```
-
-We will produce plots about the following subset of them.
-There is no particular reason to choose this subset rather than other, it's quite arbitrary.
-
-```{r}
-pars_eval <- pars %in% c("beta_rho", "beta_alpha", "beta_lambda", "beta_anc_rho", "beta_anc_alpha", "u_rho_x", "us_rho_x", "u_rho_xs")
-names(pars_eval) <- pars
-```
-
-We will write the `tmbstan` $\hat R$ and ESS for the particular parameter on each plot to help in deciding how seriously to take the MCMC results:
-
-```{r}
-rhats <- bayesplot::rhat(tmbstan$mcmc$stanfit)
-ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc$stanfit)
-niter <- 4 * tmbstan_details$niter / tmbstan_details$nthin
-ess <- ess_ratio * niter
-```
-
-The `beta_rho` and `beta_anc_rho` parameters are used in other reports, so create dataframes here to save as artefacts:
-
-```{r}
-beta_rho <- histogram_and_ecdf("beta_rho", i = 1, return_df = TRUE)
-saveRDS(beta_rho$df, "beta_rho.rds")
-
-beta_anc_rho <- histogram_and_ecdf("beta_anc_rho", return_df = TRUE)
-saveRDS(beta_anc_rho$df, "beta_anc_rho.rds")
-```
-
-## `beta_rho`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
-histogram_and_ecdf_helper("beta_rho")
-```
-
-## `beta_alpha`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_alpha")
-```
-
-## `beta_lambda`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_lambda")
-```
-
-## `beta_anc_rho`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf("beta_anc_rho")
-```
-
-## `beta_anc_alpha`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf("beta_anc_alpha")
-```
-
-## `logit`
-
-```{r eval = FALSE}
-lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf)
-```
-
-## `log_sigma`
-
-```{r eval = FALSE}
-lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
-```
-
-## `u_rho_x`
-
-```{r eval = pars_eval[["u_rho_x"]]}
-histogram_and_ecdf_helper("u_rho_x")
-```
-
-## `u_rho_x`
-
-```{r eval = pars_eval[["u_rho_x"]]}
-histogram_and_ecdf_helper("u_rho_x")
-```
-
-## `us_rho_x`
-
-```{r eval = pars_eval[["us_rho_x"]]}
-histogram_and_ecdf_helper("us_rho_x")
-```
-
-## `u_rho_xs`
-
-```{r eval = pars_eval[["u_rho_xs"]]}
-histogram_and_ecdf_helper("u_rho_xs")
-```
-
-## `us_rho_xs`
-
-```{r eval = pars_eval[["us_rho_xs"]]}
-histogram_and_ecdf_helper("us_rho_xs")
-```
-
-## `u_rho_as`
-
-```{r eval = pars_eval[["u_rho_as"]]}
-histogram_and_ecdf_helper("u_rho_as")
-```
-
-## `u_alpha_x`
-
-```{r eval = pars_eval[["u_alpha_x"]]}
-histogram_and_ecdf_helper("u_alpha_x")
-```
-
-## `us_alpha_x`
-
-```{r eval = pars_eval[["us_alpha_x"]]}
-histogram_and_ecdf_helper("us_alpha_x")
-```
-
-## `u_alpha_xs`
-
-```{r eval = pars_eval[["u_alpha_xs"]]}
-histogram_and_ecdf_helper("u_alpha_xs")
-```
-
-## `us_alpha_xs`
-
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("us_alpha_xs")
-```
-
-## `u_alpha_a`
-
-```{r eval = pars_eval[["u_alpha_a"]]}
-histogram_and_ecdf_helper("u_alpha_a")
-```
-
-## `u_alpha_as`
-
-```{r eval = pars_eval[["u_alpha_as"]]}
-histogram_and_ecdf_helper("u_alpha_as")
-```
-
-## `u_alpha_xa`
-
-```{r eval = pars_eval[["u_alpha_xa"]]}
-histogram_and_ecdf_helper("u_alpha_xa")
-```
-
-## `ui_lambda_x`
-
-```{r eval = pars_eval[["ui_lambda_x"]]}
-histogram_and_ecdf_helper("ui_lambda_x")
-```
-
-## `ui_anc_rho_x`
-
-```{r eval = pars_eval[["ui_anc_rho_x"]]}
-histogram_and_ecdf_helper("ui_anc_rho_x")
-```
-
-## `ui_anc_alpha_x`
-
-```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-histogram_and_ecdf_helper("ui_anc_alpha_x")
-```
-
-## `log_or_gamma`
-
-```{r eval = pars_eval[["log_or_gamma"]]}
-histogram_and_ecdf_helper("log_or_gamma")
-```
-
-## Outcome variables
-
-These are all of the outcome variables which we might be interested in to monitor:
-
-```{r}
-names(tmb$fit$sample)[!(names(tmb$fit$sample) %in% unique(names(tmb$fit$obj$env$par)))]
-```
-The most important ones (this has been checked with Jeff) are HIV prevalence (`rho_t1_out`), ART coverage (`alpha_t1_out`), and HIV incidence (`lambda_t1_out`).
-
-* HIV prevalence has `r dim(tmb$fit$sample$rho_t1_out)[1]` variables (rows)
-* ART coverage has `r dim(tmb$fit$sample$alpha_t1_out)[1]` variables (rows)
-* HIV incidence has `r dim(tmb$fit$sample$lambda_t1_out)[1]` variables (rows)
-
-```{r}
-dim(tmb$fit$sample$rho_t1_out)
-dim(tmb$fit$sample$alpha_t1_out)
-dim(tmb$fit$sample$lambda_t1_out)
-```
-
-Within `tmb$naomi_data` there is a dataframe called `mf_out` (modelframe output) which contains the area, age, sex mapping these `r dim(tmb$fit$sample$rho_t1_out)[1]` rows.
-As a sanity check, confirm that the number of areas times number of ages times number of sexes indeed equals the number of rows:
-
-```{r}
-mf_out <- tmb$naomi_data$mf_out
-(n_area <- length(unique(mf_out$area_id)))
-(n_age <- length(unique(mf_out$age_group)))
-(n_sex <- length(unique(mf_out$sex)))
-
-stopifnot(n_area * n_age * n_sex == dim(tmb$fit$sample$rho_t1_out)[1])
-```
-
-It makes sense only to assess inferential accuracy at the finest level so as to avoid double counting.
-
-```{r}
-mf_out$area_id %>% unique()
-mf_out$sex %>% unique()
-mf_out$age_group %>% unique()
-
-mf_out_fine <- mf_out %>%
-  tibble::rownames_to_column("id") %>%
-  mutate(id = as.numeric(id)) %>%
-  filter(
-    area_id %in% paste0("MWI_4_", 1:32, "_demo"),
-    sex %in% c("male", "female"),
-    age_group %in%
-      c(
-        "Y000_004", "Y005_009", "Y010_014", "Y015_019", "Y020_024", "Y025_029",
-        "Y025_034", "Y030_034", "Y035_039", "Y040_044", "Y045_049", "Y050_054",
-        "Y055_059", "Y060_064", "Y065_069", "Y070_074", "Y075_079", "Y080_999"
-      )
-  )
-```
-
-An `id` column has been added to `mf_out_fine` to enable merging to the samples.
-
-```{r}
-head(mf_out_fine)
-nrow(mf_out_fine)
-```
-
-```{r}
-rho_t1_out_fine <- tmb$fit$sample$rho_t1_out[mf_out_fine$id, ]
-dim(rho_t1_out_fine)
-rho_t1_out_fine[1:10, 1:10]
-```
-
-Something weird happening with `adam`, go back and check this:
-
-```{r}
-histogram_and_ecdf(par = "alpha_t1_out", i = 1)
-```
-
-# KS plots
-
-## Individual parameters {.tabset .tabset-fade}
-
-```{r}
-ks_helper <- function(par, starts_with = FALSE, ...) {
-  to_ks_df(par, starts_with = starts_with) %>% ks_plot(par, ...)
-}
-```
-
-### `beta`
-
-```{r}
-ks_helper("beta", starts_with = TRUE)
-```
-
-### `logit`
-
-```{r}
-ks_helper("logit", starts_with = TRUE)
-```
-
-### `log_sigma`
-
-```{r}
-ks_helper("log_sigma", starts_with = TRUE)
-```
-
-### `u_rho_x`
-
-```{r eval = pars_eval[["u_rho_x"]]}
-ks_helper("u_rho_x")
-```
-
-### `u_rho_xs`
-
-```{r eval = pars_eval[["u_rho_xs"]]}
-ks_helper("u_rho_xs")
-```
-
-### `us_rho_x`
-
-```{r eval = pars_eval[["us_rho_x"]]}
-ks_helper("us_rho_x")
-```
-
-### `us_rho_xs`
-
-```{r eval = pars_eval[["us_rho_xs"]]}
-ks_helper("us_rho_xs")
-```
-
-### `u_rho_a`
-
-```{r eval = pars_eval[["u_rho_a"]]}
-ks_helper("u_rho_a")
-```
-
-### `u_rho_as`
-
-```{r eval = pars_eval[["u_rho_as"]]}
-ks_helper("u_rho_as")
-```
-
-### `u_alpha_x`
-
-```{r eval = pars_eval[["u_alpha_x"]]}
-ks_helper("u_alpha_x")
-```
-
-### `u_alpha_xs`
-
-```{r eval = pars_eval[["u_alpha_xs"]]}
-ks_helper("u_alpha_xs")
-```
-
-### `us_alpha_x`
-
-```{r eval = pars_eval[["us_alpha_x"]]}
-ks_helper("us_alpha_x")
-```
-
-### `us_alpha_xs`
-
-```{r eval = pars_eval[["us_alpha_xs"]]}
-ks_helper("us_alpha_xs")
-```
-
-### `u_alpha_a`
-
-```{r eval = pars_eval[["u_alpha_a"]]}
-ks_helper("u_alpha_a")
-```
-
-### `u_alpha_as`
-
-```{r eval = pars_eval[["u_alpha_as"]]}
-ks_helper("u_alpha_as")
-```
-
-### `u_alpha_xa`
-
-```{r eval = pars_eval[["u_alpha_xa"]]}
-ks_helper("u_alpha_xa")
-```
-
-### `ui_anc_rho_x`
-
-```{r eval = pars_eval[["ui_anc_rho_x"]]}
-ks_helper("ui_anc_rho_x")
-```
-
-### `ui_anc_alpha_x`
-
-```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-ks_helper("ui_anc_alpha_x")
-```
-
-### `log_or_gamma`
-
-```{r eval = pars_eval[["log_or_gamma"]]}
-ks_helper("log_or_gamma")
-```
-
-### Outcome variables
-
-```{r}
-ks_df_out <- function(par) {
-  to_ks_df(par = par, outputs = TRUE, id = mf_out_fine$id) %>%
-  select(-par) %>%
-  left_join(
-    mf_out_fine %>%
-      rename("full_id" = "id") %>%
-      tibble::rowid_to_column("index") %>%
-      mutate(index = as.numeric(index))
-  )
-}
-
-ks_rho_t1_out <- ks_df_out(par = "rho_t1_out")
-ks_alpha_t1_out <- ks_df_out(par = "alpha_t1_out")
-ks_lambda_t1_out <- ks_df_out(par = "lambda_t1_out")
-
-ks_plot(ks_df = ks_rho_t1_out, par = "rho_t1_out", alpha = 0.2)
-ks_plot(ks_df = ks_alpha_t1_out, par = "alpha_t1_out", alpha = 0.2)
-ks_plot(ks_df = ks_lambda_t1_out, par = "lambda_t1_out", alpha = 0.2)
-```
-
-## Summary table
-
-Values in dark blue are the lowest KS difference for that particular parameter:
-
-```{r class.source = 'fold-hide'}
-options(dplyr.summarise.inform = FALSE)
-
-ks_summary <- lapply(unique(names(tmb$fit$obj$env$par)), function(x) {
-  to_ks_df(x) %>%
-    group_by(method, index) %>%
-    summarise(ks = mean(ks), par = x) %>%
-    ungroup()
-}) %>%
-  bind_rows() %>%
-  pivot_wider(names_from = "method", values_from = "ks") %>%
-  rename(
-    "Parameter" = "par",
-    "KS(adam, tmbstan)" = "adam",
-    "KS(aghq, tmbstan)" = "aghq",
-    "KS(TMB, tmbstan)" = "TMB",
-  )
-
-r <- adam$quad$obj$env$random
-x_names <- names(adam$quad$obj$env$par[r])
-theta_names <- names(adam$quad$obj$env$par[-r])
-
-dict <- data.frame(
-  Parameter = c(unique(x_names), unique(theta_names)),
-  Type = c(rep("Latent field", length(unique(x_names))), rep("Hyper", length(unique(theta_names))))
-)
-
-ks_summary <- ks_summary %>%
-  left_join(dict, by = "Parameter")
-
-ks_summary %>%
-  gt::gt() %>%
-  gt::fmt_number(
-    columns = starts_with("KS"),
-    decimals = 3
-  ) %>%
-  gt::tab_style(
-    style = cell_fill(color = "#0072B2"),
-    locations = cells_body(
-      columns = `KS(adam, tmbstan)`,
-      rows = `KS(adam, tmbstan)` < `KS(aghq, tmbstan)` & `KS(adam, tmbstan)` < `KS(TMB, tmbstan)`
-    )
-  ) %>%
-    gt::tab_style(
-    style = cell_fill(color = "#0072B2"),
-    locations = cells_body(
-      columns = `KS(aghq, tmbstan)`,
-      rows = `KS(aghq, tmbstan)` < `KS(adam, tmbstan)` & `KS(aghq, tmbstan)` < `KS(TMB, tmbstan)`
-    )
-  ) %>%
-    gt::tab_style(
-    style = cell_fill(color = "#0072B2"),
-    locations = cells_body(
-      columns = `KS(TMB, tmbstan)`,
-      rows = `KS(TMB, tmbstan)` < `KS(aghq, tmbstan)` & `KS(TMB, tmbstan)` < `KS(adam, tmbstan)`
-    )
-  )
-```
-
-```{r}
-ks_summary %>%
-  filter(Type != "Hyper") %>%
-  ks_plot_many("TMB", "aghq")
-
-(ks_summary_out <- ks_summary %>%
-  group_by(Type) %>%
-  summarise(
-    TMB = mean(`KS(TMB, tmbstan)`),
-    aghq = mean(`KS(aghq, tmbstan)`),
-    adam = mean(`KS(adam, tmbstan)`)
-  )) %>%
-  gt::gt()
-```
-
-Save some things for output:
-
-```{r}
-saveRDS(ks_summary_out, "ks_summary_out.rds")
-```
-
-## Investigation into large KS values
-
-Want to create rank order lists of largest KS differences between methods:
-
-```{r}
-ks_comparison <- function(ks_summary, method1, method2, n) {
-  ks_method1 <- paste0("KS(", method1, ", tmbstan)")
-  ks_method2 <- paste0("KS(", method2, ", tmbstan)")
-  ks_summary[["KS difference"]] <- ks_summary[[ks_method1]] - ks_summary[[ks_method2]]
-  ks_summary %>%
-    filter(Type != "Hyper") %>%
-    arrange(desc(`KS difference`)) %>%
-    head(n = n)
-}
-```
-
-Nodes where `TMB` beats `aghq`:
-
-```{r}
-ks_comparison(ks_summary, method1 = "aghq", method2 = "TMB", 10)
-```
-
-```{r}
-histogram_and_ecdf(par = "ui_anc_alpha_x", i = 18)
-histogram_and_ecdf(par = "u_alpha_xs", i = 18)
-histogram_and_ecdf(par = "ui_anc_alpha_x", i = 29)
-```
-
-Nodes where `aghq` beats `TMB`:
-
-```{r}
-ks_comparison(ks_summary, method1 = "TMB", method2 = "aghq", 10)  
-```
-
-```{r}
-histogram_and_ecdf(par = "us_alpha_xs", i = 30)
-histogram_and_ecdf(par = "us_alpha_xs", i = 31)
-histogram_and_ecdf(par = "us_alpha_xs", i = 26)
-```
-
-## Correlation between KS values and ESS
-
-Is there any correlation between the value of $\text{KS}(\texttt{method}, \texttt{tmbstan})$ for a particular parameter and the ESS of that parameter from `tmbstan` output?
-
-```{r}
-ks_summary %>%
-  pivot_longer(
-    col = starts_with("KS"),
-    names_to = "Method",
-    values_to = "KS"
-  ) %>%
-  mutate(
-    Method = fct_recode(Method,
-      "adam" = "KS(adam, tmbstan)",
-      "aghq" = "KS(aghq, tmbstan)",
-      "TMB" = "KS(TMB, tmbstan)"
-    )
-  ) %>%
-  group_by(Parameter) %>%
-  mutate(
-    par_num = case_when(
-      max(index) > 1 ~ paste0(Parameter, "[", index, "]"),
-      TRUE ~ Parameter
-    )
-  ) %>%
-  ungroup() %>%
-  left_join(data.frame(ess) %>%
-    tibble::rownames_to_column("par_num")
-  ) %>%
-  ggplot(aes(x = ess, y = KS)) +
-    geom_point(alpha = 0.5) +
-    geom_smooth(method = "lm", color = "#56B4E9") +
-    facet_grid(~ Method) +
-    theme_minimal() +
-    labs(x = "ESS", y = "KS(method, tmbstan)")
-```
-
-# MMD
-
-```{r}
-#' To write!
-```
-
-# PSIS
+# Pareto-smoothed importance sampling
 

### 2023-05-05 make/_make.R

@@ -83,6 +83,8 @@ orderly::orderly_commit(id)
 #' Results
 run_commit_push("naomi-simple_mcmc")
 run_commit_push("naomi-simple_ks")
+run_commit_push("naomi-simple_psis")
+run_commit_push("naomi-simple_mmd")
 run_commit_push("naomi-simple_model-checks")
 
 #' Checks

### 2023-05-05 src/naomi-simple_mmd/script.R

@@ -1,0 +1,5 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("naomi-simple_mmd")
+# setwd("src/naomi-simple_mmd")
+
+rmarkdown::render("mmd.Rmd")

### 2023-05-05 src/naomi-simple_psis/script.R

@@ -1,0 +1,5 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("naomi-simple_psis")
+# setwd("src/naomi-simple_psis")
+
+rmarkdown::render("psis.Rmd")

### 2023-05-05 src/naomi-simple_exceedance/exceedance.Rmd
Second 90 and high incidence exceedance #35

---
title: "Exceedance probability comparison for the simplified Naomi model"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**  We have run the simplified Naomi model using a range of inference methods: `TMB`, `aghq`, `adam` and `tmbstan`.

  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using exceedance probabilities.
---

# Background

We compare the inference results from `TMB`, `aghq`, `adam`, and `tmbstan`.
Import these inference results as follows:

```{r}
tmb <- readRDS("depends/tmb.rds")
aghq <- readRDS("depends/aghq.rds")
tmbstan <- readRDS("depends/tmbstan.rds")
```

# Second 90

The 90-90-90 treatment targets are that

* 90% of PLHIV know their status
* 90% of PLHIV who kno wtheir status are on antiretroviral therapy (ART)
* 90% of PLHIV on ART have suppressed viral load

To meet the second 90, we require that $90\% \times 90\% = 81\%$ of PLHIV are on ART.

```{r}
mf_out <- tmb$naomi_data$mf_out

mf_out_fine <- mf_out %>%
  tibble::rownames_to_column("id") %>%
  mutate(id = as.numeric(id)) %>%
  filter(
    area_id %in% paste0("MWI_4_", 1:32, "_demo"),
    sex %in% c("male", "female"),
    age_group %in%
      c(
        "Y000_004", "Y005_009", "Y010_014", "Y015_019", "Y020_024", "Y025_029",
        "Y025_034", "Y030_034", "Y035_039", "Y040_044", "Y045_049", "Y050_054",
        "Y055_059", "Y060_064", "Y065_069", "Y070_074", "Y075_079", "Y080_999"
      )
  )

tmbstan_second90 <- apply(tmbstan$mcmc$sample$alpha_t1_out[mf_out_fine$id, ], 1, function(x) sum(x > 0.81) / length(x))
tmb_second90 <- apply(tmb$fit$sample$alpha_t1_out[mf_out_fine$id, ], 1, function(x) sum(x > 0.81) / length(x))
aghq_second90 <- apply(aghq$quad$sample$alpha_t1_out[mf_out_fine$id, ], 1, function(x) sum(x > 0.81) / length(x))

plot(tmbstan_second90, tmb_second90)
plot(tmbstan_second90, aghq_second90)

cor(tmbstan_second90, tmb_second90)
cor(tmbstan_second90, aghq_second90)

mse_tmb_second90 <- mean((tmbstan_second90 - tmb_second90)^2)
mse_aghq_second90 <- mean((tmbstan_second90 - aghq_second90)^2)
round(100 * (mse_aghq_second90 - mse_tmb_second90) / mse_tmb_second90)

rmse_tmb_second90 <- sqrt(mse_tmb_second90)
rmse_aghq_second90 <- sqrt(mse_aghq_second90)
round(100 * (rmse_aghq_second90 - rmse_tmb_second90) / rmse_tmb_second90)
```

# High incidence strata

Above 1% incidence is considered high, and could be met with intensified interventions.

* Very high (>3%)
* High (1-3%)
* Moderate (0.3-1%)
* Low (<0.3%)

```{r}
tmbstan_1inc <- apply(tmbstan$mcmc$sample$lambda_t1_out[mf_out_fine$id, ], 1, function(x) sum(x > 0.01) / length(x))
tmb_1inc <- apply(tmb$fit$sample$lambda_t1_out[mf_out_fine$id, ], 1, function(x) sum(x > 0.01) / length(x))
aghq_1inc <- apply(aghq$quad$sample$lambda_t1_out[mf_out_fine$id, ], 1, function(x) sum(x > 0.01) / length(x))

plot(tmbstan_1inc, tmb_1inc)
plot(tmbstan_1inc, aghq_1inc)

cor(tmbstan_1inc, tmb_1inc)
cor(tmbstan_1inc, aghq_1inc)

mse_tmb_1inc <- mean((tmbstan_1inc - tmb_1inc)^2)
mse_aghq_1inc <- mean((tmbstan_1inc - aghq_1inc)^2)
round(100 * (mse_aghq_1inc - mse_tmb_1inc) / mse_tmb_1inc)

rmse_tmb_1inc <- sqrt(mse_tmb_1inc)
rmse_aghq_1inc <- sqrt(mse_aghq_1inc)
round(100 * (rmse_aghq_1inc - rmse_tmb_1inc) / rmse_tmb_1inc)
```

# Unmet treatment need

Number of people needing treatment is a function of ART coverage, HIV prevalence and population size.
What summaries of unmet treatment need might be important to calculate?

```{r}
#' To-do!
```

### 2023-05-05 make/_make.R

@@ -83,6 +83,7 @@ orderly::orderly_commit(id)
 #' Results
 run_commit_push("naomi-simple_mcmc")
 run_commit_push("naomi-simple_ks")
+run_commit_push("naomi-simple_exceedance")
 run_commit_push("naomi-simple_psis")
 run_commit_push("naomi-simple_mmd")
 run_commit_push("naomi-simple_model-checks")

### 2023-05-09 src/docs_paper/appendix-word.Rmd
Add word version of appendix for Rachel

--- 2023-05-04 src/docs_paper/appendix.Rmd
@@
 ---
-title: Appendix to "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
+title: "Appendix to \"Fast approximate Bayesian inference for small-area estimation
+  of HIV indicators using the Naomi model\""
 author:
@@
 - Seth R. Flaxman^[Department of Computer Science, Oxford University]
-  - Jeffrey W. Eaton^[Department of Infectious Disease Epidemiology, Imperial College London]
+- Jeffrey W. Eaton^[Department of Infectious Disease Epidemiology, Imperial College
+  London]
 output:
+  word_document:
+    toc: yes
   pdf_document:
@@
 \renewcommand{\thesection}{S\arabic{section}}
-```
-
-\newpage
-
-# Inferential algorithm
-
-```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."}
-knitr::include_graphics("depends/algorithm-flowchart.pdf")
 ```
@@
 
-```{r simplified-naomi, fig.cap="Directed acyclic graph describing the simplified Naomi model (work in progress)."}
-knitr::include_graphics("depends/simplified-naomi.pdf")
-```
-
-\newpage
-
-# \textsc{C++} `TMB` user template
-
 <!-- Use table(names(tmb$fit$par.full)) to check the below table -->
@@
 
-This section contains C++ `TMB` code for the negative log-posterior of the simplified Naomi model
-For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used in Section \ref{sec:math} and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
-For further reference on the `TMB` software see `https://kaskr.github.io/adcomp/_book`, or on the method see @kristensen2016tmb.
-
-\newpage
-
-```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
-```
-
-\newpage
-
-# MCMC convergence and suitability
-
-We assessed MCMC convergence and suitability using a range of graphical and numerical tests.
-These included the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
-
-For the time being, this analysis is available from `athowes.github.io/elgm-inf/mcmc-convergence`.
-Once the MCMC results are finalised, the analysis will be moved to this appendix and expanded upon.
-The following draft text and references may be useful in that expanded write-up.
-
-- Improved $\hat R$ statistic from @vehtari2021rank. Recommended only to use the sample if the value is less than 1.05.
-- ESS and autocorrelation related to efficiency
-- Traceplots helpful for diagnosis of problems
-- Pairs density plots helpful for understanding relationships between parameters, including possible identifiability issues
-- Energy plot from @betancourt2017conceptual
-
-\newpage
-
 # References {#references .unnumbered}

### 2023-05-09 src/naomi-simple_psis/psis.Rmd
Import objfull into PSIS notebook, and look at lp__ from tmbstan #41

@@
 aghq <- readRDS("depends/aghq.rds")
-adam <- readRDS("depends/adam.rds")
-adam_time <- adam$time
-adam <- adam$adam
 tmbstan <- readRDS("depends/tmbstan.rds")
@@
 
-## Run details {.tabset}
-
-For more information about the conditions under which these results were generated, see:
-
-### `TMB`
-
-```{r}
-dependency_details <- function(i) {
-  report_name <- names(depends[[i]])
-  print(paste0("Inference results obtained from ", report_name, " with the query ", depends[[i]][[report_name]]$id))
-  report_id <- orderly::orderly_search(query = depends[[i]][[report_name]]$id, report_name)
-  print(paste0("Obtained report had ID ", report_id, " and was run with the following parameters:"))
-  print(orderly::orderly_info(report_id, report_name)$parameters)
-}
-
-dependency_details(1)
-```
-
-### `aghq`
-
-```{r}
-dependency_details(2)
-```
-
-### `adam`
-
-```{r}
-dependency_details(3)
-```
-
-### `tmbstan`
-
-```{r}
-tmbstan_details <- dependency_details(4)
-tmbstan_details
-```
-
 # Pareto-smoothed importance sampling
@@
 For each sample we are going to want to evaluate the log-likelihood, so that we can calculate the log-likelihood ratios.
-We can extract the `TMB` objective function for the log-likelihood as follows:
+
+Although we can extract a `TMB` objective function from `tmb` directly, it would correspond to the Laplace approximation.
+As such we use a new objective function using `naomi_simple.cpp` directly:
 
 ```{r}
-tmb$fit$obj$fn()
-#' To write!
+objfull <- readRDS("depends/objfull.rds")
+objfull$fn(objfull$par)
 ```
+
+The samples from `tmbstan` have the following values of `lp__`.
+These should be equal to what would be obtained by putting the sample into `objfull$fn` (check this):
+
+```{r}
+tmbstan_samples <- rstan::extract(tmbstan$mcmc$stanfit)
+
+data.frame(
+  y = tmbstan_samples$lp__,
+  x = 1:length(tmbstan_samples$lp__)
+) %>%
+  ggplot(aes(x = x, y = y)) +
+    geom_point(alpha = 0.4) +
+    labs(y = "Log-likelihood", x = "Draw number") +
+    theme_minimal() 
+```
+
+Need to write a function which reformats `TMB` or `tmbstan` samples to be evaluated using `objfull$fn`:
+
+```{r}
+objfull$par
+
+#' TODO
+```

### 2023-05-09 src/naomi-simple_psis/psis.Rmd
Don't need naomi_simple.cpp here now that there is objfull #41

@@
 abstract: |
-  **Background**  We have run the simplified Naomi model using a range of inference methods: `TMB`, `aghq`, `adam` and `tmbstan`.
+  **Background**  We have run the simplified Naomi model using a range of inference methods: `TMB`, `aghq`, and `tmbstan`.
 
@@
 
-We compare the inference results from `TMB`, `aghq`, `adam`, and `tmbstan`.
+We compare the inference results from `TMB`, `aghq`, and `tmbstan`.
 Import these inference results as follows:

### 2023-05-09 src/naomi-simple_mmd/mmd.Rmd
Calculate MMD with kernlab #42

@@
 abstract: |
-  **Background**  We have run the simplified Naomi model using a range of inference methods: `TMB`, `aghq`, `adam` and `tmbstan`.
+  **Background**  We have run the simplified Naomi model using a range of inference methods: `TMB`, `aghq` and `tmbstan`.
 
@@
 stopifnot(names(tmb$fit$sample) == names(aghq$quad$sample))
-stopifnot(names(tmb$fit$sample) == names(adam$adam$sample))
 stopifnot(names(tmb$fit$sample) == names(tmbstan$mcmc$sample))
-```
-
-## Run details {.tabset}
-
-For more information about the conditions under which these results were generated, see:
-
-### `TMB`
-
-```{r}
-dependency_details <- function(i) {
-  report_name <- names(depends[[i]])
-  print(paste0("Inference results obtained from ", report_name, " with the query ", depends[[i]][[report_name]]$id))
-  report_id <- orderly::orderly_search(query = depends[[i]][[report_name]]$id, report_name)
-  print(paste0("Obtained report had ID ", report_id, " and was run with the following parameters:"))
-  print(orderly::orderly_info(report_id, report_name)$parameters)
-}
-
-dependency_details(1)
-```
-
-### `aghq`
-
-```{r}
-dependency_details(2)
-```
-
-### `adam`
-
-```{r}
-dependency_details(3)
-```
-
-### `tmbstan`
-
-```{r}
-tmbstan_details <- dependency_details(4)
-tmbstan_details
 ```
@@
 ```{r}
-#' To write!
+mmd1 <- kernlab::kmmd(matrix(tmb$fit$sample$beta_rho), matrix(tmbstan$mcmc$sample$beta_rho))
+mmd2 <- kernlab::kmmd(matrix(aghq$quad$sample$beta_rho), matrix(tmbstan$mcmc$sample$beta_rho))
+
+mmd1
+mmd2
 ```

### 2023-05-09 src/docs_19-05-23-turing/19-05-23-turing.Rmd
Start on ATI presentation #37

--- 2023-04-25 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
@@
 ---
-title: "Deterministic Bayesian inference methods for the Naomi model"
+title: "Fast approximate Bayesian inference of HIV indicators"
 author: "Adam Howes"
@@
 
-# The Naomi model
+#
 
-:::::::::::::: {.columns}
-
-::: {.column width=.65}
-
-* Naomi is a complicated spatio-temporal evidence synthesis model
-* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
-* Fast inference is important to allow for interactive review and development of estimates
-* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2016tmb]
-
-:::
-
-::: {.column width=.35}
-
-```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'}
-knitr::include_graphics("naomi_hex.png")
+```{r, echo=FALSE, fig.cap="Estimates of HIV indicators at a district level produced using the Naomi model.", out.width = '90%'}
+knitr::include_graphics("depends/naomi_results.png")
 ```
-
-:::
-
-::::::::::::::
 
@@
 
-# Why do we use `TMB`
+# Goal
 
-1. It runs \textcolor{hilit}{quickly}
-2. It is \textcolor{hilit}{flexible} enough to be compatible with the model
-3. We don't have better viable options
+* Fast approximate Bayesian inference for a complex, spatiotemporal, evidence synthesis model
 
-# What problem we are trying to solve
+# Approach
 
-* Ideally we want exact Bayesian inference: compute the posterior distribution of the parameters of the model given the data
-* Computationally this amounts to solving a difficult integral
-* We can't solve this, but \textcolor{hilit}{we can give approximate answers}
+1. The marginal Laplace approximation
+2. Adaptive Gauss-Hermite quadrature
+3. Principle components analysis
 
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-
-Goal: approximate this integral for Naomi better than \texttt{TMB} does, without taking too long. In doing so, more accurately reflect uncertainty over hyperparameters
-
-\end{tcolorbox}
-\end{center}
-
-# Two deterministic methods
-
-* We use two deterministic^[In contrast to the most famous approximate Bayesian inference method, Markov chain Monte Carlo, which is fundamentally stochastic.] methods to approximate our integral
-
-1. The Laplace approximation
-2. Quadrature
-
-# The Laplace approximation
-
-* If you pretend the posterior distribution is a Gaussian, computation is easy
+# The marginal Laplace approximation
 
@@
 
-```{r message=FALSE, echo=FALSE,  fig.cap="A Gamma prior with $a = 3$ and $b = 1$."}
-cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
-
-a <- 3
-b <- 1
-
-prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) +
-  stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) +
-  annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1)", col = cbpalette[1], size = 5) +
-  labs(x = "", y = "") +
-  theme_minimal()
-
-prior
-```
-
-#
-
-```{r message=FALSE, echo=FALSE,  fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."}
-set.seed(2)
-y <- rpois(3, lambda = 2)
-
-posterior <- prior +
-  geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) +
-  stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) +
-  annotate("text", x = 5, y = 0.25, label = "Gamma(9, 4)", col = cbpalette[2], size = 5)
-
-posterior
-```
-
-#
-
-```{r message=FALSE}
-fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE)
-
-# Here we are using numerical derivatives
-ff <- list(
-  fn = fn,
-  gr = function(x) numDeriv::grad(fn, x),
-  he = function(x) numDeriv::hessian(fn, x)
-)
-
-opt_bfgs <- aghq::optimize_theta(
-  ff, 1, control = aghq::default_control(method = "BFGS")
-)
-```
-
-# Laplace approximation
-
-```{r}
-laplace <- posterior +
-  stat_function(
-    data = data.frame(x = c(0, 10)),
-    aes(x),
-    fun = dnorm,
-    n = 500,
-    args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)),
-    col = cbpalette[3]
-  )
-```
-
-# Laplace approximation
-
-```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation in this case is good near the mode but not in the tails."}
-laplace +
-  annotate("text", x = 5.5, y = 0.35, label = "Laplace approximation", col = cbpalette[3], size = 5)
-```
-
-# Computation of the Laplace approximation
-
-* This computation was simple, and involved
-
-1. Optimising a function
-2. Taking the mode and the Hessian at the mode
-
-# The marginal Laplace approximation
-
-* If we don't want to pretend the whole posterior distribution is Gaussian, another option is to pretend \textcolor{hilit}{some of its marginals are}
-* This is how `TMB` works: it's up to the user to choose which parameters should be Gaussian using the `random` option
-
-# Which parameters should we treat as Gaussian?
-
-* In spatio-temporal statistics we have data indexed by space and time
-* We use random effects also indexed by space and time to model this data
-* Spatio-temporal fields can be big
-* Willing to make assumptions about how things vary over spacetime
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-
-$\implies$ treat the latent field parameters as Gaussian! That's the majority of the integral done.
-
-\end{tcolorbox}
-\end{center}
-
-# What about the hyperparameters?
-
-* `TMB` uses optimisation to find the hyperparameters which maximise the marginal Laplace approximation
-* This is the "outer" optimisation loop, where the "inner" is for computation of the Gaussian distribution
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-
-Inference for the latent field is based on a single value of the hyperparameters (the mode) -- so called empirical Bayes $\implies$ no uncertainty in the hyperparameters taken into account!
-How can you sleep at night.
-
-\end{tcolorbox}
-\end{center}
-
-# Quadrature
-
-* This brings us to our other method for solving integrals deterministically
-* Say we have a function, then quadrature has two ingredients
-
-1. Nodes: points to evaluate the function at
-2. Weights: importance of function evaluation at that point
-
-* You evaluate the function at the nodes, then do a weighted sum to calculate your integral
-
-# Trapezoid rule example
-
-* Let's compute $\int_0^\pi \sin(x) \text{d}x = 2$ using quadrature
-
-```{r}
-trapezoid_rule <- function(x, spacing) {
-  # Assumes nodes are evenly spaced
-  w <- rep(spacing, length(x)) # Weights given by space between nodes
-  w[1] <- w[1] / 2 # Apart from the first which is halved
-  w[length(x)] <- w[length(x)] / 2 # And the last, also halved
-  sum(w * x) # Compute the weighted sum
-}
-```
-
-#
-
-```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."}
-finegrid <- seq(0, pi, length.out = 1000)
-
-plot <- data.frame(x = finegrid, y = sin(finegrid)) %>%
-  ggplot(aes(x = x, y = y)) +
-   geom_line(col = cbpalette[1]) +
-   theme_minimal() +
-   labs(x = "x", y = "sin(x)")
-
-trapezoid_plot <- function(N) {
-  grid <- seq(0, pi, length.out = N)
-  int <- trapezoid_rule(x = sin(grid), spacing = grid[2] - grid[1])
-
-  plot + 
-    geom_bar(
-      data = data.frame(x = grid, y = sin(grid)),
-      aes(x = x, y = y), alpha = 0.7, stat = "identity",
-      inherit.aes = FALSE, fill = cbpalette[2]) +
-    theme_minimal() +
-    labs(
-      subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"),
-      x = "x", y = "sin(x)"
-    )
-}
-
-trapezoid_plot(N = 10)
-```
-
-#
-
-```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."}
-trapezoid_plot(N = 30)
-```
-
-#
-
-```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."}
-trapezoid_plot(N = 100)
-```
-
-# Adaptive Gauss-Hermite quadrature
-
-* \textcolor{hilit}{Gauss-Hermite quadrature} is a method for picking nodes and weights based on the theory of polynomial interpolation
-* It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution
-* The \textcolor{hilit}{adaptive} part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data
-* Implemented by the `aghq` package [@stringer2021implementing]
-
-#
-
-```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points in two dimensions with $k = 3$."}
+```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted Gauss-Hermite points in two dimensions with $k = 3$."}
 mu <- c(1, 1.5)
@@
 
-# The plan?
-
-:::::::::::::: {.columns}
-
-::: {.column width=.55}
-
-* Write and implement an algorithm for fast approximate Bayesian inference using the Laplace approximation and quadrature for the Naomi model
-* Use `TMB` for writing the model in C++, and implementing the Laplace approximation via automatic differentitation
-* Use adaptive Gauss-Hermite quadrature to integrate the hyperparameters
-
-:::
-
-::: {.column width=.45}
-
-```{r, echo=FALSE, fig.cap="One of Canada's top engineering schools (though ungraded)", out.width = '90%'}
-knitr::include_graphics("the-plan.png")
-```
-
-:::
-
-::::::::::::::
-
-# Long prophecised
-
-> My main comment is that several aspects of the computational machineery that is presented by Rue and his colleagues **could benefit from the use of a numerical technique known as automatic differentiation (AD)** ...
-By the use of AD one could obtain a system that is automatic from a user's perspective...
-the benefit would be a fast, flexible and easy-to-use system for doing Bayesian analysis in models with Gaussian latent variables
-
--- Hans J. Skaug (coauthor of `TMB`), RSS discussion of @rue2009approximate
-
-# One challenge
-
-* For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
-* One proposed solution is to use \textcolor{hilit}{principal components analysis} (PCA) and keep only the first $s < 24$ dimensions
-* We will use $k = 3$ points per dimension ($k = 2$ doesn't include the mode)
-
 #
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what PCA-AGHQ looks like for the 2D example, keeping only the first principal component, $s = 1$."}
+```{r, echo=FALSE, message=FALSE, fig.cap="Now keeping only the first principal component, $s = 1$."}
 gg4 <- mvQuad::createNIGrid(2, "GHe", level = c(3, 1))
@@
 
-```{r, echo=FALSE, fig.cap="This Scree plot suggests 10 or so dimensions is enough. We use $s = 8$. Avoiding long computation times is still important.", out.width = '80%'}
+```{r, echo=FALSE, fig.cap="Scree plot suggests 10 or so dimensions is enough. We use $s = 8$ to avoid long computation times.", out.width = '80%'}
 knitr::include_graphics("depends/tv-plot.png")
@@
 
-# Comparison of posterior distributions
-
-* We run HMC for as long as it takes to get a good answer (3 days)
-* Compare output of `TMB` with `aghq`: which is more similar to HMC?
-
-#
-
-```{r, echo=FALSE, fig.cap="Histogram of sample draws from each method for one latent field parameter.
-Which of \\texttt{TMB} and \\texttt{aghq} is closer to \\texttt{tmbstan}?
-Hard to say.", out.width = '75%'}
-knitr::include_graphics("beta-alpha-histogram.png")
-```
-
-#
-
-```{r, echo=FALSE, fig.cap="The empirical cumulative distribution is slightly more enlightening.", out.width = '80%'}
-knitr::include_graphics("ecdf-plot.png")
-```
-
-#
-
-```{r, echo=FALSE, fig.cap="The Kolmogorov-Smirnov test statistic is the maximum difference between ECDFs.", out.width = '80%'}
-knitr::include_graphics("ecdf-diff-plot.png")
-```
-
-#
-
-```{r, echo=FALSE, fig.cap="Generating the KS test statistic for all latent field parameters, on average \\texttt{aghq} does better.", out.height = '80%'}
-knitr::include_graphics("ks-density-plot.png")
-```
-
-# Conclusions
-
-* Developed an approximate Bayesian inference method more accurate than `TMB` and faster than MCMC -- we're on the Pareto frontier!
-* Scaled hyperparameter grid to >20 dimesions using PCA
-* Relatively straightforward implementation using `aghq` R package, compatible with any model that has a `TMB` template
-
-# Future work
-
-* More comprehensive inference comparison, particularly of model outputs
-  * Maximum mean discrepancy
-  * Pareto-smoothed importance sampling
-  * Simulation-based calibration
-* Extension to Laplace marginals via @wood2020simplified
-  * May be more accurate, but have to think about computational costs
-
 # Thanks for listening!
 
-* Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
-* Code and notebooks for this project are available at `athowes.github.io/elgm-inf`
-
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-
-Let me know if you'd be up for being an early reader!
-
-\end{tcolorbox}
-\end{center}
-
-# Risk group retrospective
-
-```{r, echo=FALSE, fig.cap="Available at \\texttt{athowes.github.io}", out.width = '90%'}
-knitr::include_graphics("blog-post.png")
+```{r echo=FALSE, out.width = "300px", fig.align='center'}
+knitr::include_graphics("mrc-gida-icl-mlgh.png")
 ```

### 2023-05-09 src/docs_19-05-23-turing/temp.R

@@ -1,159 +1,0 @@
-#' Temporary plots for presentation
-
-#' Histogram plot
-colours <- c("#56B4E9", "#009E73", "#E69F00", "#CC79A7")
-
-par <- "beta_alpha"
-i <- 1
-
-if(!is.null(i)) {
-  par_name <- paste0(par, "[", i, "]")
-
-  df_compare <- rbind(
-    data.frame(method = "TMB", samples = as.numeric(tmb$fit$sample[[par]][i, ])),
-    data.frame(method = "aghq", samples = as.numeric(aghq$quad$sample[[par]][i, ])),
-    data.frame(method = "adam", samples = as.numeric(adam$sample[[par]][i, ])),
-    data.frame(method = "tmbstan", samples = as.numeric(tmbstan$mcmc$sample[[par]][i, ]))
-  )
-} else {
-  par_name <- paste0(par)
-
-  df_compare <- rbind(
-    data.frame(method = "TMB", samples = as.numeric(tmb$fit$sample[[par]])),
-    data.frame(method = "aghq", samples = as.numeric(aghq$quad$sample[[par]])),
-    data.frame(method = "adam", samples = as.numeric(adam$sample[[par]])),
-    data.frame(method = "tmbstan", samples = as.numeric(tmbstan$mcmc$sample[[par]]))
-  )
-}
-
-df_compare$method <- factor(df_compare$method, levels = c("TMB", "aghq", "adam", "tmbstan"))
-
-mean <- df_compare %>%
-  filter(method == "tmbstan") %>%
-  summarise(mean = mean(samples)) %>%
-  pull(mean) %>%
-  round(digits = 3)
-
-sd <- df_compare %>%
-  filter(method == "tmbstan") %>%
-  summarise(sd = sd(samples)) %>%
-  pull(sd) %>%
-  round(digits = 3)
-
-rhat <- tryCatch(round(rhats[[par_name]], 3), error = function(e) return("Missing!"))
-ess <- tryCatch(round(ess[[par_name]], 3), error = function(e) return("Missing!"))
-
-histogram_plot <- df_compare %>%
-  filter(method != "adam") %>%
-  ggplot(aes(x = samples, fill = method, col = method)) +
[…]

### 2023-05-10 src/naomi-simple_exceedance/exceedance.Rmd
Short notes on excursions #35

@@
 ```
+
+# Excursion and contour uncertainty regions
+
+Based on @bolin2015excursion.
+
+* $D_m = \{s: P(x(s) > u) \geq 1 - \alpha\}$ runs into the problem of multiple hypothesis testing
+* Care about $P(x(s) > u, s \in D_m)$ which is smaller than $1 - \alpha$
+* You can use the marginals and adjust the threshold to control type 1 error, false discovery rate, or posteior probabilites
+* Bolin and Lindgren use sequential importance sampling to estimate the joint probabilities
+* $x(s) > u$ is called a positive excursion set, and $x(s) < u$ is called a negative excursion set
+* The `excursions` package can calculate excursion sets, contour credible regions and level avoiding sets, excursion functions, contour maps and their quality measures, and simultaneous confidence bands

### 2023-05-10 src/naomi-simple_mmd/mmd.Rmd
Update MMD report #42

@@
 tmbstan <- readRDS("depends/tmbstan.rds")
-
-depends <- yaml::read_yaml("orderly.yml")$depends
 ```
@@
 ```{r}
-mmd1 <- kernlab::kmmd(matrix(tmb$fit$sample$beta_rho), matrix(tmbstan$mcmc$sample$beta_rho))
-mmd2 <- kernlab::kmmd(matrix(aghq$quad$sample$beta_rho), matrix(tmbstan$mcmc$sample$beta_rho))
+par_samp_matrix <- function(sample) {
+  x <- sample[!(stringr::str_ends(names(sample), "_out") | stringr::str_ends(names(sample), "_ll"))]
+  do.call(rbind, x)
+}
 
-mmd1
-mmd2
+tmb_samples <- par_samp_matrix(tmb$fit$sample)
+aghq_samples <- par_samp_matrix(aghq$quad$sample)
+tmbstan_samples <- par_samp_matrix(tmbstan$mcmc$sample)
+
+#' Keep only the final 1000 samples from tmbstan
+tmbstan_samples <- tmbstan_samples[, 4001:5000]
+
+mmd_tmb <- kernlab::kmmd(tmb_samples, tmbstan_samples, kernel = "rbfdot")
+mmd_aghq <- kernlab::kmmd(aghq_samples, tmbstan_samples, kernel = "rbfdot")
+
+#' Take sigma to be the mean estimated for the two different methods
+kpar <- list(sigma = mean(mmd_tmb@kernelf@kpar$sigma, mmd_aghq@kernelf@kpar$sigma))
+
+#' Rerun with sigma the same across the different tests
+mmd_tmb <- kernlab::kmmd(tmb_samples, tmbstan_samples, kernel = "rbfdot", kpar = kpar, alpha = 0.05)
+mmd_aghq <- kernlab::kmmd(aghq_samples, tmbstan_samples, kernel = "rbfdot", kpar = kpar, alpha = 0.05)
+
+mmd_tmb
+mmd_aghq
+
+mmd_tmb@mmdstats
+mmd_aghq@mmdstats
+
+round(100 * (mmd_aghq@mmdstats[1] - mmd_tmb@mmdstats[1]) / mmd_tmb@mmdstats[1])
+round(100 * (mmd_aghq@mmdstats[2] - mmd_tmb@mmdstats[2]) / mmd_tmb@mmdstats[2])
 ```

### 2023-05-10 src/naomi-simple_psis/psis.Rmd
Get NLL for all methods, step of PSIS #41

@@
 aghq <- readRDS("depends/aghq.rds")
-tmbstan <- readRDS("depends/tmbstan.rds")
-
-depends <- yaml::read_yaml("orderly.yml")$depends
+tmbstan <- readRDS("depends/tmbstan.rds")å
 ```
@@
 stopifnot(names(tmb$fit$sample) == names(aghq$quad$sample))
-stopifnot(names(tmb$fit$sample) == names(adam$adam$sample))
 stopifnot(names(tmb$fit$sample) == names(tmbstan$mcmc$sample))
@@
 Although we can extract a `TMB` objective function from `tmb` directly, it would correspond to the Laplace approximation.
-As such we use a new objective function using `naomi_simple.cpp` directly:
+Instead we use `objfull` obtained from a previous report, and load the `naomi_simple` DLL:
 
 ```{r}
+TMB::compile("naomi_simple.cpp")
+dyn.load(TMB::dynlib("naomi_simple"))
+
 objfull <- readRDS("depends/objfull.rds")
@@
 ```{r}
-tmbstan_samples <- rstan::extract(tmbstan$mcmc$stanfit)
+tmbstan_extract <- rstan::extract(tmbstan$mcmc$stanfit)
 
 data.frame(
-  y = tmbstan_samples$lp__,
-  x = 1:length(tmbstan_samples$lp__)
+  y = tmbstan_extract$lp__,
+  x = 1:length(tmbstan_extract$lp__)
 ) %>%
@@
 ```{r}
-objfull$par
+par_samp_matrix <- function(sample) {
+  # Remove the Naomi outputs, ending with either _out or _ll
+  x <- sample[!(stringr::str_ends(names(sample), "_out") | stringr::str_ends(names(sample), "_ll"))]
+  # Reorder to be the same as objfull$par
+  x <- x[unique(names(objfull$par))]
+  # Bind rows together
+  do.call(rbind, x)
+}
 
-#' TODO
+tmb_samples <- par_samp_matrix(tmb$fit$sample)
+aghq_samples <- par_samp_matrix(aghq$quad$sample)
+tmbstan_samples <- par_samp_matrix(tmbstan$mcmc$sample)
+
+tmb_ll <- apply(tmb_samples, 2, FUN = function(x) -1 * objfull$fn(x))
+aghq_ll <- apply(aghq_samples, 2, FUN = function(x) -1 * objfull$fn(x))
+tmbstan_ll <- apply(tmbstan_samples, 2, FUN = function(x) -1 * objfull$fn(x))
+
+data.frame(
+  y = c(tmb_ll, aghq_ll, tmbstan_ll),
+  x = c(1:length(tmb_ll), 1:length(aghq_ll), 1:length(tmbstan_ll)),
+  method = c(rep("TMB", times = length(tmb_ll)), rep("aghq", times = length(aghq_ll)), rep("tmbstan", times = length(tmbstan_ll)))
+) %>%
+  ggplot(aes(x = x, y = y, col = method)) +
+    geom_point(alpha = 0.4) +
+    facet_grid(~ method, scales = "free_x") +
+    scale_colour_manual(values = c("#56B4E9", "#009E73", "#E69F00")) +
+    labs(y = "Log-likelihood", x = "Draw number", col = "Method") +
+    theme_minimal()
+
+#' Check that these are the same
+plot(sort(tmbstan_extract$lp__), sort(tmbstan_ll))
+
+-1 * mean(tmb_ll)
+-1 * mean(aghq_ll)
+-1 * mean(tmbstan_ll)
+
+#' Run PSIS
+#' Need to get log-likelihood under the proposal
+#' A Gaussian for TMB
+#' A mixture of Gaussians for aghq
+lw <- tmb_ll
+loo::psis(log_ratios = lw, r_eff = 1)
+
+lw <- aghq_ll
+loo::psis(log_ratios = lw, r_eff = 1)
 ```

### 2023-05-11 src/naomi-simple_psis/psis.Rmd
First try at producing PSIS k hat statistic #41

--- 2023-05-09 src/naomi-simple_psis/psis.Rmd
@@
 
-Need to write a function which reformats `TMB` or `tmbstan` samples to be evaluated using `objfull$fn`:
+We would like to produce evalautions of the log-likelihood for the `TMB` and `aghq` samples as well.
+To do this, we first create a function which reformats the samples to be evaluated using `objfull$fn`:
 
 ```{r}
-objfull$par
+par_samp_matrix <- function(sample, elements = unique(names(objfull$par))) {
+  # Keep and reorder to be the same as elements
+  x <- sample[elements]
+  # Bind rows together
+  do.call(rbind, x)
+}
 
-#' TODO
+tmb_samples <- par_samp_matrix(tmb$fit$sample)
+aghq_samples <- par_samp_matrix(aghq$quad$sample)
+tmbstan_samples <- par_samp_matrix(tmbstan$mcmc$sample)
 ```
+
+Now we can calculate the log-likelihood by applying this function across the rows of the sample matrices:
+
+```{r}
+tmb_ll <- apply(tmb_samples, 2, FUN = function(x) -1 * objfull$fn(x))
+aghq_ll <- apply(aghq_samples, 2, FUN = function(x) -1 * objfull$fn(x))
+tmbstan_ll <- apply(tmbstan_samples, 2, FUN = function(x) -1 * objfull$fn(x))
+
+data.frame(
+  y = c(tmb_ll, aghq_ll, tmbstan_ll),
+  x = c(1:length(tmb_ll), 1:length(aghq_ll), 1:length(tmbstan_ll)),
+  method = c(rep("TMB", times = length(tmb_ll)), rep("aghq", times = length(aghq_ll)), rep("tmbstan", times = length(tmbstan_ll)))
+) %>%
+  ggplot(aes(x = x, y = y, col = method)) +
+    geom_point(alpha = 0.4) +
+    facet_grid(~ method, scales = "free_x") +
+    scale_colour_manual(values = c("#56B4E9", "#009E73", "#E69F00")) +
+    labs(y = "Log-likelihood", x = "Draw number", col = "Method") +
+    theme_minimal()
+```
+
+To check that this is working as expected, we can compare the answers produced this way to those obtained directly from `lp__` in Stan:
+
+```{r}
+data.frame(lp__ = sort(tmbstan_extract$lp__), objfull = sort(tmbstan_ll)) %>%
+  ggplot(aes(x = lp__, y = objfull)) +
+    geom_point(alpha = 0.4) +
+    labs(title = "Appears these points lie on a straight line") +
+    theme_minimal()
+```
+
+It's interesting to note the differing levels of the log-likelihoods between the three sampling methods.
+Is there a way that this could be interpreted?
+Stan samples from the typical set, whereas `aghq` and `TMB` are sampling from closer to the mode?
+
+```{r}
+-1 * mean(tmb_ll)
+-1 * mean(aghq_ll)
+-1 * mean(tmbstan_ll)
+
+(mean(tmb_ll) - mean(tmbstan_ll)) / 24
+```
+
+To run PSIS we need to get the log-likelihoods under the proposal, as well as the log-likelihoods under the target.
+The proposal for `TMB` is a multivariate Gaussian, but only for the random effects:
+
+```{r}
+r <- tmb$fit$obj$env$random # Indices of the random effects
+par_r <- tmb$fit$par.full[r] # Mode of the random effects
+
+# Hessian of the random effects (this has crashed on me before, so being safe)
+safe <- parallel::mcparallel({tmb$fit$obj$env$spHess(tmb$fit$par.full, random = TRUE)})
+hess_r <- parallel::mccollect(safe, wait = TRUE, timeout = 0, intermediate = FALSE)
+```
+
+For `aghq` it's a mixture of Gaussians:
+
+```{r}
+aghq_modes <- aghq$quad$modesandhessians$mode
+aghq_hessians <- aghq$quad$modesandhessians$H
+aghq_lambda <- exp(aghq$quad$normalized_posterior$nodesandweights$logpost_normalized) * aghq$quad$normalized_posterior$nodesandweights$weights
+```
+
+`aghq_lambda` contains the multinomial weights, which we verify are in fact weights here:
+
+```{r}
+stopifnot(sum(aghq_lambda) - 1 < 10E-9)
+stopifnot(all(aghq_lambda > 0))
+```
+
+Now to construct the log-likelihood, we will use the log-sum-exp trick:
+
+```{r}
+aghq_chol <- lapply(aghq_hessians, Matrix::Cholesky)
+aghq_latent_samples <- par_samp_matrix(aghq$quad$sample, elements = unique(names(aghq_modes[[1]])))
+
+start <- Sys.time()
+mvn_ll <- mapply(sparseMVN::dmvn.sparse, mu = aghq_modes, CH = aghq_chol, MoreArgs = list(x = t(aghq_latent_samples), prec = TRUE, log = TRUE))
+end <- Sys.time()
+
+#' Takes around 2 minutes
+end - start
+
+aghq_q_ll <- apply(mvn_ll, 1, FUN = function(row) matrixStats::logSumExp(log(aghq_weights) + row))
+plot(aghq_q_ll)
+```
+
+Now run the PSIS with `aghq`.
+The importance ratios are $\log(p(\theta, y)) - \log(q(\theta))$:
+
+```{r}
+r <- - aghq_ll - aghq_q_ll
+plot(r)
+
+aghq_psis <- loo::psis(log_ratios = lw, r_eff = 1)
+aghq_psis$diagnostics
+```

### 2023-05-12 src/naomi-simple_exceedance/exceedance.Rmd
Further investigation into second 90 and high incidence exceedance. Finding different behaviour by sex suggests ANC model interaction #35

@@
 aghq_second90 <- apply(aghq$quad$sample$alpha_t1_out[mf_out_fine$id, ], 1, function(x) sum(x > 0.81) / length(x))
+```
 
-plot(tmbstan_second90, tmb_second90)
-plot(tmbstan_second90, aghq_second90)
+Plot the results from the different inference methods:
 
+```{r}
+n <- length(tmbstan_second90)
+stopifnot(length(tmb_second90) == n)
+stopifnot(length(aghq_second90) == n)
+
+df_second90 <- data.frame(
+  prob = c(tmbstan_second90, tmb_second90, aghq_second90),
+  method = c(rep("tmbstan", times = n), rep("TMB", times = n), rep("aghq", times = n)),
+  index = c(1:n, 1:n, 1:n)
+)
+
+df_wide_second90 <- pivot_wider(df_second90, names_from = method, values_from = prob)
+
+plot1 <- ggplot(df_wide_second90, aes(x = tmbstan, y = TMB)) +
+  geom_point(alpha = 0.3) +
+  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
+  labs(title = "Probability of 81% ART coverage in strata") +
+  theme_minimal()
+
+plot2 <- ggplot(df_wide_second90, aes(x = tmbstan, y = aghq)) +
+  geom_point(alpha = 0.3) +
+  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
+  theme_minimal()
+
+plot1 + plot2
+```
+
+Appears that: 1) systematic overestimation of probability by both approximate inference methods 2) `aghq` does better at not underestimating near zero.
+
+What is the correlation?
+
+```{r}
 cor(tmbstan_second90, tmb_second90)
 cor(tmbstan_second90, aghq_second90)
+```
 
-mse_tmb_second90 <- mean((tmbstan_second90 - tmb_second90)^2)
-mse_aghq_second90 <- mean((tmbstan_second90 - aghq_second90)^2)
-round(100 * (mse_aghq_second90 - mse_tmb_second90) / mse_tmb_second90)
+And what about the mean square error and root mean square error?
 
-rmse_tmb_second90 <- sqrt(mse_tmb_second90)
-rmse_aghq_second90 <- sqrt(mse_aghq_second90)
-round(100 * (rmse_aghq_second90 - rmse_tmb_second90) / rmse_tmb_second90)
+```{r}
+(mse_tmb_second90 <- mean((tmbstan_second90 - tmb_second90)^2))
+(mse_aghq_second90 <- mean((tmbstan_second90 - aghq_second90)^2))
+
+(rmse_tmb_second90 <- sqrt(mse_tmb_second90))
+(rmse_aghq_second90 <- sqrt(mse_aghq_second90))
+
+mse_change <- round(100 * (mse_aghq_second90 - mse_tmb_second90) / mse_tmb_second90)
+rmse_change <- round(100 * (rmse_aghq_second90 - rmse_tmb_second90) / rmse_tmb_second90)
+```
+There is a change of `r mse_change`% in the MSE moving from `TMB` to `aghq`, and a change of `r rmse_change`% in the RMSE.
+
+Which strata have exceedance probabilities which are particularly poorly estimated?
+First for `TMB`:
+
+```{r}
+tmb_fails_second90 <- df_wide_second90 %>%
+  mutate(diff = tmbstan - TMB) %>%
+  filter(diff > 0.1) %>%
+  pull(index)
+```
+And for `aghq`:
+
+```{r}
+aghq_fails_second90 <- df_wide_second90 %>%
+  mutate(diff = tmbstan - aghq) %>%
+  filter(diff > 0.1) %>%
+  pull(index)
+```
+
+There are `r sum(!(tmb_fails_second90 %in% aghq_fails_second90))` uniquely badly estimated by `TMB`, `r sum(!(aghq_fails_second90 %in% tmb_fails_second90))` uniquely badly estimated by `aghq`, and `r sum(tmb_fails_second90 %in% aghq_fails_second90)` badly estimated by both.
+Which strata do these correspond to?
+
+```{r}
+unique(mf_out_fine[tmb_fails_second90, ]$area_id)
+unique(mf_out_fine[tmb_fails_second90, ]$sex)
+unique(mf_out_fine[tmb_fails_second90, ]$age_group)
+
+unique(mf_out_fine[aghq_fails_second90, ]$area_id)
+unique(mf_out_fine[aghq_fails_second90, ]$sex)
+unique(mf_out_fine[aghq_fails_second90, ]$age_group)
+```
+Add colour by sex to the above plots:
+
+```{r}
+df_wide_second90 <- df_wide_second90 %>%
+  left_join(
+    mf_out_fine %>%
+      tibble::rowid_to_column("index"),
+    by = "index"
+  )
+
+plot1 <- ggplot(df_wide_second90, aes(x = tmbstan, y = TMB, col = sex)) +
+  geom_point(alpha = 0.3) +
+  scale_color_manual(values = c("#E69F00", "#F0E442")) +
+  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
+  labs(title = "Probability of at least 81% ART coverage in strata", col = "Sex") +
+  theme_minimal() +
+  theme(legend.position = "bottom")
+
+plot2 <- ggplot(df_wide_second90, aes(x = tmbstan, y = aghq, col = sex)) +
+  geom_point(alpha = 0.3) +
+  scale_color_manual(values = c("#E69F00", "#F0E442")) +
+  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
+  guides(col = "none") +
+  theme_minimal()
+
+plot1 + plot2
+```
+Check the RMSE stratified by sex:
+
+```{r}
+df_wide_second90 %>%
+  group_by(sex) %>%
+  summarise(
+    "RMSE TMB (percent)" = 100 * sqrt(mean((tmbstan - TMB)^2)),
+    "RMSE aghq (percent)" = 100 * sqrt(mean((tmbstan - aghq)^2))
+  )
 ```
@@
 
-plot(tmbstan_1inc, tmb_1inc)
-plot(tmbstan_1inc, aghq_1inc)
+n <- length(tmbstan_1inc)
+stopifnot(length(tmb_1inc) == n)
+stopifnot(length(aghq_1inc) == n)
+
+df_1inc <- data.frame(
+  prob = c(tmbstan_1inc, tmb_1inc, aghq_1inc),
+  method = c(rep("tmbstan", times = n), rep("TMB", times = n), rep("aghq", times = n)),
+  index = c(1:n, 1:n, 1:n)
+)
+
+df_wide_1inc <- pivot_wider(df_1inc, names_from = method, values_from = prob)
+
+plot1 <- ggplot(df_wide_1inc, aes(x = tmbstan, y = TMB)) +
+  geom_point(alpha = 0.3) +
+  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
+  labs(title = "Probability of at least 1% incidence in strata") +
+  theme_minimal()
+
+plot2 <- ggplot(df_wide_1inc, aes(x = tmbstan, y = aghq)) +
+  geom_point(alpha = 0.3) +
+  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
+  theme_minimal()
+
+plot1 + plot2
 

### 2023-05-12 src/docs_paper/paper.Rmd
Update abstract focusing on >20 hypers rather than SINLA

@@
 Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
-Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package.
-We propose a new inference method combining adaptive Gauss-Hermite quadrature together with the simplified integrated nested Laplace approximation approach of @wood2020simplified to enable fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but faster to run.
+The model is provided as a tool for countries to input their data to and generate estimates using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package.
+We propose a new inference method extending adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, thereby enabling fast and accurate inference for Naomi and other extended latent Gaussian models.
+Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is substantially faster to run to Hamiltonian Monte Carlo with the No-U-Turn sampler.
 By extending the \texttt{aghq} \textsc{R} package we facilitate easy, flexible use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
-In doing so, we enable inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
 keyword-subclass: | 

### 2023-05-12 src/docs_paper/appendix.Rmd
Move Laplace marginals part to appendix

@@
 
-# Inferential algorithm
-
-```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."}
-knitr::include_graphics("depends/algorithm-flowchart.pdf")
-```
-
-\newpage
-
 # Simplified Naomi model description \label{sec:math}
@@
 
-Let $\nu$ be the most recent national household survey with HIV testing which has taken place in the country of interest.
+Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
 Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
-Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age bands.
-As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
+Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
+As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
 We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities: HIV prevalence $\rho_{x, s, a} \in [0, 1]$, ART coverage $\alpha_{x, s, a} \in [0, 1]$, annual HIV incidence rate $\lambda_{x, s, a} > 0$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
@@
 
-For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
+For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the most recent household survey furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
 $$
@@
 For the Malawi example considered in the main text HIV prevalence and ART coverage data are not available for those aged 65+.
-As a result, there are $|\{\text{0-4}, \ldots, \text{50-54}\}| = 13$ age bands included for the age random effects.
+As a result, there are $|\{\text{0-4}, \ldots, \text{50-54}\}| = 13$ age groups included for the age random effects.
 2.
@@
 
+# Laplace marginals algorithm
+
+1. Calculate the mode, Hessian at the mode, and lower Cholesky
+\begin{align*}
+\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
+\mathbf{H} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
+\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
+\end{align*}
+of the Laplace approximation
+$$
+\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+$$
+where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+\begin{align*}
+\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
+\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
+\end{align*}
+2.
+Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
+If possible $k \geq 3$ is preferred, though the number of grid points scales exponentially with choice of $k$.
+Then, use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
+\begin{equation}
+\tilde p_{\texttt{AQ}}(\y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\btheta(\bu), \y) \omega(\bu). \label{eq:evidence1}
+\end{equation}
+4.
+For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
+A value of $l \geq 4$ is recommended to enable B-spline interpolation.
+Then, for $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
+\begin{align*}
+\hat \x_{-i}(x_i, \btheta) &= \argmax_{\x_{-i}} \log p(\y, x_i, \x_{-i}, \btheta), \\
+\mathbf{H}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
+\end{align*}
+where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
+6. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
+\begin{equation}
+\tilde p_\texttt{AQ}(x_i \, | \, \y) = \frac{\tilde p_\texttt{LA}(x_i, \y)}{\tilde p_{\texttt{AQ}}(\y)}. \label{eq:laplace-marginal}
+\end{equation}
+where
+$$
+\tilde p_\texttt{LA}(x_i, \y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(x_i, \btheta(\bu), \y) \omega(\bu).
+$$
+and
+$$
+\tilde p_\texttt{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\texttt{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
+$$
+Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence in Equation \ref{eq:evidence1} it is more numerically accurate to use the estimate
+$$
+\tilde p_{\texttt{AQ}}(\y) = \sum_{\bv \in \mathcal{Q}(1, l)} \tilde p_\texttt{LA}(x_i(\bv), \y) \omega(\bv)
+$$
+
+7.
+Given $\{x_i(\bv), \tilde p_\texttt{AQ}(x_i(\bv) \, | \, \y)\}_{\bv \in \mathcal{Q}(1, l)}$ create a spline interpolant to each posterior marginal on the log-scale.
+Samples, and thereby relevant posterior marginal summaries, may be obtained using inverse transform sampling.
+
+```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."}
+knitr::include_graphics("depends/algorithm-flowchart.pdf")
+```
+
+\newpage
+
 # References {#references .unnumbered}

### 2023-05-12 src/docs_paper/paper.Rmd
Move Laplace marginals part to appendix

@@
 To mount an effective public health response to the HIV epidemic, it is crucial to have accurate, timely estimates of HIV indicators at the geographic level at which health systems are planned and delivered.
-However, producing these estimates is challenging due to the limitations of all available data sources.
-<!-- Phrasing? -->
+However, producing these estimates is challenging, in large part due to limitations of the available data sources.
 Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they are conducted only every five years or so, with limited sample size at the district level.
@@
 
-Practical requirements for the model, in combination with its relative complexity, present a difficult Bayesian inference problem.
-Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production.
-Markov chain Monte Carlo (MCMC) approaches are prohibitively slow due to the scale of the model and challenging features of its posterior geometry [@neal2003slice].
+The complexity of the model, in combination with practical requirements for its operation, present a difficult Bayesian inference problem.
+First, due to dependence of observations on multiple structured additive predictors, Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast].
+Furthermore, as well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number typically required for use of integrated nested Laplace approximations [@rue2009approximate].
+Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production across a range of countries.
+In this setting, Markov chain Monte Carlo (MCMC) approaches are prohibitively slow, both due to the scale of the model and challenging features of its posterior geometry [@neal2003slice].
 Inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
@@
 
-To obtain fast, accurate Bayesian inferences for the Naomi model we develop an inference methodology which combines adaptive Gauss-Hermite quadrature (AGHQ) with the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified.
-INLA is an approximate inference approach based on nested Laplace approximations and numerical quadrature.
-The key innovation of @rue2009approximate is an approximation which enables accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
-Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on precision matrix of the latent field required for this approximation to be accurate.
-This extension facilitates inference for models like Naomi, which are not quite latent Gaussian models (LGMs) and so were not previously amenable to inference with INLA.
-In particular, due to dependence of observations on multiple structured additive predictors, Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast].
-We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
-Though no theory yet exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2022stochastic using AGHQ.
+To obtain fast, accurate Bayesian inferences for the Naomi model we develop an inference methodology which extends adaptive Gauss-Hermite quadrature (AGHQ) to handle a higher number of hyperparameters.
 We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
 Since `aghq` is designed to naturally interface with `TMB`, use of our method is simple when provided a \textsf{C++} user template for the log-posterior.
-
-Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package [@bachl2019inlabru], INLA within MCMC [@gomez2018markov], and importance sampling with INLA [@berild2022importance], all of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
-`inlabru` approximates non-linear predictors using linearisation, which involves making iterative calls to `R-INLA`.
-<!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
-<!-- Want to say in particular what the class of models that `inlabru` allows you to fit are -->
-<!-- Don't know the answer to this yet, but asking Janine Illian I'm told you don't have to call R-INLA that many times -->
-<!-- I think it'd obviously depend on the exact details of the problem at hand -->
-INLA within MCMC and importance sampling with INLA are suitable for models which are LGMs conditional on some subset of the parameters being fixed.
-<!-- It would or would not be possible to fit the Naomi model in these frameworks. -->
 
@@
 
-@eaton2021naomi specify a joint model linking three small-area estimation models, and defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
-We consider a simplified version defined only at $T_1$ omitting nowcasting and temporal projection, as these time points involve limited additional inference.
-An overview of this simplified model is given below, highlighting the aspects which make it a challenge for existing inferential approaches.
+@eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points $T_i, i = 1, 2, 3$.
+We consider a simplified version defined only at the time of the most recent household survey with HIV testing ($T_1$), omitting nowcasting ($T_2$) and temporal projection ($T_3$), as these time points involve limited additional inference.
+An overview of this simplified model is given below.
 A more complete mathematical description (Appendix S1) as well as a \textsf{C++} template for the log-posterior (Appendix S3) are provided in the supplementary material.
@@
 Consider a country in sub-Saharan Africa where a household survey with complex design has taken place at time $T_1$.
-Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ index five-year age band, and $s \in \mathcal{S}$ index sex.
+Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ index five-year age group, and $s \in \mathcal{S}$ index sex.
 For ease of notation, let $i$ index the finest district-age-sex division included in the model.
-The data we observe may be aggregated over indices $i$.
-Let $\mathcal{I} \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of  indices $i$ for which an observation is reported.
+Let $\mathcal{I} \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported.
 
 Let $N_i \in \mathbb{N}$ be the known, fixed population size.
-We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proporiton of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
+We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
 Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that
@@
 where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
-<!-- The design weights are given by -->
-<!-- $$ -->
-<!-- w_j = \frac{1}{\pi_j} \times \frac{1}{\omega_j}, -->
-<!-- $$ -->
-<!-- where $\pi_j$ is the probability of inclusion and $\omega_j$ is a non-response factor for the particular survey. -->
 We calculate the observed number of outcomes as $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
@@
 \end{align*}
-These processes are informed by likelihoods specified for the following aggregate ANC data from the year of the most recent survey: the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
+These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
+In particular, the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
 We use the binomial working likelihoods
@@
 \end{equation*}
-with $\Psi_i$ the number of pregnant women.
+with $\Psi_i$ the number of pregnant women, which we assume to be fixed.
 
-## ART attendance component \label{sec:art}
+<!-- ## ART attendance component \label{sec:art} -->
 
-We use a multinomial logistic regression model to account for people living with HIV choosing to access ART services outside of the district that they reside in.
-Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
-We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent such that $x \sim x'$. 
-We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a linear predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
-We model aggregate ART attendance data $\dot A_\mathcal{I}$ using a Gaussian approximation to a sum of binomials.
-The sum results both by aggregation over $i \in \mathcal{I}$ and by number of ART clients travelling from district $x'$ to $x$.
-More details regarding this part of the model are provided in Appendix S1.
+<!-- We use a multinomial logistic regression model to account for people living with HIV choosing to access ART services outside of the district that they reside in. -->
+<!-- Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$. -->
+<!-- We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent such that $x \sim x'$.  -->
+<!-- We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a linear predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
+-->
+<!-- We model aggregate ART attendance data $\dot A_\mathcal{I}$ using a Gaussian approximation to a sum of binomials. -->
+<!-- The sum results both by aggregation over $i \in \mathcal{I}$ and by number of ART clients travelling from district $x'$ to $x$. -->
+<!-- More details regarding this part of the model are provided in Appendix S1. -->
 
 # Extended Latent Gaussian models\label{sec:elgm}
+
+## Definitions
 
@@
 
-Naomi is not an LGM, and instead falls into the ELGM class, because:
+## Naomi
 
-1. In the ANC testing component (Section \ref{sec:anc}), the HIV prevalence and ART coverage depend upon the household survey component.
-Specifically, $|\mathcal{J}_i| = 2$ such that for $\theta \in \{\rho, \alpha\}$
-$$
-\mu_i = g_i(\eta^\theta_i, \eta^{\theta^{\text{ANC}}}_i) = \text{logit}^{-1}(\eta^\theta_i + \eta^{\theta^{\text{ANC}}}_i).
-$$
-2. In the household survey component (Section \ref{sec:household}), the HIV incidence rate depends on adult HIV prevalence and adult ART coverage via non-linear transformations.
-3.
+The Naomi model is not an LGM, and instead falls into the ELGM class.
+
+<!-- 1. In the ANC testing component (Section \ref{sec:anc}), the HIV prevalence and ART coverage depend upon the household survey component. -->
+<!-- Specifically, $|\mathcal{J}_i| = 2$ such that for $\theta \in \{\rho, \alpha\}$ -->
+<!-- $$ -->
+<!-- \mu_i = g_i(\eta^\theta_i, \eta^{\theta^{\text{ANC}}}_i) = \text{logit}^{-1}(\eta^\theta_i + \eta^{\theta^{\text{ANC}}}_i). -->
+<!-- $$ -->
+<!-- 2. In the household survey component (Section \ref{sec:household}), the HIV incidence rate depends on adult HIV prevalence and adult ART coverage via non-linear transformations. -->
+<!-- 3.
 Individuals who are taking ART, or have been recently infected, must be HIV positive.
 As a result observations are made taking the product of HIV prevalence with ART coverge or recent HIV infection.
-4. Recent HIV infection is non-linearly linked to both HIV incidence and HIV prevalence.
-5. The ART attendance component (Section \ref{sec:art}) uses a multinomial model.
-6. Processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations aggregated over sets of indices $\mathcal{I}$.
-7. Multiple link functions are used: there is no one $g$.
+-->
+<!-- 4. Recent HIV infection is non-linearly linked to both HIV incidence and HIV prevalence. -->
+<!-- 5. The ART attendance component (Section \ref{sec:art}) uses a multinomial model. -->
+<!-- 6. Processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations aggregated over sets of indices $\mathcal{I}$. -->
+<!-- 7. Multiple link functions are used: there is no one $g$. -->
 
@@
 ## Algorithm\label{sec:algorithm}
-
-1. Calculate the mode, Hessian at the mode, and lower Cholesky
-\begin{align*}
-\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
-\mathbf{H} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
-\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
-\end{align*}
-of the Laplace approximation
-$$
-\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
-$$
-where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
-\begin{align*}
-\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
-\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
-\end{align*}
-2.
-Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
-If possible $k \geq 3$ is preferred, though the number of grid points scales exponentially with choice of $k$.
-Then, use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
-\begin{equation}
-\tilde p_{\texttt{AQ}}(\y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\btheta(\bu), \y) \omega(\bu). \label{eq:evidence1}
-\end{equation}
-4.
-For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
-A value of $l \geq 4$ is recommended to enable B-spline interpolation.
-Then, for $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
-\begin{align*}
-\hat \x_{-i}(x_i, \btheta) &= \argmax_{\x_{-i}} \log p(\y, x_i, \x_{-i}, \btheta), \\
-\mathbf{H}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
-\end{align*}
-where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
-6. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
-\begin{equation}
-\tilde p_\texttt{AQ}(x_i \, | \, \y) = \frac{\tilde p_\texttt{LA}(x_i, \y)}{\tilde p_{\texttt{AQ}}(\y)}. \label{eq:laplace-marginal}
-\end{equation}
-where
-$$
-\tilde p_\texttt{LA}(x_i, \y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(x_i, \btheta(\bu), \y) \omega(\bu).
-$$
-and
-$$
-\tilde p_\texttt{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\texttt{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
-$$
-Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence in Equation \ref{eq:evidence1} it is more numerically accurate to use the estimate
-$$
-\tilde p_{\texttt{AQ}}(\y) = \sum_{\bv \in \mathcal{Q}(1, l)} \tilde p_\texttt{LA}(x_i(\bv), \y) \omega(\bv)
-$$
-
-7.
-Given $\{x_i(\bv), \tilde p_\texttt{AQ}(x_i(\bv) \, | \, \y)\}_{\bv \in \mathcal{Q}(1, l)}$ create a spline interpolant to each posterior marginal on the log-scale.
-Samples, and thereby relevant posterior marginal summaries, may be obtained using inverse transform sampling.
 

### 2023-05-16 src/docs_paper/appendix.Rmd
Check in paper and appendix writing

@@
 \end{align*}
-where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the lag-one correlation parameter.
+where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
 Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
@@
 where $u_{-i}$ is $u$ with the $i$th element removed i.e.
-$(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
+$(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
 We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
@@
 
-We assume the household survey was run according to a complex survey design where each individual $j \in U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$.
+We assume the household survey was run according to a complex survey design where each individual $j$ in the population $U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$.
 Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
-Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then the weighted mean
+Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean is given by
 $$

### 2023-05-16 src/docs_paper/paper.Rmd
Check in paper and appendix writing

@@
 @eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points $T_i, i = 1, 2, 3$.
-We consider a simplified version defined only at the time of the most recent household survey with HIV testing ($T_1$), omitting nowcasting ($T_2$) and temporal projection ($T_3$), as these time points involve limited additional inference.
+We consider a simplified version defined only at the time of the most recent household survey with HIV testing ($T_1$), omitting nowcasting ($T_2$) and temporal projection ($T_3$) which involve limited additional inference.
 An overview of this simplified model is given below.
@@
 Let $N_i \in \mathbb{N}$ be the known, fixed population size.
-We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
-Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that
-\begin{align*}
-\text{logit}(\rho_i) &= \eta^\rho_i, \\
-\text{logit}(\alpha_i) &= \eta^\alpha_i,
-\end{align*}
-for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
-HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
-Finally, let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
-$$
-\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
-$$
+We infer the following unknown HIV indicators using coupled regression equations: HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
+Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that $\text{logit}(\rho_i) = \eta^\rho_i$ and $\text{logit}(\alpha_i) = \eta^\alpha_i$, for certain choice of structured additive predictors.
+HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i$, and depends on adult HIV prevalence and adult ART coverage. 
+Let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
+\begin{equation}
+\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right), \label{eq:kappa}
+\end{equation}
 where the mean duration of recent infection $\Omega_T$ and the proportion of long-term HIV infections misclassified as recent $\beta_T$ are strongly informed by priors for the particular survey.
 
-For $\theta \in \{\rho, \alpha, \kappa\}$ we calculate the weighted, aggregated survey observations
+These processes are informed by household survey data.
+For $\theta \in \{\rho, \alpha, \kappa\}$ let
 $$
-\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j},
+\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}
 $$
-where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
-We calculate the observed number of outcomes as $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
+be weighted, aggregate survey observations, where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
+The observed number of outcomes are $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
 $$
@@
 
-HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset on the logit scale from the general population indicators as follows
+HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset from the general population indicators as follows
 \begin{align*}
@@
 
-<!-- ## ART attendance component \label{sec:art} -->
+## ART attendance component \label{sec:art}
 
-<!-- We use a multinomial logistic regression model to account for people living with HIV choosing to access ART services outside of the district that they reside in. -->
-<!-- Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$. -->
-<!-- We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent such that $x \sim x'$.  -->
-<!-- We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a linear predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
--->
-<!-- We model aggregate ART attendance data $\dot A_\mathcal{I}$ using a Gaussian approximation to a sum of binomials. -->
-<!-- The sum results both by aggregation over $i \in \mathcal{I}$ and by number of ART clients travelling from district $x'$ to $x$. -->
-<!-- More details regarding this part of the model are provided in Appendix S1. -->
+People living with HIV sometimes choose to access ART services outside of the district that they reside in.
+To account for this, we use multinomial logistic regression to model the probabilities of accessing services outside the home district.
+Let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring, denoted by $x \sim x'$.
+The log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ are modelled using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
+Aggregate ART attendance data $\dot A_\mathcal{I}$ is modelled using a Gaussian approximation to a sum of binomials.
+The sum results both by aggregation over $i \in \mathcal{I}$ and by number of ART clients travelling from district $x'$ to $x$.
+More details regarding this part of the model are provided in Appendix S1.
 
 # Extended Latent Gaussian models\label{sec:elgm}
+
+We now describe a popular class of models, and an extension encapsulates the complexities of Naomi.
 
@@
 The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
-Importantly, this mapping allows for the presence of more non-linearity in the model.
-ELGMs are then of the form
+Importantly, this mapping allows for the presence of non-linearity in the model.
+Put together, ELGMs are then of the form
 \begin{align*}
@@
 
-## Naomi
+## Naomi as an ELGM
 
-The Naomi model is not an LGM, and instead falls into the ELGM class.
+Naomi has a lot in common with many LGMS: it is a spatio-temporal model with a large latent field, governed by a smaller number of hyperparameters.
+However, Naomi is not an LGM, and instead falls into the ELGM class, for the following reasons:
 
-<!-- 1. In the ANC testing component (Section \ref{sec:anc}), the HIV prevalence and ART coverage depend upon the household survey component. -->
-<!-- Specifically, $|\mathcal{J}_i| = 2$ such that for $\theta \in \{\rho, \alpha\}$ -->
-<!-- $$ -->
-<!-- \mu_i = g_i(\eta^\theta_i, \eta^{\theta^{\text{ANC}}}_i) = \text{logit}^{-1}(\eta^\theta_i + \eta^{\theta^{\text{ANC}}}_i). -->
-<!-- $$ -->
-<!-- 2. In the household survey component (Section \ref{sec:household}), the HIV incidence rate depends on adult HIV prevalence and adult ART coverage via non-linear transformations. -->
-<!-- 3.
-Individuals who are taking ART, or have been recently infected, must be HIV positive.
-As a result observations are made taking the product of HIV prevalence with ART coverge or recent HIV infection.
--->
-<!-- 4. Recent HIV infection is non-linearly linked to both HIV incidence and HIV prevalence. -->
-<!-- 5. The ART attendance component (Section \ref{sec:art}) uses a multinomial model. -->
-<!-- 6. Processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations aggregated over sets of indices $\mathcal{I}$. -->
-<!-- 7. Multiple link functions are used: there is no one $g$. -->
+1.
+In the household survey component, HIV incidence depends on district-level adult HIV prevalence and ART coverage, such that $\lambda \propto \rho (1 - \omega \cdot \alpha)$, where $\omega = 0.7$ is a fixed constant.
+This reflects basic HIV epidemiology: HIV incidence is proportional to unsuppresed viral load.
+As a result, $\log(\lambda_i)$ depends on 28 structured additive predictors (2 sexes $\times$ 7 age groups $\times$ 2 indicators, HIV prevalence and ART coverage).
+2. In the household survey component, HIV incidence and HIV prevalence are linked to the proportion recently infected via Equation \ref{eq:kappa}.
+3.
+In the ANC testing component, HIV prevalence and ART coverage depend upon the respective indicators in the household survey component.
+Although $\text{logit}(\rho_i)$ and $\text{logit}(\alpha_i)$ are Gaussian, nonetheless this introduces dependence of each mean response on two structured additive predictors.
+4.
+Throughout the model components, processes are modelled at the finest distict-age-sex division, but likelihoods are defined for observations aggregated over sets of indices.
+As such, single observations are related to $|\mathcal{I}|$ structured additive predictors.
+5. Individuals taking ART, or who have been recently infected, must be HIV positive. 
+6. The ART attendance component uses a multinomial model with softmax link function which takes as input $|\{x': x' \sim x\}| + 1$ structured additive predictors.
+7. Multiple link functions are used throughout the model, such that there is no one inverse link function $g$. Instead,
 
@@
 
-To write.
+Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+\begin{align*}
+\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
+\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
+\end{align*}
+Then the Laplace approximation to $p(\btheta, \y)$ is given by
+$$
+\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+$$
 
@@
 where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
-In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials, and weights so as to interpolate functions of the form $f(\z) = \tilde f(\z) \exp(-\z^2)$.
-These nodes and weights can be shifted and scaled to suit the particular integrand using adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate].
-AGHQ with $k = 1$ recovers the Laplace approximation.
+In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials, with weights chosen so as to interpolate functions of the form $f(\z) = \tilde f(\z) \exp(-\z^2)$.
+In adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate] the nodes and weights are shifted and rotated to suit the particular integrand, based on the mode, curvature at the mode, and a matrix decomposition of the inverse curvature.
+Two possibilities for this matrix decomposition are the Cholesky decomposition and the spectral decomposition [@jackel2005note].
 
-```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension are adapted based on the mean and covariance matrix of the target via $\\btheta(\\z) = \\hat \\btheta + \\mathbf{L} \\z$."}
+```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension (a) are adapted based on the mean (b) and covariance matrix of the target via $f(\\z) = \\hat f + \\mathbf{L} \\z$, where..."}
 mu <- c(1, 1.5)
@@
 plot1 <- add_points(plot0, gg) +
-  labs(size = "", caption = "")
+  labs(size = "", subtitle = "(a)")
 
@@
 plot2 <- add_points(plot0, gg2) +
-  labs(size = "", caption = "")
+  labs(size = "", subtitle = "(b)")
 
@@
 plot3 <- add_points(plot0, gg3) +
-  labs(size = "", caption = "")
+  labs(size = "", subtitle = "(c)")
 
-plot1 + plot2 + plot3
+#' Adapt by the spectral
+gg4 <- gg
+mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)
+
+plot4 <- add_points(plot0, gg4) +
+  labs(size = "", subtitle = "(d)")
+
+(plot1 + plot2) / (plot3 + plot4)
 ```
-
-## Algorithm\label{sec:algorithm}
 
@@
 
-We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using four inferential approaches.
+We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
 For each approach, the `TMB` C++ user-template (available in the appendix) used to specify the log-posterior was the same.
 The four approaches were:
-1. `TMB`: EB combined with a Gaussian approximation via `TMB`,
-2. `aghq`: AGHQ combined with a Gaussian approximation via `aghq`,
-3. `adam`: AGHQ combined with a Laplace approximation by extending `aghq`, and
-4. `NUTS`: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
-In this instance, the dimension of the latent field is $N = 491$ and the dimension of the hyperparameters is $m = 24$.
+1. TMB: EB combined with a Gaussian approximation via `TMB`,
+2. PCA-AGQH: the PCA AGHQ grid combined with a Gaussian approximation via `aghq`,
+3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
+The dimension of the latent field is $N = 491$ and the dimension of the hyperparameters is $m = 24$.
 Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
@@
 \small
-\begin{tabularx}{\textwidth}{p{0.3\linewidth}p{0.65\linewidth}}
+\begin{tabularx}{\textwidth}{p{0.15\linewidth}p{0.15\linewidth}p{0.6\linewidth}}
 \toprule
-Inferential method & Details \\
+Name & Software & Details \\
 \midrule
-1. \texttt{TMB}: EB, Gaussian & $1000$ samples \\
-2. \texttt{aghq}: AGHQ, Gaussian & $k = 1$, $1000$ samples \\
-3. \texttt{adam}: AGHQ, Laplace & $k = 1$, $l = 5$, $1000$ samples \\
-4.
-\texttt{NUTS}: NUTS & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$.
-HMC parameters set to default for \texttt{rstan}.
+TMB & \texttt{TMB} & Empirical Bayes, $1000$ samples \\
+PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$, $1000$ samples \\
+NUTS & \texttt{tmbstan} & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$.
+HMC parameters set to default for software.
 \\
@@
 
-Inferential results from MCMC are only accurate once convergence has been reached and the chain lengths are sufficient.
-We assessed the quality of our MCMC results using the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
+We assessed the quality of our NUTS results using the potential scale reduction factor $\hat R$, bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
 Full details are provided in the appendix.

### 2023-05-16 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
ELGM for #40

@@
 ---
+
+<!-- https://github.com/athowes/elgm-inf/blob/fd4b0bb6b9ba93ac277de7aad47a94581e8fc900/src/docs_18-04-23-explainer/18-04-23-explainer.Rmd -->
 
@@
 
+# Latent Gaussian models
+
+* In an LGM, the conditional mean depends on exactly one structured additive predictor
+\begin{align*}
+y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\
+\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
+\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
+\end{align*}
+where $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors (and can be collected into the latent field $x$)
+
+# Extended latent Gaussian models
+
+* ELGM remove this requirement such that
+$$
+\mu_i = g_i(\eta_{\mathcal{J}_i})
+$$
+where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices
+* Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$
+  * $N_n < n$: more data points than structured additive predictors
+  * $N_n = n$: as many data points as structured additive predictors (LGM case)
+  * $N_n > n$: fewer data points than structured additive predictors
+* The $g_i$ allow for a higher degree of non-linearity in the model
+
+# Why is Naomi an ELGM?
+
+| | Things that make Naomi an ELGM |
+|-|---------------|
+|1.| Incidence depends on adult prevalence and coverage |
+|2.| Incidence linked non-linearly to recent infection |
+|3.| ANC offset from household survey |
+|4.| ART coverage and recent infection are products |
+|5.| Aggregation of finer processes |
+|6.| ART attendance uses a multinomial |
+|7.| Multiple link functions |
+
+# 1. Incidence depends on adult prevalence and coverage
+
+* Linear predictor for incidence contains aggregated prevalence and coverage
+$$
+\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda.
+$$
+* Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian
+
+# 2. Incidence and prevalence linked to recent infection
+
+$$
+\kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right)
+$$
+
+* Here $\Omega_T$ and $\beta_T$ are Gaussian and have strong priors depending on the particular survey
+
+# 3. ANC offset from household survey
+
+* Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators
+\begin{align*}
+\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
+\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}. 
+\end{align*}
+* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$
+
+# 3. ANC offset from household survey
+
+* Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$, so I'm unsure which is right]
+$$
+\begin{pmatrix}
+\eta_1^\star \\
+\eta_2^\star
+\end{pmatrix}
+=
+\begin{pmatrix}
+1 & 1 \\
+0 & 1
+\end{pmatrix}
+\begin{pmatrix}
+\eta_1 \\
+\eta_2
+\end{pmatrix}
+=
+\begin{pmatrix}
+\eta_1 + \eta_2 \\
+\eta_2
+\end{pmatrix}
+$$
+* By having effects only apply to a subset of indices (`idx = c(NA, 1)` say) perhaps it can work
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+
+Warning!
+
+\begin{enumerate}
+    \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA}
+    \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM
+\end{enumerate}
+
+\end{tcolorbox}
+\end{center}
+
+# 4. ART coverage and recent infection are products
+
+* In the household survey, say, individuals who are taking ART or have been recently infected must be HIV positive
+\begin{align*}
+y^{\hat \alpha}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \alpha_{x, s, a}), \\
+y^{\hat \kappa}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \kappa_{x, s, a}).
+\end{align*}
+* $\text{logit}(\rho_{x, s, a})$ and $\text{logit}(\alpha_{x, s, a})$ are Gaussian, but we're taking a product here
+* $\kappa_{x, s, a}$ is more complicated: a function of incidence, prevalence, mean duration of recent infection and false recent ratio
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+Warning! These equations as writen do not appear in Naomi: instead there are aggregated versions, more about this soon.
+\end{tcolorbox}
+\end{center}
+
+# 5. Aggregation of finer processes
+
+* There are many instances of models being placed on aggregate quantities
+\begin{align*}
+y^{\hat \theta}_{\mathcal{I}} &\sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}), \\
+\rho_{\mathcal{I}} &= \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}.
+\end{align*}
+* Here we have $|\mathcal{I}|$ linear predictors being informed by one observation
+* Known as disaggregation regression
+
+# 6. ART attendance uses the multinomial
+
+# 7. Multiple link functions
+
+* The Naomi model uses both $\text{logit}$ and $\log$ (inverse) link functions (not even considering the constructed quasi-link functions)
+* For LGMs there is only one $g$, whereas ELGMs allow $g_i$
+* In `R-INLA` it is possible for `y` to be a matrix where each column contains observations with shared likelihood family (and hyperparameters) and `family = c("family1", "family2", ...)`
+
+# Statistical software
+
+|           | Interface `y ~ 1 + x`  | General |
+|-----------|------------|---------|
+| Example   | `brms`     | Stan    |
+| Users     | Scientists | Statisticians |
+| Benefits  | Ease of use, sensible defaults, trust | Flexibility, development |
+| Drawbacks | Being fenced in | Barrier to entry, failing silently |
+
+\begin{center}
+\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+\texttt{athowes/multi-agyw} case-study: attempting to define $\text{AR}(1) \times \text{Besag} \times \text{IID}$ random effects in \texttt{R-INLA}
+\end{tcolorbox}
+\end{center}
+
 # Thanks for listening!

### 2023-05-16 src/naomi-simple_exceedance/exceedance.Rmd
Update exceedance case-study with figures that can go into manuscript #35

@@
   geom_point(alpha = 0.3) +
-  scale_color_manual(values = c("#E69F00", "#F0E442")) +
+  scale_color_manual(values = c("#56B4E9","#009E73")) +
   geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
-  labs(title = "Probability of at least 81% ART coverage in strata", col = "Sex") +
-  theme_minimal() +
-  theme(legend.position = "bottom")
+  labs(subtitle = "Probability of greater than 81% ART coverage in strata", col = "Sex") +
+  guides(col = "none") +
+  labs(x = "NUTS", y = "TMB") +
+  theme_minimal()
 
@@
   geom_point(alpha = 0.3) +
-  scale_color_manual(values = c("#E69F00", "#F0E442")) +
+  scale_colour_manual(values = c("#56B4E9","#009E73"), labels = c("Female", "Male")) +
   geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
-  guides(col = "none") +
+  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
+  labs(x = "NUTS", y = "PCA-AGHQ", col = "Sex") +
   theme_minimal()
+
+pdf("second90.pdf", h = 4, w = 6.25)
+
+plot1 + plot2
+
+dev.off()
 
@@
   geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
-  labs(title = "Probability of at least 1% incidence in strata") +
+  labs(subtitle = "Probability of at least 1% incidence in strata") +
   theme_minimal()
@@
   theme_minimal()
+
+pdf("1inc.pdf", h = 4, w = 6.25)
+
+plot1 + plot2
+
+dev.off()
 

### 2023-05-16 src/docs_paper/paper.Rmd
Start of exceedance case-study into paper #35

@@
 
-We used three methods to assess the accuracy of posterior distributions produced by each inferential method as compared with those from NUTS: (1) Kolmogorov-Smirnov tests, (2) maximum mean discrepancy, and (3) Pareto-smoothed importance sampling.
-Our primary applied interest is in comparing the accuracy of model outputs, rather than the internal hyperparameters or latent field parameters.
-That said, obtaining accurate inferences for these internal parameters is also relevant.
+We used three statistical approaches to assess the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS (1) Kolmogorov-Smirnov tests, (2) Pareto-smoothed importance sampling, and (3) maximum mean discrepancy.
 
@@
 
-Let $\{\theta_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function $F(\vartheta) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\theta_i \leq \vartheta}$.
+Let $\{\phi_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
 The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is given by the maximum absolute difference between two ECDFs.
@@
 \begin{equation*}
-D_\bullet = \sup_\vartheta | F_\texttt{NUTS}(\vartheta) - F_\bullet(\vartheta)|.
+D_\bullet = \sup_\vartheta | F_\text{NUTS}(\vartheta) - F_\bullet(\vartheta)|.
 \end{equation*}
@@
 
+### Pareto-smoothed importance sampling
+
+We are interested in assessing the accuracy of joint distributions as well as marginal distributions.
+Let $\{\bphi_i\}_{i = 1}^n$ be posterior joint samples.
+Pareto smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes].
+See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.
+
+```{r psis, fig.cap="Results of Pareto-smoothed importance sampling analysis."}
+qplot(1:10, 1:10) + theme_minimal()
+```
+
 ### Maximum mean discrepancy
@@
 
-### Pareto-smoothed importance sampling
+## Case study on exceedance probabilites
 
-To write.
-See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.
+### Meeting the second 90
 
-```{r psis, fig.cap="Results of Pareto-smoothed importance sampling analysis."}
-qplot(1:10, 1:10) + theme_minimal()
+The Joint United Nations Programme on HIV/AIDS (UNAIDS) has developed ambitious fast-track targets for scaling up ART treatment with the goal to "end the AIDS epidemic by 2030".
+The 90-90-90 by 2020 fast-track target is that 90% of PLHIV know their status, 90% of those are on antiretroviral therapy (ART) and 90% of those have suppressed viral load.
+Naomi can be used to assess the probability that the second 90 target has been met, thereby helping to find treatment gaps, by evaluating $\mathbb{P}(\alpha_i > 0.81)$ for each strata $i$.
+See a summary of the results in Figure \ref{fig:second90}.
+
+```{r second90, fig.cap="Results of second 90 case-study."}
+knitr::include_graphics("depends/second90.png")
+```
+
+### Finding strata with high incidence strata
+
+Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as above 1% incidence per year.
+Naomi can be used to assess the probability of there being high incidence within a strata by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
+See a summary of the results in Figure \ref{fig:1inc}.
+
+```{r 1inc, fig.cap="Results of high incidence case-study."}
+knitr::include_graphics("depends/1inc.png")
 ```
@@
 We developed an approximate Bayesian inference algorithm motivated by a challenging problem in small-area estimation of HIV in low resource settings.
-For the simplified Naomi model in Malawi (Section \ref{sec:results}) our method is demonstrated to be more accurate than the EB Gaussian approximation currently in use, and substantially faster than NUTS.
-We anticipate that our method could be added to the Naomi web interface as an alternative to `TMB`.
+For the simplified Naomi model in Malawi (Section \ref{sec:results}) our method is demonstrated to be more accurate than TMB, and substantially faster than NUTS.
+We anticipate that our method could be added to the Naomi web interface as an alternative to TMB.
 Analysts might quickly iterate over model options using the faster, less accurate inference approach, switching to the slower, more accurate approach once they are happy with the results.
@@
 In doing so, we hope our work enables use of deterministic inference algorithms for ELGMs in applied settings, as well as further methodological exploration of their accuracy and limitations.
-Among the ELGMs structures of particular interest in spatial epidemiology, many of which are included in the structure of Naomi, are: aggregated Gaussian process models [@nandi2020disaggregation], evidence synthesis models [@amoah2020geostatistical].
+Among the ELGM-type structures of particular interest in spatial epidemiology, many of which feature in Naomi, are: aggregated Gaussian process models [@nandi2020disaggregation], evidence synthesis models [@amoah2020geostatistical].
 Although our method is designed for ELGMs, it is possible to use it outside this class, as it is compatible with any model with a `TMB` \textsc{C++} template.
@@
 
-Future work could look to implement our algorithm (Section \ref{sec:algorithm}) within probabilistic programming languages, facilitating access by a broader user-base.
+Future work could look to implement our algorithm within probabilistic programming languages, facilitating access by a broader user-base.
 This might be possible in Stan by use of the `bridgestan` package [@bridgestan] together with the adjoint-differentiated Laplace approximation of @margossian2020hamiltonian.

### 2023-05-18 src/docs_19-05-23-turing/19-05-23-turing.Rmd
Slides up for #37

@@
 author: "Adam Howes"
-date: "April 2023"
+date: "May 2023"
 output:
@@
       in_header: preamble.tex
-subtitle: PhD Student Presentations & Networking Event
+subtitle: PhD Student Presentations & Networking Event, Alan Turing Institute
 bibliography: citations.bib
@@
 
-```{r, echo=FALSE, fig.cap="Estimates of HIV indicators at a district level produced using the Naomi model.", out.width = '90%'}
+```{r, echo=FALSE, fig.cap="Ambitious targets required to end the AIDS epidemic as a public health threat by 2030.", out.width = '70%'}
+knitr::include_graphics("fast-track.png")
+```
+
+#
+
+```{r, echo=FALSE, fig.cap="More effective intervention based on granular estimates of HIV indicators. One size does not fit all!", out.width = '90%'}
+cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
 knitr::include_graphics("depends/naomi_results.png")
@@
 
-```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'}
+```{r, echo=FALSE, fig.cap="Generation of estimates by country teams strengthens data quality, use and ownership. User interface from \\texttt{https://naomi.unaids.org/}.", out.width = '70%'}
 knitr::include_graphics("naomi_user.png")
@@
 
-# Goal
+#
 
-* Fast approximate Bayesian inference for a complex, spatiotemporal, evidence synthesis model
+:::::::::::::: {.columns}
 
-# Approach
+::: {.column width=.5}
 
-1. The marginal Laplace approximation
+\textcolor{hilit}{Want}
+
+Fast approximate Bayesian inference for a complex, spatiotemporal, evidence synthesis model
+
+:::
+
+::: {.column width=.5}
+
+\textcolor{hilit}{Strategy}
+
+1. Marginal Laplace approximation
 2. Adaptive Gauss-Hermite quadrature
-3. Principle components analysis
+3. Principal components analysis
 
-# The marginal Laplace approximation
+:::
+
+::::::::::::::
+
+#
+
+```{r message=FALSE, echo=FALSE, fig.cap="A Gamma prior with $a = 3$ and $b = 1$."}
+cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
+a <- 3
+b <- 1
+
+prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) +
+  stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) +
+  annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1) prior", col = cbpalette[1], size = 5) +
+  labs(x = "", y = "") +
+  theme_minimal()
+
+prior
+```
+
+#
+
+```{r message=FALSE, echo=FALSE, fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."}
+set.seed(2)
+y <- rpois(3, lambda = 2)
+
+posterior <- prior +
+  geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) +
+  stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) +
+  annotate("text", x = 6, y = 0.25, label = "Gamma(9, 4) posterior (known)", col = cbpalette[2], size = 5)
+
+posterior
+```
+
+#
+
+```{r message=FALSE, echo=FALSE, fig.cap="The Laplace approximation in this case is good near the mode but not in the tails."}
+fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE)
+
+ff <- list(
+  fn = fn,
+  gr = function(x) numDeriv::grad(fn, x),
+  he = function(x) numDeriv::hessian(fn, x)
+)
+
+opt_bfgs <- aghq::optimize_theta(
+  ff, 1, control = aghq::default_control(method = "BFGS")
+)
+
+laplace <- posterior +
+  stat_function(
+    data = data.frame(x = c(0, 10)),
+    aes(x),
+    fun = dnorm,
+    n = 500,
+    args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)),
+    col = cbpalette[3]
+  ) +
+  annotate("text", x = 5, y = 0.4, label = "Laplace approximation", col = cbpalette[3], size = 5)
+
+laplace
+```
+
+#
+
+* For Malawi, the model has 491 parameters
+* 467 have a joint Gaussian prior: call them the latent field $x$
+* 24 are not Gaussian: call them hyperparameters $\theta$
+* Use the \textcolor{hilit}{Laplace} approximation only for the latent field \textcolor{hilit}{marginal} posterior!
+
+#
+
+```{r echo=FALSE, warning =FALSE, out.height = '70%', fig.cap = "Spatial random effects $\\phi_i \\, | \\, \\phi_{-i} \\sim \\mathcal{N} \\left(\\frac{1}{n_{\\delta i}} \\sum_{j: j \\sim i} \\phi_j, \\frac{1}{n_{\\delta i}\\tau_\\phi} \\right)$ are included in the latent field.
+We assume that neighbouring districts are similar: first law of geography."}
+area_merged <- sf::read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
+
+sf <- area_merged %>%
+  filter(area_level == 3)
+
+nb <- sf_to_nb(sf)
+
+nb_sf <- spdep::nb2lines(nb, coords = sp::coordinates(as(sf, "Spatial"))) %>%
+  as("sf") %>%
+  sf::st_set_crs(sf::st_crs(sf))
+
+b <- ggplot2::ggplot(sf) +
+  ggplot2::geom_sf(data = nb_sf) +
+  ggplot2::theme_minimal() +
+  ggplot2::labs(subtitle = "Graph") + 
+  ggplot2::theme_void()
+
+a <- ggplot2::ggplot(sf) +
+  ggplot2::geom_sf() +
+  ggplot2::theme_minimal() +
+  ggplot2::labs(subtitle = "Geography") + 
+  ggplot2::theme_void()
+
+a + b
+```
 
@@
 
-```{r, echo=FALSE, fig.cap="With 8 dimenions, the covariance matrix is accurately reproduced.", out.width = '80%'}
+```{r, echo=FALSE, fig.cap="With 8 dimenions, the hyperparameter covariance matrix is accurately reproduced.", out.width = '80%'}
 knitr::include_graphics("depends/reduced-rank-plot.png")
@@
 
+#
+
+```{r, echo=FALSE, fig.cap="Reduced RMSE by 10%, but still a work in progress!
+Also using Kolmogorov-Smirnov tests, Pareto-smoothed importance sampling, maximum mean discrepancy...", out.width = '80%'}
+knitr::include_graphics("depends/second90.pdf")
+```
+
 # Thanks for listening!
+
+* For more about Naomi, see @eaton2021naomi
+* Joint work with Alex Stringer (Waterloo), Seth Flaxman (Oxford), and Jeff Eaton (Harvard, Imperial)
+* For more about me, see `athowes.github.io/about`
 

### 2023-05-18 src/naomi-simple_model-checks/model-checks.Rmd
Add placeholder histograms and ECDF difference plots (still need to compute "realised" data) #43

--- 2023-01-15 src/naomi-simple_model-checks/model-checks.Rmd
@@
 ```{r}
-#' Yes!
+tmb <- readRDS("depends/tmb.rds")
+
+indicators <- naomi::add_output_labels(tmb$outputs) %>%
+  left_join(tmb$outputs$meta_area %>% select(area_level, area_id, center_x, center_y)) %>%
+  sf::st_as_sf()
+
+indicators %>%
+  filter(
+    sex %in% c("male", "female"),
+    age_group == "Y015_049",
+    indicator == "prevalence",
+    area_level == 4
+  ) %>%
+  ggplot(aes(fill = mode)) +
+  geom_sf() +
+  viridis::scale_fill_viridis(labels = scales::percent_format()) +
+  naomi::th_map() +
+  facet_wrap(~sex)
+
+indicators_fine <- indicators %>%
+  filter(
+    indicator %in% c("incidence", "prevalence", "art_coverage"),
+    area_id %in% paste0("MWI_4_", 1:32, "_demo"),
+    sex %in% c("male", "female"),
+    age_group %in%
+      c(
+        "Y000_004", "Y005_009", "Y010_014", "Y015_019", "Y020_024", "Y025_029",
+        "Y025_034", "Y030_034", "Y035_039", "Y040_044", "Y045_049", "Y050_054",
+        "Y055_059", "Y060_064", "Y065_069", "Y070_074", "Y075_079", "Y080_999"
+      )
+  )
 ```
+
+```{r}
+mf_out <- tmb$naomi_data$mf_out
+
+mf_out_fine <- mf_out %>%
+  tibble::rownames_to_column("id") %>%
+  mutate(id = as.numeric(id)) %>%
+  filter(
+    area_id %in% paste0("MWI_4_", 1:32, "_demo"),
+    sex %in% c("male", "female"),
+    age_group %in%
+      c(
+        "Y000_004", "Y005_009", "Y010_014", "Y015_019", "Y020_024", "Y025_029",
+        "Y025_034", "Y030_034", "Y035_039", "Y040_044", "Y045_049", "Y050_054",
+        "Y055_059", "Y060_064", "Y065_069", "Y070_074", "Y075_079", "Y080_999"
+      )
+  )
+
+nrow(indicators_fine) / nrow(mf_out_fine)
+```
+
+```{r}
+get_prob_quantile <- function(indicator) {
+  samples <- tmb$fit$sample[[indicator]][mf_out_fine$id, ]
+  S <- ncol(samples)
+  prob_quantile <- apply(samples, MARGIN = 1, function(x) {
+    estimate <- x[1] # This is wrong! I am just using the first sample as a placeholder
+    sample_row <- x[1:S]
+    if(all(is.na(sample_row))) return(NA)
+    else ecdf(sample_row)(estimate)
+  })
+  data.frame(indicator = indicator, prob_quantile = prob_quantile)
+}
+
+df <- lapply(list("rho_t1_out", "lambda_t1_out", "alpha_t1_out"), get_prob_quantile) %>%
+  bind_rows()
+
+bins <- 20
+alpha <- 0.05
+
+ci <- qbinom(
+  p = c(alpha / 2, 0.5, (1 - alpha / 2)),
+  size = S,
+  prob = 1 / bins
+)
+
+polygon_data <- data.frame(
+  x = c(-0.05, 0, 1, 0, -0.05, 1.05, 1, 1.05, -0.05),
+  y = c(ci[1], ci[2], ci[2], ci[2], ci[3], ci[3], ci[2], ci[1], ci[1]) / S
+)
+
+fct_reorg <- function(fac, ...) {
+  fct_recode(fct_relevel(fac, ...), ...)
+}
+
+histograms <- df %>%
+  mutate(indicator = fct_reorg(indicator, "HIV Prevalence" = "rho_t1_out", "HIV Incidence" = "lambda_t1_out", "ART Coverage" = "alpha_t1_out")) %>%
+  ggplot(aes(x = prob_quantile)) +
+  facet_grid(~indicator, drop = TRUE, scales = "free", labeller = labeller(indicator = label_wrap_gen(20))) +
+  geom_histogram(aes(y = (..count..) / tapply(..count..,..PANEL..,sum)[..PANEL..]),
+                 breaks = seq(0, 1, length.out = bins + 1), fill = "#009E73", col = "black", alpha = 0.9) +
+  geom_polygon(data = polygon_data, aes(x = x, y = y), fill = "grey75", color = "grey50", alpha = 0.6) +
+  labs(x = "", y = "") +
+  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(0, 0.25, 0.5, 0.75, 1)) +
+  theme_minimal()
+
+lims <- multi.utils::get_lims(n = S, alpha, K = 100)
+
+ecdf_diff <- df %>%
+  split(.$indicator) %>%
+  lapply(function(y) {
+    empirical_coverage <- purrr::map_dbl(seq(0, 1, by = 0.01), ~ empirical_coverage(y$prob_quantile, .x))
+    data.frame(nominal_coverage = seq(0, 1, by = 0.01), empirical_coverage = empirical_coverage) %>%
+      mutate(
+        ecdf_diff = empirical_coverage - nominal_coverage,
+        ecdf_diff_lower = lims$lower / S - nominal_coverage,
+        ecdf_diff_upper = lims$upper / S - nominal_coverage,
+      )
+  }) %>%
+  purrr::map_df(~as.data.frame(.x), .id = "indicator") %>%
+  mutate(indicator = fct_reorg(indicator, "HIV Prevalence" = "rho_t1_out", "HIV Incidence" = "lambda_t1_out", "ART Coverage" = "alpha_t1_out")) %>%
+  ggplot(aes(x = nominal_coverage, y = ecdf_diff)) +
+  facet_grid(~indicator, drop = TRUE, scales = "free", labeller = labeller(indicator = label_wrap_gen(20))) +
+  geom_line(col = "#009E73") +
+  geom_step(aes(x = nominal_coverage, y = ecdf_diff_upper), alpha = 0.7, col = "grey50") +
+  geom_step(aes(x = nominal_coverage, y = ecdf_diff_lower), alpha = 0.7, col = "grey50") +
+  geom_abline(intercept = 0, slope = 0, linetype = "dashed", col = "grey75") +
+  labs(x = "", y = "ECDF difference") +
+  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(0, 0.25, 0.5, 0.75, 1)) +
+  theme_minimal()
+
+pdf("coverage.pdf", h = 4, w = 6.25)
+
+histograms / ecdf_diff
+
+dev.off()
+
+histograms / ecdf_diff
+```
+
+Next steps here are:
+
+* Add column to `indicators` with the "raw" value of the indicator and replace the placeholder `estimate <- sample[1]`
+* Should probably put the value of `prob_quantile` into the `indicators` dataframe as well

### 2023-05-19 src/docs_19-05-23-turing/19-05-23-turing.Rmd
Correct LaTeX error #37

@@
 
-```{r, echo=FALSE, fig.cap="Reduced RMSE by 10%, but still a work in progress!
-Also using Kolmogorov-Smirnov tests, Pareto-smoothed importance sampling, maximum mean discrepancy...", out.width = '80%'}
+```{r, echo=FALSE, fig.cap="Reduced RMSE by 10 percent, but still a work in progress!
+Also seeing small improvements using Kolmogorov-Smirnov tests, Pareto-smoothed importance sampling, maximum mean discrepancy.", out.width = '80%'}
 knitr::include_graphics("depends/second90.pdf")

### 2023-05-22 src/docs_paper/paper.Rmd
491 is the total number of parameters
Update paper writing

@@
 
-## Naomi as an ELGM
+## Naomi framed as an ELGM
 
@@
 
+## TMB
+
+## PCA-AGHQ
+
 # Application to data from Malawi\label{sec:results}
@@
 1. TMB: EB combined with a Gaussian approximation via `TMB`,
-2. PCA-AGQH: the PCA AGHQ grid combined with a Gaussian approximation via `aghq`,
+2. PCA-AGQH: the PCA-AGHQ grid combined with a Gaussian approximation via `aghq`,
 3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
-The dimension of the latent field is $N = 491$ and the dimension of the hyperparameters is $m = 24$.
+The dimension of the latent field is $N = 467$ and the dimension of the hyperparameters is $m = 24$.
 Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
@@
 <!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
-Example model outputs from `TMB` are illustrated in Figure \ref{fig:naomi-results}.
+Example model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
 The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
@@
 
-```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted with \\texttt{TMB}."}
+```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted with TMB."}
 knitr::include_graphics("depends/naomi_results.png")
@@
 
-We assessed the quality of our NUTS results using the potential scale reduction factor $\hat R$, bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
+We assessed the quality of our NUTS results to be satisfactory using the potential scale reduction factor $\hat R$ [@gelman1992inference; @vehtari2021rank], bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments [@betancourt2017conceptual].
 Full details are provided in the appendix.
-We  treat these results from NUTS as a gold-standard to which other inferential methods are compared.
+We thus treat the results from NUTS as a gold-standard to which other inferential methods are compared.
 
@@
 
-## Inference comparison
+## Inference comparison \label{sec:inf-comparison}
 
@@
 
-We are interested in assessing the accuracy of joint distributions as well as marginal distributions.
-Let $\{\bphi_i\}_{i = 1}^n$ be posterior joint samples.
-Pareto smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes].
+As marginal distributions, we are interested in assessing the accuracy of joint distributions.
+Let $\{\bphi_i\}_{i = 1}^n$ be samples from a joint posterior.
+Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios $r_s = p(\theta_s, y) / q(\theta)$ used in importance sampling.
+Let
+$$
+p(y \, | \, \mu, \sigma, k) =
+\begin{cases}
+\frac{1}{\sigma} \left( 1 + k \left( \frac{y - \mu}{\sigma} \right) \right)^{- \frac{1}{k} - 1}, \quad k \neq 0 \\
+\frac{1}{\sigma} \exp \left( \frac{y - \mu}{\sigma} \right), \quad k = 0
+\end{cases}
+$$
+be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
 See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.
@@
 
-To write.
+Another way to compare joint distributions is via the maximum mean discrepancy [MMD; @gretton2006kernel].
+Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of samples from a joint posterior obtained using different inference methods.
+The MMD can be empirically estimated by
+$$
+\text{MMD}(\Phi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bphi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
+$$
 See a summary of the results in Table and Figure \ref{fig:mmd}, and full results available in the appendix.
@@
 
-## Case study on exceedance probabilites
+## Case study on exceedance probabilites \label{sec:exceedance}
 
@@
 
-The Joint United Nations Programme on HIV/AIDS (UNAIDS) has developed ambitious fast-track targets for scaling up ART treatment with the goal to "end the AIDS epidemic by 2030".
-The 90-90-90 by 2020 fast-track target is that 90% of PLHIV know their status, 90% of those are on antiretroviral therapy (ART) and 90% of those have suppressed viral load.
-Naomi can be used to assess the probability that the second 90 target has been met, thereby helping to find treatment gaps, by evaluating $\mathbb{P}(\alpha_i > 0.81)$ for each strata $i$.
-See a summary of the results in Figure \ref{fig:second90}.
+The Joint United Nations Programme on HIV/AIDS has developed ambitious fast-track targets for scaling up ART treatment with the goal of "ending the AIDS epidemic by 2030".
+Specifically, the "90-90-90 by 2020" fast-track target is that:
+
+* 90% of PLHIV know their status,
+* 90% of those are on antiretroviral therapy (ART), and
+* 90% of those have suppressed viral load.
+
+Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met $\mathbb{P}(\alpha_i > 0.81)$ for each strata $i$.
+We found that both TMB and PCA-AGHQ underestimate the probability that the second 90 target has been met in women (Figure \ref{fig:second90}).
+This difficulty may be related to interactions between the household survey and ANC components of the model.
 
@@
 
-### Finding strata with high incidence strata
+### Finding strata with high incidence
 
-Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as above 1% incidence per year.
-Naomi can be used to assess the probability of there being high incidence within a strata by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
-See a summary of the results in Figure \ref{fig:1inc}.
+Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as greater than 1% incidence per year.
+Naomi can be used to assess the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
+We found that both TMB and PCA-AGHQ overestimate these exceedance probabilities (Figure \ref{fig:1inc}).
+We do not yet have a working hypothesis as to why this is.
 
@@
 
-We developed an approximate Bayesian inference algorithm motivated by a challenging problem in small-area estimation of HIV in low resource settings.
+We developed an approximate Bayesian inference algorithm motivated by a challenging problem in small-area estimation of HIV.
 For the simplified Naomi model in Malawi (Section \ref{sec:results}) our method is demonstrated to be more accurate than TMB, and substantially faster than NUTS.
 We anticipate that our method could be added to the Naomi web interface as an alternative to TMB.
-Analysts might quickly iterate over model options using the faster, less accurate inference approach, switching to the slower, more accurate approach once they are happy with the results.
+This would enable analysts to quickly iterate over model options using a faster, less accurate inference approach, before switching to a slower, more accurate approach once they are happy with the results.
 
@@
 
-In our case study we demonstrated a Bayesian workflow for deterministic inference methods.
-We retained the ability to draw samples from the posterior distributions of interest, facilitating use of posterior predictive checks (Section \ref{sec:model-assessment}).
+We demonstrated a Bayesian workflow for deterministic inference methods.
+We retained the ability to draw samples from the posterior distributions of interest, facilitating use of posterior predictive checks (Section \ref{sec:model-assessment}) and methods for inference comparison such as Pareto-smoothed importance sampling (Section \ref{sec:inf-comparison}).
+The exceedance probabilities case-study (Section \ref{sec:exceedance}) demonstrates the importance of accurate posterior inferences for realistic use-cases of the Naomi model.
 

### 2023-05-23 src/docs_bioinference-poster/bioinference-poster.Rmd
Update template for BioInference poster #36

@@
   - name: "Jeffrey W. Eaton"
-    affil: 2
+    affil: 5, 2
 affiliation:
@@
     address: Department of Computer Science, University of Oxford
-logoleft_name: "mrc-gida-icl-uw-uo.png"
+  - num: 5
+    address: Center for Communicable Disease Dynamics, Department of Epidemiology, Harvard T.H. Chan School of Public Health
+logoleft_name: "bioinference-joined.png"
 # Style & formatting
@@
 
-* Approximate Bayesian inference method using Laplace approximations and adaptive Gauss-Hermite quadrature 
+* Approximate Bayesian inference method using Laplace approximation, adaptive Gauss-Hermite quadrature and principal component analysis
 * Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
-* Looking to implement as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a `TMB` C++ user template
+* Implemented as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a `TMB` C++ user template
 
@@
   * Figure \@ref(fig:fig1) illustrates the seven stages of using the app
-* Inference conducted in minutes using empirical Bayes (EB) and a Gaussian approximation via Template Model Builder `TMB` [@kristensen2016tmb]
+* Inference conducted in minutes using empirical Bayes and a Gaussian approximation via Template Model Builder `TMB` [@kristensen2016tmb]
 * It would take days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no], and this is not practical in this setting
@@
 * Naomi is an ELGM, not an LGM, because it includes complex dependency structures:
-  1. ANC indicators offset from household survey
-  2. Incidence depends on prevalence and ART coverage
-  3. Observed data are aggregated finer processes
-  4. Allow attendance of ART clinics outside home district
-  5. ART attendance probability as product of prevalence and coverage 
+  1. Incidence depends on prevalence and ART coverage
+  2. Incidence ane prevalence linked to recent infection
+  3. ANC offset from household survey
+  4. ART coverage and recent infection are products
+  5. Observed data are aggregated finer processes
+  6. ART attendance uses the multinomial
+  7. Multiple link functions
 * We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
@@
 $$
-where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(\dim(\theta), k)$ with $k$ points per dimension is adapted based upon
-  * The mode $\hat \theta = \texttt{argmax}_{\theta \in \Theta} p(\theta)$
-  * The lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$
-* Algorithm (called `adam` for now) summarized by Figure \@ref(fig:fig2)
-  * Where possible, previously calculated quantities and quadrature rules are reused
-
-```{tikz, fig2, echo=FALSE, fig.align='center', fig.cap = "Flowchart describing the algorithm", fig.ext = 'png', cache=TRUE}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2.5cm, auto,>=latex', thick, scale = 1]
-\node (1) {$p(\theta, x, y)$};
-\node (2) [below of=1] {$\tilde p_\texttt{LA}(x_i, \theta, y)$};
-\node (3) [below of=2] {$\tilde p_\texttt{LA}(x_i, y)$};
-\node (4) [below of=3] {$\tilde p_\texttt{AQ}(x_i \, | \, y)$};
-\node (5) [below of=4] {Output};
-
-\node (6) [right of=1, xshift=1cm] {$\tilde p_\text{LA}(\theta, y)$};
-\node (7) [below of=6] {$\tilde p_\texttt{AQ}(y)$};
-\node (8) [below of=7] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$};
-\node (9) [below of=8] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$};
-\node (10) [below of=9] {Output};
-
-\draw[->] (1) to node {Laplace} (2);
-\draw[->] (2) to node {AGHQ} (3);
-\draw[->] (3) to node {Normalise} (4);
-\draw[->] (4) to node {Spline} (5);
-\draw[->] (1) to node {Laplace} (6);
-\draw[->] (6) to node {AGHQ} (7);
-\draw[->] (7) to node {Normalise} (8);
-\draw[->] (8) to node {Marginalise} (9);
-\draw[->] (9) to node {Spline} (10);
-\end{tikzpicture}
-```
+where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ with $m = \dim(\theta)$ and $k$ points per dimension is adapted based upon
+  * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$
+  * A matrix decomposition $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$
+* Keep first $s < m$ **principal components** to get size $k^s$ grid
 
@@
 * Relatively small country but still a large model: latent field $\text{dim}(x) = 491$, hyperparameters $\text{dim}(\theta) = 24$
-* Fit four inference methods (using one [!] C++ template):
-  * `TMB` (`r round(time_taken$TMB, 0)` mins)
-  * `aghq` (`r round(time_taken$aghq, 0)` mins): $k = 1$
-  * `adam` (`r round(time_taken$adam, 0)` min): $k = 1$
-  * `tmbstan` (`r round(time_taken$tmbstan, 1)` days): 4 chains of 100,000 thinned by 40 (required for good diagnostics)
+* Fit three inference methods (using one C++ template):
+  * TMB (`r round(time_taken$TMB, 0)` mins)
+  * PCA-AGHQ (`r round(time_taken$aghq, 0)` mins): $k = 3, s = 8$
+  * NUTS (`r round(time_taken$tmbstan, 1)` days): 4 chains of 100,000 thinned by 40 (required for good diagnostics)
 * Figure \@ref(fig:fig3) illustrates example model outputs: HIV prevalence, ART coverage, HIV incidence, and number of new infections, at the district level
-
-```{r fig4, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Inference reults and ECDF comparison for one $x_i$.'}
-knitr::include_graphics("fig4.png")
-```
-
-* Compare hyperparameter, latent field, and output posterior marginals based on maximum ECDF difference (Kolmogorov-Smirnov test) 
-* Figure \@ref(fig:fig4) illustrates this approach for one node in the model with $\text{KS}(\texttt{TMB}) = 0.09$ and $\text{KS}(\texttt{adam}) = 0.02$
+* Compare hyperparameter, latent field, and output posteriors based on Kolmogorov-Smirnov tests, Pareto-smoothed importance sampling, and maximum mean discrepancy
 
@@
 
-* Scaling up the hyperparameter grid beyond EB $k = 1$
-  * Any dense grid would be impractical ($k^{24}$ nodes)
-  * Alternatives: sparse grids, dense grids on a subspace
-* Add Laplace matrix algebra approximations [@wood2020simplified] to speed up latent field marginal calculations
-* More comprehensive inference comparison
-  * Maximum mean discrepancy
-  * Pareto-smoothed importance sampling
+* Laplace marginals with matrix algebra approximations [@wood2020simplified] to speed up latent field marginal calculations
+* Further methods for allocation of effort to important dimensions
 

### 2023-05-23 src/docs_bioinference-poster/figs.R

@@ -135,78 +135,3 @@ ggsave(
   height = 6,
   units = "in"
 )
-
-#' Figure 4
-fig4data <- readRDS("depends/beta_anc_rho.rds") %>%
-  filter(method != "aghq")
-
-colours <- c("#56B4E9", "#009E73", "#E69F00")
-
-mean <- fig4data %>%
-  filter(method == "tmbstan") %>%
-  summarise(mean = mean(samples)) %>%
-  pull(mean) %>%
-  round(digits = 3)
-
-sd <- fig4data %>%
-  filter(method == "tmbstan") %>%
-  summarise(sd = sd(samples)) %>%
-  pull(sd) %>%
-  round(digits = 3)
-
-fig4a <- ggplot(fig4data, aes(x = samples, fill = method, col = method)) +
-  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +
-  theme_minimal() +
-  facet_grid(method~.) +
-  labs(x = "beta_anc_rho", y = "Density", fill = "Method") +
-  scale_color_manual(values = colours) +
-  scale_fill_manual(values = colours) +
-  theme(legend.position = "none")
-
-grid <- seq(from = min(fig4data$samples), to = max(fig4data$samples), length.out = 1000)
-
-tmb_ecdf <- stats::ecdf(filter(fig4data, method == "TMB") %>% pull(samples))
-tmb_ecdf_df <- data.frame(x = grid, ecdf = tmb_ecdf(grid), method = "TMB")
-
-adam_ecdf <- stats::ecdf(filter(fig4data, method == "adam") %>% pull(samples))
-adam_ecdf_df <- data.frame(x = grid, ecdf = adam_ecdf(grid), method = "adam")
-
-tmbstan_ecdf <- stats::ecdf(filter(fig4data, method == "tmbstan") %>% pull(samples))
-tmbstan_ecdf_df <- data.frame(x = grid, ecdf = tmbstan_ecdf(grid), method = "tmbstan")
-
-# Add ECDF differences
-tmb_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - tmb_ecdf_df$ecdf
-adam_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - adam_ecdf_df$ecdf
-tmbstan_ecdf_df$ecdf_diff <- 0
-
[…]

### 2023-05-23 src/docs_paper/paper.Rmd
Rename posterior shrinkage to posterior contraction and start move into manuscript #45

@@
 For $\theta \in \{\rho, \alpha, \kappa\}$ let
-$$
+\begin{equation*}
 \hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}
-$$
+\end{equation*}
 be weighted, aggregate survey observations, where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
 The observed number of outcomes are $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
-$$
+\begin{equation*}
 m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
-$$
+\end{equation*}
 is the Kish effective sample size [@kish1965survey].
 We use a binomial working likelihood
-$$
+\begin{equation*}
 y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})
-$$
+\end{equation*}
 to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following weighted aggregates
@@
 Then the Laplace approximation to $p(\btheta, \y)$ is given by
-$$
+\begin{equation*}
 \tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
-$$
+\end{equation*}
 
@@
 A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by the following weighted sum
-$$
+\begin{equation*}
 \int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z),
-$$
+\end{equation*}
 where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
@@
 
-We performed posterior predictive checks to assess the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.
+### Posterior contraction
+
+To assess the informativeness of the data on each model parameter $\phi$ we compared (Figure \ref{fig:contraction}) the prior variance $\sigma_\text{prior}^2(\phi)$ to the posterior variance $\sigma_\text{posterior}^2(\phi)$ via the posterior contraction [@schad2021toward].
+\begin{equation*}
+c(\phi) = 1 - \frac{\sigma_\text{posterior}^2(\phi)}{\sigma_\text{prior}^2(\phi)}.
+\end{equation*}
+
+```{r contraction, fig.cap="Posterior contraction."}
+knitr::include_graphics("depends/posterior-contraction.png")
+```
+
+### Coverage
+
+We assessed the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.
+Let $\{\phi_i\}_{i = 1}^n$ be posterior marginal samples.
 
@@
 
-Let $\{\phi_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
-The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is given by the maximum absolute difference between two ECDFs.
+The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is given by the maximum absolute difference between two empirical cumulative distribution (ECDF) functions $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
 For each method we compare the KS statistics
 \begin{equation*}
-D_\bullet = \sup_\vartheta | F_\text{NUTS}(\vartheta) - F_\bullet(\vartheta)|.
+D_\bullet = \sup_\varphi | F_\text{NUTS}(\varphi) - F_\bullet(\varphi)|.
 \end{equation*}
@@
 Let
-$$
+\begin{equation*}
 p(y \, | \, \mu, \sigma, k) =
@@
 \end{cases}
-$$
+\end{equation*}
 be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
@@
 The MMD can be empirically estimated by
-$$
+\begin{equation*}
 \text{MMD}(\Phi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bphi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
-$$
+\end{equation*}
 See a summary of the results in Table and Figure \ref{fig:mmd}, and full results available in the appendix.

### 2023-05-23 src/naomi-simple_psis/psis.Rmd
Rename posterior shrinkage to posterior contraction and start move into manuscript #45

@@
 
-aghq_q_ll <- apply(mvn_ll, 1, FUN = function(row) matrixStats::logSumExp(log(aghq_weights) + row))
+aghq_q_ll <- apply(mvn_ll, 1, FUN = function(row) matrixStats::logSumExp(log(aghq_lambda) + row))
 plot(aghq_q_ll)
@@
 
-aghq_psis <- loo::psis(log_ratios = lw, r_eff = 1)
+aghq_psis <- loo::psis(log_ratios = r, r_eff = 1)
 aghq_psis$diagnostics

### 2023-05-23 make/_make.R

@@ -87,6 +87,7 @@ run_commit_push("naomi-simple_exceedance")
 run_commit_push("naomi-simple_psis")
 run_commit_push("naomi-simple_mmd")
 run_commit_push("naomi-simple_model-checks")
+run_commit_push("naomi-simple_contraction")
 
 #' Checks
 run_commit_push("check_hyper-marginals")

### 2023-05-23 src/naomi-simple_contraction/script.R

@@ -1,13 +1,13 @@
 #' Uncomment and run the two line below to resume development of this script
-# orderly::orderly_develop_start("naomi-simple_shrinkage")
-# setwd("src/naomi-simple_shrinkage")
+# orderly::orderly_develop_start("naomi-simple_contraction")
+# setwd("src/naomi-simple_contraction")
 
 tmb <- readRDS("depends/tmb.rds")
 aghq <- readRDS("depends/aghq.rds")
 tmbstan <- readRDS("depends/tmbstan.rds")
 mcmc <- tmbstan$mcmc$stanfit
 
-#' Posterior shrinkage is
+#' Posterior contraction is
 #' 1 - posterior_variance / prior_variance
 
 #' It's easy to obtain the posterior variance
@@ -32,42 +32,42 @@ prior_mcmc <- tmbstan::tmbstan(obj_prior, chains = 4, iter = 1000)
 prior_stan_summary <- rstan::summary(prior_mcmc)$summary
 prior_sd <- prior_stan_summary[, "sd"][1:491] #' Remove the lp__ at the end
 
-posterior_shrinkage <- 1 - posterior_sd^2 / prior_sd^2
+posterior_contraction <- 1 - posterior_sd^2 / prior_sd^2
 
 df_hyper <- data.frame(
-  posterior_shrinkage = posterior_shrinkage
+  posterior_contraction = posterior_contraction
 ) %>%
   tibble::rownames_to_column("par") %>%
   filter(par %in% names(tmb$fit$obj$par)) %>%
   mutate(type = "Hyper")
 
 df_latent <- data.frame(
-  posterior_shrinkage = posterior_shrinkage
+  posterior_contraction = posterior_contraction
 ) %>%
   tibble::rownames_to_column("par") %>%
   filter(!(par %in% names(tmb$fit$obj$par))) %>%
   mutate(par_name = ifelse(str_detect(par, "\\["), str_extract(par, ".*(?=\\[)"), par)) %>%
   group_by(par_name) %>%
-  summarise(posterior_shrinkage = mean(posterior_shrinkage)) %>%
+  summarise(posterior_contraction = mean(posterior_contraction)) %>%
   mutate(type = "Latent (averaged)") %>%
   rename(par = par_name)
 
-pdf("posterior-shrinkage.pdf", h = 5, w = 8)
[…]

### 2023-05-23 src/astro/astro.Rmd
Move astronomy example to its own report, add EB to astronomy example #46

---
title: "Milky way mass estimation using PCA-AGHQ"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
bibliography: citations.bib
abstract: |
  **Background**

  **Task** 
---

We reanalyse the "estimating the mass of the Milky Way" example from @bilodeau2021stochastic.

Originally Alex used the `ipoptr` package.
However, it is now not possible, or at least very difficult, to install it (as of 01/04/2023).
We will here use the `nloptr` package instead.

```{r}
data(gcdatalist, package = "aghq")
precompile()

set.seed(4365789)

globalpath <- normalizePath(tempdir(), winslash = "/")
plotpath <- normalizePath(file.path(globalpath, "astro"), winslash = "/", mustWork = FALSE)
if (!dir.exists(plotpath)) dir.create(plotpath)
```

Get the TMB template from the `aghq` package, the compile and load it:

```{r}
file.copy(normalizePath(system.file("extsrc/01_astro.cpp", package = "aghq"), winslash = "/"), globalpath)
compile(normalizePath(file.path(globalpath, "01_astro.cpp"), winslash = "/"))
dyn.load(normalizePath(dynlib(file.path(globalpath, "01_astro")), winslash = "/"))
```

Parameter transformations:

```{r}
parambounds <- list(
  Psi0 = c(1, 200),
  gamma = c(0.3, 0.7),
  alpha = c(3.0, 3.7),
  beta = c(-0.5, 1)
)

get_Psi0 <- function(theta) {
  # theta = -log( (Psi0 - 1) / (200 - 1) )
  (parambounds$Psi0[2] - parambounds$Psi0[1]) * exp(-exp(theta)) + parambounds$Psi0[1]
}

get_theta1 <- function(Psi0) {
  log(-log((Psi0 - parambounds$Psi0[1]) / (parambounds$Psi0[2] - parambounds$Psi0[1])))
}

get_gamma <- function(theta) {
  # theta = -log( (gamma - 0.3) / (0.7 - 0.3) )
  (parambounds$gamma[2] - parambounds$gamma[1]) * exp(-exp(theta)) + parambounds$gamma[1]
}

get_theta2 <- function(gamma) {
  log(-log((gamma - parambounds$gamma[1]) / (parambounds$gamma[2] - parambounds$gamma[1])))
}

get_alpha <- function(theta) {
  # theta = log(alpha - 3)
  exp(theta) + parambounds$alpha[1]
}

get_theta3 <- function(alpha) {
  log(alpha - parambounds$alpha[1])
}

get_beta <- function(theta) {
  # theta = -log( (beta - (-0.5)) / (1 - (-0.5)) )
  (parambounds$beta[2] - parambounds$beta[1]) * exp(-exp(theta)) + parambounds$beta[1]
}

get_theta4 <- function(beta) {
  log(-log((beta - parambounds$beta[1]) / (parambounds$beta[2] - parambounds$beta[1])))
}
```

# Optimisation

Optimisation using IPOPT replaced by optimisation using `nloptr`:

```{r}
nloptr_objective <- function(theta) ff$fn(theta)
nloptr_objective_gradient <- function(theta) ff$gr(theta)
```

We use box constraints, to improve stability of optimization:

```{r}
lowerbounds <- c(
  get_theta1(parambounds$Psi0[2] - 0.001),
  get_theta2(parambounds$gamma[2] - 0.001),
  get_theta3(parambounds$alpha[1] + 0.001),
  get_theta4(parambounds$beta[2] - 0.001)
)

upperbounds <- c(
  get_theta1(parambounds$Psi0[1] + 1),
  get_theta2(parambounds$gamma[1] + 0.01),
  get_theta3(parambounds$alpha[2] - 0.01),
  get_theta4(parambounds$beta[1] + 0.01)
)
```

Start optimisation at the midpoint:

```{r}
thetastart <- (upperbounds + lowerbounds) / 2
thetastart_list <- list(theta1 = thetastart[1], theta2 = thetastart[2], theta3 = thetastart[3], theta4 = thetastart[4])

data.frame(
  "param" = paste0("theta", 1:4),
  lowerbound = lowerbounds,
  upperbound = upperbounds,
  thetastart = thetastart
) %>%
  ggplot(aes(ymin = lowerbound, y = thetastart, ymax = upperbound, x = param)) +
    geom_pointrange() +
    coord_flip() +
    labs(x = "", y = "") +
    theme_minimal()
```

Nonlinear constraints, specified as a function:

```{r}
nloptr_nonlinear_constraints <- function(theta) -1 * Es$fn(theta)

nloptr_nonlinear_constraints_jacobian <- function(theta) {
  J <- -1 * Es$gr(theta)
  as.matrix(J)
}

nonlinear_lowerbound <- rep(0, nrow(gcdata) + 2)
nonlinear_upperbound <- rep(Inf, nrow(gcdata) + 2)
```

Objective function and its derivatives:

```{r}
ff <- MakeADFun(
  data = gcdatalist,
  parameters = thetastart_list,
  DLL = "01_astro",
  ADreport = FALSE,
  silent = TRUE
)
```

Nonlinear constraints and their Jacobian:

```{r}
Es <- MakeADFun(
  data = gcdatalist,
  parameters = thetastart_list,
  DLL = "01_astro",
  ADreport = TRUE,
  silent = TRUE
)
```

Check starting value satisfies constraints.
Note that constraints in `nloptr` are of the form $g(x) \leq 0$.

```{r}
stopifnot(all(nloptr_nonlinear_constraints(thetastart) <= 0))

tm <- Sys.time()

tmp <- capture.output(
  nloptr_result <- nloptr::nloptr(
    x0 = thetastart,
    eval_f = nloptr_objective,
    eval_grad_f = nloptr_objective_gradient,
    lb = lowerbounds,
    ub = upperbounds,
    eval_g_ineq = nloptr_nonlinear_constraints,
    eval_jac_g_ineq = nloptr_nonlinear_constraints_jacobian,
    opts = list("algorithm" = "NLOPT_LD_MMA", obj_scaling_factor = 1, "xtol_rel" = 1.0e-3)
  )
)

optruntime <- difftime(Sys.time(), tm, units = "secs")
cat("Run time for mass model optimisation:", optruntime, "seconds.\n")
```

Check that the optimisation results look right (based on Figure 2 of @bilodeau2021stochastic).

```{r}
get_Psi0(nloptr_result$solution[1])
get_gamma(nloptr_result$solution[2])
get_alpha(nloptr_result$solution[3])
get_beta(nloptr_result$solution[4])
```

# AGHQ

Create the template for AGHQ:

```{r}
ffa <- list(
  fn = function(x) -1 * ff$fn(x),
  gr = function(x) -1 * ff$gr(x),
  he = function(x) -1 * ff$he(x)
)

useropt <- list(
  ff = ffa,
  mode = nloptr_result$solution,
  hessian = ff$he(nloptr_result$solution)
)

#' Fit it with "reuse" first, then correct marginals later
cntrl <- aghq::default_control(negate = TRUE)

#' Do the quadrature
tm <- Sys.time()

k_dense <- 5
astroquad <- aghq::aghq(ff, k = k_dense, thetastart, optresults = useropt, control = cntrl)

#' Correct the marginals except for beta, which gave NA due to the constraints
for (j in 1:3) astroquad$marginals[[j]] <- marginal_posterior(astroquad, j, method = "correct")
quadruntime <- difftime(Sys.time(), tm, units = "secs")

cat("Run time for mass model quadrature:", quadruntime ,"seconds.\n")

#' Total runtime for AGHQ
optruntime + quadruntime
```

# EB

```{r}
tm <- Sys.time()

k_eb <- 1
astroebquad <- aghq::aghq(ff, k = k_eb, thetastart, optresults = useropt, control = cntrl)

#' Correct the marginals except for beta, which gave NA due to the constraints
for (j in 1:3) astroebquad$marginals[[j]] <- marginal_posterior(astroebquad, j, method = "correct")
quadebruntime <- difftime(Sys.time(), tm, units = "secs")

cat("Run time for mass model quadrature:", quadebruntime ,"seconds.\n")

#' Total runtime for EB
optruntime + quadebruntime
```

# PCA-AGHQ

```{r}
C <- Matrix::forceSymmetric(solve(useropt$hessian))
eigenC <- eigen(C)
lambda <- eigenC$values

data.frame(
  n = 1:length(lambda),
  tv = cumsum(lambda / sum(lambda))
) %>% ggplot(aes(x = n, y = tv)) +
  geom_point() +
  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
  annotate("text", x = 3, y = 0.875, label = "90% of total variation explained", col = "grey") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PCA dimensions included", y = "Total variation explained") +
  theme_minimal()

k_pca <- 5
s <- 1 #' Could choose this based on the Scree plot
m <- length(useropt$mode)
levels <- c(rep(k_pca, s), rep(1, m - s))
pca_grid <- mvQuad::createNIGrid(dim = m, type = "GHe", level = levels)
mvQuad::rescale(pca_grid, m = useropt$mode, C = C, dec.type = 1)

#' Check that the PCA grid has the correct number of nodes
stopifnot(nrow(mvQuad::getNodes(pca_grid)) == k_pca^s)

tm <- Sys.time()

astropcaquad <- local_aghq(ff, k_pca, thetastart, optresults = useropt, basegrid = pca_grid, adapt = FALSE, control = cntrl)

#' Correct the marginals except for beta, which gave NA due to the constraints
for (j in 1:3) astropcaquad$marginals[[j]] <- marginal_posterior(astropcaquad, j, method = "correct")
pcaquadruntime <- difftime(Sys.time(), tm, units = "secs")

cat("Run time for mass model quadrature:", pcaquadruntime ,"seconds.\n")

#' Total runtime for PCA-AGHQ
optruntime + pcaquadruntime
```

# NUTS

```{r}
stanmod <- tmbstan(
  ff,
  chains = 4,
  cores = 4,
  iter = 1e04,
  warmup = 1e03,
  init = nloptr_result$solution,
  seed = 48645,
  algorithm = "NUTS"
)

# Time
get_elapsed_time(stanmod)

standata <- as.data.frame(stanmod)
standata$Psi0 <- get_Psi0(standata[ ,1])
standata$gamma <- get_gamma(standata[ ,2])
standata$alpha <- get_alpha(standata[ ,3])
standata$beta <- get_beta(standata[ ,4])
```

# Comparison

## Plots

```{r}
#' Manual plots
parambounds <- list(
  Psi0 = c(1, 200),
  gamma = c(0.3, 0.7),
  alpha = c(3.0, 3.7),
  beta = c(-0.5, 1)
)

#' Add small buffers for stability
get_theta2_robust <- function(gamma) {
  log(-log((gamma - parambounds$gamma[1] + 1e-03) / (parambounds$gamma[2] - parambounds$gamma[1] + 1e-03)))
}

get_theta3_robust <- function(alpha) log(alpha - parambounds$alpha[1] + 1e-03)

#' Transformed PDFs
translist1 <- make_transformation(totheta = get_theta1, fromtheta = get_Psi0)
translist2 <- make_transformation(totheta = get_theta2_robust, fromtheta = get_gamma)
translist3 <- make_transformation(totheta = get_theta3_robust, fromtheta = get_alpha)
translist4 <- make_transformation(totheta = get_theta4, fromtheta = get_beta)

Psi0pdf <- compute_pdf_and_cdf(astroquad$marginals[[1]], translist1)
gammapdf <- compute_pdf_and_cdf(astroquad$marginals[[2]], translist2)
alphapdf <- compute_pdf_and_cdf(astroquad$marginals[[3]], translist3, interpolation = "polynomial")
betapdf <- compute_pdf_and_cdf(astroquad$marginals[[4]], translist4)

Psi0ebpdf <- compute_pdf_and_cdf(astroebquad$marginals[[1]], translist1)
gammaebpdf <- compute_pdf_and_cdf(astroebquad$marginals[[2]], translist2)
alphaebpdf <- compute_pdf_and_cdf(astroebquad$marginals[[3]], translist3, interpolation = "polynomial")
betaebpdf <- compute_pdf_and_cdf(astroebquad$marginals[[4]], translist4)

Psi0pcapdf <- compute_pdf_and_cdf(astropcaquad$marginals[[1]], translist1)
gammapcapdf <- compute_pdf_and_cdf(astropcaquad$marginals[[2]], translist2)
alphapcapdf <- compute_pdf_and_cdf(astropcaquad$marginals[[3]], translist3, interpolation = "polynomial")
betapcapdf <- compute_pdf_and_cdf(astropcaquad$marginals[[4]], translist4)

Psi0prior <- function(Psi0) dunif(Psi0, parambounds$Psi0[1], parambounds$Psi0[2], log = FALSE)
gammaprior <- function(gamma) dunif(gamma, parambounds$gamma[1], parambounds$gamma[2], log = FALSE)
alphaprior <- function(alpha) dgamma(alpha - parambounds$alpha[1], shape = 1, rate = 4.6, log = FALSE)
betaprior <- function(beta) dunif(beta, parambounds$beta[1], parambounds$beta[2], log = FALSE)

Psi0_postplot <- Psi0pdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = Psi0, y = after_stat(density)),
    bins = 50,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = Psi0pcapdf,
    inherit.aes = FALSE,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = Psi0prior, linetype = "dashed") +
  labs(title = "", x = "", y = "Density") +
  scale_x_continuous(breaks = seq(10, 60, by = 5)) +
  coord_cartesian(xlim = c(24, 43)) +
  theme_minimal()

gamma_postplot <- gammapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = gamma, y = after_stat(density)),
    bins = 100,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = gammapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = gammaprior, linetype = "dashed") +
  labs(title = "", x = "", y = "Density") +
  scale_x_continuous(breaks = seq(0, 0.44, by = 0.02)) +
  coord_cartesian(xlim = c(0.3, 0.4)) +
  theme_minimal()

alpha_postplot <- alphapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = alpha, y = after_stat(density)),
    bins = 2000,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = alphapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = alphaprior, linetype = "dashed") +
  labs(title = "",x = "",y = "Density") +
  scale_x_continuous(breaks = seq(3, 4, by = 0.02)) +
  coord_cartesian(xlim = c(3, 3.05)) +
  theme_minimal()

beta_postplot <- betapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = beta, y = after_stat(density)),
    bins = 50,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = betapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = betaprior, linetype = "dashed") +
  labs(title = "",x = "",y = "Density") +
  scale_x_continuous(breaks = seq(-0.5, 1, by = 0.1)) +
  coord_cartesian(xlim = c(-0.3, 0.4)) +
  theme_minimal()

{Psi0_postplot + gamma_postplot} / {alpha_postplot + beta_postplot}
```

## KS tests

```{r}
kstable <- function(quad) {
  aghqpostsamps_spline <- sample_marginal(astroquad, nrow(standata), interpolation = "spline")
  aghqpostsamps_poly <- sample_marginal(astroquad, nrow(standata), interpolation = "polynomial")

  suppressWarnings({
    kstable <- data.frame(
      Psi0 = ks.test(standata[[1]], aghqpostsamps_poly[[1]])$statistic,
      gamma = ks.test(standata[[2]], aghqpostsamps_spline[[2]])$statistic,
      alpha = ks.test(standata[[3]], aghqpostsamps_poly[[3]])$statistic,
      beta = ks.test(standata[[4]], aghqpostsamps_poly[[4]])$statistic
    )
  })

  return(kstable)
}

kstable_astroquad <- kstable(astroquad) %>%
  pivot_longer(cols = everything()) %>%
  mutate(method = "AGHQ", k = k_dense, s = NA, total_points = k^4)

kstable_astroebquad <- kstable(astroebquad) %>%
  pivot_longer(cols = everything()) %>%
  mutate(method = "EB", k = k_eb, s = NA, total_points = k_eb)

kstable_astropcaquad <- kstable(astroquad) %>%
  pivot_longer(cols = everything()) %>%
  mutate(method = "AGHQ-PCA", k = k_pca, s = s, total_points = k^s)

bind_rows(kstable_astroquad, kstable_astroebquad, kstable_astropcaquad) %>%
  gt::gt()

bind_rows(kstable_astroquad, kstable_astroebquad, kstable_astropcaquad) %>%
  ggplot(aes(x = name, y = value, col = method)) +
    geom_jitter(height = 0, width = 0.05) +
    scale_y_continuous(limits = c(0, 0.05)) +
    scale_colour_manual(values = multi.utils::cbpalette()) +
    labs(x = "Parameter", y = "KS(method, NUTS)", col = "Method") +
    theme_minimal()
```

# Original computing environment {-}

```{r}
sessionInfo()
```

# Bibliography {-}

### 2023-05-23 src/astro/functions.R

@@ -1,0 +1,288 @@
+#' A local version of naomi::make_tmb_obj, edited to work with DLL = "naomi_simple"
+local_make_tmb_obj <- function(data, par, calc_outputs = 0L, inner_verbose, progress = NULL, map = NULL, DLL = "naomi_simple") {
+  # Begin expose naomi:::make_tmb_obj
+  # https://github.com/mrc-ide/naomi/blob/e9de40f12cf2e652f78966bb351fa5718ecd7867/R/tmb-model.R#L496
+  data$calc_outputs <- as.integer(calc_outputs)
+
+  integrate_out <- c(
+    "beta_rho", "beta_alpha", "beta_alpha_t2", "beta_lambda", "beta_asfr",
+    "beta_anc_rho", "beta_anc_alpha", "beta_anc_rho_t2", "beta_anc_alpha_t2",
+    "u_rho_x", "us_rho_x", "u_rho_xs", "us_rho_xs", "u_rho_a", "u_rho_as",
+    "u_rho_xa", "u_alpha_x", "us_alpha_x", "u_alpha_xs", "us_alpha_xs",
+    "u_alpha_a", "u_alpha_as", "u_alpha_xt", "u_alpha_xa", "u_alpha_xat",
+    "u_alpha_xst", "ui_lambda_x", "logit_nu_raw", "ui_asfr_x", "ui_anc_rho_x",
+    "ui_anc_alpha_x", "ui_anc_rho_xt", "ui_anc_alpha_xt", "log_or_gamma",
+    "log_or_gamma_t1t2"
+  )
+
+  if (DLL == "naomi_simple") {
+    exclude_random_pars <- c(
+      "beta_alpha_t2", "beta_asfr", "beta_anc_rho_t2", "beta_anc_alpha_t2",
+      "u_alpha_xt", "u_alpha_xat", "u_alpha_xst", "logit_nu_raw", "ui_asfr_x",
+      "ui_anc_rho_xt", "ui_anc_alpha_xt", "log_or_gamma_t1t2"
+    )
+
+    integrate_out <- setdiff(integrate_out, exclude_random_pars)
+  }
+
+  if(DLL == "naomi_simple_x_index") {
+    integrate_out <- "x_minus_i"
+  }
+
+  obj <- TMB::MakeADFun(
+    data = data,
+    parameters = par,
+    DLL = DLL,
+    silent = !inner_verbose,
+    random = integrate_out,
+    map = map
+  )
+
+  if (!is.null(progress)) {
[…]

### 2023-05-23 src/astro/script.R

@@ -1,0 +1,6 @@
+#' Uncomment and run the two line below to resume development of this script
+# orderly::orderly_develop_start("astro")
+# setwd("astro")
+
+#' Run astronomy example
+rmarkdown::render("astro.Rmd")

### 2023-05-24 src/naomi-simple_ks/ks.Rmd
Clean up KS report

@@
 }
-
 ```
@@
 ks_helper("beta", starts_with = TRUE)
+
+#' Save off KS results for one parameter as example
+saveRDS(
+  to_ks_df(par = "beta_alpha", starts_with = TRUE),
+  file = "beta_alpha.rds"
+)
 ```
@@
 ```{r}
-ks_df_out(par = "rho_t1_out") %>%
+ks_df_out(par = "lambda_t1_out") %>%
   filter(!age_group %in% c("Y005_009", "Y010_014", "Y080_999")) %>%
@@
 
-Filter out the age group 5-9, 10-14 and 80+ here becausethere are new new infections in those age groups, and the posterior samples from each method are exactly the same: in particular draws of identically zero.
+Filter out the age group 5-9, 10-14 and 80+ here because there are new new infections in those age groups, and the posterior samples from each method are exactly the same: in particular draws of identically zero.
 
 ```{r}
-ks_lambda_t1_out %>%
-  filter(ks == 0) %>%
-  group_by(method) %>%
-  summarise(n = n())
-
-ks_lambda_t1_out %>%
-  filter(ks == 0) %>%
-  pull(age_group) %>%
-  unique()
+# ks_lambda_t1_out %>%
+#   filter(ks == 0) %>%
+#   group_by(method) %>%
+#   summarise(n = n())
+# 
+# ks_lambda_t1_out %>%
+#   filter(ks == 0) %>%
+#   pull(age_group) %>%
+#   unique()
 ```
@@
 
-saveRDS(ks_summary_out, "ks_summary_out.rds")
+saveRDS(ks_summary, "ks_summary.rds")
 
@@
 ```{r}
-(ks_summary_out <- ks_summary %>%
+ks_summary %>%
   group_by(Type) %>%
@@
     adam = mean(`KS(adam, tmbstan)`)
-  )) %>%
+  ) %>%
   gt::gt()
@@
 ```{r}
-ks_table(ks_summary, type = "Latent field")
+ks_summary %>%
+  filter(Type == "Latent field") %>%
+  group_by(Parameter) %>%
+  summarise(
+    `KS(TMB, tmbstan)` = mean(`KS(TMB, tmbstan)`),
+    `KS(aghq, tmbstan)` = mean(`KS(aghq, tmbstan)`),
+    `KS(adam, tmbstan)` = mean(`KS(adam, tmbstan)`)
+  ) %>%
+  ks_table()
 ```
@@
 ```{r}
-ks_table(ks_summary, type = "Latent field")
+# ks_summary %>%
+#   filter(Type == "Hyper") %>%
+#   group_by(Parameter) %>%
+#   summarise(
+#     `KS(TMB, tmbstan)` = mean(`KS(TMB, tmbstan)`),
+#     `KS(aghq, tmbstan)` = mean(`KS(aghq, tmbstan)`),
+#     `KS(adam, tmbstan)` = mean(`KS(adam, tmbstan)`)
+#   ) %>%
+#   ks_table()
 ```

### 2023-05-24 src/docs_paper/appendix.Rmd
Update write-up

@@
 \hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
-\mathbf{H} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
-\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
+\hat{\mathbf{H}} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
+\hat{\mathbf{H}}^{-1} &= \hat{\mathbf{L}} \hat{\mathbf{L}}^\top,
 \end{align*}
@@
 \hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
-\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
+\hat{\mathbf{H}}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
 \end{align*}
@@
 4.
-For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
+For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\hat{\mathbf{H}}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
 A value of $l \geq 4$ is recommended to enable B-spline interpolation.
@@
 \hat \x_{-i}(x_i, \btheta) &= \argmax_{\x_{-i}} \log p(\y, x_i, \x_{-i}, \btheta), \\
-\mathbf{H}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
+\hat{\mathbf{H}}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
 \end{align*}

### 2023-05-24 src/docs_paper/paper.Rmd
Update write-up

@@
     label: e1
-    addressLabel: A,D
+    addressLabel: A,E
     sepNext: ","
@@
     label: e4
-    addressLabel: D
+    addressLabel: D,E
 affiliation:
@@
     authorsLabels: e3
-  - label: D
+  - label: E
     name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
     authorsLabels: e4
+  - label: D
+    name: Center for Communicable Disease Dynamics, Harvard T.H. Chan School of Public Health, Harvard University
 abstract: |
@@
 Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they are conducted only every five years or so, with limited sample size at the district level.
-Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in more real-time but based on limited or non-representative samples of the population.
+Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available closer to real-time but based on limited or non-representative samples of the population.
 To address these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
@@
 First, due to dependence of observations on multiple structured additive predictors, Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast].
-Furthermore, as well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number typically required for use of integrated nested Laplace approximations [@rue2009approximate].
-Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production across a range of countries.
+Further, as well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number typically required for use of integrated nested Laplace approximations [@rue2009approximate].
+Second, any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production across a range of countries.
 In this setting, Markov chain Monte Carlo (MCMC) approaches are prohibitively slow, both due to the scale of the model and challenging features of its posterior geometry [@neal2003slice].
+
 Inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
@@
 Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
-In the Naomi model, this subset is the high-dimensional latent field.
+In the Naomi model, this subset is the high-dimensional latent field, leaving the smaller number of hyperparameters.
 Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
-Although this approach is fast, it has the downside that within the empirical Bayes framework hyperparameter uncertainty is not accounted for in the latent field posterior.
+Although this approach is fast, it has the downside that within the empirical Bayes framework hyperparameter uncertainty is not properly accounted for in the latent field posterior.
 This has motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
 
-To obtain fast, accurate Bayesian inferences for the Naomi model we develop an inference methodology which extends adaptive Gauss-Hermite quadrature (AGHQ) to handle a higher number of hyperparameters.
+To obtain fast, accurate Bayesian inferences for the Naomi model we develop an inference methodology which extends adaptive Gauss-Hermite quadrature to handle a higher number of hyperparameters. 
 We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
-Since `aghq` is designed to naturally interface with `TMB`, use of our method is simple when provided a \textsf{C++} user template for the log-posterior.
+As `aghq` is designed to naturally interface with `TMB`, use of our method is simple when provided a \textsf{C++} user template for the log-posterior.
 
@@
 
-@eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points $T_i, i = 1, 2, 3$.
-We consider a simplified version defined only at the time of the most recent household survey with HIV testing ($T_1$), omitting nowcasting ($T_2$) and temporal projection ($T_3$) which involve limited additional inference.
-An overview of this simplified model is given below.
-A more complete mathematical description (Appendix S1) as well as a \textsf{C++} template for the log-posterior (Appendix S3) are provided in the supplementary material.
+@eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points $\{T_i\}_{i = 1, 2, 3}$.
+We consider a simplified version defined only at the time of the most recent household survey with HIV testing ($T_1$), omitting nowcasting ($T_2$) and temporal projection ($T_3$) which involve limited inferences.
+An overview of the simplified model is given below, and a more complete mathematical description (Appendix S1) as well as a \textsf{C++} template for the log-posterior (Appendix S2) are provided in the supplementary material.
 
@@
 For ease of notation, let $i$ index the finest district-age-sex division included in the model.
-Let $\mathcal{I} \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported.
+Let $I \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported, with $I \in \mathcal{I}$.
 
@@
 HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i$, and depends on adult HIV prevalence and adult ART coverage. 
-Let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
+Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
+We link this proportion to HIV incidence via
 \begin{equation}
@@
 \begin{equation*}
-\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}
+\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}
 \end{equation*}
-be weighted, aggregate survey observations, where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
-The observed number of outcomes are $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
+be weighted, aggregate survey observations, where $j$ indexes individuals across all strata $i \in I$ and $w_j$ are design weights.
+The observed number of outcomes are $y^{\hat \theta}_{I} = m^{\hat \theta}_{I} \cdot \hat \theta_{I}$ where
 \begin{equation*}
-m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
+m^{\hat \theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
 \end{equation*}
@@
 \begin{equation*}
-y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})
+y^{\hat \theta}_{I} \sim \text{xBin}(m^{\hat \theta}_{I}, \theta_{I})
 \end{equation*}
-to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following weighted aggregates
+to model these aggregate observations, where $\theta_{I}$ are the following weighted aggregates
 \begin{equation*}
-\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \quad
-\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \quad
-\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
+\rho_{I} = \frac{\sum_{i \in I} N_i \rho_i}{\sum_{i \in I} N_i}, \quad
+\alpha_{I} = \frac{\sum_{i \in I} N_i \rho_i \alpha_i}{\sum_{i \in I} N_i \rho_i}, \quad
+\kappa_{I} = \frac{\sum_{i \in I} N_i \rho_i \kappa_i}{\sum_{i \in I} N_i \rho_i}.
 \end{equation*}
@@
 These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
-In particular, the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
+In particular, the number of ANC clients with ascertained status $x^\text{ANC}_I$, the number of those with positive status $y^\text{ANC}_I$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_I$.
 We use the binomial working likelihoods
 \begin{align*}
-y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(x^\text{ANC}_\mathcal{I}, \rho^\text{ANC}_{\mathcal{I}}), \\
-z^\text{ANC}_\mathcal{I} &\sim \text{Bin}(y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}),
+y^\text{ANC}_I &\sim \text{Bin}(x^\text{ANC}_I, \rho^\text{ANC}_{I}), \\
+z^\text{ANC}_I &\sim \text{Bin}(y^\text{ANC}_I, \alpha^\text{ANC}_{I}),
 \end{align*}
@@
 \begin{equation*}
-\rho^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in \mathcal{I}} \Psi_i}, \quad
-\alpha^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}},
+\rho^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in I} \Psi_i}, \quad
+\alpha^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}},
 \end{equation*}
@@
 The log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ are modelled using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
-Aggregate ART attendance data $\dot A_\mathcal{I}$ is modelled using a Gaussian approximation to a sum of binomials.
-The sum results both by aggregation over $i \in \mathcal{I}$ and by number of ART clients travelling from district $x'$ to $x$.
+Aggregate ART attendance data $\dot A_I$ is modelled using a Gaussian approximation to a sum of binomials.
+The sum results both by aggregation over $i \in I$ and by number of ART clients travelling from district $x'$ to $x$.
 More details regarding this part of the model are provided in Appendix S1.
+
+## Collected together
+
+Let $\mathbf{y} = (y^{\hat \theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa\}$ and $I \in \mathcal{I}$ be the vector of observations.
+Attempt to write fully the Naomi model in a small number of collected together equations here.
 
@@
 
-We now describe a popular class of models, and an extension encapsulates the complexities of Naomi.
+We now describe the popular latent Gaussian class of models, and an extension which encapsulates the complexities of Naomi.
 
@@
 
-Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\Hb}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
 \begin{align*}
 \hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
-\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
+\hat {\Hb}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
 \end{align*}
 Then the Laplace approximation to $p(\btheta, \y)$ is given by
+\begin{equation}
+\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}. \label{eq:la}
+\end{equation}
+Inference in TMB proceeds by optimising Equation \ref{eq:la} to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
+Latent field joint and marginal inferences are then direct from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
+
+## Gauss-Hermite quadrature
+
+Quadrature rules can be used to approximate the integral of $\tilde p_\texttt{LA}(\btheta, \y)$ via the weighted sum
+\begin{equation}
+p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \text{d}\btheta \approx \sum_{\z \in \mathcal{Q}} p_\texttt{LA}(\z, \y) \omega(\z), \label{eq:quad}
+\end{equation}
+where $\z \in \mathcal{Q}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
+Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule where, in the univariate case, the nodes $\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}$ are selected as zeros of the $k$th Hermite polynomial.
+The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by
 \begin{equation*}
-\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+\omega(z) = \frac{k!}{[H_{k + 1}(z)]^2 \phi(z)},
+\end{equation*}
+where $\phi(\cdot)$ is a standard univariate Gaussian density.
+Multivariate GHQ rules are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
+GHQ is attractive because it is exact for functions which are a polynomial of total order no more than $2k - 1$ multiplied by a Gaussian density.
+
+## Adaptive quadrature
+
+In adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] the nodes are shifted and rotated to suit the particular integrand.
+Repositioning the nodes is particularly important for statistical quadrature problems, where the integral depends on data $\y$ and so regions of high density are not known in advance.
+To obtain an AGHQ estimate of Equation \ref{eq:quad}, let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ a matrix decomposition of the inverse curvature, then
+\begin{equation}
+\tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z).
+\label{eq:aghq}
+\end{equation}
+Two possibilities for the matrix decomposition are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$ and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$ [@jackel2005note].
+Equation \ref{eq:aghq} may be used to normalise the Laplace approximation
+\begin{equation*}
+\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
 \end{equation*}
 
-## AGHQ
+For the latent field, the adapted nodes and weights may be reused [@rue2009approximate; @stringer2022fast] in the nested approximation
+\begin{equation}
+\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z).
+\label{eq:nest}
+\end{equation}
+Samples from this mixture of Gaussians are obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
 
-A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by the following weighted sum
-\begin{equation*}
-\int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z),
-\end{equation*}
-where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
-In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials, with weights chosen so as to interpolate functions of the form $f(\z) = \tilde f(\z) \exp(-\z^2)$.
-In adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate] the nodes and weights are shifted and rotated to suit the particular integrand, based on the mode, curvature at the mode, and a matrix decomposition of the inverse curvature.
-Two possibilities for this matrix decomposition are the Cholesky decomposition and the spectral decomposition [@jackel2005note].
+## Principal components analysis
 
-```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension (a) are adapted based on the mean (b) and covariance matrix of the target via $f(\\z) = \\hat f + \\mathbf{L} \\z$, where..."}
+Use of the product rule requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points, which becomes intractable as $m$ increases for $k > 1$.
+An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$, and define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
+Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
+This choice of nodes, taken together with use of the spectral decomposition, is analogous to a principal components analysis (PCA) approach to AGHQ
+\begin{equation}
+\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
+\end{equation}
+where $\hat{\mathbf{E}}_{\texttt{LA}, s} = \cdots$ and $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s} = \cdots$.
+
+## Sparse rules
+
+Though there are sparse rules which retain the attractive exactness properties of GHQ, with fewer than $k^m$ quadrature points, they use weighting functions which may be negative $\omega(\z) < 0$.
+This introduces a problem as we are not aware of any suitable approaches to obtain multinomial samples when $\lambda(\z)$ is negative.
+
+```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension are adapted based on the mean and covariance matrix of the target via the Cholesky decomposition or spectral decompostion of the inverse curvature at the mode."}
 mu <- c(1, 1.5)
@@
 plot1 <- add_points(plot0, gg) +
-  labs(size = "", subtitle = "(a)")
+  labs(size = "")
 
@@
 plot2 <- add_points(plot0, gg2) +
-  labs(size = "", subtitle = "(b)")
+  labs(size = "")
 
@@
 plot3 <- add_points(plot0, gg3) +
-  labs(size = "", subtitle = "(c)")
+  labs(size = "")
 
@@
 plot4 <- add_points(plot0, gg4) +
-  labs(size = "", subtitle = "(d)")
+  labs(size = "")
 
-(plot1 + plot2) / (plot3 + plot4)
+#' PCA-AGHQ
+gg5 <- mvQuad::createNIGrid(2, "GHe", level = c(3, 1))
+mvQuad::rescale(gg5, m = mu, C = cov, dec.type = 1)
+
+plot5 <- add_points(plot0, gg5) +
+  labs(size = "")
+
+(plot1 + plot2) / (plot3 + plot4 + plot5)
 ```
-
-## TMB
-
-## PCA-AGHQ
 
@@
 We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
-For each approach, the `TMB` C++ user-template (available in the appendix) used to specify the log-posterior was the same.
-The four approaches were:
-1. TMB: EB combined with a Gaussian approximation via `TMB`,
-2. PCA-AGQH: the PCA-AGHQ grid combined with a Gaussian approximation via `aghq`,
+For each approach, the `TMB` C++ user-template (Appendix S2) used to specify the log-posterior was the same.
+The three approaches were:
+1. TMB,
+2. PCA-AGHQ, and
 3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
-The dimension of the latent field is $N = 467$ and the dimension of the hyperparameters is $m = 24$.
-Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
+The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
+Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
 For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
@@
 \midrule
-TMB & \texttt{TMB} & Empirical Bayes, $1000$ samples \\
+TMB & \texttt{TMB} & $1000$ samples \\
 PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$, $1000$ samples \\
-NUTS & \texttt{tmbstan} & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$.
-HMC parameters set to default for software.
+NUTS & \texttt{tmbstan} & $4$ chains of $20000$ iterations, with the first $10000$ iterations of each chain discarded as warmup, thinned by a factor of $20$.
+Default NUTS tuning parameters \citep{hoffman2014no}.
 \\
@@
 
-We assessed the quality of our NUTS results to be satisfactory using the potential scale reduction factor $\hat R$ [@gelman1992inference; @vehtari2021rank], bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments [@betancourt2017conceptual].
-Full details are provided in the appendix.
-We thus treat the results from NUTS as a gold-standard to which other inferential methods are compared.
+We obtained satisfactory NUTS results by gradually increasing the amount of computation until all diagnostics were acceptable.
+These include the potential scale reduction factors $\hat R$ [@gelman1992inference; @vehtari2021rank] for each parameter, bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments [@betancourt2017conceptual].
+For full details see Appendix S3.
+We treat the NUTS results as a gold-standard.
+
+## PCA-AGHQ settings
+
+We used a Scree plot to select the number of principal components $s < m$ to keep.
+Figure showing Scree plot.
+Figure showing approximation to precision matrix.
+Figure showing positions of nodes on hyperparameter marginal posteriors as compared with NUTS.
 
@@
 
-To assess the informativeness of the data on each model parameter $\phi$ we compared (Figure \ref{fig:contraction}) the prior variance $\sigma_\text{prior}^2(\phi)$ to the posterior variance $\sigma_\text{posterior}^2(\phi)$ via the posterior contraction [@schad2021toward].
+Let $\phi$ be a generic model parameter.
+To assess the informativeness of the data we compared (Figure \ref{fig:contraction}) the prior variance $\sigma_\text{prior}^2(\phi)$ to the posterior variance $\sigma_\text{posterior}^2(\phi)$ via the posterior contraction [@schad2021toward]
 \begin{equation*}
@@
 \end{equation*}
+For the parameters of length greater than one, we averaged the posterior contraction.
 
@@
 
-We used three statistical approaches to assess the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS (1) Kolmogorov-Smirnov tests, (2) Pareto-smoothed importance sampling, and (3) maximum mean discrepancy.
+We used three statistical approaches to assess the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS: (1) Kolmogorov-Smirnov tests, (2) Pareto-smoothed importance sampling, and (3) maximum mean discrepancy.
 
@@
 
-The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is given by the maximum absolute difference between two empirical cumulative distribution (ECDF) functions $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
-For each method we compare the KS statistics
-\begin{equation*}
-D_\bullet = \sup_\varphi | F_\text{NUTS}(\varphi) - F_\bullet(\varphi)|.
-\end{equation*}
+The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two empirical cumulative distribution (ECDF) functions $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
+We compare the KS statistics
+\begin{align*}
+D_\text{TMB} &= \sup_\varphi | F_\text{NUTS}(\varphi) - F_\text{TMB}(\varphi)|, \\
+D_\text{PCA-AGHQ} &= \sup_\varphi | F_\text{NUTS}(\varphi) - F_\text{PCA-AGHQ}(\varphi)|.
+\end{align*}
 See a summary of the results in Table and Figure \ref{fig:ks}, and full results available in the appendix.
@@
 ```{r ks, fig.cap="Results of Kolmogorov-Smironv tests analysis."}
-qplot(1:10, 1:10) + theme_minimal()
+qplot(1:3, 1:3) + theme_minimal()
 ```
@@
 
-As marginal distributions, we are interested in assessing the accuracy of joint distributions.
-Let $\{\bphi_i\}_{i = 1}^n$ be samples from a joint posterior.
+As well as marginal distributions, we are interested in assessing the accuracy of joint distributions.
+Let $\{\bphi_i\}_{i = 1}^n$ be generic samples from a joint posterior.
 Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios $r_s = p(\theta_s, y) / q(\theta)$ used in importance sampling.
@@
 be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
-See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.
+See a summary of the results in Table and Figure \ref{fig:psis}.
 
 ```{r psis, fig.cap="Results of Pareto-smoothed importance sampling analysis."}
-qplot(1:10, 1:10) + theme_minimal()
+qplot(1:3, 1:3) + theme_minimal()
 ```
@@
 \end{equation*}
-See a summary of the results in Table and Figure \ref{fig:mmd}, and full results available in the appendix.
+See a summary of the results in Table and Figure \ref{fig:mmd}.
 
 ```{r mmd, fig.cap="Results of maximum mean discrepancy analysis."}
-qplot(1:10, 1:10) + theme_minimal()
+qplot(1:3, 1:3) + theme_minimal()
 ```

### 2023-05-24 src/docs_paper/paper.Rmd
Add KS results for latent field to manuscript

@@
 \end{align*}
-See a summary of the results in Table and Figure \ref{fig:ks}, and full results available in the appendix.
+See a summary of the results in Table.
 
 ```{r ks, fig.cap="Results of Kolmogorov-Smironv tests analysis."}
-qplot(1:3, 1:3) + theme_minimal()
-```
+ks_summary <- readRDS("depends/ks_summary.rds")
 
-```{r}
-ks_summary <- readRDS("depends/ks_summary_out.rds")
-
-# gt::gt(ks_summary) %>% gt::as_latex()
+ks_summary %>%
+  filter(Type == "Latent field") %>%
+  group_by(Parameter) %>%
+  summarise(
+    `D(TMB)` = round(mean(`KS(TMB, tmbstan)`), 3),
+    `D(PCA-AGHQ)` = round(mean(`KS(aghq, tmbstan)`), 3)
+  ) %>%
+  gt::gt(rowname_col = "Parameter") %>%
+  gt::summary_rows(
+   columns = everything(),
+   fns = list(Average = ~ mean(.x)),
+   formatter = gt::fmt_number,
+   decimals = 3
+  )
 ```
@@
 Let $\{\bphi_i\}_{i = 1}^n$ be generic samples from a joint posterior.
-Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios $r_s = p(\theta_s, y) / q(\theta)$ used in importance sampling.
-Let
-\begin{equation*}
-p(y \, | \, \mu, \sigma, k) =
-\begin{cases}
-\frac{1}{\sigma} \left( 1 + k \left( \frac{y - \mu}{\sigma} \right) \right)^{- \frac{1}{k} - 1}, \quad k \neq 0 \\
-\frac{1}{\sigma} \exp \left( \frac{y - \mu}{\sigma} \right), \quad k = 0
-\end{cases}
-\end{equation*}
-be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
+Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios used in importance sampling.
 See a summary of the results in Table and Figure \ref{fig:psis}.

### 2023-05-24 src/naomi-simple_ks/ks.Rmd
Produce beta_anc_rho output too in KS

@@
 ```{r}
-pars_eval <- pars %in% c("beta_rho", "u_rho_x")
+pars_eval <- pars %in% c("beta_rho", "beta_alpha", "beta_anc_rho", "u_rho_x")
 names(pars_eval) <- pars
@@
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_alpha")
+```{r eval = pars_eval[["beta_alpha"]]}
+beta_alpha <- histogram_and_ecdf("beta_alpha", return_df = TRUE)
+beta_alpha$plot
+
+#' Save off KS results as example
+saveRDS(
+  beta_alpha$df_compare,
+  file = "beta_alpha.rds"
+)
 ```
@@
 
-```{r eval = pars_eval[["beta_rho"]]}
+```{r eval = pars_eval[["beta_lambda"]]}
 histogram_and_ecdf_helper("beta_lambda")
@@
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf("beta_anc_rho")
+```{r eval = pars_eval[["beta_anc_rho"]]}
+beta_anc_rho <- histogram_and_ecdf("beta_anc_rho", i = 1, return_df = TRUE)
+beta_anc_rho$plot
+
+#' Save off KS results as example
+saveRDS(
+  beta_anc_rho$df_compare,
+  file = "beta_anc_rho.rds"
+)
 ```
@@
 
-```{r eval = pars_eval[["beta_rho"]]}
+```{r eval = pars_eval[["beta_anc_alpha"]]}
 histogram_and_ecdf("beta_anc_alpha")
@@
 ks_helper("beta", starts_with = TRUE)
-
-#' Save off KS results for one parameter as example
-saveRDS(
-  to_ks_df(par = "beta_alpha", starts_with = TRUE),
-  file = "beta_alpha.rds"
-)
 ```

### 2023-05-24 src/docs_paper/paper.Rmd
Example KS plot into manuscript

@@
 \end{align*}
-See a summary of the results in Table.
+See an illustration of the KS test in Figure \ref{fig:ks} and a summary of the results in Table.
 
-```{r ks, fig.cap="Results of Kolmogorov-Smironv tests analysis."}
+```{r ks, fig.height = 3, fig.cap="Example KS test for one parameter."}
+df_compare <- readRDS("depends/beta_alpha.rds") %>%
+  filter(method != "adam")
+
+colours <- c("#56B4E9", "#009E73", "#E69F00")
+
+mean <- df_compare %>%
+  filter(method == "tmbstan") %>%
+  summarise(mean = mean(samples)) %>%
+  pull(mean) %>%
+  round(digits = 3)
+
+sd <- df_compare %>%
+  filter(method == "tmbstan") %>%
+  summarise(sd = sd(samples)) %>%
+  pull(sd) %>%
+  round(digits = 3)
+
+histogram <- ggplot(df_compare, aes(x = samples, fill = method, col = method)) +
+  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +
+  theme_minimal() +
+  facet_grid(method~.) +
+  labs(x = "beta_alpha", y = "Density", fill = "Method") +
+  scale_color_manual(values = colours) +
+  scale_fill_manual(values = colours) +
+  theme(legend.position = "none")
+
+grid <- seq(from = min(df_compare$samples), to = max(df_compare$samples), length.out = 1000)
+
+tmb_ecdf <- stats::ecdf(filter(df_compare, method == "TMB") %>% pull(samples))
+tmb_ecdf_df <- data.frame(x = grid, ecdf = tmb_ecdf(grid), method = "TMB")
+
+aghq_ecdf <- stats::ecdf(filter(df_compare, method == "aghq") %>% pull(samples))
+aghq_ecdf_df <- data.frame(x = grid, ecdf = aghq_ecdf(grid), method = "aghq")
+
+tmbstan_ecdf <- stats::ecdf(filter(df_compare, method == "tmbstan") %>% pull(samples))
+tmbstan_ecdf_df <- data.frame(x = grid, ecdf = tmbstan_ecdf(grid), method = "tmbstan")
+
+# Add ECDF differences
+tmb_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - tmb_ecdf_df$ecdf
+aghq_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - aghq_ecdf_df$ecdf
+tmbstan_ecdf_df$ecdf_diff <- 0
+
+absmax <- function(x) x[which.max(abs(x))]
+
+ks_tmb <- absmax(tmb_ecdf_df$ecdf_diff)
+ks_aghq <- absmax(aghq_ecdf_df$ecdf_diff)
+
+ecdf_df <- bind_rows(tmb_ecdf_df, aghq_ecdf_df, tmbstan_ecdf_df)
+
+ecdf_df$method <- factor(ecdf_df$method, levels = c("TMB", "aghq", "tmbstan"))
+
+ks_labeller <- function(x) toString(round(abs(x), 2))
+
+ecdf_diff <- ggplot(ecdf_df, aes(x = x, y = ecdf_diff, col = method)) +
+  geom_line() +
+  geom_abline(intercept = ks_tmb, slope = 0, col = colours[1], linetype = "dashed", alpha = 0.8) +
+  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_tmb, label = ks_labeller(ks_tmb), col = colours[1], alpha = 0.8) +
+  geom_abline(intercept = ks_aghq, slope = 0, col = colours[2], linetype = "dashed", alpha = 0.8) +
+  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_aghq, label = ks_labeller(ks_aghq), col = colours[2], alpha = 0.8) +
+  scale_color_manual(values = colours) +
+  labs(x = "beta_alpha", y = "ECDF difference") +
+  guides(col = "none") +
+  coord_cartesian(xlim = c(min(ecdf_df$x), max(ecdf_df$x)), clip = "off") +
+  theme_minimal() +
+  theme(plot.margin = unit(c(1, 3, 1, 1), "lines"))
+
+histogram + ecdf_diff
+```
+
+```{r}
 ks_summary <- readRDS("depends/ks_summary.rds")

### 2023-05-24 src/naomi-simple_ks/ks.Rmd
Example KS plot into manuscript

@@
 ```{r}
-pars_eval <- pars %in% c("beta_rho", "beta_alpha", "beta_anc_rho", "u_rho_x")
+pars_eval <- pars %in% c("beta_rho", "u_rho_x")
 names(pars_eval) <- pars
@@
 ```{r eval = pars_eval[["beta_alpha"]]}
-beta_alpha <- histogram_and_ecdf("beta_alpha", return_df = TRUE)
-beta_alpha$plot
-
-#' Save off KS results as example
-saveRDS(
-  beta_alpha$df_compare,
-  file = "beta_alpha.rds"
-)
+histogram_and_ecdf_helper("beta_alpha")
 ```
@@
 ```{r eval = pars_eval[["beta_anc_rho"]]}
-beta_anc_rho <- histogram_and_ecdf("beta_anc_rho", i = 1, return_df = TRUE)
-beta_anc_rho$plot
-
-#' Save off KS results as example
-saveRDS(
-  beta_anc_rho$df_compare,
-  file = "beta_anc_rho.rds"
-)
+histogram_and_ecdf_helper("beta_anc_rho")
 ```

### 2023-05-24 src/naomi-simple_ks/script.R

@@ -4,8 +4,8 @@
 
 rmarkdown::render("ks.Rmd")
 
-beta_rho <- histogram_and_ecdf("beta_rho", i = 1, return_df = TRUE)
-saveRDS(beta_rho$df, "beta_rho.rds")
+beta_alpha <- histogram_and_ecdf("beta_alpha", return_df = TRUE)
+saveRDS(beta_alpha$df, "beta_alpha.rds")
 
 beta_anc_rho <- histogram_and_ecdf("beta_anc_rho", return_df = TRUE)
 saveRDS(beta_anc_rho$df, "beta_anc_rho.rds")

### 2023-05-25 src/docs_paper/paper.Rmd
Add plot of nodes overlaid on NUTS marginals for hypers

@@
 ```{r echo = FALSE}
+colours <- c("#56B4E9", "#009E73", "#E69F00")
+
 options(scipen = 100)
@@
 plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
-  geom_contour(col = multi.utils::cbpalette()[1]) +
+  geom_contour(col = "lightgrey") +
   coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
@@
       alpha = 0.8,
-      col = multi.utils::cbpalette()[2],
+      col = "#009E73",
       inherit.aes = FALSE
@@
 Figure showing approximation to precision matrix.
-Figure showing positions of nodes on hyperparameter marginal posteriors as compared with NUTS.
+
+Figure \ref{fig:node-posiitons} shows the positions of the generated PCA-AGHQ nodes overlaid onto the hyperparameter marginal posteriors obtained NUTS.
+
+```{r node-positions, fig.cap="PCA-AGHQ node positions (green, rug plot) overlaid onto the hyperparameter marginal posteriors (grey, histogram) for each of the 24 hyperparameters."}
+knitr::include_graphics("depends/nodes-samples-comparison.png")
+```
 
@@
 
-colours <- c("#56B4E9", "#009E73", "#E69F00")
-
 mean <- df_compare %>%
@@
 
-histogram <- ggplot(df_compare, aes(x = samples, fill = method, col = method)) +
+histogram <- df_compare %>%
+  mutate(method = fct_recode(method, "PCA-AGHQ" = "aghq", "NUTS"= "tmbstan")) %>%
+  ggplot(aes(x = samples, fill = method, col = method)) +
   geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +

### 2023-06-02 src/docs_bioinference-poster/bioinference-poster.Rmd
Poster update #36

@@
 
-* Approximate Bayesian inference method using Laplace approximation, adaptive Gauss-Hermite quadrature and principal component analysis
+* Develop an approximate Bayesian inference method using Laplace approximation, adaptive Gauss-Hermite quadrature and principal component analysis
 * Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
@@
 * It would take days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no], and this is not practical in this setting
-* Motivates looking for a fast, approximate approach, that takes uncertainty in hyperparameters into account
+* Motivates our work, looking for a fast, approximate approach, that properly takes uncertainty in hyperparameters into account
 
@@
 * In an LGM the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
-  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs, because ELGM precision matrices are not as sparse
 * Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$ and $\mathcal{J_i}$ is some set of indices
@@
 * We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
-* Though we focus on Naomi here, the HIV Inference Group (`hiv-inference.org`) works on many other complex models, challenging for existing Bayesian inference methods, which require flexible modelling tools
+* Though we focus on Naomi, the HIV Inference Group (`hiv-inference.org`) works on many other complex models, challenging for existing Bayesian inference methods, which require flexible modelling tools
 
@@
 $$
-where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ with $m = \dim(\theta)$ and $k$ points per dimension is adapted based upon
+where the Gauss-Hermite quadrature rule $\{z \in \mathcal{Q}(m, k), \omega\}$ with $m = \dim(\theta)$ and $k$ points per dimension is adapted based upon
   * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$
   * A matrix decomposition $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$
-* Keep first $s < m$ **principal components** to get size $k^s$ grid
+* Use the spectral decomposition $L = E\Lambda^{1/2}$ and keep only the first $s < m$ **principal components**
 
 $$\\[0.1in]$$
+
+```{r aghq, echo=FALSE, fig.cap='Demonstration of PCA-AGHQ.'}
+mu <- c(1, 1.5)
+cov <- matrix(c(2, 1, 1, 1), ncol = 2)
+
+obj <- function(theta) {
+  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
+}
+
+grid <- expand.grid(
+  theta1 = seq(-2, 5, length.out = 700),
+  theta2 = seq(-2, 5, length.out = 700)
+)
+
+ground_truth <- cbind(grid, pdf = obj(grid))
+
+plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = "lightgrey") +
+  coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
+  labs(x = "", y = "") +
+  theme_minimal() +
+  guides(size = FALSE) +
+  theme(
+    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
+    axis.text.y = element_blank(), axis.ticks.y = element_blank()
+  )
+
+gg <- mvQuad::createNIGrid(2, "GHe", 3)
+
+add_points <- function(plot0, gg) {
+  plot0 +
+    geom_point(
+      data = mvQuad::getNodes(gg) %>%
+              as.data.frame() %>%
+              mutate(weights = mvQuad::getWeights(gg)),
+      aes(x = V1, y = V2, size = weights),
+      alpha = 0.8,
+      col = "#009E73",
+      inherit.aes = FALSE
+    ) +
+    scale_size_continuous(range = c(1, 2))
+}
+
+plot1 <- add_points(plot0, gg) +
+  labs(size = "")
+
+#' Adapt by the mean
+gg2 <- gg
+mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2)
+
+plot2 <- add_points(plot0, gg2) +
+  labs(size = "")
+
+#' Adapt by the spectral
+gg3 <- gg
+mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1)
+
+plot3 <- add_points(plot0, gg3) +
+  labs(size = "")
+
+#' PCA-AGHQ
+gg4 <- mvQuad::createNIGrid(2, "GHe", level = c(3, 1))
+mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)
+
+plot4 <- add_points(plot0, gg4) +
+  labs(size = "")
+
+(plot1 + plot2) / (plot3 + plot4)
+```
 
@@
 
-* Relatively small country but still a large model: latent field $\text{dim}(x) = 491$, hyperparameters $\text{dim}(\theta) = 24$
+* Small country but still has latent field $\text{dim}(x) = 491$ and hyperparameters $\text{dim}(\theta) = 24$
 * Fit three inference methods (using one C++ template):
-  * TMB (`r round(time_taken$TMB, 0)` mins)
-  * PCA-AGHQ (`r round(time_taken$aghq, 0)` mins): $k = 3, s = 8$
+  * TMB (`r round(time_taken$TMB, 0)` secs)
+  * PCA-AGHQ (`r round(time_taken$aghq, 0)` hour): $k = 3, s = 8$
   * NUTS (`r round(time_taken$tmbstan, 1)` days): 4 chains of 100,000 thinned by 40 (required for good diagnostics)

### 2023-06-02 src/docs_paper/appendix.Rmd
Add MCMC convergence asssessment to appendix #48

@@
 
-# Simplified Naomi model description \label{sec:math}
+# Simplified Naomi \label{sec:math}
 
@@
 
-\newpage
+## Implementation
 
-# \textsc{C++} `TMB` user template
+The C++ `TMB` code for the negative log-posterior of the simplified Naomi model is available from the GitHub repository `athowes/elgm-inf`.
+For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used in Section \ref{sec:math} and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
+For further reference on the `TMB` software see @tmbdocumentation.
 
@@
 
-This section contains C++ `TMB` code for the negative log-posterior of the simplified Naomi model
-For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used in Section \ref{sec:math} and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
-For further reference on the `TMB` software see `https://kaskr.github.io/adcomp/_book`, or on the method see @kristensen2016tmb.
+<!-- \newpage -->
 
-\newpage
-
-```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
-```
+<!-- ```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")} -->
+<!-- ``` -->
 
@@
 We assessed MCMC convergence and suitability using a range of graphical and numerical tests.
-These included the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
+All potential scale reduction factor $\hat R$ statistics [@vehtari2021rank] were below 1.05 and therefore acceptable (Figure \ref{fig:rhat}).
+However, even thinning by a factor of 20, samples were not obtained very efficiently, resulting in the majority of effective sample size (ESS) ratios being below 0.5, with some as low as 0.1 (Figure \ref{fig:ratio}).
+As a result, the number of obtained ESS varied substantially by parameter (Figure \ref{fig:ess}).
+There were no divergent transitions.
+Could add energy plot from @betancourt2017conceptual here.
 
-For the time being, this analysis is available from `athowes.github.io/elgm-inf/mcmc-convergence`.
-Once the MCMC results are finalised, the analysis will be moved to this appendix and expanded upon.
-The following draft text and references may be useful in that expanded write-up.
+```{r}
+np <- readRDS("depends/nuts-params.rds")
 
-- Improved $\hat R$ statistic from @vehtari2021rank. Recommended only to use the sample if the value is less than 1.05.
-- ESS and autocorrelation related to efficiency
-- Traceplots helpful for diagnosis of problems
-- Pairs density plots helpful for understanding relationships between parameters, including possible identifiability issues
-- Energy plot from @betancourt2017conceptual
+n_divergent <- np %>%
+  filter(Parameter == "divergent__") %>%
+  summarise(n_divergent = sum(Value)) %>%
+  pull(n_divergent)
+
+stopifnot(n_divergent == 0)
+```
+
+```{r rhat, fig.cap="The potential scale reduction factor compared between- and within- estimates of univariate parameters.
+It is recommended against to use the results if the value is less than 1.05."}
+knitr::include_graphics("depends/rhat.png")
+```
+
+```{r ratio, fig.cap="The ESS ratio measures how efficient sampling is."}
+knitr::include_graphics("depends/ratio.png")
+```
+
+```{r ess, fig.cap="The effective number of samples varied substantially."}
+knitr::include_graphics("depends/ess.png")
+```
 
@@
 
-1. Calculate the mode, Hessian at the mode, and lower Cholesky
+In earlier work, we used a full Laplace approximation to the latent field marginals.
+The algorithm for obtaining this approximation is detailed here.
+
+1. Calculate the mode, Hessian at the mode, lower Cholesky, and Laplace approximation
 \begin{align*}
@@
 \hat{\mathbf{H}} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
-\hat{\mathbf{H}}^{-1} &= \hat{\mathbf{L}} \hat{\mathbf{L}}^\top,
+\hat{\mathbf{H}}^{-1} &= \hat{\mathbf{L}} \hat{\mathbf{L}}^\top, \\
+\tilde p_\texttt{LA}(\btheta, \y) &= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
 \end{align*}
-of the Laplace approximation
-$$
-\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
-$$
-where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\mathbf{H}}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
 \begin{align*}
@@
 Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
-If possible $k \geq 3$ is preferred, though the number of grid points scales exponentially with choice of $k$.
 Then, use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
@@
 \end{equation}
-4.
+3.
 For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\hat{\mathbf{H}}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
-A value of $l \geq 4$ is recommended to enable B-spline interpolation.
+$l \geq 4$ is recommended to enable B-spline interpolation.
 Then, for $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
@@
 where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
-6. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
+4. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
 \begin{equation}
@@
 
-7.
+5.
 Given $\{x_i(\bv), \tilde p_\texttt{AQ}(x_i(\bv) \, | \, \y)\}_{\bv \in \mathcal{Q}(1, l)}$ create a spline interpolant to each posterior marginal on the log-scale.
@@
 
-```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."}
-knitr::include_graphics("depends/algorithm-flowchart.pdf")
-```
+<!-- ```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."} -->
+<!-- knitr::include_graphics("depends/algorithm-flowchart.pdf") -->
+<!-- ``` -->
 

### 2023-06-02 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Small edits to MCMC convergence
Add MCMC convergence asssessment to appendix #48

@@
 
-This MCMC took `r out$time` to run
+This MCMC took `r round(out$time, 3)` days to run
 
 ```{r}
+cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
 bayesplot::color_scheme_set("viridisA")
@@
 rhats <- bayesplot::rhat(mcmc)
-bayesplot::mcmc_rhat(rhats)
+
+rhat_plot <- mcmc_rhat_data(rhats) %>%
+  ggplot(mapping = aes_(x = ~value, y = ~parameter, col = ~description)) +
+  geom_segment(mapping = aes_(yend = ~parameter, xend = ifelse(min(data$value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7) +
+  scale_color_manual(values = "#E69F00") +
+  geom_vline(xintercept = 1.05, linetype = "dashed", col = "grey40") +
+  labs(x = "Potential scale reduction factor", y = "NUTS parameter", col = "") +
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    panel.grid.major = element_blank()
+  )
+
+rhat_plot
+
+pdf("rhat.pdf", h = 3, w = 6.25)
+
+rhat_plot
+
+dev.off()
+
 (big_rhats <- rhats[rhats > 1.1])
 length(big_rhats) / length(rhats)
+
 ```
@@
 ratios <- bayesplot::neff_ratio(mcmc)
-bayesplot::mcmc_neff(ratios)
+
+breaks <- c(0, 0.1, 0.25, 0.5, 0.75, 1)
+
+ratio_plot <- mcmc_neff_data(ratios) %>%
+  ggplot(mapping = aes_(x = ~value, y = ~parameter, color = ~description)) +
+  geom_segment(aes_(yend = ~parameter, xend = -Inf), na.rm = TRUE, alpha = 0.7) +
+  scale_color_manual(values = c("#56B4E9", "#009E73", "#E69F00")) +
+  geom_vline(xintercept = 0.1, linetype = "dashed", col = "grey40") +
+  geom_vline(xintercept = 0.5, linetype = "dashed", col = "grey40") +
+  geom_vline(xintercept = 1, linetype = "dashed", col = "grey40") +
+  labs(x = "ESS ratio", y = "NUTS parameter", col = "") +
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    panel.grid.major = element_blank()
+  )
+
+ratio_plot
+
+pdf("ratio.pdf", h = 3, w = 6.25)
+
+ratio_plot
+
+dev.off()
 ```
@@
 
-data.frame(mcmc_summary) %>%
+ess_plot <- data.frame(mcmc_summary) %>%
   tibble::rownames_to_column("param") %>%
@@
     labs(x = "ESS", y = "Count")
+
+ess_plot
+
+pdf("ess.pdf", h = 3, w = 6.25)
+
+ess_plot
+
+dev.off()
 ```
@@
 np <- bayesplot::nuts_params(mcmc)
+
+saveRDS(np, "nuts-params.rds")
 ```
@@
 ```
+
+# Original computing environment {-}
+
+```{r}
+sessionInfo()
+```
+
+# Bibliography {-}

### 2023-06-02 src/docs_paper/paper.Rmd
Keep writing paper

@@
 Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
-The model is provided as a tool for countries to input their data to and generate estimates using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package.
-We propose a new inference method extending adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, thereby enabling fast and accurate inference for Naomi and other extended latent Gaussian models.
+The model is provided as a tool for countries to input their data to and generate estimates, during a yearly process supported by UNAIDS.
+Currently, inference is conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
+We propose a new inference method which extends adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, enabling fast and accurate inference for Naomi and other extended latent Gaussian models.
 Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is substantially faster to run to Hamiltonian Monte Carlo with the No-U-Turn sampler.
@@
 
-The complexity of the model, in combination with practical requirements for its operation, present a difficult Bayesian inference problem.
-First, due to dependence of observations on multiple structured additive predictors, Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast].
+The complexity of Naomi, in combination with practical requirements, present a difficult Bayesian inference problem.
+First, due to dependence of observations on multiple structured additive predictors, Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
 Further, as well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number typically required for use of integrated nested Laplace approximations [@rue2009approximate].
 Second, any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production across a range of countries.
-In this setting, Markov chain Monte Carlo (MCMC) approaches are prohibitively slow, both due to the scale of the model and challenging features of its posterior geometry [@neal2003slice].
+In this setting, Markov chain Monte Carlo (MCMC) approaches are prohibitively slow due to the scale of the model and challenging features of its posterior geometry [@neal2003slice].
 
 Inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgood2022statistical].
-Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
+Owing to its speed and flexibility, `TMB` is gaining popularity more broadly in spatial statistics [@osgood2022statistical].
+Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the parameters.
 In the Naomi model, this subset is the high-dimensional latent field, leaving the smaller number of hyperparameters.
-Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
-Although this approach is fast, it has the downside that within the empirical Bayes framework hyperparameter uncertainty is not properly accounted for in the latent field posterior.
-This has motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
+`TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation, Taking tnspiration from the AD Model Builder package [@fournier2012ad].
+Although the approach of `TMB` is fast, within the empirical Bayes framework hyperparameter uncertainty is not properly accounted for in the latent field posterior.
+This consideration motivates us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
 
-To obtain fast, accurate Bayesian inferences for the Naomi model we develop an inference methodology which extends adaptive Gauss-Hermite quadrature to handle a higher number of hyperparameters. 
-We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
-As `aghq` is designed to naturally interface with `TMB`, use of our method is simple when provided a \textsf{C++} user template for the log-posterior.
+To obtain fast, accurate Bayesian inferences for the Naomi model we developed an inference method which extends adaptive Gauss-Hermite quadrature to handle many hyperparameters.
+We used principal components analysis to create a quadrature grid which is computationally feasible.
+Our method is implemented as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
+As `aghq` is designed to naturally interface with `TMB`, use is simple when provided a \textsf{C++} user template for the log-posterior.
 
@@
 
-# A simplified Naomi model\label{sec:naomi}
+# Simplified Naomi model\label{sec:naomi}
 
-@eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points $\{T_i\}_{i = 1, 2, 3}$.
-We consider a simplified version defined only at the time of the most recent household survey with HIV testing ($T_1$), omitting nowcasting ($T_2$) and temporal projection ($T_3$) which involve limited inferences.
-An overview of the simplified model is given below, and a more complete mathematical description (Appendix S1) as well as a \textsf{C++} template for the log-posterior (Appendix S2) are provided in the supplementary material.
+@eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points.
+We consider a simplified version defined only at the time of the most recent household survey with HIV testing, omitting nowcasting and temporal projection, as these time points involve limited inferences.
+An overview of the simplified model is given below, and a more complete mathematical description is provided in Appendix S1.
 
@@
 
-Consider a country in sub-Saharan Africa where a household survey with complex design has taken place at time $T_1$.
+Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place.
 Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ index five-year age group, and $s \in \mathcal{S}$ index sex.
@@
 Let $N_i \in \mathbb{N}$ be the known, fixed population size.
-We infer the following unknown HIV indicators using coupled regression equations: HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
+We infer the following unknown HIV indicators using linked regression equations: HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
 Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that $\text{logit}(\rho_i) = \eta^\rho_i$ and $\text{logit}(\alpha_i) = \eta^\alpha_i$, for certain choice of structured additive predictors.
@@
 These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
-In particular, the number of ANC clients with ascertained status $x^\text{ANC}_I$, the number of those with positive status $y^\text{ANC}_I$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_I$.
-We use the binomial working likelihoods
+In particular
+
+1. $x^\text{ANC}_I$, the number of ANC clients with ascertained status,
+2. $y^\text{ANC}_I$, the number of those with positive status, and
+3. $z^\text{ANC}_I$, the number of those already on ART prior to their first ANC visit.
+
+We use the binomial likelihoods
 \begin{align*}
@@
 The log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ are modelled using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
-Aggregate ART attendance data $\dot A_I$ is modelled using a Gaussian approximation to a sum of binomials.
-The sum results both by aggregation over $i \in I$ and by number of ART clients travelling from district $x'$ to $x$.
+We then model aggregate ART attendance data $\dot A_I$ using a Gaussian approximation to a sum of binomials.
+This sum is both by over $i \in I$ and by number of ART clients travelling from district $x'$ to $x$.
 More details regarding this part of the model are provided in Appendix S1.
@@
 Attempt to write fully the Naomi model in a small number of collected together equations here.
+The ANC testing component and ART attendance component might be hard to do this for.
 
@@
 
-Latent Gaussian models (LGMs) [@rue2009approximate] are three-stage hierarchical models of the form
+Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models of the form
 \begin{align*}
@@
 
-Extended latent Gaussian models (ELGMs) [@stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
+Extended latent Gaussian models [ELGMs; @stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
 Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
@@
 
+If Naomi was written out in full mathematical detail, I might be able to make these reasons more concrete.
+It would also be good to specify how much of an impact each of these reasons is likely to have on the difficulty of inference.
+
 # Fast approximate inference method\label{sec:fastinferencemethods}
 
-The joint posterior of $(\x, \btheta)$ for an ELGM is given by
+The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ for an ELGM is given by
 \begin{equation*}
@@
 \begin{align}
-  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
-  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
+  p(x_i \, | \, \y) &\approx \tilde p(x_i \, | \, \y) = \int \tilde p(x_i \, | \, \btheta, \y) \tilde p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
+  p(\theta_j \, | \, \y) &\approx \tilde p(\theta_j \, | \, \y) = \int \tilde  p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
 \end{align}
@@
 Latent field joint and marginal inferences are then direct from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
+Hyperparameter inferences are obtained according to method which should be specified here.
 
@@
 
-Quadrature rules can be used to approximate the integral of $\tilde p_\texttt{LA}(\btheta, \y)$ via the weighted sum
+Quadrature rules can be used to approximate the integral of $\tilde p_\texttt{LA}(\btheta, \y)$ via a weighted sum
 \begin{equation}
@@
 Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule where, in the univariate case, the nodes $\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}$ are selected as zeros of the $k$th Hermite polynomial.
-The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by
-\begin{equation*}
-\omega(z) = \frac{k!}{[H_{k + 1}(z)]^2 \phi(z)},
-\end{equation*}
-where $\phi(\cdot)$ is a standard univariate Gaussian density.
+The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, where $\phi(\cdot)$ is a standard univariate Gaussian density.
+GHQ is attractive because it is exact for functions which are a polynomial of total order no more than $2k - 1$ multiplied by a Gaussian density.
 Multivariate GHQ rules are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
-GHQ is attractive because it is exact for functions which are a polynomial of total order no more than $2k - 1$ multiplied by a Gaussian density.
 
@@
 In adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] the nodes are shifted and rotated to suit the particular integrand.
-Repositioning the nodes is particularly important for statistical quadrature problems, where the integral depends on data $\y$ and so regions of high density are not known in advance.
+Repositioning the nodes is especially important in statistical quadrature problems, where the integral depends on data $\y$ such that regions of high density cannot reasonably be known in advance.
 To obtain an AGHQ estimate of Equation \ref{eq:quad}, let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ a matrix decomposition of the inverse curvature, then
@@
 \end{equation}
-Two possibilities for the matrix decomposition are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$ and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$ [@jackel2005note].
-Equation \ref{eq:aghq} may be used to normalise the Laplace approximation
+That is, the unadapted nodes have been shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
+Two options for the particular matrix decomposition are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$ and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$ [@jackel2005note].
+Figure \ref{fig:aghq} demonstrates GHQ, as well as adaption for these two choices of matrix decomposition.
+Equation \ref{eq:aghq} may then be used to normalise the Laplace approximation
 \begin{equation*}
@@
 
-For the latent field, the adapted nodes and weights may be reused [@rue2009approximate; @stringer2022fast] in the nested approximation
+To obtain inferences for the latent field (Equation \ref{eq:inla1}) the adapted nodes and weights are reused [@rue2009approximate; @stringer2022fast]
 \begin{equation}
@@
 \end{equation}
-Samples from this mixture of Gaussians are obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
-
-## Principal components analysis
-
-Use of the product rule requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points, which becomes intractable as $m$ increases for $k > 1$.
-An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$, and define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
-Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
-This choice of nodes, taken together with use of the spectral decomposition, is analogous to a principal components analysis (PCA) approach to AGHQ
-\begin{equation}
-\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
-\end{equation}
-where $\hat{\mathbf{E}}_{\texttt{LA}, s} = \cdots$ and $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s} = \cdots$.
-
-## Sparse rules
-
-Though there are sparse rules which retain the attractive exactness properties of GHQ, with fewer than $k^m$ quadrature points, they use weighting functions which may be negative $\omega(\z) < 0$.
-This introduces a problem as we are not aware of any suitable approaches to obtain multinomial samples when $\lambda(\z)$ is negative.
+Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
 
@@
 
+## Principal components analysis
+
+Use of the product rule requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points.
+This quickly becomes intractable as $m$ increases for $k > 1$.
+An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
+We may then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
+Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
+Taken together with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ
+\begin{equation}
+\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
+\end{equation}
+where $\hat{\mathbf{E}}_{\texttt{LA}, s} = \cdots$ and $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s} = \cdots$ are terms which should be described.
+The final panel of Figure \ref{fig:aghq} illustrates a case where $m = 2$ and $s = 1$ such that 
+We refer to use of this quadrature rule as PCA-AGHQ.
+
+## Sparse rules
+
+Though there are sparse rules which retain the attractive exactness properties of GHQ, using fewer than $k^m$ quadrature points, they use weighting functions which may be negative $\omega(\z) < 0$.
+This introduces a problem when trying to produce inferences about the latent field, as the resulting $\lambda(\z)$ may be negative.
+We are not aware of any suitable approaches to obtain multinomial samples in such cases.
+
 # Application to data from Malawi\label{sec:results}
@@
 We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
-For each approach, the `TMB` C++ user-template (Appendix S2) used to specify the log-posterior was the same.
+For each approach, the `TMB` C++ user-template used to specify the log-posterior was the same.
 The three approaches were:
@@
 TMB & \texttt{TMB} & $1000$ samples \\
-PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$, $1000$ samples \\
+PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-settings}), $1000$ samples \\
 NUTS & \texttt{tmbstan} & $4$ chains of $20000$ iterations, with the first $10000$ iterations of each chain discarded as warmup, thinned by a factor of $20$.
@@
 These include the potential scale reduction factors $\hat R$ [@gelman1992inference; @vehtari2021rank] for each parameter, bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments [@betancourt2017conceptual].
-For full details see Appendix S3.
+For full details see Appendix S2.
 We treat the NUTS results as a gold-standard.
 
-## PCA-AGHQ settings
+## PCA-AGHQ settings \label{sec:pca-settings}
 
@@
 
-Figure \ref{fig:node-posiitons} shows the positions of the generated PCA-AGHQ nodes overlaid onto the hyperparameter marginal posteriors obtained NUTS.
+Figure \ref{fig:node-positions} shows the positions of the generated PCA-AGHQ nodes overlaid onto the hyperparameter marginal posteriors obtained NUTS.
+For around 12 of the 24 hyperparameters, the marginal was well covered by the nodes.
+Although this is an improvement on the $s = 8$ that could be covered naively, there remain many hyperparameters uncovered.
+Furthermore, it is likely that the hyperparameters with skewed marginal posteriors are more important to cover well.
+We have not yet found a way to take this into account.
 
@@
 ```
+
+Demonstrate that the 12 hyperparameters which are well covered are the ones that you would expect given their marginal standard deviations.
+
+One crude approach for assessing appropriateness of the grid is to look at the estimate of $\log p_\texttt{PCA}(\y)$ for a range of settings.
+Convergence in $\log p_\texttt{PCA}(\y)$ as $s$ and $k$ are increased may suggest sufficiency has been reached.
+Figure shows those values which were computable in a reasonable time.
 
@@
 ```{r psis, fig.cap="Results of Pareto-smoothed importance sampling analysis."}
-qplot(1:3, 1:3) + theme_minimal()
 ```
@@
 ```{r mmd, fig.cap="Results of maximum mean discrepancy analysis."}
-qplot(1:3, 1:3) + theme_minimal()
 ```

### 2023-06-05 src/docs_bioinference-poster/bioinference-poster.Rmd
Just need to add second 90 plot then finished with BioInference poster #36

@@
     highlight: espresso
-    number_sections: false
 ---
@@
 
-* Develop an approximate Bayesian inference method using Laplace approximation, adaptive Gauss-Hermite quadrature and principal component analysis
+* We developed an approximate Bayesian inference method using Laplace approximation, adaptive Gauss-Hermite quadrature and principal component analysis
 * Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
-* Implemented as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a `TMB` C++ user template
+* Implemented as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a Template Model Builder `TMB` [@kristensen2016tmb] C++ user template
 
-## The Naomi HIV model
+## 1. The Naomi HIV model
 
-* District-level model of HIV indicators [@eaton2021naomi] which synthesises data from household surveys, antenatal care (ANC) clinics, and routine service provision of antiretroviral therapy (ART)
+* District-level model of HIV indicators [@eaton2021naomi] which synthesises data from 1) household surveys, 2) antenatal care (ANC) clinics, and 3) routine service provision of antiretroviral therapy (ART)
   * Combining evidence from multiple data sources helps overcome the limitations of any one
@@
   * Figure \@ref(fig:fig1) illustrates the seven stages of using the app
-* Inference conducted in minutes using empirical Bayes and a Gaussian approximation via Template Model Builder `TMB` [@kristensen2016tmb]
+* Inference conducted in minutes using empirical Bayes and a Gaussian approximation
 * It would take days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no], and this is not practical in this setting
-* Motivates our work, looking for a fast, approximate approach, that properly takes uncertainty in hyperparameters into account
+* We are looking for a fast, approximate approach, that properly takes uncertainty in hyperparameters into account
 
@@
 
-## Extended latent Gaussian models
+## 2. Extended latent Gaussian models
 
@@
 
-## Inference procedure
+## 3. Inference procedure
 
-* **Laplace approximation** Integrate out variables using a Gaussian approximation to the denominator
+* **Laplace approximation** Integrate out latent field using a Gaussian approximation to the denominator
 $$
-p(\theta, y) \approx \tilde p_\texttt{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\texttt{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
+p(\theta, y) \approx \tilde p_\texttt{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\texttt{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)},
 $$
-where $\tilde p_\texttt{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
+where $\tilde p_\texttt{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \hat H(\theta)^{-1})$
   * Use automatic differentiation via `CppAD` in `TMB`
-* **Adaptive Gauss-Hermite Quadrature**
+
+```{r aghq, echo=FALSE, fig.align='center', fig.cap='Demonstration of PCA-AGHQ.'}
+knitr::include_graphics("fig2.png")
+```
+
+* **Adaptive Gauss-Hermite Quadrature** (AGHQ) perform quadrature over the hyperparameters
 $$
-\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)
+\int_\Theta p_\texttt{LA}(\theta, y) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p_\texttt{LA}(\hat \theta + Lz, y) \omega(z),
 $$
 where the Gauss-Hermite quadrature rule $\{z \in \mathcal{Q}(m, k), \omega\}$ with $m = \dim(\theta)$ and $k$ points per dimension is adapted based upon
-  * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$
-  * A matrix decomposition $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$
-* Use the spectral decomposition $L = E\Lambda^{1/2}$ and keep only the first $s < m$ **principal components**
+  * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p_\texttt{LA}(\theta, y)$
+  * A matrix decomposition $LL^\top = - \partial_\theta^2 \log p_\texttt{LA}(\theta, y) \rvert_{\theta = \hat \theta}$
+* Use the spectral decomposition $L = E\Lambda^{1/2}$ and keep only the first $s < m$ **principal components** (PCA-AGHQ)
 
-$$\\[0.1in]$$
+## 4. Application to Malawi
 
-```{r aghq, echo=FALSE, fig.cap='Demonstration of PCA-AGHQ.'}
-mu <- c(1, 1.5)
-cov <- matrix(c(2, 1, 1, 1), ncol = 2)
-
-obj <- function(theta) {
-  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
-}
-
-grid <- expand.grid(
-  theta1 = seq(-2, 5, length.out = 700),
-  theta2 = seq(-2, 5, length.out = 700)
-)
-
-ground_truth <- cbind(grid, pdf = obj(grid))
-
-plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
-  geom_contour(col = "lightgrey") +
-  coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
-  labs(x = "", y = "") +
-  theme_minimal() +
-  guides(size = FALSE) +
-  theme(
-    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
-    axis.text.y = element_blank(), axis.ticks.y = element_blank()
-  )
-
-gg <- mvQuad::createNIGrid(2, "GHe", 3)
-
-add_points <- function(plot0, gg) {
-  plot0 +
-    geom_point(
-      data = mvQuad::getNodes(gg) %>%
-              as.data.frame() %>%
-              mutate(weights = mvQuad::getWeights(gg)),
-      aes(x = V1, y = V2, size = weights),
-      alpha = 0.8,
-      col = "#009E73",
-      inherit.aes = FALSE
-    ) +
-    scale_size_continuous(range = c(1, 2))
-}
-
-plot1 <- add_points(plot0, gg) +
-  labs(size = "")
-
-#' Adapt by the mean
-gg2 <- gg
-mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2)
-
-plot2 <- add_points(plot0, gg2) +
-  labs(size = "")
-
-#' Adapt by the spectral
-gg3 <- gg
-mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1)
-
-plot3 <- add_points(plot0, gg3) +
-  labs(size = "")
-
-#' PCA-AGHQ
-gg4 <- mvQuad::createNIGrid(2, "GHe", level = c(3, 1))
-mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)
-
-plot4 <- add_points(plot0, gg4) +
-  labs(size = "")
-
-(plot1 + plot2) / (plot3 + plot4)
+```{r fig3, echo=FALSE, fig.align='center', fig.cap='District-level model outputs for adults 15-49 in January 2016. Adapted from Eaton et al. 2021.'}
+knitr::include_graphics("fig3.png")
 ```
-
-## Application to Malawi data
 
@@
 
-```{r fig3, echo=FALSE, fig.align='center', out.width='100%', fig.cap='District-level model outputs for adults 15-49 in January 2016. Adapted from Eaton et al. 2021.'}
-knitr::include_graphics("fig3.png")
+* Malawi is a relatively small country but still has latent field $\text{dim}(x) = 491$ and hyperparameters $\text{dim}(\theta) = 24$
+
+| Method | Description | Time |
+|-----------|----------|------|
+| TMB | Baseline | `r round(time_taken$TMB, 0)` secs |
+| PCA-AGHQ | Ours | `r round(time_taken$aghq, 0)` hour |
+| NUTS | Gold-standard | `r round(time_taken$tmbstan, 1)` days |
+
+* For PCA-AGHQ $k = 3$ and $s = 8$ chosen using Scree plot to explain ~90% of variance
+* For NUTS 4 chains of 100,000 thinned by 40 were required for good diagnostics
+* Kolmogorov-Smirnov (KS) test based on the maximum difference between marginal ECDFs
+  * Reduces average KS distance from 0.08 to 0.075
+* Also considering joints via Pareto-smoothed importance sampling and maximum mean discrepancy
+* Naomi can be used to assess probabilites targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90")
+  * Reduces RMSE for estimating second 90 exceedance probabilities by 10%
+
+```{r}
+# knitr::include_graphics("fig4.png")
+# Figure of Second 90 here
 ```
 
-* Small country but still has latent field $\text{dim}(x) = 491$ and hyperparameters $\text{dim}(\theta) = 24$
-* Fit three inference methods (using one C++ template):
-  * TMB (`r round(time_taken$TMB, 0)` secs)
-  * PCA-AGHQ (`r round(time_taken$aghq, 0)` hour): $k = 3, s = 8$
-  * NUTS (`r round(time_taken$tmbstan, 1)` days): 4 chains of 100,000 thinned by 40 (required for good diagnostics)
-* Figure \@ref(fig:fig3) illustrates example model outputs: HIV prevalence, ART coverage, HIV incidence, and number of new infections, at the district level
-* Compare hyperparameter, latent field, and output posteriors based on Kolmogorov-Smirnov tests, Pareto-smoothed importance sampling, and maximum mean discrepancy
+## 5. Future directions
 
-## Future directions
-
-* Laplace marginals with matrix algebra approximations [@wood2020simplified] to speed up latent field marginal calculations
-* Further methods for allocation of effort to important dimensions
+* Laplace marginals with matrix algebra approximations [@wood2020simplified] to speed up calculations
+* Further methods for allocation of effort to "important" dimensions of hyperparameter grid
 

### 2023-06-05 src/docs_bioinference-poster/bioinference-poster.Rmd
Save exceedance probability output and put into poster #36

@@
 
-$$\\[0.1in]$$
-
 ```{r fig1, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model fitting occurs interactively in stages.'}
@@
 ```
-
-$$\\[0.1in]$$
 
@@
 
+* Malawi is a relatively small country but still has latent field $\text{dim}(x) = 491$ and hyperparameters $\text{dim}(\theta) = 24$
+
 ```{r fig3, echo=FALSE, fig.align='center', fig.cap='District-level model outputs for adults 15-49 in January 2016. Adapted from Eaton et al. 2021.'}
@@
 
-* Malawi is a relatively small country but still has latent field $\text{dim}(x) = 491$ and hyperparameters $\text{dim}(\theta) = 24$
-
 | Method | Description | Time |
@@
 
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+ks <- readRDS("depends/ks_summary.rds")
+
+ks_reduction <- ks %>%
+  filter(Type == "Latent field") %>%
+  summarise(
+    ks_tmb = mean(`KS(TMB, tmbstan)`),
+    ks_aghq = mean(`KS(aghq, tmbstan)`)
+  ) %>%
+  mutate(reduction = 100 * (ks_aghq - ks_tmb) / ks_tmb) %>%
+  pull(reduction) %>%
+  round()
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+second90 <- readr::read_csv("depends/second90.csv")
+
+second90_reduction <- second90 %>%
+  summarise(
+    rmse_tmb = 100 * sqrt(mean((tmbstan - TMB)^2)),
+    rmse_aghq = 100 * sqrt(mean((tmbstan - aghq)^2))
+  ) %>%
+  mutate(reduction = 100 * (rmse_aghq - rmse_tmb) / rmse_tmb) %>%
+  pull(reduction) %>%
+  round()
+```
+
 * For PCA-AGHQ $k = 3$ and $s = 8$ chosen using Scree plot to explain ~90% of variance
@@
 * Kolmogorov-Smirnov (KS) test based on the maximum difference between marginal ECDFs
-  * Reduces average KS distance from 0.08 to 0.075
-* Also considering joints via Pareto-smoothed importance sampling and maximum mean discrepancy
-* Naomi can be used to assess probabilites targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90")
-  * Reduces RMSE for estimating second 90 exceedance probabilities by 10%
+  * Average KS distance from NUTS reduced by `r abs(ks_reduction)`%
+  * Also considering joint posteriors via Pareto-smoothed importance sampling and maximum mean discrepancy
+* Naomi can be used to assess probabilities targets have been met e.g.
+90% of those who know their HIV status are on ART ("second 90").
+Both TMB and PCA-AGHQ have biased inferences (Figure \@ref(fig:fig4))
+  * Reduced RMSE for estimating second 90 exceedance probabilities by `r abs(second90_reduction)`%
 
-```{r}
-# knitr::include_graphics("fig4.png")
-# Figure of Second 90 here
+```{r fig4, echo=FALSE, fig.align='center', fig.cap='TMB is biased at estimating second 90 exceedance probabilities for women.'}
+knitr::include_graphics("fig4.png")
 ```
@@
 
+* **Can we do any better than modest improvements?**
 * Laplace marginals with matrix algebra approximations [@wood2020simplified] to speed up calculations
@@
 
-**Interested?** Working notebooks and R code available from `github.com/athowes/elgm-inf`. Or get in touch:
+**Funding** AH was supported by the EPSRC and Bill & Melinda Gates Foundation. This research was supported by the MRC Centre for Global Infectious Disease Analysis.
 
@@
 * `r fontawesome::fa("envelope", fill = "black")` `ath19@ic.ac.uk`
-* `r fontawesome::fa("twitter", fill = "black")` `adamhowes`
-
-**Funding** AH was supported by the EPSRC and Bill & Melinda Gates Foundation. This research was supported by the MRC Centre for Global Infectious Disease Analysis.
+* `r fontawesome::fa("twitter", fill = "black")` `@adamhowes`
 
@@
 <font size="5"> <div id="refs" class="references"></div> </font>
+
+<!-- To-do: -->
+<!-- * Add sex legend to Fig 4 -->
+<!-- * Get "3. Inference procedure" to be on next page -->
+<!-- * Try to reduce whitespace around Fig 3 -->

### 2023-06-06 src/docs_bioinference-poster/bioinference-poster.Rmd
Finish BioInference poster #36

@@
 
+$$\\[0.1in]$$
+
 ## 3. Inference procedure
@@
 
-```{r fig4, echo=FALSE, fig.align='center', fig.cap='TMB is biased at estimating second 90 exceedance probabilities for women.'}
+```{r fig4, echo=FALSE, fig.align='center', fig.cap='Both approximate methods are meaningfully incorrect for policy.'}
 knitr::include_graphics("fig4.png")
@@
 
-* `r fontawesome::fa("home", fill = "black")` `athowes.github.io`
-* `r fontawesome::fa("envelope", fill = "black")` `ath19@ic.ac.uk`
-* `r fontawesome::fa("twitter", fill = "black")` `@adamhowes`
+**Contact** `r fontawesome::fa("home", fill = "black")` `athowes.github.io`; `r fontawesome::fa("envelope", fill = "black")` `ath19@ic.ac.uk`
 
@@
 <!-- To-do: -->
-<!-- * Add sex legend to Fig 4 -->
 <!-- * Get "3. Inference procedure" to be on next page -->

### 2023-06-12 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
Start "Trees, maps, and theorems" approach to slides (as much as is possible using .Rmd...) #40

--- 2023-04-25 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
@@
 ---
-title: "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
+title: "Fast approximate Bayesian inference for the Naomi model"
 author: "Adam Howes"
@@
 
-# The Naomi model
+# Doing precision public health requires granular data
 
-:::::::::::::: {.columns}
+<!-- https://github.com/athowes/elgm-inf/blob/fd4b0bb6b9ba93ac277de7aad47a94581e8fc900/src/docs_18-04-23-explainer/18-04-23-explainer.Rmd -->
 
-::: {.column width=.65}
-
-* Naomi is a complicated spatio-temporal evidence synthesis model
-* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
-* Fast inference is important to allow for interactive review and development of estimates
-* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2016tmb]
-
-:::
-
-::: {.column width=.35}
-
-```{r, echo=FALSE, fig.cap="A supermodel", out.width = '65%'}
-knitr::include_graphics("naomi_hex.png")
-```
-
-:::
-
-::::::::::::::
+1. $\color{lolit}{\text{The}}$ right interventions
+2. $\color{lolit}{\text{in the}}$ right place
+3. $\color{lolit}{\text{to the}}$ right populations
+4. $\color{lolit}{\text{at the}}$ right time
 
@@
 
-```{r, echo=FALSE, fig.cap="Example of the user interface from \\texttt{https://naomi.unaids.org/}", out.width = '70%'}
+```{r, echo=FALSE, fig.cap="Naomi is a district-level model of HIV indicators.", out.width = '90%'}
+knitr::include_graphics("depends/naomi_results.png")
+```
+
+#
+
+```{r, echo=FALSE, fig.cap="Web interface promotes data ownership, data use, and data quality. From \\texttt{https://naomi.unaids.org/.}", out.width = '70%'}
 knitr::include_graphics("naomi_user.png")
@@
 
-# Thanks for listening!
+# Better estimates by triangulating information
 
-* Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
-* Code and notebooks for this project are available at `athowes.github.io/elgm-inf`
+<!-- https://github.com/athowes/elgm-inf/blob/fd4b0bb6b9ba93ac277de7aad47a94581e8fc900/src/docs_18-04-23-explainer/18-04-23-explainer.Rmd -->
 
-\begin{center}
-\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
+1. Household surveys $\color{lolit}{\text{infreqent, but gold-standard}}$
+2. Antenatal care clinic data $\color{lolit}{\text{frequent, only for pregnant women}}$
+3. Treatment service provision data $\color{lolit}{\text{frequent, but hard to interpret}}$
 
-Let me know if you'd be up for being an early reader!
+# Put together, it's a challenging Bayesian inference problem
 
-\end{tcolorbox}
-\end{center}
+We want our inference procedure to be
+
+1. Fast $\color{lolit}{\text{enough for interactive review of estimates}}$
+2. Accurate $\color{lolit}{\text{enough for precision public health}}$
+3. Flexible $\color{lolit}{\text{enough for compatibility with a complex model}}$
+
+# The model has a big Gaussian latent field
+
+# Approximate the marginal posterior of $\x$ by a Gaussian
+
+Given hyperparameters $\btheta$ we compute this as
+
+$$
+\tilde{p}_{\texttt{G}}(\x \, | \, \btheta, \y) = \mathcal{N}(\hat \x(\btheta), \hat{\mathbf{H}} (\btheta))
+$$
+
+\lolit{
+
+If you input $\theta_1, \ldots, \theta_{24}$ then it'll return a 467 length mean vector $\x(\btheta)$ and $467 \times 467$ length covariance matrix $\hat{\mathbf{H}} (\btheta)$
+
+}
+
+# Get in touch to chat about
+
+1. Further directions for this research $\color{lolit}{\text{including my suggestions for short masters or PhD projects}}$  
+2. Impactful academic or industry jobs using Bayesian statistics $\color{lolit}{\text{to begin around the end of this year when I graduate (I hope!)}}$
+
+# For more information
+
+* Code and notebooks: `github.com/athowes/elgm-inf`
+* Working paper on the way^[For sufficiently vague definition of "on the way"]: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model $\color{lolit}{\text{Adam Howes, Alex Stringer, Seth Flaxman, Jeff Eaton}}$
+
+#
+
+```{r, echo=FALSE, fig.cap="Much of this work done in Waterloo, Canada visiting Alex Stringer last fall! Would definitely recommend the SAS department.", out.width = '60%'}
+knitr::include_graphics("uw.png")
+```
 
 # References {.allowframebreaks}
+
+<!-- # Latent Gaussian models -->
+
+<!-- * In an LGM [@rue2009approximate] the conditional mean depends on exactly one structured additive predictor -->
+<!-- \begin{align*} -->
+<!-- y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\ -->
+<!-- \mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\ -->
+<!-- \eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), -->
+<!-- \end{align*} -->
+<!-- * $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors and can be collected into the latent field $\x$ -->
+
+<!-- # Extended latent Gaussian models -->
+
+<!-- * ELGM remove this requirement such that -->
+<!-- $$ -->
+<!-- \mu_i = g_i(\eta_{\mathcal{J}_i}) -->
+<!-- $$ -->
+<!-- where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices -->
+<!-- * Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$ -->
+<!--   * $N_n < n$: more data points than structured additive predictors -->
+<!--   * $N_n = n$: as many data points as structured additive predictors (LGM case) -->
+<!--   * $N_n > n$: fewer data points than structured additive predictors -->
+<!-- * The $g_i$ allow for a higher degree of non-linearity in the model -->
+
+<!-- # Why is Naomi an ELGM? -->
+
+<!-- 1. Incidence depends on adult prevalence and coverage -->
+<!-- 2. Incidence linked non-linearly to recent infection -->
+<!-- 3. ANC offset from household survey -->
+<!-- 4. ART coverage and recent infection are products -->
+<!-- 5. Aggregation of finer processes -->
+<!-- 6. ART attendance uses a multinomial -->
+<!-- 7. Multiple link functions -->
+
+<!-- # 1. Incidence depends on adult prevalence and coverage -->
+
+<!-- * Linear predictor for incidence contains aggregated prevalence and coverage -->
+<!-- $$ -->
+<!-- \log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda.
+-->
+<!-- $$ -->
+<!-- * Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian -->
+
+<!-- # 2. Incidence and prevalence linked to recent infection -->
+
+<!-- $$ -->
+<!-- \kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right) -->
+<!-- $$ -->
+
+<!-- * Here $\Omega_T$ and $\beta_T$ are Gaussian and have strong priors depending on the particular survey -->
+
+<!-- # 3. ANC offset from household survey -->
+
+<!-- * Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators -->
+<!-- \begin{align*} -->
+<!-- \text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\ -->
+<!-- \text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}.  -->
+<!-- \end{align*} -->
+<!-- * Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$ -->
+
+<!-- # 3. ANC offset from household survey -->
+
+<!-- * Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$?] -->
+<!-- $$ -->
+<!-- \begin{pmatrix} -->
+<!-- \eta_1^\star \\ -->
+<!-- \eta_2^\star -->
+<!-- \end{pmatrix} -->
+<!-- = -->
+<!-- \begin{pmatrix} -->
+<!-- 1 & 1 \\ -->
+<!-- 0 & 1 -->
+<!-- \end{pmatrix} -->
+<!-- \begin{pmatrix} -->
+<!-- \eta_1 \\ -->
+<!-- \eta_2 -->
+<!-- \end{pmatrix} -->
+<!-- = -->
+<!-- \begin{pmatrix} -->
+<!-- \eta_1 + \eta_2 \\ -->
+<!-- \eta_2 -->
+<!-- \end{pmatrix} -->
+<!-- $$ -->
+<!-- * By having effects only apply to a subset of indices (`idx = c(NA, 1)` say) perhaps it can work -->
+
+<!-- \begin{center} -->
+<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
+
+<!-- Warning! -->
+
+<!-- \begin{enumerate} -->
+<!--     \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA} -->
+<!--     \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM -->
+<!-- \end{enumerate} -->
+
+<!-- \end{tcolorbox} -->
+<!-- \end{center} -->
+
+<!-- # 4. ART coverage and recent infection are products -->
+
+<!-- * In the household survey, say, individuals who are taking ART or have been recently infected must be HIV positive -->
+<!-- \begin{align*} -->
+<!-- y^{\hat \alpha}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \alpha_{x, s, a}), \\ -->
+<!-- y^{\hat \kappa}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \kappa_{x, s, a}). -->
+<!-- \end{align*} -->
+<!-- * $\text{logit}(\rho_{x, s, a})$ and $\text{logit}(\alpha_{x, s, a})$ are Gaussian, but we're taking a product here -->
+<!-- * $\kappa_{x, s, a}$ is more complicated: a function of incidence, prevalence, mean duration of recent infection and false recent ratio -->
+
+<!-- \begin{center} -->
+<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
+<!-- Warning! These equations as written do not appear in Naomi: instead there are aggregated versions, as we will see next slide. -->
+<!-- \end{tcolorbox} -->
+<!-- \end{center} -->
+
+<!-- # 5. Aggregation of finer processes -->
+
+<!-- * There are many instances of likelihoods being placed on aggregate quantities -->
+<!-- \begin{align*} -->
+<!-- y^{\hat \theta}_{I} &\sim \text{xBin}(m^{\hat \theta}_{I}, \theta_{I}), \\ -->
+<!-- \rho_{I} &= \frac{\sum_{i \in I} N_i \rho_i}{\sum_{i \in I} N_i}. -->
+<!-- \end{align*} -->
+<!-- * Here we have $|I|$ linear predictors being informed by one observation -->
+<!-- * Known as disaggregation regression -->
+
+<!-- # 6. ART attendance uses the multinomial -->
+
+<!-- # 7. Multiple link functions -->
+
+<!-- * The Naomi model uses both $\text{logit}$ and $\log$ (inverse) link functions (not even considering the constructed quasi-link functions) -->
+<!-- * For LGMs there is only one $g$, whereas ELGMs allow $g_i$ -->
+<!-- * In `R-INLA` it is possible for `y` to be a matrix where each column contains observations with shared likelihood family (and hyperparameters) and `family = c("family1", "family2", ...)` -->
+
+<!-- # Two deterministic methods -->
+
+<!-- * We use two deterministic^[In contrast to the most famous approximate Bayesian inference method, Markov chain Monte Carlo, which is fundamentally stochastic.] methods to approximate our integral -->
+
+<!-- 1. The Laplace approximation -->
+<!-- 2. Quadrature -->
+
+<!-- # The Laplace approximation -->
+
+<!-- Pretend $p(\vartheta \, | \, y)$ is Gaussian -->
+
+<!-- * Mode $\hat \vartheta = \argmax_\vartheta \log p(y, \vartheta)$ -->
+<!-- * Hessian $H(\hat \vartheta) = - \partial_\vartheta^2 \log p(y, \vartheta) \rvert_{\vartheta = \hat \vartheta}$ -->
+<!-- * Gaussian approximation $\implies \tilde p_{\texttt{G}}(\vartheta \, | \, y) = \mathcal{N}(\vartheta \, | \, \hat \vartheta, H(\hat \vartheta)^{-1})$ -->
+
+<!-- # -->
+
+<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="A Gamma prior with $a = 3$ and $b = 1$."} -->
+<!-- cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999") -->
+
+<!-- a <- 3 -->
+<!-- b <- 1 -->
+
+<!-- prior <- ggplot(data = data.frame(x = c(0, 10)), aes(x)) + -->
+<!--   stat_function(fun = dgamma, n = 500, args = list(shape = a, rate = b), col = cbpalette[1]) + -->
+<!--   annotate("text", x = 6, y = 0.15, label = "Gamma(3, 1)", col = cbpalette[1], size = 5) + -->
+<!--   labs(x = "", y = "") + -->
+<!--   theme_minimal() -->
+
+<!-- prior -->
+<!-- ``` -->
+
+<!-- # -->
+
+<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="Draw 3 points from $\\text{Poisson}(3)$, then compute the posterior."} -->
+<!-- set.seed(2) -->
+<!-- y <- rpois(3, lambda = 2) -->
+
+<!-- posterior <- prior + -->
+<!--   geom_point(data = data.frame(x = y, y = 0), aes(x = x, y = y), inherit.aes = FALSE, alpha = 0.7, size = 2) + -->
+<!--   stat_function(data = data.frame(x = c(0, 10)), aes(x), fun = dgamma, n = 500, args = list(shape = a + sum(y), rate = b + length(y)), col = cbpalette[2]) + -->
+<!--   annotate("text", x = 5, y = 0.25, label = "Gamma(9, 4)", col = cbpalette[2], size = 5) -->
+
+<!-- posterior -->
+<!-- ``` -->
+
+<!-- # -->
+
+<!-- ```{r message=FALSE} -->
+<!-- fn <- function(x) dgamma(x, a + sum(y), b + length(y), log = TRUE) -->
+
+<!-- # Here we are using numerical derivatives -->
+<!-- ff <- list( -->
+<!--   fn = fn, -->
+<!--   gr = function(x) numDeriv::grad(fn, x), -->
+<!--   he = function(x) numDeriv::hessian(fn, x) -->
+<!-- ) -->
+
+<!-- opt_bfgs <- aghq::optimize_theta( -->
+<!--   ff, 1, control = aghq::default_control(method = "BFGS") -->
+<!-- ) -->
+<!-- ``` -->
+
+<!-- # Laplace approximation -->
+
+<!-- ```{r} -->
+<!-- laplace <- posterior + -->
+<!--   stat_function( -->
+<!--     data = data.frame(x = c(0, 10)), -->
+<!--     aes(x), -->
+<!--     fun = dnorm, -->
+<!--     n = 500, -->
+<!--     args = list(mean = opt_bfgs$mode, sd = sqrt(1 / opt_bfgs$hessian)), -->
+<!--     col = cbpalette[3] -->
+<!--   ) -->
+<!-- ``` -->
+
+<!-- # Laplace approximation -->
+
+<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="The Laplace approximation in this case is good near the mode but not in the tails."} -->
+<!-- laplace + -->
+<!--   annotate("text", x = 5.5, y = 0.35, label = "Laplace approximation", col = cbpalette[3], size = 5) -->
+<!-- ``` -->
+
+<!-- # Computation of the Laplace approximation -->
+
+<!-- * This computation was simple, and involved -->
+
+<!-- 1. Optimising a function -->
+<!-- 2. Taking the mode and the Hessian at the mode -->
+
+<!-- # The marginal Laplace approximation -->
+
+<!-- * If we don't want to pretend the whole posterior distribution is Gaussian, another option is to pretend \textcolor{hilit}{some of its marginals are} -->
+<!-- * This is how `TMB` works: it's up to the user to choose which parameters should be Gaussian using the `random` option -->
+
+<!-- # Which parameters should we treat as Gaussian? -->
+
+<!-- * In spatio-temporal statistics we have data indexed by space and time -->
+<!-- * We use random effects also indexed by space and time to model this data -->
+<!-- * Spatio-temporal fields can be big -->
+<!-- * Willing to make assumptions about how things vary over spacetime -->
+
+<!-- \begin{center} -->
+<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
+
+<!-- $\implies$ treat the latent field parameters as Gaussian! That's the majority of the integral done. -->
+
+<!-- \end{tcolorbox} -->
+<!-- \end{center} -->
+
+<!-- # What about the hyperparameters? -->
+
+<!-- * `TMB` uses optimisation to find the hyperparameters which maximise the marginal Laplace approximation -->
+<!-- * This is the "outer" optimisation loop, where the "inner" is for computation of the Gaussian distribution -->
+
+<!-- \begin{center} -->
+<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
+
+<!-- Inference for the latent field is based on a single value of the hyperparameters (the mode) -- so called empirical Bayes $\implies$ no uncertainty in the hyperparameters taken into account!
+How can you sleep at night.
+-->
+
+<!-- \end{tcolorbox} -->
+<!-- \end{center} -->
+
+<!-- # Quadrature -->
+
+<!-- * This brings us to our other method for solving integrals deterministically -->
+<!-- * Say we have a function, then quadrature has two ingredients -->
+
+<!-- 1. Nodes: points to evaluate the function at -->
+<!-- 2. Weights: importance of function evaluation at that point -->
+
+<!-- * You evaluate the function at the nodes, then do a weighted sum to calculate your integral -->
+
+<!-- # Trapezoid rule example -->
+
+<!-- * Let's compute $\int_0^\pi \sin(x) \text{d}x = 2$ using quadrature -->
+
+<!-- ```{r} -->
+<!-- trapezoid_rule <- function(x, spacing) { -->
+<!--   # Assumes nodes are evenly spaced -->
+<!--   w <- rep(spacing, length(x)) # Weights given by space between nodes -->
+<!--   w[1] <- w[1] / 2 # Apart from the first which is halved -->
+<!--   w[length(x)] <- w[length(x)] / 2 # And the last, also halved -->
+<!--   sum(w * x) # Compute the weighted sum -->
+<!-- } -->
+<!-- ``` -->
+
+<!-- # -->
+
+<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."} -->
+<!-- finegrid <- seq(0, pi, length.out = 1000) -->
+
+<!-- plot <- data.frame(x = finegrid, y = sin(finegrid)) %>% -->
+<!--   ggplot(aes(x = x, y = y)) + -->
+<!--    geom_line(col = cbpalette[1]) + -->
+<!--    theme_minimal() + -->
+<!--    labs(x = "x", y = "sin(x)") -->
+
+<!-- trapezoid_plot <- function(N) { -->
+<!--   grid <- seq(0, pi, length.out = N) -->
+<!--   int <- trapezoid_rule(x = sin(grid), spacing = grid[2] - grid[1]) -->
+
+<!--   plot +  -->
+<!--     geom_bar( -->
+<!--       data = data.frame(x = grid, y = sin(grid)), -->
+<!--       aes(x = x, y = y), alpha = 0.7, stat = "identity", -->
+<!--       inherit.aes = FALSE, fill = cbpalette[2]) + -->
+<!--     theme_minimal() + -->
+<!--     labs( -->
+<!--       subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"), -->
+<!--       x = "x", y = "sin(x)" -->
+<!--     ) -->
+<!-- } -->
+
+<!-- trapezoid_plot(N = 10) -->
+<!-- ``` -->
+
+<!-- # -->
+
+<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."} -->
+<!-- trapezoid_plot(N = 30) -->
+<!-- ``` -->
+
+<!-- # -->
+
+<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."} -->
+<!-- trapezoid_plot(N = 100) -->
+<!-- ``` -->
+
+<!-- # Adaptive Gauss-Hermite quadrature -->
+
+<!-- * \textcolor{hilit}{Gauss-Hermite quadrature} is a method for picking nodes and weights based on the theory of polynomial interpolation -->
+<!-- * It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution -->
+<!-- * The \textcolor{hilit}{adaptive} part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data -->
+<!-- * Implemented by the `aghq` package [@stringer2021implementing] -->
+
+<!-- #  -->
+
+<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points in two dimensions with $k = 3$."} -->
+<!-- mu <- c(1, 1.5) -->
+<!-- cov <- matrix(c(2, 1, 1, 1), ncol = 2) -->
+
+<!-- obj <- function(theta) { -->
+<!--   mvtnorm::dmvnorm(theta, mean = mu, sigma = cov) -->
+<!-- } -->
+
+<!-- grid <- expand.grid( -->
+<!--   theta1 = seq(-2, 5, length.out = 700), -->
+<!--   theta2 = seq(-2, 5, length.out = 700) -->
+<!-- ) -->
+
+<!-- ground_truth <- cbind(grid, pdf = obj(grid)) -->
+
+<!-- plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) + -->
+<!--   geom_contour(col = cbpalette[1]) + -->
+<!--   coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) + -->
+<!--   labs(x = "", y = "") + -->
+<!--   theme_minimal() -->
+
+<!-- gg <- mvQuad::createNIGrid(2, "GHe", 3) -->
+
+<!-- add_points <- function(plot0, gg) { -->
+<!--   plot0 + -->
+<!--     geom_point( -->
+<!--       data = mvQuad::getNodes(gg) %>% -->
+<!--               as.data.frame() %>% -->
+<!--               mutate(weights = mvQuad::getWeights(gg)),  -->
+<!--       aes(x = V1, y = V2, size = weights), -->
+<!--       alpha = 0.8, -->
+<!--       col = cbpalette[2], -->
+<!--       inherit.aes = FALSE -->
+<!--     ) + -->
+<!--     scale_size_continuous(range = c(1, 2)) -->
+<!-- } -->
+
+<!-- add_points(plot0, gg) + -->
+<!--   labs(size = "Weight", caption = "") -->
+<!-- ``` -->
+
+<!-- # -->
+
+<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean $z + \\hat \\theta$."} -->
+<!-- gg2 <- gg -->
+<!-- mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2) -->
+
+<!-- add_points(plot0, gg2) + -->
+<!--   labs(size = "Weight", caption = "") -->
+<!-- ``` -->
+
+<!-- # -->
+
+<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $Lz + \\hat \\theta$."} -->
+<!-- gg3 <- gg -->
+<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2) -->
+
+<!-- add_points(plot0, gg3) + -->
+<!--   labs(size = "Weight", caption = "") -->
+<!-- ``` -->
+
+<!-- # -->
+
+<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2} z + \\hat \\theta$."} -->
+<!-- gg3 <- gg -->
+<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1) -->
+
+<!-- add_points(plot0, gg3) + -->
+<!--   labs(size = "Weight", caption = "") -->
+<!-- ``` -->

### 2023-06-12 src/dev_scale-grid/scale-grid.Rmd
Temporarily keep version of tmb.rds from last archived orderly repo that ran without this bug #50 #49

@@
 ```{r, class.source = 'fold-hide'}
-compile("naomi_simple.cpp")
-dyn.load(dynlib("naomi_simple"))
+TMB::compile("naomi_simple.cpp")
+dyn.load(TMB::dynlib("naomi_simple"))
 
-tmb <- readRDS("depends/tmb.rds")
+# tmb <- readRDS("depends/tmb.rds")
+tmb <- readRDS("tmb.rds")
 fit <- tmb$fit
@@
 ```{r}
-sd_out <- sdreport(obj = fit$obj, par.fixed = fit$par, getJointPrecision = TRUE)
+testrun <- parallel::mcparallel({TMB::sdreport(obj = fit$obj, par.fixed = fit$par, getJointPrecision = TRUE)})
+sd_out <- parallel::mccollect(testrun, wait = TRUE, timeout = 0, intermediate = FALSE)
 ```

### 2023-06-12 src/dev_scale-grid/scale-grid.Rmd
Add marginal SD plot output to dev_scale-grid and rerun #49

@@
 sd_out <- parallel::mccollect(testrun, wait = TRUE, timeout = 0, intermediate = FALSE)
+sd_out <- sd_out[[1]]
 ```
@@
 ```{r sd, class.source = 'fold-hide', fig.cap="Marginal standard deviations."}
+cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
 sd <- sqrt(diag(sd_out$cov.fixed))
 
-data.frame(par = names(sd), sd = unname(sd)) %>%
-  ggplot(aes(x = reorder(par, sd), y = sd)) +
-    geom_point(alpha = 0.6) +
+(marginal_sd_plot <- data.frame(par = names(sd), sd = unname(sd)) %>%
+  mutate(
+    start = str_extract(par, "^[^_]+"),
+    scale = fct_recode(start, "Log" = "log", "Logit" = "logit", "Other" = "OmegaT")
+  ) %>%
+  ggplot(aes(x = reorder(par, sd), y = sd, col = scale)) +
+    geom_point(size = 2) +
     coord_flip() +
+    scale_color_manual(values = c("#56B4E9","#009E73", "#E69F00")) +
     lims(y = c(0, 2)) +
     theme_minimal() +
-    labs(x = "", y = "SD from TMB")
+    labs(x = "", y = "Marginal standard deviation", col = "Scale"))
+
+ggsave("marginal-sd.png", plot = marginal_sd_plot, height = 3.5, w = 6.25)
 ```
@@
 
-ggsave("tv-plot.png", plot = tv_plot, h = 4, w = 6.25)
+ggsave("tv-plot.png", plot = tv_plot, h = 3, w = 6.25)
 ```
@@
 
-ggsave("reduced-rank-plot.png", reduced_rank_plot, h = 4, w = 6.25)
+ggsave("reduced-rank-plot.png", reduced_rank_plot, h = 3.5, w = 6.25)
 ```

### 2023-06-13 src/docs_paper/paper.Rmd
Update abstract
Paper drafting

@@
 Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
-The model is provided as a tool for countries to input their data to and generate estimates, during a yearly process supported by UNAIDS.
+The model is provided as a tool for countries to input their data to and generate estimates during a yearly process supported by UNAIDS.
 Currently, inference is conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
@@
 <!-- * [ ] Tying together Naomi model within LGM / ELGM mathematical notation -->
-<!-- * [ ] Create DAG representing Naomi model. This would likely go into the appenxix, so use that notation -->
-<!-- * [ ] Rewrite section on why Naomi is an ELGM, drawing from slides -->
-<!-- * [ ] Write section on Laplace approximation -->
-<!-- * [ ] Write section on AGHQ, tying together connection to Laplace for k = 1 -->
-<!-- * [ ] Move algorithm to algorithm environment -->
+<!-- * [ ] Create DAG representing Naomi model. This would likely go into the appendix, so use that notation -->
+<!-- * [ ] Tie together connection to Laplace for AGHQ k = 1 -->
 
@@
 
-To mount an effective public health response to the HIV epidemic, it is crucial to have accurate, timely estimates of HIV indicators at the geographic level at which health systems are planned and delivered.
-However, producing these estimates is challenging, in large part due to limitations of the available data sources.
-Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they are conducted only every five years or so, with limited sample size at the district level.
-Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available closer to real-time but based on limited or non-representative samples of the population.
-To address these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
-Modelling multiple data sources jointly has many benefits, including mitigating the limitations of any single source, increasing statistical power, and prompting investigation into any conflicts of information between sources.
-Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates in a yearly process supported by UNAIDS.
-Creation of estimates by country teams, rather than external agencies, is a noteworthy feature of the HIV response.
-Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust and ownership of the resulting estimates.
+Accurate estimates of HIV indicators are crucial for mounting an effective public health response to the HIV epidemic.
+These estimates should be timely, and at a geographic level at which health systems are planned and delivered.
+Producing granular estimates is challenging, in large part due to limitations of the available data sources.
+Nationally-representative household surveys provide the most statistically reliable data, but are only conducted every five years or so in most countries, with limited sample size at the district level, due to their high costs to run.
+Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in closer to real-time, but based on limited or non-representative samples of the population.
+To address these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
+Modelling multiple data sources jointly mitigates the limitations of any single source, increases statistical power, and can prompts investigation into any conflicts of information.
 
-The complexity of Naomi, in combination with practical requirements, present a difficult Bayesian inference problem.
-First, due to dependence of observations on multiple structured additive predictors, Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
-Further, as well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number typically required for use of integrated nested Laplace approximations [@rue2009approximate].
-Second, any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production across a range of countries.
-In this setting, Markov chain Monte Carlo (MCMC) approaches are prohibitively slow due to the scale of the model and challenging features of its posterior geometry [@neal2003slice].
+Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates during workshops as a part of a yearly process supported by UNAIDS.
+Creation of estimates by country teams, rather than external agencies or researchers, is an important and distinctive feature of the HIV response.
+Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust in the resulting estimates, creating a virtuous cycle of data quality, use and ownership.
 
-Inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-Owing to its speed and flexibility, `TMB` is gaining popularity more broadly in spatial statistics [@osgood2022statistical].
+Naomi is a complex model, and as such presents a challenging Bayesian inference problem.
+As well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number typically required for use of integrated nested Laplace approximations [INLA; @rue2009approximate].
+Moreover, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
+As such, the model is not easily specified by any software which requires a formula interface.
+
+To allow for interactive review and iteration of model results, the inference procedure should be fast for workshop participants to fit the model on a laptop.
+Due to the scale of the model and features of its posterior geometry [@neal2003slice], Markov chain Monte Carlo (MCMC) approaches are therefore prohibitively slow.
+Furthermore, it is preferable that use of the inference method across countries to be automatic, and not require substantial statistical expertise, as would be the case for monitoring MCMC convergence and suitability.
+
+To meet these requirements, inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
+Owing to its speed and flexibility, `TMB` is gaining popularity more broadly, particularly in spatial statistics [@osgood2022statistical].
 Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the parameters.
-In the Naomi model, this subset is the high-dimensional latent field, leaving the smaller number of hyperparameters.
-`TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation, Taking tnspiration from the AD Model Builder package [@fournier2012ad].
+For the Naomi model, this subset is the high-dimensional latent field, leaving a smaller number of hyperparameters.
+`TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation, Taking inspiration from the AD Model Builder package [@fournier2012ad].
+
 Although the approach of `TMB` is fast, within the empirical Bayes framework hyperparameter uncertainty is not properly accounted for in the latent field posterior.
-This consideration motivates us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
-
+This consideration motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
 To obtain fast, accurate Bayesian inferences for the Naomi model we developed an inference method which extends adaptive Gauss-Hermite quadrature to handle many hyperparameters.
-We used principal components analysis to create a quadrature grid which is computationally feasible.
+We used principal components analysis to create a computationally feasible quadrature grid.
 Our method is implemented as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
 As `aghq` is designed to naturally interface with `TMB`, use is simple when provided a \textsf{C++} user template for the log-posterior.
+
+Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package [@bachl2019inlabru], INLA within MCMC [@gomez2018markov], and importance sampling with INLA [@berild2022importance], all of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
+`inlabru` approximates non-linear predictors using linearisation, which involves making iterative calls to `R-INLA`.
+<!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
+<!-- Want to say in particular what the class of models that `inlabru` allows you to fit are -->
+<!-- Don't know the answer to this yet, but asking Janine Illian I'm told you don't have to call R-INLA that many times -->
+<!-- I think it'd obviously depend on the exact details of the problem at hand -->
+INLA within MCMC and importance sampling with INLA are suitable for models which are LGMs conditional on some subset of the parameters being fixed.
+<!-- It would or would not be possible to fit the Naomi model in these frameworks. -->
 
@@
 
-@eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points.
+@eaton2021naomi specify a joint model linking three small-area estimation models.
 We consider a simplified version defined only at the time of the most recent household survey with HIV testing, omitting nowcasting and temporal projection, as these time points involve limited inferences.
@@
 \end{align*}
-These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
-In particular
-
-1. $x^\text{ANC}_I$, the number of ANC clients with ascertained status,
-2. $y^\text{ANC}_I$, the number of those with positive status, and
-3. $z^\text{ANC}_I$, the number of those already on ART prior to their first ANC visit.
-
+These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey:
+$m^{\rho^\text{ANC}}_I$, the number of ANC clients with ascertained status,
+$y^{\rho^\text{ANC}}_I$, the number of those with positive status, and
+$y^{\alpha^\text{ANC}}_I$, the number of those already on ART prior to their first ANC visit.
 We use the binomial likelihoods
 \begin{align*}
-y^\text{ANC}_I &\sim \text{Bin}(x^\text{ANC}_I, \rho^\text{ANC}_{I}), \\
-z^\text{ANC}_I &\sim \text{Bin}(y^\text{ANC}_I, \alpha^\text{ANC}_{I}),
+y^{\rho^\text{ANC}}_I &\sim \text{Bin}(m^{\rho^\text{ANC}}_I, \rho^\text{ANC}_{I}), \\
+y^{\alpha^\text{ANC}}_I &\sim \text{Bin}(y^{\rho^\text{ANC}}_I, \alpha^\text{ANC}_{I}),
 \end{align*}
@@
 People living with HIV sometimes choose to access ART services outside of the district that they reside in.
-To account for this, we use multinomial logistic regression to model the probabilities of accessing services outside the home district.
-Let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring, denoted by $x \sim x'$.
+To account for this, we use multinomial logistic regressions to model the probabilities of accessing services outside the home district.
+Briefly, let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring, denoted by $x \sim x'$.
 The log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ are modelled using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
-We then model aggregate ART attendance data $\dot A_I$ using a Gaussian approximation to a sum of binomials.
+We then model aggregate ART attendance data $y^{\text{ART}}_I$ using a Gaussian approximation to a sum of binomials.
 This sum is both by over $i \in I$ and by number of ART clients travelling from district $x'$ to $x$.
@@
 
-Let $\mathbf{y} = (y^{\hat \theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa\}$ and $I \in \mathcal{I}$ be the vector of observations.
-Attempt to write fully the Naomi model in a small number of collected together equations here.
-The ANC testing component and ART attendance component might be hard to do this for.
+Let $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\hat \rho, \hat \alpha, \hat \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, \text{ART}\}$ and $I \in \mathcal{I}$ be a concatenated vector of observations.
+Here, we write the Naomi model fully in a small number of equations.
+The ANC testing component and ART attendance component may be challenging.
 
@@
 \end{equation}
-Inference in TMB proceeds by optimising Equation \ref{eq:la} to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
-Latent field joint and marginal inferences are then direct from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
-Hyperparameter inferences are obtained according to method which should be specified here.
+Inference in TMB proceeds by optimising Equation \ref{eq:la} using a gradient-based routine to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
+Latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
+Hyperparameter inferences are obtained according to some method which should be specified here.
 
@@
 In adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] the nodes are shifted and rotated to suit the particular integrand.
-Repositioning the nodes is especially important in statistical quadrature problems, where the integral depends on data $\y$ such that regions of high density cannot reasonably be known in advance.
-To obtain an AGHQ estimate of Equation \ref{eq:quad}, let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ a matrix decomposition of the inverse curvature, then
+Repositioning the nodes is especially important in statistical quadrature problems, where the integral depends on data $\y$ such that regions of high density are not known in advance.
+To obtain an AGHQ estimate of Equation \ref{eq:quad}, let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then
 \begin{equation}
@@
 That is, the unadapted nodes have been shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
-Two options for the particular matrix decomposition are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$ and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$ [@jackel2005note].
+Two alternatives for the decomposition are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$ and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$ [@jackel2005note].
 Figure \ref{fig:aghq} demonstrates GHQ, as well as adaption for these two choices of matrix decomposition.
@@
 
-ground_truth <- cbind(grid, pdf = obj(grid))
+ground_truth <- cbind(grid, pdf = apply(grid, 1, obj))
 

### 2023-06-13 src/docs_paper/paper.Rmd
First shot at an interesting 2D non-Gaussian distribution

@@
 ```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension are adapted based on the mean and covariance matrix of the target via the Cholesky decomposition or spectral decompostion of the inverse curvature at the mode."}
-mu <- c(1, 1.5)
-cov <- matrix(c(2, 1, 1, 1), ncol = 2)
+TMB::compile("2d.cpp")
+dyn.load(TMB::dynlib("2d"))
 
-obj <- function(theta) {
-  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
-}
+obj <- TMB::MakeADFun(data = list(), parameters = list(theta1 = 0, theta2 = 0), DLL = "2d")
 
 grid <- expand.grid(
-  theta1 = seq(-2, 5, length.out = 700),
-  theta2 = seq(-2, 5, length.out = 700)
+  theta1 = seq(-5, 5, length.out = 400),
+  theta2 = seq(-5, 5, length.out = 400)
 )
 
-ground_truth <- cbind(grid, pdf = apply(grid, 1, obj))
+ground_truth <- cbind(grid, pdf = apply(grid, 1, function(x) exp(-1 * obj$fn(x))))
+
+opt <- nlminb(
+  start = obj$par,
+  objective = obj$fn,
+  gradient = obj$gr,
+  control = list(iter.max = 1000, trace = 0)
+)
+
+sd_out <- TMB::sdreport(
+  obj,
+  par.fixed = opt$par,
+  getJointPrecision = TRUE
+)
+
+mu <- opt$par
+cov <- as.matrix(sd_out$cov.fixed)
 
@@
   geom_contour(col = "lightgrey") +
-  coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
   labs(x = "", y = "") +

### 2023-06-13 src/docs_paper/2d.cpp

@@ -8,9 +8,14 @@ Type objective_function<Type>::operator()()
   PARAMETER(theta1);
   PARAMETER(theta2);
 
-  // Initialise negative log-likelihood
   Type nll;
   nll = Type(0.0);
+
+  Type phi1 = 0.2 * theta1 + 0.3 * theta2;
+  Type phi2 = 0.6 * theta1 + 0.1 * theta2;
+
+  nll -= dsn(phi1, Type(2.0), true);
+  nll -= dsn(phi2, Type(-2.0), true);
 
   return(nll);
 }

### 2023-06-15 src/docs_paper/paper.Rmd
Iterate paper, appendix

@@
 \end{equation*}
-be weighted, aggregate survey observations, where $j$ indexes individuals across all strata $i \in I$ and $w_j$ are design weights.
-The observed number of outcomes are $y^{\hat \theta}_{I} = m^{\hat \theta}_{I} \cdot \hat \theta_{I}$ where
+be weighted, aggregate survey observations, with individual responses $\theta_j \in \{0, 1\}$ and design weights $w_j$.
+The index $j$ is across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
+<!-- Relevant means e.g. for alpha it's only the HIV positive individuals -->
+The observed number of outcomes are then $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
 \begin{equation*}
-m^{\hat \theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
+m^{\theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
 \end{equation*}
@@
 \begin{equation*}
-y^{\hat \theta}_{I} \sim \text{xBin}(m^{\hat \theta}_{I}, \theta_{I})
+y^{\theta}_{I} \sim \text{xBin}(m^{\theta}_{I}, \theta_{I})
 \end{equation*}
@@
 \end{align*}
-These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey:
-$m^{\rho^\text{ANC}}_I$, the number of ANC clients with ascertained status,
-$y^{\rho^\text{ANC}}_I$, the number of those with positive status, and
-$y^{\alpha^\text{ANC}}_I$, the number of those already on ART prior to their first ANC visit.
-We use the binomial likelihoods
+These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
+Of the number of ANC clients with ascertained status $m^{\rho^\text{ANC}}_I$, we model the number of those with positive status $y^{\rho^\text{ANC}}_I$ and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ using nested binomial likelihoods
 \begin{align*}
@@
 \end{align*}
-where, again, we use weighted aggregates
+As above, we use weighted aggregates
 \begin{equation*}
@@
 The log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ are modelled using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
-We then model aggregate ART attendance data $y^{\text{ART}}_I$ using a Gaussian approximation to a sum of binomials.
-This sum is both by over $i \in I$ and by number of ART clients travelling from district $x'$ to $x$.
+We then model aggregate ART attendance data $y^{N^\text{ART}}_I$ using a Gaussian approximation to a sum of binomials.
+This sum is over both $i \in I$ and the number of ART clients travelling from district $x'$ to $x$.
 More details regarding this part of the model are provided in Appendix S1.
@@
 
-Let $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\hat \rho, \hat \alpha, \hat \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, \text{ART}\}$ and $I \in \mathcal{I}$ be a concatenated vector of observations.
-Here, we write the Naomi model fully in a small number of equations.
-The ANC testing component and ART attendance component may be challenging.
+Let $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$ be a concatenated vector of observations.
+Here, we could attempt to write the Naomi model fully in a small number of equations, but likely it'd be difficult.
 
@@
 
-If Naomi was written out in full mathematical detail, I might be able to make these reasons more concrete.
-It would also be good to specify how much of an impact each of these reasons is likely to have on the difficulty of inference.
+If Naomi was written out in full mathematical detail (in the section above), it would be easier to make these reasons more concrete.
+We should also specify how much of an impact each of these reasons is likely to have on the difficulty of inference.
 
@@
 \end{align}
-Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the above posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$ via nested applications of the Laplace approximation and AGHQ.
+Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the above posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^N$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$ via nested applications of the Laplace approximation and AGHQ.
 
@@
 Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\Hb}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
-\begin{align*}
-\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
-\hat {\Hb}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
-\end{align*}
+\begin{align}
+\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \label{eq:mode} \\
+\hat {\Hb}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}. \label{eq:precision}
+\end{align}
 Then the Laplace approximation to $p(\btheta, \y)$ is given by
 \begin{equation}
-\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}. \label{eq:la}
+\tilde p_\texttt{LA}(\btheta, \y)
+= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+= \sqrt{\frac{\det(\hat {\Hb}(\btheta))}{(2 \pi)^{N}}} p(\y, \hat \x(\btheta), \btheta). \label{eq:la}
 \end{equation}
 Inference in TMB proceeds by optimising Equation \ref{eq:la} using a gradient-based routine to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
+Each evaluation in the optimisation requires an inner optimisation to obtain $\hat \x(\btheta)$ via Equation \ref{eq:mode}.
 Latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
-Hyperparameter inferences are obtained according to some method which should be specified here.
+Hyperparameter inferences are obtained according to some method which should be specified here as well.
 
@@
 
-Quadrature rules can be used to approximate the integral of $\tilde p_\texttt{LA}(\btheta, \y)$ via a weighted sum
+Quadrature rules can be used to approximate the integral of $\tilde p_\texttt{LA}(\btheta, \y)$ via the weighted sum
 \begin{equation}
@@
 where $\z \in \mathcal{Q}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
-Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule where, in the univariate case, the nodes $\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}$ are selected as zeros of the $k$th Hermite polynomial.
+Gauss-Hermite quadrature [GHQ; @davis1975methods] is one such quadrature rule, where in the univariate case the nodes $\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}$ are selected as zeros of the $k$th Hermite polynomial.
 The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, where $\phi(\cdot)$ is a standard univariate Gaussian density.
 GHQ is attractive because it is exact for functions which are a polynomial of total order no more than $2k - 1$ multiplied by a Gaussian density.
-Multivariate GHQ rules are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
+We typically expected posterior distributions to be relatively well described by this class of functions.
+Multivariate GHQ rules, required to integrate over $\btheta$, are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
 
@@
 
-```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension are adapted based on the mean and covariance matrix of the target via the Cholesky decomposition or spectral decompostion of the inverse curvature at the mode."}
+```{r aghq, results='hide', fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension.
+Adaption occurs based on the mean and covariance matrix of the target via the Cholesky decomposition or spectral decompostion of the inverse curvature at the mode.
+The integrand is $f(\\btheta) = \\text{sn}(0.3 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.5 \\theta_1 - 0.3 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
 TMB::compile("2d.cpp")
@@
 grid <- expand.grid(
-  theta1 = seq(-5, 5, length.out = 400),
-  theta2 = seq(-5, 5, length.out = 400)
+  theta1 = seq(-8, 8, length.out = 400),
+  theta2 = seq(-8, 8, length.out = 400)
 )
@@
 mu <- opt$par
-cov <- as.matrix(sd_out$cov.fixed)
+cov <- sd_out$cov.fixed
 
@@
   geom_contour(col = "lightgrey") +
+  coord_fixed(xlim = c(-8, 8), ylim = c(-8, 8), ratio = 1) +
   labs(x = "", y = "") +
@@
 add_points <- function(plot0, gg) {
+
+  points <- mvQuad::getNodes(gg) %>%
+    as.data.frame() %>%
+    mutate(weights = mvQuad::getWeights(gg))
+
+  colnames(points) <- c("theta1", "theta2", "weights")
+
   plot0 +
     geom_point(
-      data = mvQuad::getNodes(gg) %>%
-              as.data.frame() %>%
-              mutate(weights = mvQuad::getWeights(gg)),
-      aes(x = V1, y = V2, size = weights),
+      data = points,
+      aes(x = theta1, y = theta2, size = weights),
       alpha = 0.8,
@@
 gg2 <- gg
-mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2)
+mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 1)
 
@@
 where $\hat{\mathbf{E}}_{\texttt{LA}, s} = \cdots$ and $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s} = \cdots$ are terms which should be described.
-The final panel of Figure \ref{fig:aghq} illustrates a case where $m = 2$ and $s = 1$ such that 
+The final panel of Figure \ref{fig:aghq} illustrates a case where $m = 2$ and $s = 1$. 
 We refer to use of this quadrature rule as PCA-AGHQ.
@@
 We are not aware of any suitable approaches to obtain multinomial samples in such cases.
+It is however, still possible to calculate the normalising constant using a sparse rule, which we do so in Appendix S3, finding it to be related to that from PCA-AGHQ as follows.
+
+## Laplace marginals
+
+See Appendix S4.
 
@@
 We obtained satisfactory NUTS results by gradually increasing the amount of computation until all diagnostics were acceptable.
-These include the potential scale reduction factors $\hat R$ [@gelman1992inference; @vehtari2021rank] for each parameter, bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments [@betancourt2017conceptual].
+All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] attained $\hat R < 1.05$.
 For full details see Appendix S2.
-We treat the NUTS results as a gold-standard.
+On this basis, though inaccuracies were still possible, we considered the NUTS results to be a gold-standard.
 
@@
 
-We used a Scree plot to select the number of principal components $s < m$ to keep.
-Figure showing Scree plot.
-Figure showing approximation to precision matrix.
+We used a Scree plot (Figure \ref{fig:scree}) based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})$ to select the number of principal components $s < m$ to keep.
+We choose $s = 8$, which was sufficient to explain close to 90% of total variation.
+Reconstruction of the Hessian using this reduced rank approximation is shown in Figure \ref{fig:reduced-rank}.
 
-Figure \ref{fig:node-positions} shows the positions of the generated PCA-AGHQ nodes overlaid onto the hyperparameter marginal posteriors obtained NUTS.
-For around 12 of the 24 hyperparameters, the marginal was well covered by the nodes.
-Although this is an improvement on the $s = 8$ that could be covered naively, there remain many hyperparameters uncovered.
-Furthermore, it is likely that the hyperparameters with skewed marginal posteriors are more important to cover well.
-We have not yet found a way to take this into account.
+```{r scree, fig.cap="Most variation could be explained by fewer than the full 24 dimensions."}
+knitr::include_graphics("depends/tv-plot.png")
+```
+
+```{r reduced-rank, fig.cap="The reduced rank approximation to the Hessian is visually similar."}
+knitr::include_graphics("depends/reduced-rank-plot.png")
+```
 
@@
 
-Demonstrate that the 12 hyperparameters which are well covered are the ones that you would expect given their marginal standard deviations.
+```{r marginal-sd, fig.cap="Hyperparameters on the logit-scale had higher marginal standard deviations than those on the log-scale."}
+knitr::include_graphics("depends/marginal-sd.png")
+```
 
-One crude approach for assessing appropriateness of the grid is to look at the estimate of $\log p_\texttt{PCA}(\y)$ for a range of settings.
-Convergence in $\log p_\texttt{PCA}(\y)$ as $s$ and $k$ are increased may suggest sufficiency has been reached.
-Figure shows those values which were computable in a reasonable time.
+### Visual assessment
+
+Figure \ref{fig:node-positions} shows the positions of the generated PCA-AGHQ nodes overlaid onto the hyperparameter marginal posteriors obtained using NUTS.
+For approximately 12 of the 24 hyperparameters, the marginal was well covered by the nodes.
+Although this is an improvement in so far as $12 > s = 8$ there do remain many hyperparameters uncovered.
+Coverage was positively associated with marginal standard deviation (Figure \ref{fig:marginal-sd}), which varied particuarly according to the scale of the hyperparameter.
+All constrained hyperparameters $\theta$ were transformed to the real line, using either a log ($\theta > 0$) or logit ($\theta \in [0, 1]$) transformation.
+As a result, marginal standard deviations for log transformed hyperparameters were systematically smaller than those which were logit transformed.
+
+### Normalising constant assessment
+
+We assessed appropriateness of the grid by comparing the estimate of $\log p_\texttt{PCA}(\y)$ for a range of settings.
+Convergence in $\log p_\texttt{PCA}(\y)$ as $s$ and $k$ are increased may suggest a suitable grid has been reached.
+Appendix S3 shows those values which were we could compute in a reasonable time.
 
@@
 Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios used in importance sampling.
-See a summary of the results in Table and Figure \ref{fig:psis}.
-
-```{r psis, fig.cap="Results of Pareto-smoothed importance sampling analysis."}
-```
 
@@
 \end{equation*}
-See a summary of the results in Table and Figure \ref{fig:mmd}.
-
-```{r mmd, fig.cap="Results of maximum mean discrepancy analysis."}
-```
 
@@
 
-The Joint United Nations Programme on HIV/AIDS has developed ambitious fast-track targets for scaling up ART treatment with the goal of "ending the AIDS epidemic by 2030".
-Specifically, the "90-90-90 by 2020" fast-track target is that:
-
-* 90% of PLHIV know their status,
-* 90% of those are on antiretroviral therapy (ART), and
-* 90% of those have suppressed viral load.
-
+UNAIDS have developed ambitious fast-track targets for scaling up ART treatment with the goal of "ending the AIDS epidemic by 2030".
+Specifically, the "90-90-90" fast-track target requires that 90% of PLHIV know their status, 90% of those are on ART, and 90% of those have suppressed viral load.
 Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met $\mathbb{P}(\alpha_i > 0.81)$ for each strata $i$.
-We found that both TMB and PCA-AGHQ underestimate the probability that the second 90 target has been met in women (Figure \ref{fig:second90}).
-This difficulty may be related to interactions between the household survey and ANC components of the model.
+We found that both TMB and PCA-AGHQ underestimate these exceedance probabilities, specifically for women (Figure \ref{fig:second90}).
+We hypothesise that this discrepancy is related to interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.
 
@@
 We found that both TMB and PCA-AGHQ overestimate these exceedance probabilities (Figure \ref{fig:1inc}).
-We do not yet have a working hypothesis as to why this is.
+We are yet to formulate hypothesis as to why this is the case.
 
@@
 
+The exceedance probabilities case-study (Section \ref{sec:exceedance}) demonstrates the importance of accurate posterior inferences for realistic use-cases of the Naomi model.
+
 We provide a flexible implementation of the algorithm, building on the `TMB` and `aghq` \textsc{R} packages.
 In doing so, we hope our work enables use of deterministic inference algorithms for ELGMs in applied settings, as well as further methodological exploration of their accuracy and limitations.
-Among the ELGM-type structures of particular interest in spatial epidemiology, many of which feature in Naomi, are: aggregated Gaussian process models [@nandi2020disaggregation], evidence synthesis models [@amoah2020geostatistical].
+Among the ELGM-type structures of particular interest in spatial epidemiology and feature in Naomi, are aggregated Gaussian process models [@nandi2020disaggregation] and evidence synthesis models [@amoah2020geostatistical].
 Although our method is designed for ELGMs, it is possible to use it outside this class, as it is compatible with any model with a `TMB` \textsc{C++} template.
 
-We demonstrated a Bayesian workflow for deterministic inference methods.
-We retained the ability to draw samples from the posterior distributions of interest, facilitating use of posterior predictive checks (Section \ref{sec:model-assessment}) and methods for inference comparison such as Pareto-smoothed importance sampling (Section \ref{sec:inf-comparison}).
-The exceedance probabilities case-study (Section \ref{sec:exceedance}) demonstrates the importance of accurate posterior inferences for realistic use-cases of the Naomi model.
+## Future work
+
+Hyperparameters with skewed marginal posteriors are more important to cover well using quadrature nodes.
+
+Parallel.
 

### 2023-06-15 src/docs_paper/2d.cpp

@@ -11,8 +11,8 @@ Type objective_function<Type>::operator()()
   Type nll;
   nll = Type(0.0);
 
-  Type phi1 = 0.2 * theta1 + 0.3 * theta2;
-  Type phi2 = 0.6 * theta1 + 0.1 * theta2;
+  Type phi1 = 0.3 * theta2;
+  Type phi2 = 0.5 * theta1 - 0.3 * theta2;
 
   nll -= dsn(phi1, Type(2.0), true);
   nll -= dsn(phi2, Type(-2.0), true);

### 2023-06-16 src/naomi-simple_exceedance/exceedance.Rmd
Save exceedance probability output and put into poster #36
Create manuscript ready plot for exceedance, general plot improvements, and rerun report

--- 2023-05-12 src/naomi-simple_exceedance/exceedance.Rmd
@@
 plot1 <- ggplot(df_wide_second90, aes(x = tmbstan, y = TMB)) +
-  geom_point(alpha = 0.3) +
-  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
-  labs(title = "Probability of 81% ART coverage in strata") +
+  geom_point(shape = 1, alpha = 0.6, col = "#999999") +
+  coord_fixed(ratio = 1) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  labs(x = "NUTS", y = "TMB") +
   theme_minimal()
@@
 plot2 <- ggplot(df_wide_second90, aes(x = tmbstan, y = aghq)) +
-  geom_point(alpha = 0.3) +
-  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
+  geom_point(shape = 1, alpha = 0.6, col = "#999999") +
+  coord_fixed(ratio = 1) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  labs(x = "NUTS", y = "PCA-AGHQ") +
   theme_minimal()
@@
 plot1 <- ggplot(df_wide_second90, aes(x = tmbstan, y = TMB, col = sex)) +
-  geom_point(alpha = 0.3) +
-  scale_color_manual(values = c("#E69F00", "#F0E442")) +
-  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
-  labs(title = "Probability of at least 81% ART coverage in strata", col = "Sex") +
-  theme_minimal() +
-  theme(legend.position = "bottom")
+  geom_point(shape = 1, alpha = 0.6) +
+  coord_fixed(ratio = 1) +
+  scale_color_manual(values = c("#E69F00", "#999999")) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  labs(col = "Sex") +
+  guides(col = "none") +
+  labs(x = "NUTS", y = "TMB") +
+  theme_minimal()
 
 plot2 <- ggplot(df_wide_second90, aes(x = tmbstan, y = aghq, col = sex)) +
-  geom_point(alpha = 0.3) +
-  scale_color_manual(values = c("#E69F00", "#F0E442")) +
-  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
-  guides(col = "none") +
+  geom_point(shape = 1, alpha = 0.6) +
+  coord_fixed(ratio = 1) +
+  scale_colour_manual(values = c("#E69F00", "#999999"), labels = c("Female", "Male")) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
+  labs(x = "NUTS", y = "PCA-AGHQ", col = "Sex") +
   theme_minimal()
+
+ggsave("second90.png", plot1 + plot2, h = 4, w = 6.25)
 
@@
 
+df_wide_1inc <- df_wide_1inc %>%
+  left_join(
+    mf_out_fine %>%
+      tibble::rowid_to_column("index"),
+    by = "index"
+  )
+
 plot1 <- ggplot(df_wide_1inc, aes(x = tmbstan, y = TMB)) +
-  geom_point(alpha = 0.3) +
-  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
-  labs(title = "Probability of at least 1% incidence in strata") +
+  geom_point(shape = 1, alpha = 0.6, col = "#999999") +
+  coord_fixed(ratio = 1) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  labs(x = "NUTS", y = "TMB") +
   theme_minimal()
@@
 plot2 <- ggplot(df_wide_1inc, aes(x = tmbstan, y = aghq)) +
-  geom_point(alpha = 0.3) +
-  geom_abline(slope = 1, intercept = 0, col = "#CC79A7", linetype = "dashed") +
+  geom_point(shape = 1, alpha = 0.6, col = "#999999") +
+  coord_fixed(ratio = 1) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  labs(x = "NUTS", y = "PCA-AGHQ") +
   theme_minimal()
+
+ggsave("1inc.png", plot1 + plot2, h = 4, w = 6.25)
 
@@
 
+# Plot for manuscript
+
+Combine the above plots into one nice one.
+ABF (always be facetting):
+
+```{r}
+df_wide <- bind_rows(
+  df_wide_second90 %>%
+    mutate(type = "Second 90"),
+  df_wide_1inc %>%
+    mutate(type = "High incidence")
+)
+
+write_csv(df_wide, "exceedance.csv")
+
+fct_reorg <- function(fac, ...) {
+  fct_recode(fct_relevel(fac, ...), ...)
+}
+
+exceedance_manuscript_plot <- df_wide %>%
+  pivot_longer(cols = c("TMB", "aghq"), names_to = "method", values_to = "estimate") %>%
+  mutate(
+    method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq"),
+    type = fct_relevel(type, "Second 90", "High incidence")
+  ) %>%
+  ggplot(aes(x = tmbstan, y = estimate, col = sex)) +
+    geom_point(shape = 1, alpha = 0.6) +
+    coord_fixed(ratio = 1) +
+    scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    facet_grid(type ~ method) +
+    labs(x = "", y = "", col = "Sex") +
+    theme_minimal() +
+    theme(
+      legend.position = "bottom"
+    )
+
+exceedance_manuscript_plot
+
+ggsave("exceedance.png", exceedance_manuscript_plot, h = 6.5, w = 6.25, bg = "white")
+```
+
 # Excursion and contour uncertainty regions

### 2023-06-16 src/naomi-simple_exceedance/exceedance.Rmd
Should be captioned NUTS!

@@
     facet_grid(type ~ method) +
-    labs(x = "", y = "", col = "Sex") +
+    labs(x = "NUTS", y = "", col = "Sex") +
     theme_minimal() +

### 2023-06-16 src/docs_paper/appendix.Rmd
Iterate paper, appendix
Lots more paper writing! Getting there eh

@@
   - Seth R. Flaxman^[Department of Computer Science, Oxford University]
-  - Jeffrey W. Eaton^[Department of Infectious Disease Epidemiology, Imperial College London]
+  - Jeffrey W. Eaton^[Harvard T.H. Chan School of Public Health, Harvard University]
 output:
@@
 
-# Simplified Naomi \label{sec:math}
+# Simplified Naomi model description \label{sec:math}
 
@@
 
+# AGHQ and PCA-AGHQ
+
+```{r marginal-sd, fig.cap="Hyperparameters on the logit-scale had systematically higher marginal standard deviations than those on the log-scale.
+This is because variation on the real scale is compressed by the inverse logit and expanded by the inverse log (exponential)."}
+knitr::include_graphics("depends/marginal-sd.png")
+```
+
+```{r scree, fig.cap="Most variation could be explained by fewer than the full 24 dimensions."}
+knitr::include_graphics("depends/tv-plot.png")
+```
+
+```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank (24) matrix."}
+knitr::include_graphics("depends/reduced-rank-plot.png")
+```
+
+\newpage
+
+# Further inference comparison results
+
+## KS tests
+
+```{r}
+ks_summary <- readRDS("depends/ks_summary.rds")
+
+ks_summary %>%
+  filter(Type == "Latent field") %>%
+  group_by(Parameter) %>%
+  summarise(
+    `D(TMB)` = round(mean(`KS(TMB, tmbstan)`), 3),
+    `D(PCA-AGHQ)` = round(mean(`KS(aghq, tmbstan)`), 3)
+  ) %>%
+  gt::gt(rowname_col = "Parameter") %>%
+  gt::summary_rows(
+   columns = everything(),
+   fns = list(Average = ~ mean(.x)),
+   formatter = gt::fmt_number,
+   decimals = 3
+  )
+```
+
+\newpage
+
 # Laplace marginals algorithm

### 2023-06-16 src/docs_paper/paper.Rmd
Lots more paper writing! Getting there eh

@@
 Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference for ELGMs using simplified INLA and AGHQ.
-As a case-study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:naomi}.
+As a case study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:naomi}.
 We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
@@
 We obtained satisfactory NUTS results by gradually increasing the amount of computation until all diagnostics were acceptable.
-All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] attained $\hat R < 1.05$.
+This requried chains of length 20,000, thinned by a factor of 20 for ease-of-storage.
+All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$.
 For full details see Appendix S2.
-On this basis, though inaccuracies were still possible, we considered the NUTS results to be a gold-standard.
+Though inaccuracies remain possible, we considered the NUTS results to be a gold-standard.
 
@@
 
-We used a Scree plot (Figure \ref{fig:scree}) based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})$ to select the number of principal components $s < m$ to keep.
-We choose $s = 8$, which was sufficient to explain close to 90% of total variation.
-Reconstruction of the Hessian using this reduced rank approximation is shown in Figure \ref{fig:reduced-rank}.
-
-```{r scree, fig.cap="Most variation could be explained by fewer than the full 24 dimensions."}
-knitr::include_graphics("depends/tv-plot.png")
-```
-
-```{r reduced-rank, fig.cap="The reduced rank approximation to the Hessian is visually similar."}
-knitr::include_graphics("depends/reduced-rank-plot.png")
-```
+We used a Scree plot (Figure S$X$) based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})$ to select the number of principal components $s = 8 < m = 24$ to keep, sufficient to explain close to 90% of total variation.
+This resulted in a close reconstruction of the Hessian using the reduced rank approximation (Figure S$X$).
 
@@
 
-```{r marginal-sd, fig.cap="Hyperparameters on the logit-scale had higher marginal standard deviations than those on the log-scale."}
-knitr::include_graphics("depends/marginal-sd.png")
-```
-
 ### Visual assessment
 
-Figure \ref{fig:node-positions} shows the positions of the generated PCA-AGHQ nodes overlaid onto the hyperparameter marginal posteriors obtained using NUTS.
-For approximately 12 of the 24 hyperparameters, the marginal was well covered by the nodes.
-Although this is an improvement in so far as $12 > s = 8$ there do remain many hyperparameters uncovered.
-Coverage was positively associated with marginal standard deviation (Figure \ref{fig:marginal-sd}), which varied particuarly according to the scale of the hyperparameter.
+Overlaying the generated PCA-AGHQ nodes overlaid onto the hyperparameter marginal posteriors obtained using NUTS, we found that approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
+Though $12 > s = 8$ there remain many hyperparameters poorly covered.
+Coverage was associated with marginal standard deviation (Figure S$X$), which varied particularly according to the hyperparameter scale.
 All constrained hyperparameters $\theta$ were transformed to the real line, using either a log ($\theta > 0$) or logit ($\theta \in [0, 1]$) transformation.
@@
 Convergence in $\log p_\texttt{PCA}(\y)$ as $s$ and $k$ are increased may suggest a suitable grid has been reached.
-Appendix S3 shows those values which were we could compute in a reasonable time.
+Appendix S3 shows those values which we could compute in a reasonable time (less than 24 hours using a high performance computing cluster).
 
@@
 Let $\phi$ be a generic model parameter.
-To assess the informativeness of the data we compared (Figure \ref{fig:contraction}) the prior variance $\sigma_\text{prior}^2(\phi)$ to the posterior variance $\sigma_\text{posterior}^2(\phi)$ via the posterior contraction [@schad2021toward]
-\begin{equation*}
-c(\phi) = 1 - \frac{\sigma_\text{posterior}^2(\phi)}{\sigma_\text{prior}^2(\phi)}.
-\end{equation*}
-For the parameters of length greater than one, we averaged the posterior contraction.
+To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\phi)$ to the posterior variance $\sigma_\text{posterior}^2(\phi)$ via the posterior contraction $c(\phi) = 1 - (\sigma_\text{posterior}^2(\phi) / \sigma_\text{prior}^2(\phi))$ [@schad2021toward].
+We found that (Figure \ref{fig:contraction})) something something.
+For greater interpretability, facet parameters in this plot according to model component.
 
@@
 The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two empirical cumulative distribution (ECDF) functions $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
-We compare the KS statistics
-\begin{align*}
-D_\text{TMB} &= \sup_\varphi | F_\text{NUTS}(\varphi) - F_\text{TMB}(\varphi)|, \\
-D_\text{PCA-AGHQ} &= \sup_\varphi | F_\text{NUTS}(\varphi) - F_\text{PCA-AGHQ}(\varphi)|.
-\end{align*}
+We compare the KS statistics $D_\text{TMB} = \sup_\varphi | F_\text{NUTS}(\varphi) - F_\text{TMB}(\varphi)|$ and $D_\text{PCA-AGHQ} = \sup_\varphi | F_\text{NUTS}(\varphi) - F_\text{PCA-AGHQ}(\varphi)|$.
 See an illustration of the KS test in Figure \ref{fig:ks} and a summary of the results in Table.
 
-```{r ks, fig.height = 3, fig.cap="Example KS test for one parameter."}
+```{r ks, fig.height = 4, fig.cap="Example KS test for one parameter."}
 df_compare <- readRDS("depends/beta_alpha.rds") %>%
@@
 
-```{r}
-ks_summary <- readRDS("depends/ks_summary.rds")
-
-ks_summary %>%
-  filter(Type == "Latent field") %>%
-  group_by(Parameter) %>%
-  summarise(
-    `D(TMB)` = round(mean(`KS(TMB, tmbstan)`), 3),
-    `D(PCA-AGHQ)` = round(mean(`KS(aghq, tmbstan)`), 3)
-  ) %>%
-  gt::gt(rowname_col = "Parameter") %>%
-  gt::summary_rows(
-   columns = everything(),
-   fns = list(Average = ~ mean(.x)),
-   formatter = gt::fmt_number,
-   decimals = 3
-  )
-```
-
 ### Pareto-smoothed importance sampling
 
-As well as marginal distributions, we are interested in assessing the accuracy of joint distributions.
-Let $\{\bphi_i\}_{i = 1}^n$ be generic samples from a joint posterior.
-Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios used in importance sampling.
+Alongside marginal distributions, we are interested in assessing the accuracy of joint distributions.
+Let $\{\bphi_i\}_{i = 1}^n$ be joint posterior samples.
+Pareto-smoothed importance sampling [PSIS; @vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios used in importance sampling.
+Results for the PSIS analysis are pending.
 
@@
 
-Another way to compare joint distributions is via the maximum mean discrepancy [MMD; @gretton2006kernel].
-Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of samples from a joint posterior obtained using different inference methods.
+Joint distributions can also be compared via the maximum mean discrepancy [MMD; @gretton2006kernel].
+Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples obtained from different inference methods, and $k$ be a kernel.
 The MMD can be empirically estimated by
@@
 \end{equation*}
+We set $k(\bphi_i, \bphi_j) = \exp(-\sigma \lVert \bphi_i - \bphi_j \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
+We the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080) as compared with NUTS.
 
@@
 
-UNAIDS have developed ambitious fast-track targets for scaling up ART treatment with the goal of "ending the AIDS epidemic by 2030".
-Specifically, the "90-90-90" fast-track target requires that 90% of PLHIV know their status, 90% of those are on ART, and 90% of those have suppressed viral load.
-Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met $\mathbb{P}(\alpha_i > 0.81)$ for each strata $i$.
-We found that both TMB and PCA-AGHQ underestimate these exceedance probabilities, specifically for women (Figure \ref{fig:second90}).
-We hypothesise that this discrepancy is related to interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.
-
-```{r second90, fig.cap="Results of second 90 case-study."}
-knitr::include_graphics("depends/second90.png")
-```
+Ambitious fast-track targets for scaling up ART treatment have been developed by UNAIDS, with the goal of "ending the AIDS epidemic by 2030".
+Specifically, meeting the "90-90-90" fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have suppressed viral load.
+Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
+We found that for women both TMB and PCA-AGHQ underestimate these exceedance probabilities (Figure \ref{fig:exceedance}, first row).
+We hypothesise that the sex related discrepancy occurs due to interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.
 
@@
 Naomi can be used to assess the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
-We found that both TMB and PCA-AGHQ overestimate these exceedance probabilities (Figure \ref{fig:1inc}).
-We are yet to formulate hypothesis as to why this is the case.
+We found that both TMB and PCA-AGHQ overestimate these exceedance probabilities (Figure \ref{fig:exceedance}, second row).
+This is surprising, in that we expect inferences from NUTS to be more heavy-tailed than those from TMB or PCA-AGHQ.
 
-```{r 1inc, fig.cap="Results of high incidence case-study."}
-knitr::include_graphics("depends/1inc.png")
+```{r exceedance, fig.cap="Both TMB and PCA-AGHQ are meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities.
+For the second 90 target (top row) the inaccuracy varies substantially by sex."}
+knitr::include_graphics("depends/exceedance.png")
 ```
@@
 
-The exceedance probabilities case-study (Section \ref{sec:exceedance}) demonstrates the importance of accurate posterior inferences for realistic use-cases of the Naomi model.
+In the exceedance probabilities case study (Section \ref{sec:exceedance}) we found that posterior inferences from the approximate methods were systematically inaccurate, and have the potential to meaningfully mislead policy.
 
-We provide a flexible implementation of the algorithm, building on the `TMB` and `aghq` \textsc{R} packages.
-In doing so, we hope our work enables use of deterministic inference algorithms for ELGMs in applied settings, as well as further methodological exploration of their accuracy and limitations.
-Among the ELGM-type structures of particular interest in spatial epidemiology and feature in Naomi, are aggregated Gaussian process models [@nandi2020disaggregation] and evidence synthesis models [@amoah2020geostatistical].
-Although our method is designed for ELGMs, it is possible to use it outside this class, as it is compatible with any model with a `TMB` \textsc{C++} template.
+PCA-AGHQ was flexibly implemented, building on the `TMB` and `aghq` \textsc{R} packages.
+We hope that our work further encourages use of deterministic inference algorithms for ELGMs in applied settings, as well as methodological exploration of their accuracy and limitations.
+Among the ELGM-type structures of particular interest in spatial epidemiology are aggregated Gaussian process models [@nandi2020disaggregation] and evidence synthesis models [@amoah2020geostatistical].
 
-## Future work
+## Future directions
 
-Hyperparameters with skewed marginal posteriors are more important to cover well using quadrature nodes.
+Part of our motivation for developing PCA-AGHQ was to place more quadrature nodes, and as a result computation, in those dimensions which are more important, in some sense.
+While PCA is a sensible approach, there are avenues where it overlooks potential benefits, or does not behave as one might hope.
+First, we identified a challenge using PCA when the dimensions have different scales, here specifically log and logit.
+Second, the amount of variation explained for the Hessian matrix is not directly what we care about, rather the effect of the different dimensions on the outputs of interest.
+Instead, using measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] may be preferable.
+Third, it is more important to allocate quadrature nodes to those marginals which are non-Gaussian.
+This is because the Laplace approximation is exact when the integrand is Gaussian, so a single quadrature node is sufficiently.
+The difficulty is of course knowing in advance which marginals will be non-Gaussian.
+A potential approach could be based measuring fit using PSIS, or similar algorithms, with samples from a cheap approximation, like TMB.
+This particular approach would not work however, as PSIS operates on joint distributions rather than marginals [@yao2018yes].
 
-Parallel.
-
-Future work could look to implement our algorithm within probabilistic programming languages, facilitating access by a broader user-base.
-This might be possible in Stan by use of the `bridgestan` package [@bridgestan] together with the adjoint-differentiated Laplace approximation of @margossian2020hamiltonian.
+Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], may be better suited than NUTS to sampling from Naomi.
+These algorithms are available, and customisable, including e.g. choice of block structure within the `NIMBLE` probabilistic programming language [@de2017programming].
+The recently released version 1.0.0 of `NIMBLE` also includes functionality for automatic differentiation, Laplace approximation and HMC, built using `CppAD` as with `TMB`.
+As such, it would be possible to implement AGHQ within `NIMBLE`, would facilitating access by a broader user-base.
+Similarly, AGHQ could be implemented in Stan by use of the `bridgestan` package [@bridgestan] together with the adjoint-differentiated Laplace approximation of @margossian2020hamiltonian.
 <!-- Note: can fit hyperparameters with Laplace on latent field using HMC via tmbstan::tmbstan(laplace = TRUE) -->
 <!-- i.e. what Charles Margossian has implemented for Stan is done similarly by default in TMB -->
-As well, statistical theory for the algorithm could be established by extension of Theorem 1 in @stringer2022fast.
+
+Statistical theory for PCA-AGHQ could be established by extension of Theorem 1 in @stringer2022fast.
 

### 2023-06-18 src/docs_paper/paper.Rmd
Point estimates into manuscript

@@
     surname: Eaton
-    email: jeffrey.eaton@imperial.ac.uk
+    email: jeaton@hsph.harvard.edu
     label: e4
@@
     authorsLabels: e3
+  - label: D
+    name: Center for Communicable Disease Dynamics, Harvard T.H. Chan School of Public Health, Harvard University
+    authorsLabels: e4
   - label: E
     name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
-    authorsLabels: e4
-  - label: D
-    name: Center for Communicable Disease Dynamics, Harvard T.H. Chan School of Public Health, Harvard University
+
 abstract: |
@@
 keywords: 
+  - Bayesian statistics
   - spatial statistics
+  - evidence synthesis
   - small-area estimation
+  - approximate inference
   - INLA
@@
 
-We obtained satisfactory NUTS results by gradually increasing the amount of computation until all diagnostics were acceptable.
-This requried chains of length 20,000, thinned by a factor of 20 for ease-of-storage.
-All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$.
+To obtain satisfactory NUTS results we increased the Markov chain lengths until all diagnostics were acceptable.
+The results were chains of length 20,000, thinned by a factor of 20 for ease-of-storage.
+All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were then $\hat R < 1.05$.
 For full details see Appendix S2.
@@
 
-We used three statistical approaches to assess the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS: (1) Kolmogorov-Smirnov tests, (2) Pareto-smoothed importance sampling, and (3) maximum mean discrepancy.
+To compare the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS we assessed (1) marginal point estimates, (2) marginal Kolmogorov-Smirnov and Anderson-Darling tests using the empirical cumulative distribution function (ECDF), (3) joint Pareto-smoothed importance sampling results, and (4) joint maximum mean discrepancy results.
 
-### Kolmogorov-Smirnov tests
+### Point estimates
 
-The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two empirical cumulative distribution (ECDF) functions $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
-We compare the KS statistics $D_\text{TMB} = \sup_\varphi | F_\text{NUTS}(\varphi) - F_\text{TMB}(\varphi)|$ and $D_\text{PCA-AGHQ} = \sup_\varphi | F_\text{NUTS}(\varphi) - F_\text{PCA-AGHQ}(\varphi)|$.
+```{r}
+df_point <- read_csv("depends/mean-sd.csv")
+
+df_point_pct <- df_point %>%
+  ungroup() %>%
+  group_by(indicator) %>%
+  summarise(
+    rmse_diff = 100 * diff(rmse) / max(rmse),
+    mae_diff = 100 * diff(mae) / max(mae)
+  )
+
+rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate") %>% pull(rmse) %>% round(2)
+rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate") %>% pull(rmse) %>% round(2)
+rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate") %>% pull(rmse_diff) %>% round(0)
+
+rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate") %>% pull(rmse) %>% round(2)
+rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate") %>% pull(rmse) %>% round(2)
+rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate") %>% pull(rmse_diff) %>% round(0)
+```
+
+The root mean square error (RMSE) between posterior mean estimates from PCA-AGHQ and NUTS (`r rmse_ahgq_mean`) was `r abs(rmse_diff_mean)`% lower than that between TMB and NUTS (`r rmse_tmb_mean`).
+For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_ahgq_sd)` (PCA-AGHQ).
+These results, alongside those for the mean absolute error (MAE), are presented in Figure \ref{fig:mean-sd}.
+
+```{r mean-sd, fig.cap="TMT."}
+knitr::include_graphics("depends/mean-sd-comparison.png")
+```
+
+### Distribution tests
+
+The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
 See an illustration of the KS test in Figure \ref{fig:ks} and a summary of the results in Table.
@@
 
-Alongside marginal distributions, we are interested in assessing the accuracy of joint distributions.
 Let $\{\bphi_i\}_{i = 1}^n$ be joint posterior samples.
@@
 
-Joint distributions can also be compared via the maximum mean discrepancy [MMD; @gretton2006kernel].
-Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples obtained from different inference methods, and $k$ be a kernel.
-The MMD can be empirically estimated by
+Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
+The maximum mean discrepancy [MMD; @gretton2006kernel] can be empirically estimated by
 \begin{equation*}

### 2023-06-18 src/naomi-simple_exceedance/exceedance.Rmd
Add RMSE and MAE to exceedance plot

--- 2023-06-16 src/naomi-simple_exceedance/exceedance.Rmd
@@
   df_wide_second90 %>%
-    mutate(type = "Second 90"),
+    mutate(indicator = "Second 90"),
   df_wide_1inc %>%
-    mutate(type = "High incidence")
+    mutate(indicator = "High incidence")
 )
-
-write_csv(df_wide, "exceedance.csv")
 
@@
 
+df_metrics <- df_wide %>%
+  select(TMB, aghq, truth = tmbstan, indicator) %>%
+  pivot_longer(
+    cols = c("TMB", "aghq"),
+    names_to = "method",
+    values_to = "approximate"
+  ) %>%
+  group_by(indicator, method) %>%
+  summarise(
+    rmse = sqrt(mean((truth - approximate)^2)),
+    mae = mean(abs(truth - approximate))
+  ) %>%
+  mutate(method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq"))
+
+df_metrics_pct <- df_metrics %>%
+  ungroup() %>%
+  group_by(indicator) %>%
+  summarise(
+    rmse_diff = 100 * (min(rmse) - max(rmse)) / max(rmse),
+    mae_diff = 100 * (min(mae) - max(mae)) / max(mae)
+  )
+
+df_metrics <- df_metrics %>%
+  left_join(
+    mutate(df_metrics_pct, method = "PCA-AGHQ"),
+    by = c("indicator", "method")
+  ) %>%
+  mutate(
+    label = ifelse(
+      method == "PCA-AGHQ",
+      paste0("RMSE: ", signif(rmse, 2), " (", signif(rmse_diff, 2), "%)", "\nMAE: ", signif(mae, 2), " (", signif(mae_diff, 2), "%)"),
+      paste0("RMSE: ", signif(rmse, 2), "\nMAE: ", signif(mae, 2))
+    )
+  ) %>%
+  ungroup()
+
+write_csv(df_metrics, "exceedance.csv")
+
 exceedance_manuscript_plot <- df_wide %>%
@@
     method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq"),
-    type = fct_relevel(type, "Second 90", "High incidence")
+    indicator = fct_relevel(indicator, "Second 90", "High incidence")
   ) %>%
-  ggplot(aes(x = tmbstan, y = estimate, col = sex)) +
-    geom_point(shape = 1, alpha = 0.6) +
+  ggplot(aes(x = tmbstan, y = estimate)) +
+    geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
     coord_fixed(ratio = 1) +
@@
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
-    facet_grid(type ~ method) +
-    labs(x = "", y = "", col = "Sex") +
+    facet_grid(indicator ~ method) +
+    labs(x = "NUTS", y = "", col = "Sex") +
     theme_minimal() +
@@
 
-exceedance_manuscript_plot
+exceedance_manuscript_plot <- exceedance_manuscript_plot +
+  geom_text(
+    data = df_metrics %>% mutate(method = fct_relevel(method, "TMB", "PCA-AGHQ")),
+    aes(x = -Inf, y = Inf, label = label),
+    size = 3, hjust = 0, vjust = 1.5
+  )
 

### 2023-06-18 src/docs_paper/paper.Rmd
Title change

@@
 ---
-title: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model
+title: Fast approximate Bayesian inference of HIV indicators using the Naomi small-area estimation model
 author:

### 2023-06-18 src/docs_paper/paper.Rmd
Short paragraph on parallel and GPUs

@@
 
+Integration of a moderate number of hyperparameters posed a particular challenge, and led to the use of quadrature grids with large number of nodes.
+However, computation at each node is independent, such that the run-time of the algorithm could potentially be significantly improved by parallel computing.
+Further computational speed-ups might be obtained using graphics processing units (GPUs) speciallised for the relevant matrix operations.
+
 Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], may be better suited than NUTS to sampling from Naomi.

### 2023-06-19 src/docs_paper/appendix.Rmd
Reworking appendix, some main text

@@
 ---
-title: Appendix to "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
+title: Appendix to "Fast approximate Bayesian inference of HIV indicators using the Naomi small-area estimation model"
 author:
@@
 
-```{=tex}
 \setcounter{table}{0}  
@@
 \renewcommand{\thesection}{S\arabic{section}}
-```
 
@@
 
-In this section we describe the simplified Naomi model [@eaton2021naomi] considered in the main text in more complete detail.
+Here, we describe the simplified Naomi model [@eaton2021naomi] in more detail.
 
@@
 Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
-In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
-We let $\sum_{\{x\}}$ be used as shorthand for $\sum_{x \in \{x\}}$, and likewise for $s$ and $a$.
-In the main text we use a more concise, but less descriptive, approach.
+We then use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
+Let $\sum_{\{x\}}$ be shorthand for $\sum_{x \in \{x\}}$, analogously for $s$ and $a$.
 
@@
 
-We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
+We use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
 Let $u$ be a generic random effect, with length $\dim(u)$.
@@
 Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
-Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean is given by
+Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean, calculated using the `survey` \textsc{R} package [@lumley2004analysis], is given by
 $$
@@
 $$
-m^{\hat \theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
+m^{\theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
 $$
-The observed number of indicator cases is then $y^{\hat \theta} = m^{\hat \theta} \cdot \hat \theta$.
-To make these computations we use the `survey` \textsc{R} package [@lumley2004analysis].
+The observed number of indicator cases is then $y^{\theta} = m^{\theta} \cdot \hat \theta$.
 
@@
 
-```{=tex}
 \begin{table}[h]
@@
 \end{tabularx}
+\caption{The numer of latent field parameters and hyperparameters in each model section.}
 \label{tab:latent-hyper-size}
-\caption{The numer of latent field parameters and hyperparameters in each model section.}
 \end{table}
-```
 
@@
 We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
-$$
+\begin{equation}
 \text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a}.
-$$
-The term $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept, and $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women.
-The terms $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, and $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$.
-The terms $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, and $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$.
-The terms $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence.
-Finally, $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
-The two BYM2 random effects are comprised of the following unit scaled spatially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components
+\label{eq:prev}
+\end{equation}
+Table \ref{tab:prev} provides a description of the terms included in Equation \ref{eq:prev}.
+
+\begin{table}[h]
+\centering
+\begin{tabularx}{\textwidth}{llX}
+\toprule
+Term & Distribution & Description \\ 
+\midrule
+$\beta^\rho_0$ & $\mathcal{N}(0, 5)$ & Intercept \\
+$\beta_{s}^{\rho, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The difference in logit prevalence for men compared to women \\
+$u^\rho_a$ & $\text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ & Age random effects for women \\
+$u_a^{\rho, s = \text{M}}$ & $\text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ & Age random effects for the difference in logit prevalence for men compared to women age $a$ \\
+$u^\rho_x$ & $\text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ & Spatial random effects for women \\
+$u_x^{\rho, s = \text{M}}$ & $\text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ & Spatial random effects for the difference in logit prevalence for men compared to women in district $x$  \\
+$u_x^{\rho, a < 15}$ & $\text{ICAR}(\sigma_{XA}^\rho)$ & Spatial random effects for the ratio of paediatric prevalence to adult women prevalence \\
+$\eta^\rho_{R_x, s, a}$ & $-$ & Fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available \\
+\bottomrule
+\end{tabularx}
+\caption{Terms included in the linear predictor for HIV prevalence (Equation \ref{eq:prev}).}
+\label{tab:prev}
+\end{table}
+
+The two BYM2 random effects $u^\rho_x$ and $u_x^{\rho, s = \text{M}}$ are comprised of the following unit scaled spatially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components, respectively
 \begin{align*}
@@
 We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
-$$
-\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda, \label{eq:hiv-inc}
-$$
-where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults and $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women.
-The term
-$$
-\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}
-$$
-is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences, and
-$$
-\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}
-$$
-is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages.
-The term $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation].
-The terms $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects.
-Finally, $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output^[Note that these outputs are the only source of age structure for this part of the model.
-Furthermore, Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80.
-A consequence is that the posterior over new infections in these age groups is exactly zero, by definition.
-We remove these identically zero posteriors from any inference comparisons.]
+\begin{equation}
+\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda.
+\label{eq:inc}
+\end{equation}
+Table \ref{tab:inc} provides a description of the terms included in Equation \ref{eq:inc}.
+
+<!-- \citet{eaton2019estimation} -->
+
+\begin{table}[h]
+\centering
+\begin{tabularx}{\textwidth}{llX}
+\toprule
+Term & Distribution & Description \\ 
+\midrule
+$\beta^\lambda_0$ & $\mathcal{N}(0, 5)$ & Intercept term proportional to the average HIV transmission rate for untreated HIV positive adults \\
+$\beta_S^{\lambda, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The log incidence rate ratio for men compared to women \\
+$\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}$ & $-$ & The HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences \\
+$\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}$ & $-$ & The ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages \\
+$\omega = 0.7$ & $-$ & Average reduction in HIV transmission rate per increase in population ART coverage fixed based on inputs to the Estimation and Projection Package (EPP) model \\
+$u_x^\lambda$ & $\mathcal{N}(0, \sigma^\lambda)$  & IID spatial random effects with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ \\
+$\eta^\lambda_{R_x, s, a}$ & $-$ & Fixed log incidence rate ratios by sex and age group calculated from Spectrum model output \\
+\bottomrule
+\end{tabularx}
+\caption{Terms included in the linear predictor for HIV incidence (Equation \ref{eq:inc}).
+Note that the only source age structure for this part of the model are $\eta^\lambda_{R_x, s, a}$.
+As Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80, the posterior over new infections in these age groups is exactly zero, by definition.
+We remove these identically zero posteriors from any later inference comparisons.}
+\label{tab:inc}
+\end{table}
 
@@
 $$
-\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
+\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi,
 $$
@@
 
-For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the most recent household survey furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
+For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the most recent household survey furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
 $$
-y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}.
+y^{\theta}_{\{x\}, \{s\}, \{a\}} = m^{\theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}.
 $$
@@
 \begin{align*}
-y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
+y^{\rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
 \rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}, \\
-y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
+y^{\alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
 \alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}, \\
-y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
+y^{\kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
 \kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
@@
 
+<!-- Create DAG representing Naomi model. This would likely go into the appendix, so use that notation -->
+
 ```{r simplified-naomi, fig.cap="Directed acyclic graph describing the simplified Naomi model (work in progress)."}
@@
 
-```{=tex}
 \begin{table}[p!]
@@
 \end{tabularx}
-\label{tab:tmb-lookup}
 \caption{Correspondence between mathematical notation and variable names used in our \texttt{TMB} code.
@@
 We use the notation \cmark \ to refer to direct dependence of the parameter on the variable, \xmark \ to refer to no dependence, and a blank entry to refer to dependence conditional on the data.}
+\label{tab:tmb-lookup}
 \end{table}
-```
 
@@
 
-```{r rhat, fig.cap="The potential scale reduction factor compared between- and within- estimates of univariate parameters.
-It is recommended against to use the results if the value is less than 1.05."}
+```{r rhat, fig.cap="The potential scale reduction factor compares between- and within- estimates of univariate parameters.
+It is recommended only to use NUTS results if the value is less than 1.05, which it is for all parameters."}
 knitr::include_graphics("depends/rhat.png")
@@
 
-```{r ratio, fig.cap="The ESS ratio measures how efficient sampling is."}
+```{r ratio, fig.cap="The efficiently, as measured by the ESS ratio, of the NUTS sampler was poor."}
 knitr::include_graphics("depends/ratio.png")
@@
 
-```{r ess, fig.cap="The effective number of samples varied substantially."}
+```{r ess, fig.cap="The effective number of samples we obtained varied substantially between parameters."}
 knitr::include_graphics("depends/ess.png")
@@
 
-# AGHQ and PCA-AGHQ
+# AGHQ and PCA-AGHQ details
+
+## Additional figures
 
@@
 
-```{r scree, fig.cap="Most variation could be explained by fewer than the full 24 dimensions."}
+```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
 knitr::include_graphics("depends/tv-plot.png")
@@
 
-```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank (24) matrix."}
+```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank matrix."}
 knitr::include_graphics("depends/reduced-rank-plot.png")
 ```
+
+## Estimated normalising constant comparison
 
@@
 
-## KS tests
+## Point estimates
+
+## Distribution tests
 
@@
 
+## Pareto-smoothed importance sampling
+
+## Maximum mean discrepancy
+
 \newpage
 
-# Laplace marginals algorithm
-
-In earlier work, we used a full Laplace approximation to the latent field marginals.
-The algorithm for obtaining this approximation is detailed here.
+# Algorithm using Laplace latent field marginals
 
@@
 \hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
-\hat{\mathbf{H}} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
+\hat{\mathbf{H}} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta}, \\
 \hat{\mathbf{H}}^{-1} &= \hat{\mathbf{L}} \hat{\mathbf{L}}^\top, \\
-\tilde p_\texttt{LA}(\btheta, \y) &= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+\tilde p_\texttt{LA}(\btheta, \y) &= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)},
 \end{align*}
@@
 \end{align*}
-2.
-Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
-Then, use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
+2. Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension.
+Adapt these nodes based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
+Use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
 \begin{equation}
@@
 3.
-For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\hat{\mathbf{H}}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
-$l \geq 4$ is recommended to enable B-spline interpolation.
-Then, for $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
+For $i \in [N]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\hat{\mathbf{H}}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
+A value of $l \geq 4$ is recommended to enable B-spline interpolation.
+For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
 \begin{align*}
@@
 \end{align*}
-where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
+where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ can be initialised at $\hat \x(\btheta)_{-i}$.
 4. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
 \begin{equation}
-\tilde p_\texttt{AQ}(x_i \, | \, \y) = \frac{\tilde p_\texttt{LA}(x_i, \y)}{\tilde p_{\texttt{AQ}}(\y)}. \label{eq:laplace-marginal}
 \end{equation}
@@
 $$
-Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence in Equation \ref{eq:evidence1} it is more numerically accurate to use the estimate
+Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence given in Equation \ref{eq:evidence1} it is more numerically accurate, and requires little extra computation, to use the estimate
 $$

### 2023-06-19 src/docs_paper/paper.Rmd
Reworking appendix, some main text

@@
 ---
-
-<!-- Remaining things to add, and or desiderata: -->
-
-<!-- * [ ] Tying together Naomi model within LGM / ELGM mathematical notation -->
-<!-- * [ ] Create DAG representing Naomi model. This would likely go into the appendix, so use that notation -->
-<!-- * [ ] Tie together connection to Laplace for AGHQ k = 1 -->
 
@@
 The index $j$ is across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
-<!-- Relevant means e.g. for alpha it's only the HIV positive individuals -->
 The observed number of outcomes are then $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
@@
 We then model aggregate ART attendance data $y^{N^\text{ART}}_I$ using a Gaussian approximation to a sum of binomials.
-This sum is over both $i \in I$ and the number of ART clients travelling from district $x'$ to $x$.
+This sum is over both strata $i \in I$ and the number of ART clients travelling from district $x'$ to $x$.
 More details regarding this part of the model are provided in Appendix S1.
@@
 We should also specify how much of an impact each of these reasons is likely to have on the difficulty of inference.
+Perhaps a table format could be good here.
 
@@
 Multivariate GHQ rules, required to integrate over $\btheta$, are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
+
+<!-- Tie together connection to Laplace for AGHQ k = 1 -->
 
@@
 
-(plot1 + plot2) / (plot3 + plot4 + plot5)
+(plot1 + plot2) / (plot4 + plot5)
 ```
@@
 
-## Laplace marginals
-
-See Appendix S4.
-
 # Application to data from Malawi\label{sec:results}
@@
 We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
-For each approach, the `TMB` C++ user-template used to specify the log-posterior was the same.
 The three approaches were:
@@
 3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
+The `TMB` C++ user-template used to specify the log-posterior was the same for each.
 The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
@@
 knitr::include_graphics("depends/nodes-samples-comparison.png")
+
+# This plot could be improved with a scale for "coverage"
 ```
 
-### Visual assessment
+### Visual inspection
 
-Overlaying the generated PCA-AGHQ nodes overlaid onto the hyperparameter marginal posteriors obtained using NUTS, we found that approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
+Overlaying the generated PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had marginals which were well covered (Figure \ref{fig:node-positions}).
 Though $12 > s = 8$ there remain many hyperparameters poorly covered.
@@
 
-```{r contraction, fig.cap="Posterior contraction."}
+```{r contraction, fig.cap="A posterior contraction tending toward 1.0 corresponds to a posterior tending towards a Dirac delta function.
+A posterior contraction of 0.0 corresponds to no change in standard deviation between prior and posterior.
+A posterior contraction less than 0.0 corresponds to a wider posterior than prior."}
 knitr::include_graphics("depends/posterior-contraction.png")
@@
 
-We developed an approximate Bayesian inference algorithm motivated by a challenging problem in small-area estimation of HIV.
-For the simplified Naomi model in Malawi (Section \ref{sec:results}) our method is demonstrated to be more accurate than TMB, and substantially faster than NUTS.
-We anticipate that our method could be added to the Naomi web interface as an alternative to TMB.
-This would enable analysts to quickly iterate over model options using a faster, less accurate inference approach, before switching to a slower, more accurate approach once they are happy with the results.
+We developed an approximate Bayesian inference algorithm, combining AGHQ with PCA, motivated by a challenging problem in small-area estimation of HIV.
+For the simplified Naomi model in Malawi (Section \ref{sec:results}) we demonstrated the method to be more accurate, across a broad range of metrics, than TMB, and substantially faster than NUTS.
+PCA-AGHQ could be added to the Naomi web interface as an alternative to TMB, enabling analysts to quickly iterate over model options using a fast inference approach, before switching to a more accurate approach once they are happy with the results.
 
-In the exceedance probabilities case study (Section \ref{sec:exceedance}) we found that posterior inferences from the approximate methods were systematically inaccurate, and have the potential to meaningfully mislead policy.
+We found that posterior exceedance probabilities (Section \ref{sec:exceedance}) from the approximate methods to be systematically inaccurate, with the potential to meaningfully mislead policy.
 
-PCA-AGHQ was flexibly implemented, building on the `TMB` and `aghq` \textsc{R} packages.
+PCA-AGHQ was implemented using the `TMB` and `aghq` \textsc{R} packages.
+
 We hope that our work further encourages use of deterministic inference algorithms for ELGMs in applied settings, as well as methodological exploration of their accuracy and limitations.
@@
 
-Part of our motivation for developing PCA-AGHQ was to place more quadrature nodes, and as a result computation, in those dimensions which are more important, in some sense.
-While PCA is a sensible approach, there are avenues where it overlooks potential benefits, or does not behave as one might hope.
-First, we identified a challenge using PCA when the dimensions have different scales, here specifically log and logit.
-Second, the amount of variation explained for the Hessian matrix is not directly what we care about, rather the effect of the different dimensions on the outputs of interest.
-Instead, using measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] may be preferable.
+### Improving the quadrature grid
+
+Part of our motivation for PCA-AGHQ was to place more quadrature nodes in those dimensions which are in-some-sense more important.
+While PCA is a sensible way to approach this goal, there are avenues where it overlooks potential benefits, or does not behave as one might hope.
+The first challenge we identified was using PCA when the dimensions have different scales. 
+Specifically, we found logit-scale hyperparamters to be systematically favoured over those on the log-scale.
+Second, the amount of variation explained for the Hessian matrix is not of directly interest, rather the effect of the different dimensions on the relevant outputs.
+Using measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] may be preferable.
 Third, it is more important to allocate quadrature nodes to those marginals which are non-Gaussian.
 This is because the Laplace approximation is exact when the integrand is Gaussian, so a single quadrature node is sufficiently.
-The difficulty is of course knowing in advance which marginals will be non-Gaussian.
-A potential approach could be based measuring fit using PSIS, or similar algorithms, with samples from a cheap approximation, like TMB.
-This particular approach would not work however, as PSIS operates on joint distributions rather than marginals [@yao2018yes].
+The difficulty is, of course, knowing in advance which marginals will be non-Gaussian.
+This could be done if there were a cheap way to obtain posterior means, which could then be compared to posterior modes obtained using optimisation.
+Another approach would be to measure the fit of marginal samples from a cheap approximation, like TMB.
+The main challenge is that the measurements have to for marginals, ruling out approaches like PSIS which operate on joint distributions [@yao2018yes].
 
-Integration of a moderate number of hyperparameters posed a particular challenge, and led to the use of quadrature grids with large number of nodes.
+### Computational speed-ups
+
+Integration over a moderate number of hyperparameters posed a challenge, and led us to use a quadrature grids with a large number of nodes.
 However, computation at each node is independent, such that the run-time of the algorithm could potentially be significantly improved by parallel computing.
@@
 
-Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], may be better suited than NUTS to sampling from Naomi.
+### Comparison to other MCMC algorithms
+
+Blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], may be better suited than NUTS to sampling from Naomi.
 These algorithms are available, and customisable, including e.g. choice of block structure within the `NIMBLE` probabilistic programming language [@de2017programming].
-The recently released version 1.0.0 of `NIMBLE` also includes functionality for automatic differentiation, Laplace approximation and HMC, built using `CppAD` as with `TMB`.
-As such, it would be possible to implement AGHQ within `NIMBLE`, would facilitating access by a broader user-base.
-Similarly, AGHQ could be implemented in Stan by use of the `bridgestan` package [@bridgestan] together with the adjoint-differentiated Laplace approximation of @margossian2020hamiltonian.
+
+### Implementation into probabilistic programming languages
+
+Though gaining in popularity, the user-base of `TMB` remains relatively small.
+Furthermore, for users unfamiliar with C++, it can be challenging to use.
+As such, it could be beneficial to implement AGHQ within other probabilistic programming languages.
+Implementation in `NIMBLE` could be relatively straightforward, as it (for version >1.0.0) includes functionality for automatic differentiation and Laplace approximation, built using `CppAD` like `TMB`.
+Similarly, implementation in Stan could be possible by use of the `bridgestan` package [@bridgestan] together with the adjoint-differentiated Laplace approximation of @margossian2020hamiltonian.
 <!-- Note: can fit hyperparameters with Laplace on latent field using HMC via tmbstan::tmbstan(laplace = TRUE) -->
-<!-- i.e. what Charles Margossian has implemented for Stan is done similarly by default in TMB -->
+<!-- What Charles Margossian has implemented for Stan is done similarly by default in TMB -->
 
-Statistical theory for PCA-AGHQ could be established by extension of Theorem 1 in @stringer2022fast.
+### Statistical theory
+
+@stringer2022fast (Theorem 1) bound the total variation error of AGHQ, establishing convergence in probability of coverage probabilities under the approximate posterior to those under the true posterior.
+It's possible that similar theory could be established for PCA-AGHQ, or more generally AGHQ with varying numbers of nodes per dimension.
+
+### Laplace marginals
+
+See Appendix S4.
 

### 2023-06-19 src/docs_paper/paper.Rmd
Create figures.R and move from paper.Rmd

@@
 ```{r echo = FALSE}
-colours <- c("#56B4E9", "#009E73", "#E69F00")
+cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
 
@@
 )
+
+source("figures.R")
 ```
@@
 The integrand is $f(\\btheta) = \\text{sn}(0.3 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.5 \\theta_1 - 0.3 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
-TMB::compile("2d.cpp")
-dyn.load(TMB::dynlib("2d"))
-
-obj <- TMB::MakeADFun(data = list(), parameters = list(theta1 = 0, theta2 = 0), DLL = "2d")
-
-grid <- expand.grid(
-  theta1 = seq(-8, 8, length.out = 400),
-  theta2 = seq(-8, 8, length.out = 400)
-)
-
-ground_truth <- cbind(grid, pdf = apply(grid, 1, function(x) exp(-1 * obj$fn(x))))
-
-opt <- nlminb(
-  start = obj$par,
-  objective = obj$fn,
-  gradient = obj$gr,
-  control = list(iter.max = 1000, trace = 0)
-)
-
-sd_out <- TMB::sdreport(
-  obj,
-  par.fixed = opt$par,
-  getJointPrecision = TRUE
-)
-
-mu <- opt$par
-cov <- sd_out$cov.fixed
-
-plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
-  geom_contour(col = "lightgrey") +
-  coord_fixed(xlim = c(-8, 8), ylim = c(-8, 8), ratio = 1) +
-  labs(x = "", y = "") +
-  theme_minimal() +
-  guides(size = FALSE) +
-  theme(
-    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
-    axis.text.y = element_blank(), axis.ticks.y = element_blank()
-  )
-
-gg <- mvQuad::createNIGrid(2, "GHe", 3)
-
-add_points <- function(plot0, gg) {
-
-  points <- mvQuad::getNodes(gg) %>%
-    as.data.frame() %>%
-    mutate(weights = mvQuad::getWeights(gg))
-
-  colnames(points) <- c("theta1", "theta2", "weights")
-
-  plot0 +
-    geom_point(
-      data = points,
-      aes(x = theta1, y = theta2, size = weights),
-      alpha = 0.8,
-      col = "#009E73",
-      inherit.aes = FALSE
-    ) +
-    scale_size_continuous(range = c(1, 2))
-}
-
-plot1 <- add_points(plot0, gg) +
-  labs(size = "")
-
-#' Adapt by the mean
-gg2 <- gg
-mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 1)
-
-plot2 <- add_points(plot0, gg2) +
-  labs(size = "")
-
-#' Adapt by the lower Cholesky
-gg3 <- gg
-mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)
-
-plot3 <- add_points(plot0, gg3) +
-  labs(size = "")
-
-#' Adapt by the spectral
-gg4 <- gg
-mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)
-
-plot4 <- add_points(plot0, gg4) +
-  labs(size = "")
-
-#' PCA-AGHQ
-gg5 <- mvQuad::createNIGrid(2, "GHe", level = c(3, 1))
-mvQuad::rescale(gg5, m = mu, C = cov, dec.type = 1)
-
-plot5 <- add_points(plot0, gg5) +
-  labs(size = "")
-
-(plot1 + plot2) / (plot4 + plot5)
+figA
 ```
@@
 
-rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate") %>% pull(rmse) %>% round(2)
-rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate") %>% pull(rmse) %>% round(2)
-rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate") %>% pull(rmse_diff) %>% round(0)
+rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate") %>% pull(rmse) %>% signif(2)
+rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate") %>% pull(rmse) %>% signif(2)
+rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate") %>% pull(rmse_diff) %>% signif(2)
 
-rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate") %>% pull(rmse) %>% round(2)
-rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate") %>% pull(rmse) %>% round(2)
-rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate") %>% pull(rmse_diff) %>% round(0)
+rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate") %>% pull(rmse) %>% signif(2)
+rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate") %>% pull(rmse) %>% signif(2)
+rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate") %>% pull(rmse_diff) %>% signif(0)
 ```
@@
 ```{r ks, fig.height = 4, fig.cap="Example KS test for one parameter."}
-df_compare <- readRDS("depends/beta_alpha.rds") %>%
-  filter(method != "adam")
-
-mean <- df_compare %>%
-  filter(method == "tmbstan") %>%
-  summarise(mean = mean(samples)) %>%
-  pull(mean) %>%
-  round(digits = 3)
-
-sd <- df_compare %>%
-  filter(method == "tmbstan") %>%
-  summarise(sd = sd(samples)) %>%
-  pull(sd) %>%
-  round(digits = 3)
-
-histogram <- df_compare %>%
-  mutate(method = fct_recode(method, "PCA-AGHQ" = "aghq", "NUTS"= "tmbstan")) %>%
-  ggplot(aes(x = samples, fill = method, col = method)) +
-  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +
-  theme_minimal() +
-  facet_grid(method~.) +
-  labs(x = "beta_alpha", y = "Density", fill = "Method") +
-  scale_color_manual(values = colours) +
-  scale_fill_manual(values = colours) +
-  theme(legend.position = "none")
-
-grid <- seq(from = min(df_compare$samples), to = max(df_compare$samples), length.out = 1000)
-
-tmb_ecdf <- stats::ecdf(filter(df_compare, method == "TMB") %>% pull(samples))
-tmb_ecdf_df <- data.frame(x = grid, ecdf = tmb_ecdf(grid), method = "TMB")
-
-aghq_ecdf <- stats::ecdf(filter(df_compare, method == "aghq") %>% pull(samples))
-aghq_ecdf_df <- data.frame(x = grid, ecdf = aghq_ecdf(grid), method = "aghq")
-
-tmbstan_ecdf <- stats::ecdf(filter(df_compare, method == "tmbstan") %>% pull(samples))
-tmbstan_ecdf_df <- data.frame(x = grid, ecdf = tmbstan_ecdf(grid), method = "tmbstan")
-
-# Add ECDF differences
-tmb_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - tmb_ecdf_df$ecdf
-aghq_ecdf_df$ecdf_diff <- tmbstan_ecdf_df$ecdf - aghq_ecdf_df$ecdf
-tmbstan_ecdf_df$ecdf_diff <- 0
-
-absmax <- function(x) x[which.max(abs(x))]
-
-ks_tmb <- absmax(tmb_ecdf_df$ecdf_diff)
-ks_aghq <- absmax(aghq_ecdf_df$ecdf_diff)
-
-ecdf_df <- bind_rows(tmb_ecdf_df, aghq_ecdf_df, tmbstan_ecdf_df)
-
-ecdf_df$method <- factor(ecdf_df$method, levels = c("TMB", "aghq", "tmbstan"))
-
-ks_labeller <- function(x) toString(round(abs(x), 2))
-
-ecdf_diff <- ggplot(ecdf_df, aes(x = x, y = ecdf_diff, col = method)) +
-  geom_line() +
-  geom_abline(intercept = ks_tmb, slope = 0, col = colours[1], linetype = "dashed", alpha = 0.8) +
-  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_tmb, label = ks_labeller(ks_tmb), col = colours[1], alpha = 0.8) +
-  geom_abline(intercept = ks_aghq, slope = 0, col = colours[2], linetype = "dashed", alpha = 0.8) +
-  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_aghq, label = ks_labeller(ks_aghq), col = colours[2], alpha = 0.8) +
-  scale_color_manual(values = colours) +
-  labs(x = "beta_alpha", y = "ECDF difference") +
-  guides(col = "none") +
-  coord_cartesian(xlim = c(min(ecdf_df$x), max(ecdf_df$x)), clip = "off") +
-  theme_minimal() +
-  theme(plot.margin = unit(c(1, 3, 1, 1), "lines"))
-
-histogram + ecdf_diff
+figB
 ```
@@
 
+```{r}
+exceedance <- read_csv("depends/exceedance.csv")
+
+```
+
 ### Meeting the second 90
@@
 Ambitious fast-track targets for scaling up ART treatment have been developed by UNAIDS, with the goal of "ending the AIDS epidemic by 2030".
-Specifically, meeting the "90-90-90" fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have suppressed viral load.
-Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
+Meeting the "90-90-90" fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have suppressed viral load.
+Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met, that is $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
 We found that for women both TMB and PCA-AGHQ underestimate these exceedance probabilities (Figure \ref{fig:exceedance}, first row).
-We hypothesise that the sex related discrepancy occurs due to interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.
+We hypothesise this discrepancy in accuracy by sex is related to interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.
 
@@
 
-Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as greater than 1% incidence per year.
+Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as higher than 1% incidence per year.
 Naomi can be used to assess the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
@@
 
-```{r exceedance, fig.cap="Both TMB and PCA-AGHQ are meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities.
-For the second 90 target (top row) the inaccuracy varies substantially by sex."}
+```{r exceedance, fig.cap="Though PCA-AGHQ does perform slightly better, both approximate inference methods are meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities.
+For the second 90 target the inaccuracy varies substantially by sex."}
 knitr::include_graphics("depends/exceedance.png")

### 2023-06-19 src/naomi-simple_exceedance/exceedance.Rmd
Should have second 90 as top row rather than high incidence

@@
   geom_text(
-    data = df_metrics %>% mutate(method = fct_relevel(method, "TMB", "PCA-AGHQ")),
+    data = df_metrics %>% mutate(
+      method = fct_relevel(method, "TMB", "PCA-AGHQ"),
+      indicator = fct_relevel(indicator, "Second 90", "High incidence")    
+    ),
     aes(x = -Inf, y = Inf, label = label),

### 2023-06-19 src/docs_paper/paper.Rmd
Update abstract

@@
 The model is provided as a tool for countries to input their data to and generate estimates during a yearly process supported by UNAIDS.
-Currently, inference is conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
+Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
 We propose a new inference method which extends adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, enabling fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, our method provides more accurate inferences than \texttt{TMB}, and is substantially faster to run to Hamiltonian Monte Carlo with the No-U-Turn sampler.
+Using data from Malawi, our method improves the accuracy of inferences across a range of metrics, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
 By extending the \texttt{aghq} \textsc{R} package we facilitate easy, flexible use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
@@
 
-```{r echo = FALSE}
+```{r results='hide', echo = FALSE}
 cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
@@
 These estimates should be timely, and at a geographic level at which health systems are planned and delivered.
-Producing granular estimates is challenging, in large part due to limitations of the available data sources.
+Producing granular estimates is challenging, in large part due to limitations of the data from available sources.
 Nationally-representative household surveys provide the most statistically reliable data, but are only conducted every five years or so in most countries, with limited sample size at the district level, due to their high costs to run.
@@
 
-```{r aghq, results='hide', fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension.
+```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension.
 Adaption occurs based on the mean and covariance matrix of the target via the Cholesky decomposition or spectral decompostion of the inverse curvature at the mode.
@@
 We set $k(\bphi_i, \bphi_j) = \exp(-\sigma \lVert \bphi_i - \bphi_j \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
-We the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080) as compared with NUTS.
+As compared with NUTS, the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080).
 
-## Case study on exceedance probabilites \label{sec:exceedance}
+<!-- This should be calculated inline! -->
+
+## Case study on exceedance probabilites\label{sec:exceedance}
 
@@
 exceedance <- read_csv("depends/exceedance.csv")
-
 ```

### 2023-06-20 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
Mostly working on MLGH presentation, some other writing

@@
 
-```{r, echo=FALSE, fig.cap="Naomi is a district-level model of HIV indicators.", out.width = '90%'}
+```{r, echo=FALSE, fig.cap="Naomi is a district-level model of HIV indicators, like prevalence, treatment coverage, and incidence.", out.width = '90%'}
 knitr::include_graphics("depends/naomi_results.png")
@@
 
-```{r, echo=FALSE, fig.cap="Web interface promotes data ownership, data use, and data quality. From \\texttt{https://naomi.unaids.org/.}", out.width = '70%'}
+```{r, echo=FALSE, fig.cap="Estimate generation via a web interface promotes data ownership, data use, and data quality. From \\texttt{https://naomi.unaids.org/.}", out.width = '70%'}
 knitr::include_graphics("naomi_user.png")
@@
 
-# Better estimates by triangulating information
+# Better estimates by integration information from many sources
 
@@
 
-# The model has a big Gaussian latent field
+# The model has a big, structured, Gaussian latent field $\x$
+
+* Fixed effects $\color{lolit}{\text{Gaussian}}$
+* Age structure $\color{lolit}{\text{AR1}}$
+* Spatial structure $\color{lolit}{\text{IID, ICAR, BYM2}}$
+
+Concatenate together as $\x$, length 467
+
+# Smaller, non-Gaussian, hyperparameters $\btheta$
+
+* Standard deviations $\color{lolit}{\text{Half-Gaussian}}$
+* BYM2 proportion parameters $\color{lolit}{\text{Beta}}$
+* AR1 autocorrelation parameters $\color{lolit}{\text{Uniform}}$
+
+Concatenate together as $\btheta$, length 24
 
@@
 
-If you input $\theta_1, \ldots, \theta_{24}$ then it'll return a 467 length mean vector $\x(\btheta)$ and $467 \times 467$ length covariance matrix $\hat{\mathbf{H}} (\btheta)$
+If you input 24 length $\btheta$ then it'll return a 467 length mean vector $\hat \x(\btheta)$ and $467 \times 467$ length covariance matrix $\hat{\mathbf{H}} (\btheta)$.
+Mean calculated using gradient based optimisation.
+Gradients, and Hessian, obtained using automatic differentiation.
 
@@
 
+# Use adaptive Gauss-Hermite quadrature to integrate over $\btheta$
+
+Quadrature method based on the theory of polynomial interpolation which:
+
+1. Works well when the integrand looks like a polynomial times a Gaussian
+2. Adapts to the particular integrand based on the mode and Hessian
+3. Is implemented by the `aghq` package [@stringer2021implementing]
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted Gauss-Hermite nodes in two dimensions with $k = 3$."}
+figA1
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Add the mode $\\mathbf{z} + \\hat \\btheta$."}
+figA2
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="You could rotate by the lower Cholesky $\\mathbf{Lz} + \\hat \\btheta$."}
+figA3
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Or you could rotate using the eigendecomposition $\\mathbf{E} \\bm{\\Lambda}^{1/2} \\mathbf{z} + \\hat \\btheta$."}
+figA4
+```
+
+# 24 hyperparameters is too many for a dense grid
+
+$k = 3$ points in 24 dimensions is not feasible
+$$
+3 \times 3 \times \cdots \times 3 = 3^{24}
+$$
+So we need to find something smaller $\color{lolit}{\text{(that still does a good job!)}}$
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="An obvious thing to try is only keeping points from the largest eigenvectors: we call this PCA-AGHQ. Corresponds to variable choice of $k$ by dimension."}
+figA5
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap=".", out.width = '90%'}
+knitr::include_graphics("depends/tv-plot.png")
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap=".", out.width = '90%'}
+knitr::include_graphics("depends/reduced-rank-plot.png")
+```
+
+# Yes, but did it work?
+
+Run NUTS^[For 3 days!] as gold-standard, then compare to TMB to PCA-AGHQ using:
+
+1. Marginal distributions $\color{lolit}{\text{point estimates, ECDF e.g. KS or AD}}$  
+2. Joint distributions $\color{lolit}{\text{PSIS, MMD}}$ 
+3. Policy relevant outcomes $\color{lolit}{\text{second 90, high incidence}}$
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="PCA-AGHQ modestly improves estimation of the posterior mean.", out.width = '80%'}
+knitr::include_graphics("depends/mean.png")
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="PCA-AGHQ substantially improves estimation of the posterior standard deviation. TMB systematically underestimates, which you'd expect.", out.width = '80%'}
+knitr::include_graphics("depends/sd.png")
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Strata probabilities of having greater than 81\\% ART coverage, and as such meeting the second 90 target.
+Both approximate methods are inaccurate for females.", out.width = '85%'}
+knitr::include_graphics("depends/exceedance-second90.png")
+```
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Strata probabilities of having greater than 1\\% HIV incidence, and as such being classified high incidence.
+Again, both approximate methods are inaccurate.", out.width = '85%'}
+knitr::include_graphics("depends/exceedance-1inc.png")
+```
+
+# Further improvements look possible
+
+1. Fix issues with scaling $\color{lolit}{\text{Logit scaled not uniformly more important than log scaled}}$  
+2. Take into account importance for outputs of interest $\color{lolit}{\text{Variance of inputs isn't really what we care about}}$  
+3. Take into account marginal skewness $\color{lolit}{\text{The more skewed, the more we should be placing lots of points}}$
+
+#
+
+```{r, echo=FALSE, message=FALSE, fig.cap="On the real scale, $[0, 1]$ hyperparameters appear to have more marginal variance than $[0, \\infty)$ hyperparameters.
+This doesn't really make them more important though.", out.width = '80%'}
+knitr::include_graphics("depends/marginal-sd.png")
+```
+
 # Get in touch to chat about
 
-1. Further directions for this research $\color{lolit}{\text{including my suggestions for short masters or PhD projects}}$  
-2. Impactful academic or industry jobs using Bayesian statistics $\color{lolit}{\text{to begin around the end of this year when I graduate (I hope!)}}$
+1. Further directions for this research $\color{lolit}{\text{e.g. suggestions for short masters or PhD projects}}$  
+2. Impactful academic or industry jobs using Bayesian statistics $\color{lolit}{\text{to begin around the end of this year when I graduate (hopefully!)}}$
+
+| Method      | Details          |
+| ----------- | ---------------- |
+| Zulip       | Adam Howes       |
+| Email       | `ath19@ic.ac.uk` |
+| Calendly    | `adamthowes`     |
 
@@
 * Code and notebooks: `github.com/athowes/elgm-inf`
-* Working paper on the way^[For sufficiently vague definition of "on the way"]: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model $\color{lolit}{\text{Adam Howes, Alex Stringer, Seth Flaxman, Jeff Eaton}}$
+* Working paper on the way^[For sufficiently vague definition of "on the way"], Any early readers greatly appreciated!
+  * Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model $\color{lolit}{\text{Adam Howes, Alex Stringer, Seth Flaxman, Jeff Eaton}}$
 
@@
 
-<!-- # Two deterministic methods -->
-
-<!-- * We use two deterministic^[In contrast to the most famous approximate Bayesian inference method, Markov chain Monte Carlo, which is fundamentally stochastic.] methods to approximate our integral -->
-
-<!-- 1. The Laplace approximation -->
-<!-- 2. Quadrature -->
-
 <!-- # The Laplace approximation -->
@@
 
-<!-- # Which parameters should we treat as Gaussian? -->
-
-<!-- * In spatio-temporal statistics we have data indexed by space and time -->
-<!-- * We use random effects also indexed by space and time to model this data -->
-<!-- * Spatio-temporal fields can be big -->
-<!-- * Willing to make assumptions about how things vary over spacetime -->
-
-<!-- \begin{center} -->
-<!-- \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}] -->
-
-<!-- $\implies$ treat the latent field parameters as Gaussian! That's the majority of the integral done. -->
-
-<!-- \end{tcolorbox} -->
-<!-- \end{center} -->
-
 <!-- # What about the hyperparameters? -->
@@
 <!-- \end{center} -->
-
-<!-- # Quadrature -->
-
-<!-- * This brings us to our other method for solving integrals deterministically -->
-<!-- * Say we have a function, then quadrature has two ingredients -->
-
-<!-- 1. Nodes: points to evaluate the function at -->
-<!-- 2. Weights: importance of function evaluation at that point -->
-
-<!-- * You evaluate the function at the nodes, then do a weighted sum to calculate your integral -->
-
-<!-- # Trapezoid rule example -->
-
-<!-- * Let's compute $\int_0^\pi \sin(x) \text{d}x = 2$ using quadrature -->
-
-<!-- ```{r} -->
-<!-- trapezoid_rule <- function(x, spacing) { -->
-<!--   # Assumes nodes are evenly spaced -->
-<!--   w <- rep(spacing, length(x)) # Weights given by space between nodes -->
-<!--   w[1] <- w[1] / 2 # Apart from the first which is halved -->
-<!--   w[length(x)] <- w[length(x)] / 2 # And the last, also halved -->
-<!--   sum(w * x) # Compute the weighted sum -->
-<!-- } -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."} -->
-<!-- finegrid <- seq(0, pi, length.out = 1000) -->
-
-<!-- plot <- data.frame(x = finegrid, y = sin(finegrid)) %>% -->
-<!--   ggplot(aes(x = x, y = y)) + -->
-<!--    geom_line(col = cbpalette[1]) + -->
-<!--    theme_minimal() + -->
-<!--    labs(x = "x", y = "sin(x)") -->
-
-<!-- trapezoid_plot <- function(N) { -->
-<!--   grid <- seq(0, pi, length.out = N) -->
-<!--   int <- trapezoid_rule(x = sin(grid), spacing = grid[2] - grid[1]) -->
-
-<!--   plot +  -->
-<!--     geom_bar( -->
-<!--       data = data.frame(x = grid, y = sin(grid)), -->
-<!--       aes(x = x, y = y), alpha = 0.7, stat = "identity", -->
-<!--       inherit.aes = FALSE, fill = cbpalette[2]) + -->
-<!--     theme_minimal() + -->
-<!--     labs( -->
-<!--       subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"), -->
-<!--       x = "x", y = "sin(x)" -->
-<!--     ) -->
-<!-- } -->
-
-<!-- trapezoid_plot(N = 10) -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."} -->
-<!-- trapezoid_plot(N = 30) -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."} -->
-<!-- trapezoid_plot(N = 100) -->
-<!-- ``` -->
-
-<!-- # Adaptive Gauss-Hermite quadrature -->
-
-<!-- * \textcolor{hilit}{Gauss-Hermite quadrature} is a method for picking nodes and weights based on the theory of polynomial interpolation -->
-<!-- * It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution -->
-<!-- * The \textcolor{hilit}{adaptive} part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data -->
-<!-- * Implemented by the `aghq` package [@stringer2021implementing] -->
-
-<!-- #  -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points in two dimensions with $k = 3$."} -->
-<!-- mu <- c(1, 1.5) -->
-<!-- cov <- matrix(c(2, 1, 1, 1), ncol = 2) -->
-
-<!-- obj <- function(theta) { -->
-<!--   mvtnorm::dmvnorm(theta, mean = mu, sigma = cov) -->
-<!-- } -->
-
-<!-- grid <- expand.grid( -->
-<!--   theta1 = seq(-2, 5, length.out = 700), -->
-<!--   theta2 = seq(-2, 5, length.out = 700) -->
-<!-- ) -->
-
-<!-- ground_truth <- cbind(grid, pdf = obj(grid)) -->
-
-<!-- plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) + -->
-<!--   geom_contour(col = cbpalette[1]) + -->
-<!--   coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) + -->
-<!--   labs(x = "", y = "") + -->
-<!--   theme_minimal() -->
-
-<!-- gg <- mvQuad::createNIGrid(2, "GHe", 3) -->
-
-<!-- add_points <- function(plot0, gg) { -->
-<!--   plot0 + -->
-<!--     geom_point( -->
-<!--       data = mvQuad::getNodes(gg) %>% -->
-<!--               as.data.frame() %>% -->
-<!--               mutate(weights = mvQuad::getWeights(gg)),  -->
-<!--       aes(x = V1, y = V2, size = weights), -->
-<!--       alpha = 0.8, -->
-<!--       col = cbpalette[2], -->
-<!--       inherit.aes = FALSE -->
-<!--     ) + -->
-<!--     scale_size_continuous(range = c(1, 2)) -->
-<!-- } -->
-
-<!-- add_points(plot0, gg) + -->
-<!--   labs(size = "Weight", caption = "") -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean $z + \\hat \\theta$."} -->
-<!-- gg2 <- gg -->
-<!-- mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2) -->
-
-<!-- add_points(plot0, gg2) + -->
-<!--   labs(size = "Weight", caption = "") -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $Lz + \\hat \\theta$."} -->
-<!-- gg3 <- gg -->
-<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2) -->
-
-<!-- add_points(plot0, gg3) + -->
-<!--   labs(size = "Weight", caption = "") -->
-<!-- ``` -->
-
-<!-- # -->
-
-<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2} z + \\hat \\theta$."} -->
-<!-- gg3 <- gg -->
-<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1) -->
-
-<!-- add_points(plot0, gg3) + -->
-<!--   labs(size = "Weight", caption = "") -->
-<!-- ``` -->

### 2023-06-20 src/docs_paper/paper.Rmd
Mostly working on MLGH presentation, some other writing

@@
 
-```{r results='hide', echo = FALSE}
+```{r echo = FALSE}
 cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
@@
 )
+```
 
+```{r results='hide'}
 source("figures.R")
@@
 ```{r mean-sd, fig.cap="TMT."}
-knitr::include_graphics("depends/mean-sd-comparison.png")
+knitr::include_graphics("depends/mean-sd.png")
 ```

### 2023-06-20 src/naomi-simple_exceedance/exceedance.Rmd
Mostly working on MLGH presentation, some other writing

@@
 
-ggsave("exceedance.png", exceedance_manuscript_plot, h = 6.5, w = 6.25, bg = "white")
+ggsave("exceedance.png", exceedance_manuscript_plot, h = 6.5, w = 6.25)
+
+#' Separate versions are nice for presentations etc.
+
+exceedance_manuscript_plot_second90 <- df_wide %>%
+  filter(indicator == "Second 90") %>%
+  pivot_longer(cols = c("TMB", "aghq"), names_to = "method", values_to = "estimate") %>%
+  mutate(
+    method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq")
+  ) %>%
+  ggplot(aes(x = tmbstan, y = estimate)) +
+    geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
+    coord_fixed(ratio = 1) +
+    scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    geom_text(
+    data = df_metrics %>%
+      filter(indicator == "Second 90") %>%
+      mutate(method = fct_relevel(method, "TMB", "PCA-AGHQ")),
+      aes(x = -Inf, y = Inf, label = label),
+      size = 3, hjust = 0, vjust = 1.5
+    ) +
+    facet_grid(indicator ~ method) +
+    labs(x = "NUTS", y = "", col = "Sex") +
+    theme_minimal()
+
+ggsave("exceedance-second90.png", exceedance_manuscript_plot_second90, h = 4, w = 6.25)
+
+exceedance_manuscript_plot_1inc <- df_wide %>%
+  filter(indicator == "High incidence") %>%
+  pivot_longer(cols = c("TMB", "aghq"), names_to = "method", values_to = "estimate") %>%
+  mutate(
+    method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq")
+  ) %>%
+  ggplot(aes(x = tmbstan, y = estimate)) +
+    geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
+    coord_fixed(ratio = 1) +
+    scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    geom_text(
+    data = df_metrics %>%
+      filter(indicator == "High incidence") %>%
+      mutate(method = fct_relevel(method, "TMB", "PCA-AGHQ")),
+      aes(x = -Inf, y = Inf, label = label),
+      size = 3, hjust = 0, vjust = 1.5
+    ) +
+    facet_grid(indicator ~ method) +
+    labs(x = "NUTS", y = "", col = "Sex") +
+    theme_minimal()
+
+ggsave("exceedance-1inc.png", exceedance_manuscript_plot_1inc, h = 4, w = 6.25)
 ```

### 2023-06-20 src/naomi-simple_contraction/script.R

@@ -131,11 +131,42 @@ mean_sd_plot <- ggplot(df_plot, aes(x = truth, y = approximate)) +
   facet_grid(indicator ~ method) +
   coord_fixed(ratio = 1) +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  geom_text(data = df_metrics, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
   labs(x = "NUTS", y = "") +
   theme_minimal()
 
-mean_sd_plot <- mean_sd_plot +
-  geom_text(data = df_metrics, aes(x = -Inf, y = Inf, label = label),
-            size = 3, hjust = 0, vjust = 1.5)
+ggsave("mean-sd.png", mean_sd_plot, h = 6, w = 6.25)
 
-ggsave("mean-sd-comparison.png", mean_sd_plot, h = 6.25, w = 6.25, bg = "white")
+#' Split into two plots for presentations etc.
+
+mean_plot <- df_plot %>%
+  filter(indicator == "Posterior mean estimate") %>%
+  ggplot(aes(x = truth, y = approximate)) +
+  geom_point(shape = 1, alpha = 0.4) +
+  facet_grid(indicator ~ method) +
+  coord_fixed(ratio = 1) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  geom_text(
+    data = filter(df_metrics, indicator == "Posterior mean estimate"),
+    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
+  ) +
+  labs(x = "NUTS", y = "") +
+  theme_minimal()
+
+ggsave("mean.png", mean_plot, h = 4, w = 6.25)
+
+sd_plot <- df_plot %>%
+  filter(indicator == "Posterior SD estimate") %>%
+  ggplot(aes(x = truth, y = approximate)) +
+  geom_point(shape = 1, alpha = 0.4) +
+  facet_grid(indicator ~ method) +
+  coord_fixed(ratio = 1) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  geom_text(
+    data = filter(df_metrics, indicator == "Posterior SD estimate"),
+    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
+  ) +
+  labs(x = "NUTS", y = "") +
+  theme_minimal()
+
+ggsave("sd.png", sd_plot, h = 4, w = 6.25)

### 2023-06-20 src/docs_paper/figures.R
Increment uploaded paper and appendix

@@ -1,3 +1,5 @@
+cols <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
 #' Fig A
 
 TMB::compile("2d.cpp")
@@ -117,8 +119,8 @@ histogram <- df_compare %>%
   theme_minimal() +
   facet_grid(method~.) +
   labs(x = "beta_alpha", y = "Density", fill = "Method") +
-  scale_color_manual(values = cbpalette) +
-  scale_fill_manual(values = cbpalette) +
+  scale_color_manual(values = cols) +
+  scale_fill_manual(values = cols) +
   theme(legend.position = "none")
 
 grid <- seq(from = min(df_compare$samples), to = max(df_compare$samples), length.out = 1000)
@@ -150,11 +152,11 @@ ks_labeller <- function(x) toString(signif(abs(x), 2))
 
 ecdf_diff <- ggplot(ecdf_df, aes(x = x, y = ecdf_diff, col = method)) +
   geom_line() +
-  geom_abline(intercept = ks_tmb, slope = 0, col = cbpalette[1], linetype = "dashed", alpha = 0.8) +
-  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_tmb, label = ks_labeller(ks_tmb), col = cbpalette[1], alpha = 0.8) +
-  geom_abline(intercept = ks_aghq, slope = 0, col = cbpalette[2], linetype = "dashed", alpha = 0.8) +
-  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_aghq, label = ks_labeller(ks_aghq), col = cbpalette[2], alpha = 0.8) +
-  scale_color_manual(values = cbpalette) +
+  geom_abline(intercept = ks_tmb, slope = 0, col = cols[1], linetype = "dashed", alpha = 0.8) +
+  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_tmb, label = ks_labeller(ks_tmb), col = cols[1], alpha = 0.8) +
+  geom_abline(intercept = ks_aghq, slope = 0, col = cols[2], linetype = "dashed", alpha = 0.8) +
+  annotate("text", x = 1.1 * max(ecdf_df$x), y = ks_aghq, label = ks_labeller(ks_aghq), col = cols[2], alpha = 0.8) +
+  scale_color_manual(values = cols) +
[…]

### 2023-06-21 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
Finish slides #40

@@
 
-```{r, echo=FALSE, fig.cap="Naomi is a district-level model of HIV indicators, like prevalence, treatment coverage, and incidence.", out.width = '90%'}
+```{r, echo=FALSE, fig.cap="Naomi is a age-sex-district model of HIV indicators, like prevalence, treatment coverage, and incidence, for countries in sub-Saharan Africa.", out.width = '90%'}
 knitr::include_graphics("depends/naomi_results.png")
@@
 
-```{r, echo=FALSE, fig.cap="Estimate generation via a web interface promotes data ownership, data use, and data quality. From \\texttt{https://naomi.unaids.org/.}", out.width = '70%'}
+```{r, echo=FALSE, fig.cap="Estimates are generated yearly via a web interface. This promotes data ownership, data use, and data quality. From \\texttt{https://naomi.unaids.org/.}", out.width = '70%'}
 knitr::include_graphics("naomi_user.png")
@@
 
-# Better estimates by integration information from many sources
+# Integration of information from many sources
 
@@
 
-# Put together, it's a challenging Bayesian inference problem
+# A challenging Bayesian inference problem
 
@@
 \lolit{
-
 If you input 24 length $\btheta$ then it'll return a 467 length mean vector $\hat \x(\btheta)$ and $467 \times 467$ length covariance matrix $\hat{\mathbf{H}} (\btheta)$.
@@
 Gradients, and Hessian, obtained using automatic differentiation.
+}
 
-}
+Optimise the resulting Laplace approximation
+
+$$
+\hat{\btheta}_{\texttt{LA}} = \argmax_{\btheta} \tilde p_{\texttt{LA}}(\btheta, \y) = \argmax_{\btheta} \frac{p(\y, \x, \btheta)}{\tilde p_{\texttt{G}}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat{\x}(\btheta)}
+$$
 
@@
 3. Is implemented by the `aghq` package [@stringer2021implementing]
+
+\lolit{
+
+$$
+p(\y) \approx \int_{\btheta} \tilde p_{\texttt{LA}}(\btheta, \y) \text{d}\btheta \approx \sum_{\z \in \mathcal{Q}} p_{\texttt{LA}}(\z, \y) \omega(\z)
+$$
+where $\z \in \mathcal{Q}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
+
+}
 
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap=".", out.width = '90%'}
+```{r, echo=FALSE, message=FALSE, fig.cap="Including 8 dimensions you can explain close to 90\\% of the total variation.", out.width = '90%'}
 knitr::include_graphics("depends/tv-plot.png")
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap=".", out.width = '90%'}
+```{r, echo=FALSE, message=FALSE, fig.cap="If you reconstruct the Hessian with just those 8 dimensions, it looks pretty similar.", out.width = '90%'}
 knitr::include_graphics("depends/reduced-rank-plot.png")
@@
 
-1. Further directions for this research $\color{lolit}{\text{e.g. suggestions for short masters or PhD projects}}$  
-2. Impactful academic or industry jobs using Bayesian statistics $\color{lolit}{\text{to begin around the end of this year when I graduate (hopefully!)}}$
+1. You have suggestions to improve the work done so far!
+2. Further directions for this research $\color{lolit}{\text{e.g. suggestions for short masters or PhD projects}}$  
+3. Impactful academic or industry jobs using Bayesian statistics $\color{lolit}{\text{to begin around the end of this year when I graduate (hopefully!)}}$
 
@@
   * Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model $\color{lolit}{\text{Adam Howes, Alex Stringer, Seth Flaxman, Jeff Eaton}}$
+
+#
+
+```{r, echo=FALSE, fig.cap="Custom version of Laplace marginals implemented using \\texttt{TMB}.
+Speed-up possible using Simon Wood's method?
+For the Epilepsy example from Rue (2009), see \\texttt{athowes/elgm-inf/src/epil}.", out.width = '60%'}
+knitr::include_graphics("laplace.png")
+```
 

### 2023-06-22 src/docs_paper/appendix.Rmd
Mostly working on MLGH presentation, some other writing
Lots of Seth fixes, change paper name, change repo name #55

@@
 ---
-title: Appendix to "Fast approximate Bayesian inference of HIV indicators using the Naomi small-area estimation model"
+title: Appendix to "Fast approximate Bayesian inference of HIV indicators using principal components analysis adaptive Gauss-Hermite quadrature"
 author:
@@
 
-<!-- Had previously been getting a figure of 491 somewhere -- which is 51 + 14n + 24. Something going on there? -->
+<!-- 51 + 13n + 467 is the value I get from the code. 467 + 24 = 491 total -->
+<!-- Need to check the discrepency -->
 

### 2023-06-22 src/docs_paper/paper.Rmd
Increment uploaded paper and appendix
Lots of Seth fixes, change paper name, change repo name #55

@@
 ---
-title: Fast approximate Bayesian inference of HIV indicators using the Naomi small-area estimation model
+title: Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature
 author:
@@
 Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
-We propose a new inference method which extends adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, enabling fast and accurate inference for Naomi and other extended latent Gaussian models.
+We propose a new inference method extending adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, enabling fast and accurate inference for Naomi and other extended latent Gaussian models.
 Using data from Malawi, our method improves the accuracy of inferences across a range of metrics, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
-By extending the \texttt{aghq} \textsc{R} package we facilitate easy, flexible use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
+Our implementation uses the \texttt{aghq} \textsc{R} package, facilitating easy, flexible use of the method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
 keyword-subclass: | 
@@
 ```{r echo = FALSE}
-cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
-
 options(scipen = 100)
@@
 Nationally-representative household surveys provide the most statistically reliable data, but are only conducted every five years or so in most countries, with limited sample size at the district level, due to their high costs to run.
-Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in closer to real-time, but based on limited or non-representative samples of the population.
+Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in closer to real-time, but are not representative of the entire population.
 To address these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
-Modelling multiple data sources jointly mitigates the limitations of any single source, increases statistical power, and can prompts investigation into any conflicts of information.
+Modelling multiple data sources jointly mitigates the limitations of any single source, increases statistical power, and can prompt investigation into any conflicts of information.
 
@@
 Naomi is a complex model, and as such presents a challenging Bayesian inference problem.
-As well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number typically required for use of integrated nested Laplace approximations [INLA; @rue2009approximate].
+As well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number that can typically be handled by integrated nested Laplace approximations [INLA; @rue2009approximate].
 Moreover, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
@@
 
-To allow for interactive review and iteration of model results, the inference procedure should be fast for workshop participants to fit the model on a laptop.
-Due to the scale of the model and features of its posterior geometry [@neal2003slice], Markov chain Monte Carlo (MCMC) approaches are therefore prohibitively slow.
-Furthermore, it is preferable that use of the inference method across countries to be automatic, and not require substantial statistical expertise, as would be the case for monitoring MCMC convergence and suitability.
+To allow for interactive review and iteration of model results, the inference procedure should be fast and low memory usage enough for workshop participants to fit the model.
+Due to the scale of the model and features of its posterior geometry [@neal2003slice], Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
+Furthermore, it is preferable that use of the inference method across countries is automatic, and doesn't require substantial statistical expertise, as would be the case for monitoring MCMC convergence and suitability.
 
 To meet these requirements, inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-Owing to its speed and flexibility, `TMB` is gaining popularity more broadly, particularly in spatial statistics [@osgood2022statistical].
+Owing to its speed and flexibility, `TMB` is gaining popularity, particularly in spatial statistics [@osgood2022statistical], and via the `glmmTMB` \textsc{R} package [@brooks2017glmmTMB] which provides a user-friendly formula interface for fitting common models.
 Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the parameters.
@@
 This consideration motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
-To obtain fast, accurate Bayesian inferences for the Naomi model we developed an inference method which extends adaptive Gauss-Hermite quadrature to handle many hyperparameters.
-We used principal components analysis to create a computationally feasible quadrature grid.
+To obtain fast, accurate Bayesian inferences for the Naomi model we developed an inference method which extends adaptive Gauss-Hermite quadrature (AGHQ) to handle many hyperparameters.
+AGHQ is a quadrature method based on the theory of polynomial interpolation, which is well suited to statistical estimation problems.
+@bilodeau2022stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
+It is not computationally feasible to use AGHQ in high dimensions directly, as exponentially many nodes would be required.
+Instead, we use principal components analysis (PCA) of the Hessian at the mode to find a smaller number of dimensions which explain most of the variance.
+In our application, this results in a grid which has millions of times fewer nodes the corresponding dense grid.
 Our method is implemented as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
@@
 Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference for ELGMs using simplified INLA and AGHQ.
-As a case study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:naomi}.
+As a case study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:results}.
 We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
@@
 For ease of notation, let $i$ index the finest district-age-sex division included in the model.
-Let $I \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported, with $I \in \mathcal{I}$.
+Let $I \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported, and $\mathcal{I}$ be the set of all $I$.
 
 Let $N_i \in \mathbb{N}$ be the known, fixed population size.
-We infer the following unknown HIV indicators using linked regression equations: HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
+We infer the following unknown HIV indicators using linked regression equations:
+
+* HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive;
+* antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and
+* annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
+
 Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that $\text{logit}(\rho_i) = \eta^\rho_i$ and $\text{logit}(\alpha_i) = \eta^\alpha_i$, for certain choice of structured additive predictors.
@@
 Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
-The vector $\btheta_1 \in \mathbb{R}^s$, with $s_1$ assumed small, are additional parameters of the likelihood.
+The vector $\btheta_1 \in \mathbb{R}^{s_1}$, with $s_1$ assumed small, are additional parameters of the likelihood.
 The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
@@
 = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
-= \sqrt{\frac{\det(\hat {\Hb}(\btheta))}{(2 \pi)^{N}}} p(\y, \hat \x(\btheta), \btheta). \label{eq:la}
+= \sqrt{\frac{ \lvert \hat {\Hb}(\btheta) \rvert}{(2 \pi)^{N}}} p(\y, \hat \x(\btheta), \btheta). \label{eq:la}
 \end{equation}
@@
 GHQ is attractive because it is exact for functions which are a polynomial of total order no more than $2k - 1$ multiplied by a Gaussian density.
-We typically expected posterior distributions to be relatively well described by this class of functions.
-Multivariate GHQ rules, required to integrate over $\btheta$, are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
+Posterior distributions can typically be expected to be relatively well described by this class of functions.
+Multivariate GHQ rules, required to integrate over $\btheta$, are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ [@jackel2005note].
 
@@
 
-```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension.
-Adaption occurs based on the mean and covariance matrix of the target via the Cholesky decomposition or spectral decompostion of the inverse curvature at the mode.
+```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension (A).
+Adaption occurs based on the mode (B) and covariance matrix of the target via the Cholesky decomposition (C) or spectral decompostion (D) of the inverse curvature at the mode.
+In PCA-AGHQ (E) only nodes along the first $s$ principal components are kept.
 The integrand is $f(\\btheta) = \\text{sn}(0.3 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.5 \\theta_1 - 0.3 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
@@
 It is however, still possible to calculate the normalising constant using a sparse rule, which we do so in Appendix S3, finding it to be related to that from PCA-AGHQ as follows.
+
+<!-- Pruning (Jackel, 2005): is this something we could be doing? -->
 
@@
 To obtain satisfactory NUTS results we increased the Markov chain lengths until all diagnostics were acceptable.
-The results were chains of length 20,000, thinned by a factor of 20 for ease-of-storage.
+This required chains of length 20,000, thinned by a factor of 20 for ease-of-storage.
 All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were then $\hat R < 1.05$.
@@
 We used a Scree plot (Figure S$X$) based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})$ to select the number of principal components $s = 8 < m = 24$ to keep, sufficient to explain close to 90% of total variation.
-This resulted in a close reconstruction of the Hessian using the reduced rank approximation (Figure S$X$).
+With this choice of $s$, the reduced rank approximation to the Hessian is visually very similar (Figure S$X$).
 
-```{r node-positions, fig.cap="PCA-AGHQ node positions (green, rug plot) overlaid onto the hyperparameter marginal posteriors (grey, histogram) for each of the 24 hyperparameters."}
+```{r node-positions, fig.cap="The 6561 PCA-AGHQ node positions (green, rug plot) projected onto the hyperparameter marginal posteriors (grey, histogram) for each of the 24 hyperparameters.
+Some hyperparameters, such as \\texttt{logit_phi_alpha_x}, are well covered, where as others, such as \\texttt{log_sigma_lambda_x}, are near only being covered by one unique node."}
 knitr::include_graphics("depends/nodes-samples-comparison.png")
@@
 
-Overlaying the generated PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had marginals which were well covered (Figure \ref{fig:node-positions}).
-Though $12 > s = 8$ there remain many hyperparameters poorly covered.
+Overlaying the generated $3^8 = 6561$ PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
+Though 12 is an improvment on the 8 that would be naively achieved using a dense grid, there remained many hyperparameters poorly covered.
 Coverage was associated with marginal standard deviation (Figure S$X$), which varied particularly according to the hyperparameter scale.
@@
 
-```{r mean-sd, fig.cap="TMT."}
-knitr::include_graphics("depends/mean-sd.png")
+(ref:mean-sd) Barplot displaying frequency of dependencies for all of the R packages in the CRAN repository to date `r Sys.Date()
+
+```{r mean-sd, fig.cap="."}
+knitr::include_graphics("depends/mean-sd-alt.png")
 ```
@@
 
-## Case study on exceedance probabilites\label{sec:exceedance}
+## Case study on exceedance probabilities\label{sec:exceedance}
 
@@
 For the simplified Naomi model in Malawi (Section \ref{sec:results}) we demonstrated the method to be more accurate, across a broad range of metrics, than TMB, and substantially faster than NUTS.
-PCA-AGHQ could be added to the Naomi web interface as an alternative to TMB, enabling analysts to quickly iterate over model options using a fast inference approach, before switching to a more accurate approach once they are happy with the results.
+
+PCA-AGHQ could be added to the Naomi web interface as an alternative to TMB.
+Analysts may then quickly iterate over model options using a fast inference approach, before switching to a more accurate approach once they are happy with the results.
+By selecting $s$ and $k$, PCA-AGHQ can be adjusted to suit the computational budget available. 
+We selected $s$ based on the Scree plot, and for the most part fixed $k = 3$.
+Whether it is preferable, for a given computational budget, to increase $s$ or increase $k$ remains an open question.
+Further strategies, such as gradually lowering $k$ over the principal components, could also be considered.
+
+@bilodeau2022stochastic highlight developing computationally feasible quadrature methods for high dimensions as a challenging open problem.
 
@@
 
-Part of our motivation for PCA-AGHQ was to place more quadrature nodes in those dimensions which are in-some-sense more important.
-While PCA is a sensible way to approach this goal, there are avenues where it overlooks potential benefits, or does not behave as one might hope.
+We aimed to develop a quadrature grid which allocates more effort to more important dimensions.
+While PCA is a sensible approach, there are avenues where it does not behave as one might hope, or otherwise overlooks potential benefits.
 The first challenge we identified was using PCA when the dimensions have different scales. 

### 2023-06-22 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Work on slides for HIV Lab Group #25
Replace all elgm-inf with naomi-aghq

@@
 \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-Notebook for this example at \texttt{athowes.github.io/elgm-inf/epil}
+Notebook for this example at \texttt{athowes.github.io/naomi-aghq/epil}
 \end{tcolorbox}
@@
 \begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
-Notebook for this example at \texttt{athowes.github.io/elgm-inf/prev-anc-art}
+Notebook for this example at \texttt{athowes.github.io/naomi-aghq/prev-anc-art}
 \end{tcolorbox}
@@
 * Joint work with Alex Stringer (Waterloo) and my PhD supervisors Seth Flaxman (Oxford) and Jeff Eaton (Imperial)
-* The code for this project is at `github.com/athowes/elgm-inf`
+* The code for this project is at `github.com/athowes/naomi-aghq`
 * You can find me online at `athowes.github.io`

### 2023-06-22 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Replace all elgm-inf with naomi-aghq

@@
 * Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
-* Code and notebooks for this project are available at `athowes.github.io/elgm-inf`
+* Code and notebooks for this project are available at `athowes.github.io/naomi-aghq`
 

### 2023-06-22 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
Replace all elgm-inf with naomi-aghq

@@
 
-<!-- https://github.com/athowes/elgm-inf/blob/fd4b0bb6b9ba93ac277de7aad47a94581e8fc900/src/docs_18-04-23-explainer/18-04-23-explainer.Rmd -->
+<!-- https://github.com/athowes/naomi-aghq/blob/fd4b0bb6b9ba93ac277de7aad47a94581e8fc900/src/docs_18-04-23-explainer/18-04-23-explainer.Rmd -->
 
@@
 
-* Code and notebooks: `github.com/athowes/elgm-inf`
+* Code and notebooks: `github.com/athowes/naomi-aghq`
 * Working paper on the way^[For sufficiently vague definition of "on the way"], Any early readers greatly appreciated!
@@
 Speed-up possible using Simon Wood's method?
-For the Epilepsy example from Rue (2009), see \\texttt{athowes/elgm-inf/src/epil}.", out.width = '60%'}
+For the Epilepsy example from Rue (2009), see \\texttt{athowes/naomi-aghq/src/epil}.", out.width = '60%'}
 knitr::include_graphics("laplace.png")

### 2023-06-22 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Replace all elgm-inf with naomi-aghq

@@
 
-**Interested?** Working notebooks and R code available from `github.com/athowes/elgm-inf`. Or get in touch:
+**Interested?** Working notebooks and R code available from `github.com/athowes/naomi-aghq`. Or get in touch:
 

### 2023-06-22 src/example_inla-replication/inla-replication.Rmd
Replace all elgm-inf with naomi-aghq

@@
 Under this approach, all weights would be equal $\Delta_k = 1 / K$.
-For a more detailed run through of the grid integration method, see the [INLA two dimensional grid replication](https://athowes.github.io/elgm-inf/inla-grid.html) notebook.
+For a more detailed run through of the grid integration method, see the [INLA two dimensional grid replication](https://athowes.github.io/naomi-aghq/inla-grid.html) notebook.
 

### 2023-06-22 src/prev-anc-art_results/prev-anc-art.Rmd
Slides up!
Replace all elgm-inf with naomi-aghq

@@
 |:------|:-------|:-----------|
-| 0 | [`prev-anc-art_model0`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model0) | Prevalence (no random effects) |
-| 1 | [`prev-anc-art_model1`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model1) | Prevalence |
-| 2 | [`prev-anc-art_model2`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model2) | Prevalence, ANC |
-| 3 | [`prev-anc-art_model3`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model3) | Prevalence, ART |
-| 4 | [`prev-anc-art_model4`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_model4) | Prevalence, ANC, ART |
+| 0 | [`prev-anc-art_model0`](https://github.com/athowes/naomi-aghq/tree/master/src/prev-anc-art_model0) | Prevalence (no random effects) |
+| 1 | [`prev-anc-art_model1`](https://github.com/athowes/naomi-aghq/tree/master/src/prev-anc-art_model1) | Prevalence |
+| 2 | [`prev-anc-art_model2`](https://github.com/athowes/naomi-aghq/tree/master/src/prev-anc-art_model2) | Prevalence, ANC |
+| 3 | [`prev-anc-art_model3`](https://github.com/athowes/naomi-aghq/tree/master/src/prev-anc-art_model3) | Prevalence, ART |
+| 4 | [`prev-anc-art_model4`](https://github.com/athowes/naomi-aghq/tree/master/src/prev-anc-art_model4) | Prevalence, ANC, ART |
 
-We perform inference on simulated data from Model 4 generated ([`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim)) with the following parameter values:
+We perform inference on simulated data from Model 4 generated ([`prev-anc-art_sim`](https://github.com/athowes/naomi-aghq/tree/master/src/prev-anc-art_sim)) with the following parameter values:
 
@@
 draw_ksplots_D(model4) + labs(title = "Model 4")
+
+ggsave("ks-example.pdf", draw_ksplots_D_params(model4, params = "phi_prev[5]"), h = 4, w = 6.25)
 ```

### 2023-06-22 src/dev_sinla/sinla.Rmd
Add warning disclaimer
Missed from previous commit

@@
 
+**Warning: this report was created during the process of learning, and for that reason is not concise or well-structured.
+I would not recommend you reading it, but if you do proceed at your own caution!**
+
 ```{r echo=FALSE}
@@
 `aghq::sample_marginal` draws samples from this Gaussian, given values of $z$ drawn from a multinomial distribution, in order to evaluate Equation \@ref(eq:pxigiveny).
-For a line-by-line walkthrough of how these functions work, see the notebook [Code walkthrough](https://athowes.github.io/elgm-inf/walkthrough.html) (which uses the model from the notebook [Epilepsy example](https://athowes.github.io/elgm-inf/epil.html)).
-
-The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/elgm-inf/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $\tilde p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
+For a line-by-line walkthrough of how these functions work, see the notebook [Code walkthrough](https://athowes.github.io/naomi-aghq/walkthrough.html) (which uses the model from the notebook [Epilepsy example](https://athowes.github.io/naomi-aghq/epil.html)).
+
+The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/naomi-aghq/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $\tilde p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
 $$
@@
 
-We will use Model 1 from the [Prevalence, ANC, ART model inference case-study](https://athowes.github.io/elgm-inf/prev-anc-art.html) as a running example.
+We will use Model 1 from the [Prevalence, ANC, ART model inference case-study](https://athowes.github.io/naomi-aghq/prev-anc-art.html) as a running example.
 To provide a ground truth to compare the methods that follow to, we generate "gold-standard" inferences using a relatively long run of MCMC as implemented by `tmbstan`.
 
-We begin by loading in testing data, as generated by [`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim).
+We begin by loading in testing data, as generated by [`prev-anc-art_sim`](https://github.com/athowes/naomi-aghq/tree/master/src/prev-anc-art_sim).
 `sim_data` contains many replicates of simulated data.

### 2023-06-22 src/docs_paper/paper.Rmd
Replace all elgm-inf with naomi-aghq
Upload paper and appendix new versions #55

@@
 Moreover, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
-As such, the model is not easily specified by any software which requires a formula interface.
 
@@
 Example model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
-The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
+The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/naomi-aghq`.
 We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
@@
 ```{r node-positions, fig.cap="The 6561 PCA-AGHQ node positions (green, rug plot) projected onto the hyperparameter marginal posteriors (grey, histogram) for each of the 24 hyperparameters.
-Some hyperparameters, such as \\texttt{logit_phi_alpha_x}, are well covered, where as others, such as \\texttt{log_sigma_lambda_x}, are near only being covered by one unique node."}
+Some hyperparameters, such as logit\\_phi\\_alpha\\_x, are well covered where as others, such as log\\_sigma\\_lambda\\_x, are near only being covered by one unique node."}
 knitr::include_graphics("depends/nodes-samples-comparison.png")
@@
 
-(ref:mean-sd) Barplot displaying frequency of dependencies for all of the R packages in the CRAN repository to date `r Sys.Date()
-
-```{r mean-sd, fig.cap="."}
+```{r mean-sd, fig.cap="PCA-AGHQ modestly improves estimation of the posterior mean, and substantially improves estimation of the posterior standard deviation, as compared with TMB."}
 knitr::include_graphics("depends/mean-sd-alt.png")

### 2023-06-22 src/naomi-simple_exceedance/exceedance.Rmd
Upload paper and appendix new versions #55

@@
   ) %>%
-  ggplot(aes(x = tmbstan, y = estimate)) +
+  ggplot(aes(x = tmbstan, y = estimate - tmbstan)) +
     geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
-    coord_fixed(ratio = 1) +
     scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
-    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
     facet_grid(indicator ~ method) +
-    labs(x = "NUTS", y = "", col = "Sex") +
-    theme_minimal() +
-    theme(
-      legend.position = "bottom"
-    )
-
-exceedance_manuscript_plot <- exceedance_manuscript_plot +
+    labs(x = "NUTS", y = "Approximate - NUTS", col = "Sex") +
     geom_text(
@@
       size = 3, hjust = 0, vjust = 1.5
+    ) +
+    theme_minimal() +
+    theme(
+      legend.position = "bottom"
     )
 
-ggsave("exceedance.png", exceedance_manuscript_plot, h = 6.5, w = 6.25)
+ggsave("exceedance.png", exceedance_manuscript_plot, h = 6.5, w = 6.25, bg = "white")
 

### 2023-06-22 src/docs_paper/appendix.Rmd
Replace all elgm-inf with naomi-aghq
PCA in title #55

@@
 ---
-title: Appendix to "Fast approximate Bayesian inference of HIV indicators using principal components analysis adaptive Gauss-Hermite quadrature"
+title: Appendix to "Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature"
 author:
@@
 
-The C++ `TMB` code for the negative log-posterior of the simplified Naomi model is available from the GitHub repository `athowes/elgm-inf`.
+The C++ `TMB` code for the negative log-posterior of the simplified Naomi model is available from the GitHub repository `athowes/naomi-aghq`.
 For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used in Section \ref{sec:math} and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.

### 2023-06-26 src/docs_paper/paper.Rmd
Easy Alex fixes

@@
 Instead, we use principal components analysis (PCA) of the Hessian at the mode to find a smaller number of dimensions which explain most of the variance.
-In our application, this results in a grid which has millions of times fewer nodes the corresponding dense grid.
-Our method is implemented as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
-As `aghq` is designed to naturally interface with `TMB`, use is simple when provided a \textsf{C++} user template for the log-posterior.
+For the Naomi model in Malawi, this results in a grid which has millions of times fewer nodes the corresponding dense grid.
+Our implementation of the method makes use of the existing Naomi `TMB` template, and is immediately compatible with any model with such a template.
 
@@
 
-Let $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$ be a concatenated vector of observations.
-Here, we could attempt to write the Naomi model fully in a small number of equations, but likely it'd be difficult.
+In all, we model the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
+Let $\x$ be the latent field, comprised of all intercept effects, age random effects, and spatial random effects in the model.
+Let $\btheta$ be hyperparameters, comprised of all standard deviations, and BYM2 proportion parameters.
 
@@
 
-Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models of the form
+Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models with likelihood
 \begin{align*}
@@
 \mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
-\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n].
+\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n],
 \end{align*}
+with latent field and hyperparameter priors as in the LGM case.
 
@@
 We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
-The three approaches were:
+These were:
 1. TMB,
-2. PCA-AGHQ, and
+2. PCA-AGHQ, building upon the `aghq` package [@stringer2021implementing],
 3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
-The `TMB` C++ user-template used to specify the log-posterior was the same for each.
+Our goal in this analysis was to determine the accuracy of the approximate methods TMB and PCA-AGHQ as compared with the gold-standard NUTS.
+The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
+Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
 The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
-Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
 For the deterministic methods, following inference we simulated hyperparameter and latent field samples.

### 2023-06-28 src/docs_paper/paper.Rmd
Edit to abstract

@@
 Using data from Malawi, our method improves the accuracy of inferences across a range of metrics, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
-Our implementation uses the \texttt{aghq} \textsc{R} package, facilitating easy, flexible use of the method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
+Our implementation is based on the existing \texttt{TMB} \textsf{C++} template for the model's log-posterior, and is compatible with any model with such a template.
 keyword-subclass: | 

### 2023-06-29 src/docs_paper/paper.Rmd
Reshuffle some material

@@
 Section \ref{sec:naomi} outlines the version of the Naomi model that we consider in this paper, and  Section \ref{sec:elgm} describes how it falls within the ELGM framework.
-Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference for ELGMs using simplified INLA and AGHQ.
-As a case study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:results}.
+Section \ref{sec:inferencenaomi} outlines a deterministic inference method for ELGMs based on nested application of AGHQ and Laplace approximation, and how we modify it using PCA to be applicable to Naomi.
+In a case study, we evaluate the accuracy of our PCA-AGHQ method for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:results}.
 We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
-Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how we anticipate our method might be useful for other models, and directions for future research.
+Finally, in Section \ref{sec:conclusions} we discuss our conclusions, and directions for future research.
 
@@
 
-# Fast approximate inference method\label{sec:fastinferencemethods}
+# Inference methods for Naomi \label{sec:inferencenaomi}
+
+We begin by describing (Section \ref{sec:inferenceelgm}) the inference method for ELGMs based on nested applications of the Laplace approximation and AGHQ used by @stringer2022fast.
+We then propose (Section \ref{sec:pca}) an extension of the method which uses PCA to facilitates inference for Naomi, which otherwise would be intractable.
+
+## Inference for ELGMs \label{sec:inferenceelgm}
 
@@
 \end{equation*}
-We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
+Consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
 \begin{align}
@@
 \end{align}
-Given the negative unnormalised log posterior $- \log p(\y, \x, \btheta)$, we obtain the above posterior marginal approximations $\{ \tilde p(x_i \, | \, \y) \}_{i = 1}^N$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$ via nested applications of the Laplace approximation and AGHQ.
 
-## Laplace approximation
+### Laplace approximation
 
@@
 
-## Gauss-Hermite quadrature
+### Gauss-Hermite quadrature
 
@@
 
-## Adaptive quadrature
+### Adaptive quadrature
 
@@
 
+## Principal components analysis \label{sec:pca}
+
 ```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension (A).
@@
 
-## Principal components analysis
-
-Use of the product rule requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points.
+Use of the product rule grid above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points.
 This quickly becomes intractable as $m$ increases for $k > 1$.
@@
 
-## Sparse rules
-
-Though there are sparse rules which retain the attractive exactness properties of GHQ, using fewer than $k^m$ quadrature points, they use weighting functions which may be negative $\omega(\z) < 0$.
-This introduces a problem when trying to produce inferences about the latent field, as the resulting $\lambda(\z)$ may be negative.
-We are not aware of any suitable approaches to obtain multinomial samples in such cases.
-It is however, still possible to calculate the normalising constant using a sparse rule, which we do so in Appendix S3, finding it to be related to that from PCA-AGHQ as follows.
-
-<!-- Pruning (Jackel, 2005): is this something we could be doing? -->
-
 # Application to data from Malawi\label{sec:results}
@@
 These were:
+
 1. TMB,
-2. PCA-AGHQ, building upon the `aghq` package [@stringer2021implementing],
+2. PCA-AGHQ, via extension of the `aghq` package [@stringer2021implementing],
 3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
-Our goal in this analysis was to determine the accuracy of the approximate methods TMB and PCA-AGHQ as compared with the gold-standard NUTS.
+
+Our goal in this analysis was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
 The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
@@
 Example model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
-The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/naomi-aghq`.
-We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
 
@@
 ```
+
+The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/naomi-aghq`.
+We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
 
@@
 TMB & \texttt{TMB} & $1000$ samples \\
-PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-settings}), $1000$ samples \\
+PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-use}), $1000$ samples \\
 NUTS & \texttt{tmbstan} & $4$ chains of $20000$ iterations, with the first $10000$ iterations of each chain discarded as warmup, thinned by a factor of $20$.
@@
 
-## PCA-AGHQ settings \label{sec:pca-settings}
+## Use of PCA-AGHQ \label{sec:pca-use}
 
@@
 This research was supported by the MRC Centre for Global Infectious Disease Analysis (MR/R015600/1), jointly funded by the UK Medical Research Council (MRC) and the UK Foreign, Commonwealth \& Development Office (FCDO), under the MRC/FCDO Concordat program and is also part of the EDCTP2 programme supported by the European Union.
+
+<!-- # Graveyard -->
+
+<!-- ## Sparse rules -->
+
+<!-- Though there are sparse rules which retain the attractive exactness properties of GHQ, using fewer than $k^m$ quadrature points, they use weighting functions which may be negative $\omega(\z) < 0$.
+-->
+<!-- This introduces a problem when trying to produce inferences about the latent field, as the resulting $\lambda(\z)$ may be negative. -->
+<!-- We are not aware of any suitable approaches to obtain multinomial samples in such cases. -->
+<!-- It is however, still possible to calculate the normalising constant using a sparse rule, which we do so in Appendix S3, finding it to be related to that from PCA-AGHQ as follows. -->
+
+<!-- Pruning (Jackel, 2005): is this something we could be doing? -->

### 2023-06-30 src/naomi-simple_exceedance/exceedance.Rmd
Point estimates rescue completed, you're safe now #56

@@
 * The `excursions` package can calculate excursion sets, contour credible regions and level avoiding sets, excursion functions, contour maps and their quality measures, and simultaneous confidence bands
+
+# Original computing environment {-}
+
+```{r}
+sessionInfo()
+```
+
+# Bibliography {-}

### 2023-06-30 src/naomi-simple_point-estimates/point-estimates.Rmd
Point estimates rescue completed, you're safe now #56

---
title: "Point estimates comparison for the simplified Naomi model"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**  We have run the simplified Naomi model using a range of inference methods: `TMB`, `aghq`, and `tmbstan`.

  **Task** In this report, we compare the accuracy of the posterior distributions obtained from these inference methods using point estimates.
---

# Background

We compare the inference results from `TMB`, `aghq`, and `tmbstan`.
Import these inference results as follows:

```{r}
tmb <- readRDS("depends/tmb.rds")
aghq <- readRDS("depends/aghq.rds")
tmbstan <- readRDS("depends/tmbstan.rds")
```

# Means and standard deviations

```{r}
latent_pars <- unique(names(tmb$fit$par.full))[!(unique(names(tmb$fit$par.full)) %in% names(tmb$fit$par))]
output_pars <- c("rho_t1_out", "lambda_t1_out", "alpha_t1_out")

type_key <- data.frame(
  par = c(latent_pars, output_pars),
  type = c(rep("latent", times = length(latent_pars)), rep("output", times = length(output_pars)))
)

df_tmb <- lapply(tmb$fit$sample[type_key$par], function(row) {
  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "TMB")

df_aghq <- lapply(aghq$quad$sample[type_key$par], function(row) {
  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "PCA-AGHQ")

df_tmbstan <- lapply(tmbstan$mcmc$sample[type_key$par], function(row) {
  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "NUTS")

df <- bind_rows(df_tmb, df_aghq, df_tmbstan) %>%
  left_join(
    type_key,
    by = "par"
  )

df_plot <- df %>%
  pivot_longer(cols = c("sd", "mean"), names_to = "indicator", values_to = "estimate") %>%
  pivot_wider(values_from = "estimate", names_from = "method") %>%
  pivot_longer(cols = c("TMB", "PCA-AGHQ"), names_to = "method", values_to = "approximate") %>%
  rename("truth" = "NUTS") %>%
  mutate(
    indicator = fct_recode(indicator, "Posterior mean estimate" = "mean", "Posterior SD estimate" = "sd"),
    method = fct_relevel(method, "TMB", "PCA-AGHQ")
  )

df_metrics <- df_plot %>%
  group_by(method, indicator) %>%
  summarise(
    rmse = sqrt(mean((truth - approximate)^2)),
    mae = mean(abs(truth - approximate))
  )

df_metrics_pct <- df_metrics %>%
  ungroup() %>%
  group_by(indicator) %>%
  summarise(
    rmse_diff = 100 * diff(rmse) / max(rmse),
    mae_diff = 100 * diff(mae) / max(mae)
  )

df_metrics <- df_metrics %>%
  mutate(
    label = ifelse(
      method == "PCA-AGHQ",
      paste0("RMSE: ", round(rmse, 2), " (", round(df_metrics_pct$rmse_diff), "%)", "\nMAE: ", round(mae, 2), " (", round(df_metrics_pct$mae_diff), "%)"),
      paste0("RMSE: ", round(rmse, 2), "\nMAE: ", round(mae, 2))
    )
  )

write_csv(df_metrics, "mean-sd.csv")

mean_sd_plot <- ggplot(df_plot, aes(x = truth, y = approximate - truth)) +
  geom_point(shape = 1, alpha = 0.4) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
  geom_text(data = df_metrics, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal()

ggsave("mean-sd.png", mean_sd_plot, h = 6, w = 6.25)

#' Split into two plots for presentations etc.

jitter_amount <- 0.02

mean_plot <- df_plot %>%
  filter(indicator == "Posterior mean estimate") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.4, 0.4)) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
  geom_text(
    data = filter(df_metrics, indicator == "Posterior mean estimate"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal()

ggsave("mean.png", mean_plot, h = 4, w = 6.25)

sd_plot <- df_plot %>%
  filter(indicator == "Posterior SD estimate") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.6, 0.6)) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
  geom_text(
    data = filter(df_metrics, indicator == "Posterior SD estimate"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal()

ggsave("sd.png", sd_plot, h = 4, w = 6.25)

sd_plot_alt <- sd_plot +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )

y_axis <- ggplot(data.frame(l = mean_plot$labels$y, x = 1, y = 1)) +
  geom_text(aes(x, y, label = l), angle = 90) +
  theme_void() +
  coord_cartesian(clip = "off")

mean_plot_alt <- mean_plot
mean_plot_alt$labels$x <- ""
mean_plot_alt$labels$y <- sd_plot_alt$labels$y <- ""

mean_sd_plot_alt <- y_axis + (mean_plot_alt / sd_plot_alt) +
  plot_layout(widths = c(1, 30))

ggsave("mean-sd-alt.png", mean_sd_plot_alt, h = 6, w = 6.25)
```

# Original computing environment {-}

```{r}
sessionInfo()
```

# Bibliography {-}

### 2023-06-30 src/naomi-simple_contraction/script.R

@@ -73,129 +73,3 @@ ggsave("posterior-contraction.png", posterior_contraction_plot, h = 6.5, w = 6.2
 
 #' Which have lower amounts of posterior contraction?
 names(subset(posterior_contraction, posterior_contraction < 0.5))
-
-#' How does the size of standard deviations compare in TMB to AGHQ?
-df_tmb <- lapply(tmb$fit$sample[df_latent$par], function(row) {
-  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
-    tibble::rowid_to_column("id")
-  }) %>%
-  dplyr::bind_rows(.id = "par") %>%
-  mutate(method = "TMB")
-
-df_aghq <- lapply(aghq$quad$sample[df_latent$par], function(row) {
-  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
-    tibble::rowid_to_column("id")
-  }) %>%
-  dplyr::bind_rows(.id = "par") %>%
-  mutate(method = "PCA-AGHQ")
-
-df_tmbstan <- lapply(tmbstan$mcmc$sample[df_latent$par], function(row) {
-  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
-    tibble::rowid_to_column("id")
-  }) %>%
-  dplyr::bind_rows(.id = "par") %>%
-  mutate(method = "NUTS")
-
-df <- bind_rows(df_tmb, df_aghq, df_tmbstan)
-
-df_plot <- df %>%
-  pivot_longer(cols = c("sd", "mean"), names_to = "indicator", values_to = "estimate") %>%
-  pivot_wider(values_from = "estimate", names_from = "method") %>%
-  pivot_longer(cols = c("TMB", "PCA-AGHQ"), names_to = "method", values_to = "approximate") %>%
-  rename("truth" = "NUTS") %>%
-  mutate(
-    indicator = fct_recode(indicator, "Posterior mean estimate" = "mean", "Posterior SD estimate" = "sd"),
-    method = fct_relevel(method, "TMB", "PCA-AGHQ")
-  )
-
-df_metrics <- df_plot %>%
-  group_by(method, indicator) %>%
-  summarise(
-    rmse = sqrt(mean((truth - approximate)^2)),
-    mae = mean(abs(truth - approximate))
-  )
[…]

### 2023-07-03 src/naomi-simple_point-estimates/point-estimates.Rmd
Add point estimates notebook (and split out latent and output analysis)

@@
 
-# Means and standard deviations
+# Data manipulation
+
+We're interested in the latent field and certain model outputs (prevalence, incidence and treatment coverage):
 
@@
 )
+```
 
+Extract the mean and standard deviation for each parameter for each inference method, then bind them together:
+
+```{r}
 df_tmb <- lapply(tmb$fit$sample[type_key$par], function(row) {
-  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
+  data.frame("mean" = rowMeans(row), "sd" = matrixStats::rowSds(row)) %>%
     tibble::rowid_to_column("id")
@@
 df_aghq <- lapply(aghq$quad$sample[type_key$par], function(row) {
-  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
+  data.frame("mean" = rowMeans(row), "sd" = matrixStats::rowSds(row)) %>%
     tibble::rowid_to_column("id")
@@
 df_tmbstan <- lapply(tmbstan$mcmc$sample[type_key$par], function(row) {
-  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
+  data.frame("mean" = rowMeans(row), "sd" = matrixStats::rowSds(row)) %>%
     tibble::rowid_to_column("id")
@@
 
+head(df)
+```
+
+Pivot so that the data is ready to plot:
+
+```{r}
 df_plot <- df %>%
@@
 
+head(df_plot)
+```
+Calculate the root mean square and mean absolute errors between the approximate methods and NUTS (taken to be the truth):
+
+```{r}
 df_metrics <- df_plot %>%
-  group_by(method, indicator) %>%
+  group_by(method, indicator, type) %>%
   summarise(
@@
     mae = mean(abs(truth - approximate))
-  )
+  ) %>%
+  ungroup()
 
+head(df_metrics)
+```
+
+Add difference in metric between TMB and PCA-AGHQ:
+
+```{r}
 df_metrics_pct <- df_metrics %>%
-  ungroup() %>%
-  group_by(indicator) %>%
-  summarise(
+  split(~ indicator + type) %>%
+  lapply(function(x) {
+    x %>% summarise(
       rmse_diff = 100 * diff(rmse) / max(rmse),
@@
     )
+  }) %>%
+  bind_rows(.id = "id") %>%
+  separate(id, c("indicator", "type"), sep = "\\.")
 
+#' Add as a naming column for plot
 df_metrics <- df_metrics %>%
+  left_join(
+    df_metrics_pct,
+    by = c("indicator", "type")
+  ) %>%
   mutate(
@@
       method == "PCA-AGHQ",
-      paste0("RMSE: ", round(rmse, 2), " (", round(df_metrics_pct$rmse_diff), "%)", "\nMAE: ", round(mae, 2), " (", round(df_metrics_pct$mae_diff), "%)"),
-      paste0("RMSE: ", round(rmse, 2), "\nMAE: ", round(mae, 2))
+      paste0("RMSE: ", signif(rmse, 2), " (", signif(rmse_diff, 2), "%)", "\nMAE: ", signif(mae, 2), " (", signif(mae_diff, 2), "%)"),
+      paste0("RMSE: ", signif(rmse, 2), "\nMAE: ", signif(mae, 2))
     )
@@
 write_csv(df_metrics, "mean-sd.csv")
+```
 
-mean_sd_plot <- ggplot(df_plot, aes(x = truth, y = approximate - truth)) +
+# Plots
+
+## Latent field
+
+```{r}
+df_plot_latent <- filter(df_plot, type == "latent")
+df_metrics_latent <- filter(df_metrics, type == "latent")
+
+(mean_sd_plot_latent <- df_plot_latent %>%
+  ggplot(aes(x = truth, y = approximate - truth)) +
   geom_point(shape = 1, alpha = 0.4) +
@@
   geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
-  geom_text(data = df_metrics, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
+  geom_text(data = df_metrics_latent, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
   labs(x = "NUTS", y = "Approximation - NUTS") +
-  theme_minimal()
+  theme_minimal())
 
-ggsave("mean-sd.png", mean_sd_plot, h = 6, w = 6.25)
+ggsave("mean-sd-latent.png", mean_sd_plot_latent, h = 6, w = 6.25)
+```
 
-#' Split into two plots for presentations etc.
+Version split into two for presentations
 
+```{r}
 jitter_amount <- 0.02
 
-mean_plot <- df_plot %>%
+mean_plot_latent <- df_plot_latent %>%
   filter(indicator == "Posterior mean estimate") %>%
@@
   geom_text(
-    data = filter(df_metrics, indicator == "Posterior mean estimate"),
+    data = filter(df_metrics_latent, indicator == "Posterior mean estimate"),
     aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
@@
 
-ggsave("mean.png", mean_plot, h = 4, w = 6.25)
+ggsave("mean-latent.png", mean_plot_latent, h = 4, w = 6.25)
 
-sd_plot <- df_plot %>%
+sd_plot_latent <- df_plot_latent %>%
   filter(indicator == "Posterior SD estimate") %>%
@@
   geom_text(
-    data = filter(df_metrics, indicator == "Posterior SD estimate"),
+    data = filter(df_metrics_latent, indicator == "Posterior SD estimate"),
     aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
@@
 
-ggsave("sd.png", sd_plot, h = 4, w = 6.25)
+ggsave("sd-latent.png", sd_plot_latent, h = 4, w = 6.25)
 
-sd_plot_alt <- sd_plot +
+sd_plot_latent_alt <- sd_plot_latent +
   theme(
@@
 
-y_axis <- ggplot(data.frame(l = mean_plot$labels$y, x = 1, y = 1)) +
+y_axis <- ggplot(data.frame(l = mean_plot_latent$labels$y, x = 1, y = 1)) +
   geom_text(aes(x, y, label = l), angle = 90) +
@@
 
-mean_plot_alt <- mean_plot
-mean_plot_alt$labels$x <- ""
-mean_plot_alt$labels$y <- sd_plot_alt$labels$y <- ""
+mean_plot_latent_alt <- mean_plot_latent
+mean_plot_latent_alt$labels$x <- ""
+mean_plot_latent_alt$labels$y <- sd_plot_latent_alt$labels$y <- ""
 
-mean_sd_plot_alt <- y_axis + (mean_plot_alt / sd_plot_alt) +
-  plot_layout(widths = c(1, 30))
+(mean_sd_plot_latent_alt <- y_axis + (mean_plot_latent_alt / sd_plot_latent_alt) +
+  plot_layout(widths = c(1, 30)))
 
-ggsave("mean-sd-alt.png", mean_sd_plot_alt, h = 6, w = 6.25)
+ggsave("mean-sd-alt-latent.png", mean_sd_plot_latent_alt, h = 6, w = 6.25)
+```
+
+## Outputs
+
+```{r}
+df_plot_output <- filter(df_plot, type == "output")
+df_metrics_output <- filter(df_metrics, type == "output")
+cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
+(mean_sd_plot_output <- df_plot_output %>%
+  ggplot() +
+  geom_point(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4) +
+  scale_color_manual(values = cbpalette) +
+  facet_grid(indicator ~ method) +
+  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
+  geom_text(data = df_metrics_output, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
+  labs(x = "NUTS", y = "Approximation - NUTS") +
+  theme_minimal())
+
+ggsave("mean-sd-output.png", mean_sd_plot_output, h = 6, w = 6.25)
+```
+
+Version split into two for presentations
+
+```{r}
+jitter_amount <- 0.02
+
+mean_plot_output <- df_plot_output %>%
+  filter(indicator == "Posterior mean estimate") %>%
+  ggplot() +
+  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
+  scale_color_manual(values = cbpalette) +
+  facet_grid(indicator ~ method) +
+  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  geom_text(
+    data = filter(df_metrics_output, indicator == "Posterior mean estimate"),
+    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
+  ) +
+  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
+  theme_minimal()
+
+ggsave("mean-output.png", mean_plot_output, h = 4, w = 6.25)
+
+sd_plot_output <- df_plot_output %>%
+  filter(indicator == "Posterior SD estimate") %>%
+  ggplot() +
+  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
+  scale_color_manual(values = cbpalette) +
+  facet_grid(indicator ~ method) +
+  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  geom_text(
+    data = filter(df_metrics_output, indicator == "Posterior SD estimate"),
+    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
+  ) +
+  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
+  theme_minimal()
+
+ggsave("sd-output.png", sd_plot_output, h = 4, w = 6.25)
+
+sd_plot_output_alt <- sd_plot_output +
+  theme(
+    strip.background = element_blank(),
+    strip.text.x = element_blank(),
+    legend.position = "bottom"
+  )
+
+y_axis <- ggplot(data.frame(l = mean_plot_output$labels$y, x = 1, y = 1)) +
+  geom_text(aes(x, y, label = l), angle = 90) +
+  theme_void() +
+  coord_cartesian(clip = "off")
+
+mean_plot_output_alt <- mean_plot_output +
+  theme(legend.position = "none")
+
+mean_plot_output_alt$labels$x <- ""
+mean_plot_output_alt$labels$y <- sd_plot_output_alt$labels$y <- ""
+
+(mean_sd_plot_output_alt <- y_axis + (mean_plot_output_alt / sd_plot_output_alt) +
+  plot_layout(widths = c(1, 30)))
+
+ggsave("mean-sd-alt-output.png", mean_sd_plot_output_alt, h = 6, w = 6.25)
 ```

### 2023-07-03 src/naomi-simple_point-estimates/point-estimates.Rmd
Add largest error data tables in point estimate notebook

@@
 
+Where are the greatest errors occurring for the mean?
+
+```{r}
+df_plot_latent %>%
+  filter(indicator == "Posterior mean estimate") %>%
+  mutate(diff = abs(approximate - truth)) %>%
+  arrange(desc(diff)) %>%
+  select(indicator, par, id, method, truth, approximate, diff) %>%
+  DT::datatable()
+```
+
+What about for the standard deviations?
+
+```{r}
+df_plot_latent %>%
+  filter(indicator == "Posterior SD estimate") %>%
+  mutate(diff = abs(approximate - truth)) %>%
+  arrange(desc(diff)) %>%
+  select(indicator, par, id, method, truth, approximate, diff) %>%
+  DT::datatable()
+```
+
 ## Outputs

### 2023-07-03 src/docs_paper/appendix.Rmd
Add plot of node marginal quantiles within NUTS posterior samples

@@
 
+```{r nodes-quantiles-sd, fig.cap="Standard deviations of the quantiles of the quadrature nodes within the NUTS posterior draws varied substantially, in accordance with the marginal standard deviations shown in Figure \\ref{fig:marginal-sd}."}
+knitr::include_graphics("depends/nodes-quantiles-sd.png")
+```
+
 ```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
@@
 ## Estimated normalising constant comparison
+
+Add here plots and description of the estimated normalising constant for different PCA-AGHQ settings.
 

### 2023-07-03 src/docs_paper/paper.Rmd
Add plot of node marginal quantiles within NUTS posterior samples

@@
 ```{r mean-sd, fig.cap="PCA-AGHQ modestly improves estimation of the posterior mean, and substantially improves estimation of the posterior standard deviation, as compared with TMB."}
-knitr::include_graphics("depends/mean-sd-alt.png")
+knitr::include_graphics("depends/mean-sd-alt-latent.png")
 ```

### 2023-07-03 src/naomi-simple_ks/ks.Rmd
Remove adam method from ks report (for now?)

@@
 
-### `adam`
-
-```{r}
-dependency_details(3)
-```
-
 ### `tmbstan`
@@
 ```{r}
-tmbstan_details <- dependency_details(4)
+tmbstan_details <- dependency_details(3)
 tmbstan_details
@@
   "aghq" = aghq$time,
-  "adam" = adam_time,
   "tmbstan" = tmbstan$time
@@
     "Parameter" = "par",
-    "KS(adam, tmbstan)" = "adam",
     "KS(aghq, tmbstan)" = "aghq",
@@
 
-r <- adam$quad$obj$env$random
-x_names <- names(adam$quad$obj$env$par[r])
-theta_names <- names(adam$quad$obj$env$par[-r])
+r <- tmb$obj$env$random
+x_names <- names(tmb$obj$env$par[r])
+theta_names <- names(tmb$obj$env$par[-r])
 
@@
     TMB = mean(`KS(TMB, tmbstan)`),
-    aghq = mean(`KS(aghq, tmbstan)`),
-    adam = mean(`KS(adam, tmbstan)`)
+    aghq = mean(`KS(aghq, tmbstan)`)
   ) %>%
@@
     `KS(TMB, tmbstan)` = mean(`KS(TMB, tmbstan)`),
-    `KS(aghq, tmbstan)` = mean(`KS(aghq, tmbstan)`),
-    `KS(adam, tmbstan)` = mean(`KS(adam, tmbstan)`)
+    `KS(aghq, tmbstan)` = mean(`KS(aghq, tmbstan)`)
   ) %>%
@@
 #     `KS(TMB, tmbstan)` = mean(`KS(TMB, tmbstan)`),
-#     `KS(aghq, tmbstan)` = mean(`KS(aghq, tmbstan)`),
-#     `KS(adam, tmbstan)` = mean(`KS(adam, tmbstan)`)
+#     `KS(aghq, tmbstan)` = mean(`KS(aghq, tmbstan)`)
 #   ) %>%
@@
     Method = fct_recode(Method,
-      "adam" = "KS(adam, tmbstan)",
       "aghq" = "KS(aghq, tmbstan)",

### 2023-07-03 src/naomi-simple_ks/ks.Rmd
Simplify KS comparison, ready to use AD as well if required

@@
 
-write_csv(time_taken, "time_taken.csv")
+write_csv(time_taken, "time-taken.csv")
 
@@
 
-```{r}
-histogram_and_ecdf_helper <- function(par) lapply(1:sum(names(tmb$fit$obj$env$par) == par), histogram_and_ecdf, par = par)
-```
-
 ## `beta_rho`
@@
 ```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("beta_rho")
+histogram_and_ecdf_list("beta_rho")
 ```
@@
 ```{r eval = pars_eval[["beta_alpha"]]}
-histogram_and_ecdf_helper("beta_alpha")
+histogram_and_ecdf_list("beta_alpha")
 ```
@@
 ```{r eval = pars_eval[["beta_lambda"]]}
-histogram_and_ecdf_helper("beta_lambda")
+histogram_and_ecdf_list("beta_lambda")
 ```
@@
 ```{r eval = pars_eval[["beta_anc_rho"]]}
-histogram_and_ecdf_helper("beta_anc_rho")
+histogram_and_ecdf_list("beta_anc_rho")
 ```
@@
 ```{r eval = pars_eval[["u_rho_x"]]}
-histogram_and_ecdf_helper("u_rho_x")
+histogram_and_ecdf_list("u_rho_x")
 ```
@@
 ```{r eval = pars_eval[["us_rho_x"]]}
-histogram_and_ecdf_helper("us_rho_x")
+histogram_and_ecdf_list("us_rho_x")
 ```
@@
 ```{r eval = pars_eval[["u_rho_xs"]]}
-histogram_and_ecdf_helper("u_rho_xs")
+histogram_and_ecdf_list("u_rho_xs")
 ```
@@
 ```{r eval = pars_eval[["us_rho_xs"]]}
-histogram_and_ecdf_helper("us_rho_xs")
+histogram_and_ecdf_list("us_rho_xs")
 ```
@@
 ```{r eval = pars_eval[["u_rho_as"]]}
-histogram_and_ecdf_helper("u_rho_as")
+histogram_and_ecdf_list("u_rho_as")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_x"]]}
-histogram_and_ecdf_helper("u_alpha_x")
+histogram_and_ecdf_list("u_alpha_x")
 ```
@@
 ```{r eval = pars_eval[["us_alpha_x"]]}
-histogram_and_ecdf_helper("us_alpha_x")
+histogram_and_ecdf_list("us_alpha_x")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_xs"]]}
-histogram_and_ecdf_helper("u_alpha_xs")
+histogram_and_ecdf_list("u_alpha_xs")
 ```
@@
 ```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_helper("us_alpha_xs")
+histogram_and_ecdf_list("us_alpha_xs")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_a"]]}
-histogram_and_ecdf_helper("u_alpha_a")
+histogram_and_ecdf_list("u_alpha_a")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_as"]]}
-histogram_and_ecdf_helper("u_alpha_as")
+histogram_and_ecdf_list("u_alpha_as")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_xa"]]}
-histogram_and_ecdf_helper("u_alpha_xa")
+histogram_and_ecdf_list("u_alpha_xa")
 ```
@@
 ```{r eval = pars_eval[["ui_lambda_x"]]}
-histogram_and_ecdf_helper("ui_lambda_x")
+histogram_and_ecdf_list("ui_lambda_x")
 ```
@@
 ```{r eval = pars_eval[["ui_anc_rho_x"]]}
-histogram_and_ecdf_helper("ui_anc_rho_x")
+histogram_and_ecdf_list("ui_anc_rho_x")
 ```
@@
 ```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-histogram_and_ecdf_helper("ui_anc_alpha_x")
+histogram_and_ecdf_list("ui_anc_alpha_x")
 ```
@@
 ```{r eval = pars_eval[["log_or_gamma"]]}
-histogram_and_ecdf_helper("log_or_gamma")
+histogram_and_ecdf_list("log_or_gamma")
 ```
@@
 
+```{r}
+r <- tmb$fit$obj$env$random
+x_names <- names(tmb$fit$obj$env$par[r])
+theta_names <- names(tmb$fit$obj$env$par[-r])
+
+dict <- data.frame(
+  parname = c(unique(x_names), unique(theta_names)),
+  type = c(rep("Latent field", length(unique(x_names))), rep("Hyper", length(unique(theta_names))))
+)
+
+ks_df <- lapply(unique(names(tmb$fit$obj$env$par)), to_ks_df) %>%
+  bind_rows() %>%
+  mutate(parname = str_extract(par, ".*(?=\\[)")) %>%
+  left_join(dict, by = "parname")
+```
+
 ## Individual parameters {.tabset .tabset-fade}
@@
 ```{r}
-ks_helper <- function(par, starts_with = FALSE, ...) {
-  to_ks_df(par, starts_with = starts_with) %>% ks_plot(par, ...)
+ks_plot_starts_with <- function(param) {
+  ks_df %>%
+    filter(startsWith(par, param)) %>%
+    ks_plot(par = param)
 }
@@
 ```{r}
-ks_helper("beta", starts_with = TRUE)
+ks_plot_starts_with("beta")
 ```
@@
 ```{r}
-ks_helper("logit", starts_with = TRUE)
+ks_plot_starts_with("logit")
 ```
@@
 ```{r}
-ks_helper("log_sigma", starts_with = TRUE)
+ks_plot_starts_with("log_sigma")
 ```
@@
 ```{r eval = pars_eval[["u_rho_x"]]}
-ks_helper("u_rho_x")
+ks_plot_starts_with("u_rho_x")
 ```
@@
 ```{r eval = pars_eval[["u_rho_xs"]]}
-ks_helper("u_rho_xs")
+ks_plot_starts_with("u_rho_xs")
 ```
@@
 ```{r eval = pars_eval[["us_rho_x"]]}
-ks_helper("us_rho_x")
+ks_plot_starts_with("us_rho_x")
 ```
@@
 ```{r eval = pars_eval[["us_rho_xs"]]}
-ks_helper("us_rho_xs")
+ks_plot_starts_with("us_rho_xs")
 ```
@@
 ```{r eval = pars_eval[["u_rho_a"]]}
-ks_helper("u_rho_a")
+ks_plot_starts_with("u_rho_a")
 ```
@@
 ```{r eval = pars_eval[["u_rho_as"]]}
-ks_helper("u_rho_as")
+ks_plot_starts_with("u_rho_as")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_x"]]}
-ks_helper("u_alpha_x")
+ks_plot_starts_with("u_alpha_x")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_xs"]]}
-ks_helper("u_alpha_xs")
+ks_plot_starts_with("u_alpha_xs")
 ```
@@
 ```{r eval = pars_eval[["us_alpha_x"]]}
-ks_helper("us_alpha_x")
+ks_plot_starts_with("us_alpha_x")
 ```
@@
 ```{r eval = pars_eval[["us_alpha_xs"]]}
-ks_helper("us_alpha_xs")
+ks_plot_starts_with("us_alpha_xs")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_a"]]}
-ks_helper("u_alpha_a")
+ks_plot_starts_with("u_alpha_a")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_as"]]}
-ks_helper("u_alpha_as")
+ks_plot_starts_with("u_alpha_as")
 ```
@@
 ```{r eval = pars_eval[["u_alpha_xa"]]}
-ks_helper("u_alpha_xa")
+ks_plot_starts_with("u_alpha_xa")
 ```
@@
 ```{r eval = pars_eval[["ui_anc_rho_x"]]}
-ks_helper("ui_anc_rho_x")
+ks_plot_starts_with("ui_anc_rho_x")
 ```
@@
 ```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-ks_helper("ui_anc_alpha_x")
+ks_plot_starts_with("ui_anc_alpha_x")
 ```
@@
 ```{r eval = pars_eval[["log_or_gamma"]]}
-ks_helper("log_or_gamma")
+ks_plot_starts_with("log_or_gamma")
 ```
@@
 
-```{r}
-# ks_lambda_t1_out %>%
-#   filter(ks == 0) %>%
-#   group_by(method) %>%
-#   summarise(n = n())
-# 
-# ks_lambda_t1_out %>%
-#   filter(ks == 0) %>%
-#   pull(age_group) %>%
-#   unique()
-```
-
 ## Summary {.tabset .tabset-fade}
-
-Values in dark blue are the lowest KS difference for that particular parameter:
 
@@
 
-ks_summary <- lapply(unique(names(tmb$fit$obj$env$par)), function(x) {
-  to_ks_df(x) %>%
-    group_by(method, index) %>%
-    summarise(ks = mean(ks), par = x) %>%
-    ungroup()
-}) %>%
-  bind_rows() %>%
+ks_summary <- ks_df %>%
+  group_by(method, parname, type) %>%
+  summarise(ks = signif(mean(ks), 4)) %>%
   pivot_wider(names_from = "method", values_from = "ks") %>%
   rename(
-    "Parameter" = "par",
+    "Parameter" = "parname",
     "KS(aghq, tmbstan)" = "aghq",
@@
 
-r <- tmb$obj$env$random
-x_names <- names(tmb$obj$env$par[r])
-theta_names <- names(tmb$obj$env$par[-r])
-
-dict <- data.frame(
-  Parameter = c(unique(x_names), unique(theta_names)),
-  Type = c(rep("Latent field", length(unique(x_names))), rep("Hyper", length(unique(theta_names))))
-)
-
-ks_summary <- ks_summary %>%
-  left_join(dict, by = "Parameter")
-
-saveRDS(ks_summary, "ks_summary.rds")
+saveRDS(ks_summary, "ks-summary.rds")
 
 ks_summary %>%
-  filter(Type != "Hyper") %>%
+  filter(type != "Hyper") %>%
   ks_plot_many("TMB", "aghq")
@@
 ks_summary %>%
-  group_by(Type) %>%
+  group_by(type) %>%
   summarise(
@@
 ks_summary %>%
-  filter(Type == "Latent field") %>%
-  group_by(Parameter) %>%
-  summarise(
-    `KS(TMB, tmbstan)` = mean(`KS(TMB, tmbstan)`),
-    `KS(aghq, tmbstan)` = mean(`KS(aghq, tmbstan)`)
-  ) %>%
-  ks_table()
-```
-
-### Hyperparameters
-
-```{r}
-# ks_summary %>%
-#   filter(Type == "Hyper") %>%
-#   group_by(Parameter) %>%
-#   summarise(
-#     `KS(TMB, tmbstan)` = mean(`KS(TMB, tmbstan)`),
-#     `KS(aghq, tmbstan)` = mean(`KS(aghq, tmbstan)`)
-#   ) %>%
-#   ks_table()
-```
-
-### Outputs
-
-```{r}
-#' To-do
+  filter(type == "Latent field") %>%
+  mutate(diff = signif(`KS(TMB, tmbstan)` - `KS(aghq, tmbstan)`, 3)) %>%
+  DT::datatable()
 ```
@@
 ```{r}
-ks_comparison <- function(ks_summary, method1, method2, n) {
-  ks_method1 <- paste0("KS(", method1, ", tmbstan)")
-  ks_method2 <- paste0("KS(", method2, ", tmbstan)")
-  ks_summary[["KS difference"]] <- ks_summary[[ks_method1]] - ks_summary[[ks_method2]]
-  ks_summary %>%
-    filter(Type != "Hyper") %>%
-    arrange(desc(`KS difference`)) %>%
-    head(n = n)
-}
+ks_df_wide <- ks_df %>%
+  pivot_wider(names_from = "method", values_from = "ks") %>%
+  mutate(diff = TMB - aghq)
 ```
@@
 ```{r}
-ks_comparison(ks_summary, method1 = "aghq", method2 = "TMB", 10)
+ks_df_wide %>%
+  arrange(desc(diff)) %>%
+  head(n = 10)
 
-histogram_and_ecdf(par = "us_rho_x", i = 19)
-histogram_and_ecdf(par = "ui_anc_rho_x", i = 2)
-histogram_and_ecdf(par = "u_rho_x", i = 2)
+histogram_and_ecdf(par = "us_alpha_xs", i = 18)
+histogram_and_ecdf(par = "us_alpha_xs", i = 16)
+histogram_and_ecdf(par = "u_alpha_xs", i = 13)
 ```
@@
 ```{r}
-ks_comparison(ks_summary, method1 = "TMB", method2 = "aghq", 10)
+ks_df_wide %>%
+  arrange(diff) %>%
+  head(n = 10)
 
-histogram_and_ecdf(par = "us_alpha_xs", i = 18)
-histogram_and_ecdf(par = "us_alpha_xs", i = 30)
-histogram_and_ecdf(par = "u_alpha_xa", i = 10)
+histogram_and_ecdf(par = "ui_anc_rho_x", i = 27)
+histogram_and_ecdf(par = "ui_anc_alpha_x", i = 19)
+histogram_and_ecdf(par = "u_alpha_x", i = 18)
 ```
@@
 
-ks_summary %>%
-  pivot_longer(
-    col = starts_with("KS"),
-    names_to = "Method",
-    values_to = "KS"
+ks_df %>%
+  filter(type != "Hyper") %>%
+  left_join(data.frame(ess) %>%
+    tibble::rownames_to_column("par"),
   ) %>%
-  mutate(
-    Method = fct_recode(Method,
-      "aghq" = "KS(aghq, tmbstan)",
-      "TMB" = "KS(TMB, tmbstan)"
-    )
-  ) %>%
-  group_by(Parameter) %>%
-  mutate(
-    par_num = case_when(
-      max(index) > 1 ~ paste0(Parameter, "[", index, "]"),
-      TRUE ~ Parameter
-    )
-  ) %>%
-  ungroup() %>%
-  left_join(data.frame(ess) %>%
-    tibble::rownames_to_column("par_num")
-  ) %>%
-  ggplot(aes(x = ess, y = KS)) +
+  ggplot(aes(x = ess, y = ks)) +
     geom_point(alpha = 0.5) +
     geom_smooth(method = "lm", color = "#56B4E9") +
-    facet_grid(~ Method) +
+    facet_grid(~ method) +
     theme_minimal() +

### 2023-07-03 src/docs_paper/appendix.Rmd
Add time taken to draft #55

@@
 ```{r}
-ks_summary <- readRDS("depends/ks_summary.rds")
+ks_summary <- readRDS("depends/ks-summary.rds")
 
 ks_summary %>%
-  filter(Type == "Latent field") %>%
-  group_by(Parameter) %>%
-  summarise(
-    `D(TMB)` = round(mean(`KS(TMB, tmbstan)`), 3),
-    `D(PCA-AGHQ)` = round(mean(`KS(aghq, tmbstan)`), 3)
-  ) %>%
+  ungroup() %>%
+  filter(type == "Latent field") %>%
+  select(-type) %>%
   gt::gt(rowname_col = "Parameter") %>%

### 2023-07-03 src/docs_paper/paper.Rmd
Add time taken to draft #55

@@
 
+```{r}
+time_taken <- read_csv("depends/time-taken.csv")
+```
+
 We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
@@
 
-1. TMB,
-2. PCA-AGHQ, via extension of the `aghq` package [@stringer2021implementing],
-3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
+1. TMB (`r signif(time_taken$TMB, 2)` seconds)
+2. PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), implemented via extension of the `aghq` package [@stringer2021implementing],
+3.
+NUTS (`r signif(time_taken$tmbstan, 2)` days), the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] implemented via the `tmbstan` package [@monnahan2018no].
 

### 2023-07-03 src/docs_paper/figures.R

@@ -99,8 +99,7 @@ figA <- (figA1 + figA2) / (figA3 + figA4 + figA5) +
 
 #' Fig B
 
-df_compare <- readRDS("depends/beta_alpha.rds") %>%
-  filter(method != "adam")
+df_compare <- readRDS("depends/beta_alpha.rds")
 
 mean <- df_compare %>%
   filter(method == "tmbstan") %>%

### 2023-07-04 src/docs_paper/paper.Rmd
Update FigA #55

@@
 Adaption occurs based on the mode (B) and covariance matrix of the target via the Cholesky decomposition (C) or spectral decompostion (D) of the inverse curvature at the mode.
-In PCA-AGHQ (E) only nodes along the first $s$ principal components are kept.
-The integrand is $f(\\btheta) = \\text{sn}(0.3 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.5 \\theta_1 - 0.3 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
+In PCA-AGHQ (E) only nodes along the first $s = 1$ principal components are kept.
+The Scree plot (F) shows that over 90% of variation is explained by keeping just the first principal component.
+The integrand is $f(\\btheta) = \\text{sn}(0.5 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.8 \\theta_1 - 0.5 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
 figA

### 2023-07-04 src/docs_paper/figures.R

@@ -1,15 +1,18 @@
-cols <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+library(tidyverse)
+library(patchwork)
 
-#' Fig A
+cols <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
 
 TMB::compile("2d.cpp")
 dyn.load(TMB::dynlib("2d"))
 
 obj <- TMB::MakeADFun(data = list(), parameters = list(theta1 = 0, theta2 = 0), DLL = "2d")
 
+box_size <- 10
+
 grid <- expand.grid(
-  theta1 = seq(-8, 8, length.out = 400),
-  theta2 = seq(-8, 8, length.out = 400)
+  theta1 = seq(-box_size, box_size, length.out = box_size * 50),
+  theta2 = seq(-box_size, box_size, length.out = box_size * 50)
 )
 
 ground_truth <- cbind(grid, pdf = apply(grid, 1, function(x) exp(-1 * obj$fn(x))))
@@ -32,7 +35,7 @@ cov <- sd_out$cov.fixed
 
 figA0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
   geom_contour(col = "lightgrey") +
-  coord_fixed(xlim = c(-8, 8), ylim = c(-8, 8), ratio = 1) +
+  coord_fixed(xlim = c(-box_size, box_size), ylim = c(-box_size, box_size), ratio = 1) +
   labs(x = "", y = "") +
   theme_minimal() +
   guides(size = "none") +
@@ -93,7 +96,22 @@ mvQuad::rescale(gg5, m = mu, C = cov, dec.type = 1)
 figA5 <- add_points(figA0, gg5) +
   labs(size = "")
 
-figA <- (figA1 + figA2) / (figA3 + figA4 + figA5) +
+#' Scree
+lambda <- eigen(cov)$values
+
+figA6 <- data.frame(
+  n = 1:length(lambda),
+  tv = cumsum(lambda / sum(lambda))
+) %>%
+  ggplot(aes(x = as.factor(n), y = tv)) +
+  geom_point() +
+  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
+  annotate("text", x = 1, y = 0.85, label = "90% of total variation explained", col = "grey") +
[…]

### 2023-07-04 src/docs_paper/script.R

@@ -1,16 +1,8 @@
 # orderly::orderly_develop_start("docs_paper")
 # setwd("src/docs_paper")
 
-#' Conversion of figures from .pdf to .jpeg at specified resolution
+#' Conversion of figures from .pdf at specified resolution
 Sys.setenv(PATH = paste(Sys.getenv("PATH"), "/opt/homebrew/bin", sep = ":"))
-
-convert_pdf_jpeg <- function(name) {
-  command <- paste0(
-    "convert -density 300 depends/", name, ".pdf -scene 1 -background white",
-    " -alpha remove -alpha off -quality 80 depends/", name, ".jpeg"
-  )
-  system(command)
-}
 
 convert_pdf_png <- function(name) {
   command <- paste0(

### 2023-07-04 src/astro/astro.Rmd
Fixes to astro example -- now understand where we are (needing a method for hyper marginals from PCA-AGHQ and EB) #46

@@
 abstract: |
-  **Background**
-
-  **Task** 
+  **Task** We reanalyse the "estimating the mass of the Milky Way" example from @bilodeau2021stochastic.
 ---
-
-We reanalyse the "estimating the mass of the Milky Way" example from @bilodeau2021stochastic.
 
@@
 
-```{r}
+```{r message=FALSE}
 ffa <- list(
@@
 
-```{r}
+```{r message=FALSE}
 tm <- Sys.time()
@@
 
-```{r}
+```{r message=FALSE}
 C <- Matrix::forceSymmetric(solve(useropt$hessian))
@@
   geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
-  annotate("text", x = 3, y = 0.875, label = "90% of total variation explained", col = "grey") +
+  annotate("text", x = 3, y = 0.89, label = "90% of total variation explained", col = "grey") +
   scale_y_continuous(labels = scales::percent) +
@@
 k_pca <- 5
-s <- 1 #' Could choose this based on the Scree plot
+s <- 1 #' Chosen based on the Scree plot
 m <- length(useropt$mode)
@@
 
-```{r}
+```{r warning=FALSE}
 #' Manual plots
@@
 
-Psi0_postplot <- Psi0pdf %>%
+(Psi0_postplot <- Psi0pdf %>%
   ggplot(aes(x = transparam, y = pdf_transparam)) +
@@
   coord_cartesian(xlim = c(24, 43)) +
-  theme_minimal()
+  theme_minimal())
 
-gamma_postplot <- gammapdf %>%
+(gamma_postplot <- gammapdf %>%
   ggplot(aes(x = transparam, y = pdf_transparam)) +
@@
   coord_cartesian(xlim = c(0.3, 0.4)) +
-  theme_minimal()
+  theme_minimal())
 
-alpha_postplot <- alphapdf %>%
+(alpha_postplot <- alphapdf %>%
   ggplot(aes(x = transparam, y = pdf_transparam)) +
@@
   coord_cartesian(xlim = c(3, 3.05)) +
-  theme_minimal()
+  theme_minimal())
 
-beta_postplot <- betapdf %>%
+(beta_postplot <- betapdf %>%
   ggplot(aes(x = transparam, y = pdf_transparam)) +
@@
   coord_cartesian(xlim = c(-0.3, 0.4)) +
-  theme_minimal()
+  theme_minimal())
+```
 
-{Psi0_postplot + gamma_postplot} / {alpha_postplot + beta_postplot}
+## Normalising constant
+
+```{r}
+data.frame(
+  method = c("AGHQ", "PCA-AGHQ", "EB"),
+  lognormconst = c(astroquad$normalized_posterior$lognormconst, astropcaquad$normalized_posterior$lognormconst, astroebquad$normalized_posterior$lognormconst)
+) %>%
+  gt::gt()
 ```
@@
 
-```{r}
+```{r warning=FALSE}
 kstable <- function(quad) {
-  aghqpostsamps_spline <- sample_marginal(astroquad, nrow(standata), interpolation = "spline")
-  aghqpostsamps_poly <- sample_marginal(astroquad, nrow(standata), interpolation = "polynomial")
+  aghqpostsamps_spline <- sample_marginal(quad, nrow(standata), interpolation = "spline")
+  aghqpostsamps_poly <- sample_marginal(quad, nrow(standata), interpolation = "polynomial")
 
@@
       Psi0 = ks.test(standata[[1]], aghqpostsamps_poly[[1]])$statistic,
-      gamma = ks.test(standata[[2]], aghqpostsamps_spline[[2]])$statistic,
+      gamma = ks.test(standata[[2]], aghqpostsamps_poly[[2]])$statistic,
       alpha = ks.test(standata[[3]], aghqpostsamps_poly[[3]])$statistic,
@@
 
-kstable_astroebquad <- kstable(astroebquad) %>%
-  pivot_longer(cols = everything()) %>%
-  mutate(method = "EB", k = k_eb, s = NA, total_points = k_eb)
+#' Not working (yet?)
+# kstable_astropcaquad <- kstable(astropcaquad) %>%
+#   pivot_longer(cols = everything()) %>%
+#   mutate(method = "AGHQ-PCA", k = k_pca, s = s, total_points = k^s)
 
-kstable_astropcaquad <- kstable(astroquad) %>%
-  pivot_longer(cols = everything()) %>%
-  mutate(method = "AGHQ-PCA", k = k_pca, s = s, total_points = k^s)
-
-bind_rows(kstable_astroquad, kstable_astroebquad, kstable_astropcaquad) %>%
+bind_rows(kstable_astroquad) %>%
   gt::gt()
 
-bind_rows(kstable_astroquad, kstable_astroebquad, kstable_astropcaquad) %>%
+bind_rows(kstable_astroquad) %>%
   ggplot(aes(x = name, y = value, col = method)) +

### 2023-07-04 src/docs_paper/appendix.Rmd
Table, fixes to bold and format of appendix #55

@@
   out.width = "95%",
-  fig.align = 'center'
+  fig.align = "center",
+  fig.pos = "!h",
+  out.extra = ""
 )
+
+knitr::opts_chunk$set()
 ```
 
-\newpage
+\clearpage
 
@@
 
-\newpage
+\clearpage
 
@@
 
-Here, we describe the simplified Naomi model [@eaton2021naomi] in more detail.
+In this section we describe the simplified Naomi model [@eaton2021naomi] in more complete mathematical detail than the main text.
 
@@
 As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
-We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities: HIV prevalence $\rho_{x, s, a} \in [0, 1]$, ART coverage $\alpha_{x, s, a} \in [0, 1]$, annual HIV incidence rate $\lambda_{x, s, a} > 0$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
+
+We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities:
+
+* HIV prevalence $\rho_{x, s, a} \in [0, 1]$,
+* ART coverage $\alpha_{x, s, a} \in [0, 1]$,
+* annual HIV incidence rate $\lambda_{x, s, a} > 0$, and
+* proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
+
 Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
-We then use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
-Let $\sum_{\{x\}}$ be shorthand for $\sum_{x \in \{x\}}$, analogously for $s$ and $a$.
+We use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
+The notation $\sum_{\{x\}}$ is used as shorthand for $\sum_{x \in \{x\}}$, and analogously for $s$ and $a$.
 
@@
 We use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
-Let $u$ be a generic random effect, with length $\dim(u)$.
-Three types of structured random effects are used in the model.
-First, we specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
-\begin{align*}
+Let $\bu$ be a generic random effect, with length $\dim(\bu) > 1$.
+Three types of structured random effects are used in the model:
+
+1. We specify the first order auto-regressive model by $\bu \sim \text{AR}1(\sigma, \phi)$ such that
+\begin{align}
 u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
-u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(u)
-\end{align*}
+u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(\bu)
+\end{align}
 where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
-Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
-$$
-u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
-$$
-where $u_{-i}$ is $u$ with the $i$th element removed i.e.
-$(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
+2. We use $\bu \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
+\begin{equation}
+u_i \, | \, \bu_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
+\end{equation}
+where $\bu_{-i}$ is $\bu$ with the $i$th element removed i.e.
+$(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(\bu)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
 We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
-Third, for the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
-$$
-u = \sigma \left( \sqrt\phi \cdot v + \sqrt{1 - \phi} \cdot w \right).
-$$
+3.
+For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $\bu \sim \text{BYM}2(\sigma, \phi)$, where $\bu$ is comprised of a spatially structured ICAR component $\bv$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $\w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
+\begin{equation}
+\bu = \sigma \left( \sqrt\phi \cdot \bv + \sqrt{1 - \phi} \cdot \w \right).
+\end{equation}
 
@@
 Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
-Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean, calculated using the `survey` \textsc{R} package [@lumley2004analysis], is given by
-$$
+Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean is given by
+\begin{equation}
 \hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
-$$
+\end{equation}
 We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
-$$
+\begin{equation}
 m^{\theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
-$$
+\end{equation}
 The observed number of indicator cases is then $y^{\theta} = m^{\theta} \cdot \hat \theta$.
+To make these calculations we used the `survey` \textsc{R} package [@lumley2004analysis]
 
@@
 \begin{equation}
-\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a}.
+\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + \bu^\rho_a + \bu_a^{\rho, s = \text{M}} + \bu^\rho_x + \bu_x^{\rho, s = \text{M}} + \bu_x^{\rho, a < 15} + \bmeta^\rho_{R_x, s, a}.
 \label{eq:prev}
@@
 $\beta_{s}^{\rho, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The difference in logit prevalence for men compared to women \\
-$u^\rho_a$ & $\text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ & Age random effects for women \\
-$u_a^{\rho, s = \text{M}}$ & $\text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ & Age random effects for the difference in logit prevalence for men compared to women age $a$ \\
-$u^\rho_x$ & $\text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ & Spatial random effects for women \\
-$u_x^{\rho, s = \text{M}}$ & $\text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ & Spatial random effects for the difference in logit prevalence for men compared to women in district $x$  \\
-$u_x^{\rho, a < 15}$ & $\text{ICAR}(\sigma_{XA}^\rho)$ & Spatial random effects for the ratio of paediatric prevalence to adult women prevalence \\
-$\eta^\rho_{R_x, s, a}$ & $-$ & Fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available \\
+$\bu^\rho_a$ & $\text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ & Age random effects for women \\
+$\bu_a^{\rho, s = \text{M}}$ & $\text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ & Age random effects for the difference in logit prevalence for men compared to women age $a$ \\
+$\bu^\rho_x$ & $\text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ & Spatial random effects for women \\
+$\bu_x^{\rho, s = \text{M}}$ & $\text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ & Spatial random effects for the difference in logit prevalence for men compared to women in district $x$  \\
+$\bu_x^{\rho, a < 15}$ & $\text{ICAR}(\sigma_{XA}^\rho)$ & Spatial random effects for the ratio of paediatric prevalence to adult women prevalence \\
+$\bmeta^\rho_{R_x, s, a}$ & $-$ & Fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available \\
 \bottomrule
@@
 
-The two BYM2 random effects $u^\rho_x$ and $u_x^{\rho, s = \text{M}}$ are comprised of the following unit scaled spatially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components, respectively
-\begin{align*}
-u_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot v_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot w_x^{\rho} \right), \\
-u_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot v_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot w_x^{\rho, s = \text{M}} \right).
-\end{align*}
+The two BYM2 random effects $\bu^\rho_x$ and $\bu_x^{\rho, s = \text{M}}$ are comprised of the following unit scaled spatially structured $\{\bv_x^{\rho}, \bv_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{\w_x^{\rho}, \w_x^{\rho, s = \text{M}}\}$ components, respectively
+\begin{align}
+\bu_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot \bv_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot \w_x^{\rho} \right), \\
+\bu_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot \bv_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot \w_x^{\rho, s = \text{M}} \right).
+\end{align}
 We use half-normal priors for the standard deviation terms
-$$
+\begin{equation}
 \{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5),
-$$
+\end{equation}
 uniform priors for the AR1 correlation parameters
-$$
+\begin{equation}
 \{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1),
-$$
+\end{equation}
 and beta priors for the BYM2 proportion parameters
-$$
+\begin{equation}
 \{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5).
-$$
+\end{equation}
 
@@
 We model ART coverage $\alpha_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
-$$
-\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a}
-$$
+\begin{equation}
+\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + \bu^\alpha_a + \bu_a^{\alpha, s = \text{M}} + \bu^\alpha_x + \bu_x^{\alpha, s = \text{M}} + \bu_x^{\alpha, a < 15} + \bmeta^\alpha_{R_x, s, a}
+\end{equation}
 with terms and priors analogous to the HIV prevalence process model in Section \ref{sec:hiv-prev} above.
@@
 \begin{equation}
-\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda.
+\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + \bu_x^\lambda + \bmeta_{R_x, s, a}^\lambda.
 \label{eq:inc}
@@
 $\omega = 0.7$ & $-$ & Average reduction in HIV transmission rate per increase in population ART coverage fixed based on inputs to the Estimation and Projection Package (EPP) model \\
-$u_x^\lambda$ & $\mathcal{N}(0, \sigma^\lambda)$  & IID spatial random effects with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ \\
-$\eta^\lambda_{R_x, s, a}$ & $-$ & Fixed log incidence rate ratios by sex and age group calculated from Spectrum model output \\
+$\bu_x^\lambda$ & $\mathcal{N}(0, \sigma^\lambda)$  & IID spatial random effects with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ \\
+$\bmeta^\lambda_{R_x, s, a}$ & $-$ & Fixed log incidence rate ratios by sex and age group calculated from Spectrum model output \\
 \bottomrule
@@
 \caption{Terms included in the linear predictor for HIV incidence (Equation \ref{eq:inc}).
-Note that the only source age structure for this part of the model are $\eta^\lambda_{R_x, s, a}$.
+Note that the only source age structure for this part of the model are $\bmeta^\lambda_{R_x, s, a}$.
 As Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80, the posterior over new infections in these age groups is exactly zero, by definition.
@@
 We model the proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ as
-$$
+\begin{equation}
 \kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right),
-$$
+\end{equation}
 where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection, and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio.
 We use an informative prior for $\Omega_T$ based on the characteristics of the recent infection testing algorithm.
-For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$, and further we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.
+For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$, and further there is assumed to be no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.
 
@@
 HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
-\begin{align*}
-\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
-\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}},
-\end{align*}
-where, for $\theta \in \{\rho, \alpha\}$, $\beta^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, 5)$ are the average differences between population and ANC outcomes, $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ are IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$, and $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$.
+\begin{align}
+\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + \bu_x^{\rho^{\text{ANC}}} + \bmeta_{R_x, a}^{\rho^{\text{ANC}}}, \label{eq:anc1} \\
+\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + \bu_x^{\alpha^{\text{ANC}}} + \bmeta_{R_x, a}^{\alpha^{\text{ANC}}} \label{eq:anc2}.
+\end{align}
+Table \ref{tab:anc} provides a description of the terms included in Equation \ref{eq:anc1} and Equation \ref{eq:anc2}.
+
+\begin{table}[h]
+\centering
+\begin{tabularx}{\textwidth}{llX}
+\toprule
+Term & Distribution & Description \\ 
+\midrule
+$\beta^{\theta^{\text{ANC}}}$ & $\mathcal{N}(0, 5)$ & Intercept giving the average difference between population and ANC outcomes \\
+$\bu_x^{\theta^{\text{ANC}}}$ & $\mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ & IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$ \\ 
+$\bmeta_{R_x, a}^{\theta^{\text{ANC}}}$ & $-$ & Offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$ \\
+\bottomrule
+\end{tabularx}
+\caption{Terms included in the linear predictors for ANC testing, with $\theta \in \{\rho, \alpha\}$ (Equation \ref{eq:prev}).}
+\label{tab:anc}
+\end{table}
 
@@
 $$
-\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi,
+\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + \bu_x^\psi,
 $$
-where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
-Here we fix $\beta^\psi = u_x^\psi = 0$ such that $\Psi_{x, a}$ are simply constants.
+where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $\bu_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
+Here we fix $\beta^\psi = 0$ and $\bu_x^\psi = \mathbf{0}$ such that $\Psi_{x, a}$ are simply constants.
 
@@
 \begin{equation}
-\tilde \gamma_{x, x'} = \log \left( \frac{\gamma_{x, x'}}{1 - \gamma_{x, x'}} \right) = \tilde \gamma_0 + u_x^{\tilde \gamma}, \label{eq:log-or}
+\tilde \gamma_{x, x'} = \log \left( \frac{\gamma_{x, x'}}{1 - \gamma_{x, x'}} \right) = \tilde \gamma_0 + \bu_x^{\tilde \gamma}, \label{eq:log-or}
 \end{equation}
-where $\tilde \gamma_0 = -4$ is a fixed intercept, and $u_x^{\tilde \gamma} \sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})$ are district random effects with $\sigma_X^{\tilde \gamma} \sim \mathcal{N}^+(0, 2.5)$.
+where $\tilde \gamma_0 = -4$ is a fixed intercept, and $\bu_x^{\tilde \gamma} \sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})$ are district random effects with $\sigma_X^{\tilde \gamma} \sim \mathcal{N}^+(0, 2.5)$.
 Note that Equation \ref{eq:log-or} does not depend on $x'$, such that $\gamma_{x, x'}$ is only a function of $x$.
@@
 We fix $\tilde \gamma_{x, x} = 0$ and recover the multinomial probabilities using the softmax
-$$
+\begin{equation}
 \gamma_{x, x'} = \frac{\exp(\tilde \gamma_{x, x'})}{\sum_{x^\star \in \{x, \text{ne}(x)\}} \exp(\tilde \gamma_{x, x^\star})}.
-$$
+\end{equation}
 Given the total number of PLHIV on ART $A_{x, s, a} = N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}$, the number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a} = A_{x, s, a} \cdot \gamma_{x, x'}$, and the total attending ART facilities in district $x'$ are
-$$
+\begin{equation}
 \tilde A_{x', s, a} = \sum_{x \in \{x', \text{ne}(x')\}} A_{x, x', s, a}.
-$$
+\end{equation}
 
@@
 For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the most recent household survey furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
-$$
+\begin{equation}
 y^{\theta}_{\{x\}, \{s\}, \{a\}} = m^{\theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}.
-$$
-To model these observations, we use following three binomial working likelihoods
-\begin{align*}
-y^{\rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
-\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}, \\
-y^{\alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
-\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}, \\
-y^{\kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
-\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
-\end{align*}
-The generalised binomial $y \sim \text{xBin}(m, p)$ is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
-$$
+\end{equation}
+To model these observations, we use the following three binomial working likelihoods:
+
+1. The number of people living with HIV (PLHIV) is
+\begin{equation}
+y^{\rho}_{\{x\}, \{s\}, \{a\}} \sim \text{xBin}(m^{\rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}),
+\end{equation}
+where
+\begin{equation}
+\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}
+\end{equation}
+has denominator given by the total population size.
+2. The number of PLHIV on ART is
+\begin{equation}
+y^{\alpha}_{\{x\}, \{s\}, \{a\}} \sim \text{xBin}(m^{\alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}),
+\end{equation}
+where
+\begin{equation}
+\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}
+\end{equation}
+has denominator given by the total PLHIV.
+3. The number of PLHIV recently infected
+\begin{equation}
+y^{\kappa}_{\{x\}, \{s\}, \{a\}} \sim \text{xBin}(m^{\kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}),
+\end{equation}
+where
+\begin{equation}
+\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}
+\end{equation}
+has denominator given by the total PLHIV.
+
+The generalised binomial $y \sim \text{xBin}(m, p)$ used above is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
+\begin{equation}
 \log p(y) = \log \Gamma(m + 1) - \log \Gamma(y + 1) - \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
-$$
+\end{equation}
 where the gamma function $\Gamma$ is such that $\forall n \in \mathbb{N}$, $\Gamma(n) = (n - 1)!$.
@@
 Let $W^\text{ANC}_{\{x\}}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}}$ the number of ANC clients already on ART prior to first ANC, such that
-$$
+\begin{equation}
 W^\text{ANC}_{x} \geq X^\text{ANC}_{x} \geq Y^\text{ANC}_{x} \geq Z^\text{ANC}_{x},
-$$
+\end{equation}
 for all $x \in \{x\}$.
@@
 The observed number of HIV positive and already on ART among ANC clients is modelled by
-\begin{align*}
+\begin{align}
 Y_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} }^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right), \\
 Z_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} }^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right),
-\end{align*}
+\end{align}
 where prevalence and ART coverage are aggregated by the number of pregnant women $\Psi_{x, a}$
-\begin{align*}
+\begin{align}
 \rho_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a}}, \\
 \alpha_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}.
-\end{align*}
+\end{align}
 
@@
 1.
-If survey data on HIV prevalence or ART coverage by age and sex are not available then we set $u_a^\theta = 0$ and $u_{a, s = \text{M}}^\theta = 0$ and use the average age-sex pattern of from the Spectrum offset $\eta_{R_x, s, a}^\theta$.
+If survey data on HIV prevalence or ART coverage by age and sex are not available then we set $\bu_a^\theta = 0$ and $\bu_{a, s = \text{M}}^\theta = 0$ and use the average age-sex pattern of from the Spectrum offset $\bmeta_{R_x, s, a}^\theta$.
 For the Malawi example considered in the main text HIV prevalence and ART coverage data are not available for those aged 65+.
@@
 If no ART data, either survey or ART programme, are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
-In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha$.
+In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = \bu_x^\alpha + \bmeta_{R_x, s, a}^\alpha$.
 3.
-If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$.
+If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = 0$ and $\bu_x^\lambda = \mathbf{0}$.
 The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.
@@
 \texttt{beta\_anc\_alpha} & $\beta^{\alpha^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ & & & \\
-\texttt{u\_rho\_x} & $w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{us\_rho\_x} & $v^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{u\_rho\_xs} & $w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{us\_rho\_xs} & $v_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{u\_rho\_a} & $u^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
-\texttt{u\_rho\_as} & $u_a^{\rho, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
-\texttt{u\_rho\_xa} & $u_x^{\rho, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{u\_alpha\_x} & $w^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ && \cmark & \\
-\texttt{us\_alpha\_x} & $v^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
-\texttt{u\_alpha\_xs} & $w_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
-\texttt{us\_alpha\_xs} & $v_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
-\texttt{u\_alpha\_a} & $u^\alpha_a$ & Latent & 13 & $\mathbb{R}^{13}$ & & \cmark & \\
-\texttt{u\_alpha\_as} & $u_a^{\alpha, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ & & \cmark & \\
-\texttt{u\_alpha\_xa} & $u_x^{\alpha, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
-\texttt{ui\_lambda\_x} & $u_x^\lambda$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \cmark \\
-\texttt{ui\_anc\_rho\_x} & $u_x^{\rho^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
-\texttt{ui\_anc\_alpha\_x} & $u_x^{\alpha^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
-\texttt{log\_or\_gamma} & $u_x^{\tilde \gamma}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
+\texttt{u\_rho\_x} & $\w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{us\_rho\_x} & $\bv^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{u\_rho\_xs} & $\w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{us\_rho\_xs} & $\bv_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{u\_rho\_a} & $\bu^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
+\texttt{u\_rho\_as} & $\bu_a^{\rho, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
+\texttt{u\_rho\_xa} & $\bu_x^{\rho, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
+\texttt{u\_alpha\_x} & $\w^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ && \cmark & \\
+\texttt{us\_alpha\_x} & $\bv^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
+\texttt{u\_alpha\_xs} & $\w_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
+\texttt{us\_alpha\_xs} & $\bv_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
+\texttt{u\_alpha\_a} & $\bu^\alpha_a$ & Latent & 13 & $\mathbb{R}^{13}$ & & \cmark & \\
+\texttt{u\_alpha\_as} & $\bu_a^{\alpha, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ & & \cmark & \\
+\texttt{u\_alpha\_xa} & $\bu_x^{\alpha, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
+\texttt{ui\_lambda\_x} & $\bu_x^\lambda$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \cmark \\
+\texttt{ui\_anc\_rho\_x} & $\bu_x^{\rho^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
+\texttt{ui\_anc\_alpha\_x} & $\bu_x^{\alpha^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
+\texttt{log\_or\_gamma} & $\bu_x^{\tilde \gamma}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
 \bottomrule
@@
 
-<!-- \newpage -->
+<!-- \clearpage -->
 
@@
 
-\newpage
+\clearpage
+
+# AGHQ and PCA-AGHQ details
+
+## Additional figures
+
+```{r marginal-sd, fig.cap="Hyperparameters on the logit-scale had systematically higher marginal standard deviations than those on the log-scale.
+This is because variation on the real scale is compressed by the inverse logit and expanded by the inverse log (exponential)."}
+knitr::include_graphics("depends/marginal-sd.png")
+```
+
+```{r nodes-quantiles-sd, fig.cap="Standard deviations of the quantiles of the quadrature nodes within the NUTS posterior draws varied substantially, in accordance with the marginal standard deviations shown in Figure \\ref{fig:marginal-sd}."}
+knitr::include_graphics("depends/nodes-quantiles-sd.png")
+```
+
+```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
+knitr::include_graphics("depends/tv-plot.png")
+```
+
+```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank matrix."}
+knitr::include_graphics("depends/reduced-rank-plot.png")
+```
+
+## Estimated normalising constant comparison
+
+Add here plots and description of the estimated normalising constant for different PCA-AGHQ settings.
+
+\clearpage
+
+# Comprehensive inference comparison
+
+In this section we give further inference comparison results, parameter by parameter.
+
+## Point estimates
+
+## Distribution tests
+
+```{r}
+ks_summary <- readRDS("depends/ks-summary.rds")
+
+ks_summary %>%
+  ungroup() %>%
+  filter(type == "Latent field") %>%
+  select(-type) %>%
+  rename(`PCA-AGHQ` = `KS(aghq, tmbstan)`, `TMB` = `KS(TMB, tmbstan)`) %>%
+  gt::gt(rowname_col = "Parameter") %>%
+  gt::fmt_number(
+    columns = c(`PCA-AGHQ`, `TMB`),
+    decimals = 3
+  ) %>%
+  gt::summary_rows(
+   columns = everything(),
+   fns = list(Average = ~ mean(.x)),
+   formatter = gt::fmt_number,
+   decimals = 3
+  )
+```
+
+## Pareto-smoothed importance sampling
+
+## Maximum mean discrepancy
+
+\clearpage
 
@@
 
-\newpage
-
-# AGHQ and PCA-AGHQ details
-
-## Additional figures
-
-```{r marginal-sd, fig.cap="Hyperparameters on the logit-scale had systematically higher marginal standard deviations than those on the log-scale.
-This is because variation on the real scale is compressed by the inverse logit and expanded by the inverse log (exponential)."}
-knitr::include_graphics("depends/marginal-sd.png")
-```
-
-```{r nodes-quantiles-sd, fig.cap="Standard deviations of the quantiles of the quadrature nodes within the NUTS posterior draws varied substantially, in accordance with the marginal standard deviations shown in Figure \\ref{fig:marginal-sd}."}
-knitr::include_graphics("depends/nodes-quantiles-sd.png")
-```
-
-```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
-knitr::include_graphics("depends/tv-plot.png")
-```
-
-```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank matrix."}
-knitr::include_graphics("depends/reduced-rank-plot.png")
-```
-
-## Estimated normalising constant comparison
-
-Add here plots and description of the estimated normalising constant for different PCA-AGHQ settings.
-
-\newpage
-
-# Further inference comparison results
-
-## Point estimates
-
-## Distribution tests
-
-```{r}
-ks_summary <- readRDS("depends/ks-summary.rds")
-
-ks_summary %>%
-  ungroup() %>%
-  filter(type == "Latent field") %>%
-  select(-type) %>%
-  gt::gt(rowname_col = "Parameter") %>%
-  gt::summary_rows(
-   columns = everything(),
-   fns = list(Average = ~ mean(.x)),
-   formatter = gt::fmt_number,
-   decimals = 3
-  )
-```
-
-## Pareto-smoothed importance sampling
-
-## Maximum mean discrepancy
-
-\newpage
+\clearpage
 
@@
 1. Calculate the mode, Hessian at the mode, lower Cholesky, and Laplace approximation
-\begin{align*}
+\begin{align}
 \hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
@@
 \tilde p_\texttt{LA}(\btheta, \y) &= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)},
-\end{align*}
+\end{align}
 where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\mathbf{H}}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
-\begin{align*}
+\begin{align}
 \hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
 \hat{\mathbf{H}}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
-\end{align*}
+\end{align}
 2. Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension.
@@
 For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
-\begin{align*}
+\begin{align}
 \hat \x_{-i}(x_i, \btheta) &= \argmax_{\x_{-i}} \log p(\y, x_i, \x_{-i}, \btheta), \\
 \hat{\mathbf{H}}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
-\end{align*}
+\end{align}
 where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ can be initialised at $\hat \x(\btheta)_{-i}$.

### 2023-07-04 src/docs_paper/paper.Rmd
Correct caption error #55

@@
 In PCA-AGHQ (E) only nodes along the first $s = 1$ principal components are kept.
-The Scree plot (F) shows that over 90% of variation is explained by keeping just the first principal component.
+The Scree plot (F) shows that over 90\\% of variation is explained by keeping just the first principal component.
 The integrand is $f(\\btheta) = \\text{sn}(0.5 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.8 \\theta_1 - 0.5 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
@@
 ```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted with TMB."}
-knitr::include_graphics("depends/naomi_results.png")
+knitr::include_graphics("depends/naomi-results.png")
 ```

### 2023-07-04 src/dev_scale-grid/scale-grid.Rmd
Simplify and improve scale grid notebook! Add PC loadings plot! Start moving unit tests out into check report

@@
 
-The names of the hyperparameters are:
-
 ```{r}
-(hypers <- names(fit$par))
+hypers <- names(fit$par)
 ```
 
-There are `r length(hypers)` of them, if you'd like to count.
-Including `r data.frame(hypers) %>% filter(str_detect(hypers, "^logit_")) %>% nrow()` on the logit scale, and `r data.frame(hypers) %>% filter(str_detect(hypers, "^log_")) %>% nrow()` on the log scale.
+There are `r length(hypers)` hyperparameters, comprised ofa `r data.frame(hypers) %>% filter(str_detect(hypers, "^logit_")) %>% nrow()` on the logit scale, and `r data.frame(hypers) %>% filter(str_detect(hypers, "^log_")) %>% nrow()` on the log scale.
 
@@
 
-(marginal_sd_plot <- data.frame(par = names(sd), sd = unname(sd)) %>%
+data.frame(par = names(sd), sd = unname(sd)) %>%
   mutate(
@@
     theme_minimal() +
-    labs(x = "", y = "Marginal standard deviation", col = "Scale"))
+    labs(x = "", y = "Marginal standard deviation", col = "Scale")
 
-ggsave("marginal-sd.png", plot = marginal_sd_plot, height = 3.5, w = 6.25)
+ggsave("marginal-sd.png", height = 3.5, w = 6.25)
 ```
@@
 
-The total number of points for this grid is 2:
-
-```{r}
-prod(base_grid$level)
-```
-
+The total number of points for this grid is `r prod(base_grid$level)`.
 Let's try fitting this, and see how long it takes.
@@
     geom_point(alpha = 0.8) +
+    coord_flip() +
     labs(x = "", y = "Value") + 
-    theme_minimal() +
-    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))
+    theme_minimal()
 ```
 
-Interesting to note that for the Cholesky adapted grid with `levels = c(1, 1, ..., 2, 1, ..., 1)` we see variation in all of the dimensions greater than or equal to the 16th.
+Interesting to note that for the Cholesky adapted grid with `levels = c(1, 1, ..., 2, 1, ..., 1)` there is variation in all of the dimensions greater than or equal to the 16th.
 This is related to the Cholesky adaption being based on a lower triangular matrix.
-To verify this, we can look at the differences between the two grid points:
+To verify this, we can look at the differences between the two grid points.
+For many of the dimensions the difference is too small to see on the plot above.
 
 ```{r}
-diff(as.matrix(quad$modesandhessians[, 1:24]))
+t(unname(diff(as.matrix(quad$modesandhessians[, 1:24]))))
 ```
@@
 ```{r}
-#' This is what the base grid looks like
-mvQuad::getNodes(base_grid)
-
 base_grid_copy <- rlang::duplicate(base_grid)
@@
 mvQuad::rescale(base_grid_copy, m = quad$optresults$mode, C = Matrix::forceSymmetric(solve(quad$optresults$hessian)), dec.type = 1)
-diff(mvQuad::getNodes(base_grid_copy))
+t(diff(mvQuad::getNodes(base_grid_copy)))
 ```
@@
 Perhaps ideally we would like to know how much variation in each hyperparameter effects variation in model outputs, and then take that into account in the importance we place on each dimension.
-
-## Points of confusion
-
-How does this approach interact with the rescaling which occurs in adaptive quadrature?
-That is to say $z \mapsto \hat \theta + Lz$ with $LL^\top = H$.
-Do each of the dimensions of $z$ reflect dimensions of $Lz$?
 
@@
 
-(tv_plot <- ggplot(tv_df, aes(x = n, y = tv)) +
+ggplot(tv_df, aes(x = n, y = tv)) +
   geom_point() +
@@
   labs(x = "PCA dimensions included", y = "Total variation explained") +
-  theme_minimal())
+  theme_minimal()
 
-ggsave("tv-plot.png", plot = tv_plot, h = 3, w = 6.25)
+ggsave("total-variation.png", h = 3, w = 6.25)
 ```
@@
 ```{r pc-loadings, class.source = 'fold-hide', fig.cap="PC loadings."}
+rownames(E) <- hypers
+
 reshape2::melt(as.matrix(E)) %>%
-  ggplot(aes(x = Var1, y = Var2, fill = value)) +
-  labs(x = "", y = "", fill = "E[i, j]") +
+  ggplot(aes(x = Var1, y = factor(Var2), fill = value)) +
+  labs(x = "", y = "", fill = "") +
   geom_tile() +
-  scale_fill_viridis_c() +
-  labs(x = "PC loading", y = "Hyper") +
-  theme_minimal()
+  coord_flip() +
+  scale_fill_gradientn(
+    colours = c("#E69F00", "white", "#009E73"),
+    rescaler = ~ scales::rescale_mid(.x, mid = 0),
+    limits = c(-1.1, 1.1)
+  ) +
+  labs(x = "Hyper", y = "Principal component loading") +
+  theme_minimal() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+
+ggsave("pc-loadings.png", h = 3.5, w = 6.25)
 ```
@@
 
-```{r cov-reconstruction, class.source = 'fold-hide', fig.cap="PC loadings."}
+```{r cov-reconstruction, class.source = 'fold-hide', fig.cap="Hessian matrix recons."}
 E_s <- E[, 1:s]
@@
 
-(reduced_rank_plot <- {C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + labs(subtitle = "Reduced rank")})
+{C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + labs(subtitle = "Reduced rank")}
 
-ggsave("reduced-rank-plot.png", reduced_rank_plot, h = 3.5, w = 6.25)
-```
-
-Create a function to do the PCA rescaling, which also adapts according to the mean and reweights the nodes:
-
-```{r}
-#' @param m Mean vector
-#' @param C Covariance matrix
-#' @param s Small grid dimension
-#' @param k Number of points per small grid dimension
-pca_rescale <- function(m, C, s, k) {
-  d <- nrow(C)
-  stopifnot(d == length(m))
-  eigenC <- eigen(C)
-  lambda <- eigenC$values
-  Lambda <- diag(lambda)
-  E <- eigenC$vectors
-  E_s <- E[, 1:s] 
-  gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = k) 
-  nodes_out <- t(E_s %*% diag(lambda[1:s]^{0.5}, ncol = s) %*% t(mvQuad::getNodes(gg_s)))
-  for(j in 1:d) nodes_out[, j] <- nodes_out[, j] + m[j]
-  weights_out <- mvQuad::getWeights(gg_s) * as.numeric(mvQuad::getWeights(mvQuad::createNIGrid(dim = d - s, type = "GHe", level = 1)))
-
-  # Putting things into a mvQuad format manually
-  gg <- mvQuad::createNIGrid(dim = d, type = "GHe", level = 1)
-  gg$level <- rep(NA, times = d)
-  gg$ndConstruction <- "PCA"
-  gg$nodes <- nodes_out
-  gg$weights <- weights_out
-  return(gg)
-}
-```
-
-## Testing with simple examples
-
-Here we run some unit tests to make sure, visually and numerically, that the methods are working as expected.
-
-### Gaussian 2D function
-
-A simple 2D Gaussian example:
-
-$$
-\begin{pmatrix}
-x_1 \\
-x_2
-\end{pmatrix}
-\sim
-\mathcal{N}
-\left(
-\begin{pmatrix}
-1 \\
-3/2
-\end{pmatrix},
-\begin{pmatrix}
-2 & 1 \\
-1 & 1
-\end{pmatrix}
-\right)
-$$
-
-```{r gaussian-2d, class.source = 'fold-hide', fig.cap="Gaussian 2D objective function."}
-m <- c(1, 1.5)
-C <- matrix(c(2, 1, 1, 1), ncol = 2)
-
-obj <- function(theta) {
-  mvtnorm::dmvnorm(theta, mean = m, sigma = C, log = TRUE)
-}
-
-ff <- list(
-  fn = obj,
-  gr = function(theta) numDeriv::grad(obj, theta),
-  he = function(theta) numDeriv::hessian(obj, theta)
-)
-
-grid <- expand.grid(
-  theta1 = seq(-2, 5, length.out = 700),
-  theta2 = seq(-2, 5, length.out = 700)
-)
-
-ground_truth <- cbind(grid, pdf = exp(obj(grid)))
-
-plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
-  geom_contour(col = multi.utils::cbpalette()[1]) +
-  coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
-  labs(x = "", y = "") +
-  theme_minimal()
-
-plot
-```
-
-Define four AGHQ-PCA grids, with `k` running from 1 to 7:
-
-```{r}
-pca_grid_1 <- pca_rescale(m = m, C = C, s = 1, k = 1)
-pca_grid_3 <- pca_rescale(m = m, C = C, s = 1, k = 3)
-pca_grid_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
-pca_grid_7 <- pca_rescale(m = m, C = C, s = 1, k = 7)
-```
-
-For example, the version with 3 points looks as follows:
-
-```{r pca-grid-3, class.source = 'fold-hide', fig.cap="AGHQ-PCA grid for Gaussian with $k = 3$."}
-plot_points <- function(gg) {
-  plot +
-    geom_point(
-      data = mvQuad::getNodes(gg) %>%
-        as.data.frame() %>%
-        mutate(weights = mvQuad::getWeights(gg)),
-      aes(x = V1, y = V2, size = weights),
-      alpha = 0.8,
-      col = multi.utils::cbpalette()[2],
-      inherit.aes = FALSE
-    ) +
-    scale_size_continuous(range = c(1, 2)) +
-    labs(x = "", y = "", size = "Weight") +
-    theme_minimal()
-}
-
-plot_points(pca_grid_3)
-```
-
-We can optimise the objective function using the BFGS algorithm to recover the mode and Hessian at the mode.
-Of course, for this example, these are known in advance as we are dealing with a Gaussian, but in general this is how they will be computed.
-
-```{r}
-opt_bfgs <- aghq::optimize_theta(ff, m, control = default_control(method = "BFGS"))
-opt_bfgs$mode
-opt_bfgs$hessian
-Matrix::forceSymmetric(solve(opt_bfgs$hessian))
-```
-
-Now we can perform regular AGHQ with a product grid (again with a range of values for `k`):
-
-```{r}
-norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, k = 1)
-norm_bfgs_3 <- aghq::normalize_logpost(opt_bfgs, k = 3)
-norm_bfgs_5 <- aghq::normalize_logpost(opt_bfgs, k = 5)
-norm_bfgs_7 <- aghq::normalize_logpost(opt_bfgs, k = 7)
-norm_bfgs <- list(norm_bfgs_1, norm_bfgs_3, norm_bfgs_5, norm_bfgs_7)
-```
-
-As well as the the AGHQ-PCA grids:
-
-```{r}
-norm_pca_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_1, adapt = FALSE)
-norm_pca_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_3, adapt = FALSE)
-norm_pca_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_5, adapt = FALSE)
-norm_pca_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_7, adapt = FALSE)
-norm_pca_bfgs <- list(norm_pca_bfgs_1, norm_pca_bfgs_3, norm_pca_bfgs_5, norm_pca_bfgs_7)
-```
-
-And a second way of producing the AGHQ-PCA grids directly using `mvQuad`:
-
-```{r, fig.cap="Alternative approach AGHQ-PCA grid for Gaussian with $k = 3$."}
-pca2_grid_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(1, 1))
-mvQuad::rescale(pca2_grid_1, m = m, C = C, dec.type = 1)
-
-pca2_grid_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(3, 1))
-mvQuad::rescale(pca2_grid_3, m = m, C = C, dec.type = 1)
-
-pca2_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(5, 1))
-mvQuad::rescale(pca2_grid_5, m = m, C = C, dec.type = 1)
-
-pca2_grid_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(7, 1))
-mvQuad::rescale(pca2_grid_7, m = m, C = C, dec.type = 1)
-
-plot_points(pca2_grid_3)
-
-norm_pca2_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_1, adapt = FALSE)
-norm_pca2_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_3, adapt = FALSE)
-norm_pca2_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_5, adapt = FALSE)
-norm_pca2_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_7, adapt = FALSE)
-norm_pca2_bfgs <- list(norm_pca2_bfgs_1, norm_pca2_bfgs_3, norm_pca2_bfgs_5, norm_pca2_bfgs_7)
-```
-
-The density is already normalised, so it has a normalising constant of 1, and a log normalising constant of $\log(1) = 0$:
-
-```{r, class.source = 'fold-hide', fig.cap="Unit test for the Gaussian example"}
-truelognormconst <- 0
-
-results <- data.frame(
-  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7)), paste0("aghq-pca2, k = ", c(1, 3, 5, 7))),
-  "lognormconst" = c(truelognormconst, sapply(c(norm_bfgs, norm_pca_bfgs, norm_pca2_bfgs), function(x) x$lognormconst)),
-  "type" = c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4), rep("AGHQ-PCA2", 4))
-)
-
-ggplot(results, aes(x = method, y = lognormconst, col = as.factor(type))) +
-  geom_point(size = 1) +
-  labs(x = "Method", y = "Log normalising constant", col = "") +
-  scale_color_manual(values = multi.utils::cbpalette()) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 70, vjust = 0.5, hjust = 0.5))
-```
-
-### Non-Gaussian 2D function
-
-A more complicated two dimensional function from Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) for the `aghq` package.
-
-```{r non-gaussian, class.source = 'fold-hide', fig.cap="Non-Gaussian 2D objective function."}
-set.seed(84343124)
-
-logfteta <- function(eta, y) {
-  n <- length(y)
-  n1 <- ceiling(n / 2)
-  n2 <- floor(n / 2)
-  y1 <- y[1:n1]
-  y2 <- y[(n1 + 1):(n1 + n2)]
-  eta1 <- eta[1]
-  eta2 <- eta[2]
-  sum(y1) * eta1 - (length(y1) + 1) * exp(eta1) - sum(lgamma(y1+1)) + eta1 +
-    sum(y2) * eta2 - (length(y2) + 1) * exp(eta2) - sum(lgamma(y2+1)) + eta2
-}
-
-n1 <- 5
-n2 <- 5
-n <- n1 + n2
-y1 <- rpois(n1, 5)
-y2 <- rpois(n2, 5)
-
-truemode <- c(log((sum(y1) + 1) / (length(y1) + 1)), log((sum(y2) + 1) / (length(y2) + 1)))
-truelogint <- function(y) lgamma(1 + sum(y)) - (1 + sum(y)) * log(length(y) + 1) - sum(lgamma(y + 1))
-truelognormconst <- truelogint(y1) + truelogint(y2)
-
-obj <- function(x) logfteta(x, c(y1, y2))
-
-ff <- list(
-  fn = obj,
-  gr = function(x) numDeriv::grad(obj, x),
-  he = function(x) numDeriv::hessian(obj, x)
-)
-
-grid <- expand.grid(
-  theta1 = seq(0, 4, length.out = 400),
-  theta2 = seq(0, 4, length.out = 400)
-)
-
-ground_truth <- data.frame(grid) %>%
-  rowwise() %>%
-  mutate(pdf = exp(obj(x = c(theta1, theta2))))
-
-plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
-  geom_contour(col = multi.utils::cbpalette()[1]) +
-  coord_fixed(xlim = c(0.8, 2), ylim = c(0.8, 2), ratio = 1) +
-  labs(x = "", y = "") +
-  theme_minimal()
-
-plot
-```
-
-Optimise using BFSG:
-
-```{r}
-opt_bfgs <- aghq::optimize_theta(ff, c(1.5, 1.5), control = default_control(method = "BFGS"))
-(m <- opt_bfgs$mode)
-(C <- Matrix::forceSymmetric(solve(opt_bfgs$hessian)))
-eigen(C)
-```
-
-```{r, fig.cap="AGHQ grid for non-Gaussian with $k = 3$."}
-grid_bfgs_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
-mvQuad::rescale(grid_bfgs_1, m = m, C = C)
-
-grid_bfgs_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 3)
-mvQuad::rescale(grid_bfgs_3, m = m, C = C)
-
-grid_bfgs_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 5)
-mvQuad::rescale(grid_bfgs_5, m = m, C = C)
-
-grid_bfgs_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 7)
-mvQuad::rescale(grid_bfgs_7, m = m, C = C)
-
-norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, 1, 1)
-norm_bfgs_3 <- aghq::normalize_logpost(opt_bfgs, 3, 1)
-norm_bfgs_5 <- aghq::normalize_logpost(opt_bfgs, 5, 1)
-norm_bfgs_7 <- aghq::normalize_logpost(opt_bfgs, 7, 1)
-norm_bfgs <- list(norm_bfgs_1, norm_bfgs_3, norm_bfgs_5, norm_bfgs_7)
-
-plot_points(grid_bfgs_3)
-```
-
-```{r, fig.cap="AGHQ-PCA grid for non-Gaussian with $k = 3$."}
-pca_grid_bfgs_1 <- pca_rescale(m = m, C = C, s = 1, k = 1)
-pca_grid_bfgs_3 <- pca_rescale(m = m, C = C, s = 1, k = 3)
-pca_grid_bfgs_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
-pca_grid_bfgs_7 <- pca_rescale(m = m, C = C, s = 1, k = 7)
-
-norm_pca_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_1, adapt = FALSE)
-norm_pca_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_3, adapt = FALSE)
-norm_pca_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_5, adapt = FALSE)
-norm_pca_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_7, adapt = FALSE)
-norm_pca_bfgs <- list(norm_pca_bfgs_1, norm_pca_bfgs_3, norm_pca_bfgs_5, norm_pca_bfgs_7)
-
-plot_points(pca_grid_bfgs_3)
-```
-
-```{r, fig.cap="Alternative approach AGHQ-PCA grid for non-Gaussian with $k = 3$."}
-pca2_grid_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(1, 1))
-mvQuad::rescale(pca2_grid_1, m = m, C = C, dec.type = 1)
-
-pca2_grid_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(3, 1))
-mvQuad::rescale(pca2_grid_3, m = m, C = C, dec.type = 1)
-
-pca2_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(5, 1))
-mvQuad::rescale(pca2_grid_5, m = m, C = C, dec.type = 1)
-
-pca2_grid_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = c(7, 1))
-mvQuad::rescale(pca2_grid_7, m = m, C = C, dec.type = 1)
-
-norm_pca2_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_1, adapt = FALSE)
-norm_pca2_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_3, adapt = FALSE)
-norm_pca2_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_5, adapt = FALSE)
-norm_pca2_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca2_grid_7, adapt = FALSE)
-norm_pca2_bfgs <- list(norm_pca2_bfgs_1, norm_pca2_bfgs_3, norm_pca2_bfgs_5, norm_pca2_bfgs_7)
-
-plot_points(pca2_grid_3)
-```
-
-```{r, class.source = 'fold-hide', fig.cap="Unit test for the non-Gaussian example"}
-results <- data.frame(
-  "method" = c("Truth", paste0("aghq, k = ", c(1, 3, 5, 7)), paste0("aghq-pca, k = ", c(1, 3, 5, 7)), paste0("aghq-pca2, k = ", c(1, 3, 5, 7))),
-  "lognormconst" = c(truelognormconst, sapply(c(norm_bfgs, norm_pca_bfgs, norm_pca2_bfgs), function(x) x$lognormconst)),
-  "type" = as.factor(c("Truth", rep("AGHQ", 4), rep("AGHQ-PCA", 4), rep("AGHQ-PCA2", 4)))
-)
-
-ggplot(results, aes(x = method, y = lognormconst, col = type)) +
-  geom_point(size = 1) +
-  labs(x = "Method", y = "Log normalising constant", col = "") +
-  scale_color_manual(values = multi.utils::cbpalette()) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 70, vjust = 0.5, hjust = 0.5))
+ggsave("reduced-rank.png", h = 3.5, w = 6.25)
 ```
@@
 
-gg3 <- pca_rescale(m = m, C = C, s = 3, k = 2)
+gg3 <- pca_rescale(m = m, C = C, s = 3, k = 3)
 nrow(mvQuad::getNodes(gg3))
 
-gg5 <- pca_rescale(m = m, C = C, s = 5, k = 2)
+gg5 <- pca_rescale(m = m, C = C, s = 5, k = 3)
 nrow(mvQuad::getNodes(gg5))
 
-gg7 <- pca_rescale(m = m, C = C, s = 7, k = 2)
+gg7 <- pca_rescale(m = m, C = C, s = 7, k = 3)
 nrow(mvQuad::getNodes(gg7))
 
-gg9 <- pca_rescale(m = m, C = C, s = 9, k = 2)
+gg9 <- pca_rescale(m = m, C = C, s = 9, k = 3)
 nrow(mvQuad::getNodes(gg9))
 ```
+
+# Dealing with the scales issue
+
+Put `logit` and `log` on parity somehow!
+
+```{r}
+#' TODO!
+```

### 2023-07-04 src/docs_paper/appendix.Rmd
Add PC loadings plot to appendix #55

@@
 ```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
-knitr::include_graphics("depends/tv-plot.png")
+knitr::include_graphics("depends/total-variation.png")
+```
+
+```{r pc-loadings, fig.cap="Principal component loadings."}
+knitr::include_graphics("depends/pc-loadings.png")
 ```
@@
 ```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank matrix."}
-knitr::include_graphics("depends/reduced-rank-plot.png")
+knitr::include_graphics("depends/reduced-rank.png")
 ```

### 2023-07-05 src/dev_scale-grid/scale-grid.Rmd
Extend and simplify scale grid development, including first try at dealing with scaling #58

@@
 The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical.
-In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using principle components analysis.
+In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using different varieties of principal components analysis.
 editor_options: 
@@
 C <- sd_out$cov.fixed
-C_df <- reshape2::melt(as.matrix(C))
-
-C_plot <- C_df %>%
-  ggplot(aes(x = Var1, y = Var2, fill = value)) +
-  labs(x = "", y = "", fill = "C[i, j]") +
-  geom_tile() +
-  scale_fill_viridis_c() +
-  labs(x = "i", y = "j") +
-  theme_minimal() +
-  theme(
-    axis.text.x = element_blank(),
-    axis.text.y = element_blank()
-  )
-
-C_plot
+(C_plot <- plot_matrix(C))
 ```
@@
   ggplot(aes(x = reorder(name, id), y = value)) +
-    geom_point(alpha = 0.8) +
+    geom_jitter(shape = 1, height = 0, width = 0.2) +
     coord_flip() +
@@
 ```{r scree, class.source = 'fold-hide', fig.cap="Scree plot."}
-tv_df <- data.frame(
-  n = 1:length(lambda),
-  tv = cumsum(lambda / sum(lambda))
-)
-
-ggplot(tv_df, aes(x = n, y = tv)) +
-  geom_point() +
-  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
-  annotate("text", x = 20, y = 0.875, label = "90% of total variation explained", col = "grey") +
-  scale_y_continuous(labels = scales::percent) +
-  labs(x = "PCA dimensions included", y = "Total variation explained") +
-  theme_minimal()
+plot_total_variation(eigenC, label_x = 20)
 
 ggsave("total-variation.png", h = 3, w = 6.25)
+
+tv <- cumsum(eigenC$values) / sum(eigenC$values)
 ```
 
-Based on Figure \@ref(fig:scree) with 5 dimensions included, we can explain `r round(100 * filter(tv_df, n == 5) %>% pull(tv), 1)`% of the total variation.
-Or, with 10 dimensions included, that percentage increases to `r round(100 * filter(tv_df, n == 10) %>% pull(tv), 1)`%.
+Based on Figure \@ref(fig:scree) with 5 dimensions included, we can explain `r signif(100 * tv[5], 3)`% of the total variation.
+Or, with 10 dimensions included, that percentage increases to `r signif(100 * tv[10], 3)`%.
 
@@
 rownames(E) <- hypers
-
-reshape2::melt(as.matrix(E)) %>%
-  ggplot(aes(x = Var1, y = factor(Var2), fill = value)) +
-  labs(x = "", y = "", fill = "") +
-  geom_tile() +
-  coord_flip() +
-  scale_fill_gradientn(
-    colours = c("#E69F00", "white", "#009E73"),
-    rescaler = ~ scales::rescale_mid(.x, mid = 0),
-    limits = c(-1.1, 1.1)
-  ) +
-  labs(x = "Hyper", y = "Principal component loading") +
-  theme_minimal() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+plot_pc_loadings(eigenC)
 
@@
 
-C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
-  ggplot(aes(x = as.factor(Var1), y = as.factor(Var2), fill = value)) +
-  labs(x = "", y = "", fill = "C_s[i, j]") +
-  geom_tile() +
-  scale_fill_viridis_c() +
-  labs(x = "i", y = "j") +
-  theme_minimal() +
-  theme(
-    axis.text.x = element_blank(),
-    axis.text.y = element_blank(),
-    legend.position = "none"
-  )
-
-{C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + labs(subtitle = "Reduced rank")}
+C_s_plot <- plot_matrix(C_s)
+{C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + theme(legend.position = "none") + labs(subtitle = "Reduced rank")}
 
@@
 
-Put `logit` and `log` on parity somehow!
+Eventually we want to put parameters on the log and logit scales in parity.
+
+## Using the correlation matrix
+
+```{r}
+C <- sd_out$cov.fixed
+d <- 1 / (sqrt(diag(C)))
+R <- diag(d) %*% C %*% diag(d)
+plot_matrix(R)
+```
+
+```{r}
+eigenR <- eigen(R)
+plot_total_variation(eigenR, label_x = 5)
+```
+
+```{r}
+rownames(eigenR$vectors) <- hypers
+plot_pc_loadings(eigenR)
+```
+
+## Scale standardised system
+
+```{r}
+C <- sd_out$cov.fixed
+var <- diag(C)
+
+df_std <- data.frame(var = var) %>%
+  tibble::rownames_to_column("par") %>%
+  mutate(
+    scale = fct_recode(str_extract(par, "^[^_]+"), "Log" = "log", "Logit" = "logit", "Other" = "OmegaT")
+  ) %>%
+  group_by(scale) %>%
+  mutate(mean_var = mean(var)) %>%
+  ungroup() %>%
+  mutate(var_std = var / mean_var)
+
+df_std %>%  
+  pivot_longer(
+    cols = c("var", "var_std"),
+    names_to = "method",
+    values_to = "value"
+  ) %>%
+  mutate(
+    method = fct_recode(method, "Default" = "var", "Scale standardised" = "var_std"),
+    start = str_extract(par, "^[^_]+"),
+    scale = fct_recode(start, "Log" = "log", "Logit" = "logit", "Other" = "OmegaT") 
+  ) %>%
+  ggplot(aes(x = reorder(par, value), y = value, col = scale)) +
+    geom_point() +
+    facet_grid(~ method) +
+    coord_flip() +
+    scale_color_manual(values = c("#56B4E9","#009E73", "#E69F00")) +
+    labs(x = "Hyperparameter", y = "Marginal variance", col = "Scale") +
+    theme_minimal()
+
+Rs <- diag(1 / sqrt(df_std$mean_var)) %*% C %*% diag(1 / sqrt(df_std$mean_var))
+plot_matrix(Rs)
+
+eigenRs <- eigen(Rs)
+plot_total_variation(eigenRs, label_x = 5)
+
+rownames(eigenRs$vectors) <- hypers
+plot_pc_loadings(eigenRs)
+```
+
+# Testing pruning
+
+Remove any nodes which have small enough weight.
 

### 2023-07-05 src/dev_scale-grid/functions.R

@@ -1,3 +1,51 @@
+#' A local version of naomi::make_tmb_obj, edited to work with DLL = "naomi_simple"
+local_make_tmb_obj <- function(data, par, calc_outputs = 0L, inner_verbose, progress = NULL, map = NULL, DLL = "naomi_simple") {
+  # Begin expose naomi:::make_tmb_obj
+  # https://github.com/mrc-ide/naomi/blob/e9de40f12cf2e652f78966bb351fa5718ecd7867/R/tmb-model.R#L496
+  data$calc_outputs <- as.integer(calc_outputs)
+
+  integrate_out <- c(
+    "beta_rho", "beta_alpha", "beta_alpha_t2", "beta_lambda", "beta_asfr",
+    "beta_anc_rho", "beta_anc_alpha", "beta_anc_rho_t2", "beta_anc_alpha_t2",
+    "u_rho_x", "us_rho_x", "u_rho_xs", "us_rho_xs", "u_rho_a", "u_rho_as",
+    "u_rho_xa", "u_alpha_x", "us_alpha_x", "u_alpha_xs", "us_alpha_xs",
+    "u_alpha_a", "u_alpha_as", "u_alpha_xt", "u_alpha_xa", "u_alpha_xat",
+    "u_alpha_xst", "ui_lambda_x", "logit_nu_raw", "ui_asfr_x", "ui_anc_rho_x",
+    "ui_anc_alpha_x", "ui_anc_rho_xt", "ui_anc_alpha_xt", "log_or_gamma",
+    "log_or_gamma_t1t2"
+  )
+
+  if (DLL == "naomi_simple") {
+    exclude_random_pars <- c(
+      "beta_alpha_t2", "beta_asfr", "beta_anc_rho_t2", "beta_anc_alpha_t2",
+      "u_alpha_xt", "u_alpha_xat", "u_alpha_xst", "logit_nu_raw", "ui_asfr_x",
+      "ui_anc_rho_xt", "ui_anc_alpha_xt", "log_or_gamma_t1t2"
+    )
+
+    integrate_out <- setdiff(integrate_out, exclude_random_pars)
+  }
+
+  if(DLL == "naomi_simple_x_index") {
+    integrate_out <- "x_minus_i"
+  }
+
+  obj <- TMB::MakeADFun(
+    data = data,
+    parameters = par,
+    DLL = DLL,
+    silent = !inner_verbose,
+    random = integrate_out,
+    map = map
+  )
+
[…]

### 2023-07-06 src/docs_paper/paper.Rmd
Improve Fig A by a lot! #55

@@
 )
-```
-
-```{r results='hide'}
-source("figures.R")
 ```
@@
 
-The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ for an ELGM is given by
+The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ is given by
 \begin{equation*}
@@
 \end{equation*}
-Consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
+We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
 \begin{align}
@@
 where $\z \in \mathcal{Q}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
-Gauss-Hermite quadrature [GHQ; @davis1975methods] is one such quadrature rule, where in the univariate case the nodes $\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}$ are selected as zeros of the $k$th Hermite polynomial.
-The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, where $\phi(\cdot)$ is a standard univariate Gaussian density.
-GHQ is attractive because it is exact for functions which are a polynomial of total order no more than $2k - 1$ multiplied by a Gaussian density.
-Posterior distributions can typically be expected to be relatively well described by this class of functions.
+Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule, where the nodes $\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}$ are selected as zeros of the $k$th Hermite polynomial in the univariate case.
+The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, where $\phi(\cdot)$ is a standard Gaussian density.
+GHQ is attractive because it is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$, and we expect posterior distributions to be relatively well described by this class of functions.
 Multivariate GHQ rules, required to integrate over $\btheta$, are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ [@jackel2005note].
@@
 In adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] the nodes are shifted and rotated to suit the particular integrand.
-Repositioning the nodes is especially important in statistical quadrature problems, where the integral depends on data $\y$ such that regions of high density are not known in advance.
-To obtain an AGHQ estimate of Equation \ref{eq:quad}, let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then
+Repositioning the nodes is especially important in statistical quadrature problems, where the integral depends on data $\y$ and regions of high density are not known in advance.
+Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then an AGHQ estimate of Equation \ref{eq:quad} is given by
 \begin{equation}
@@
 \end{equation}
-That is, the unadapted nodes have been shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
+The unadapted nodes are shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
 Two alternatives for the decomposition are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$ and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$ [@jackel2005note].
-Figure \ref{fig:aghq} demonstrates GHQ, as well as adaption for these two choices of matrix decomposition.
 Equation \ref{eq:aghq} may then be used to normalise the Laplace approximation
@@
 
-## Principal components analysis \label{sec:pca}
+## Principal components analysis\label{sec:pca}
 
 ```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension (A).
-Adaption occurs based on the mode (B) and covariance matrix of the target via the Cholesky decomposition (C) or spectral decompostion (D) of the inverse curvature at the mode.
-In PCA-AGHQ (E) only nodes along the first $s = 1$ principal components are kept.
-The Scree plot (F) shows that over 90\\% of variation is explained by keeping just the first principal component.
+Adaption occurs based on the mode and covariance matrix of the target via the Cholesky decomposition of the inverse curvature at the mode (B).
+In PCA-AGHQ (C) only nodes along the first $s$ principal components are kept.
+Here, 95\\% of variation is explained by the first principal component.
 The integrand is $f(\\btheta) = \\text{sn}(0.5 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.8 \\theta_1 - 0.5 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
-figA
+knitr::include_graphics("figA.png")
 ```
@@
 
-```{r ks, fig.height = 4, fig.cap="Example KS test for one parameter."}
-figB
+```{r ks, fig.cap="Example KS test for one parameter."}
+knitr::include_graphics("figB.png")
 ```

### 2023-07-06 src/docs_paper/figures.R

@@ -8,11 +8,13 @@ dyn.load(TMB::dynlib("2d"))
 
 obj <- TMB::MakeADFun(data = list(), parameters = list(theta1 = 0, theta2 = 0), DLL = "2d")
 
-box_size <- 10
+box_lower <- -5
+box_upper <- 10
+box_size <- box_upper - box_lower
 
 grid <- expand.grid(
-  theta1 = seq(-box_size, box_size, length.out = box_size * 50),
-  theta2 = seq(-box_size, box_size, length.out = box_size * 50)
+  theta1 = seq(box_lower, box_upper, length.out = box_size * 50),
+  theta2 = seq(box_lower, box_upper, length.out = box_size * 50)
 )
 
 ground_truth <- cbind(grid, pdf = apply(grid, 1, function(x) exp(-1 * obj$fn(x))))
@@ -35,7 +37,7 @@ cov <- sd_out$cov.fixed
 
 figA0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
   geom_contour(col = "lightgrey") +
-  coord_fixed(xlim = c(-box_size, box_size), ylim = c(-box_size, box_size), ratio = 1) +
+  coord_fixed(xlim = c(box_lower, box_upper), ylim = c(box_lower, box_upper), ratio = 1) +
   labs(x = "", y = "") +
   theme_minimal() +
   guides(size = "none") +
@@ -66,54 +68,41 @@ add_points <- function(figA0, gg) {
 }
 
 figA1 <- add_points(figA0, gg) +
-  labs(size = "")
-
-#' Adapt by the mean
-gg2 <- gg
-mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 1)
-
-figA2 <- add_points(figA0, gg2) +
-  labs(size = "")
-
-#' Adapt by the lower Cholesky
-gg3 <- gg
-mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)
-
-figA3 <- add_points(figA0, gg3) +
-  labs(size = "")
+  labs(subtitle = "A", size = "")
 
 #' Adapt by the spectral
-gg4 <- gg
-mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)
+gg2 <- gg
+mvQuad::rescale(gg2, m = mu, C = cov, dec.type = 1)
 
-figA4 <- add_points(figA0, gg4) +
-  labs(size = "")
[…]

### 2023-07-06 src/docs_paper/paper.Rmd
Write in updated formula for PCA-AGHQ #55

@@
 
-Use of the product rule grid above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points.
+Use of the product rule grid described above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points.
 This quickly becomes intractable as $m$ increases for $k > 1$.
@@
 Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
-Taken together with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ
+Using the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ
 \begin{equation}
-\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
+\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
 \end{equation}
-where $\hat{\mathbf{E}}_{\texttt{LA}, s} = \cdots$ and $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s} = \cdots$ are terms which should be described.
+where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
 The final panel of Figure \ref{fig:aghq} illustrates a case where $m = 2$ and $s = 1$. 

### 2023-07-06 src/docs_paper/figures.R

@@ -102,7 +102,7 @@ figA3 <- add_points(figA0, gg3) +
 
 cowplot::plot_grid(figA1, figA2, figA3, ncol = 3)
 
-ggsave("figA.png", h = 3, w = 6.25, bg = "white")
+ggsave("figA.png", h = 2.5, w = 6.25, bg = "white")
 
 #' Fig B
 

### 2023-07-07 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Add logit_phi and log_sigma paris plot to the MCMC, and put into appendix

--- 2023-04-19 src/naomi-simple_mcmc/mcmc-convergence.Rmd
@@
 rhats <- bayesplot::rhat(mcmc)
-bayesplot::mcmc_rhat(rhats)
+
+bayesplot::mcmc_rhat_data(rhats) %>%
+  ggplot(aes(x = value, y = parameter, col = description)) +
+  geom_segment(aes(yend = parameter, xend = ifelse(min(value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7) +
+  scale_color_manual(values = "#E69F00") +
+  geom_vline(xintercept = 1.05, linetype = "dashed", col = "grey40") +
+  labs(x = "Potential scale reduction factor", y = "NUTS parameter", col = "") +
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    panel.grid.major = element_blank()
+  )
+
+ggsave("rhat.png", h = 3, w = 6.25)
+
 (big_rhats <- rhats[rhats > 1.1])
@@
 ratios <- bayesplot::neff_ratio(mcmc)
-bayesplot::mcmc_neff(ratios)
+
+breaks <- c(0, 0.1, 0.25, 0.5, 0.75, 1)
+
+bayesplot::mcmc_neff_data(ratios) %>%
+  ggplot(mapping = aes(x = value, y = parameter, color = description)) +
+  geom_segment(aes(yend = parameter, xend = -Inf), na.rm = TRUE, alpha = 0.7) +
+  scale_color_manual(values = c("#56B4E9", "#009E73", "#E69F00")) +
+  geom_vline(xintercept = 0.1, linetype = "dashed", col = "grey40") +
+  geom_vline(xintercept = 0.5, linetype = "dashed", col = "grey40") +
+  geom_vline(xintercept = 1, linetype = "dashed", col = "grey40") +
+  labs(x = "ESS ratio", y = "NUTS parameter", col = "") +
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    panel.grid.major = element_blank()
+  )
+
+ggsave("ratio.png", h = 3, w = 6.25)
 ```
@@
     labs(x = "ESS", y = "Count")
+
+ggsave("ess.png", h = 3, w = 6.25)
 ```
@@
 
+## Paired PCA parameters
+
+Variation between units can be explained by high correlation and high standard deviation, or low correlation and low standard deviation.
+Hence there is an unidentifiabiility here that leads to correlated posteriors.
+
+```{r}
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_as", "logit_phi_alpha_as"), diag_fun = "hist", off_diag_fun = "hex")
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_as", "logit_phi_rho_as"), diag_fun = "hist", off_diag_fun = "hex")
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_a", "logit_phi_alpha_a"), diag_fun = "hist", off_diag_fun = "hex")
+(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex"))
+
+ggsave("rho_a.png", plot, h = 4, w = 6.25)
+```
+
+## ART attendance
+
 There is a prior suspicion (from Jeff, Tim, Rachel) that the ART attendance model is unidentifiable.

### 2023-07-07 src/docs_paper/appendix.Rmd
Add rho_a plot to appendix

@@
 
+```{r rho_a, fig.cap="Variation between units can be explained either by high correlation and high variance, or by low correlation and low variance.
+As such, some hyperparameters has correlated posteriors.
+It is this correlation which is made use of by PCA-AGHQ."}
+knitr::include_graphics("depends/rho_a.png")
+```
+
 \clearpage

### 2023-07-07 src/naomi-simple_point-estimates/point-estimates.Rmd
Fix adding jitter to plot which shouldn't have jitter added to it

@@
 
-(mean_sd_plot_latent <- df_plot_latent %>%
+df_plot_latent %>%
   ggplot(aes(x = truth, y = approximate - truth)) +
@@
   labs(x = "NUTS", y = "Approximation - NUTS") +
-  theme_minimal())
+  theme_minimal()
 
-ggsave("mean-sd-latent.png", mean_sd_plot_latent, h = 6, w = 6.25)
+ggsave("mean-sd-latent.png", h = 6, w = 6.25)
 ```
@@
 
-(mean_sd_plot_output <- df_plot_output %>%
+df_plot_output %>%
   ggplot() +
@@
   geom_text(data = df_metrics_output, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
-  labs(x = "NUTS", y = "Approximation - NUTS") +
-  theme_minimal())
+  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
+  theme_minimal() +
+  theme(legend.position = "bottom")
 
-ggsave("mean-sd-output.png", mean_sd_plot_output, h = 6, w = 6.25)
+ggsave("mean-sd-output.png", h = 6, w = 6.25)
 ```
@@
 ```{r}
-jitter_amount <- 0.02
+jitter_amount <- 0
 
@@
 
-ggsave("mean-sd-alt-output.png", mean_sd_plot_output_alt, h = 6, w = 6.25)
+ggsave("mean-sd-alt-output.png", mean_sd_plot_output_alt, h = 6, w = 6.25, bg = "white")
 ```

### 2023-07-11 src/dev_scale-grid/scale-grid.Rmd
Work towards arbitrary orthogonal partial basis AGHQ #58

@@
     wrap: sentence
+bibliography: citations.bib
 ---
@@
 
-# Testing pruning
+## Creating quadrature grids with different coordinate systems
 
-Remove any nodes which have small enough weight.
+Want to use a particular coordinate system, but still the covariance matrix scaling as in adaptive quadrature.
+
+```{r}
+m <- c(1, 1.5)
+C <- matrix(c(2, 1, 1, 1), ncol = 2)
+
+obj <- function(theta) {
+  mvtnorm::dmvnorm(theta, mean = m, sigma = C, log = TRUE)
+}
+
+ff <- list(
+  fn = obj,
+  gr = function(theta) numDeriv::grad(obj, theta),
+  he = function(theta) numDeriv::hessian(obj, theta)
+)
+
+grid <- expand.grid(
+  theta1 = seq(-6, 6, length.out = 600),
+  theta2 = seq(-6, 6, length.out = 600)
+)
+
+ground_truth <- cbind(grid, pdf = exp(obj(grid)))
+
+plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = multi.utils::cbpalette()[1]) +
+  coord_fixed(xlim = c(-6, 6), ylim = c(-6, 6), ratio = 1) +
+  labs(x = "", y = "") +
+  theme_minimal()
+
+grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 5)
+mvQuad::rescale(grid_5, m = m, C = C, dec.type = 1)
+
+pca_grid_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
+
+plot_points <- function(gg) {
+  plot +
+    geom_point(
+      data = mvQuad::getNodes(gg) %>%
+        as.data.frame() %>%
+        mutate(weights = mvQuad::getWeights(gg)),
+      aes(x = V1, y = V2, size = weights),
+      alpha = 0.8,
+      col = multi.utils::cbpalette()[2],
+      inherit.aes = FALSE
+    ) +
+    scale_size_continuous(range = c(1, 2)) +
+    labs(x = "", y = "", size = "Weight") +
+    theme_minimal()
+}
+
+plot_points(grid_5) + plot_points(pca_grid_5) + plot_annotation(tag_levels = "A")
+```
+
+Try to create a grid similar to A above but using the second principal component rather than the first:
+
+```{r}
+eigenC <- eigen(C)
+E <- eigenC$vectors
+lambda <- eigenC$values
+
+gg_s <- mvQuad::createNIGrid(dim = 1, type = "GHe", level = 5)
+nodes_out <- t(E[, 2] %*% diag(sqrt(lambda[2]), ncol = 1) %*% t(mvQuad::getNodes(gg_s)))
+for(j in 1:2) nodes_out[, j] <- nodes_out[, j] + m[j] # Adaption
+weights_out <- mvQuad::getWeights(gg_s) * as.numeric(mvQuad::getWeights(mvQuad::createNIGrid(dim = 1, type = "GHe", level = 1)))
+weights_out <- det(chol(C)) * weights_out # Adaption
+
+custom_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
+custom_grid_5$level <- rep(NA, times = 2)
+custom_grid_5$ndConstruction <- "custom"
+custom_grid_5$nodes <- nodes_out
+custom_grid_5$weights <- weights_out
+
+plot_points(grid_5) / plot_points(pca_grid_5) / plot_points(custom_grid_5) + plot_annotation(tag_levels = "A")
+
+ggsave("pca-demo.png", h = 7, w = 6.25)
+```
+
+We know that $C = E \Lambda E^\top = \sum \lambda_i \mathbf{e}_i \mathbf{e}_i^\top$
+
+```{r}
+lambda[1] * E[, 1] %*% t(E[, 1]) + lambda[2] * E[, 2] %*% t(E[, 2])
+```
+
+We also know that any vector can be written in terms of the eigendecomposition.
+For example, the vector `c(1, 1)` can be written as:
+
+```{r}
+a <- c(1, 1) %*% E
+
+a[1] * E[, 1] + a[2] * E[, 2]
+```
+
+# Testing prerotation
+
+See @jackel2005note.
+
+```{r}
+R45 <- matrix(
+  c(cos(pi / 2), sin(pi / 2), -sin(pi / 2), cos(pi / 2)), ncol = 2, nrow = 2
+)
+```
 
@@
 ```
+
+# Testing pruning
+
+Remove any nodes which have small enough weight [@jackel2005note].
+
+```{r}
+#' TODO!
+```
+
+# Original computing environment {-}
+
+```{r}
+sessionInfo()
+```
+
+# Bibliography {-}

### 2023-07-11 src/dev_scale-grid/scale-grid.Rmd
Custom coordinate system AGHQ done #58

@@
 }
-
-plot_points(grid_5) + plot_points(pca_grid_5) + plot_annotation(tag_levels = "A")
 ```
 
-Try to create a grid similar to A above but using the second principal component rather than the first:
+Create a grid similar to A above but using the second principal component rather than the first:
 
@@
 
-custom_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
-custom_grid_5$level <- rep(NA, times = 2)
-custom_grid_5$ndConstruction <- "custom"
-custom_grid_5$nodes <- nodes_out
-custom_grid_5$weights <- weights_out
-
-plot_points(grid_5) / plot_points(pca_grid_5) / plot_points(custom_grid_5) + plot_annotation(tag_levels = "A")
-
-ggsave("pca-demo.png", h = 7, w = 6.25)
+pc2_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
+pc2_grid_5$level <- rep(NA, times = 2)
+pc2_grid_5$ndConstruction <- "custom"
+pc2_grid_5$nodes <- nodes_out
+pc2_grid_5$weights <- weights_out
 ```
@@
 
-We also know that any vector can be written in terms of the eigendecomposition.
-For example, the vector `c(1, 1)` can be written as:
+We also know that any vector can be written in terms of the eigendecomposition:
 
@@
 a <- c(1, 1) %*% E
+a[1] * E[, 1] + a[2] * E[, 2]
+```
 
-a[1] * E[, 1] + a[2] * E[, 2]
+We also have the definition of an eigenvalue and vector:
+
+```{r}
+sum(t(lambda[1] %*% E[, 1]) - C %*% E[, 1]) < 10e-12
+```
+
+The definition of the unit eigenvectors is that they maximise the amount of variation explained, which is given by this function `variance_explained`.
+By definition `E[, 1]` maximises this!
+
+```{r}
+variance_explained <- function(v) {
+  v %*% C %*% v / v %*% v
+}
+
+variance_explained(c(1, 1))
+variance_explained(E[, 1])
+```
+
+```{r}
+v <- c(1, 1)
+v_norm <- v / sqrt(c(v %*% v))
+
+gg_s <- mvQuad::createNIGrid(dim = 1, type = "GHe", level = 5)
+nodes_out <- t(v_norm %*% sqrt(variance_explained(v_norm)) %*% t(mvQuad::getNodes(gg_s)))
+for(j in 1:2) nodes_out[, j] <- nodes_out[, j] + m[j] # Adaption
+weights_out <- mvQuad::getWeights(gg_s) * as.numeric(mvQuad::getWeights(mvQuad::createNIGrid(dim = 1, type = "GHe", level = 1)))
+weights_out <- det(chol(C)) * weights_out # Adaption
+
+v_norm_grid_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
+v_norm_grid_5$level <- rep(NA, times = 2)
+v_norm_grid_5$ndConstruction <- "custom"
+v_norm_grid_5$nodes <- nodes_out
+v_norm_grid_5$weights <- weights_out
+
+{plot_points(grid_5) + plot_points(pca_grid_5)} / {plot_points(pc2_grid_5) + plot_points(v_norm_grid_5)} + plot_annotation(tag_levels = "A")
+
+ggsave("pca-demo.png", h = 6, w = 6.25)
+```
+
+## Application of custom coordinate system to scale standardised system
+
+Now we apply what we've learnt about making grids with custom coordinate systems to make a grid with coordinate system defined by the eigendecomposition of the scale standardised variance matrix!
+
+```{r}
+#' TODO
 ```

### 2023-07-13 src/dev_scale-grid/scale-grid.Rmd
Rename "pca_rescale" as "create_pca_grid" #58

@@
 
-gg3 <- pca_rescale(m = m, C = C, s = 3, k = 3)
+gg3 <- create_pca_grid(m = m, C = C, s = 3, k = 3)
 nrow(mvQuad::getNodes(gg3))
 
-gg5 <- pca_rescale(m = m, C = C, s = 5, k = 3)
+gg5 <- create_pca_grid(m = m, C = C, s = 5, k = 3)
 nrow(mvQuad::getNodes(gg5))
 
-gg7 <- pca_rescale(m = m, C = C, s = 7, k = 3)
+gg7 <- create_pca_grid(m = m, C = C, s = 7, k = 3)
 nrow(mvQuad::getNodes(gg7))
 
-gg9 <- pca_rescale(m = m, C = C, s = 9, k = 3)
+gg9 <- create_pca_grid(m = m, C = C, s = 9, k = 3)
 nrow(mvQuad::getNodes(gg9))
@@
 
-pca_grid_5 <- pca_rescale(m = m, C = C, s = 1, k = 5)
+pca_grid_5 <- create_pca_grid(m = m, C = C, s = 1, k = 5)
 

### 2023-07-13 src/check_pca-aghq/functions.R

@@ -17,13 +17,13 @@ compare_nodes_and_weights <- function(quad1, quad2) {
     theme_minimal()
 }
 
-#' Create a function to do the PCA rescaling, which also adapts according to the mean and reweights the nodes:
+#' Create a PCA-AGHQ grid
 #'
 #' @param m Mean vector
 #' @param C Covariance matrix
 #' @param s Small grid dimension
 #' @param k Number of points per small grid dimension
-pca_rescale <- function(m, C, s, k) {
+create_pca_grid <- function(m, C, s, k) {
   d <- nrow(C)
   stopifnot(d == length(m))
   eigenC <- eigen(C)

### 2023-07-13 src/dev_scale-grid/functions.R

@@ -112,13 +112,13 @@ sd_levels_ghe_grid <- function(dim, level, cut_off, sd) {
   grid
 }
 
-#' Create a function to do the PCA rescaling, which also adapts according to the mean and reweights the nodes
+#' Create a PCA-AGHQ grid
 #'
 #' @param m Mean vector
 #' @param C Covariance matrix
 #' @param s Small grid dimension
 #' @param k Number of points per small grid dimension
-pca_rescale <- function(m, C, s, k) {
+create_pca_grid <- function(m, C, s, k) {
   d <- nrow(C)
   stopifnot(d == length(m))
   eigenC <- eigen(C)

### 2023-07-13 src/dev_scale-grid/scale-grid.Rmd
Rename Rs as Cs (stands for "covariance scaled") #58

@@
 
-Rs <- diag(1 / sqrt(df_std$mean_var)) %*% C %*% diag(1 / sqrt(df_std$mean_var))
-plot_matrix(Rs)
+Cs <- diag(1 / sqrt(df_std$mean_var)) %*% C %*% diag(1 / sqrt(df_std$mean_var))
+plot_matrix(Cs)
 
-eigenRs <- eigen(Rs)
-plot_total_variation(eigenRs, label_x = 5)
+eigenCs <- eigen(Cs)
+plot_total_variation(eigenCs, label_x = 5)
 
-rownames(eigenRs$vectors) <- hypers
-plot_pc_loadings(eigenRs)
+rownames(eigenCs$vectors) <- hypers
+plot_pc_loadings(eigenCs)
 ```

### 2023-07-13 src/dev_scale-grid/scale-grid.Rmd
It's the covariance matrix not Hessian that is reconstructed

@@
 
-```{r cov-reconstruction, class.source = 'fold-hide', fig.cap="Hessian matrix recons."}
+```{r cov-reconstruction, class.source = 'fold-hide', fig.cap="Covariance matrix reconstruction."}
 E_s <- E[, 1:s]

### 2023-07-13 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
It's the covariance matrix not Hessian that is reconstructed

@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="If you reconstruct the Hessian with just those 8 dimensions, it looks pretty similar.", out.width = '90%'}
+```{r, echo=FALSE, message=FALSE, fig.cap="If you reconstruct the covariance matrix with just those 8 dimensions, it looks pretty similar.", out.width = '90%'}
 knitr::include_graphics("depends/reduced-rank-plot.png")

### 2023-07-13 src/docs_paper/appendix.Rmd
It's the covariance matrix not Hessian that is reconstructed

@@
 
-```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank matrix."}
+```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the covariance matrix is visually similar to the full rank original."}
 knitr::include_graphics("depends/reduced-rank.png")

### 2023-07-14 src/naomi-simple_point-estimates/point-estimates.Rmd
Create point estimate plots of output split by parameter

@@
 
+Version split by output type:
+
+```{r}
+df_metrics_output_split <- df_plot_output %>%
+  group_by(method, indicator, type, par) %>%
+  summarise(
+    rmse = sqrt(mean((truth - approximate)^2)),
+    mae = mean(abs(truth - approximate))
+  ) %>%
+  ungroup()
+
+df_metrics_output_split_pct <- df_metrics_output_split %>%
+  split(~ indicator + type + par) %>%
+  lapply(function(x) {
+    x %>% summarise(
+      rmse_diff = 100 * diff(rmse) / max(rmse),
+      mae_diff = 100 * diff(mae) / max(mae)
+    )
+  }) %>%
+  bind_rows(.id = "id") %>%
+  separate(id, c("indicator", "type", "par"), sep = "\\.")
+
+df_metrics_output_split <- df_metrics_output_split %>%
+  left_join(
+    df_metrics_output_split_pct,
+    by = c("indicator", "type", "par")
+  ) %>%
+  mutate(
+    label = ifelse(
+      method == "PCA-AGHQ",
+      paste0("RMSE: ", signif(rmse, 2), " (", signif(rmse_diff, 2), "%)", "\nMAE: ", signif(mae, 2), " (", signif(mae_diff, 2), "%)"),
+      paste0("RMSE: ", signif(rmse, 2), "\nMAE: ", signif(mae, 2))
+    )
+  )
+
+plot_outputs <- function(.indicator, .par, col) {
+  df_plot_output %>%
+  filter(indicator == .indicator, par == .par) %>%
+  ggplot() +
+  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
+  scale_color_manual(values = c(col, "white", "white")) +
+  facet_grid(indicator + par ~ method) +
+  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  geom_text(
+    data = filter(df_metrics_output_split, indicator == .indicator, par == .par),
+    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
+  ) +
+  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
+  theme_minimal() +
+  theme(legend.position = "none")
+}
+
+plot_outputs(.indicator = "Posterior mean estimate", .par = "rho_t1_out", col = "#E69F00")
+plot_outputs(.indicator = "Posterior mean estimate", .par = "lambda_t1_out", col = "#009E73")
+plot_outputs(.indicator = "Posterior mean estimate", .par = "alpha_t1_out", col = "#56B4E9")
+plot_outputs(.indicator = "Posterior SD estimate", .par = "rho_t1_out", col = "#E69F00")
+plot_outputs(.indicator = "Posterior SD estimate", .par = "lambda_t1_out", col = "#009E73")
+plot_outputs(.indicator = "Posterior SD estimate", .par = "alpha_t1_out", col = "#56B4E9")
+```
+
 # Original computing environment {-}

### 2023-07-14 make/_make.R

@@ -96,7 +96,7 @@ run_commit_push("naomi-simple_mcmc")
 run_commit_push("naomi-simple_ks")
 run_commit_push("naomi-simple_exceedance")
 run_commit_push("naomi-simple_point-estimates")
-run_commit_push("naomi-simple_psis")
+  run_commit_push("naomi-simple_psis")
 run_commit_push("naomi-simple_mmd")
 run_commit_push("naomi-simple_model-checks")
 run_commit_push("naomi-simple_contraction")

### 2023-07-17 src/docs_paper/appendix.Rmd
Keep writing paper! #55

@@
 
-We use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
+We use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age groups.
 Let $\bu$ be a generic random effect, with length $\dim(\bu) > 1$.
@@
 
-```{r marginal-sd, fig.cap="Hyperparameters on the logit-scale had systematically higher marginal standard deviations than those on the log-scale.
-This is because variation on the real scale is compressed by the inverse logit and expanded by the inverse log (exponential)."}
-knitr::include_graphics("depends/marginal-sd.png")
+```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
+knitr::include_graphics("depends/total-variation.png")
+```
+
+```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the covariance matrix is visually similar to the full rank original."}
+knitr::include_graphics("depends/reduced-rank.png")
 ```
@@
 
-```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
-knitr::include_graphics("depends/total-variation.png")
+```{r marginal-sd, fig.cap="Hyperparameters on the logit-scale had systematically higher marginal standard deviations than those on the log-scale.
+This is because variation on the real scale is compressed by the inverse logit and expanded by the inverse log (exponential)."}
+knitr::include_graphics("depends/marginal-sd.png")
 ```
@@
 knitr::include_graphics("depends/pc-loadings.png")
-```
-
-```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the covariance matrix is visually similar to the full rank original."}
-knitr::include_graphics("depends/reduced-rank.png")
 ```

### 2023-07-17 src/docs_paper/paper.Rmd
Keep writing paper! #55

@@
 Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
-We propose a new inference method extending adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, enabling fast and accurate inference for Naomi and other extended latent Gaussian models.
-Using data from Malawi, our method improves the accuracy of inferences across a range of metrics, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
+We propose a new inference method extending adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters.
+Using data from Malawi, our method improves the accuracy of inferences for model parameters, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
+However, for model ouptuts, we found the simpler empirical Bayes approach to achieve similar performance.
 Our implementation is based on the existing \texttt{TMB} \textsf{C++} template for the model's log-posterior, and is compatible with any model with such a template.
@@
 Producing granular estimates is challenging, in large part due to limitations of the data from available sources.
-Nationally-representative household surveys provide the most statistically reliable data, but are only conducted every five years or so in most countries, with limited sample size at the district level, due to their high costs to run.
+Nationally-representative household surveys provide the most statistically reliable data, but are costly to run and so only conducted every five years or so in most countries, with limited sample size at the district level.
 Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in closer to real-time, but are not representative of the entire population.
 To address these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
-Modelling multiple data sources jointly mitigates the limitations of any single source, increases statistical power, and can prompt investigation into any conflicts of information.
+Modelling multiple data sources jointly in this way mitigates the limitations of any single source, increases statistical power, and can prompt investigation into possible conflicts of information.
 
-Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates during workshops as a part of a yearly process supported by UNAIDS.
+Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing over 35 countries to input their data and interactively generate estimates during workshops as a part of a yearly process supported by UNAIDS.
 Creation of estimates by country teams, rather than external agencies or researchers, is an important and distinctive feature of the HIV response.
-Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust in the resulting estimates, creating a virtuous cycle of data quality, use and ownership.
+Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust in the resulting estimates, creating a virtuous cycle of data quality, use and ownership [@noor2022country].
 
 Naomi is a complex model, and as such presents a challenging Bayesian inference problem.
-As well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number that can typically be handled by integrated nested Laplace approximations [INLA; @rue2009approximate].
+As well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number that can typically be handled by approaches like integrated nested Laplace approximations [INLA; @rue2009approximate].
 Moreover, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
 
-To allow for interactive review and iteration of model results, the inference procedure should be fast and low memory usage enough for workshop participants to fit the model.
+To allow for interactive review and iteration of model results by workshop participants, the inference procedure should be fast and have low memory usage.
 Due to the scale of the model and features of its posterior geometry [@neal2003slice], Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
-Furthermore, it is preferable that use of the inference method across countries is automatic, and doesn't require substantial statistical expertise, as would be the case for monitoring MCMC convergence and suitability.
+Furthermore, use of the inference method across countries should be effortless, without requiring substantial statistical expertise, as would be the case for monitoring MCMC convergence and suitability.
 
@@
 
-Although the approach of `TMB` is fast, within the empirical Bayes framework hyperparameter uncertainty is not properly accounted for in the latent field posterior.
+Although the approach of `TMB` is fast, within the empirical Bayes framework hyperparameter uncertainty is not accounted for in the latent field posterior.
 This consideration motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
-To obtain fast, accurate Bayesian inferences for the Naomi model we developed an inference method which extends adaptive Gauss-Hermite quadrature (AGHQ) to handle many hyperparameters.
+We developed an inference method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over many hyperparameters.
 AGHQ is a quadrature method based on the theory of polynomial interpolation, which is well suited to statistical estimation problems.
 @bilodeau2022stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
-It is not computationally feasible to use AGHQ in high dimensions directly, as exponentially many nodes would be required.
-Instead, we use principal components analysis (PCA) of the Hessian at the mode to find a smaller number of dimensions which explain most of the variance.
-For the Naomi model in Malawi, this results in a grid which has millions of times fewer nodes the corresponding dense grid.
+However, it is not computationally feasible to use AGHQ in high dimensions directly, as exponentially many nodes would be required.
+Instead, we use principal components analysis (PCA) of the inverse curvature at the mode to find a smaller number of dimensions which explain most of the variance.
+For the Naomi model in Malawi, this results in a grid which has millions of times fewer nodes the corresponding dense grid and is tractable to run.
 Our implementation of the method makes use of the existing Naomi `TMB` template, and is immediately compatible with any model with such a template.
@@
 For ease of notation, let $i$ index the finest district-age-sex division included in the model.
-Let $I \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported, and $\mathcal{I}$ be the set of all $I$.
+Let $I \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported, and $\mathcal{I}$ be the set of all $I$ such that $I \in \mathcal{I}$.
 
@@
 
-Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that $\text{logit}(\rho_i) = \eta^\rho_i$ and $\text{logit}(\alpha_i) = \eta^\alpha_i$, for certain choice of structured additive predictors.
+We specify independent logistic regression models for HIV prevalence and ART coverage in the general population such that $\text{logit}(\rho_i) = \eta^\rho_i$ and $\text{logit}(\alpha_i) = \eta^\alpha_i$.
 HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i$, and depends on adult HIV prevalence and adult ART coverage.
+The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix S1.
 Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
@@
 
-These processes are informed by household survey data.
-For $\theta \in \{\rho, \alpha, \kappa\}$ let
+These processes are each informed by household survey data.
+We first calculate the weighted aggregate survey observations
 \begin{equation*}
-\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}
+\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j},
 \end{equation*}
-be weighted, aggregate survey observations, with individual responses $\theta_j \in \{0, 1\}$ and design weights $w_j$.
-The index $j$ is across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
-The observed number of outcomes are then $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
+with individual responses $\theta_j \in \{0, 1\}$ and design weights $w_j$ for each of $\theta \in \{\rho, \alpha, \kappa\}$.
+The design weights are provided by the survey and aim to reduce bias by decreasing possible correlation between response and recording mechanism [@meng2018statistical].
+The index $j$ runs across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
+We take the weighted observed number of outcomes to be $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
 \begin{equation*}
@@
 \end{equation*}
-is the Kish effective sample size [@kish1965survey].
-We use a binomial working likelihood
+is the Kish effective sample size (ESS) [@kish1965survey].
+As the Kish ESS is maximised by constant design weights, in exchange for reducing bias we reduce our ESS and thereby increase variance.
+We use a binomial working likelihood defined to operate on the reals
 \begin{equation*}
@@
 \end{equation*}
+Though our approach accounts for survey weights, it does not take into account sample correlation structure, for example due to cluster sampling [@wakefield2020small].
 
@@
 These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
-Of the number of ANC clients with ascertained status $m^{\rho^\text{ANC}}_I$, we model the number of those with positive status $y^{\rho^\text{ANC}}_I$ and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ using nested binomial likelihoods
+We take the number of ANC clients with ascertained status to be fixed as $m^{\rho^\text{ANC}}_I$. 
+We then model the number of those with positive status $y^{\rho^\text{ANC}}_I$, and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ using nested binomial likelihoods
 \begin{align*}
 y^{\rho^\text{ANC}}_I &\sim \text{Bin}(m^{\rho^\text{ANC}}_I, \rho^\text{ANC}_{I}), \\
-y^{\alpha^\text{ANC}}_I &\sim \text{Bin}(y^{\rho^\text{ANC}}_I, \alpha^\text{ANC}_{I}),
+y^{\alpha^\text{ANC}}_I &\sim \text{Bin}(y^{\rho^\text{ANC}}_I, \alpha^\text{ANC}_{I}).
 \end{align*}
-As above, we use weighted aggregates
+As in the household survey component, we use weighted aggregates
 \begin{equation*}
@@
 People living with HIV sometimes choose to access ART services outside of the district that they reside in.
-To account for this, we use multinomial logistic regressions to model the probabilities of accessing services outside the home district.
-Briefly, let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring, denoted by $x \sim x'$.
-The log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ are modelled using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
+To account for this, we model the probabilities of accessing services outside the home district using multinomial logistic regressions.
+Briefly, let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring such that $x \sim x'$.
+We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$.
+As such, we assume travel to each neighbouring district, for all age-sex strata, is equally likely.
 We then model aggregate ART attendance data $y^{N^\text{ART}}_I$ using a Gaussian approximation to a sum of binomials.
@@
 
-## Collected together
+## Summary
 
-In all, we model the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
-Let $\x$ be the latent field, comprised of all intercept effects, age random effects, and spatial random effects in the model.
-Let $\btheta$ be hyperparameters, comprised of all standard deviations, and BYM2 proportion parameters.
+In all, Naomi is a joint model on the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
+The structured additive predictors contain intercept effects, age random effects, and spatial random effects which we collectively describe as the latent field $\x$.
+The latent field is controlled by hyperparamters $\btheta$ which include standard deviations, first-order autoregressive model correlation parameters, and reparameterised Besag-York-Mollie model [@simpson2017penalising] proportion parameters.
 
@@
 
-We now describe the popular latent Gaussian class of models, and an extension which encapsulates the complexities of Naomi.
+We now describe the latent Gaussian class of models, and an extension which encapsulates the complexities of Naomi.
 
@@
 
-## Naomi framed as an ELGM
+## Naomi as an ELGM
 
-Naomi has a lot in common with many LGMS: it is a spatio-temporal model with a large latent field, governed by a smaller number of hyperparameters.
-However, Naomi is not an LGM, and instead falls into the ELGM class, for the following reasons:
+Naomi is a spatio-temporal model with a large Gaussian latent field, governed by a smaller number of hyperparameters.
+However, it is an ELGM rather than an LGM, for the reasons below.
+Note that when dependence on a specific number of structured additive predictors is given, it is for that factor in isolation, and as such should be considered illustrative.
 
 1.
-In the household survey component, HIV incidence depends on district-level adult HIV prevalence and ART coverage, such that $\lambda \propto \rho (1 - \omega \cdot \alpha)$, where $\omega = 0.7$ is a fixed constant.
-This reflects basic HIV epidemiology: HIV incidence is proportional to unsuppresed viral load.
-As a result, $\log(\lambda_i)$ depends on 28 structured additive predictors (2 sexes $\times$ 7 age groups $\times$ 2 indicators, HIV prevalence and ART coverage).
+In the household survey component, HIV incidence depends on district-level adult HIV prevalence and ART coverage.
+This reflects basic HIV epidemiology: HIV incidence is proportional to unsuppresed viral load such that such that $\lambda \propto \rho (1 - \omega \cdot \alpha)$, with $\omega = 0.7$ a fixed constant.
+As a result, each $\log(\lambda_i)$ depends on 28 structured additive predictors (where 28 arises from the product of 2 sexes [male and female], 7 age groups, [$\{\text{15-19}, \ldots, \text{45-49}\}$], and 2 indicators [HIV prevalence and ART coverage]).
 2. In the household survey component, HIV incidence and HIV prevalence are linked to the proportion recently infected via Equation \ref{eq:kappa}.
@@
 In the ANC testing component, HIV prevalence and ART coverage depend upon the respective indicators in the household survey component.
-Although $\text{logit}(\rho_i)$ and $\text{logit}(\alpha_i)$ are Gaussian, nonetheless this introduces dependence of each mean response on two structured additive predictors.
+Though $\text{logit}(\rho_i)$ and $\text{logit}(\alpha_i)$ are Gaussian, this nonetheless introduces dependence of each mean response on two structured additive predictors.
 4.
-Throughout the model components, processes are modelled at the finest distict-age-sex division, but likelihoods are defined for observations aggregated over sets of indices.
-As such, single observations are related to $|\mathcal{I}|$ structured additive predictors.
+Throughout the model components, processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations aggregated over sets of indices $i \in I$.
+As such, all observations are related to $|I|$ structured additive predictors.
 5. Individuals taking ART, or who have been recently infected, must be HIV positive. 
-6. The ART attendance component uses a multinomial model with softmax link function which takes as input $|\{x': x' \sim x\}| + 1$ structured additive predictors.
-7. Multiple link functions are used throughout the model, such that there is no one inverse link function $g$. Instead,
-
-If Naomi was written out in full mathematical detail (in the section above), it would be easier to make these reasons more concrete.
-We should also specify how much of an impact each of these reasons is likely to have on the difficulty of inference.
-Perhaps a table format could be good here.
+6.
+The ART attendance component uses a multinomial model with softmax link function which takes as input $|\{x': x' \sim x\}| + 1$ structured additive predictors, one for each neighbouring district plus one for remaining in the home district.
+7. Multiple link functions are used throughout the model, such that there is no one inverse link function $g$.
 
@@
 We begin by describing (Section \ref{sec:inferenceelgm}) the inference method for ELGMs based on nested applications of the Laplace approximation and AGHQ used by @stringer2022fast.
-We then propose (Section \ref{sec:pca}) an extension of the method which uses PCA to facilitates inference for Naomi, which otherwise would be intractable.
+We then propose (Section \ref{sec:pca}) an extension of the method which uses PCA to facilitate inference for Naomi, which otherwise would be intractable.
 
-## Inference for ELGMs \label{sec:inferenceelgm}
+## Inference for ELGMs\label{sec:inferenceelgm}
 
@@
 
-### Laplace approximation
+### Laplace approximation\label{sec:la}
 
@@
 \end{equation}
-Inference in TMB proceeds by optimising Equation \ref{eq:la} using a gradient-based routine to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
+Inference proceeds by optimising Equation \ref{eq:la} using a gradient-based routine to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
 Each evaluation in the optimisation requires an inner optimisation to obtain $\hat \x(\btheta)$ via Equation \ref{eq:mode}.
-Latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
-Hyperparameter inferences are obtained according to some method which should be specified here as well.
+Supposing the hyperparameters are to be considered fixed, then latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
+This is the approach used currently for Naomi, which we refer to as TMB (distinguished from the software \texttt{TMB}).
+Hyperparameter inferences can be obtained according to some method which should be specified here as well.
 
-### Gauss-Hermite quadrature
+### Adaptive Gauss-Hermite quadrature
 
-Quadrature rules can be used to approximate the integral of $\tilde p_\texttt{LA}(\btheta, \y)$ via the weighted sum
+Quadrature rules can be used to approximate integrals using a set of nodes $\z \in \mathcal{Q}$ and a weighting function $\omega: \mathcal{Q} \to \mathbb{R}$.
+Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule, where the nodes are selected as zeros of the $k$th Hermite polynomial
 \begin{equation}
-p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \text{d}\btheta \approx \sum_{\z \in \mathcal{Q}} p_\texttt{LA}(\z, \y) \omega(\z), \label{eq:quad}
+\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}.
 \end{equation}
-where $\z \in \mathcal{Q}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
-Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule, where the nodes $\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}$ are selected as zeros of the $k$th Hermite polynomial in the univariate case.
 The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, where $\phi(\cdot)$ is a standard Gaussian density.
 GHQ is attractive because it is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$, and we expect posterior distributions to be relatively well described by this class of functions.
-Multivariate GHQ rules, required to integrate over $\btheta$, are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ [@jackel2005note].
-
-<!-- Tie together connection to Laplace for AGHQ k = 1 -->
-
-### Adaptive quadrature
-
+Multivariate GHQ rules, required to integrate over $\btheta$, are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
 In adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] the nodes are shifted and rotated to suit the particular integrand.
 Repositioning the nodes is especially important in statistical quadrature problems, where the integral depends on data $\y$ and regions of high density are not known in advance.
-Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then an AGHQ estimate of Equation \ref{eq:quad} is given by
+
+Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then an AGHQ estimate of the normalising constant $p(\y)$ based on the Laplace approximation is
 \begin{equation}
-\tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z).
+p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z).
 \label{eq:aghq}
@@
 The unadapted nodes are shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
-Two alternatives for the decomposition are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$ and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$ [@jackel2005note].
+Two alternatives [@jackel2005note] are
+
+1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
+2.
+the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
+
 Equation \ref{eq:aghq} may then be used to normalise the Laplace approximation
-\begin{equation*}
+\begin{equation}
 \tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
-\end{equation*}
-
-To obtain inferences for the latent field (Equation \ref{eq:inla1}) the adapted nodes and weights are reused [@rue2009approximate; @stringer2022fast]
+\end{equation}
+To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
 \begin{equation}
@@
 \end{equation}
-Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
+Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
 
@@
 An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
-We may then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
+We may then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ to be a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
 Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
-Using the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ
+For example, for $m = 2$ and $s = 1$ then $\bk = (k, 1)$.
+In combination with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ.
+We term this approach PCA-AGHQ, with estimate of the normalising constant given by
 \begin{equation}
@@
 where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
-The final panel of Figure \ref{fig:aghq} illustrates a case where $m = 2$ and $s = 1$. 
-We refer to use of this quadrature rule as PCA-AGHQ.
+The final panel of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
 
@@
 
-1. TMB (`r signif(time_taken$TMB, 2)` seconds)
-2. PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), implemented via extension of the `aghq` package [@stringer2021implementing],
+1. TMB (`r signif(time_taken$TMB, 2)` seconds), using a Gaussian approximation at $\hat{\btheta}_\texttt{LA}$.
+2.
+PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), using a mixture of Gaussians approximation at a PCA-AGHQ quadrature grid, as described in Section \ref{sec:pca}, and implemented via extension of the `aghq` package [@stringer2021implementing].
 3.
@@
 
-Our goal in this analysis was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
+Our goal was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
+Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
 The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
-Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
 The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
@@
 <!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
-Example model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
+To provide intuition, model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
 
-```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted with TMB."}
+```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted using TMB."}
 knitr::include_graphics("depends/naomi-results.png")
@@
 \toprule
-Name & Software & Details \\
+Method & Software & Details \\
 \midrule
@@
 
-To obtain satisfactory NUTS results we increased the Markov chain lengths until all diagnostics were acceptable.
-This required chains of length 20,000, thinned by a factor of 20 for ease-of-storage.
-All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were then $\hat R < 1.05$.
+We increased the chain lengths until all NUTS diagnostics were acceptable.
+This required chains of length 20,000, which we thinned by a factor of 20 for ease-of-storage.
+All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$.
 For full details see Appendix S2.
@@
 
-We used a Scree plot (Figure S$X$) based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})$ to select the number of principal components $s = 8 < m = 24$ to keep, sufficient to explain close to 90% of total variation.
-With this choice of $s$, the reduced rank approximation to the Hessian is visually very similar (Figure S$X$).
+We used a Scree plot (Figure S2) based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})^{-1}$ to select the number of principal components to keep.
+We found that $s = 8$ principal components were sufficient to explain close to 90% of total variation.
+With this choice of $s$, the reduced rank approximation to the inverse curvature was visually similar (Figure S3).
 
@@
 
-Overlaying the generated $3^8 = 6561$ PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
-Though 12 is an improvment on the 8 that would be naively achieved using a dense grid, there remained many hyperparameters poorly covered.
-Coverage was associated with marginal standard deviation (Figure S$X$), which varied particularly according to the hyperparameter scale.
+Overlaying the resulting $3^8 = 6561$ PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
+Though 12 is an improvement on the 8 that would be naively achieved using a dense grid, there remained many hyperparameters poorly covered.
+Coverage was associated with marginal standard deviation (Figure S4), which varied particularly according to the hyperparameter scale.
 All constrained hyperparameters $\theta$ were transformed to the real line, using either a log ($\theta > 0$) or logit ($\theta \in [0, 1]$) transformation.
-As a result, marginal standard deviations for log transformed hyperparameters were systematically smaller than those which were logit transformed.
+As a result, marginal standard deviations for log transformed hyperparameters were systematically smaller than those which were logit transformed (Figure S5).
+
+### Interpreation of eigenvectors
+
+The ordered eigenvectors can be interpreted to inform us about the directions of greatest variation in the inverse curvature (Figure S6)
+The first and second eigenvectors are the weakly identified AR1 standard deviation and correlation parameters.
+Variability can be explained both by low correlation or by high standard deviation.
 
@@
 
-We assessed appropriateness of the grid by comparing the estimate of $\log p_\texttt{PCA}(\y)$ for a range of settings.
+We assessed appropriateness of the quadrature grid by comparing the estimate of $\log p_\texttt{PCA}(\y)$ for a range of settings.
 Convergence in $\log p_\texttt{PCA}(\y)$ as $s$ and $k$ are increased may suggest a suitable grid has been reached.
@@
 
-## Model assessment \label{sec:model-assessment}
+## Model assessment\label{sec:model-assessment}
 
@@
 We developed an approximate Bayesian inference algorithm, combining AGHQ with PCA, motivated by a challenging problem in small-area estimation of HIV.
-For the simplified Naomi model in Malawi (Section \ref{sec:results}) we demonstrated the method to be more accurate, across a broad range of metrics, than TMB, and substantially faster than NUTS.
+For the simplified Naomi model in Malawi (Section \ref{sec:results}) we demonstrated the method to be more accurate for model parameters, across a broad range of metrics, than TMB, and substantially faster than NUTS.
+However, this improvement in accuracy for model parameters did not translate into model outputs.
+Indeed, we found posterior exceedance probabilities (Section \ref{sec:exceedance}) from both TMB and PCA-AGHQ to have systematically inaccurates, with the potential to meaningfully mislead policy.
 
@@
 @bilodeau2022stochastic highlight developing computationally feasible quadrature methods for high dimensions as a challenging open problem.
-
-We found that posterior exceedance probabilities (Section \ref{sec:exceedance}) from the approximate methods to be systematically inaccurate, with the potential to meaningfully mislead policy.
 

### 2023-07-18 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Add check of BYM2 correlation between standard deviation and proprotion, and save AR1 and BYM2 results for inclusion in manuscript

@@
 ```{r}
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_a", "logit_phi_alpha_a"), diag_fun = "hist", off_diag_fun = "hex")
 bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_as", "logit_phi_alpha_as"), diag_fun = "hist", off_diag_fun = "hex")
 bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_as", "logit_phi_rho_as"), diag_fun = "hist", off_diag_fun = "hex")
-bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_a", "logit_phi_alpha_a"), diag_fun = "hist", off_diag_fun = "hex")
-(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex"))
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex")
 
-ggsave("rho_a.png", plot, h = 4, w = 6.25)
+ggsave("rho_a.png", h = 4, w = 6.25)
+
+get_correlation <- function(par) {
+  cor(as.data.frame(rstan::extract(mcmc, c(paste0("log_sigma_", par), paste0("logit_phi_", par)))))[1, 2]
+}
+
+ar1_cor_df <- data.frame(
+  par = c("alpha_a", "alpha_as", "rho_as", "rho_a"),
+  type = "ar1"
+) %>%
+  mutate(cor = purrr::map_dbl(get_correlation, .x = par))
+
+ar1_cor_df
+```
+
+## BYM2 orthogonality
+
+Does the supposed orthogonality of the BYM2 model play out?
+Looks like the answer is mostly yes.
+
+```{r}
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_x", "logit_phi_rho_x"), diag_fun = "hist", off_diag_fun = "hex")
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_xs", "logit_phi_rho_xs"), diag_fun = "hist", off_diag_fun = "hex")
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_xs", "logit_phi_alpha_xs"), diag_fun = "hist", off_diag_fun = "hex")
+bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x"), diag_fun = "hist", off_diag_fun = "hex")
+
+ggsave("alpha_x.png", h = 4, w = 6.25)
+
+bym2_cor_df <- data.frame(
+  par = c("rho_x", "rho_xs", "alpha_x", "alpha_xs"),
+  type = "bym2"
+) %>%
+  mutate(cor = purrr::map_dbl(get_correlation, .x = par))
+
+bym2_cor_df
+
+write_csv(bind_rows(ar1_cor_df, bym2_cor_df), file = "ar1-bym2-cor.csv")
 ```

### 2023-07-18 make/_make.R

@@ -96,7 +96,7 @@ run_commit_push("naomi-simple_mcmc")
 run_commit_push("naomi-simple_ks")
 run_commit_push("naomi-simple_exceedance")
 run_commit_push("naomi-simple_point-estimates")
-  run_commit_push("naomi-simple_psis")
+run_commit_push("naomi-simple_psis")
 run_commit_push("naomi-simple_mmd")
 run_commit_push("naomi-simple_model-checks")
 run_commit_push("naomi-simple_contraction")

### 2023-07-18 src/docs_paper/appendix.Rmd
Adding BYM2 and AR1 discussion to manuscript #55

@@
 ```{r rho_a, fig.cap="Variation between units can be explained either by high correlation and high variance, or by low correlation and low variance.
-As such, some hyperparameters has correlated posteriors.
-It is this correlation which is made use of by PCA-AGHQ."}
+As such, the AR1 hyperparameters had correlated posteriors.
+It is this correlation which we made use of with PCA-AGHQ."}
 knitr::include_graphics("depends/rho_a.png")
+```
+
+```{r rho_a, fig.cap="In contrast to the AR1 hyperparameters, the BYM2 hyperparameter posteriors are much less correlated."}
+knitr::include_graphics("depends/alpha_x.png")
 ```

### 2023-07-18 src/docs_paper/paper.Rmd
Adding BYM2 and AR1 discussion to manuscript #55

@@
 The structured additive predictors contain intercept effects, age random effects, and spatial random effects which we collectively describe as the latent field $\x$.
-The latent field is controlled by hyperparamters $\btheta$ which include standard deviations, first-order autoregressive model correlation parameters, and reparameterised Besag-York-Mollie model [@simpson2017penalising] proportion parameters.
+The latent field is controlled by hyperparamters $\btheta$ which include standard deviations, first-order autoregressive model correlation parameters, and reparameterised Besag-York-Mollie model [BYM2; @simpson2017penalising] proportion parameters.
 
@@
 ```{r}
+#' It's a shame here that I have to put "seconds", "hours" and "days" manually and it's not stored in the time taken
 time_taken <- read_csv("depends/time-taken.csv")
@@
 This required chains of length 20,000, which we thinned by a factor of 20 for ease-of-storage.
-All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$.
+All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$ (Figure S7).
 For full details see Appendix S2.
@@
 knitr::include_graphics("depends/nodes-samples-comparison.png")
-
-# This plot could be improved with a scale for "coverage"
 ```
@@
 
-The ordered eigenvectors can be interpreted to inform us about the directions of greatest variation in the inverse curvature (Figure S6)
-The first and second eigenvectors are the weakly identified AR1 standard deviation and correlation parameters.
-Variability can be explained both by low correlation or by high standard deviation.
+```{r}
+cor_df <- read_csv("depends/ar1-bym2-cor.csv")
+
+summary_cor_df <- cor_df %>%
+  group_by(type) %>%
+  summarise(mean_cor = mean(abs(cor)))
+
+ar1_mean_cor <- signif(filter(summary_cor_df, type == "ar1")$mean_cor, 2)
+bym2_mean_cor <- signif(filter(summary_cor_df, type == "bym2")$mean_cor, 2)
+```
+
+The ordered eigenvectors correspond to the directions of greatest variation in the inverse curvature.
+The first and second eigenvectors each contain coupled AR1 standard deviation and correlation parameters (Figure S6).
+As variability can be explained either by high standard deviation and high correlation or low standard deviation and low correlation these parameters are weakly identified.
+This is reflected in high correlation in the the NUTS posterior pairs plot (Figure S10, average absolute correlation across all four pairs of `r ar1_mean_cor`).
+The BYM2 standard deviation and proportion parameters on the other hand are designed to be orthogonal, and as such did not display posterior correlation (Figure S11, average absolute correlation across all four pairs of `r bym2_mean_cor`) or appear prominently in the eigenvectors.
 
@@
 
-Let $\phi$ be a generic model parameter.
-To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\phi)$ to the posterior variance $\sigma_\text{posterior}^2(\phi)$ via the posterior contraction $c(\phi) = 1 - (\sigma_\text{posterior}^2(\phi) / \sigma_\text{prior}^2(\phi))$ [@schad2021toward].
+To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\psi)$ to the posterior variance $\sigma_\text{posterior}^2(\psi)$ via the posterior contraction $c(\psi) = 1 - (\sigma_\text{posterior}^2(\psi) / \sigma_\text{prior}^2(\psi))$ [@schad2021toward], where $\psi$ is a model parameter.
 We found that (Figure \ref{fig:contraction})) something something.
@@
 We assessed the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.
-Let $\{\phi_i\}_{i = 1}^n$ be posterior marginal samples.
+Let $\{\psi_i\}_{i = 1}^n$ be posterior marginal samples.
 
@@
 
-The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
+The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\omega) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\psi_i \leq \omega}$.
 See an illustration of the KS test in Figure \ref{fig:ks} and a summary of the results in Table.
@@
 
-Let $\{\bphi_i\}_{i = 1}^n$ be joint posterior samples.
+Let $\{\bpsi_i\}_{i = 1}^n$ be joint posterior samples.
 Pareto-smoothed importance sampling [PSIS; @vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios used in importance sampling.
@@
 
-Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
+Let $\Psi = \{\bpsi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
 The maximum mean discrepancy [MMD; @gretton2006kernel] can be empirically estimated by
 \begin{equation*}
-\text{MMD}(\Phi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bphi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
+\text{MMD}(\Psi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
 \end{equation*}
-We set $k(\bphi_i, \bphi_j) = \exp(-\sigma \lVert \bphi_i - \bphi_j \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
+We set $k(\bpsi_i, \bpsi_j) = \exp(-\sigma \lVert \bpsi_i - \bpsi_j \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
 As compared with NUTS, the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080).

### 2023-07-18 src/docs_paper/paper.Rmd
Eventually make map figure with PCA-AGHQ rather than TMB #55

@@
 Two alternatives [@jackel2005note] are
-
 1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
@@
 the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
-
 Equation \ref{eq:aghq} may then be used to normalise the Laplace approximation
@@
 For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
-<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
 To provide intuition, model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
@@
 knitr::include_graphics("depends/naomi-results.png")
+
+# Eventually this figure should be generated using PCA-AGHQ
+# knitr::include_graphics("depends/figB.png")
 ```
@@
   ungroup() %>%
-  group_by(indicator) %>%
+  group_by(indicator, type) %>%
   summarise(
@@
 
-rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate") %>% pull(rmse) %>% signif(2)
-rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate") %>% pull(rmse) %>% signif(2)
-rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate") %>% pull(rmse_diff) %>% signif(2)
+rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate", type == "latent") %>% pull(rmse_diff) %>% signif(2)
 
-rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate") %>% pull(rmse) %>% signif(2)
-rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate") %>% pull(rmse) %>% signif(2)
-rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate") %>% pull(rmse_diff) %>% signif(0)
+rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate", type == "latent") %>% pull(rmse_diff) %>% signif(0)
 ```
@@
 ```{r ks, fig.cap="Example KS test for one parameter."}
-knitr::include_graphics("figB.png")
+knitr::include_graphics("figC.png")
 ```

### 2023-07-18 src/docs_paper/appendix.Rmd
Add MCMC output numerics to manuscript and appendix, plus traceplots for worst parameters #55

@@
 
+```{r}
+mcmc_out <- readRDS("depends/mcmc-out.rds")
+```
+
 We assessed MCMC convergence and suitability using a range of graphical and numerical tests.
-All potential scale reduction factor $\hat R$ statistics [@vehtari2021rank] were below 1.05 and therefore acceptable (Figure \ref{fig:rhat}).
-However, even thinning by a factor of 20, samples were not obtained very efficiently, resulting in the majority of effective sample size (ESS) ratios being below 0.5, with some as low as 0.1 (Figure \ref{fig:ratio}).
-As a result, the number of obtained ESS varied substantially by parameter (Figure \ref{fig:ess}).
+The largest scale reduction factor $\hat R$ was `r signif(mcmc_out$max_rhat, 3)` (Figure \ref{fig:rhat}).
+As such, all values of $\hat R < 1.05$.
+Even thinning by a factor of 20, samples were not obtained very efficiently, resulting in the majority of effective sample size (ESS) ratios being below 0.5, with some as low as 0.1 (Figure \ref{fig:ratio}).
+As a result, the number of obtained ESS varied substantially by parameter (Figure \ref{fig:ess}): the minimum was `r round(mcmc_out$ess_min, 1)`, 2.5% quantile was `r round(mcmc_out$ess_lower, 1)`, median was `r round(mcmc_out$ess_median, 1)`, 97.5% quantile was `r round(mcmc_out$ess_upper, 1)` and maximum was `r round(mcmc_out$ess_max, 1)`.
+Traceplots for the parameters with the lowest ESS (`log_sigma_alpha_xs`) and highest $\hat R$ (the 10th index of `ui_lambda_x`) are shown in Figure \ref{fig:worst_trace}.
 There were no divergent transitions.
-Could add energy plot from @betancourt2017conceptual here.
+<!-- Could add energy plot from @betancourt2017conceptual here. -->
 
@@
 
+```{r worst_trace, fig.cap="Traceplots of the parameters with the lowest ESS and highest scale reduction factor."}
+knitr::include_graphics("depends/worst-trace.png")
+```
+
 ```{r rho_a, fig.cap="Variation between units can be explained either by high correlation and high variance, or by low correlation and low variance.
@@
 
-```{r rho_a, fig.cap="In contrast to the AR1 hyperparameters, the BYM2 hyperparameter posteriors are much less correlated."}
+```{r alpha_x, fig.cap="In contrast to the AR1 hyperparameters, the BYM2 hyperparameter posteriors are much less correlated."}
 knitr::include_graphics("depends/alpha_x.png")

### 2023-07-18 src/docs_paper/paper.Rmd
Add MCMC output numerics to manuscript and appendix, plus traceplots for worst parameters #55

@@
 PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-use}), $1000$ samples \\
-NUTS & \texttt{tmbstan} & $4$ chains of $20000$ iterations, with the first $10000$ iterations of each chain discarded as warmup, thinned by a factor of $20$.
+NUTS & \texttt{tmbstan} & $4$ chains of $100,000$ iterations, with the first $50,000$ iterations of each chain discarded as warmup, thinned by a factor of $40$, to give a total of $5000$ samples kept.
 Default NUTS tuning parameters \citep{hoffman2014no}.
@@
 
-We increased the chain lengths until all NUTS diagnostics were acceptable.
-This required chains of length 20,000, which we thinned by a factor of 20 for ease-of-storage.
-All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$ (Figure S7).
+```{r}
+mcmc_out <- readRDS("depends/mcmc-out.rds")
+# signif(mcmc_out$max_rhat, 4)
+```
+
+Obtaining acceptable NUTS diagnostics required four chains run in parallel for 100,000 iterations, thinned by a factor of 20 for ease-of-storage.
+The largest potential scale reduction factor [@gelman1992inference; @vehtari2021rank] was $\hat R = 1.021$ (Figure S7), and there were no divergent transitions.
+We considered the NUTS results a gold-standard, though inaccuracies remain possible.
 For full details see Appendix S2.
-Though inaccuracies remain possible, we considered the NUTS results to be a gold-standard.
 
@@
 Overlaying the resulting $3^8 = 6561$ PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
-Though 12 is an improvement on the 8 that would be naively achieved using a dense grid, there remained many hyperparameters poorly covered.
+Though 12 improves on the 8 naively achievable using a dense grid, there remained many hyperparameters poorly covered.
 Coverage was associated with marginal standard deviation (Figure S4), which varied particularly according to the hyperparameter scale.
@@
 
-To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\psi)$ to the posterior variance $\sigma_\text{posterior}^2(\psi)$ via the posterior contraction $c(\psi) = 1 - (\sigma_\text{posterior}^2(\psi) / \sigma_\text{prior}^2(\psi))$ [@schad2021toward], where $\psi$ is a model parameter.
+To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\psi)$ to the posterior variance $\sigma_\text{posterior}^2(\psi)$ via the posterior contraction [@schad2021toward]
+\begin{equation}
+c(\psi) = 1 - (\sigma_\text{posterior}^2(\psi) / \sigma_\text{prior}^2(\psi)),
+\end{equation}
+where $\psi$ is a model parameter.
 We found that (Figure \ref{fig:contraction})) something something.
@@
 
-Let $\Psi = \{\bpsi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
+Let $\Psi^{1} = \{\bpsi^1_i\}_{i = 1}^n$ and $\Psi^2 = \{\bpsi^2_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
 The maximum mean discrepancy [MMD; @gretton2006kernel] can be empirically estimated by
 \begin{equation*}
-\text{MMD}(\Psi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
+\text{MMD}(\Psi^1, \Psi^2) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi^1_i, \bpsi^1_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bpsi_i^1, \bpsi_j^2) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi^2_i, \bpsi^2_j)}.
 \end{equation*}
-We set $k(\bpsi_i, \bpsi_j) = \exp(-\sigma \lVert \bpsi_i - \bpsi_j \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
+We set $k(\bpsi^1, \bpsi^2) = \exp(-\sigma \lVert \bpsi^1 - \bpsi^2 \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
 As compared with NUTS, the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080).
@@
 
-```{r exceedance, fig.cap="Though PCA-AGHQ does perform slightly better, both approximate inference methods are meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities.
-For the second 90 target the inaccuracy varies substantially by sex."}
+```{r exceedance, fig.cap="Though PCA-AGHQ was marginally better, both approximate inference methods were meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities.
+For the second 90 target the inaccuracy varied substantially by sex."}
 knitr::include_graphics("depends/exceedance.png")

### 2023-07-18 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Add MCMC output numerics to manuscript and appendix, plus traceplots for worst parameters #55

@@
 mcmc <- out$mcmc$stanfit
+
+depends <- yaml::read_yaml("orderly.yml")$depends
+
+dependency_details <- function(i) {
+  report_name <- names(depends[[i]])
+  print(paste0("Inference results obtained from ", report_name, " with the query ", depends[[i]][[report_name]]$id))
+  report_id <- orderly::orderly_search(query = depends[[i]][[report_name]]$id, report_name)
+  print(paste0("Obtained report had ID ", report_id, " and was run with the following parameters:"))
+  print(orderly::orderly_info(report_id, report_name)$parameters)
+}
+
+dependency_details(1)
 ```
@@
 
-bayesplot::color_scheme_set("viridisA")
+color_scheme_set("viridis")
 ggplot2::theme_set(theme_minimal())
@@
 
-We are looking for values of $\hat R$ less than 1.1 here.
+We are looking for values of $\hat R$ less than 1.05 here.
 
@@
 
-(big_rhats <- rhats[rhats > 1.1])
+(big_rhats <- rhats[rhats > 1.05])
 length(big_rhats) / length(rhats)
+
+(max_rhat <- max(rhats))
 ```
@@
 ggsave("ratio.png", h = 3, w = 6.25)
+
+(average_ess_ratio <- mean(ratios))
 ```
@@
 ggsave("ess.png", h = 3, w = 6.25)
+
+(ess_min <- min(mcmc_summary[, "n_eff"]))
+(ess_lower <- quantile(mcmc_summary[, "n_eff"], 0.025))
+(ess_median <- quantile(mcmc_summary[, "n_eff"], 0.50))
+(ess_upper <- quantile(mcmc_summary[, "n_eff"], 0.975))
+(ess_max <- max(mcmc_summary[, "n_eff"]))
+```
+
+Save outputs for use in manuscript:
+
+```{r}
+out <- list(
+  "max_rhat" = max_rhat,
+  "average_ess_ratio" = average_ess_ratio, 
+  "ess_min" = ess_min,
+  "ess_lower" = ess_lower,
+  "ess_median" = ess_median,
+  "ess_upper" = ess_upper,
+  "ess_max" = ess_max
+)
+
+saveRDS(out, "out.rds")
 ```
@@
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_sigma")))
+```
+
+The parameters with the worst ESS and the worst $\hat R$:
+
+```{r}
+(plot <- bayesplot::mcmc_trace(mcmc, pars = c(names(which.min(mcmc_summary[, "n_eff"])), names(which.max(rhats)))))
+
+ggsave("worst-trace.png", plot, h = 4, w = 6.25)
 ```
@@
 bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_as", "logit_phi_rho_as"), diag_fun = "hist", off_diag_fun = "hex")
-bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex")
+(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex"))
 
-ggsave("rho_a.png", h = 4, w = 6.25)
+ggsave("rho_a.png", plot, h = 4, w = 6.25)
 
@@
 bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_xs", "logit_phi_alpha_xs"), diag_fun = "hist", off_diag_fun = "hex")
-bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x"), diag_fun = "hist", off_diag_fun = "hex")
+(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x"), diag_fun = "hist", off_diag_fun = "hex"))
 
-ggsave("alpha_x.png", h = 4, w = 6.25)
+ggsave("alpha_x.png", plot, h = 4, w = 6.25)
 
@@
 }
-
-# area_merged %>%
-#   filter(area_level == max(area_level)) %>%
-#   print(n = Inf)
 ```

### 2023-07-20 src/dev_scale-grid/scale-grid.Rmd
Add hyperparameter names to PC loadings plot, add variance explained number (not just near 90), work on formatting of tables #55

@@
 tv <- cumsum(eigenC$values) / sum(eigenC$values)
+
+tv_df <- data.frame(
+  s = 1:length(m),
+  tv = tv
+)
+
+write_csv(tv_df, "total-variation.csv")
 ```
@@
 ```{r pc-loadings, class.source = 'fold-hide', fig.cap="PC loadings."}
-rownames(E) <- hypers
+rownames(eigenC$vectors) <- hypers
 plot_pc_loadings(eigenC)

### 2023-07-20 src/docs_paper/appendix.Rmd
Add hyperparameter names to PC loadings plot, add variance explained number (not just near 90), work on formatting of tables #55

@@
 
-In this section we describe the simplified Naomi model [@eaton2021naomi] in more complete mathematical detail than the main text.
+In this section we describe the simplified Naomi model [@eaton2021naomi] in complete detail.
 
@@
 Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
-Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
-Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
+Let $x \in \{1, \ldots, n\}$ be district, $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
+Each district is located within a correspodning Spectrum [@stover2010spectrum] region $R_x$.
 As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
@@
 We use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
-The notation $\sum_{\{x\}}$ is used as shorthand for $\sum_{x \in \{x\}}$, and analogously for $s$ and $a$.
+The notation $\sum_{\{x\}}$ is used as shorthand for $\sum_{x \in \{x\}}$.
+Analogous notion is used for sums over $\sum_{a \in \{a\}}$ and $\sum_{s \in \{s\}}$.
 
@@
 u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
-u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(\bu)
+u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(\bu),
 \end{align}
@@
 3.
-For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $\bu \sim \text{BYM}2(\sigma, \phi)$, where $\bu$ is comprised of a spatially structured ICAR component $\bv$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $\w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
+For the reparameterised Besag-York-Mollie model [BYM2; @simpson2017penalising], we write $\bu \sim \text{BYM}2(\sigma, \phi)$, where $\bu$ is comprised of a spatially structured ICAR component $\bv$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $\w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
 \begin{equation}
@@
 We now describe the hyperparameter and latent field process specification for the model.
-Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood} to follow.
+Whereas in the main text process and likelihood specifications are given together, here we consider the likelihood equations separately (Section \ref{sec:likelihood}).
 Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each component of the model.
@@
 
-```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
+```{r scree, fig.cap="Most variation in the inverse curvature could be explained using far fewer than the full 24 dimensions."}
 knitr::include_graphics("depends/total-variation.png")
@@
 
-```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the covariance matrix is visually similar to the full rank original."}
+```{r reduced-rank, fig.cap="The reduced rank ($s = 8$) matrix approximation to the covariance matrix is visually similar to the full rank original."}
 knitr::include_graphics("depends/reduced-rank.png")
@@
 
-```{r pc-loadings, fig.cap="Principal component loadings."}
+```{r pc-loadings, fig.cap="Principal component loadings based on the eigendecomposition of the inverse curvature."}
 knitr::include_graphics("depends/pc-loadings.png")
@@
   select(-type) %>%
-  rename(`PCA-AGHQ` = `KS(aghq, tmbstan)`, `TMB` = `KS(TMB, tmbstan)`) %>%
-  gt::gt(rowname_col = "Parameter") %>%
-  gt::fmt_number(
-    columns = c(`PCA-AGHQ`, `TMB`),
-    decimals = 3
-  ) %>%
+  rename(TMB = `KS(TMB, tmbstan)`, `PCA-AGHQ` = `KS(aghq, tmbstan)`) %>%
+  select(Parameter, TMB, `PCA-AGHQ`) %>%
+  mutate(Difference = TMB - `PCA-AGHQ`) %>%
+  gt::gt() %>%
   gt::summary_rows(
-   columns = everything(),
+   columns = c(`TMB`, `PCA-AGHQ`, Difference),
    fns = list(Average = ~ mean(.x)),
@@
    decimals = 3
-  )
+  ) %>%
+  gt::fmt_number(
+    columns = c(`TMB`, `PCA-AGHQ`, Difference),
+    decimals = 3
+  ) %>%
+  gt::text_transform(
+    locations = gt::cells_body("Parameter"),
+    fn = function(x) paste0("\\texttt{", x, "}")
+  ) %>%
+  gt::text_transform(
+    locations = gt::cells_body("Difference"),
+    fn = function(x) ifelse(x > 0, paste0("\\textbf{", x, "}"), paste0(x))
+  ) %>%
+  gt::as_latex() %>%
+  as.character() %>%
+  cat(file = "ks-summary.tex")
 ```
+
+\input{ks-summary.tex}
 

### 2023-07-20 src/docs_paper/paper.Rmd
Add hyperparameter names to PC loadings plot, add variance explained number (not just near 90), work on formatting of tables #55

@@
 Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
-The model is provided as a tool for countries to input their data to and generate estimates during a yearly process supported by UNAIDS.
+The model is provided as a tool for countries to input their data to and generate estimates with during a yearly process supported by UNAIDS.
 Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
@@
 
-To meet these requirements, inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
-Owing to its speed and flexibility, `TMB` is gaining popularity, particularly in spatial statistics [@osgood2022statistical], and via the `glmmTMB` \textsc{R} package [@brooks2017glmmTMB] which provides a user-friendly formula interface for fitting common models.
+To meet these requirements, inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb], which we refer to as TMB (distinguished from the package `TMB`).
+Owing to its speed and flexibility, `TMB` is gaining popularity, particularly in spatial statistics [@osgood2022statistical] and via the user-friendly `glmmTMB` \textsc{R} package [@brooks2017glmmTMB].
 Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the parameters.
 For the Naomi model, this subset is the high-dimensional latent field, leaving a smaller number of hyperparameters.
-`TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation, Taking inspiration from the AD Model Builder package [@fournier2012ad].
+Taking inspiration from the AD Model Builder package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
 
-Although the approach of `TMB` is fast, within the empirical Bayes framework hyperparameter uncertainty is not accounted for in the latent field posterior.
-This consideration motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
-We developed an inference method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over many hyperparameters.
+Although the TMB approach is fast, within the empirical Bayes framework hyperparameter uncertainty is not accounted for in the latent field posterior.
+We suspected this would result in underestimation of posterior variances, which may have implications 
+As such, we were motivated to look for an more fully Bayesian inference approach, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
+We developed an inference method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over large numbers of hyperparameters.
 AGHQ is a quadrature method based on the theory of polynomial interpolation, which is well suited to statistical estimation problems.
@@
 Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package [@bachl2019inlabru], INLA within MCMC [@gomez2018markov], and importance sampling with INLA [@berild2022importance], all of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
-`inlabru` approximates non-linear predictors using linearisation, which involves making iterative calls to `R-INLA`.
+The approach of `inlabru` is to approximate non-linear predictors using linearisation, by making iterative calls to `R-INLA`.
 <!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
@@
 Section \ref{sec:naomi} outlines the version of the Naomi model that we consider in this paper, and Section \ref{sec:elgm} describes how it falls within the ELGM framework.
-Section \ref{sec:inferencenaomi} outlines a deterministic inference method for ELGMs based on nested application of AGHQ and Laplace approximation, and how we modify it using PCA to be applicable to Naomi.
-In a case study, we evaluate the accuracy of our PCA-AGHQ method for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:results}.
-We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
-Finally, in Section \ref{sec:conclusions} we discuss our conclusions, and directions for future research.
+In Section \ref{sec:inferencenaomi} we review the deterministic inference method for ELGMs used by @stringer2022fast based on nested application of AGHQ and the Laplace approximation, before introducing the PCA-based modification we use to enable application to Naomi.
+In a case study (Section \ref{sec:results}) we evaluate the accuracy of PCA-AGHQ for the simplified Naomi model fit to data from Malawi, as compared with TMB and gold-standard MCMC.
+Finally, we discuss our conclusions, and directions for future research in Section \ref{sec:conclusions}.
 
@@
 
-We begin by describing (Section \ref{sec:inferenceelgm}) the inference method for ELGMs based on nested applications of the Laplace approximation and AGHQ used by @stringer2022fast.
+We first describe (Section \ref{sec:inferenceelgm}) the inference method for ELGMs based on nested applications of the Laplace approximation and AGHQ used by @stringer2022fast.
 We then propose (Section \ref{sec:pca}) an extension of the method which uses PCA to facilitate inference for Naomi, which otherwise would be intractable.
@@
 
-The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ is given by
+The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ in an ELGM is given by
 \begin{equation*}
@@
 Each evaluation in the optimisation requires an inner optimisation to obtain $\hat \x(\btheta)$ via Equation \ref{eq:mode}.
-Supposing the hyperparameters are to be considered fixed, then latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
-This is the approach used currently for Naomi, which we refer to as TMB (distinguished from the software \texttt{TMB}).
+Supposing the hyperparameters are to be considered fixed, as with the TMB approach used currently for Naomi, then latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
 Hyperparameter inferences can be obtained according to some method which should be specified here as well.
@@
 Quadrature rules can be used to approximate integrals using a set of nodes $\z \in \mathcal{Q}$ and a weighting function $\omega: \mathcal{Q} \to \mathbb{R}$.
-Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule, where the nodes are selected as zeros of the $k$th Hermite polynomial
+Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule, where in the univariate case the nodes are selected as zeros of the $k$th Hermite polynomial
 \begin{equation}
@@
 The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, where $\phi(\cdot)$ is a standard Gaussian density.
-GHQ is attractive because it is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$, and we expect posterior distributions to be relatively well described by this class of functions.
-Multivariate GHQ rules, required to integrate over $\btheta$, are typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
+GHQ is attractive because it is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.
+We expect posterior distributions to be relatively well described by this class of functions.
+To integrate over $\btheta$ we require a multivariate GHQ rule, which is typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
 In adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] the nodes are shifted and rotated to suit the particular integrand.
@@
 In combination with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ.
-We term this approach PCA-AGHQ, with estimate of the normalising constant given by
+We refer to this approach as PCA-AGHQ, with corresponding estimate of the normalising constant given by
 \begin{equation}
@@
 where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
-The final panel of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
+Panel C of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
 
@@
 
-1. TMB (`r signif(time_taken$TMB, 2)` seconds), using a Gaussian approximation at $\hat{\btheta}_\texttt{LA}$.
+1. TMB (`r signif(time_taken$TMB, 2)` seconds), based on a Gaussian approximation at $\hat{\btheta}_\texttt{LA}$.
 2.
-PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), using a mixture of Gaussians approximation at a PCA-AGHQ quadrature grid, as described in Section \ref{sec:pca}, and implemented via extension of the `aghq` package [@stringer2021implementing].
+PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), based on a Gaussian approximation mixture at the nodes of a PCA-AGHQ quadrature grid, as described in Section \ref{sec:pca}, and implemented via extension of the `aghq` package [@stringer2021implementing].
 3.
-NUTS (`r signif(time_taken$tmbstan, 2)` days), the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] implemented via the `tmbstan` package [@monnahan2018no].
+NUTS (`r signif(time_taken$tmbstan, 2)` days), the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling using Stan [@carpenter2017stan] implemented via the `tmbstan` package [@monnahan2018no].
 
@@
 PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-use}), $1000$ samples \\
-NUTS & \texttt{tmbstan} & $4$ chains of $100,000$ iterations, with the first $50,000$ iterations of each chain discarded as warmup, thinned by a factor of $40$, to give a total of $5000$ samples kept.
+NUTS & \texttt{tmbstan} & $4$ chains of $100000$ iterations, with the first $50000$ iterations of each chain discarded as warmup, thinned by a factor of $40$, to give a total of $5000$ samples kept.
 Default NUTS tuning parameters \citep{hoffman2014no}.
@@
 
-Obtaining acceptable NUTS diagnostics required four chains run in parallel for 100,000 iterations, thinned by a factor of 20 for ease-of-storage.
-The largest potential scale reduction factor [@gelman1992inference; @vehtari2021rank] was $\hat R = 1.021$ (Figure S7), and there were no divergent transitions.
+Due to low effective sample size ratios (Figure S$X$), obtaining acceptable NUTS diagnostics required four chains run in parallel for 100000 iterations, thinned by a factor of 20 for ease-of-storage.
+There were no divergent transitions, and the largest potential scale reduction factor [@gelman1992inference; @vehtari2021rank] was $\hat R = 1.021$ (Figure S7).
 We considered the NUTS results a gold-standard, though inaccuracies remain possible.
@@
 
-We used a Scree plot (Figure S2) based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})^{-1}$ to select the number of principal components to keep.
-We found that $s = 8$ principal components were sufficient to explain close to 90% of total variation.
-With this choice of $s$, the reduced rank approximation to the inverse curvature was visually similar (Figure S3).
+```{r}
+tv_df <- read_csv("depends/total-variation.csv")
+```
+
+We used a Scree plot based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})^{-1}$ to select the number of principal components to keep (Figure S2).
+We found that $s = 8$ principal components were sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
+This choice of $s$ gave a visually similar reduced rank approximation to the inverse curvature (Figure S3).
 
 ```{r node-positions, fig.cap="The 6561 PCA-AGHQ node positions (green, rug plot) projected onto the hyperparameter marginal posteriors (grey, histogram) for each of the 24 hyperparameters.
-Some hyperparameters, such as logit\\_phi\\_alpha\\_x, are well covered where as others, such as log\\_sigma\\_lambda\\_x, are near only being covered by one unique node."}
+Some hyperparameters, such as \\texttt{logit\\_phi\\_alpha\\_x}, are well covered where as others, such as \\texttt{log\\_sigma\\_lambda\\_x}, are near only being covered by one unique node."}
 knitr::include_graphics("depends/nodes-samples-comparison.png")
@@
 Overlaying the resulting $3^8 = 6561$ PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
-Though 12 improves on the 8 naively achievable using a dense grid, there remained many hyperparameters poorly covered.
-Coverage was associated with marginal standard deviation (Figure S4), which varied particularly according to the hyperparameter scale.
+Though 12 does improve on the 8 naively obtained with a dense grid, there remained poorly covered hyperparameters.
+Coverage was associated with marginal standard deviation (Figure S4).
 All constrained hyperparameters $\theta$ were transformed to the real line, using either a log ($\theta > 0$) or logit ($\theta \in [0, 1]$) transformation.
@@
 The first and second eigenvectors each contain coupled AR1 standard deviation and correlation parameters (Figure S6).
-As variability can be explained either by high standard deviation and high correlation or low standard deviation and low correlation these parameters are weakly identified.
-This is reflected in high correlation in the the NUTS posterior pairs plot (Figure S10, average absolute correlation across all four pairs of `r ar1_mean_cor`).
+These parameters are weakly identified, and have high correlation in the the NUTS posterior pairs plot (Figure S10, average absolute correlation across all four pairs of `r ar1_mean_cor`).
+The reason why is that the same amount of variation can equally be explained by high standard deviation and high correlation or low standard deviation and low correlation.
 The BYM2 standard deviation and proportion parameters on the other hand are designed to be orthogonal, and as such did not display posterior correlation (Figure S11, average absolute correlation across all four pairs of `r bym2_mean_cor`) or appear prominently in the eigenvectors.
 
-### Normalising constant assessment
+### Normalising constant estimation
 
@@
 where $\psi$ is a model parameter.
-We found that (Figure \ref{fig:contraction})) something something.
+We found that (Figure \ref{fig:contraction}) something something.
 For greater interpretability, facet parameters in this plot according to model component.
 
-```{r contraction, fig.cap="A posterior contraction tending toward 1.0 corresponds to a posterior tending towards a Dirac delta function.
-A posterior contraction of 0.0 corresponds to no change in standard deviation between prior and posterior.
-A posterior contraction less than 0.0 corresponds to a wider posterior than prior."}
+```{r contraction, fig.cap="A posterior contraction of 1 corresponds to a Dirac delta function.
+A posterior contraction of 0 corresponds to no change in standard deviation between prior and posterior.
+A posterior contraction less than 0 corresponds to a wider posterior than prior."}
 knitr::include_graphics("depends/posterior-contraction.png")

### 2023-07-20 src/naomi-simple_point-estimates/point-estimates.Rmd
combine_two_plots() and combine_three_plots() for point estimate latent and output plots #55

@@
 
-sd_plot_latent_alt <- sd_plot_latent +
+combine_two_plots <- function(plot1, plot2) {
+  plot1alt <- plot1 +
+    theme(legend.position = "none")
+
+  plot2alt <- plot2 +
   theme(
     strip.background = element_blank(),
-    strip.text.x = element_blank()
+    strip.text.x = element_blank(),
+    legend.position = "none"
   )
 
-y_axis <- ggplot(data.frame(l = mean_plot_latent$labels$y, x = 1, y = 1)) +
+  y_axis <- ggplot(data.frame(l = plot1$labels$y, x = 1, y = 1)) +
     geom_text(aes(x, y, label = l), angle = 90) +
@@
 
-mean_plot_latent_alt <- mean_plot_latent
-mean_plot_latent_alt$labels$x <- ""
-mean_plot_latent_alt$labels$y <- sd_plot_latent_alt$labels$y <- ""
+  plot1alt$labels$x <- ""
+  plot1alt$labels$y <- plot2alt$labels$y <- ""
 
-(mean_sd_plot_latent_alt <- y_axis + (mean_plot_latent_alt / sd_plot_latent_alt) +
-  plot_layout(widths = c(1, 30)))
+  y_axis + (plot1alt / plot2alt) +
+    plot_layout(widths = c(1, 35))
+}
 
-ggsave("mean-sd-alt-latent.png", mean_sd_plot_latent_alt, h = 6, w = 6.25)
+combine_two_plots(mean_plot_latent, sd_plot_latent)
+
+ggsave("mean-sd-alt-latent.png", h = 6, w = 6.25)
 ```
@@
 cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
+df_plot_output <- df_plot_output %>%
+  mutate(par = fct_recode(par, "HIV prevalence" = "rho_t1_out", "ART coverage" = "alpha_t1_out", "HIV incidence" = "lambda_t1_out"))
 
@@
 
-sd_plot_output_alt <- sd_plot_output +
-  theme(
-    strip.background = element_blank(),
-    strip.text.x = element_blank(),
-    legend.position = "bottom"
-  )
+combine_two_plots(mean_plot_output, sd_plot_output)
 
-y_axis <- ggplot(data.frame(l = mean_plot_output$labels$y, x = 1, y = 1)) +
-  geom_text(aes(x, y, label = l), angle = 90) +
-  theme_void() +
-  coord_cartesian(clip = "off")
-
-mean_plot_output_alt <- mean_plot_output +
-  theme(legend.position = "none")
-
-mean_plot_output_alt$labels$x <- ""
-mean_plot_output_alt$labels$y <- sd_plot_output_alt$labels$y <- ""
-
-(mean_sd_plot_output_alt <- y_axis + (mean_plot_output_alt / sd_plot_output_alt) +
-  plot_layout(widths = c(1, 30)))
-
-ggsave("mean-sd-alt-output.png", mean_sd_plot_output_alt, h = 6, w = 6.25, bg = "white")
+ggsave("mean-sd-alt-output.png", h = 6, w = 6.25)
 ```
@@
 
-plot_outputs(.indicator = "Posterior mean estimate", .par = "rho_t1_out", col = "#E69F00")
-plot_outputs(.indicator = "Posterior mean estimate", .par = "lambda_t1_out", col = "#009E73")
-plot_outputs(.indicator = "Posterior mean estimate", .par = "alpha_t1_out", col = "#56B4E9")
-plot_outputs(.indicator = "Posterior SD estimate", .par = "rho_t1_out", col = "#E69F00")
-plot_outputs(.indicator = "Posterior SD estimate", .par = "lambda_t1_out", col = "#009E73")
-plot_outputs(.indicator = "Posterior SD estimate", .par = "alpha_t1_out", col = "#56B4E9")
+mean_prev_plot <- plot_outputs(.indicator = "Posterior mean estimate", .par = "HIV prevalence", col = "#E69F00")
+mean_inc_plot <- plot_outputs(.indicator = "Posterior mean estimate", .par = "HIV incidence", col = "#009E73")
+mean_art_plot <- plot_outputs(.indicator = "Posterior mean estimate", .par = "ART coverage", col = "#56B4E9")
+sd_prev_plot <- plot_outputs(.indicator = "Posterior SD estimate", .par = "HIV prevalence", col = "#E69F00")
+sd_inc_plot <- plot_outputs(.indicator = "Posterior SD estimate", .par = "HIV incidence", col = "#009E73")
+sd_art_plot <- plot_outputs(.indicator = "Posterior SD estimate", .par = "ART coverage", col = "#56B4E9")
+
+#' Put together
+combine_three_plots <- function(plot1, plot2, plot3) {
+  plot1alt <- plot1 +
+    theme(legend.position = "none")
+
+  plot2alt <- plot2 +
+  theme(
+    strip.background = element_blank(),
+    strip.text.x = element_blank(),
+    legend.position = "none"
+  )
+
+  plot3alt <- plot3 +
+  theme(
+    strip.background = element_blank(),
+    strip.text.x = element_blank(),
+    legend.position = "none"
+  )
+
+  y_axis <- ggplot(data.frame(l = plot1$labels$y, x = 1, y = 1)) +
+    geom_text(aes(x, y, label = l), angle = 90) +
+    theme_void() +
+    coord_cartesian(clip = "off")
+
+  plot1alt$labels$x <- plot2alt$labels$x <- ""
+  plot1alt$labels$y <- plot2alt$labels$y <- plot3alt$labels$y <- ""
+
+  three_plots <- cowplot::plot_grid(plot1alt, plot2alt, plot3alt, ncol = 1)
+  cowplot::plot_grid(y_axis, three_plots, ncol = 2, rel_widths = c(1, 35))
+}
+
+combine_three_plots(mean_prev_plot, mean_inc_plot, mean_art_plot)
+ggsave("mean-alt-output.png", h = 7, w = 6.25)
+
+combine_three_plots(sd_prev_plot, sd_inc_plot, sd_art_plot)
+ggsave("sd-alt-output.png", h = 7, w = 6.25)
 ```

### 2023-07-20 src/docs_paper/appendix.Rmd
Add tables of point estimate estimation results to appendix #55

@@
 
-## Estimated normalising constant comparison
+## Normalising constant estimation
 
@@
 
-# Comprehensive inference comparison
+# Inference comparison
 
-In this section we give further inference comparison results, parameter by parameter.
+In this section we give more detailed inference comparison results for each parameter or group of parameters.
+For any parameter with length greater than one, the results presented are an average.
 
@@
 
+### Posterior mean
+
+```{r}
+point_summary <- read_csv("depends/point-estimates.csv")
+
+point_summary <- point_summary %>%
+  mutate(type = fct_recode(type, "Output" = "output", "Latent" = "latent"))
+
+latex_table_point_estimates <- function(.indicator) {
+  point_summary %>%
+  group_by(par, type, indicator, method) %>%
+  summarise(
+    rmse = sqrt(mean((truth - approximate)^2)),
+    mae = mean(abs(truth - approximate))
+  ) %>%
+  ungroup() %>%
+  pivot_wider(
+    names_from = "method",
+    values_from = c("rmse", "mae")
+  ) %>%
+  filter(indicator == .indicator) %>%
+  select(Parameter = par, Type = type, `TMB` = `rmse_TMB`, `PCA-AGHQ` = `rmse_PCA-AGHQ`) %>%
+  mutate(
+    Difference = `PCA-AGHQ` - TMB,
+    `Percent Difference` =  (`PCA-AGHQ` - TMB) / TMB
+  ) %>%
+  gt::gt(groupname_col = "Type") %>%
+  gt::fmt_number(
+    columns = c(`TMB`, `PCA-AGHQ`, Difference),
+    n_sigfig = 2
+  ) %>%
+ gt::fmt_percent(
+    columns = `Percent Difference`,
+    decimals = 1
+  ) %>%
+  gt::text_transform(
+    locations = gt::cells_body("Parameter"),
+    fn = function(x) paste0("\\texttt{", x, "}")
+  ) %>%
+  gt::as_latex() %>%
+  as.character()
+}
+
+latex_table_point_estimates("Posterior mean estimate") %>%
+  cat(file = "point-summary-mean.tex")
+
+latex_table_point_estimates("Posterior SD estimate") %>%
+  cat(file = "point-summary-sd.tex")
+```
+
+\input{point-summary-mean.tex}
+
+### Posterior standard deviation
+
+\input{point-summary-sd.tex}
+
 ## Distribution tests
+
+### Kolmogorov-Smirnov
 
@@
   select(-type) %>%
-  rename(TMB = `KS(TMB, tmbstan)`, `PCA-AGHQ` = `KS(aghq, tmbstan)`) %>%
-  select(Parameter, TMB, `PCA-AGHQ`) %>%
-  mutate(Difference = TMB - `PCA-AGHQ`) %>%
+  select(Parameter, TMB = `KS(TMB, tmbstan)`, `PCA-AGHQ` = `KS(aghq, tmbstan)`) %>%
+  mutate(Difference = `PCA-AGHQ` - TMB) %>%
   gt::gt() %>%
@@
     locations = gt::cells_body("Difference"),
-    fn = function(x) ifelse(x > 0, paste0("\\textbf{", x, "}"), paste0(x))
+    fn = function(x) ifelse(x < 0, paste0("\\textbf{", x, "}"), paste0(x))
   ) %>%

### 2023-07-20 src/naomi-simple_point-estimates/point-estimates.Rmd
Add tables of point estimate estimation results to appendix #55

@@
 ```{r}
+write_csv(df_plot, "point-estimates.csv")
+
 df_metrics <- df_plot %>%

### 2023-07-20 src/docs_paper/paper.Rmd
Improve writing of AGHQ section #55

@@
 
-Quadrature rules can be used to approximate integrals using a set of nodes $\z \in \mathcal{Q}$ and a weighting function $\omega: \mathcal{Q} \to \mathbb{R}$.
-Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule, where in the univariate case the nodes are selected as zeros of the $k$th Hermite polynomial
+Let $\z \in \mathcal{Q}(m, k)$ be $m$-dimensional Gauss-Hermite quadrature [GHQ; @davis1975methods] rule with $k$ nodes per dimension constructed using the product rule such that $\mathcal{Q}(m, k) = \mathcal{Q}(1, k) \times \cdots \times \mathcal{Q}(1, k)$ where
 \begin{equation}
-\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}.
+\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}, \\
 \end{equation}
-The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, where $\phi(\cdot)$ is a standard Gaussian density.
-GHQ is attractive because it is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.
-We expect posterior distributions to be relatively well described by this class of functions.
-To integrate over $\btheta$ we require a multivariate GHQ rule, which is typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
-In adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] the nodes are shifted and rotated to suit the particular integrand.
-Repositioning the nodes is especially important in statistical quadrature problems, where the integral depends on data $\y$ and regions of high density are not known in advance.
+with $\phi(\cdot)$ is a standard Gaussian density.
+The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$.
+For $k = 1$ GHQ corresponds to a Laplace approximation.
+Further GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$, a class of functions we expect the posterior to be close to.
 
-Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then an AGHQ estimate of the normalising constant $p(\y)$ based on the Laplace approximation is
+Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature.
+An adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] estimate of the normalising constant $p(\y)$ based on the Laplace approximation is given by
 \begin{equation}
@@
 \end{equation}
-The unadapted nodes are shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
-Two alternatives [@jackel2005note] are
+The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
+Repositioning the nodes is crucial for statistical quadrature problems like ours, where the integral depends on data $\y$ and regions of high density are not known in advance.
+Two alternatives for the matrix decomposition [@jackel2005note] are
 1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
@@
 the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
-Equation \ref{eq:aghq} may then be used to normalise the Laplace approximation
+This estimate may be used to normalise the Laplace approximation
 \begin{equation}
@@
 Panel C of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
+As AGHQ with $k = 1$ corresponds to the Laplace approximation, PCA-AGHQ can be interpreted as performing AGHQ on the first $s$ principal components of the inverse curvature, and a Laplace approximation on the remaining $m - s$ principal components.
+Inference for the latent field follows analogously to Equation \ref{eq:nest}.
 

### 2023-07-20 src/docs_paper/appendix.Rmd
Cross-linked section and figure references between main text and appendix #55

@@
 cls: imsart.cls
+header-includes:
+  \usepackage{xr} \externaldocument{paper}
 ---
@@
 
+# Model assessment\label{sec:model-assessment}
+
+## Additional figures
+
+```{r contraction, fig.cap="A posterior contraction of 1 corresponds to a Dirac delta function.
+A posterior contraction of 0 corresponds to no change in standard deviation between prior and posterior.
+A posterior contraction less than 0 corresponds to a wider posterior than prior."}
+knitr::include_graphics("depends/posterior-contraction.png")
+```
+
 # AGHQ and PCA-AGHQ details
@@
 
-## Normalising constant estimation
+## Normalising constant estimation\label{sec:normalising}
 
@@
 
-# MCMC convergence and suitability
+# MCMC convergence and suitability\label{sec:mcmc}
 
@@
 
-# Algorithm using Laplace latent field marginals
+# Algorithm using Laplace latent field marginals\label{sec:laplace-marginals}
 

### 2023-07-20 src/docs_paper/paper.Rmd
Cross-linked section and figure references between main text and appendix #55

@@
 biblio-style: imsart-nameyear # alternative: imsart-number
+header-includes:
+  \usepackage{xr} \externaldocument{appendix}
 output:
@@
 We consider a simplified version defined only at the time of the most recent household survey with HIV testing, omitting nowcasting and temporal projection, as these time points involve limited inferences.
-An overview of the simplified model is given below, and a more complete mathematical description is provided in Appendix S1.
+An overview of the simplified model is given below, and a more complete mathematical description is provided in Appendix \ref{sec:math}.
 
@@
 HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i$, and depends on adult HIV prevalence and adult ART coverage.
-The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix S1.
+The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix \ref{sec:math}.
 Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
@@
 This sum is over both strata $i \in I$ and the number of ART clients travelling from district $x'$ to $x$.
-More details regarding this part of the model are provided in Appendix S1.
+More details regarding this part of the model are provided in Appendix \ref{sec:math}.
 
@@
 
-Due to low effective sample size ratios (Figure S$X$), obtaining acceptable NUTS diagnostics required four chains run in parallel for 100000 iterations, thinned by a factor of 20 for ease-of-storage.
-There were no divergent transitions, and the largest potential scale reduction factor [@gelman1992inference; @vehtari2021rank] was $\hat R = 1.021$ (Figure S7).
+Due to low effective sample size ratios (Figure \ref{fig:ratio}), obtaining acceptable NUTS diagnostics required four chains run in parallel for 100000 iterations, thinned by a factor of 20 for ease-of-storage.
+There were no divergent transitions, and the largest potential scale reduction factor [@gelman1992inference; @vehtari2021rank] was $\hat R = 1.021$ (Figure \ref{fig:rhat}).
 We considered the NUTS results a gold-standard, though inaccuracies remain possible.
-For full details see Appendix S2.
+For full details see Appendix \ref{sec:mcmc}.
 
@@
 
-We used a Scree plot based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})^{-1}$ to select the number of principal components to keep (Figure S2).
+We used a Scree plot based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})^{-1}$ to select the number of principal components to keep (Figure \ref{fig:scree}).
 We found that $s = 8$ principal components were sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
-This choice of $s$ gave a visually similar reduced rank approximation to the inverse curvature (Figure S3).
+This choice of $s$ gave a visually similar reduced rank approximation to the inverse curvature (Figure \ref{fig:reduced-rank}).
 
@@
 Though 12 does improve on the 8 naively obtained with a dense grid, there remained poorly covered hyperparameters.
-Coverage was associated with marginal standard deviation (Figure S4).
+Coverage was associated with marginal standard deviation (Figure \ref{fig:nodes-quantiles-sd}).
 All constrained hyperparameters $\theta$ were transformed to the real line, using either a log ($\theta > 0$) or logit ($\theta \in [0, 1]$) transformation.
-As a result, marginal standard deviations for log transformed hyperparameters were systematically smaller than those which were logit transformed (Figure S5).
+As a result, marginal standard deviations for log transformed hyperparameters were systematically smaller than those which were logit transformed (Figure \ref{fig:marginal-sd}).
 
@@
 The ordered eigenvectors correspond to the directions of greatest variation in the inverse curvature.
-The first and second eigenvectors each contain coupled AR1 standard deviation and correlation parameters (Figure S6).
-These parameters are weakly identified, and have high correlation in the the NUTS posterior pairs plot (Figure S10, average absolute correlation across all four pairs of `r ar1_mean_cor`).
+The first and second eigenvectors each contain coupled AR1 standard deviation and correlation parameters (Figure \ref{fig:pc-loadings}).
+These parameters are weakly identified, and have high correlation in the the NUTS posterior pairs plot (Figure \ref{fig:rho_a}, average absolute correlation across all four pairs of `r ar1_mean_cor`).
 The reason why is that the same amount of variation can equally be explained by high standard deviation and high correlation or low standard deviation and low correlation.
-The BYM2 standard deviation and proportion parameters on the other hand are designed to be orthogonal, and as such did not display posterior correlation (Figure S11, average absolute correlation across all four pairs of `r bym2_mean_cor`) or appear prominently in the eigenvectors.
+The BYM2 standard deviation and proportion parameters on the other hand are designed to be orthogonal, and as such did not display posterior correlation (Figure \ref{fig:alpha_x}, average absolute correlation across all four pairs of `r bym2_mean_cor`) or appear prominently in the eigenvectors.
 
@@
 Convergence in $\log p_\texttt{PCA}(\y)$ as $s$ and $k$ are increased may suggest a suitable grid has been reached.
-Appendix S3 shows those values which we could compute in a reasonable time (less than 24 hours using a high performance computing cluster).
+Appendix \ref{sec:normalising} shows those values which we could compute in a reasonable time (less than 24 hours using a high performance computing cluster).
 
@@
 For greater interpretability, facet parameters in this plot according to model component.
-
-```{r contraction, fig.cap="A posterior contraction of 1 corresponds to a Dirac delta function.
-A posterior contraction of 0 corresponds to no change in standard deviation between prior and posterior.
-A posterior contraction less than 0 corresponds to a wider posterior than prior."}
-knitr::include_graphics("depends/posterior-contraction.png")
-```
 
@@
 
-See Appendix S4.
+See Appendix \ref{sec:laplace-marginals}.
 

### 2023-07-20 src/docs_paper/script.R

@@ -18,8 +18,15 @@ convert_pdf_png <- function(name) {
 convert_pdf_png("nodes-samples-comparison")
 
 #' Render documents
-rmarkdown::render("paper.Rmd")
-rmarkdown::render("appendix.Rmd")
+#' Found a way to cross-link references: https://stackoverflow.com/questions/52531637/knitr-rmarkdown-latex-how-to-cross-reference-figures-and-tables-in-2-different/52532269#52532269
+rmarkdown::render("paper.Rmd", clean = FALSE)
+rmarkdown::render("appendix.Rmd", clean = FALSE)
+
+tinytex::pdflatex("paper.tex", clean = FALSE)
+tinytex::pdflatex("appendix.tex", clean = FALSE)
+
+tinytex::pdflatex("paper.tex")
+tinytex::pdflatex("appendix.tex")
 
 #' I just used this function to check some things as I was writing!
 kish_ess <- function(w) {

### 2023-07-20 src/docs_paper/paper.Rmd
Custom version of the Naomi output figure #55

@@
 
-```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted using TMB."}
-knitr::include_graphics("depends/naomi-results.png")
-
-# Eventually this figure should be generated using PCA-AGHQ
-# knitr::include_graphics("depends/figB.png")
+```{r naomi-results, fig.cap="District-level HIV prevalence (A), ART coverage (B), and HIV incidence and new cases (C) for adults 15-49 in Malawi. Inference conducted using TMB."}
+knitr::include_graphics("figB.png")
 ```

### 2023-07-20 src/docs_paper/figures.R

@@ -1,184 +1,0 @@
-library(tidyverse)
-library(patchwork)
-
-cols <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
-
-#' Fig A
-
-TMB::compile("2d.cpp")
-dyn.load(TMB::dynlib("2d"))
-
-obj <- TMB::MakeADFun(data = list(), parameters = list(theta1 = 0, theta2 = 0), DLL = "2d")
-
-box_lower <- -5
-box_upper <- 10
-box_size <- box_upper - box_lower
-
-grid <- expand.grid(
-  theta1 = seq(box_lower, box_upper, length.out = box_size * 50),
-  theta2 = seq(box_lower, box_upper, length.out = box_size * 50)
-)
-
-ground_truth <- cbind(grid, pdf = apply(grid, 1, function(x) exp(-1 * obj$fn(x))))
-
-opt <- nlminb(
-  start = obj$par,
-  objective = obj$fn,
-  gradient = obj$gr,
-  control = list(iter.max = 1000, trace = 0)
-)
-
-sd_out <- TMB::sdreport(
-  obj,
-  par.fixed = opt$par,
-  getJointPrecision = TRUE
-)
-
-mu <- opt$par
-cov <- sd_out$cov.fixed
-
-figA0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
-  geom_contour(col = "lightgrey") +
-  coord_fixed(xlim = c(box_lower, box_upper), ylim = c(box_lower, box_upper), ratio = 1) +
-  labs(x = "", y = "") +
-  theme_minimal() +
-  guides(size = "none") +
-  theme(
-    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
-    axis.text.y = element_blank(), axis.ticks.y = element_blank()
-  )
-
-gg <- mvQuad::createNIGrid(2, "GHe", 3)
-
-add_points <- function(figA0, gg) {
-
-  points <- mvQuad::getNodes(gg) %>%
-    as.data.frame() %>%
-    mutate(weights = mvQuad::getWeights(gg))
-
-  colnames(points) <- c("theta1", "theta2", "weights")
-
-  figA0 +
-    geom_point(
-      data = points,
-      aes(x = theta1, y = theta2, size = weights),
[…]

### 2023-07-20 src/docs_paper/script.R

@@ -1,7 +1,7 @@
 # orderly::orderly_develop_start("docs_paper")
 # setwd("src/docs_paper")
 
-source("figures.R")
+source("figs.R")
 
 #' Conversion of figures from .pdf at specified resolution
 Sys.setenv(PATH = paste(Sys.getenv("PATH"), "/opt/homebrew/bin", sep = ":"))
@@ -27,6 +27,8 @@ tinytex::pdflatex("appendix.tex", clean = FALSE)
 
 tinytex::pdflatex("paper.tex")
 tinytex::pdflatex("appendix.tex")
+
+pdftools::pdf_combine(c("paper.pdf", "appendix.pdf"), output = "naomi-aghq.pdf")
 
 #' I just used this function to check some things as I was writing!
 kish_ess <- function(w) {

### 2023-07-20 src/docs_paper/figs.R
Add redundancy to HIV outputs plot using captions #55

@@ -1,0 +1,326 @@
+library(tidyverse)
+library(patchwork)
+
+cols <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
+#' Fig A
+
+TMB::compile("2d.cpp")
+dyn.load(TMB::dynlib("2d"))
+
+obj <- TMB::MakeADFun(data = list(), parameters = list(theta1 = 0, theta2 = 0), DLL = "2d")
+
+box_lower <- -5
+box_upper <- 10
+box_size <- box_upper - box_lower
+
+grid <- expand.grid(
+  theta1 = seq(box_lower, box_upper, length.out = box_size * 50),
+  theta2 = seq(box_lower, box_upper, length.out = box_size * 50)
+)
+
+ground_truth <- cbind(grid, pdf = apply(grid, 1, function(x) exp(-1 * obj$fn(x))))
+
+opt <- nlminb(
+  start = obj$par,
+  objective = obj$fn,
+  gradient = obj$gr,
+  control = list(iter.max = 1000, trace = 0)
+)
+
+sd_out <- TMB::sdreport(
+  obj,
+  par.fixed = opt$par,
+  getJointPrecision = TRUE
+)
+
+mu <- opt$par
+cov <- sd_out$cov.fixed
+
+figA0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
+  geom_contour(col = "lightgrey") +
+  coord_fixed(xlim = c(box_lower, box_upper), ylim = c(box_lower, box_upper), ratio = 1) +
+  labs(x = "", y = "") +
+  theme_minimal() +
+  guides(size = "none") +
+  theme(
+    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
+    axis.text.y = element_blank(), axis.ticks.y = element_blank()
+  )
+
+gg <- mvQuad::createNIGrid(2, "GHe", 3)
+
+add_points <- function(figA0, gg) {
+
+  points <- mvQuad::getNodes(gg) %>%
+    as.data.frame() %>%
+    mutate(weights = mvQuad::getWeights(gg))
+
+  colnames(points) <- c("theta1", "theta2", "weights")
+
+  figA0 +
+    geom_point(
+      data = points,
+      aes(x = theta1, y = theta2, size = weights),
+      alpha = 0.8,
+      col = "#009E73",
+      inherit.aes = FALSE
+    ) +
[…]

### 2023-07-20 src/docs_paper/paper.Rmd
Add redundancy to HIV outputs plot using captions #55
Import point estimates for outputs plot #55

@@
 Two alternatives for the matrix decomposition [@jackel2005note] are
-1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
-2.
-the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
+(1) the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
+(2) the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
 This estimate may be used to normalise the Laplace approximation
@@
 2.
-PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), based on a Gaussian approximation mixture at the nodes of a PCA-AGHQ quadrature grid, as described in Section \ref{sec:pca}, and implemented via extension of the `aghq` package [@stringer2021implementing].
+PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), based on a Gaussian approximation mixture at the adapted nodes $\z \in \mathcal{Q}(m, s, k)$, as described in Section \ref{sec:pca}, and implemented via extension of the `aghq` package [@stringer2021implementing].
 3.
@@
 
-```{r naomi-results, fig.cap="District-level HIV prevalence (A), ART coverage (B), and HIV incidence and new cases (C) for adults 15-49 in Malawi. Inference conducted using TMB."}
+```{r naomi-results, fig.cap="District-level HIV prevalence (A), ART coverage (B), and new HIV cases and HIV incidence (C) for adults 15-49 in Malawi. Inference conducted using TMB."}
 knitr::include_graphics("figB.png")
@@
 
-## Inference comparison \label{sec:inf-comparison}
+## Inference comparison\label{sec:inf-comparison}
 
-To compare the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS we assessed (1) marginal point estimates, (2) marginal Kolmogorov-Smirnov and Anderson-Darling tests using the empirical cumulative distribution function (ECDF), (3) joint Pareto-smoothed importance sampling results, and (4) joint maximum mean discrepancy results.
+We compared the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS for latent field parameters and model outputs.
+The metrics we used were (1) marginal point estimates, (2) marginal Kolmogorov-Smirnov and Anderson-Darling tests using the empirical cumulative distribution function (ECDF), (3) joint Pareto-smoothed importance sampling results, and (4) joint maximum mean discrepancy results.
 
@@
 
-The root mean square error (RMSE) between posterior mean estimates from PCA-AGHQ and NUTS (`r rmse_ahgq_mean`) was `r abs(rmse_diff_mean)`% lower than that between TMB and NUTS (`r rmse_tmb_mean`).
+For the latent field, the root mean square error (RMSE) between posterior mean estimates from PCA-AGHQ and NUTS (`r rmse_ahgq_mean`) was `r abs(rmse_diff_mean)`% lower than that between TMB and NUTS (`r rmse_tmb_mean`).
 For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_ahgq_sd)` (PCA-AGHQ).
-These results, alongside those for the mean absolute error (MAE), are presented in Figure \ref{fig:mean-sd}.
+These results, alongside those for the mean absolute error (MAE), are presented in Figure \ref{fig:mean-sd-latent}.
 
-```{r mean-sd, fig.cap="PCA-AGHQ modestly improves estimation of the posterior mean, and substantially improves estimation of the posterior standard deviation, as compared with TMB."}
+These improvements did not transfer to the model outputs (Figure \ref{fig:mean-sd-output}).
+
+```{r mean-sd-latent, fig.cap="For the latent field PCA-AGHQ modestly improves estimation of the posterior mean, and substantially improves estimation of the posterior standard deviation, as compared with TMB."}
 knitr::include_graphics("depends/mean-sd-alt-latent.png")
+```
+
+```{r mean-sd-output, fig.cap="PCA-AGHQ doesn't do so well for the model outputs."}
+knitr::include_graphics("depends/mean-sd-alt-output.png")
 ```

### 2023-07-21 src/docs_statml-poster/statml-poster.Rmd
Start work on StatML poster #57

--- 2023-06-06 src/docs_bioinference-poster/bioinference-poster.Rmd
@@
 # Essentials
-title: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model
+title: Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature
 # titlebox_height: "10%"
@@
     address: Center for Communicable Disease Dynamics, Department of Epidemiology, Harvard T.H. Chan School of Public Health
-logoleft_name: "bioinference-joined.png"
+logoleft_name: "mrc-gida-icl-uw-hthc-mlgh-uo.png"
 # Style & formatting
@@
 * We developed an approximate Bayesian inference method using Laplace approximation, adaptive Gauss-Hermite quadrature and principal component analysis
-* Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
-* Implemented as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a Template Model Builder `TMB` [@kristensen2016tmb] C++ user template
+* Motivation: an evidence synthesis model for small-area estimation of HIV indicators in >35 countries in sub-Saharan Africa
+* Implemented using the `aghq` package [@stringer2021implementing], and compatible with any model with a Template Model Builder `TMB` [@kristensen2016tmb] C++ user template
 
@@
 
-* District-level model of HIV indicators [@eaton2021naomi] which synthesises data from 1) household surveys, 2) antenatal care (ANC) clinics, and 3) routine service provision of antiretroviral therapy (ART)
+* District-level model of HIV indicators which synthesises data from 1) household surveys, 2) antenatal care (ANC) clinics, and 3) routine service provision of antiretroviral therapy (ART) [@eaton2021naomi]
   * Combining evidence from multiple data sources helps overcome the limitations of any one
-  * Small-area estimation methods to overcome small district-level sample sizes
-* Yearly estimation process: model run interactively by country teams using a web-app `naomi.unaids.org`
+  * Small-area estimation methods (structured random effects) to overcome limited district-level sample sizes
+* Yearly estimation process: model run interactively by by country teams using `naomi.unaids.org` web-app
   * Figure \@ref(fig:fig1) illustrates the seven stages of using the app
 * Inference conducted in minutes using empirical Bayes and a Gaussian approximation
-* It would take days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no], and this is not practical in this setting
-* We are looking for a fast, approximate approach, that properly takes uncertainty in hyperparameters into account
+* Days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no]: not practical!
+* Naomi has a large latent field $x$ controlled by a smaller number of hyperparameters $\theta$
+* Extended latent Gaussian model [@stringer2022fast]: more complex dependency structures than a latent Gaussian model [@rue2009approximate]
+* Looking for a fast, approximate approach, that properly takes uncertainty in hyperparameters into account
 
-```{r fig1, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model fitting occurs interactively in stages.'}
+```{r fig1, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model fitting occurs interactively in stages'}
 knitr::include_graphics("fig1.png")
@@
 
-## 2. Extended latent Gaussian models
-
-* Latent Gaussian models (LGMs) [@rue2009approximate] are three stage hierarchical models with observations $y$, Gaussian latent field $x$ and hyperparameters $\theta$
-* In an LGM the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
-* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$ and $\mathcal{J_i}$ is some set of indices
-  * Allows a higher degree of non-linearity in the model
-* Naomi is an ELGM, not an LGM, because it includes complex dependency structures:
-  1. Incidence depends on prevalence and ART coverage
-  2. Incidence ane prevalence linked to recent infection
-  3. ANC offset from household survey
-  4. ART coverage and recent infection are products
-  5. Observed data are aggregated finer processes
-  6. ART attendance uses the multinomial
-  7. Multiple link functions
-* We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
-* Though we focus on Naomi, the HIV Inference Group (`hiv-inference.org`) works on many other complex models, challenging for existing Bayesian inference methods, which require flexible modelling tools
-
-$$\\[0.1in]$$
-
-## 3. Inference procedure
+## 2. Inference procedure
 
@@
 
-## 4. Application to Malawi
+## 3. Application to Malawi
 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-time_taken <- readr::read_csv("depends/time_taken.csv")
+time_taken <- readr::read_csv("depends/time-taken.csv")
 ```
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-ks <- readRDS("depends/ks_summary.rds")
+ks <- readRDS("depends/ks-summary.rds")
 
 ks_reduction <- ks %>%
-  filter(Type == "Latent field") %>%
+  filter(type == "Latent field") %>%
   summarise(
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-second90 <- readr::read_csv("depends/second90.csv")
-
-second90_reduction <- second90 %>%
-  summarise(
-    rmse_tmb = 100 * sqrt(mean((tmbstan - TMB)^2)),
-    rmse_aghq = 100 * sqrt(mean((tmbstan - aghq)^2))
-  ) %>%
-  mutate(reduction = 100 * (rmse_aghq - rmse_tmb) / rmse_tmb) %>%
-  pull(reduction) %>%
-  round()
+# second90 <- readr::read_csv("depends/second90.csv")
+# 
+# second90_reduction <- second90 %>%
+#   summarise(
+#     rmse_tmb = 100 * sqrt(mean((tmbstan - TMB)^2)),
+#     rmse_aghq = 100 * sqrt(mean((tmbstan - aghq)^2))
+#   ) %>%
+#   mutate(reduction = 100 * (rmse_aghq - rmse_tmb) / rmse_tmb) %>%
+#   pull(reduction) %>%
+#   round()
 ```
@@
   * Also considering joint posteriors via Pareto-smoothed importance sampling and maximum mean discrepancy
-* Naomi can be used to assess probabilities targets have been met e.g.
-90% of those who know their HIV status are on ART ("second 90").
-Both TMB and PCA-AGHQ have biased inferences (Figure \@ref(fig:fig4))
-  * Reduced RMSE for estimating second 90 exceedance probabilities by `r abs(second90_reduction)`%
+* Naomi can be used to assess probabilities targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90"). Both TMB and PCA-AGHQ have biased inferences
 
-```{r fig4, echo=FALSE, fig.align='center', fig.cap='Both approximate methods are meaningfully incorrect for policy.'}
-knitr::include_graphics("fig4.png")
-```
+<!-- ```{r fig4, echo=FALSE, fig.align='center', fig.cap='Both approximate methods are meaningfully incorrect for policy.'} -->
+<!-- knitr::include_graphics("fig4.png") -->
+<!-- ``` -->
 
-## 5. Future directions
+## 4. Future directions
 

### 2023-07-21 src/docs_paper/appendix.Rmd
Add plot with ESS against KS result for TMB and PCA-AGHQ #55

@@
 
+```{r ks-ess, fig.cap="The ESS was negatively correlated with KS test statistic for both TMB and PCA-AGHQ.
+This could be explained in two ways.
+First, harder to sample parameters could be more difficult to estimate with approximate methods.
+Second, as more effective samples are collected, the NUTS posterior could become closer to the approximate posteriors."}
+knitr::include_graphics("depends/ks-ess.png")
+```
+
 \clearpage

### 2023-07-21 src/naomi-simple_ks/ks.Rmd
Add plot with ESS against KS result for TMB and PCA-AGHQ #55

@@
 ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc$stanfit)
-niter <- 4 * tmbstan_details$niter / tmbstan_details$nthin
+niter <- 0.5 * 4 * tmbstan_details$niter / tmbstan_details$nthin
 ess <- ess_ratio * niter
+
+fct_reorg <- function(fac, ...) {
+  fct_recode(fct_relevel(fac, ...), ...)
+}
 
@@
   filter(type != "Hyper") %>%
+  mutate(method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq")) %>%
   left_join(data.frame(ess) %>%
@@
   ggplot(aes(x = ess, y = ks)) +
-    geom_point(alpha = 0.5) +
-    geom_smooth(method = "lm", color = "#56B4E9") +
+    geom_point(shape = 1, alpha = 0.5) +
+    geom_smooth(method = "lm", color = "#56B4E9", fullrange = TRUE) +
+    stat_poly_eq(use_label("eq"), label.x.npc = "right") +
+    stat_poly_eq(label.y = 0.9, label.x.npc = "right") +
+    scale_x_continuous(limits = c(0, NA)) +
     facet_grid(~ method) +
     theme_minimal() +
-    labs(x = "ESS", y = "KS(method, tmbstan)")
+    labs(x = "ESS", y = "KS(method, NUTS)")
+
+ggsave("ks-ess.png", h = 4, w = 6.25)
 ```

### 2023-07-21 src/naomi-simple_ks/ks.Rmd
Better KS summary plot (still some workshopping to go), worst parameters automatically plotted #55

@@
   parname = c(unique(x_names), unique(theta_names)),
-  type = c(rep("Latent field", length(unique(x_names))), rep("Hyper", length(unique(theta_names))))
+  type = c(rep("Latent", length(unique(x_names))), rep("Hyper", length(unique(theta_names))))
 )
@@
   group_by(method, parname, type) %>%
-  summarise(ks = signif(mean(ks), 4)) %>%
-  pivot_wider(names_from = "method", values_from = "ks") %>%
-  rename(
-    "Parameter" = "parname",
-    "KS(aghq, tmbstan)" = "aghq",
-    "KS(TMB, tmbstan)" = "TMB",
-  )
+  summarise(
+    ks = mean(ks),
+    size = n()
+  ) %>%
+  pivot_wider(names_from = "method", values_from = "ks")
 
@@
 
-ks_summary %>%
-  filter(type != "Hyper") %>%
-  ks_plot_many("TMB", "aghq")
+extended_cbpalette <- colorRampPalette(multi.utils::cbpalette())
+
+ks_summary_latent <- filter(ks_summary, type == "Latent field")
+
+xy_length <- min(1, max(ks_summary_latent$aghq, ks_summary_latent$TMB) + 0.03)
+
+ks_summary_latent %>%
+  ggplot(aes(x = TMB, y = aghq, col = parname, size = size)) +
+  geom_point(alpha = 0.4) +
+  xlim(0, xy_length) +
+  ylim(0, xy_length) +
+  scale_color_manual(values = extended_cbpalette(n = 20)) +
+  scale_size_continuous(breaks = c(2, 10, 32), labels = c(2, 10, 32)) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  labs(x = "KS(TMB, NUTS)", y = "KS(PCA-AGHQ, NUTS)", col = "Parameter", size = "Length", subtitle = "Smaller KS values indicate higher accuracy") +
+  theme_minimal() +
+  guides(size = guide_legend(title.position = "top", direction = "vertical"), col = guide_legend(title.position = "top")) +
+  theme(legend.position = "bottom", legend.title = element_text(size = rel(0.9)), legend.text = element_text(size = rel(0.7)))
+
+ggsave("ks-summary.png", h = 5, w = 6.25, bg = "white")
 ```
@@
   summarise(
-    TMB = mean(`KS(TMB, tmbstan)`),
-    aghq = mean(`KS(aghq, tmbstan)`)
+    TMB = mean(TMB),
+    aghq = mean(aghq)
   ) %>%
@@
 ks_summary %>%
-  filter(type == "Latent field") %>%
-  mutate(diff = signif(`KS(TMB, tmbstan)` - `KS(aghq, tmbstan)`, 3)) %>%
+  filter(type == "Latent") %>%
+  mutate(diff = signif(TMB - aghq, 3)) %>%
   DT::datatable()
@@
 ```{r}
-ks_df_wide %>%
-  arrange(desc(diff)) %>%
-  head(n = 10)
+(tmb_beats_aghq <- ks_df_wide %>%
+  filter(!is.na(type)) %>%
+  arrange(diff) %>%
+  head(n = 10))
 
-histogram_and_ecdf(par = "us_alpha_xs", i = 18)
-histogram_and_ecdf(par = "us_alpha_xs", i = 16)
-histogram_and_ecdf(par = "u_alpha_xs", i = 13)
+histogram_and_ecdf(par = tmb_beats_aghq$parname[1], i = tmb_beats_aghq$index[1])
+histogram_and_ecdf(par = tmb_beats_aghq$parname[2], i = tmb_beats_aghq$index[2])
+histogram_and_ecdf(par = tmb_beats_aghq$parname[3], i = tmb_beats_aghq$index[3])
 ```
@@
 ```{r}
-ks_df_wide %>%
-  arrange(diff) %>%
-  head(n = 10)
+(aghq_beats_tmb <- ks_df_wide %>%
+  filter(!is.na(type)) %>%
+  arrange(desc(diff)) %>%
+  head(n = 10))
 
-histogram_and_ecdf(par = "ui_anc_rho_x", i = 27)
-histogram_and_ecdf(par = "ui_anc_alpha_x", i = 19)
-histogram_and_ecdf(par = "u_alpha_x", i = 18)
+histogram_and_ecdf(par = aghq_beats_tmb$parname[1], i = aghq_beats_tmb$index[1])
+histogram_and_ecdf(par = aghq_beats_tmb$parname[2], i = aghq_beats_tmb$index[2])
+histogram_and_ecdf(par = aghq_beats_tmb$parname[3], i = aghq_beats_tmb$index[3])
 ```

### 2023-07-21 src/naomi-simple_ks/functions.R

@@ -199,40 +199,3 @@ ks_plot <- function(ks_df, par, method1 = "TMB", method2 = "aghq", alpha = 0.5) 
   scatterplot + jitterplot +
     plot_layout(widths = c(2, 1))
 }
-
-#' Create a density plot and ridgeplot of the KS test statistics
-#'
-#' @param ks_df The output of `to_ks_df`
-#' @param par Parameter name (only used for labelling)
-#' @param method1 Samples from this method will be used as the first entry in the KS test
-#' @param method2 Samples from this method will be used as the second entry in the KS test
-#' @return A `ggplot2` object
-ks_plot_many <- function(ks_summary, method1, method2) {
-  ks_method1 <- paste0("KS(", method1, ", tmbstan)")
-  ks_method2 <- paste0("KS(", method2, ", tmbstan)")
-
-  xy_length <- min(1, max(ks_summary[[ks_method1]], ks_summary[[ks_method2]]) + 0.03)
-
-  densityplot <- ggplot() +
-    stat_density_2d(data = ks_summary, aes(x = .data[[ks_method1]], y = .data[[ks_method2]], linetype = type), col = "black") +
-    xlim(0, xy_length) +
-    ylim(0, xy_length) +
-    geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
-    scale_linetype_manual(values = c("solid", "dashed")) +
-    labs(x = ks_method1, y = ks_method2) +
-    theme_minimal() +
-    guides(fill = "none") +
-    theme(legend.position = "bottom")
-
-  ks_summary[["KS difference"]] <- ks_summary[[ks_method1]] - ks_summary[[ks_method2]]
-
-  ridgeplot <- ggplot(ks_summary, aes(y = type, x = `KS difference`)) +
-    ggridges::geom_density_ridges(alpha = 0.7, fill = NA, aes(linetype = type)) +
-    coord_flip() +
-    scale_linetype_manual(values = c("solid", "dashed")) +
-    labs(y = "", x = paste0(ks_method1, " - ", ks_method2)) +
-    guides(linetype = "none") +
-    theme_minimal()
-
-  densityplot + ridgeplot
-}

### 2023-07-21 src/docs_statml-poster/statml-poster.Rmd
Poster update! Nicer figure 2, get rid of big table, better PCA-AGHQ description. Need just to add some new results in now #57

@@
 * Naomi has a large latent field $x$ controlled by a smaller number of hyperparameters $\theta$
-* Extended latent Gaussian model [@stringer2022fast]: more complex dependency structures than a latent Gaussian model [@rue2009approximate]
+* Extended latent Gaussian model [@stringer2022fast]: more complex dependency structures than a latent Gaussian model
 * Looking for a fast, approximate approach, that properly takes uncertainty in hyperparameters into account
@@
 
+```{r aghq, echo=FALSE, fig.align='center', fig.cap='Demonstration of Gauss-Hermite quadrature, adaption, and our principal components approach.'}
+knitr::include_graphics("fig2.png")
+```
+
 * **Laplace approximation** Integrate out latent field using a Gaussian approximation to the denominator
@@
   * Use automatic differentiation via `CppAD` in `TMB`
-
-```{r aghq, echo=FALSE, fig.align='center', fig.cap='Demonstration of PCA-AGHQ.'}
-knitr::include_graphics("fig2.png")
-```
-
 * **Adaptive Gauss-Hermite Quadrature** (AGHQ) perform quadrature over the hyperparameters
@@
 $$
-where the Gauss-Hermite quadrature rule $\{z \in \mathcal{Q}(m, k), \omega\}$ with $m = \dim(\theta)$ and $k$ points per dimension is adapted based upon
+where $m = \dim(\theta)$ and the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ with weights $\omega:\mathcal{Q} \mapsto \mathbb{R}$ and $k$ points per dimension is adapted based upon
   * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p_\texttt{LA}(\theta, y)$
   * A matrix decomposition $LL^\top = - \partial_\theta^2 \log p_\texttt{LA}(\theta, y) \rvert_{\theta = \hat \theta}$
-* Use the spectral decomposition $L = E\Lambda^{1/2}$ and keep only the first $s < m$ **principal components** (PCA-AGHQ)
+* **PCA-AGHQ** To integrate over large spaces, use the spectral decomposition $L = E\Lambda^{1/2}$ and keep only the first $s < m$ principal components
 
@@
 
-| Method | Description | Time |
-|-----------|----------|------|
-| TMB | Baseline | `r round(time_taken$TMB, 0)` secs |
-| PCA-AGHQ | Ours | `r round(time_taken$aghq, 0)` hour |
-| NUTS | Gold-standard | `r round(time_taken$tmbstan, 1)` days |
-
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
@@
 ks_reduction <- ks %>%
-  filter(type == "Latent field") %>%
+  filter(type == "Latent") %>%
   summarise(
-    ks_tmb = mean(`KS(TMB, tmbstan)`),
-    ks_aghq = mean(`KS(aghq, tmbstan)`)
+    ks_tmb = mean(TMB),
+    ks_aghq = mean(aghq)
   ) %>%
@@
 ```
-
+* Run three methods: 1. TMB (baseline, `r round(time_taken$TMB, 0)` secs), 2. PCA-AGHQ (new, `r round(time_taken$aghq, 0)` hour), 3. NUTS (gold-standard, `r round(time_taken$tmbstan, 1)` days)
   * For PCA-AGHQ $k = 3$ and $s = 8$ chosen using Scree plot to explain ~90% of variance
-* For NUTS 4 chains of 100,000 thinned by 40 were required for good diagnostics
-* Kolmogorov-Smirnov (KS) test based on the maximum difference between marginal ECDFs
-  * Average KS distance from NUTS reduced by `r abs(ks_reduction)`%
-  * Also considering joint posteriors via Pareto-smoothed importance sampling and maximum mean discrepancy
+  * For NUTS 4 chains of 100,000 thinned by 40 were required for good diagnostics e.g. all $\hat R < 1.025$
 * Naomi can be used to assess probabilities targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90"). Both TMB and PCA-AGHQ have biased inferences
-
-<!-- ```{r fig4, echo=FALSE, fig.align='center', fig.cap='Both approximate methods are meaningfully incorrect for policy.'} -->
-<!-- knitr::include_graphics("fig4.png") -->
-<!-- ``` -->
 

### 2023-07-22 src/docs_statml-poster/statml-poster.Rmd
Add custom plot of second 90 exceedance to poster #57

@@
 
-```{r aghq, echo=FALSE, fig.align='center', fig.cap='Demonstration of Gauss-Hermite quadrature, adaption, and our principal components approach.'}
+```{r fig2, echo=FALSE, fig.align='center', fig.cap='Demonstration of Gauss-Hermite quadrature, adaption, and our principal components approach.'}
 knitr::include_graphics("fig2.png")
@@
 $$
-where $m = \dim(\theta)$ and the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ with weights $\omega:\mathcal{Q} \mapsto \mathbb{R}$ and $k$ points per dimension is adapted based upon
+where $m = \dim(\theta)$ and the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ with weights $\omega:\mathcal{Q} \mapsto \mathbb{R}$ and $k$ points per dimension is adapted (Figure 2) based upon
   * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p_\texttt{LA}(\theta, y)$
@@
 
-```{r fig3, echo=FALSE, fig.align='center', fig.cap='District-level model outputs for adults 15-49 in January 2016. Adapted from Eaton et al. 2021.'}
+```{r fig3, echo=FALSE, fig.align='center', fig.cap='District-level model outputs for adults 15-49.'}
 knitr::include_graphics("fig3.png")
@@
   round()
+
+exceedance_summary <- read_csv("depends/exceedance-summary.csv")
+second90_rmse <- filter(exceedance_summary, indicator == "Second 90", method == "PCA-AGHQ")$rmse_diff
 ```
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
-# second90 <- readr::read_csv("depends/second90.csv")
-# 
-# second90_reduction <- second90 %>%
-#   summarise(
-#     rmse_tmb = 100 * sqrt(mean((tmbstan - TMB)^2)),
-#     rmse_aghq = 100 * sqrt(mean((tmbstan - aghq)^2))
-#   ) %>%
-#   mutate(reduction = 100 * (rmse_aghq - rmse_tmb) / rmse_tmb) %>%
-#   pull(reduction) %>%
-#   round()
-```
 * Run three methods: 1. TMB (baseline, `r round(time_taken$TMB, 0)` secs), 2. PCA-AGHQ (new, `r round(time_taken$aghq, 0)` hour), 3. NUTS (gold-standard, `r round(time_taken$tmbstan, 1)` days)
@@
   * For NUTS 4 chains of 100,000 thinned by 40 were required for good diagnostics e.g. all $\hat R < 1.025$
-* Naomi can be used to assess probabilities targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90"). Both TMB and PCA-AGHQ have biased inferences
+* Naomi can be used to assess probabilities targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90")
+  * Though PCA-AGHQ is better (`r round(second90_rmse, 0)`% RMSE), both TMB and PCA-AGHQ are biased (Figure 5)
+
+```{r fig5, echo=FALSE, fig.align='center', fig.cap='PCA-AGHQ somewhat improves the second 90 bias.'}
+knitr::include_graphics("fig5.png")
+```
 
@@
 <font size="5"> <div id="refs" class="references"></div> </font>
-
-<!-- To-do: -->
-<!-- * Get "3. Inference procedure" to be on next page -->
-<!-- * Try to reduce whitespace around Fig 3 -->

### 2023-07-22 src/docs_statml-poster/statml-poster.Rmd
Complete StatML poster #57

@@
 
-* Malawi is a relatively small country but still has latent field $\text{dim}(x) = 491$ and hyperparameters $\text{dim}(\theta) = 24$
+* Malawi is a relatively small country but still has latent field $\text{dim}(x) = 467$ and hyperparameters $\text{dim}(\theta) = 24$
 
@@
 
-ks_reduction <- ks %>%
+ks_summary <- ks %>%
   filter(type == "Latent") %>%
+  group_by(type) %>%
   summarise(
-    ks_tmb = mean(TMB),
-    ks_aghq = mean(aghq)
+    ks_tmb = mean(size * TMB) / sum(size),
+    ks_aghq = mean(size * aghq) / sum(size)
   ) %>%
@@
   * For NUTS 4 chains of 100,000 thinned by 40 were required for good diagnostics e.g. all $\hat R < 1.025$
+
+```{r fig4, echo=FALSE, fig.align='center', fig.cap='PCA-AGHQ moderately improves the posterior mean and substantially improves the posterior standard deviation'}
+knitr::include_graphics("depends/mean-sd-alt-latent.png")
+```
+
+* PCA-AGHQ improves latent field point estimates (Figure 4) and distributional quantities like the Kolmogorov-Smirnov (KS, Figure 5) test statistic (`r ks_summary`% on average)
+
+```{r fig5, echo=FALSE, fig.align='center', fig.cap='KS test demonstration.'}
+knitr::include_graphics("fig5.png")
+```
+
 * Naomi can be used to assess probabilities targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90")
-  * Though PCA-AGHQ is better (`r round(second90_rmse, 0)`% RMSE), both TMB and PCA-AGHQ are biased (Figure 5)
+  * Though PCA-AGHQ is better (`r round(second90_rmse, 0)`% RMSE), both TMB and PCA-AGHQ are biased (Figure 6)
 
-```{r fig5, echo=FALSE, fig.align='center', fig.cap='PCA-AGHQ somewhat improves the second 90 bias.'}
-knitr::include_graphics("fig5.png")
+```{r fig6, echo=FALSE, fig.align='center', fig.cap='PCA-AGHQ moderately improves the second 90 bias.'}
+knitr::include_graphics("fig6.png")
 ```
@@
 
-* **Can we do any better than modest improvements?**
+* Can we do better than moderate improvements?
+  * Especially for the quantities of policy interest
 * Laplace marginals with matrix algebra approximations [@wood2020simplified] to speed up calculations
@@
 
+**Contact**
+
+* `r fontawesome::fa("code", fill = "black")` `github.com/athowes/naomi-aghq`
+* `r fontawesome::fa("home", fill = "black")` `athowes.github.io`
+* `r fontawesome::fa("envelope", fill = "black")` `ath19@ic.ac.uk`
+
 **Funding** AH was supported by the EPSRC and Bill & Melinda Gates Foundation. This research was supported by the MRC Centre for Global Infectious Disease Analysis.
-
-**Contact** `r fontawesome::fa("home", fill = "black")` `athowes.github.io`; `r fontawesome::fa("envelope", fill = "black")` `ath19@ic.ac.uk`
 

### 2023-07-22 utils/archive_to_docs.R

@@ -37,6 +37,7 @@ archive_to_docs("example_naomi")
 archive_to_docs("explore_posterior-comparison")
 archive_to_docs("docs_bayescomp-poster")
 archive_to_docs("docs_bioinference-poster")
+archive_to_docs("docs_statml-poster")
 archive_to_docs("plot-tikz_algorithm-flowchart")
 archive_to_docs("naomi-simple_mcmc")
 archive_to_docs("naomi-simple_ks")

### 2023-07-24 src/docs_paper/appendix.Rmd
Update paper and appendix #55

@@
   ungroup() %>%
-  filter(type == "Latent field") %>%
+  filter(type == "Latent") %>%
   select(-type) %>%
-  select(Parameter, TMB = `KS(TMB, tmbstan)`, `PCA-AGHQ` = `KS(aghq, tmbstan)`) %>%
+  select(Parameter = parname, TMB, `PCA-AGHQ` = aghq) %>%
   mutate(Difference = `PCA-AGHQ` - TMB) %>%

### 2023-07-27 src/docs_paper/paper.Rmd
Add plot with ESS against KS result for TMB and PCA-AGHQ #55
Chen citation for the xBinomial weighting approach, improve introduction section a little #55

@@
 
-To allow for interactive review and iteration of model results by workshop participants, the inference procedure should be fast and have low memory usage.
-Due to the scale of the model and features of its posterior geometry [@neal2003slice], Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
-Furthermore, use of the inference method across countries should be effortless, without requiring substantial statistical expertise, as would be the case for monitoring MCMC convergence and suitability.
+To allow interactive review and iteration of model results by workshop participants, the inference procedure should be fast and have low memory usage.
+Because of the scale of the model and challenging features of its posterior geometry [@neal2003slice], Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
+In addition, we require the inference procedure to be automatic and produce estimates reliably, across a range of country settings, with specific data not known in advance.
+As such, approaches requiring substantial statistical expertise, as would be the case for monitoring MCMC convergence and suitability, are not viable.
 
@@
 Although the TMB approach is fast, within the empirical Bayes framework hyperparameter uncertainty is not accounted for in the latent field posterior.
-We suspected this would result in underestimation of posterior variances, which may have implications 
-As such, we were motivated to look for an more fully Bayesian inference approach, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
-We developed an inference method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over large numbers of hyperparameters.
-AGHQ is a quadrature method based on the theory of polynomial interpolation, which is well suited to statistical estimation problems.
+We expected this would result in underestimation of posterior variances, which may have practical implications for use of the resulting estimates.
+As such, we were motivated to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
+To meet these requirements, we developed an inference method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over large numbers of hyperparameters.
+AGHQ is a quadrature method based on the theory of polynomial interpolation, and is well suited to statistical estimation problems in which the integrand is well approximated by a Gaussian multiplied by a polynomial.
 @bilodeau2022stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
-However, it is not computationally feasible to use AGHQ in high dimensions directly, as exponentially many nodes would be required.
-Instead, we use principal components analysis (PCA) of the inverse curvature at the mode to find a smaller number of dimensions which explain most of the variance.
-For the Naomi model in Malawi, this results in a grid which has millions of times fewer nodes the corresponding dense grid and is tractable to run.
+However, it is not computationally feasible to use AGHQ in >20 dimensions directly, as exponentially many nodes are required.
+Instead, we used principal components analysis (PCA) of the inverse curvature at the mode to find a subspace which explained most of the hyperparameter variance.
+For the Naomi model in Malawi, this resulted in a grid which was tractable as it had millions of times fewer nodes the corresponding dense grid.
 Our implementation of the method makes use of the existing Naomi `TMB` template, and is immediately compatible with any model with such a template.
@@
 As the Kish ESS is maximised by constant design weights, in exchange for reducing bias we reduce our ESS and thereby increase variance.
-We use a binomial working likelihood defined to operate on the reals
+We use a binomial working likelihood [@chen2014use] defined to operate on the reals
 \begin{equation*}
@@
 The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\omega) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\psi_i \leq \omega}$.
-See an illustration of the KS test in Figure \ref{fig:ks} and a summary of the results in Table.
+See Figure \ref{fig:ks} for an example.
+Additionally, we found that KS test results were negatively correlated with ESS (Figure \ref{fig:ks-ess}).
 

### 2023-07-31 src/docs_paper/paper.Rmd
Update paper #55

@@
 Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
-We propose a new inference method extending adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters.
+We propose a new inference method based on an extension of adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters.
 Using data from Malawi, our method improves the accuracy of inferences for model parameters, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
@@
 To address these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
-Modelling multiple data sources jointly in this way mitigates the limitations of any single source, increases statistical power, and can prompt investigation into possible conflicts of information.
+Modelling multiple data sources jointly mitigates the limitations of any single source, increases statistical power, and can prompt investigation into conflicts of information.
 
-Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing over 35 countries to input their data and interactively generate estimates during workshops as a part of a yearly process supported by UNAIDS.
+Software has been developed for Naomi (\url{https://naomi.unaids.org}), allowing over 35 countries to input their data and interactively generate estimates during workshops as a part of a yearly process supported by UNAIDS.
 Creation of estimates by country teams, rather than external agencies or researchers, is an important and distinctive feature of the HIV response.
@@
 
-Naomi is a complex model, and as such presents a challenging Bayesian inference problem.
-As well as hundreds of latent field parameters, Naomi has >20 hyperparameters: substantially more than the small number that can typically be handled by approaches like integrated nested Laplace approximations [INLA; @rue2009approximate].
+Naomi is a complex model, comprised of multiple linked generalized linear mixed models (GLMMs), and presents a challenging Bayesian inference problem.
+As well as hundreds of fixed and random effect parameters, Naomi has >20 hyperparameters: substantially more than the small number that can typically be handled by approaches like integrated nested Laplace approximations [INLA; @rue2009approximate].
 Moreover, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
@@
 Because of the scale of the model and challenging features of its posterior geometry [@neal2003slice], Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
-In addition, we require the inference procedure to be automatic and produce estimates reliably, across a range of country settings, with specific data not known in advance.
-As such, approaches requiring substantial statistical expertise, as would be the case for monitoring MCMC convergence and suitability, are not viable.
+In addition, we require the inference procedure to produce estimates reliably and automatically, across a range of country settings, without knowledge of the specific data used in advance.
+As such, approaches requiring substantial statistical expertise or tuning, as would be the case for monitoring MCMC convergence and suitability, are not viable.
 
-To meet these requirements, inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb], which we refer to as TMB (distinguished from the package `TMB`).
+To meet these requirements, inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
+We refer to this approach as TMB, distinguished from the package `TMB`.
 Owing to its speed and flexibility, `TMB` is gaining popularity, particularly in spatial statistics [@osgood2022statistical] and via the user-friendly `glmmTMB` \textsc{R} package [@brooks2017glmmTMB].
@@
 For the Naomi model, this subset is the high-dimensional latent field, leaving a smaller number of hyperparameters.
-Taking inspiration from the AD Model Builder package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
+Taking inspiration from the AD Model Builder package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for gradient-based numerical optimisation routines and the Laplace approximation.
 
 Although the TMB approach is fast, within the empirical Bayes framework hyperparameter uncertainty is not accounted for in the latent field posterior.
-We expected this would result in underestimation of posterior variances, which may have practical implications for use of the resulting estimates.
+We had concerns that this would result in underestimation of posterior variances, and may have important practical implications.
 As such, we were motivated to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
-To meet these requirements, we developed an inference method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over large numbers of hyperparameters.
+To meet these requirements, we developed an inference method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over many hyperparameters.
 AGHQ is a quadrature method based on the theory of polynomial interpolation, and is well suited to statistical estimation problems in which the integrand is well approximated by a Gaussian multiplied by a polynomial.
@@
 
-Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package [@bachl2019inlabru], INLA within MCMC [@gomez2018markov], and importance sampling with INLA [@berild2022importance], all of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
-The approach of `inlabru` is to approximate non-linear predictors using linearisation, by making iterative calls to `R-INLA`.
 <!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
@@
 <!-- Don't know the answer to this yet, but asking Janine Illian I'm told you don't have to call R-INLA that many times -->
-<!-- I think it'd obviously depend on the exact details of the problem at hand -->
+Other work aiming to extend the scope of INLA-like methods includes the `inlabru` \textsc{R} package [@bachl2019inlabru], INLA within MCMC [@gomez2018markov], and importance sampling with INLA [@berild2022importance], all of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
+The approach of `inlabru` is to approximate non-linear predictors using linearisation, by making iterative calls to `R-INLA`.
 INLA within MCMC and importance sampling with INLA are suitable for models which are LGMs conditional on some subset of the parameters being fixed.
-<!-- It would or would not be possible to fit the Naomi model in these frameworks. -->
+It's not clear the extent to which any of these methods would be compatible with Naomi.
 
@@
 \end{equation*}
-Though our approach accounts for survey weights, it does not take into account sample correlation structure, for example due to cluster sampling [@wakefield2020small].
+Though our approach accounts for effect of survey weighting on the variance, it does not take into account correlation structure in the samples, for example due to cluster sampling [@wakefield2020small].
 
@@
 \end{align}
-Then the Laplace approximation to $p(\btheta, \y)$ is given by
+Then the Laplace approximation [@tierney1986accurate] to $p(\btheta, \y)$ is given by
 \begin{equation}
@@
 The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$.
-For $k = 1$ GHQ corresponds to a Laplace approximation.
-Further GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$, a class of functions we expect the posterior to be close to.
+GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.
 
@@
 \end{equation}
+When $k = 1$ Equation \ref{eq:aghq} corresponds to a Laplace approximation.
 The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
@@
 
-We developed an approximate Bayesian inference algorithm, combining AGHQ with PCA, motivated by a challenging problem in small-area estimation of HIV.
-For the simplified Naomi model in Malawi (Section \ref{sec:results}) we demonstrated the method to be more accurate for model parameters, across a broad range of metrics, than TMB, and substantially faster than NUTS.
-However, this improvement in accuracy for model parameters did not translate into model outputs.
-Indeed, we found posterior exceedance probabilities (Section \ref{sec:exceedance}) from both TMB and PCA-AGHQ to have systematically inaccurates, with the potential to meaningfully mislead policy.
+We developed an approximate Bayesian inference algorithm, combining AGHQ with PCA, motivated by a challenging problem in small-area estimation of HIV indicators.
+For the simplified Naomi model in Malawi (Section \ref{sec:results}) we demonstrated the method to be more accurate at inferring posterior distributions of model parameters, across a broad range of metrics, than TMB, and substantially faster than NUTS.
+However, improvements in accuracy for model parameters did not translate into model outputs.
+Indeed, we found posterior exceedance probabilities (Section \ref{sec:exceedance}) from both TMB and PCA-AGHQ to have systematically inaccuracies, with the potential to meaningfully mislead policy.
+If possible, though not a desirable situation, it could be advisable to provide gold-standard NUTS results after the workshop has concluded.
+However, running NUTS for Naomi took days, in countries with 100s of districts it may simply not be feasible.
 
@@
 We selected $s$ based on the Scree plot, and for the most part fixed $k = 3$.
-Whether it is preferable, for a given computational budget, to increase $s$ or increase $k$ remains an open question.
+Whether it is preferable, for a given computational budget, to increase $s$ or increase $k$ is an open question.
 Further strategies, such as gradually lowering $k$ over the principal components, could also be considered.
-
-@bilodeau2022stochastic highlight developing computationally feasible quadrature methods for high dimensions as a challenging open problem.
-
-PCA-AGHQ was implemented using the `TMB` and `aghq` \textsc{R} packages.
 
@@
 
-## Future directions
+## Suggestions for future work
 
-### Improving the quadrature grid
+### Improved quadrature grids for moderate dimensions
 
@@
 @stringer2022fast (Theorem 1) bound the total variation error of AGHQ, establishing convergence in probability of coverage probabilities under the approximate posterior to those under the true posterior.
-It's possible that similar theory could be established for PCA-AGHQ, or more generally AGHQ with varying numbers of nodes per dimension.
-
-### Laplace marginals
-
-See Appendix \ref{sec:laplace-marginals}.
+Similar theory might be established for PCA-AGHQ, or more generally AGHQ with varying numbers of nodes per dimension.
+The challenge of connecting this theory to use of the quadrature rule within nested compuations remains an open question.
 

### 2023-12-04 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Update NUTS diagnostics for thesis

@@
 rhats <- bayesplot::rhat(mcmc)
+rhats <- bayesplot:::drop_NAs_and_warn(bayesplot:::new_rhat(rhats))
+rhats_df <- bayesplot:::diagnostic_data_frame(rhats)
 
-bayesplot::mcmc_rhat_data(rhats) %>%
-  ggplot(aes(x = value, y = parameter, col = description)) +
-  geom_segment(aes(yend = parameter, xend = ifelse(min(value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7) +
-  scale_color_manual(values = "#E69F00") +
-  geom_vline(xintercept = 1.05, linetype = "dashed", col = "grey40") +
-  labs(x = "Potential scale reduction factor", y = "NUTS parameter", col = "") +
+ggplot(rhats_df, aes(x = value, y = parameter)) +
+  geom_segment(aes(yend = parameter, xend = ifelse(min(value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7, col = "#56B4E9") +
+  lims(x = c(NA, 1.03)) +
+  labs(x = "Potential scale reduction factor", y = "Parameter", col = "") +
   theme_minimal() +
@@
     axis.ticks.y = element_blank(),
-    panel.grid.major = element_blank()
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
   )
@@
 
-(max_rhat <- max(rhats))
+#' I think that this $summary should be all of the chains grouped together
+mcmc_summary <- summary(mcmc)$summary
+
+(rhat_min <- min(mcmc_summary[, "Rhat"]))
+(rhat_lower <- quantile(mcmc_summary[, "Rhat"], 0.025))
+(rhat_median <- quantile(mcmc_summary[, "Rhat"], 0.50))
+(rhat_upper <- quantile(mcmc_summary[, "Rhat"], 0.975))
+(rhat_max <- max(mcmc_summary[, "Rhat"]))
+(rhat_mean <- mean(mcmc_summary[, "Rhat"]))
 ```
 
-# ESS ratio
+# ESS
 
@@
 
-bayesplot::mcmc_neff_data(ratios) %>%
-  ggplot(mapping = aes(x = value, y = parameter, color = description)) +
+ess_fig <- bayesplot::mcmc_neff_data(ratios) %>%
+  mutate(
+    description_new = case_when(
+      value < 0.1 ~ "< 0.1",
+      value < 0.5 ~ "< 0.5",
+      value > 0.5 ~ "> 0.5"
+    )
+  ) %>%
+  ggplot(mapping = aes(x = value, y = parameter, color = description_new)) +
   geom_segment(aes(yend = parameter, xend = -Inf), na.rm = TRUE, alpha = 0.7) +
-  scale_color_manual(values = c("#56B4E9", "#009E73", "#E69F00")) +
-  geom_vline(xintercept = 0.1, linetype = "dashed", col = "grey40") +
-  geom_vline(xintercept = 0.5, linetype = "dashed", col = "grey40") +
-  geom_vline(xintercept = 1, linetype = "dashed", col = "grey40") +
-  labs(x = "ESS ratio", y = "NUTS parameter", col = "") +
+  scale_color_manual(values = c("#E69F00", "#009E73", "#56B4E9")) +
+  geom_vline(xintercept = 0.1, linetype = "dashed", col = "grey40", alpha = 0.5) +
+  geom_vline(xintercept = 0.5, linetype = "dashed", col = "grey40", alpha = 0.5) +
+  geom_vline(xintercept = 1, linetype = "dashed", col = "grey40", alpha = 0.5) +
+  labs(x = "ESS ratio", y = "Parameter", col = "", tag = "A") +
   theme_minimal() +
@@
     axis.ticks.y = element_blank(),
-    panel.grid.major = element_blank()
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank()
   )
 
-ggsave("ratio.png", h = 3, w = 6.25)
-
-(average_ess_ratio <- mean(ratios))
 ```
-
-# ESS
 
@@
 ```{r}
-#' I think that this $summary should be all of the chains grouped together
-mcmc_summary <- summary(mcmc)$summary
-
-data.frame(mcmc_summary) %>%
+ess_histogram <- data.frame(mcmc_summary) %>%
   tibble::rownames_to_column("param") %>%
@@
     geom_histogram(alpha = 0.8) +
-    labs(x = "ESS", y = "Count")
+    labs(x = "ESS", y = "Count", tag = "B")
 
-ggsave("ess.png", h = 3, w = 6.25)
+ess_fig / ess_histogram
+
+ggsave("ess.png", h = 5, w = 6.25)
 
@@
 (ess_max <- max(mcmc_summary[, "n_eff"]))
+(ess_mean <- mean(mcmc_summary[, "n_eff"]))
 ```
@@
 out <- list(
-  "max_rhat" = max_rhat,
-  "average_ess_ratio" = average_ess_ratio, 
+  "rhat_min" = rhat_min,
+  "rhat_lower" = rhat_lower,
+  "rhat_median" = rhat_median,
+  "rhat_upper" = rhat_upper,
+  "rhat_max" = rhat_max,
+  "rhat_mean" = rhat_mean,
   "ess_min" = ess_min,
@@
   "ess_upper" = ess_upper,
-  "ess_max" = ess_max
+  "ess_max" = ess_max,
+  "ess_mean" = ess_mean
 )
@@
 ```{r}
+cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#E69F00", "#E69F00")
+bayesplot::color_scheme_set(rev(cbpalette))
+
 bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("beta")))
@@
 
-ggsave("worst-trace.png", plot, h = 4, w = 6.25)
+ggsave("worst-trace.png", plot, h = 3, w = 6.25)
 ```
@@
 bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_as", "logit_phi_rho_as"), diag_fun = "hist", off_diag_fun = "hex")
-(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex"))
 
-ggsave("rho_a.png", plot, h = 4, w = 6.25)
+pairs_df <- extract(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a")) %>%
+  as.data.frame()
+
+fig_top_left <- ggplot(pairs_df, aes(x = log_sigma_rho_a)) +
+  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
+  labs(x = "", y = "", tag = "A") + 
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+fig_top_right <- ggplot(pairs_df, aes(x = logit_phi_rho_a, y = log_sigma_rho_a)) +
+  geom_point(shape = 1, alpha = 0.5, col = "#009E73") +
+  labs(x = "", y = "", tag = "B") + 
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+fig_bot_left <- ggplot(pairs_df, aes(x = log_sigma_rho_a, y = logit_phi_rho_a)) +
+  geom_point(shape = 1, alpha = 0.5, col = "#009E73") +
+  labs(x = "log_sigma_rho_a", y = "", tag = "C") + 
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+fig_bot_right <- ggplot(pairs_df, aes(x = logit_phi_rho_a)) +
+  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
+  labs(x = "logit_phi_rho_a", y = "", tag = "D") + 
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+(fig_top_left + fig_top_right) / (fig_bot_left + fig_bot_right)
+
+ggsave("rho_a.png", h = 4, w = 6.25)
 
@@
 bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_xs", "logit_phi_alpha_xs"), diag_fun = "hist", off_diag_fun = "hex")
-(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x"), diag_fun = "hist", off_diag_fun = "hex"))
 
-ggsave("alpha_x.png", plot, h = 4, w = 6.25)
+pairs_df <- extract(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x")) %>%
+  as.data.frame()
+
+fig_top_left <- ggplot(pairs_df, aes(x = log_sigma_alpha_x)) +
+  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
+  labs(x = "", y = "", tag = "A") + 
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+fig_top_right <- ggplot(pairs_df, aes(x = logit_phi_alpha_x, y = log_sigma_alpha_x)) +
+  geom_point(shape = 1, alpha = 0.5, col = "#009E73") +
+  labs(x = "", y = "", tag = "B") + 
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+fig_bot_left <- ggplot(pairs_df, aes(x = log_sigma_alpha_x, y = logit_phi_alpha_x)) +
+  geom_point(shape = 1, alpha = 0.5, col = "#009E73") +
+  labs(x = "log_sigma_alpha_x", y = "", tag = "C") + 
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+fig_bot_right <- ggplot(pairs_df, aes(x = logit_phi_alpha_x)) +
+  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
+  labs(x = "logit_phi_alpha_x", y = "", tag = "D") + 
+  theme_minimal() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+
+(fig_top_left + fig_top_right) / (fig_bot_left + fig_bot_right)
+
+ggsave("alpha_x.png", h = 4, w = 6.25)
 

### 2023-12-05 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Final improvements to NUTS figures

@@
 (rhat_mean <- mean(mcmc_summary[, "Rhat"]))
+(rhat_above_threshold <- mean(mcmc_summary[, "Rhat"] < 1.01))
 ```
@@
   geom_vline(xintercept = 1, linetype = "dashed", col = "grey40", alpha = 0.5) +
-  labs(x = "ESS ratio", y = "Parameter", col = "", tag = "A") +
+  labs(x = "Effective sample size ratio", y = "Parameter", col = "", tag = "A") +
   theme_minimal() +
@@
   ggplot(aes(x = n_eff)) +
-    geom_histogram(alpha = 0.8) +
-    labs(x = "ESS", y = "Count", tag = "B")
+    geom_histogram(col = "grey60", fill = "grey80") +
+    labs(x = "Effective sample size", y = "Count", tag = "B")
 
@@
   "rhat_mean" = rhat_mean,
+  "rhat_above_threshold" = rhat_above_threshold,
   "ess_min" = ess_min,
@@
 ```{r}
-(plot <- bayesplot::mcmc_trace(mcmc, pars = c(names(which.min(mcmc_summary[, "n_eff"])), names(which.max(rhats)))))
+worst_eff <- bayesplot::mcmc_trace(mcmc, pars = names(which.min(mcmc_summary[, "n_eff"]))) + labs(tag = "A") + guides(col = "none")
+worst_rhat <- bayesplot::mcmc_trace(mcmc, pars = names(which.max(rhats))) + labs(tag = "B")
 
-ggsave("worst-trace.png", plot, h = 3, w = 6.25)
+worst_eff + worst_rhat
+
+ggsave("worst-trace.png", h = 3, w = 6.25)
 ```

### 2023-12-05 src/dev_scale-grid/scale-grid.Rmd
Improvements to PCA-AGHQ plots for thesis

@@
 
-ggsave("marginal-sd.png", height = 3.5, w = 6.25)
+ggsave("marginal-sd.png", height = 4, w = 6.25)
 ```
@@
 C <- sd_out$cov.fixed
-(C_plot <- plot_matrix(C))
+(C_plot <- plot_matrix(unname(C), legend = FALSE))
 ```
@@
 
-ggsave("pc-loadings.png", h = 3.5, w = 6.25)
+ggsave("pc-loadings.png", h = 4.5, w = 6.25)
 ```
@@
 
-C_s_plot <- plot_matrix(C_s)
-{C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + theme(legend.position = "none") + labs(subtitle = "Reduced rank")}
+C_s_plot <- plot_matrix(C_s, legend = FALSE)
+
+{C_plot + labs(tag = "A", caption = "Full rank")} + {C_s_plot + labs(tag = "B", caption = "Reduced rank")}
 

### 2023-12-05 src/dev_scale-grid/functions.R

@@ -139,14 +139,14 @@ create_pca_grid <- function(m, C, s, k) {
   return(gg)
 }
 
-plot_matrix <- function(M) {
-  name <- deparse(substitute(M))
+plot_matrix <- function(M, legend = TRUE) {
   M_df <- reshape2::melt(as.matrix(M))
 
-  ggplot(M_df, aes(x = Var1, y = Var2, fill = value)) +
-    labs(x = "", y = "", fill = paste0(name, "[i, j]")) +
+  fig <- ggplot(M_df, aes(x = Var1, y = Var2, fill = value)) +
+    labs(x = "", y = "") +
     geom_tile() +
     scale_fill_viridis_c() +
+    coord_fixed() +
     labs(x = "i", y = "j") +
     theme_minimal() +
     theme(
@@ -155,30 +155,37 @@ plot_matrix <- function(M) {
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank()
     )
+
+  if(legend) {
+    name <- deparse(substitute(M))
+    fig <- fig + labs(fill = paste0(name, "[i, j]"))
+  } else {
+    fig <- fig + theme(legend.position = "none")
+  }
 }
 
 plot_total_variation <- function(eigen, label_x) {
   ggplot(data = NULL, aes(x = 1:length(eigen$values), y = cumsum(eigen$values) / sum(eigen$values))) +
     geom_point() +
     geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
-    annotate("text", x = label_x, y = 0.875, label = "90% of total variation explained", col = "grey") +
+    annotate("text", x = label_x, y = 0.85, label = "90% of total variation explained", col = "grey") +
     scale_y_continuous(labels = scales::percent) +
-    labs(x = "PCA dimensions included", y = "Total variation explained") +
+    labs(x = "PCA dimensions included", y = "Proportion of total variation explained") +
     theme_minimal()
 }
 
 plot_pc_loadings <- function(eigen) {
   reshape2::melt(as.matrix(eigen$vectors)) %>%
-    ggplot(aes(x = Var1, y = factor(Var2), fill = value)) +
+    ggplot(aes(x = factor(Var2), y = Var1, fill = value)) +
     labs(x = "", y = "", fill = "") +
     geom_tile() +
-    coord_flip() +
+    coord_fixed() +
     scale_fill_gradientn(
       colours = c("#E69F00", "white", "#009E73"),
[…]

### 2023-12-21 src/naomi-simple_point-estimates/point-estimates.Rmd
Update point estimate plots for thesis. Just need to add LEB into these

@@
   dplyr::bind_rows(.id = "par") %>%
-  mutate(method = "TMB")
+  mutate(method = "GEB")
 
@@
   dplyr::bind_rows(.id = "par") %>%
-  mutate(method = "PCA-AGHQ")
+  mutate(method = "GPCA-AGHQ")
 
@@
   pivot_wider(values_from = "estimate", names_from = "method") %>%
-  pivot_longer(cols = c("TMB", "PCA-AGHQ"), names_to = "method", values_to = "approximate") %>%
+  pivot_longer(cols = c("GEB", "GPCA-AGHQ"), names_to = "method", values_to = "approximate") %>%
   rename("truth" = "NUTS") %>%
   mutate(
-    indicator = fct_recode(indicator, "Posterior mean estimate" = "mean", "Posterior SD estimate" = "sd"),
-    method = fct_relevel(method, "TMB", "PCA-AGHQ")
+    indicator = fct_recode(indicator, "Posterior mean" = "mean", "Posterior SD" = "sd"),
+    method = fct_relevel(method, "GEB", "GPCA-AGHQ")
   )
@@
 
-Add difference in metric between TMB and PCA-AGHQ:
+Add difference in metric between GEB and GPCA-AGHQ:
 
@@
     label = ifelse(
-      method == "PCA-AGHQ",
+      method == "GPCA-AGHQ",
       paste0("RMSE: ", signif(rmse, 2), " (", signif(rmse_diff, 2), "%)", "\nMAE: ", signif(mae, 2), " (", signif(mae_diff, 2), "%)"),
@@
   facet_grid(indicator ~ method) +
-  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
+  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
   geom_text(data = df_metrics_latent, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
-  labs(x = "NUTS", y = "Approximation - NUTS") +
+  labs(x = "Estimate", y = "Absolute difference to NUTS") +
   theme_minimal()
@@
 mean_plot_latent <- df_plot_latent %>%
-  filter(indicator == "Posterior mean estimate") %>%
+  filter(indicator == "Posterior mean") %>%
   ggplot(aes(x = truth, y = approximate - truth)) +
@@
   facet_grid(indicator ~ method) +
-  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
   geom_text(
-    data = filter(df_metrics_latent, indicator == "Posterior mean estimate"),
+    data = filter(df_metrics_latent, indicator == "Posterior mean"),
     aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
   ) +
-  labs(x = "NUTS", y = "Approximation - NUTS") +
+  labs(x = "Estimate", y = "Absolute difference to NUTS") +
   theme_minimal()
@@
 sd_plot_latent <- df_plot_latent %>%
-  filter(indicator == "Posterior SD estimate") %>%
+  filter(indicator == "Posterior SD") %>%
   ggplot(aes(x = truth, y = approximate - truth)) +
@@
   facet_grid(indicator ~ method) +
-  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
   geom_text(
-    data = filter(df_metrics_latent, indicator == "Posterior SD estimate"),
+    data = filter(df_metrics_latent, indicator == "Posterior SD"),
     aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
   ) +
-  labs(x = "NUTS", y = "Approximation - NUTS") +
+  labs(x = "Estimate", y = "Absolute difference to NUTS") +
   theme_minimal()
@@
 df_plot_latent %>%
-  filter(indicator == "Posterior mean estimate") %>%
+  filter(indicator == "Posterior mean") %>%
   mutate(diff = abs(approximate - truth)) %>%
@@
 df_plot_latent %>%
-  filter(indicator == "Posterior SD estimate") %>%
+  filter(indicator == "Posterior SD") %>%
   mutate(diff = abs(approximate - truth)) %>%
@@
 df_metrics_output <- filter(df_metrics, type == "output")
-cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
 
@@
   geom_point(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4) +
-  scale_color_manual(values = cbpalette) +
+  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
   facet_grid(indicator ~ method) +
-  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
+  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
   geom_text(data = df_metrics_output, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
-  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
+  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
   theme_minimal() +
@@
 mean_plot_output <- df_plot_output %>%
-  filter(indicator == "Posterior mean estimate") %>%
+  filter(indicator == "Posterior mean") %>%
   ggplot() +
   geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
-  scale_color_manual(values = cbpalette) +
+  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
   facet_grid(indicator ~ method) +
-  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
   geom_text(
-    data = filter(df_metrics_output, indicator == "Posterior mean estimate"),
+    data = filter(df_metrics_output, indicator == "Posterior mean"),
     aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
   ) +
-  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
+  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
   theme_minimal()
@@
 sd_plot_output <- df_plot_output %>%
-  filter(indicator == "Posterior SD estimate") %>%
+  filter(indicator == "Posterior SD") %>%
   ggplot() +
   geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
-  scale_color_manual(values = cbpalette) +
+  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
   facet_grid(indicator ~ method) +
-  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
   geom_text(
-    data = filter(df_metrics_output, indicator == "Posterior SD estimate"),
+    data = filter(df_metrics_output, indicator == "Posterior SD"),
     aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
   ) +
-  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
+  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
   theme_minimal()
@@
   scale_color_manual(values = c(col, "white", "white")) +
-  facet_grid(indicator + par ~ method) +
-  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  facet_grid(indicator + par ~ method, labeller = labeller(indicator = function(x) {rep("", length(x))})) +
+  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 4)) +
   geom_text(
@@
   ) +
-  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
+  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
   theme_minimal() +
-  theme(legend.position = "none")
+  theme(legend.position = "none", panel.spacing.x = unit(6, "mm"))
 }
 
-mean_prev_plot <- plot_outputs(.indicator = "Posterior mean estimate", .par = "HIV prevalence", col = "#E69F00")
-mean_inc_plot <- plot_outputs(.indicator = "Posterior mean estimate", .par = "HIV incidence", col = "#009E73")
-mean_art_plot <- plot_outputs(.indicator = "Posterior mean estimate", .par = "ART coverage", col = "#56B4E9")
-sd_prev_plot <- plot_outputs(.indicator = "Posterior SD estimate", .par = "HIV prevalence", col = "#E69F00")
-sd_inc_plot <- plot_outputs(.indicator = "Posterior SD estimate", .par = "HIV incidence", col = "#009E73")
-sd_art_plot <- plot_outputs(.indicator = "Posterior SD estimate", .par = "ART coverage", col = "#56B4E9")
+mean_prev_plot <- plot_outputs(.indicator = "Posterior mean", .par = "HIV prevalence", col = "#CC79A7")
+mean_inc_plot <- plot_outputs(.indicator = "Posterior mean", .par = "HIV incidence", col = "#009E73")
+mean_art_plot <- plot_outputs(.indicator = "Posterior mean", .par = "ART coverage", col = "#56B4E9")
+sd_prev_plot <- plot_outputs(.indicator = "Posterior SD", .par = "HIV prevalence", col = "#CC79A7")
+sd_inc_plot <- plot_outputs(.indicator = "Posterior SD", .par = "HIV incidence", col = "#009E73")
+sd_art_plot <- plot_outputs(.indicator = "Posterior SD", .par = "ART coverage", col = "#56B4E9")
 
-#' Put together
-combine_three_plots <- function(plot1, plot2, plot3) {
-  plot1alt <- plot1 +
-    theme(legend.position = "none")
+combine_three_plots <- function(plot1, plot2, plot3, subtitle) {
+
+  plot1alt <- plot1 + theme(legend.position = "none") + labs(subtitle = subtitle)
 
@@
       strip.text.x = element_blank(),
+
       legend.position = "none"
@@
 
-  three_plots <- cowplot::plot_grid(plot1alt, plot2alt, plot3alt, ncol = 1)
+  three_plots <- cowplot::plot_grid(plot1alt, plot2alt, plot3alt, ncol = 1, rel_heights = c(1.25, 1, 1))
   cowplot::plot_grid(y_axis, three_plots, ncol = 2, rel_widths = c(1, 35))
@@
 
-combine_three_plots(mean_prev_plot, mean_inc_plot, mean_art_plot)
+combine_three_plots(mean_prev_plot, mean_inc_plot, mean_art_plot, subtitle = "Posterior mean")
 ggsave("mean-alt-output.png", h = 7, w = 6.25)
 
-combine_three_plots(sd_prev_plot, sd_inc_plot, sd_art_plot)
+combine_three_plots(sd_prev_plot, sd_inc_plot, sd_art_plot, subtitle = "Posterior SD")
 ggsave("sd-alt-output.png", h = 7, w = 6.25)

### 2023-12-28 src/naomi-simple_exceedance/exceedance.Rmd
Add custom plot of second 90 exceedance to poster #57
Update naming in exceedance notebook (GEB!)

@@
   prob = c(tmbstan_second90, tmb_second90, aghq_second90),
-  method = c(rep("tmbstan", times = n), rep("TMB", times = n), rep("aghq", times = n)),
+  method = c(rep("NUTS", times = n), rep("GEB", times = n), rep("GPCA-AGHQ", times = n)),
   index = c(1:n, 1:n, 1:n)
@@
 
-plot1 <- ggplot(df_wide_second90, aes(x = tmbstan, y = TMB)) +
-  geom_point(shape = 1, alpha = 0.6, col = "#999999") +
+plot1 <- ggplot(df_wide_second90, aes(x = NUTS, y = GEB)) +
+  geom_point(shape = 1, alpha = 0.6) +
   coord_fixed(ratio = 1) +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
-  labs(x = "NUTS", y = "TMB") +
+  labs(x = "NUTS", y = "GEB") +
   theme_minimal()
 
-plot2 <- ggplot(df_wide_second90, aes(x = tmbstan, y = aghq)) +
-  geom_point(shape = 1, alpha = 0.6, col = "#999999") +
+plot2 <- ggplot(df_wide_second90, aes(x = NUTS, y = `GPCA-AGHQ`)) +
+  geom_point(shape = 1, alpha = 0.6) +
   coord_fixed(ratio = 1) +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
-  labs(x = "NUTS", y = "PCA-AGHQ") +
+  labs(x = "NUTS", y = "GPCA-AGHQ") +
   theme_minimal()
@@
 ```
+
 There is a change of `r mse_change`% in the MSE moving from `TMB` to `aghq`, and a change of `r rmse_change`% in the RMSE.
@@
 tmb_fails_second90 <- df_wide_second90 %>%
-  mutate(diff = tmbstan - TMB) %>%
+  mutate(diff = NUTS - GEB) %>%
   filter(diff > 0.1) %>%
@@
 ```
+
 And for `aghq`:
@@
 aghq_fails_second90 <- df_wide_second90 %>%
-  mutate(diff = tmbstan - aghq) %>%
+  mutate(diff = NUTS - `GPCA-AGHQ`) %>%
   filter(diff > 0.1) %>%
@@
 ```
+
 Add colour by sex to the above plots:
@@
 
-plot1 <- ggplot(df_wide_second90, aes(x = tmbstan, y = TMB, col = sex)) +
+plot1 <- ggplot(df_wide_second90, aes(x = NUTS, y = GEB, col = sex)) +
   geom_point(shape = 1, alpha = 0.6) +
   coord_fixed(ratio = 1) +
-  scale_color_manual(values = c("#E69F00", "#999999")) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  scale_colour_manual(values = c("#0072B2", "black")) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "#E69F00") +
   labs(col = "Sex") +
   guides(col = "none") +
-  labs(x = "NUTS", y = "TMB") +
+  labs(x = "NUTS", y = "GEB") +
   theme_minimal()
 
-plot2 <- ggplot(df_wide_second90, aes(x = tmbstan, y = aghq, col = sex)) +
+plot2 <- ggplot(df_wide_second90, aes(x = NUTS, y = `GPCA-AGHQ`, col = sex)) +
   geom_point(shape = 1, alpha = 0.6) +
   coord_fixed(ratio = 1) +
-  scale_colour_manual(values = c("#E69F00", "#999999"), labels = c("Female", "Male")) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    scale_colour_manual(values = c("#0072B2", "black"), labels = c("Female", "Male")) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "#E69F00") +
   guides(colour = guide_legend(override.aes = list(alpha = 1))) +
-  labs(x = "NUTS", y = "PCA-AGHQ", col = "Sex") +
+  labs(x = "NUTS", y = "GPCA-AGHQ", col = "Sex") +
   theme_minimal()
 
-ggsave("second90.png", plot1 + plot2, h = 4, w = 6.25)
+plot1 + plot2
 
-plot1 + plot2
+ggsave("second90.png", h = 4, w = 6.25)
 ```
+
 Check the RMSE stratified by sex:
@@
   summarise(
-    "RMSE TMB (percent)" = 100 * sqrt(mean((tmbstan - TMB)^2)),
-    "RMSE aghq (percent)" = 100 * sqrt(mean((tmbstan - aghq)^2))
+    "RMSE GEB (percent)" = 100 * sqrt(mean((NUTS - GEB)^2)),
+    "RMSE GPCA-AGHQ (percent)" = 100 * sqrt(mean((NUTS - `GPCA-AGHQ`)^2))
   )
@@
   prob = c(tmbstan_1inc, tmb_1inc, aghq_1inc),
-  method = c(rep("tmbstan", times = n), rep("TMB", times = n), rep("aghq", times = n)),
+  method = c(rep("NUTS", times = n), rep("GEB", times = n), rep("GPCA-AGHQ", times = n)),
   index = c(1:n, 1:n, 1:n)
@@
 
-plot1 <- ggplot(df_wide_1inc, aes(x = tmbstan, y = TMB)) +
-  geom_point(shape = 1, alpha = 0.6, col = "#999999") +
+plot1 <- ggplot(df_wide_1inc, aes(x = NUTS, y = GEB)) +
+  geom_point(shape = 1, alpha = 0.6) +
   coord_fixed(ratio = 1) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
-  labs(x = "NUTS", y = "TMB") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "#E69F00") +
+  labs(x = "NUTS", y = "GEB") +
   theme_minimal()
 
-plot2 <- ggplot(df_wide_1inc, aes(x = tmbstan, y = aghq)) +
-  geom_point(shape = 1, alpha = 0.6, col = "#999999") +
+plot2 <- ggplot(df_wide_1inc, aes(x = NUTS, y = `GPCA-AGHQ`)) +
+  geom_point(shape = 1, alpha = 0.6) +
   coord_fixed(ratio = 1) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
-  labs(x = "NUTS", y = "PCA-AGHQ") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "#E69F00") +
+  labs(x = "NUTS", y = "GPCA-AGHQ") +
   theme_minimal()
 
-ggsave("1inc.png", plot1 + plot2, h = 4, w = 6.25)
+plot1 + plot2
 
-plot1 + plot2
+ggsave("1inc.png", h = 4, w = 6.25)
 
@@
 
+write_csv(df_wide, "exceedance.csv")
+
 fct_reorg <- function(fac, ...) {
@@
 df_metrics <- df_wide %>%
-  select(TMB, aghq, truth = tmbstan, indicator) %>%
+  select(GEB, `GPCA-AGHQ`, truth = NUTS, indicator) %>%
   pivot_longer(
-    cols = c("TMB", "aghq"),
+    cols = c("GEB", "GPCA-AGHQ"),
     names_to = "method",
@@
     mae = mean(abs(truth - approximate))
-  ) %>%
-  mutate(method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq"))
+  )
 
@@
   left_join(
-    mutate(df_metrics_pct, method = "PCA-AGHQ"),
+    mutate(df_metrics_pct, method = "GPCA-AGHQ"),
     by = c("indicator", "method")
@@
 
-write_csv(df_metrics, "exceedance.csv")
+write_csv(df_metrics, "exceedance-summary.csv")
 
 exceedance_manuscript_plot <- df_wide %>%
-  pivot_longer(cols = c("TMB", "aghq"), names_to = "method", values_to = "estimate") %>%
+  pivot_longer(cols = c("GEB", "GPCA-AGHQ"), names_to = "method", values_to = "estimate") %>%
   mutate(
-    method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq"),
     indicator = fct_relevel(indicator, "Second 90", "High incidence")
   ) %>%
-  ggplot(aes(x = tmbstan, y = estimate - tmbstan)) +
-    geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
-    scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
-    geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
+  ggplot(aes(x = NUTS, y = estimate - NUTS)) +
+    geom_point(aes(col = sex), shape = 1, alpha = 0.4) +
+    scale_colour_manual(values = c("#0072B2", "black"), labels = c("Female", "Male")) +
+    geom_abline(slope = 0, intercept = 0, linetype = "dashed", col = "#E69F00") +
     facet_grid(indicator ~ method) +
-    labs(x = "NUTS", y = "Approximate - NUTS", col = "Sex") +
+    labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Sex") +
     geom_text(
       data = df_metrics %>% mutate(
-        method = fct_relevel(method, "TMB", "PCA-AGHQ"),
         indicator = fct_relevel(indicator, "Second 90", "High incidence")    
@@
   filter(indicator == "Second 90") %>%
-  pivot_longer(cols = c("TMB", "aghq"), names_to = "method", values_to = "estimate") %>%
-  mutate(
-    method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq")
-  ) %>%
-  ggplot(aes(x = tmbstan, y = estimate)) +
-    geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
+  pivot_longer(cols = c("GEB", "GPCA-AGHQ"), names_to = "method", values_to = "estimate") %>%
+  ggplot(aes(x = NUTS, y = estimate)) +
+    geom_point(aes(col = sex), shape = 1, alpha = 0.4) +
     coord_fixed(ratio = 1) +
-    scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
-    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    scale_colour_manual(values = c("#0072B2", "black"), labels = c("Female", "Male")) +
+    geom_abline(slope = 0, intercept = 0, linetype = "dashed", col = "#E69F00") +
     geom_text(
     data = df_metrics %>%
-      filter(indicator == "Second 90") %>%
-      mutate(method = fct_relevel(method, "TMB", "PCA-AGHQ")),
+        filter(indicator == "Second 90"),
       aes(x = -Inf, y = Inf, label = label),
@@
     facet_grid(indicator ~ method) +
-    labs(x = "NUTS", y = "", col = "Sex") +
+    labs(x = "NUTS", y = "Absolute difference to NUTS", col = "Sex") +
     theme_minimal()
 
-ggsave("exceedance-second90.png", exceedance_manuscript_plot_second90, h = 4, w = 6.25)
+ggsave("exceedance-second90.png", exceedance_manuscript_plot_second90, h = 3, w = 6.25)
 
@@
   filter(indicator == "High incidence") %>%
-  pivot_longer(cols = c("TMB", "aghq"), names_to = "method", values_to = "estimate") %>%
-  mutate(
-    method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq")
-  ) %>%
-  ggplot(aes(x = tmbstan, y = estimate)) +
+  pivot_longer(cols = c("GEB", "GPCA-AGHQ"), names_to = "method", values_to = "estimate") %>%
+  ggplot(aes(x = NUTS, y = estimate)) +
     geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
     coord_fixed(ratio = 1) +
-    scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
-    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    scale_colour_manual(values = c("#0072B2", "black"), labels = c("Female", "Male")) +
+    geom_abline(slope = 0, intercept = 0, linetype = "dashed", col = "#E69F00") +
     geom_text(
       data = df_metrics %>%
-      filter(indicator == "High incidence") %>%
-      mutate(method = fct_relevel(method, "TMB", "PCA-AGHQ")),
+        filter(indicator == "High incidence"),
       aes(x = -Inf, y = Inf, label = label),
@@
     facet_grid(indicator ~ method) +
-    labs(x = "NUTS", y = "", col = "Sex") +
+    labs(x = "NUTS", y = "Absolute difference to NUTS", col = "Sex") +
     theme_minimal()
 
-ggsave("exceedance-1inc.png", exceedance_manuscript_plot_1inc, h = 4, w = 6.25)
+ggsave("exceedance-1inc.png", exceedance_manuscript_plot_1inc, h = 3, w = 6.25)
 ```

### 2023-12-28 src/naomi-simple_exceedance/exceedance.Rmd
Make the exceedance plots difference on the y-axis

@@
   pivot_longer(cols = c("GEB", "GPCA-AGHQ"), names_to = "method", values_to = "estimate") %>%
-  ggplot(aes(x = NUTS, y = estimate)) +
+  ggplot(aes(x = NUTS, y = estimate - NUTS)) +
     geom_point(aes(col = sex), shape = 1, alpha = 0.4) +
-    coord_fixed(ratio = 1) +
     scale_colour_manual(values = c("#0072B2", "black"), labels = c("Female", "Male")) +
@@
     facet_grid(indicator ~ method) +
-    labs(x = "NUTS", y = "Absolute difference to NUTS", col = "Sex") +
+    labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Sex") +
     theme_minimal()
@@
   pivot_longer(cols = c("GEB", "GPCA-AGHQ"), names_to = "method", values_to = "estimate") %>%
-  ggplot(aes(x = NUTS, y = estimate)) +
+  ggplot(aes(x = NUTS, y = estimate - NUTS)) +
     geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
-    coord_fixed(ratio = 1) +
     scale_colour_manual(values = c("#0072B2", "black"), labels = c("Female", "Male")) +

### 2023-12-28 src/naomi-simple_ks/ks.Rmd
Called "Latent" not "Latent field" now
Update KS plots for thesis

@@
 
-## `beta_rho`
+<!-- ## `beta_rho` -->
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_list("beta_rho")
-```
+<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
+<!-- histogram_and_ecdf_list("beta_rho") -->
+<!-- ``` -->
 
-## `beta_alpha`
+<!-- ## `beta_alpha` -->
 
-```{r eval = pars_eval[["beta_alpha"]]}
-histogram_and_ecdf_list("beta_alpha")
-```
+<!-- ```{r eval = pars_eval[["beta_alpha"]]} -->
+<!-- histogram_and_ecdf_list("beta_alpha") -->
+<!-- ``` -->
 
-## `beta_lambda`
+<!-- ## `beta_lambda` -->
 
-```{r eval = pars_eval[["beta_lambda"]]}
-histogram_and_ecdf_list("beta_lambda")
-```
+<!-- ```{r eval = pars_eval[["beta_lambda"]]} -->
+<!-- histogram_and_ecdf_list("beta_lambda") -->
+<!-- ``` -->
 
-## `beta_anc_rho`
+<!-- ## `beta_anc_rho` -->
 
-```{r eval = pars_eval[["beta_anc_rho"]]}
-histogram_and_ecdf_list("beta_anc_rho")
-```
+<!-- ```{r eval = pars_eval[["beta_anc_rho"]]} -->
+<!-- histogram_and_ecdf_list("beta_anc_rho") -->
+<!-- ``` -->
 
-## `beta_anc_alpha`
+<!-- ## `beta_anc_alpha` -->
 
-```{r eval = pars_eval[["beta_anc_alpha"]]}
-histogram_and_ecdf("beta_anc_alpha")
-```
+<!-- ```{r eval = pars_eval[["beta_anc_alpha"]]} -->
+<!-- histogram_and_ecdf("beta_anc_alpha") -->
+<!-- ``` -->
 
-## `logit`
+<!-- ## `logit` -->
 
-```{r eval = FALSE}
-lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf)
-```
+<!-- ```{r eval = FALSE} -->
+<!-- lapply(pars[stringr::str_starts(pars, "logit")], histogram_and_ecdf) -->
+<!-- ``` -->
 
-## `log_sigma`
+<!-- ## `log_sigma` -->
 
-```{r eval = FALSE}
-lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf)
-```
+<!-- ```{r eval = FALSE} -->
+<!-- lapply(pars[stringr::str_starts(pars, "log_sigma")], histogram_and_ecdf) -->
+<!-- ``` -->
 
-## `u_rho_x`
+<!-- ## `u_rho_x` -->
 
-```{r eval = pars_eval[["u_rho_x"]]}
-histogram_and_ecdf_list("u_rho_x")
-```
+<!-- ```{r eval = pars_eval[["u_rho_x"]]} -->
+<!-- histogram_and_ecdf_list("u_rho_x") -->
+<!-- ``` -->
 
-## `us_rho_x`
+<!-- ## `us_rho_x` -->
 
-```{r eval = pars_eval[["us_rho_x"]]}
-histogram_and_ecdf_list("us_rho_x")
-```
+<!-- ```{r eval = pars_eval[["us_rho_x"]]} -->
+<!-- histogram_and_ecdf_list("us_rho_x") -->
+<!-- ``` -->
 
-## `u_rho_xs`
+<!-- ## `u_rho_xs` -->
 
-```{r eval = pars_eval[["u_rho_xs"]]}
-histogram_and_ecdf_list("u_rho_xs")
-```
+<!-- ```{r eval = pars_eval[["u_rho_xs"]]} -->
+<!-- histogram_and_ecdf_list("u_rho_xs") -->
+<!-- ``` -->
 
-## `us_rho_xs`
+<!-- ## `us_rho_xs` -->
 
-```{r eval = pars_eval[["us_rho_xs"]]}
-histogram_and_ecdf_list("us_rho_xs")
-```
+<!-- ```{r eval = pars_eval[["us_rho_xs"]]} -->
+<!-- histogram_and_ecdf_list("us_rho_xs") -->
+<!-- ``` -->
 
-## `u_rho_as`
+<!-- ## `u_rho_as` -->
 
-```{r eval = pars_eval[["u_rho_as"]]}
-histogram_and_ecdf_list("u_rho_as")
-```
+<!-- ```{r eval = pars_eval[["u_rho_as"]]} -->
+<!-- histogram_and_ecdf_list("u_rho_as") -->
+<!-- ``` -->
 
-## `u_alpha_x`
+<!-- ## `u_alpha_x` -->
 
-```{r eval = pars_eval[["u_alpha_x"]]}
-histogram_and_ecdf_list("u_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_x"]]} -->
+<!-- histogram_and_ecdf_list("u_alpha_x") -->
+<!-- ``` -->
 
-## `us_alpha_x`
+<!-- ## `us_alpha_x` -->
 
-```{r eval = pars_eval[["us_alpha_x"]]}
-histogram_and_ecdf_list("us_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["us_alpha_x"]]} -->
+<!-- histogram_and_ecdf_list("us_alpha_x") -->
+<!-- ``` -->
 
-## `u_alpha_xs`
+<!-- ## `u_alpha_xs` -->
 
-```{r eval = pars_eval[["u_alpha_xs"]]}
-histogram_and_ecdf_list("u_alpha_xs")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_xs"]]} -->
+<!-- histogram_and_ecdf_list("u_alpha_xs") -->
+<!-- ``` -->
 
-## `us_alpha_xs`
+<!-- ## `us_alpha_xs` -->
 
-```{r eval = pars_eval[["beta_rho"]]}
-histogram_and_ecdf_list("us_alpha_xs")
-```
+<!-- ```{r eval = pars_eval[["beta_rho"]]} -->
+<!-- histogram_and_ecdf_list("us_alpha_xs") -->
+<!-- ``` -->
 
-## `u_alpha_a`
+<!-- ## `u_alpha_a` -->
 
-```{r eval = pars_eval[["u_alpha_a"]]}
-histogram_and_ecdf_list("u_alpha_a")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_a"]]} -->
+<!-- histogram_and_ecdf_list("u_alpha_a") -->
+<!-- ``` -->
 
-## `u_alpha_as`
+<!-- ## `u_alpha_as` -->
 
-```{r eval = pars_eval[["u_alpha_as"]]}
-histogram_and_ecdf_list("u_alpha_as")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_as"]]} -->
+<!-- histogram_and_ecdf_list("u_alpha_as") -->
+<!-- ``` -->
 
-## `u_alpha_xa`
+<!-- ## `u_alpha_xa` -->
 
-```{r eval = pars_eval[["u_alpha_xa"]]}
-histogram_and_ecdf_list("u_alpha_xa")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_xa"]]} -->
+<!-- histogram_and_ecdf_list("u_alpha_xa") -->
+<!-- ``` -->
 
-## `ui_lambda_x`
+<!-- ## `ui_lambda_x` -->
 
-```{r eval = pars_eval[["ui_lambda_x"]]}
-histogram_and_ecdf_list("ui_lambda_x")
-```
+<!-- ```{r eval = pars_eval[["ui_lambda_x"]]} -->
+<!-- histogram_and_ecdf_list("ui_lambda_x") -->
+<!-- ``` -->
 
-## `ui_anc_rho_x`
+<!-- ## `ui_anc_rho_x` -->
 
-```{r eval = pars_eval[["ui_anc_rho_x"]]}
-histogram_and_ecdf_list("ui_anc_rho_x")
-```
+<!-- ```{r eval = pars_eval[["ui_anc_rho_x"]]} -->
+<!-- histogram_and_ecdf_list("ui_anc_rho_x") -->
+<!-- ``` -->
 
-## `ui_anc_alpha_x`
+<!-- ## `ui_anc_alpha_x` -->
 
-```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-histogram_and_ecdf_list("ui_anc_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["ui_anc_alpha_x"]]} -->
+<!-- histogram_and_ecdf_list("ui_anc_alpha_x") -->
+<!-- ``` -->
 
-## `log_or_gamma`
+<!-- ## `log_or_gamma` -->
 
-```{r eval = pars_eval[["log_or_gamma"]]}
-histogram_and_ecdf_list("log_or_gamma")
-```
+<!-- ```{r eval = pars_eval[["log_or_gamma"]]} -->
+<!-- histogram_and_ecdf_list("log_or_gamma") -->
+<!-- ``` -->
 
-## `rho_t1_out`
+<!-- ## `rho_t1_out` -->
 
-There are too many variables to plot this here!
+<!-- There are too many variables to plot this here! -->
 
-## `alpha_t1_out`
+<!-- ## `alpha_t1_out` -->
 
-There are too many variables to plot this here!
+<!-- There are too many variables to plot this here! -->
 
-## `lambda_t1_out`
+<!-- ## `lambda_t1_out` -->
 
-There are too many variables to plot this here!
+<!-- There are too many variables to plot this here! -->
 
@@
 
-### `beta`
+<!-- ### `beta` -->
 
-```{r}
-ks_plot_starts_with("beta")
-```
+<!-- ```{r} -->
+<!-- ks_plot_starts_with("beta") -->
+<!-- ``` -->
 
-### `logit`
+<!-- ### `logit` -->
 
-```{r}
-ks_plot_starts_with("logit")
-```
+<!-- ```{r} -->
+<!-- ks_plot_starts_with("logit") -->
+<!-- ``` -->
 
-### `log_sigma`
+<!-- ### `log_sigma` -->
 
-```{r}
-ks_plot_starts_with("log_sigma")
-```
+<!-- ```{r} -->
+<!-- ks_plot_starts_with("log_sigma") -->
+<!-- ``` -->
 
-### `u_rho_x`
+<!-- ### `u_rho_x` -->
 
-```{r eval = pars_eval[["u_rho_x"]]}
-ks_plot_starts_with("u_rho_x")
-```
+<!-- ```{r eval = pars_eval[["u_rho_x"]]} -->
+<!-- ks_plot_starts_with("u_rho_x") -->
+<!-- ``` -->
 
-### `u_rho_xs`
+<!-- ### `u_rho_xs` -->
 
-```{r eval = pars_eval[["u_rho_xs"]]}
-ks_plot_starts_with("u_rho_xs")
-```
+<!-- ```{r eval = pars_eval[["u_rho_xs"]]} -->
+<!-- ks_plot_starts_with("u_rho_xs") -->
+<!-- ``` -->
 
-### `us_rho_x`
+<!-- ### `us_rho_x` -->
 
-```{r eval = pars_eval[["us_rho_x"]]}
-ks_plot_starts_with("us_rho_x")
-```
+<!-- ```{r eval = pars_eval[["us_rho_x"]]} -->
+<!-- ks_plot_starts_with("us_rho_x") -->
+<!-- ``` -->
 
-### `us_rho_xs`
+<!-- ### `us_rho_xs` -->
 
-```{r eval = pars_eval[["us_rho_xs"]]}
-ks_plot_starts_with("us_rho_xs")
-```
+<!-- ```{r eval = pars_eval[["us_rho_xs"]]} -->
+<!-- ks_plot_starts_with("us_rho_xs") -->
+<!-- ``` -->
 
-### `u_rho_a`
+<!-- ### `u_rho_a` -->
 
-```{r eval = pars_eval[["u_rho_a"]]}
-ks_plot_starts_with("u_rho_a")
-```
+<!-- ```{r eval = pars_eval[["u_rho_a"]]} -->
+<!-- ks_plot_starts_with("u_rho_a") -->
+<!-- ``` -->
 
-### `u_rho_as`
+<!-- ### `u_rho_as` -->
 
-```{r eval = pars_eval[["u_rho_as"]]}
-ks_plot_starts_with("u_rho_as")
-```
+<!-- ```{r eval = pars_eval[["u_rho_as"]]} -->
+<!-- ks_plot_starts_with("u_rho_as") -->
+<!-- ``` -->
 
-### `u_alpha_x`
+<!-- ### `u_alpha_x` -->
 
-```{r eval = pars_eval[["u_alpha_x"]]}
-ks_plot_starts_with("u_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_x"]]} -->
+<!-- ks_plot_starts_with("u_alpha_x") -->
+<!-- ``` -->
 
-### `u_alpha_xs`
+<!-- ### `u_alpha_xs` -->
 
-```{r eval = pars_eval[["u_alpha_xs"]]}
-ks_plot_starts_with("u_alpha_xs")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_xs"]]} -->
+<!-- ks_plot_starts_with("u_alpha_xs") -->
+<!-- ``` -->
 
-### `us_alpha_x`
+<!-- ### `us_alpha_x` -->
 
-```{r eval = pars_eval[["us_alpha_x"]]}
-ks_plot_starts_with("us_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["us_alpha_x"]]} -->
+<!-- ks_plot_starts_with("us_alpha_x") -->
+<!-- ``` -->
 
-### `us_alpha_xs`
+<!-- ### `us_alpha_xs` -->
 
-```{r eval = pars_eval[["us_alpha_xs"]]}
-ks_plot_starts_with("us_alpha_xs")
-```
+<!-- ```{r eval = pars_eval[["us_alpha_xs"]]} -->
+<!-- ks_plot_starts_with("us_alpha_xs") -->
+<!-- ``` -->
 
-### `u_alpha_a`
+<!-- ### `u_alpha_a` -->
 
-```{r eval = pars_eval[["u_alpha_a"]]}
-ks_plot_starts_with("u_alpha_a")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_a"]]} -->
+<!-- ks_plot_starts_with("u_alpha_a") -->
+<!-- ``` -->
 
-### `u_alpha_as`
+<!-- ### `u_alpha_as` -->
 
-```{r eval = pars_eval[["u_alpha_as"]]}
-ks_plot_starts_with("u_alpha_as")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_as"]]} -->
+<!-- ks_plot_starts_with("u_alpha_as") -->
+<!-- ``` -->
 
-### `u_alpha_xa`
+<!-- ### `u_alpha_xa` -->
 
-```{r eval = pars_eval[["u_alpha_xa"]]}
-ks_plot_starts_with("u_alpha_xa")
-```
+<!-- ```{r eval = pars_eval[["u_alpha_xa"]]} -->
+<!-- ks_plot_starts_with("u_alpha_xa") -->
+<!-- ``` -->
 
-### `ui_anc_rho_x`
+<!-- ### `ui_anc_rho_x` -->
 
-```{r eval = pars_eval[["ui_anc_rho_x"]]}
-ks_plot_starts_with("ui_anc_rho_x")
-```
+<!-- ```{r eval = pars_eval[["ui_anc_rho_x"]]} -->
+<!-- ks_plot_starts_with("ui_anc_rho_x") -->
+<!-- ``` -->
 
-### `ui_anc_alpha_x`
+<!-- ### `ui_anc_alpha_x` -->
 
-```{r eval = pars_eval[["ui_anc_alpha_x"]]}
-ks_plot_starts_with("ui_anc_alpha_x")
-```
+<!-- ```{r eval = pars_eval[["ui_anc_alpha_x"]]} -->
+<!-- ks_plot_starts_with("ui_anc_alpha_x") -->
+<!-- ``` -->
 
-### `log_or_gamma`
+<!-- ### `log_or_gamma` -->
 
-```{r eval = pars_eval[["log_or_gamma"]]}
-ks_plot_starts_with("log_or_gamma")
-```
+<!-- ```{r eval = pars_eval[["log_or_gamma"]]} -->
+<!-- ks_plot_starts_with("log_or_gamma") -->
+<!-- ``` -->
 
-### `rho_t1_out`
+<!-- ### `rho_t1_out` -->
 
-```{r}
-ks_df_out(par = "rho_t1_out") %>%
-  ks_plot(par = "rho_t1_out", alpha = 0.2)
-```
+<!-- ```{r} -->
+<!-- ks_df_out(par = "rho_t1_out") %>% -->
+<!--   ks_plot(par = "rho_t1_out", alpha = 0.2) -->
+<!-- ``` -->
 
-### `alpha_t1_out`
+<!-- ### `alpha_t1_out` -->
 
-```{r}
-ks_df_out(par = "alpha_t1_out") %>%
-  ks_plot(par = "rho_t1_out", alpha = 0.2)
-```
+<!-- ```{r} -->
+<!-- ks_df_out(par = "alpha_t1_out") %>% -->
+<!--   ks_plot(par = "rho_t1_out", alpha = 0.2) -->
+<!-- ``` -->
 
-### `lambda_t1_out`
+<!-- ### `lambda_t1_out` -->
 
-```{r}
-ks_df_out(par = "lambda_t1_out") %>%
-  filter(!age_group %in% c("Y005_009", "Y010_014", "Y080_999")) %>%
-  ks_plot(par = "lambda_t1_out", alpha = 0.2)
-```
+<!-- ```{r} -->
+<!-- ks_df_out(par = "lambda_t1_out") %>% -->
+<!--   filter(!age_group %in% c("Y005_009", "Y010_014", "Y080_999")) %>% -->
+<!--   ks_plot(par = "lambda_t1_out", alpha = 0.2) -->
+<!-- ``` -->
 
@@
 
-ks_summary_latent <- filter(ks_summary, type == "Latent field")
+ks_summary_latent <- filter(ks_summary, type == "Latent")
 
@@
 ks_summary_latent %>%
-  ggplot(aes(x = TMB, y = aghq, col = parname, size = size)) +
-  geom_point(alpha = 0.4) +
-  xlim(0, xy_length) +
-  ylim(0, xy_length) +
+  ggplot(aes(x = TMB, y = aghq - TMB, col = parname, size = size)) +
+  geom_point(alpha = 0.7) +
   scale_color_manual(values = extended_cbpalette(n = 20)) +
   scale_size_continuous(breaks = c(2, 10, 32), labels = c(2, 10, 32)) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
-  labs(x = "KS(TMB, NUTS)", y = "KS(PCA-AGHQ, NUTS)", col = "Parameter", size = "Length", subtitle = "Smaller KS values indicate higher accuracy") +
+  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
+  labs(x = "KS(GEB, NUTS)", y = "KS(GPCA-AGHQ, NUTS) - KS(GEB, NUTS)", col = "Parameter", size = "Length") +
   theme_minimal() +
   guides(size = guide_legend(title.position = "top", direction = "vertical"), col = guide_legend(title.position = "top")) +
-  theme(legend.position = "bottom", legend.title = element_text(size = rel(0.9)), legend.text = element_text(size = rel(0.7)))
+  theme(legend.position = "bottom", legend.title = element_text(size = rel(0.9)), legend.text = element_text(size = rel(0.65)))
 
@@
   filter(type != "Hyper") %>%
-  mutate(method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq")) %>%
+  mutate(method = fct_reorg(method, "GEB" = "TMB", "GPCA-AGHQ" = "aghq")) %>%
   left_join(data.frame(ess) %>%
@@
     geom_point(shape = 1, alpha = 0.5) +
-    geom_smooth(method = "lm", color = "#56B4E9", fullrange = TRUE) +
-    stat_poly_eq(use_label("eq"), label.x.npc = "right") +
-    stat_poly_eq(label.y = 0.9, label.x.npc = "right") +
+    geom_smooth(method = "lm", color = "#CC79A7", fullrange = TRUE) +
+    ggpmisc::stat_poly_eq(ggpmisc::use_label("eq"), label.x.npc = "right") +
+    ggpmisc::stat_poly_eq(label.y = 0.9, label.x.npc = "right") +
     scale_x_continuous(limits = c(0, NA)) +
@@
     theme_minimal() +
-    labs(x = "ESS", y = "KS(method, NUTS)")
+    labs(x = "NUTS ESS", y = "KS(approach, NUTS)")
 

### 2023-12-31 src/naomi-simple_increase-s-k/increase-s-k.Rmd
Small change to MCMC, and make increase s, k output a figure for thesis appendix

@@
 ```{r}
-ggplot(df, aes(x = s, y = lognormconst)) +
+multi.utils::cbpalette()
+
+df %>%
+  pivot_longer(
+    cols = c("lognormconst", "time"),
+    values_to = "value",
+    names_to = "indicator"
+  ) %>%
+  mutate(
+    indicator = fct_recode(indicator,
+      "Log normalising\nconstant estimate" = "lognormconst",
+      "Time taken (m)" = "time"
+    )
+  ) %>%
+  ggplot(aes(x = s, y = value, col = indicator)) +
     geom_point() +
-  facet_grid(. ~ k) +
-  labs(x = "Dimensions kept of the PCA", y = "Log normalising constant") +
+    facet_grid(indicator ~ k, scales = "free") +
+    labs(x = " PCA-AGHQ dimensions kept", y = "") +
+    scale_color_manual(values = c("#009E73", "#E69F00")) +
+    guides(col = "none") +
     theme_minimal()
+
+ggsave("lognormconst.png", h = 4, w = 6.25, bg = "white")
 ```
-
-```{r}
-ggplot(df, aes(x = s, y = time)) +
-  geom_point() +
-  facet_grid(. ~ k) +
-  scale_y_continuous(limits = c(0, NA)) +
-  labs(x = "Dimensions kept of the PCA", y = "Time (mins)") +
-  theme_minimal()
-```

### 2023-12-31 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Small change to MCMC, and make increase s, k output a figure for thesis appendix

@@
 fig_top_left <- ggplot(pairs_df, aes(x = log_sigma_rho_a)) +
-  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
+  geom_histogram(fill = "#56B4E9") +
   labs(x = "", y = "", tag = "A") + 
@@
 fig_bot_right <- ggplot(pairs_df, aes(x = logit_phi_rho_a)) +
-  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
+  geom_histogram(fill = "#56B4E9") +
   labs(x = "logit_phi_rho_a", y = "", tag = "D") + 
@@
 fig_top_left <- ggplot(pairs_df, aes(x = log_sigma_alpha_x)) +
-  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
+  geom_histogram(fill = "#56B4E9") +
   labs(x = "", y = "", tag = "A") + 
@@
 fig_bot_right <- ggplot(pairs_df, aes(x = logit_phi_alpha_x)) +
-  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
+  geom_histogram(fill = "#56B4E9") +
   labs(x = "logit_phi_alpha_x", y = "", tag = "D") + 

### 2023-12-31 src/naomi-simple_ks/ks.Rmd
Final KS figure touches

@@
   geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
-  labs(x = "KS(GEB, NUTS)", y = "KS(GPCA-AGHQ, NUTS) - KS(GEB, NUTS)", col = "Parameter", size = "Length") +
+  labs(x = "KS(GEB)", y = "KS(GPCA-AGHQ) - KS(GEB)", col = "Parameter", size = "Length") +
   theme_minimal() +
-  guides(size = guide_legend(title.position = "top", direction = "vertical"), col = guide_legend(title.position = "top")) +
-  theme(legend.position = "bottom", legend.title = element_text(size = rel(0.9)), legend.text = element_text(size = rel(0.65)))
+  guides(size = guide_legend(title.position = "top", direction = "vertical"), col = guide_legend(title.position = "top", nrow = 5, byrow = TRUE)) +
+  theme(legend.position = "bottom", legend.title = element_text(size = rel(0.9)), legend.text = element_text(size = rel(0.9)))
 

### 2023-12-31 src/naomi-simple_ks/script.R

@@ -9,3 +9,79 @@ saveRDS(beta_alpha$df, "beta_alpha.rds")
 
 beta_anc_rho <- histogram_and_ecdf("beta_anc_rho", return_df = TRUE)
 saveRDS(beta_anc_rho$df, "beta_anc_rho.rds")
+
+#' Quick plot for thesis!
+
+par <- "ui_lambda_x"
+i <- 26
+
+colours <- c("#56B4E9", "#009E73", "#E69F00", "#CC79A7")
+
+par_name <- paste0(par, "[", i, "]")
+
+df_compare <- rbind(
+  data.frame(method = "GEB", samples = as.numeric(tmb$fit$sample[[par]][i, ])),
+  data.frame(method = "GPCA-AGHQ", samples = as.numeric(aghq$quad$sample[[par]][i, ])),
+  data.frame(method = "NUTS", samples = as.numeric(tmbstan$mcmc$sample[[par]][i, ]))
+)
+
+rhat <- signif(rstan::Rhat(tmbstan$mcmc$sample[[par]][i, ]), 3)
+ess <- signif(rstan::ess_bulk(tmbstan$mcmc$sample[[par]][i, ]), 3)
+
+df_compare$method <- factor(df_compare$method, levels = c("GEB", "GPCA-AGHQ", "NUTS"))
+
+mean <- signif(mean(filter(df_compare, method == "NUTS")$samples), digits = 3)
+sd <- signif(sd(filter(df_compare, method == "NUTS")$samples), digits = 3)
+
+histogram_plot <- ggplot(df_compare, aes(x = samples, fill = method, col = method)) +
+  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, position = "identity", bins = 30) +
+  theme_minimal() +
+  facet_grid(method~.) +
+  labs(y = "Density", fill = "Method") +
+  scale_color_manual(values = colours) +
+  scale_fill_manual(values = colours) +
+  theme(legend.position = "none") +
+  labs(x = "")
+
+grid <- seq(from = min(df_compare$samples), to = max(df_compare$samples), length.out = 1000)
+
+tmb_ecdf <- stats::ecdf(filter(df_compare, method == "GEB") %>% pull(samples))
[…]

### 2023-12-31 src/naomi-simple_mmd/mmd.Rmd
Add MMD output for thesis

@@
 round(100 * (mmd_aghq@mmdstats[2] - mmd_tmb@mmdstats[2]) / mmd_tmb@mmdstats[2])
+
+saveRDS(list(mmd_tmb = mmd_tmb, mmd_aghq = mmd_aghq), file = "mmd.rds")
 ```

### 2024-12-07 src/docs_paper-jtb/paper.Rmd
Add JTB template

---
title: Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature
author:
  - name: Adam Howes
    email: ath19@ic.ac.uk
    affiliation: A
    correspondingauthor: true
  - name: Alex Stringer
    email: alex.stringer@uwaterloo.ca
    affiliation: B
  - name: Seth R. Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    affiliation: C
  - name: Jeffrey W. Imai-Eaton
    email: jeaton@hsph.harvard.edu
    affiliation: D
address:
  - code: A
    organization: Department of Mathematics
    country: Imperial College London
  - code: B
    organization: Department of Statistics and Actuarial Science
    country: University of Waterloo
  - code: C
    organization: Department of Computer Science
    country: University of Oxford
  - code: D
    organization: Harvard T.H. Chan School of Public Health
    country: Harvard University
abstract: |
  This is the abstract.
keywords: 
  - Bayesian statistics
  - spatial statistics
  - evidence synthesis
  - small-area estimation
  - approximate inference
  - INLA
  - AGHQ
  - HIV epidemiology
journal: "Journal of Theoretical Biology"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: citations.bib
biblio-style: elsarticle-harv # author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme
classoption: preprint, 3p, authoryear # remove authoryear is not using `elsarticle-harv`
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

# References {-}

### 2024-12-07 src/docs_paper-jtb/appendix.Rmd
Copy over from old paper

--- 2023-07-24 src/docs_paper/appendix.Rmd
@@
 
-```{r echo = FALSE}
+```{r echo = FALSE, message = FALSE}
+library(dplyr)
+
 options(scipen = 100)
@@
 ```{r}
-point_summary <- read_csv("depends/point-estimates.csv")
+point_summary <- readr::read_csv("depends/point-estimates.csv")
 
-point_summary <- point_summary %>%
-  mutate(type = fct_recode(type, "Output" = "output", "Latent" = "latent"))
+point_summary <- point_summary |>
+  mutate(type = forcats::fct_recode(type, "Output" = "output", "Latent" = "latent"))
 
 latex_table_point_estimates <- function(.indicator) {
-  point_summary %>%
-  group_by(par, type, indicator, method) %>%
+  point_summary |>
+  group_by(par, type, indicator, method) |>
   summarise(
@@
     mae = mean(abs(truth - approximate))
-  ) %>%
-  ungroup() %>%
-  pivot_wider(
+  ) |>
+  ungroup() |>
+  tidyr::pivot_wider(
     names_from = "method",
     values_from = c("rmse", "mae")
-  ) %>%
-  filter(indicator == .indicator) %>%
-  select(Parameter = par, Type = type, `TMB` = `rmse_TMB`, `PCA-AGHQ` = `rmse_PCA-AGHQ`) %>%
+  ) |>
+  filter(indicator == .indicator) |>
+  select(Parameter = par, Type = type, `TMB` = `rmse_TMB`, `PCA-AGHQ` = `rmse_PCA-AGHQ`) |>
   mutate(
@@
     `Percent Difference` =  (`PCA-AGHQ` - TMB) / TMB
-  ) %>%
-  gt::gt(groupname_col = "Type") %>%
+  ) |>
+  gt::gt(groupname_col = "Type") |>
   gt::fmt_number(
@@
     n_sigfig = 2
-  ) %>%
+  ) |>
  gt::fmt_percent(
@@
     decimals = 1
-  ) %>%
+  ) |>
   gt::text_transform(
@@
     fn = function(x) paste0("\\texttt{", x, "}")
-  ) %>%
-  gt::as_latex() %>%
+  ) |>
+  gt::as_latex() |>
   as.character()
@@
 
-latex_table_point_estimates("Posterior mean estimate") %>%
+latex_table_point_estimates("Posterior mean estimate") |>
   cat(file = "point-summary-mean.tex")
 
-latex_table_point_estimates("Posterior SD estimate") %>%
+latex_table_point_estimates("Posterior SD estimate") |>
   cat(file = "point-summary-sd.tex")
@@
 
-ks_summary %>%
-  ungroup() %>%
-  filter(type == "Latent") %>%
-  select(-type) %>%
-  select(Parameter = parname, TMB, `PCA-AGHQ` = aghq) %>%
-  mutate(Difference = `PCA-AGHQ` - TMB) %>%
-  gt::gt() %>%
+ks_summary |>
+  ungroup() |>
+  filter(type == "Latent") |>
+  select(-type) |>
+  select(Parameter = parname, TMB, `PCA-AGHQ` = aghq) |>
+  mutate(Difference = `PCA-AGHQ` - TMB) |>
+  gt::gt() |>
   gt::summary_rows(
@@
    decimals = 3
-  ) %>%
+  ) |>
   gt::fmt_number(
@@
     decimals = 3
-  ) %>%
+  ) |>
   gt::text_transform(
@@
     fn = function(x) paste0("\\texttt{", x, "}")
-  ) %>%
+  ) |>
   gt::text_transform(
@@
     fn = function(x) ifelse(x < 0, paste0("\\textbf{", x, "}"), paste0(x))
-  ) %>%
-  gt::as_latex() %>%
-  as.character() %>%
+  ) |>
+  gt::as_latex() |>
+  as.character() |>
   cat(file = "ks-summary.tex")
@@
 
-n_divergent <- np %>%
-  filter(Parameter == "divergent__") %>%
-  summarise(n_divergent = sum(Value)) %>%
+n_divergent <- np |>
+  filter(Parameter == "divergent__") |>
+  summarise(n_divergent = sum(Value)) |>
   pull(n_divergent)
@@
 
-<!-- ```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."} -->
-<!-- knitr::include_graphics("depends/algorithm-flowchart.pdf") -->
-<!-- ``` -->
-
 \clearpage

### 2024-12-07 src/docs_paper-jtb/paper.Rmd
Copy over from old paper

@@
     citation_package: natbib
+    includes:
+      in_header: preamble.tex
 ---
 
+```{r echo = FALSE, message = FALSE}
+library(dplyr)
+
+options(scipen = 100)
+
+knitr::opts_chunk$set(
+  echo = FALSE,
+  warning = FALSE,
+  message = FALSE,
+  dpi = 320,
+  cache = TRUE,
+  out.width = "95%",
+  fig.align = 'center'
+)
+```
+
+# Introduction
+
+Accurate estimates of HIV indicators are crucial for mounting an effective public health response to the HIV epidemic.
+These estimates should be timely, and at a geographic level at which health systems are planned and delivered.
+Producing granular estimates is challenging, in large part due to limitations of the data from available sources.
+Nationally-representative household surveys provide the most statistically reliable data, but are costly to run and so only conducted every five years or so in most countries, with limited sample size at the district level.
+Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available in closer to real-time, but are not representative of the entire population.
+To address these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
+Modelling multiple data sources jointly mitigates the limitations of any single source, increases statistical power, and can prompt investigation into conflicts of information.
+
+Software has been developed for Naomi (\url{https://naomi.unaids.org}), allowing over 35 countries to input their data and interactively generate estimates during workshops as a part of a yearly process supported by UNAIDS.
+Creation of estimates by country teams, rather than external agencies or researchers, is an important and distinctive feature of the HIV response.
+Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust in the resulting estimates, creating a virtuous cycle of data quality, use and ownership [@noor2022country].
+
+Naomi is a complex model, comprised of multiple linked generalized linear mixed models (GLMMs), and presents a challenging Bayesian inference problem.
+As well as hundreds of fixed and random effect parameters, Naomi has >20 hyperparameters: substantially more than the small number that can typically be handled by approaches like integrated nested Laplace approximations [INLA; @rue2009approximate].
+Moreover, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
+
+To allow interactive review and iteration of model results by workshop participants, the inference procedure should be fast and have low memory usage.
+Because of the scale of the model and challenging features of its posterior geometry [@neal2003slice], Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
+In addition, we require the inference procedure to produce estimates reliably and automatically, across a range of country settings, without knowledge of the specific data used in advance.
+As such, approaches requiring substantial statistical expertise or tuning, as would be the case for monitoring MCMC convergence and suitability, are not viable.
+
+To meet these requirements, inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
+We refer to this approach as TMB, distinguished from the package `TMB`.
+Owing to its speed and flexibility, `TMB` is gaining popularity, particularly in spatial statistics [@osgood2022statistical] and via the user-friendly `glmmTMB` \textsc{R} package [@brooks2017glmmTMB].
+Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the parameters.
+For the Naomi model, this subset is the high-dimensional latent field, leaving a smaller number of hyperparameters.
+Taking inspiration from the AD Model Builder package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for gradient-based numerical optimisation routines and the Laplace approximation.
+
+Although the TMB approach is fast, within the empirical Bayes framework hyperparameter uncertainty is not accounted for in the latent field posterior.
+We had concerns that this would result in underestimation of posterior variances, and may have important practical implications.
+As such, we were motivated to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
+To meet these requirements, we developed an inference method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over many hyperparameters.
+AGHQ is a quadrature method based on the theory of polynomial interpolation, and is well suited to statistical estimation problems in which the integrand is well approximated by a Gaussian multiplied by a polynomial.
+@bilodeau2022stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
+However, it is not computationally feasible to use AGHQ in >20 dimensions directly, as exponentially many nodes are required.
+Instead, we used principal components analysis (PCA) of the inverse curvature at the mode to find a subspace which explained most of the hyperparameter variance.
+For the Naomi model in Malawi, this resulted in a grid which was tractable as it had millions of times fewer nodes the corresponding dense grid.
+Our implementation of the method makes use of the existing Naomi `TMB` template, and is immediately compatible with any model with such a template.
+
+<!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
+<!-- Want to say in particular what the class of models that `inlabru` allows you to fit are -->
+<!-- Don't know the answer to this yet, but asking Janine Illian I'm told you don't have to call R-INLA that many times -->
+Other work aiming to extend the scope of INLA-like methods includes the `inlabru` \textsc{R} package [@bachl2019inlabru], INLA within MCMC [@gomez2018markov], and importance sampling with INLA [@berild2022importance], all of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
+The approach of `inlabru` is to approximate non-linear predictors using linearisation, by making iterative calls to `R-INLA`.
+INLA within MCMC and importance sampling with INLA are suitable for models which are LGMs conditional on some subset of the parameters being fixed.
+It's not clear the extent to which any of these methods would be compatible with Naomi.
+
+The remainder of this paper is organised as follows.
+Section \ref{sec:naomi} outlines the version of the Naomi model that we consider in this paper, and Section \ref{sec:elgm} describes how it falls within the ELGM framework.
+In Section \ref{sec:inferencenaomi} we review the deterministic inference method for ELGMs used by @stringer2022fast based on nested application of AGHQ and the Laplace approximation, before introducing the PCA-based modification we use to enable application to Naomi.
+In a case study (Section \ref{sec:results}) we evaluate the accuracy of PCA-AGHQ for the simplified Naomi model fit to data from Malawi, as compared with TMB and gold-standard MCMC.
+Finally, we discuss our conclusions, and directions for future research in Section \ref{sec:conclusions}.
+
+# Simplified Naomi model\label{sec:naomi}
+
+@eaton2021naomi specify a joint model linking three small-area estimation models.
+We consider a simplified version defined only at the time of the most recent household survey with HIV testing, omitting nowcasting and temporal projection, as these time points involve limited inferences.
+An overview of the simplified model is given below, and a more complete mathematical description is provided in Appendix \ref{sec:math}.
+
+## Household survey component \label{sec:household}
+
+Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place.
+Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ index five-year age group, and $s \in \mathcal{S}$ index sex.
+For ease of notation, let $i$ index the finest district-age-sex division included in the model.
+Let $I \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported, and $\mathcal{I}$ be the set of all $I$ such that $I \in \mathcal{I}$.
+
+Let $N_i \in \mathbb{N}$ be the known, fixed population size.
+We infer the following unknown HIV indicators using linked regression equations:
+
+* HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive;
+* antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and
+* annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
+
+We specify independent logistic regression models for HIV prevalence and ART coverage in the general population such that $\text{logit}(\rho_i) = \eta^\rho_i$ and $\text{logit}(\alpha_i) = \eta^\alpha_i$.
+HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i$, and depends on adult HIV prevalence and adult ART coverage.
+The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix \ref{sec:math}.
+Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
+We link this proportion to HIV incidence via
+\begin{equation}
+\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right), \label{eq:kappa}
+\end{equation}
+where the mean duration of recent infection $\Omega_T$ and the proportion of long-term HIV infections misclassified as recent $\beta_T$ are strongly informed by priors for the particular survey.
+
+These processes are each informed by household survey data.
+We first calculate the weighted aggregate survey observations
+\begin{equation*}
+\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j},
+\end{equation*}
+with individual responses $\theta_j \in \{0, 1\}$ and design weights $w_j$ for each of $\theta \in \{\rho, \alpha, \kappa\}$.
+The design weights are provided by the survey and aim to reduce bias by decreasing possible correlation between response and recording mechanism [@meng2018statistical].
+The index $j$ runs across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
+We take the weighted observed number of outcomes to be $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
+\begin{equation*}
+m^{\theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
+\end{equation*}
+is the Kish effective sample size (ESS) [@kish1965survey].
+As the Kish ESS is maximised by constant design weights, in exchange for reducing bias we reduce our ESS and thereby increase variance.
+We use a binomial working likelihood [@chen2014use] defined to operate on the reals
+\begin{equation*}
+y^{\theta}_{I} \sim \text{xBin}(m^{\theta}_{I}, \theta_{I})
+\end{equation*}
+to model these aggregate observations, where $\theta_{I}$ are the following weighted aggregates
+\begin{equation*}
+\rho_{I} = \frac{\sum_{i \in I} N_i \rho_i}{\sum_{i \in I} N_i}, \quad
+\alpha_{I} = \frac{\sum_{i \in I} N_i \rho_i \alpha_i}{\sum_{i \in I} N_i \rho_i}, \quad
+\kappa_{I} = \frac{\sum_{i \in I} N_i \rho_i \kappa_i}{\sum_{i \in I} N_i \rho_i}.
+\end{equation*}
+Though our approach accounts for effect of survey weighting on the variance, it does not take into account correlation structure in the samples, for example due to cluster sampling [@wakefield2020small].
+
+## ANC testing component \label{sec:anc}
+
+HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset from the general population indicators as follows
+\begin{align*}
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
+\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\alpha^\text{ANC}}_i.
+\end{align*}
+These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
+We take the number of ANC clients with ascertained status to be fixed as $m^{\rho^\text{ANC}}_I$. 
+We then model the number of those with positive status $y^{\rho^\text{ANC}}_I$, and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ using nested binomial likelihoods
+\begin{align*}
+y^{\rho^\text{ANC}}_I &\sim \text{Bin}(m^{\rho^\text{ANC}}_I, \rho^\text{ANC}_{I}), \\
+y^{\alpha^\text{ANC}}_I &\sim \text{Bin}(y^{\rho^\text{ANC}}_I, \alpha^\text{ANC}_{I}).
+\end{align*}
+As in the household survey component, we use weighted aggregates
+\begin{equation*}
+\rho^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in I} \Psi_i}, \quad
+\alpha^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}},
+\end{equation*}
+with $\Psi_i$ the number of pregnant women, which we assume to be fixed.
+
+## ART attendance component \label{sec:art}
+
+People living with HIV sometimes choose to access ART services outside of the district that they reside in.
+To account for this, we model the probabilities of accessing services outside the home district using multinomial logistic regressions.
+Briefly, let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring such that $x \sim x'$.
+We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$.
+As such, we assume travel to each neighbouring district, for all age-sex strata, is equally likely.
+We then model aggregate ART attendance data $y^{N^\text{ART}}_I$ using a Gaussian approximation to a sum of binomials.
+This sum is over both strata $i \in I$ and the number of ART clients travelling from district $x'$ to $x$.
+More details regarding this part of the model are provided in Appendix \ref{sec:math}.
+
+## Summary
+
+In all, Naomi is a joint model on the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
+The structured additive predictors contain intercept effects, age random effects, and spatial random effects which we collectively describe as the latent field $\x$.
+The latent field is controlled by hyperparamters $\btheta$ which include standard deviations, first-order autoregressive model correlation parameters, and reparameterised Besag-York-Mollie model [BYM2; @simpson2017penalising] proportion parameters.
+
+# Extended Latent Gaussian models\label{sec:elgm}
+
+We now describe the latent Gaussian class of models, and an extension which encapsulates the complexities of Naomi.
+
+## Definitions
+
+Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models with likelihood
+\begin{align*}
+y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
+\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
+\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
+\end{align*}
+where $[n] = \{1, \ldots, n\}$.
+The response variable is $\y = (y)_{i \in [n]}$ with likelihood $p(\y \, | \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1)$, where $\bmeta = (\eta)_{i \in [n]}$.
+Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
+The vector $\btheta_1 \in \mathbb{R}^{s_1}$, with $s_1$ assumed small, are additional parameters of the likelihood.
+The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
+The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
+It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-1})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
+Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.
+
+Extended latent Gaussian models [ELGMs; @stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
+Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
+Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
+The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
+Importantly, this mapping allows for the presence of non-linearity in the model.
+Put together, ELGMs are then of the form
+\begin{align*}
+y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
+\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
+\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n],
+\end{align*}
+with latent field and hyperparameter priors as in the LGM case.
+
+## Naomi as an ELGM
+
+Naomi is a spatio-temporal model with a large Gaussian latent field, governed by a smaller number of hyperparameters.
+However, it is an ELGM rather than an LGM, for the reasons below.
+Note that when dependence on a specific number of structured additive predictors is given, it is for that factor in isolation, and as such should be considered illustrative.
+
+1.
+In the household survey component, HIV incidence depends on district-level adult HIV prevalence and ART coverage.
+This reflects basic HIV epidemiology: HIV incidence is proportional to unsuppresed viral load such that such that $\lambda \propto \rho (1 - \omega \cdot \alpha)$, with $\omega = 0.7$ a fixed constant.
+As a result, each $\log(\lambda_i)$ depends on 28 structured additive predictors (where 28 arises from the product of 2 sexes [male and female], 7 age groups, [$\{\text{15-19}, \ldots, \text{45-49}\}$], and 2 indicators [HIV prevalence and ART coverage]).
+2. In the household survey component, HIV incidence and HIV prevalence are linked to the proportion recently infected via Equation \ref{eq:kappa}.
+3.
+In the ANC testing component, HIV prevalence and ART coverage depend upon the respective indicators in the household survey component.
+Though $\text{logit}(\rho_i)$ and $\text{logit}(\alpha_i)$ are Gaussian, this nonetheless introduces dependence of each mean response on two structured additive predictors.
+4.
+Throughout the model components, processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations aggregated over sets of indices $i \in I$.
+As such, all observations are related to $|I|$ structured additive predictors.
+5. Individuals taking ART, or who have been recently infected, must be HIV positive. 
+6.
+The ART attendance component uses a multinomial model with softmax link function which takes as input $|\{x': x' \sim x\}| + 1$ structured additive predictors, one for each neighbouring district plus one for remaining in the home district.
+7. Multiple link functions are used throughout the model, such that there is no one inverse link function $g$.
+
+# Inference methods for Naomi \label{sec:inferencenaomi}
+
+We first describe (Section \ref{sec:inferenceelgm}) the inference method for ELGMs based on nested applications of the Laplace approximation and AGHQ used by @stringer2022fast.
+We then propose (Section \ref{sec:pca}) an extension of the method which uses PCA to facilitate inference for Naomi, which otherwise would be intractable.
+
+## Inference for ELGMs\label{sec:inferenceelgm}
+
+The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ in an ELGM is given by
+\begin{equation*}
+  p(\x, \btheta \, | \, \y)
+  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
+\end{equation*}
+We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
+\begin{align}
+  p(x_i \, | \, \y) &\approx \tilde p(x_i \, | \, \y) = \int \tilde p(x_i \, | \, \btheta, \y) \tilde p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
+  p(\theta_j \, | \, \y) &\approx \tilde p(\theta_j \, | \, \y) = \int \tilde  p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
+\end{align}
+
+### Laplace approximation\label{sec:la}
+
+Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\Hb}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+\begin{align}
+\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \label{eq:mode} \\
+\hat {\Hb}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}. \label{eq:precision}
+\end{align}
+Then the Laplace approximation [@tierney1986accurate] to $p(\btheta, \y)$ is given by
+\begin{equation}
+\tilde p_\texttt{LA}(\btheta, \y)
+= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+= \sqrt{\frac{ \lvert \hat {\Hb}(\btheta) \rvert}{(2 \pi)^{N}}} p(\y, \hat \x(\btheta), \btheta). \label{eq:la}
+\end{equation}
+Inference proceeds by optimising Equation \ref{eq:la} using a gradient-based routine to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
+Each evaluation in the optimisation requires an inner optimisation to obtain $\hat \x(\btheta)$ via Equation \ref{eq:mode}.
+Supposing the hyperparameters are to be considered fixed, as with the TMB approach used currently for Naomi, then latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
+Hyperparameter inferences can be obtained according to some method which should be specified here as well.
+
+### Adaptive Gauss-Hermite quadrature
+
+Let $\z \in \mathcal{Q}(m, k)$ be $m$-dimensional Gauss-Hermite quadrature [GHQ; @davis1975methods] rule with $k$ nodes per dimension constructed using the product rule such that $\mathcal{Q}(m, k) = \mathcal{Q}(1, k) \times \cdots \times \mathcal{Q}(1, k)$ where
+\begin{equation}
+\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}, \\
+\end{equation}
+with $\phi(\cdot)$ is a standard Gaussian density.
+The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$.
+GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.
+
+Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature.
+An adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] estimate of the normalising constant $p(\y)$ based on the Laplace approximation is given by
+\begin{equation}
+p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z).
+\label{eq:aghq}
+\end{equation}
+When $k = 1$ Equation \ref{eq:aghq} corresponds to a Laplace approximation.
+The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
+Repositioning the nodes is crucial for statistical quadrature problems like ours, where the integral depends on data $\y$ and regions of high density are not known in advance.
+Two alternatives for the matrix decomposition [@jackel2005note] are
+(1) the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
+(2) the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
+This estimate may be used to normalise the Laplace approximation
+\begin{equation}
+\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
+\end{equation}
+To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
+\begin{equation}
+\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z).
+\label{eq:nest}
+\end{equation}
+Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
+
+## Principal components analysis\label{sec:pca}
+
+```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension (A).
+Adaption occurs based on the mode and covariance matrix of the target via the Cholesky decomposition of the inverse curvature at the mode (B).
+In PCA-AGHQ (C) only nodes along the first $s$ principal components are kept.
+Here, 95\\% of variation is explained by the first principal component.
+The integrand is $f(\\btheta) = \\text{sn}(0.5 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.8 \\theta_1 - 0.5 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
+knitr::include_graphics("figA.png")
+```
+
+Use of the product rule grid described above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points.
+This quickly becomes intractable as $m$ increases for $k > 1$.
+An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
+We may then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ to be a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
+Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
+For example, for $m = 2$ and $s = 1$ then $\bk = (k, 1)$.
+In combination with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ.
+We refer to this approach as PCA-AGHQ, with corresponding estimate of the normalising constant given by
+\begin{equation}
+\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
+\end{equation}
+where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
+Panel C of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
+As AGHQ with $k = 1$ corresponds to the Laplace approximation, PCA-AGHQ can be interpreted as performing AGHQ on the first $s$ principal components of the inverse curvature, and a Laplace approximation on the remaining $m - s$ principal components.
+Inference for the latent field follows analogously to Equation \ref{eq:nest}.
+
+# Application to data from Malawi\label{sec:results}
+
+```{r}
+#' It's a shame here that I have to put "seconds", "hours" and "days" manually and it's not stored in the time taken
+time_taken <- readr::read_csv("depends/time-taken.csv")
+```
+
+We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
+These were:
+
+1. TMB (`r signif(time_taken$TMB, 2)` seconds), based on a Gaussian approximation at $\hat{\btheta}_\texttt{LA}$.
+2.
+PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), based on a Gaussian approximation mixture at the adapted nodes $\z \in \mathcal{Q}(m, s, k)$, as described in Section \ref{sec:pca}, and implemented via extension of the `aghq` package [@stringer2021implementing].
+3.
+NUTS (`r signif(time_taken$tmbstan, 2)` days), the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling using Stan [@carpenter2017stan] implemented via the `tmbstan` package [@monnahan2018no].
+
+Our goal was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
+Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
+The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
+The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
+For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
+For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
+To provide intuition, model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
+
+```{r naomi-results, fig.cap="District-level HIV prevalence (A), ART coverage (B), and new HIV cases and HIV incidence (C) for adults 15-49 in Malawi. Inference conducted using TMB."}
+knitr::include_graphics("figB.png")
+```
+
+The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/naomi-aghq`.
+We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
+
+\begin{table}[]
+\small
+\begin{tabularx}{\textwidth}{p{0.15\linewidth}p{0.15\linewidth}p{0.6\linewidth}}
+\toprule
+Method & Software & Details \\
+\midrule
+TMB & \texttt{TMB} & $1000$ samples \\
+PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-use}), $1000$ samples \\
+NUTS & \texttt{tmbstan} & $4$ chains of $100000$ iterations, with the first $50000$ iterations of each chain discarded as warmup, thinned by a factor of $40$, to give a total of $5000$ samples kept.
+Default NUTS tuning parameters \citep{hoffman2014no}.
+\\
+\bottomrule
+\end{tabularx}
+\caption{A summary of settings used for each inferential method.}
+\label{tab:inference-methods}
+\small
+\end{table}
+
+## NUTS convergence
+
+```{r}
+mcmc_out <- readRDS("depends/mcmc-out.rds")
+# signif(mcmc_out$max_rhat, 4)
+```
+
+Due to low effective sample size ratios (Figure \ref{fig:ratio}), obtaining acceptable NUTS diagnostics required four chains run in parallel for 100000 iterations, thinned by a factor of 20 for ease-of-storage.
+There were no divergent transitions, and the largest potential scale reduction factor [@gelman1992inference; @vehtari2021rank] was $\hat R = 1.021$ (Figure \ref{fig:rhat}).
+We considered the NUTS results a gold-standard, though inaccuracies remain possible.
+For full details see Appendix \ref{sec:mcmc}.
+
+## Use of PCA-AGHQ \label{sec:pca-use}
+
+```{r}
+tv_df <- readr::read_csv("depends/total-variation.csv")
+```
+
+We used a Scree plot based on the spectral decomposition of $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})^{-1}$ to select the number of principal components to keep (Figure \ref{fig:scree}).
+We found that $s = 8$ principal components were sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
+This choice of $s$ gave a visually similar reduced rank approximation to the inverse curvature (Figure \ref{fig:reduced-rank}).
+
+```{r node-positions, fig.cap="The 6561 PCA-AGHQ node positions (green, rug plot) projected onto the hyperparameter marginal posteriors (grey, histogram) for each of the 24 hyperparameters.
+Some hyperparameters, such as \\texttt{logit\\_phi\\_alpha\\_x}, are well covered where as others, such as \\texttt{log\\_sigma\\_lambda\\_x}, are near only being covered by one unique node."}
+knitr::include_graphics("depends/nodes-samples-comparison.png")
+```
+
+### Visual inspection
+
+Overlaying the resulting $3^8 = 6561$ PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
+Though 12 does improve on the 8 naively obtained with a dense grid, there remained poorly covered hyperparameters.
+Coverage was associated with marginal standard deviation (Figure \ref{fig:nodes-quantiles-sd}).
+All constrained hyperparameters $\theta$ were transformed to the real line, using either a log ($\theta > 0$) or logit ($\theta \in [0, 1]$) transformation.
+As a result, marginal standard deviations for log transformed hyperparameters were systematically smaller than those which were logit transformed (Figure \ref{fig:marginal-sd}).
+
+### Interpreation of eigenvectors
+
+```{r}
+cor_df <- readr::read_csv("depends/ar1-bym2-cor.csv")
+
+summary_cor_df <- cor_df |>
+  group_by(type) |>
+  summarise(mean_cor = mean(abs(cor)))
+
+ar1_mean_cor <- signif(filter(summary_cor_df, type == "ar1")$mean_cor, 2)
+bym2_mean_cor <- signif(filter(summary_cor_df, type == "bym2")$mean_cor, 2)
+```
+
+The ordered eigenvectors correspond to the directions of greatest variation in the inverse curvature.
+The first and second eigenvectors each contain coupled AR1 standard deviation and correlation parameters (Figure \ref{fig:pc-loadings}).
+These parameters are weakly identified, and have high correlation in the the NUTS posterior pairs plot (Figure \ref{fig:rho_a}, average absolute correlation across all four pairs of `r ar1_mean_cor`).
+The reason why is that the same amount of variation can equally be explained by high standard deviation and high correlation or low standard deviation and low correlation.
+The BYM2 standard deviation and proportion parameters on the other hand are designed to be orthogonal, and as such did not display posterior correlation (Figure \ref{fig:alpha_x}, average absolute correlation across all four pairs of `r bym2_mean_cor`) or appear prominently in the eigenvectors.
+
+### Normalising constant estimation
+
+We assessed appropriateness of the quadrature grid by comparing the estimate of $\log p_\texttt{PCA}(\y)$ for a range of settings.
+Convergence in $\log p_\texttt{PCA}(\y)$ as $s$ and $k$ are increased may suggest a suitable grid has been reached.
+Appendix \ref{sec:normalising} shows those values which we could compute in a reasonable time (less than 24 hours using a high performance computing cluster).
+
+## Model assessment\label{sec:model-assessment}
+
+### Posterior contraction
+
+To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\psi)$ to the posterior variance $\sigma_\text{posterior}^2(\psi)$ via the posterior contraction [@schad2021toward]
+\begin{equation}
+c(\psi) = 1 - (\sigma_\text{posterior}^2(\psi) / \sigma_\text{prior}^2(\psi)),
+\end{equation}
+where $\psi$ is a model parameter.
+We found that (Figure \ref{fig:contraction}) something something.
+For greater interpretability, facet parameters in this plot according to model component.
+
+### Coverage
+
+We assessed the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.
+Let $\{\psi_i\}_{i = 1}^n$ be posterior marginal samples.
+
+## Inference comparison\label{sec:inf-comparison}
+
+We compared the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS for latent field parameters and model outputs.
+The metrics we used were (1) marginal point estimates, (2) marginal Kolmogorov-Smirnov and Anderson-Darling tests using the empirical cumulative distribution function (ECDF), (3) joint Pareto-smoothed importance sampling results, and (4) joint maximum mean discrepancy results.
+
+### Point estimates
+
+```{r}
+df_point <- readr::read_csv("depends/mean-sd.csv")
+
+df_point_pct <- df_point |>
+  ungroup() |>
+  group_by(indicator, type) |>
+  summarise(
+    rmse_diff = 100 * diff(rmse) / max(rmse),
+    mae_diff = 100 * diff(mae) / max(mae)
+  )
+
+rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate", type == "latent") |> pull(rmse) |> signif(2)
+rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate", type == "latent") |> pull(rmse) |> signif(2)
+rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate", type == "latent") |> pull(rmse_diff) |> signif(2)
+
+rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate", type == "latent") |> pull(rmse) |> signif(2)
+rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate", type == "latent") |> pull(rmse) |> signif(2)
+rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate", type == "latent") |> pull(rmse_diff) |> signif(0)
+```
+
+For the latent field, the root mean square error (RMSE) between posterior mean estimates from PCA-AGHQ and NUTS (`r rmse_ahgq_mean`) was `r abs(rmse_diff_mean)`% lower than that between TMB and NUTS (`r rmse_tmb_mean`).
+For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_ahgq_sd)` (PCA-AGHQ).
+These results, alongside those for the mean absolute error (MAE), are presented in Figure \ref{fig:mean-sd-latent}.
+
+These improvements did not transfer to the model outputs (Figure \ref{fig:mean-sd-output}).
+
+```{r mean-sd-latent, fig.cap="For the latent field PCA-AGHQ modestly improves estimation of the posterior mean, and substantially improves estimation of the posterior standard deviation, as compared with TMB."}
+knitr::include_graphics("depends/mean-sd-alt-latent.png")
+```
+
+```{r mean-sd-output, fig.cap="PCA-AGHQ doesn't do so well for the model outputs."}
+knitr::include_graphics("depends/mean-sd-alt-output.png")
+```
+
+### Distribution tests
+
+The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\omega) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\psi_i \leq \omega}$.
+See Figure \ref{fig:ks} for an example.
+Additionally, we found that KS test results were negatively correlated with ESS (Figure \ref{fig:ks-ess}).
+
+```{r ks, fig.cap="Example KS test for one parameter."}
+knitr::include_graphics("figC.png")
+```
+
+### Pareto-smoothed importance sampling
+
+Let $\{\bpsi_i\}_{i = 1}^n$ be joint posterior samples.
+Pareto-smoothed importance sampling [PSIS; @vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios used in importance sampling.
+Results for the PSIS analysis are pending.
+
+### Maximum mean discrepancy
+
+Let $\Psi^{1} = \{\bpsi^1_i\}_{i = 1}^n$ and $\Psi^2 = \{\bpsi^2_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
+The maximum mean discrepancy [MMD; @gretton2006kernel] can be empirically estimated by
+\begin{equation*}
+\text{MMD}(\Psi^1, \Psi^2) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi^1_i, \bpsi^1_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bpsi_i^1, \bpsi_j^2) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi^2_i, \bpsi^2_j)}.
+\end{equation*}
+We set $k(\bpsi^1, \bpsi^2) = \exp(-\sigma \lVert \bpsi^1 - \bpsi^2 \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
+As compared with NUTS, the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080).
+
+<!-- This should be calculated inline! -->
+
+## Case study on exceedance probabilities\label{sec:exceedance}
+
+```{r}
+exceedance <- readr::read_csv("depends/exceedance.csv")
+```
+
+### Meeting the second 90
+
+Ambitious fast-track targets for scaling up ART treatment have been developed by UNAIDS, with the goal of "ending the AIDS epidemic by 2030".
+Meeting the "90-90-90" fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have suppressed viral load.
+Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met, that is $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
+We found that for women both TMB and PCA-AGHQ underestimate these exceedance probabilities (Figure \ref{fig:exceedance}, first row).
+We hypothesise this discrepancy in accuracy by sex is related to interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.
+
+### Finding strata with high incidence
+
+Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as higher than 1% incidence per year.
+Naomi can be used to assess the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
+We found that both TMB and PCA-AGHQ overestimate these exceedance probabilities (Figure \ref{fig:exceedance}, second row).
+This is surprising, in that we expect inferences from NUTS to be more heavy-tailed than those from TMB or PCA-AGHQ.
+
+```{r exceedance, fig.cap="Though PCA-AGHQ was marginally better, both approximate inference methods were meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities.
+For the second 90 target the inaccuracy varied substantially by sex."}
+knitr::include_graphics("depends/exceedance.png")
+```
+
+# Discussion\label{sec:conclusions}
+
+We developed an approximate Bayesian inference algorithm, combining AGHQ with PCA, motivated by a challenging problem in small-area estimation of HIV indicators.
+For the simplified Naomi model in Malawi (Section \ref{sec:results}) we demonstrated the method to be more accurate at inferring posterior distributions of model parameters, across a broad range of metrics, than TMB, and substantially faster than NUTS.
+However, improvements in accuracy for model parameters did not translate into model outputs.
+Indeed, we found posterior exceedance probabilities (Section \ref{sec:exceedance}) from both TMB and PCA-AGHQ to have systematically inaccuracies, with the potential to meaningfully mislead policy.
+If possible, though not a desirable situation, it could be advisable to provide gold-standard NUTS results after the workshop has concluded.
+However, running NUTS for Naomi took days, in countries with 100s of districts it may simply not be feasible.
+
+PCA-AGHQ could be added to the Naomi web interface as an alternative to TMB.
+Analysts may then quickly iterate over model options using a fast inference approach, before switching to a more accurate approach once they are happy with the results.
+By selecting $s$ and $k$, PCA-AGHQ can be adjusted to suit the computational budget available. 
+We selected $s$ based on the Scree plot, and for the most part fixed $k = 3$.
+Whether it is preferable, for a given computational budget, to increase $s$ or increase $k$ is an open question.
+Further strategies, such as gradually lowering $k$ over the principal components, could also be considered.
+
+We hope that our work further encourages use of deterministic inference algorithms for ELGMs in applied settings, as well as methodological exploration of their accuracy and limitations.
+Among the ELGM-type structures of particular interest in spatial epidemiology are aggregated Gaussian process models [@nandi2020disaggregation] and evidence synthesis models [@amoah2020geostatistical].
+
+## Suggestions for future work
+
+### Improved quadrature grids for moderate dimensions
+
+We aimed to develop a quadrature grid which allocates more effort to more important dimensions.
+While PCA is a sensible approach, there are avenues where it does not behave as one might hope, or otherwise overlooks potential benefits.
+The first challenge we identified was using PCA when the dimensions have different scales. 
+Specifically, we found logit-scale hyperparamters to be systematically favoured over those on the log-scale.
+Second, the amount of variation explained for the Hessian matrix is not of directly interest, rather the effect of the different dimensions on the relevant outputs.
+Using measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] may be preferable.
+Third, it is more important to allocate quadrature nodes to those marginals which are non-Gaussian.
+This is because the Laplace approximation is exact when the integrand is Gaussian, so a single quadrature node is sufficiently.
+The difficulty is, of course, knowing in advance which marginals will be non-Gaussian.
+This could be done if there were a cheap way to obtain posterior means, which could then be compared to posterior modes obtained using optimisation.
+Another approach would be to measure the fit of marginal samples from a cheap approximation, like TMB.
+The main challenge is that the measurements have to for marginals, ruling out approaches like PSIS which operate on joint distributions [@yao2018yes].
+
+### Computational speed-ups
+
+Integration over a moderate number of hyperparameters posed a challenge, and led us to use a quadrature grids with a large number of nodes.
+However, computation at each node is independent, such that the run-time of the algorithm could potentially be significantly improved by parallel computing.
+Further computational speed-ups might be obtained using graphics processing units (GPUs) speciallised for the relevant matrix operations.
+
+### Comparison to other MCMC algorithms
+
+Blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], may be better suited than NUTS to sampling from Naomi.
+These algorithms are available, and customisable, including e.g. choice of block structure within the `NIMBLE` probabilistic programming language [@de2017programming].
+
+### Implementation into probabilistic programming languages
+
+Though gaining in popularity, the user-base of `TMB` remains relatively small.
+Furthermore, for users unfamiliar with C++, it can be challenging to use.
+As such, it could be beneficial to implement AGHQ within other probabilistic programming languages.
+Implementation in `NIMBLE` could be relatively straightforward, as it (for version >1.0.0) includes functionality for automatic differentiation and Laplace approximation, built using `CppAD` like `TMB`.
+Similarly, implementation in Stan could be possible by use of the `bridgestan` package [@bridgestan] together with the adjoint-differentiated Laplace approximation of @margossian2020hamiltonian.
+<!-- Note: can fit hyperparameters with Laplace on latent field using HMC via tmbstan::tmbstan(laplace = TRUE) -->
+<!-- What Charles Margossian has implemented for Stan is done similarly by default in TMB -->
+
+### Statistical theory
+
+@stringer2022fast (Theorem 1) bound the total variation error of AGHQ, establishing convergence in probability of coverage probabilities under the approximate posterior to those under the true posterior.
+Similar theory might be established for PCA-AGHQ, or more generally AGHQ with varying numbers of nodes per dimension.
+The challenge of connecting this theory to use of the quadrature rule within nested compuations remains an open question.
+
+# Acknowledgements {-}
+
+AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1), and conducted part of this research while an International Visiting Graduate Student at the University of Waterloo.
+AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).
+SRF was supported by the EPSRC (EP/V002910/2).
+JWE was supported by UNAIDS and National Institute of Allergy and Infectious Disease of the National Institutes of Health (R01AI136664).
+This research was supported by the MRC Centre for Global Infectious Disease Analysis (MR/R015600/1), jointly funded by the UK Medical Research Council (MRC) and the UK Foreign, Commonwealth \& Development Office (FCDO), under the MRC/FCDO Concordat program and is also part of the EDCTP2 programme supported by the European Union.
+
+<!-- # Graveyard -->
+
+<!-- ## Sparse rules -->
+
+<!-- Though there are sparse rules which retain the attractive exactness properties of GHQ, using fewer than $k^m$ quadrature points, they use weighting functions which may be negative $\omega(\z) < 0$.
+-->
+<!-- This introduces a problem when trying to produce inferences about the latent field, as the resulting $\lambda(\z)$ may be negative. -->
+<!-- We are not aware of any suitable approaches to obtain multinomial samples in such cases. -->
+<!-- It is however, still possible to calculate the normalising constant using a sparse rule, which we do so in Appendix S3, finding it to be related to that from PCA-AGHQ as follows. -->
+
+<!-- Pruning (Jackel, 2005): is this something we could be doing? -->
+
 # References {-}

### 2024-12-07 src/docs_paper-jtb/paper.Rmd
Cleaning up draft

--- 2024-12-07 src/docs_paper-jtb/paper.Rmd
@@
 abstract: |
-  This is the abstract.
+  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa.
+Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data.
+The model is provided as a tool for countries to input their data to and generate estimates with during a yearly process supported by UNAIDS.
+Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
+We propose a new inference method based on an extension of adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters.
+Using data from Malawi, our method improves the accuracy of inferences for model parameters, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
+Our implementation is based on the existing \texttt{TMB} \textsf{C++} template for the model's log-posterior, and is compatible with any model with such a template.
 keywords: 
@@
     citation_package: natbib
+    includes:
+      in_header: preamble.tex
 ---
 
+```{r echo = FALSE, message = FALSE}
+library(dplyr)
+
+options(scipen = 100)
+
+knitr::opts_chunk$set(
+  echo = FALSE,
+  warning = FALSE,
+  message = FALSE,
+  dpi = 320,
+  cache = TRUE,
+  out.width = "95%",
+  fig.align = 'center'
+)
+```
+
+# Introduction
+
+Accurate estimates of HIV indicators are crucial for mounting an effective public health response to the epidemic.
+Producing timely estimates at the district level, where health systems are planned and delivered, is challenging.
+Nationally-representative household surveys, while providing the most statistically reliable data, are costly to run and typically conducted only every five years.
+Furthermore, sample sizes at the district level are limited.
+Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are available nearer to real-time, but lack population representativeness.
+
+The Naomi small-area estimation model [@eaton2021naomi] addresses these challenges by synthesising data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
+Modelling multiple data sources jointly mitigates the limitations of any single source and increases statistical power.
+Software developed for Naomi (\url{https://naomi.unaids.org}) facilitates over 35 countries inputting their data and interactively generating estimates during workshops as a part of a yearly estimates process supported by UNAIDS.
+Generation of estimates by country teams is an important and distinctive feature of the HIV response.
+Drawing on expert knowledge of the data improves the accuracy of the process, as well as strengthening trust in the resulting estimates, creating a virtuous cycle of data quality, use and ownership [@noor2022country].
+
+Naomi is comprised of multiple linked generalized linear mixed models (GLMMs) and presents a challenging Bayesian inference problem.
+The model contains hundreds of fixed and random effect parameters and over 20 hyperparameters.
+This is substantially more than the small number that can typically be handled by approaches like integrated nested Laplace approximations [INLA; @rue2009approximate].
+Moreover, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
+Inference must be fast and have low memory usage, to allow interactive review and iteration of model results by workshop participants.
+The scale of the model and features of its posterior geometry [@neal2003slice] make Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
+Inference must be reliable and automatic across a range of country settings, and shouldn't require close monitoring and diagnostics (e.g. for MCMC convergence).
+
+Inference for Naomi is currently conducted using an empirical Bayes (EB) approach.
+A Gaussian approximation to the latent field is used via the Template Model Builder (`TMB`) [@kristensen2016tmb] \textsf{R} package.
+`TMB` is gaining popularity due to its speed and flexibility, especially in spatial statistics [@osgood2022statistical] and via the user-friendly `glmmTMB` \textsc{R} package [@brooks2017glmmTMB].
+Inference in `TMB` occurs via optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the parameters.
+`TMB` uses automatic differentiation [@fournier2012ad; @baydin2017automatic] to calculate the derivatives required for gradient-based numerical optimisation routines and the Laplace approximation.
+
+Although the EB approach is fast, it does not take into account hyperparameter uncertainty in the latent field posterior, potentially resutling in underestimated posterior variances.
+This concern could have important practical implications for use of the estimates from the Naomi model, and motivated us to look for an approach closer full Bayesian inference.
+We developed a method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over many hyperparameters.
+AGHQ is a quadrature method based on the theory of polynomial interpolation, and is well suited to statistical estimation problems in which the integrand is well approximated by a Gaussian multiplied by a polynomial.
+For example, @bilodeau2022stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
+However, it is not computationally feasible to use AGHQ in >20 dimensions directly, as exponentially many nodes are required.
+Instead, we used principal components analysis (PCA) of the inverse curvature at the mode to find a subspace which explained most of the hyperparameter variance.
+In an application to Malawi, this resulted in a grid which was tractable as it had millions of times fewer nodes the corresponding dense grid.
+Our implementation of the method makes use of the existing Naomi `TMB` template, and is immediately compatible with any model with such a template.
+
+<!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
+<!-- Want to say in particular what the class of models that `inlabru` allows you to fit are -->
+<!-- Don't know the answer to this yet, but asking Janine Illian I'm told you don't have to call R-INLA that many times -->
+<!-- Other work aiming to extend the scope of INLA-like methods includes the `inlabru` \textsc{R} package [@bachl2019inlabru], INLA within MCMC [@gomez2018markov], and importance sampling with INLA [@berild2022importance], all of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
+-->
+<!-- The approach of `inlabru` is to approximate non-linear predictors using linearisation, by making iterative calls to `R-INLA`. -->
+<!-- INLA within MCMC and importance sampling with INLA are suitable for models which are LGMs conditional on some subset of the parameters being fixed. -->
+
+The remainder of this paper is organised as follows.
+In Section \ref{sec:background} we give background on latent Gaussian and extended latent Gaussian models.
+Section \ref{sec:naomi} outlines the version of the Naomi model that we consider in this paper.
+In Section \ref{sec:inferencenaomi} we review the deterministic inference method for ELGMs used by @stringer2022fast based on nested application of AGHQ and the Laplace approximation, before introducing the PCA-based modification we use to enable application to Naomi.
+In Section \ref{sec:results} we evaluate the accuracy of PCA-AGHQ for the simplified Naomi model fit to data from Malawi, as compared with TMB and gold-standard MCMC.
+Finally, we discuss our conclusions, and directions for future research in Section \ref{sec:conclusions}.
+
+# Background\label{sec:background}
+
+## Latent Gaussian model
+
+Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models with
+\begin{align*}
+y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
+\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
+\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
+\end{align*}
+where $[n] = \{1, \ldots, n\}$.
+The response variable is $\y = (y)_{i \in [n]}$ with likelihood $p(\y \, | \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1)$, where $\bmeta = (\eta)_{i \in [n]}$.
+Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
+The vector $\btheta_1 \in \mathbb{R}^{s_1}$, with $s_1$ assumed small, are additional parameters of the likelihood.
+The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
+The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
+It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-1})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
+Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.
+
+## Extended latent Gaussian model
+
+Extended latent Gaussian models [ELGMs; @stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
+Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
+Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
+The inverse link function $g$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
+Importantly, this mapping allows for the presence of non-linearity in the model.
+Put together, ELGMs are then of the form
+\begin{align*}
+y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
+\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
+\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n],
+\end{align*}
+with latent field and hyperparameter priors as in the LGM case.
+
+# Simplified Naomi model\label{sec:naomi}
+
+```{r naomi-results, fig.cap="District-level HIV prevalence (A), ART coverage (B), and new HIV cases and HIV incidence (C) for adults 15-49 in Malawi. Inference conducted using EB."}
+knitr::include_graphics("figB.png")
+```
+
+@eaton2021naomi introduce a joint ELGM linking three small-area estimation models.
+We consider a simplified version defined only at the time of the most recent household survey with HIV testing.
+While this version omits nowcasting and temporal projection, as these time points involve limited inference we expect conclusions for the simplified model to be applicable to the complete model.
+An overview of the simplified model is given below, and Appendix \ref{sec:math} provides a full specification.
+
+## Notation
+
+Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place.
+Let $x \in \mathcal{X}$ index district, $a \in \mathcal{A}$ index five-year age group, and $s \in \mathcal{S}$ index sex.
+For ease of notation, let $i$ index the finest district-age-sex division included in the model.
+Let $I \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported, and $\mathcal{I}$ be the set of all $I$ such that $I \in \mathcal{I}$.
+
+Let $N_i \in \mathbb{N}$ be the known, fixed population size, $\rho_i \in [0, 1]$ be the HIV prevalence, $\alpha_i \in [0, 1]$ be the antiretroviral therapy (ART) coverage, $\kappa_i \in [0, 1]$ be the proportion recently infected among HIV positive persons, and $\lambda_i > 0$ be the annual HIV incidence rate.
+
+Some observations are made at an aggregate level over a collection of strata $i$ rather than for a single $i$.
+Let $I \subseteq \mathcal{X} \times \mathcal{A} \times \mathcal{S}$ be a set of indices $i$ for which an aggregate observation is reported.
+The set of all $I$ is denoted $\mathcal{I}$ such that $I \in \mathcal{I}$.
+
+Naomi is a joint model on the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
+The structured additive predictors contain intercept effects, age random effects, and spatial random effects which we collectively describe as the latent field $\x$.
+The latent field is controlled by hyperparamters $\btheta$ which include standard deviations, first-order autoregressive model correlation parameters, and reparameterised Besag-York-Mollie model [BYM2; @simpson2017penalising] proportion parameters.
+
+## Household survey component \label{sec:household}
+
+Independent logistic regression models are specified for HIV prevalence and ART coverage in the general population such that $\text{logit}(\rho_i) = \eta^\rho_i$ and $\text{logit}(\alpha_i) = \eta^\alpha_i$.
+HIV incidence rate is modelled by $\log(\lambda_i) = \eta^\lambda_i$ and depends on adult HIV prevalence and adult ART coverage.
+The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix \ref{sec:math}.
+The proportion recently infected $\kappa_i$ is linked to HIV incidence via
+\begin{equation}
+\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right), \label{eq:kappa}
+\end{equation}
+where the mean duration of recent infection $\Omega_T$ and the proportion of long-term HIV infections misclassified as recent $\beta_T$ are strongly informed by priors for the particular survey.
+
+These processes are each informed by household survey data.
+Weighted aggregate survey observations are calculated based on individual responses $\theta_j \in \{0, 1\}$ as
+\begin{equation*}
+\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}.
+\end{equation*}
+Design weights $w_j$ for each of $\theta \in \{\rho, \alpha, \kappa\}$ are supplied by the survey provider and aim to reduce bias by decreasing possible correlation between response and recording mechanism [@meng2018statistical].
+The index $j$ runs across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
+We take the weighted observed number of outcomes to be $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
+\begin{equation*}
+m^{\theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
+\end{equation*}
+is the Kish effective sample size [ESS; @kish1965survey].
+The weighted aggregated number of outcomes are modelled using a binomial working likelihood [@chen2014use] defined to operate on the reals
+\begin{equation*}
+y^{\theta}_{I} \sim \text{xBin}(m^{\theta}_{I}, \theta_{I})
+\end{equation*}
+The terms $\theta_{I}$ are the following weighted aggregates
+\begin{equation*}
+\rho_{I} = \frac{\sum_{i \in I} N_i \rho_i}{\sum_{i \in I} N_i}, \quad
+\alpha_{I} = \frac{\sum_{i \in I} N_i \rho_i \alpha_i}{\sum_{i \in I} N_i \rho_i}, \quad
+\kappa_{I} = \frac{\sum_{i \in I} N_i \rho_i \kappa_i}{\sum_{i \in I} N_i \rho_i}.
+\end{equation*}
+
+## ANC testing component \label{sec:anc}
+
+Women attending ANC clinics are routinely tested for HIV, to help prevent mother-to-child transmission.
+HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset from the general population indicators as follows
+\begin{align*}
+\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
+\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\alpha^\text{ANC}}_i.
+\end{align*}
+These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
+We take the number of ANC clients with ascertained status to be fixed as $m^{\rho^\text{ANC}}_I$. 
+We then model the number of those with positive status $y^{\rho^\text{ANC}}_I$, and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ using nested binomial likelihoods
+\begin{align*}
+y^{\rho^\text{ANC}}_I &\sim \text{Bin}(m^{\rho^\text{ANC}}_I, \rho^\text{ANC}_{I}), \\
+y^{\alpha^\text{ANC}}_I &\sim \text{Bin}(y^{\rho^\text{ANC}}_I, \alpha^\text{ANC}_{I}).
+\end{align*}
+As in the household survey component, we use weighted aggregates
+\begin{equation*}
+\rho^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in I} \Psi_i}, \quad
+\alpha^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}},
+\end{equation*}
+with $\Psi_i$ the number of pregnant women, which we assume to be fixed.
+
+## ART attendance component \label{sec:art}
+
+Data on attendance of ART clinics are routinely collected.
+These data provide helpful information about HIV prevalence and coverage of ART, but are challenging to use because people living with HIV sometimes choose to access ART services outside of the district that they reside in.
+We model the probabilities of accessing services outside the home district using multinomial logistic regressions.
+
+Briefly, let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring such that $x \sim x'$.
+We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$.
+As such, we assume travel to each neighbouring district, for all age-sex strata, is equally likely.
+We then model aggregate ART attendance data $y^{N^\text{ART}}_I$ using a Gaussian approximation to a sum of binomials.
+This sum is over both strata $i \in I$ and the number of ART clients travelling from district $x'$ to $x$.
+
+# Inference methods for Naomi \label{sec:inferencenaomi}
+
+Section \ref{sec:inferenceelgm} gives the inference method for ELGMs of @stringer2022fast based on nested applications of the Laplace approximation and AGHQ.
+In Section \ref{sec:pca} we propose an extension of the method which uses PCA to facilitate inference for Naomi, which otherwise would be intractable.
+
+## Inference for ELGMs\label{sec:inferenceelgm}
+
+The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ in an ELGM is given by
+\begin{equation*}
+  p(\x, \btheta \, | \, \y)
+  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
+\end{equation*}
+We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
+\begin{align}
+  p(x_i \, | \, \y) &\approx \tilde p(x_i \, | \, \y) = \int \tilde p(x_i \, | \, \btheta, \y) \tilde p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
+  p(\theta_j \, | \, \y) &\approx \tilde p(\theta_j \, | \, \y) = \int \tilde  p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
+\end{align}
+
+### Laplace approximation\label{sec:la}
+
+Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\Hb}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+\begin{align}
+\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \label{eq:mode} \\
+\hat {\Hb}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}. \label{eq:precision}
+\end{align}
+Then the Laplace approximation [@tierney1986accurate] to $p(\btheta, \y)$ is given by
+\begin{equation}
+\tilde p_\texttt{LA}(\btheta, \y)
+= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
+= \sqrt{\frac{ \lvert \hat {\Hb}(\btheta) \rvert}{(2 \pi)^{N}}} p(\y, \hat \x(\btheta), \btheta). \label{eq:la}
+\end{equation}
+Inference proceeds by optimising Equation \ref{eq:la} using a gradient-based routine to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
+Each evaluation in the optimisation requires an inner optimisation to obtain $\hat \x(\btheta)$ via Equation \ref{eq:mode}.
+Supposing the hyperparameters are to be considered fixed, as with the TMB approach used currently for Naomi, then latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
+
+### Adaptive Gauss-Hermite quadrature
+
+Let $\z \in \mathcal{Q}(m, k)$ be $m$-dimensional Gauss-Hermite quadrature [GHQ; @davis1975methods] rule with $k$ nodes per dimension constructed using the product rule $\mathcal{Q}(m, k) = \mathcal{Q}(1, k) \times \cdots \times \mathcal{Q}(1, k)$ where
+\begin{equation}
+\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}, \\
+\end{equation}
+with $\phi(\cdot)$ is a standard Gaussian density.
+The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$.
+GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.
+
+Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature.
+An adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] estimate of the normalising constant $p(\y)$ based on the Laplace approximation is given by
+\begin{equation}
+p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z).
+\label{eq:aghq}
+\end{equation}
+When $k = 1$ Equation \ref{eq:aghq} corresponds to a Laplace approximation.
+The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
+Repositioning the nodes is crucial for statistical quadrature problems like ours, where the integral depends on data $\y$ and regions of high density are not known in advance.
+Two alternatives for the matrix decomposition [@jackel2005note] are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
+This estimate may be used to normalise the Laplace approximation
+\begin{equation}
+\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
+\end{equation}
+To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
+\begin{equation}
+\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z).
+\label{eq:nest}
+\end{equation}
+Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
+
+## Principal components analysis\label{sec:pca}
+
+```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension (A).
+Adaption occurs based on the mode and covariance matrix of the target via the Cholesky decomposition of the inverse curvature at the mode (B).
+In PCA-AGHQ (C) only nodes along the first $s$ principal components are kept.
+Here, 95\\% of variation is explained by the first principal component.
+The integrand is $f(\\btheta) = \\text{sn}(0.5 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.8 \\theta_1 - 0.5 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
+knitr::include_graphics("figA.png")
+```
+
+Use of the product rule grid described above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points which quickly becomes intractable as $m$ increases for $k > 1$.
+An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
+We may then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ to be a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
+Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
+For example, for $m = 2$ and $s = 1$ then $\bk = (k, 1)$.
+In combination with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ.
+We describe this approach as PCA-AGHQ, with corresponding estimate of the normalising constant given by
+\begin{equation}
+\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
+\end{equation}
+where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
+Panel C of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
+As AGHQ with $k = 1$ corresponds to the Laplace approximation, PCA-AGHQ can be interpreted as performing AGHQ on the first $s$ principal components of the inverse curvature, and a Laplace approximation on the remaining $m - s$ principal components.
+Inference for the latent field follows analogously to Equation \ref{eq:nest}.
+
+# Application to data from Malawi\label{sec:results}
+
+```{r}
+#' It's a shame here that I have to put "seconds", "hours" and "days" manually and it's not stored in the time taken
+time_taken <- readr::read_csv("depends/time-taken.csv")
+```
+
+We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi.
+Data from Malawi has previously been used to demonstrate the Naomi model, including as a part of the `naomi` R package vignette.
+Malawi was chosen in part because it has a small number of districts, $n = 30$, limiting the computational demand of the model.
+Three Bayesian inference approaches were considered:
+
+1. Gaussian marginals and EB with `TMB`. This approach was previously used in production for Naomi. As short-hand, this approach is referred to as GEB.
+2. Gaussian marginals and PCA-AGHQ with `TMB`. This is a novel approach. As short-hand, this approach is referred to as GPCA-AGHQ.
+3.
+The Hamiltonian Monte Carlo algorithm No-U-Turn Sampling (NUTS) with `tmbstan` [@monnahan2018no].
+Conditional on assessing chain convergence and suitability, inferences from NUTS represent a gold-standard.
+
+\begin{table}[]
+\small
+\begin{tabularx}{\textwidth}{p{0.15\linewidth}p{0.15\linewidth}p{0.6\linewidth}}
+\toprule
+Method & Software & Details \\
+\midrule
+TMB & \texttt{TMB} & $1000$ samples \\
+PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-use}), 1000 samples \\
+NUTS & \texttt{tmbstan} & 4 chains of 100000 iterations, with the first 50000 iterations of each chain discarded as warmup, thinned by a factor of 20, to give a total of 10000 samples kept.
+Default NUTS tuning parameters \citep{hoffman2014no}.
+\\
+\bottomrule
+\end{tabularx}
+\caption{A summary of settings used for each inferential method.}
+\label{tab:inference-methods}
+\small
+\end{table}
+
+Our goal was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
+Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
+The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
+The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
+For GEB and GPCA-AGHQ, hyperparameter and latent field samples were simulated following deterministic inference.
+For all methods, age-sex-district specific HIV prevalence, ART coverage and HIV incidence were simulated from the latent field and hyperparameter posteriors.
+
+The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/naomi-aghq`.
+We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
+
+# Conclusion\label{sec:conclusions}
+
+# Acknowledgements {-}
+
+AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1), and conducted part of this research while an International Visiting Graduate Student at the University of Waterloo.
+AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).
+SRF was supported by the EPSRC (EP/V002910/2).
+JWE was supported by UNAIDS and National Institute of Allergy and Infectious Disease of the National Institutes of Health (R01AI136664).
+This research was supported by the MRC Centre for Global Infectious Disease Analysis (MR/R015600/1), jointly funded by the UK Medical Research Council (MRC) and the UK Foreign, Commonwealth \& Development Office (FCDO), under the MRC/FCDO Concordat program and is also part of the EDCTP2 programme supported by the European Union.
+
 # References {-}

### 2024-12-07 src/docs_paper-jtb/appendix.Rmd
Adding in appendix and making progress

@@
   - Seth R. Flaxman^[Department of Computer Science, Oxford University]
-  - Jeffrey W. Eaton^[Harvard T.H. Chan School of Public Health, Harvard University]
+  - Jeffrey W. Imai-Eaton^[Harvard T.H. Chan School of Public Health, Harvard University]
 output:
-  pdf_document:
+  bookdown::pdf_document2:
     toc: yes
@@
     keep_tex: yes
-    includes:
-      in_header: preamble.tex
 bibliography: citations.bib
-cls: imsart.cls
 header-includes:
@@
 
-\setcounter{table}{0}  
-\renewcommand{\thetable}{S\arabic{table}} 
-\setcounter{figure}{0} 
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{section}{0} 
-\renewcommand{\thesection}{S\arabic{section}}
+# Simplified Naomi model description {#naomi-math}
 
-\clearpage
+This section describes the simplified version of the Naomi model [@eaton2021naomi] in more detail.
+The concise $i$ indexing used in Section \@ref(naomi) is replaced by a more complete $x, s, a$ indexing.
+There are four sections:
 
-# Simplified Naomi model description \label{sec:math}
+1. Section \@ref(naomi-process) gives the process specifications, giving the terms in each structured additive predictor, along with their distributions.
+2. Section \@ref(naomi-likelihood) gives additional details about the likelihood terms not provided in Section \@ref(naomi).
+3. Section \@ref(naomi-identifiability) gives identifiability constraints used in circumstances where incomplete data is available for the country.
+4. Section \@ref(naomi-implementation) provides details of the `TMB` implementation.
 
-In this section we describe the simplified Naomi model [@eaton2021naomi] in complete detail.
+## Process specification {#naomi-process}
 
-## Background
+|                   | Model component    | Latent field | Hyperparameter |
+| ----------------- | ------------------ | ------------ | -------------- |
+| Section \@ref(hiv-prev)   | HIV prevalence     | $22 + 5n$    | 9              |
+| Section \@ref(art-cov)    | ART coverage       | $25 + 5n$    | 9              |
+| Section \@ref(hiv-inc)    | HIV incidence rate | $2 + n$      | 3              |
+| Section \@ref(anc-test)   | ANC testing        | $2 + 2n$     | 2              |
+| Section \@ref(art-attend) | ART attendance     | $n$          | 1              |
+|                   | Total              | $51 + 14n$   | 24             |
 
-### Indexing
+Table: (\#tab:process) The Naomi model can be conceptualised as having five processes.
+This table gives the number of latent field parameters and hyperparameters in each process, where $n$ is the number of districts in the country.
 
-Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
-Let $x \in \{1, \ldots, n\}$ be district, $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
-Each district is located within a correspodning Spectrum [@stover2010spectrum] region $R_x$.
-As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
+### HIV prevalence {#hiv-prev}
 
-We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities:
-
-* HIV prevalence $\rho_{x, s, a} \in [0, 1]$,
-* ART coverage $\alpha_{x, s, a} \in [0, 1]$,
-* annual HIV incidence rate $\lambda_{x, s, a} > 0$, and
-* proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
-
-Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
-We use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
-The notation $\sum_{\{x\}}$ is used as shorthand for $\sum_{x \in \{x\}}$.
-Analogous notion is used for sums over $\sum_{a \in \{a\}}$ and $\sum_{s \in \{s\}}$.
-
-### Structured random effects
-
-We use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age groups.
-Let $\bu$ be a generic random effect, with length $\dim(\bu) > 1$.
-Three types of structured random effects are used in the model:
-
-1. We specify the first order auto-regressive model by $\bu \sim \text{AR}1(\sigma, \phi)$ such that
-\begin{align}
-u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
-u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(\bu),
-\end{align}
-where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
-2. We use $\bu \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
+HIV prevalence $\rho_{x, s, a} \in [0, 1]$ was modelled on the logit scale using the structured additive predictor
 \begin{equation}
-u_i \, | \, \bu_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
+\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + \mathbf{u}^\rho_a + \mathbf{u}_a^{\rho, s = \text{M}} + \mathbf{u}^\rho_x + \mathbf{u}_x^{\rho, s = \text{M}} + \mathbf{u}_x^{\rho, a < 15} + \boldsymbol{\mathbf{\eta}}^\rho_{R_x, s, a}.
+(\#eq:prev)
 \end{equation}
-where $\bu_{-i}$ is $\bu$ with the $i$th element removed i.e.
-$(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(\bu)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
-We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
-3.
-For the reparameterised Besag-York-Mollie model [BYM2; @simpson2017penalising], we write $\bu \sim \text{BYM}2(\sigma, \phi)$, where $\bu$ is comprised of a spatially structured ICAR component $\bv$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $\w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
-\begin{equation}
-\bu = \sigma \left( \sqrt\phi \cdot \bv + \sqrt{1 - \phi} \cdot \w \right).
-\end{equation}
-
-### Complex survey design
-
-We assume the household survey was run according to a complex survey design where each individual $j$ in the population $U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$.
-Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
-Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean is given by
-\begin{equation}
-\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
-\end{equation}
-We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
-\begin{equation}
-m^{\theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
-\end{equation}
-The observed number of indicator cases is then $y^{\theta} = m^{\theta} \cdot \hat \theta$.
-To make these calculations we used the `survey` \textsc{R} package [@lumley2004analysis]
-
-## Process specification
-
-\begin{table}[h]
-\centering
-\begin{tabularx}{\textwidth}{llXX}
-\toprule
-& Model component & Latent field & Hyperparameter \\ 
-\midrule
-\ref{sec:hiv-prev} & HIV prevalence & $22 + 5n$ & 9 \\ 
-\ref{sec:art-cov} & ART coverage & $25 + 5n$ & 9 \\ 
-\ref{sec:hiv-inc} & HIV incidence rate & $2 + n$ & 3 \\ 
-\ref{sec:anc-test} & ANC testing & $2 + 2n$ & 2 \\ 
-\ref{sec:art-attend} & ART attendance & $n$ & 1 \\ 
-& Total & $51 + 14n$ & 24 \\ 
-\bottomrule
-\end{tabularx}
-\caption{The numer of latent field parameters and hyperparameters in each model section.}
-\label{tab:latent-hyper-size}
-\end{table}
-
-We now describe the hyperparameter and latent field process specification for the model.
-Whereas in the main text process and likelihood specifications are given together, here we consider the likelihood equations separately (Section \ref{sec:likelihood}).
-Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each component of the model.
-In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $51 + 14 \cdot 32 = 499$ and the total number of hyperparameters is $24$.
-
-<!-- 51 + 13n + 467 is the value I get from the code. 467 + 24 = 491 total -->
-<!-- Need to check the discrepency -->
-
-### HIV prevalence \label{sec:hiv-prev}
-
-We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
-\begin{equation}
-\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + \bu^\rho_a + \bu_a^{\rho, s = \text{M}} + \bu^\rho_x + \bu_x^{\rho, s = \text{M}} + \bu_x^{\rho, a < 15} + \bmeta^\rho_{R_x, s, a}.
-\label{eq:prev}
-\end{equation}
-Table \ref{tab:prev} provides a description of the terms included in Equation \ref{eq:prev}.
-
-\begin{table}[h]
-\centering
-\begin{tabularx}{\textwidth}{llX}
-\toprule
-Term & Distribution & Description \\ 
-\midrule
-$\beta^\rho_0$ & $\mathcal{N}(0, 5)$ & Intercept \\
-$\beta_{s}^{\rho, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The difference in logit prevalence for men compared to women \\
-$\bu^\rho_a$ & $\text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ & Age random effects for women \\
-$\bu_a^{\rho, s = \text{M}}$ & $\text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ & Age random effects for the difference in logit prevalence for men compared to women age $a$ \\
-$\bu^\rho_x$ & $\text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ & Spatial random effects for women \\
-$\bu_x^{\rho, s = \text{M}}$ & $\text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ & Spatial random effects for the difference in logit prevalence for men compared to women in district $x$  \\
-$\bu_x^{\rho, a < 15}$ & $\text{ICAR}(\sigma_{XA}^\rho)$ & Spatial random effects for the ratio of paediatric prevalence to adult women prevalence \\
-$\bmeta^\rho_{R_x, s, a}$ & $-$ & Fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available \\
-\bottomrule
-\end{tabularx}
-\caption{Terms included in the linear predictor for HIV prevalence (Equation \ref{eq:prev}).}
-\label{tab:prev}
-\end{table}
-
-The two BYM2 random effects $\bu^\rho_x$ and $\bu_x^{\rho, s = \text{M}}$ are comprised of the following unit scaled spatially structured $\{\bv_x^{\rho}, \bv_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{\w_x^{\rho}, \w_x^{\rho, s = \text{M}}\}$ components, respectively
-\begin{align}
-\bu_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot \bv_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot \w_x^{\rho} \right), \\
-\bu_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot \bv_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot \w_x^{\rho, s = \text{M}} \right).
-\end{align}
-We use half-normal priors for the standard deviation terms
+Table \@ref(tab:prev) provides a description of the terms included in Equation \@ref(eq:prev).
+Independent half-normal prior distributions were chosen for the five standard deviation terms
 \begin{equation}
@@
 \end{equation}
-uniform priors for the AR1 correlation parameters
+independent uniform prior distributions for the two AR1 correlation parameters
 \begin{equation}
@@
 \end{equation}
-and beta priors for the BYM2 proportion parameters
+and independent beta prior distributions for the two BYM2 proportion parameters
 \begin{equation}
@@
 
-<!-- Latent field size -->
-<!-- 1 + 1 + 10 + 10 + n + n + n = 36 + 3n -->
-<!-- Hyperparameter size -->
-<!-- 5 + 2 + 2 -->
+| Term                             | Distribution                                    | Description                                                                                                     |
+|----------------------------------|-------------------------------------------------|-----------------------------------------------------------------------------------------------------------------|
+| $\beta^\rho_0$                   | $\mathcal{N}(0, 5)$                             | Intercept                                                                                                       |
+| $\beta_{s}^{\rho, s = \text{M}}$ | $\mathcal{N}(0, 5)$                             | The difference in logit prevalence for men compared to women                                                    |
+| $\mathbf{u}^\rho_a$              | $\text{AR}1(\sigma_A^\rho, \phi_A^\rho)$        | Age random effects for women                                                                                    |
+| $\mathbf{u}_a^{\rho, s = \text{M}}$ | $\text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$  | Age random effects for the difference in logit prevalence for men compared to women age $a$                  |
+| $\mathbf{u}^\rho_x$              | $\text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$       | Spatial random effects for women                                                                                |
+| $\mathbf{u}_x^{\rho, s = \text{M}}$ | $\text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ | Spatial random effects for the difference in logit prevalence for men compared to women in district $x$      |
+| $\mathbf{u}_x^{\rho, a < 15}$    | $\text{ICAR}(\sigma_{XA}^\rho)$                 | Spatial random effects for the difference in logit paediatric prevalence to adult women prevalence in district $x$ |
+| $\boldsymbol{\mathbf{\eta}}^\rho_{R_x, s, a}$ | $-$                                | Fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data were available.
+Calculated from Spectrum model [@stover2019updates] outputs for region $R_x$ |
 
-### ART coverage \label{sec:art-cov}
+Table: (\#tab:prev) Each term in Equation \@ref(eq:prev) together with, where applicable, its prior distribution and a written description of its role.
 
-We model ART coverage $\alpha_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
+### ART coverage {#art-cov}
+
+ART coverage $\alpha_{x, s, a} \in [0, 1]$ was modelled on the logit scale using the structured additive predictor
 \begin{equation}
-\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + \bu^\alpha_a + \bu_a^{\alpha, s = \text{M}} + \bu^\alpha_x + \bu_x^{\alpha, s = \text{M}} + \bu_x^{\alpha, a < 15} + \bmeta^\alpha_{R_x, s, a}
+\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + \mathbf{u}^\alpha_a + \mathbf{u}_a^{\alpha, s = \text{M}} + \mathbf{u}^\alpha_x + \mathbf{u}_x^{\alpha, s = \text{M}} + \mathbf{u}_x^{\alpha, a < 15} + \boldsymbol{\mathbf{\eta}}^\alpha_{R_x, s, a}
 \end{equation}
-with terms and priors analogous to the HIV prevalence process model in Section \ref{sec:hiv-prev} above.
+with terms and prior distributions analogous to the HIV prevalence process model in Section \@ref(hiv-prev) above.
 
-<!-- Latent field size -->
-<!-- 1 + 1 + 13 + 10 + n + n + n = 36 + 3n -->
-<!-- Hyperparameter size -->
-<!-- 5 + 2 + 2 -->
+### HIV incidence rate {#hiv-inc}
 
-### HIV incidence rate \label{sec:hiv-inc}
+HIV incidence rate $\lambda_{x, s, a} > 0$ was modelled on the log scale using the structured additive predictor
+\begin{equation}
+\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + \mathbf{u}_x^\lambda + \boldsymbol{\mathbf{\eta}}_{R_x, s, a}^\lambda.
+(\#eq:inc)
+\end{equation}
+Table \@ref(tab:inc) provides a description of the terms included in Equation \@ref(eq:inc).
 
-We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
-\begin{equation}
-\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + \bu_x^\lambda + \bmeta_{R_x, s, a}^\lambda.
-\label{eq:inc}
-\end{equation}
-Table \ref{tab:inc} provides a description of the terms included in Equation \ref{eq:inc}.
+| Term                              | Distribution                     | Description                                                                                                                                                   |
+|-----------------------------------|----------------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------|
+| $\beta^\lambda_0$                 | $\mathcal{N}(0, 5)$              | Intercept term proportional to the average HIV transmission rate for untreated HIV positive adults                                                            |
+| $\beta_S^{\lambda, s = \text{M}}$ | $\mathcal{N}(0, 5)$              | The log incidence rate ratio for men compared to women                                                                                                        |
+| $\rho_{x}^{\text{15-49}}$         | $-$                              | The HIV prevalence among adults 15-49 in district $x$ calculated by aggregating age-specific HIV prevalences                                                                  |
+| $\alpha_{x}^{\text{15-49}}$       | $-$                              | The ART coverage among adults 15-49 in district $x$ calculated by aggregating age-specific ART coverages                                                                      |
+| $\omega = 0.7$                    | $-$                              | Average reduction in HIV transmission rate per increase in population ART coverage fixed based on inputs to the Estimation and Projection Package (EPP) model |
+| $\mathbf{u}_x^\lambda$                   | $\mathcal{N}(0, \sigma^\lambda)$ | IID spatial random effects with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$                                                                              |
+| $\boldsymbol{\mathbf{\eta}}^\lambda_{R_x, s, a}$      | $-$          | Fixed log incidence rate ratios by sex and age group calculated from Spectrum model outputs for region $R_x$                                                                    |
 
-<!-- \citet{eaton2019estimation} -->
+Table: (\#tab:inc) Each term in Equation \@ref(eq:inc) together with, where applicable, its prior distribution and a written description of its role.
 
-\begin{table}[h]
-\centering
-\begin{tabularx}{\textwidth}{llX}
-\toprule
-Term & Distribution & Description \\ 
-\midrule
-$\beta^\lambda_0$ & $\mathcal{N}(0, 5)$ & Intercept term proportional to the average HIV transmission rate for untreated HIV positive adults \\
-$\beta_S^{\lambda, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The log incidence rate ratio for men compared to women \\
-$\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}$ & $-$ & The HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences \\
-$\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}$ & $-$ & The ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages \\
-$\omega = 0.7$ & $-$ & Average reduction in HIV transmission rate per increase in population ART coverage fixed based on inputs to the Estimation and Projection Package (EPP) model \\
-$\bu_x^\lambda$ & $\mathcal{N}(0, \sigma^\lambda)$  & IID spatial random effects with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ \\
-$\bmeta^\lambda_{R_x, s, a}$ & $-$ & Fixed log incidence rate ratios by sex and age group calculated from Spectrum model output \\
-\bottomrule
-\end{tabularx}
-\caption{Terms included in the linear predictor for HIV incidence (Equation \ref{eq:inc}).
-Note that the only source age structure for this part of the model are $\bmeta^\lambda_{R_x, s, a}$.
-As Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80, the posterior over new infections in these age groups is exactly zero, by definition.
-We remove these identically zero posteriors from any later inference comparisons.}
-\label{tab:inc}
-\end{table}
-
-We model the proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ as
+The proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ was modelled as
 \begin{equation}
@@
 where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection, and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio.
-We use an informative prior for $\Omega_T$ based on the characteristics of the recent infection testing algorithm.
-For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$, and further there is assumed to be no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.
+The prior distribution for $\Omega_T$ was informed by the characteristics of the recent infection testing algorithm.
+For PHIA surveys this was $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$.
+For PHIA surveys there was assumed to be no false recency, such that $\beta_{T_0} = 0.0$, $\sigma^{\beta_T} = 0.0$, and $\beta_T = 0$.
 
-<!-- Latent field size -->
-<!-- 1 + 1 + n + 1 + 1 = 4 + n -->
-<!-- Hyperparameter size -->
-<!-- 3 -->
+### ANC testing {#anc-test}
 
-### ANC testing \label{sec:anc-test}
+HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women were modelled as being offset on the logit scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
+\begin{align}
+\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + \mathbf{u}_x^{\rho^{\text{ANC}}} + \boldsymbol{\mathbf{\eta}}_{R_x, a}^{\rho^{\text{ANC}}}, (\#eq:anc1) \\
+\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + \mathbf{u}_x^{\alpha^{\text{ANC}}} + \boldsymbol{\mathbf{\eta}}_{R_x, a}^{\alpha^{\text{ANC}}} (\#eq:anc2).
+\end{align}
+Table \@ref(tab:anc) provides a description of the terms included in Equation \@ref(eq:anc1) and Equation \@ref(eq:anc2).
 
-HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
-\begin{align}
-\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + \bu_x^{\rho^{\text{ANC}}} + \bmeta_{R_x, a}^{\rho^{\text{ANC}}}, \label{eq:anc1} \\
-\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + \bu_x^{\alpha^{\text{ANC}}} + \bmeta_{R_x, a}^{\alpha^{\text{ANC}}} \label{eq:anc2}.
-\end{align}
-Table \ref{tab:anc} provides a description of the terms included in Equation \ref{eq:anc1} and Equation \ref{eq:anc2}.
+| Term                                    | Distribution                                     | Description                                                                                                                                                                                                    |
+|-----------------------------------------|--------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
+| $\beta^{\theta^{\text{ANC}}}$           | $\mathcal{N}(0, 5)$                              | Intercept giving the average difference between population and ANC outcomes                                                                                                                                    |
+| $\mathbf{u}_x^{\theta^{\text{ANC}}}$    | $\mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ | IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$                                                                                                                     |
+| $\boldsymbol{\mathbf{\eta}}_{R_x, a}^{\theta^{\text{ANC}}}$ | $-$                          | Offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$ |
 
-\begin{table}[h]
-\centering
-\begin{tabularx}{\textwidth}{llX}
-\toprule
-Term & Distribution & Description \\ 
-\midrule
-$\beta^{\theta^{\text{ANC}}}$ & $\mathcal{N}(0, 5)$ & Intercept giving the average difference between population and ANC outcomes \\
-$\bu_x^{\theta^{\text{ANC}}}$ & $\mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ & IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$ \\ 
-$\bmeta_{R_x, a}^{\theta^{\text{ANC}}}$ & $-$ & Offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$ \\
-\bottomrule
-\end{tabularx}
-\caption{Terms included in the linear predictors for ANC testing, with $\theta \in \{\rho, \alpha\}$ (Equation \ref{eq:prev}).}
-\label{tab:anc}
-\end{table}
+Table: (\#tab:anc) Each term in Equations \@ref(eq:anc1) and \@ref(eq:anc2) together with (where applicable) its prior distribution and a written description of its role.
+The notation $\theta$ is used as stand in for $\theta \in \{\rho, \alpha\}$.
 
-In the full Naomi model, for adult women 15-49 the number of ANC clients $\Psi_{x, a} > 0$ are modelled as
-$$
-\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + \bu_x^\psi,
-$$
-where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $\bu_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
-Here we fix $\beta^\psi = 0$ and $\bu_x^\psi = \mathbf{0}$ such that $\Psi_{x, a}$ are simply constants.
+In the full Naomi model, for adult women 15-49 the number of ANC clients $\Psi_{x, a} > 0$ were modelled as
+\begin{equation}
+\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + \mathbf{u}_x^\psi,
+\end{equation}
+where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $\mathbf{u}_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
+Here these terms are fixed to $\beta^\psi = 0$ and $\mathbf{u}_x^\psi = \mathbf{0}$ such that $\Psi_{x, a}$ are simply constants.
 
-<!-- Latent field size -->
-<!-- 1 + n + 1 + n = 2 + 2n -->
-<!-- Hyperparameter size -->
-<!-- 2 -->
-
-### ART attendance \label{sec:art-attend}
+### ART attendance {#art-attend}
 
@@
 Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
-We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district or its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation, and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.
-To model $\gamma_{x, x'}$ for $x \sim x'$ we use a multinomial logistic regression model, based on the log-odds ratios
+Assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district or its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation, and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.
+
+The probabilities $\gamma_{x, x'}$ for $x \sim x'$ were modelled using multinomial logistic regression model, based on the log-odds ratios
 \begin{equation}
-\tilde \gamma_{x, x'} = \log \left( \frac{\gamma_{x, x'}}{1 - \gamma_{x, x'}} \right) = \tilde \gamma_0 + \bu_x^{\tilde \gamma}, \label{eq:log-or}
+\tilde \gamma_{x, x'} = \log \left( \frac{\gamma_{x, x'}}{1 - \gamma_{x, x'}} \right) = \tilde \gamma_0 + \mathbf{u}_x^{\tilde \gamma}. (\#eq:log-or)
 \end{equation}
-where $\tilde \gamma_0 = -4$ is a fixed intercept, and $\bu_x^{\tilde \gamma} \sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})$ are district random effects with $\sigma_X^{\tilde \gamma} \sim \mathcal{N}^+(0, 2.5)$.
-Note that Equation \ref{eq:log-or} does not depend on $x'$, such that $\gamma_{x, x'}$ is only a function of $x$.
-Choice of $\tilde \gamma_0 = -4$ implies a prior mean on $\gamma_{x, x'}$ of 1.8%, such that $(100 - 1.8 \times \text{ne}(x))\%$ of ART clients in district $x$ obtain treatment in their home district, a-priori.
-We fix $\tilde \gamma_{x, x} = 0$ and recover the multinomial probabilities using the softmax
+Table \@ref(tab:log-or) provides a description of the terms included in Equation \@ref(eq:log-or).
+Fixing $\tilde \gamma_{x, x} = 0$ then the multinomial probabilities may be recovered using the softmax
 \begin{equation}
@@
 \end{equation}
-Given the total number of PLHIV on ART $A_{x, s, a} = N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}$, the number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a} = A_{x, s, a} \cdot \gamma_{x, x'}$, and the total attending ART facilities in district $x'$ are
-\begin{equation}
-\tilde A_{x', s, a} = \sum_{x \in \{x', \text{ne}(x')\}} A_{x, x', s, a}.
-\end{equation}
 
-## Likelihood specification \label{sec:likelihood}
+| Term                           | Distribution                               | Description                                                                         |
+|--------------------------------|--------------------------------------------|-------------------------------------------------------------------------------------|
+| $\tilde \gamma_0$              | $-$                                        | Fixed intercept $\tilde \gamma_0 = -4$.
+Implies a prior mean on $\gamma_{x, x'}$ of 1.8%, such that a-priori $(100 - 1.8 \times \text{ne}(x))\%$ of ART clients in district $x$ obtain treatment in their home district |
+| $\mathbf{u}_x^{\tilde \gamma}$ | $\mathcal{N}(0, \sigma_X^{\tilde \gamma})$ | District random effects, with $\sigma_X^{\tilde \gamma} \sim \mathcal{N}^+(0, 2.5)$ |
+
+Table: (\#tab:log-or) Each term in Equation \@ref(eq:log-or) together with, where applicable, its prior distribution and a written description of its role.
+As no terms include $x'$, $\gamma_{x, x'}$ is only a function of $x$.
+
+## Additional likelihood specification {#naomi-likelihood}
+
+Additional useful details regarding Naomi's likelihood specification are provided here.
 
@@
 
-For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the most recent household survey furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
-\begin{equation}
-y^{\theta}_{\{x\}, \{s\}, \{a\}} = m^{\theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}.
-\end{equation}
-To model these observations, we use the following three binomial working likelihoods:
-
-1. The number of people living with HIV (PLHIV) is
-\begin{equation}
-y^{\rho}_{\{x\}, \{s\}, \{a\}} \sim \text{xBin}(m^{\rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}),
-\end{equation}
-where
-\begin{equation}
-\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}
-\end{equation}
-has denominator given by the total population size.
-2. The number of PLHIV on ART is
-\begin{equation}
-y^{\alpha}_{\{x\}, \{s\}, \{a\}} \sim \text{xBin}(m^{\alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}),
-\end{equation}
-where
-\begin{equation}
-\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}
-\end{equation}
-has denominator given by the total PLHIV.
-3. The number of PLHIV recently infected
-\begin{equation}
-y^{\kappa}_{\{x\}, \{s\}, \{a\}} \sim \text{xBin}(m^{\kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}),
-\end{equation}
-where
-\begin{equation}
-\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}
-\end{equation}
-has denominator given by the total PLHIV.
-
-The generalised binomial $y \sim \text{xBin}(m, p)$ used above is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
-\begin{equation}
-\log p(y) = \log \Gamma(m + 1) - \log \Gamma(y + 1) - \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
-\end{equation}
+The generalised binomial $y \sim \text{xBin}(m, p)$ is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
+\begin{align}
+\log p(y) = &\log \Gamma(m + 1) - \log \Gamma(y + 1) \\
+            &- \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
+\end{align}
 where the gamma function $\Gamma$ is such that $\forall n \in \mathbb{N}$, $\Gamma(n) = (n - 1)!$.
 
-### ANC testing data
-
-We include ANC testing data for the year of the most recent survey.
-Let $W^\text{ANC}_{\{x\}}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}}$ the number of ANC clients already on ART prior to first ANC, such that
-\begin{equation}
-W^\text{ANC}_{x} \geq X^\text{ANC}_{x} \geq Y^\text{ANC}_{x} \geq Z^\text{ANC}_{x},
-\end{equation}
-for all $x \in \{x\}$.
-When ANC testing data are only available for part of a given year, we denote $m^\text{ANC} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for that year.
-The observed number of HIV positive and already on ART among ANC clients is modelled by
-\begin{align}
-Y_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} }^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right), \\
-Z_{\{x\} }^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} }^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}}^{\text{ANC}}\right),
-\end{align}
-where prevalence and ART coverage are aggregated by the number of pregnant women $\Psi_{x, a}$
-\begin{align}
-\rho_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a}}, \\
-\alpha_{\{x\}\{a\}}^{\text{ANC}} &= \frac{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{\{x\}} \sum_{\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}.
-\end{align}
-
-### Number receiving ART
-
-Let $\dot A_{\{x\}, \{s\}, \{a\}}$ be data for the number receiving ART
-$$
-\dot A_{\{x\}, \{s\}, \{a\}} = \sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a}.
-$$
-We model the unobserved numbers of ART clients travelling from $x'$ to $x$ as
-$$
-\dot A_{x', x, s, a} \sim \text{Bin}(N_{x', s, a}, \pi_{x', x, s, a})
-$$
-where $\pi_{x', x, s, a} = \rho_{x', s, a} \cdot \alpha_{x', s, a} \cdot \gamma_{x', x, s, a}$. This likelihood is approximated using a normal for the sum of binomials by
-$$
-\dot A_{\{x\}, \{s\}, \{a\}} \sim \mathcal{N}(\tilde A_{\{x\}, \{s\}, \{a\}}, \sigma^{\tilde A}_{\{x\}, \{s\}, \{a\}})
-$$
-where the mean is
-$$
-\tilde{A}_{\{x\},\{s\}\{a\}} = \sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a},
-$$
-and the variance is
-$$
-\left(\sigma_{\{x\},\{s\}\{a\}}^{\tilde{A}}\right)^2 = \sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} \sum_{x \sim x', x = x'} N_{x', s, a} \cdot \pi_{x', x, s, a} \cdot \left(1-\pi_{x', x, s, a}\right).
-$$
-
-## Identifiability constaints
+## Identifiability constraints {#naomi-identifiability}
 
@@
 
-1.
-If survey data on HIV prevalence or ART coverage by age and sex are not available then we set $\bu_a^\theta = 0$ and $\bu_{a, s = \text{M}}^\theta = 0$ and use the average age-sex pattern of from the Spectrum offset $\bmeta_{R_x, s, a}^\theta$.
-For the Malawi example considered in the main text HIV prevalence and ART coverage data are not available for those aged 65+.
+1. If survey data on HIV prevalence or ART coverage by age and sex are not available then $\mathbf{u}_a^\theta = 0$ and $\mathbf{u}_{a, s = \text{M}}^\theta = 0$.
+In this case, the average age-sex pattern from the Spectrum is used.
+For the Malawi case-study (Section \@ref(malawi)), HIV prevalence and ART coverage data are not available for those aged 65+.
 As a result, there are $|\{\text{0-4}, \ldots, \text{50-54}\}| = 13$ age groups included for the age random effects.
@@
 If no ART data, either survey or ART programme, are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable.
-In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = \bu_x^\alpha + \bmeta_{R_x, s, a}^\alpha$.
-3.
-If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = 0$ and $\bu_x^\lambda = \mathbf{0}$.
-The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.
+In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that
+\begin{equation}
+\text{logit} \left(\alpha_{x, s, a} \right) = \mathbf{u}_x^\alpha + \boldsymbol{\mathbf{\eta}}_{R_x, s, a}^\alpha.
+\end{equation}
+3. If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = 0$ and $\mathbf{u}_x^\lambda = \mathbf{0}$.
+The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum, and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district-level estimates of HIV prevalence and ART coverage.
 
-<!-- Create DAG representing Naomi model. This would likely go into the appendix, so use that notation -->
+## Implementation {#naomi-implementation}
 
-```{r simplified-naomi, fig.cap="Directed acyclic graph describing the simplified Naomi model (work in progress)."}
-knitr::include_graphics("depends/simplified-naomi.pdf")
-```
-
-## Implementation
-
-The C++ `TMB` code for the negative log-posterior of the simplified Naomi model is available from the GitHub repository `athowes/naomi-aghq`.
-For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used in Section \ref{sec:math} and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
+The `TMB` C++ code for the negative log-posterior of the simplified Naomi model is available from [`https://github.com/athowes/naomi-aghq`](https://github.com/athowes/naomi-aghq).
+For ease of understanding, Table \@ref(tab:tmb-lookup) provides correspondence between the mathematical notation used in Section \@ref(naomi-math) and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
 For further reference on the `TMB` software see @tmbdocumentation.
 
-<!-- Use table(names(tmb$fit$par.full)) to check the below table -->
+| Variable name              | Notation    | Type   | Domain  | $\rho$ | $\alpha$ | $\lambda$ |
+|----------------------------|-------------|--------|---------|--------|----------|-----------|
+| `logit_phi_rho_x`      | $\text{logit}(\phi_X^\rho)$                          | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `log_sigma_rho_x`      | $\log(\sigma_X^\rho)$                                | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `logit_phi_rho_xs`     | $\text{logit}(\phi_{XS}^\rho)$                       | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `log_sigma_rho_xs`     | $\log(\sigma_{XS}^\rho)$                             | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `logit_phi_rho_a`      | $\text{logit}(\phi_A^\rho)$                          | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `log_sigma_rho_a`      | $\log(\sigma_A^\rho)$                                | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `logit_phi_rho_as`     | $\text{logit}(\phi_{AS}^\rho)$                       | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `log_sigma_rho_as`     | $\log(\sigma_{AS}^\rho)$                             | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `log_sigma_rho_xa`     | $\log(\sigma_{XA}^\rho)$                             | Hyper  | $\mathbb{R}$      | Yes |     |     |
+| `logit_phi_alpha_x`    | $\text{logit}(\phi_X^\alpha)$                        | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `log_sigma_alpha_x`    | $\log(\sigma_X^\alpha)$                              | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `logit_phi_alpha_xs`   | $\text{logit}(\phi_{XS}^\alpha)$                     | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `log_sigma_alpha_xs`   | $\log(\sigma_{XS}^\alpha)$                           | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `logit_phi_alpha_a`    | $\text{logit}(\phi_A^\alpha)$                        | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `log_sigma_alpha_a`    | $\log(\sigma_A^\alpha)$                              | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `logit_phi_alpha_as`   | $\text{logit}(\phi_{AS}^\alpha)$                     | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `log_sigma_alpha_as`   | $\log(\sigma_{AS}^\alpha)$                           | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `log_sigma_alpha_xa`   | $\log(\sigma_{XA}^\alpha)$                           | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `OmegaT_raw`           | $\Omega_T$                                           | Hyper  | $\mathbb{R}$      |     |     | Yes |
+| `log_betaT`            | $\log(\beta_T)$                                      | Hyper  | $\mathbb{R}$      |     |     | Yes |
+| `log_sigma_lambda_x`   | $\log(\sigma^\lambda)$                               | Hyper  | $\mathbb{R}$      |     |     | Yes |
+| `log_sigma_ancrho_x`   | $\log(\sigma_X^{\rho^{\text{ANC}}})$                 | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `log_sigma_ancalpha_x` | $\log(\sigma_X^{\alpha^{\text{ANC}}})$               | Hyper  | $\mathbb{R}$      |     | Yes |     |
+| `log_sigma_or_gamma`   | $\log(\sigma_X^{\tilde \gamma})$                     | Hyper  | $\mathbb{R}$      |     |     |     |
+| `beta_rho`             | $(\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}})$     | Latent | $\mathbb{R}^2$    | Yes |     |     |
+| `beta_alpha`           | $(\beta^\alpha_0, \beta_{S}^{\alpha, s = \text{M}})$ | Latent | $\mathbb{R}^2$    |     | Yes |     |
+| `beta_lambda`          | $(\beta_0^\lambda, \beta_S^{\lambda, s = \text{M}})$ | Latent | $\mathbb{R}^2$    |     |     | Yes |
+| `beta_anc_rho`         | $\beta^{\rho^{\text{ANC}}}$                          | Latent | $\mathbb{R}$      |     | Yes |     |
+| `beta_anc_alpha`       | $\beta^{\alpha^{\text{ANC}}}$                        | Latent | $\mathbb{R}$      |     | Yes |     |
+| `u_rho_x`              | $\mathbf{w}^\rho_x$                                  | Latent | $\mathbb{R}^{n}$  | Yes |     |     |
+| `us_rho_x`             | $\mathbf{v}^\rho_x$                                  | Latent | $\mathbb{R}^{n}$  | Yes |     |     |
+| `u_rho_xs`             | $\mathbf{w}_x^{\rho, s = \text{M}}$                  | Latent | $\mathbb{R}^{n}$  | Yes |     |     |
+| `us_rho_xs`            | $\mathbf{v}_x^{\rho, s = \text{M}}$                  | Latent | $\mathbb{R}^{n}$  | Yes |     |     |
+| `u_rho_a`              | $\mathbf{u}^\rho_a$                                  | Latent | $\mathbb{R}^{10}$ | Yes |     |     |
+| `u_rho_as`             | $\mathbf{u}_a^{\rho, s = \text{M}}$                  | Latent | $\mathbb{R}^{10}$ | Yes |     |     |
+| `u_rho_xa`             | $\mathbf{u}_x^{\rho, a < 15}$                        | Latent | $\mathbb{R}^{n}$  | Yes |     |     |
+| `u_alpha_x`            | $\mathbf{w}^\alpha_x$                                | Latent | $\mathbb{R}^{n}$  |     | Yes |     |
+| `us_alpha_x`           | $\mathbf{v}^\alpha_x$                                | Latent | $\mathbb{R}^{n}$  |     | Yes |     |
+| `u_alpha_xs`           | $\mathbf{w}_x^{\alpha, s = \text{M}}$                | Latent | $\mathbb{R}^{n}$  |     | Yes |     |
+| `us_alpha_xs`          | $\mathbf{v}_x^{\alpha, s = \text{M}}$                | Latent | $\mathbb{R}^{n}$  |     | Yes |     |
+| `u_alpha_a`            | $\mathbf{u}^\alpha_a$                                | Latent | $\mathbb{R}^{13}$ |     | Yes |     |
+| `u_alpha_as`           | $\mathbf{u}_a^{\alpha, s = \text{M}}$                | Latent | $\mathbb{R}^{10}$ |     | Yes |     |
+| `u_alpha_xa`           | $\mathbf{u}_x^{\alpha, a < 15}$                      | Latent | $\mathbb{R}^{n}$  |     | Yes |     |
+| `ui_lambda_x`          | $\mathbf{u}_x^\lambda$                               | Latent | $\mathbb{R}^{n}$  |     |     | Yes |
+| `ui_anc_rho_x`         | $\mathbf{u}_x^{\rho^{\text{ANC}}}$                   | Latent | $\mathbb{R}^{n}$  |     | Yes |     |
+| `ui_anc_alpha_x`       | $\mathbf{u}_x^{\alpha^{\text{ANC}}}$                 | Latent | $\mathbb{R}^{n}$  |     | Yes |     |
+| `log_or_gamma`         | $\mathbf{u}_x^{\tilde \gamma}$                       | Latent | $\mathbb{R}^{n}$  |     |     |
 
-\begin{table}[p!]
-\centering
-\begin{tabularx}{\textwidth}{llllllll}
-\toprule
-Variable name & Notation & Type & Size & Domain & $\rho$ input? & $\alpha$ input? & $\lambda$ input? \\
-\midrule
-\texttt{logit\_phi\_rho\_x} & $\text{logit}(\phi_X^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{log\_sigma\_rho\_x} & $\log(\sigma_X^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{logit\_phi\_rho\_xs} & $\text{logit}(\phi_{XS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{log\_sigma\_rho\_xs} & $\log(\sigma_{XS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{logit\_phi\_rho\_a} & $\text{logit}(\phi_A^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{log\_sigma\_rho\_a} & $\log(\sigma_A^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{logit\_phi\_rho\_as} & $\text{logit}(\phi_{AS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{log\_sigma\_rho\_as} & $\log(\sigma_{AS}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{log\_sigma\_rho\_xa} & $\log(\sigma_{XA}^\rho)$ & Hyper & 1 & $\mathbb{R}$ & \cmark & & \\
-\texttt{logit\_phi\_alpha\_x} & $\text{logit}(\phi_X^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{log\_sigma\_alpha\_x} & $\log(\sigma_X^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{logit\_phi\_alpha\_xs} & $\text{logit}(\phi_{XS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{log\_sigma\_alpha\_xs} & $\log(\sigma_{XS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{logit\_phi\_alpha\_a} & $\text{logit}(\phi_A^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{log\_sigma\_alpha\_a} & $\log(\sigma_A^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{logit\_phi\_alpha\_as} & $\text{logit}(\phi_{AS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{log\_sigma\_alpha\_as} & $\log(\sigma_{AS}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{log\_sigma\_alpha\_xa} & $\log(\sigma_{XA}^\alpha)$ & Hyper & 1 & $\mathbb{R}$ & & \cmark & \\
-\texttt{OmegaT\_raw} & $\Omega_T$ & Hyper & 1 & $\mathbb{R}$ & & & \cmark \\
-\texttt{log\_betaT} & $\log(\beta_T)$ & Hyper & 1 & $\mathbb{R}$ & & & \cmark \\
-\texttt{log\_sigma\_lambda\_x} & $\log(\sigma^\lambda)$ & Hyper & 1 & $\mathbb{R}$ & & & \cmark \\
-\texttt{log\_sigma\_ancrho\_x} & $\log(\sigma_X^{\rho^{\text{ANC}}})$ & Hyper & 1 & $\mathbb{R}$ & & & \\
-\texttt{log\_sigma\_ancalpha\_x} & $\log(\sigma_X^{\alpha^{\text{ANC}}})$ & Hyper & 1 & $\mathbb{R}$ & & & \\
-\texttt{log\_sigma\_or\_gamma} & $\log(\sigma_X^{\tilde \gamma})$ & Hyper & 1 & $\mathbb{R}$ & & & \\
-\texttt{beta\_rho} & $(\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ & \cmark & & \\
-\texttt{beta\_alpha} & $(\beta^\alpha_0, \beta_{S}^{\alpha, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ & & \cmark & \\
-\texttt{beta\_lambda} & $(\beta_0^\lambda, \beta_S^{\lambda, s = \text{M}})$ & Latent & 2 & $\mathbb{R}^2$ & & & \cmark \\
-\texttt{beta\_anc\_rho} & $\beta^{\rho^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ & & & \\
-\texttt{beta\_anc\_alpha} & $\beta^{\alpha^{\text{ANC}}}$ & Latent & 1 & $\mathbb{R}$ & & & \\
-\texttt{u\_rho\_x} & $\w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{us\_rho\_x} & $\bv^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{u\_rho\_xs} & $\w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{us\_rho\_xs} & $\bv_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{u\_rho\_a} & $\bu^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
-\texttt{u\_rho\_as} & $\bu_a^{\rho, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
-\texttt{u\_rho\_xa} & $\bu_x^{\rho, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
-\texttt{u\_alpha\_x} & $\w^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ && \cmark & \\
-\texttt{us\_alpha\_x} & $\bv^\alpha_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
-\texttt{u\_alpha\_xs} & $\w_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
-\texttt{us\_alpha\_xs} & $\bv_x^{\alpha, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
-\texttt{u\_alpha\_a} & $\bu^\alpha_a$ & Latent & 13 & $\mathbb{R}^{13}$ & & \cmark & \\
-\texttt{u\_alpha\_as} & $\bu_a^{\alpha, s = \text{M}}$ & Latent & 10 & $\mathbb{R}^{10}$ & & \cmark & \\
-\texttt{u\_alpha\_xa} & $\bu_x^{\alpha, a < 15}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & \cmark & \\
-\texttt{ui\_lambda\_x} & $\bu_x^\lambda$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \cmark \\
-\texttt{ui\_anc\_rho\_x} & $\bu_x^{\rho^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
-\texttt{ui\_anc\_alpha\_x} & $\bu_x^{\alpha^{\text{ANC}}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
-\texttt{log\_or\_gamma} & $\bu_x^{\tilde \gamma}$ & Latent & $n$ & $\mathbb{R}^{n}$ & & & \\
-\bottomrule
-\end{tabularx}
-\caption{Correspondence between mathematical notation and variable names used in our \texttt{TMB} code.
-The total number of hyperparameters is 24, and the total number of latent field parameters is $51 + 14n$, where $n$ is the number of districts.
-We use the notation \cmark \ to refer to direct dependence of the parameter on the variable, \xmark \ to refer to no dependence, and a blank entry to refer to dependence conditional on the data.}
-\label{tab:tmb-lookup}
-\end{table}
-
-<!-- Latent field size -->
-<!-- 2 + 2 + 2 + 1 + 1+ 10 + 10 + 13 + 10 + 14n = 51 + 14n -->
-<!-- Hyperparameter size -->
-<!-- 24 -->
-
-<!-- \clearpage -->
-
-<!-- ```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")} -->
-<!-- ``` -->
+Table: (\#tab:tmb-lookup) Correspondence between the variable name used in the Naomi `TMB` template and the mathematical notation used in Appendix \@ref(naomi-math).
+The parameter type, either a hyperparameter or element of the latent field, is also given.
+All of the parameters are defined on the real-scale in some dimension.
+In the final three columns ($\rho$, $\alpha$, and $\lambda$) indication is given as to which component of the model the parameter is primarily used in.
 
@@
 
-# Model assessment\label{sec:model-assessment}
+# NUTS convergence and suitability
 
-## Additional figures
+```{r}
+naomi_nuts <- readRDS("depends/mcmc-out.rds")
+```
 
-```{r contraction, fig.cap="A posterior contraction of 1 corresponds to a Dirac delta function.
-A posterior contraction of 0 corresponds to no change in standard deviation between prior and posterior.
-A posterior contraction less than 0 corresponds to a wider posterior than prior."}
+(ref:rhat) For NUTS run on the Naomi ELGM, the maximum potential scale reduction factor was `r round(naomi_nuts$rhat_max, 3)`, below the value of 1.05 typically used as a cutoff for acceptable chain mixing, indicating that the results are acceptable to use.
+Additionally, the vast majority (`r round(100 * naomi_nuts$rhat_above_threshold, 1)`%) of $\hat R$ values were less than 1.1.
+
+```{r rhat, fig.cap="(ref:rhat)"}
+knitr::include_graphics("depends/rhat.png")
+```
+
+(ref:ess) The efficiency of the NUTS, as measured by the ratio of effective sample size to total number of iterations run, was low for most parameters (Panel A).
+As a result, the number of iterations required for the the effective number of samples (mean `r round(naomi_nuts$ess_mean)`) to be satisfactory was high (Panel B).
+
+```{r ess, fig.cap="(ref:ess)"}
+knitr::include_graphics("depends/ess.png")
+```
+
+(ref:worst-trace) Traceplots for the parameter with the lowest ESS which was `log_sigma_alpha_xs` (an $\text{ESS}$ of `r round(naomi_nuts$ess_min)`, Panel A) and highest potential scale reduction factor which was `ui_lambda_x[10]` (an $\hat R$ of `r round(naomi_nuts$rhat_max, 3)`, Panel B).
+
+```{r worst-trace, fig.cap="(ref:worst-trace)"}
+knitr::include_graphics("depends/worst-trace.png")
+```
+
+(ref:rho-a) Pairs plots for the parameters $\log(\sigma_{A}^\rho)$ and $\text{logit}(\phi_{A}^\rho)$, or `log_sigma_rho_a` and `logit_phi_rho_a` as implemented in code.
+These parameters are the log standard deviation and logit lag-one correlation parameter of an AR1 process.
+In the posterior distribution obtained with NUTS, they have a high degree of correlation.
+
+```{r rho-a, fig.cap="(ref:rho-a)"}
+knitr::include_graphics("depends/rho_a.png")
+```
+
+(ref:alpha-x) Pairs plots for the parameters $\log(\sigma_X^\alpha)$ and $\text{logit}(\phi_X^\alpha)$, or `log_sigma_alpha_x` and `logit_phi_alpha_x` as implemented in code.
+These parameters are the log standard deviation and logit BYM2 proportion parameter of a BYM2 process.
+In the posterior distribution obtained with NUTS, they are close to uncorrelated.
+
+```{r alpha-x, fig.cap="(ref:alpha-x)"}
+knitr::include_graphics("depends/alpha_x.png")
+```
+
+(ref:nuts-hand-comparison) Prior standard deviations were calculated by using NUTS to simulate from the prior distribution.
+This approach is more convenient than simulating directly from the model, but can lead to inaccuracies.
+
+```{r nuts-hand-comparison, fig.cap="(ref:nuts-hand-comparison)"}
+knitr::include_graphics("depends/nuts-hand-comparison.png")
+```
+
+(ref:posterior-contraction) The posterior contraction for each parameter in the model.
+Values are averaged for parameters of length greater than one.
+The posterior contraction is zero when the prior distribution and posterior distribution have the same standard deviation.
+This could indicate that the data is not informative about the parameter.
+The closer the posterior contraction is to one, the more than the marginal posterior distribution has concentrated about a single point.
+
+```{r posterior-contraction, fig.cap="(ref:posterior-contraction)"}
 knitr::include_graphics("depends/posterior-contraction.png")
@@
 
-# AGHQ and PCA-AGHQ details
+\clearpage
 
-## Additional figures
+# Use of PCA-AGHQ
 
-```{r scree, fig.cap="Most variation in the inverse curvature could be explained using far fewer than the full 24 dimensions."}
-knitr::include_graphics("depends/total-variation.png")
+```{r}
+u <- runif(10000)
 ```
 
-```{r reduced-rank, fig.cap="The reduced rank ($s = 8$) matrix approximation to the covariance matrix is visually similar to the full rank original."}
-knitr::include_graphics("depends/reduced-rank.png")
-```
+(ref:nodes-quantiles-sd) The standard deviation of the quadrature nodes can be used as a measure of coverage of the posterior marginal distribution.
+Nodes spaced evenly within the marginal distribution would be expected to uniformly distributed quantile, corresponding to a standard deviation of `r signif(sd(u), 4)`, shown as a dashed line.
 
-```{r nodes-quantiles-sd, fig.cap="Standard deviations of the quantiles of the quadrature nodes within the NUTS posterior draws varied substantially, in accordance with the marginal standard deviations shown in Figure \\ref{fig:marginal-sd}."}
+```{r nodes-quantiles-sd, fig.cap="(ref:nodes-quantiles-sd)"}
 knitr::include_graphics("depends/nodes-quantiles-sd.png")
@@
 
-```{r marginal-sd, fig.cap="Hyperparameters on the logit-scale had systematically higher marginal standard deviations than those on the log-scale.
-This is because variation on the real scale is compressed by the inverse logit and expanded by the inverse log (exponential)."}
+(ref:marginal-sd) The estimated posterior marginal standard deviation of each hyperparameter varied substantially based on its scale, either logarithmic or logistic.
+
+```{r marginal-sd, fig.cap="(ref:marginal-sd)"}
 knitr::include_graphics("depends/marginal-sd.png")
@@
 
-```{r pc-loadings, fig.cap="Principal component loadings based on the eigendecomposition of the inverse curvature."}
-knitr::include_graphics("depends/pc-loadings.png")
+(ref:lognormconst) The logarithm of the normalising constant estimated using PCA-AGHQ and a range of possible values of $k = 2, 3, 5$ and $s \leq 8$.
+Using this range of settings, there was not convergence of the logarithm of the normalising constant estimate.
+The time taken by GPCA-AGHQ increases exponentially with number of PCA-AGHQ dimensions kept.
+
+```{r lognormconst, fig.cap="(ref:lognormconst)"}
+knitr::include_graphics("depends/lognormconst.png")
 ```
-
-## Normalising constant estimation\label{sec:normalising}
-
-Add here plots and description of the estimated normalising constant for different PCA-AGHQ settings.
 
@@
 
-In this section we give more detailed inference comparison results for each parameter or group of parameters.
-For any parameter with length greater than one, the results presented are an average.
-
 ## Point estimates
 
-### Posterior mean
+(ref:mean-alt-output) Differences in Naomi model output posterior means as estimated by GEB and GPCA-AGHQ compared to NUTS.
+Each point is an estimate of the indicator for a particular strata.
+In all cases, error is reduced by GPCA-AGHQ, most of all for ART coverage.
 
-```{r}
-point_summary <- readr::read_csv("depends/point-estimates.csv")
-
-point_summary <- point_summary |>
-  mutate(type = forcats::fct_recode(type, "Output" = "output", "Latent" = "latent"))
-
-latex_table_point_estimates <- function(.indicator) {
-  point_summary |>
-  group_by(par, type, indicator, method) |>
-  summarise(
-    rmse = sqrt(mean((truth - approximate)^2)),
-    mae = mean(abs(truth - approximate))
-  ) |>
-  ungroup() |>
-  tidyr::pivot_wider(
-    names_from = "method",
-    values_from = c("rmse", "mae")
-  ) |>
-  filter(indicator == .indicator) |>
-  select(Parameter = par, Type = type, `TMB` = `rmse_TMB`, `PCA-AGHQ` = `rmse_PCA-AGHQ`) |>
-  mutate(
-    Difference = `PCA-AGHQ` - TMB,
-    `Percent Difference` =  (`PCA-AGHQ` - TMB) / TMB
-  ) |>
-  gt::gt(groupname_col = "Type") |>
-  gt::fmt_number(
-    columns = c(`TMB`, `PCA-AGHQ`, Difference),
-    n_sigfig = 2
-  ) |>
- gt::fmt_percent(
-    columns = `Percent Difference`,
-    decimals = 1
-  ) |>
-  gt::text_transform(
-    locations = gt::cells_body("Parameter"),
-    fn = function(x) paste0("\\texttt{", x, "}")
-  ) |>
-  gt::as_latex() |>
-  as.character()
-}
-
-latex_table_point_estimates("Posterior mean estimate") |>
-  cat(file = "point-summary-mean.tex")
-
-latex_table_point_estimates("Posterior SD estimate") |>
-  cat(file = "point-summary-sd.tex")
+```{r mean-alt-output, fig.cap="(ref:mean-alt-output)"}
+knitr::include_graphics("depends/mean-alt-output.png")
 ```
 
-\input{point-summary-mean.tex}
+(ref:sd-alt-output) Differences in Naomi model output posterior standard deviations as estimated by GEB and GPCA-AGHQ compared to NUTS.
+Each point is an estimate of the indicator for a particular strata.
+Error is increased by GPCA-AGHQ for HIV prevalence and HIV incidence, and reduced for ART coverage.
 
-### Posterior standard deviation
-
-\input{point-summary-sd.tex}
-
-## Distribution tests
-
-### Kolmogorov-Smirnov
-
-```{r}
-ks_summary <- readRDS("depends/ks-summary.rds")
-
-ks_summary |>
-  ungroup() |>
-  filter(type == "Latent") |>
-  select(-type) |>
-  select(Parameter = parname, TMB, `PCA-AGHQ` = aghq) |>
-  mutate(Difference = `PCA-AGHQ` - TMB) |>
-  gt::gt() |>
-  gt::summary_rows(
-   columns = c(`TMB`, `PCA-AGHQ`, Difference),
-   fns = list(Average = ~ mean(.x)),
-   formatter = gt::fmt_number,
-   decimals = 3
-  ) |>
-  gt::fmt_number(
-    columns = c(`TMB`, `PCA-AGHQ`, Difference),
-    decimals = 3
-  ) |>
-  gt::text_transform(
-    locations = gt::cells_body("Parameter"),
-    fn = function(x) paste0("\\texttt{", x, "}")
-  ) |>
-  gt::text_transform(
-    locations = gt::cells_body("Difference"),
-    fn = function(x) ifelse(x < 0, paste0("\\textbf{", x, "}"), paste0(x))
-  ) |>
-  gt::as_latex() |>
-  as.character() |>
-  cat(file = "ks-summary.tex")
+```{r sd-alt-output, fig.cap="(ref:sd-alt-output)"}
+knitr::include_graphics("depends/sd-alt-output.png")
 ```
 
-\input{ks-summary.tex}
+## Distributional quantities
 
-## Pareto-smoothed importance sampling
+(ref:ks-ess) The Kolmogorov-Smirnov (KS) test statistic for each latent field parameter is correlated with the effective sample size (ESS) from NUTS, for both GEB and GPCA-AGHQ.
+This may be because parameters which are harder to estimate with INLA-like methods also have posterior distributions which are more difficult to sample from.
+Alternatively, it may be that high KS values are caused by inaccurate NUTS estimates generated by limited effective samples.
 
-## Maximum mean discrepancy
-
-\clearpage
-
-# MCMC convergence and suitability\label{sec:mcmc}
-
-```{r}
-mcmc_out <- readRDS("depends/mcmc-out.rds")
-```
-
-We assessed MCMC convergence and suitability using a range of graphical and numerical tests.
-The largest scale reduction factor $\hat R$ was `r signif(mcmc_out$max_rhat, 3)` (Figure \ref{fig:rhat}).
-As such, all values of $\hat R < 1.05$.
-Even thinning by a factor of 20, samples were not obtained very efficiently, resulting in the majority of effective sample size (ESS) ratios being below 0.5, with some as low as 0.1 (Figure \ref{fig:ratio}).
-As a result, the number of obtained ESS varied substantially by parameter (Figure \ref{fig:ess}): the minimum was `r round(mcmc_out$ess_min, 1)`, 2.5% quantile was `r round(mcmc_out$ess_lower, 1)`, median was `r round(mcmc_out$ess_median, 1)`, 97.5% quantile was `r round(mcmc_out$ess_upper, 1)` and maximum was `r round(mcmc_out$ess_max, 1)`.
-Traceplots for the parameters with the lowest ESS (`log_sigma_alpha_xs`) and highest $\hat R$ (the 10th index of `ui_lambda_x`) are shown in Figure \ref{fig:worst_trace}.
-There were no divergent transitions.
-<!-- Could add energy plot from @betancourt2017conceptual here. -->
-
-```{r}
-np <- readRDS("depends/nuts-params.rds")
-
-n_divergent <- np |>
-  filter(Parameter == "divergent__") |>
-  summarise(n_divergent = sum(Value)) |>
-  pull(n_divergent)
-
-stopifnot(n_divergent == 0)
-```
-
-```{r rhat, fig.cap="The potential scale reduction factor compares between- and within- estimates of univariate parameters.
-It is recommended only to use NUTS results if the value is less than 1.05, which it is for all parameters."}
-knitr::include_graphics("depends/rhat.png")
-```
-
-```{r ratio, fig.cap="The efficiently, as measured by the ESS ratio, of the NUTS sampler was poor."}
-knitr::include_graphics("depends/ratio.png")
-```
-
-```{r ess, fig.cap="The effective number of samples we obtained varied substantially between parameters."}
-knitr::include_graphics("depends/ess.png")
-```
-
-```{r worst_trace, fig.cap="Traceplots of the parameters with the lowest ESS and highest scale reduction factor."}
-knitr::include_graphics("depends/worst-trace.png")
-```
-
-```{r rho_a, fig.cap="Variation between units can be explained either by high correlation and high variance, or by low correlation and low variance.
-As such, the AR1 hyperparameters had correlated posteriors.
-It is this correlation which we made use of with PCA-AGHQ."}
-knitr::include_graphics("depends/rho_a.png")
-```
-
-```{r alpha_x, fig.cap="In contrast to the AR1 hyperparameters, the BYM2 hyperparameter posteriors are much less correlated."}
-knitr::include_graphics("depends/alpha_x.png")
-```
-
-```{r ks-ess, fig.cap="The ESS was negatively correlated with KS test statistic for both TMB and PCA-AGHQ.
-This could be explained in two ways.
-First, harder to sample parameters could be more difficult to estimate with approximate methods.
-Second, as more effective samples are collected, the NUTS posterior could become closer to the approximate posteriors."}
+```{r ks-ess, fig.cap="(ref:ks-ess)"}
 knitr::include_graphics("depends/ks-ess.png")
@@
 
-# Algorithm using Laplace latent field marginals\label{sec:laplace-marginals}
-
-1. Calculate the mode, Hessian at the mode, lower Cholesky, and Laplace approximation
-\begin{align}
-\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
-\hat{\mathbf{H}} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta}, \\
-\hat{\mathbf{H}}^{-1} &= \hat{\mathbf{L}} \hat{\mathbf{L}}^\top, \\
-\tilde p_\texttt{LA}(\btheta, \y) &= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)},
-\end{align}
-where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\mathbf{H}}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
-\begin{align}
-\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
-\hat{\mathbf{H}}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
-\end{align}
-2. Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension.
-Adapt these nodes based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
-Use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
-\begin{equation}
-\tilde p_{\texttt{AQ}}(\y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\btheta(\bu), \y) \omega(\bu). \label{eq:evidence1}
-\end{equation}
-3.
-For $i \in [N]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\hat{\mathbf{H}}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
-A value of $l \geq 4$ is recommended to enable B-spline interpolation.
-For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
-\begin{align}
-\hat \x_{-i}(x_i, \btheta) &= \argmax_{\x_{-i}} \log p(\y, x_i, \x_{-i}, \btheta), \\
-\hat{\mathbf{H}}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
-\end{align}
-where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ can be initialised at $\hat \x(\btheta)_{-i}$.
-4. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
-\begin{equation}
-\end{equation}
-where
-$$
-\tilde p_\texttt{LA}(x_i, \y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(x_i, \btheta(\bu), \y) \omega(\bu).
-$$
-and
-$$
-\tilde p_\texttt{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\texttt{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
-$$
-Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence given in Equation \ref{eq:evidence1} it is more numerically accurate, and requires little extra computation, to use the estimate
-$$
-\tilde p_{\texttt{AQ}}(\y) = \sum_{\bv \in \mathcal{Q}(1, l)} \tilde p_\texttt{LA}(x_i(\bv), \y) \omega(\bv)
-$$
-
-5.
-Given $\{x_i(\bv), \tilde p_\texttt{AQ}(x_i(\bv) \, | \, \y)\}_{\bv \in \mathcal{Q}(1, l)}$ create a spline interpolant to each posterior marginal on the log-scale.
-Samples, and thereby relevant posterior marginal summaries, may be obtained using inverse transform sampling.
-
-\clearpage
-
 # References {#references .unnumbered}

### 2024-12-15 src/docs_paper-jtb/appendix.Rmd
Add more sections and figures!

@@
 This section describes the simplified version of the Naomi model [@eaton2021naomi] in more detail.
-The concise $i$ indexing used in Section \@ref(naomi) is replaced by a more complete $x, s, a$ indexing.
+The concise $i$ indexing is replaced by a more complete $x, s, a$ indexing.
 There are four sections:
@@
 1. Section \@ref(naomi-process) gives the process specifications, giving the terms in each structured additive predictor, along with their distributions.
-2. Section \@ref(naomi-likelihood) gives additional details about the likelihood terms not provided in Section \@ref(naomi).
+2. Section \@ref(naomi-likelihood) gives additional details about the likelihood terms.
 3. Section \@ref(naomi-identifiability) gives identifiability constraints used in circumstances where incomplete data is available for the country.
@@
 In this case, the average age-sex pattern from the Spectrum is used.
-For the Malawi case-study (Section \@ref(malawi)), HIV prevalence and ART coverage data are not available for those aged 65+.
+For the Malawi case-study, HIV prevalence and ART coverage data are not available for those aged 65+.
 As a result, there are $|\{\text{0-4}, \ldots, \text{50-54}\}| = 13$ age groups included for the age random effects.
@@
 
+```{r}
+ar1_bym2_cor <- readr::read_csv("depends/ar1-bym2-cor.csv")
+
+ar1_bym2_cor_grouped <- ar1_bym2_cor %>%
+  dplyr::group_by(type) %>%
+  dplyr::summarise(cor_mean = mean(abs(cor)))
+```
+
+Correlation structure in the posterior can result in sampler inefficiency.
+Each of the four pairs of AR1 log standard deviation $\log(\sigma)$ and logit lag-one autocorrelation parameter $\text{logit}(\phi)$ posteriors were positively correlated (mean absolute correlation `r round(ar1_bym2_cor_grouped$cor_mean[1], 2)`, Figure \@ref(fig:rho-a)).
+These parameters are partially identifiable as variation can either be explained by high standard deviation and high autocorrelation or low standard deviation and low autocorrelation.
+On the other hand, the BYM2 log standard deviation $\log(\sigma)$ and logit proportion parameter $\text{logit}(\phi)$ were, as designed, more orthogonal (mean absolute correlation `r round(ar1_bym2_cor_grouped$cor_mean[2], 2)`, Figure \@ref(fig:alpha-x)).
+
 (ref:rho-a) Pairs plots for the parameters $\log(\sigma_{A}^\rho)$ and $\text{logit}(\phi_{A}^\rho)$, or `log_sigma_rho_a` and `logit_phi_rho_a` as implemented in code.
@@
 ```
+
+The informativeness of data about a parameter can be summarised by the posterior contraction [@schad2021toward] which compares the prior variance $\mathbb{V}_\text{prior}(\phi)$ to posterior variance $\mathbb{V}_\text{post}(\phi)$ via
+\begin{equation}
+c(\phi) = 1 - \frac{\mathbb{V}_\text{prior}}{\mathbb{V}_\text{post}(\phi)}.
+\end{equation}
+Posterior variances were extracted from NUTS results, and prior variances obtained by simulating from a model with the likelihood component removed (Figure \@ref(fig:nuts-hand-comparison)).
+The average posterior contraction was positive for all latent field parameter vectors, and for the majority of hyperparameters (Figure \@ref(fig:posterior-contraction)).
+However, for seven hyperparameters the posterior contraction was very close to zero.
+Furthermore, for some latent field parameter vectors, the average contraction was small.
+Based on this findings, these parameters may not be identifiable.
 
@@
 
+(ref:pc-loadings) Each principal component loading, obtained by the eigendecomposition of the inverse curvature, gives the direction of maximum variation conditional on inclusion of each previous principal component loading.
+For example, the first principal component loading is a sum of `log_sigma_alpha_as` and `logit_phi_alpha_as`.
+
+```{r pc-loadings, fig.cap="(ref:pc-loadings)"}
+knitr::include_graphics("depends/pc-loadings.png")
+```
+
 \clearpage

### 2024-12-15 src/docs_paper-jtb/paper.Rmd
Add more sections and figures!

@@
 
-The Naomi small-area estimation model [@eaton2021naomi] addresses these challenges by synthesising data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
+The Naomi small-area estimation model [@eaton2021naomi; @esra2024improved] addresses these challenges by synthesising data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
 Modelling multiple data sources jointly mitigates the limitations of any single source and increases statistical power.
@@
 
+### Template Model Builder
+
+Template Model Builder [`TMB`; @kristensen2016tmb] is an R package which implements the Laplace approximation.
+In `TMB`, derivatives are obtained using automatic differentiation [AD; @baydin2017automatic].
+The approach of AD is to decompose any function into a sequence of elementary operations with known derivatives.
+The known derivatives of the elementary operations may then be composed by repeat use of the chain rule to obtain the function's derivative.
+A review of AD and how it can be efficiently implemented is provided by @margossian2019review.
+`TMB` uses the C++ package `CppAD` [@cppaddocumentation] for AD [Section 3; @kristensen2016tmb].
+The development of `TMB` was strongly inspired by the Automatic Differentiation Model Builder [ADMB; @fournier2012ad; @bolker2013strategies] project.
+An algorithm is used in `TMB` to automatically determine matrix sparsity structure [Section 4.2; @kristensen2016tmb].
+The R package `Matrix` and C++ package `Eigen` are then used for sparse and dense matrix calculations.
+@kristensen2016tmb highlight the modular design philosophy of `TMB`.
+A review of the use of `TMB` for spatial modelling, including comparison to `R-INLA`, is provided by @osgood2023statistical.
+
+Models are specified in `TMB` using a C++ template file which evaluates $\log p(\mathbf{y}, \mathbf{x}, \boldsymbol{\mathbf{\theta}})$ in a Bayesian context or $\log p(\mathbf{y} \, | \, \mathbf{x}, \boldsymbol{\mathbf{\theta}})$ in a frequentist setting.
+Other software packages have been developed which also use `TMB` C++ templates.
+The `tmbstan` R package [@monnahan2018no] allows running the Hamiltonian Monte Carlo (HMC) algorithm via `Stan`.
+The `aghq` R package [@stringer2021implementing] allows use of AGHQ, and AGHQ over the marginal Laplace approximation, via the `mvQuad` R package [@weiser2016mvquad].
+The `glmmTMB` R package [@brooks2017glmmTMB] allows specification of common GLMM models via a formula interface.
+It is also possible to extract the `TMB` objective function used by `glmmTMB`, which may then be passed into `aghq` or `tmbstan`.
+
 # Simplified Naomi model\label{sec:naomi}
@@
 While this version omits nowcasting and temporal projection, as these time points involve limited inference we expect conclusions for the simplified model to be applicable to the complete model.
-An overview of the simplified model is given below, and Appendix \ref{sec:math} provides a full specification.
+An overview of the simplified model is given below, and a full specification is provided in the Appendix.
 
@@
 HIV incidence rate is modelled by $\log(\lambda_i) = \eta^\lambda_i$ and depends on adult HIV prevalence and adult ART coverage.
-The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix \ref{sec:math}.
+The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in the Appendix.
 The proportion recently infected $\kappa_i$ is linked to HIV incidence via
@@
 
-```{r}
-#' It's a shame here that I have to put "seconds", "hours" and "days" manually and it's not stored in the time taken
-time_taken <- readr::read_csv("depends/time-taken.csv")
-```
-
 We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi.
@@
 \midrule
-TMB & \texttt{TMB} & $1000$ samples \\
-PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-use}), 1000 samples \\
+GEB & \texttt{TMB} & $1000$ samples \\
+GPCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$, 1000 samples \\
 NUTS & \texttt{tmbstan} & 4 chains of 100000 iterations, with the first 50000 iterations of each chain discarded as warmup, thinned by a factor of 20, to give a total of 10000 samples kept.
@@
 
+## Time taken
+
+```{r naomi-time}
+time_taken <- readr::read_csv("depends/time-taken.csv")
+
+time_taken$aghq <- time_taken$aghq * 60 * 60 
+time_taken$tmbstan <- time_taken$tmbstan * 60 * 60 * 24
+
+time_taken_df <- as.data.frame(t(time_taken)) %>%
+  tibble::rownames_to_column("method") %>%
+  mutate(
+    mins = V1 / 60,
+    hours = mins / 60,
+    days = hours / 24,
+    method = case_when(
+      method == "TMB" ~ "GEB",
+      method == "aghq" ~ "GPCA-AGHQ",
+      method == "tmbstan" ~ "NUTS"
+    ),
+    software = case_when(
+      method == "NUTS" ~ "tmbstan",
+      TRUE ~ "TMB"
+    )
+  )
+
+naomi_time_tmbstan_hours <- signif(filter(time_taken_df, method == "NUTS")$hours, 2)
+naomi_time_aghq_hours <- signif(filter(time_taken_df, method == "GPCA-AGHQ")$hours, 2)
+naomi_time_tmb_minutes <- signif(filter(time_taken_df, method == "GEB")$mins, 1)
+```
+
+Inference with NUTS took `r naomi_time_tmbstan_hours` hours, while inference with GPCA-AGHQ took `r naomi_time_aghq_hours` hours and GEB just `r naomi_time_tmb_minutes` minutes.
+Both the NUTS and GPCA-AGHQ algorithms can be run under a range of settings, trading off accuracy and runtime.
+
+## NUTS convergence and suitability
+
+```{r}
+naomi_nuts <- readRDS("depends/mcmc-out.rds")
+```
+
+The Naomi model was difficult to efficiently sample from using NUTS via `tmbstan`.
+The lowest effective sample size was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Panel \@ref(fig:worst-trace)A).
+The largest potential scale reduction factor was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Panel \@ref(fig:worst-trace)B).
+Though inaccuracies remain possible, these diagnostics are sufficient to treat inferences obtained from NUTS as a gold-standard.
+
+## Use of PCA-AGHQ
+
+```{r}
+tv_df <- readr::read_csv("depends/total-variation.csv")
+```
+
+(ref:total-variation) Under PCA, the proportion of total variation explained is given by the sum of the first $s$ eigenvalues over the sum of all eigenvalues.
+A typical rule-of-thumb is to include dimensions sufficient to explain 90% of total variation.
+In this case, for computational reasons, `r round(100 * tv_df$tv[8])`% was considered sufficient.
+
+```{r total-variation, fig.cap="(ref:total-variation)"}
+knitr::include_graphics("depends/total-variation.png")
+```
+
+```{r reduced-rank, fig.cap="The full rank original covariance matrix (Panel A) was closely reproduced by its reduced rank ($s = 8$) matrix approximation (Panel B)."}
+knitr::include_graphics("depends/reduced-rank.png")
+```
+
+For the PCA-AGHQ quadrature grid, a Scree plot based on the spectral decomposition of $\hat {\mathbf{H}}_\texttt{LA}^{-1}$ (as defined in Equation \@ref(eq:hessian2)) was used to select the number of principal components to keep (Figure \ref{fig:total-variation}).
+Keeping $s = 8$ principal components was sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
+The reduced rank approximation to the inverse curvature with this choice of $s$ was visually similar to the full rank matrix (Figure \@ref(fig:reduced-rank)).
+
+The principal component (PC) loadings (Figure \{fig:pc-loadings}) provide interpretable information about which directions had the greatest variation.
+Many of the first PC loadings are sums of two hyperparameters.
+As such, there is some redundancy in the hyperparameter parameterisation, supporting the findings of Section \@ref(naomi-nuts) regarding correlation structure in the hyperparameter posterior.
+It is exactly this correlation structure that PCA, and PCA-AGHQ, looks to utilise.
+
+```{r nodes-samples-comparison, fig.cap="The grey histograms show the 24 hyperparameter marginal distributions obtained with NUTS.
+The green lines indicate the position of the 6561 PCA-AGHQ nodes projected onto each hyperparameter marginal.
+For some hyperparameters, the PCA-AGHQ nodes vary over the domain of the posterior marginal distribution, while for others they concentrate at the mode."}
+knitr::include_graphics("depends/nodes-samples-comparison.png")
+```
+
+Projecting the $3^8 = 6561$ PCA-AGHQ quadrature nodes onto each hyperparameter dimension, there was substantial variation in coverage by hyperparameter (Figure \ref{fig:nodes-samples-comparison}).
+Approximately 12 hyperparameters had well covered marginals: greater than the 8 naively obtained with a dense grid, but nonetheless far fewer than the full 24.
+Coverage was higher among hyperparameters on the logistic scale, and lower among hyperparameters on the logarithmic scale.
+This discrepancy occurred due to logistic hyperparameters naturally having higher posterior marginal standard deviation than logarithmic hyperparameters (Figure \@ref(fig:marginal-sd)).
+
+## Inference comparison
+
+Posterior inferences from GEB, GPCA-AGHQ and NUTS were compared using point estimates (Section \@ref(point-estimates)), distributional quantities (Section \@ref(distributional-quantities)), and exceedance probabilities (Section \@ref(exceedance))
+
+### Point estimates \label{sec:point-estimates}
+
+### Distributional quantities\label{sec:distributional-quantities}
+
+### Exceedance probabilities\label{sec:exceedance}
+
 # Conclusion\label{sec:conclusions}

### 2024-12-15 src/docs_paper-jtb/paper.Rmd
Fix to figure caption

@@
 The Naomi model was difficult to efficiently sample from using NUTS via `tmbstan`.
-The lowest effective sample size was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Panel \@ref(fig:worst-trace)A).
-The largest potential scale reduction factor was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Panel \@ref(fig:worst-trace)B).
+The lowest effective sample size was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Panel XA).
+The largest potential scale reduction factor was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Panel XB).
 Though inaccuracies remain possible, these diagnostics are sufficient to treat inferences obtained from NUTS as a gold-standard.
@@
 
-(ref:total-variation) Under PCA, the proportion of total variation explained is given by the sum of the first $s$ eigenvalues over the sum of all eigenvalues.
+```{r total-variation, fig.cap=paste0("Under PCA, the proportion of total variation explained is given by the sum of the first $s$ eigenvalues over the sum of all eigenvalues.
 A typical rule-of-thumb is to include dimensions sufficient to explain 90% of total variation.
-In this case, for computational reasons, `r round(100 * tv_df$tv[8])`% was considered sufficient.
-
-```{r total-variation, fig.cap="(ref:total-variation)"}
+In this case, for computational reasons, ", round(100 * tv_df$tv[8]), "% was considered sufficient.")}
 knitr::include_graphics("depends/total-variation.png")
@@
 
-For the PCA-AGHQ quadrature grid, a Scree plot based on the spectral decomposition of $\hat {\mathbf{H}}_\texttt{LA}^{-1}$ (as defined in Equation \@ref(eq:hessian2)) was used to select the number of principal components to keep (Figure \ref{fig:total-variation}).
-Keeping $s = 8$ principal components was sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
-The reduced rank approximation to the inverse curvature with this choice of $s$ was visually similar to the full rank matrix (Figure \@ref(fig:reduced-rank)).
-
-The principal component (PC) loadings (Figure \{fig:pc-loadings}) provide interpretable information about which directions had the greatest variation.
-Many of the first PC loadings are sums of two hyperparameters.
-As such, there is some redundancy in the hyperparameter parameterisation, supporting the findings of Section \@ref(naomi-nuts) regarding correlation structure in the hyperparameter posterior.
-It is exactly this correlation structure that PCA, and PCA-AGHQ, looks to utilise.
-
 ```{r nodes-samples-comparison, fig.cap="The grey histograms show the 24 hyperparameter marginal distributions obtained with NUTS.
@@
 
+For the PCA-AGHQ quadrature grid, a Scree plot based on the spectral decomposition of $\hat {\mathbf{H}}_\texttt{LA}^{-1}$ was used to select the number of principal components to keep (Figure \ref{fig:total-variation}).
+Keeping $s = 8$ principal components was sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
+The reduced rank approximation to the inverse curvature with this choice of $s$ was visually similar to the full rank matrix (Figure \ref{fig:reduced-rank}).
+
+The principal component (PC) loadings (Figure Appendix) provide interpretable information about which directions had the greatest variation.
+Many of the first PC loadings are sums of two hyperparameters.
+As such, there is some redundancy in the hyperparameter parameterisation, supporting the findings of the Appendix regarding correlation structure in the hyperparameter posterior.
+It is exactly this correlation structure that PCA, and PCA-AGHQ, looks to utilise.
+
 Projecting the $3^8 = 6561$ PCA-AGHQ quadrature nodes onto each hyperparameter dimension, there was substantial variation in coverage by hyperparameter (Figure \ref{fig:nodes-samples-comparison}).
@@
 Coverage was higher among hyperparameters on the logistic scale, and lower among hyperparameters on the logarithmic scale.
-This discrepancy occurred due to logistic hyperparameters naturally having higher posterior marginal standard deviation than logarithmic hyperparameters (Figure \@ref(fig:marginal-sd)).
+This discrepancy occurred due to logistic hyperparameters naturally having higher posterior marginal standard deviation than logarithmic hyperparameters (Figure Appendix).
 
@@
 
-Posterior inferences from GEB, GPCA-AGHQ and NUTS were compared using point estimates (Section \@ref(point-estimates)), distributional quantities (Section \@ref(distributional-quantities)), and exceedance probabilities (Section \@ref(exceedance))
+Posterior inferences from GEB, GPCA-AGHQ and NUTS were compared using point estimates (Section \ref{sec:point-estimates}), distributional quantities (Section \ref{sec:distributional-quantities}), and exceedance probabilities (Section \ref{sec:exceedance}).
 

### 2024-12-22 src/docs_paper-jtb/paper.Rmd
Some Alex comments

@@
 AGHQ is a quadrature method based on the theory of polynomial interpolation, and is well suited to statistical estimation problems in which the integrand is well approximated by a Gaussian multiplied by a polynomial.
-For example, @bilodeau2022stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
+For example, @bilodeau2024stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
 However, it is not computationally feasible to use AGHQ in >20 dimensions directly, as exponentially many nodes are required.
 Instead, we used principal components analysis (PCA) of the inverse curvature at the mode to find a subspace which explained most of the hyperparameter variance.
-In an application to Malawi, this resulted in a grid which was tractable as it had millions of times fewer nodes the corresponding dense grid.
+In an application to Malawi, this resulted in a grid which was tractable as it had millions of times fewer nodes than the corresponding dense grid.
 Our implementation of the method makes use of the existing Naomi `TMB` template, and is immediately compatible with any model with such a template.
@@
 
-Models are specified in `TMB` using a C++ template file which evaluates $\log p(\mathbf{y}, \mathbf{x}, \boldsymbol{\mathbf{\theta}})$ in a Bayesian context or $\log p(\mathbf{y} \, | \, \mathbf{x}, \boldsymbol{\mathbf{\theta}})$ in a frequentist setting.
+Models are specified in `TMB` using a C++ template file which evaluates $\log p(\mathbf{y}, \mathbf{x}, \boldsymbol{\mathbf{\theta}})$.
 Other software packages have been developed which also use `TMB` C++ templates.

### 2024-12-28 src/docs_paper-jtb/appendix.Rmd
Editing and further material in

@@
 The concise $i$ indexing is replaced by a more complete $x, s, a$ indexing.
-There are four sections:
+There are four sections, as follows:
 
-1. Section \@ref(naomi-process) gives the process specifications, giving the terms in each structured additive predictor, along with their distributions.
+1. Section \@ref(naomi-process) gives the process specifications, including the terms in each structured additive predictor, along with their distributions.
 2. Section \@ref(naomi-likelihood) gives additional details about the likelihood terms.
@@
 Table: (\#tab:process) The Naomi model can be conceptualised as having five processes.
-This table gives the number of latent field parameters and hyperparameters in each process, where $n$ is the number of districts in the country.
+This table gives the number of latent field parameters and hyperparameters in each model component, where $n$ is the number of districts in the country.
 
@@
 \begin{equation}
-\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + \mathbf{u}^\alpha_a + \mathbf{u}_a^{\alpha, s = \text{M}} + \mathbf{u}^\alpha_x + \mathbf{u}_x^{\alpha, s = \text{M}} + \mathbf{u}_x^{\alpha, a < 15} + \boldsymbol{\mathbf{\eta}}^\alpha_{R_x, s, a}
+\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + \mathbf{u}^\alpha_a + \mathbf{u}_a^{\alpha, s = \text{M}} + \mathbf{u}^\alpha_x + \mathbf{u}_x^{\alpha, s = \text{M}} + \mathbf{u}_x^{\alpha, a < 15} + \boldsymbol{\mathbf{\eta}}^\alpha_{R_x, s, a},
 \end{equation}
@@
 The `TMB` C++ code for the negative log-posterior of the simplified Naomi model is available from [`https://github.com/athowes/naomi-aghq`](https://github.com/athowes/naomi-aghq).
-For ease of understanding, Table \@ref(tab:tmb-lookup) provides correspondence between the mathematical notation used in Section \@ref(naomi-math) and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
+Table \@ref(tab:tmb-lookup) maps the mathematical notation used in Section \@ref(naomi-math) to the variable names used in the `TMB` code for all hyperparameters and latent field parameters.
 For further reference on the `TMB` software see @tmbdocumentation.
@@
 
+(ref:ess) The efficiency of the NUTS, as measured by the ratio of effective sample size to total number of iterations run, was low for most parameters (Panel A).
+As a result, the number of iterations required for the the effective number of samples (mean `r round(naomi_nuts$ess_mean)`) to be satisfactory was high (Panel B).
+
+```{r ess, fig.cap="(ref:ess)"}
+knitr::include_graphics("depends/ess.png")
+```
+
 (ref:rhat) For NUTS run on the Naomi ELGM, the maximum potential scale reduction factor was `r round(naomi_nuts$rhat_max, 3)`, below the value of 1.05 typically used as a cutoff for acceptable chain mixing, indicating that the results are acceptable to use.
@@
 
-(ref:ess) The efficiency of the NUTS, as measured by the ratio of effective sample size to total number of iterations run, was low for most parameters (Panel A).
-As a result, the number of iterations required for the the effective number of samples (mean `r round(naomi_nuts$ess_mean)`) to be satisfactory was high (Panel B).
-
-```{r ess, fig.cap="(ref:ess)"}
-knitr::include_graphics("depends/ess.png")
-```
-
 (ref:worst-trace) Traceplots for the parameter with the lowest ESS which was `log_sigma_alpha_xs` (an $\text{ESS}$ of `r round(naomi_nuts$ess_min)`, Panel A) and highest potential scale reduction factor which was `ui_lambda_x[10]` (an $\hat R$ of `r round(naomi_nuts$rhat_max, 3)`, Panel B).
@@
 ```
+
+## Posterior correlation structure
 
@@
 ```
+
+## Posterior contraction
 
@@
 
+(ref:pc-loadings) Each principal component loading, obtained by the eigendecomposition of the inverse curvature, gives the direction of maximum variation conditional on inclusion of each previous principal component loading.
+For example, the first principal component loading is a sum of `log_sigma_alpha_as` and `logit_phi_alpha_as`.
+
+```{r pc-loadings, fig.cap="(ref:pc-loadings)"}
+knitr::include_graphics("depends/pc-loadings.png")
+```
+
+(ref:marginal-sd) The estimated posterior marginal standard deviation of each hyperparameter varied substantially based on its scale, either logarithmic or logistic.
+
+```{r marginal-sd, fig.cap="(ref:marginal-sd)"}
+knitr::include_graphics("depends/marginal-sd.png")
+```
+
 ```{r}
@@
 
-(ref:marginal-sd) The estimated posterior marginal standard deviation of each hyperparameter varied substantially based on its scale, either logarithmic or logistic.
-
-```{r marginal-sd, fig.cap="(ref:marginal-sd)"}
-knitr::include_graphics("depends/marginal-sd.png")
-```
-
 (ref:lognormconst) The logarithm of the normalising constant estimated using PCA-AGHQ and a range of possible values of $k = 2, 3, 5$ and $s \leq 8$.
@@
 knitr::include_graphics("depends/lognormconst.png")
-```
-
-(ref:pc-loadings) Each principal component loading, obtained by the eigendecomposition of the inverse curvature, gives the direction of maximum variation conditional on inclusion of each previous principal component loading.
-For example, the first principal component loading is a sum of `log_sigma_alpha_as` and `logit_phi_alpha_as`.
-
-```{r pc-loadings, fig.cap="(ref:pc-loadings)"}
-knitr::include_graphics("depends/pc-loadings.png")
 ```

### 2024-12-28 src/docs_paper-jtb/paper.Rmd
Editing and further material in

@@
 The model is provided as a tool for countries to input their data to and generate estimates with during a yearly process supported by UNAIDS.
-Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package.
-We propose a new inference method based on an extension of adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters.
+Previously, inference has been conducted using empirical Bayes and a Gaussian approximation, implemented via the \texttt{TMB} \textsc{R} package.
+We propose a new inference method based on an extension of adaptive Gauss-Hermite quadrature to deal with more than 20 hyperparameters.
 Using data from Malawi, our method improves the accuracy of inferences for model parameters, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
@@
 Generation of estimates by country teams is an important and distinctive feature of the HIV response.
-Drawing on expert knowledge of the data improves the accuracy of the process, as well as strengthening trust in the resulting estimates, creating a virtuous cycle of data quality, use and ownership [@noor2022country].
+Drawing on expert knowledge of the data improves the accuracy of the process and strengthens trust in the resulting estimates, creating a virtuous cycle of data quality, use, and ownership [@noor2022country].
 
-Naomi is comprised of multiple linked generalized linear mixed models (GLMMs) and presents a challenging Bayesian inference problem.
+Naomi comprises multiple linked generalized linear mixed models (GLMMs) and presents a challenging Bayesian inference problem.
 The model contains hundreds of fixed and random effect parameters and over 20 hyperparameters.
 This is substantially more than the small number that can typically be handled by approaches like integrated nested Laplace approximations [INLA; @rue2009approximate].
-Moreover, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
+Additionally, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
 Inference must be fast and have low memory usage, to allow interactive review and iteration of model results by workshop participants.
 The scale of the model and features of its posterior geometry [@neal2003slice] make Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
-Inference must be reliable and automatic across a range of country settings, and shouldn't require close monitoring and diagnostics (e.g. for MCMC convergence).
+Inference must be reliable and automatic across a range of country settings, and should not require close monitoring and diagnostics (e.g. for MCMC convergence).
 
@@
 
-Although the EB approach is fast, it does not take into account hyperparameter uncertainty in the latent field posterior, potentially resutling in underestimated posterior variances.
+Although the EB approach is fast, it does not take into account hyperparameter uncertainty in the latent field posterior, potentially resulting in underestimated posterior variances.
 This concern could have important practical implications for use of the estimates from the Naomi model, and motivated us to look for an approach closer full Bayesian inference.
@@
 For example, @bilodeau2024stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
-However, it is not computationally feasible to use AGHQ in >20 dimensions directly, as exponentially many nodes are required.
+However, it is not computationally feasible to use AGHQ in greater than 20 dimensions directly, as exponentially many nodes are required.
 Instead, we used principal components analysis (PCA) of the inverse curvature at the mode to find a subspace which explained most of the hyperparameter variance.
@@
 The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
-It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-1})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
+It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
 Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.
@@
 
-### Template Model Builder
+## Template Model Builder
 
@@
 The index $j$ runs across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
-We take the weighted observed number of outcomes to be $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
+The weighted observed number of outcomes is taken to be $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
 \begin{equation*}
@@
 These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
-We take the number of ANC clients with ascertained status to be fixed as $m^{\rho^\text{ANC}}_I$. 
-We then model the number of those with positive status $y^{\rho^\text{ANC}}_I$, and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ using nested binomial likelihoods
+The number of ANC clients with ascertained status to be fixed as $m^{\rho^\text{ANC}}_I$. 
+The number of those with positive status $y^{\rho^\text{ANC}}_I$, and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ are modelled using nested binomial likelihoods
 \begin{align*}
@@
 \end{align*}
-As in the household survey component, we use weighted aggregates
+As in the household survey component, weighted aggregates
 \begin{equation*}
@@
 \end{equation*}
-with $\Psi_i$ the number of pregnant women, which we assume to be fixed.
+are used, with $\Psi_i$ the number of pregnant women assumed here to be fixed.
 
@@
 These data provide helpful information about HIV prevalence and coverage of ART, but are challenging to use because people living with HIV sometimes choose to access ART services outside of the district that they reside in.
-We model the probabilities of accessing services outside the home district using multinomial logistic regressions.
+These probabilities of accessing services outside the home district are modelled using multinomial logistic regressions.
 
 Briefly, let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring such that $x \sim x'$.
-We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$.
-As such, we assume travel to each neighbouring district, for all age-sex strata, is equally likely.
-We then model aggregate ART attendance data $y^{N^\text{ART}}_I$ using a Gaussian approximation to a sum of binomials.
+The log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ are modelled using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$.
+As such, travel to each neighbouring district, for all age-sex strata, is assumed to be equally likely.
+The aggregate ART attendance data $y^{N^\text{ART}}_I$ are modelled using a Gaussian approximation to a sum of binomials.
 This sum is over both strata $i \in I$ and the number of ART clients travelling from district $x'$ to $x$.
@@
 
-The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ in an ELGM is given by
+The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ and hyperparameter prior distribution $p(\btheta)$ in an ELGM is given by
 \begin{equation*}
@@
 \end{align}
-Then the Laplace approximation [@tierney1986accurate] to $p(\btheta, \y)$ is given by
+Then the Laplace approximation [@naylor1982applications] to $p(\btheta, \y)$ is given by
 \begin{equation}
@@
 
-Let $\z \in \mathcal{Q}(m, k)$ be $m$-dimensional Gauss-Hermite quadrature [GHQ; @davis1975methods] rule with $k$ nodes per dimension constructed using the product rule $\mathcal{Q}(m, k) = \mathcal{Q}(1, k) \times \cdots \times \mathcal{Q}(1, k)$ where
+Let $\z \in \mathcal{Q}(m, k)$ be an $m$-dimensional Gauss-Hermite quadrature [GHQ; @davis1975methods] rule with $k$ nodes per dimension constructed using the product rule.
+In particular, $\mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ where
 \begin{equation}
-\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}, \\
+\mathcal{Q}(1, k) = \{z: H_k(z) = 0\} \\
 \end{equation}
-with $\phi(\cdot)$ is a standard Gaussian density.
-The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$.
+are the zeroes of the $k$th Hermite polynomial $H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2)$.
+The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k!
+/ [H_{k + 1}(z)]^2 \phi(z)$, and $\phi(\cdot)$ is a standard Gaussian density.
 GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.
@@
 Use of the product rule grid described above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points which quickly becomes intractable as $m$ increases for $k > 1$.
-An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
-We may then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ to be a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
+We propose to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
+Then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ as a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
 Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
@@
 We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi.
-Data from Malawi has previously been used to demonstrate the Naomi model, including as a part of the `naomi` R package vignette.
-Malawi was chosen in part because it has a small number of districts, $n = 30$, limiting the computational demand of the model.
+Data from Malawi, which has $n = 30$ districts, has previously been used to demonstrate the Naomi model, including as a part of the `naomi` R package vignette.
 Three Bayesian inference approaches were considered:
@@
 The Naomi model was difficult to efficiently sample from using NUTS via `tmbstan`.
-The lowest effective sample size was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Panel XA).
-The largest potential scale reduction factor was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Panel XB).
+The lowest effective sample size (ESS) was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Supplementary Figure 1).
+The largest potential scale reduction factor ($\hat R$) was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Supplementary Figure 2A).
+Supplementary Figure 3 shows traceplots for the parameters with the lowest ESS and $\hat R$ respectively.
 Though inaccuracies remain possible, these diagnostics are sufficient to treat inferences obtained from NUTS as a gold-standard.
@@
 
+For the PCA-AGHQ quadrature grid, a Scree plot based on the spectral decomposition of $\hat {\mathbf{H}}_\texttt{LA}^{-1}$ was used to select the number of principal components to keep (Figure \ref{fig:total-variation}).
+Keeping $s = 8$ principal components was sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
+The reduced rank approximation to the inverse curvature with this choice of $s$ was visually similar to the full rank matrix (Figure \ref{fig:reduced-rank}).
+
 ```{r reduced-rank, fig.cap="The full rank original covariance matrix (Panel A) was closely reproduced by its reduced rank ($s = 8$) matrix approximation (Panel B)."}
@@
 
-For the PCA-AGHQ quadrature grid, a Scree plot based on the spectral decomposition of $\hat {\mathbf{H}}_\texttt{LA}^{-1}$ was used to select the number of principal components to keep (Figure \ref{fig:total-variation}).
-Keeping $s = 8$ principal components was sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
-The reduced rank approximation to the inverse curvature with this choice of $s$ was visually similar to the full rank matrix (Figure \ref{fig:reduced-rank}).
-
-The principal component (PC) loadings (Figure Appendix) provide interpretable information about which directions had the greatest variation.
-Many of the first PC loadings are sums of two hyperparameters.
-As such, there is some redundancy in the hyperparameter parameterisation, supporting the findings of the Appendix regarding correlation structure in the hyperparameter posterior.
-It is exactly this correlation structure that PCA, and PCA-AGHQ, looks to utilise.
+The principal component (PC) loadings (Supplementary Figure 8) provide interpretable information about which directions had the greatest variation.
+Several of the first $s = 8$ PC loadings are sums of two hyperparameters, indicating some redundancy in the hyperparameter parameterisation.
+This finding is supported analysis of correlation structure in the NUTS hyperparameter posterior (Appendix 2.1).
+It is exactly this correlation structure that PCA, and PCA-AGHQ, makes use of.
 
@@
 Coverage was higher among hyperparameters on the logistic scale, and lower among hyperparameters on the logarithmic scale.
-This discrepancy occurred due to logistic hyperparameters naturally having higher posterior marginal standard deviation than logarithmic hyperparameters (Figure Appendix).
+This discrepancy occurred due to logistic hyperparameters naturally having higher posterior marginal standard deviation than logarithmic hyperparameters (Supplementary Figures 9 and 10).
 
@@
 
+#### Meeting the second 90
+
+Ambitious targets for scaling up ART treatment have been developed by UNAIDS, with the goal of ending the AIDS epidemic by 2030 [@unaids2014target].
+Meeting the 90-90-90 fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have a suppressed viral load.
+Inferences from Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met, that is $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
+
+(ref:exceedance-second90) The probability each strata has met the second 90 (ART coverage above 81\%) calculated using each inference method, as compared with NUTS.
+The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
+
+```{r exceedance-second90, fig.cap="(ref:exceedance-second90)"}
+knitr::include_graphics("depends/exceedance-second90.png")
+```
+
+Strata probabilities of having met the second 90 target were more accurately estimated by GPCA-AGHQ than GEB (Figure \@ref(fig:exceedance-second90)).
+Both GPCA-AGHQ and GEB had substantial error as compared to results from NUTS, however, particularly for girls and women.
+This discrepancy in accuracy by sex may be caused by interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.
+
+#### Finding strata with high incidence {#one-inc}
+
+Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as higher than 1% incidence per year.
+Inferences from Naomi can be used to calculate the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
+
+(ref:exceedance-1inc) The probability each strata has high HIV incidence (above 1\% per year) calculated using each inference method, as compared with NUTS.
+The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
+
+```{r exceedance-1inc, fig.cap="(ref:exceedance-1inc)"}
+knitr::include_graphics("depends/exceedance-1inc.png")
+```
+
+GPCA-AGHQ gave more accurate estimates of the probability that a strata has high HIV incidence than GEB (Figure \@ref(fig:exceedance-1inc)).
+Again, both methods had significant error.
+Unlike in Section \@ref(second90), there was little difference in error by sex.
+
 # Conclusion\label{sec:conclusions}
+
+For the simplified Naomi model applied to data from Malawi, GPCA-AGHQ more accurately inferred latent field posterior marginal distributions than GEB.
+However, model output posterior marginals did not see the same improvements.
+Approximate posterior exceedance probabilities from both GEB and GPCA-AGHQ had systematic inaccuracies as compared with NUTS.
+GEB and GPCA-AGHQ were substantially faster than NUTS, which took over two days to reach convergence.
+
+Inaccuracies in model outputs from GEB and GPCA-AGHQ do have potential to meaningfully mislead policy.
+As such, where possible, gold-standard NUTS results should be computed.
+Though NUTS is to slow too run during a workshop, it could be run afterwards.
+As the UNAIDS HIV estimates process occurs annually, requiring days to compute more accurate estimates is viable.
+That said, Malawi is one of the countries with the fewest number of districts.
+As NUTS took days to run in Malawi, for larger countries, with hundreds of districts, it may be impossible to run NUTS to convergence, and approximate methods may be required.
+
+To empower users, GPCA-AGHQ and NUTS could be added to the Naomi web interface as alternatives to GEB.
+Analysts would be able to quickly iterate over model options using GEB, before switching to a more accurate approach once they are happy with the results.
+
+PCA-AGHQ can be adjusted to suit the computational budget available by choice of the number of dimensions kept in the PCA $s$ and the number of points per dimension $k$.
+The scree plot is a well established heuristic for choosing $s$.
+Heuristics for choosing $k$ are less well established.
+Whether it is preferable for a given computational budget to increase $s$ or increase $k$ is an open question.
+Further strategies, such as gradually lowering $k$ over the principal components, could also be considered.
+
+## Suggestions for future work
+
+### Further comparisons {#further-comparisons}
+
+Comparison to further Bayesian inference methods could be included.
+Four possibilities stand out as being particularly valuable:
+
+1. There exist other quadrature rules for moderate dimension, such as the CCD.
+It would be of interest to compare INLA with a PCA-AGHQ rule to INLA with other such quadrature rules.
+2. Rather than use quadrature to integrate the marginal Laplace approximation, an alternative approach is to run HMC [@monnahan2018no; @margossian2020hamiltonian].
+When run to convergence, inferential error of this method would solely be due to the Laplace approximation, helping to clarify the extent to which the inferential error of INLA is attributable to the quadrature grid.
+Preliminary testing of this approach, using `tmbstan` and setting `laplace = TRUE`, did not show immediate success but likely could be worked on.
+3. NUTS is not especially well suited to sampling from Gaussian latent field models like Naomi.
+Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], could be considered.
+It may be difficult to implement such algorithms using `TMB`.
+Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language [@de2017programming].
+Requiring rewriting the Naomi model log-posterior outside of `TMB` would be a substantial downside.
+4. Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
+However, as `inlabru`, like `R-INLA`, is based on a formula interface, it may not be possible to do so directly.
+
+### Better quadrature grids
+
+PCA-AGHQ is a sensible approach to allocating more computational to dimensions which contribute more to the integral in question.
+However, its application to Naomi surfaced instances where it overlooked potential benefits, or otherwise did not behave as one might wish:
+
+1. The amount of variation explained in the Hessian matrix may not be of direct interest.
+For the Naomi model, interest is in the effect of including each dimension on the relevant model outputs.
+As such, using alternative measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] or Sobol indices, could be preferable.
+2. Use of PCA is challenging when the dimensions have different scales. 
+For the Naomi model, logit-scale hyperparameters were systematically favoured over those on the log-scale.
+3. When the quadrature rule is used within an INLA algorithm, it is more important to allocate quadrature nodes to those hyperparameter marginals which are non-Gaussian.
+This is because the Laplace approximation is exact when the integrand is Gaussian, so a single quadrature node is sufficient.
+The difficulty is, of course, knowing in advance which marginals will be non-Gaussian.
+This could be done if there were a cheap way to obtain posterior means, which could then be compared to posterior modes obtained using optimisation.
+Another approach would be to measure the fit of marginal samples from a cheap approximation, like EB.
+The measures of fit would have to be for marginals, ruling out approaches like PSIS [@yao2018yes] which operate on joint distributions.
+4. Finally, it may be possible to achieve better performance by pruning and prerotation, as discussed by @jackel2005note.
+
+#### Statistical theory
+
+The class of functions which are integrated exactly by PCA-AGHQ remains to be shown.
+Theorem 1 of @stringer2022fast bounds the total variation error of AGHQ, establishing convergence in probability of coverage probabilities under the approximate posterior distribution to those under the true posterior distribution.
+Similar theory could be established for PCA-AGHQ, or more generally AGHQ with varying levels.
+The challenge of connecting this theory to nested use of any quadrature rule, like that in the INLA algorithm, remains an important open question.
+
+#### Testing quadrature assumptions
+
+It may be possible to test the assumptions made by use of AGHQ grids, allowing their suitability for a particular integral to be assessed.
+Specifically, AGHQ assumes that the integrand is closely approximated by a polynomial multiplied by a Gaussian density.
+Given NUTS hyperparameter samples (or better yet, hyperparameter samples from the Laplace NUTS hybrid discussed in Section \@ref(further-comparisons)) this assumption could be tested by fitting a model using a polynomial times Gaussian kernel.
+This approach could be generalised to also test the suitability of PCA-AGHQ grids.
 
@@
 AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).
+AS was supported by NSERC Discovery Grant RGPIN-2023-03331.
 SRF was supported by the EPSRC (EP/V002910/2).

### 2024-12-28 src/docs_paper-jtb/paper.Rmd
Add unformatted results content

@@
 
+```{r}
+df_point <- readr::read_csv(paste0("resources/naomi-aghq/", resource_version, "/depends/mean-sd.csv"))
+
+df_point_pct <- df_point %>%
+  ungroup() %>%
+  group_by(indicator, type) %>%
+  summarise(
+    rmse_diff = 100 * diff(rmse) / max(rmse),
+    mae_diff = 100 * diff(mae) / max(mae)
+  )
+
+rmse_aghq_mean <- filter(df_point, method == "GPCA-AGHQ", indicator == "Posterior mean", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_tmb_mean <- filter(df_point, method == "GEB", indicator == "Posterior mean", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean", type == "latent") %>% pull(rmse_diff) %>% signif(2)
+
+rmse_aghq_sd <- filter(df_point, method == "GPCA-AGHQ", indicator == "Posterior SD", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_tmb_sd <- filter(df_point, method == "GEB", indicator == "Posterior SD", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD", type == "latent") %>% pull(rmse_diff) %>% signif(0)
+```
+
+(ref:mean-sd-alt-latent) The latent field posterior mean and posterior standard deviation point estimates from each inference method as compared with those from NUTS.
+The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
+For both the posterior mean and posterior standard deviation, GPCA-AGHQ reduced RMSE and MAE as compared with GEB.
+
+```{r mean-sd-alt-latent, fig.cap="(ref:mean-sd-alt-latent)"}
+knitr::include_graphics(paste0("resources/naomi-aghq/", resource_version, "/depends/mean-sd-alt-latent.png"))
+```
+
+Latent field point estimates obtained from GPCA-AGHQ were closer to the gold-standard results from NUTS than those obtained from GEB (Figure \@ref(fig:mean-sd-alt-latent)).
+The root mean square error (RMSE) between posterior mean estimates from GPCA-AGHQ and NUTS (`r rmse_aghq_mean`) was `r abs(rmse_diff_mean)`% lower than that between GEB and NUTS (`r rmse_tmb_mean`).
+For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_aghq_sd)` (PCA-AGHQ).
+However, puzzlingly, improvements in latent field estimate accuracy only transferred to model outputs to a limited extent (Figures \@ref(fig:mean-alt-output) and \@ref(fig:sd-alt-output)).
+
 ### Distributional quantities\label{sec:distributional-quantities}
+
+##### Kolmogorov-Smirnov
+
+```{r}
+ks_summary <- readRDS(paste0("resources/naomi-aghq/", resource_version, "/depends/ks-summary.rds"))
+
+ks_summary_summarise <- ks_summary %>%
+  ungroup() %>%
+  filter(type == "Latent") %>%
+  summarise(
+    tmb = sum(size * TMB) / sum(size),
+    aghq = sum(size * aghq) / sum(size)
+  )
+
+ks_pct <- (ks_summary_summarise$tmb - ks_summary_summarise$aghq) / ks_summary_summarise$tmb
+```
+
+(ref:ks-summary) The average Kolmogorov-Smirnov (KS) test statistic for each latent field parameter of the Naomi model.
+Vectors of parameters were grouped together.
+For points above the dashed line at zero, performance of GEB was better.
+For points below the dashed line, performance of GPCA-AGHQ was better.
+Most notably, for the latent field parameters `ui_lambda_x` the test statistic for GEB was substantially higher than for GPCA-AGHQ.
+This parameter, of length 32, corresponds to $\mathbf{u}_x^\lambda$ and plays a key role in the ART attendance component of the Naomi (Section \@ref(art)).
+
+```{r ks-summary, fig.cap="(ref:ks-summary)"}
+knitr::include_graphics(paste0("resources/naomi-aghq/", resource_version, "/depends/ks-summary.png"))
+```
+
+```{r}
+ui_lambda_x <- readRDS(paste0("resources/naomi-aghq/", resource_version, "/depends/ui-lambda-x.rds"))
+```
+
+(ref:ui-lambda-x) The parameter ``r paste0(ui_lambda_x$par)`` had the greatest difference in KS test statistics between GEB and GPCA-AGHQ to NUTS.
+For this parameter, the potential scale reduction factor was `r ui_lambda_x$rhat` and effective sample size was `r ui_lambda_x$ess`.
+
+```{r ui-lambda-x, fig.cap="(ref:ui-lambda-x)"}
+knitr::include_graphics(paste0("resources/naomi-aghq/", resource_version, "/depends/ui-lambda-x.png"))
+```
+
+The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\omega) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \omega}$.
+It is a relatively stringent, worst case, measure of distance between empirical distributions.
+The average KS test statistic for GPCA-AGHQ (`r signif(ks_summary_summarise$aghq, 2)`) was `r signif(100 * ks_pct, 2)`% less than the average KS test statistic for GEB (`r signif(ks_summary_summarise$tmb, 2)`).
+
+For both GEB and GPCA-AGHQ the KS test statistic for a parameter was correlated with low NUTS ESS (Figure \@ref(fig:ks-ess)).
+This may be due to by difficulties estimating particular parameters for all inference methods, or high KS values caused by NUTS inaccuracies.
+
+##### Maximum mean discrepancy
+
+```{r}
+mmd <- readRDS("depends/mmd.rds")
+
+first_order_pct <- signif(100 * (mmd$mmd_tmb@mmdstats[1] - mmd$mmd_aghq@mmdstats[1]) / mmd$mmd_tmb@mmdstats[1], 1)
+third_order_pct <- signif(100 * (mmd$mmd_tmb@mmdstats[2] - mmd$mmd_aghq@mmdstats[2]) / mmd$mmd_tmb@mmdstats[2], 1)
+```
+
+Let $\Phi^{1} = \{\boldsymbol{\mathbf{\phi}}^1_i\}_{i = 1}^n$ and $\Phi^2 = \{\boldsymbol{\mathbf{\phi}}^2_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
+The maximum mean discrepancy [MMD; @gretton2006kernel] is a measure of distance between joint distributions, and can be estimated empirically by samples
+\begin{equation}
+\text{MMD}(\Phi^1, \Phi^2) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\boldsymbol{\mathbf{\phi}}^1_i, \boldsymbol{\mathbf{\phi}}^1_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\boldsymbol{\mathbf{\phi}}_i^1, \boldsymbol{\mathbf{\phi}}_j^2) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\boldsymbol{\mathbf{\phi}}^2_i, \boldsymbol{\mathbf{\phi}}^2_j)}.
+\end{equation}
+The kernel was set to $k(\boldsymbol{\mathbf{\phi}}^1, \boldsymbol{\mathbf{\phi}}^2) = \exp(-\sigma \lVert \boldsymbol{\mathbf{\phi}}^1 - \boldsymbol{\mathbf{\phi}}^2 \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
+The first and third order MMD statistics for GEB were `r signif(mmd$mmd_tmb@mmdstats[1], 2)` and `r signif(mmd$mmd_tmb@mmdstats[2], 2)`.
+Those of GPCA-AGHQ (`r signif(mmd$mmd_aghq@mmdstats[1], 2)` and `r signif(mmd$mmd_aghq@mmdstats[2], 2)`) were `r first_order_pct`% and `r third_order_pct`% lower.
 

### 2024-12-29 src/docs_paper-jtb/paper.Rmd
Format Section 5.4.1

--- 2024-12-28 src/docs_paper-jtb/paper.Rmd
@@
 
-### Point estimates \label{sec:point-estimates}
+### Point estimates\label{sec:point-estimates}
+
+```{r}
+df_point <- readr::read_csv("depends/mean-sd.csv")
+
+df_point_pct <- df_point %>%
+  ungroup() %>%
+  group_by(indicator, type) %>%
+  summarise(
+    rmse_diff = 100 * diff(rmse) / max(rmse),
+    mae_diff = 100 * diff(mae) / max(mae)
+  )
+
+rmse_aghq_mean <- filter(df_point, method == "GPCA-AGHQ", indicator == "Posterior mean", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_tmb_mean <- filter(df_point, method == "GEB", indicator == "Posterior mean", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean", type == "latent") %>% pull(rmse_diff) %>% signif(2)
+
+rmse_aghq_sd <- filter(df_point, method == "GPCA-AGHQ", indicator == "Posterior SD", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_tmb_sd <- filter(df_point, method == "GEB", indicator == "Posterior SD", type == "latent") %>% pull(rmse) %>% signif(2)
+rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD", type == "latent") %>% pull(rmse_diff) %>% signif(0)
+```
+
+```{r mean-sd-alt-latent, fig.cap="The latent field posterior mean and posterior standard deviation point estimates from each inference method as compared with those from NUTS.
+The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
+For the posterior mean and posterior standard deviation, GPCA-AGHQ reduced RMSE and MAE as compared with GEB."}
+knitr::include_graphics("depends/mean-sd-alt-latent.png")
+```
+
+Latent field mean and standard deviation point estimates obtained from GPCA-AGHQ were more accurate than those from GEB (Figure \ref{fig:mean-sd-alt-latent}).
+The root mean square error (RMSE) between posterior mean estimates from GPCA-AGHQ and NUTS (`r rmse_aghq_mean`) was `r abs(rmse_diff_mean)`% lower than that between GEB and NUTS (`r rmse_tmb_mean`).
+For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_aghq_sd)` (PCA-AGHQ).
+However, improvements in latent field estimate accuracy transferred to model outputs to a more limited and mixed extent (Supplementary Figures 12 and 13).
 
@@
 
-### Exceedance probabilities\label{sec:exceedance}
+#### Kolmogorov-Smirnov
 
-#### Meeting the second 90
+```{r}
+ks_summary <- readRDS("depends/ks-summary.rds")
 
-Ambitious targets for scaling up ART treatment have been developed by UNAIDS, with the goal of ending the AIDS epidemic by 2030 [@unaids2014target].
-Meeting the 90-90-90 fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have a suppressed viral load.
-Inferences from Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met, that is $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
+ks_summary_summarise <- ks_summary %>%
+  ungroup() %>%
+  filter(type == "Latent") %>%
+  summarise(
+    tmb = sum(size * TMB) / sum(size),
+    aghq = sum(size * aghq) / sum(size)
+  )
 
-(ref:exceedance-second90) The probability each strata has met the second 90 (ART coverage above 81\%) calculated using each inference method, as compared with NUTS.
-The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
-
-```{r exceedance-second90, fig.cap="(ref:exceedance-second90)"}
-knitr::include_graphics("depends/exceedance-second90.png")
+ks_pct <- (ks_summary_summarise$tmb - ks_summary_summarise$aghq) / ks_summary_summarise$tmb
 ```
 
-Strata probabilities of having met the second 90 target were more accurately estimated by GPCA-AGHQ than GEB (Figure \@ref(fig:exceedance-second90)).
-Both GPCA-AGHQ and GEB had substantial error as compared to results from NUTS, however, particularly for girls and women.
-This discrepancy in accuracy by sex may be caused by interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.
-
-#### Finding strata with high incidence {#one-inc}
-
-Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as higher than 1% incidence per year.
-Inferences from Naomi can be used to calculate the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
-
-(ref:exceedance-1inc) The probability each strata has high HIV incidence (above 1\% per year) calculated using each inference method, as compared with NUTS.
-The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
-
-```{r exceedance-1inc, fig.cap="(ref:exceedance-1inc)"}
-knitr::include_graphics("depends/exceedance-1inc.png")
+```{r ks-summary, fig.cap="The average Kolmogorov-Smirnov (KS) test statistic for each latent field parameter of the Naomi model.
+Vectors of parameters were grouped together.
+For points above the dashed line at zero, performance of GEB was better.
+For points below the dashed line, performance of GPCA-AGHQ was better.
+Most notably, for the latent field parameters ui_lambda_x the test statistic for GEB was substantially higher than for GPCA-AGHQ.
+This parameter, of length 32, corresponds to and plays a key role in the ART attendance component of the Naomi."}
+knitr::include_graphics("depends/ks-summary.png")
 ```
-
-GPCA-AGHQ gave more accurate estimates of the probability that a strata has high HIV incidence than GEB (Figure \@ref(fig:exceedance-1inc)).
-Again, both methods had significant error.
-Unlike in Section \@ref(second90), there was little difference in error by sex.
 

### 2024-12-29 src/docs_paper-jtb/paper.Rmd
Format 5.5

@@
 The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/naomi-aghq`.
-We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
+We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] to write this paper via `rmarkdown` [@allaire2022rmarkdown].
 
@@
 For points above the dashed line at zero, performance of GEB was better.
-For points below the dashed line, performance of GPCA-AGHQ was better.
-Most notably, for the latent field parameters ui_lambda_x the test statistic for GEB was substantially higher than for GPCA-AGHQ.
-This parameter, of length 32, corresponds to and plays a key role in the ART attendance component of the Naomi."}
+For points below the dashed line, performance of GPCA-AGHQ was better."}
 knitr::include_graphics("depends/ks-summary.png")
 ```
+
+#### Maximum mean discrepancy
+
+## Exceedance probabilities\label{sec:exceedance}
+
+As realistic use cases for Naomi model outputs, we considered the following two case-studies based on exceedance probabilities.
+
+### Meeting the second 90
+
+```{r exceedance-second90, fig.cap="The probability each strata has met the second 90 (ART coverage above 81\\%) calculated using each inference method, as compared with NUTS.
+The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left."}
+knitr::include_graphics("depends/exceedance-second90.png")
+```
+
+Ambitious targets for scaling up ART treatment have been developed by UNAIDS, with the goal of ending the AIDS epidemic by 2030 [@unaids2014target].
+Meeting the 90-90-90 fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have a suppressed viral load.
+Inferences from Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met, that is $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
+Strata probabilities of having met the second 90 target were more accurately estimated by GPCA-AGHQ than GEB (Figure \ref{fig:exceedance-second90}).
+However, both GPCA-AGHQ and GEB had substantial error as compared to results from NUTS, particularly for women and girls.
+The discrepancy in accuracy by sex may be due to a more challenging posterior geometry caused by interactions between the household survey and ANC components of the model.
+
+### Finding strata with high incidence
+
+```{r exceedance-1inc, fig.cap="The probability each strata has high HIV incidence (above 1\\% per year) calculated using each inference method, as compared with NUTS.
+The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left."}
+knitr::include_graphics("depends/exceedance-1inc.png")
+```
+
+Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as higher than 1% incidence per year.
+Inferences from Naomi can be used to calculate the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
+GPCA-AGHQ gave more accurate estimates of the probability that a strata has high HIV incidence than GEB (Figure \ref{fig:exceedance-1inc}).
+Again, both methods had significant error as compared with NUTS.
 

### 2024-12-29 src/docs_paper-jtb/paper.Rmd
Use bookdown::pdf_book and complete first plausible version

@@
 output:
-  rticles::elsevier_article:
+  bookdown::pdf_book:
+    base_format: rticles::elsevier_article
+    number_sections: yes
     keep_tex: true
@@
 The remainder of this paper is organised as follows.
-In Section \ref{sec:background} we give background on latent Gaussian and extended latent Gaussian models.
-Section \ref{sec:naomi} outlines the version of the Naomi model that we consider in this paper.
-In Section \ref{sec:inferencenaomi} we review the deterministic inference method for ELGMs used by @stringer2022fast based on nested application of AGHQ and the Laplace approximation, before introducing the PCA-based modification we use to enable application to Naomi.
-In Section \ref{sec:results} we evaluate the accuracy of PCA-AGHQ for the simplified Naomi model fit to data from Malawi, as compared with TMB and gold-standard MCMC.
-Finally, we discuss our conclusions, and directions for future research in Section \ref{sec:conclusions}.
+In Section \@ref(sec:background) we give background on latent Gaussian and extended latent Gaussian models.
+Section \@ref(sec:naomi) outlines the version of the Naomi model that we consider in this paper.
+In Section \@ref(sec:inferencenaomi) we review the deterministic inference method for ELGMs used by @stringer2022fast based on nested application of AGHQ and the Laplace approximation, before introducing the PCA-based modification we use to enable application to Naomi.
+In Section \@ref(sec:results) we evaluate the accuracy of PCA-AGHQ for the simplified Naomi model fit to data from Malawi, as compared with TMB and gold-standard MCMC.
+Finally, we discuss our conclusions, and directions for future research in Section \@ref(sec:conclusions).
 
-# Background\label{sec:background}
+# Background {#sec:background}
 
@@
 
-# Simplified Naomi model\label{sec:naomi}
+# Simplified Naomi model {#sec:naomi}
 
-```{r naomi-results, fig.cap="District-level HIV prevalence (A), ART coverage (B), and new HIV cases and HIV incidence (C) for adults 15-49 in Malawi. Inference conducted using EB."}
+(ref:naomi-results) District-level HIV prevalence (A), ART coverage (B), and new HIV cases and HIV incidence (C) for adults 15-49 in Malawi. Inference conducted using EB.
+
+```{r naomi-results, fig.cap="(ref:naomi-results)"}
 knitr::include_graphics("figB.png")
@@
 
-## Household survey component \label{sec:household}
+## Household survey component {#sec:household}
 
@@
 
-## ANC testing component \label{sec:anc}
+## ANC testing component {#sec:anc}
 
@@
 
-## ART attendance component \label{sec:art}
+## ART attendance component {#sec:art}
 
@@
 
-# Inference methods for Naomi \label{sec:inferencenaomi}
+# Inference methods for Naomi {#sec:inferencenaomi}
 
-Section \ref{sec:inferenceelgm} gives the inference method for ELGMs of @stringer2022fast based on nested applications of the Laplace approximation and AGHQ.
-In Section \ref{sec:pca} we propose an extension of the method which uses PCA to facilitate inference for Naomi, which otherwise would be intractable.
+Section \@ref(sec:inferenceelgm) gives the inference method for ELGMs of @stringer2022fast based on nested applications of the Laplace approximation and AGHQ.
+In Section \@ref(sec:pca) we propose an extension of the method which uses PCA to facilitate inference for Naomi, which otherwise would be intractable.
 
-## Inference for ELGMs\label{sec:inferenceelgm}
+## Inference for ELGMs {#sec:inferenceelgm}
 
@@
 \end{equation}
-Inference proceeds by optimising Equation \ref{eq:la} using a gradient-based routine to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
-Each evaluation in the optimisation requires an inner optimisation to obtain $\hat \x(\btheta)$ via Equation \ref{eq:mode}.
+Inference proceeds by optimising Equation \@ref(eq:la) using a gradient-based routine to obtain $\hat{\btheta}_\texttt{LA} = \argmax_{\btheta} \tilde p_\texttt{LA}(\btheta, \y)$.
+Each evaluation in the optimisation requires an inner optimisation to obtain $\hat \x(\btheta)$ via Equation \@ref(eq:mode).
 Supposing the hyperparameters are to be considered fixed, as with the TMB approach used currently for Naomi, then latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
@@
 \end{equation}
-When $k = 1$ Equation \ref{eq:aghq} corresponds to a Laplace approximation.
+When $k = 1$ Equation \@ref(eq:aghq) corresponds to a Laplace approximation.
 The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
@@
 \end{equation}
-To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
+To obtain inferences for the latent field (Equation \@ref(eq:inla1)) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
 \begin{equation}
@@
 
-```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension (A).
+(ref:aghq) The Gauss-Hermite quadrature nodes $\z \in \mathcal{Q}(2, 3)$ for a two dimensional integral with three nodes per dimension (A).
 Adaption occurs based on the mode and covariance matrix of the target via the Cholesky decomposition of the inverse curvature at the mode (B).
 In PCA-AGHQ (C) only nodes along the first $s$ principal components are kept.
-Here, 95\\% of variation is explained by the first principal component.
-The integrand is $f(\\btheta) = \\text{sn}(0.5 \\theta_1, \\alpha = 2) \\cdot \\text{sn}(0.8 \\theta_1 - 0.5 \\theta_2, \\alpha = -2)$, where $\\text{sn}(\\cdot)$ is the standard skewnormal probability density function with shape parameter $\\alpha \\in \\mathbb{R}$."}
+Here, 95\% of variation is explained by the first principal component.
+The integrand is $f(\btheta) = \text{sn}(0.5 \theta_1, \alpha = 2) \cdot \text{sn}(0.8 \theta_1 - 0.5 \theta_2, \alpha = -2)$, where $\text{sn}(\cdot)$ is the standard skewnormal probability density function with shape parameter $\alpha \in \mathbb{R}$.
+
+```{r aghq, fig.cap="(ref:aghq)"}
 knitr::include_graphics("figA.png")
@@
 where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
-Panel C of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
+Panel C of Figure \@ref(fig:aghq) illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
 As AGHQ with $k = 1$ corresponds to the Laplace approximation, PCA-AGHQ can be interpreted as performing AGHQ on the first $s$ principal components of the inverse curvature, and a Laplace approximation on the remaining $m - s$ principal components.
-Inference for the latent field follows analogously to Equation \ref{eq:nest}.
+Inference for the latent field follows analogously to Equation \@ref(eq:nest).
 
-# Application to data from Malawi\label{sec:results}
+# Application to data from Malawi {#sec:results}
 
-We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi.
+We fit the simplified Naomi model (Section \@ref(sec:naomi)) to data from Malawi.
 Data from Malawi, which has $n = 30$ districts, has previously been used to demonstrate the Naomi model, including as a part of the `naomi` R package vignette.
@@
 Our goal was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
-Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
+Settings used for each inferential method are provided in Table \@ref(tab:inference-methods), and, where relevant, discussed further below.
 The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
@@
 
-```{r total-variation, fig.cap=paste0("Under PCA, the proportion of total variation explained is given by the sum of the first $s$ eigenvalues over the sum of all eigenvalues.
+(ref:total-variation) Under PCA, the proportion of total variation explained is given by the sum of the first $s$ eigenvalues over the sum of all eigenvalues.
 A typical rule-of-thumb is to include dimensions sufficient to explain 90% of total variation.
-In this case, for computational reasons, ", round(100 * tv_df$tv[8]), "% was considered sufficient.")}
+In this case, for computational reasons `r round(100 * tv_df$tv[8])`\% was considered sufficient.
+
+```{r total-variation, fig.cap="(ref:total-variation)"}
 knitr::include_graphics("depends/total-variation.png")
@@
 
-For the PCA-AGHQ quadrature grid, a Scree plot based on the spectral decomposition of $\hat {\mathbf{H}}_\texttt{LA}^{-1}$ was used to select the number of principal components to keep (Figure \ref{fig:total-variation}).
-Keeping $s = 8$ principal components was sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
-The reduced rank approximation to the inverse curvature with this choice of $s$ was visually similar to the full rank matrix (Figure \ref{fig:reduced-rank}).
+(ref:reduced-rank) The full rank original covariance matrix (Panel A) was closely reproduced by its reduced rank ($s = 8$) matrix approximation (Panel B).
 
-```{r reduced-rank, fig.cap="The full rank original covariance matrix (Panel A) was closely reproduced by its reduced rank ($s = 8$) matrix approximation (Panel B)."}
+```{r reduced-rank, fig.cap="(ref:reduced-rank)"}
 knitr::include_graphics("depends/reduced-rank.png")
@@
 
-```{r nodes-samples-comparison, fig.cap="The grey histograms show the 24 hyperparameter marginal distributions obtained with NUTS.
+(ref:nodes-samples-comparison) The grey histograms show the 24 hyperparameter marginal distributions obtained with NUTS.
 The green lines indicate the position of the 6561 PCA-AGHQ nodes projected onto each hyperparameter marginal.
-For some hyperparameters, the PCA-AGHQ nodes vary over the domain of the posterior marginal distribution, while for others they concentrate at the mode."}
+For some hyperparameters, the PCA-AGHQ nodes vary over the domain of the posterior marginal distribution, while for others they concentrate at the mode.
+
+```{r nodes-samples-comparison, fig.cap="(ref:nodes-samples-comparison)"}
 knitr::include_graphics("depends/nodes-samples-comparison.png")
 ```
+
+For the PCA-AGHQ quadrature grid, a Scree plot based on the spectral decomposition of $\hat {\mathbf{H}}_\texttt{LA}^{-1}$ was used to select the number of principal components to keep (Figure \@ref(fig:total-variation)).
+Keeping $s = 8$ principal components was sufficient to explain `r round(100 * tv_df$tv[8])`% of total variation.
+The reduced rank approximation to the inverse curvature with this choice of $s$ was visually similar to the full rank matrix (Figure \@ref(fig:reduced-rank)).
 
@@
 
-Projecting the $3^8 = 6561$ PCA-AGHQ quadrature nodes onto each hyperparameter dimension, there was substantial variation in coverage by hyperparameter (Figure \ref{fig:nodes-samples-comparison}).
+Projecting the $3^8 = 6561$ PCA-AGHQ quadrature nodes onto each hyperparameter dimension, there was substantial variation in coverage by hyperparameter (Figure \@ref(fig:nodes-samples-comparison)).
 Approximately 12 hyperparameters had well covered marginals: greater than the 8 naively obtained with a dense grid, but nonetheless far fewer than the full 24.
@@
 
-Posterior inferences from GEB, GPCA-AGHQ and NUTS were compared using point estimates (Section \ref{sec:point-estimates}), distributional quantities (Section \ref{sec:distributional-quantities}), and exceedance probabilities (Section \ref{sec:exceedance}).
+Posterior inferences from GEB, GPCA-AGHQ and NUTS were compared using point estimates (Section \@ref(sec:point-estimates)), distributional quantities (Section \@ref(sec:distributional-quantities)), and exceedance probabilities (Section \@ref(sec:exceedance)).
 
-### Point estimates\label{sec:point-estimates}
+### Point estimates {#sec:point-estimates}
 
@@
 
-```{r mean-sd-alt-latent, fig.cap="The latent field posterior mean and posterior standard deviation point estimates from each inference method as compared with those from NUTS.
+(ref:mean-sd-alt-latent) The latent field posterior mean and posterior standard deviation point estimates from each inference method as compared with those from NUTS.
 The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
-For the posterior mean and posterior standard deviation, GPCA-AGHQ reduced RMSE and MAE as compared with GEB."}
+For the posterior mean and posterior standard deviation, GPCA-AGHQ reduced RMSE and MAE as compared with GEB.
+
+```{r mean-sd-alt-latent, fig.cap="(ref:mean-sd-alt-latent)"}
 knitr::include_graphics("depends/mean-sd-alt-latent.png")
@@
 
-Latent field mean and standard deviation point estimates obtained from GPCA-AGHQ were more accurate than those from GEB (Figure \ref{fig:mean-sd-alt-latent}).
+Latent field mean and standard deviation point estimates obtained from GPCA-AGHQ were more accurate than those from GEB (Figure \@ref(fig:mean-sd-alt-latent)).
 The root mean square error (RMSE) between posterior mean estimates from GPCA-AGHQ and NUTS (`r rmse_aghq_mean`) was `r abs(rmse_diff_mean)`% lower than that between GEB and NUTS (`r rmse_tmb_mean`).
@@
 
-### Distributional quantities\label{sec:distributional-quantities}
+### Distributional quantities {#sec:distributional-quantities}
 
@@
 
-```{r ks-summary, fig.cap="The average Kolmogorov-Smirnov (KS) test statistic for each latent field parameter of the Naomi model.
+(ref:ks-summary) The average Kolmogorov-Smirnov (KS) test statistic for each latent field parameter of the Naomi model.
 Vectors of parameters were grouped together.
 For points above the dashed line at zero, performance of GEB was better.
-For points below the dashed line, performance of GPCA-AGHQ was better."}
+For points below the dashed line, performance of GPCA-AGHQ was better.
+Most notably, for the latent field parameters $\texttt{ui\_lambda\_x}$ the average test statistic for GEB was substantially higher than for GPCA-AGHQ.
+This length 32 parameter $\mathbf{u}_x^{\lambda}$ plays a key role in the ART attendance component of the Naomi.
+
+```{r ks-summary, fig.cap="(ref:ks-summary)"}
 knitr::include_graphics("depends/ks-summary.png")
@@
 
+```{r}
+ui_lambda_x <- readRDS("depends/ui-lambda-x.rds")
+```
+
+(ref:ui-lambda-x) The parameter ``r paste0(ui_lambda_x$par)`` had the greatest difference in KS test statistics between GEB and GPCA-AGHQ to NUTS.
+For this parameter, the potential scale reduction factor was `r ui_lambda_x$rhat` and effective sample size was `r ui_lambda_x$ess`.
+
+```{r ui-lambda-x, fig.cap="(ref:ui-lambda-x)"}
+knitr::include_graphics("depends/ui-lambda-x.png")
+```
+
+The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is the maximum absolute difference between two ECDFs $F(\omega) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \omega}$.
+It is a relatively stringent worst-case measure of distance between empirical distributions.
+The average KS test statistic for GPCA-AGHQ (`r signif(ks_summary_summarise$aghq, 2)`) was `r signif(100 * ks_pct, 2)`% less than the average KS test statistic for GEB (`r signif(ks_summary_summarise$tmb, 2)`).
+For both GEB and GPCA-AGHQ the KS test statistic for a parameter was correlated with low NUTS ESS (Supplementary Figure 14).
+This correlation may be due to by difficulties estimating particular parameters for all inference methods, or high KS values caused by NUTS inaccuracies.
+
 #### Maximum mean discrepancy
 
-## Exceedance probabilities\label{sec:exceedance}
+```{r}
+mmd <- readRDS("depends/mmd.rds")
+
+first_order_pct <- signif(100 * (mmd$mmd_tmb@mmdstats[1] - mmd$mmd_aghq@mmdstats[1]) / mmd$mmd_tmb@mmdstats[1], 1)
+third_order_pct <- signif(100 * (mmd$mmd_tmb@mmdstats[2] - mmd$mmd_aghq@mmdstats[2]) / mmd$mmd_tmb@mmdstats[2], 1)
+```
+
+The maximum mean discrepancy [MMD; @gretton2006kernel] is a measure of distance between joint distributions and can be estimated empirically using samples.
+Let $\Phi^{1} = \{\boldsymbol{\mathbf{\phi}}^1_s\}_{s = 1}^S$ and $\Phi^2 = \{\boldsymbol{\mathbf{\phi}}^2_s\}_{s = 1}^S$ be two sets of joint posterior samples and $k$ be a kernel, then
+\begin{equation}
+\text{MMD}(\Phi^1, \Phi^2) = \sqrt{\frac{1}{S^2} \sum_{s, l = 1}^S k(\boldsymbol{\mathbf{\phi}}^1_s, \boldsymbol{\mathbf{\phi}}^1_l) - \frac{2}{S^2} \sum_{s, l = 1}^S k(\boldsymbol{\mathbf{\phi}}_s^1, \boldsymbol{\mathbf{\phi}}_l^2) + \frac{1}{S^2} \sum_{s, l = 1}^S k(\boldsymbol{\mathbf{\phi}}^2_i, \boldsymbol{\mathbf{\phi}}^2_l)}.
+\end{equation}
+The kernel was set to $k(\boldsymbol{\mathbf{\phi}}^1, \boldsymbol{\mathbf{\phi}}^2) = \exp(-\sigma \lVert \boldsymbol{\mathbf{\phi}}^1 - \boldsymbol{\mathbf{\phi}}^2 \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
+Then, the first and third order MMD statistics for GEB were `r signif(mmd$mmd_tmb@mmdstats[1], 2)` and `r signif(mmd$mmd_tmb@mmdstats[2], 2)` while those of GPCA-AGHQ (`r signif(mmd$mmd_aghq@mmdstats[1], 2)` and `r signif(mmd$mmd_aghq@mmdstats[2], 2)`) were `r first_order_pct`% and `r third_order_pct`% lower.
+
+## Exceedance probabilities {#sec:exceedance}
 
@@
 
-```{r exceedance-second90, fig.cap="The probability each strata has met the second 90 (ART coverage above 81\\%) calculated using each inference method, as compared with NUTS.
-The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left."}
+(ref:exceedance-second90) The probability each strata has met the second 90 (ART coverage above 81\\%) calculated using each inference method, as compared with NUTS.
+The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
+
+```{r exceedance-second90, fig.cap="(ref:exceedance-second90)"}
 knitr::include_graphics("depends/exceedance-second90.png")
@@
 Inferences from Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met, that is $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.
-Strata probabilities of having met the second 90 target were more accurately estimated by GPCA-AGHQ than GEB (Figure \ref{fig:exceedance-second90}).
+Strata probabilities of having met the second 90 target were more accurately estimated by GPCA-AGHQ than GEB (Figure \@ref(fig:exceedance-second90)).
 However, both GPCA-AGHQ and GEB had substantial error as compared to results from NUTS, particularly for women and girls.
@@
 
-```{r exceedance-1inc, fig.cap="The probability each strata has high HIV incidence (above 1\\% per year) calculated using each inference method, as compared with NUTS.
-The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left."}
+(ref:exceedance-1inc) The probability each strata has high HIV incidence (above 1\\% per year) calculated using each inference method, as compared with NUTS.
+The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
+
+```{r exceedance-1inc, fig.cap="(ref:exceedance-1inc)"}
 knitr::include_graphics("depends/exceedance-1inc.png")
@@
 Inferences from Naomi can be used to calculate the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
-GPCA-AGHQ gave more accurate estimates of the probability that a strata has high HIV incidence than GEB (Figure \ref{fig:exceedance-1inc}).
+GPCA-AGHQ gave more accurate estimates of the probability that a strata has high HIV incidence than GEB (Figure \@ref(fig:exceedance-1inc)).
 Again, both methods had significant error as compared with NUTS.
 
-# Conclusion\label{sec:conclusions}
+# Conclusion {#sec:conclusions}
 

### 2024-12-29 src/docs_paper-jtb/preamble.tex
Add JTB template
Use bookdown::pdf_book and complete first plausible version

--- 2022-07-31 src/docs_paper/preamble.tex
@@
 \RequirePackage{bm}
-\RequirePackage{enumerate}
-\RequirePackage{enumitem}
+% \RequirePackage{enumerate}
+% \RequirePackage{enumitem}
 \RequirePackage{tabularx}
-\RequirePackage{adjustbox}
-\RequirePackage{tikz}
-\RequirePackage{bbm}
-\RequirePackage{lscape}
-\RequirePackage{booktabs}
-\RequirePackage{longtable}
-\RequirePackage{lscape}
-\RequirePackage{pifont}
-\RequirePackage[margin=1in]{geometry}
-\usepackage{footnote}
+% \RequirePackage{adjustbox}
+% \RequirePackage{tikz}
+% \RequirePackage{bbm}
+% \RequirePackage{lscape}
+% \RequirePackage{booktabs}
+% \RequirePackage{longtable}
+% \RequirePackage{lscape}
+% \RequirePackage{pifont}
+% \usepackage{footnote}
+% \usepackage{algorithm}
+% \usepackage{algpseudocode}
 
@@
 \newcommand{\m}{\mathbf{m}}
+\newcommand{\bk}{\mathbf{k}}
 \newcommand{\bu}{\mathbf{u}}
@@
 \newcommand{\bb}{\mathbf{b}}
+\newcommand{\Hb}{\mathbf{H}}
 \newcommand{\bphi}{\bm{\phi}}
+\newcommand{\bpsi}{\bm{\psi}}
 \newcommand{\brho}{\bm{\rho}}
 \newcommand{\btheta}{\bm{\theta}}
+\newcommand{\bmeta}{\bm{\eta}}
+\newcommand{\bmu}{\bm{\mu}}
 
@@
 
-\definecolor{lightblue}{HTML}{56B4E9}
-\definecolor{lightgreen}{HTML}{009E73}
-\definecolor{lightorange}{HTML}{E69F00}
-\definecolor{lightgrey}{HTML}{E6E6E6}
-
-\newcommand{\lightblueline}{\raisebox{2pt}{\tikz{\draw[-,lightblue,solid,line width = 3pt](0,0) -- (2mm,0);}}}
-\newcommand{\lightgreenline}{\raisebox{2pt}{\tikz{\draw[-,lightgreen,solid,line width = 3pt](0,0) -- (2mm,0);}}}
-\newcommand{\lightorangeline}{\raisebox{2pt}{\tikz{\draw[-,lightorange,solid,line width = 3pt](0,0) -- (2mm,0);}}}
-\newcommand{\lightgreyline}{\raisebox{2pt}{\tikz{\draw[-,lightgrey,solid,line width = 3pt](0,0) -- (2mm,0);}}}
-
 \usepackage{amsmath}
 \DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator*{\argmax}{arg\,max}

### 2024-12-29 src/docs_paper-jtb/appendix.Rmd
Format Section 5.4.1
Label equations by default and some edits

@@
 
-The proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ was modelled as
+The proportion who were recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ was modelled as
 \begin{equation}
@@
 (ref:ess) The efficiency of the NUTS, as measured by the ratio of effective sample size to total number of iterations run, was low for most parameters (Panel A).
-As a result, the number of iterations required for the the effective number of samples (mean `r round(naomi_nuts$ess_mean)`) to be satisfactory was high (Panel B).
+As a result, the number of iterations required for the effective number of samples (mean `r round(naomi_nuts$ess_mean)`) to be satisfactory was high (Panel B).
 
@@
 Each point is an estimate of the indicator for a particular strata.
-Error is increased by GPCA-AGHQ for HIV prevalence and HIV incidence, and reduced for ART coverage.
+Error is increased by GPCA-AGHQ for ART coverage and HIV incidence, and reduced for HIV prevalence.
 
@@
 ```
+
+\clearpage
 

### 2024-12-29 src/docs_paper-jtb/paper.Rmd
Label equations by default and some edits

@@
 Using data from Malawi, our method improves the accuracy of inferences for model parameters, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler.
-Our implementation is based on the existing \texttt{TMB} \textsf{C++} template for the model's log-posterior, and is compatible with any model with such a template.
+Our implementation leverages the existing \texttt{TMB} \textsf{C++} template for the model's log-posterior, and is compatible with any model with such a template.
 keywords: 
@@
 Generation of estimates by country teams is an important and distinctive feature of the HIV response.
-Drawing on expert knowledge of the data improves the accuracy of the process and strengthens trust in the resulting estimates, creating a virtuous cycle of data quality, use, and ownership [@noor2022country].
+Drawing on expert knowledge about the data improves the accuracy of the process and strengthens trust in the resulting estimates, creating a virtuous cycle of data quality, use, and ownership [@noor2022country].
 
@@
 The scale of the model and features of its posterior geometry [@neal2003slice] make Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
-Inference must be reliable and automatic across a range of country settings, and should not require close monitoring and diagnostics (e.g. for MCMC convergence).
+Inference should also be reliable, automatic across various country settings, and require minimal manual monitoring (e.g. of MCMC convergence).
 
@@
 Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models with
-\begin{align*}
+\begin{align}
 y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
@@
 \eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
-\end{align*}
+\end{align}
 where $[n] = \{1, \ldots, n\}$.
@@
 Put together, ELGMs are then of the form
-\begin{align*}
+\begin{align}
 y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
@@
 \eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n],
-\end{align*}
+\end{align}
 with latent field and hyperparameter priors as in the LGM case.
@@
 Weighted aggregate survey observations are calculated based on individual responses $\theta_j \in \{0, 1\}$ as
-\begin{equation*}
+\begin{equation}
 \hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}.
-\end{equation*}
+\end{equation}
 Design weights $w_j$ for each of $\theta \in \{\rho, \alpha, \kappa\}$ are supplied by the survey provider and aim to reduce bias by decreasing possible correlation between response and recording mechanism [@meng2018statistical].
@@
 The weighted observed number of outcomes is taken to be $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
-\begin{equation*}
+\begin{equation}
 m^{\theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
-\end{equation*}
+\end{equation}
 is the Kish effective sample size [ESS; @kish1965survey].
 The weighted aggregated number of outcomes are modelled using a binomial working likelihood [@chen2014use] defined to operate on the reals
-\begin{equation*}
+\begin{equation}
 y^{\theta}_{I} \sim \text{xBin}(m^{\theta}_{I}, \theta_{I})
-\end{equation*}
+\end{equation}
 The terms $\theta_{I}$ are the following weighted aggregates
-\begin{equation*}
+\begin{equation}
 \rho_{I} = \frac{\sum_{i \in I} N_i \rho_i}{\sum_{i \in I} N_i}, \quad
@@
 \kappa_{I} = \frac{\sum_{i \in I} N_i \rho_i \kappa_i}{\sum_{i \in I} N_i \rho_i}.
-\end{equation*}
+\end{equation}
 
@@
 HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset from the general population indicators as follows
-\begin{align*}
+\begin{align}
 \text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
 \text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\alpha^\text{ANC}}_i.
-\end{align*}
+\end{align}
 These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
@@
 The number of those with positive status $y^{\rho^\text{ANC}}_I$, and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ are modelled using nested binomial likelihoods
-\begin{align*}
+\begin{align}
 y^{\rho^\text{ANC}}_I &\sim \text{Bin}(m^{\rho^\text{ANC}}_I, \rho^\text{ANC}_{I}), \\
 y^{\alpha^\text{ANC}}_I &\sim \text{Bin}(y^{\rho^\text{ANC}}_I, \alpha^\text{ANC}_{I}).
-\end{align*}
+\end{align}
 As in the household survey component, weighted aggregates
-\begin{equation*}
+\begin{equation}
 \rho^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in I} \Psi_i}, \quad
 \alpha^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}},
-\end{equation*}
+\end{equation}
 are used, with $\Psi_i$ the number of pregnant women assumed here to be fixed.
@@
 The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ and hyperparameter prior distribution $p(\btheta)$ in an ELGM is given by
-\begin{equation*}
+\begin{equation}
   p(\x, \btheta \, | \, \y)
   \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
-\end{equation*}
+\end{equation}
 We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
@@
 
-AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1), and conducted part of this research while an International Visiting Graduate Student at the University of Waterloo.
+AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1).
+AH conducted part of this research while an International Visiting Graduate Student at the University of Waterloo.
 AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).
-AS was supported by NSERC Discovery Grant RGPIN-2023-03331.
+AS was supported by NSERC Discovery Grant (RGPIN-2023-03331).
 SRF was supported by the EPSRC (EP/V002910/2).

### 2024-12-29 src/docs_paper-jtb/paper.Rmd
Shorten sentence

@@
 GPCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$, 1000 samples \\
-NUTS & \texttt{tmbstan} & 4 chains of 100000 iterations, with the first 50000 iterations of each chain discarded as warmup, thinned by a factor of 20, to give a total of 10000 samples kept.
+NUTS & \texttt{tmbstan} & 4 chains of 100000 iterations, with the first half of each chain discarded as warmup, thinned by a factor of 20, to give a total of 10000 samples.
 Default NUTS tuning parameters \citep{hoffman2014no}.

### 2024-12-31 src/docs_paper-jtb/paper.Rmd
Improvements to conclusion section

@@
 
-(ref:naomi-results) District-level HIV prevalence (A), ART coverage (B), and new HIV cases and HIV incidence (C) for adults 15-49 in Malawi. Inference conducted using EB.
+(ref:naomi-results) District-level HIV prevalence (A), ART coverage (B), and new HIV cases and HIV incidence (C) for adults 15-49 in Malawi.
+Inference here was conducted using and Gaussian approximation and EB via `TMB`.
 
@@
 
-For the simplified Naomi model applied to data from Malawi, GPCA-AGHQ more accurately inferred latent field posterior marginal distributions than GEB.
-However, model output posterior marginals did not see the same improvements.
-Approximate posterior exceedance probabilities from both GEB and GPCA-AGHQ had systematic inaccuracies as compared with NUTS.
-GEB and GPCA-AGHQ were substantially faster than NUTS, which took over two days to reach convergence.
+In the Malawi case-study, GPCA-AGHQ more accurately inferred latent field posterior marginal distributions than GEB.
+Benefits were particularly large for latent field posterior marginal standard deviations (Figure \@ref(fig:mean-sd-alt-latent)).
+However, the benefit of using GPCA-AGHQ for model outputs, obtained by a link function applied to a linear transformation of latent field parameters, was more limited.
+Although posterior exceedance probabilities based on model outputs from GPCA-AGHQ were more accurate (Figures \@ref(fig:exceedance-second90) and \@ref(fig:exceedance-1inc)), both GEB and GPCA-AGHQ showed systematic inaccuracies as compared with NUTS.
 
 Inaccuracies in model outputs from GEB and GPCA-AGHQ do have potential to meaningfully mislead policy.
-As such, where possible, gold-standard NUTS results should be computed.
-Though NUTS is to slow too run during a workshop, it could be run afterwards.
-As the UNAIDS HIV estimates process occurs annually, requiring days to compute more accurate estimates is viable.
+As such, where it is possible to do so, gold-standard NUTS results should be computed.
+Although NUTS is too slow to run during a workshop, as the UNAIDS HIV estimates process occurs annually, it would be viable to compute more accurate estimates afterwards.
 That said, Malawi is one of the countries with the fewest number of districts.
-As NUTS took days to run in Malawi, for larger countries, with hundreds of districts, it may be impossible to run NUTS to convergence, and approximate methods may be required.
+As NUTS took over two days to reach convergence in Malawi, for larger countries, with hundreds of districts, it may be impossible to run NUTS to convergence, and faster approximate methods such as GEB and GPCA-AGHQ may be necessary.
 
@@
 Analysts would be able to quickly iterate over model options using GEB, before switching to a more accurate approach once they are happy with the results.
-
-PCA-AGHQ can be adjusted to suit the computational budget available by choice of the number of dimensions kept in the PCA $s$ and the number of points per dimension $k$.
-The scree plot is a well established heuristic for choosing $s$.
-Heuristics for choosing $k$ are less well established.
+One benefit of PCA-AGHQ is that it can be adjusted to suit the computational budget available by choice of the number of dimensions kept in the PCA $s$ and the number of points per dimension $k$.
+The scree plot is a well established heuristic for choosing $s$, while heuristics for choosing $k$ are less well established.
 Whether it is preferable for a given computational budget to increase $s$ or increase $k$ is an open question.
@@
 
-### Further comparisons {#further-comparisons}
+### Further comparisons {#sec:further-comparisons}
 
-Comparison to further Bayesian inference methods could be included.
-Four possibilities stand out as being particularly valuable:
-
-1. There exist other quadrature rules for moderate dimension, such as the CCD.
+Comparison to further Bayesian inference methods could prove beneficial.
+Four possibilities stand out as being particularly valuable.
+First, there exist other quadrature rules for moderate dimension, such as the Box-Wilson central composite design [@box1992experimental].
 It would be of interest to compare INLA with a PCA-AGHQ rule to INLA with other such quadrature rules.
-2. Rather than use quadrature to integrate the marginal Laplace approximation, an alternative approach is to run HMC [@monnahan2018no; @margossian2020hamiltonian].
+Second, rather than use quadrature to integrate the marginal Laplace approximation, an alternative approach is to run HMC [@monnahan2018no; @margossian2020hamiltonian].
+This method can be run by using `tmbstan` and setting `laplace = TRUE`.
 When run to convergence, inferential error of this method would solely be due to the Laplace approximation, helping to clarify the extent to which the inferential error of INLA is attributable to the quadrature grid.
-Preliminary testing of this approach, using `tmbstan` and setting `laplace = TRUE`, did not show immediate success but likely could be worked on.
-3. NUTS is not especially well suited to sampling from Gaussian latent field models like Naomi.
+Third, NUTS is not especially well suited to sampling from Gaussian latent field models like Naomi.
 Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], could be considered.
 It may be difficult to implement such algorithms using `TMB`.
-Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language [@de2017programming].
-Requiring rewriting the Naomi model log-posterior outside of `TMB` would be a substantial downside.
-4. Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
+Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language [@de2017programming], however requiring rewriting the Naomi model log-posterior outside of `TMB` would be a downside.
+Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
 However, as `inlabru`, like `R-INLA`, is based on a formula interface, it may not be possible to do so directly.
@@
 
-PCA-AGHQ is a sensible approach to allocating more computational to dimensions which contribute more to the integral in question.
-However, its application to Naomi surfaced instances where it overlooked potential benefits, or otherwise did not behave as one might wish:
-
-1. The amount of variation explained in the Hessian matrix may not be of direct interest.
+PCA-AGHQ is a sensible approach to allocating more computational effort to dimensions which contribute more to the integral in question.
+Its application to Naomi surfaced several ideas for improvement and challenges to be overcome.
+First, the amount of variation explained in the Hessian matrix may not be of direct interest.
 For the Naomi model, interest is in the effect of including each dimension on the relevant model outputs.
 As such, using alternative measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] or Sobol indices, could be preferable.
-2. Use of PCA is challenging when the dimensions have different scales.
+Second, use of PCA is challenging when the dimensions have different scales.
 For the Naomi model, logit-scale hyperparameters were systematically favoured over those on the log-scale.
-3. When the quadrature rule is used within an INLA algorithm, it is more important to allocate quadrature nodes to those hyperparameter marginals which are non-Gaussian.
+Third, when the quadrature rule is used within an INLA algorithm, it is more important to allocate quadrature nodes to those hyperparameter marginals which are non-Gaussian.
 This is because the Laplace approximation is exact when the integrand is Gaussian, so a single quadrature node is sufficient.
@@
 The measures of fit would have to be for marginals, ruling out approaches like PSIS [@yao2018yes] which operate on joint distributions.
-4. Finally, it may be possible to achieve better performance by pruning and prerotation, as discussed by @jackel2005note.
+Finally, it may be possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.
 
-#### Statistical theory
+### Statistical theory
 
@@
 Similar theory could be established for PCA-AGHQ, or more generally AGHQ with varying levels.
-The challenge of connecting this theory to nested use of any quadrature rule, like that in the INLA algorithm, remains an important open question.
+The challenge of connecting this theory to nested use of any quadrature rule, like GPCA-AGHQ, remains an important open question.
 
-#### Testing quadrature assumptions
+### Testing quadrature assumptions
 
-It may be possible to test the assumptions made by use of AGHQ grids, allowing their suitability for a particular integral to be assessed.
-Specifically, AGHQ assumes that the integrand is closely approximated by a polynomial multiplied by a Gaussian density.
-Given NUTS hyperparameter samples (or better yet, hyperparameter samples from the Laplace NUTS hybrid discussed in Section \@ref(further-comparisons)) this assumption could be tested by fitting a model using a polynomial times Gaussian kernel.
+It may be possible to test the assumptions underlying use of AGHQ grids, enabling assessment of their suitability for specific integrals.
+In particular, AGHQ assumes that the integrand is well approximated by a polynomial multiplied by a Gaussian density.
+This assumption could be tested by fitting a model using a polynomial times Gaussian kernel, given NUTS hyperparameter samples (or even better, hyperparameter samples obtained from the Laplace NUTS approach mentioned in Section \@ref(sec:further-comparisons)).
 This approach could be generalised to also test the suitability of PCA-AGHQ grids.

### 2024-12-31 src/docs_paper-jtb/paper.Rmd
Final fixes

@@
 A Gaussian approximation to the latent field is used via the Template Model Builder (`TMB`) [@kristensen2016tmb] \textsf{R} package.
-`TMB` is gaining popularity due to its speed and flexibility, especially in spatial statistics [@osgood2022statistical] and via the user-friendly `glmmTMB` \textsc{R} package [@brooks2017glmmTMB].
+`TMB` is gaining popularity due to its speed and flexibility, especially in spatial statistics [@osgood2023statistical.] and via the user-friendly `glmmTMB` \textsc{R} package [@brooks2017glmmTMB].
 Inference in `TMB` occurs via optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the parameters.
@@
 In Section \@ref(sec:inferencenaomi) we review the deterministic inference method for ELGMs used by @stringer2022fast based on nested application of AGHQ and the Laplace approximation, before introducing the PCA-based modification we use to enable application to Naomi.
-In Section \@ref(sec:results) we evaluate the accuracy of PCA-AGHQ for the simplified Naomi model fit to data from Malawi, as compared with TMB and gold-standard MCMC.
+In Section \@ref(sec:results) we evaluate the accuracy of a PCA-AGHQ grid for the simplified Naomi model fit to data from Malawi, as compared with EB and gold-standard MCMC.
 Finally, we discuss our conclusions, and directions for future research in Section \@ref(sec:conclusions).
@@
 It is also possible to extract the `TMB` objective function used by `glmmTMB`, which may then be passed into `aghq` or `tmbstan`.
+This discussion highlights the utility of `TMB` templates, emphasising that our approach, built on `TMB`, is both general and broadly applicable.
 
@@
 While this version omits nowcasting and temporal projection, as these time points involve limited inference we expect conclusions for the simplified model to be applicable to the complete model.
-An overview of the simplified model is given below, and a full specification is provided in the Appendix.
+An overview of the simplified model is given below, and a full specification is provided in Appendix 1.
 
-## Notation
+## Notation and overview
 
@@
 
-Let $N_i \in \mathbb{N}$ be the known, fixed population size, $\rho_i \in [0, 1]$ be the HIV prevalence, $\alpha_i \in [0, 1]$ be the antiretroviral therapy (ART) coverage, $\kappa_i \in [0, 1]$ be the proportion recently infected among HIV positive persons, and $\lambda_i > 0$ be the annual HIV incidence rate.
+Let $N_i \in \mathbb{N}$ be the known, fixed population size, $\rho_i \in [0, 1]$ be the HIV prevalence, $\alpha_i \in [0, 1]$ be the antiretroviral therapy (ART) coverage, $\kappa_i \in [0, 1]$ be the proportion who are recently infected among HIV positive persons, and $\lambda_i > 0$ be the annual HIV incidence rate.
 
@@
 
-Naomi is a joint model on the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
+Naomi is a joint model on the observations $\mathbf{y} = (y^\theta_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
+A superscript is used here to refer to components of the model (and does not refer to a power, unless made clear).
 The structured additive predictors contain intercept effects, age random effects, and spatial random effects which we collectively describe as the latent field $\x$.
 The latent field is controlled by hyperparamters $\btheta$ which include standard deviations, first-order autoregressive model correlation parameters, and reparameterised Besag-York-Mollie model [BYM2; @simpson2017penalising] proportion parameters.
+The number of hyperparameters is a key feature making inference challenging.
 
@@
 HIV incidence rate is modelled by $\log(\lambda_i) = \eta^\lambda_i$ and depends on adult HIV prevalence and adult ART coverage.
-The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in the Appendix.
+The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix 1.
 The proportion recently infected $\kappa_i$ is linked to HIV incidence via
@@
   p(\x, \btheta \, | \, \y)
-  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
+  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{N/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
 \end{equation}
@@
   p(x_i \, | \, \y) &\approx \tilde p(x_i \, | \, \y) = \int \tilde p(x_i \, | \, \btheta, \y) \tilde p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
-  p(\theta_j \, | \, \y) &\approx \tilde p(\theta_j \, | \, \y) = \int \tilde  p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
+  p(\theta_j \, | \, \y) &\approx \tilde p(\theta_j \, | \, \y) = \int \tilde  p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m], \label{eq:inla2}
 \end{align}
+where the approximations $\tilde p(x_i \, | \, \btheta, \y)$ and $\tilde  p(\btheta \, | \, \y)$ remain to be defined.
 
@@
 
-Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\Hb}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
+Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\Hb}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ defined by mode and precision matrix
 \begin{align}
@@
 Each evaluation in the optimisation requires an inner optimisation to obtain $\hat \x(\btheta)$ via Equation \@ref(eq:mode).
-Supposing the hyperparameters are to be considered fixed, as with the TMB approach used currently for Naomi, then latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
+Supposing the hyperparameters are to be considered fixed, as with the `TMB` approach used currently for Naomi, then latent field joint and marginal inferences then follow directly from the Gaussian approximation $\tilde p_\texttt{G}(\x \, | \, \hat{\btheta}_\texttt{LA}, \y)$.
 
@@
 We fit the simplified Naomi model (Section \@ref(sec:naomi)) to data from Malawi.
-Data from Malawi, which has $n = 30$ districts, has previously been used to demonstrate the Naomi model, including as a part of the `naomi` R package vignette.
+Malawi, which has $n = 30$ districts, has previously been used to demonstrate the Naomi model, including as a part of the `naomi` R package vignette.
 Three Bayesian inference approaches were considered:
@@
 
-Our goal was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
+Our goal was to determine the accuracy of the approximate methods (GEB and GPCA-AGHQ) as compared with the gold-standard (NUTS).
 Settings used for each inferential method are provided in Table \@ref(tab:inference-methods), and, where relevant, discussed further below.
@@
 We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] to write this paper via `rmarkdown` [@allaire2022rmarkdown].
-
-## Time taken
-
-```{r naomi-time}
-time_taken <- readr::read_csv("depends/time-taken.csv")
-
-time_taken$aghq <- time_taken$aghq * 60 * 60 
-time_taken$tmbstan <- time_taken$tmbstan * 60 * 60 * 24
-
-time_taken_df <- as.data.frame(t(time_taken)) %>%
-  tibble::rownames_to_column("method") %>%
-  mutate(
-    mins = V1 / 60,
-    hours = mins / 60,
-    days = hours / 24,
-    method = case_when(
-      method == "TMB" ~ "GEB",
-      method == "aghq" ~ "GPCA-AGHQ",
-      method == "tmbstan" ~ "NUTS"
-    ),
-    software = case_when(
-      method == "NUTS" ~ "tmbstan",
-      TRUE ~ "TMB"
-    )
-  )
-
-naomi_time_tmbstan_hours <- signif(filter(time_taken_df, method == "NUTS")$hours, 2)
-naomi_time_aghq_hours <- signif(filter(time_taken_df, method == "GPCA-AGHQ")$hours, 2)
-naomi_time_tmb_minutes <- signif(filter(time_taken_df, method == "GEB")$mins, 1)
-```
-
-Inference with NUTS took `r naomi_time_tmbstan_hours` hours, while inference with GPCA-AGHQ took `r naomi_time_aghq_hours` hours and GEB just `r naomi_time_tmb_minutes` minutes.
-Both the NUTS and GPCA-AGHQ algorithms can be run under a range of settings, trading off accuracy and runtime.
 
@@
 Several of the first $s = 8$ PC loadings are sums of two hyperparameters, indicating some redundancy in the hyperparameter parameterisation.
-This finding is supported analysis of correlation structure in the NUTS hyperparameter posterior (Appendix 2.1).
-It is exactly this correlation structure that PCA, and PCA-AGHQ, makes use of.
+This finding is supported by analysis of correlation structure in the NUTS hyperparameter posterior (Appendix 2.1).
 
-Projecting the $3^8 = 6561$ PCA-AGHQ quadrature nodes onto each hyperparameter dimension, there was substantial variation in coverage by hyperparameter (Figure \@ref(fig:nodes-samples-comparison)).
+Projecting the $3^8 = 6561$ PCA-AGHQ quadrature nodes onto each univariate hyperparameter dimension, there was substantial variation in coverage by hyperparameter (Figure \@ref(fig:nodes-samples-comparison)).
 Approximately 12 hyperparameters had well covered marginals: greater than the 8 naively obtained with a dense grid, but nonetheless far fewer than the full 24.
-Coverage was higher among hyperparameters on the logistic scale, and lower among hyperparameters on the logarithmic scale.
-This discrepancy occurred due to logistic hyperparameters naturally having higher posterior marginal standard deviation than logarithmic hyperparameters (Supplementary Figures 9 and 10).
+Coverage was higher among hyperparameters on the logistic scale, and lower among hyperparameters on the logarithmic scale (Supplementary Figure 9).
+This discrepancy occurred due to logistic hyperparameters naturally having higher posterior marginal standard deviation than logarithmic hyperparameters (Supplementary Figure 10).
+
+## Time taken
+
+```{r naomi-time}
+time_taken <- readr::read_csv("depends/time-taken.csv")
+
+time_taken$aghq <- time_taken$aghq * 60 * 60 
+time_taken$tmbstan <- time_taken$tmbstan * 60 * 60 * 24
+
+time_taken_df <- as.data.frame(t(time_taken)) %>%
+  tibble::rownames_to_column("method") %>%
+  mutate(
+    mins = V1 / 60,
+    hours = mins / 60,
+    days = hours / 24,
+    method = case_when(
+      method == "TMB" ~ "GEB",
+      method == "aghq" ~ "GPCA-AGHQ",
+      method == "tmbstan" ~ "NUTS"
+    ),
+    software = case_when(
+      method == "NUTS" ~ "tmbstan",
+      TRUE ~ "TMB"
+    )
+  )
+
+naomi_time_tmbstan_hours <- signif(filter(time_taken_df, method == "NUTS")$hours, 2)
+naomi_time_aghq_hours <- signif(filter(time_taken_df, method == "GPCA-AGHQ")$hours, 2)
+naomi_time_tmb_hours <- signif(filter(time_taken_df, method == "GEB")$hours, 2)
+naomi_time_tmb_minutes <- signif(filter(time_taken_df, method == "GEB")$mins, 2)
+```
+
+Inference with NUTS took `r naomi_time_tmbstan_hours` hours, while inference with GPCA-AGHQ took `r naomi_time_aghq_hours` hours and GEB just `r naomi_time_tmb_hours` hours (equivalent to `r naomi_time_tmb_minutes` minutes).
+Both the NUTS and GPCA-AGHQ algorithms can be run under a range of settings, trading off accuracy and runtime.
 
@@
 The root mean square error (RMSE) between posterior mean estimates from GPCA-AGHQ and NUTS (`r rmse_aghq_mean`) was `r abs(rmse_diff_mean)`% lower than that between GEB and NUTS (`r rmse_tmb_mean`).
-For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_aghq_sd)` (PCA-AGHQ).
+For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (GEB) to `r abs(rmse_aghq_sd)` (GPCA-AGHQ).
 However, improvements in latent field estimate accuracy transferred to model outputs to a more limited and mixed extent (Supplementary Figures 12 and 13).
@@
 Most notably, for the latent field parameters $\texttt{ui\_lambda\_x}$ the average test statistic for GEB was substantially higher than for GPCA-AGHQ.
-This length 32 parameter $\mathbf{u}_x^{\lambda}$ plays a key role in the ART attendance component of the Naomi.
+This length 32 parameter $\mathbf{u}_x^{\lambda}$ plays a key role in the ART attendance component of the Naomi (Section \@ref(sec:art)).
 
@@
 (ref:ui-lambda-x) The parameter ``r paste0(ui_lambda_x$par)`` had the greatest difference in KS test statistics between GEB and GPCA-AGHQ to NUTS.
-For this parameter, the potential scale reduction factor was `r ui_lambda_x$rhat` and effective sample size was `r ui_lambda_x$ess`.
+For this parameter, the potential scale reduction factor was `r ui_lambda_x$rhat` and effective sample size was `r ui_lambda_x$ess`, indicating that NUTS results are likely to be accurate.
 
@@
 The average KS test statistic for GPCA-AGHQ (`r signif(ks_summary_summarise$aghq, 2)`) was `r signif(100 * ks_pct, 2)`% less than the average KS test statistic for GEB (`r signif(ks_summary_summarise$tmb, 2)`).
+Figure \@ref(fig:ks-summary) shows the differences in average KS test statistic for each latent field parameter.
+GCPA-AGHQ showed the greatest average improvement in KS test statistic for a vector of parameters in the ART attendance component of Naomi (Figure \@ref(fig:ui-lambda-x)).
 For both GEB and GPCA-AGHQ the KS test statistic for a parameter was correlated with low NUTS ESS (Supplementary Figure 14).
@@
 
-## Exceedance probabilities {#sec:exceedance}
+### Exceedance probabilities {#sec:exceedance}
 
@@
 
-### Meeting the second 90
+#### Meeting the second 90
 
-(ref:exceedance-second90) The probability each strata has met the second 90 (ART coverage above 81\\%) calculated using each inference method, as compared with NUTS.
+(ref:exceedance-second90) The probability each strata has met the second 90 (ART coverage above 81\%) calculated using each inference method, as compared with NUTS.
 The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
@@
 
-### Finding strata with high incidence
+#### Finding strata with high incidence
 
-(ref:exceedance-1inc) The probability each strata has high HIV incidence (above 1\\% per year) calculated using each inference method, as compared with NUTS.
+(ref:exceedance-1inc) The probability each strata has high HIV incidence (above 1\% per year) calculated using each inference method, as compared with NUTS.
 The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.
@@
 This is because the Laplace approximation is exact when the integrand is Gaussian, so a single quadrature node is sufficient.
-The difficulty is, of course, knowing in advance which marginals will be non-Gaussian.
+The difficulty is knowing in advance which marginals will be non-Gaussian.
 This could be done if there were a cheap way to obtain posterior means, which could then be compared to posterior modes obtained using optimisation.
@@
 The measures of fit would have to be for marginals, ruling out approaches like PSIS [@yao2018yes] which operate on joint distributions.
-Finally, it may be possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.
+Finally, it is likely possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.
 

### 2024-12-31 src/docs_paper-jtb/paper.Rmd
Final fixes (again)

@@
 The Naomi model was difficult to efficiently sample from using NUTS via `tmbstan`.
-The lowest effective sample size (ESS) was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Supplementary Figure 1).
-The largest potential scale reduction factor ($\hat R$) was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Supplementary Figure 2A).
+The lowest effective sample size (ESS) was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Supplementary Figure 1A).
+The largest potential scale reduction factor ($\hat R$) was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Supplementary Figure 2).
 Supplementary Figure 3 shows traceplots for the parameters with the lowest ESS and $\hat R$ respectively.
@@
 
-## Time taken
+## Time taken comparison
 

### 2025-05-10 src/docs_paper-jtb/paper.Rmd
Start with revisions (a typo!)

@@
 Although the EB approach is fast, it does not take into account hyperparameter uncertainty in the latent field posterior, potentially resulting in underestimated posterior variances.
-This concern could have important practical implications for use of the estimates from the Naomi model, and motivated us to look for an approach closer full Bayesian inference.
+This concern could have important practical implications for use of the estimates from the Naomi model, and motivated us to look for an approach closer to full Bayesian inference.
 We developed a method based on adaptive Gauss-Hermite quadrature (AGHQ) extended to handle integration over many hyperparameters.

### 2025-05-10 src/docs_paper-jtb/revisions.Rmd
Start with revisions (a typo!)

---
title: Response to reviewers
author:
  - Adam Howes^[Department of Mathematics, Imperial College London]
  - Alex Stringer^[Department of Statistics and Actuarial Science, University of Waterloo]
  - Seth R. Flaxman^[Department of Computer Science, Oxford University]
  - Jeffrey W. Imai-Eaton^[Harvard T.H. Chan School of Public Health, Harvard University]
output:
  bookdown::pdf_document2:
    toc: no
    number_sections: yes
    keep_tex: yes
bibliography: citations.bib
header-includes:
  \usepackage{xr} \externaldocument{paper}
---

```{r echo = FALSE, message = FALSE}
library(dplyr)

options(scipen = 100)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = "center",
  fig.pos = "!h",
  out.extra = ""
)

knitr::opts_chunk$set()
```

Thank you for taking the time to read and review our paper "Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature".
In what follows we address your comments and suggestions point by point.
We appreciate the feedback to improve the work and have made the relevant changes to the manuscript.

\clearpage

# Reviewer 1 {-}

> Typos spotted

Thank you for spotting these two typographic errors.
We have corrected them in the manuscript.

# Reviewer 2 {-}

# References {#references .unnumbered}

### 2025-05-13 src/docs_paper-jtb/revisions.Rmd
Start to address revisions!

@@
 
+> On `tmbstan` as compared with e.g. `rstan`
+
+`tmbstan` does use the Stan NUTS algorithm directly.
+That said, the performance can differ depending on the model and parameterisation.
+@monnahan2018no discuss a binomial GLMM implemented in `tmbstan` and Stan directly in which there is essentially no performance difference.
+That said, in the supplementary information, two models "zdiag" and "growth" are discussed where it does appear that models implemented directly in Stan are faster (in the figure "TMB" refers to `tmbstan`).
+For this reason, I do think that implementing the model directly in Stan would be likely to have some performance benefit.
+As @monnahan2018no write:
+
+> Based on our performance tests (S1 Table, S1 Fig), substantial speed improvements by converting models to the Stan language is not guaranteed.
+
+```{r fig.height = 4}
+knitr::include_graphics("revisions/monnahan2028no_s1.png")
+```
+
+> Bayesian optimisation as a competitor in the space
+
+The goal of Bayesian optimisation (BO) is to find the maximum of a (black box) function.
+For this reason, we do not see BO as a competitor to algorithms (like ours) intended to perform Bayesian inference.
+One could imagine using BO to optimise the hyperparameters, however it would likely be preferable to use gradients (as they are available).
+
 # Reviewer 2 {-}
 
+> ...the method is tunable with how many PCA components are selected and what value of $k$ to use...
+
+We agree that it would be of interest to explore the effect of choosing a different number of PCA components ($s$) and number of quadrature points per dimension ($s$).
+Furthermore, one could consider a varying number of quadrature points.
+A particular instance of this approach could look like "gradually lowering $k$ over the principal components".
+
+> ...a useful intermediate scheme to me seems to be to drop the corners of the grid...
+
+We agree that this could be a useful approach.
+Indeed Section 6.1.2. reads "Finally, it is likely possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.".
+We believe the suggested approach is what we refer to here (very briefly) as "pruning".
+
 # References {#references .unnumbered}

### 2025-05-18 src/docs_paper-jtb/revisions.Rmd
Figure related edits

@@
 
+> Questions about the HIV model
+
+Jeff broadly agrees with the reviewer.
+(This is an area of current work on Naomi extension.)
+
+> Selecting dimensions in proportion to variance in some outcome explained
+
 > Bayesian optimisation as a competitor in the space
@@
 # Reviewer 2 {-}
+
+> On Figure 5
+
+Thank you for the suggestion.
+We have adjusted the `alpha` value from 0.5 to 0.1 and added a caption to the figure explaining the possible overlap to make the figure clearer.
+
+```{r}
+knitr::include_graphics("revisions/fig5_edit.png")
+```
+
+> On Figure 6
+
+Thank you for noting that the $y$-axis here is ambiguous.
+We have clarified this here and in other instances throughout the manuscript.
+(Remains to do this!)
+
+> On Figure 7
+
+We agree that there are too many colours here for this figure to be clear.
+Thank you for the good suggestion to split the plot into multiple panels.
+We have done this, choosing to group the parameters by the component of the model they are most associated with.
+(Remains to do this!)
 

### 2025-05-26 src/docs_paper-jtb/paper.Rmd
Cafe revisions!

@@
 Other software packages have been developed which also use `TMB` C++ templates.
-The `tmbstan` R package [@monnahan2018no] allows running the Hamiltonian Monte Carlo (HMC) algorithm via `Stan`.
+The `tmbstan` R package [@monnahan2018no] allows running the Hamiltonian Monte Carlo (HMC) algorithm via Stan [@rstan].
 The `aghq` R package [@stringer2021implementing] allows use of AGHQ, and AGHQ over the marginal Laplace approximation, via the `mvQuad` R package [@weiser2016mvquad].
@@
 It may be difficult to implement such algorithms using `TMB`.
-Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language [@de2017programming], however requiring rewriting the Naomi model log-posterior outside of `TMB` would be a downside.
+Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language (PPL) [@de2017programming].
+Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
+However, requiring rewriting the Naomi model log-posterior outside of `TMB` would be a subtantial downside.
 Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
@@
 The measures of fit would have to be for marginals, ruling out approaches like PSIS [@yao2018yes] which operate on joint distributions.
-Finally, it is likely possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.
+Finally, it is likely possible to achieve better performance by pruning (dropping points where the weight is below some threshold) or prerotation (before adaption, in order to e.g.
+achieve better coverage along the principal axis) of the quadrature grid [@jackel2005note].
 

### 2025-05-26 src/docs_paper-jtb/revisions.Rmd
Cafe revisions!

--- 2025-05-10 src/docs_paper-jtb/revisions.Rmd
@@
 
+> On `tmbstan` as compared with e.g. `rstan`
+
+`tmbstan` passes the objective function and its gradient (calculated in `TMB`) directly to `rstan` via the [`op_matrix_vari`](https://mc-stan.org/math/classstan_1_1math_1_1op__matrix__vari.html).
+As such, the NUTS algorithm used should be the same.
+Nonetheless, it could be that there are speed differences between function evaluations in `TMB` and Stan.
+We have added a sentence to the discussion noting this possible comparison: "Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder]."
+
+> Questions about the HIV model
+
+Jeff broadly agrees with the reviewer.
+(This is an area of current work on Naomi extension.)
+
+> Selecting dimensions in proportion to variance in some outcome explained
+
+It is possible to obtain derivative information about a quantity in a `TMB` model using `ADREPORT`.
+
+> Bayesian optimisation as a competitor in the space
+
+The goal of Bayesian optimisation (BO) is to find the maximum of a (black box) function.
+For this reason, we do not see BO as a competitor to algorithms (like ours) intended to perform Bayesian inference.
+(One could imagine using BO to optimise the hyperparameters, however it would be preferable to use gradients [as they are available].)
+
 > Typos spotted
@@
 
+> On Figure 5
+
+Thank you for the suggestion.
+We have adjusted the `alpha` value from 0.5 to 0.1 and added a caption to the figure explaining the possible overlap to make the figure clearer:
+
+```{r}
+knitr::include_graphics("revisions/fig5_edit.png")
+```
+
+> On Figure 6
+
+Yes that is correct, it's the NUTS estimate minus the GEB or GPCA-AGHQ estimate.
+We have clarified the $y$-axis here (and where relevant throughout the manuscript) to read "Absolute difference (NUTS - method)".
+
+> On Figure 7
+
+We agree that there are too many colours here for this figure to be clear.
+Thank you for the suggestion to split the plot into multiple panels.
+We have done this, choosing to group the parameters by the component of the model they are most associated with.
+(Remains to do this!)
+
+> ...the method is tunable with how many PCA components are selected and what value of $k$ to use...
+
+We agree that it would be of interest to explore the effect of choosing a different number of PCA components ($s$) and number of quadrature points per dimension ($k$).
+One could also consider a varying number of quadrature points.
+For example "gradually lowering $k$ over the principal components" presumably would be the most efficient approach.
+
+> ...a useful intermediate scheme to me seems to be to drop the corners of the grid...
+
+We agree that this could be a useful approach.
+Indeed Section 6.1.2. reads "Finally, it is likely possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.".
+We believe the suggested approach is what we refer to here (very briefly) as "pruning".
+To clarify this point, we have rewritten this sentence as: "Finally, it is likely possible to achieve better performance by pruning (dropping points where the weight is below some threshold) or prerotation (before adaption, in order to e.g.
+achieve better coverage along the principal axis) of the quadrature grid [@jackel2005note].".
+
 # References {#references .unnumbered}

### 2025-05-26 src/naomi-simple_point-estimates/point-estimates.Rmd
Cafe revisions!

@@
   geom_text(data = df_metrics_latent, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
-  labs(x = "Estimate", y = "Absolute difference to NUTS") +
+  labs(x = "Estimate", y = "Absolute difference (NUTS - method)") +
   theme_minimal()
@@
   ) +
-  labs(x = "Estimate", y = "Absolute difference to NUTS") +
+  labs(x = "Estimate", y = "Absolute difference (NUTS - method)") +
   theme_minimal()
@@
   geom_text(data = df_metrics_output, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
-  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
+  labs(x = "Estimate", y = "Absolute difference (NUTS - method)", col = "Output") +
   theme_minimal() +
@@
   ) +
-  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
+  labs(x = "Estimate", y = "Absolute difference (NUTS - method)", col = "Output") +
   theme_minimal()
@@
   ) +
-  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
+  labs(x = "Estimate", y = "Absolute difference (NUTS - method)", col = "Output") +
   theme_minimal() +

### 2025-05-26 src/naomi-simple_ks/ks.Rmd
Improve KS summary figure

--- 2023-12-28 src/naomi-simple_ks/ks.Rmd
@@
 
-ks_summary_latent <- filter(ks_summary, type == "Latent")
+fct_case_when <- function(...) {
+  args <- as.list(match.call())
+  levels <- sapply(args[-1], function(f) f[[3]])  # extract RHS of formula
+  levels <- levels[!is.na(levels)]
+  factor(dplyr::case_when(...), levels=levels)
+}
+
+# Include group information
+ks_summary_latent <- filter(ks_summary, type == "Latent") %>%
+  mutate(
+    component = fct_case_when(
+      parname %in% c("beta_rho", "u_rho_a", "u_rho_as", "u_rho_x", "u_rho_xs", "us_rho_x", "us_rho_xs") ~ "HIV prevalence",
+      parname %in% c("beta_lambda", "ui_lambda_x") ~ "HIV incidence",
+      parname %in% c("beta_alpha", "u_alpha_a", "u_alpha_as", "u_alpha_x", "u_alpha_xs", "us_alpha_x", "us_alpha_xs", "u_alpha_xa", "ui_anc_alpha_x", "ui_anc_rho_x", "log_or_gamma") ~ "ART coverage",
+      TRUE ~ "Other"
+    ),
+    label = fct_case_when(
+      parname %in% c("beta_alpha", "beta_rho", "beta_lambda") ~ "Intercept",
+      parname %in% c("u_alpha_a", "u_rho_a") ~ "Age effect",
+      parname %in% c("u_alpha_x", "u_alpha_xs", "u_rho_x", "us_rho_x", "us_alpha_x", "ui_lambda_x", "ui_anc_alpha_x", "ui_anc_rho_x") ~ "Space effect",
+      parname %in% c("u_alpha_as", "u_rho_as", "u_alpha_xa") ~ "Age-sex effect",
+      parname %in% c("us_rho_xs", "us_alpha_xs", "u_rho_xs") ~ "Space-sex effect",
+      parname %in% c("log_or_gamma") ~ "Log odds ratio",
+      TRUE ~ "Other"
+    )
+  )
 
@@
 ks_summary_latent %>%
-  ggplot(aes(x = TMB, y = aghq - TMB, col = parname, size = size)) +
-  geom_point(alpha = 0.7) +
-  scale_color_manual(values = extended_cbpalette(n = 20)) +
-  scale_size_continuous(breaks = c(2, 10, 32), labels = c(2, 10, 32)) +
-  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
-  labs(x = "KS(GEB, NUTS)", y = "KS(GPCA-AGHQ, NUTS) - KS(GEB, NUTS)", col = "Parameter", size = "Length") +
+  ggplot(aes(x = TMB, y = aghq - TMB, col = label, size = size)) +
+  facet_wrap(~ component, ncol = 1) +
+  geom_point(shape = 1) +
+  scale_color_manual(values = multi.utils::cbpalette()) +
+  geom_abline(slope = 0, intercept = 0, linetype = "dashed", col = "grey") +
+  scale_size_continuous(breaks = c(2, 10, 32), labels = c(2, 10, 32), range = c(2, 4)) +
+  scale_y_continuous(limits = c(-0.08, 0.08)) +
+  labs(x = "Average of KS(GEB)", y = "Average of KS(GPCA-AGHQ) - KS(GEB)", col = "Parameter type", size = "Parameter length") +
   theme_minimal() +
-  guides(size = guide_legend(title.position = "top", direction = "vertical"), col = guide_legend(title.position = "top")) +
-  theme(legend.position = "bottom", legend.title = element_text(size = rel(0.9)), legend.text = element_text(size = rel(0.65)))
+  guides(size = guide_legend(title.position = "top", direction = "horizontal"), col = guide_legend(title.position = "top", nrow = 6, byrow = TRUE)) +
+  theme(
+    legend.position = "right",
+    legend.title = element_text(size = rel(0.9)),
+    legend.text = element_text(size = rel(0.9)),
+    panel.spacing.y = unit(1.2, "lines")
+  )
 
-ggsave("ks-summary.png", h = 5, w = 6.25, bg = "white")
+ggsave("ks-summary.png", h = 5.25, w = 6.25, bg = "white")
 ```

### 2025-05-26 src/docs_paper-jtb/revisions.Rmd
Figure update text

@@
 
-We agree that there are too many colours here for this figure to be clear.
-Thank you for the suggestion to split the plot into multiple panels.
-We have done this, choosing to group the parameters by the component of the model they are most associated with.
-(Remains to do this!)
+We agree that there are too many colours here for this figure to be clear, and that the suggestion to split the figure across panels is a good one.
+To do so, we have grouped the parameters by the model component (e.g. HIV prevalence), and used colour to represent the type of parameter (e.g. intecept).
+
+```{r}
+knitr::include_graphics("depends/ks-summary.png")
+```
 

### 2025-06-01 src/docs_paper-jtb/revisions.Rmd
Almost there!

@@
 
-> Questions about the HIV model
+> On the proportion recently infected and ANC offsets
 
-Jeff broadly agrees with the reviewer.
-(This is an area of current work on Naomi extension.)
+Some HIV surveys provide bio-marker data on whether or not the infection was recent.
+The equation in the manuscript links the probability of recent infection to HIV prevalence, HIV incidence, and properties of the recent infection assay.
+This allows us to back-out estimates of the HIV incidence.
+<!-- Is there something to add to the manuscript here? -->
+
+We agree that tightening the bounds on the ANC offsets would be an important step for improving estimates of HIV incidence.
+Work is ongoing to improve the Naomi model [@esra2024improved].
+<!-- Has this work been included yet, or it's still to be done in future? -->
 
@@
 We agree that there are too many colours here for this figure to be clear, and that the suggestion to split the figure across panels is a good one.
-To do so, we have grouped the parameters by the model component (e.g. HIV prevalence), and used colour to represent the type of parameter (e.g. intecept).
+To do so, we have grouped the parameters by the model component (e.g. HIV prevalence), and used colour to represent the type of parameter (e.g. intercept).
 

### 2025-06-01 src/naomi-simple_ks/ks.Rmd
Almost there!

@@
 
-fct_reorg <- function(fac, ...) {
-  fct_recode(fct_relevel(fac, ...), ...)
-}
-
 ks_df %>%
   filter(type != "Hyper") %>%
-  mutate(method = fct_reorg(method, "GEB" = "TMB", "GPCA-AGHQ" = "aghq")) %>%
+  mutate(
+    method = fct_recode(
+      method,
+      "GEB" = "TMB",
+      "GPCA-AGHQ" = "aghq",
+    ),
+    method = fct_relevel(method, "GEB", "GPCA-AGHQ")
+  ) %>%
   left_join(data.frame(ess) %>%

### 2025-06-09 src/docs_paper-jtb/revisions.Rmd
Changes required for running and putting on Overleaf

@@
   bookdown::pdf_document2:
-    toc: no
-    number_sections: yes
-    keep_tex: yes
+    toc: true
+    number_sections: true
+    keep_tex: true
+    includes:
+      in_header: preamble.tex
+    extra_dependencies: awesomebox
 bibliography: citations.bib
@@
 
+\clearpage
+
+# Introduction
+
 Thank you for taking the time to read and review our paper "Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature".
@@
 
-\clearpage
+# Reviewer 1
 
-# Reviewer 1 {-}
+\begin{reviewer}
+The class of models to which the present model is assigned, namely Extended Latent Gaussian Models (ELGMs), is very broad, and in my (perhaps rather curmudgeonly) opinion too broad to be of much value.
+On page 3 the authors write about the new mapping structure for ELGMs which is said to "relax the restriction that there is a one-to-one mapping between the mean response mu and structured additive predictor eta." and "Importantly, this mapping allows for the presence of non-linearity in the model." What exactly is meant by this?
+The original link function of LGMs can be non-linear of course, though it must sets up invertibility which is easy if it is one-to-one in that sense.
+Here the many-to-one refers to a mapping from multiple latent parameters to a combined latent parameter, so for example if we're talking about direct summation then yes we can have equality on a set of zero-measure for the many (the pull back of the 'one'), but I bet that for the model to work (in the sense of obeying the conditions of the ELGM convergence proofs) we couldn't, say, define the mapping such that its pull-back is sometimes a set of measure zero and sometimes a set of non-zero measure.
+It probably wouldn't work if we did a non-linear mapping like the product of the many.
+But who knows, because all this is left to the user of the ELGMs to figure out: that is, the original ELGM paper is constructed backwards: rather than defining a very strict model class with conditions that ensure a shared computability and set of theoretical results (like the LGM or GLM class) it is instead defined very loosely, meaning that finding out if it's computable using the given algorithms is left up to the user.
+Rant over.
+What can the authors do in the present paper?
+First, I would suggest to be clear on page 3 about what kind of mappings are being done in this model and see if these also match to a named smaller class within the ELGM class, and/or identify fuzzy matches to existing model structures of this form (such as Katie Wilson's "pointless spatial models"); especially those use in spatial statistics and/or epi.
+\end{reviewer}
 
-> On `tmbstan` as compared with e.g. `rstan`
+Response here.
+
+\begin{reviewer}
+I'm not an expert on the various flavours of NUTS algorithms but from my experience I have always been disappointed in the performance of tmbstan.
+Are there any comparisons between tmbstan and ordinary stan not conducted by the tmbstan package authors?
+Do people use tmbstan much in research?
+The software paper is well-cited but a lot of citations seem to be from other people just talking about software or HMC generally.
+Perhaps the best way to ally my concern that tmbstan might not be a good benchmark would be to check and explain in the paper whether its flavour of NUTS implementation is comparable to that now used in Stan?
+If not, perhaps see if a caveat is needed, something along the lines of 'tmbstan was tested because it's compatible with the idea of repurposing existing software (i.e., the original TMB template) but this doesn't mean that rebuilding the model from scratch in Stan, for instance, would create a sampler just as impractically slow'.
+\end{reviewer}
 
@@
 Nonetheless, it could be that there are speed differences between function evaluations in `TMB` and Stan.
-We have added a sentence to the discussion noting this possible comparison: "Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder]."
+We have added a sentence to the discussion noting this possible comparison:
 
-> On the proportion recently infected and ANC offsets
+> Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
+
+\begin{reviewer}
+The equation for the fraction recently infected on page 17 puzzles me; though I don't come from the HIV field.
+It seems to me that this is a convenience 'guesstimation' used only as an output for end users, rather than being fitted to data or offering some kind structural constraint on the fits?
+Can these model outputs be pushed into a compartmental model later to give 'proper' incidence estimates, i.e., ones inferred from changes to prevalence and death rates?
+Also, presumably the epidemiology prescribes some tight bounds on the ANC offsets given the demography of birth and sexual partnership; are these pre-built into the priors for the model?
+\end{reviewer}
 
@@
 
-> Selecting dimensions in proportion to variance in some outcome explained
+\begin{reviewer}
+You anticipated my very clever idea for improving the performance of the PCA algorithm: namely, to consider the relevance of each dimension to a measure of uncertainty relevant to the end predictive purposes.
+Are calculations of this nature possible with TMB as it stands, or would one need to dig into the bowels of the c++ code and create a TMB package extension?
+I think this is an interesting area for future work.
+\end{reviewer}
 
@@
 
-> Bayesian optimisation as a competitor in the space
+\begin{reviewer}
+Finally, on page 17 it feels like Bayesian 'optimisation' is another natural competitor in this space, especially since it's quite straightforward to just run this kind of algorithm to explore hyper parameters until you get bored or run out of time, i.e., you get what you get from stopping early, unlike HMC for instance.
+\end{reviewer}
 
@@
 
-> Typos spotted
+\begin{reviewer}
+Typos spotted:
+p2 "closer [to]"
+p4. "hyperparmters"
+\end{reviewer}
 
@@
 
-# Reviewer 2 {-}
+# Reviewer 2
 
-> On Figure 5
+\begin{reviewer}
+The NUTS sampler was run for 4 chains for 100,000 iterations, with the first half discarded.
+This gives a lowest ESS of 208 and a largest Rhat of 1.021.
+This run took 79 hours.
+While this run is pretty reasonable for applied work, I do worry about it being used as a gold standard to compare against.
+Given that the authors are comparing the full posterior distribution, the tails are quite important which will be poorly characterised with ESS of 208.
+The Rhat of 1.021 is just about ok, so only slightly longer runs, but many more chains would, at fairly little expense, really improve the ESS.
+Can the model be run with 16 chains on an HPC or something?
+\end{reviewer}
+
+Response here.
+
+\begin{reviewer}
+I really like Figure 5 but just a few more words explaining the green lines would be useful.
+My first thought was that there was 1 or 3 green lines and that the width meant something.
+I realise now it's 6561 green lines that either overlap or join together to make thick lines.
+Perhaps some alpha would help communicate this.
+\end{reviewer}
 
@@
 
-> On Figure 6
+\begin{reviewer}
+In Figure 6, is the y axis NUTS estimate - GEB estimate?
+Or the other way around?
+For the posterior SD, you can interpret positive and negative errors as being conservative or anti-conservative.
+The new method moves the 3 little clusters at 1.1, 1.4 and 1.8 from negative to positive which seems it could be important and worth noting.
+\end{reviewer}
 
@@
 
-> On Figure 7
+\begin{reviewer}
+Figure 7 is very hard to read.
+It's an interesting and useful plot.
+However, I would much prefer the labels to be proper greek characters.
+And there is just too many colours, so I think it needs to be split into 4 or so groups on seperate panels.
+\end{reviewer}
 
@@
 
-> ...the method is tunable with how many PCA components are selected and what value of $k$ to use...
+\begin{reviewer}
+The authors mention that the method is tunable with how many PCA components are selected and what value of k to use.
+It is a bit of a shame that one or two other options are not examined.
+But I understand it is already a lot of work to describe all the results.
+Given that this model is now on the slow side, I would prioritise examining a method that accounts for 80% of the variance, i.e.
+a faster model rather than a more accurate model.
+So I will leave it to the authors whether or not to do more work here.
+\end{reviewer}
 
@@
 
-> ...a useful intermediate scheme to me seems to be to drop the corners of the grid...
+\begin{reviewer}
+Just looking at Figure 2, a useful intermediate scheme to me seems to be to drop the corners of the grid.
+So you would be left with 5 points.
+That said, I think the benefit of this reduces as you get to higher dimensions.
+So please don't take this comments as work to be done, but rather as an interesting idea for future work.
+\end{reviewer}
 
@@
 We believe the suggested approach is what we refer to here (very briefly) as "pruning".
-To clarify this point, we have rewritten this sentence as: "Finally, it is likely possible to achieve better performance by pruning (dropping points where the weight is below some threshold) or prerotation (before adaption, in order to e.g.
-achieve better coverage along the principal axis) of the quadrature grid [@jackel2005note].".
+To clarify this point, we have rewritten this sentence as:
+
+> Finally, it is likely possible to achieve better performance by pruning (dropping points where the weight is below some threshold) or prerotation (before adaption, in order to e.g.
+achieve better coverage along the principal axis) of the quadrature grid [@jackel2005note].
 

### 2025-06-09 src/docs_paper-jtb/script.R

@@ -20,6 +20,7 @@ convert_pdf_png("nodes-samples-comparison")
 #' Render documents
 rmarkdown::render("paper.Rmd", clean = FALSE)
 rmarkdown::render("appendix.Rmd", clean = FALSE)
+rmarkdown::render("revisions.Rmd", clean = FALSE)
 
 pdftools::pdf_combine(c("paper.pdf", "appendix.pdf"), output = "naomi-aghq.pdf")
 

### 2025-07-04 src/docs_paper-jtb/revisions.Rmd
Refer to figures by number and add NUTS rerun details

@@
 Some HIV surveys provide bio-marker data on whether or not the infection was recent.
-The equation in the manuscript links the probability of recent infection to HIV prevalence, HIV incidence, and properties of the recent infection assay.
+Equation 7 links the probability of recent infection to HIV prevalence, HIV incidence, and properties of the recent infection assay.
 This allows us to back-out estimates of the HIV incidence.
@@
 
-Response here.
+We agree that running more iterations of the NUTS sampler is a good idea, and could improve the minimum ESS and accuracy of any comparisons, especially distributional ones.
+We have rerun the NUTS sampler using a HPC with the 18 chains (across 18 cores) with 120,000 iterations each.
 
@@
 Thank you for the suggestion.
-We have adjusted the `alpha` value from 0.5 to 0.1 and added a caption to the figure explaining the possible overlap to make the figure clearer:
+We have adjusted the `alpha` value from 0.5 to 0.1 and added a caption to the figure explaining the possible overlap to make Figure 5 clearer:
 
@@
 We agree that there are too many colours here for this figure to be clear, and that the suggestion to split the figure across panels is a good one.
-To do so, we have grouped the parameters by the model component (e.g. HIV prevalence), and used colour to represent the type of parameter (e.g. intercept).
+To do this, we have grouped the parameters by the model component (e.g. HIV prevalence), and used colour to represent the type of parameter (e.g. intercept).
+The updated version of Figure 7 is shown below:
 

### 2025-07-04 src/docs_paper-jtb/revisions.Rmd
Draft ELGM response

@@
 
-Response here.
+We have added the following text in Section 2.2 with the aim of making the mapping between the latent field and structured additive predictor of the Naomi model clearer:
+
+> Equations...
+
+We are not aware of any named smaller class of ELGMs to which the Naomi model belongs.
+The available convergence theory for ELGMs is very restrictive and it is not clear whether or how the available results apply to the Naomi model; accordingly, we do not make any claims about theoretical performance in the present paper.
+Our empirical analysis shows that we obtain results that are practically similar to gold-standard NUTS results.
 

### 2025-07-04 src/docs_paper-jtb/paper.Rmd
Figure update text
Move ADREPORT into main text and update reply

@@
 Most notably, for the latent field parameters $\texttt{ui\_lambda\_x}$ the average test statistic for GEB was substantially higher than for GPCA-AGHQ.
-This length 32 parameter $\mathbf{u}_x^{\lambda}$ plays a key role in the ART attendance component of the Naomi (Section \@ref(sec:art)).
+This length 32 parameter $\mathbf{u}_x^{\lambda}$ plays a key role in modelling ART attendance (Section \@ref(sec:art)).
 
@@
 Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
-However, requiring rewriting the Naomi model log-posterior outside of `TMB` would be a subtantial downside.
+However, requiring rewriting the Naomi model log-posterior outside of `TMB` would be a substantial downside.
 Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
@@
 As such, using alternative measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] or Sobol indices, could be preferable.
+This may be achievable by applying the `ADREPORT` function to model outputs within the `TMB` template to obtain gradients.
 Second, use of PCA is challenging when the dimensions have different scales.

### 2025-07-04 src/docs_paper-jtb/revisions.Rmd
Move ADREPORT into main text and update reply

@@
 
-It is possible to obtain derivative information about a quantity in a `TMB` model using `ADREPORT`.
+We agree this is an interesting area for future work, and have added a sentence to Section 6.1.2. discussing the possibility of doing this in `TMB`:
+
+> This may be achievable by applying the `ADREPORT` function to model outputs within the `TMB` template to obtain gradients.
 

### 2025-07-04 src/docs_paper-jtb/revisions.Rmd
Soften reply

@@
 
-The goal of Bayesian optimisation (BO) is to find the maximum of a (black box) function.
-For this reason, we do not see BO as a competitor to algorithms (like ours) intended to perform Bayesian inference.
-(One could imagine using BO to optimise the hyperparameters, however it would be preferable to use gradients [as they are available].)
+Our understanding is that the goal of Bayesian optimisation (BO) is to find the maximum of a (black box) function, rather than to perform Bayesian inference.
+One could imagine using BO to optimise the hyperparameters, however it would be preferable to use gradients as they are available.
 

### 2025-07-04 src/docs_paper-jtb/revisions.Rmd
Alter response on s and k

@@
 
-We agree that it would be of interest to explore the effect of choosing a different number of PCA components ($s$) and number of quadrature points per dimension ($k$).
-One could also consider a varying number of quadrature points.
-For example "gradually lowering $k$ over the principal components" presumably would be the most efficient approach.
+In the current work we focused on results which were as accurate as possible while still running in a practical amount of time.
+We agree that it is possible that faster results that are nearly as accurate could be obtained using or procedure.
 

### 2025-07-04 src/docs_paper-jtb/paper.Rmd
Add clarification about LGM

@@
 
-## Latent Gaussian model
+## Latent Gaussian model {#sec:lgm}
 
@@
 with latent field and hyperparameter priors as in the LGM case.
+
+Section \@ref(sec:naomi) show that the Naomi model involves many-to-one mappings of latent field components to the structured additive predictor for each response.
+Because of this, the Naomi model cannot be written as an LGM of the form described in Section \@ref(sec:lgm), but can be written as an ELGM.
 
@@
 \begin{equation}
-\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}.
+\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}. \label{eq:weighted-aggregate}
 \end{equation}
@@
 \begin{equation}
-m^{\theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
+m^{\theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2}, \label{eq:kiss}
 \end{equation}
@@
 \begin{equation}
-y^{\theta}_{I} \sim \text{xBin}(m^{\theta}_{I}, \theta_{I})
+y^{\theta}_{I} \sim \text{xBin}(m^{\theta}_{I}, \theta_{I}) \label{eq:xbin}
 \end{equation}
@@
 \alpha_{I} = \frac{\sum_{i \in I} N_i \rho_i \alpha_i}{\sum_{i \in I} N_i \rho_i}, \quad
-\kappa_{I} = \frac{\sum_{i \in I} N_i \rho_i \kappa_i}{\sum_{i \in I} N_i \rho_i}.
+\kappa_{I} = \frac{\sum_{i \in I} N_i \rho_i \kappa_i}{\sum_{i \in I} N_i \rho_i}. \label{eq:weighted-aggregate2}
 \end{equation}

### 2025-07-04 src/docs_paper-jtb/paper.Rmd
Include equation numbers

@@
 
-Section \@ref(sec:naomi) show that the Naomi model involves many-to-one mappings of latent field components to the structured additive predictor for each response.
+Section \@ref(sec:naomi) (e.g.
+Equations \@ref(eq:weighted-aggregate) to \@ref(eq:weighted-aggregate2)) shows that the Naomi model involves many-to-one mappings of latent field components to the structured additive predictor for each response.
 Because of this, the Naomi model cannot be written as an LGM of the form described in Section \@ref(sec:lgm), but can be written as an ELGM.
@@
 Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language (PPL) [@de2017programming].
-Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
+Direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
 However, requiring rewriting the Naomi model log-posterior outside of `TMB` would be a substantial downside.

### 2025-07-04 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Small updates to convergence report

@@
 
-color_scheme_set("viridis")
+bayesplot::color_scheme_set("viridis")
 ggplot2::theme_set(theme_minimal())
+```
+
+```{r}
+# This is code I tried when one chain was not converging -- trying for a rerun instead!
+remove_chains = function(fit, drop) {
+  sim <- fit@sim
+  sim$samples <- sim$samples[-drop]
+  sim$chains <- sim$chains - length(drop)
+  sim$warmup2 <- sim$warmup2[-drop]
+  sim$n_save <- sim$n_save[-drop]
+  fit@sim <- sim
+  fit
+}
 ```
@@
 (rhat_mean <- mean(mcmc_summary[, "Rhat"]))
-(rhat_above_threshold <- mean(mcmc_summary[, "Rhat"] < 1.01))
+(rhat_above_threshold <- mean(mcmc_summary[, "Rhat"] > 1.01))
 ```
@@
   )
-
 ```
@@
 ```{r}
-cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#E69F00", "#E69F00")
+cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00")
 bayesplot::color_scheme_set(rev(cbpalette))

### 2025-07-04 src/docs_paper-jtb/revisions.Rmd
Add clarification about LGM
Include equation numbers
Better grammar

@@
 
-> Equations...
+> Section 3 (e.g.
+Equations(8) to (11)) shows that the Naomi model involves many-to-one mappings of latent field components to the structured additive predictor for each response.
+Because of this, the Naomi model cannot be written as an LGM of the form described in Section 2.1, but can be written as an ELGM.
 
@@
 
-We agree this is an interesting area for future work, and have added a sentence to Section 6.1.2. discussing the possibility of doing this in `TMB`:
+We agree that this is an interesting area for future work, and have added a sentence to Section 6.1.2. discussing the possibility of doing this in `TMB`:
 

### 2025-08-05 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Run through of MCMC convergence report with new results

@@
 
-```{r}
-# This is code I tried when one chain was not converging -- trying for a rerun instead!
-remove_chains = function(fit, drop) {
-  sim <- fit@sim
-  sim$samples <- sim$samples[-drop]
-  sim$chains <- sim$chains - length(drop)
-  sim$warmup2 <- sim$warmup2[-drop]
-  sim$n_save <- sim$n_save[-drop]
-  fit@sim <- sim
-  fit
-}
-```
-
 # $\hat R$
@@
   geom_segment(aes(yend = parameter, xend = ifelse(min(value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7, col = "#56B4E9") +
-  lims(x = c(NA, 1.03)) +
+  lims(x = c(NA, 1.012)) +
   labs(x = "Potential scale reduction factor", y = "Parameter", col = "") +
@@
     geom_histogram(col = "grey60", fill = "grey80") +
-    labs(x = "Effective sample size", y = "Count", tag = "B")
+    scale_x_log10() +
+    labs(x = "Effective sample size (log 10 scale)", y = "Count", tag = "B")
 
@@
 ```{r}
-worst_eff <- bayesplot::mcmc_trace(mcmc, pars = names(which.min(mcmc_summary[, "n_eff"]))) + labs(tag = "A") + guides(col = "none")
-worst_rhat <- bayesplot::mcmc_trace(mcmc, pars = names(which.max(rhats))) + labs(tag = "B")
+worst_ess_par <- names(which.min(mcmc_summary[, "n_eff"]))
+worst_rhat_par <- names(which.max(rhats))
 
-worst_eff + worst_rhat
+worst_df <- bayesplot::mcmc_trace_data(mcmc, pars = c(worst_ess_par, worst_rhat_par)) |>
+  mutate(
+    chain_group = case_when(
+      chain %in% 1:6 ~ "Chains 1–6",
+      chain %in% 7:12 ~ "Chains 7–12",
+      chain %in% 13:18 ~ "Chains 13–18"
+    ),
+    chain_id = as.numeric(chain) %% 6 + 1
+  )
 
-ggsave("worst-trace.png", h = 3, w = 6.25)
+worst <- ggplot(worst_df, aes(x = iteration, y = value, col = as.factor(chain_id))) +
+  geom_line() +
+  scale_color_manual(values = cbpalette) +
+  facet_grid(chain_group ~ parameter) +
+  guides(col = "none") +
+  labs(x = "Iteration", y = "") +
+  theme_minimal()
+
+ggsave("worst-trace.png", worst, h = 4.5, w = 6.25)
 ```
@@
 
-pairs_df <- extract(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a")) %>%
+pairs_df <- rstan::extract(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a")) %>%
   as.data.frame()
@@
 
+sf_to_nb <- function(sf){
+  nb <- spdep::poly2nb(as(sf, "Spatial"))
+  return(nb)
+}
+
 nb <- area_merged %>%
   filter(area_level == max(area_level)) %>%
-  bsae::sf_to_nb()
+  sf_to_nb()
 

### 2025-08-05 src/naomi-simple_mmd/mmd.Rmd
Run MMD

@@
 stopifnot(names(tmb$fit$sample) == names(aghq$quad$sample))
-stopifnot(names(tmb$fit$sample) == names(tmbstan$mcmc$sample))
+stopifnot(setequal(names(tmb$fit$sample), names(tmbstan$mcmc$sample)))
 ```
@@
 
-#' Keep only the final 1000 samples from tmbstan
-tmbstan_samples <- tmbstan_samples[, 4001:5000]
+#' Thin down to 1000 samples from tmbstan
+tmbstan_samples <- tmbstan_samples[, seq(from = 1, to = ncol(tmbstan_samples), length.out = 1000)]
 

### 2025-08-05 src/naomi-simple_ks/ks.Rmd
Rerun of KS with correct MCMC samples

@@
 ess_ratio <- bayesplot::neff_ratio(tmbstan$mcmc$stanfit)
-niter <- 0.5 * 4 * tmbstan_details$niter / tmbstan_details$nthin
+niter <- 0.5 * tmbstan_details$nchains * tmbstan_details$niter / tmbstan_details$nthin
 ess <- ess_ratio * niter

### 2025-08-09 src/docs_paper-jtb/paper.Rmd
Update text regarding the NUTS diagnostics

@@
 GPCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$, 1000 samples \\
-NUTS & \texttt{tmbstan} & 4 chains of 100000 iterations, with the first half of each chain discarded as warmup, thinned by a factor of 20, to give a total of 10000 samples.
-Default NUTS tuning parameters \citep{hoffman2014no}.
+NUTS & \texttt{tmbstan} & 18 chains of 40000 iterations, with the first half of each chain discarded as warmup, thinned by a factor of 20, to give a total of 18000 samples.
+The target average acceptance probability was set to 0.95 and the maximum treedepth to 12 \citep{hoffman2014no}.
 \\
@@
 
-The Naomi model was difficult to efficiently sample from using NUTS via `tmbstan`.
-The lowest effective sample size (ESS) was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Supplementary Figure 1A).
-The largest potential scale reduction factor ($\hat R$) was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Supplementary Figure 2).
+The Naomi model was challenging to efficiently sample from using NUTS via `tmbstan`.
+We used 18 chains of 40000 iterations run on a high-performance computing cluster (Table \@ref(tab:inference-methods)).
+The lowest resulting effective sample size (ESS) was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Supplementary Figure 1A).
+The largest resulting potential scale reduction factor ($\hat R$) was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Supplementary Figure 2).
 Supplementary Figure 3 shows traceplots for the parameters with the lowest ESS and $\hat R$ respectively.
-Though inaccuracies remain possible, these diagnostics are sufficient to treat inferences obtained from NUTS as a gold-standard.
+Based on these diagnostics, we considered the NUTS samples to be of a high quality and suitable for use as a gold-standard.
 

### 2025-08-09 src/docs_paper-jtb/revisions.Rmd
Update text regarding the NUTS diagnostics

@@
 We agree that running more iterations of the NUTS sampler is a good idea, and could improve the minimum ESS and accuracy of any comparisons, especially distributional ones.
-We have rerun the NUTS sampler using a HPC with the 18 chains (across 18 cores) with 120,000 iterations each.
+We have rerun the NUTS sampler using a HPC with 18 chains (across 18 cores) with 40,000 per chain and a thinning factor of 20.
+To improve the sampling also increased the `adapt_delta` parameter from the default of 0.8 to 0.95, and the `max_treedepth` parameter from the default of 10 to 12.
+The resulting diagnostics are much improved, as described in an update to Section 5.1:
+
+```{r}
+naomi_nuts <- readRDS("depends/mcmc-out.rds")
+```
+
+> The Naomi model was challenging to efficiently sample from using NUTS via `tmbstan`.
+We used 18 chains of 40000 iterations run on a high-performance computing cluster (Table \@ref(tab:inference-methods)).
+The lowest resulting effective sample size (ESS) was `r round(naomi_nuts$ess_min, 0)` (2.5% quantile `r round(naomi_nuts$ess_lower, 0)`, 50% quantile `r round(naomi_nuts$ess_median, 0)`, and 97.5% quantile `r round(naomi_nuts$ess_upper, 0)`; Supplementary Figure 1A).
+The largest resulting potential scale reduction factor ($\hat R$) was `r round(naomi_nuts$rhat_max, 3)` (2.5% quantile `r round(naomi_nuts$rhat_lower, 3)`, 50% quantile `r round(naomi_nuts$rhat_median, 3)`, and 97.5% quantile `r round(naomi_nuts$rhat_upper, 3)`; Supplementary Figure 2).
+Supplementary Figure 3 shows traceplots for the parameters with the lowest ESS and $\hat R$ respectively.
+Based on these diagnostics, we considered the NUTS samples to be of a high quality and suitable for use as a gold-standard.
 

### 2025-08-09 src/docs_paper-jtb/appendix.Rmd
Final fixes
Updates to the plots in the appendix regarding NUTS results

@@
 
-(ref:ess) The efficiency of the NUTS, as measured by the ratio of effective sample size to total number of iterations run, was low for most parameters (Panel A).
+(ref:ess) The efficiency of the NUTS, as measured by the ratio of effective sample size to total number of iterations run, was low for some parameters (Panel A).
 As a result, the number of iterations required for the effective number of samples (mean `r round(naomi_nuts$ess_mean)`) to be satisfactory was high (Panel B).
@@
 
-(ref:rhat) For NUTS run on the Naomi ELGM, the maximum potential scale reduction factor was `r round(naomi_nuts$rhat_max, 3)`, below the value of 1.05 typically used as a cutoff for acceptable chain mixing, indicating that the results are acceptable to use.
-Additionally, the vast majority (`r round(100 * naomi_nuts$rhat_above_threshold, 1)`%) of $\hat R$ values were less than 1.1.
+(ref:rhat) For NUTS run on the Naomi ELGM, the maximum potential scale reduction factor was `r round(naomi_nuts$rhat_max, 3)`, substantially below the value of 1.05 typically used as a cutoff for acceptable chain mixing, indicating that the results are of high quality.
+Additionally, all $\hat R$ values were less than 1.1.
 
@@
 
-(ref:worst-trace) Traceplots for the parameter with the lowest ESS which was `log_sigma_alpha_xs` (an $\text{ESS}$ of `r round(naomi_nuts$ess_min)`, Panel A) and highest potential scale reduction factor which was `ui_lambda_x[10]` (an $\hat R$ of `r round(naomi_nuts$rhat_max, 3)`, Panel B).
+(ref:worst-trace) Traceplots for the parameter with the lowest ESS which was `log_sigma_lambda_x` (an $\text{ESS}$ of `r round(naomi_nuts$ess_min)`, Panel A) and highest potential scale reduction factor which was `log_sigma_alpha_x` (an $\hat R$ of `r round(naomi_nuts$rhat_max, 3)`, Panel B).
 
@@
 
-(ref:marginal-sd) The estimated posterior marginal standard deviation of each hyperparameter varied substantially based on its scale, either logarithmic or logistic.
-
-```{r marginal-sd, fig.cap="(ref:marginal-sd)"}
-knitr::include_graphics("depends/marginal-sd.png")
-```
-
 ```{r}
@@
 knitr::include_graphics("depends/nodes-quantiles-sd.png")
+```
+
+(ref:marginal-sd) The estimated posterior marginal standard deviation of each hyperparameter varied substantially based on its scale, either logarithmic or logistic.
+
+```{r marginal-sd, fig.cap="(ref:marginal-sd)"}
+knitr::include_graphics("depends/marginal-sd.png")
 ```

### 2025-08-10 src/docs_paper-jtb/paper.Rmd
Updates to paper and reply with grid results fix

@@
 
-Inference with NUTS took `r naomi_time_tmbstan_hours` hours, while inference with GPCA-AGHQ took `r naomi_time_aghq_hours` hours and GEB just `r naomi_time_tmb_hours` hours (equivalent to `r naomi_time_tmb_minutes` minutes).
-Both the NUTS and GPCA-AGHQ algorithms can be run under a range of settings, trading off accuracy and runtime.
+Inference with NUTS took `r naomi_time_tmbstan_hours` hours (across 18 cores), while inference with GPCA-AGHQ took `r naomi_time_aghq_hours` hours and GEB just `r naomi_time_tmb_hours` hours (equivalent to `r naomi_time_tmb_minutes` minutes) both on a laptop.
+The NUTS and GPCA-AGHQ algorithms can be run under a range of settings, trading off accuracy and run-time.
 
@@
 For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (GEB) to `r abs(rmse_aghq_sd)` (GPCA-AGHQ).
-However, improvements in latent field estimate accuracy transferred to model outputs to a more limited and mixed extent (Supplementary Figures 12 and 13).
+Improvements in latent field estimate accuracy transferred to model outputs to a more limited and mixed extent (Supplementary Figures 12 and 13).
 
@@
 For points below the dashed line, performance of GPCA-AGHQ was better.
-Most notably, for the latent field parameters $\texttt{ui\_lambda\_x}$ the average test statistic for GEB was substantially higher than for GPCA-AGHQ.
-This length 32 parameter $\mathbf{u}_x^{\lambda}$ plays a key role in modelling ART attendance (Section \@ref(sec:art)).
+The latent field parameters informing the ART coverage component of the model (Section \@ref(sec:art)) had the greatest reduction in average test statistic.
 
@@
 ```{r}
-ui_lambda_x <- readRDS("depends/ui-lambda-x.rds")
+us_alpha_xs <- readRDS("depends/us-alpha-xs.rds")
 ```
 
-(ref:ui-lambda-x) The parameter ``r paste0(ui_lambda_x$par)`` had the greatest difference in KS test statistics between GEB and GPCA-AGHQ to NUTS.
-For this parameter, the potential scale reduction factor was `r ui_lambda_x$rhat` and effective sample size was `r ui_lambda_x$ess`, indicating that NUTS results are likely to be accurate.
+(ref:us-alpha-xs) The parameters `us_alpha_xs` had the greatest difference in KS test statistics between GEB and GPCA-AGHQ to NUTS.
+This figure shows the element ``r paste0(ui_lambda_x$par)``.
+For this element, the NUTS potential scale reduction factor was `r ui_lambda_x$rhat` and effective sample size was `r ui_lambda_x$ess`.
 
-```{r ui-lambda-x, fig.cap="(ref:ui-lambda-x)"}
-knitr::include_graphics("depends/ui-lambda-x.png")
+```{r us-alpha-xs, fig.cap="(ref:us-alpha-xs)"}
+knitr::include_graphics("depends/us-alpha-xs.png")
 ```
@@
 Figure \@ref(fig:ks-summary) shows the differences in average KS test statistic for each latent field parameter.
-GCPA-AGHQ showed the greatest average improvement in KS test statistic for a vector of parameters in the ART attendance component of Naomi (Figure \@ref(fig:ui-lambda-x)).
-For both GEB and GPCA-AGHQ the KS test statistic for a parameter was correlated with low NUTS ESS (Supplementary Figure 14).
-This correlation may be due to by difficulties estimating particular parameters for all inference methods, or high KS values caused by NUTS inaccuracies.
+GCPA-AGHQ showed the greatest average improvement in KS test statistic for parameters in the ART coverage component of Naomi (Figure \@ref(fig:us-alpha-xs)).
+For both GEB and GPCA-AGHQ the KS test statistic for a parameter was correlated with lower NUTS ESS (Supplementary Figure 14).
+This correlation may be due to by difficulties estimating particular parameters for all inference methods.
 
@@
 That said, Malawi is one of the countries with the fewest number of districts.
-As NUTS took over two days to reach convergence in Malawi, for larger countries, with hundreds of districts, it may be impossible to run NUTS to convergence, and faster approximate methods such as GEB and GPCA-AGHQ may be necessary.
+As NUTS took around three days to reach convergence in Malawi and required substantial computational resources.
+For larger countries, with hundreds of districts, it may be impossible to run NUTS to convergence, and faster approximate methods such as GEB and GPCA-AGHQ may be necessary.
 

### 2025-08-10 src/docs_paper-jtb/revisions.Rmd
Updates to paper and reply with grid results fix

@@
 We appreciate the feedback to improve the work and have made the relevant changes to the manuscript.
+
+**Important**
+
+While addressing the comments, we noticed an issue with the presented results.
+Some of the AGHQ results in the manuscript used the PCA grid, and others used an experimental "scaled" PCA grid.
+This "scaled" grid involved scaling all hyperparameters to have the same marginal variance, which we found not to work well in practice.
+We have now corrected the results in the manuscript to all be from the PCA grid described.
+This change impacts the KS test, MMD, and exceedance probabilities.
+We are grateful to have had a chance to find and correct this issue before publication.
 

### 2025-08-12 src/docs_paper-jtb/paper.Rmd
Correction to figure caption

@@
 (ref:us-alpha-xs) The parameters `us_alpha_xs` had the greatest difference in KS test statistics between GEB and GPCA-AGHQ to NUTS.
-This figure shows the element ``r paste0(ui_lambda_x$par)``.
-For this element, the NUTS potential scale reduction factor was `r ui_lambda_x$rhat` and effective sample size was `r ui_lambda_x$ess`.
+This figure shows the element ``r paste0(us_alpha_xs$par)``.
+For this element, the NUTS potential scale reduction factor was `r us_alpha_xs$rhat` and effective sample size was `r us_alpha_xs$ess`.
 

### 2025-08-12 src/docs_paper-jtb/paper.Rmd
Clearer exposition about compute time

@@
 
-Inference with NUTS took `r naomi_time_tmbstan_hours` hours (across 18 cores), while inference with GPCA-AGHQ took `r naomi_time_aghq_hours` hours and GEB just `r naomi_time_tmb_hours` hours (equivalent to `r naomi_time_tmb_minutes` minutes) both on a laptop.
+Inference with NUTS took `r naomi_time_tmbstan_hours` hours (across 18 cores, i.e. over 50 days of compute time) running on a high performance computing cluster.
+Meanwhile, inference with GPCA-AGHQ took `r naomi_time_aghq_hours` hours and GEB just `r naomi_time_tmb_hours` hours (equivalent to `r naomi_time_tmb_minutes` minutes), both running on a laptop.
 The NUTS and GPCA-AGHQ algorithms can be run under a range of settings, trading off accuracy and run-time.

### 2025-08-12 src/docs_paper-jtb/paper.Rmd

@@
 are the zeroes of the $k$th Hermite polynomial $H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2)$.
-The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k!
+The corresponding weighting function $\omega_k: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega_k(\z) = \prod_{j = 1}^m \omega_k(z_j)$ where $\omega_k(z) = k!
 / [H_{k + 1}(z)]^2 \phi(z)$, and $\phi(\cdot)$ is a standard Gaussian density.
@@
 \begin{equation}
-p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z).
+p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega_k(\z).
 \label{eq:aghq}
@@
 \begin{equation}
-\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z).
+\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega_k(\z).
 \label{eq:nest}
 \end{equation}
-Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
+Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega_k(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.
 
@@
 \begin{equation}
-\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
+\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \widetilde{\omega}(\z),
 \end{equation}
-where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
+where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\widetilde{\omega}(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
 Panel C of Figure \@ref(fig:aghq) illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.

### 2025-08-20 src/docs_paper-jtb/paper.Rmd
JIE suggestions

@@
 The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix 1.
-The proportion recently infected $\kappa_i$ is linked to HIV incidence via
+The proportion recently infected $\kappa_i$ is linked to HIV incidence via a reformulation of the Kassanjee \textit{et al.} biomarker-based incidence estimator (INSERT REFERENCE: https://pubmed.ncbi.nlm.nih.gov/22627902/):
+
 \begin{equation}
@@
 Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language (PPL) [@de2017programming].
-Direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
+Direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50%  based on experiements conducted by others [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
 However, requiring rewriting the Naomi model log-posterior outside of `TMB` would be a substantial downside.

### 2025-08-20 src/docs_paper-jtb/revisions.Rmd
JIE suggestions

--- 2025-08-09 src/docs_paper-jtb/revisions.Rmd
@@
 ---
-title: Response to reviewers
+title: 
 author:
-  - Adam Howes^[Department of Mathematics, Imperial College London]
-  - Alex Stringer^[Department of Statistics and Actuarial Science, University of Waterloo]
-  - Seth R. Flaxman^[Department of Computer Science, Oxford University]
-  - Jeffrey W. Imai-Eaton^[Harvard T.H. Chan School of Public Health, Harvard University]
 output:
   bookdown::pdf_document2:
-    toc: true
+    toc: false
     number_sections: true
@@
 
-# Introduction
+Dear Dr. XXXX,
 
-Thank you for taking the time to read and review our paper "Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature".
-In what follows we address your comments and suggestions point by point.
-We appreciate the feedback to improve the work and have made the relevant changes to the manuscript.
+Many thanks for your consideration of our paper "Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature" and to the reviewers for their thoughtful comments on the manuscript.
+
+Below we address the reviewers' comments and suggestions point by point.
+While updating and extending the simulations to address some of the reviewers' comments, we noticed that the results reported arose from two different versions of the model, with some from an alternative experimental algorithm that was not retained for the final method.
+Along with updating all results with longer MCMC chain simulations, we have now corrected the results in the manuscript to all be from the PCA grid described.
+This change impacts numerical results reported for the KS test, MMD, and exceedance probabilities, but does not change the overall conclusions or recommendations from the paper.
+
+We appreciate the feedback to improve the work and have made the relevant changes to the manuscript as described below.
+
+Sincerely,  
+Adam Howes on behalf of Alex Stringer, Seth Flaxman, and Jeff Imai-Eaton
 
@@
 \begin{reviewer}
-The class of models to which the present model is assigned, namely Extended Latent Gaussian Models (ELGMs), is very broad, and in my (perhaps rather curmudgeonly) opinion too broad to be of much value.
+\begin{itemize}
+\item [R1.1] The class of models to which the present model is assigned, namely Extended Latent Gaussian Models (ELGMs), is very broad, and in my (perhaps rather curmudgeonly) opinion too broad to be of much value.
 On page 3 the authors write about the new mapping structure for ELGMs which is said to "relax the restriction that there is a one-to-one mapping between the mean response mu and structured additive predictor eta." and "Importantly, this mapping allows for the presence of non-linearity in the model." What exactly is meant by this?
@@
 First, I would suggest to be clear on page 3 about what kind of mappings are being done in this model and see if these also match to a named smaller class within the ELGM class, and/or identify fuzzy matches to existing model structures of this form (such as Katie Wilson's "pointless spatial models"); especially those use in spatial statistics and/or epi.
+\end{itemize}
 \end{reviewer}
 
+\textbf{Response:} We thank the reviewer for the thoughtful and insightful critique of the ELGM framing, and also for the concrete and practical suggestions for how we can address this to improve the current paper.
 We have added the following text in Section 2.2 with the aim of making the mapping between the latent field and structured additive predictor of the Naomi model clearer:
@@
 \begin{reviewer}
-I'm not an expert on the various flavours of NUTS algorithms but from my experience I have always been disappointed in the performance of tmbstan.
+\begin{itemize}
+\item [R1.2] I'm not an expert on the various flavours of NUTS algorithms but from my experience I have always been disappointed in the performance of tmbstan.
 Are there any comparisons between tmbstan and ordinary stan not conducted by the tmbstan package authors?
@@
 If not, perhaps see if a caveat is needed, something along the lines of 'tmbstan was tested because it's compatible with the idea of repurposing existing software (i.e., the original TMB template) but this doesn't mean that rebuilding the model from scratch in Stan, for instance, would create a sampler just as impractically slow'.
+\end{itemize}
 \end{reviewer}
 
-`tmbstan` passes the objective function and its gradient (calculated in `TMB`) directly to `rstan` via the [`op_matrix_vari`](https://mc-stan.org/math/classstan_1_1math_1_1op__matrix__vari.html).
-As such, the NUTS algorithm used should be the same.
-Nonetheless, it could be that there are speed differences between function evaluations in `TMB` and Stan.
+\textbf{Response:} `tmbstan` passes the objective function and its gradient (calculated in `TMB`) directly to `rstan` via the [`op_matrix_vari`](https://mc-stan.org/math/classstan_1_1math_1_1op__matrix__vari.html).
+As such, the NUTS algorithm used is exactly the same (as the same underlying `rstan` code is called.
+It could be the case that there are speed differences between function evaluations in `TMB` and Stan.
 We have added a sentence to the discussion noting this possible comparison:
 
-> Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
+> Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% based on experiements conducted by others [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
 
 \begin{reviewer}
-The equation for the fraction recently infected on page 17 puzzles me; though I don't come from the HIV field.
+\begin{itemize}
+\item [R1.3] The equation for the fraction recently infected on page 17 puzzles me; though I don't come from the HIV field.
 It seems to me that this is a convenience 'guesstimation' used only as an output for end users, rather than being fitted to data or offering some kind structural constraint on the fits?
@@
 Also, presumably the epidemiology prescribes some tight bounds on the ANC offsets given the demography of birth and sexual partnership; are these pre-built into the priors for the model?
+\end{itemize}
 \end{reviewer}
 
-Some HIV surveys provide bio-marker data on whether or not the infection was recent.
-Equation 7 links the probability of recent infection to HIV prevalence, HIV incidence, and properties of the recent infection assay.
-This allows us to back-out estimates of the HIV incidence.
-<!-- Is there something to add to the manuscript here? -->
+\textbf{Response:} Some HIV surveys, including the example data from Malawi used in the analysis reported here, provide bio-marker data on whether or not HIV infection was recently acquired (approximately within the previous 4 months; the $\Omega_T$ parameter in Equation 7).
+This information on the proportion recently infected along with characteristics of the assay used to measure recent infection provides a direct survey estimator for HIV incidence, as derived by Kassanjee \textit{et al.} (https://pubmed.ncbi.nlm.nih.gov/22627902/).
+The formula in Equation 7 is a reformulation of the Kassanjee \textit{et al.} estimator to express a likelihood for the survey observations of the number recenlty infected, enabling Bayesian inference on the $\lambda_i$ parameters (HIV incidence rate).
+Details of the derivation of Equation 7 are described in the supplementary materials of Eaton \text{et al.} (and also Bao \text{et al.} (https://pubmed.ncbi.nlm.nih.gov/25406754/)).
+We have revised the manuscript here to clarify that Equation 7 is derived from the classical Kassanjee \textit{et al.} estimator:
 
-We agree that tightening the bounds on the ANC offsets would be an important step for improving estimates of HIV incidence.
-Work is ongoing to improve the Naomi model [@esra2024improved].
-<!-- Has this work been included yet, or it's still to be done in future? -->
+> The proportion recently infected $\kappa_i$ is linked to HIV incidence via a reformulation of the Kassanjee \textit{et al.} biomarker-based incidence estimator...
+
+The proposed approach by the reviewer to extend the model described here to a compartmental model that makes inference on HIV incidence from fitting to time series of HIV prevalence is indeed promising and an active area of research for our group and others.
+However, we feel this is separate from and outside the scope of the methodological inference questions addressed in this paper and describing this potential extension adds complexity that distracts from the primary research questions addressed here.
+
+The reviewer is also correct to intuit that there are well characterised relationships between fertility and HIV infection that shape the pattern of HIV prevalence among pregnant women versus all women over the life course.
+However, at the ecological level the relationship can be somewhat more complex because of other confounding factors that affect both fertility and HIV risk, resulting in variation in the ANC versus population offsets across countries (for example, higher fertility and lower HIV prevalence in rural areas compared to urban areas).
+For this reason, we opted for a diffuse prior on the offset parameters, which in practice is well identified with the data available in applications.
 
 \begin{reviewer}
-You anticipated my very clever idea for improving the performance of the PCA algorithm: namely, to consider the relevance of each dimension to a measure of uncertainty relevant to the end predictive purposes.
+\begin{itemize}
+\item [R1.4] You anticipated my very clever idea for improving the performance of the PCA algorithm: namely, to consider the relevance of each dimension to a measure of uncertainty relevant to the end predictive purposes.
 Are calculations of this nature possible with TMB as it stands, or would one need to dig into the bowels of the c++ code and create a TMB package extension?
 I think this is an interesting area for future work.
+\end{itemize}
 \end{reviewer}
 
-We agree that this is an interesting area for future work, and have added a sentence to Section 6.1.2. discussing the possibility of doing this in `TMB`:
+\textbf{Response:} We agree that this is an interesting area for future work.
+Regarding the question about whether this is possible within the existing `TMB` infrastructure, we believe it would be.
+We have added the following sentence to Section 6.1.2.
+discussing how:
 
@@
 \begin{reviewer}
-Finally, on page 17 it feels like Bayesian 'optimisation' is another natural competitor in this space, especially since it's quite straightforward to just run this kind of algorithm to explore hyper parameters until you get bored or run out of time, i.e., you get what you get from stopping early, unlike HMC for instance.
+\begin{itemize}
+\item [R1.5] Finally, on page 17 it feels like Bayesian 'optimisation' is another natural competitor in this space, especially since it's quite straightforward to just run this kind of algorithm to explore hyper parameters until you get bored or run out of time, i.e., you get what you get from stopping early, unlike HMC for instance.
+\end{itemize}
 \end{reviewer}
 
-Our understanding is that the goal of Bayesian optimisation (BO) is to find the maximum of a (black box) function, rather than to perform Bayesian inference.
-One could imagine using BO to optimise the hyperparameters, however it would be preferable to use gradients as they are available.
+\textbf{Response:} Our understanding of the goal of Bayesian optimisation (BO) is to find the maximum of a (black box) function, rather than to perform Bayesian inference.
+(For example, one could imagine using BO to optimise the hyperparameters.)
 
 \begin{reviewer}
-Typos spotted:
+\begin{itemize}
+\item [R1.6] Typos spotted:
 p2 "closer [to]"
 p4. "hyperparmters"
+\end{itemize}
 \end{reviewer}
 
-Thank you for spotting these two typographic errors.
-We have corrected them in the manuscript.
+\textbf{Response:} We thank the reviewer for spotting these two typographic errors. We have corrected these in the manuscript.
 
@@
 \begin{reviewer}
-The NUTS sampler was run for 4 chains for 100,000 iterations, with the first half discarded.
+\begin{itemize}
+\item [R2.1] The NUTS sampler was run for 4 chains for 100,000 iterations, with the first half discarded.
 This gives a lowest ESS of 208 and a largest Rhat of 1.021.
@@
 Can the model be run with 16 chains on an HPC or something?
+\end{itemize}
 \end{reviewer}
 
-We agree that running more iterations of the NUTS sampler is a good idea, and could improve the minimum ESS and accuracy of any comparisons, especially distributional ones.
+\textbf{Response:} We agree that running more iterations of the NUTS sampler is a good idea, and could improve the minimum ESS and accuracy of any comparisons, especially distributional ones.
 We have rerun the NUTS sampler using a HPC with 18 chains (across 18 cores) with 40,000 per chain and a thinning factor of 20.
 To improve the sampling also increased the `adapt_delta` parameter from the default of 0.8 to 0.95, and the `max_treedepth` parameter from the default of 10 to 12.
-The resulting diagnostics are much improved, as described in an update to Section 5.1:
+The resulting diagnostics are much improved. Updated sampler diagnostic results are reported in Section 5.1:
 
@@
 \begin{reviewer}
-I really like Figure 5 but just a few more words explaining the green lines would be useful.
+\begin{itemize}
+\item [R2.2] I really like Figure 5 but just a few more words explaining the green lines would be useful.
 My first thought was that there was 1 or 3 green lines and that the width meant something.
@@
 Perhaps some alpha would help communicate this.
+\end{itemize}
 \end{reviewer}
 
-Thank you for the suggestion.
+\textbf{Response:} We thank the reviewer for the suggestion.
 We have adjusted the `alpha` value from 0.5 to 0.1 and added a caption to the figure explaining the possible overlap to make Figure 5 clearer:
@@
 \begin{reviewer}
-In Figure 6, is the y axis NUTS estimate - GEB estimate?
+\begin{itemize}
+\item [R2.3] In Figure 6, is the y axis NUTS estimate - GEB estimate?
 Or the other way around?
@@
 The new method moves the 3 little clusters at 1.1, 1.4 and 1.8 from negative to positive which seems it could be important and worth noting.
+\end{itemize}
 \end{reviewer}
@@
 \begin{reviewer}
-Figure 7 is very hard to read.
+\begin{itemize}
+\item [R2.4] Figure 7 is very hard to read.
 It's an interesting and useful plot.
@@
 And there is just too many colours, so I think it needs to be split into 4 or so groups on seperate panels.
+\end{itemize}
 \end{reviewer}
 
-We agree that there are too many colours here for this figure to be clear, and that the suggestion to split the figure across panels is a good one.
+\textbf{Response:} We agree that the number of colours in the figure impeded clarity, and have adopted the suggestion to split the figure across panels.
 To do this, we have grouped the parameters by the model component (e.g. HIV prevalence), and used colour to represent the type of parameter (e.g. intercept).
@@
 \begin{reviewer}
-The authors mention that the method is tunable with how many PCA components are selected and what value of k to use.
+\begin{itemize}
+\item [R2.5] The authors mention that the method is tunable with how many PCA components are selected and what value of k to use.
 It is a bit of a shame that one or two other options are not examined.
@@
 So I will leave it to the authors whether or not to do more work here.
+\end{itemize}
 \end{reviewer}
 
-In the current work we focused on results which were as accurate as possible while still running in a practical amount of time.
+\textbf{Response:} In the current work we focused on results which were as accurate as possible while still running in a practical amount of time.
 We agree that it is possible that faster results that are nearly as accurate could be obtained using or procedure.
@@
 \begin{reviewer}
-Just looking at Figure 2, a useful intermediate scheme to me seems to be to drop the corners of the grid.
+\begin{itemize}
+\item [R2.6] Just looking at Figure 2, a useful intermediate scheme to me seems to be to drop the corners of the grid.
 So you would be left with 5 points.
@@
 So please don't take this comments as work to be done, but rather as an interesting idea for future work.
+\end{itemize}
 \end{reviewer}
 
-We agree that this could be a useful approach.
-Indeed Section 6.1.2. reads "Finally, it is likely possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.".
+\textbf{Response:} We agree that this could be a useful approach.
+This is highlighted in Section 6.1.2., which reads "Finally, it is likely possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.".
 We believe the suggested approach is what we refer to here (very briefly) as "pruning".

### 2025-08-20 src/docs_paper-jtb/paper.Rmd
Add cite

@@
 The structured additive predictors $\eta^\theta_i$ for $\theta \in \{\rho, \alpha, \lambda\}$ are given in Appendix 1.
-The proportion recently infected $\kappa_i$ is linked to HIV incidence via a reformulation of the Kassanjee \textit{et al.} biomarker-based incidence estimator (INSERT REFERENCE: https://pubmed.ncbi.nlm.nih.gov/22627902/):
+The proportion recently infected $\kappa_i$ is linked to HIV incidence via a reformulation of the biomarker-based incidence estimator of @kassanjee2012new
 

### 2025-08-20 src/docs_paper-jtb/revisions.Rmd
JIE suggestions (#61)
* JIE suggestions

* Add cite

* Integrate JIE changes

---------

Co-authored-by: athowes <adamthowes@gmail.com>

@@
 
-Dear Dr. XXXX,
+Dear Dr. Chaplain,
 
@@
 
-Below we address the reviewers' comments and suggestions point by point.
-While updating and extending the simulations to address some of the reviewers' comments, we noticed that the results reported arose from two different versions of the model, with some from an alternative experimental algorithm that was not retained for the final method.
+We appreciate the feedback to improve the work and have made the relevant changes to the manuscript as described below.
+
+While updating and extending the simulations to address some of the reviewers' comments, we noticed that the results reported arose from two different versions of the method, with some from an alternative experimental algorithm that was not retained for the final method.
 Along with updating all results with longer MCMC chain simulations, we have now corrected the results in the manuscript to all be from the PCA grid described.
 This change impacts numerical results reported for the KS test, MMD, and exceedance probabilities, but does not change the overall conclusions or recommendations from the paper.
-
-We appreciate the feedback to improve the work and have made the relevant changes to the manuscript as described below.
 
@@
 \textbf{Response:} `tmbstan` passes the objective function and its gradient (calculated in `TMB`) directly to `rstan` via the [`op_matrix_vari`](https://mc-stan.org/math/classstan_1_1math_1_1op__matrix__vari.html).
-As such, the NUTS algorithm used is exactly the same (as the same underlying `rstan` code is called.
+As such, the NUTS algorithm used is exactly the same (as the same underlying `rstan` code is called).
 It could be the case that there are speed differences between function evaluations in `TMB` and Stan.
@@
 \textbf{Response:} Some HIV surveys, including the example data from Malawi used in the analysis reported here, provide bio-marker data on whether or not HIV infection was recently acquired (approximately within the previous 4 months; the $\Omega_T$ parameter in Equation 7).
-This information on the proportion recently infected along with characteristics of the assay used to measure recent infection provides a direct survey estimator for HIV incidence, as derived by Kassanjee \textit{et al.} (https://pubmed.ncbi.nlm.nih.gov/22627902/).
-The formula in Equation 7 is a reformulation of the Kassanjee \textit{et al.} estimator to express a likelihood for the survey observations of the number recenlty infected, enabling Bayesian inference on the $\lambda_i$ parameters (HIV incidence rate).
-Details of the derivation of Equation 7 are described in the supplementary materials of Eaton \text{et al.} (and also Bao \text{et al.} (https://pubmed.ncbi.nlm.nih.gov/25406754/)).
-We have revised the manuscript here to clarify that Equation 7 is derived from the classical Kassanjee \textit{et al.} estimator:
+This information on the proportion recently infected along with characteristics of the assay used to measure recent infection provides a direct survey estimator for HIV incidence, as derived by @kassanjee2012new.
+The formula in Equation 7 is a reformulation of the @kassanjee2012new estimator to express a likelihood for the survey observations of the number recently infected, enabling Bayesian inference on the $\lambda_i$ parameters (HIV incidence rate).
+Details of the derivation of Equation 7 are described in the supplementary materials of @eaton2021naomi (and also @bao2014incorporating).
+We have revised the manuscript here to clarify that Equation 7 is derived from the classical @kassanjee2012new estimator:
 
-> The proportion recently infected $\kappa_i$ is linked to HIV incidence via a reformulation of the Kassanjee \textit{et al.} biomarker-based incidence estimator...
+> The proportion recently infected $\kappa_i$ is linked to HIV incidence via a reformulation of the biomarker-based incidence estimator of @kassanjee2012new...
 
@@
 Regarding the question about whether this is possible within the existing `TMB` infrastructure, we believe it would be.
-We have added the following sentence to Section 6.1.2.
-discussing how:
+We have added the following sentence to Section 6.1.2. discussing how:
 
@@
 To improve the sampling also increased the `adapt_delta` parameter from the default of 0.8 to 0.95, and the `max_treedepth` parameter from the default of 10 to 12.
-The resulting diagnostics are much improved. Updated sampler diagnostic results are reported in Section 5.1:
+The resulting diagnostics are much improved.
+Updated sampler diagnostic results are reported in Section 5.1:
 
@@
 \textbf{Response:} In the current work we focused on results which were as accurate as possible while still running in a practical amount of time.
-We agree that it is possible that faster results that are nearly as accurate could be obtained using or procedure.
+We agree that it is possible that faster results that are nearly as accurate could be obtained using our procedure.
 
